### 2015-04-12 Algebraische Geometrie/AlgebraischeGeometrie.tex
Add Algebraische Geometrie

\documentclass[11pt,b5paper,openany]{memoir}
\usepackage[utf8]{inputenc}
\usepackage[german]{babel}
\usepackage{amsmath, amsfonts, amssymb}
\usepackage{paralist}
\usepackage{makeidx}
\usepackage{graphicx}
\usepackage{dsfont}
\usepackage{hyperref}
\hypersetup{
    bookmarksopen=true,
    pdfstartview=FitH
}
\usepackage[all]{xy}
\usepackage[left=1.3cm,right=1.3cm,top=2cm,bottom=2cm]{geometry}
\author{Yichuan Shen}
\title{Algebraische Geometrie}
\linespread{1.1}
\makeindex
\chapterstyle{tandh}

\def \: {\mathbin:}
\def \qed {$\hfill\square$}
\def \qedhere {\tag*{$\square$}}

\begin{document}

\frontmatter
\pagenumbering{gobble}

\begin{center}
\vspace*{0cm}
\vfill {
    \begin{Huge}Algebraische Geometrie\end{Huge}\\
    \vspace{1mm}
    Vorlesung von \textsc{prof. dr. kay wingberg}
}
\vfill {
    \includegraphics[scale=.2]{seal.pdf}\\
    gesetzt von\\\textsc{yichuan shen} \\\ \\
    2014/2015\\
    \today
}
\end{center}

\clearpage

\vspace*{0cm}
\vfill
\paragraph{Lizenz \& Quellcode.} Dieses Dokument steht unter einer Creative Commons Attribution Lizenz. Weitere Informationen findet man auf \url{http://creativecommons.org/licenses/by/3.0/de/}. Der Quellcode ist auf \url{http://github.com/yishn/Uni} zu finden.

\mainmatter
\pagenumbering{arabic}
\setcounter{page}{3}
\tableofcontents

\chapter{Varietäten}

Sei $k$ ein algebraisch abgeschlossener Körper.

\section{Affine Varietäten}

\paragraph{Definition.} \begin{enumerate}[(i)]
\item Die Menge aller $n$-Tupeln über $k$
\[\mathbf{A}^n = \mathbf{A}_k^n = \{(a_1,\ldots,a_n)\mid a_i\in k\} \]
heißt \textit{affiner $n$-dimensionaler Raum}\index{affiner Raum} über $k$. Ein Element $P=(a_1,\ldots,a_n)\in\mathbf{A}^n$ heißt \textit{Punkt}\index{Punkt} und die $a_i$ heißen \textit{Koordinaten}\index{Koordinate} von $P$.
\item Der Polynomring über $k$ in $n$ Variablen bezeichnen wir mit $A=k[X_1,\ldots,X_n]$. Für $T\subset A$ definieren wir die \textit{Nullstellenmenge}\index{Nullstellenmenge} von $T$ wie folgt:
\[Z(T)=\{P\in\mathbf{A}^n\mid \forall f\in T\colon f(P)=0\} \]
Es gilt $Z(T)=Z(\mathfrak{a})$, wobei $\mathfrak{a}$ das von $T$ erzeugte Ideal in $A$ ist.
\item Eine Teilmenge $Y\subset\mathbf{A}^n$ der Form $Y=Z(T)$ für ein $T\subset A$ heißt \textit{affine algebraische Menge}\index{algebraische Menge!affin}. Für algebraische Mengen $Y_i=Z(\mathfrak{a}_i)$ mit Ideale $\mathfrak{a}_i\subset A,\ i\in I$ gilt:
\[Y_1\cup Y_2=Z(\mathfrak{a}_1\cap\mathfrak{a}_2),\quad \bigcap_{i\in I}Y_i = Z\Big(\sum_{i\in I}\mathfrak{a}_i\Big) \]
Ferner gilt $\mathbf{A}^n=Z(0)$ und $\varnothing = Z(1)$. Wir statten $\mathbf{A}^n$ mit der sogenannten \textit{Zariski-Topologie}\index{Zariski-Topologie} aus, in dem wir eine Menge $U\subset\mathbf{A}^n$ genau dann offen nennen, wenn $\mathbf{A}^n\setminus U$ eine algebraische Menge ist.
\item Eine \textit{affine Varietät}\index{Varietät!affin} $V$ ist eine \textit{irreduzible}\index{irreduzibel} abgeschlossene Menge in $\mathbf{A}^n$, d.h. aus $V=V_1\cup V_2$ mit abgeschlossenen Mengen $V_1,V_2\subset\mathbf{A}^n$ folgt $V_1=\varnothing$ oder $V_2=\varnothing$.

Eine offene Teilmenge einer affinen Varietät bzgl. der induzierten Topologie nennt man \textit{quasi-affine Varietät}\index{Varietät!quasi-affin}. 
\item Sei $Y\subset\mathbf{A}^n$ eine algebraische Menge. Dann definieren wir das Ideal:
\[I(Y)=\{f\in A\mid \forall P\in Y\colon f(P)=0\} \]
Der \textit{Koordinatenring}\index{Koordinatenring} von $Y$ ist definiert als $A(Y)=A/I(Y)$.
\end{enumerate}

\paragraph{Definition.} Sei $\mathfrak{a}\subset A$ ein Ideal. Das \textit{Radikal}\index{Radikal} von $\mathfrak{a}$ ist definiert als:
\[\operatorname{Rad}(\mathfrak{a})=\{f\in A\mid\exists r>0\colon f^r\in\mathfrak{a} \} \]
Ein Ideal $\mathfrak{a}\subset A$ heißt \textit{Radikalideal}\index{Radikalideal}, wenn $\mathfrak{a}=\operatorname{Rad}(\mathfrak{a})$ gilt.

\paragraph{Satz.} Es gibt eine inklusionsumkehrende Bijektion:
\[\{\text{Algebraische Mengen in }\mathbf{A}^n\} \to \{\text{Radikalideale in }k[X_1,\ldots,X_n]\},\ Y\mapsto I(Y) \]
mit der Umkehrabbildung $\mathfrak{a}\mapsto Z(\mathfrak{a})$. Eine algebraische Menge $Y$ ist genau dann irreduzibel, wenn $I(Y)\subset A$ ein Primideal ist.

\section{Projektive Varietäten}

\paragraph{Definition.}\begin{enumerate}[(i)]
\item Zwei Punkte $(a_0,\ldots,a_n),(b_0,\ldots,b_n)\in \mathbf{A}^{n+1}$ heißen äquivalent, wenn ein $\lambda\in k^\times$ existiert, so dass $a_i=\lambda b_i$ für alle $i$ gilt. Die Äquivalenzklasse von $(a_0,\ldots,a_n)$ wird mit $(a_0\: \ldots\: a_n)$ bezeichnet. Der \textit{projektiver $n$-dimensionaler Raum}\index{projektiver Raum} über $k$ wird definiert als:
\[\mathbf{P}^n=\mathbf{P}_k^n = \{(a_0\: \ldots\: a_n) \mid a_i\in k\text{ nicht alle $0$} \} \]
Ein Element $P=(a_0\: \ldots\: a_n)\in\mathbf{P}^n$ heißt \textit{Punkt}\index{Punkt} und die $a_i$ heißen \textit{homogene Koordinaten} von $P$.\index{Koordinate!homogen}
\item Der Polynomring über $k$ in $n+1$ Variablen $S=k[X_0,\ldots,X_n]$ wird mit der folgenden Zerlegung zu einem graduierten Ring:
\[S=\bigoplus_{d\geq 0}S_d,\quad S_d=\Big\{ \sum a_{i_0,\ldots,i_n} X_0^{i_0}\cdots X_n^{i_n}\Bigm| a_{i_0,\ldots,i_n}\in k,\ \sum_{j=0}^n i_j=d \Big\} \]
Die Elemente in $S_d$ heißen \textit{homogene Elemente vom Grad $d$}\index{homogene Elemente}. Ein Ideal $\mathfrak{a}\subset S$ heißt \textit{homogenes Ideal}\index{homogenes Ideal}, wenn $\mathfrak{a}=\bigoplus_{d\geq 0}(S_d\cap\mathfrak{a})$ gilt.
\item Sei $\mathfrak{a}\subset S$ ein homogenes Ideal. Dann setzen wir:
\[Z(\mathfrak{a}) = \{P\in\mathbf{P}^n\mid \forall f\in\mathfrak{a}\text{ homogen}\colon f(P)=0 \} \]
Diese ist wohldefiniert, da $f(\lambda a_0,\ldots,\lambda a_n)=\lambda^d f(a_0,\ldots,a_n)$ für $f\in S_d$. Eine Menge $Y\subset\mathbf{P}^n$ heißt \textit{projektive algebraische Menge}\index{algebraische Menge!projektiv}, wenn $Y=Z(\mathfrak{a})$ für ein homogenes Ideal $\mathfrak{a}\subset S$ gilt.

Analog wie im affinen Fall, können wir auch $\mathbf{P}^n$ mit der \textit{Zariski-Topologie} ausstatten, d.h. eine Menge $U\subset\mathbf{P}^n$ ist genau dann offen, wenn $\mathbf{P}^n\setminus U$ eine projektive algebraische Menge ist.\index{Zariski-Topologie}
\item Eine \textit{projektive Varietät}\index{Varietät!projektiv} $V$ ist eine irreduzible abgeschlossene Menge in $\mathbf{P}^n$. Eine offene Teilmenge einer projektiven Varietät bzgl. der induzierten Topologie nennt man \textit{quasi-projektive Varietät}\index{Varietät!quasi-projektiv}. 
\item Sei $Y\subset\mathbf{P}^n$ eine algebraische Menge. Dann setzen wir $I(Y)$ als das Ideal in $S$, das von der folgenden Menge erzeugt wird:
\[\{f\in S \text{ homogen}\mid \forall P\in Y\colon f(P)=0\} \]
$I(Y)$ ist ein homogenes Ideal in $S$. Der \textit{homogene Koordinatenring}\index{Koordinatenring!homogen} von $Y$ ist definiert als $S(Y)=S/I(Y)$.
\end{enumerate}

\paragraph{Satz.} Wir haben eine inklusionsumkehrende Bijektion:
\[\{\text{Algebraische Mengen in }\mathbf{P}^n\}\to\{\text{Radikalideale in }k[X_0,\ldots,X_n]\},\ Y\mapsto I(Y) \]
mit der Umkehrabbildung $\mathfrak{a}\mapsto Z(\mathfrak{a})$.

\paragraph{Satz.} Sei $Y$ eine (quasi-)projektive Varietät. Dann wird $Y$ von offenen Mengen der Form $Y\cap U_i,\ i=0,\ldots,n$ überdeckt mit:
\[U_i=\{(a_0\: \ldots\: a_n)\in \mathbf{P}^n\mid a_i\neq 0\} \]
Die Abbildungen $\varphi_i: U_i\to\mathbf{A}^n,\ (a_0\: \ldots\: a_n)\mapsto \big(\frac{a_0}{a_i},\ldots,\widehat{\frac{a_i}{a_i}},\ldots,\frac{a_n}{a_i}\big)$ sind wohldefiniert und Ho\-möo\-mor\-phis\-men, d.h. die $Y\cap U_i$ sind (quasi-)affine Varietäten.

\section{Morphismen}

\paragraph{Definition.} \begin{enumerate}[(i)]
\item Sei $Y\subset\mathbf{A}^n$ eine quasi-affine Varietät. Eine Abbildung $f:Y\to k$ heißt \textit{reguläre Funktion} in $P\in Y$, wenn eine offene Umgebung $U\subset Y$ mit $P\in U$ existiert, so dass $f=\frac{g}{h}$ auf $U$ für gewisse $g,h\in A$ gilt.

Sei $Y\subset\mathbf{P}^n$ eine quasi-projektive Varietät. Eine Abbildung $f:Y\to k$ heißt \textit{reguläre Funktion}\index{reguläre Funktion} in $P\in Y$, wenn eine offene Umgebung $U\subset Y$ mit $P\in U$ existiert, so dass $f=\frac{g}{h}$ auf $U$ für gewisse homogene Polynome $g,h\in S$ vom gleichen Grad.

Identifizieren wir $k\cong \mathbf{A}^1$ so ist eine reguläre Funktion notwendigerweise stetig.
\item Eine stetige Abbildung $\varphi:X\to Y$ zwischen zwei (quasi-projektive) Varietäten heißt \textit{Morphismus}\index{Morphismus!Varietäten}, wenn für jede offene Menge $V\subset Y$ und reguläre Funktion $f:V\to k$ auch $f\circ\varphi:\varphi^{-1}(V)\to k$ regulär ist.

Damit erhält man die Kategorie $\textbf{Var$(k)$}$ aller Varietäten auf $k$.
\end{enumerate}

%\paragraph{Bemerkung.} Sei $Y$ eine Varietät und $f,g:Y\to k$ reguläre Funktionen. Existiert eine nichtleere offene Menge $U\subset Y$ mit $f=g$ auf $U$, so folgt $f=g$ auf ganz $Y$: Sei dazu o.B.d.A. $U$ so klein, dass $f$ und $g$ als Brüche homogener Polynome vom gleichen Grad dargestellt werden kann. Dann ist die Nullstellenmenge $\{P\in Y\mid (f-g)(P)=0\}$ abgeschlossen und dicht in $Y$.

\paragraph{Definition.} Sei $Y$ eine Varietät und $P\in Y$ ein Punkt.
\begin{enumerate}[(i)]
\item Wir bezeichnen den Ring aller regulären Funktionen auf $Y$ mit $\mathcal{O}(Y)$.
\item $\mathcal{O}_{P,Y}=\{\langle U,f\rangle\mid P\in U\subset_\text{o} Y,\ f\text{ ist auf $U$ regulär}\}$ heißt der Ring der \textit{Keime} regulärer Funktionen auf $Y$ in $P$.\index{Keim} Wir identifizieren zwei Keime $\langle U,f\rangle=\langle V,g\rangle$, wenn $f=g$ auf $U\cap V$ gilt.

$\mathcal{O}_{P,Y}$ ist ein lokaler Ring, dessen Maximalideal wir mit $\mathfrak{m}_P$ bezeichnen.
\item $K(Y)=\{\langle U,f\rangle\mid\varnothing\neq U\subset_\text{o}Y,\ f\text{ ist auf $U$ regulär}\}$ heißt der \textit{Funktionenkörper}\index{Funktionenkörper}  von $Y$. Die Elemente von $K(Y)$ heißen \textit{rationale Funktionen} auf $Y$\index{rationale Funktion}.
\end{enumerate}
Es gilt $\mathcal{O}(Y)\subset\mathcal{O}_{P,Y}\subset K(Y)$.

\paragraph{Theorem.} Sei $Y\subset\mathbf{A}^n$ eine affine Varietät. Dann gilt:
\begin{enumerate}[(i)]
\item $\mathcal{O}(Y)\cong A(Y)$
\item Die Abbildung $Y\to \{\text{Maximale Ideale in }A(Y)\},\ P\mapsto \mathfrak{m}_P\subset\mathcal{O}_{P,Y}$ ist eine Bijektion.
\item $\mathcal{O}_{P,Y}\cong A(Y)_{\mathfrak{m}_P}$ und $\dim\mathcal{O}_{P,Y}=\dim Y$.
\item $K(Y)\cong\operatorname{Quot}(A(Y))$
\end{enumerate}

\paragraph{Theorem.} Sei $Y\subset\mathbf{P}^n$ eine projektive Varietät. Dann gilt:
\begin{enumerate}[(i)]
\item $\mathcal{O}(Y)=k$
\item $\mathcal{O}_{P,Y}\cong S(Y)_{(\mathfrak{m}_P)}$
\item $K(Y)\cong S(Y)_{((0))}$
\end{enumerate}

\paragraph{Theorem.} Sei $X$ eine beliebige Varietät und $Y$ eine affine Varietät. Dann haben wir eine Bijektion:
\[\operatorname{Mor}_{\textbf{Var}}(X,Y) \to \operatorname{Hom}_{\textbf{$k$-Alg}}(A(Y),\mathcal{O}(X)),\ f\mapsto(\varphi\mapsto \varphi\circ f) \]
Ist $X$ ebenfalls affin, so gilt $X\cong Y$, genau dann wenn $A(X)\cong A(Y)$. Der Funktor $\textbf{affine Var$(k)$}\to\textbf{nullteilerfreie $k$-Alg},\ X\mapsto A(X)$ ist eine pfeilumkehrende Äquivalenz von Kategorien.

\chapter{Schemata}
\section{Garben}

\paragraph{Definition 1.1.}\label{1.1} Sei $X$ ein topologischer Raum.
\begin{enumerate}[(i)]
\item Die Menge aller offenen Teilmengen in $X$ bilden zusammen mit den natürlichen Inklusionen eine Kategorie $\textbf{Top$(X)$}$.
\item Eine \textit{Prägarbe}\index{Prägarbe} $F$ abelscher Gruppen ist nichts anderes als ein kontravarianter Funk\-tor $F:\textbf{Top$(X)$}\to\textbf{Ab}$ mit $F(\varnothing)=0$.
\end{enumerate}

\paragraph{Bemerkung.}\begin{enumerate}
\item Eine Prägarbe besteht also aus abelschen Gruppen $F(U),\ U\subset_\text{o}X$ und Homomorphismen abelscher Gruppen $\operatorname{res}_V^U:F(U)\to F(V)$ für alle offenen $V\subset U$, so dass $\operatorname{res}_U^U=\operatorname{id}_{F(U)}$. Für offene Mengen $W\subset V\subset U$ gelte ferner $\operatorname{res}_W^U=\operatorname{res}_W^V\circ\operatorname{res}_V^U$.
\item Ebenso können Prägarben in eine beliebige Kategorie gebildet werden, z.B. $\textbf{Ringe}$ und $\textbf{Mengen}$.
\item Die Elemente von $F(U)$ heißen \textit{Schnitte} von $F$ über $U$.\index{Schnitt} Manchmal schreiben wir auch $\Gamma(U,F)=F(U)$. Die $\operatorname{res}_V^U$ heißen \textit{Restriktionsabbildungen}\index{Restriktionsabbildung}. Wir schreiben auch $\operatorname{res}_V^U(s)=s|_V$.
\end{enumerate}

\paragraph{Definition 1.2.}\label{1.2} Eine Prägarbe $F$ auf einem topologischen Raum $X$ heißt \textit{Garbe}\index{Garbe}, falls die folgenden Diagramme exakt sind:
\[\xymatrix{0\ar[r] & F(U)\ar[r]_-{\operatorname{res}} & \displaystyle\prod_i F(U_i)\ar@<.5ex>[r]\ar@<-.5ex>[r]_-{\operatorname{res}} & \displaystyle\prod_{i,j}F(U_i\cap U_j)} \]
für alle $U\subset_\text{o}X$ und jede offene Überdeckung $U=\bigcup_i U_i$, d.h:
\begin{enumerate}[(i)]
\item $s\mapsto (\operatorname{res}^U_{U_i}(s))_i$ ist injektiv, d.h. aus $s|_{U_i}=0$ für alle $i$ folgt $s=0$.
\item Sei $s_i\in F(U_i)$ für alle $i$ gegeben, so dass $s_i|_{U_i\cap U_j}=s_j|_{U_i\cap U_j}$ für alle $i,j$. Dann gibt es ein $s\in F(U)$ mit $s|_{U_i}=s_i$ für alle $i$.
\end{enumerate}

\paragraph{Definition 1.3.}\label{1.3} \begin{enumerate}[(i)]
\item Ein \textit{Morphismus}\index{Morphismus!Prägarben} $\varphi:F\to G$ von Prägarben auf $X$ ist ein Morphismus kontravarianter Funktoren, d.h. eine Kollektion von Morphismen $(\varphi(U))_{U\subset_\text{o}X}$, so dass für alle offenen Mengen $V\subset U$ folgendes Diagramm kommutativ ist:
\[\xymatrix{
F(U)\ar[rr]^{\varphi(U)}\ar[d]_{\operatorname{res}^U_V} && G(U)\ar[d]^{\operatorname{res}^U_V}\\
F(V)\ar[rr]_{\varphi(V)} && G(V)
} \]
\item Ein \textit{Morphismus}\index{Morphismus!Garben} von Garben ist ein Morphismus von Prägarben. Die (Prä-)Garben bilden eine Kategorie.
\end{enumerate}

\paragraph{Beispiel 1.4.}\label{1.4} \begin{enumerate}
\item Sei $X$ eine Varietät über $k$. Betrachte den Funktor $\mathcal{O}:\textbf{Top$(X)$}\to\textbf{komm Ringe}$ mit den gewöhnlichen Restriktionsabbildungen $\operatorname{res}^U_V:\mathcal{O}(U)\to\mathcal{O}(V)$ ist offensichtlich eine Prägarbe von Ringen. Da ferner reguläre Funktionen $0$ ist, wenn sie lokal $0$ ist, und eine lokal reguläre Funktion auch global regulär ist, ist $\mathcal{O}$ auch eine Garbe.
\item Sei $X$ ein topologischer Raum und $A$ eine abelsche Gruppe. Die \textit{konstante Garbe}\index{Garbe!konstant} $\mathcal{A}$ auf $X$ ist folgendermaßen definiert: Wir statten $A$ mit der diskreten Topologie aus. Für jedes $U\subset_\text{o}X$ setze:
\[\mathcal{A}(U)=\{f:U\to A\mid f\text{ stetig}\} \]
Ist $U$ zusammenhängend, so gilt $\mathcal{A}(U)\stackrel{\sim}{\to} A,\ f\mapsto f(x)$, wobei $x\in U$ beliebig.
\end{enumerate}

\paragraph{Definition 1.5.}\label{1.5} Sei $F$ eine Prägarbe auf $X$ und $P\in X$. Der \textit{Halm}\index{Halm} $F_P$ von $F$ in $P$ ist definiert als:
\[F_P =\varinjlim_{\substack{U\subset_\text{o}X\\ P\in U}} F(U)=\coprod_{\substack{U\subset_\text{o}X\\ P\in U}} F(U)\Big/{\sim} \]
wobei zwei Elemente $s\in F(U),\ t\in F(V)$ genau dann äquivalent $s\sim t$ sind, wenn es ein $\varnothing \neq W\subset_\text{o}X$ mit $W\subset U\cap V$ existiert, so dass $s|_W=t|_W$ gilt. Die Elemente eines Halms heißen \textit{Keime}\index{Keim} der Schnitte von $F$ in $P$.

\paragraph{Beispiel.} Sei $X$ eine Varietät, $P\in X$ ein Punkt und $\mathcal{O}$ die Garbe der regulären Funktionen. Dann ist der Halm in $P$ gerade der lokale Ring $\mathcal{O}_{P,X}$.

\paragraph{Bemerkung.} Ein Morphismus $\phi:F\to G$ von Prägarben induziert für alle $P\in X$ ein Gruppenhomomorphismus $\phi_P:F_P\to G_P$.

\paragraph{Satz 1.6.}\label{1.6} Sei $\phi:F\to G$ ein Morphismus von Garben auf einem topologischen Raum $X$. Dann gilt:
\[\phi:F\to G\text{ ist Isomorphismus}\iff \phi_P:F_P\to G_P\text{ ist Isomorphismus für alle }P\in X \]
Für Prägarben gilt dieser Satz im Allgemeinen nicht.

\paragraph{Beweis.} Ist $\phi$ ein Isomorphismus, so auch alle $\phi_P,\ P\in X$. Sei umgekehrt $\phi_P$ Isomorphismen für alle $P\in X$. Es genügt zu zeigen, dass $\phi(U):F(U)\to G(U)$ für alle $U\subset_\text{o}X$ ein Isomorphismus ist. Sei $U\subset_\text{o}X$ und setze $\varphi=\phi(U)$.
\begin{itemize}
\item \textit{Injektivität von $\varphi$:} Sei $s\in F(U)$ mit $0=\varphi(s)\in G(U)$. Dann gilt für das Bild $\varphi(s)_P$ von $\varphi(s)$ im Halm $0=\varphi(s)_P\in F_P$. Wegen $\varphi(s)_P=\phi_P(s_P)$ für das Bild $s_P\in F_P$ von $s$, folgt wegen der Injektivität von $\phi_P$ nun $s_P=0$ für alle $P\in U$.

Per Definition gibt es für jedes $P\in U$ eine offene Umgebung $W_P\subset_\text{o}X$ von $P$ mit $W_P\subset U$, so dass $s|_{W_P}=0$ gilt. Dann bilden die $W_P$ eine offene Überdeckung von $U=\bigcup_{P\in U}W_P$. Da $F$ eine Garbe ist, folgt $s=0$.

Wir haben gezeigt, dass $\phi(U)$ für alle $U\subset_\text{o}X$ injektiv ist, genau dann wenn $\phi_P$ für alle $P\in X$ injektiv ist.
\item \textit{Surjektivität von $\varphi$:} Sei $t\in G(U)$ ein Schnitt und $t_P\in G_P$ sein Keim in $P$. Da $\phi_P$ surjektiv ist, existiert ein $s_P\in F_P$ mit $\phi_P(s_P)=t_P$. Sei $s_P$ durch den Schnitt $s(P)\in F(V_P)$ mit $V_P\subset_\text{o}U,\ P\in V_P$ repräsentiert. Dann sind $\phi(V_P)(s(P))$ und $t|_{V_P}$ zwei Elemente aus $G(V_P)$ mit demselben Keim. Durch das Verkleinern von $V_P$ folgt $\phi(V_P)(s(P))=t|_{V_P}$ in $G(V_P)$.

Dann bilden die $V_P$ eine offene Überdeckung von $U=\bigcup_{P\in U}V_P$. Es gilt außerdem $s(P)|_{V_P\cap V_Q}=s(Q)|_{V_P\cap V_Q}$ für alle $P,Q\in U$, denn beide Elemente sind Schnitte aus $F(V_P\cap V_Q)$, die durch $\phi(V_P\cap V_Q)$ auf $t|_{V_P\cap V_Q}$ abgebildet werden, und $\phi(V_P\cap V_Q)$ aus dem ersten Teil injektiv ist.

Da $F$ eine Garbe ist, existiert ein $s\in F(U)$ mit $s|_{V_P}=s(P)$ für alle $P\in U$. Schließlich gilt $\varphi(s)|_{V_P}=t|_{V_P}$ für alle $P\in U$, d.h. $(\varphi(s)-t)|_{V_P}=0$. Da $G$ eine Garbe ist, folgt $\varphi(s)=t$.\qed
\end{itemize}

\paragraph{Definition 1.7.}\label{1.7} Sei $\varphi:F\to G$ ein Morphismus von Prägarben. Die Prägarben
\[U\mapsto \ker\varphi(U),\quad U\mapsto\operatorname{coker}\varphi(U),\quad U\mapsto\operatorname{im}\varphi(U) \]
heißen \textit{Prägarbenkern}, \textit{-kokern} und \textit{-bild} von $\varphi$.\index{Kern}\index{Kokern}\index{Bild} Sind $F$ und $G$ Garben, so sind Kokern und Bild nicht notwendig Garben.

\paragraph{Satz \& Definition 1.8.}\label{1.8} Sei $F$ eine Prägarbe. Dann existiert eine Garbe $F^+$ und ein Morphismus von Prägarben $\theta:F\to F^+$ mit folgender Universaleigenschaft:

Sei $G$ eine Garbe und $\phi:F\to G$ ein Morphismus von Prägarben. Dann existiert ein eindeutig bestimmter Morphismus $\psi:F^+\to G$, so dass das folgende Diagramm kommutiert:
\[\xymatrix{ 
F\ar[rr]^-\theta\ar[d]_\phi && F^+\ar@{-->}[dll]^\psi\\
G
} \]
$F^+$ ist somit eindeutig bestimmt und heißt die zu $F$ \textit{assoziierte Garbe}\index{assoziierte Garbe!Prägarbe}.

\paragraph{Beweis.} Für jede offene Menge $U\subset X$ setze $F^+(U)$ als die Menge aller Abbildungen $s:U\to\coprod_{P\in U}F_P$, so dass:
\begin{enumerate}[(i)]
\item Für alle $P\in U$ gilt $s(P)\in F_P$.
\item Für alle $P\in U$ gibt es eine offene Umgebung $V$ von $P$ mit $V\subset U$ und ein Element $t\in F(V)$, so dass für alle $Q\in V$ der Keim $t_Q$ von $t$ in $Q$ gleich $s(Q)$ ist.
\end{enumerate}
Somit wird $F^+$ zu einer Garbe bzgl. der natürlichen Restriktionsabbildungen und besitzt die verlangte Universaleigenschaft. Für jeden Punkt $P\in X$ gilt $F^+_P=F_P$. Ist $F$ eine Garbe, so ist $F^+\cong F$ via $\theta$.\qed

\paragraph{Definition 1.9.}\label{1.9} 
\begin{enumerate}[(i)]
\item Eine \textit{Untergarbe}\index{Garbe!Untergarbe} von $F$ ist eine Garbe $F'$ derart, dass:
\begin{enumerate}[(a)]
\item $F'(U)\subset F(U)$ ist eine Untergruppe für alle $U\subset_\text{o}X$.
\item Für offene Mengen $V\subset U$ gilt $\operatorname{res}'{}^U_V=\operatorname{res}^U_V|_{F'(U)}$.
\end{enumerate}
Insbesondere ist $F'_P\subset F_P$ eine Untergruppe.
\item Der \textit{Kern}\index{Kern} von $\varphi$ ist die Prägarbe $\ker(\varphi)$, die bereits eine Garbe ist. \textit{Grund:}

Sei $U\subset_\text{o}X$ und $U=\bigcup U_i$ eine offene Überdeckung. Sei $s\in\ker\varphi(U)$ mit $s|_{U_i}=0$ für alle $i$. Da $F$ eine Garbe ist und $s\in F(U)$, folgt $s=0$. Sei nun $s_i\in \ker\varphi(U_i)$ für alle $i$ gegeben mit $s_i|_{U_i\cap U_j}=s_j|_{U_i\cap U_j}$ für alle $i,j$. Da $F$ eine Garbe ist, existiert ein $s\in F(U)$ mit $s|_{U_i}=s_i$ für alle $i$. Zu zeigen ist noch $s\in \ker\varphi(U)$. Es gilt für alle $i$:
\[0=\varphi(U_i)(s_i)=\varphi(U_i)(s|_{U_i}) = \varphi(U)(s)|_{U_i}\in G(U_i) \]
Da nun auch $G$ eine Garbe ist, folgt $\varphi(U)(s)=0$.
\item $\varphi$ heißt \textit{injektiv}\index{Injektivität}, falls $\ker(\varphi)=0$.

Mit anderen Worten: $\varphi$ ist genau dann injektiv, wenn $\varphi(U):F(U)\to G(U)$ für alle $U\subset_\text{o}X$ injektiv ist.
\item Das \textit{Bild}\index{Bild} $\operatorname{im}(\varphi)$ von $\varphi$ ist die assoziierte Garbe des Prägarbenbilds von $\varphi$.

Nach der Universaleigenschaft gibt es einen natürlichen Morphismus $\psi:\operatorname{im}(\varphi)\to G$. Dieser ist injektiv, da $(\operatorname{im}\varphi)_P:\operatorname{im}(\varphi_P) \to G_P$ für alle $P\in X$ injektiv ist.
\item $\varphi$ heißt \textit{surjektiv}\index{Surjektivität}, wenn $\operatorname{im}(\varphi)=G$.
\item Eine Garbensequenz
\[\xymatrix{\cdots \ar[r] & F^i\ar[r]^{\varphi^i} & F^{i+1}\ar[r]^{\varphi^{i+1}} & F^{i+2}\ar[r] & \cdots } \]
heißt \textit{exakt}\index{Exaktheit}, falls $\ker(\varphi^{i+1})=\operatorname{im}(\varphi^i)$ für alle $i$ gilt.
\item Sei $F'$ eine Untergarbe von $F$. Die \textit{Quotientengarbe}\index{Quotientengarbe} $F/F'$ ist die assoziierte Garbe zur Prägarbe $U\mapsto F(U)/F'(U)$. Offensichtlich gilt $(F/F')_P=F_P/F'_P$ für alle $P\in X$.
\item Der \textit{Kokern}\index{Kokern} von $\varphi$ ist die assoziierte Garbe zum Prägarbenkokern von $\varphi$.
\end{enumerate}

\paragraph{Regeln 1.10.}\label{1.10} Seien $F,G$ Garben auf $X$ und $\varphi:F\to G$ ein Morphismus von Garben.
\begin{enumerate}[(i)]
\item $\varphi$ ist genau dann injektiv, wenn $0\to F\to G$ exakt ist und genau dann surjektiv, wenn $F\to G\to 0$ exakt ist.
\item Eine Garbensequenz $\cdots \to F^i\to F^{i+1}\to F^{i+2}\to\cdots$ ist genau dann exakt, wenn ihre entsprechenden Halmsequenzen in allen Punkten $P\in X$ exakt ist. \textit{Grund:}
\[(\operatorname{im} \varphi^i)_P=\operatorname{im}(\varphi^i_P),\quad (\ker\varphi^{i+1})_P=\ker(\varphi^{i+1}_P) \]
Insbesondere ist ein Garbenmorphismus genau dann injektiv bzw. surjektiv falls alle Halmabbildungen injektiv bzw. surjektiv sind.
\item $\varphi$ ist genau dann surjektiv, wenn für alle $U\subset_\text{o}X$ und $s\in G(U)$ eine Überdeckung $U=\bigcup_{i\in I} U_i$ mit Schnitten $t_i\in F(U_i),\ i\in I$ existieren, so dass $\varphi(t_i)=s|_{U_i}$.

Ist $\varphi$ surjektiv, so muss im Allgemeinen $\varphi(U):F(U)\to G(U)$ nicht surjektiv sein.
\item Sei $0\to F'\to F\to F''$ eine exakte Garbensequenz und $U\subset_\text{o}X$. Dann ist auch die folgende Sequenz exakt:
\[\xymatrix{
0 \ar[r] & \Gamma(U,F') \ar[r]& \Gamma(U,F) \ar[r]& \Gamma(U,F'')
} \]
Der Funktor $\Gamma(U,-)$ ist linksexakt, aber nicht exakt.
\item Sei $\varphi:F\to G$ ein injektiver Morphismus von Prägarben. Dann ist der induzierte Morphismus $\varphi^+:F^+\to G^+$ der assoziierten Garben auch injektiv. Der Funktor $-^+$ ist sogar exakt.
\end{enumerate}

\paragraph{Regeln 1.11.}\label{1.11} \begin{enumerate}[(i)]
\item Sei $F'$ eine Untergarbe von $F$. Dann ist die Sequenz $0\to F'\to F\to F/F'\to 0$ exakt, da sie halmweise exakt ist.
\item Sei $\varphi:F\to G$ ein Garbenmorphismus. Dann gilt:
\[\operatorname{im}(\varphi)\cong F/\ker(\varphi),\quad \operatorname{coker}(\varphi)\cong G/ \operatorname{im}(\varphi) \]
d.h. die Folgen $0\to\ker(\varphi)\to F\to\operatorname{im}(\varphi)\to 0$ und $0\to\operatorname{im}(\varphi)\to G\to\operatorname{coker}(\varphi)\to 0$ sind exakt.
\end{enumerate}

\paragraph{Definition 1.12.}\label{1.12} \begin{enumerate}[(i)]
\item Seien $F$ und $G$ Garben auf $X$. Die \textit{Summe}\index{Summe} $F\oplus G$ von $F$ und $G$ ist die Garbe $U\mapsto F(U)\oplus G(U)$.
\item Sei $(F_i,\varphi_{i,j})$ ein direktes System von Garben auf $X$. Der \textit{direkte Limes}\index{Limes!direkt} $(\varinjlim F_i,\varphi_i)$ ist die assoziierte Garbe zur Prägarbe $U\mapsto \varinjlim F_i(U)$.

Der direkte Limes besitzt die übliche Universaleigenschaft: Sei $G$ eine Garbe und $\psi_i:F_i\to G$ Morphismen mit $\varphi_k\varphi_{ik}=\psi_i$ für alle $i\leq k$. Dann existiert ein eindeutig bestimmter Morphismus $\psi:\varinjlim F_i\to G$, so dass das folgende Diagramm kommutiert:
\[\xymatrix{
F_i\ar[rr]^{\varphi_i}\ar[d]_{\psi_i} & & \varinjlim F_i\ar@{-->}[lld]^\psi\\
G
} \]
\item Ebenso wird der \textit{projektive Limes}\index{Limes!projektiv} definiert, wobei alle Pfeile umgedreht werden. Es ist außerdem $U\mapsto\varprojlim F_i(U)$ bereits eine Garbe.
\item Sei $F$ eine Garbe auf $X$ und $s\in F(U)$ ein Schnitt über $U\subset_\text{o}X$. Dann heißt
\[\operatorname{Supp}(s)=\{P\in U\mid s_P\neq 0\} \]
wobei $s_P\in F_P$ der Keim von $s$ in $P$ bezeichnet, der \textit{Support}\index{Support} von $s$. $\operatorname{Supp}(s)$ ist abgeschlossen in $U$.
\[\operatorname{Supp}(F)=\{P\in X\mid F_P\neq 0\} \]
heißt \textit{Support} von $F$. Dieser ist nicht notwendigerweise abgeschlossen.
\item Seien $F,G$ Garben abelscher Gruppen auf $X$. Für ein $U\subset_\text{o}X$ sei $F|_U$ die Einschränkung von $F$ auf $U$, d.h. $F|_U(V)=F(V)$ für alle $V\subset_\text{o}U$. Dann ist die Menge $\operatorname{Hom}(F|_U,G|_U)$ der Morphismen von $F|_U$ nach $G|_U$ eine abelsche Gruppe.
\[U\mapsto\operatorname{Hom}(F|_U,G|_U)\]
definiert eine Garbe und wird die \textit{Hom-Garbe}\index{Hom-Garbe} genannt. Sie wird mit $\operatorname{\mathcal{H}om}(F,G)$ bezeichnet.
\end{enumerate}

\paragraph{Definition 1.13.}\label{1.13} Sei $f:X\to Y$ eine stetige Abbildung topologischer Räume, $F$ eine Garbe auf $X$ und $G$ eine Garbe auf $Y$.
\begin{enumerate}[(i)]
\item Die \textit{direkte Bildgarbe}\index{direkte Bildgarbe} $f_\ast F$ von $F$ auf $Y$ ist die Garbe
\[V\mapsto (f_\ast F)(V)=F(f^{-1}(V)) \]
\item Die \textit{Urbildgarbe}\index{Urbildgarbe} $f^{-1}G$ von $G$ auf $X$ ist die assoziierte Garbe zur Prägarbe
\[U\mapsto (f^{-1}G)(U)= \varinjlim_{\substack{V\subset_\text{o}Y\\ f(U)\subset V}}G(V) \]
\end{enumerate}

\paragraph{Regeln 1.14.}\label{1.14} Seien $X,Y$ topologische Räume.
\begin{enumerate}[(i)]
\item Sei $Z\subset X$ ein Teilraum mit der Inklusionsabbildung $i:Z\hookrightarrow X$ und $F$ eine Garbe auf $X$. Dann gilt:
\[i^{-1}F = F|_Z \]
Offensichtlich gilt $(F|_Z)_P=F_P$ für $P\in Z$.
\item Seien $\mathbf{Ab}(X)$ und $\mathbf{Ab}(Y)$ die Kategorien der Garben auf $X$ bzw. $Y$. Dann sind $f_\ast:\mathbf{Ab}(X)\to\mathbf{Ab}(Y)$ und $f^{-1}:\mathbf{Ab}(Y)\to\mathbf{Ab}(X)$ Funktoren.
\item Sei $f:X\to Y$ stetig. Dann sind die \textit{Adjunktionsabbildungen}\index{Adjunktionsabbildung} 
\[\operatorname{ad}:f^{-1}f_\ast F\to F,\quad\operatorname{ad} :G\to f_\ast f^{-1}G \]
für Garben $F$ auf $X$ bzw. $G$ auf $Y$ Garbenmorphismen und wie folgt definiert:
\begin{itemize}
\item Sei $U\subset_\text{o}X$ und $s\in (f^{-1}f_\ast F)(U)$ ein Schnitt, das durch $s'\in (f_\ast F)(V)$ mit $f(U)\subset V\subset_\text{o}Y$ repräsentiert wird, d.h. $s'\in F(f^{-1}(V))$ mit $U\subset f^{-1}(V)$. Dann setzen wir: 
\[(f^{-1}f_\ast F)(U)\to F(U),\ s\mapsto \operatorname{res}_U^{f^{-1}(V)}s'\in F(U)\]
\item Sei $V\subset_\text{o}Y$. Es besteht $(f_\ast f^{-1}G)(V)=(f^{-1}G)(f^{-1}(V))$ aus Abbildungen der Form $f^{-1}(V)\to\coprod_{P\in f^{-1}(V)}(f^{-1}G)_P$. Setze nun:
\[G(V)\to (f_\ast f^{-1}G)(V),\ s\mapsto (s\circ f: f^{-1}(V)\to\coprod  f^{-1}(G)_P,\ P\mapsto s_{f(P)}) \]
\end{itemize}
Es existiert eine natürliche Bijektion:
\[\operatorname{Hom}_X(f^{-1}G,F)\cong\operatorname{Hom}_Y(G,f_\ast F) \]
in dem wir ein $\varphi:f^{-1}(G)\to F$ auf $\psi: G\stackrel{\operatorname{ad}}{\longrightarrow} f_\ast f^{-1}G\stackrel{f_\ast(\varphi)}{\longrightarrow} f_\ast F$ schicken und ein $\psi:G\to f_\ast F$ auf $\varphi:f^{-1}G \stackrel{f^{-1}(\psi)}{\longrightarrow} f^{-1}f_\ast F\stackrel{\operatorname{ad}}{\longrightarrow}F$ schicken. Somit ist $f^{-1}$ linksadjungiert zu $f_\ast$.
\end{enumerate}

\paragraph{Definition 1.15.}\label{1.15} Sei $X$ ein topologischer Raum mit $P\in X$ und $A$ eine abelsche Gruppe. Sei $\mathcal{A}$ die konstante Garbe auf $\overline{\{P\}}$ und $i:\overline{\{P\}}\hookrightarrow X$ die natürliche Inklusion. Dann heißt die Garbe $i_\ast \mathcal{A}$ \textit{Wolkenkratzergarbe}. Es gilt:
\[(i_\ast\mathcal{A})(U) = \begin{cases}
A, &\text{wenn } P\in U\\
0, &\text{sonst}
\end{cases},\quad (i_\ast \mathcal{A})_Q = \begin{cases}
A,&\text{wenn } Q\in\overline{\{P\}}\\
0,&\text{sonst}
\end{cases} \]

\paragraph{Definition 1.16.}\label{1.16} Sei $X$ ein topologischer Raum, $Z\subset X$ abgeschlossen und $U=X\setminus Z$. Seien $j:U\hookrightarrow X$, $i:Z\hookrightarrow X$ die Inklusionsabbildungen. Ist $F$ eine Garbe auf $Z$, so gilt:
\[(i_\ast F)_P=\begin{cases}
F_P, &\text{wenn }P\in Z\\
0, &\text{sonst}
\end{cases} \]
Sei $F$ eine Garbe auf $U$ und $j_!(F)$ die Garbe auf $X$, die zur Prägarbe
\[V\mapsto\begin{cases}
F(V),&\text{wenn }V\subset U\\
0,&\text{sonst}
\end{cases} \]
assoziiert ist. Sie heißt die \textit{außerhalb $U$ durch Null fortgesetzte Garbe}\index{Garbe!Fortsetzung} von $F$. Es gilt:
\[(j_!F)_P=\begin{cases}
F_P,&\text{wenn }P\in U\\
0,&\text{sonst}
\end{cases} \]
Sei $F$ eine Garbe auf $X$. Dann ist die folgende Sequenz exakt:
\[\xymatrix{
0\ar[r] & j_!(F|_U)\ar[r] & F \ar[r]& i_\ast(F|_Z)\ar[r] & 0
} \]

\section{Schemata}

Sei $A$ ein kommutativer Ring mit Eins und $\operatorname{Spec}(A)$ die Menge aller Primideale von $A$. Für ein Ideal $\mathfrak{a}\subset A$, setzen wir:
\[V(\mathfrak{a}) = \{\mathfrak{p}\in\operatorname{Spec}(A)\mid \mathfrak{a}\subset\mathfrak{p}\} \]

\paragraph{Lemma 2.1.}\label{2.1} \begin{enumerate}[(i)]
\item Sind $\mathfrak{a},\mathfrak{b}$ zwei Ideale von $A$, so gilt $V(\mathfrak{ab})=V(\mathfrak{a})\cup V(\mathfrak{b})$.
\item Sind $\mathfrak{a}_i\subset A,\ i\in I$ Ideale, so gilt $V\big(\sum\mathfrak{a}_i\big)=\bigcap V(\mathfrak{a}_i)$.
\item Sind $\mathfrak{a},\mathfrak{b}\subset A$ Ideale, gilt: $V(\mathfrak{a})\subset V(\mathfrak{b}) \iff \operatorname{Rad}(\mathfrak{a})\supset\operatorname{Rad}(\mathfrak{b})$
\end{enumerate}
Wegen $V(A)=\varnothing$ und $V(0)=\operatorname{Spec}(A)$ sehen wir, dass wir Teilmengen der Form $V(\mathfrak{a})$ zu abgeschlossene Mengen in $\operatorname{Spec}(A)$ erklären können. Somit erhalten wir die \textit{Zariski-Topologie}\index{Zariski-Topologie} auf $\operatorname{Spec}(A)$.

Setzen wir $D(f)=\operatorname{Spec}(A)\setminus V(f)$ für $f\in A$, so bilden diese offene Mengen eine Basis der Topologie auf $\operatorname{Spec}$.

\paragraph{Definition 2.2.}\label{2.2} Wir definieren eine Ringgarbe $\mathcal{O}$ auf $\operatorname{Spec}(A)$ wie folgt: Sei $U\subset_\text{o}\operatorname{Spec}(A)$. Setze $\mathcal{O}(U)$ als die Menge aller Abbildungen $s:U\to\coprod_{\mathfrak{p}\in U}A_\mathfrak{p}$, so dass:
\begin{enumerate}[(i)]
\item Für alle $\mathfrak{p}\in U$ gilt $s(\mathfrak{p})\in A_\mathfrak{p}$.
\item Für alle $\mathfrak{p}\in U$ gibt es eine offene Umgebung $V$ von $\mathfrak{p}$ mit $V\subset U$ und Elemente $a,f\in A$, so dass für alle $\mathfrak{q}\in V$ stets $f\not\in\mathfrak{q}$ und $s(\mathfrak{q})=\frac{a}{f}$ in $A_\mathfrak{q}$ gilt.
\end{enumerate}
Offensichtlich ist mit $s,t\in\mathcal{O}(U)$ auch $s+t,st\in\mathcal{O}(U)$. Ferner ist für $V\subset U$ offen $\operatorname{res}^U_V:\mathcal{O}(U)\to\mathcal{O}(V)$ ein Ringhomomorphismus. $\mathcal{O}$ heißt \textit{Strukturgarbe}\index{Strukturgarbe}. Das \textit{Spektrum}\index{Spektrum} von $A$ ist das Paar $(\operatorname{Spec}A,\mathcal{O})$.

\paragraph{Satz 2.3.}\label{2.3} Sei $A$ ein Ring und $(\operatorname{Spec} A,\mathcal{O})$ sein Spektrum. Dann gilt:
\begin{enumerate}[(i)]
\item Der Halm $\mathcal{O}_\mathfrak{p}$ ist isomorph zu $A_\mathfrak{p}$ für alle $\mathfrak{p}\in\operatorname{Spec}(A)$.
\item $\mathcal{O}(D(f))\cong A_f$ für alle $f\in A$
\item $\Gamma(\operatorname{Spec}(A),\mathcal{O})=A$
\end{enumerate}

\paragraph{Beweis.} (iii) folgt aus (ii) mit $f=1$.
\begin{enumerate}[(i)]
\item Die Abbildungen $\mathcal{O}(U)\to A_\mathfrak{p},\ s\mapsto s(\mathfrak{p})$ mit $\mathfrak{p}\in U\subset_\text{o}\operatorname{Spec}(A)$ sind kompatibel und induzieren einen Homomorphismus $\varphi:\mathcal{O}_\mathfrak{p}\to A_\mathfrak{p}$.
\begin{itemize}
\item \textit{Surjektivität:} Sei $\frac{a}{f}\in A_\mathfrak{p}$ mit $a,f\in A,\ f\not\in\mathfrak{p}$. Dann ist $D(f)$ eine offene Umgebung von $\mathfrak{p}$ und es gibt ein $s\in\mathcal{O}(D(f))$ mit $s(\mathfrak{p})=\frac{a}{f}$.
\item \textit{Injektivität:} Sei $U$ eine Umgebung von $\mathfrak{p}$ und $s,t\in\mathcal{O}(U)$ mit $s(\mathfrak{p})=t(\mathfrak{p})$. Verkleinern wir $U$ wenn nötig, so können wir o.B.d.A. annehmen, dass:
\[s=\frac{a}{f},\quad t=\frac{b}{g}\quad \text{ für gewisse } a,b,g,f\in A,\ g,f\not\in\mathfrak{p} \]
Wegen $\frac{a}{f}=\frac{b}{g}$ in $A_\mathfrak{p}$, gibt es ein $h\not\in\mathfrak{p}$, so dass $h(ga-bf)=0$ in $A$. Insbesondere ist $\frac{a}{f}=\frac{b}{g}$ in $A_\mathfrak{q}$ für alle $\mathfrak{q}\in\operatorname{Spec}(A)$ mit $g,f,h\not\in\mathfrak{q}$. Somit ist $s=t$ auf der offenen Umgebung $D(f)\cap D(g)\cap D(h)$ von $\mathfrak{p}$ und haben daher denselben Keim.
\end{itemize}
\item Sei $f\in A$ und $\mathfrak{p}\in D(f)$, d.h. $(f)\subset A\setminus\mathfrak{p}$. Betrachte die kanonische Abbildung $\lambda_\mathfrak{p}:A_f\to A_\mathfrak{p}$. Setze:
\[\psi: A_f \to\mathcal{O}(D(f)),\ \frac{a}{f^n}\mapsto \Big(\mathfrak{p}\mapsto \lambda_\mathfrak{p}\Big(\frac{a}{f^n}\Big)\Big) \]
\begin{itemize}
\item \textit{Injektivität:} Sei $\psi\big(\frac{a}{f^n}\big)=\psi \big(\frac{b}{f^m}\big)$ und $\mathfrak{p}\in D(f)$. Dann ist $\frac{a}{f^n}=\frac{b}{f^m}$ in $A_\mathfrak{p}$, d.h. es gibt ein $h\not\in\mathfrak{p}$ mit $h(f^ma-f^nb)=0$. Setze $\mathfrak{a}=\operatorname{Ann}(f^ma-f^nb)$. Dann ist $\mathfrak{a}\not\subset\mathfrak{p}$, da $h\in\mathfrak{a}$, also folgt $\mathfrak{p}\not\in V(\mathfrak{a})$. Wir haben also $V(\mathfrak{a})\subset V(f)$ gezeigt. Nach \hyperref[2.1]{Lemma 2.1 (iii)} folgt $f\in\operatorname{Rad}(\mathfrak{a})$, d.h. $f^e\in\mathfrak{a}$ für ein $e>0$. Per Definition gilt $f^e(f^na-f^mb)=0$, also $\frac{a}{f^n}=\frac{b}{f^m}$ in $A_f$.
\item \textit{Surjektivität:} Sei $s\in\mathcal{O}(D(f))$. Nach Definition von $\mathcal{O}$ ist $D(f)=\bigcup V_i$ mit $s=\frac{a_i}{g_i}$ auf $V_i$ für gewisse $a_i,g_i\in A,\ g_i\not\in\mathfrak{p}$ für alle $\mathfrak{p}\in V_i$. Insbesondere gilt $V_i\subset D(g_i)$. Da die $D(h)$ eine Basis der Topologie bilden, können wir o.B.d.A. $V_i=D(h_i)$ annehmen, also $D(h_i)\subset D(g_i)$. Es folgt $V(h_i)\supset V(g_i)$ und nach \hyperref[2.1]{Lemma 2.1 (iii)} auch $\operatorname{Rad}(h_i)\subset\operatorname{Rad}(g_i)$. Wähle ein $n$, so dass $h_i^n\in (g_i)$ für alle $i$, d.h. $h_i^n=c_ig_i$ für ein $c_i\in A$. Es folgt:
\[\frac{a_i}{g_i}=\frac{c_ia_i}{h_i^n}\]
Ersetzt man $a_i$ durch $c_ia_i$ und $h_i$ durch $h_i^n$, so können wir o.B.d.A. $D(f)\subset\bigcup D(h_i)$ und $s=\frac{a_i}{h_i}$ auf $D(h_i)$ annehmen.

Wir zeigen nun, dass $D(f)$ durch endlich viele $D(h_i)$ überdeckt werden kann. Wir haben mit \hyperref[2.1]{Lemma 2.1} Äquivalenzen:
\begin{align*}
D(f)\subset \bigcup_{i}D(h_i) &\iff V(f)\supset\bigcap_i V(h_i) = V\Big(\sum_i (h_i)\Big)\\
&\iff f\in\operatorname{Rad}\Big(\sum_i (h_i)\Big)\\
&\iff \exists n\in\mathbb{N}\colon f^n\in\sum_i (h_i)
\end{align*}
Daher ist $f^n$ eine endliche Summe der Form $f^n=\sum_{i=1}^r b_ih_i$ für gewisse $b_i\in A$, d.h. $D(f)\subset D(h_1)\cup\cdots\cup D(h_r)$.

Nun gilt: \[D(h_i)\cap D(h_j)=\operatorname{Spec}(A)\setminus (V(h_i)\cup V(h_j))=\operatorname{Spec}(A)\setminus V(h_ih_j)=D(h_ih_j)\]
Auf $D(h_ih_j)$ wird $s$ repräsentiert durch $\frac{a_i}{h_i}$ und $\frac{a_j}{h_j}$ in $A_{h_ih_j}$. Wenden wir die Injektivität von $\psi$ auf $D(h_ih_j)$ an, so erhalten wir $\frac{a_i}{h_i}=\frac{a_j}{h_j}$ in $A_{h_ih_j}$. Es folgt $(h_ih_j)^n(h_ja_i-h_ia_j)=0$ für ein $m$. Sei $m$ so groß, dass dies für alle endlich vielen $i,j$ gilt, also gilt für alle $i,j$:
\[h_j^{m+1}(h_i^ma_i)-h_i^{m+1}(h_j^ma_j)=0 \]
Ersetzen wir nun $h_i$ durch $h_i^{m+1}$ und $a_i$ durch $a_ih_i^m$, so wird $s$ auf $D(h_i)$ immer noch durch $\frac{a_i}{h_i}$ repräsentiert und es gilt $h_ja_i=h_ia_j$ für alle $i,j$.

Schreibe nun $f^n=\sum b_ih_i$ für ein $n$ und setze $a=\sum b_ia_i$. Es folgt für alle $j$:
\[h_ja = \sum_i b_ia_ih_j=\sum _i b_ih_ia_j=f^na_j \]
Also gilt $\frac{a}{f^n}=\frac{a_j}{h_j}$ auf $D(h_j)$ für alle $j$, d.h. $\psi\big(\frac{a}{f^n}\big)=s$.\qed
\end{itemize}
\end{enumerate}

\paragraph{Definition 2.4.}\label{2.4} \begin{enumerate}[(i)]
\item Ein \textit{geringter Raum}\index{geringter Raum} $(X,\mathcal{O}_X)$ besteht aus einem topologischen Raum $X$ und einer Ringgarbe $\mathcal{O}_X$ auf $X$. Ein \textit{Morphismus von geringten Räumen}\index{Morphismus!geringte Räume} ist ein Paar
\[(f,f^\sharp):(X,\mathcal{O}_X)\to (Y,\mathcal{O}_Y) \]
wobei $f:X\to Y$ eine stetige Abbildung und $f^\sharp:\mathcal{O}_Y\to f_\ast\mathcal{O}_X$ ein Morphismus von Ringgarben auf $Y$ ist.
\item Ein geringter Raum $(X,\mathcal{O}_X)$ heißt \textit{lokal}\index{geringter Raum!lokal}, falls für alle $P\in X$ der Halm $\mathcal{O}_{X,P}$ ein lokaler Ring ist. Ein \textit{Morphismus von lokal geringten Räumen}\index{Morphismus!lokal geringte Räume} ist ein Morphismus $(f,f^\sharp)$ von geringten Räumen derart, dass für alle $P\in X$ die induzierte Abbildung $f_P^\sharp:\mathcal{O}_{Y,f(P)}\to\mathcal{O}_{X,P}$ lokale Homomorphismen sind.
\end{enumerate}

\paragraph{Bemerkung 2.5.}\label{2.5}\begin{itemize}
\item Die (lokal) geringte Räume bilden eine Kategorie.
\item Ein Morphismus $(f,f^\sharp)$ von (lokal) geringten Räumen ist genau dann ein Isomorphismus, wenn $f$ ein Homöomorphismus ist und $f^\sharp$ ein Garbenisomorphismus ist.
\end{itemize}

\paragraph{Satz 2.6.}\label{2.6} Seien $A,B$ Ringe.
\begin{enumerate}[(i)]
\item $(\operatorname{Spec}A,\mathcal{O})$ ist ein lokal geringter Raum.
\item Sei $\varphi:A\to B$ ein Ringhomomorphismus. Dann induziert $\varphi$ einen natürlichen Morphismus von lokal geringten Räumen:
\[(f,f^\sharp):(\operatorname{Spec}B,\mathcal{O}_B)\to (\operatorname{Spec}A,\mathcal{O}_A),\ f(\mathfrak{p})=\varphi^{-1}(\mathfrak{p}) \]
\item Jeder Morphismus $(f,f^\sharp):(\operatorname{Spec}B,\mathcal{O}_B)\to(\operatorname{Spec}A,\mathcal{O}_A)$ von lokal geringten Räu\-men ist induziert von einem Ringhomomorphismus $\varphi:A\to B$.
\end{enumerate}

\paragraph{Beweis.} \begin{enumerate}[(i)]
\item folgt aus \hyperref[2.3]{Satz 2.3 (i)}.
\item Definiere $f$ durch $f(\mathfrak{p})=\varphi^{-1}(\mathfrak{p})$ für alle $\mathfrak{p}\in\operatorname{Spec}(B)$. Sei $\mathfrak{a}\subset A$ ein Ideal. Dann ist $f^{-1}(V(\mathfrak{a}))=V(\varphi(\mathfrak{a}))$. Daher ist $f$ stetig. Sei $\mathfrak{p}\in\operatorname{Spec}(B)$. Dann liefert $\varphi$ einen lokalen Homomorphismus $\varphi_\mathfrak{p}:A_{\varphi^{-1}(\mathfrak{p})}\to B_\mathfrak{p}$. Das liefert für $V\subset_\text{o}\operatorname{Spec}(A)$ einen Ringhomomorphismus:
\[f^\sharp(V):\mathcal{O}_A(V)\to\mathcal{O}_B(f^{-1}(V)) \]
indem man eine Abbildung $s:V\to\coprod_{\mathfrak{q}\in V} A_\mathfrak{q}$ auf die folgende Abbildung $f^\sharp(s): f^{-1}(V)\to\coprod_{\mathfrak{p}\in f^{-1}(V)}B_\mathfrak{p}$ schickt:
\[ \xymatrix{
f^{-1}(V) \ar[r]^-f\ar@{-}[d] & V\ar[r]^-s\ar@{-}[d] & A_{\varphi^{-1}(\mathfrak{p})}\ar[r]^-{\varphi_\mathfrak{p}}\ar@{-}[d] & B_\mathfrak{p}\ar@{-}[d]\\
\mathfrak{p}\ar@{|->}[r] & f(\mathfrak{p}) = \varphi^{-1}(\mathfrak{p})\ar@{|->}[r] & s(\varphi^{-1}(\mathfrak{p}))\ar@{|->}[r] & \varphi_\mathfrak{p}(s(\varphi^{-1}(\mathfrak{p})))
} \]
Die $f^\sharp(V)$ gibt uns einen Garbenmorphismus $f^\sharp:\mathcal{O}_A\to f_\ast \mathcal{O}_B$. Die durch $f^\sharp$ induzierte Abbildungen auf den Halmen sind gerade die $\varphi_\mathfrak{p}$. Somit ist $(f,f^\sharp)$ ein Morphismus lokal geringten Räumen.
\item Sei $(f,f^\sharp):(\operatorname{Spec}B,\mathcal{O}_B)\to(\operatorname{Spec}A,\mathcal{O}_A)$ ein Morphismus von lokal geringten Räu\-men. $f^\sharp$ induziert einen Ringhomomorphismus:
\[\varphi: A=\Gamma(\operatorname{Spec}A,\mathcal{O}_A)\to\Gamma(\operatorname{Spec}B,\mathcal{O}_B) =B \]
Sei $\mathfrak{p}\in\operatorname{Spec}(B)$. Dann haben wir induzierte lokale Homomorphismen mit kommutativem Diagramm:
\[\xymatrix{
A_{f(\mathfrak{p})} = \mathcal{O}_{A,f(\mathfrak{p})}  \ar[r]^-{f^\sharp_\mathfrak{p}} & \mathcal{O}_{B,\mathfrak{p}} = B_\mathfrak{p}\\
A \ar[r]_\varphi\ar[u] & B\ar[u]
} \]
Da die $f^\sharp_\mathfrak{p}$ lokale Homomorphismen sind, folgt $f(\mathfrak{p})=\varphi^{-1}(\mathfrak{p})$. Somit ist $f^\sharp$ von dem Ringhomomorphismus $\varphi$ induziert.\qed
\end{enumerate}

\paragraph{Definition 2.7.}\label{2.7} \begin{enumerate}[(i)]
\item Ein \textit{affines Schema}\index{Schema!affin} ist ein lokal geringter Raum $(X,\mathcal{O}_X)$, der als lokal geringter Raum isomorph zum Spektrum eines Rings ist.
\item Ein \textit{Schema}\index{Schema} ist ein lokal geringter Raum $(X,\mathcal{O}_X)$ derart, dass jeder Punkt $P\in X$ eine offene Umgebung $U\subset X$ besitzt, so dass $(U,\mathcal{O}_X|_U)$ ein affines Schema ist.
\end{enumerate}
$\mathcal{O}_X$ heißt \textit{Strukturgarbe}\index{Strukturgarbe}. Ein \textit{Morphismus}\index{Morphismus!Schemata} von Schemata ist ein Morphismus von lokal geringten Räumen.

\paragraph{Beispiel 2.8.}\label{2.8}\begin{enumerate}
\item Sei $k$ ein Körper. $\operatorname{Spec}(k)$ ist ein affines Schema, dessen topologischer Raum aus einem Punkt besteht.
\end{enumerate}

\paragraph{Definition 2.9.}\label{2.9} Sei $K$ ein Körper. Eine \textit{diskrete Bewertung}\index{Bewertung!diskret} von $K$ ist eine Abbildung $v:K\to\mathbb{Z}\cup\{\infty\}$, so dass für alle $x,y\in K$ gilt:
\begin{enumerate}[(i)]
\item $v(xy)=v(x)+v(y)$
\item $v(x+y)\geq \min\{v(x),v(y)\}$
\item $v(x)=\infty\iff x=0$
\end{enumerate}
$R=\{x\in K\mid v(x)\geq 0\}$ definiert einen Teilring von $K$ und heißt \textit{diskreter Be\-wer\-tungs\-ring}\index{Bewertungsring!diskret} von $v$. $R$ ist ein lokaler Hauptidealring mit Maximalideal $\mathfrak{m}=\{x\in K\mid v(x) >0\}$. $R/\mathfrak{m}$ heißt \textit{Restklassenkörper}\index{Restklassenkörper} von $v$. Ein \textit{diskreter Bewertungsring}\index{Bewertungsring!diskret} $A$ ist ein nullteilerfreier Ring, der diskreter Bewertungsring für eine Bewertung seines Quotientenkörpers ist.

\paragraph{Beispiel 2.8}\begin{enumerate}
\item[2.] Sei $R$ ein diskreter Bewertungsring. Es ist $T=\operatorname{Spec}(R)$ ein affines Schema, bestehend aus zwei Punkten:
\begin{itemize}
\item Der Punkt $t_0=\mathfrak{m}\in\operatorname{Spec}(R)$ ist abgeschlossen, da $V(\mathfrak{m})=\{\mathfrak{m}\}$ und besitzt $R=R_{t_0}$ als lokalen Ring.
\item Der Punkt $t_1=(0)\in\operatorname{Spec}(R)$ ist offen und dicht in $\operatorname{Spec}(R)$, da $V(0)=\operatorname{Spec}(R)$. $t_1$ besitzt $K=\operatorname{Quot}(R)=R_{t_1}$ als lokalen Ring.
\end{itemize}
\[\xymatrix{
\operatorname{Spec}(K)\ar[r] & \operatorname{Spec}(R) &\ar[l] \operatorname{Spec}(R/\mathfrak{m})\\
(0)\ar@{|->}[r]\ar@{-}[u] & t_1\quad t_0 \ar@{-}@<-1.8ex>[u]\ar@{-}@<2.2ex>[u]& \ar@{|->}[l] (0)\ar@{-}[u]
} \]
\item[3.] Sei $k$ ein Körper. Die \textit{affine Gerade}\index{affine Gerade} $\mathbf{A}_k^1$ über $k$ ist $\operatorname{Spec}k[X]$. Sei $\xi$ das Nullideal in $\operatorname{Spec}k[X]$. Dann ist $\overline{\{\xi\}}=\mathbf{A}_k^1$. Ein solcher Punkt heißt \textit{generischer Punkt}\index{generischer Punkt}. Alle anderen Punkte sind abgeschlossen, da diese den maximalen Idealen in $k[X]$ entsprechen. Es besteht eine Bijektion zwischen den irreduziblen, nicht-konstanten, normierten Polynomen aus $k[X]$ und den abgeschlossenen Punkten von $\mathbf{A}_k^1$.

Ist $k$ algebraisch abgeschlossen, so besteht eine Bijektion zwischen den Elementen aus $k$ und den abgeschlossenen Punkten von $\mathbf{A}_k^1$.
\item[4.] Allgemeiner definieren wir den \textit{affinen $n$-dimensionalen Raum}\index{affiner Raum} über $k$ als: \[\mathbf{A}_k^n=\operatorname{Spec}k[X_1,\ldots,X_n]\]
\item[5.] Sei $k$ algebraisch abgeschlossen. Dann entsprechen die abgeschlossenen Punkte von $\mathbf{A}_k^n$ nach dem hilbertschen Nullstellensatz bijektiv den $n$-Tupeln von Elementen aus $k$. Ferner gibt es einen generischen Punkt $\xi$, der dem Nullideal in $k[X_1,\ldots,X_n]$ entspricht, d.h. $\overline{\{\xi\}}=\mathbf{A}_k^n$.
\end{enumerate}

\paragraph{Definition 2.10.}\label{2.10} \textit{(Offene Unterschemata)}\index{Schema!Unterschema!offen} Sei $(X,\mathcal{O}_X)$ ein Schema und $U\subset X$ offen. Dann ist $(U,\mathcal{O}_X|_U)$ ein Schema. Diese Aussage ist nichttrivial und wir werden sie \hyperref[2.10-beweis]{später} zeigen. Die offene Menge $U$ besitzt die \textit{induzierte Unterschemastruktur}.

\paragraph{Definition 2.11.}\label{2.11} \textit{(Verkleben von Schemata)}\index{Verklebung!Schema} Sei $\{X_i\}$ eine Familie von Schemata und $U_{ij}\subset X_i,\ i\neq j$ offene Teilmengen mit induzierter Struktur. Ferner haben wir für  $i\neq j$ Isomorphismen von Schemata:
\[\varphi_{ij}: (U_{ij},\mathcal{O}_{X_i}|_{U_{ij}}) \stackrel{\sim}{\to} (U_{ji},\mathcal{O}_{X_j}|_{U_{ji}}) \]
mit $\varphi_{ij}=\varphi_{ji}^{-1}$ und $\varphi_{ij}(U_{ij}\cap U_{ik})=U_{ji}\cap U_{jk}$ und $\varphi_{jk}=\varphi_{ik}\circ\varphi_{ij}$ auf $U_{ij}\cap U_{ik}$ für alle paar\-wei\-se verschiedene $i,j,k$. Wir erhalten ein Schema $X$ durch \textit{Verkleben} der $X_i$ längst $U_{ij}$ bzgl. $\varphi_{ij}$:
\[X=\dot{\bigcup_i}\ X_i\Big/{\sim},\quad x_i\sim\varphi_{ij}(x_i)\text{ für alle }x_i\in U_{ij},\ i\neq j \]
$X$ besitze die Quotiententopologie. Es existiert für jedes $j$ ein Morphismus $\psi_j:X_j\to X$ von Schemata, das ein Isomorphismus auf einem offenen Unterschema in $X$ induziert mit $X=\bigcup\psi_j(X_j)$ und $\psi_i(U_{ij})=\psi_i(X_i)\cap \psi_j(X_j)$ und $\psi_i=\psi_j\circ\varphi_{ij}$ auf $U_{ij}$ für alle $i\neq j$. Die Strukturgarbe auf $X$ ist folgendermaßen gegeben: Sei $V\subset_\text{o} X$. Setze:
\[\mathcal{O}_X(V)=\Big\{ (s_i)\in\prod_i \mathcal{O}_{X_i}(\psi_i^{-1}(V))\Bigm| \varphi_{ij}(s_i|_{\psi_i^{-1}(V)\cap U_{ij}}) = s_j|_{\psi_j^{-1}(V)\cap U_{ji}} \Big\} \]
Somit ist $(X,\mathcal{O}_X)$ ein lokal geringter Raum. Da alle $X_i$ Schemata sind, besitzt jeder Punkt von $X$ eine affine Umgebung. Also ist $X$ ein Schema.

\paragraph{Beispiel 2.12.}\label{2.12} Sei $k$ ein Körper, $X_1=X_2=\mathbf{A}_k^1$ und $U_1=U_2=\mathbf{A}_k^1\setminus\{P\}$ mit einem abgeschlossenen Punkt $P$. Ist $\varphi:U_1\to U_2$ die Identität, so ist die Verklebung $X$ von $X_1$ und $X_2$ längst $\varphi$ die affine Gerade, wobei der Punkt $P$ verdoppelt wurde. $X$ ist selbst nicht mehr affin.

\paragraph{Satz 2.13.}\label{2.13} Sei $A$ ein Ring und $(X,\mathcal{O}_X)$ ein Schema. Dann ist die Abbildung bijektiv:
\[\alpha:\operatorname{Hom}_\mathbf{Sch}(X,\operatorname{Spec}A)\to \operatorname{Hom}_\textbf{Ringe}(A,\Gamma(X,\mathcal{O}_X)) \]
wobei wir $(f:X\to\operatorname{Spec}A ,\ f^\sharp:\mathcal{O}_A\to f_\ast\mathcal{O}_X)$ durch das Nehmen der globalen Schnitte auf den Ringhomomorphismus $A=\Gamma(\operatorname{Spec}A,\mathcal{O}_A)\to\Gamma(X,\mathcal{O}_X)$ schicken.

\paragraph{Beweis.} Sei $X=\bigcup_\nu X_\nu$ eine affine Überdeckung. Ein Morphismus $(f,f^\sharp)$ ist eindeutig durch seine Einschränkungen $(f_\nu,f_\nu^\sharp)$ auf $X_\nu$ bestimmt. Diese sind nach \hyperref[2.6]{Satz 2.6} wiederum eindeutig bestimmt durch $\alpha(f_\nu)$. Ferner kommutiert das folgende Diagramm:
\[\xymatrix{
A\ar[d]_{\alpha(f)}\ar[rr]^-{\alpha(f_\nu)} && \Gamma(X_\nu,\mathcal{O}_{X_\nu})\\
\Gamma(X,\mathcal{O}_X)\ar[rru]_-{\operatorname{res}}
} \]
weshalb $\alpha(f)$ schon alle $\alpha(f_\nu)$ eindeutig bestimmt. Somit ist $\alpha$ injektiv. Sei nun ein Ringhomomorphismus $h:A\to\Gamma(X,\mathcal{O}_X)$ gegeben und $h_\nu:A\to\Gamma(X,\mathcal{O}_X)\to\Gamma(X_\nu,\mathcal{O}_X)$. Nach \hyperref[2.6]{Satz 2.6 (iii)} existiert ein $f_\nu:X_\nu\to\operatorname{Spec}(A)$ mit $\alpha(f_\nu)=h_\nu$. Für alle $\nu,\mu$ ist das folgende Diagramm kommutativ:
\[\xymatrix{
& \Gamma(X_\nu,\mathcal{O}_X)\ar[dr]^{\operatorname{res}} &\\
A\ar[ur]^{h_\nu} \ar[dr]_{h_\mu} & & \Gamma(X_\nu\cap X_\mu,\mathcal{O}_X)\\
& \Gamma(X_\mu,\mathcal{O}_X)\ar[ur]_{\operatorname{res}} &
} \]
Aus der Injektivität von $\alpha$ folgt $f_\nu=f_\mu$ auf $X_\nu\cap X_\mu$. Kleben wir die $f_\nu$ nun zusammen, so erhalten wir ein $f:X\to\operatorname{Spec}(A)$ mit $\alpha(f)=h$.\qed

\paragraph{Korollar 2.14.}\label{2.14} Es gibt eine pfeilumkehrende Kategorienäquivalenz zwischen der Kategorie der affinen Schemata und der Kategorie der kommutativen Ringe mit $1$.

\paragraph{Korollar 2.15.}\label{2.15} $\operatorname{Spec}(\mathbb{Z})$ ist Endobjekt in der Kategorie der Schemata, d.h. für jedes Schema $X$ existiert ein eindeutiger Morphismus $f:X\to\operatorname{Spec}(\mathbb{Z})$.

\paragraph{Satz 2.16.}\label{2.16} Sei $A$ ein Ring, $X=\operatorname{Spec}(A)$ und $f\in A$. Dann gilt:
\[(D(f),\mathcal{O}_X|_{D(f)})\cong \operatorname{Spec}(A_f) \]

\paragraph{Beweis.} Die natürliche Abbildung $\varphi:A\to A_f$ induziert einen Homöomorphismus:
\[\psi:\operatorname{Spec}(A_f) \to \{\mathfrak{p}\in\operatorname{Spec}(A)\mid f\not\in\mathfrak{p} \}=D(f),\ \mathfrak{p}\mapsto\varphi^{-1}(\mathfrak{p})\qedhere \]

\paragraph{Beweis zu \hyperref[2.10]{Definition 2.10}.}\label{2.10-beweis} Sei $(X,\mathcal{O}_X)$ ein Schema und $U\subset X$ offen. Dann ist $(U,\mathcal{O}_X|_U)$ ein lokal geringter Raum. Sei $P\in U$. Wir zeigen, dass eine Umgebung $V\subset_\text{o} U$ von $P$ existiert, so dass $(V,\mathcal{O}_X|_V)$ affin ist. Da $(X,\mathcal{O}_X)$ ein Schema ist, existiert ein $V'\subset_\text{o}X,\ P\in V'$ mit $(V',\mathcal{O}_X|_{V'})$ affin. Sei also $V'=\operatorname{Spec}(A)$ für einen Ring $A$. Da die $D(f),\ f\in A$ eine Basis der Topologie auf $X$ bilden, existiert ein $f\in A$, so dass $P\in D(f)\subset V'\cap U$. Wegen \hyperref[2.16]{Satz 2.16} ist $(D(f),\mathcal{O}_X|_{D(f)})\cong\operatorname{Spec}(A_f)$ affin.\qed

\paragraph{Satz 2.17.}\label{2.17} Sei $X$ ein Schema und $Z\subset X$ irreduzibel und abgeschlossen. Dann existiert genau ein Punkt $\xi\in Z$ derart, dass $\overline{\{\xi\}}=Z$. $\xi$ heißt \textit{generischer Punkt}\index{generischer Punkt} von $Z$.

\paragraph{Beweis.}\begin{itemize}
\item \textit{Existenz:} Sei $U\subset_\text{o}X$ affin mit $Z\cap U\neq\varnothing$. Da $Z\cap U$ abgeschlossen in $U$ ist, gibt es ein Radikalideal $\mathfrak{p}$ mit $Z\cap U=V(\mathfrak{p})$.

Nun ist $Z\cap U$ irreduzibel, da für offene Mengen $V_1,V_2\subset_\text{o} Z\cap U$ stets $V_1,V_2\subset_\text{o}Z$ gilt und aus der Irreduzibilität von $Z$ stets $V_1\cap V_2\neq\varnothing$ folgt.

Ferner ist $\mathfrak{p}$ ein Primideal, denn ist $\mathfrak{ab}\subset\mathfrak{p}$ für gewisse Ideale $\mathfrak{a}$ und $\mathfrak{b}$, so folgt $\operatorname{Rad}(\mathfrak{ab})\subset\operatorname{Rad}(\mathfrak{p})=\mathfrak{p}$. Dies ist äquivalent zu $V(\mathfrak{p})\subset V(\mathfrak{ab})=V(\mathfrak{a})\cup V(\mathfrak{b})$, daher:
\[Z\cap U =V(\mathfrak{p})=(V(\mathfrak{p})\cap V(\mathfrak{a}))\cup (V(\mathfrak{p})\cap V(\mathfrak{b})) \]
Da $Z\cap U$ irreduzibel ist, folgt o.B.d.A. $V(\mathfrak{p})=V(\mathfrak{p})\cap V(\mathfrak{a})\subset V(\mathfrak{a})$, d.h. $\mathfrak{a}\subset\operatorname{Rad}(\mathfrak{a})\subset\operatorname{Rad}(\mathfrak{p})=\mathfrak{p}$.

$\mathfrak{p}$ ist nun generischer Punkt von $Z\cap U$, da:
\[\overline{\{\mathfrak{p}\}}=\bigcap_{\mathfrak{p}\in V(\mathfrak{a})}V(\mathfrak{a})=\bigcap_{\mathfrak{a}\subset\mathfrak{p}} V(\mathfrak{a})\subset V(\mathfrak{p}) \]
Die andere Inklusion folgt aus $\mathfrak{a}\subset\mathfrak{p}$. Da $\mathfrak{p}$ prim ist, folgt $\operatorname{Rad}(\mathfrak{a})\subset\mathfrak{p}$ und somit $V(\mathfrak{p})\subset V(\mathfrak{a})$. Somit ist $Z\cap U=\overline{\{\mathfrak{p}\}}$.

Bezeichne mit $\overline{\{\mathfrak{p}\}}^X$ den topologischen Abschluss in $X$. Dann ist $Z\cap U\subset\overline{\{\mathfrak{p}\}}^X$. Da $Z$ irreduzibel ist, gilt $Z=\overline{Z\cap U}\subset\overline{\{\mathfrak{p}\}}^X$, also $Z=\overline{\{\mathfrak{p}\}}^X$.
\item \textit{Eindeutigkeit:} Sei $\overline{\{\xi\}}^X=Z=\overline{\{\xi'\}}^X$. Sei $U\subset_\text{o}X$ eine affine Umgebung von $\xi\in U$. Dann ist $\xi'\in U$, da aus $\xi'\in Z\setminus U$ der Widerspruch $\xi\in Z=\overline{\{\xi'\}}^X\subset X\setminus U$ folgt. Wähle nun Radikalideale $\mathfrak{p}'$ und $\mathfrak{p}$ mit:
\[V(\mathfrak{p}')= \overline{\{\xi'\}}^U=U\cap \overline{\{\xi'\}}^X=U\cap\overline{\{\xi\}}^X=\overline{\{\xi\}}^U=V(\mathfrak{p}) \]
Es folgt $\xi'=\mathfrak{p}'=\operatorname{Rad}(\mathfrak{p}')=\operatorname{Rad}(\mathfrak{p})=\mathfrak{p}=\xi$.\qed
\end{itemize}

\paragraph{Satz 2.18.}\label{2.18} Sei $X$ ein Schema und $K$ ein Körper. Dann gibt es eine Bijektion:
\[\operatorname{Hom}_\mathbf{Sch}(\operatorname{Spec}K,X)\stackrel{\sim}{\to} \{(x,i)\mid x\in X,\ i:\kappa(x)\hookrightarrow K\text{ Ringhomomorphismus} \} \]
wobei $\kappa(x)=\mathcal{O}_{X,x}/\mathfrak{m}_x$ der Restklassenkörper von $\mathcal{O}_{X,x}$ bezeichnet. Die Elemente heißen \textit{$K$-wertige Punkte} von $X$.\index{Punkt!$K$-wertig}

\paragraph{Beweis.} Sei ein Morphismus $f:\operatorname{Spec}(K)\to X$ gegeben. Setze $x=f((0))\in X$. $f^\sharp$ induziert einen lokalen Homomorphismus $\mathcal{O}_{X,x}\to\mathcal{O}_{K,(0)}=K$. $f^\sharp$ faktorisiert daher über $\mathcal{O}_{X,x}/\mathfrak{m}_x=\kappa(x)$ und induziert einen Homomorphismus $i:\kappa(x)\hookrightarrow K$.

Sei nun umgekehrt ein $x\in X$ und $i:\kappa(x)\hookrightarrow K$ gegeben. $i$ definiert nach \hyperref[2.6]{Satz 2.6} ein Schemamorphismus:
\[f:\operatorname{Spec}(K)\to\operatorname{Spec}\kappa(x)\to\operatorname{Spec}(\mathcal{O}_{X,x})\stackrel{\psi}{\to} X \]
wobei $\psi$ die folgende kanonische Abbildung ist:

Sei $U\subset_\text{o}X$ eine affine Umgebung von $x$ mit $U=\operatorname{Spec}(A)$. Nach \hyperref[2.3]{Satz 2.3 (i)} gilt $\mathcal{O}_{X,x}=\mathcal{O}_{U,x}=A_x$. Die kanonische Abbildung $A\to A_x$ induziert $\psi:\operatorname{Spec}(\mathcal{O}_{X,x})\to U\hookrightarrow X$. Diese Abbildung ist unabhängig von $U$:

Sei $U'\subset_\text{o}X$ eine weitere affine Umgebung von $x$. Dann existiert eine affine Umgebung $U''\subset_\text{o}U\cap U'$ mit $x\in U''$, also können wir o.B.d.A. $\operatorname{Spec}(A)= U\subset U'=\operatorname{Spec}(A')$ annehmen. Es existiert ein kanonischer Homomorphismus $A'\to A$ derart, dass das folgende Diagramm kommutiert:
\[\xymatrix{
A'\ar[rd]\ar[r] & A\ar[d]\\
& A'_x= \mathcal{O}_{X,x}= A_x &
} \]
Somit kommutiert:
\[\xymatrix{
X & U'\ar[l] & U\ar[l]\\
 && \operatorname{Spec}(\mathcal{O}_{X,x})\ar[lu]\ar[u]
}\qedhere\]

\paragraph{Satz 2.19.}\label{2.19} \begin{enumerate}[(i)]
\item Sei $A$ ein Ring und $f\in A$. Dann gilt:
\[D(f)=\varnothing\iff f\text{ nilpotent} \]
\item Sei $\varphi:A\to B$ ein Ringhomomorphismus und $f: Y=\operatorname{Spec}(B)\to\operatorname{Spec}(A)=X$ der durch $\varphi$ induzierte Morphismus affiner Schemata. Dann gilt:
\begin{enumerate}[(a)]
\item $\varphi$ ist genau dann injektiv, wenn $f^\sharp:\mathcal{O}_X\to f_\ast\mathcal{O}_Y$ injektiv ist. In diesem Fall ist $f$ \textit{dominant}\index{dominant}, d.h. $f(Y)\subset X$ ist dicht.
\item $\varphi$ ist genau dann surjektiv, wenn $f^\sharp$ surjektiv ist und $f$ ein Homöomorphismus auf eine abgeschlossene Teilmenge von $X$ ist.
\end{enumerate}
\end{enumerate}

\paragraph{Beweis.} Übung.\qed

% \paragraph{Beispiel.} Das Spektrum $T=\operatorname{Spec}( k[X,Y]/(X^2))$ entspricht der $Y$-Achse, also besitzt der Ring $\mathcal{O}_T(T)= k[X,Y]/(X^2)$ nilpotente Elemente.

\paragraph{Definition 2.20.}\label{2.20} Ein Schema $(X,\mathcal{O}_X)$ heißt \textit{reduziert}\index{Schema!reduziert}, falls $\mathcal{O}_X(U)$ für alle $U\subset_\text{o}X$ reduziert sind, d.h. keine nilpotente Elemente besitzt.

\paragraph{Regeln 2.21.}\label{2.21} Sei $(X,\mathcal{O}_X)$ ein Schema.
\begin{enumerate}[(i)]
\item $(X, \mathcal{O}_X)$ ist genau dann reduziert, wenn $\mathcal{O}_{X,P}$ für alle $P\in X$ keine nilpotente Elemente besitzt.
\item Sei $\mathcal{O}_X^\text{red}$ die assoziierte Garbe zur folgenden Prägarbe:
\[ U \mapsto \mathcal{O}_X(U)/\mathfrak{N}_U \]
wobei $\mathfrak{N}_U$ das Nilradikal von $\mathcal{O}_X(U)$ bezeichnet. Dann ist $(X,\mathcal{O}_X^\text{red})$ ein Schema, das zu $X$ assoziierte \textit{reduzierte Schema}\index{Schema!reduziert} $X_\text{red}$. Es gibt einen Morphismus $f:X_\text{red}\to X$ mit dem Homöomorphismus $\operatorname{id}$ auf den unterliegenden topologischen Räumen und $f^\sharp:\mathcal{O}_X\to f_\ast\mathcal{O}_X^\text{red}$ gegeben durch:
\[\mathcal{O}_X(U)\to\mathcal{O}_X^\text{red}(U),\ s\mapsto \Big(U\stackrel{s}{\to}\coprod_{P\in U}\mathcal{O}_{X,P} \to\coprod_{P\in U}\mathcal{O}_{X,P}^\text{red}\Big) \]
\item Sei $f:X\to Y$ ein Morphismus von Schemata mit $X$ reduziert. Setze $g=f$ auf den unterliegenden topologischen Räumen. Da $X$ reduziert ist, faktorisiert $f^\sharp:\mathcal{O}_Y\to f_\ast\mathcal{O}_X$ über $\mathcal{O}_Y^\text{red}$. $f^\sharp$ induziert $g^\sharp:\mathcal{O}_Y^\text{red}\to g_\ast\mathcal{O}_X=f_\ast\mathcal{O}_X$. Es gibt also einen eindeutig bestimmten Morphismus $g:X\to Y_\text{red}$ mit kommutativem Diagramm:
\[\xymatrix{
X\ar[rr]^f\ar@{-->}[drr]_g && Y\\
&& Y_\text{red}\ar[u]
} \]
\end{enumerate}

\paragraph{} Sei $S=\bigoplus_{d\geq 0} S_d$ ein graduierter Ring und $S_+=\bigoplus_{d>0}S_d$. Ein Primideal $\mathfrak{p}\subset S$ ist genau dann homogen, wenn aus $fg\in\mathfrak{p}$ für gewisse homogene Elemente $f,g\in S$ stets $f\in\mathfrak{p}$ oder $g\in\mathfrak{p}$ folgt. Für ein homogenes Ideal $\mathfrak{a}\subset S$ setze:
\[\operatorname{Proj}(S)=\{\mathfrak{p}\subset S\text{ homogenes Primideal}\mid S_+\not\subset\mathfrak{p}\},\quad V(\mathfrak{a})=\{\mathfrak{p}\in\operatorname{Proj}(S)\mid\mathfrak{a}\subset\mathfrak{p}\} \]

\paragraph{Lemma 2.22.}\label{2.22}\begin{enumerate}[(i)]
\item Sind $\mathfrak{a},\mathfrak{b}\subset S$ homogene Ideale, so gilt $V(\mathfrak{ab})=V(\mathfrak{a})\cup V(\mathfrak{b})$.
\item Ist $(\mathfrak{a}_i)_i$ eine Familie homogener Ideale in $S$, so folgt $V\big(\sum\mathfrak{a}_i\big)=\bigcap V(\mathfrak{a}_i)$.
\end{enumerate}
Damit wird auf $\operatorname{Proj}(S)$ eine Topologie definiert. Die abgeschlossenen Mengen sind genau die Mengen der Form $V(\mathfrak{a})$ für ein homogenes Ideal $\mathfrak{a}\subset S$.

\paragraph{Beweis.} Wie in \hyperref[2.1]{Lemma 2.1} unter Beachtung, dass homogene Ideale von homogene Elemente erzeugt werden.\qed

\paragraph{Definition.} Sei $T\subset S$ multiplikativ abgeschlossen, die aus homogenen Elementen besteht. Dann wird $T^{-1}S = \bigoplus_{i\geq 0} (T^{-1}S)_i$ zu einem graduierten Ring:
\[(T^{-1}S)_i=\Big\{\frac{s}{t}\in T^{-1}S\Bigm| s\in S\text{ homogen},\ t\in T,\ \deg(s)-\deg(t)=i \Big\} \]
Ist $\mathfrak{p}\subset S$ ein homogenes Primideal und $f\in S$ ein homogenes Element, so ist die \textit{homogene Lokalisierung}\index{homogene Lokalisierung} bzgl. $\mathfrak{p}$ bzw. $f$ definiert als:
\[S_{(\mathfrak{p})} = (S_\mathfrak{p})_0,\quad S_{(f)} = (S_f)_0 \]

\paragraph{Definition.} Wir definieren eine Ringgarbe $\mathcal{O}$ auf $\operatorname{Proj}(S)$ wie folgt: Sei $U\subset_\text{o}\operatorname{Proj}(S)$ und setze $\mathcal{O}(U)$ als die Menge aller Abbildungen $s:U\to\coprod_{\mathfrak{p}\in U}S_{(\mathfrak{p})}$, so dass:
\begin{enumerate}[(i)]
\item Für alle $\mathfrak{p}\in U$ gilt $s(\mathfrak{p})\in S_{(\mathfrak{p})}$.
\item Für alle $\mathfrak{p}\in U$ existiert eine offene Umgebung $V$ von $\mathfrak{p}$ mit $V\subset U$ und homogene Elemente $a,f\in S$ mit $\deg(a)=\deg(f)$ derart, dass für alle $\mathfrak{q}\in V$ stets $f\not\in\mathfrak{q}$ und $s(\mathfrak{q})=\frac{a}{f}$ in $S_{(\mathfrak{q})}$ gilt.
\end{enumerate}

\paragraph{Satz 2.23.}\label{2.23} Sei $S$ ein graduierter Ring. Dann gilt:
\begin{enumerate}[(i)]
\item $\mathcal{O}_\mathfrak{p}\cong S_{(\mathfrak{p})}$ für alle $\mathfrak{p}\in\operatorname{Proj}(S)$.
\item Für ein homogenes $f\in S_+$ setze $D_+(f)=\{\mathfrak{p}\in\operatorname{Proj}(S)\mid f\not\in\mathfrak{p}\}$. Dann ist $D_+(f)$ offen in $\operatorname{Proj}(S)$ und es gilt:
\[\operatorname{Proj}(S) =\bigcup_{f\in S_+\text{ homogen}}D_+(f) \]
Es gibt einen Isomorphismus lokal geringter Räume $(D_+(f),\mathcal{O}|_{D_+(f)})\cong\operatorname{Spec}S_{(f)}$.
\item $(\operatorname{Proj}S,\mathcal{O})$ ist ein Schema.
\end{enumerate}

\paragraph{Beweis.}\begin{enumerate}[(i)]
\item Die Abbildung $\mathcal{O}_\mathfrak{p}\to S_{(\mathfrak{p})},\ s_\mathfrak{p}\mapsto s(\mathfrak{p})$, wobei $s$ ein Repräsentant von $s_\mathfrak{p}$ ist, ist ein Isomorphismus. Beweis analog wie \hyperref[2.3]{Satz 2.3 (i)}.
\item Da $D_+(f)=\operatorname{Proj}(S)\setminus V(f)$, ist $D_+(f)$ offen. Sei $\mathfrak{p}\in\operatorname{Proj}(S)$, d.h. $\mathfrak{p}\subset S$ ist ein homogenes Primideal mit $S_+\not\subset\mathfrak{p}$. Sei $f\in S_+\setminus\mathfrak{p}$. Dann ist $\mathfrak{p}\not\in V(f)$, also $\mathfrak{p}\in D_+(f)$. Daher ist $\operatorname{Proj}(S)=\bigcup D_+(f)$.

Sei $f\in S_+$. Wir definieren ein Morphismus lokal geringter Räume $(\phi,\phi^\sharp): D_+(f)\to\operatorname{Spec}S_{(f)}$ wie folgt: Sei $S\rightarrow S_f$ der natürliche Homomorphismus. Sei $\mathfrak{a}\subset S$ ein homogenes Ideal und setze $\phi(\mathfrak{a})=\mathfrak{a}S_f\cap S_{(f)}$. Beachte, dass $S_{(f)}=(S_f)_0\subset S_f$ ein Teilring ist. Für $\mathfrak{p}\in D_+(f)$ ist $\phi(\mathfrak{p})\in\operatorname{Spec}S_{(f)}$, siehe \hyperref[2.16]{Satz 2.16}, und $\phi$ ist bijektiv. Sei $\mathfrak{a}\subset S$ ein homogenes Ideal. Dann ist $\mathfrak{p}\supset\mathfrak{a}$ genau dann, wenn $\phi(\mathfrak{p})\supset\phi(\mathfrak{a})$. Daher ist $\phi$ ein Homöomorphismus.

$\phi^\sharp:\mathcal{O}_{\operatorname{Spec} S_{(f)}}\to\phi_\ast \big(\mathcal{O}_{\operatorname{Proj}(S)}|_{D_+(f)}\big)$ wird wie folgt definiert:
\[\mathcal{O}_{\operatorname{Spec}S_{(f)}}(U)\to\mathcal{O}_{\operatorname{Proj}(S)}(\phi^{-1}(U)),\ s\mapsto \Big(\phi^{-1}(U)\stackrel{s\circ\phi}{\longrightarrow}\coprod (S_{(f)})_{\phi(\mathfrak{p})} \cong\coprod S_{(\mathfrak{p})}\Big) \]
Dieses ist ein Isomorphismus.
\item folgt aus (i) und (ii).\qed
\end{enumerate}

\paragraph{Beispiel 2.24.}\label{2.24} Sei $A$ ein Ring. Dann heißt
\[\mathbf{P}_A^n = \operatorname{Proj} A[X_0,\ldots,X_n] \]
der \textit{$n$-dimensionaler projektiver Raum}\index{projektiver Raum} über $A$. Ist speziell $A=k$ ein algebraisch abgeschlossener Körper, so ist $\mathbf{P}_k^n$ ein Schema. Dessen Teilraum aller abgeschlossenen Punkte ist homöomorph zur projektiven $n$-dimensionalen Varietät.

\paragraph{Definition 2.25.} \label{2.25} Sei $S$ ein beliebiges Schema. Ein \textit{Schema über $S$}\index{Schema!über $S$} ist ein Schema $X$ zusammen mit einem Morphismus $X\to S$, der sogenannte \textit{Strukturmorphismus}\index{Strukturmorphismus}. Ein Morphismus zweier Schemata $X$ und $Y$ über $S$ ist ein Morphismus $f:X\to X'$, so dass das folgende Diagramm kommutiert:
\[\xymatrix{
X\ar[rr]^f\ar[d] && X'\ar[lld]\\
S
} \]
So ein Morphismus nennt man auch \textit{$S$-Morphismus}\index{Morphismus!$S$-Schemata}. Bezeichne die Kategorie aller Schemata über $S$ mit $S$-Morphismen mit $\mathbf{Sch}(S)$. Für einen Ring $A$ setzen wir auch $\mathbf{Sch}(A)=\mathbf{Sch}(\operatorname{Spec}A)$.

\paragraph{Satz 2.26.}\label{2.26} Sei $k$ ein algebraisch abgeschlossener Körper. Dann gibt es einen natürlichen Funktor
\[t:\mathbf{Var}(k)\to\mathbf{Sch}(k) \]
der \textit{volltreu}\index{volltreuer Funktor} ist, d.h. für zwei Varietäten $V,W$ ist die durch $t$ auf den Morphismen induzierte Abbildung $\operatorname{Hom}_{\mathbf{Var}(k)}(V,W)\to \operatorname{Hom}_{\mathbf{Sch}(k)}(tV,tW)$ bijektiv. Für eine Varietät $V$ setze $\mathfrak{M}(V)$ als die Menge aller abgeschlossenen Punkte des Schemas $tV$ mit der Teilraumtopologie. Es gibt einen Homöomorphismus topologischer Räume $V\cong\mathfrak{M}(V)$. Die Garbe der regulären Funktionen ist via diesen Homöomorphismus isomorph zu $\mathcal{O}_{tV}|_{\mathfrak{M}(V)}$.

\paragraph{Beweis.} Siehe z.B. Hartshorne Kapitel II, Proposition 2.6 oder Mumford, Theorem 2 auf Seite 168.

\section{Erste Eigenschaften von Schemata}

\paragraph{Definition 3.1.}\label{3.1} \begin{enumerate}[(i)]
\item Ein Schema heißt \textit{zusammenhängend}\index{Schema!zusammenhängend}, falls es als topologischer Raum zu\-sam\-men\-hän\-gend ist.
\item Ein Schema heißt \textit{irreduzibel}\index{Schema!irreduzibel}, falls es als topologischer Raum irreduzibel ist.
\item Ein Schema heißt \textit{reduziert}\index{Schema!reduziert} falls für alle $U\subset_\text{o}X$ der Ring $\mathcal{O}_X(U)$ reduziert ist.
\item Ein Schema heißt \textit{integer}\index{Schema!integer}, falls für alle $U\subset_\text{o}X$ der Ring $\mathcal{O}_X(U)$ nullteilerfrei ist.
\end{enumerate}

\paragraph{Beispiel 3.2.}\label{3.2} Sei $X=\operatorname{Spec}(A)$ ein affines Schema. Dann gilt:
\begin{enumerate}[(i)]
\item $X$ ist irreduzibel $\iff \mathfrak{N}(A)$ ist ein Primideal
\item $X$ ist reduziert $\iff A$ ist reduziert $\iff\mathfrak{N}(A)=0$
\item $X$ ist integer $\iff A$ ist nullteilerfrei
\end{enumerate}

\paragraph{Beweis.} (i) und (ii) sind klar. Ist $X$ integer, so ist $A=\mathcal{O}_X(X)$ nullteilerfrei. Sei nun umgekehrt $A$ nullteilerfrei, d.h. $\mathfrak{N}(A)=0$ ist ein Primideal. Nach (i) und (ii) ist $X$ irreduzibel und reduziert. Daher folgt die Aussage aus dem nächsten Satz.\qed

\paragraph{Satz 3.3.}\label{3.3} Ein Schema $X$ ist genau dann integer, wenn $X$ reduziert und irreduzibel ist.

\paragraph{Beweis.} Sei $X$ integer. Dann ist $X$ offensichtlich reduziert. Wäre $X$ nicht irreduzibel, so gäbe es $\varnothing\neq U_1,U_2\subset_\text{o}X$ mit $U_1\cap U_2=\varnothing$. Dann ist:
\[\mathcal{O}_X(U_1\cup U_2)=\mathcal{O}_X(U_1)\times\mathcal{O}_X(U_2) \]
Somit ist $\mathcal{O}_X(U_1\cup U_2)$ nicht nullteilerfrei. Sei nun umgekehrt $X$ irreduzibel und reduziert.

Sei $V\subset_\text{o}X$ affin mit $V=\operatorname{Spec}(A)$. Sei $a,b\in A$ mit $ab=0$. Es folgt:
\[\operatorname{Spec}(A)=V(ab)=V(a)\cup V(b) \]
Da $V$ irreduzibel ist, folgt o.B.d.A. $\operatorname{Spec}(A)=V(a)$. Da $A=\mathcal{O}_X(V)$ reduziert ist, folgt $\operatorname{Rad}(a)=(0)$, also $a=0$. Daher ist $\mathcal{O}_X(V)$ für jedes affine $V\subset_\text{o}X$ nullteilerfrei.

Sei nun $U\subset_\text{o}X$ beliebig und $f,g\in\mathcal{O}_X(U)$ mit $fg=0$. Für $V\subset_\text{o}U$ affin, folgt aus $f|_V\cdot g|_V=0$ o.B.d.A. $f|_V=0$. Nun ist $U$ der Abschluss von $V$ in $U$. Sei $x\in U$ und $U(x)\subset_\text{o}U$ eine affine Umgebung von $x$ mit $U(x)=\operatorname{Spec}(B)$. Sei $f(x)=\frac{a}{h}\in B_x$ für alle $x\in U(x)$. Es ist $U(x)\cap V\neq\varnothing$, da $U$ irreduzibel ist. Wähle ein $y\in U(x)\cap V$; es folgt $0=f(y)=\frac{a}{h}\in B_y$, also gibt es ein $k\in B\setminus y$ mit $ka=0$. Da $B$ nullteilerfrei ist, folgt $a=0$ und $f=0$ auf $U(x)$. Somit folgt $f=0$.\qed

\paragraph{Definition.} \begin{enumerate}[(i)]
\item Ein Schema $X$ heißt \textit{quasikompakt}\index{quasikompakt}, wenn sein unterliegender topologischer Raum quasikompakt ist.
\item Ein topologischer Raum $X$ heißt \textit{noethersch}\index{noethersch}, wenn jede absteigende Kette von abgeschlossenen Teilmengen in $X$ stationär wird.
\end{enumerate}

\paragraph{Bemerkung.} Sei $X$ ein noetherscher Raum. Nach Zorns Lemma besitzt jede nichtleere Menge $\Sigma$ von abgeschlossenen Mengen in $X$ ein minimales Element, da jede Kette in $\Sigma$ ein minimales Element besitzt.

\paragraph{Satz 3.4.}\label{3.4} \begin{enumerate}[(i)]
\item Sei $X$ ein topologischer Raum. Dann ist $X$ genau dann noethersch, wenn alle offenen Teilmengen $U\subset X$ quasikompakt sind.
\item Sei $X$ ein affines Schema. Dann ist $X$ quasikompakt, aber nicht notwendig noethersch.
\item Sei $A$ ein noetherscher Ring. Dann ist der $\operatorname{Spec}(A)$ unterliegender Raum noethersch.
\end{enumerate}

\paragraph{Beweis.}\begin{enumerate}[(i)]
\item Sei $U\subset_\text{o}X$ und $U=\bigcup_{i\in I}U_i$ eine offene Überdeckung. Für eine endliche Teilmenge $J\subset I$ setze $V_J=\bigcup_{i\in J}U_i$. Dann ist $V_J\subset X$ offen und es gilt:
\[U=\bigcup_{J\subset I\text{ endlich}}V_J \]
Wählt man aus $\Sigma=\{X\setminus V_J\mid J\subset I\text{ endlich}\}$ ein minimales Element $X\setminus V_{J'}$. Dann gilt $V_{J'}\supset V_J$ für alle $J$. Also ist $U=V_{J'}=\bigcup_{i\in J'}U_i$ eine endliche Teilüberdeckung.

Sei umgekehrt $Y_1\supset Y_2\supset\ldots$ eine Kette abgeschlossener Mengen in $X$, so ist die Menge $U=\bigcup_{j\geq 1}X\setminus Y_j$ offen in $X$. Wir erhalten eine endliche Teilüberdeckung $U=\bigcup_{r\geq j\geq 1}X\setminus Y_j=X\setminus Y_r$, also folgt $Y_s=Y_r$ für alle $s\geq r$.
\item Sei $X=\operatorname{Spec}(A)=\bigcup_{i\in I}U_i$ eine offene Überdeckung. Wir können o.B.d.A. $U_i=D(f_i)$ für gewisse $f_i\in A$ annehmen. Sei $\mathfrak{a}=(f_i\mid i\in I)\subset A$. Dann gilt:
\[X=\bigcup_{i\in I}X\setminus V(f_i) = X\setminus\bigcap_{i\in I}V(f_i)=X\setminus V(\mathfrak{a}) \]
Es folgt $V(\mathfrak{a})=\varnothing$, also $1\in\mathfrak{a}$. Somit gibt es endlich viele $g_j\in A$ und $i_j\in I$ mit $1=\sum_j g_jf_{i_j}$. Wir erhalten die endliche Teilüberdeckung $X=\bigcup_j D(f_{i_j})$.
\item Sei $V(\mathfrak{a}_1)\supset V(\mathfrak{a}_2)\supset\ldots$ eine Kette abgeschlossener Mengen in $\operatorname{Spec}(A)$ mit Ra\-di\-kal\-idealen $\mathfrak{a}_i\subset A$. Sie wird stationär, da die Kette $\mathfrak{a}_1\subset \mathfrak{a}_2\subset\ldots$ stationär wird.\qed
\end{enumerate}

\paragraph{Definition 3.5.}\label{3.5} Sei $X$ ein Schema.
\begin{enumerate}[(i)]
\item $X$ heißt \textit{lokal noethersch}\index{noethersch!lokal}, falls $X$ von offenen, affinen Teilmengen $\operatorname{Spec}(A_i)$ mit noetherschen Ringen $A_i$ überdeckt werden kann.
\item $X$ heißt \textit{noethersch}\index{noethersch}, falls $X$ lokal noethersch und quasikompakt ist. Dies ist äquivalent dazu, dass $X$ von endlich vielen offenen, affinen Teilmengen $\operatorname{Spec}(A_i)$ mit noetherschen Ringen $A_i$ überdeckt werden kann.
\end{enumerate}

\paragraph{Bemerkung.} Ist ein Schema $X$ noethersch, so ist nach \hyperref[3.4]{Satz 3.4 (iii)} der unterliegender Raum von $X$ noethersch. Die Umkehrung gilt im Allgemeinen nicht.

\paragraph{Satz 3.6.}\label{3.6} Sei $X$ ein Schema. Dann ist $X$ genau dann lokal noethersch, wenn für alle offenen, affinen Teilmengen $U=\operatorname{Spec}(A)$ stets $A$ ein noetherscher Ring ist.

\paragraph{Beweis.} Die Rückrichtung ist trivial. Sei also $X$ lokal noethersch und $U=\operatorname{Spec}(A)$ offen in $X$. Wir haben eine offene affine Überdeckung $X=\bigcup_i\operatorname{Spec}(B_i)$ mit noetherschen Ringen $B_i$. Da die offenen Mengen $D(f)$ eine offene Basis der Topologie bilden, haben wir eine Darstellung:
\[U=\bigcup_{i,j} D(f_{ij}),\quad f_{ij}\in B_i,\ D(f_{ij})\subset U\cap \operatorname{Spec}(B_i) \]
mit $D(f_{ij})=\operatorname{Spec}(B_i)_{f_{ij}}$. Da $B_i$ noethersch sind, sind die Lokalisierungen $(B_i)_{f_{ij}}$ ebenfalls noethersche Ringe. Da $U$ affin und somit quasikompakt ist, kann $U$ von endlich vielen Spektren noetherscher Ringe überdeckt werden.

Sei $V=\operatorname{Spec}(B)$ offen in $U$ mit noetherschen Ring $B$. Sei $f\in A$ und betrachte $D(f)\subset V$. Die natürliche Inklusion $V\hookrightarrow U$ induziert einen Ringhomomorphismus $A\to B$. Sei $\overline{f}$ das Bild von $f$ in $B$. Es gilt:
\[\operatorname{Spec}(A_f)=D(f)=D(\overline{f})=\operatorname{Spec}(B_{\overline{f}})\]
Es folgt $A_f\cong B_{\overline{f}}$ und $A_f$ ist noethersch. Wir haben nun gezeigt, dass $U$ von endlich vielen offenen Mengen der Form $D(f)=\operatorname{Spec}(A_f)$ überdeckt werden kann mit noetherschen Ringen $A_f$. Somit folgt die Aussage aus dem nächsten Lemma.\qed

\paragraph{Lemma 1.}\label{3.6-lemma1} Sei $A$ ein Ring und $f_1,\ldots,f_r\in A$ mit $1=(f_1,\ldots,f_r)$. Sind alle $A_{f_i}$ noethersch, so ist auch $A$ noethersch.

\paragraph{Lemma 2.}\label{3.6-lemma2} Sei $A$ ein Ring und $f_1,\ldots,f_r\in A$ mit $1=(f_1,\ldots,f_r)$. Sei $\mathfrak{a}\subset A$ ein Ideal und $\varphi_i:A\to A_{f_i}$ die Lokalisierungsabbildung. Dann gilt:
\[\mathfrak{a} = \bigcap_{i=1}^r \varphi_i^{-1}(\varphi_i(\mathfrak{a}) A_{f_i}) \]

\paragraph{Beweis.} Für die nichttriviale Inklusion sei $b\in A$ mit $\varphi_i(b)\in \varphi(\mathfrak{a})A_{f_i}$ für alle $i$. Schreibe:
\[\varphi_i(b)=\frac{a_i}{f_i^{n_i}}\in A_{f_i},\quad a_i\in\mathfrak{a},\ n_i>0 \]
Sei o.B.d.A. $n=n_1=\ldots=n_r$. Somit gibt es für alle $i$ ein $m_i\geq 0$ mit:
\[f_i^{m_i}(f_i^nb-a_i)=0 \]
Sei o.B.d.A. $m=m_1=\ldots=m_r$. Es folgt $f_i^{m+n}b\in\mathfrak{a}$ für alle $i$. Aus $1=(f_1,\ldots,f_r)$ folgt $1=(f_1^N,\ldots,f_r^N)$ für alle $N\geq 0$, insbesondere für $N=m+n$. Sei also $1=\sum_{i=1}^rc_if_i^N$ für gewisse $c_i\in A$. Dann gilt:
\[b=\sum_{i=1}^rc_if_i^Nb\in\mathfrak{a}\qedhere \]

\paragraph{Beweis von \hyperref[3.6-lemma1]{Lemma 1}.} Sei $\mathfrak{a}_1\subset\mathfrak{a}_2\subset\ldots$ eine Kette von Idealen in $A$. Diese induziert für alle $i$ eine Kette von Idealen in $A_{f_i}$:
\[\varphi_i(\mathfrak{a}_1)A_{f_i}\subset\varphi_i(\mathfrak{a}_2) A_{f_i}\subset\ldots \]
Da $A_{f_i}$ noethersch ist, wird diese Kette stationär für alle $i$. Es existiert also ein $s$ mit $\varphi_i(\mathfrak{a}_s)A_{f_i}=\varphi(\mathfrak{a}_{s+1})A_{f_i}=\ldots$ für alle $i$. Mit \hyperref[3.6-lemma2]{Lemma 2} wird auch die ursprüngliche Kette von Idealen in $A$ stationär.\qed

\paragraph{Definition.} Sei $f:X\to Y$ ein Morphismus von Schemata.
\begin{enumerate}[(i)]
\item $f$ heißt \textit{lokal von endlichem Typ}\index{von endlichem Typ!lokal}, falls $Y$ eine offene affine Überdeckung $\bigcup_i\operatorname{Spec}(B_i)$ besitzt, so dass $f^{-1}(\operatorname{Spec}B_i)$ für alle $i$ eine offene affine Überdeckung $\bigcup_j\operatorname{Spec}(A_{ij})$ besitzt, wobei alle $A_{ij}$ endlich erzeugte $B_i$-Algebren sind.
\item $f$ heißt \textit{von endlichem Typ}\index{von endlichem Typ}, falls $Y$ eine offene affine Überdeckung $\bigcup_i\operatorname{Spec}(B_i)$ besitzt, so dass $f^{-1}(\operatorname{Spec}B_i)$ für alle $i$ eine endliche, offene affine Überdeckung $\bigcup_j\operatorname{Spec}(A_{ij})$ besitzt, wobei alle $A_{ij}$ endlich erzeugte $B_i$-Algebren sind.
\item $f$ heißt \textit{endlich}\index{endlich}, falls eine offene affine Überdeckung $Y=\bigcup_i\operatorname{Spec}(B_i)$ existiert, so dass $f^{-1}(\operatorname{Spec}B_i)=\operatorname{Spec}(A_i)$ für alle $i$ ist, wobei $A_i$ eine $B_i$-Algebra ist, die als $B_i$-Modul endlich erzeugt ist.
\end{enumerate}

\paragraph{Bemerkung 3.7.}\label{3.7} Sei $f:X\to Y$ ein Morphismus von Schemata.
\begin{enumerate}[(i)]
\item $f$ ist genau dann lokal von endlichem Typ, wenn für alle offene, affine $V=\operatorname{Spec}(B)$ in $Y$ es eine offene affine Überdeckung $f^{-1}(V)=\bigcup_j\operatorname{Spec}(A_j)$ gibt, wobei $A_j$ endlich erzeugte $B$-Algebren sind.
\item $f$ ist genau dann von endlichem Typ, wenn für alle offene, affine $V=\operatorname{Spec}(B)$ in $Y$ es eine endliche, offene affine Überdeckung $f^{-1}(V)=\bigcup_j\operatorname{Spec}(A_j)$ gibt, wobei $A_j$ endlich erzeugte $B$-Algebren sind.
\item $f$ ist genau dann endlich, wenn für alle offene, affine $V=\operatorname{Spec}(B)$ in $Y$ die Menge $f^{-1}(V)=\operatorname{Spec}(A)$ affin ist, wobei $A$ ein endlich erzeugter $B$-Modul ist.
\end{enumerate}

\paragraph{Beweis.} Dies werden wir später zeigen.

\paragraph{Beispiel 3.8.}\label{3.8} Sei $V$ eine Varietät über einem algebraisch abgeschlossenen Körper $k$. Dann ist das Schema $t(V)$ in \hyperref[2.26]{Satz 2.26} ein integres, noethersches Schema von endlichem Typ über $k$.

\paragraph{Beispiel 3.9.}\label{3.9} Sei $P$ ein Punkt einer Varietät und $\mathcal{O}_P$ der zugehörige Halm. Dann ist $\operatorname{Spec}(\mathcal{O}_P)$ ein integres, noethersches Schema, aber nicht von endlichem Typ über $k$.

\paragraph{Definition 3.10.}\label{3.10} Ein \textit{offenes Unterschema}\index{Schema!Unterschema!offen} eines Schemas $X$ ist ein Schema $U$, dessen unterliegender topologischer Raum eine offene Teilmenge von $X$ ist und dessen Strukturgarbe $\mathcal{O}_U$ isomorph zu $\mathcal{O}_X|_U$ ist.

Ein Schemamorphismus $f:X\to Y$ heißt \textit{offene Immersion}\index{Immersion!offen}, falls $f$ ein Isomorphismus auf ein offenes Unterschema in $Y$ induziert.

\paragraph{Bemerkung.} Jede offene Teilmenge eines Schemas $X$ trägt eine eindeutig bestimmte Struktur als offenes Unterschema, siehe auch \hyperref[2.10]{Definition 2.10}.

\paragraph{Definition 3.11.}\label{3.11} Ein \textit{abgeschlossenes Unterschema}\index{Schema!Unterschema!abgeschlossen} eines Schemas $X$ ist ein Schema $Y$, zusammen mit einem Morphismus $(i,i^\sharp):Y\to X$, so dass:
\begin{enumerate}[(i)]
\item Der $Y$ unterliegender Raum ist eine abgeschlossene Teilmenge von $X$.
\item $i:Y\hookrightarrow X$ ist die natürliche Inklusion.
\item $i^\sharp:\mathcal{O}_X\to i_\ast\mathcal{O}_Y$ ist surjektiv.
\end{enumerate}

Ein Schemamorphismus $f:X\to Y$ heißt \textit{abgeschlossene Immersion}\index{Immersion!abgeschlossen}, falls $f$ ein Isomorphismus auf ein abgeschlossenes Unterschema in $Y$ induziert.

\paragraph{Beispiel 3.12.}\label{3.12} Sei $A$ ein Ring, $\mathfrak{a}\subset A$ ein Ideal und $X=\operatorname{Spec}(A),\ Y=\operatorname{Spec}(A/\mathfrak{a})$. Der Ringhomomorphismus $\varphi:A\to A/\mathfrak{a}$ induziert $f:Y\to X,\ \mathfrak{p}\mapsto\varphi^{-1}(\mathfrak{p})$. Dies induziert ein Morphismus von Schemata. $f$ ist ein Homöomorphismus von $Y$ auf $V(\mathfrak{a})$ und die Abbildung $f^\sharp:\mathcal{O}_X\to f_\ast\mathcal{O}_Y$ induziert: 
\[f_\mathfrak{p}^\sharp:\mathcal{O}_{X, f(\mathfrak{p})}=A_{\varphi^{-1}(\mathfrak{p})}\to (A/\mathfrak{a})_\mathfrak{p}=\mathcal{O}_{Y,\mathfrak{p}}\]
$f_\mathfrak{p}^\sharp$ ist surjektiv für alle $\mathfrak{p}\in Y$, also ist nach \hyperref[1.10]{1.10 (ii)} auch $f^\sharp$ surjektiv. Somit erhält man für jedes $\mathfrak{a}\subset A$ auf $V(\mathfrak{a})\subset X$ eine Struktur als abgeschlossenes Unterschema in $X$.

\paragraph{Bemerkung.} Ist $Y\subset X=\operatorname{Spec}(A)$ eine abgeschlossene Teilmenge, so existieren auf $Y$ viele abgeschlossene Unterschamestrukturen. Wir werden später sehen, das sie genau den Idealen $\mathfrak{a}\subset A$ mit $Y=V(\mathfrak{a})$ entsprechen.

\paragraph{Satz 3.14.}\label{3.14} Sei $X$ ein Schema und $Y$ eine abgeschlossene Teilmenge von $X$. Dann besitzt $Y$ eine eindeutig bestimmte induzierte Struktur als reduziertes, abgeschlossenes Unterschema.

\iffalse
\paragraph{Beweis.}
\begin{enumerate}
\item Fall: Sei zunächst $X=\operatorname{Spec}(A)$ affin. Betrachte das folgende Ideal in $A$:
\[\mathfrak{a}=\bigcap_{\mathfrak{p}\in Y}\mathfrak{p} \]
Dann ist $V(\mathfrak{a})=Y$ und $\mathfrak{a}$ ist maximal unter allen Idealen $\mathfrak{a}'$ mit $V(\mathfrak{a}')=Y$. Die Unterschemastruktur $Y=V(\mathfrak{a})=\operatorname{Spec}(A/\mathfrak{a})$ auf $Y$ ist reduziert, da $A/\mathfrak{a}$ reduziert ist, wegen $\mathfrak{a}=\operatorname{Rad}(\mathfrak{a})$.

Für die Eindeutigkeit sei $Y'=V(\mathfrak{a}')=\operatorname{Spec}(A/ \mathfrak{a}')$ eine weitere abgeschlossene Unterschemastruktur. Dann ist $\mathfrak{a}'\subset\mathfrak{a}$. Die kanonische Abbildung $A/\mathfrak{a}'\to A/\mathfrak{a}$ induziert ein Morphismus $Y\to Y'$. Da $Y$ reduziert ist, gibt es einen eindeutigen Morphismus $Y\to Y'_\text{red}$, so dass folgendes Diagramm kommutiert:
\[\xymatrix{
Y\ar[rr]\ar@{-->}[rrd] && Y'\\
&& Y'_\text{red}\ar[u]
} \]
\end{enumerate}
\fi

\paragraph{Lemma 3.15.}\label{3.15} Sei $X$ ein topologischer Raum und $X=\bigcup_i U_i$ eine offene Überdeckung. Ferner sei für jedes $i$ eine Garbe $F_i$ auf $U_i$ gegeben und für alle $i,j$ seien Isomorphismen gegeben:
\[\varphi_{ij}:F_i|_{U_i\cap U_j}\stackrel{\sim}{\to} F_j|_{U_i\cap U_j} \]
so dass $\varphi_{ii}=\operatorname{id}$ und $\varphi_{ik}=\varphi_{jk}\varphi_{ij}$ auf $U_i\cap U_j\cap U_k$ gilt. Dann existiert eine eindeutig bestimmte Garbe $F$ auf $X$ und Isomorphismen $\psi_i:F|_{U_i}\stackrel{\sim}{\to}F_i$ mit $\psi_j=\varphi_{ij}\psi_i$ auf $U_i\cap U_j$. Wir sagen auch, dass $F$ durch \textit{Verkleben} der $F_i$ längst $\varphi_{ij}$ entsteht.\index{Verklebung!Garbe}

\paragraph{Beweis.} Folgt direkt aus der Definition einer Garbe.\qed

\paragraph{Definition 3.16.}\label{3.16} \begin{enumerate}[(i)]
\item Die \textit{Dimension} $\dim(X)$ eines Schemas $X$ ist die Dimension von $X$ als topologischer Raum, d.h:\index{Dimension}
\begin{align*}
\dim(X)=\sup\{n\mid&\text{ es gibt eine Kette }Z_0\subsetneq Z_1\subsetneq\ldots\subsetneq Z_n\\
&\text{ von irreduziblen, abgeschlossenen Teilmengen in }X \} 
\end{align*}
\item Sei $Z$ eine irreduzible, abgeschlossene Teilmenge eines Schemas $X$. Dann ist die \textit{Kodimension}\index{Kodimension} $\operatorname{codim}(Z,X)$ von $Z$ in $X$ definiert als:
\begin{align*}
\operatorname{codim}(Z,X)=\sup\{n\mid & \text{ es gibt eine Kette }Z=Z_0\subsetneq Z_1\subsetneq\ldots\subsetneq Z_n\\
&\text{ von irreduziblen, abgeschlossenen Teilmengen in }X\}
\end{align*}
Ist $Y$ eine abgeschlossene Teilmenge von $X$, so setzen wir:
\[\operatorname{codim}(Y,X)=\inf \{\operatorname{codim}(Z,X)\mid Z\subset Y\text{ irreduzibel, abgeschlossen} \} \]
\end{enumerate}

\paragraph{Definition 3.17.}\label{3.17} Sei $S$ ein Schema und $X,Y$ $S$-Schemata. Das \textit{Faserprodukt}\index{Faserprodukt} $X\times_SY$ von $X$ und $Y$ über $S$ ist ein Schema, zusammen mit Projektionsmorphismen $p_1:X\times_SY\to X$ und $p_2:X\times_SY\to Y$ derart, dass:
\begin{enumerate}[(i)]
\item Das folgende Diagramm kommutiert:
\[ \xymatrix{
X\times_S Y\ar[r]^-{p_2}\ar[d]_-{p_1} & Y\ar[d]\\
X\ar[r] & S
}\]
\item Ist $Z$ ein $S$-Schema und $f:Z\to X,\ g:Z\to Y$ Morphismen derart, dass das folgende Diagramm kommutiert:
\[\xymatrix{
Z\ar@/^1pc/[rrd]^g\ar@/_1pc/[ddr]_f\ar@{-->}[rd]^\theta &&\\
&X\times_S Y\ar[r]_-{p_2}\ar[d]^-{p_1} & Y\ar[d]\\
&X\ar[r] & S
} \]
so existiert einen eindeutig bestimmten Morphismus $\theta:Z\to X\times_SY$ mit $f=p_1\theta$ und $g=p_2\theta$.
\end{enumerate}
Wird für Schemata $X$ und $Y$ kein Bezug zu einer Basis angegeben, so ist immer das Endobjekt $S=\operatorname{Spec}(\mathbb{Z})$ gemeint, d.h. $X\times Y=X\times_{\operatorname{Spec}(\mathbb{Z})}Y$.

\paragraph{Theorem 3.18.}\label{3.18} Seien $X$ und $Y$ $S$-Schemata. Dann existiert das Faserprodukt $X\times_S Y$ und ist auf Isomorphie eindeutig.

\paragraph{Lemma 3.19.}\label{3.19} \textit{(Verkleben von Morphismen, vgl. \hyperref[2.13]{Satz 2.13})}\index{Verklebung!Morphismus} Seien $X,Y$ Schemata und $\{U_i\}$ eine offene Überdeckung von $X$. Ferner seien $f_i:U_i\to Y$ Morphismen gegeben, wobei $U_i$ mit der offenen Unterschemastruktur versehen ist. Es gelte $f_i|_{U_i\cap U_j}=f_j|_{U_i\cap U_j}$ für alle $i,j$. Dann gibt es einen Morphismus $f:X\to Y$ mit $f|_{U_i}=f_i$ für alle $i$.

\paragraph{Beweis von \hyperref[3.18]{Theorem 3.18}.} Die Eindeutigkeit ist klar.
\begin{enumerate}
\item Schritt: Seien $X=\operatorname{Spec}(A),\ Y=\operatorname{Spec}(B),\ S=\operatorname{Spec}(R)$ affin. Somit sind $A$ und $B$ $R$-Algebren. Wir zeigen $X\times_S Y=\operatorname{Spec}(A\otimes_R B)$. Die Projektionsabbildung $p_1:\operatorname{Spec}(A\otimes_RB)\to \operatorname{Spec}(A)$ ist durch die natürliche Abbildung $\tilde{p}_1:A\to A\otimes_RB$ gegeben, analog für $p_2$. Offensichtlich kommutiert:
\[\xymatrix{
 A\ar[r]^-{\tilde{p}_1} & A\otimes_RB\\
 R\ar[r]\ar[u] & B\ar[u]_{\tilde{p}_2}
} \]
Sei also $Z$ ein $S$-Schema und Morphismen $f:Z\to X,\ g:Z\to Y$ gegeben, die über $S$ gleich sind. Diese entsprechen Ringhomomorphismen $\tilde{f}:A\to\Gamma(Z,\mathcal{O}_Z)$ und $\tilde{g}:B\to\Gamma(Z,\mathcal{O}_Z)$ nach \hyperref[2.13]{Satz 2.13}. Es kommutiert:
\[\xymatrix{
&&\Gamma(Z,\mathcal{O}_Z)\\
A\ar[r]_-{\tilde{p}_2}\ar@/^1pc/[rru]^{\tilde{f}}& A\otimes_RB\ar@{-->}[ru]_{\tilde{\theta}} &\\
R\ar[r]\ar[u] & B\ar[u]^{\tilde{p}_2}\ar@/_1pc/[ruu]_{\tilde{g}}
} \]
Wegen der Universaleigenschaft des Tensorprodukts gibt es genau einen Ringhomomorphismus $\tilde{\theta}:A\otimes_RB\to\Gamma(Z,\mathcal{O}_Z)$ mit $\tilde{f}\tilde{\theta}=\tilde{p}_1$ und $\tilde{g}\tilde{\theta}=\tilde{p}_2$. \hyperref[2.13]{Satz 2.13} liefert ein eindeutiges $\theta:Z\to\operatorname{Spec}(A\otimes_RB)$ mit $f=p_1\theta$ und $g=p_2\theta$.
\item Schritt: Seien $X,Y$ beliebige $S$-Schemata und $U\subset_\text{o}X$. Wir nehmen an, dass das Faserprodukt $X\times_SY$ mit Projektionen $p_1,p_2$ existiert. Wir zeigen, dass für die offene Teilmenge $p_1^{-1}(U)\subset X\times_SY$ stets $p_1^{-1}(U)=U\times_SY$ gilt.

Da $p_1^{-1}(U)\subset X\times_SY$, kommutiert das Diagramm:
\[\xymatrix{
p_1^{-1}(U)\ar[r]^-{p_2}\ar[d]_{p_1} & Y\ar[d]\\
U\ar[r] & S
} \]
Sei $Z$ ein $S$-Schema und Morphismen $f:Z\to U,\ g:Z\to Y$ gegeben, so dass $(Z\stackrel{f}{\to} U\stackrel{i}{\hookrightarrow} X\to S)=(Z\stackrel{g}{\to} Y\to S)$. Nach der Universaleigenschaft von $X\times_SY$ existiert ein eindeutiges $\theta:Z\to X\times_SY$ mit $if=p_1\theta$ und $g=p_2\theta$. Insbesondere gilt $\theta(Z)\subset p_1^{-1}(U)$, also $\theta:Z\to p_1^{-1}(U)$. Somit erfüllt $p_1^{-1}(U)$ die Universaleigenschaft von $U\times_SY$.
\item Schritt: Seien $X,Y$ $S$-Schemata und $\{X_i\}$ eine offene Überdeckung von $X$. Wir nehmen an, dass alle Faserprodukte $X_i\times_SY$ mit Projektionen $p_{1i},p_{2i}$ existieren. Wir zeigen, dass in diesem Fall auch $X\times_SY$ existiert.

Setze $X_{ij}=X_i\cap X_j$ und $U_{ij}=p_{1i}^{-1}(X_{ij})\subset X_i\times_SY$. Nach Schritt 2 folgt nun $U_{ij}=X_{ij}\times_SY$. Wegen Eindeutigkeit existieren nun Isomorphismen $\varphi_{ij}:U_{ij}\stackrel{\sim}{\to}U_{ji}$ mit $\varphi_{ij}=\varphi_{ji}^{-1}$, $\varphi_{ij}(U_{ij}\cap U_{ik})=U_{ji}\cap U_{jk}$ und $\varphi_{ik}=\varphi_{jk}\circ\varphi_{ij}$ auf $U_{ij}\cap U_{jk}$. Mithilfe \hyperref[2.11]{2.11} verkleben wir die $X_i\times_SY$ via $\varphi_{ij}$ und erhalten so ein Schema $Z$ mit Morphismen $\psi_j:X_j\times_SY\to Z$, die Isomorphismen auf einem offenen Unterschema induzieren. Seien $p_1,p_2$ die Morphismen, die durch Verkleben der $p_{1i}$ bzw. $p_{2i}$ entstehen, siehe \hyperref[3.19]{Lemma 3.19}. Wir zeigen nun, dass $Z$ gerade das Faserprodukt $X\times_SY$ mit Projektionsmorphismen $p_1,p_2$ ist.

Es gilt $Z=\bigcup_j\psi_j(X_j\times_SY)$. Also folgt die Kommutativität des zweiten Diagramms aus dem ersten:
\[\xymatrix{
X_j\times_S Y\ar[r]^-{p_{1j}}\ar[d]_{p_{2j}} & X_j\ar[d]\\
Y\ar[r] & S
}\qquad\qquad \xymatrix{
Z\ar[d]_{p_2}\ar[r]^-{p_1} & X\ar[d]\\
Y\ar[r] & S
}\]
Sei nun $Z'$ ein weiteres $S$-Schema und $f:Z'\to X,\ g:Z'\to Y$ gegeben, die über $S$ gleich sind. Setze $Z_i'=f^{-1}(X_i)$ für alle $i$. Zu jedem $i$ existiert genau ein Morphismus ${\theta}_i:Z_i'\to X_i\times_SY\hookrightarrow Z$ mit $f|_{Z_i'}=p_{1i}\circ{\theta}_i$ und $g|_{Z_i'}=p_{2i}\circ{\theta}_i$. Es kommutiert:
\[\xymatrix{
X_i\times_SY\ar@{^{(}->}[r]\ar[d]_{p_{1i}} & Z\ar[d]^{p_1}\\
X_i\ar@{^{(}->}[r] & X
} \]
Es gilt $Z_i'\cap Z_j'=f^{-1}(X_i\cap X_j)=f^{-1}(X_{ij})$ und daher $f|_{Z'_i\cap Z'_j}=p_{1i}\circ\theta_i|_{Z'_i\cap Z'_j}=p_{1j}\circ\theta_j|_{Z_i'\cap Z_j'}$, entsprechend für $g$. Wegen Eindeutigkeit folgt $\theta_i|_{Z_i'\cap Z_j'}=\theta_j|_{Z'_i\cap Z'_j}$. Daher können wir die $\theta_i$ zu einem Morphismus $\theta:Z'\to Z$ verkleben mit $f=p_1\theta$ und $g=p_2\theta$. $\theta$ ist eindeutig, da $\theta|_{Z_i'}=\theta_i$ und alle $\theta_i$ eindeutig sind.
\item Schritt: Seien $X,Y$ $S$-Schemata und $S$ affin. Wir zeigen, dass $X\times_SY$ existiert.

Seien $X=\bigcup_iX_i$ und $Y=\bigcup_jY_j$ offene affine Überdeckungen. Nach Schritt 1 existieren $X_i\times_S Y_j$ für alle $i,j$. Nach Schritt 3 existieren $X\times_SY_j$ für alle $j$. Wegen Symmetrie, existiert somit $ X\times_SY$ nach Schritt 3.
\item Schritt: Seien $X,Y$ $S$-Schemata mit $S$ beliebig. Wir zeigen, dass $X\times_SY$ existiert.

Seien $q:X\to S$ und $r:Y\to S$ die Strukturmorphismen und $S=\bigcup_iS_i$ eine offene affine Überdeckung. Setze $X_i=q^{-1}(S_i)$ und $Y_i=r^{-1}(S_i)$. Nach Schritt 4 existieren $X_i\times_{S_i}Y_i$. Wir zeigen, dass $X_i\times_{S_i}Y_i$ die Universaleigenschaft von $X_i\times_SY$ erfüllt.

Seien $f:Z\to X_i$ und $g:Z\to Y$ gegeben, die über $S$ gleich sind. Dann gilt $rg(Z)=qf(Z)\subset q(X_i)\subset S_i$, also $g(Z)\subset Y_i$. Wir erhalten kommutatives Diagramm:
\[\xymatrix{
Z\ar@/_1pc/[ddr]_f\ar@/^1pc/[rrd]_g\ar@/^1pc/[rrrd]^g\ar@{-->}[rd]_\theta\\
& X_i\times_{S_i}Y_i\ar[r]_-{p_2}\ar[d]^{p_1} & Y_i\ar@{^{(}->}[r]\ar[d] & Y\ar[dd]^r\\
& X_i\ar@{^{(}->}[d]\ar[r] & S_i\ar@{^{(}->}[dr] &\\
& X\ar[rr]_q & & S
} \]
Es existiert genau ein $\theta:Z\to X_i\times_{S_i} Y_i$ mit $f=p_1\theta$ und $g=p_2\theta$. Somit existieren auch $X_i\times_SY$. Nach Schritt 3 existiert auch $X\times_SY$.\qed
\end{enumerate}

\paragraph{Lemma 3.20.}\label{3.20} Seien $X,Y$ $S$-Schemata mit Strukturmorphismen $\xi:X\to S,\ \eta:Y\to S$ und $U,V,W$ offen in $X,Y$ bzw. $S$, so dass $\xi(U)\subset W$ und $\eta(V)\subset W$. Ferner seien $p_1,p_2$ die Projektionsmorphismen von $X\times_SY$. Dann gibt es einen $S$-Schemaisomorphismus:
\[p_1^{-1}(U)\cap p_2^{-1}(V)\cong U\times_WV=U\times_SV \]

\paragraph{Beweis.} $E=p_1^{-1}(U)\cap p_2^{-1}(V)$ ist offenes Unterschema von $X\times_SY$. Die Projektionsmorphismen von $X\times_SY$ induzieren $p_1:E\to U,\ p_2:E\to V$. Sei $Z$ ein $W$-Schema und Morphismen $\varphi:Z\to U,\ \psi:Z\to V$ gegeben, die über $W$ gleich sind. Wir erhalten Morphismen $\varphi':Z\to U\hookrightarrow X,\ \psi':Z\to V\hookrightarrow Y$, die über $S$ gleich sind. Es existiert einen eindeutigen Morphismus $\theta:Z\to X\times_SY$ mit $\varphi'=p_1\theta$ und $\psi'=p_2\theta$. Nun gilt $\theta(Z)\subset E$, da $p_1\theta(Z)=\varphi'(Z)\subset U$ und $p_2\theta(Z)=\psi'(Z)\subset V$. Somit ist $\theta:Z\to E$ mit $\varphi=p_1\theta,\ \psi=p_2\theta$. Es folgt $E\cong U\times_WV$. Da $W$ beliebig war, folgt auch $E\cong U\times_SV$.\qed

\paragraph{Definition.} Ein \textit{Unterschema}\index{Schema!Unterschema} $Y$ eines Schemas $X$ ist ein abgeschlossenes Unterschema eines offenen Unterschemas von $X$. Mit anderen Worten haben wir eine abgeschlossene Immersion $\phi$ und eine offene Immersion $\psi$:
\[Y\stackrel{\phi}{\hookrightarrow} U\stackrel{\psi}{\hookrightarrow}X \]
Ein Morphismus $i:Y\to X$ heißt \textit{Immersion}\index{Immersion}, wenn $i$ einen Isomorphismus von $Y$ auf ein Unterschema in $X$ induziert.

\paragraph{Satz 3.21.}\label{3.21} Seien $X,Y,Z$ $S$-Schemata und $W$ ein $Z$-Schema. Dann gilt:
\begin{enumerate}[(i)]
\item $X\times_SS\cong X$
\item $X\times_SY\cong Y\times_SX$
\item $(X\times_SY)\times_SZ\cong X\times_S(Y\times_SZ)$
\item $(X\times_SZ)\times_ZW\cong X\times_SW$
\item $(X\times_SY)\times_SZ\cong (X\times_SZ)\times_Z(Y\times_SZ)$
\item Ist $\sigma:S\to T$ eine Immersion, so ist $X\times_SY\cong X\times_TY$.
\end{enumerate}

\paragraph{Beweis.} (i) bis (v) folgen direkt aus der Universaleigenschaft des Faserprodukts. Für (vi) seien $\varphi:Z\to X,\ \psi:Z\to Y$ Morphismen über $T$ und $\xi:X\to S,\ \eta:Y\to S$ die Strukturmorphismen. Wir erhalten folgendes kommutative Diagramm:
\[ \xymatrix{
Z\ar@/^1pc/[rrd]^\varphi\ar@{-->}[rd]_\theta\ar@/_1pc/[ddr]_\psi\\
& X\ar[r]\ar[d]\times_SY & X\ar[d]_\xi\ar@/^1pc/[ddr] &\\
& Y\ar[r]^\eta\ar@/_1pc/[rrd] & S\ar[dr]^\sigma&\\
&&&T
}\]
Wegen der Injektivität von $\sigma$, folgt aus $\sigma\xi\varphi=\sigma\eta\psi$ stets $\xi\varphi=\eta\psi$. Es existiert ein eindeutiges, kompatibles $\theta:Z\to X\times_SY$. Somit erfüllt $X\times_SY$ die Universaleigenschaft von $X\times_TY$.\qed

\paragraph{Definition 3.22.}\label{3.22} Seien $f:X_1\to X_2$ und $g:Y_1\to Y_2$ $S$-Morphismen mit kommutativen Diagramm:
\[\xymatrix{
&X_1 \ar[r]^f& X_2\\
X_1\times_SY_1\ar[dr]_{p_2}\ar[ur]^{p_1}\ar@{-->}[rrr]^{f\times_Sg} &&& X_2\times_SY_2\ar[lu]_{q_1}\ar[ld]^{q_2}\\
& Y_1\ar[r]_g&Y_2
} \]
Es existiert genau ein Morphismus $f\times_Sg: X_1\times_SY_1\to X_2\times_SY_2$ mit $fp_1=q_1(f\times_Sg)$ und $gp_2=q_2(f\times_Sg)$.

\paragraph{Definition 3.23.}\label{3.23} Sei $f:X\to S$ ein Morphismus von Schemata. Für ein Morphismus $g:T\to S$ erhält man ein $T$-Schema $X_T=X\times_ST$ mit:
\[\xymatrix{
X_T\ar[r]\ar[d] & T\ar[d]^g\\
X\ar[r]_f & S
} \]
Man sagt, dass $X_T$ durch \textit{Basiswechsel} von $X$ über $S$ durch $T$ erhalten bleibt.\index{Basiswechsel}

\paragraph{Lemma 3.24.}\label{3.24} Sei $U\subset S$ ein offenes Unterschema. Dann gilt für den Basiswechsel $X_U$ von $f:X\to S$:
\[X_U=X\times_SU\cong f^{-1}(U) \]

\paragraph{Beweis.} Nach \hyperref[3.21]{Satz 3.21 (i)} ist die Projektion $p_1:X\times_SS\stackrel{\sim}{\to} X$ ein Isomorphismus. Sei in \hyperref[3.20]{Lemma 3.20} $Y=S$ und setze $f=p_2:X\to S$. Wir erhalten: \[f^{-1}(U)=p_1^{-1}(U)\cap p_2^{-1}(U)\cong X\times_SU\qedhere\]

\paragraph{Definition.} Sei $\mathcal{P}$ eine Eigenschaft eines Morphismus' $f:X\to S$. Man sagt, dass $\mathcal{P}$ bei Basiswechsel \textit{erhalten} bleibt, wenn $f_T:X_T\to T$ die Eigenschaft $\mathcal{P}$ besitzt für alle Morphismen $g:T\to S$.

\paragraph{Bemerkung.} Hat $f:X \to S$ eine Eigenschaft $\mathcal{P}$, die stabil unter Basiswechsel ist, so hat insbesondere $f|_{f^{-1}(U)}:f^{-1}(U)\to U$ für alle offenen $U\subset S$ die Eigenschaft $\mathcal{P}$.

\paragraph{Satz 3.25.}\label{3.25}\begin{enumerate}[(i)]
\item Die Eigenschaft (abgeschlossene, offene) Immersion zu sein bleibt bei Basiswechsel erhalten.
\item Die Eigenschaft von endlichem Typ zu sein bleibt stabil unter Basiswechsel.
\end{enumerate}

\paragraph{Beweis.} Wird ausgelassen.

\paragraph{Bemerkung.} Irreduzibilität, Reduziertheit und Integrität bleiben unter Basiswechsel nicht notwendig erhalten.

\paragraph{Lemma 3.26.}\label{3.26} Sei $\psi:A\to B$ ein Ringhomomorphismus, so dass $\alpha(\psi):\operatorname{Spec}(B)\to\operatorname{Spec}(A)$ eine abgeschlossene Immersion ist. Sei ferner $C$ ein beliebiger Ring. Dann ist $\alpha(\psi\otimes\operatorname{id}_C):\operatorname{Spec}(B\otimes_AC)\to\operatorname{Spec}(A\otimes_AC)=\operatorname{Spec}(C)$ eine abgeschlossene Immersion, wobei $\alpha$ die Abbildung aus \hyperref[2.13]{Satz 2.13} bezeichnet.

\paragraph{Beweis.} Da $\tilde{\psi}=\alpha(\psi)$ eine abgeschlossene Immersion ist, ist $\tilde{\psi}^\sharp:\mathcal{O}_A\to\tilde{\psi}_\ast\mathcal{O}_B$ surjektiv. $\tilde{\psi}$ ist Homöomorphismus auf eine abgeschlossene Teilmenge, daher ist $(\tilde{\psi}_\ast\mathcal{O}_B)_{\tilde{\psi}(\mathfrak{p})}=\mathcal{O}_{B,\mathfrak{p}}$ für alle $\mathfrak{p}\in\operatorname{Spec}(B)$. Nach \hyperref[3.12]{Beispiel 3.12} ist die auf den Halmen induzierte Abbildung $\tilde{\psi}^\sharp_\mathfrak{p}$ surjektiv:
\[\tilde{\psi}^\sharp_\mathfrak{p} : A_{\psi^{-1}(\mathfrak{p})}=\mathcal{O}_{A,\tilde{\psi}(\mathfrak{p})}\to (\tilde{\psi}_\ast\mathcal{O}_B)_{\tilde{\psi}(\mathfrak{p})}=B_\mathfrak{p} \]
Betrachte nun die exakte $A$-Modulsequenz $A\stackrel{\psi}{\to}B\to B/\psi(A)\to 0$. Diese induziert unter $-\otimes_AA_\mathfrak{q}$ für alle $\mathfrak{q}\in\operatorname{Spec}(A)$ die exakte Folge:
\[A_\mathfrak{q}\to B\otimes_A A_\mathfrak{q}\to B/\psi(A)\otimes_A A_\mathfrak{q}\to 0 \]
Sei das Bild von $\tilde{\psi}$ von der Form $V(\mathfrak{a})\cong\operatorname{Spec}(B)$ für ein Ideal $\mathfrak{a}\subset A$. Für ein Primideal $\mathfrak{q}\in\operatorname{Spec}(A)$ gilt:
\begin{align*}
B\otimes_A A_\mathfrak{q}\neq 0 &\iff \mathfrak{q}\in\operatorname{Supp}(B)\text{, wobei $B$ als $A$-Modul aufgefasst wird}\\
&\iff \mathfrak{q}\supset\mathfrak{a}\\
&\iff \mathfrak{q}\in\operatorname{im}(\tilde{\psi})\\
&\iff\mathfrak{q}=\psi^{-1}(\mathfrak{p})\text{ für ein }\mathfrak{p}\in\operatorname{Spec}(B)\\
&\implies B/\psi(A)\otimes_AA_\mathfrak{q}=0
\end{align*}
Somit ist $A_\mathfrak{q}\to B\otimes A_\mathfrak{q}$ surjektiv für alle $\mathfrak{q}\in\operatorname{Spec}(A)$. Es folgt die Surjektivität von $\psi$. Wir erhalten unter $-\otimes_AC$ die surjektive Abbildung $C\to B\otimes_AC$. Nach \hyperref[3.12]{Beispiel 3.12} folgt die Behauptung.\qed

\paragraph{Lemma 3.27.}\label{3.27} Sei $f:X\to Y$ ein Morphismus und $Y=\bigcup_\lambda Y_\lambda$ eine offene Überdeckung. Dann ist $f$ genau dann eine (abgeschlossene, offene) Immersion, wenn $f_\lambda=f|_{X_\lambda}:X_\lambda \to Y_\lambda$ mit $X_\lambda=f^{-1}(Y_\lambda)$ die Eigenschaft hat.

\paragraph{Beweis.} Kein Beweis.

\paragraph{Bemerkung 3.28.}\label{3.28} Sind $Y_1,Y_2$ Unterschemata von $X$, so ist $Y_1\times_XY_2\to X$ ein Unterschema von $X$, d.h. eine Immersion.

\paragraph{Definition 3.29.}\label{3.29} Sei $f:X\to Y$ ein $S$-Morphismus von $S$-Schemata. Der Morphismus $\Gamma_f=(\operatorname{id}_X,f)_S:X\to X\times_SY$ heißt \textit{$S$-Graph}\index{Graph} von $f$ und $\Gamma_f(X)$ heißt \textit{Graph} von $f$. Ist $f=\operatorname{id}_X$ die Identität, so heißt $\Delta_{X/S}=\Gamma_{\operatorname{id}_X}$ der \textit{Diagonalmorphismus}\index{Diagonalmorphismus}.

\paragraph{Satz 3.30.}\label{3.30} Sei $f:X\to Y$ ein $S$-Morphismus von $S$-Schemata. \begin{enumerate}[(i)]
\item Sind $Y$ und $S$ affine Schemata, so ist $\Gamma_f$ eine abgeschlossene Immersion. Insbesondere ist für $X,S$ affin $\Delta_{X/S}$ eine abgeschlossene Immersion.
\item $\Gamma_f$ ist allgemein eine Immersion.
\end{enumerate}

\paragraph{Beweis.} \begin{enumerate}[(i)]
\item Sei $X=\bigcup_\alpha X_\alpha$ eine offene affine Überdeckung und $X\times_SY$ das Faserprodukt. Nach \hyperref[3.24]{Lemma 3.24} ist $p_1^{-1}(X_\alpha)\cong X_\alpha\times_SY$ affin. Da $p_1\circ\Gamma_f=\operatorname{id}_X$, gilt $\Gamma_f^{-1}(p_1^{-1}(X_\alpha))=X_\alpha$. Da $\Gamma_f: \bigcup_\alpha\Gamma_f^{-1}(X_\alpha\times_SY) \to \bigcup_\alpha(X_\alpha\times_SY)=X\times_SY$, können wir nach \hyperref[3.27]{Lemma 3.27} o.B.d.A. annehmen, dass $X$ affin ist.

Sei also $S=\operatorname{Spec}(R),\ X=\operatorname{Spec}(B)$ und $Y=\operatorname{Spec}(A)$. Sei $f$ induziert von $\varphi:A\to B$ und $\Gamma_f$ induziert von $\psi:B\otimes_RA\to A,\ b\otimes a\mapsto b\varphi(a)$. $\psi$ ist offensichtlich surjektiv, daher ist $\Gamma_f$ nach \hyperref[2.19]{Satz 2.19 (ii)} eine abgeschlossene Immersion.
\item Sei zunächst $S$ affin und $Y=\bigcup_\alpha Y_\alpha$ eine offene affine Überdeckung. Setze $X_\alpha=f^{-1}(Y_\alpha)$, $\Gamma_{f\alpha}=\Gamma_f|_{X_\alpha}$ und $f_\alpha=f|_{X_\alpha}:X_\alpha\to Y_\alpha$. Es ist $\Gamma_f(X_\alpha)\subset X_\alpha\times_SY_\alpha$ nach (i) ein abgeschlossenes Unterschema und $X_\alpha\times_SY_\alpha\subset X\times_SY_\alpha$ nach \hyperref[3.25]{Satz 3.25 (i)} ein offenes Unterschema ist. Nach (i) ist $\Gamma_{f\alpha}:X_\alpha\to X_\alpha\times_SY_\alpha$ eine Immersion. Nach \hyperref[3.27]{Lemma 3.27} ist auch $\Gamma_f$ eine Immersion.

Sei nun $S$ beliebig und $S=\bigcup_\lambda S_\lambda$ eine offene affine Überdeckung. Nach \hyperref[3.21]{Satz 3.21 (v)} ist $(X\times_SY)\times_SS_\lambda=X_\lambda\times_{S_\lambda}Y_\lambda$, wobei $X_\lambda=X\times_SS_\lambda$ und $Y_\lambda=Y\times_SS_\lambda$. Ferner ist $\Gamma_{f_\lambda}$ der Graph von $f_\lambda:X_\lambda\to Y_\lambda$ eine Immersion. Nach \hyperref[3.27]{Lemma 3.27} ist auch $\Gamma_f$ eine Immersion.\qed
\end{enumerate}

\paragraph{Definition 3.31.}\label{3.31} Sei $f:X\to Y$ ein Morphismus von Schemata und $y\in Y$ ein Punkt mit Restklassenkörper $\kappa(y)=\mathcal{O}_{Y,y}/\mathfrak{m}_{Y,y}$. Sei weiter $i:\operatorname{Spec}\kappa(y)\to Y$ der natürliche Morphismus gegeben durch $(y,\operatorname{id}_{\kappa(y)})$ (siehe \hyperref[2.18]{Satz 2.18}). Dann heißt das Faserprodukt $X_y=X\times_Y\operatorname{Spec}\kappa(y)$ die \textit{Faser}\index{Faser} von $f$ über dem Punkt $y$.
\[\xymatrix{
X_y\ar[r]^-{p_2}\ar[d]_{p_1} & \operatorname{Spec}\kappa(y)\ar[d]^i\\
X\ar[r]_f & Y
}\]

\paragraph{Satz.} $X_y$ ist ein $\kappa(y)$-Schema mit $f^{-1}(y)$ als unterliegender topologischer Raum.

\paragraph{Beweis.} Es ist $fp_1(X_y)=ip_2(X_y)=y$, also folgt $p_1(X_y)\subset f^{-1}(y)$ und somit $X_y=p_1^{-1}(f^{-1}(y))$. Sei $V\subset_\text{o}X$ affin mit $V\cap f^{-1}(y)\neq\varnothing$. Dann gilt:
\[p_1^{-1}(V)=V_y\cong X\times_Y\operatorname{Spec}\kappa(y)\times_XV=X_y\times_XV \]
Wir können somit o.B.d.A. $X$ und $Y$ als affin annehmen, sei also $X=\operatorname{Spec}(B),\ Y=\operatorname{Spec}(A),\ y=\mathfrak{p}$. Sei $\varphi:A\to B$ assoziiert zu $f$. Setze $\varphi':A_\mathfrak{p}\to B'=B\otimes_AA_\mathfrak{p}=\varphi(A\setminus\mathfrak{p})^{-1}B$. Es gibt eine Folge von Homöomorphismen:
\begin{align*}
f^{-1}(y) &= \{\mathfrak{q}\in\operatorname{Spec}(B)\mid\varphi^{-1}(\mathfrak{q})=\mathfrak{p} \}\\
&\cong \{\mathfrak{q}\in\operatorname{Spec}(B')\mid\varphi'^{-1}(\mathfrak{q})=\mathfrak{p}A_\mathfrak{p} \}\\
&\cong \{ \mathfrak{q}\in\operatorname{Spec}(B')\mid\mathfrak{q}\supset (\mathfrak{p}A_\mathfrak{p})B' \} \\
&=\operatorname{Spec}(B'/(\mathfrak{p}A_\mathfrak{p})B')\\
&=\operatorname{Spec}(B'\otimes_{A_\mathfrak{p}}A_\mathfrak{p}/\mathfrak{p}A_\mathfrak{p})\\
&=\operatorname{Spec}((B\otimes_AA_\mathfrak{p})\otimes_{A_\mathfrak{p}}A_\mathfrak{p}/\mathfrak{p}A_\mathfrak{p}) =\operatorname{Spec}(B\otimes_A\kappa(\mathfrak{p}))\qedhere
\end{align*}

\paragraph{Definition 3.32.}\label{3.32} Sei $f:X\to Y$ ein Morphismus von Schemata, so dass $f(X)\subset Y$ dicht liegt, und $Y$ irreduzibel mit generischer Punkt $\xi$ mit $\xi\in f(X)$. Dann heißt $f^{-1}(\xi)$ \textit{generische Faser}\index{generische Faser} von $f$.

\paragraph{Beispiel 3.33.}\label{3.33} Sei $k$ ein algebraisch abgeschlossener Körper und:
\[X=\operatorname{Spec}k[X,Y,t]/(tY-X^2),\quad Y=\operatorname{Spec}k[T] \]
Sei $f:X\to Y$ gegeben durch die kanonische Abbildung $\varphi:k[t]\to k[X,Y,t]/(tY-X^2)$. $X$ und $Y$ sind integre Schemata von endlichem Typ über $k$. Ferner ist $f$ surjektiv, da $\varphi$ injektiv ist (vgl. \hyperref[2.19]{Satz 2.19 (ii)}). Abgeschlossene Punkte von $Y$ entsprechen $k$. Sei $a\in Y$ ein abgeschlossener Punkt und betrachte die Faser: \[X_a=X\times_Y\operatorname{Spec}\kappa(a)=\operatorname{Spec}k[X,Y]/(aY-X^2)\]
Im Fall $a\neq 0$ ist $X_a$ eine ebene Kurve in $\mathbf{A}_k^2$, die irreduzibel und reduziert ist. Für $a=0$ ist $X_0=\operatorname{Spec}k[X,Y]/(X^2)$ die $Y$-Achse. Diese ist nicht reduziert.

\section{Separierte \& eigentliche Morphismen}

\paragraph{Definition 4.1.}\label{4.1} Sei $f:X\to Y$ ein Morphismus von Schemata. $f$ heißt \textit{separiert}, falls $\Delta_{X/Y}:X\to X\times_YX$ eine abgeschlossene Immersion ist. In diesem Fall sagen wir auch, dass $X$ über $Y$ separiert ist.

Ein Schema $X$ heißt \textit{separiert}\index{separiert}, falls $X\to\operatorname{Spec}(\mathbb{Z})$ separiert ist.

\paragraph{Beispiel 4.2.}\label{4.2} Sei $k$ ein Körper und $X$ die affine Gerade über $k$ mit doppeltem Nullpunkt wie in \hyperref[2.12]{Beispiel 2.12}. Dann ist $X\times_kX$ die affine Ebene mit vierfachen Nullpunkt und $\Delta(X)$ die gewöhnliche Diagonale, die zwei Nullpunkte besitzt. $\Delta(X)$ ist nicht abgeschlossen, da $\overline{\Delta(X)}$ vier Nullpunkte besitzt.

\paragraph{Beispiel 4.3.}\label{4.3} Ist $V$ eine Varietät über eine algebraisch abgeschlossenen Körper $k$, so ist $t(V)$ separiert über $k$. Dies werden wir später zeigen.

\paragraph{Beispiel 4.4.}\label{4.4} Ist $f:X\to Y$ ein Morphismus affiner Schemata, so ist $f$ separiert nach \hyperref[3.30]{Satz 3.30 (i)}.

\paragraph{Beispiel 4.5.}\label{4.5} Ist $f:X\to Y$ eine Immersion, so ist $f$ separiert, da nach \hyperref[3.21]{Satz 3.21 (vi)} $X\cong X\times_XX\cong X\times_YX$ gilt, d.h. $\Delta_{X/Y}:X\to X$ ist ein Isomorphismus.

\paragraph{Satz 4.6.}\label{4.6} Sei $f:X\to Y$ ein Morphismus von Schemata. Dann ist $f$ genau dann separiert, wenn $\Delta(X)\subset X\times_YX$ abgeschlossen ist.

\paragraph{Beweis.} Für die nicht-triviale Richtung sei $\Delta(X)\subset X\times_YX$ abgeschlossen. Wir müssen zeigen, dass $X\to \Delta(X)$ ein Homöomorphismus ist, und dass der Morphismus $\mathcal{O}_{X\times_YX}\to\Delta_\ast\mathcal{O}_X$ surjektiv ist.
\begin{itemize}
\item Sei $p_1:X\times_YX\to X$ die erste Projektion. Da $p_1\circ\Delta=\operatorname{id}_X$, induziert $\Delta$ ein Homöomorphismus auf sein Bild. 
\item Sei $P\in X$ und $V\subset_\text{o} Y$ affin. Wähle $U\subset_\text{o}X$ affin mit $P\in U$ und $f(U)\subset V$. Dann ist $U\times_VU$ eine offene, affine Umgebung von $\Delta(P)$. Nach \hyperref[4.4]{Beispiel 4.4} ist $\Delta:U\to U\times_VU$ eine abgeschlossene Immersion. Somit ist $\mathcal{O}_{X\times_YX}|_{U\times_VU}\to \Delta_\ast\mathcal{O}_X|_U$ surjektiv.\qed
\end{itemize}

\paragraph{Definition 4.7.}\label{4.7} Ein Ring $R$ heißt \textit{Bewertungsring}\index{Bewertungsring}, wenn $R$ nullteilerfrei ist, wenn von der folgenden Form ist:
\[R=\{x\in K^\times\mid v(x)\geq 0\} \]
wobei $v:K^\times\to G$ eine Bewertung ist, d.h. $v(xy)=v(x)+v(y)$ und $v(x+y)\geq\min\{v(x),v(y)\}$, und $G$ eine total geordnete abelsche Gruppe ist. $R$ ist dann lokaler Ring mit Maximalideal $\mathfrak{m}=\{x\in K^\times\mid v(x)>0\}\cup\{0\}$ und $K=\operatorname{Quot}(R)$.

\paragraph{Theorem 4.8.}\label{4.8} \textit{(Bewertungstheoretisches Kriterium für Separiertheit)}\index{Bewertungstheoretisches Kriterium} Sei $f:X\to Y$ ein Morphismus von Schemata mit $X$ noethersch. Es sind folgende Aussagen äquivalent:
\begin{enumerate}[(i)]
\item $f$ ist separiert.
\item Sei $R$ ein Bewertungsring des Körpers $K=\operatorname{Quot}(R)$ und $i:U=\operatorname{Spec}(K)\hookrightarrow T=\operatorname{Spec}(R)$ die kanonische Inklusion. Seien ferner Morphismen $T\to Y,\ U\to X$ derart gegeben, dass das folgende Diagramm kommutativ ist:
\[\xymatrix{
U\ar[r]\ar[d]_i & X\ar[d]^f\\
T\ar[r]\ar@{-->}[ur]^h & Y
}\]
Dann gibt es höchstens eine Abbildung $h:T\to X$, die das obige Diagramm kommutativ macht. Mit anderen Worten: Die folgende kanonische Abbildung ist für jeden Bewertungsring $R$ über $Y$ injektiv:
\[\operatorname{Hom}_Y(\operatorname{Spec}(R),X)\to\operatorname{Hom}_Y(\operatorname{Spec}(K),X),\ h\mapsto h\circ i \]
\end{enumerate}

\paragraph{Definition 4.9.}\label{4.9}  Sei $X$ ein topologischer Raum und $x\in X$. Ein Punkt $x'\in X$ heißt \textit{Spezialisierung}\index{Spezialisierung} von $x$, wenn $x'\in\overline{\{x\}}$ gilt. Ist $x''$ Spezialisierung von $x'$ und $x'$ Spezialisierung von $x$, so ist $x''$ auch Spezialisierung von $x$.

Eine Teilmenge $Z\subset X$ heißt \textit{stabil unter Spezialisierungen}, falls mit $x\in Z$ auch jede Spezialisierung in $Z$ ist. Abgeschlossene Mengen sind stabil unter Spezialisierungen. Die Umkehrung gilt im Allgemeinen nicht.

Ein Morphismus von Schemata $f$ heißt \textit{quasikompakt}\index{quasikompakt}, falls es eine offene affine Ü\-ber\-deckung $Y=\bigcup_\alpha Y_\alpha$ existiert, so dass alle $f^{-1}(Y_\alpha)$ quasikompakt sind. Es gilt:
\[f\text{ quasikompakt}\iff\forall V\subset_\text{o}Y\text{ affin}\colon f^{-1}(V)\text{ quasikompakt} \]

\paragraph{Satz 4.10.}\label{4.10} Sei $f:X\to Y$ ein quasikompakter Morphismus von Schemata. Dann sind folgende Aussagen äquivalent:
\begin{enumerate}[(i)]
\item $f$ ist abgeschlossen.
\item Ist $x\in X$ und $y'$ eine Spezialisierung von $y=f(x)$, so gibt es eine Spezialisierung $x'$ von $x$ mit $f(x')=y'$. Mit anderen Worten: $f(\overline{\{x\}})$ ist stabil unter Spezialisierungen für alle $x\in X$.
\end{enumerate}

\paragraph{Beweis.} (i)$\implies$(ii) ist trivial. Sei also $X'\subset X$ abgeschlossen und setze $Y'=\overline{f(X')}$. Wir zeigen $Y'=f(X')$. Wir versehen $X'$ und $Y'$ mit der eindeutig bestimmten, reduzierte abgeschlossene Unterschemastruktur in $X$ bzw. $Y$. Seien $i:X'\hookrightarrow X$ und $j:Y'\hookrightarrow Y$ die natürlichen Inklusionen. Das Urbildschema $(f\circ i)^{-1}(Y')$ ist reduziert, besitzt $X'$ als unterliegender topologischer Raum und ist als Schema gleich $X'$, da die reduzierte Unterschemastrukturen eindeutig sind. Somit existiert ein eindeutig bestimmter Morphismus $f':X'\to Y'$:
\[\xymatrix{
X'\ar[r]^{f'}\ar[d]_i & Y'\ar[d]^j\\
X\ar[r]_f & Y
} \]
$f'$ erfüllt ebenfalls die Voraussetzung (ii). Ferner ist $f'$ quasikompakt. $i$ ist als abgeschlossene Immersion quasikompakt, also auch $f\circ i=j\circ f':X'\to Y$. Unter Basiswechsel sehen wir, dass $X'\times_YY'\to Y'$ quasikompakt ist. Da $j$ eine Immersion ist, ist $X'\cong X'\times_{Y'}Y' = X'\times_YY'$, also ist $X'\to Y'$ quasikompakt.

Wir müssen noch zeigen: Ist $f:X\to Y$ ein quasikompakter, dominanter Morphismus reduzierter Schema, welcher (ii) erfüllt, so ist $f$ surjektiv. Sei dazu $y'\in Y$ gegeben und $y$ ein generischer Punkt der irreduziblen Komponente von $Y$, in der $y'$ liegt. Somit ist $y'$ eine Spezialisierung von $y$. Gilt nun $f^{-1}(y)\neq\varnothing$, so gibt es ein $x\in X$ mit $f(x)=y$. Wegen (ii) gibt es eine Spezialisierung $x'$ von $x$ mit $f(x')=y'$ und wir sind fertig. Somit ist noch das nächste Lemma zu zeigen.\qed

\paragraph{Lemma 4.12.}\label{4.12} Für einen quasikompakten Morphismus $f:X\to Y$ ist äquivalent:
\begin{enumerate}[(i)]
\item $f$ ist dominant.
\item Für die generischen Punkte $y$ von $Y$ der irreduziblen Komponenten gilt $f^{-1}(y)\neq\varnothing$.
\end{enumerate}

\iffalse
\paragraph{Beweis.} (ii)$\implies$(i) ist trivial. Für die andere Richtung ersetzen wir $f$ durch $f_\text{red}$ und nehmen $X,Y$ als reduziert an. Sei $y\in Y$ ein generischer Punkt einer irreduziblen Komponente von $Y$ und $U\subset_\text{o}Y$ eine affine Umgebung von $y$. $U$ ist insbesondere quasikompakt, also auch $f^{-1}(U)$, d.h. $f^{-1}(U)=\bigcup_iV_i$ hat eine endliche Überdeckung durch offene, affine $V_i$. Da $f$ nach Voraussetzung dominant ist, folgt $U=\overline{ff^{-1}(U)}=\bigcup_i\overline{f(V_i)}$. Daher gibt es ein $i$ mit $y\in\overline{f(V_i)}$. $\overline{f(V_i)}\subset U$ wird mit der reduzierten, abgeschlossenen Unterschemastruktur versehen. Da alle $V_i$ reduziert sind, faktorisiert $f|_{V_i}:V_i\to U$ eindeutig in der Form $V_i\stackrel{g}{\to} \overline{f(V_i)}\hookrightarrow U$.

Somit genügt es zu zeigen, dass $g^{-1}(y)\neq \varnothing$, da $g^{-1}(y)=f^{-1}(y)\cap V_i$. $y$ ist generischer Punkt einer irreduziblen Komponente von $Y$, daher gilt $\mathcal{O}_{U,y}=\mathcal{O}_{Y,y}$.
\fi

\paragraph{Korollar 4.13.}\label{4.13} Eine quasikompakte Immersion $f:X\to Y$ ist genau dann eine abgeschlossene Immersion, wenn $f(X)$ in $Y$ stabil unter Spezialisierungen ist.

\paragraph{Beweis.} $f$ faktorisiert in eindeutiger Weise in der Form $X\stackrel{\sim}{\to}Z\stackrel{i}{\hookrightarrow} Y$, wobei $i$ die kanonische Inklusion des Unterschema $Z\subset Y$ ist. Es ist $i$ quasikompakt. Ist $f(X)=Z$ stabil unter Spezialisierungen, ist (ii) in \hyperref[4.10]{Satz 4.10} erfüllt, daher ist $i:Z\to Y$ abgeschlossen und $Z$ ein abgeschlossenes Unterschema in $Y$.\qed

\paragraph{Lemma 4.14.}\label{4.14} Sei $R$ ein Bewertungsring eines Körpers $K$ und $T=\operatorname{Spec}(R),\ U=\operatorname{Spec}(K)$. Sei $X$ ein Schema. Dann gilt:
\begin{enumerate}[(i)]
\item $\operatorname{Hom}(U,X)$ ist die Menge aller Paare $(x,i)$ mit $x\in X$ und Körperhomomorphismus $i:\kappa(x)\hookrightarrow K$, wobei $\kappa(x)$ der Restklassenkörper in $x$ bezeichnet.
\item $\operatorname{Hom}(T,X)$ ist die Menge aller Tripel $(x_0,x_1,i)$ mit $x_0,x_1\in X,\ x_0\in\overline{\{x_1\}}$ und Körperhomomorphismus $\kappa(x_1)\hookrightarrow K$, so dass $R$ über $\mathcal{O}_{\overline{\{x_1\}},x_0}$ \textit{dominiert}\index{dominiert}, wobei $\overline{\{x_1\}}$ mit der reduzierten Unterschemastruktur ausgestattet ist.
\end{enumerate}

\paragraph{Definition 4.15.}\label{4.15} Seien $(A,\mathfrak{m}_A),(B,\mathfrak{m}_B)$ lokale Ringe, die in einem Körper $K$ eingebettet sind. Wir sagen $B$ \textit{dominiert}\index{dominiert} $A$, wenn $A\subset B$ und $\mathfrak{m}_B\cap A=\mathfrak{m}_A$ gilt. Mit anderen Worten, wenn die natürliche Inklusion $A\hookrightarrow B$ ein lokaler Homomorphismus ist.

\paragraph{Beweis von \hyperref[4.14]{Lemma 4.14}.} (i) ist gerade \hyperref[2.18]{Satz 2.18}. Setze $t_0=\mathfrak{m}_R\in T$ und $t_1=(0)\in T$. Für (ii) sei $f:T\to X$ ein Morphismus. Setze $x_i=f(t_i)$ und $Z=\overline{\{x_1\}}$ mit der reduzierten Unterschemastruktur. Dann gilt $f^{-1}(Z)=T$ als Mengen. Nach \hyperref[2.21]{Regeln 2.21 (iii)} faktorisiert $f$ in der Form:
\[\xymatrix{
T\ar[rr]^f\ar[rrd] &&X\\
&& Z\ar[u]
} \]
Beachte, dass für $U\subset_\text{o} Z$ mit $x_0\in U$ auch $x_1\in U$ gilt. Das induziert $\mathcal{O}_{Z,x_0}\to\mathcal{O}_{Z,x_1}$, analog haben wir $\mathcal{O}_{T,t_0}\to\mathcal{O}_{T,t_1}$. Wir erhalten folgendes kommutative Diagramm:
\[\xymatrix{
\mathcal{O}_{Z,x_1} \ar[r]& \mathcal{O}_{T,t_1}=K\\
\mathcal{O}_{Z,x_0} \ar[r]\ar[u]& \mathcal{O}_{T,t_0}=R\ar@{^{(}->}[u]\\
\mathfrak{m}_{Z,x_0}\ar[r]\ar@{^{(}->}[u] & \mathfrak{m}_{T,t_0}\ar@{^{(}->}[u]
}\]
Da $Z$ reduziert und irreduzibel ist, ist $Z$ integer. Daher ist $\mathcal{O}_{Z,x_1}=\kappa(x_1)$, alle Abbildungen sind injektiv und wir sehen, dass $R$ über $\mathcal{O}_{Z,x_0}$ dominiert.

Sei umgekehrt $x_0,x_1\in X$ mit $x_0\in\overline{\{x_1\}}= Z$ und $\mathcal{O}_{Z,x_0}\hookrightarrow R$ ein lokaler Homomorphismus. Das induziert $T=\operatorname{Spec}(R)\to\operatorname{Spec}(\mathcal{O}_{Z,x_0})\to Z\to X$. Diese beiden Konstruktionen sind invers zueinander.\qed

\paragraph{Beweis zu \hyperref[4.8]{Theorem 4.8}.} Sei $f:X\to Y$ ein Morphismus von Schemata mit $X$ noethersch.
\begin{itemize}
\item Sei zunächst $f$ separiert und $R$ ein Bewertungsring des Körpers $K=\operatorname{Quot}(R)$ mit Inklusion $i:U=\operatorname{Spec}(K)\hookrightarrow\operatorname{Spec}(R)=T$. Seien $T\to Y,\ U\to X$ gegeben und $h_1,h_2:T\to X$ mit kommutativem Diagramm:
\[\xymatrix{
U\ar[d]_i\ar[r] & X\ar[d]^f\\
T\ar[r]\ar@<.5ex>[ur]^{h_1}\ar@<-.5ex>[ur]_{h_2} & Y
} \]
Setze $t_0=\mathfrak{m}_R\in T$ und $t_1=(0)\in T$. Aus der Kommutativität $h_1i=h_2i$ folgt insbesondere $h_1(t_1)=h_2(t_1)$. Setze $h''=(h_1,h_2)_Y:T\to X\times_Y X$. Betrachte nun folgendes kommutative Diagramm:
\[\xymatrix{
U\ar[r]\ar[d]^i\ar@/_3pc/[rdd]_{h_2i}\ar@/^3pc/[rrd]^{h_1i} & X\ar[d]^-\Delta &\\
T\ar[r]_-{h''} & X\times_Y X\ar[r]^-{p_1}\ar[d]_-{p_2} & X\ar[d]^f\\
& X\ar[r]_f & Y
} \]
Es gilt $\{h''(t_1)\}= h''i(U)\subset\Delta(X)$. Da $f$ separiert ist, ist $\Delta(X)$ abgeschlossen und es folgt $h''(t_0)\in h''(\overline{\{t_1\}})\subset\overline{\{h''(t_1)\}}\subset\Delta(X)$. Es folgt:
\[h_1(t_0)=p_1h''(t_0)=p_2h''(t_0)=h_2(t_0) \]
Aus der Kommutativität folgt, dass $h_1^\sharp$ und $h_2^\sharp$ dieselbe Abbildung $\kappa(x_1)\hookrightarrow K$ mit $x_1=h_1(t_1)=h_2(t_1)$ induzieren. Mit \hyperref[4.14]{Lemma 4.14 (ii)} folgt $h_1=h_2$.
\item Für die andere Richtung genügt es nach \hyperref[4.6]{Satz 4.6} zu zeigen, dass $\Delta(X)\subset X\times_YX$ abgeschlossen ist. Nach \hyperref[3.30]{Satz 3.30 (ii)} ist $\Delta$ eine Immersion. Da $X$ noethersch ist, ist $\Delta$ quasikompakt. Nach \hyperref[4.13]{Korollar 4.13} genügt es zu zeigen, dass $\Delta(X)$ stabil unter Spezialisierungen ist.

Sei also $\xi_1\in\Delta(X)$ und $\xi_0\in\overline{\{\xi_1\}}$. Sei $K=\kappa(\xi_1)$ und $\mathcal{O}=\mathcal{O}_{\overline{\{\xi_1\}},\xi_0}$, wobei $\overline{\{\xi_1\}}$ mit der reduzierten Unterschemastruktur versehen ist, d.h. $\mathcal{O}$ ist ein lokaler Ring in $K$. Mit \hyperref[4.16]{Satz 4.16} und dem Lemma von Zorn sehen wir, dass für jeden lokalen Ring $\mathcal{O}$ in $K$ einen Bewertungsring $R$ gibt, der $\mathcal{O}$ dominiert. Sei $R$ ein solcher Bewertungsring. Nach \hyperref[4.14]{4.14 (ii)} haben wir einen Morphismus $h:T=\operatorname{Spec}(R)\to X\times_YX$ mit $t_i\mapsto\xi_i$, wobei $t_0=\mathfrak{m}_R\in T,\ t_1=(0)\in T$. Betrachte das kommutative Diagramm:
\[\xymatrix{
\operatorname{Spec}(K)\ar[r]^-i & T\ar[dr]^h\ar@/^1pc/[rrd]^{p_1h}\ar@/_1pc/[ddr]_{p_2h}\\
& & X\times_YX\ar[r]_-{p_1}\ar[d]^{p_2} & X\ar[d]^f\\
&& X\ar[r]_f & Y
}\]
Ist $x\in X$, so ist $\kappa(x)\cong\kappa(\Delta(x))$, wegen:
\[\xymatrix{
X\ar[rr]^-\Delta\ar[rrd]_{\operatorname{id}} & & X\times_Y X\ar[d]^{p_i}\\
&& X
} \]
Betrachte nun das folgende Diagramm:
\[\xymatrix{
\operatorname{Spec}(K)\ar[rd]_{hi}\ar@{-->}[r]^-g & X\ar[d]^\Delta \\
& X\times_YX 
}\]
Wähle ein $x_1\in X$ mit $\Delta(x_1)=\xi_1$ und $j:\kappa(x_1)\hookrightarrow K$ als den von $hi$ induzierten $\kappa(\Delta(x_1))\hookrightarrow K$. Nach \hyperref[4.14]{Lemma 4.14 (i)} erhalten wir den zu $(x_1, j)$ passenden Morphismus $g:\operatorname{Spec}(K)\to X$, das das obige Diagramm kommutativ macht. Also faktorisiert $\operatorname{Spec}(K)\to X\times_YX$ über $\operatorname{Spec}(K)\to X\stackrel{\Delta}{\to}X\times_YX$, d.h. $p_1hi=p_2hi$. Nach Voraussetzung folgt $p_1h=p_2h$, d.h. auch $T\to X\times_YX$ faktorisiert über $T\to X\stackrel{\Delta}{\to}X\times_YX$. Somit folgt $\xi_0\in \Delta(X)$.\qed
\end{itemize}

\paragraph{Satz 4.16.}\label{4.16} Sei $K$ ein Körper und $R\subset K$ ein lokaler Ring. Dann ist $R$ genau dann ein Bewertungsring, wenn für jeden lokalen Ring $R\subset S\subset K$ mit lokalen Homomorphismus $R\hookrightarrow S$ stets $R=S$ folgt.

\paragraph{Beweis.} Siehe z.B. Bourbaki, Algebra VI §13.

\paragraph{Satz 4.17.}\label{4.17} Alle involvierten Schemata seine noethersch. Dann gilt:
\begin{enumerate}[(i)]
\item Offene und abgeschlossene Immersionen sind separiert.
\item Die Komposition separierter Morphismen ist separiert.
\item Separiertheit ist stabil unter Basiswechsel.
\item Sind $f:X\to Y,\ f':X'\to Y'$ separierte Morphismen von $S$-Schemata, so ist auch $f\times_Sf':X\times_SX'\to Y\times_SY'$ separiert.
\item Ist $f\circ f$ separiert, so ist auch $f$ separiert.
\item $f:X\to Y$ ist genau dann separiert, wenn es eine offene Überdeckung $Y=\bigcup_\alpha Y_\alpha$ gibt, so dass alle $f^{-1}(Y_\alpha)\to Y_\alpha$ separiert sind.
\end{enumerate}

\paragraph{Beweis.} Folgt alles aus \hyperref[4.8]{Theorem 4.8}.\qed

\paragraph{Definition 4.18.}\label{4.18} Ein Morphismus von Schemata $f:X\to Y$ heißt \textit{eigentlich}\index{eigentlich}, wenn gilt:
\begin{enumerate}[(i)]
\item $f$ ist von endlichem Typ.
\item $f$ ist separiert.
\item $f$ ist \textit{universell abgeschlossen}\index{universell abgeschlossen}, d.h. für jeden Morphismus $Y'\to Y$ ist der Morphismus $f':X'=X\times_YY'\to Y'$ abgeschlossen.
\end{enumerate}
In diesem Fall heißt $X$ auch \textit{eigentlich über} $Y$.

\paragraph{Beispiel 4.19.}\label{4.19} Sei $k$ ein Körper und $X=\mathbf{A}_k^1$ die affine Gerade. Dann ist $X$ separiert und von endlichem Typ. Basiserweiterung mit $X\to k$ ergibt $X\times_kX\to X$, die Projektionsabbildung $\mathbf{A}_k^2\to\mathbf{A}_k^1$. Sei $Y=\operatorname{Spec}(k[X,Y]/(XY-1))$ die hyperbolische Kurve in $\mathbf{A}_k^2$. $Y\subset\mathbf{A}_k^2$. Diese ist abgeschlossen in $\mathbf{A}_k^2$ und projeziert sich in $\mathbf{A}_k^1$ auf $\mathbf{A}_k^1\setminus\{0\}$, die in $\mathbf{A}_k^1$ nicht abgeschlossen ist. Daher ist die Projektionsabbildung nicht abgeschlossen. In diesem Beispiel fehlt der unendliche Punkt von $Y$. Wir werden später zeigen, dass $X$ eigentlich über $k$ ist, wenn $X$ eine sogenannte projektive Varietät.

\paragraph{Theorem 4.20.}\label{4.20} \textit{(Bewertungstheoretisches Kriterium für Eigentlichkeit)}\index{Bewertungstheoretisches Kriterium} Sei $f:X\to Y$ ein Morphismus von endlichem Typ mit $X$ noethersch. Dann sind äquivalent:
\begin{enumerate}[(i)]
\item $f$ ist eigentlich.
\item Sei $R$ ein Bewertungsring des Körpers $K=\operatorname{Quot}(R)$ und $i:U=\operatorname{Spec}(K)\hookrightarrow T=\operatorname{Spec}(R)$ die kanonische Inklusion. Seien ferner Morphismen $T\to Y$ und $U\to X$ derart gegeben, dass das folgende Diagramm kommutativ ist:
\[\xymatrix{
U\ar[r]^u\ar[d]_i & X\ar[d]^f\\
T\ar[r]_t\ar@{-->}[ru]^h & Y
}\]
Dann existiert ein eindeutig bestimmter Morphismus $h:T\to X$, der das Diagramm kommutativ macht. Mit anderen worten ist die folgende Abbildung bijektiv:
\[\operatorname{Hom}_Y(T,X)\to\operatorname{Hom}_Y(U,X),\ h\mapsto h\circ i \]
\end{enumerate}

\paragraph{Lemma 4.21.}\label{4.21} \begin{enumerate}[(i)]
\item Abgeschlossene Immersinoen sind von endlichem Typ. Quasikompakte offene Immersionen sind von endlichem Typ.
\item Kompositum zweier Morphismen von endlichem Typ ist von endlichem Typ.
\item Ein Morphismus $f:X\to Y$ ist genau dann von endlichem Typ, wenn $f$ lokal von endlichem Typ und quasikompakt ist.
\item Seien $X\stackrel{f}{\to}Y\stackrel{g}{\to}Z$ Morphismen mit $f$ quasikompakt und $g\circ f$ von endlichem Typ. Dann ist $f$ von endlichem Typ.
\end{enumerate}

\paragraph{Beweis von \hyperref[4.20]{Theorem 4.20}.}\begin{itemize}
\item Sei $f$ eigentlich. Nach Definition ist $f$ separiert, somit ist ein Morphismus $h:T\to X$ wie oben eindeutig bestimmt nach \hyperref[4.8]{Theorem 4.8}. Es bleibt die Existenz zu zeigen. Betrachte die Basiserweiterung $X_T=X\times_YT$ von $X$ mit $T\to Y$:
\[\xymatrix{
U \ar@{-->}[dr]_\theta\ar@/^1pc/[rrd]^u\ar@/_1pc/[ddr]_i &&\\
& X_T\ar[r]_-{p_1}\ar[d]^{f'} & X\ar[d]^f\\
& T\ar[r]_t & Y
} \]
Wir erhalten ein $\theta:U\to X_T$, so dass das obige Diagramm kommutiert. Sei $\xi_1\in X_T$ das Bild des einzigen Punktes $t_1\in U$ und $Z=\overline{\{\xi_1\}}\subset X_T$ mit der reduzierten Unterschemastruktur. Da $f$ universell abgeschlossen ist, ist $f'$ abgeschlossen, somit ist $f'(Z)\subset T$ abgeschlossen. Da $f'(\xi_1)=t_1$ und $t_1$ der generische Punkt von $T$ ist, gilt $f'(Z)=T$. Es existiert also ein $\xi_0\in Z$ mit $f'(\xi_0)=t_0$, wobei $t_0$ der abgeschlossene Punkt in $T$ ist.

Betrachte den zu $f'$ gehörigen lokalen Homomorphismus $R\hookrightarrow\mathcal{O}_{Z,\xi_0}$. Da $\xi_1$ der generische Punkt von $Z$ ist, gilt $\kappa(\xi_1)=\mathcal{O}_{Z,\xi_1}$ und nach \hyperref[4.14]{Lemma 4.14 (i)} erhalten wir $\kappa(\xi_1)\subset K$, der durch $U\to Z$ induziert wird. Nach \hyperref[4.16]{Satz 4.16} ist $R$ als Bewertungsring maximal unter allen lokalen Ringen in $K$ bzgl. Dominanz. Ferner gilt $\mathcal{O}_{Z,\xi_0}\subset\mathcal{O}_{Z,\xi_1}=\kappa(\xi_1)\subset K$ und $\mathcal{O}_{Z,\xi_0}\subsetneq K$, da $\xi_0\neq\xi_1$. Wegen $f'(\xi_0)=t_0$ dominiert $\mathcal{O}_{Z,\xi_0}$ den Ring $R$, d.h. $R\cong\mathcal{O}_{Z,\xi_0}$, insbesondere dominiert $R$ über $\mathcal{O}_{Z,\xi_0}$. Nach \hyperref[4.14]{Lemma 4.14 (ii)} erhalten wir Morphismus $h':T\cong\operatorname{Spec}(\mathcal{O}_{Z,\xi_0})\to X_T,\ t_i\mapsto \xi_i$. Wir erhalten $h:T\stackrel{h'}{\to}X_T\stackrel{p_1}{\to}X$ mit $fh=fp_1h'=tf'h'=t$ und $hi=p_1h'i=p_1h'f'\theta = p_1\theta = u$.
\item Es gelte (ii). $f$ ist nach Voraussetzung von endlichem Typ und separiert nach \hyperref[4.8]{Theorem 4.8}. Sei also $Y'\to Y$ ein Morphismus und $f':X'=X\times_YY'\to Y'$ die Basiserweiterung von $f$. Wir zeigen, dass $f'$ abgeschlossen ist. Sei $Z\subset X'$ abgeschlossen mit der reduzierten Unterschemastruktur. Betrachte das kommutative Diagramm:
\[ \xymatrix{
Z\ar@{^{(}->}[r]\ar[dr] & X'\ar[r]\ar[d]^{f'} & X\ar[d]^f\\
& Y'\ar[r] & Y
} \]
Da $f$ von endlichem Typ ist, ist nach \hyperref[3.25]{Satz 3.25 (ii)} auch $f'$ von endlichem Typ. Es folgt, dass $f'|_Z:Z\to Y'$ von endlichem Typ ist und somit quasikompakt. $f'|_Z$ faktorisiert über $Z\to f'(Z)\hookrightarrow Y'$. Wir sehen, dass $f'(Z)\hookrightarrow Y'$ quasikompakt ist. Nach \hyperref[4.13]{Korollar 4.13} genügt es zu zeigen, dass $f'(Z)$ stabil unter Spezialisierungen ist.

Sei also $z_1\in Z$ und $y_1=f'(z_1),\ y_0\in\overline{\{y_1\}}$. Wir versehen $\overline{\{y_1\}}$ mit der reduzierten Unterschemastruktur. Sei $\mathcal{O}=\mathcal{O}_{\overline{\{y_1\}},y_0 }$. Dann ist $\operatorname{Quot}(\mathcal{O})=\kappa(y_1)\hookrightarrow \kappa(z_1)=:K$.  Sei $R$ ein Bewertungsring von $K$, der $\mathcal{O}$ dominiert. Nach \hyperref[4.14]{Lemma 4.14} haben wir Morphismen:
\[U=\operatorname{Spec}(K)\to Z,\ t_1\mapsto z_1 \]
\[T=\operatorname{Spec}(R)\to Y',\ t_i\mapsto y_i \]
Betrachte folgendes kommutative Diagramm:
\[\xymatrix{
U\ar[r] \ar[d]_i & Z\ar@{^{(}->}[r] \ar[dr] & X'\ar[r]\ar[d]^{f'} & X\ar[d]^f\\
T\ar[rr]\ar@{-->}[rrru]\ar@{-->}[rru] & & Y'\ar[r] & Y
} \]
Nach Voraussetzung gibt es ein Morphismus $h:T\to X$, der das Diagramm kommutativ macht. Aus der Universaleigenschaft des Faserprodukts $X'$ erhalten wir ein Morphismus $h':T\to X'$, der $h:T\to X$ liftet. Nach Voraussetzung ist $Z$ abgeschlossen und da der generische Punkt $t_1\in T$ auf $z_1\in X'$ abgebildet wird, faktorisiert $h':T\to X'$ über $h':T\to Z\to X'$. Setze $z_0=h'(t_0)\in Z$. Dann ist $f'(z_0)=y_0$, also $y_0\in f'(Z)$.\qed
\end{itemize}

\paragraph{Korollar 4.22.}\label{4.22} Alle vorkommenden Schemata seien noethersch. \begin{enumerate}[(i)]
\item Abgeschlossene Immersionen sind eigentlich.
\item Kompositum zweier eigentlicher Morphismen ist eigentlich.
\item Eigentliche Morphismen sind stabil unter Basiserweiterung.
\item Seien $f:X\to Y$ und $f':X'\to Y'$ eigentliche Morphismen von $S$-Schemata. Dann ist $f\times f':X\times_SX'\to Y\times_SY'$ eigentlich.
\item Seien $f:X\to Y,\ g:Y\to Z$ Morphismen. Ist $g\circ f$ eigentlich und $g$ separabel, so ist $f$ eigentlich.
\item Ein Morphismus $f:X\to Y$ ist genau dann eigentlich, wenn es eine offene Ü\-ber\-deckung $Y=\bigcup_\alpha Y_\alpha$ gibt, so dass $f^{-1}(Y_\alpha)\to Y_\alpha$ für alle $\alpha$ eigentlich ist.
\end{enumerate}

\paragraph{Beweis.} Wir zeigen nur (v). Da $X$ noethersch ist, ist $f$ quasikompakt. Nach \hyperref[4.21]{Lemma 4.21} ist $f$ von endlichem Typ. Nach \hyperref[4.17]{Satz 4.17} ist $f$ separiert. Sei nun $R$ ein Bewertungsring und seien Morphismen $U\to X,\ T\to Y$ gegeben mit kommutativem Diagramm:
\[\xymatrix{
U \ar[r]^u\ar[d]_i& X\ar[d]^f\\
T\ar[r]_t\ar[dr]_{t'} & Y\ar[d]^g\\
& Z
} \]
Wir wollen ein $h:T\to X$ konstruieren mit $fh=t$ und $hi=u$. Setze $t'=gt$. Da $gf$ eigentlich ist, gibt es genau ein $h:T\to X$ mit $gfh=t'$ und $hi=u$. Sei $t''=fh$. Nun ist $g$ separabel und $ti=fu=fhi=t''i$ und $gt=t'=gfh=gt''$. Nach \hyperref[4.8]{Theorem 4.8} folgt $t=t''$ und somit $fh=t$.\qed

\paragraph{} Ist $\varphi:A\to B$ ein Ringhomomorphismus und $\operatorname{Spec}(B)\to\operatorname{Spec}(A)$ der zugehörige Morphismus, so gilt:
\[\mathbf{P}_B^n\cong\mathbf{P}_A^n\times_{\operatorname{Spec}(A)}\operatorname{Spec}(B) \]
Insbesondere gilt $\mathbf{P}_A^n\cong\mathbf{P}_\mathbb{Z}^n\times_\mathbb{Z}A$.

\paragraph{Definition 4.23.}\label{4.23} \begin{enumerate}[(i)]
\item Sei $Y$ ein Schema. Dann heißt $\mathbf{P}_Y^n=\mathbf{P}_\mathbb{Z}^n\times_\mathbb{Z}Y$ der \textit{$n$-dimensionale projektive Raum über $Y$}\index{projektiver Raum}.
\item Ein Morphismus $f:X\to Y$ von Schemata heißt \textit{projektiv}, wenn er eine Faktorisierung $f:X\stackrel{i}{\hookrightarrow} \mathbf{P}_Y^n\stackrel{p_2}{\to} Y$ besitzt, wobei $i$ eine abgeschlossene Immersion ist.
\item Ein Morphismus $f:X\to Y$ von Schemata heißt \textit{quasiprojektiv}\index{quasiprojektiv}, falls er eine Faktorisierung der Form $f:X\stackrel{i}{\hookrightarrow} X'\stackrel{p}{\to}Y$ besitzt, wobei $i$ eine offene Immersion und $p$ projektiv ist.
\end{enumerate}

\paragraph{Theorem 4.24.}\label{4.24} Ein projektiver Morphismus von noetherschen Schemata ist eigentlich. Ein quasiprojektiver Morphismus von noetherschen Schemata ist von endlichem Typ und separiert.

\paragraph{Beweis.} Es reicht zu zeigen, dass $\mathbf{P}_\mathbb{Z}^n$ eigentlich über $\mathbb{Z}$ ist, da der Basiswechsel nach \hyperref[4.22]{Korollar 4.22 (iii)} eigentlich über $Y$ ist. Ist $f:X\hookrightarrow \mathbf{P}^n_Y \to Y$ projektiv, so ist er als Verkettung von eigentlichen Morphismen wieder eigentlich, siehe \hyperref[4.22]{Korollar 4.22}. Ist $f:X\hookrightarrow X'\to Y$ quasiprojektiv, so ist $f$ als Verkettung separierter Morphismen separiert, siehe \hyperref[4.17]{Korollar 4.17}. Da $X$ noethersch ist, ist $X\hookrightarrow X'$ eine quasikompakte offene Immersion, also nach \hyperref[4.21]{Lemma 4.21 (i)} vom endlichem Typ.

Sei also $X=\mathbf{P}_\mathbb{Z}^n$ und $X=\bigcup V_i$ eine offene affine Überdeckung mit $V_i=D_+(x_i)=\operatorname{Spec}\mathbb{Z}\big[\frac{x_0}{x_i},\ldots,\frac{x_n}{x_i} \big]$. Es ist also $X$ von endlichem Typ. Sei nun $R$ ein Bewertungsring mit Quotientenkörper $K=\operatorname{Quot}(R)$ und $U\to X,\ T\to\operatorname{Spec}(\mathbb{Z})$ Morphismen mit kommutativem Diagramm:
\[\xymatrix{
U\ar[r]^u\ar[d]_i & X\ar[d]\\
T\ar[r]_-t & \operatorname{Spec}(\mathbb{Z})
}\]
Wir zeigen nun per Induktion, dass genau ein Morphismus $h:T\to X$ existiert, der das obige Diagramm kommutativ ergänzt. Für $n=0$ ist $X=\operatorname{Spec}(\mathbb{Z})$ und die Aussage ist klar. Sei $n\geq 1$ und $\xi_1$ das Bild des einzigen Punktes aus $U$ in $X$. Ist $\xi_1\in X\setminus V_i$ für ein $i$, so folgt die Behauptung nach Induktionsvoraussetzung, da die Hyperebene $X\setminus V_i$ isomorph zu $\mathbf{P}_\mathbb{Z}^{n-1}$ ist.

Sei also $\xi_1\in\bigcap V_i$, d.h. alle Funktionen der Form $\frac{x_i}{x_j}$ sind invertierbar in $\mathcal{O}_{\xi_1}$. Der Morphismus $U\to X$ liefert Inklusion $\kappa(\xi_1)\hookrightarrow K$. Sei weiter $f_{ij}\in K^\times$ das Bild von $\frac{x_i}{x_j}$ unter $\mathcal{O}_{\xi_1}\to \kappa(\xi_1)\hookrightarrow K$. Dann folgt $f_{ik}=f_{ij}f_{jk}$ für alle $i,j,k$. Sei $v:K\to G$ die zu $R$ gehörige Bewertung und setze $g_i=v(f_{i0})$ für alle $i$. Sei $k$ derart, dass $g_k$ minimal unter $g_0,\ldots,g_n$ ist. Es gilt für alle $i$:
\[v(f_{ik})=v(f_{i0})-v(f_{k0})=g_i-g_k\geq 0 \]
Es folgt $f_{ik}\in R$ für alle $i$. Es gibt Homomorphismen:
\[\xymatrix{
\mathbb{Z}\big[\frac{x_0}{x_k},\ldots,\frac{x_n}{x_k}\big]\ar[rr]^-\varphi_-{\frac{x_i}{x_k}\mapsto f_{ik} } \ar[d]&& R\ar[d]\\
\mathcal{O}_{\xi_1}\ar[r] & \kappa(\xi_1)\ar[r] & K
}\]
Wir erhalten den zu $\varphi$ gehörige Morphismus $T\to V_k$ und somit $h:T\to V_k\hookrightarrow X$. Offensichtlich ist $T\to X\to\operatorname{Spec}(\mathbb{Z})$ der Morphismus $t$ unt $U\to T\to X$ der Morphismus $u$. Ferner ist $h$ eindeutig nach Konstruktion.\qed

\paragraph{Satz 4.25.}\label{4.25} Sei $k$ ein algebraisch abgeschlossener Körper. Das Bild des Funktors $t:\mathbf{Var}(k)\to \mathbf{Sch}(k)$ ist die Menge aller quasiprojektiven, integren Schemata über $k$. Insbesondere ist $t(V)$ integer, separiert und von endlichem Typ für jede Varietät $V$.

\paragraph{Beweis.} Ohne Beweis.

\paragraph{Satz 4.26.}\label{4.26} Sei $X$ ein Schema von endlichem Typ über einem Körper $k$. Dann ist die Menge der abgeschlossenen Punkte in $X$ dicht in $X$.

\iffalse
\paragraph{Beweis.} Es genügt die Aussage für affine Schemata zu zeigen. Sei $X=\operatorname{Spec}(A)$ mit einer endlich erzeugten $k$-Algebra $A$ und $\mathfrak{M}$ die Menge der abgeschlossenen Punkte von $\operatorname{Spec}(A)$. Sei $f\in A$ mit $D(f)\neq\varnothing$. Wir zeigen $\mathfrak{M}\cap D(f)\neq\varnothing$. Es ist $D(f)\cong\operatorname{Spec}(A_f)$. Es existiert ein abgeschlossener Punkt $y\in \operatorname{Spec}(A_f)$, d.h. ein Maximalideal in $A_f$. Nun ist $A_f=A\big[\frac{1}{f}\big]$ endlich erzeugt über $k$, also $\kappa(y)=A_f/y$ eine algebraische Erweiterung von $k$ nach dem Hilbertschen Nullstellensatz. Sei $x$ der zu $y$ gehörige Punkt in $D(f)$.
\fi

\paragraph{Definition 4.27.}\label{4.27} Eine \textit{(abstrakte) Varietät}\index{Varietät!abstrakt} ist ein integres, separiertes Schema $X$ von endlichem Typ über einem algebraisch abgeschlossenen Körper $k$. Ist $X$ über $k$ eigentlich, so heißt $X$ \textit{vollständig}\index{Varietät!eigentlich}. Quasiprojektive abstrakte Varietäten entsprechen den klassischen Varietäten.

\paragraph{Bemerkung 4.28.}\label{4.28}\begin{enumerate}[(i)]
\item Eine projektive, abstrakte Varietät ist vollständig nach \hyperref[4.24]{Theorem 4.24}.
\item Es gibt vollständige Varietäten, die nicht projektiv sind, d.h. die Klasse der abstrakten Varietäten ist größer als die Klasse der klassischen Varietäten.
\item Jede vollständige abstrakte Varietät der Dimension $1$, d.h. eine vollständige Kurve, ist projektiv.
\item Jede Varietät kann als offene Menge in eine vollständige Varietät eingebettet werden.
\end{enumerate}

\section{Modulgarben}

\paragraph{Definition 5.1.}\label{5.1} Sei $(X,\mathcal{O}_X)$ ein geringter Raum.
\begin{enumerate}[(i)]
\item Eine \textit{Garbe von $\mathcal{O}_X$"~Moduln} bzw. \textit{$\mathcal{O}_X$"~Modul}\index{$\mathcal{O}_X$-Modul} ist eine Garbe $\mathcal{F}$ auf $X$ derart, dass $\mathcal{F}(U)$ ein $\mathcal{O}_X(U)$"~Modul für alle $U\subset_\text{o}X$ ist und alle $\operatorname{res}:\mathcal{F}(U)\to\mathcal{F}(V),\ V\subset U$ verträglich mit den Modulstrukturen via $\operatorname{res}:\mathcal{O}_X(U)\to\mathcal{O}_X(V)$ ist.
\item Ein \textit{Morphismus}\index{Morphismus!$\mathcal{O}_X$-Moduln} $\mathcal{F}\to\mathcal{G}$ von $\mathcal{O}_X$-Moduln ist ein Garbenmorphismus, so dass $\mathcal{F}(U)\to\mathcal{G}(U)$ ein $\mathcal{O}_X(U)$"~Modulhomomorphismus für alle $U\subset_\text{o}X$ ist.
\item Eine Sequenz von $\mathcal{O}_X$"~Moduln heißt \textit{exakt}\index{Exaktheit}, wenn sie exakt als Garbensequenz abelscher Garben ist.
\item Sind $\mathcal{F},\mathcal{G}$ $\mathcal{O}_X$"~Moduln, so bezeichnet $\operatorname{Hom}_{\mathcal{O}_X}(\mathcal{F},\mathcal{G})$ die Gruppe der Morphismen von $\mathcal{F}$ nach $\mathcal{G}$.
\item Sind $\mathcal{F},\mathcal{G}$ $\mathcal{O}_X$"~Moduln, so heißt die Garbe $U\mapsto \operatorname{Hom}_{\mathcal{O}_X|_U}(\mathcal{F}|_U,\mathcal{G}|_U)$ die \textit{Hom-Garbe}\index{Hom-Garbe} und wird mit $\operatorname{\mathcal{H}om}(\mathcal{F},\mathcal{G})$ bezeichnet. Diese ist ein $\mathcal{O}_X$"~Modul.
\item Seien $\mathcal{F},\mathcal{G}$ $\mathcal{O}_X$"~Moduln. Dann heißt die zur Prägarbe $U\mapsto\mathcal{F}(U)\otimes_{\mathcal{O}_X(U)}\mathcal{G}(U)$ assoziierte Garbe das \textit{Tensorprodukt}\index{Tensorprodukt} von $\mathcal{F}$ und $\mathcal{G}$. Diese wird  mit $\mathcal{F}\otimes_{\mathcal{O}_X}\mathcal{G}$ bezeichnet.
\item Ein $\mathcal{O}_X$"~Modul $\mathcal{F}$ heißt \textit{frei}\index{$\mathcal{O}_X$-Modul!frei}, wenn $\mathcal{F}$ isomorph zu einer direkten Summe von Exemplaren von $\mathcal{O}_X$ ist.
\item Ein $\mathcal{O}_X$"~Modul $\mathcal{F}$ heißt \textit{lokal frei}\index{$\mathcal{O}_X$-Modul!lokal frei}, wenn $X$ durch offene Mengen $U$ überdeckt werden kann, so dass $\mathcal{F}|_U$ freier $\mathcal{O}_X|_U$"~Modul ist.

Der \textit{Rang} $r$\index{Rang} von $\mathcal{F}$ auf $U$ ist gerade die Anzahl der Kopien von $\mathcal{O}_X|_U$. Wir schreiben $\operatorname{rang}(\mathcal{F}|_U)=r$. Ist $X$ zusammenhängend, so ist dieser Rang überall gleich. 
\item Ein lokal freier $\mathcal{O}_X$"~Modul vom Rang $1$ heißt \textit{invertierbare Garbe}\index{Garbe!invertierbar}.
\item Eine \textit{Idealgarbe} auf $X$\index{Idealgarbe} ist ein $\mathcal{O}_X$"~Modul $\mathcal{J}$, der Untergarbe von $\mathcal{O}_X$ ist, d.h. $\mathcal{J}(U)$ ist ein Ideal von $\mathcal{O}_X(U)$ für $U\subset_\text{o}X$.
\item Sei $f:(X,\mathcal{O}_X)\to (Y,\mathcal{O}_Y)$ ein Morphismus von geringten Räumen und $\mathcal{F}$ ein $\mathcal{O}_X$"~Modul. Dann ist $f_\ast\mathcal{F}$ ein $f_\ast\mathcal{O}_X$"~Modul und somit auch ein $\mathcal{O}_Y$"~Modul unter dem Garbenmorphismus $f^\sharp:\mathcal{O}_Y\to f_\ast\mathcal{O}_X$. $f_\ast\mathcal{F}$ heißt \textit{direktes Bild} von $\mathcal{F}$ unter $f$.\index{direktes Bild}

Sei $\mathcal{G}$ ein $\mathcal{O}_Y$"~Modul. Dann ist $f^{-1}\mathcal{G}$ ein $f^{-1}\mathcal{O}_Y$"~Modul. Betrachte das Bild $\theta$ von $f^\sharp$ unter dem Adjunktionsisomorphismus $\operatorname{Hom}_Y(\mathcal{O}_Y,f_\ast\mathcal{O}_X)\to\operatorname{Hom}_X(f^{-1}\mathcal{O}_Y,\mathcal{O}_X)$. Durch $\theta$ wird $\mathcal{O}_X$ zu einem $f^{-1}\mathcal{O}_Y$"~Modul. Wir definieren: 
\[f^\ast\mathcal{G}= f^{-1}\mathcal{G}\otimes_{f^{-1}\mathcal{O}_Y}\mathcal{O}_X\]
Somit ist $f^\ast\mathcal{G}$ ein $\mathcal{O}_X$"~Modul, das \textit{Urbild}\index{Urbild} von $\mathcal{G}$ unter $f$.

Für jeden $\mathcal{O}_X$"~Modul $\mathcal{F}$ und $\mathcal{O}_Y$"~Modul $\mathcal{G}$ gilt:
\[\operatorname{Hom}_{\mathcal{O}_X}(f^\ast\mathcal{G},\mathcal{F})\cong\operatorname{Hom}_{\mathcal{O}_Y}(\mathcal{G},f_\ast\mathcal{F}) \]
Somit ist $f^\ast$ linksadjungiert zu $f_\ast$.
\end{enumerate}

\paragraph{Bemerkung.}\begin{enumerate}[(i)]
\item Kern, Bild und Kokern eines Morphismus von $\mathcal{O}_X$"~Moduln ist wieder ein $\mathcal{O}_X$"~Modul.
\item Sind $\mathcal{F},\mathcal{F}'$ $\mathcal{O}_X$"~Moduln und $\mathcal{F}'$ eine Untergarbe von $\mathcal{F}$, so ist $\mathcal{F}/\mathcal{F}'$ ein $\mathcal{O}_X$-Modul.
\item Direkte Summen, direkte Produkte und projektive Limiten von $\mathcal{O}_X$"~Moduln sind wieder $\mathcal{O}_X$"~Moduln.
\end{enumerate}

\paragraph{Definition.} Sei $A$ ein Ring und $M$ ein $A$-Modul. Wir definieren die \textit{zu $M$ assoziierte Garbe}\index{assoziierte Garbe!Modul} $\widetilde{M}$ auf $\operatorname{Spec}(A)$ wie folgt: Für $U\subset_\text{o}\operatorname{Spec}(A)$ setzen wir $\widetilde{M}(U)$ als die Menge aller Abbildungen $s:U\to\coprod_{\mathfrak{p}\in U}M_\mathfrak{p}$, so dass:
\begin{enumerate}[(i)]
\item Für alle $\mathfrak{p}\in U$ gilt $s(\mathfrak{p})\in M_\mathfrak{p}$.
\item Für alle $\mathfrak{p}\in U$ gibt es eine offene Umgebung $V$ von $\mathfrak{p}$ mit $V\subset U$ und Elemente $m\in M,\ f\in A$, so dass für alle $\mathfrak{q}\in V$ stets $f\not\in\mathfrak{q}$ und $s(\mathfrak{q})=\frac{m}{f}\in M_\mathfrak{q}$ gilt.
\end{enumerate}
Dann ist $\widetilde{M}$ mit den gewöhnlichen Restriktionsabbildungen eine Garbe.

\paragraph{Satz 5.2.}\label{5.2} Sei $A$ ein Ring und $M$ ein $A$-Modul mit der assoziierten Garbe $\widetilde{M}$ auf $X=\operatorname{Spec}(A)$. Dann gilt:\begin{enumerate}[(i)]
\item $\widetilde{M}$ ist ein $\mathcal{O}_X$"~Modul.
\item Für jedes $\mathfrak{p}\in X$ gilt für den Halm von $\widetilde{M}$ in $\mathfrak{p}$ stets $\widetilde{M}_\mathfrak{p}\cong M_\mathfrak{p}$.
\item Für alle $f\in A$ gibt es einen $A_f$-Modulisomorphismus $\widetilde{M}(D(f))\cong M_f$.
\item Insbesondere gilt $\Gamma(X,\widetilde{M})\cong M$.
\end{enumerate}

\paragraph{Beweis.} (i) ist klar. (iv) folgt aus (iii) mit $f=1$. (ii) und (iii) gehen analog zu \hyperref[2.3]{2.3}.\qed

\paragraph{Satz 5.3.}\label{5.3} Sei $A$ ein Ring und $X=\operatorname{Spec}(A)$. Ferner sei $\varphi:A\to B$ ein Ringhomomorphismus und $f:\operatorname{Spec}(B)\to \operatorname{Spec}(A)$ der entsprechende Morphismus. Dann gilt:
\begin{enumerate}[(i)]
\item Die Abbildung $M\mapsto\widetilde{M}$ liefert einen exakten, volltreuen Funktor von der Kategorie der $A$"~Moduln in die Kategorie der $\mathcal{O}_X$"~Moduln.
\item Seien $M,N$ $A$"~Moduln. Dann gilt $\widetilde{M\otimes_AN}\cong \widetilde{M}\otimes_{\mathcal{O}_X}\widetilde{N}$.
\item Sei $(M_i)_i$ eine Familie von $A$-Moduln. Dann gilt $\widetilde{\bigoplus_i M_i}\cong\bigoplus_i\widetilde{M_i}$.
\item Sei $N$ ein $B$-Modul. Dann gilt $f_\ast\widetilde{N}\cong \widetilde{_AN}$, wobei $_AN$ den Modul $N$ als $A$"~Modul via $\varphi$ bezeichnet.
\item Sei $M$ ein $A$"~Modul. Dann gilt $f^\ast\widetilde{M}\cong \widetilde{M\otimes_A B}$.
\end{enumerate}

\paragraph{Beweis.} Für (i) zeigen wir: \begin{itemize}
\item \textit{Funktorialität:} Sei $\psi:M\to N$ ein $A$"~Modulhomomorphismus. Dieser induziert für alle $f\in A$ einen $A_f$"~Modulhomomorphismus $\psi_f:M_f\to N_f$. Ist $D(f)\supset D(g)$, so erhalten wir das folgende kommutative Diagramm:
\[\xymatrix{
M_f\ar[rr]^{\psi_f}\ar[d] && N_f\ar[d]\\
M_g\ar[rr]_{\psi_g} && N_g
} \]
Da $D(f),\ f\in A$ eine Basis der Topologie bilden, induzieren $\psi_f,\ f\in A$ einen $\widetilde{A}$"~Homomorphismus $\widetilde{\psi}:\widetilde{M}\to\widetilde{N}$ mit $\widetilde{\psi}|_{D(f)}=\psi_f$.
\item \textit{Volltreu:} Die Umkehrabbildung zu $\operatorname{Hom}_A(M,N)\to\operatorname{Hom}_{\widetilde{A}}(\widetilde{M},\widetilde{N})$ ist gegeben durch das Bilden der globalen Schnitte: 
\[\operatorname{Hom}_{\widetilde{A}}(\widetilde{M},\widetilde{N})\to\operatorname{Hom}_A(\Gamma(X,\widetilde{M}),\Gamma(X,\widetilde{N}))=\operatorname{Hom}_A(M,N)\]
\item \textit{Exaktheit:} Sei $0\to M'\to M\to M''\to 0$ eine kurze exakte Folge von $A$"~Moduln. Da Lokalisierungen exakt sind, ist $0\to M'_\mathfrak{p}\to M_\mathfrak{p}\to M''_\mathfrak{p}\to 0$ exakt. Nach \hyperref[5.2]{Satz 5.2 (ii)} ist $\widetilde{M}_\mathfrak{p}=M_\mathfrak{p}$. Da jede Halmsequenz exakt ist, folgt nach \hyperref[1.10]{1.10} die Exaktheit von $0\to\widetilde{M'}\to\widetilde{M}\to\widetilde{M''}\to 0$.
\end{itemize}
(ii) und (iii) folgen, da direkte Summen und Tensorprodukte mit Lokalisierungen kommutieren. Für (iv) sei $g\in A$. Es gilt:
\[\Gamma(D(g),f_\ast\widetilde{N}) = \Gamma(f^{-1}(D(g)),\widetilde{N}) = \Gamma(D(\varphi(g)),\widetilde{N}) \cong N_{\varphi(g)} = N_g\cong \Gamma(D(g),\widetilde{N})\]
Für (v) sei $N$ ein $B$-Modul. Dann gilt:
\begin{align*}
\operatorname{Hom}_{\widetilde{B}}(f^\ast\widetilde{M},\widetilde{N})&=\operatorname{Hom}_{\widetilde{A}}(\widetilde{M},f_\ast \widetilde{N}) \stackrel{\text{(iv)}}{=}\operatorname{Hom}_{\widetilde{A}}(\widetilde{M}, \widetilde{_AN})\\
&= \operatorname{Hom}_A(M, {_AN}) \stackrel{\star}{\cong}\operatorname{Hom}_B(M\otimes_A B,N)\cong\operatorname{Hom}_{\widetilde{B}}(\widetilde{M\otimes_A B},\widetilde{N})
\end{align*}
wobei $\star$ durch die Abbildung $\eta\mapsto (m\otimes b\mapsto \eta(m)b)$ gegeben ist.\qed

\paragraph{Definition 5.4.}\label{5.4} \begin{enumerate}[(i)]
\item Sei $(X,\mathcal{O}_X)$ ein Schema. Ein $\mathcal{O}_X$"~Modul $\mathcal{F}$ heißt \textit{quasikohärent}\index{$\mathcal{O}_X$-Modul!quasikohärent}, falls es eine offene affine Überdeckung $U_i=\operatorname{Spec}(A_i),\ i\in I$ von $X$ gibt, so dass für jedes $i$ ein $A_i$"~Modul $M_i$ existiert mit $\mathcal{F}|_{U_i}\cong\widetilde{M_i}$.
\item $\mathcal{F}$ heißt \textit{kohärent}\index{kohärent}, falls $\mathcal{F}$ quasikohärent ist und alle vorkommenden $M_i$ in (i) endlich erzeugte $A_i$"~Moduln sind.
\end{enumerate}

\paragraph{Beispiel 5.5.}\label{5.5} Für jedes Schema $X$ ist $\mathcal{O}_X$ kohärent, da $\mathcal{O}_X|_{\operatorname{Spec}(A)}=\widetilde{A}$.

\paragraph{Beispiel 5.6.}\label{5.6} Sei $X=\operatorname{Spec}(A)$ affin und $Y\subset X$ ein abgeschlossenes Unterschema, das durch das Ideal $\mathfrak{a}\subset A$ definiert ist. Sei $i:Y\hookrightarrow X$ die natürliche Inklusion. Es ist $\mathcal{O}_Y\cong\widetilde{A/\mathfrak{a}}$ und somit $i_\ast\mathcal{O}_Y=\widetilde{A/\mathfrak{a}}$, wobei hier $A/\mathfrak{a}$ als $A$"~Modul aufgefasst wird. Somit ist $i_\ast\mathcal{O}_Y$ ein kohärenter $\mathcal{O}_X$"~Modul.

\paragraph{Beispiel 5.7.}\label{5.7} Sei $X=\operatorname{Spec}(A)$ affin und $U\subsetneq_\text{o}X$ mit der natürlichen Inklusion $j:U\hookrightarrow X$. Betrachte die Garbe $j_!\mathcal{O}_U$, die außerhalb $U$ durch Null fortgesetzte Garbe von $\mathcal{O}_U$. $j_!\mathcal{O}_U$ ist nicht quasikohärent:

Sei $X$ irreduzibel und $V=\operatorname{Spec}(A)\subset_\text{o}X$ mit $V\subsetneq U$. Wäre $j_!\mathcal{O}_U|_V\cong\widetilde{M}$ für einen $A$"~Modul $M$., so ist $(j_!\mathcal{O}_U|_V)(V)=M$, aber $(j_!\mathcal{O}_U|_V)(V)=0$ und $j_!\mathcal{O}_U|_V\neq 0$.

\paragraph{Lemma 5.8.}\label{5.8} Sei $X=\operatorname{Spec}(A)$ ein affines Schema, $f\in A$ und $\mathcal{F}$ ein quasikohärenter $\mathcal{O}_X$"~Modul.
\begin{enumerate}[(i)]
\item Sei $s\in\Gamma(X,\mathcal{F})$ mit $s|_{D(f)}=0$. Dann existiert ein $n>0$ mit $f^ns=0$.
\item Sei $t\in\Gamma(D(f),\mathcal{F})$. Dann existiert ein $n>0$ und $t'\in\mathcal{F}(X)$ mit $f^nt=t'|_{D(f)}$.
\end{enumerate}

\paragraph{Beweis.} Wir zeigen zunächst, dass es eine Überdeckung der Form $X=\bigcup_{i=1}^m D(g_i)$ gibt, so dass $\mathcal{F}|_{D(g_i)}\cong\widetilde{M_i}$ für einen $A_{g_i}$"~Modul $M_i$.

Da $\mathcal{F}$ quasikohärent ist, existiert eine offene affine Überdeckung aus Mengen der Form $V=\operatorname{Spec}(B)$ mit $\mathcal{F}|_V=\widetilde{M}$ für einen $B$"~Modul $M$. Wir schreiben $V=\bigcup_{\text{gewisse }g\in A}D(g)$. Die natürlichen Morphismen $D(g)\hookrightarrow V$ liefern Ringhomomorphismen $B\to A_g$. Nach \hyperref[5.3]{Satz 5.3 (v)} ist $\mathcal{F}|_{D(g)}\cong \widetilde{M\otimes_B A_g}$. Da $X$ affin und somit quasikompakt ist, kann $X$ durch solche Mengen endlich überdeckt werden.
\begin{enumerate}[(i)]
\item Setze $s_i$ als das Bild von $s|_{D(g_i)}$ unter $\Gamma(D(g_i),\mathcal{F})=\Gamma(D(g_i),\widetilde{M_i})\cong M_i$. Wegen $D(fg_i)=D(f)\cap D(g_i)$ folgt nach \hyperref[5.2]{Satz 5.2 (iii)} $\Gamma(D(fg_i),\mathcal{F})=(M_i)_f$. Also ist $s_i=0$ in $(M_i)_f$. Nach Definition gibt es ein $n_i\in\mathbb{N}$ mit $f^{n_i}s_i=0$ in $M_i$. Sei $n$ das Maximum aller $n_i,\ i=1,\ldots,m$. Dann folgt $f^ns=0$ aus der ersten Garbeneigenschaft.
\item Betrachte die Einschränkungen $t\in\Gamma(D(fg_i),\mathcal{F})=(M_i)_f$. Für alle $i$ gibt es ein $n_i\in\mathbb{N}$, so dass $f^{n_i}t=t_i|_{D(fg_i)}$ für ein $t_i\in \Gamma(D(g_i),\mathcal{F})=M_i$. Sei $n$ das Maximum aller $n_i,\ i=1,\ldots,m$. Dann gibt es für alle $i$ ein $t_i\in\Gamma(D(g_i),\mathcal{F})$ mit $f^n t = t_i|_{D(fg_i)}$.

Auf $D(g_i)\cap D(g_j)=D(g_ig_j)$ haben wir Schnitte $t_i,t_j$ konstruiert, die auf $D(fg_ig_j)$ übereinstimmen. Nach (i) gibt es ein $m_{ij}$, so dass $f^{m_{ij}}(t_i-t_j)=0$ auf $D(g_ig_j)$ gilt. Sei $m$ das Maximum aller $m_{ij}$, so dass $f^m(t_i-t_j)=0$ auf $D(g_ig_j)$ für alle $i,j$ gilt. Die lokalen Schnitte $f^mt_i$ in $\Gamma(D(g_i),\mathcal{F})$ verkleben sich somit zu einem globalen Schnitt $t'$ von $\mathcal{F}$ zusammen mit $t'|_{D(f)}=f^{n+m}t$.\qed
\end{enumerate}

\paragraph{Satz 5.9.}\label{5.9} Sei $X$ ein Schema und $\mathcal{F}$ ein $\mathcal{O}_X$"~Modul. Dann ist $\mathcal{F}$ genau dann quasikohärent, wenn für alle affinen $U=\operatorname{Spec}(A)\subset_\text{o}X$ ein $A$"~Modul $M$ existiert mit $\mathcal{F}|_U\cong\widetilde{M}$.

Ist $X$ noethersch, so ist $\mathcal{F}$ genau dann kohärent, wenn für alle affinen $U=\operatorname{Spec}(A)\subset_\text{o}X$ ein endlich erzeugter $A$"~Modul $M$ existiert mit $\mathcal{F}|_U\cong\widetilde{M}$.

\paragraph{Lemma.} Sei $X=\operatorname{Spec}(A)$ affin, $M$ ein $A$"~Modul und $\mathcal{F}$ ein $\mathcal{O}_X$"~Modul. Dann ist
\[\operatorname{Hom}_A(M,\Gamma(X,\mathcal{F}))\to\operatorname{Hom}_{\widetilde{A}}(\widetilde{M},\mathcal{F}),\ \varphi\mapsto\widetilde{\varphi} \]
ein Isomorphismus mit Umkehrabbildung $\theta\mapsto\Gamma(X,\theta)$.

\paragraph{Beweis von \hyperref[5.9]{Satz 5.9}.} Die Rückrichtungen der beiden Aussagen sind trivial. Sei $U\subset_\text{o}X$ affin. Nach dem Beweis von \hyperref[5.8]{Lemma 5.8} gibt es eine Basis der Topologie von $U$, bestehend aus affinen Teilmengen $V_i$ derart, dass $\mathcal{F}|_{V_i}\cong\widetilde{N_i}$ mit einem Modul $N_i$ ist. Somit ist $\mathcal{F}|_U$ quasikohärent. Wir können somit o.B.d.A. $X=U=\operatorname{Spec}(A)$ als affin annehmen.

Setze $M=\Gamma(X,\mathcal{F})$ und $\alpha:\widetilde{M}\to\mathcal{F}$ als das Bild von $\operatorname{id}_M$ unter der Abbildung im vorherigen Lemma. Wie im Beweis von \hyperref[5.8]{Lemma 5.8} gezeigt, gibt es eine Überdeckung der Form $X=\bigcup_{i=1}^mD(g_i)$ mit $\mathcal{F}|_{D(g_i)}\cong\widetilde{M_i}$ für einen $A_{g_i}$"~Modul $M_i$. Es gilt $M_i=\mathcal{F}(D(g_i))\cong M_{g_i}$ und wir haben ein kommutatives Diagramm:
\[\xymatrix{
\mathcal{F}(X)_{g_i}\ar[rr]^\cong && \mathcal{F}(D(g_i))\\
\mathcal{F}(X)\ar[u]\ar[urr]_{\operatorname{res}}
} \]
Somit ist $\alpha|_{D(g_i)}$ ein Isomorphismus für alle $i$. Da die $D(g_i)$ ganz $X$ überdecken, ist $\alpha$ ein Isomorphismus.

Sei nun $X$ zusätzlich noethersch. Dann sind die $A_{g_i}$"~Moduln $M_{g_i}$ endlich erzeugt. Wir zeigen, dass $M$ endlich erzeugt ist. Da $A$ noethersch ist, sind alle $A_{g_i}$ noethersch. Also sind die endlich erzeugten $A_{g_i}$"~Moduln $M_{g_i}$ noethersch. Analog zu \hyperref[3.6]{Satz 3.6} folgt $M$ noethersch. Insbesondere ist $M$ endlich erzeugt. \qed

\paragraph{Korollar 5.10.}\label{5.10} Sei $A$ ein Ring und $X=\operatorname{Spec}(A)$. Dann ist
\[\{\text{Kategorie der $A$-Moduln}\}\to\{\text{Kategorie der quasikohärenten $\mathcal{O}_X$-Moduln}\},\ M\mapsto\widetilde{M} \]
ist eine Kategorienäquivalenz mit der Umkehrabbildung $\mathcal{F}\mapsto\Gamma(X,\mathcal{F})$. Ist $A$ noethersch, so geben dieselben Funktoren eine Kategorienäquivalenz zwischen den endlich erzeugten $A$-Moduln und den kohärenten $\mathcal{O}_X$"~Moduln.

\paragraph{Beweis.} Sei $\mathcal{F}$ ein quasikohärenter $\mathcal{O}_X$"~Modul. Nach \hyperref[5.9]{Satz 5.9} existiert ein $A$"~Modul $M$ mit $\mathcal{F}\cong\widetilde{M}$. Nach \hyperref[5.2]{Satz 5.2 (iv)} ist $\Gamma(X,\mathcal{F})=M$, also:
\[M\mapsto\widetilde{M}\mapsto \Gamma(X,\widetilde{M})=M,\quad\mathcal{F}=\widetilde{M}\mapsto \Gamma(X,\widetilde{M})=M\mapsto\widetilde{M}\qedhere \]

\paragraph{Satz 5.11.}\label{5.11} Sei $X=\operatorname{Spec}(A)$ ein affines Schema und sei $0\to\mathcal{F}'\to\mathcal{F}\to\mathcal{F}''\to 0$ eine exakte Sequenz von $\mathcal{O}_X$"~Moduln, wobei $\mathcal{F}'$ quasikohärent ist. Dann ist die Sequenz der globalen Schnitte ebenfalls exakt:
\[0\to\Gamma(X,\mathcal{F}')\to\Gamma(X,\mathcal{F})\to\Gamma(X,\mathcal{F}'')\to 0 \]

\paragraph{Beweis.} Da $\Gamma(X,-)$ linksexakt ist, bleibt nur die Surjektivität von $\Gamma(X,\mathcal{F})\to\Gamma(X,\mathcal{F}'')$ zu zeigen. Sei $s\in\Gamma(X,\mathcal{F}'')$ gegeben. Für $x\in X$ gibt es ein $f\in A$ mit $x\in D(f) \subset X$, so dass $s|_{D(f)}$ sich zu einem $t\in\mathcal{F}(D(f))$ liftet. Wir zeigen nun, dass es ein $r>0$ existiert, so dass sich $f^rs$ zu einem $t''\in\mathcal{F}(X)$ liftet.

Sei $X=\bigcup_i D(g_i)$ eine endliche offene Überdeckung, so dass sich $s|_{D(g_i)}$ zu einem $t_i\in\mathcal{F}(D(g_i))$ liftet. Auf $D(f)\cap D(g_i)=D(fg_i)$ liften $t_i,t\in\mathcal{F}(D(fg_i))$ beide $s$. Aus der Linksexaktheit von $\Gamma(D(fg_i), -)$ folgt $t-t_i\in\mathcal{F}'(D(fg_i))$. Da $\mathcal{F}'$ quasikohärent ist, folgt aus \hyperref[5.8]{Lemma 5.8 (ii)} die Existenz eines $n>0$ und $u_i\in\mathcal{F}'(D(g_i))$ mit $u_i|_{D(fg_i)}=f^n(t-t_i)$. Wir können $n$ unabhängig von $i$ wählen. Setze $t_i'=f^nt_i+u_i\in\mathcal{F}(D(g_i))$. Dann ist $t'_i$ ein Lift von $f^ns|_{D(g_i)}$ und $\star\ t_i'=f^nt$ auf $\mathcal{F}(D(fg_i))$. Auf $D(g_ig_j)$ liften $t_i'$ und $t_j'$ beide $f^ns$, also $t_i'-t_j'\in\mathcal{F}'(D(g_ig_j))$ und daher $t'_i=t'_j$ auf $\mathcal{F}(D(fg_ig_j))$ wegen $\star$. Nach \hyperref[5.8]{Lemma 5.8 (i)} existiert ein $m>0$, so dass $f^m(t_i'-t_j')=0$ auf $\mathcal{F}(D(g_ig_j))$. Wir können $m$ unabhängig von $i,j$ wählen. Somit verkleben sich die $f^mt_i'$ zu einem $t''\in\mathcal{F}(X)$ zusammen und $t''$ ist ein Lift von $f^{n+m}s$.

Sei nun $X=\bigcup_i D(f_i)$ eine endliche offene Überdeckung, so dass sich $s|_{D(f_i)}$  zu einem Schnitt aus $\mathcal{F}(D(f_i))$ liften lässt. Für alle $i$ existiert ein $n$ und ein Lift $t_i\in\Gamma(X,\mathcal{F})$ von $f_i^ns$. Wir können o.B.d.A. $n$ unabhängig von $i$ annehmen. Da $X=\bigcup_i D(f_i)$, gilt $(f_1^n,\ldots,f_k^n)=A$, also $1=\sum_{i=1}^k a_if_i^n$ für gewisse $a_i\in A$. Setze $t=\sum_{i=1}^k a_it_i\in\mathcal{F}(X)$. Dann ist $t$ ein Lift von $\sum_{i=1}^k a_if_i^ns=s\in\mathcal{F}''(X)$.\qed

\paragraph{Satz 5.12.}\label{5.12} Sei $X$ ein Schema.
\begin{enumerate}[(i)]
\item Kern, Kokern und Bild eines Morphismus von quasikohärenten Garben ist wieder quasikohärent.
\item Ist $0\to\mathcal{F}'\to\mathcal{F}\to\mathcal{F}''\to 0$ eine kurze exakte Sequenz von $\mathcal{O}_X$"~Moduln und $\mathcal{F}',\mathcal{F}''$ quasikohärent, so ist $\mathcal{F}$ quasikohärent.
\item Ist $X$ noethersch, so gilt (ii) auch für kohärente Garben.
\end{enumerate}

\paragraph{Beweis.} Da Quasikohärenz bzw. Kohärenz eine lokale Eigenschaft ist, können wir ohne Einschränkung $X=\operatorname{Spec}(A)$ als affin annehmen. Nach \hyperref[5.10]{Korollar 5.10} gelten (i) und (ii) für Modulgarben der Form $\widetilde{M}$. Da $M\mapsto \widetilde{M}$ nach \hyperref[5.3]{5.3 (i)} ein exakter, volltreuer Funktor ist, folgt die Aussage über Kern, Kokern und Bild.

Sei nun $0\to\mathcal{F}'\to\mathcal{F}\to\mathcal{F}''\to 0$ eine kurze exakte Sequenz von $\mathcal{O}_X$"~Moduln mit $\mathcal{F}',\mathcal{F}''$ quasikohärent. Nach \hyperref[5.11]{Satz 5.11} ist die folgende Folge exakt:
\[0\to\Gamma(X,\mathcal{F}')\to\Gamma(X,\mathcal{F})\to\Gamma(X,\mathcal{F}'')\to 0 \]
Da $M\mapsto\widetilde{M}$ ein exakter Funktor ist, folgt die Exaktheit von:
\[\xymatrix{
0  \ar[r]& \widetilde{\Gamma(X,\mathcal{F}')} \ar[d] \ar[r]& \widetilde{\Gamma(X,\mathcal{F})} \ar[d] \ar[r]& \widetilde{\Gamma(X,\mathcal{F}'')} \ar[d] \ar[r]& 0\\
0 \ar[r] & \mathcal{F}'  \ar[r]& \mathcal{F}  \ar[r]&  \ar[r]\mathcal{F}''  \ar[r]& 0
} \]
Die vertikalen Pfeile links und rechts sind nach \hyperref[5.10]{Korollar 5.10} Isomorphismen. Nach dem 5er Lemma ist somit auch der mittlere Pfeil ein Isomorphismus und $\mathcal{F}$ ist quasikohärent.

Sei nun $X$ noethersch und $\mathcal{F}',\mathcal{F}''$ kohärent. Dann sind $\Gamma(X,\mathcal{F}'),\Gamma(X,\mathcal{F}'')$ endlich erzeugt und somit auch $\Gamma(X,\mathcal{F})$. Es folgt die Kohärenz von $\mathcal{F}$.\qed

\paragraph{Satz 5.13.}\label{5.13} Sei $f:X\to Y$ ein Morphismus von Schemata.
\begin{enumerate}[(i)]
\item Sei $\mathcal{G}$ ein quasikohärenter $\mathcal{O}_Y$"~Modul. Dann ist $f^\ast\mathcal{G}$ quasikohärenter $\mathcal{O}_X$"~Modul.
\item Seien $X,Y$ noethersch und $\mathcal{G}$ kohärenter $\mathcal{O}_Y$"~Modul. Dann ist $f^\ast\mathcal{G}$ kohärent.
\item Sei $X$ noethersch oder $f$ quasikompakt und separiert. Ist $\mathcal{F}$ ein quasikohärenter $\mathcal{O}_X$"~Modul, so ist $f_\ast\mathcal{F}$ quasikohärenter $\mathcal{O}_Y$"~Modul.
\end{enumerate}

\paragraph{Lemma 5.14.}\label{5.14} Sei $Y$ ein affines Schema, $f:X\to Y$ ein separierter Morphismus und $U,V\subset_\text{o}X$ affin. Dann ist $U\cap V$ ein abgeschlossenes Unterschema eines affinen Schemas. Wir werden in \hyperref[5.18]{Korollar 5.18} sehen, dass $U\cap V$ sogar affin ist. Insbesondere ist $U\cap V$ quasikompakt.

\paragraph{Beweis.} Betrachte das kartesische Diagramm:
\[\xymatrix{
X\times_Y X\ar[r]^-{p_2}\ar[d]_{p_1} & X\ar[d]^f\\
X\ar[r]_f & Y
} \]
Nach \hyperref[3.20]{3.20} ist $p_1^{-1}(U)\cap p_2^{-1}(V)=U\times_YV$ affin. Es ist $U\cap V\cong \Delta_{X/Y}(X)\cap U\times_YV\subset U\times_YV$. Nun ist $\Delta_{X/Y}(X)$ abgeschlossen in $X\times_YX$, d.h. $U\cap V$ ist ein abgeschlossenes Unterschema in $U\times_YV$.\qed

\paragraph{Beweis von \hyperref[5.13]{Satz 5.13}.} Für (i) und (ii) ist die Aussage lokal in $X$ und in $Y$. Daher können wir o.B.d.A. $X=\operatorname{Spec}(B)$ und $Y=\operatorname{Spec}(A)$ als affin annehmen. Dann folgt die Behauptung aus der Kategorienäquivalenz \hyperref[5.10]{Korollar 5.10} und \hyperref[5.3]{Satz 5.3 (v)} $f^\ast\widetilde{M} = \widetilde{M\otimes_AB}$.

Für (iii) können wir nur $Y$ als affin annehmen. Sei $X=\bigcup_i U_i$ eine endliche, offene affine Überdeckung, da in beiden Fällen $X$ quasikompakt ist. Setze $U_{ij}=U_i\cap U_j$. In beiden Fällen sind $U_{ij}$ quasikompakt, siehe \hyperref[5.14]{Lemma 5.14} und \hyperref[3.5]{3.5}, \hyperref[3.4]{3.4}. Sei also $U_{ij}=\bigcup_k U_{ijk}$ eine endliche, offene affine Überdeckung. Nach der Garbeneigenschaft haben wir die folgende exakte Sequenz:
\[0\to f_\ast\mathcal{F} \to\bigoplus_if_\ast(\mathcal{F}|_{U_i}) \to\bigoplus_{i,j,k}f_\ast(\mathcal{F}|_{U_{ijk}}) \]
Wegen \hyperref[5.3]{Satz 5.3 (iv)} sind $f_\ast(\mathcal{F}|_{U_i})$ und $f_\ast(\mathcal{F}|_{U_{ijk}})$ quasikohärent. Nach \hyperref[5.12]{Satz 5.12 (ii)} sind $\bigoplus_i f_\ast(\mathcal{F}|_{U_i})$ und $\bigoplus_{i,j,k}f_\ast(\mathcal{F}|_{U_{ijk}})$ quasikohärent. Nach \hyperref[5.12]{5.12 (i)} ist daher auch $f_\ast\mathcal{F}$ quasikohärent.\qed

\paragraph{Bemerkung.} Sind $X,Y$ noethersch und $\mathcal{F}$  kohärent, so muss $f_\ast\mathcal{F}$ nicht notwendigerweise kohärent sein.

\paragraph{Definition 5.16.}\label{5.16} Sei $Y$ ein abgeschlossenes Unterschema von $X$ und $i:Y\hookrightarrow X$ der Inklusionsmorphismus. Dann ist die \textit{zu $Y$ gehörige Idealgarbe} auf $X$.\index{Idealgarbe} wie folgt definiert:
\[\mathcal{J}_Y=\ker(i^\sharp:\mathcal{O}_X\to i_\ast\mathcal{O}_Y) \]

\paragraph{Satz 5.17.}\label{5.17} Sei $X$ ein Schema. Dann gilt:
\begin{enumerate}[(i)]
\item Ist $Y$ ein abgeschlossenes Unterschema von $X$, so ist $\mathcal{J}_Y$ eine quasikohärente Idealgarbe auf $X$.
\item Ist $X$ zusätzlich noethersch, so ist $\mathcal{J}_Y$ kohärent.
\item Jede quasikohärente Idealgarbe auf $X$ bestimmt in eindeutiger Weise ein abgeschlossenes Unterschema.
\end{enumerate}

\paragraph{Beweis.}\begin{enumerate}[(i)]
\item Sei $Y\subset X$ abgeschlossen. Dann ist $i:Y\hookrightarrow X$ ein quasikompakter Morphismus. Wegen \hyperref[4.17]{Satz 4.17 (i)} ist $i$ separiert. Nach \hyperref[5.13]{Satz 5.13 (iii)} ist $i_\ast\mathcal{O}_Y$ quasikohärent, also $\mathcal{J}_Y$ quasikohärent nach \hyperref[5.12]{Satz 5.12 (i)}.
\item Ist $X$ noethersch und $U\subset_\text{o}X$ affin mit $U=\operatorname{Spec}(A)$, so ist auch $A$ noethersch nach \hyperref[3.6]{Satz 3.6}. Daher ist $I=\Gamma(U,\mathcal{J}_Y|_U)$ ein endlich erzeugtes Ideal in $A$. Nach \hyperref[5.9]{Satz 5.9} ist $\mathcal{J}_Y$ kohärent.
\item Sei $\mathcal{J}$ eine quasikohärente Idealgarbe auf $X$. Setze:
\[Y=\operatorname{Supp}(\mathcal{O}_X/\mathcal{J})=\{x\in X\mid (\mathcal{O}_X/\mathcal{J})_x\neq 0 \}\subset X \]
Wir zeigen, dass $(Y,\mathcal{O}_X/\mathcal{J})$ ein abgeschlossenes Unterschema von $X$ ist. Dies ist eine lokale Frage, sei o.B.d.A. $X=\operatorname{Spec}(A)$ affin. Da $\mathcal{J}$ quasikohärent ist, folgt $\mathcal{J}=\widetilde{\mathfrak{a}}$ für ein Ideal $\mathfrak{a}\subset A$. Es gilt:
\begin{align*}
Y&=\operatorname{Supp}(\mathcal{O}_X/\mathcal{J})=\operatorname{Supp}(\widetilde{A/\mathfrak{a}})\\
&= \{\mathfrak{p}\in X\mid (A/\mathfrak{a})_\mathfrak{p}\neq 0\}\\
&=\{\mathfrak{p}\in X\mid\mathfrak{a}\subset\mathfrak{p}\}=\operatorname{Spec}(A/\mathfrak{a})
\end{align*}
Die Eindeutigkeit ist klar.\qed
\end{enumerate}

\paragraph{Korollar 5.18.}\label{5.18} Sei $X=\operatorname{Spec}(A)$ affin. Dann haben wir eine Bijektion:
\[\{\mathfrak{a}\mid \mathfrak{a}\subset A \text{ Ideal}\}\to \{Y\mid Y\subset X \text{ abgeschlossenes Unterschema}\},\ \mathfrak{a}\mapsto \operatorname{Spec}(A/\mathfrak{a}) \]
Insbesondere ist jedes abgeschlossenes Unterschema eines affinen Schemas wieder affin.

\paragraph{Definition 5.19.}\label{5.19} Sei $S=\bigoplus_d S_d$ ein graduierter Ring. Ein $S$"~Modul heißt \textit{graduierter $S$"~Modul}\index{Modul!graduiert}, falls $M=\bigoplus_d M_d$ mit $S_d\cdot M_e\subset S_{d+e}$ gilt. Sei $\ell\in\mathbb{Z}$. Dann definieren wir den \textit{getwisteten $S$"~Modul}\index{Modul!getwistet} $M(\ell)$ von $M$ durch:
\[M(\ell)_d = M_{d+\ell} \]

\paragraph{Definition 5.20.}\label{5.20} Sei $S$ ein graduierter Ring und $M$ ein graduierter $S$"~Modul. Die zu $M$ assoziierte Garbe $\widetilde{M}$ auf $\operatorname{Proj}(S)$ ist wie folgt definiert: Sei $U\subset_\text{o}\operatorname{Proj}(S)$ und setze $\widetilde{M}(U)$ als die Menge aller Abbildungen $s:U\to\coprod_{\mathfrak{p}\in U}M_{(\mathfrak{p})}$, so dass:
\begin{enumerate}[(i)]
\item Für alle $\mathfrak{p}\in U$ gilt $s(\mathfrak{p})\in M_{(\mathfrak{p})}$.
\item Für alle $\mathfrak{p}\in U$ existiert eine offene Umgebung $V$ von $\mathfrak{p}$ mit $V\subset U$ und homogene Elemente $m\in M,\ f\in S$ mit $\deg(m)=\deg(f)$ derart, dass für alle $\mathfrak{q}\in V$ stets $f\not\in\mathfrak{q}$ und $s(\mathfrak{q})=\frac{m}{f}$ in $M_{(\mathfrak{q})}$ gilt.
\end{enumerate}
$\widetilde{M}$ wird zu einer Garbe mit den gewöhnlichen Restriktionsabbildungen.

\paragraph{Satz 5.21.}\label{5.21} Sei $S$ ein graduierter Ring, $M$ ein graduierter $S$"~Modul und $X=\operatorname{Proj}(S)$. Dann gilt:
\begin{enumerate}[(i)]
\item $(\widetilde{M})_\mathfrak{p}=M_{(\mathfrak{p})}$ für alle $\mathfrak{p}\in X$.
\item Für alle homogene Elemente $f\in S_+$ ist $\widetilde{M}|_{D_+(f)}\cong\widetilde{M_{(f)}}$ bzgl. des Isomorphismus' $D_+(f)\cong\operatorname{Spec}S_{(f)}$.
\item $\widetilde{M}$ ist ein quasikohärenter $\mathcal{O}_X$"~Modul. Ist $X$ noethersch und $M$ endlich erzeugt, so ist $\widetilde{M}$ kohärent.
\end{enumerate}

\paragraph{Beweis.} (i) und (ii) sind analog zu \hyperref[2.23]{Satz 2.23}. (iii) folgt aus (ii).\qed

\paragraph{Definition 5.22.}\label{5.22} Sei $S$ ein graduierter Ring, $X=\operatorname{Proj}(S)$ und $\mathcal{F}$ ein $\mathcal{O}_X$"~Modul. Für $n\in\mathbb{Z}$ definieren wir die \textit{getwistete Garbe}\index{Garbe!getwistet} $\mathcal{F}(n)$ von $\mathcal{F}$ wie folgt:
\[\mathcal{F}(n)=\mathcal{F}\otimes_{\mathcal{O}_X}\mathcal{O}_X(n),\quad \mathcal{O}_X(n)=\widetilde{S(n)} \]

\paragraph{Satz 5.23.}\label{5.23} Sei $S$ ein graduierter Ring und $X=\operatorname{Proj}(S)$, wobei $S$ als $S_0$"~Algebra von $S_1$ erzeugt wird. Dann gilt:
\begin{enumerate}[(i)]
\item $\mathcal{O}_X(n)$ ist eine invertierbare Garbe auf $X$.
\item Sind $M,N$ graduierte $S$"~Moduln, so ist $\widetilde{M\otimes_SN}\cong\widetilde{M}\otimes_{\mathcal{O}_X}\widetilde{N}$. Insbesondere gilt $\widetilde{M(n)}\cong\widetilde{M}(n)$ und $\mathcal{O}_X(n+m)\cong\mathcal{O}_X(n)\otimes\mathcal{O}_X(m)$.
\item Sei $T$ ein weiterer graduierter Ring, der von $T_1$ als $T_0$"~Algebra erzeugt wird und $\varphi:S\to T$ ein Homomorphismus graduierter Ringe. Sei $U\subset_\text{o} Y=\operatorname{Proj}(T)$ und $f:U\to X$ der durch $\varphi$ induzierte Morphismus. Dann gilt für jeder graduierte $S$"~Modul $M$ und jeder graduierte $T$"~Modul $N$:
\[f^\ast(\widetilde{M})\cong\widetilde{M\otimes_ST}|_U,\quad f_\ast(\widetilde{N}|_U)\cong (\widetilde{_SN}) \]
Insbesondere gilt $f^\ast(\mathcal{O}_X(n))\cong\mathcal{O}_Y(n)|_U$ und $f_\ast(\mathcal{O}_X(n)|_U)=(f_\ast\mathcal{O}_U)(n)$.
\end{enumerate}

\paragraph{Beweis.}\begin{enumerate}[(i)]
\item Sei $f\in S_1$ und betrachte $\mathcal{O}_X(n)|_{D_+(f)}\cong\widetilde{S(n)_{(f)}}$ auf $\operatorname{Spec}S_{(f)}$. Es ist $S(n)_{(f)}$ freier $S_{(f)}$"~Modul vom Rang $1$ via dem Isomorphismus $(S_f)_0\to (S_f)_n,\ s\mapsto f^ns$ für alle $n$. Da $S$ von $S_1$ als $S_0$"~Algebra erzeugt wird, gilt $X=\bigcup_{f\in S_1}D_+(f)$. Daher ist $\mathcal{O}_X(n)$ invertierbar.
\item Sei $f\in S_1$. Es gilt $(M\otimes_SN)_{(f)}=M_{(f)}\otimes_{S_{(f)}}N_{(f)}$. Da $S$ von $S_1$ erzeugt wird, folgt $\widetilde{M\otimes_SN}\cong\widetilde{M}\otimes_{\mathcal{O}_X}\widetilde{N}$.
\item Analog wie im affinen Fall.\qed
\end{enumerate}

\paragraph{Definition 5.24.}\label{5.24} Sei $S$ ein graduierter Ring, $X=\operatorname{Proj}(S)$ und $\mathcal{F}$ ein $\mathcal{O}_X$"~Modul. Der zu $\mathcal{F}$ \textit{assoziierte graduierte $S$"~Modul}\index{Modul!assoziiert} $\Gamma_\ast(\mathcal{F})$ ist definiert als die Gruppe
\[\Gamma_\ast(\mathcal{F})=\bigoplus_{n\in\mathbb{Z}}\Gamma(X,\mathcal{F}(n)) \]
mit der folgenden $S$"~Wirkung: Für $s\in S_d\cong\Gamma(X,\mathcal{O}_X(d))$ und $t\in\Gamma(X,\mathcal{F}(n))$ setze $st= s\otimes t\in\Gamma(X,\mathcal{F}(n)\otimes_{\mathcal{O}_X}\mathcal{O}_X(d))=\Gamma(X,\mathcal{F}(n+d))$.

\paragraph{Satz 5.25.}\label{5.25} Sei $A$ ein Ring und $X=\mathbf{P}_A^r$ mit $r\geq 1$ und $S=A[X_0,\ldots,X_r]$. Dann gilt:
\[\Gamma_\ast(\mathcal{O}_X)= S \]

\paragraph{Beweis.} Sei $X=\bigcup_{i=0}^r D_+(X_i)$. Ein $t\in\Gamma(X,\mathcal{O}_X(n))$ entspricht eine Familie $t_i\in\Gamma(D_+(X_i),\mathcal{O}_X(n)),\ i=1,\ldots,r$ mit $t_i=t_j$ auf $D_+(X_iX_j)$. $t_i$ ist ein homogenes Element $s_i\in S_{X_i}$ vom Grad $n$ und $t_i|_{D_+(X_iX_j)}$ entspricht dem Bild von $s_i$ in $S_{X_iX_j}$. Es folgt:
\[\Gamma_\ast(\mathcal{O}_X)=\Big\{ (t_0,\ldots,t_r)\in\prod_{i=0}^r S_{X_i}\mid t_i=t_j\text{ auf } S_{X_iX_j}\text{ für alle }i,j \Big\} \]
Da keine $X_i$ Nullteiler sind, haben wir Inklusionen $S\hookrightarrow S_{X_i}\hookrightarrow S_{X_iX_j}\hookrightarrow S_{X_0\cdots X_r}$ und:
\[\Gamma_\ast(\mathcal{O}_X)=\bigcap_{i=0}^r S_{X_i}\subset S_{X_0\cdots X_r} \]
Jedes homogene $t\in S_{X_0\cdots X_r}$ lässt sich eindeutig in der folgenden Form schreiben:
\[t=X_0^{i_0}\cdots X_r^{i_r}f,\quad i_j\in\mathbb{Z}\]
wobei $f\in S$ ein homogenes Element ist, das durch kein $X_i$ teilbar ist. Es ist $t$ genau dann in $S_{X_i}$, wenn $i_j\geq 0$ für alle $j\neq i$ gilt. Also ist $\bigcap S_{X_i}=S$.\qed

\paragraph{Lemma 5.26.}\label{5.26} Sei $X$ ein Schema, $\mathcal{L}$ eine invertierbare Garbe auf $X$ und $f\in\Gamma(X,\mathcal{L})$. Setze $X_f=\{x\in X\mid f_x\not\in\mathfrak{m}_x\mathcal{L}_x\}\subset_\text{o}X$ und sei $\mathcal{F}$ eine quasikohärente Garbe auf $X$.
\begin{enumerate}[(i)]
\item Sei $X$ quasikompakt und $s\in\Gamma(X,\mathcal{F})$ mit $s|_{X_f}=0$. Dann gibt es ein $n>0$, so dass $f^ns=0\in\Gamma(X,\mathcal{F}\otimes\mathcal{L}^{\otimes n})$.
\item Sei $X=\bigcup U_i$ eine endliche, offene affine Überdeckung, so dass $\mathcal{L}|_{U_i}$ für alle $i$ frei und $U_i\cap U_j$ für alle $i,j$ quasikompakt sind. Zu $t\in\Gamma(X_f,\mathcal{F})$ gibt es ein $n>0$, so dass sich $f^nt\in\Gamma(X_f,\mathcal{F}\otimes\mathcal{L}^{\otimes n})$ zu einem globalen Schnitt auf ganz $X$ fortsetzen lässt.
\end{enumerate}

\paragraph{Bemerkung 5.27.}\label{5.27} Voraussetzungen in \hyperref[5.26]{Lemma 5.26 (i) und (ii)} sind erfüllt, wenn $X$ noethersch ist, oder wenn $X$ quasikompakt und separiert ist.

\paragraph{Beweis von \hyperref[5.26]{Lemma 5.26}.} \begin{enumerate}[(i)]
\item Sei $X=\bigcup U_i$ eine endliche, offene affine Überdeckung mit $\mathcal{L}|_{U_i}$ frei. Betrachte $U=U_i$ und sei $\psi:\mathcal{L}|_U\stackrel{\sim}{\to}\mathcal{O}_U$ ein Isomorphismus. Da $\mathcal{F}$ quasikohärent ist, folgt $\mathcal{F}|_U\cong\widetilde{M}$ für ein $A$"~Modul $M$, wobei $U=\operatorname{Spec}(A)$. Für ein $s\in\Gamma(X,\mathcal{F})$ ist $s|_U\in M$. Setze $g:=\psi(f|_U)\in A$. Es ist $X_f\cap U=D(g)$ und $s|_{X_f}=0$. Nach \hyperref[5.8]{Lemma 5.8 (i)} gibt es ein $n>0$ mit $g^ns=0\in M$. Der Isomorphismus
\[\operatorname{id}\otimes \psi^{\otimes n}:\mathcal{F}\otimes \mathcal{L}^{\otimes n}|_U\to\mathcal{F}|_U \]
liefert $0=f^ns=\Gamma(U, \mathcal{F}\otimes\mathcal{L}^{\otimes n})$ für alle $U=U_i$. Wählt man $n$ so groß, dass die obige Aussage für alle $U_i$ gilt, so folgt $f^ns=0$ auf $X$.
\item Analog zu (i) mit \hyperref[5.8]{Lemma 5.8 (ii)}.\qed
\end{enumerate}

\paragraph{Satz 5.28.}\label{5.28} Sei $S$ ein graduierter Ring, der durch $S_1$ als $S_0$"~Algebra endlich erzeugt wird. Sei $X=\operatorname{Proj}(S)$ und $\mathcal{F}$ ein quasikohärenter $\mathcal{O}_X$"~Modul. Dann gibt es einen natürlichen Isomorphismus:
\[\beta:\widetilde{\Gamma_\ast(\mathcal{F})}\stackrel{\sim}{\to}\mathcal{F} \]

\paragraph{Beweis.} Es ist $X=\bigcup_{f\in S_1}D_+(f)$ eine endliche Vereinigung. Für $f\in S_1$ definiere:
\[\beta_f:\widetilde{\Gamma_\ast(\mathcal{F})_{(f)}}=\widetilde{\Gamma_\ast(\mathcal{F})}|_{D_+(f)}\to\mathcal{F}|_{D_+(f)} \]
durch $\bigoplus_d\Gamma(D_+(f),\mathcal{F}(d))\to\Gamma(D_+(f),\mathcal{F})$, das induziert wird durch:
\[\Gamma(D_+(f),\mathcal{F}(d))\to\Gamma(D_+(f),\mathcal{F}(d)\otimes \mathcal{O}_X(-d))= \Gamma(D_+(f),\mathcal{F}),\ \frac{m}{f^d}\mapsto m\otimes f^{-d} \]
Wir erhalten eine Abbildung $\beta:\widetilde{\Gamma_\ast(\mathcal{F})}\to\mathcal{F}$. Wir zeigen nun, dass alle $\beta_f$ Isomorphismen sind. Es genügt zu zeigen, dass $\Gamma_\ast(\mathcal{F})_{(f)}\to\Gamma(D_+(f),\mathcal{F})$ Isomorphismen sind. \hyperref[5.26]{Lemma 5.26 (i)} liefert die Injektivität und \hyperref[5.26]{Lemma 5.26 (ii)} die Surjektivität.\qed

\paragraph{Korollar 5.29.}\label{5.29} Sei $A$ ein Ring. Dann gilt:
\begin{enumerate}[(i)]
\item Ist $Y\hookrightarrow\mathbf{P}_A^r$ ein abgeschlossenes Unterschema, so existiert ein homogenes Ideal $I\subset S=A[X_0,\ldots,X_r]$, so dass $Y=\operatorname{Proj}(S/I)\hookrightarrow\operatorname{Proj}(S)=X$.
\item Sei $Y$ ein Schema über $\operatorname{Spec}(A)$. Dann ist $Y$ genau dann projektiv, wenn $Y\cong\operatorname{Proj}(S)$ für einen graduierten Ring $S$, der von $S_1$ als $S_0=A$"~Algebra endlich erzeugt wird.
\end{enumerate}

\paragraph{Beweis.} \begin{enumerate}[(i)]
\item Sei $\mathcal{J}_Y\subset\mathcal{O}_X$ die Idealgarbe von $Y$ auf $X=\mathbf{P}_A^r$. Da $\mathcal{J}_Y(d)\subset\mathcal{O}_X(d)$, folgt $\Gamma_\ast(\mathcal{J}_Y)\subset\Gamma_\ast(\mathcal{O}_X)$. Nach \hyperref[5.25]{Satz 5.25} ist $\Gamma_\ast(\mathcal{O}_X)=S$, d.h. $I=\Gamma_\ast(\mathcal{J}_Y)$ ist ein homogenes Ideal in $S$. Setze $Y'=\operatorname{Proj}(S/I)$. $Y'$ ist ein abgeschlossenes Unterschema von $X$ mit Idealgarbe $\mathcal{J}_{Y'}=\widetilde{I}$. Da $\mathcal{J}_Y$ nach \hyperref[5.17]{Satz 5.17 (i)} quasikohärent ist, folgt $\mathcal{J}_Y\cong\widetilde{\Gamma_\ast(\mathcal{J}_Y)}$ nach \hyperref[5.28]{Satz 5.28}. Nun gilt:
\[\mathcal{J}_Y\cong\widetilde{\Gamma_\ast(\mathcal{J}_Y)}=\widetilde{I}=\mathcal{J}_{Y'} \]
Nach \hyperref[5.17]{Satz 5.17} folgt $Y=Y'$, also ist $Y$ das abgeschlossene Unterschema, das durch $I$ definiert ist.
\item Es gilt:
\begin{align*}
Y\text{ projektiv} &\iff Y\text{ ist abgeschlossenes Unterschema von }\mathbf{P}_A^r\text{ für ein }r\\
&\iff Y\cong\operatorname{Proj}(S'/I)\text{ für ein homogenes Ideal }I\subset S' = A[X_0,\ldots,X_r]
\end{align*}
Wir zeigen zunächst, dass $I$ und $I'=\bigoplus_{d\geq d_0}I_d$ dasselbe abgeschlossene Unterschema bestimmen. Dafür zeigen wir $\mathfrak{p}\supset I$, wenn $\mathfrak{p}\supset I'$ für alle $\mathfrak{p}\in\operatorname{Proj}(S')$. Wegen $\mathfrak{p}\not\supset S_+$ gibt es ein $x_i\not\in\mathfrak{p}$. Sei nun $x\in I_r,\ r<d_0$, dann ist $x_i^{d_0-r}x\in I_{d_0}\in\mathfrak{p}$. Da $\mathfrak{p}$ ein Primideal ist, folgt $x\in\mathfrak{p}$.

Somit können wir o.B.d.A. $I\subset S'_+$ annehmen. Also ist $A=(S'/I)_0$ und $S=S'/I$ wird als $A$"~Algebra von $S_1$ endlich erzeugt.

Umgekehrt ist jeder graduierte Ring $S$, der von $S_1$ als $S_0=A$"~Algebra endlich erzeugt ist, Quotient des Polynomrings und $\operatorname{Proj}(S)$ ist projektiv.\qed
\end{enumerate}

\paragraph{Definition 5.30.}\label{5.30} Sei $Y$ ein Schema. Der kanonische Morphismus $g:\mathbf{P}_Y^r=\mathbf{P}_\mathbb{Z}^r\times Y\to\mathbf{P}_\mathbb{Z}^r$ definiert die \textit{getwistete Garbe} $\mathcal{O}(1)$ auf $\mathbf{P}_Y^r$ durch:
\[\mathcal{O}(1)=g^\ast\mathcal{O}(1) \]

\paragraph{Bemerkung.} Ist $Y=\operatorname{Spec}(A)$ affin, so ist $\mathcal{O}(1)$ die bereits in \hyperref[5.22]{Definition 5.22} definierte Garbe auf $\mathbf{P}_A^r$.

\paragraph{Definition 5.31.}\label{5.31} Sei $X$ ein Schema über $Y$. Eine invertierbare Garbe $\mathcal{L}$ auf $X$ heißt \textit{sehr ampel}\index{$\mathcal{O}_X$-Modul!sehr ampel} bzgl. $Y$, wenn es eine Immersion $i:X\hookrightarrow\mathbf{P}_Y^r$ für ein $r$ gibt, so dass $i^\ast(\mathcal{O}(1))\cong\mathcal{L}$.

\paragraph{Satz 5.32.}\label{5.32} Sei $Y$ ein noethersches Schema und $X$ ein Schema über $Y$. Dann ist $X$ genau dann projektiv über $Y$, wenn:
\begin{enumerate}[(i)]
\item $X$ ist eigentlich über $Y$.
\item Es gibt eine sehr ample Garbe auf $X$ bzgl. $Y$.
\end{enumerate}

\paragraph{Beweis.} Sei $X$ projektiv. Dann folgt (i) aus \hyperref[4.24]{Theorem~4.24} und es gibt eine abgeschlossene Immersion $i:X\to\mathbf{P}_Y^r$ für ein $r$, so dass $i^\ast\mathcal{O}(1)$ sehr ampel ist.

Sei umgekehrt $i:X\hookrightarrow\mathbf{P}_Y^r$ eine Immersion und $\mathcal{L}\cong i^\ast(\mathcal{O}(1))$ eine sehr ample Garbe auf $X$ bzgl. $Y$. Betrachte das kartesische Quadrat:
\[\xymatrix{
\mathbf{P}^r_Y \ar@{=}[r] & Y\times \mathbf{P}_\mathbb{Z}^r \ar[r]\ar[d]& Y\ar[d]\\
&\mathbf{P}_\mathbb{Z}^r\ar[r] & \mathbb{Z}
} \]
$\mathbf{P}_\mathbb{Z}^r\to\mathbb{Z}$ ist separiert, da projektiv, also ist auch der Basiswechsel $\mathbf{P}_Y^r\to Y$ separiert. Ferner ist $X\to Y$ eigentlich, nach \hyperref[4.22]{Korollar 4.22 (v)} ist auch $X\hookrightarrow \mathbf{P}_Y^r$ eigentlich, also insbesondere abgeschlossen.\qed

\paragraph{Definition 5.33.}\label{5.33} Sei $X$ ein Schema und $\mathcal{F}$ ein $\mathcal{O}_X$"~Modul. $\mathcal{F}$ heißt \textit{von globalen Schnitten erzeugt}\index{von globalen Schnitten erzeugt}, wenn es eine Familie von globalen Schnitten $s_i\in\Gamma(X,\mathcal{F}),\ i\in I$ gibt, so dass für alle $x\in X$ die Bilder der $s_i$ den Halm $\mathcal{F}_x$ als $\mathcal{O}_X$"~Modul erzeugen. Dies ist äquivalent zu: Es gibt einen surjektiven Garbenmorphismus $\bigoplus_{i\in I}\mathcal{O}_X\to \mathcal{F}$.

\paragraph{Beispiel 5.34.}\label{3.34} Sei $X=\operatorname{Spec}(A)$ und $\mathcal{F}=\widetilde{M}$ für ein $A$"~Modul $M$. Dann wird $\mathcal{F}$ von globalen Schnitten erzeugt; jedes Erzeugendensystem von $M$ als $A$"~Modul liefern solche Schnitte. Die Surjektion $A^{(I)}\to M$ induziert surjektives $\mathcal{O}_X^{(I)}\to\mathcal{F}$.

\paragraph{Beispiel 5.35.}\label{3.35} Sei $X=\operatorname{Proj}(S)$ mit einem graduierten Ring $S$, der von $S_1$ als $S_0$"~Algebra erzeugt wird. Dann liefern die Elemente aus $S_1$ globale Schnitte von $\mathcal{O}_X(1)$ und erzeugen diesen quasikohärenten Modul.

\paragraph{Lemma 5.36.}\label{5.36} \begin{enumerate}[(i)]
\item Abgeschlossene Immersionen sind endliche Morphismen.
\item Sei $f:X\to Y$ ein endlicher Morphismues noetherscher Schemata und $\mathcal{F}$ ein kohärenter $\mathcal{O}_X$"~Modul. Dann ist $f_\ast\mathcal{F}$ kohärent.
\end{enumerate}

\paragraph{Beweis.}\begin{enumerate}[(i)]
\item Sei $f:X\to Y$ eine abgeschlossene Immersion. Sei $V=\operatorname{Spec}(B)\subset_\text{o}Y$. Es ist $f^{-1}(V)=X\times_YV\to V$ als Basiswechsel von $f$ eine abgeschlossene Immersion. Somit ist $f^{-1}(V)\cong\operatorname{Spec}(B/I)$ für ein Ideal $I\subset B$ affin und ferner $B/I$ ein endlich erzeugter $B$"~Modul.
\item Nach \hyperref[5.13]{Satz 5.13 (iii)} ist $f_\ast\mathcal{F}$ quasikohärent. Sei $V=\operatorname{Spec}(B)\subset_\text{o}Y$. Da $f$ endlich ist, ist $f^{-1}(V)=\operatorname{Spec}(A)$ affin und es gilt $\mathcal{F}|_{f^{-1}(V)}=\widetilde{N}$ für ein endlich erzeugter $A$"~Modul $N$. Nach \hyperref[5.3]{Satz 5.3 (iv)} gilt:
\[f_\ast\mathcal{F}|_V = f_\ast\widetilde{N} = \widetilde{_BN} \]
Da $f$ endlich ist, ist $A$ ein endlich erzeugter $B$"~Modul und somit auch $_BN$.\qed
\end{enumerate}

\paragraph{Satz 5.37.}\label{5.37} \textit{(Serre)} Sei $X$ ein projektives Schema über $\operatorname{Spec}(A)$ mit $A$ noethersch. Sei $\mathcal{O}_X(1)$ eine sehr ample Garbe auf $X$ und $\mathcal{F}$ ein kohärenter $\mathcal{O}_X$"~Modul. Dann existiert ein $n_0\in\mathbb{Z}$, so dass für alle $n\geq n_0$ der getwistete $\mathcal{O}_X$"~Modul $\mathcal{F}(n)$ von endlich vielen globalen Schnitten erzeugt wird.

\paragraph{Beweis.} Sei $i:X\hookrightarrow \mathbf{P}_A^r$ eine abgeschlossene Immersion mit $\mathcal{O}_X(1)\cong i^\ast(\mathcal{O}(1))$. Nach \hyperref[5.36]{Lemma 5.36} ist $i_\ast\mathcal{F}$ kohärent auf $\mathbf{P}_A^r$ und nach \hyperref[5.23]{Satz 5.23 (iii)} gilt $(i_\ast\mathcal{F})(n)=i_\ast(\mathcal{F}(n))$. Wird nun $i_\ast\mathcal{F}(n)$ von endlich vielen globalen Schnitten erzeugt, so ist dies auch für $\mathcal{F}(n)$ der Fall, betrachte dafür:
\[\Gamma(\mathbf{P}_A^r, i_\ast\mathcal{F}(n)) = \mathcal{F}(n)(i^{-1}(\mathbf{P}_A^r))=\Gamma(X,\mathcal{F}(n)) \]
\[(i_\ast\mathcal{F}(n))_x = \begin{cases}
\mathcal{F}(n)_x, & x\in i(X)\\
0, &\text{sonst}
\end{cases} \]
Sei also o.B.d.A. $X=\mathbf{P}_A^r=\operatorname{Proj}A[X_0,\ldots,X_r]$. Es ist $X=\bigcup_{i=0}^r D_+(X_i)$. Für jedes $i$ ist $\mathcal{F}|_{D_+(X_i)}\cong\widetilde{M_i}$ für ein endlich erzeugter Modul $M_i$ über $B_i=A\big[\frac{X_0}{X_i},\ldots,\widehat{\frac{X_i}{X_i}},\ldots,\frac{X_r}{X_i}\big]$. Sei $(s_{ij})_j$ ein Erzeugendensystem von $M_i$. Wegen \hyperref[5.26]{Lemma 5.26} existiert ein $n_0>0$, so dass für alle $n\geq n_0$ die Schnitte $X_i^ns_{ij}$ sich zu globalen Schnitten $t_{ij}$ von $\mathcal{F}(n)$ liften lassen. Wir können $n_0$ unabhängig von $i$ und $j$ wählen. Sei $\mathcal{F}(n)|_{D_+(X_i)}\cong\widetilde{M_i'}$ für ein $B_i$"~Modul $\widetilde{M_i'}$. Die Abbildungen $X_i^n:\mathcal{F}\to\mathcal{F}(n)$ induzieren Isomorphismen $M_i\to M_i'$. Da $\{X_i^ns_{ij}\mid j\}$ ganz $M_i'$ erzeugen, erzeugen $t_{ij}\in\Gamma(X,\mathcal{F}(n))$ als globale Schnitte ganz $\mathcal{F}(n)$.\qed

\paragraph{Korollar 5.38.}\label{5.38} Sei $X$ ein projektives Schema über $\operatorname{Spec}(A)$ mit $A$ noethersch. Dann gibt es für jeden kohärenten $\mathcal{O}_X$"~Modul $\mathcal{F}$ eine Surjektion $\mathcal{O}_X(n)^N\to\mathcal{F}$ mit $n,N\in\mathbb{Z}$.

\paragraph{Beweis.} Nach \hyperref[5.37]{Satz 5.37} gibt es eine Surjektion $\mathcal{O}_X^N\to\mathcal{F}(n)$. Tensorieren mit $\mathcal{O}_X(-n)$ gibt die Behauptung.\qed

\paragraph{Satz 5.39.}\label{5.39} Sei $k$ ein Körper, $A$ eine endlich erzeugte $k$"~Algebra und $X$ projektives Schema über $\operatorname{Spec}(A)$. Ferner sei $\mathcal{F}$ ein kohärenter $\mathcal{O}_X$"~Modul. Dann ist $\Gamma(X,\mathcal{F})$ ein endlich erzeugter $A$"~Modul.

\paragraph{Beweis.} Wir werden diesen Satz später kohomologisch beweisen.

\paragraph{Korollar 5.40.}\label{5.40} Sei $f:X\to Y$ ein projektiver Morphismus von Schemata von endlichem Typ über einem Körper $k$. Ist $\mathcal{F}$ kohärent auf $X$, so ist auch $f_\ast\mathcal{F}$ kohärent auf $Y$. Insbesondere ist für $A=k$ der Modul $\Gamma(X,\mathcal{F})$ ein endlich dimensionierter $k$"~Vektorraum.

\paragraph{Beweis.} Sei o.B.d.A. $Y=\operatorname{Spec}(A)$ affin, wobei $A$ eine endlich erzeugte $k$"~Algebra ist. Da $f$ projektiv ist, ist $f$ eigentlich und somit separiert und von endlichem Typ, also quasikompakt. Wegen \hyperref[5.13]{Satz 5.13 (iii)} ist $f_\ast\mathcal{F}$ quasikohärent. Es gilt:
\[f_\ast\mathcal{F} = \widetilde{\Gamma(Y,f_\ast\mathcal{F})}=\widetilde{\Gamma(X,\mathcal{F})} \]
Aber $\Gamma(X,\mathcal{F})$ ist endlich erzeugter $A$"~Modul nach \hyperref[5.39]{Satz 5.39}.\qed

\section{Divisoren}

\paragraph{Definition 6.1.}\label{6.1} \begin{enumerate}[(i)]
\item Ein noetherscher lokaler Ring $(R,\mathfrak{m})$ heißt \textit{regulär}\index{regulär}, falls für $k=R/\mathfrak{m}$ gilt:
\[\dim(R)=\dim_k (\mathfrak{m}/\mathfrak{m}^2)\]
Ist $R$ ein noetherscher lokaler Ring, so gilt stets $\dim(R)\leq\dim_k(\mathfrak{m}/\mathfrak{m}^2)$.
\item Ein Schema $X$ heißt \textit{regulär in Kodimension $1$}\index{regulär}, falls jeder Halm $\mathcal{O}_{X,x}$ von $X$ mit $\dim(\mathcal{O}_{X,x})=1$ regulär ist.
\end{enumerate}

\paragraph{Definition 6.3.}\label{6.3}
Ein Ring heißt \textit{normal}\index{Ring!normal}, wenn er ganzabgeschlossen und nullteilerfrei ist.
Ein Schema heißt \textit{normal}\index{Schema!normal}, wenn seine Halme normal sind.

\paragraph{Theorem 6.4.}\label{6.4} Sei $R$ ein noetherscher, normaler Ring und $\mathfrak{p}$ ein Primideal der Hö\-he $1$. Dann ist $R_\mathfrak{p}$ regulär. Genauer: Sei $(R,\mathfrak{m})$ ein noetherscher lokaler Ring der Dimension $1$. Dann sind folgende Aussagen äquivalent:
\begin{enumerate}[(i)]
\item $R$ ist ein diskreter Bewertungsring.
\item $R$ ist ganzabgeschlossen.
\item $R$ ist regulär.
\item $\mathfrak{m}$ ist ein Hauptideal.
\end{enumerate}

\paragraph{Beweis.} Siehe z.B. Matsumura: \glqq Commutative Algebra\grqq, Theorem 3.9 und Ati\-yah-Mac\-Do\-nald: \glqq Introduction to Commutative Algebra\grqq, Proposition 9.2.\qed

\paragraph{Definition.} Ein Schema habe die Eigenschaft $(\star)$, wenn es noethersch, separiert, integer und regulär in Kodimension $1$ ist.

\paragraph{Definition 6.5.}\label{6.5} Sei $X$ ein Schema mit $(\star)$. Dann gilt:
\begin{enumerate}[(i)]
\item Ein \textit{Primdivisor}\index{Primdivisor} auf $X$ ist ein abgeschlossenes, integres Unterschema der Kodimension $1$.
\item Ein \textit{Weil-Divisor}\index{Divisor!Weil} ist ein Element der freien abelschen Gruppe $\operatorname{Div}(X)$, die von den Primdivisoren erzeugt wird. Wir schreiben ein Divisor als $D=\sum_i n_iY_i$ mit Primdivisoren $Y_i$ und $n_i\in\mathbb{Z}$ mit $n_i=0$ für fast alle $i$. Ein solcher Divisor heißt \textit{effektiv}\index{Divisor!effektiv}, falls alle $n_i\geq 0$ sind.
\item Sei $Y$ ein Primdivisor auf $X$ und $\eta$ ein generischer Punkt in $Y$. Dann ist $\mathcal{O}_{X,\eta}$ nach \hyperref[6.4]{Theorem 6.4} ein diskreter Bewertungsring mit Quotientenkörper $K$, der \textit{Funktionenkörper}\index{Funktionenkörper} von $X$. Sei $v_Y:K^\times\to\mathbb{Z}$ die zugehörige diskrete Bewertung. Sei $f\in K^\times$. Ist $v_Y(f)>0$, so sagen wir, dass $f$ eine \textit{Nullstelle}\index{Nullstelle} entlang $Y$ von der Ordnung $v_Y(f)$ hat. Ist $v_Y(f)<0$, so sagen wir, dass $f$ ein \textit{Pol}\index{Pol} entlang $Y$ von der Ordnung $-v_Y(f)$ besitzt.
\end{enumerate}

\paragraph{Lemma 6.6.}\label{6.6} Sei $X$ ein Schema mit $(\star)$ und $f\in K^\times$. Dann ist $v_Y(f)=0$ für fast alle Primdivisoren $Y$.

\paragraph{Beweis.} Sei $\varnothing \neq U = \operatorname{Spec}(A)\subset_\text{o}X$ affin mit $f|_U$ \textit{regulär}, d.h. $f|_U\in\mathcal{O}_X(U)$. Sei $Z=X\setminus U\subsetneq X$ abgeschlossen. Da $X$ noethersch ist, enthält $Z$ höchstens endlich viele Primdivisoren von $X$, alle anderen treffen $U$. Es genügt also zu zeigen, dass es nur endlich viele Primdivisoren $Y$ in $U$ gibt mit $v_Y(f)\neq 0$, d.h. $v_Y(f)>0$. Es gilt mit $Y=\overline{\{ \eta\}}$:
\[v_Y(f)>0\iff f\in\mathcal{O}_{U,\eta}=A_\eta\iff f\in\eta\iff Y=V(\eta)\subset V(f)\subsetneq U \]
$V(f)$ enthält aber nur endlich viele abgeschlossene irreduzible Teilmengen der Kodimension $1$.\qed

\paragraph{Definition 6.7.}\label{6.7} Sei $X$ ein Schema mit $(\star)$ und $f\in K^\times$. Der Divisor $\operatorname{div}(f)$ von $f$ ist definiert als:
\[\operatorname{div}(f)=\sum_Y v_Y(f)Y \]
wobei $Y$ über die Primdivisoren in $X$ läuft. Diese Summe ist nach \hyperref[6.6]{Lemma 6.6} endlich und somit wohldefiniert. Jeder Divisor der Form $\operatorname{div}(f)$ heißt \textit{Hauptdivisor} oder \textit{prinzipal}.\index{Hauptdivisor}\index{Divisor!prinzipal}.

\paragraph{Bemerkung 6.8.}\label{6.8} Sei $f,g\in K^\times$. Dann gilt:
\[\operatorname{div}\big(\tfrac{f}{g}\big)=\operatorname{div}(f)-\operatorname{div}(g) \]
Somit ist $K^\times\to\operatorname{Div}(X),\ f\mapsto\operatorname{div}(f)$ ein Gruppenhomomorphismus, dessen Bild gerade die Gruppe der Hauptdivisoren in $X$ sind.

\paragraph{Definition 6.9.}\label{6.9} Sei $X$ ein Schema mit $(\star)$. Zwei Divisoren $D,D'$ heißen \textit{linear äquivalent}\index{linear äquivalent} $D\sim D'$, wenn $D-D'$ ein Hauptdivisor ist. Die Gruppe der zugehörigen Äquivalenzklassen $\operatorname{Cl}(X)$ heißt \textit{Divisorenklassengruppe}\index{Divisorenklassengruppe}. Wir haben eine exakte Folge:
\[K^\times\to\operatorname{Div}(X)\to\operatorname{Cl}(X)\to 0 \]

\paragraph{Satz 6.10.}\label{6.10} Sei $A$ ein noetherscher, nullteilerfreier Ring. Dann gilt:
\[A\text{ ist faktoriell}\iff\operatorname{Cl}(\operatorname{Spec}(A))=0 \]

\paragraph{Beweis.} Siehe z.B. Bourbaki: Algèbre Commutative, Chapitre 7 §3 Proposition 2.\qed

\paragraph{Beispiel 6.11.}\label{6.11} \begin{enumerate}
\item Sei $X=\mathbf{A}_k^n$ für ein Körper $k$. Dann gilt $\operatorname{Cl}(X)=0$, da $k[X_1,\ldots,X_n]$ faktoriell ist.
\item Sei $A$ ein Dedekindring. Dann ist $\operatorname{Cl}(\operatorname{Spec}(A))$ gerade die Idealklassengruppe.
\end{enumerate}

\paragraph{Satz 6.12.}\label{6.12} Sei $k$ ein Körper.
\begin{enumerate}[(i)]
\item Sei $X=\mathbf{A}_k^n$ und $Y\subset X$ ein abgeschlossenes Unterschema. Dann ist $Y$ genau dann ein Primdivisor, wenn:
\[Y=V(f)\quad\text{für ein irreduzibles, nicht-konstantes }f\in k[X_1,\ldots,X_n]  \]
\item Sei $X=\mathbf{P}_k^n$ und $Y\subset X$ ein abgeschlossenes Unterschema. Dann ist $Y$ genau dann ein Primdivisor, wenn:
\[Y=V_+(f)\quad\text{für ein homogenes, irreduzibles }f\in k[X_0,\ldots,X_n],\ \deg(f)=r>0 \]
\end{enumerate}

\iffalse
\paragraph{Definition 6.13.}\label{6.13} Sei $X=\mathbf{P}^n_k$. Jeder Primdivisor $Y$ in $X$ hat die Form $Y=V_+(f_Y)$. Betrachte die Abbildung $Y\mapsto\deg(f_Y)$. Diese induziert ein Gruppenhomomorphismus $\operatorname{Div}(X)\to\mathbb{Z}$. Wir zeigen, dass dieser über $\operatorname{Cl}(X)$ faktorisiert. Sei $f\in K^\times$. Dann gilt:
\[\deg\operatorname{div}(f)=\sum_Y v_Y(f) \deg(Y)=\sum_Y v_Y(f) \deg(f_Y) \]
$f=\frac{g}{h}$ mit homogenen $g,h$ vom selben Grad $d$. Sei $g=g_1^{n_1}\cdots g_r^{n_r}$ eine Zerlegung in irreduzible Elemente $g_i$ vom Grad $d_i$. Dann sind $Y_i=\operatorname{div}(g_i)$ nach \hyperref[6.12]{Satz 6.12 (ii)} Primdivisoren.
\fi

\printindex
\end{document}

### 2015-04-12 Algebraische Geometrie/AlgebraischeGeometrie.tex
Add README.md
Closes #1

@@
 \usepackage[utf8]{inputenc}
-\usepackage[german]{babel}
+\usepackage[ngerman]{babel}
 \usepackage{amsmath, amsfonts, amssymb}
@@
 \usepackage{makeidx}
-\usepackage{graphicx}
-\usepackage{dsfont}
+\usepackage{graphicx, tikz}
+\usetikzlibrary{cd, babel}
 \usepackage{hyperref}
@@
 \paragraph{Definition 1.2.}\label{1.2} Eine Prägarbe $F$ auf einem topologischen Raum $X$ heißt \textit{Garbe}\index{Garbe}, falls die folgenden Diagramme exakt sind:
-\[\xymatrix{0\ar[r] & F(U)\ar[r]_-{\operatorname{res}} & \displaystyle\prod_i F(U_i)\ar@<.5ex>[r]\ar@<-.5ex>[r]_-{\operatorname{res}} & \displaystyle\prod_{i,j}F(U_i\cap U_j)} \]
+\[\begin{tikzcd}
+0\ar[r] & F(U)\ar[r, "\operatorname{res}"] & \displaystyle\prod_i F(U_i)\ar[r, "\operatorname{res}", shift left] \ar[r, shift right] & \displaystyle\prod_{i,j}F(U_i\cap U_j)
+\end{tikzcd} \]
 für alle $U\subset_\text{o}X$ und jede offene Überdeckung $U=\bigcup_i U_i$, d.h:
@@
 eine Kollektion von Morphismen $(\varphi(U))_{U\subset_\text{o}X}$, so dass für alle offenen Mengen $V\subset U$ folgendes Diagramm kommutativ ist:
-\[\xymatrix{
-F(U)\ar[rr]^{\varphi(U)}\ar[d]_{\operatorname{res}^U_V} && G(U)\ar[d]^{\operatorname{res}^U_V}\\
-F(V)\ar[rr]_{\varphi(V)} && G(V)
-} \]
+\[\begin{tikzcd}
+F(U)\ar[rr, "\varphi(U)"]\ar[d, "\operatorname{res}^U_V"'] && G(U)\ar[d, "\operatorname{res}^U_V"]\\
+F(V)\ar[rr, "\varphi(V)"'] && G(V)
+\end{tikzcd} \]
 \item Ein \textit{Morphismus}\index{Morphismus!Garben} von Garben ist ein Morphismus von Prägarben. Die (Prä-)Garben bilden eine Kategorie.
@@
 Sei $G$ eine Garbe und $\phi:F\to G$ ein Morphismus von Prägarben. Dann existiert ein eindeutig bestimmter Morphismus $\psi:F^+\to G$, so dass das folgende Diagramm kommutiert:
-\[\xymatrix{ 
-F\ar[rr]^-\theta\ar[d]_\phi && F^+\ar@{-->}[dll]^\psi\\
+\[\begin{tikzcd} 
+F\ar[rr, "\theta"]\ar[d, "\phi"'] && F^+\ar[dll, "\psi", dashed]\\
 G
-} \]
+\end{tikzcd} \]
 $F^+$ ist somit eindeutig bestimmt und heißt die zu $F$ \textit{assoziierte Garbe}\index{assoziierte Garbe!Prägarbe}.
@@
 \item Eine Garbensequenz
-\[\xymatrix{\cdots \ar[r] & F^i\ar[r]^{\varphi^i} & F^{i+1}\ar[r]^{\varphi^{i+1}} & F^{i+2}\ar[r] & \cdots } \]
+\[\begin{tikzcd}
+\cdots \ar[r] & F^i\ar[r, "\varphi^i"] & F^{i+1}\ar[r, "\varphi^{i+1}"] & F^{i+2}\ar[r] & \cdots 
+\end{tikzcd} \]
 heißt \textit{exakt}\index{Exaktheit}, falls $\ker(\varphi^{i+1})=\operatorname{im}(\varphi^i)$ für alle $i$ gilt.
@@
 \item Sei $0\to F'\to F\to F''$ eine exakte Garbensequenz und $U\subset_\text{o}X$. Dann ist auch die folgende Sequenz exakt:
-\[\xymatrix{
+\[\begin{tikzcd}[cramped, sep=small]
 0 \ar[r] & \Gamma(U,F') \ar[r]& \Gamma(U,F) \ar[r]& \Gamma(U,F'')
-} \]
+\end{tikzcd} \]
 Der Funktor $\Gamma(U,-)$ ist linksexakt, aber nicht exakt.
@@
 Dann existiert ein eindeutig bestimmter Morphismus $\psi:\varinjlim F_i\to G$, so dass das folgende Diagramm kommutiert:
-\[\xymatrix{
-F_i\ar[rr]^{\varphi_i}\ar[d]_{\psi_i} & & \varinjlim F_i\ar@{-->}[lld]^\psi\\
+\[\begin{tikzcd}
+F_i\ar[rr, "\varphi_i"]\ar[d, "\psi_i"'] & & \varinjlim F_i\ar[lld, "\psi", dashed]\\
 G
-} \]
+\end{tikzcd} \]
 \item Ebenso wird der \textit{projektive Limes}\index{Limes!projektiv} definiert, wobei alle Pfeile umgedreht werden. Es ist außerdem $U\mapsto\varprojlim F_i(U)$ bereits eine Garbe.
@@
 Sei $F$ eine Garbe auf $X$. Dann ist die folgende Sequenz exakt:
-\[\xymatrix{
+\[\begin{tikzcd}[cramped, sep=small]
 0\ar[r] & j_!(F|_U)\ar[r] & F \ar[r]& i_\ast(F|_Z)\ar[r] & 0
-} \]
+\end{tikzcd} \]
 
@@
 indem man eine Abbildung $s:V\to\coprod_{\mathfrak{q}\in V} A_\mathfrak{q}$ auf die folgende Abbildung $f^\sharp(s): f^{-1}(V)\to\coprod_{\mathfrak{p}\in f^{-1}(V)}B_\mathfrak{p}$ schickt:
-\[ \xymatrix{
-f^{-1}(V) \ar[r]^-f\ar@{-}[d] & V\ar[r]^-s\ar@{-}[d] & A_{\varphi^{-1}(\mathfrak{p})}\ar[r]^-{\varphi_\mathfrak{p}}\ar@{-}[d] & B_\mathfrak{p}\ar@{-}[d]\\
-\mathfrak{p}\ar@{|->}[r] & f(\mathfrak{p}) = \varphi^{-1}(\mathfrak{p})\ar@{|->}[r] & s(\varphi^{-1}(\mathfrak{p}))\ar@{|->}[r] & \varphi_\mathfrak{p}(s(\varphi^{-1}(\mathfrak{p})))
-} \]
+\[ \begin{tikzcd}[row sep=.1em]
+f^{-1}(V) \ar[r, "f"] & V\ar[r, "s"] & A_{\varphi^{-1}(\mathfrak{p})}\ar[r, "\varphi_\mathfrak{p}"] & B_\mathfrak{p}\\
+\mathfrak{p}\ar[r, mapsto] & f(\mathfrak{p}) = \varphi^{-1}(\mathfrak{p})\ar[r, mapsto] & s(\varphi^{-1}(\mathfrak{p}))\ar[r, mapsto] & \varphi_\mathfrak{p}(s(\varphi^{-1}(\mathfrak{p})))
+\end{tikzcd} \]
 Die $f^\sharp(V)$ gibt uns einen Garbenmorphismus $f^\sharp:\mathcal{O}_A\to f_\ast \mathcal{O}_B$.
@@
 Sei $\mathfrak{p}\in\operatorname{Spec}(B)$. Dann haben wir induzierte lokale Homomorphismen mit kommutativem Diagramm:
-\[\xymatrix{
-A_{f(\mathfrak{p})} = \mathcal{O}_{A,f(\mathfrak{p})}  \ar[r]^-{f^\sharp_\mathfrak{p}} & \mathcal{O}_{B,\mathfrak{p}} = B_\mathfrak{p}\\
-A \ar[r]_\varphi\ar[u] & B\ar[u]
-} \]
+\[\begin{tikzcd}
+A_{f(\mathfrak{p})} = \mathcal{O}_{A,f(\mathfrak{p})}  \ar[r, "f^\sharp_\mathfrak{p}"] & \mathcal{O}_{B,\mathfrak{p}} = B_\mathfrak{p}\\
+A \ar[r, "\varphi"']\ar[u] & B\ar[u]
+\end{tikzcd} \]
 Da die $f^\sharp_\mathfrak{p}$ lokale Homomorphismen sind, folgt $f(\mathfrak{p})=\varphi^{-1}(\mathfrak{p})$. Somit ist $f^\sharp$ von dem Ringhomomorphismus $\varphi$ induziert.\qed
@@
 \end{itemize}
-\[\xymatrix{
+\[\begin{tikzcd}[row sep=.1em]
 \operatorname{Spec}(K)\ar[r] & \operatorname{Spec}(R) &\ar[l] \operatorname{Spec}(R/\mathfrak{m})\\
-(0)\ar@{|->}[r]\ar@{-}[u] & t_1\quad t_0 \ar@{-}@<-1.8ex>[u]\ar@{-}@<2.2ex>[u]& \ar@{|->}[l] (0)\ar@{-}[u]
-} \]
+(0)\ar[r, mapsto] & t_1\quad t_0 & \ar[l, mapsto] (0)
+\end{tikzcd} \]
 \item[3.] Sei $k$ ein Körper.
@@
 Ferner kommutiert das folgende Diagramm:
-\[\xymatrix{
-A\ar[d]_{\alpha(f)}\ar[rr]^-{\alpha(f_\nu)} && \Gamma(X_\nu,\mathcal{O}_{X_\nu})\\
-\Gamma(X,\mathcal{O}_X)\ar[rru]_-{\operatorname{res}}
-} \]
+\[\begin{tikzcd}
+A\ar[d, "\alpha(f)"']\ar[rr, "\alpha(f_\nu)"] && \Gamma(X_\nu,\mathcal{O}_{X_\nu})\\
+\Gamma(X,\mathcal{O}_X)\ar[rru, "\operatorname{res}"']
+\end{tikzcd} \]
 weshalb $\alpha(f)$ schon alle $\alpha(f_\nu)$ eindeutig bestimmt.
@@
 Für alle $\nu,\mu$ ist das folgende Diagramm kommutativ:
-\[\xymatrix{
-& \Gamma(X_\nu,\mathcal{O}_X)\ar[dr]^{\operatorname{res}} &\\
-A\ar[ur]^{h_\nu} \ar[dr]_{h_\mu} & & \Gamma(X_\nu\cap X_\mu,\mathcal{O}_X)\\
-& \Gamma(X_\mu,\mathcal{O}_X)\ar[ur]_{\operatorname{res}} &
-} \]
+\[\begin{tikzcd}
+& \Gamma(X_\nu,\mathcal{O}_X)\ar[dr, "\operatorname{res}"] &\\
+A\ar[ur, "h_\nu"] \ar[dr, "h_\mu"'] & & \Gamma(X_\nu\cap X_\mu,\mathcal{O}_X)\\
+& \Gamma(X_\mu,\mathcal{O}_X)\ar[ur, "\operatorname{res}"'] &
+\end{tikzcd} \]
 Aus der Injektivität von $\alpha$ folgt $f_\nu=f_\mu$ auf $X_\nu\cap X_\mu$. Kleben wir die $f_\nu$ nun zusammen, so erhalten wir ein $f:X\to\operatorname{Spec}(A)$ mit $\alpha(f)=h$.\qed
@@
 Es existiert ein kanonischer Homomorphismus $A'\to A$ derart, dass das folgende Diagramm kommutiert:
-\[\xymatrix{
+\[\begin{tikzcd}
 A'\ar[rd]\ar[r] & A\ar[d]\\
 & A'_x= \mathcal{O}_{X,x}= A_x &
-} \]
+\end{tikzcd} \]
 Somit kommutiert:
-\[\xymatrix{
+\[\begin{tikzcd}
 X & U'\ar[l] & U\ar[l]\\
  && \operatorname{Spec}(\mathcal{O}_{X,x})\ar[lu]\ar[u]
-}\qedhere\]
+\end{tikzcd}\qedhere\]
 
@@
 Es gibt also einen eindeutig bestimmten Morphismus $g:X\to Y_\text{red}$ mit kommutativem Diagramm:
-\[\xymatrix{
-X\ar[rr]^f\ar@{-->}[drr]_g && Y\\
+\[\begin{tikzcd}
+X\ar[rr, "f"]\ar[drr, "g"'] && Y\\
 && Y_\text{red}\ar[u]
-} \]
+\end{tikzcd} \]
 \end{enumerate}
@@
 Ein Morphismus zweier Schemata $X$ und $Y$ über $S$ ist ein Morphismus $f:X\to X'$, so dass das folgende Diagramm kommutiert:
-\[\xymatrix{
-X\ar[rr]^f\ar[d] && X'\ar[lld]\\
+\[\begin{tikzcd}
+X\ar[rr, "f"]\ar[d] && X'\ar[lld]\\
 S
-} \]
+\end{tikzcd} \]
 So ein Morphismus nennt man auch \textit{$S$-Morphismus}\index{Morphismus!$S$-Schemata}.
@@
 \item Das folgende Diagramm kommutiert:
-\[ \xymatrix{
-X\times_S Y\ar[r]^-{p_2}\ar[d]_-{p_1} & Y\ar[d]\\
+\[ \begin{tikzcd}
+X\times_S Y\ar[r, "p_2"]\ar[d, "p_1"'] & Y\ar[d]\\
 X\ar[r] & S
-}\]
+\end{tikzcd}\]
 \item Ist $Z$ ein $S$-Schema und $f:Z\to X,\ g:Z\to Y$ Morphismen derart, dass das folgende Diagramm kommutiert:
-\[\xymatrix{
-Z\ar@/^1pc/[rrd]^g\ar@/_1pc/[ddr]_f\ar@{-->}[rd]^\theta &&\\
-&X\times_S Y\ar[r]_-{p_2}\ar[d]^-{p_1} & Y\ar[d]\\
+\[\begin{tikzcd}
+Z\ar[rrd, bend left, "g"]\ar[ddr, "f"', bend right]\ar[rd, "\theta", dashed] &&\\
+&X\times_S Y\ar[r, "p_2"']\ar[d, "p_1"] & Y\ar[d]\\
 &X\ar[r] & S
-} \]
+\end{tikzcd} \]
 so existiert einen eindeutig bestimmten Morphismus $\theta:Z\to X\times_SY$ mit $f=p_1\theta$ und $g=p_2\theta$.
@@
 Offensichtlich kommutiert:
-\[\xymatrix{
- A\ar[r]^-{\tilde{p}_1} & A\otimes_RB\\
- R\ar[r]\ar[u] & B\ar[u]_{\tilde{p}_2}
-} \]
+\[\begin{tikzcd}
+ A\ar[r, "\tilde{p}_1"] & A\otimes_RB\\
+ R\ar[r]\ar[u] & B\ar[u, "\tilde{p}_2"']
+\end{tikzcd} \]
 Sei also $Z$ ein $S$-Schema und Morphismen $f:Z\to X,\ g:Z\to Y$ gegeben, die über $S$ gleich sind.
@@
 Es kommutiert:
-\[\xymatrix{
+\[\begin{tikzcd}
 &&\Gamma(Z,\mathcal{O}_Z)\\
-A\ar[r]_-{\tilde{p}_2}\ar@/^1pc/[rru]^{\tilde{f}}& A\otimes_RB\ar@{-->}[ru]_{\tilde{\theta}} &\\
-R\ar[r]\ar[u] & B\ar[u]^{\tilde{p}_2}\ar@/_1pc/[ruu]_{\tilde{g}}
-} \]
+A\ar[r, "\tilde{p}_2"']\ar[rru, "\tilde{f}", bend left]& A\otimes_RB\ar[ru, "\tilde{\theta}"', dashed] &\\
+R\ar[r]\ar[u] & B\ar[u, "\tilde{p}_2"]\ar[ruu, "\tilde{g}"', bend right]
+\end{tikzcd} \]
 Wegen der Universaleigenschaft des Tensorprodukts gibt es genau einen Ringhomomorphismus $\tilde{\theta}:A\otimes_RB\to\Gamma(Z,\mathcal{O}_Z)$ mit $\tilde{f}\tilde{\theta}=\tilde{p}_1$ und $\tilde{g}\tilde{\theta}=\tilde{p}_2$.
@@
 Da $p_1^{-1}(U)\subset X\times_SY$, kommutiert das Diagramm:
-\[\xymatrix{
-p_1^{-1}(U)\ar[r]^-{p_2}\ar[d]_{p_1} & Y\ar[d]\\
+\[\begin{tikzcd}
+p_1^{-1}(U)\ar[r, "p_2"]\ar[d, "p_1"'] & Y\ar[d]\\
 U\ar[r] & S
-} \]
+\end{tikzcd} \]
 Sei $Z$ ein $S$-Schema und Morphismen $f:Z\to U,\ g:Z\to Y$ gegeben, so dass $(Z\stackrel{f}{\to} U\stackrel{i}{\hookrightarrow} X\to S)=(Z\stackrel{g}{\to} Y\to S)$.
@@
 Es gilt $Z=\bigcup_j\psi_j(X_j\times_SY)$. Also folgt die Kommutativität des zweiten Diagramms aus dem ersten:
-\[\xymatrix{
-X_j\times_S Y\ar[r]^-{p_{1j}}\ar[d]_{p_{2j}} & X_j\ar[d]\\
+\[\begin{tikzcd}
+X_j\times_S Y\ar[r, "p_{1j}"]\ar[d, "p_{2j}"'] & X_j\ar[d]\\
 Y\ar[r] & S
-}\qquad\qquad \xymatrix{
-Z\ar[d]_{p_2}\ar[r]^-{p_1} & X\ar[d]\\
+\end{tikzcd}\qquad\qquad \begin{tikzcd}
+Z\ar[d, "p_2"']\ar[r, "p_1"] & X\ar[d]\\
 Y\ar[r] & S
-}\]
+\end{tikzcd}\]
 Sei nun $Z'$ ein weiteres $S$-Schema und $f:Z'\to X,\ g:Z'\to Y$ gegeben, die über $S$ gleich sind.
@@
 Es kommutiert:
-\[\xymatrix{
-X_i\times_SY\ar@{^{(}->}[r]\ar[d]_{p_{1i}} & Z\ar[d]^{p_1}\\
-X_i\ar@{^{(}->}[r] & X
-} \]
+\[\begin{tikzcd}
+X_i\times_SY\ar[r, hook]\ar[d, "p_{1i}"'] & Z\ar[d, "p_1"]\\
+X_i\ar[r, hook] & X
+\end{tikzcd} \]
 Es gilt $Z_i'\cap Z_j'=f^{-1}(X_i\cap X_j)=f^{-1}(X_{ij})$ und daher $f|_{Z'_i\cap Z'_j}=p_{1i}\circ\theta_i|_{Z'_i\cap Z'_j}=p_{1j}\circ\theta_j|_{Z_i'\cap Z_j'}$, entsprechend für $g$.
@@
 Seien $f:Z\to X_i$ und $g:Z\to Y$ gegeben, die über $S$ gleich sind. Dann gilt $rg(Z)=qf(Z)\subset q(X_i)\subset S_i$, also $g(Z)\subset Y_i$. Wir erhalten kommutatives Diagramm:
-\[\xymatrix{
-Z\ar@/_1pc/[ddr]_f\ar@/^1pc/[rrd]_g\ar@/^1pc/[rrrd]^g\ar@{-->}[rd]_\theta\\
-& X_i\times_{S_i}Y_i\ar[r]_-{p_2}\ar[d]^{p_1} & Y_i\ar@{^{(}->}[r]\ar[d] & Y\ar[dd]^r\\
-& X_i\ar@{^{(}->}[d]\ar[r] & S_i\ar@{^{(}->}[dr] &\\
+\[\begin{tikzcd}
+Z\ar[ddr, "f"', bend right]\ar[rrd, "g"', bend left]\ar[rrrd, "g", bend left]\ar[rd, "\theta"', dashed]\\
+& X_i\times_{S_i}Y_i\ar[r, "p_2"']\ar[d, "p_1"] & Y_i\ar[r, hook]\ar[d] & Y\ar[dd, "r"]\\
+& X_i\ar[d, hook]\ar[r] & S_i\ar[dr, hook] &\\
 & X\ar[rr]_q & & S
-} \]
+\end{tikzcd} \]
 Es existiert genau ein $\theta:Z\to X_i\times_{S_i} Y_i$ mit $f=p_1\theta$ und $g=p_2\theta$. Somit existieren auch $X_i\times_SY$. Nach Schritt 3 existiert auch $X\times_SY$.\qed
@@
 Wir erhalten folgendes kommutative Diagramm:
-\[ \xymatrix{
-Z\ar@/^1pc/[rrd]^\varphi\ar@{-->}[rd]_\theta\ar@/_1pc/[ddr]_\psi\\
-& X\ar[r]\ar[d]\times_SY & X\ar[d]_\xi\ar@/^1pc/[ddr] &\\
-& Y\ar[r]^\eta\ar@/_1pc/[rrd] & S\ar[dr]^\sigma&\\
+\[ \begin{tikzcd}
+Z\ar[rrd, "\varphi", bend left]\ar[rd, "\theta"', dashed]\ar[ddr, "\psi"', bend right]\\
+& X\ar[r]\ar[d]\times_SY & X\ar[d, "\xi"']\ar[ddr, bend left] &\\
+& Y\ar[r, "\eta"]\ar[rrd, bend right] & S\ar[dr, "\sigma"]&\\
 &&&T
-}\]
+\end{tikzcd}\]
 Wegen der Injektivität von $\sigma$, folgt aus $\sigma\xi\varphi=\sigma\eta\psi$ stets $\xi\varphi=\eta\psi$.
@@
 \paragraph{Definition 3.22.}\label{3.22} Seien $f:X_1\to X_2$ und $g:Y_1\to Y_2$ $S$-Morphismen mit kommutativen Diagramm:
-\[\xymatrix{
-&X_1 \ar[r]^f& X_2\\
-X_1\times_SY_1\ar[dr]_{p_2}\ar[ur]^{p_1}\ar@{-->}[rrr]^{f\times_Sg} &&& X_2\times_SY_2\ar[lu]_{q_1}\ar[ld]^{q_2}\\
-& Y_1\ar[r]_g&Y_2
-} \]
+\[\begin{tikzcd}
+&X_1 \ar[r, "f"]& X_2\\
+X_1\times_SY_1\ar[dr, "p_2"']\ar[ur, "p_1"]\ar[rrr, "f\times_Sg", dashed] &&& X_2\times_SY_2\ar[lu, "q_1"']\ar[ld, "q_2"]\\
+& Y_1\ar[r, "g"']&Y_2
+\end{tikzcd} \]
 Es existiert genau ein Morphismus $f\times_Sg: X_1\times_SY_1\to X_2\times_SY_2$ mit $fp_1=q_1(f\times_Sg)$ und $gp_2=q_2(f\times_Sg)$.
@@
 \paragraph{Definition 3.23.}\label{3.23} Sei $f:X\to S$ ein Morphismus von Schemata. Für ein Morphismus $g:T\to S$ erhält man ein $T$-Schema $X_T=X\times_ST$ mit:
-\[\xymatrix{
-X_T\ar[r]\ar[d] & T\ar[d]^g\\
-X\ar[r]_f & S
-} \]
+\[\begin{tikzcd}
+X_T\ar[r]\ar[d] & T\ar[d, "g"]\\
+X\ar[r, "f"'] & S
+\end{tikzcd} \]
 Man sagt, dass $X_T$ durch \textit{Basiswechsel} von $X$ über $S$ durch $T$ erhalten bleibt.\index{Basiswechsel}
@@
 Dann heißt das Faserprodukt $X_y=X\times_Y\operatorname{Spec}\kappa(y)$ die \textit{Faser}\index{Faser} von $f$ über dem Punkt $y$.
-\[\xymatrix{
-X_y\ar[r]^-{p_2}\ar[d]_{p_1} & \operatorname{Spec}\kappa(y)\ar[d]^i\\
-X\ar[r]_f & Y
-}\]
+\[\begin{tikzcd}
+X_y\ar[r, "p_2"]\ar[d, "p_1"'] & \operatorname{Spec}\kappa(y)\ar[d, "i"]\\
+X\ar[r, "f"'] & Y
+\end{tikzcd}\]
 
@@
 Seien ferner Morphismen $T\to Y,\ U\to X$ derart gegeben, dass das folgende Diagramm kommutativ ist:
-\[\xymatrix{
-U\ar[r]\ar[d]_i & X\ar[d]^f\\
-T\ar[r]\ar@{-->}[ur]^h & Y
-}\]
+\[\begin{tikzcd}
+U\ar[r]\ar[d, "i"'] & X\ar[d, "f"]\\
+T\ar[r]\ar[ur, "h", dashed] & Y
+\end{tikzcd}\]
 Dann gibt es höchstens eine Abbildung $h:T\to X$, die das obige Diagramm kommutativ macht. Mit anderen Worten: Die folgende kanonische Abbildung ist für jeden Bewertungsring $R$ über $Y$ injektiv:
@@
 Somit existiert ein eindeutig bestimmter Morphismus $f':X'\to Y'$:
-\[\xymatrix{
-X'\ar[r]^{f'}\ar[d]_i & Y'\ar[d]^j\\
-X\ar[r]_f & Y
-} \]
+\[\begin{tikzcd}
+X'\ar[r, "f'"]\ar[d, "i"'] & Y'\ar[d, "j"]\\
+X\ar[r, "f"'] & Y
+\end{tikzcd} \]
 $f'$ erfüllt ebenfalls die Voraussetzung (ii).
@@
 Dann gilt $f^{-1}(Z)=T$ als Mengen.
-Nach \hyperref[2.21]{Regeln 2.21 (iii)} faktorisiert $f$ in der Form:
-\[\xymatrix{
-T\ar[rr]^f\ar[rrd] &&X\\
+Nach \hyperref[2.21]{Regeln~2.21~(iii)} faktorisiert $f$ in der Form:
+\[\begin{tikzcd}
+T\ar[rr, "f"]\ar[rrd] &&X\\
 && Z\ar[u]
-} \]
+\end{tikzcd} \]
 Beachte, dass für $U\subset_\text{o} Z$ mit $x_0\in U$ auch $x_1\in U$ gilt.
@@
 Wir erhalten folgendes kommutative Diagramm:
-\[\xymatrix{
+\[\begin{tikzcd}
 \mathcal{O}_{Z,x_1} \ar[r]& \mathcal{O}_{T,t_1}=K\\
-\mathcal{O}_{Z,x_0} \ar[r]\ar[u]& \mathcal{O}_{T,t_0}=R\ar@{^{(}->}[u]\\
-\mathfrak{m}_{Z,x_0}\ar[r]\ar@{^{(}->}[u] & \mathfrak{m}_{T,t_0}\ar@{^{(}->}[u]
-}\]
+\mathcal{O}_{Z,x_0} \ar[r]\ar[u]& \mathcal{O}_{T,t_0}=R\ar[u, hook]\\
+\mathfrak{m}_{Z,x_0}\ar[r]\ar[u, hook] & \mathfrak{m}_{T,t_0}\ar[u, hook]
+\end{tikzcd}\]
 Da $Z$ reduziert und irreduzibel ist, ist $Z$ integer. Daher ist $\mathcal{O}_{Z,x_1}=\kappa(x_1)$, alle Abbildungen sind injektiv und wir sehen, dass $R$ über $\mathcal{O}_{Z,x_0}$ dominiert.
@@
 Seien $T\to Y,\ U\to X$ gegeben und $h_1,h_2:T\to X$ mit kommutativem Diagramm:
-\[\xymatrix{
-U\ar[d]_i\ar[r] & X\ar[d]^f\\
-T\ar[r]\ar@<.5ex>[ur]^{h_1}\ar@<-.5ex>[ur]_{h_2} & Y
-} \]
+\[\begin{tikzcd}
+U\ar[d, "i"']\ar[r] & X\ar[d, "f"]\\
+T\ar[r]\ar[ur, "h_1", shift left]\ar[ur, "h_2"', shift right] & Y
+\end{tikzcd} \]
 Setze $t_0=\mathfrak{m}_R\in T$ und $t_1=(0)\in T$.
@@
 Betrachte nun folgendes kommutative Diagramm:
-\[\xymatrix{
-U\ar[r]\ar[d]^i\ar@/_3pc/[rdd]_{h_2i}\ar@/^3pc/[rrd]^{h_1i} & X\ar[d]^-\Delta &\\
-T\ar[r]_-{h''} & X\times_Y X\ar[r]^-{p_1}\ar[d]_-{p_2} & X\ar[d]^f\\
-& X\ar[r]_f & Y
-} \]
+\[\begin{tikzcd}
+U\ar[r]\ar[d, "i"]\ar[rdd, "h_2i"', bend right=90]\ar[rrd, "h_1i", bend left=50] & X\ar[d, "\Delta"] &\\
+T\ar[r, "h''"'] & X\times_Y X\ar[r, "p_1"]\ar[d, "p_2"'] & X\ar[d, "f"]\\
+& X\ar[r, "f"'] & Y
+\end{tikzcd}\]
 Es gilt $\{h''(t_1)\}= h''i(U)\subset\Delta(X)$.
@@
 Betrachte das kommutative Diagramm:
-\[\xymatrix{
-\operatorname{Spec}(K)\ar[r]^-i & T\ar[dr]^h\ar@/^1pc/[rrd]^{p_1h}\ar@/_1pc/[ddr]_{p_2h}\\
-& & X\times_YX\ar[r]_-{p_1}\ar[d]^{p_2} & X\ar[d]^f\\
-&& X\ar[r]_f & Y
-}\]
+\[\begin{tikzcd}
+\operatorname{Spec}(K)\ar[r, "i"] & T\ar[dr, "h"]\ar[rrd, "p_1h", bend left]\ar[ddr, "p_2h"', bend right]\\
+& & X\times_YX\ar[r, "p_1"']\ar[d, "p_2"] & X\ar[d, "f"]\\
+&& X\ar[r, "f"'] & Y
+\end{tikzcd}\]
 Ist $x\in X$, so ist $\kappa(x)\cong\kappa(\Delta(x))$, wegen:
-\[\xymatrix{
-X\ar[rr]^-\Delta\ar[rrd]_{\operatorname{id}} & & X\times_Y X\ar[d]^{p_i}\\
+\[\begin{tikzcd}
+X\ar[rr, "\Delta"]\ar[rrd, "\operatorname{id}"'] & & X\times_Y X\ar[d, "p_i"]\\
 && X
-} \]
+\end{tikzcd} \]
 Betrachte nun das folgende Diagramm:
-\[\xymatrix{
-\operatorname{Spec}(K)\ar[rd]_{hi}\ar@{-->}[r]^-g & X\ar[d]^\Delta \\
+\[\begin{tikzcd}
+\operatorname{Spec}(K)\ar[rd, "hi"']\ar[r, "g", dashed] & X\ar[d, "\Delta"] \\
 & X\times_YX 
-}\]
+\end{tikzcd}\]
 Wähle ein $x_1\in X$ mit $\Delta(x_1)=\xi_1$ und $j:\kappa(x_1)\hookrightarrow K$ als den von $hi$ induzierten $\kappa(\Delta(x_1))\hookrightarrow K$.
@@
 Seien ferner Morphismen $T\to Y$ und $U\to X$ derart gegeben, dass das folgende Diagramm kommutativ ist:
-\[\xymatrix{
-U\ar[r]^u\ar[d]_i & X\ar[d]^f\\
-T\ar[r]_t\ar@{-->}[ru]^h & Y
-}\]
+\[\begin{tikzcd}
+U\ar[r, "u"]\ar[d, "i"'] & X\ar[d, "f"]\\
+T\ar[r, "t"']\ar[ru, "h", dashed] & Y
+\end{tikzcd}\]
 Dann existiert ein eindeutig bestimmter Morphismus $h:T\to X$, der das Diagramm kommutativ macht. Mit anderen worten ist die folgende Abbildung bijektiv:
@@
 Betrachte die Basiserweiterung $X_T=X\times_YT$ von $X$ mit $T\to Y$:
-\[\xymatrix{
-U \ar@{-->}[dr]_\theta\ar@/^1pc/[rrd]^u\ar@/_1pc/[ddr]_i &&\\
-& X_T\ar[r]_-{p_1}\ar[d]^{f'} & X\ar[d]^f\\
-& T\ar[r]_t & Y
-} \]
+\[\begin{tikzcd}
+U \ar[dr, "\theta"', dashed]\ar[rrd, "u", bend left]\ar[ddr, "i"', bend right] &&\\
+& X_T\ar[r, "p_1"']\ar[d, "f'"] & X\ar[d, "f"]\\
+& T\ar[r, "t"'] & Y
+\end{tikzcd} \]
 Wir erhalten ein $\theta:U\to X_T$, so dass das obige Diagramm kommutiert.
@@
 Betrachte das kommutative Diagramm:
-\[ \xymatrix{
-Z\ar@{^{(}->}[r]\ar[dr] & X'\ar[r]\ar[d]^{f'} & X\ar[d]^f\\
+\[ \begin{tikzcd}
+Z\ar[r, hook]\ar[dr] & X'\ar[r]\ar[d, "f'"] & X\ar[d, "f"]\\
 & Y'\ar[r] & Y
-} \]
+\end{tikzcd} \]
 Da $f$ von endlichem Typ ist, ist nach \hyperref[3.25]{Satz 3.25 (ii)} auch $f'$ von endlichem Typ.
@@
 Betrachte folgendes kommutative Diagramm:
-\[\xymatrix{
-U\ar[r] \ar[d]_i & Z\ar@{^{(}->}[r] \ar[dr] & X'\ar[r]\ar[d]^{f'} & X\ar[d]^f\\
-T\ar[rr]\ar@{-->}[rrru]\ar@{-->}[rru] & & Y'\ar[r] & Y
-} \]
+\[\begin{tikzcd}
+U\ar[r] \ar[d, "i"'] & Z\ar[r, hook] & X'\ar[r]& X\ar[d, "f"]\\
+T\ar[rr]\ar[rrru, dashed]\ar[rru, dashed] & & Y'\ar[r]\ar[from=u, "f'", crossing over]\ar[from=ul, crossing over] & Y
+\end{tikzcd} \]
 Nach Voraussetzung gibt es ein Morphismus $h:T\to X$, der das Diagramm kommutativ macht.
@@
 Sei nun $R$ ein Bewertungsring und seien Morphismen $U\to X,\ T\to Y$ gegeben mit kommutativem Diagramm:
-\[\xymatrix{
-U \ar[r]^u\ar[d]_i& X\ar[d]^f\\
-T\ar[r]_t\ar[dr]_{t'} & Y\ar[d]^g\\
+\[\begin{tikzcd}
+U \ar[r, "u"]\ar[d, "i"']& X\ar[d, "f"]\\
+T\ar[r, "t"]\ar[dr, "t'"'] & Y\ar[d, "g"]\\
 & Z
-} \]
+\end{tikzcd} \]
 Wir wollen ein $h:T\to X$ konstruieren mit $fh=t$ und $hi=u$.
@@
 Sei nun $R$ ein Bewertungsring mit Quotientenkörper $K=\operatorname{Quot}(R)$ und $U\to X,\ T\to\operatorname{Spec}(\mathbb{Z})$ Morphismen mit kommutativem Diagramm:
-\[\xymatrix{
-U\ar[r]^u\ar[d]_i & X\ar[d]\\
-T\ar[r]_-t & \operatorname{Spec}(\mathbb{Z})
-}\]
+\[\begin{tikzcd}
+U\ar[r, "u"]\ar[d, "i"'] & X\ar[d]\\
+T\ar[r, "t"'] & \operatorname{Spec}(\mathbb{Z})
+\end{tikzcd}\]
 Wir zeigen nun per Induktion, dass genau ein Morphismus $h:T\to X$ existiert, der das obige Diagramm kommutativ ergänzt.
@@
 Es folgt $f_{ik}\in R$ für alle $i$. Es gibt Homomorphismen:
-\[\xymatrix{
-\mathbb{Z}\big[\frac{x_0}{x_k},\ldots,\frac{x_n}{x_k}\big]\ar[rr]^-\varphi_-{\frac{x_i}{x_k}\mapsto f_{ik} } \ar[d]&& R\ar[d]\\
+\[\begin{tikzcd}
+\mathbb{Z}\big[\frac{x_0}{x_k},\ldots,\frac{x_n}{x_k}\big]\ar[rr, "\varphi", "\frac{x_i}{x_k}\mapsto f_{ik}"'] \ar[d]&& R\ar[d]\\
 \mathcal{O}_{\xi_1}\ar[r] & \kappa(\xi_1)\ar[r] & K
-}\]
+\end{tikzcd}\]
 Wir erhalten den zu $\varphi$ gehörige Morphismus $T\to V_k$ und somit $h:T\to V_k\hookrightarrow X$.
@@
 Ist $D(f)\supset D(g)$, so erhalten wir das folgende kommutative Diagramm:
-\[\xymatrix{
-M_f\ar[rr]^{\psi_f}\ar[d] && N_f\ar[d]\\
-M_g\ar[rr]_{\psi_g} && N_g
-} \]
+\[\begin{tikzcd}
+M_f\ar[rr, "\psi_f"]\ar[d] && N_f\ar[d]\\
+M_g\ar[rr, "\psi_g"'] && N_g
+\end{tikzcd} \]
 Da $D(f),\ f\in A$ eine Basis der Topologie bilden, induzieren $\psi_f,\ f\in A$ einen $\widetilde{A}$"~Homomorphismus $\widetilde{\psi}:\widetilde{M}\to\widetilde{N}$ mit $\widetilde{\psi}|_{D(f)}=\psi_f$.
@@
 Es gilt $M_i=\mathcal{F}(D(g_i))\cong M_{g_i}$ und wir haben ein kommutatives Diagramm:
-\[\xymatrix{
-\mathcal{F}(X)_{g_i}\ar[rr]^\cong && \mathcal{F}(D(g_i))\\
-\mathcal{F}(X)\ar[u]\ar[urr]_{\operatorname{res}}
-} \]
+\[\begin{tikzcd}
+\mathcal{F}(X)_{g_i}\ar[rr, "\cong"] && \mathcal{F}(D(g_i))\\
+\mathcal{F}(X)\ar[u]\ar[urr, "\operatorname{res}"']
+\end{tikzcd} \]
 Somit ist $\alpha|_{D(g_i)}$ ein Isomorphismus für alle $i$. Da die $D(g_i)$ ganz $X$ überdecken, ist $\alpha$ ein Isomorphismus.
@@
 Da $M\mapsto\widetilde{M}$ ein exakter Funktor ist, folgt die Exaktheit von:
-\[\xymatrix{
+\[\begin{tikzcd}
 0  \ar[r]& \widetilde{\Gamma(X,\mathcal{F}')} \ar[d] \ar[r]& \widetilde{\Gamma(X,\mathcal{F})} \ar[d] \ar[r]& \widetilde{\Gamma(X,\mathcal{F}'')} \ar[d] \ar[r]& 0\\
 0 \ar[r] & \mathcal{F}'  \ar[r]& \mathcal{F}  \ar[r]&  \ar[r]\mathcal{F}''  \ar[r]& 0
-} \]
+\end{tikzcd} \]
 Die vertikalen Pfeile links und rechts sind nach \hyperref[5.10]{Korollar 5.10} Isomorphismen.
@@
 \paragraph{Beweis.} Betrachte das kartesische Diagramm:
-\[\xymatrix{
-X\times_Y X\ar[r]^-{p_2}\ar[d]_{p_1} & X\ar[d]^f\\
-X\ar[r]_f & Y
-} \]
-Nach \hyperref[3.20]{3.20} ist $p_1^{-1}(U)\cap p_2^{-1}(V)=U\times_YV$ affin.
-Es ist $U\cap V\cong \Delta_{X/Y}(X)\cap U\times_YV\subset U\times_YV$.
-Nun ist $\Delta_{X/Y}(X)$ abgeschlossen in $X\times_YX$, d.h.
-$U\cap V$ ist ein abgeschlossenes Unterschema in $U\times_YV$.\qed
+\[\begin{tikzcd}
+X\times_Y X\ar[r, "p_2"]\ar[d, "p_1"'] & X\ar[d, "f"]\\
+X\ar[r, "f"'] & Y
+\end{tikzcd} \]
+Nach \hyperref[3.20]{Lemma~3.20} ist $p_1^{-1}(U)\cap p_2^{-1}(V)=U\times_YV$ affin.
+Es gilt: \[U\cap V\cong \Delta_{X/Y}(X)\cap U\times_YV\subset U\times_YV\] Nun ist $\Delta_{X/Y}(X)$ abgeschlossen in $X\times_YX$, d.h.
+$U\cap V$ ist ein abgeschlossenes Un\-ter\-sche\-ma in $U\times_YV$.\qed
 
@@
 Sei umgekehrt $i:X\hookrightarrow\mathbf{P}_Y^r$ eine Immersion und $\mathcal{L}\cong i^\ast(\mathcal{O}(1))$ eine sehr ample Garbe auf $X$ bzgl. $Y$. Betrachte das kartesische Quadrat:
-\[\xymatrix{
-\mathbf{P}^r_Y \ar@{=}[r] & Y\times \mathbf{P}_\mathbb{Z}^r \ar[r]\ar[d]& Y\ar[d]\\
-&\mathbf{P}_\mathbb{Z}^r\ar[r] & \mathbb{Z}
-} \]
+\[\begin{tikzcd}
+\mathbf{P}^r_Y = Y\times \mathbf{P}_\mathbb{Z}^r \ar[r]\ar[d]& Y\ar[d]\\
+\mathbf{P}_\mathbb{Z}^r\ar[r] & \mathbb{Z}
+\end{tikzcd} \]
 $\mathbf{P}_\mathbb{Z}^r\to\mathbb{Z}$ ist separiert, da projektiv, also ist auch der Basiswechsel $\mathbf{P}_Y^r\to Y$ separiert.
@@
 
-\iffalse
 \paragraph{Definition 6.13.}\label{6.13} Sei $X=\mathbf{P}^n_k$.
@@
 Dann sind $Y_i=\operatorname{div}(g_i)$ nach \hyperref[6.12]{Satz 6.12 (ii)} Primdivisoren.
-\fi
 

### 2015-04-12 Algebraische Geometrie/AlgebraischeGeometrie.tex
Use fontspec package
Update Algebraische Geometrie

@@
 \documentclass[11pt,b5paper,openany]{memoir}
-\usepackage[utf8]{inputenc}
+\usepackage{fontspec}
 \usepackage[ngerman]{babel}
@@
 Dann ist $\mathcal{O}_{X,\eta}$ nach \hyperref[6.4]{Theorem 6.4} ein diskreter Bewertungsring mit Quotientenkörper $K$, der \textit{Funktionenkörper}\index{Funktionenkörper} von $X$.
-Sei $v_Y:K^\times\to\mathbb{Z}$ die zugehörige diskrete Bewertung.
+Wir bezeichnen die zugehörige diskrete Bewertung mit  $v_Y:K^\times\to\mathbb{Z}$.
+
 Sei $f\in K^\times$.
@@
 Es gilt mit $Y=\overline{\{ \eta\}}$:
-\[v_Y(f)>0\iff f\in\mathcal{O}_{U,\eta}=A_\eta\iff f\in\eta\iff Y=V(\eta)\subset V(f)\subsetneq U \]
-$V(f)$ enthält aber nur endlich viele abgeschlossene irreduzible Teilmengen der Kodimension $1$.\qed
+\[v_Y(f)>0\iff f\not\in\mathcal{O}_{U,\eta}^\times=A_\eta^\times\iff f\in\eta\iff Y=V(\eta)\subset V(f)\subset U \]
+Da $f\neq 0$, ist $V(f)\subsetneq U$ eine echte abgeschlossene Teilmenge, und enthält daher nur endlich viele abgeschlossene irreduzible Teilmengen der Kodimension $1$ in $U$.\qed
 
@@
 \[\deg\operatorname{div}(f)=\sum_Y v_Y(f) \deg(Y)=\sum_Y v_Y(f) \deg(f_Y) \]
-$f=\frac{g}{h}$ mit homogenen $g,h$ vom selben Grad $d$.
+Sei $f=\frac{g}{h}$ mit homogenen $g,h$ vom selben Grad $d$.
 Sei $g=g_1^{n_1}\cdots g_r^{n_r}$ eine Zerlegung in irreduzible Elemente $g_i$ vom Grad $d_i$.
 Dann sind $Y_i=\operatorname{div}(g_i)$ nach \hyperref[6.12]{Satz 6.12 (ii)} Primdivisoren.
+Es gilt:
+\[\deg\operatorname{div}(g)=\sum_i n_i\deg(g_i) = \sum_i n_id_i = d \]
+Analog ist $\deg\operatorname{div}(h)=d$. Somit ist $\deg\operatorname{div}(f)=\deg\operatorname{div}(g)-\deg\operatorname{div}(h)=0$.\qed
+
+\paragraph{Satz 6.15.}\label{6.15} Sei $X=\mathbf{P}_k^n$. Dann gilt:
+\begin{enumerate}[(i)]
+\item Ist $D$ ein Divisor auf $X$, so ist $D\sim\deg(D)\cdot V_+(T_0)$
+\item $\deg:\operatorname{Cl}(X)\to\mathbb{Z}$ ist ein Isomorphismus.
+\end{enumerate}
+
+\paragraph{Beweis.} \begin{enumerate}[(i)]
+\item Sei $D=\sum_Y n_YY$ ein Divisor mit Primdivisoren $Y$.
+Nach \hyperref[6.12]{Satz 6.12 (ii)} ist $Y=V_+(f_Y)$ mit irreduziblen, homogenen Polynomen $f_Y$ vom Grad $r_Y$.
+Wir schreiben $f_Y=T_0^{r_Y} g_Y$, wobei $g_Y$ ein Polynom in $\frac{T_1}{T_0},\ldots,\frac{T_n}{T_0}$ ist.
+Die $g_Y$ sind rationale Funktionen auf $\mathbf{P}_k^n$ und es gilt:
+\[\operatorname{div}(g_Y) = V_+(f_Y)-r_YV_+(T_0) \implies V_+(f_Y)\sim r_YV_+(T_0) \]
+Also gilt $D=\sum_Yn_Y V_+(f_Y)\sim \big(\sum n_Y r_Y\big) V_+(T_0) = \deg(D)\cdot V_+(T_0)$.
+\item folgt aus (i) und wegen $\deg V_+(T_0)=1$.\qed
+\end{enumerate}
+
+\paragraph{Satz 6.16.}\label{6.16} Sei $X$ ein Schema mit $(\star)$, $Z\subsetneq X$ eine abgeschlossene Teilmenge und $U=X\setminus Z$. Dann gilt:
+\begin{enumerate}[(i)]
+\item Es gibt einen surjektiven Homomorphismus $\operatorname{Cl}(X)\to\operatorname{Cl}(U)$, der durch $\sum n_iY_i\mapsto \sum n_i(Y_i\cap U)$ gegeben ist, wobei wir die leeren $Y_i\cap U$ ignorieren.
+\item Ist $\operatorname{codim}(Z,X)\geq 2$, dann ist die obige Abbildung ein Isomorphismus.
+\item Ist $Z$ irreduzibel und $\operatorname{codim}(Z,X)=1$, so gibt es eine exakte Folge:
+\[\mathbb{Z}\to\operatorname{Cl}(X)\to\operatorname{Cl}(U)\to 0 \]
+wobei die erste Abbildung durch $1\mapsto Z$ gegeben ist.
+\end{enumerate}
+
+\paragraph{Beweis.}\begin{enumerate}[(i)]
+\item Ist $Y$ ein Primdivisor auf $X$, so ist $Y\cap U$ leer oder ein Primdivisor auf $U$.
+Sei $f\in K^\times$ mit $\operatorname{div}(f)=\sum n_iY_i$.
+Fassen wir $f$ als rationale Funktion auf $U$ auf, so erhalten wir $\operatorname{div}(f|_U)=\sum n_i(Y_i\cap U)$.
+Somit ist die Abbildung $\operatorname{Cl}(X)\to\operatorname{Cl}(U)$ wohldefiniert.
+
+Die Surjektivität sieht man wie folgt: Sei $Y$ ein Primdivisor auf $U$ und $\overline{Y}$ der Abschluss von $Y$ in $X$. Dann ist $\overline{Y}$ ein Primdivisor auf $X$ mit $\overline{Y}\cap U=Y$.
+\item $\operatorname{Div}(X)$ bzw. $\operatorname{Cl}(X)$ hängen nur von Teilmengen der Kodimension $1$ ab.
+\item Es ist $\ker(\operatorname{Cl}(X)\to\operatorname{Cl}(U)) = \{ [D]\in\operatorname{Cl}(X)\mid \operatorname{supp}(D)\subset Z\}=\langle[Z]\rangle$, da $Z$ irreduzibel ist.\qed
+\end{enumerate}
+
+\paragraph{Beispiel 6.17.}\label{6.17} Sei $Y$ eine irreduzible Kurve vom Grad $d$ in $\mathbf{P}_k^2$. Wir haben ein kommutatives Diagramm:
+\[\begin{tikzcd}
+\mathbb{Z} \rar\dar["\cong"'] & \operatorname{Cl}(\mathbf{P}_k^2) \rar\dar["\cong"', "\deg"]& \operatorname{Cl}(\mathbf{P}_k^2\setminus Y) \rar\dar["\cong"]& 0\\
+d\mathbb{Z} \rar[hook] & \mathbb{Z}\rar & \mathbb{Z}/d\mathbb{Z}\rar & 0
+\end{tikzcd}\]
+
+\paragraph{Definition 6.18.}\label{6.18} Sei $k$ ein algebraisch abgeschlossener Körper.
+Eine \textit{Kurve}\index{Kurve} über $k$ ist ein integres, separiertes Schema $X$ von endlichem Typ über $k$ der Dimension $1$.
+\begin{itemize}
+\item $X$ heißt \textit{vollständig}\index{Kurve!vollständig}, wenn $X\to k$ eigentlich ist.
+\item $X$ heißt \textit{nicht-singulär}\index{Kurve!nicht-singulär}, falls alle lokalen Ringe von $X$ regulär sind.
+\end{itemize}
+
+\paragraph{Satz 6.19.}\label{6.19} Sei $X$ eine vollständige, nicht-singuläre Kurve über $k$ und $Y$ eine beliebige Kurve über $k$ mit einem Morphismus $f:X\to Y$. Dann gilt:
+\[f(X)\text{ ist ein Punkt} \quad\text{oder}\quad f(X)=Y \]
+Im zweiten Fall gilt:
+\begin{enumerate}[(i)]
+\item Die Funktionenkörpererweiterung $K(X)/K(Y)$ ist endlich.
+\item $f$ ist endlich.
+\item $Y$ ist vollständig.
+\end{enumerate}
+
+\paragraph{Lemma 6.20.}\label{6.20} Sei $f:X\to Y$ surjektiv, $g:Y\to Z$ separiert und von endlichem Typ, und $g\circ f:X\to Z$ eigentlich. Dann ist $g$ eigentlich.
+
+\paragraph{Beweis von \hyperref[6.19]{6.19}.} TODO
+
+\paragraph{Theorem 6.21.}\label{6.21} Sei $B$ ein Integritätsring und eine endlich erzeugte Algebra über einem Körper $k$.
+Setze $K=\operatorname{Quot}(B)$ und sei $L/K$ eine endliche Erweiterung.
+Dann ist der Ganzabschluss $A$ von $B$ in $L$ ein endlich erzeugter $B$"~Modul.
+
+\paragraph{Beweis.} Siehe z.B. Zariski-Samuel: Commutative Algebra I, Chapter V, Theorem 9.\qed
+
+\paragraph{Beweis von \hyperref[6.20]{6.20}.} TODO
+
+\paragraph{Definition 6.22.}\label{6.22} Sei $f:X\to Y$ ein dominanter Morphismus von Kurven.
+Dann heißt $\deg(f)=[K(X):K(Y)]$ der \textit{Grad} von $f$\index{Grad}.
+Sei $X$ eine nicht-singuläre Kurve, erfüllt also insbesondere $(\star)$.
+Ein Primdivisor von $X$ ist genau ein abgeschlossener Punkt.
+Ein Divisor $D$ ist somit von der folgenden Form:
+\[D=\sum_i n_iP_i,\quad n_i\in\mathbb{Z},\ P_i\text{ abgeschlossene Punkte} \]
+Der \textit{Grad}\index{Grad} von $D$ ist definiert als $\deg(D)=\sum n_i$.
 

### 2015-04-13 Algebraische Geometrie/AlgebraischeGeometrie.tex
Beweis von 6.19

@@
 \paragraph{Satz 6.19.}\label{6.19} Sei $X$ eine vollständige, nicht-singuläre Kurve über $k$ und $Y$ eine beliebige Kurve über $k$ mit einem Morphismus $f:X\to Y$. Dann gilt:
-\[f(X)\text{ ist ein Punkt} \quad\text{oder}\quad f(X)=Y \]
+\[f(X)=\operatorname{Pt} \quad\text{oder}\quad f(X)=Y \]
 Im zweiten Fall gilt:
@@
 
-\paragraph{Beweis von \hyperref[6.19]{6.19}.} TODO
+\paragraph{Beweis von \hyperref[6.19]{6.19}.} Da $X$ eigentlich über $k$ ist, ist $f(X)\subset Y$ ein abgeschlossenes Unterschema.
+Da $X$ irreduzibel ist, ist $f(X)$ auch irreduzibel.
+Nun ist $\dim(Y)=1$, folgt entweder $f(X)=\operatorname{Pt}$ oder $f(X)=Y$.
+Betrachte das folgende kommutative Diagramm:
+\[\begin{tikzcd}
+X \rar["f"] & f(X)\rar[hook]\ar[dr] & Y\dar\\
+ & & k
+\end{tikzcd} \]
+Es ist $f(X)\hookrightarrow Y$ eine abgeschlossene Immersion, also separiert nach \hyperref[4.17]{Satz 4.17 (i)}.
+Somit ist auch $f(X)/ k$ separiert.
+Nach \hyperref[6.20]{Lemma 6.20} ist $f(X)/k$ eigentlich.
+
+Sei nun $f(X)=Y$, also ist $Y$ vollständig.
+Sei $Y=\overline{\{\eta\}}$ und $X=\overline{\{\xi\}}$.
+Da $f:X\to Y$ dominant ist, folgt $f(\xi)=\eta$ und wir erhalten eine Inklusion $K(Y)=\mathcal{O}_{Y,\eta}\to\mathcal{O}_{X,\xi}=K(X)$.
+Da beide Körper endlich erzeugt über $k$ vom Transzendenzgrad $\dim(X)=1$ sind, ist $K(X)/K(Y)$ endlich.
+
+Es bleibt zu zeigen, dass $f$ endlich ist.
+Sei $V=\operatorname{Spec}(B)\subset_\text{o}Y$ affin.
+Dann ist $B=\mathcal{O}_Y(V)\subset K(Y)\subset K(X)$.
+Es ist zu zeigen, dass $f^{-1}(V)=\operatorname{Spec}(A)$ affin ist, wobei $A$ ein endlich erzeugter $B$"~Modul ist.
+In der Tat folgt aus der Normalität von $X$ stets $f^{-1}(V)=\operatorname{Spec}(A)$, wobei $A$ der Ganzabschluss von $B$ in $K(X)$ ist.
+Alles folgt nun aus \hyperref[6.21]{Theorem 6.21}.\qed
 

### 2015-04-13 Algebraische Geometrie/AlgebraischeGeometrie.tex
Definition 6.23

@@
 In der Tat folgt aus der Normalität von $X$ stets $f^{-1}(V)=\operatorname{Spec}(A)$, wobei $A$ der Ganzabschluss von $B$ in $K(X)$ ist.
-Alles folgt nun aus \hyperref[6.21]{Theorem 6.21}.\qed
+Alles folgt nun aus dem folgenden \hyperref[6.21]{Theorem 6.21}.\qed
 
@@
 
+\paragraph{Definition 6.23.}\label{6.23} Sei $f:X\to Y$ ein endlicher Morphismus nicht-singulärer Kurven.
+Wir definieren ein Homomorphismus $f^\ast:\operatorname{Div}(Y)\to\operatorname{Div}(X)$ wie folgt:
+
+Sei $Q\in Y$ ein abgeschlossener Punkt und wähle ein $t\in\mathcal{O}_{Y,Q}\subset K(Y)$ mit $v_Q(t)=1$, wobei $v_Q$ die diskrete Bewertung zu $\mathcal{O}_{Y,Q}$ ist.
+$t$ heißt \textit{lokaler Parameter}\index{lokaler Parameter} an der Stelle $Q$.
+Wir setzen:
+\[f^\ast(Q) = \sum_{P\in f^{-1}(Q)}v_P(t)P \]
+Da $f$ endlich ist, ist obige Summe nach \hyperref[6.24]{Lemma 6.24} endlich.
+
+\paragraph{Lemma 6.24.}\label{6.24} \begin{enumerate}[(i)]
+\item Die Eigenschaft endlich zu sein ist stabil unter Basiswechsel.
+\item Ist $f:X\to Y$ ein endlicher Morphismus und $y\in Y$, so ist $f^{-1}(y)$ endlich.
+\end{enumerate}
+
 \printindex

### 2015-04-14 Algebraische Geometrie/AlgebraischeGeometrie.tex
Beweis von 6.20

@@
 
-\paragraph{Beweis von \hyperref[6.20]{6.20}.} TODO
+\paragraph{Beweis von \hyperref[6.20]{6.20}.} Es reicht zu zeigen, dass $g$ universell abgeschlossen ist. Sei $h:Y'\to Y$ ein beliebiger Morphismus und betrachte den Basiswechsel:
+\[\begin{tikzcd}
+X\times_Y Y' \ar[r, "f'"]\ar[d, "h'"'] & Y' \ar[d, "h"]\\
+X\ar[r, "f"'] & Y
+\end{tikzcd} \]
+Wir zeigen nun, dass $f'$ surjektiv ist.
+Sei $y'\in Y'$ und sei $x\in X$ mit $f(x)=h(y')=:y$.
+Behauptung: Es gibt ein $z\in X\times_YY'$ mit $h'(z)=x$ und $f'(z)=y'$.
+Seien $i_x:\{x\}\hookrightarrow X,\ i_{y'}:\{y'\}\hookrightarrow Y'$ die Inklusionen.
+Diese liefern einen $Y$"~Morphismus:
+\[\delta : \operatorname{Spec}\kappa(x)\times_{\kappa(y)}\operatorname{Spec}\kappa(y') = \operatorname{Spec}(\kappa(x)\otimes_{\kappa(y)}\kappa(y'))\hookrightarrow X\times_YY' \]
+Da $\kappa(x)\otimes_{\kappa(y)}\kappa(y')\neq 0$, gilt für jedes $z\in\operatorname{im}(\delta)$ stets $h'(z)=x$ und $f'(z)=y'$. Dies zeigt die Surjektivität von $f'$.
+
+Da alle anderen Eigenschaften in den Voraussetzungen stabil unter Basiswechsel sind (siehe \hyperref[4.17]{4.17 (iii)} und \hyperref[4.22]{4.22}), reicht es zu zeigen, dass $g$ abgeschlossen ist.
+Sei $Y'\subset Y$ eine abgeschlossene Teilmengen.
+Dann ist wegen der Surjektivität von $g$
 

### 2015-04-14 Algebraische Geometrie/AlgebraischeGeometrie.tex
14.04.2015

@@
 
+Also ist die Surjektivität von $f$ stabil unter Basiswechsel.
 Da alle anderen Eigenschaften in den Voraussetzungen stabil unter Basiswechsel sind (siehe \hyperref[4.17]{4.17 (iii)} und \hyperref[4.22]{4.22}), reicht es zu zeigen, dass $g$ abgeschlossen ist.
 Sei $Y'\subset Y$ eine abgeschlossene Teilmengen.
-Dann ist wegen der Surjektivität von $g$
+Dann ist wegen der Surjektivität von $f$:
+\[g(Y')=(g\circ f)\circ f^{-1}(Y') \]
+$g(Y')$ ist abgeschlossen, da $g\circ f$ abgeschlossen ist.\qed
 
@@
 
+\paragraph{Satz 6.25.}\label{6.25} Sei $f:X\to Y$ ein endlicher Morphismus nicht-singulärer Kurven. Dann:
+\[\deg(f^\ast D)=\deg(f)\cdot \deg(D) \]
+
+\paragraph{Korollar 6.26.}\label{6.26} Hauptdivisoren auf einer vollständigen, nicht-singulären Kurve haben Grad $0$. Wir erhalten eine surjektive Gradabbildung:
+\[\deg:\operatorname{Cl}(X)\to\mathbb{Z} \]
+
+\paragraph{Lemma 6.27.}\label{6.27} Sei $X$ eine normale Kurve über $k$, $U\subset X$ offen, $\varphi:X\to\mathbf{P}^n$ eine rationale Abbildung.
+Dann lässt sich $\varphi$ zu einem einem Morphismus auf $X$ fortsetzen.
+
+\paragraph{Definition 6.28.}\label{6.28} Sei $X$ ein Schema.
+\begin{enumerate}[(i)]
+\item Sei $U=\operatorname{Spec}(A)\subset_\text{o}X$ affin und $S$ die Menge aller Nicht-Nullteiler in $A$.
+Dann heißt $K(U)=A_S$ \textit{totaler Quotientenring} von $A$.\index{totaler Quotientenring}
+\item Sei $\mathcal{K}$ die Ringgarbe, die zur Prägarbe $U\mapsto\varprojlim_{V\subset U\text{ affin}} K(V)$ assoziiert ist.
+$\mathcal{K}$ heißt \textit{Garbe der totalen Quotientenringe} von $\mathcal{O}$.
+\item $\mathcal{K}^\times:U\mapsto \mathcal{K}(U)^\times$ sei die Garbe von multiplikativen Gruppe der invertierbaren Elemente in $\mathcal{K}$.
+$\mathcal{O}^\times$ sei die Garbe der invertierbaren Elemente in $\mathcal{O}$.
+\end{enumerate}
+
+\paragraph{Definition 6.29.}\label{6.29} Sei $X$ ein Schema.
+Ein Element aus $\Gamma(X,\mathcal{K}^\times/\mathcal{O}^\times)$ heißt \textit{Cartier-Divisor}\index{Divisor!Cartier}.
+Ein Cartier-Divisor kann also durch ein System $(U_i,f_i)_i$ beschrieben werden, wobei $(U_i)_i$ eine offene Überdeckung von $X$ ist, und $f_i\in\Gamma(U_i,\mathcal{K}^\times)$, so dass für alle $i,j$ stets $\frac{f_i}{f_j}\in\Gamma(U_i\cap U_j,\mathcal{O}^\times)$ gilt.
+
+Ein Cartier-Divisor heißt \textit{prinzipal}\index{Divisor!prinzipal} oder \textit{Hauptdivisor}\index{Hauptdivisor}, wenn er im Bild der kanonischen Abbildung $\Gamma(X,\mathcal{K}^\times)\to\Gamma(X,\mathcal{K}^\times/\mathcal{O}^\times)$ ist.
+Zwei Cartier-Divisoren heißen \textit{linear äquivalent}\index{linear äquivalent}, falls ihre Differenz prinzipal ist.
+
+\paragraph{Satz 6.30.}\label{6.30} Sei $X$ ein integres, separiertes, noethersches Schema.
+Ferner sei $X$ \textit{lokal faktoriell}\index{lokal faktoriell}, d.h.
+alle lokalen Ringe sind faktoriell.
+Dann ist die Gruppe $\operatorname{Div}(X)$ von Weil-Divisoren isomorph zur Gruppe der Cartier-Divisoren $\Gamma(X,\mathcal{K}^\times/\mathcal{O}^\times)$.
+Prinzipale Weil-Divisoren entsprechen prinzipale Cartier-Divisoren.
+
 \printindex

### 2015-04-14 Algebraische Geometrie/AlgebraischeGeometrie.tex
Beweis von 6.6 näher erläutert

@@
 Sei $Z=X\setminus U\subsetneq X$ abgeschlossen.
-Da $X$ noethersch ist, enthält $Z$ höchstens endlich viele Primdivisoren von $X$, alle anderen treffen $U$.
+Da $X$ noethersch und irreduzibel ist, ist $Z$ noethersch mit $\operatorname{codim}(Z,X)\geq 1$ und besitzt nur endlich viele irreduzible Komponenten.
+Daher enthält $Z$ höchstens endlich viele Primdivisoren von $X$, alle anderen treffen $U$.
 Es genügt also zu zeigen, dass es nur endlich viele Primdivisoren $Y$ in $U$ gibt mit $v_Y(f)\neq 0$, d.h.
@@
 \[v_Y(f)>0\iff f\not\in\mathcal{O}_{U,\eta}^\times=A_\eta^\times\iff f\in\eta\iff Y=V(\eta)\subset V(f)\subset U \]
-Da $f\neq 0$, ist $V(f)\subsetneq U$ eine echte abgeschlossene Teilmenge, und enthält daher nur endlich viele abgeschlossene irreduzible Teilmengen der Kodimension $1$ in $U$.\qed
+Da $f\neq 0$, ist $V(f)\subsetneq U$ eine echte abgeschlossene Teilmenge, und enthält daher nur endlich viele irreduzible Komponenten, also Primdivisoren in $U$.\qed
 
@@
 Also ist die Surjektivität von $f$ stabil unter Basiswechsel.
-Da alle anderen Eigenschaften in den Voraussetzungen stabil unter Basiswechsel sind (siehe \hyperref[4.17]{4.17 (iii)} und \hyperref[4.22]{4.22}), reicht es zu zeigen, dass $g$ abgeschlossen ist.
+Da alle anderen Eigenschaften in den Voraussetzungen ebenfalls stabil unter Basiswechsel sind (siehe \hyperref[4.17]{4.17 (iii)} und \hyperref[4.22]{4.22}), reicht es zu zeigen, dass $g$ abgeschlossen ist.
 Sei $Y'\subset Y$ eine abgeschlossene Teilmengen.

### 2015-04-14 Algebraische Geometrie/AlgebraischeGeometrie.tex
Beweis von 6.30

@@
 
+\paragraph{Beweis.} TODO
+
 \paragraph{Satz 6.25.}\label{6.25} Sei $f:X\to Y$ ein endlicher Morphismus nicht-singulärer Kurven. Dann:
@@
 
-\paragraph{Definition 6.28.}\label{6.28} Sei $X$ ein Schema.
+\paragraph{Definition 6.28.}\label{6.28} Sei $(X,\mathcal{O})$ ein Schema.
 \begin{enumerate}[(i)]
@@
 
-\paragraph{Definition 6.29.}\label{6.29} Sei $X$ ein Schema.
+\paragraph{Definition 6.29.}\label{6.29} Sei $(X,\mathcal{O})$ ein Schema.
 Ein Element aus $\Gamma(X,\mathcal{K}^\times/\mathcal{O}^\times)$ heißt \textit{Cartier-Divisor}\index{Divisor!Cartier}.
@@
 
-\paragraph{Satz 6.30.}\label{6.30} Sei $X$ ein integres, separiertes, noethersches Schema.
+\paragraph{Satz 6.30.}\label{6.30} Sei $(X,\mathcal{O})$ ein integres, separiertes, noethersches Schema.
 Ferner sei $X$ \textit{lokal faktoriell}\index{lokal faktoriell}, d.h.
@@
 
+\paragraph{Beweis.} $X$ ist normal, da faktorielle Ringe insbesondere ganzabgeschlossen sind.
+Nach \hyperref[6.4]{Theorem 6.4} erfüllt $X$ $(\star)$.
+Also können wir von Weil-Divisoren sprechen.
+Da $X$ integer ist, ist $K(U)=\operatorname{Quot}(A)= K$ der Funktionenkörper von $X$ für alle $U=\operatorname{Spec}(A)\subset_\text{o}X$.
+Somit ist $\mathcal{K}$ konstante Garbe.
+
+Sei $(U_i,f_i)_i$ ein Cartier-Divisor mit einer offenen Überdeckung $(U_i)_i$ von $X$ und $f_i\in\Gamma(U_i,\mathcal{K}^\times)=K^\times$.
+Wir ordnen diesen Cartier-Divisor den folgenden Weil-Divisor zu:
+\[D=\sum_Y v_Y(f_{i_Y})Y \]
+wobei $Y$ die Primdivisoren von $X$ durchläuft und $i_Y$ ein Index mit $U_i\cap Y\neq\varnothing$. Die Summe ist endlich, da $X$ noethersch ist. $D$ ist unabhängig von Wahl der Indizes $i_Y$:
+
+Seien $i,j$ mit $U_i\cap Y\neq\varnothing$ und $U_j\cap Y\neq\varnothing$.
+Dann ist $\frac{f_i}{f_j}$ auf $U_i\cap U_j$ invertierbar, d.h.
+$\frac{f_i}{f_j}\in\Gamma(U_i\cap U_j,\mathcal{O}^\times)$.
+Es folgt $v_Y\big(\frac{f_i}{f_j}\big)=0$, also $v_Y(f_i)=v_Y(f_j)$.
+
+Sei nun umgekehrt $D=\sum_Yn_YY$ ein Weil-Divisor auf $X$ und $x\in X$. Dann induziert $D$ ein Weil-Divisor $D_x$ auf $\operatorname{Spec}(\mathcal{O}_x)$, nämlich:
+\[D_x=\sum_Y n_Y(Y\cap\operatorname{Spec}(\mathcal{O}_x)) \]
+Da $\mathcal{O}_x$ als faktoriell vorausgesetzt wird, ist nach \hyperref[6.10]{Satz 6.10} $\operatorname{Cl}(\operatorname{Spec}\mathcal{O}_x)=0$ und $D_x$ ein Hauptdivisor, d.h.
+$D_x=(f_x)$ für ein $f_x\in K^\times$.
+Fassen wir $(f_x)$ als Weil-Divisor in $X$ auf, so sehen wir, dass sich $(f_x)$ und $D$ nur bei Primdivisoren, die nicht durch $x$ gehen, unterscheiden.
+Davon gibt es nur endlich viele, deren Koeffizienten nicht verschwinden.
+Daher existiert eine offene Umgebung $U_x$ von $x$ mit $(f_x)|_{U_x}=D|_{U_x}$.
+Das System $(U_x,f_x)_{x\in X}$ liefert einen Cartier-Divisor.
+
+Geben $f,f'$ denselben Weil-Divisor auf $U\subset_\text{o}X$ offen, so ist $\frac{f}{f'}\in\Gamma(U,\mathcal{O}^\times)$, da $X$ normal ist.
+Daher ist die Konstruktion wohldefiniert.
+Die obigen Konstruktionen sind invers zueinander.\qed
+
 \printindex

### 2015-04-14 Algebraische Geometrie/AlgebraischeGeometrie.tex
Beweis von 6.24

@@
 
-\paragraph{Beweis.} TODO
+\paragraph{Beweis.} \begin{enumerate}[(i)]
+\item Sei $f:X\to Y$ ein endlicher Morphismus und $g:Z\to Y$ beliebig. Betrachte den folgenden Basiswechsel:
+\[\begin{tikzcd}
+X\times_YZ \dar\rar["f_Z"] & Z\dar["g"]\\
+X \rar["f"'] & Y
+\end{tikzcd} \]
+Sei $V=\operatorname{Spec}(B)\subset_\text{o}Y$ affin.
+Dann existiert ein affines $W=\operatorname{Spec}(C)\subset_\text{o}Z$ mit $g(W)\subset V$.
+Es ist $f^{-1}(V)=\operatorname{Spec}(A)$ affin, wobei $A$ ein endlich erzeugter $B$"~Modul ist.
+Es gilt:
+\begin{align*}
+f_Z^{-1}(W)= (X\times_YZ)\times_ZW &\cong X\times_YW \\
+&\cong (X\times_YV)\times_VW\\
+&\cong f^{-1}(V)\times_VW=\operatorname{Spec}(A\otimes_BC)
+\end{align*}
+und $A\otimes_BC$ ist ein endlich erzeugter $C$"~Modul.
+\item Sei $f^{-1}(y)\neq\varnothing$. Betrachte den Morphismus:
+\[f^{-1}(y)= X\times_Y\operatorname{Spec}\kappa(y)\to\operatorname{Spec}\kappa(y) \]
+Dieser ist endlich nach (i).
+Also ist $f^{-1}(y)=\operatorname{Spec}(A)$ affin, wobei $A$ ein endlich dimensionierter $\kappa(y)$"~Vektorraum ist.
+Somit ist $\dim f^{-1}(y)=0$, als topologischer Raum ist $f^{-1}(y)$ also eine endliche Menge an Punkten.
+\end{enumerate}
 

### 2015-04-14 Algebraische Geometrie/AlgebraischeGeometrie.tex
Bemerkung nach 6.24

@@
 Also ist $f^{-1}(y)=\operatorname{Spec}(A)$ affin, wobei $A$ ein endlich dimensionierter $\kappa(y)$"~Vektorraum ist.
-Somit ist $\dim f^{-1}(y)=0$, als topologischer Raum ist $f^{-1}(y)$ also eine endliche Menge an Punkten.
+Somit ist $\dim f^{-1}(y)=\dim\kappa(y)= 0$, als topologischer Raum ist $f^{-1}(y)$ also eine endliche Menge an Punkten.\qed
+\end{enumerate}
+
+\paragraph{Bemerkung.} \begin{enumerate}[(i)]
+\item In \hyperref[6.23]{Definition 6.23} ist $f^\ast Q$ unabhängig von der Wahl des lokalen Parameters, da ein lokaler Parameter eindeutig bis auf eine Einheit in $\mathcal{O}_{Y,Q}$ bestimmt ist.
+\item $f^\ast$ \textit{respektiert lineare Äquivalenz}, d.h. für $h\in K(Y)^\times,\ \operatorname{div}(h)=\sum_Qv_Q(h)Q$ gilt:
+\[f^\ast(\operatorname{div}(h))=\sum_Qv_Q(h)\sum_{P\in f^{-1}(Q)}v_P(t)P=\sum_Pv_P(h)P=\operatorname{div}(h) \]
+da $v_P(h)=v_P(t)\cdot v_Q(h)$ ist. Somit induziert $f^\ast$ eine Abbildung $\operatorname{Cl}(Y)\to\operatorname{Cl}(X)$.\qed
 \end{enumerate}

### 2015-04-17 Algebraische Geometrie/AlgebraischeGeometrie.tex
Ende §2.6

@@
 
+\paragraph{Satz 6.31.}\label{6.31} Seien $\mathcal{L},\mathcal{M}$ invertierbare Garben auf einem Schema $X$. Dann gilt:
+\begin{enumerate}[(i)]
+\item $\mathcal{L}\otimes \mathcal{M}$ ist invertierbar.
+\item $\operatorname{\mathcal{H}om}_{\mathcal{O}_X}(\mathcal{L},\mathcal{O}_X)$ ist invertierbar und wird mit $\mathcal{L}^{-1}$ bezeichnet.
+\item $\mathcal{L}\otimes\mathcal{L}^{-1}\cong\mathcal{O}_X$
+\item $\operatorname{\mathcal{H}om}_{\mathcal{O}_X}(\mathcal{L},\mathcal{L})=\mathcal{O}_X$
+\end{enumerate}
+
+\paragraph{Definition 6.32.}\label{6.32}\begin{enumerate}[(i)]
+\item Wir setzen:
+\[\mathcal{L}^{\otimes m}=\underbrace{\mathcal{L}\otimes\cdots\otimes\mathcal{L}}_{m\text{-mal}},\quad\quad \mathcal{L}^{\otimes -m}=(\mathcal{L}^{-1})^{\otimes m},\quad m\geq 0 \]
+Also gilt $\mathcal{L}^{\otimes n}\otimes\mathcal{L}^{\otimes m}=\mathcal{L}^{\otimes (n+m)}$ für alle $n,m\in\mathbb{Z}$.
+\item Die \textit{Picard-Gruppe} $\operatorname{Pic}(X)$\index{Picard-Gruppe} eines Schemas $X$ ist die abelsche Gruppe von Isomorphieklassen invertierbarer Garben auf $X$ mit der Operation $\otimes$.
+\end{enumerate}
+
+\paragraph{Definition 6.33.}\label{6.33} Sei $D$ ein Cartier-Divisor auf $X$ repräsentiert durch $(U_i,f_i)_i$. Dann ist die Untergarbe $\mathcal{L}(D)$ von $\mathcal{K}$ definiert durch:
+\[\mathcal{L}(D)|_{U_i}=\mathcal{O}_{U_i}f_i^{-1}\subset\mathcal{K}|_{U_i} \]
+Diese ist wohldefiniert, denn auf $U_i\cap U_j$ haben wir $\mathcal{L}(D)|_{U_i\cap U_j}=\mathcal{O}_{U_i\cap U_j}f_i^{-1}=\mathcal{O}_{U_i\cap U_j}f_j^{-1}$, da $\frac{f_i}{f_j}\in\Gamma(U_i\cap U_j,\mathcal{O}^\times)$.
+$\mathcal{L}(D)$ heißt die zum Divisor $D$ \textit{assoziierte Garbe}\index{assoziierte Garbe!zum Divisor}.
+
+\paragraph{Satz 6.34.}\label{6.34} Sei $X$ ein Schema.
+\begin{enumerate}[(i)]
+\item Für jeden Cartier-Divisor $D$ ist $\mathcal{L}(D)$ eine invertierbare Garbe auf $X$. Die Abbildung $D\mapsto\mathcal{L}(D)$ induziert eine Bijektion:
+\[\{\text{Cartier-Divisoren auf }X \}\to\{\text{Invertierbare Untergarben von }\mathcal{K} \} \]
+\item $\mathcal{L}(D_1-D_2)\cong\mathcal{L}(D_1)\otimes\mathcal{L}(D_2)^{-1}$ für Cartier-Divisoren $D_1,D_2$
+\item $D_1\sim D_2\iff\mathcal{L}(D_1)\cong\mathcal{L}(D_2)$
+\end{enumerate}
+
+\paragraph{Beweis.} \begin{enumerate}[(i)]
+\item Sei repräsentiert durch $(U_i,f_i)_i$ mit $f_i\in\Gamma(U_i,\mathcal{K}^\times)$.
+Dann ist $\mathcal{O}_{U_i}\to\mathcal{L}(D)|_{U_i},\ 1\mapsto f_i^{-1}$ ein Isomorphismus, also $\mathcal{L}(D)$ invertierbar.
+Sei umgekehrt $\varphi:\mathcal{L}\hookrightarrow\mathcal{K}$ eine invertierbare Untergarbe und $\{U_i\}$ eine Überdeckung von $X$ mit $\tilde{\varphi}:\mathcal{O}_{U_i}\stackrel{\sim}{\to}\mathcal{L}|_{U_i}\hookrightarrow\mathcal{K}|_{U_i}$.
+Setze $f_i=\tilde{\varphi}(U_i)(1)^{-1}$.
+Dann definiert $(U_i,f_i)_i$ ein Cartier-Divisor $D$ mit $\mathcal{L}(D)=\mathcal{L}$.
+\item Seien $D_1, D_2$ repräsentiert durch $(U_i,f_i)_i$ bzw. $(U_i,g_i)_i$. Dann wird $D_1-D_2$ repräsentiert durch $(U_i,f_ig_i^{-1})_i$. Also gilt:
+\[\mathcal{L}(D_1-D_2)|_{U_i}=\mathcal{O}_{U_i}(f_i^{-1}g_i)=\mathcal{L}(D_1)|_{U_i}\otimes \mathcal{L}(D_2)|_{U_i}^{-1}\subset\mathcal{K}|_{U_i} \]
+Somit folgt $\mathcal{L}(D_1-D_2)=\mathcal{L}(D_1)\otimes\mathcal{L}(D_2)^{-1}$.
+\item Wegen (ii) reicht es zu zeigen, dass $D$ genau dann ein Hauptdivisor ist, wenn $\mathcal{L}(D)\cong\mathcal{O}_X$.
+Sei $D$ prinzipal, d.h.
+$D=(f)$ für ein $f\in\Gamma(X,\mathcal{K}^\times)$.
+Somit ist $\mathcal{L}(D)=\mathcal{O}_Xf^{-1}\cong\mathcal{O}_X$.
+Sei umgekehrt $\varphi:\mathcal{O}_X\to\mathcal{L}(D)$ ein Isomorphismus.
+Setze $f$ als das Bild von $1$ der folgenden Komposition:
+\[\Gamma(X,\mathcal{O}_X)\stackrel{\sim}{\to} \Gamma(X,\mathcal{L}(D))\hookrightarrow\Gamma(X,\mathcal{K}^\times) \]
+Dann ist $\mathcal{L}(D)=\mathcal{L}((f^{-1}))$, also $D=(f^{-1})$.\qed
+\end{enumerate}
+
+\paragraph{Korollar 6.35.}\label{6.35} Sei $X$ ein Schema. Dann induziert die Abbildung $D\mapsto\mathcal{L}(D)$ einen injektiven Gruppenhomomorphismus:
+\[\operatorname{CaCl}(X)\hookrightarrow\operatorname{Pic}(X)\]
+wobei $\operatorname{CaCl}(X)$ die Cartier-Divisorenklassengruppe bezeichnet. Im Allgemeinen ist diese Abbildung nicht surjektiv.
+
+\paragraph{Satz 6.36.}\label{6.36} Sei $X$ ein integres Schema. Dann ist die Abbildung $\operatorname{CaCl}(X)\to\operatorname{Pic}(X)$ ein Gruppenisomorphismus.
+
+\paragraph{Beweis.} Es genügt zu zeigen, dass jede invertierbare Garbe isomorph zu einer Untergarbe von $\mathcal{K}$ ist.
+Da $X$ integer ist, ist $\mathcal{K}$ konstante Garbe mit $\mathcal{K}(U)=K$ der Funktionenkörper von $X$ für alle $U$.
+Sei $\mathcal{L}$ eine invertierbare Garbe auf $X$ und betrachte $\mathcal{L}\otimes\mathcal{K}$.
+Sei $X=\bigcup_i U_i$ eine offene Überdeckung mit $\mathcal{L}|_{U_i}\cong\mathcal{O}_{U_i}$.
+Dann gilt:
+\[(\mathcal{L}\otimes_{\mathcal{O}_X}\mathcal{K})|_{U_i}=\mathcal{L}|_{U_i}\otimes_{\mathcal{O}_{U_i}}\mathcal{K}|_{U_i}=\mathcal{K}|_{U_i} \]
+Da $X$ irreduzibel ist, folgt $\mathcal{L}\otimes\mathcal{K}\cong\mathcal{K}$.
+Die kanonische Abbildung $\mathcal{L}\to\mathcal{L}\otimes\mathcal{K}\cong\mathcal{K}$ zeigt, dass $\mathcal{L}$ isomorph zu einer Untergarbe von $\mathcal{K}$ ist.\qed
+
+\paragraph{Korollar 6.37.}\label{6.37} Sei $X$ ein noethersches, integres, separiertes und lokal faktorielles Schema. Dann existiert ein Isomorphismus:
+\[\operatorname{Cl}(X)\cong\operatorname{Pic}(X) \]
+
+\paragraph{Beweis.} Folgt aus \hyperref[6.30]{Satz 6.30} und \hyperref[6.36]{Satz 6.36}.\qed
+
+\paragraph{Korollar 6.38.}\label{6.38} Sei $k$ ein Körper und $X=\mathbf{P}_k^n$. Dann ist jede invertierbare Garbe auf $X$ isomorph zu einem $\mathcal{O}(\ell)$ für ein $\ell\in\mathbb{Z}$.
+
+\paragraph{Beweis.} Nach \hyperref[6.37]{Korollar 6.37} und \hyperref[6.15]{Satz 6.15} ist $\operatorname{Pic}(X)\cong \operatorname{Cl}(X)\cong\mathbb{Z}$.
+Ferner wird $\operatorname{Cl}(X)$ von der Hyperebene $D=V_+(T_0)$ erzeugt.
+Dann gilt $\mathcal{L}(D)T_0=\mathcal{O}(1)$, da $\mathcal{O}(1)|_{D_+(T_i)}=\mathcal{O}_{D_+(T_i)}T_i$ und $\mathcal{L}(D)|_{D_+(T_i)}=\mathcal{O}_{D_+(T_i)}\frac{T_i}{T_0}$.\qed
+
 \printindex

### 2015-04-17 Algebraische Geometrie/AlgebraischeGeometrie.tex
Update Index: von globalen Schnitten erzeugt

@@
 \paragraph{Definition 5.33.}\label{5.33} Sei $X$ ein Schema und $\mathcal{F}$ ein $\mathcal{O}_X$"~Modul.
-$\mathcal{F}$ heißt \textit{von globalen Schnitten erzeugt}\index{von globalen Schnitten erzeugt}, wenn es eine Familie von globalen Schnitten $s_i\in\Gamma(X,\mathcal{F}),\ i\in I$ gibt, so dass für alle $x\in X$ die Bilder der $s_i$ den Halm $\mathcal{F}_x$ als $\mathcal{O}_X$"~Modul erzeugen.
+$\mathcal{F}$ heißt \textit{von globalen Schnitten erzeugt}\index{$\mathcal{O}_X$-Modul!von globalen Schnitten erzeugt}, wenn es eine Familie von globalen Schnitten $s_i\in\Gamma(X,\mathcal{F}),\ i\in I$ gibt, so dass für alle $x\in X$ die Bilder der $s_i$ den Halm $\mathcal{F}_x$ als $\mathcal{O}_X$"~Modul erzeugen.
 Dies ist äquivalent zu: Es gibt einen surjektiven Garbenmorphismus $\bigoplus_{i\in I}\mathcal{O}_X\to \mathcal{F}$.

### 2015-04-17 Algebraische Geometrie/AlgebraischeGeometrie.tex
Anfang §2.7

@@
 
+\section{Projektive Morphismen}
+
+\paragraph{Satz 7.1.}\label{7.1} Sei $A$ ein Ring und $X$ ein Schema über $A$.
+\begin{enumerate}[(i)]
+\item Ist $\varphi:X\to\mathbf{P}_A^n$ ein $A$-Morphismus.
+Dann ist $\varphi^\ast(\mathcal{O}(1))$ eine invertierbare Garbe auf $X$, die von globalen Schnitten $s_i=\varphi^\ast(x_i),\ i=0,\ldots,n$ erzeugt wird, wobei $x_i\in\Gamma(\operatorname{Proj}(A[x_0,\ldots,x_n]),\mathcal{O}(1))$.
+\item Sei $\mathcal{L}$ eine invertierbare Garbe auf $X$ und $s_0,\ldots,s_n\in\Gamma(X,\mathcal{L})$ globale Schnitte, die $\mathcal{L}$ erzeugen.
+Dann existiert ein eindeutig bestimmter $A$"~Morphismus $\varphi:X\to\mathbf{P}_A^n$ derart, dass $\mathcal{L}\cong\varphi^\ast(\mathcal{O}(1))$ mit $s_i=\varphi^\ast(x_i)$ für gewisse $x_0,\ldots,x_n\in X$ gibt.
+\end{enumerate}
+
+\paragraph{Satz 7.2.}\label{7.2} Sei $\varphi:X\to\mathbf{P}_A^n$ ein $A$"~Morphismus, der zu einer invertierbaren Garbe $Z$ auf $X$ und globale Schnitte $s_0,\ldots,x_n\in\Gamma(X,Z)$ gehört, die $\mathcal{L}$ erzeugen.
+Dann ist $\varphi$ genau dann eine abgeschlossene Immersion, wenn die folgenden Bedingungen gelten:
+\begin{enumerate}[(i)]
+\item Alle offenen Teilmengen $X_i$ wie im Beweis von \hyperref[7.1]{Satz 7.1 (ii)} sind affin.
+\item Für alle $i$ ist die Abbildung $A[Y_0,\ldots,\widehat{Y_i},\ldots,Y_n]\to\Gamma(X_i,\mathcal{O}_{X_i}),\ Y_j\mapsto \frac{s_j}{s_i}$ surjektiv.
+\end{enumerate}
+
 \printindex

### 2015-04-17 Algebraische Geometrie/AlgebraischeGeometrie.tex
Fix labels

@@
 }
-\usepackage[all]{xy}
 \usepackage[left=1.3cm,right=1.3cm,top=2cm,bottom=2cm]{geometry}
@@
 
-\paragraph{Beispiel 5.34.}\label{3.34} Sei $X=\operatorname{Spec}(A)$ und $\mathcal{F}=\widetilde{M}$ für ein $A$"~Modul $M$.
+\paragraph{Beispiel 5.34.}\label{5.34} Sei $X=\operatorname{Spec}(A)$ und $\mathcal{F}=\widetilde{M}$ für ein $A$"~Modul $M$.
 Dann wird $\mathcal{F}$ von globalen Schnitten erzeugt; jedes Erzeugendensystem von $M$ als $A$"~Modul liefern solche Schnitte.
@@
 
-\paragraph{Beispiel 5.35.}\label{3.35} Sei $X=\operatorname{Proj}(S)$ mit einem graduierten Ring $S$, der von $S_1$ als $S_0$"~Algebra erzeugt wird.
+\paragraph{Beispiel 5.35.}\label{5.35} Sei $X=\operatorname{Proj}(S)$ mit einem graduierten Ring $S$, der von $S_1$ als $S_0$"~Algebra erzeugt wird.
 Dann liefern die Elemente aus $S_1$ globale Schnitte von $\mathcal{O}_X(1)$ und erzeugen diesen quasikohärenten Modul.

### 2015-04-18 Algebraische Geometrie/AlgebraischeGeometrie.tex
Beweis von 2.7.1

@@
 
-\paragraph{Satz 7.1.}\label{7.1} Sei $A$ ein Ring und $X$ ein Schema über $A$.
+\paragraph{Satz 7.1.}\label{7.1} Sei $A$ ein Ring, $X$ ein Schema über $A$ und $\mathbf{P}_A^n=\operatorname{Proj}A[x_0,\ldots,x_n]$.
 \begin{enumerate}[(i)]
 \item Ist $\varphi:X\to\mathbf{P}_A^n$ ein $A$-Morphismus.
-Dann ist $\varphi^\ast(\mathcal{O}(1))$ eine invertierbare Garbe auf $X$, die von globalen Schnitten $s_i=\varphi^\ast(x_i),\ i=0,\ldots,n$ erzeugt wird, wobei $x_i\in\Gamma(\operatorname{Proj}(A[x_0,\ldots,x_n]),\mathcal{O}(1))$.
+Dann ist $\varphi^\ast(\mathcal{O}(1))$ eine invertierbare Garbe auf $X$, die von globalen Schnitten $s_i=\varphi^\ast(x_i),\ i=0,\ldots,n$ erzeugt wird, wobei $x_i\in\Gamma(\mathbf{P}_A^n,\mathcal{O}(1))$.
 \item Sei $\mathcal{L}$ eine invertierbare Garbe auf $X$ und $s_0,\ldots,s_n\in\Gamma(X,\mathcal{L})$ globale Schnitte, die $\mathcal{L}$ erzeugen.
-Dann existiert ein eindeutig bestimmter $A$"~Morphismus $\varphi:X\to\mathbf{P}_A^n$ derart, dass $\mathcal{L}\cong\varphi^\ast(\mathcal{O}(1))$ mit $s_i=\varphi^\ast(x_i)$ für gewisse $x_0,\ldots,x_n\in X$ gibt.
+Dann existiert ein eindeutig bestimmter $A$"~Morphismus $\varphi:X\to\mathbf{P}_A^n$ derart, dass $\mathcal{L}\cong\varphi^\ast(\mathcal{O}(1))$ mit $s_i=\varphi^\ast(x_i)$.
+\end{enumerate}
+
+\paragraph{Beweis.} \begin{enumerate}[(i)]
+\item Nach \hyperref[5.35]{5.35} erzeugen die globalen Schnitte $x_0,\ldots,x_n$ die Garbe $\mathcal{O}(1)$.
+Ferner ist $\varphi^\ast(\mathcal{O}(1))$ invertierbar und werden von $\varphi^\ast(x_i)=s_i$ erzeugt.
+\item Setze $X_i=\{P\in X\mid (s_i)_P\not\in\mathfrak{m}_P\mathcal{L}_P\}\subset_\text{o}X$.
+Da die $s_i$ die Garbe $\mathcal{L}$ erzeugen, gilt $X=\bigcup_iX_i$.
+Sei $U_i=D_+(x_i)\cong\operatorname{Spec}A[Y_0,\ldots,\widehat{Y_i},\ldots,Y_n],\ Y_j=\frac{x_j}{x_i}$.
+Betrachte den Ringhomomorphismus $\phi:A[Y_0,\ldots,\widehat{Y_i},\ldots,Y_n]\to\Gamma(X_i,\mathcal{O}_{X_i}),\ Y_j\mapsto \frac{s_j}{s_i}$.
+Dieser ist wohldefiniert, da  $(s_i)_P\not\in \mathfrak{m}_P\mathcal{L}_P\cong\mathfrak{m}_P\mathcal{O}_{X,P}$ für alle $P\in X_i$ gilt und daher $\frac{s_j}{s_i}\in\Gamma(X_i,\mathcal{O}_{X_i})$.
+Sei $X_i\to U_i$ der zu $\phi$ gehörige $A$"~Morphismus von $A$"~Schemata.
+Verkleben ergibt ein $A$"~Morphismus $\varphi:X\to\mathbf{P}_A^n$.
+Nach Konstruktion ist $\varphi$ eindeutig bestimmt durch $\mathcal{L}\cong\varphi^\ast(\mathcal{O}(1))$ und $s_i=\varphi^\ast(x_i)$.\qed
 \end{enumerate}
@@
 \begin{enumerate}[(i)]
-\item Alle offenen Teilmengen $X_i$ wie im Beweis von \hyperref[7.1]{Satz 7.1 (ii)} sind affin.
+\item Alle offenen Teilmengen $X_i$ im Beweis von \hyperref[7.1]{Satz 7.1 (ii)} sind affin.
 \item Für alle $i$ ist die Abbildung $A[Y_0,\ldots,\widehat{Y_i},\ldots,Y_n]\to\Gamma(X_i,\mathcal{O}_{X_i}),\ Y_j\mapsto \frac{s_j}{s_i}$ surjektiv.

### 2015-04-21 Algebraische Geometrie/AlgebraischeGeometrie.tex
Fix typo

@@
 
-\paragraph{Bemerkung.} Ist $Y\subset X=\operatorname{Spec}(A)$ eine abgeschlossene Teilmenge, so existieren auf $Y$ viele abgeschlossene Unterschamestrukturen.
+\paragraph{Bemerkung.} Ist $Y\subset X=\operatorname{Spec}(A)$ eine abgeschlossene Teilmenge, so existieren auf $Y$ viele abgeschlossene Unterschemastrukturen.
 Wir werden später sehen, das sie genau den Idealen $\mathfrak{a}\subset A$ mit $Y=V(\mathfrak{a})$ entsprechen.

### 2015-04-21 Algebraische Geometrie/AlgebraischeGeometrie.tex
21.04.2015 w/o Beweis von 7.8

@@
 
+\paragraph{Beweis.} Sei $\varphi$ eine abgeschlossene Immersion.
+Dann ist $X_i=U_i\cap X$ ein abgeschlossenes Unterschema von $U_i=D_+(x_i)\subset\mathbf{P}_A^n=\operatorname{Proj}A[x_0,\ldots,x_n]$.
+Da $U_i$ affin ist, ist nach \hyperref[5.18]{Korollar 5.18} auch $X_i$ affin und wir erhalten einen surjektiven Ringhomomorphismus $A[Y_0,\ldots,\widehat{Y_i},\ldots,Y_n]\to \Gamma(X_i,\mathcal{O}_{X_i})$.
+
+Sei nun umgekehrt (i) und (ii) erfüllt.
+Dann ist $X_i$ nach (ii) ein abgeschlossenes Unterschema von $U_i$.
+Da $X=\bigcup_i X_i$ und $X_i=\varphi^{-1}(U_i)$, ist $X$ ein abgeschlossenes Unterschema von $\mathbf{P}_A^n$.\qed
+
+\paragraph{Definition 7.3.}\label{7.3} Eine invertierbare Garbe $\mathcal{L}$ auf einem noetherschen Schema $X$ heißt \textit{ampel}\index{$\mathcal{O}_X$-Modul!ampel}, falls für jede kohärente Garbe $\mathcal{F}$ auf $X$ ein $n_0>0$ existiert, so dass für alle $n\geq n_0$ der $\mathcal{O}_X$"~Modul $\mathcal{F}\otimes\mathcal{L}^n$ von globalen Schnitten erzeugt wird.
+
+\paragraph{Beispiel 7.4.}\label{7.4} Ist $X$ affin, so ist jede invertierbare Garbe ampel: Jede kohärente Garbe wird von globalen Schnitten endlich erzeugt, vgl. \hyperref[5.10]{Korollar 5.10}.
+
+\paragraph{Bemerkung 7.5.}\label{7.5} \hyperref[5.37]{Serres Theorem 5.37} besagt, dass jede sehr ample Garbe $\mathcal{L}$ auf einem Schema $X$ über einem noetherschen Ring ampel ist.
+Die Umkehrung ist im Allgemeinen falsch.
+
+\paragraph{Satz 7.6.}\label{7.6} Sei $\mathcal{L}$ eine invertierbare Garbe auf einem noetherschen Schema $X$. Dann sind folgende Aussagen äquivalent:
+\begin{enumerate}[(i)]
+\item $\mathcal{L}$ ist ampel.
+\item $\mathcal{L}^m$ ist ampel für alle $m>0$.
+\item $\mathcal{L}^m$ ist ampel für ein $m>0$.
+\end{enumerate}
+
+\paragraph{Beweis.} (i)$\implies$(ii)$\implies$(iii) ist klar.
+Sei nun $\mathcal{L}^m$ ampel für ein $m>0$ und $\mathcal{F}$ eine kohärente Garbe auf $X$.
+Nach Definition existiert ein $n_0>0$, so dass für alle $n\geq n_0$ die Garbe $\mathcal{F}\otimes(\mathcal{L}^m)^n$ von globalen Schnitten erzeugt wird.
+Betrachte die Garben $\mathcal{F}\otimes\mathcal{L}^k$ für $k=1,\ldots,m-1$.
+Diese sind kohärent, d.h.
+es gibt für jedes $k$ ein $n_k>0$, so dass für alle $n\geq n_k$ die Garbe $\mathcal{F}\otimes\mathcal{L}^k\otimes(\mathcal{L}^m)^n$ von globalen Schnitten erzeugt wird.
+Setzen wir nun $N=m\cdot\max\{n_k\mid k=1,\ldots,m-1\}$, so wird $\mathcal{F}\otimes\mathcal{L}^n$ von globalen Schnitten erzeugt für alle $n\geq N$.
+Somit ist $\mathcal{L}$ ampel.\qed
+
+\paragraph{Satz 7.7.}\label{7.7} Sei $X$ ein Schema von endlichem Typ über einem noetherschen Ring $A$ und $\mathcal{L}$ eine invertierbare Garbe auf $X$.
+Dann ist $\mathcal{L}$ genau dann ampel, wenn $\mathcal{L}^m$ sehr ampel bzgl.
+$\operatorname{Spec}(A)$ für ein $m>0$ ist.
+
+\paragraph{Lemma 7.8.}\label{7.8} Sei $X$ ein noethersches Schema, $U\subset_\text{o}X$ und $\mathcal{F}$ eine kohärente Garbe auf $U$.
+Dann existiert eine kohärente Garbe $\mathcal{F}'$ auf $X$ mit $\mathcal{F}'|_U=\mathcal{F}$.
+
+\paragraph{Beweis.} Sei o.B.d.A.
+$X$ affin.
+Sei $j:U\hookrightarrow X$ die Immersion.
+Dann ist $\mathcal{F}'=j_\ast\mathcal{F}$ nach \hyperref[5.13]{Satz 5.13 (iii)} quasikohärent.
+Ferner sind quasikohärente Garben auf einem noetherschen Schema Vereinigung seiner kohärenten Untergarben.
+Daher gilt:
+\[j_\ast\mathcal{F} =\mathcal{F}' = \bigcup_\lambda\mathcal{F}'_\lambda \]
+mit kohärenten Untergarben $\mathcal{F}'_\lambda$.
+Nach \hyperref[5.13]{Satz 5.13 (ii)} sind $j^\ast\mathcal{F}'_\lambda$ kohärent.
+Somit ist $\mathcal{F}=\bigcup_\lambda j^\ast\mathcal{F}'_\lambda$, wobei $j^\ast\mathcal{F}'_\lambda$ die kohärenten Untergarben von $\mathcal{F}$ durchläuft.
+Da $\mathcal{F}$ kohärent ist, gilt $\mathcal{F}=j^\ast\mathcal{F}'_\lambda$ für ein $\lambda$.
+Somit ist $\mathcal{F}'=\mathcal{F}'_\lambda$ kohärent.\qed
+
+\paragraph{Beweis von \hyperref[7.8]{Lemma 7.8}.}
+
 \printindex

### 2015-04-21 Algebraische Geometrie/AlgebraischeGeometrie.tex
Eine Richtung von Beweis 7.7

@@
 
-\paragraph{Beweis von \hyperref[7.8]{Lemma 7.8}.}
+\paragraph{Beweis von \hyperref[7.7]{Lemma 7.7}.} Sei zunächst $\mathcal{L}^m$ sehr ampel für ein $m>0$.
+Es existiert also eine Immersion $i:X\to\mathbf{P}_A^n$ mit $\mathcal{L}^m\cong i^\ast(\mathcal{O}(1))$.
+Sei $\overline{X}$ der Abschluss von $X$ in $\mathbf{P}_A^n$.
+Dann ist $\overline{X}$ ein projektives Schema über $\operatorname{Spec}(A)$.
+Nach \hyperref[7.5]{7.5} ist $\mathcal{O}_{\overline{X}}(1)$ ampel auf $\overline{X}$.
+Sei $\mathcal{F}$ eine kohärente Garbe auf $X$ und $\overline{\mathcal{F}}$ die kohärente Garbe in \hyperref[7.8]{Lemma 7.8} auf $\overline{X}$ mit $\overline{F}|_X=\mathcal{F}$.
+Wird $\overline{\mathcal{F}}\otimes\mathcal{O}_{\overline{X}}(\ell)$ von globalen Schnitten erzeugt, so ist dies auch für $\mathcal{F}\otimes\mathcal{O}_X(\ell)$ der Fall.
+Daher ist $\mathcal{L}^m$ ampel auf $X$ und nach \hyperref[7.6]{Satz 7.6} auch $\mathcal{L}$.
+Die andere Richtung ist schwierig.
 

### 2015-04-25 Algebraische Geometrie/AlgebraischeGeometrie.tex
23.04.2015 w/o Beweise

@@
 
+\paragraph{Definition.} Sei $X$ ein Schema.
+Ein quasikohärenter $\mathcal{O}_X$"~Modul $\mathcal{S}$ habe die Eigenschaft $(\star)$, wenn $X$ noethersch ist und wenn er eine Struktur einer graduierten $\mathcal{O}_X$"~Algebra trägt, d.h.
+$\mathcal{S}=\bigoplus_{d\geq 0}\mathcal{S}_d$ mit homogenen Teilen $\mathcal{S}_d$, so dass:
+\begin{enumerate}[(i)]
+\item $\mathcal{S}_0=\mathcal{O}_X$
+\item $\mathcal{S}_1$ ist ein kohärenter $\mathcal{O}_X$"~Modul.
+\item $\mathcal{S}$ wird lokal von $\mathcal{S}_1$ als $\mathcal{O}_X$"~Modul erzeugt.
+\end{enumerate}
+
+\paragraph{Definition 7.9.}\label{7.9} Sei $X$ ein Schema und $\mathcal{S}$ ein graduierter $\mathcal{O}_X$"~Modul mit $(\star)$.
+Sei $U=\operatorname{Spec}(A)\subset_\text{o}X$, $\mathcal{S}(U)=\Gamma(U,\mathcal{S}|_U)$ die graduierte $A$"~Algebra und $\pi_U:\operatorname{Proj}\mathcal{S}(U)\to U$ der kanonische Morphismus.
+Für $U,V\subset_\text{o}X$ affin gilt $\pi_U^{-1}(U\cap V)\cong\pi_V^{-1}(U\cap V)$.
+Verkleben von $\pi_U$ liefert ein Schema $\operatorname{\mathbf{Proj}}(\mathcal{S})$ und ein Morphismus $\pi:\operatorname{\mathbf{Proj}}(\mathcal{S})\to X$, so dass für alle affinen $U\subset_\text{o}X$ stets $\pi^{-1}(U)\cong\operatorname{Proj}\mathcal{S}(U)$ gilt.
+Die invertierbaren Garben $\mathcal{O}(1)$ auf $\operatorname{Proj}\mathcal{S}(U)$ lassen sich zu einer invertierbaren Garbe $\mathcal{O}(1)$ auf $\operatorname{\mathbf{Proj}}(\mathcal{S})$ verkleben.
+
+\paragraph{Beispiel 7.10.}\label{7.10} Sei $\mathcal{O}_X[T_0,\ldots,T_n]=\mathcal{S}$.
+Dann ist $\operatorname{\mathbf{Proj}}(\mathcal{S})=\mathbf{P}_X^n=\mathbf{P}_\mathbb{Z}^n\times X$ mit der getwisteten Garbe $\mathcal{O}(1)$ wie in \hyperref[5.30]{Definition 5.30}.
+
+\paragraph{Lemma 7.11.}\label{7.11} Sei $\mathcal{S}$ eine Garbe graduierter Algebren mit $(\star)$ auf $X$.
+Sei $\mathcal{L}$ eine invertierbare Garbe auf $X$ und $\mathcal{S}'$ die folgende Garbe graduierter Algebren:
+\[\mathcal{S}_d' =\mathcal{S}_1\otimes\mathcal{L}^d,\quad d\geq 0 \]
+Dann erfüllt $\mathcal{S}'$ ebenfalls $(\star)$ und es gibt einen natürlichen Isomorphismus:
+\[\begin{tikzcd}
+P'=\operatorname{\mathbf{Proj}}(\mathcal{S}')\ar[d, "\pi'"']\ar[r, "\varphi", "\cong"'] & P=\operatorname{\mathbf{Proj}}(\mathcal{S})\ar[dl, "\pi"]\\
+X&
+\end{tikzcd} \]
+Es gilt $\mathcal{O}_{P'}(1)\cong \varphi^\ast\mathcal{O}_P(1)\otimes\pi'^\ast\mathcal{L}$. $\mathcal{S}'$ wird auch mit $\mathcal{S}\ast\mathcal{L}$ bezeichnet.
+
+\paragraph{Satz 7.12.}\label{7.12} Sei $X$ und $\mathcal{S}$ mit $(\star)$. Sei $P=\operatorname{\mathbf{Proj}}(\mathcal{S})$ mit $\pi:P\to X$. Dann gilt:
+\begin{enumerate}[(i)]
+\item $\pi$ ist eigentlicher Morphismus.
+\item Besitzt $X$ eine ample Garbe $\mathcal{L}$, so ist $\pi$ projektiv und $\mathcal{O}_P(1)\otimes\pi^\ast\mathcal{L}^n$ ist eine sehr ample invertierbare Garbe von $P$ über $X$.
+\end{enumerate}
+
+\paragraph{Definition 7.13.}\label{7.13} \begin{enumerate}[(i)]
+\item Sei $A$ ein Ring und $M$ ein $A$"~Modul. Setze:
+\[T^0(M)=A,\quad T^n(M)= M^{\otimes n},\ n\geq 1 \]
+Dann ist $T(M)=\bigoplus_{n\geq 0}T^n(M)$ eine $A$"~Algebra mit $\otimes$ als Multiplikation, die sogenannte \textit{Tensor"~Algebra}\index{Tensor-Algebra} von $M$.
+\item Die \textit{symmetrische (Tensor-)Algebra}\index{Tensor-Algebra!symmetrisch} von $M$ ist definiert als der Quotient von $T(M)$ nach dem Ideal, das von den Elementen $x\otimes y-y\otimes x,\ x,y\in M$ erzeugt wird.
+$S(M)=\bigoplus_{n\geq 0}S^n(M)$ ist dann eine kommutative, graduierte $A$"~Algebra.
+$S^n(M)$ heißt \textit{$n$-tes symmetrisches Produkt}\index{symmetrisches Produkt} von $M$.
+\item Sei $X$ ein Schema und $\mathcal{F}$ ein $\mathcal{O}_X$"~Modul.
+Dann wird die Tensor-Algebra $T(\mathcal{F})$ bzw.
+die symmetrische Algebra $S(\mathcal{F})$ von $\mathcal{F}$ als assoziierte Garbe zur Prägarbe $U\mapsto T(\mathcal{F}(U))$ bzw.
+$U\mapsto S(\mathcal{F}(U))$ definiert.
+$T(\mathcal{F})$ und $S(\mathcal{F})$ sind $\mathcal{O}_X$"~Algebren.
+\end{enumerate}
+
+\paragraph{Beispiel.} Sei $M$ ein freier $A$"~Modul vom Rang $r$. Dann gilt $S(M)\cong A[X_1,\ldots,X_r]$.
+
+\paragraph{Definition 7.14.}\label{7.14} Sei $X$ ein noethersches Schema und $\mathcal{E}$ eine lokal freie, kohärente Garbe auf $X$.
+Sei $\mathcal{S}=S(\mathcal{E})$ die symmetrische Algebra von $\mathcal{E}$ und $\mathbf{P}(\mathcal{E})=\operatorname{\mathbf{Proj}}(\mathcal{S})$.
+Dann ist $\mathcal{S}$ eine Garbe von graduierten $\mathcal{O}_X$"~Algebren, die $(\star)$ erfüllt.
+$\pi:\mathbf{P}(\mathcal{E})\to X$ heißt \textit{projektives Vektorbündel}\index{Vektorbündel!projektiv} über $X$.
+
+\paragraph{Bemerkung.} Ist $\mathcal{E}$ frei vom Rang $n+1$ über $U\subset_\text{o}X$, so ist $\pi^{-1}(U)\cong\mathbf{P}_U^n$.
+
+\paragraph{Satz 7.15.}\label{7.15} Sei $\pi:\mathbf{P}(\mathcal{E})\to X$ ein projektives Vektorbündel über $X$. Dann gilt mit $\mathcal{S}=S(\mathcal{E})$:
+\begin{enumerate}[(i)]
+\item Ist Rang von $\mathcal{E}$ größer gleich $2$, so gibt es einen kanonischen Isomorphismus von graduierten $\mathcal{O}_X$"~Algebren:
+\[\mathcal{S}\cong\bigoplus_{\ell\in\mathbb{Z}}\pi_\ast\mathcal{O}(\ell)\]
+Insbesondere folgt $\pi_\ast(\mathcal{O}(\ell))=0$ für $\ell<0$ und $\pi_\ast(\mathcal{O})=\mathcal{O}_X$, $\pi_\ast(\mathcal{O}(1))=\mathcal{E}$.
+\item Es gibt einen natürliche surjektiven Morphismus:
+\[\pi^\ast\mathcal{E}\to\mathcal{O}(1) \]
+\end{enumerate}
+
+\paragraph{Satz 7.16.}\label{7.16} Sei $\pi:\mathbf{P}(\mathcal{E})\to X$ ein projektives Vektorbündel von $X$ und $g:Y\to X$ ein Morphismus.
+Dann gibt es genau dann einen Morphismus $f$ mit kommutativen Diagramm
+\[\begin{tikzcd}
+Y\rar[dashed, "f"]\dar["g"'] &\mathbf{P}(\mathcal{E})\ar[dl, "\pi"]\\
+X&
+\end{tikzcd}\]
+wenn es eine invertierbare Garbe $\mathcal{L}$ auf $Y$ zusammen mit einem surjektiven Garbenmorphismus $g^\ast\mathcal{E}\to\mathcal{L}$ gibt.
+
 \printindex

### 2015-05-02 Algebraische Geometrie/AlgebraischeGeometrie.tex
Aufblasung

@@
 
+\subsection{Ample Garben}
+
 \paragraph{Satz 7.1.}\label{7.1} Sei $A$ ein Ring, $X$ ein Schema über $A$ und $\mathbf{P}_A^n=\operatorname{Proj}A[x_0,\ldots,x_n]$.
@@
 
+\subsection{Projektive Vektorbündel}
+
 \paragraph{Definition.} Sei $X$ ein Schema.
@@
 
-\paragraph{Satz 7.16.}\label{7.16} Sei $\pi:\mathbf{P}(\mathcal{E})\to X$ ein projektives Vektorbündel von $X$ und $g:Y\to X$ ein Morphismus.
+\paragraph{Satz 7.16.}\label{7.16} \textit{(Universaleigenschaft des projektiven Vektorbündels)} Sei $\pi:\mathbf{P}(\mathcal{E})\to X$ ein projektives Vektorbündel von $X$ und $g:Y\to X$ ein Morphismus.
 Dann gibt es genau dann einen Morphismus $f$ mit kommutativen Diagramm
@@
 
+\subsection{Aufblasung}
+
+\paragraph{Definition 7.17.}\label{7.17} Sei $X$ ein noethersches Schema und $\mathcal{J}$ eine kohärente Idealgarbe auf $X$.
+Setze $\mathcal{J}^0=\mathcal{O}_X$ und $\mathcal{J}^d$ als das $d$-fache Produkt des Ideals $\mathcal{J}$.
+Sei $\mathcal{S}=\bigoplus_{d\geq 0}\mathcal{J}^d$.
+Dann erfüllen $X$ und $\mathcal{S}$ die Bedingung $(\star)$.
+$\widetilde{X}=\mathbf{Proj}(\mathcal{S})$ heißt die \textit{Aufblasung} von $X$ bzgl.
+$\mathcal{J}$.\index{Aufblasung}
+
+Ist $Y$ das abgeschlossene Unterschema von $X$, das zu $\mathcal{J}$ gehört, so sagen wir auch, dass $\widetilde{X}$ die Aufblasung von $X$ entlang $Y$ ist.
+
+\paragraph{Beispiel 7.18.}\label{7.18} Sei $X=\mathbf{A}^n_k$ und $P\in X$ der Nullpunkt, d.h.
+$X=\operatorname{Spec}(A),\ A=k[X_1,\ldots,X_n],\ I=(X_1,\ldots,X_n)$.
+Dann ist die Aufblasung von $X$ bei $P$ gegeben durch:
+\[\widetilde{X}=\operatorname{Proj}(S),\quad S=\bigoplus_{d\geq 0}I^d\]
+Betrachte die surjektive Abbildung $\psi:A[Y_1,\ldots,Y_n]\to S,\ Y_i\mapsto X_i$.
+Wir sehen, dass $\widetilde{X}$ isomorph zu dem abgeschlossenen Unterschema von $\mathbf{P}_A^{n-1}$ ist, das durch die homogenen Polynome in den $Y_i$, die $\ker(\psi)$ erzeugen, definiert ist:
+\[\ker(\psi)=\langle X_iY_j-X_jY_i\mid i,j=1,\ldots,n\rangle\]
+
+\paragraph{Definition 7.19.}\label{7.19} Sei $f:X\to Y$ ein Morphismus und $\mathcal{J}\subset\mathcal{O}_Y$ eine Idealgarbe auf $Y$.
+Wir definieren die \textit{inverse Idealgarbe}\index{Idealgarbe!invers} $\mathcal{J}'\subset\mathcal{O}_X$ auf $X$ wie folgt:
+
+$\mathcal{J}\hookrightarrow\mathcal{O}_Y\stackrel{f^\sharp}{\to}f_\ast\mathcal{O}_X$ liefert nach \hyperref[5.1]{5.1 (xi)} ein $\mathcal{O}_X$"~Modulmorphismus $f^\ast\mathcal{J}\to\mathcal{O}_X$.
+Setze $\mathcal{J}'=\operatorname{im}(f^\ast\mathcal{J}\to\mathcal{O}_X)$.
+
+Alternativ: Betrachte die Idealgarbe $f^{-1}\mathcal{J}$ auf $f^{-1}\mathcal{O}_Y$ und den Morphismus $f^{-1}\mathcal{O}_Y\to \mathcal{O}_Y$, der zu $f^\sharp$ adjungiert ist.
+Dann ist $\mathcal{J}'$ die vom Bild von $f^{-1}\mathcal{J}$ in $\mathcal{O}_X$ erzeugte Idealgarbe.
+Wir schreiben daher auch $\mathcal{J}'=f^{-1}\mathcal{J}\cdot\mathcal{O}_X$.
+
+\paragraph{Bemerkung.} Im Allgemeinen sind $f^\ast\mathcal{J}$ und $f^{-1}\mathcal{J}$ verschiedene Idealgarben.
+
+\paragraph{Satz 7.20.}\label{7.20} Sei $X$ ein noethersches Schema und $\mathcal{J}$ eine kohärente Idealgarbe auf $X$.
+Sei $\pi:\widetilde{X}\to X$ die Aufblasung von $X$ bzgl.
+$\mathcal{J}$.
+Dann gilt:
+\begin{enumerate}[(i)]
+\item Die inverse Idealgarbe $\widetilde{J}=\pi^{-1}\mathcal{J}\cdot\mathcal{O}_{\widetilde{X}}$ ist invertierbar auf $\mathcal{O}_{\widetilde{X}}$.
+\item Sei $Y$ das zu $\mathcal{J}$ gehörige abgeschlossene Unterschema und $U=X\setminus Y$. Dann ist $\pi:\pi^{-1}(U)\to U$ ein Isomorphismus.
+\end{enumerate}
+
+\paragraph{Beweis.}\begin{enumerate}[(i)]
+\item Sei $\widetilde{X}=\mathbf{Proj}(\mathcal{S}),\ \mathcal{S}=\bigoplus_{d\geq 0}\mathcal{J}^d$ und $V\subset_\text{o}X$ affin.
+Dann ist $\mathcal{O}(1)|_{\operatorname{Proj}\mathcal{S}(V)}$ die assoziierte Garbe zum graduierten $\mathcal{S}(V)$"~Modul $\mathcal{S}(V)(1)=\bigoplus_{d\geq 0}\mathcal{J}^{d+1}(V)=\mathcal{J}(V)\cdot \mathcal{S}(V)$ das von $\mathcal{J}(V)$ in $\mathcal{S}(V)$ erzeugte Ideal.
+Es folgt $\widetilde{\mathcal{J}}=\pi^{-1}\mathcal{J}\cdot\mathcal{O}_{\widetilde{X}}=\mathcal{O}_{\widetilde{X}}(1)$, also invertierbar.
+\item Es ist $\mathcal{J}|_U\cong\mathcal{O}_U$, da $0\to\mathcal{J}\to\mathcal{O}_X\stackrel{i^\sharp}{\to }i_\ast\mathcal{O}_Y\to 0$ exakt ist.
+Daher gilt $\pi^{-1}(U)=\operatorname{\mathbf{Proj}}\mathcal{O}_U[T]=U$.
+Beachte, dass für ein allgemeines $A$ stets $\operatorname{Proj}A[T]=D_+(T)=\operatorname{Spec}A[T]_{(T)}=\operatorname{Spec}(A)$ gilt.\qed
+\end{enumerate}
+
+\paragraph{Satz 7.21.}\label{7.21} \textit{(Universaleigenschaft des Aufblasens)} Sei $X$ ein noethersches Schema, $\mathcal{J}$ eine kohärente Idealgarbe auf $X$ und $\pi:\widetilde{X}\to X$ die Aufblasung von $X$ bzgl.
+$\mathcal{J}$.
+Ist $f:Z\to X$ ein Morphismus, so dass $f^{-1}\mathcal{J}\cdot\mathcal{O}_Z$ eine invertierbare Idealgarbe auf $Z$ ist, so existiert ein eindeutig bestimmter Morphismus $g:Z\to\widetilde{X}$ mit kommutativem Diagramm:
+\[\begin{tikzcd}
+Z\rar["g", dashed]\ar[rd, "f"'] & \widetilde{X}\dar["\pi"]\\
+& X
+\end{tikzcd} \]
+
+\paragraph{Lemma 7.22.}\label{7.22} Sei $X$ ein Schema, $f:\mathcal{L}\to\mathcal{M}$ ein surjektiver Morphismus invertierbarer Garben. Dann ist $f$ ein Isomorphismus.
+
+\paragraph{Beweis.} Siehe Bourbaki: Commutative Algebra II, §3.2 Korollar zu Proposition 6.\qed
+
+\paragraph{Korollar 7.23.}\label{7.23} Sei $f:Y\to X$ ein Morphismus noetherscher Schemata und $\mathcal{J}$ eine kohärente Idealgarbe auf $X$.
+Sei $\widetilde{X}$ die Aufblasung bzgl.
+$\mathcal{J}$ und $\widetilde{Y}$ die Aufblasung bzgl.
+$\mathcal{J}_Y=f^{-1}\mathcal{J}\cdot\mathcal{O}_Y$.
+Dann existiert ein eindeutig bestimmter Morphismus $\widetilde{f}$ mit kommutativem Diagramm:
+\[\begin{tikzcd}
+\widetilde{Y}\ar[rr, "\widetilde{f}"]\dar["\pi_Y"'] && \widetilde{X}\dar["\pi_X"]\\
+Y\ar[rr, "f"'] && X
+\end{tikzcd}\]
+Ist $f$ eine abgeschlossene Immersion, so auch $\widetilde{f}$.
+
+\paragraph{Beweis.} $\mathcal{J}_Y$ ist kohärent, da nach \hyperref[5.13]{Satz 5.13 (ii)} $f^\ast\mathcal{J}$ kohärent ist und nach \hyperref[5.12]{Satz 5.12 (ii)} auch $\operatorname{im}(f^\ast\mathcal{J}\to\mathcal{O}_Y)$.
+Nach \hyperref[7.20]{Satz 7.20} ist $\pi_Y^{-1}\mathcal{J}_Y\cdot\mathcal{O}_{\widetilde{Y}}$ eine invertierbare Garbe auf $\widetilde{Y}$.
+Nach \hyperref[7.21]{Satz 7.21} existiert $\widetilde{f}$ und ist eindeutig bestimmt.
+
+Sei nun $f$ eine abgeschlossene Immersion und:
+\[\widetilde{X}=\operatorname{\mathbf{Proj}}(\mathcal{S}),\quad \mathcal{S}=\bigoplus_{d\geq 0}\mathcal{J}^d,\qquad\widetilde{Y}=\mathbf{Proj}(\mathcal{S}_Y),\quad\mathcal{S}_Y=\bigoplus_{d\geq 0}\mathcal{J}_Y^d \]
+Ist $Y\hookrightarrow X$ eine abgeschlossene Immersion, so ist $\mathcal{J}\to\mathcal{J}_Y$ surjektiv und somit ist auch $\mathcal{S}\to\mathcal{S}_Y$ als Homomorphismus graduierter Ringe surjektiv.
+Daher ist $\widetilde{Y}\hookrightarrow\widetilde{X}$ eine abgeschlossene Immersion.\qed
+
+\paragraph{Definition 7.24.}\label{7.24} Ist $Y\hookrightarrow X$ eine abgeschlossene Immersion noetherscher Schemata.
+Dann heißt das abgeschlossene Unterschema $\widetilde{Y}$ von $\widetilde{X}$ auch \textit{strikte Transformation} von $Y$ bzgl.
+der Aufblasung $\pi:\widetilde{X}\to X$.\index{strikte Transformation}
+
 \printindex

### 2015-05-02 Algebraische Geometrie/AlgebraischeGeometrie.tex
Differentiale

@@
 
+\section{Differentiale}
+
+Sei $A$ ein Ring, $B$ eine $A$"~Algebra und $M$ ein $B$"~Modul.
+
+\paragraph{Definition 8.1.}\label{8.1} Eine \textit{$A$"~Derivation}\index{Derivation} von $B$ nach $M$ ist eine Abbildung $\mathrm{d}:B\to M$ mit:
+\begin{enumerate}[(i)]
+\item $\mathrm{d}$ ist additiv
+\item $\mathrm{d}(b\cdot b')=b\cdot \mathrm{d}(b')+ b'\cdot \mathrm{d}(b)$
+\item $\mathrm{d}(a)=0$ für alle $a\in A$
+\end{enumerate}
+Der $B$"~Modul aller $A$"~Derivationen von $B$ nach $M$ bezeichnen wir mit $D_A(B,M)$.
+
+\paragraph{Definition 8.2.}\label{8.2} Der $B$"~Modul $\Omega_{B/A}$ der \textit{relativen Differentialformen}\index{Differentialform!relativ} von $B$ über $A$ zusammen mit einer $A$"~Derivation $\mathrm{d}:B\to\Omega_{B/A}$ ist definiert durch folgende Universaleigenschaft:
+
+Für alle $B$"~Moduln $M$ und $A$"~Derivationen $\mathrm{d}':B\to M$ gibt es genau ein $B$"~Ho\-mo\-mor\-phis\-mus $f:\Omega_{B/A}\to M$, so dass folgendes Diagramm kommutiert:
+\[\begin{tikzcd}
+B\ar[r, "\mathrm{d}"]\ar[rd, "\mathrm{d}'"'] & \Omega_{B/A}\ar[d, "f", dashed]\\
+& M
+\end{tikzcd}\]
+Die Eindeutigkeit von $\Omega_{B/A}$ ist klar. Für die Existenz betrachte den freien $B$"~Modul $F$ mit Basis $\{d_b\mid b\in B\}$. Dann gilt:
+\[\Omega_{B/A}=F / \langle d_{b+b'}-d_b-d_{b'},\ d_{bb'}-bd_{b'}-b'd_b,\ d_a\rangle \]
+Zusammen mit der $A$"~Derivation $\mathrm{d}:B\to\Omega_{B/A},\ b\mapsto d_b$ erfüllt diese Konstruktion die Universaleigenschaft.
+
+\paragraph{Satz 8.3.}\label{8.3} Sei $B$ eine $A$"~Algebra und $f:B\otimes_A B\to B,\ b\otimes b'\mapsto bb'$.
+Betrachte $B\otimes_AB$ als $B$"~Modul via $b(b_1\otimes b_2)=(bb_1)\otimes b_2$.
+$B\otimes_A B$ wird zur $B$"~Algebra durch $(b_1\otimes b_1')(b_2\otimes b_2')=(b_1b_2)\otimes (b_1'b_2')$.
+Sei $I=\ker(f)$.
+Dann ist $I/I^2$ ein $B$"~Modul.
+Setze $\mathrm{d}:B\to I/I^2,\ \mathrm{d}b=1\otimes b-b\otimes 1\mod I^2$.
+Dann ist $\Omega_{B/A}= (I/I^2,\mathrm{d})$.
+
+\paragraph{Beweis.} Wir zeigen zunächst $I=\sum_{b\in B}B\mathrm{d}b$. Sei $\beta=\sum b_i\otimes c_i\in I$, also $\sum b_ic_i=0$. Dann gilt:
+\begin{align*}
+\beta &= \sum_i (b_i(1\otimes c_i-c_i\otimes 1) + b_ic_i\otimes 1)\\
+&= \sum_i b_i \mathrm{d}c_i + \Big(\sum_i b_ic_i\Big)\otimes 1 \in\sum_{b\in B} B\mathrm{d}b
+\end{align*}
+Wir zeigen nun, dass $\mathrm{d}$ eine $A$"~Derivation ist. $\mathrm{d}(A)=0$ und die Additivität ist klar.
+\begin{align*}
+\mathrm{d}b\mathrm{d}c &= (1\otimes b- b\otimes 1)(1\otimes c-c\otimes 1)\\
+&= bc\otimes 1 - b\otimes c- c\otimes b+1\otimes bc\\
+&=\mathrm{d}(bc) + 2bc\otimes 1- b(\mathrm{d}c+c\otimes 1) - c (\mathrm{d}b + b\otimes 1)\\
+&=\mathrm{d}(bc)-b\mathrm{d}c - c \mathrm{d}b
+\end{align*}
+Also ist $\mathrm{d}(bc)\equiv b\mathrm{d}c+c\mathrm{d}b\mod I^2$.
+Somit müssen wir noch zeigen, dass $(I/I^2,\mathrm{d})$ die Universaleigenschaft von $\Omega_{B/A}$ erfüllt.
+Sei $M$ ein $B$"~Modul und $\mathrm{D}:B\to M$ eine $A$"~Derivation.
+Zu zeigen ist die Existenz und Eindeutigkeit von einem $B$"~Homomorphismus $f:I/I^2\to M$ mit $\mathrm{D}=f\circ \mathrm{d}$.
+Da $I=\sum B\mathrm{d}b$ ist, folgt die Eindeutigkeit.
+
+Betrachte die \textit{triviale Erweiterung} $B\ast M$ von $B$ mit $M$, d.h.
+$B\oplus M$ mit der Multiplikation $(b_1,m_1)(b_2,m_2)=(b_1b_2,b_1m_2+b_2m_1)$.
+Dann ist $B\ast M$ ein Ring mit Einselement $(1,0)$.
+$B\ast M$ trägt eine $B$"~Modulstruktur via $b'(b,m)=(bb',bm)$, d.h.
+$B\ast M$ ist eine $B$"~Algebra.
+Wir haben kanonische $B$"~Homomorphismen:
+\begin{itemize}
+\item Die Einbettung $M\hookrightarrow B\ast M,\ m\mapsto (0,m)$
+\item Die Projektion $B\ast M\to B,\ (b,m)\mapsto b$
+\end{itemize}
+Betrachte den $B$"~Algebrenhomomorphismus $\varphi:B\otimes_AB\to B\ast M,\ x\otimes y\mapsto (xy,x\mathrm{D}y)$.
+Es gilt $\varphi(I)\subset M\subset B\ast M$, da $\varphi(\mathrm{d}b)=\varphi(1\otimes b- b\otimes 1)=(b,\mathrm{D}b)-(b,b\mathrm{D}1)=(0,\mathrm{D}b)$.
+Da $M^2=0$ in $B\ast M$, induziert $\varphi$ eine Abbildung $\phi: B\otimes_A B/I^2\to B\ast M$.
+Dann erfüllt $f=\phi|_{I/I^2}:I/I^2\to M$ die gewünschte Eigenschaft $f\circ\mathrm{d}=\mathrm{D}$.\qed
+
+\paragraph{Satz 8.4.}\label{8.4} Es gibt für jeden $B$"~Modul $M$ einen kanonischen $B$"~Modulisomorphismus:
+\[\operatorname{Hom}_B(\Omega_{B/A},M)\to\operatorname{Der}_A(B,M),\ f\mapsto f\circ\mathrm{d} \]
+
+\paragraph{Beweis.} Folgt direkt aus der Universaleigenschaft.\qed
+
+\paragraph{Beispiel 8.5.}\label{8.5} Sei $B=A[X_1,\ldots,X_n]$.
+Dann ist $\Omega_{B/A}=\bigoplus_{i=1}^n B\mathrm{d}X_i$ ein freier $B$"~Modul vom Rang $n$.
+Allgemein gilt: Wird $B$ als $A$"~Algebra von $y_1,\ldots,y_m$ erzeugt, so ist $\Omega_{B/A}=\sum_{i=1}^m B\mathrm{d}y_1$, da:
+\[\mathrm{d}\Big(\prod_i y_i^{n_i}\Big) = \sum_i n_i\prod_{j\neq i} y_j^{n_j}y_i^{n_i - 1}\mathrm{d}y_i \]
+Angenommen, $\sum_i b_i\mathrm{d}X_i=0$.
+Betrachte die Derivationen $\frac{\partial}{\partial X_i}\in\operatorname{Der}_A(B)$.
+Nach \hyperref[8.4]{Satz 8.4} existiert für jedes $k$ ein $f_k\in\operatorname{Hom}_B(\Omega_{B/A},B)$ mit $\frac{\partial}{\partial X_k}=f_k\circ\mathrm{d}$.
+Es gilt für alle $k$:
+\[ 0= f_k\Big(\sum_{i=1}^n b_i\mathrm{d}X_i\Big)= \sum_{i=1}^n b_i f_k(\mathrm{d}X_i) = \sum_{i=1}^n b_i \frac{\partial X_i}{\partial X_k} = \sum_{i=1}^n b_i\delta_{ik}=b_k \]
+Somit ist $\Omega_{B/A}=\bigoplus_{i=1}^nB\mathrm{d}X_i$
+
+\paragraph{Satz 8.6.}\label{8.6} \begin{enumerate}[(i)]
+\item Sei $S$ eine multiplikative Teilmenge der $A$"~Algebra $B$. Dann gilt:
+\[S^{-1}\Omega_{B/A}\cong\Omega_{S^{-1}B/A} \]
+\item Seien $A',B$ $A$"~Algebren und $B'=B\otimes_AA'$. Dann gilt:
+\[\Omega_{B'/A'}\cong\Omega_{B/A}\otimes_BB' \]
+\end{enumerate}
+
+\paragraph{Satz 8.7.}\label{8.7} Seien $\phi:A\to B,\ \psi:B\to C$ Ringhomomorphismen. Dann gilt:
+\begin{enumerate}[(i)]
+\item \textit{(Erste fundamentale exakte Sequenz)} Es gibt eine kurze exakte Sequenz:
+\[\begin{tikzcd}
+\Omega_{B/A}\otimes_BC \ar[r, "v"] & \Omega_{C/A} \ar[r, "u"] & \Omega_{C/B}\rar & 0
+\end{tikzcd} \]
+\item $v$ ist injektiv und die Sequenz zerfällt, genau dann wenn für alle $C$"~Moduln $M$ und jedes $\mathrm{D}\in \operatorname{Der}_A(B,M)$ zu einer Derivation aus $\operatorname{Der}_A(C,M)$ fortgesetzt werden kann.
+\end{enumerate}
+
 \printindex

### 2015-05-08 Algebraische Geometrie/AlgebraischeGeometrie.tex
1. + 2. fundamentale Sequenz

@@
 \begin{enumerate}[(i)]
-\item \textit{(Erste fundamentale exakte Sequenz)} Es gibt eine kurze exakte Sequenz:
+\item \textit{(Erste fundamentale exakte Sequenz)}\index{Fundamentale exakte Sequenz} Es gibt eine kurze exakte Sequenz:
 \[\begin{tikzcd}
@@
 \end{enumerate}
+
+\paragraph{Beweis.} Definiere $u$ und $v$ durch:
+\[v(\mathrm{d}b\otimes c)= c\mathrm{d}\psi(b),\quad u(c\mathrm{d}c') = c\mathrm{d}c',\qquad b\in B,\ c,c'\in C \]
+Äquivalent zu (i) bzw. (ii) ist die Exaktheit der $4$-Term- bzw. $5$-Termsequenzen:
+\[0\to\operatorname{Hom}_C(\Omega_{C/B},M)\to\operatorname{Hom}_C(\Omega_{C/A},M)\to\operatorname{Hom}_C(\Omega_{B/A}\otimes_B C,M)\dashrightarrow 0  \]
+Da $\operatorname{Hom}_C(\Omega_{B/A}\otimes_B C,M)\cong\operatorname{Hom}_B(\Omega_{B/A},M)$, ist nach \hyperref[8.4]{Satz 8.4} ist die Exaktheit nachzuweisen von:
+\[0\to\operatorname{Der}_B(C,M)\to\operatorname{Der}_A(C,M)\to\operatorname{Der}_A(B,M)\dashrightarrow 0 \]
+Die vorletzte Abbildung ist gegeben durch $(\delta:C\to M)\mapsto (\delta\circ\psi: B\to M)$.
+Ohne die $0$ rechts ist diese Folge per Definition exakt.
+Mit der $0$ rechts ist die Folge genau dann exakt, wenn die Forderung in (ii) gilt.\qed
+
+\paragraph{Satz 8.9.}\label{8.9}\textit{(Zweite fundamentale exakte Sequenz)}\index{Fundamentale exakte Sequenz} Sei $B$ eine $A$"~Algebra, $I$ ein Ideal in $B$ und $C=B/I$.
+Es gibt eine kanonische, exakte Folge von $C$"~Moduln:
+\[\begin{tikzcd}
+I/I^2\rar["\delta"] & \Omega_{B/A}\otimes_BC\rar["v"] & \Omega_{C/A}\rar & 0
+\end{tikzcd} \]
+wobei $v(\mathrm{d}b\otimes c)= c\mathrm{d}\overline{b}$ und $\delta$ die durch die Abbildung $I\to\Omega_{B/A}\otimes_BC,\ x\mapsto \mathrm{d}x\otimes 1$ induzierte $C$"~lineare Abbildung ist.
+
+\paragraph{Beweis.} Für $x,x'\in I$ gilt:
+\[\mathrm{d}(xx')\otimes 1 = \mathrm{d}x\otimes x' + \mathrm{d}x'\otimes x =0 \]
+Daher ist $\delta$ wohldefiniert. Nun ist für alle $C$"~Moduln $M$ die folgenden Folgen exakt:
+\[\begin{tikzcd}
+0 \rar & \operatorname{Hom}_C(\Omega_{C/A},M)\dar["\cong"] \rar & \operatorname{Hom}_C(\Omega_{B/A}\otimes_BC,M)\dar["\cong"] \rar & \operatorname{Hom}_C(I/I^2,M)\dar["\cong"]\\
+0 \rar & \operatorname{Der}_A(B/I,M) \rar & \operatorname{Der}_A(B,M) \rar["\varphi"'] & \operatorname{Hom}_B(I,M)
+\end{tikzcd} \]
+wobei $\varphi(D:B\to M) = (D|_I:I\to M)$.\qed
 

### 2015-05-08 Algebraische Geometrie/AlgebraischeGeometrie.tex
Beispiel 8.8
Bis 8.12

@@
 
+\paragraph{Beispiel 8.8.}\label{8.8} Sei $B$ eine $A$"~Algebra und $C=B[x_1,\ldots,x_n]$.
+Sei $M$ ein beliebiger $C$"~Modul und $D\in\operatorname{Der}_A(B,M)$.
+Sei $\widetilde{D}:C\to M$ die einem Polynom $F$, den Polynom $F^D$ zuordnet, wobei $F^D$ das Polynom ist, das man durch Anwenden von $D$ auf die Koeffizienten von $F$ erhält.
+Dann ist $\widetilde{D}$ eine Derivation mit $\widetilde{D}|_B=D$.
+Nach \hyperref[8.7]{Satz 8.7 (ii)} gilt:
+\[\Omega_{C/A}\cong (\Omega_{B/A}\otimes_BC)\oplus C\mathrm{d}x_1\oplus\cdots\oplus C\mathrm{d}x_n \]
+
 \paragraph{Satz 8.9.}\label{8.9}\textit{(Zweite fundamentale exakte Sequenz)}\index{Fundamentale exakte Sequenz} Sei $B$ eine $A$"~Algebra, $I$ ein Ideal in $B$ und $C=B/I$.
@@
 
+\paragraph{Korollar 8.10.}\label{8.10} Ist $B$ eine endlich erzeugte $A$"~Algebra oder Lokalisierung einer endlich erzeugten $A$"~Algebra, so ist auch $\Omega_{B/A}$ endlich erzeugt.
+
+\paragraph{Beweis.} Es ist $B \cong A[x_1,\ldots,x_n]/I$ für ein Ideal $I$.
+Wegen \hyperref[8.5]{Beispiel 8.5} ist $\Omega_{A[x_1,\ldots,x_n]/A}$ ein endlich erzeugter $A[x_1,\ldots,x_n]$"~Modul.
+Nach \hyperref[8.9]{Satz 8.9} ist $\Omega_{B/A}$ ein endlich erzeugter $B$"~Modul.
+Die Aussage über die Lokalisierung folgt aus \hyperref[8.6]{Satz 8.6 (i)}.\qed
+
+\paragraph{Definition 8.11.}\label{8.11} Eine Körpererweiterung $K/k$ heißt \textit{separabel erzeugt}\index{separabel erzeugt}, falls es eine Transzendenzbasis $\{x_\lambda\}_{\lambda\in\Lambda}$ für $K/k$ gibt, so dass $K/k[x_\lambda\mid \lambda\in\Lambda]$ separabel ist.
+
+\paragraph{Satz 8.12.}\label{8.12} Sei $K/k$ eine endlich erzeugte Körpererweiterung. Dann gilt:
+\begin{enumerate}[(i)]
+\item $\dim_K\Omega_{K/k}\geq \operatorname{trdeg}(K/k)$
+\item Gleichheit in (i) gilt genau dann, wenn $K/k$ separabel erzeugt ist.
+\end{enumerate}
+
 \printindex

### 2015-05-08 Algebraische Geometrie/AlgebraischeGeometrie.tex
05.05.2015

@@
 \begin{enumerate}[(i)]
-\item \textit{(Erste fundamentale exakte Sequenz)}\index{Fundamentale exakte Sequenz} Es gibt eine kurze exakte Sequenz:
+\item \textit{(Erste fundamentale exakte Sequenz)}\index{Fundamentale exakte Sequenz} Es gibt eine exakte Sequenz:
 \[\begin{tikzcd}
@@
 
-\paragraph{Korollar 8.10.}\label{8.10} Ist $B$ eine endlich erzeugte $A$"~Algebra oder Lokalisierung einer endlich erzeugten $A$"~Algebra, so ist auch $\Omega_{B/A}$ endlich erzeugt.
+\paragraph{Korollar 8.10.}\label{8.10} Ist $B$ eine endlich erzeugte $A$"~Algebra oder Lokalisierung einer endlich erzeugten $A$"~Algebra, so ist $\Omega_{B/A}$ endlich erzeugter $B$"~Modul.
 
@@
 
+\paragraph{Korollar 8.13.}\label{8.13} Sei $K/k$ eine endliche Erweiterung. Dann gilt:
+\[\Omega_{K/k}=0\iff K/k\text{ ist separabel} \]
+
+\paragraph{Satz 8.14.}\label{8.14} Sei $B$ ein lokaler Ring mit Maximalideal $\mathfrak{m}$ und Restklassenkörper $k$.
+Ferner gebe es einen Schnitt von $B\twoheadrightarrow k$, also ist $B$ eine $k$"~Algebra.
+Dann ist die Abbildung $\delta$ in der \hyperref[8.9]{zweiten fundamentalen Sequenz} ein Isomorphismus:
+\[\delta:\mathfrak{m}/\mathfrak{m}^2\to \Omega_{B/k}\otimes_B k \]
+
+\paragraph{Beweis.} Es ist $\operatorname{coker}(\delta)=\Omega_{k/k}=0$. Die Injektivität von $\delta$ ist äquivalent zur Surjektivität von:
+\[\delta^\ast :\operatorname{Hom}_k(\Omega_{B/k}\otimes_B k,k)\to\operatorname{Hom}_k(\mathfrak{m}/\mathfrak{m}^2,k) \]
+Nun ist $\operatorname{Hom}_k(\Omega_{B/k}\otimes_B k,k) = \operatorname{Hom}_B(\Omega_{B/k},k)=\operatorname{Der}_k(B,k)$.
+Die Abbildung $\operatorname{Der}_k(B,k)\to\operatorname{Hom}_k(\mathfrak{m}/\mathfrak{m}^2,k),\ (d:B\to k)\mapsto (d|_\mathfrak{m}:\mathfrak{m}\to k)$ kommutiert mit $\delta^\ast$.
+Beachte, dass für $d\in\operatorname{Der}_k(B,k)$ stets $d(\mathfrak{m}^2)=0\in k$ gilt.
+
+Sei nun $h\in\operatorname{Hom}_k(\mathfrak{m}/\mathfrak{m}^2,k)$ und $b\in B$.
+Wegen des Schnitts $k\to B$, hat $b$ eine eindeutige Darstellung $b=\lambda+c,\ \lambda\in k,\ c\in\mathfrak{m}$.
+Setze $d:B\to k,\ d(b) =h(\overline{c})$.
+Dann ist $d(k)=0$, $d$ additiv und es gilt:
+\begin{align*}
+d(bb') &= d(\lambda\lambda'+\lambda c'+\lambda' c+cc')\\
+&= h(\overline{\lambda c'+\lambda' c})\\
+&= \lambda h(\overline{c'}) + \lambda' h(\overline{c})\\
+&= \lambda d(b')+\lambda' d(b)= bd(b')+b'd(b)\in k
+\end{align*}
+für alle $b=\lambda+c,\ b'=\lambda'+c'\in B$. Es gilt $\delta^\ast (d)=h$.\qed
+
+\paragraph{Satz 8.15.}\label{8.15} Sei $B$ ein lokaler Ring und es gebe einen Schnitt von $B\to k$.
+Ferner sei der Restklassenkörper $k$ vollkommen.
+Sei $B$ die Lokalisierung einer endlich erzeugten $k$"~Algebra.
+Dann ist $\Omega_{B/k}$ genau dann ein freier $B$"~Modul vom Rang $\dim(B)$, wenn $B$ regulär ist.
+
+\paragraph{Beweis.} Sei $\Omega_{B/k}$ frei vom Rang $\dim(B)$. Nach \hyperref[8.14]{Satz 8.14} ist $\dim_k\mathfrak{m}/\mathfrak{m}^2=\dim(B)$, also ist $B$ per Definition regulär.
+
+Sei umgekehrt $B$ regulär von Dimension $r$. Nach \hyperref[8.14]{Satz 8.14} gilt:
+\[r =\dim_k\mathfrak{m}/\mathfrak{m}^2=\dim_k\Omega_{B/k}\otimes_B k \]
+Sei $K$ der Quotientenkörper von $B$.
+Nach \hyperref[8.6]{Satz 8.6} ist $\Omega_{B/k}\otimes_B K=\Omega_{K/k}$.
+Nun ist $K/k$ separabel erzeugt, da $k$ vollkommen (siehe z.B.
+Matsumura Chapter 10, Corollary p.
+194).
+Nach \hyperref[8.12]{Satz 8.12} gilt:
+\[\dim_K\Omega_{K/k}=\operatorname{trdeg}(K/k)=\dim(B) = r \]
+Nach \hyperref[8.10]{Korollar 8.10} ist $\Omega_{B/k}$ endlich erzeugter $B$"~Modul.\qed
+
+\paragraph{Lemma 8.16.}\label{8.16} Sei $A$ ein noetherscher, lokaler, nullteilerfreier Ring mit Restklassenkörper $k$ und Quotientenkörper $K$. Sei $M$ ein endlich erzeugter $A$"~Modul mit:
+\[\dim_k(M\otimes_Ak)=\dim_K(M\otimes_AK)=r \]
+Dann ist $M$ frei vom Rang $r$.
+
+\paragraph{Beweis.} Wegen $\dim_k(M\otimes_Ak)=r$, gibt es nach Nakayama eine Surjektion $\varphi:A^r\to M$. Setze $N=\ker(\varphi)$. Da $K$ flach über $A$ ist, haben wir eine kurze exakte Folge:
+\[0\longrightarrow N\otimes_AK\longrightarrow K^r\longrightarrow M\otimes_AK\longrightarrow 0 \]
+Da $\dim_K(M\otimes_AK)=\dim_KK^r$, folgt $N\otimes_AK=0$, also $N=0$, da $N\subset A^r$ und $A$ torsionsfrei ist.\qed
+
 \printindex

### 2015-05-08 Algebraische Geometrie/AlgebraischeGeometrie.tex
Bis 8.18

@@
 
+\subsection{Differentialgarben}
+
+Sei $f:X\to Y$ ein Morphismus von Schemata und $\Delta:X\to X\times_YX$ die Diagonale.
+$\Delta$ ist nach \hyperref[3.30]{Satz 3.30} eine Immersion und liefert ein Isomorphismus auf $\Delta(X)$, wobei $\Delta(X)$ ein abgeschlossenes Unterschema eines offenen Unterschema $W\subset X\times_YX$ ist.
+
+\paragraph{Definition 8.17.}\label{8.17} Sei $\mathcal{J}$ die Idealgarbe von $\Delta(X)$ in $W$. Dann heißt
+\[\Omega_{X/Y}=\Delta^\ast(\mathcal{J}/\mathcal{J}^2) \]
+die Garbe der \textit{relativen Differentiale}\index{Differentialform!relativ} von $X$ über $Y$.
+
+\paragraph{Bemerkung 8.18.}\label{8.18}\begin{enumerate}[(i)]
+\item $\mathcal{J}/\mathcal{J}^2$ ist ein $\mathcal{O}_{\Delta(X)}$"~Modul und somit ist über den Isomorphismus $X\to\Delta(X)$ die Modulgarbe $\Omega_{X/Y}$ ein $\mathcal{O}_X$"~Modul.
+Wegen \hyperref[5.13]{Satz 5.13 (i)} ist $\Omega_{X/Y}$ quasikohärent.
+Ist $Y$ noethersch und $f:X\to Y$ von endlichem Typ, und somit $X$ und $X\times_YX$ noethersch, dann ist $\Omega_{X/Y}$ kohärent.
+\item Sei $U=\operatorname{Spec}(A)\subset_\text{o}Y$ und $V=\operatorname{Spec}(B)\subset_\text{o}X$ mit $f(V)\subset U$.
+Dann ist $V\times_UV$ affin offen in $X\times_YX$ mit $V\times_UV= \operatorname{Spec}(B\otimes_AB)$.
+Es ist
+\[\Delta(X)\cap V\times_UV = \operatorname{Spec}(B\otimes_AB/\ker(B\otimes_AB\to B)) \]
+ein abgeschlossenes Unterschema von $V\times_UV$. Sei $I=\ker(B\otimes_AB\to B)$, also gilt:
+\[\mathcal{J}/\mathcal{J}^2|_{\Delta(X)\cap V\times_UV} = \widetilde{I/I^2} \]
+Nach \hyperref[8.3]{Satz 8.3} folgt $\Omega_{V/U}=\widetilde{\Omega_{B/A}}$.
+Somit ist \hyperref[8.17]{Definitino 8.17} verträglich mit dem affinen Fall.
+Ist $X=\bigcup V$ eine Überdeckung durch affin offenen $V$ und $Y=\bigcup U$ eine Überdeckung durch affin offenen $U$, so liefert die Verklebung von $\widetilde{\Omega_{B/A}},\ \mathrm{d}:B\to\Omega_{B/A}$ gerade $\Omega_{X/Y},\ \mathrm{d}:\mathcal{O}_X\to\Omega_{X/Y}$.
+$\mathrm{d}$ ist ein Garbenmorphismus und ist in jedem Punkt eine Derivation von lokalen Ringen.
+\end{enumerate}
+
 \printindex

### 2015-05-09 Algebraische Geometrie/AlgebraischeGeometrie.tex
Typo
07.05.2015

@@
 Die Eindeutigkeit von $\Omega_{B/A}$ ist klar. Für die Existenz betrachte den freien $B$"~Modul $F$ mit Basis $\{d_b\mid b\in B\}$. Dann gilt:
-\[\Omega_{B/A}=F / \langle d_{b+b'}-d_b-d_{b'},\ d_{bb'}-bd_{b'}-b'd_b,\ d_a\rangle \]
+\[\Omega_{B/A}=F / \langle d_{b+b'}-d_b-d_{b'},\ d_{bb'}-bd_{b'}-b'd_b,\ d_a\mid b,b'\in B,\ a\in A\rangle \]
 Zusammen mit der $A$"~Derivation $\mathrm{d}:B\to\Omega_{B/A},\ b\mapsto d_b$ erfüllt diese Konstruktion die Universaleigenschaft.
@@
 Äquivalent zu (i) bzw. (ii) ist die Exaktheit der $4$-Term- bzw. $5$-Termsequenzen:
-\[0\to\operatorname{Hom}_C(\Omega_{C/B},M)\to\operatorname{Hom}_C(\Omega_{C/A},M)\to\operatorname{Hom}_C(\Omega_{B/A}\otimes_B C,M)\dashrightarrow 0  \]
+\[0\longrightarrow \operatorname{Hom}_C(\Omega_{C/B},M)\longrightarrow \operatorname{Hom}_C(\Omega_{C/A},M)\longrightarrow \operatorname{Hom}_C(\Omega_{B/A}\otimes_B C,M)\dashrightarrow 0  \]
 Da $\operatorname{Hom}_C(\Omega_{B/A}\otimes_B C,M)\cong\operatorname{Hom}_B(\Omega_{B/A},M)$, ist nach \hyperref[8.4]{Satz 8.4} ist die Exaktheit nachzuweisen von:
-\[0\to\operatorname{Der}_B(C,M)\to\operatorname{Der}_A(C,M)\to\operatorname{Der}_A(B,M)\dashrightarrow 0 \]
+\[0\longrightarrow \operatorname{Der}_B(C,M)\longrightarrow \operatorname{Der}_A(C,M)\longrightarrow \operatorname{Der}_A(B,M)\dashrightarrow 0 \]
 Die vorletzte Abbildung ist gegeben durch $(\delta:C\to M)\mapsto (\delta\circ\psi: B\to M)$.
@@
 Nach \hyperref[8.3]{Satz 8.3} folgt $\Omega_{V/U}=\widetilde{\Omega_{B/A}}$.
-Somit ist \hyperref[8.17]{Definitino 8.17} verträglich mit dem affinen Fall.
+Somit ist \hyperref[8.17]{Definition 8.17} verträglich mit dem affinen Fall.
 Ist $X=\bigcup V$ eine Überdeckung durch affin offenen $V$ und $Y=\bigcup U$ eine Überdeckung durch affin offenen $U$, so liefert die Verklebung von $\widetilde{\Omega_{B/A}},\ \mathrm{d}:B\to\Omega_{B/A}$ gerade $\Omega_{X/Y},\ \mathrm{d}:\mathcal{O}_X\to\Omega_{X/Y}$.
@@
 
+\paragraph{Satz 8.19.}\label{8.19} Sei $f:X\to Y$ ein Morphismus und $g:Y'\to Y$ ein weiterer Morphismus. Betrachte den Basiswechsel:
+\[\begin{tikzcd}
+X\times_YY' \rar["f'"]\dar["g'"'] & Y'\dar["g"]\\
+X\rar["f"] & Y
+\end{tikzcd} \]
+Dann gilt $\Omega_{X'/Y'}\cong (g')^\ast(\Omega_{X/Y})$.
+
+\paragraph{Beweis.} Folgt aus \hyperref[8.18]{Bemerkung 8.18 (ii)} und \hyperref[8.6]{Satz 8.6 (ii)}.\qed
+
+\paragraph{Satz 8.20.}\label{8.20} Seien $f:X\to Y$ und $g:Y\to Z$ Morphismen. Dann gibt es eine exakte Garbensequenz auf $X$:
+\[f^\ast\Omega_{Y/Z}\longrightarrow\Omega_{X/Z}\longrightarrow\Omega_{X/Y}\longrightarrow 0 \]
+
+\paragraph{Beweis.} Es reicht, die Exaktheit lokal zu überprüfen. Dies folgt aus der \hyperref[8.7]{ersten fundamentalen exakten Sequenz}.\qed
+
+\paragraph{Satz 8.21.}\label{8.21} Sei $f:X\to Y$ ein Morphismus und $Z$ ein abgeschlossenes Unterschema von $X$ zum Ideal $\mathcal{J}$. Dann gibt es eine exakte Sequenz von Garben auf $Z$:
+\[\mathcal{J}/\mathcal{J}^2 \longrightarrow \Omega_{X/Y}\otimes_{\mathcal{O}_X}\mathcal{O}_Z\longrightarrow\Omega_{Z/Y}\longrightarrow 0 \]
+
+\paragraph{Beweis.} Folgt aus der \hyperref[8.9]{zweiten fundamentalen exakten Sequenz}.\qed
+
+\paragraph{Satz 8.22.}\label{8.22} Sei $X=\mathbf{A}_Y^n$.
+Dann ist $\Omega_{X/Y}$ ein freier $\mathcal{O}_X$"~Modul vom Rang $n$, erzeugt von globalen Schnitten $\mathrm{d}x_1,\ldots,\mathrm{d}x_n$, wobei $x_1,\ldots,x_n$ die affinen Koordinaten von $\mathbf{A}_Y^n$ sind.
+
+\paragraph{Beweis.} Folgt aus \hyperref[8.5]{Beispiel 8.5}.\qed
+
+\paragraph{Satz 8.23.}\label{8.23} Sei $A$ ein Ring und $Y=\operatorname{Spec}(A)$. Sei $X=\mathbf{P}_A^n$. Dann gibt es eine exakte Sequenz von Garben auf $X$:
+\[0\longrightarrow\Omega_{X/Y}\longrightarrow \bigoplus_{i=0}^n\mathcal{O}_X(-1)\longrightarrow \mathcal{O}_X\longrightarrow 0 \]
+
+\paragraph{Beweis.} Sei $S=A[x_0,\ldots,x_n]$ der homogene Koordinatenring von $X$ und setze:
+\[E=\bigoplus_{i=0}^n S(-1)e_i\]
+als den graduierten $S$"~Modul mit Basis $e_0,\ldots,e_n$ im Grad $1$.
+Betrachte den $S$"~Mo\-dul\-ho\-mo\-mor\-phis\-mus $\varphi:E\to S,\ e_i\mapsto x_i$ im Grad $0$ und $M=\ker(\varphi)$.
+Wir erhalten eine exakte Garbensequenz:
+\[0\longrightarrow\widetilde{M}\longrightarrow\bigoplus_{i=0}^n\mathcal{O}_X(-1)\stackrel{\tilde{\varphi}}{\longrightarrow}\mathcal{O}_X\longrightarrow 0 \]
+$\varphi:E\to S$ ist in allen Graden größer gleich $1$ surjektiv, also ist $\tilde{\varphi}$ surjektiv. Wir zeigen nun $\widetilde{M}=\Omega_{X/Y}$.
+
+Lokalisierung nach $x_i$ liefert eine Surjektion von freien $S_{x_i}$"~Moduln $E_{x_i}\to S_{x_i}$.
+$M_{x_i}$ ist frei vom Rang $n$ mit Basis $\{e_j-\frac{x_j}{x_i}e_i\mid j\neq i\}$.
+Multiplikation mit $\frac{1}{x_i}$ liefert Elemente vom Grad $0$ in $M_{x_i}$.
+Also ist $\widetilde{M}|_{D_+(x_i)}$ ein freier $\mathcal{O}_{D_+(x_i)}$"~Modul, der von den globalen Schnitten $\{\frac{1}{x_i}e_j-\frac{x_j}{x_i^2}e_i\mid j\neq i \}$ erzeugt wird.
+Definiere $\varphi_i:\Omega_{X/Y}|_{D_+(x_i)}\to\widetilde{M}|_{D_+(x_i)}$ wie folgt:
+
+Es ist $D_+(x_i)=\operatorname{Spec}A[\frac{x_0}{x_i},\ldots,\widehat{\frac{x_i}{x_i}},\ldots,\frac{x_n}{x_i}]$. Nach \hyperref[8.22]{Satz 8.22} gilt:
+\[\Omega_{X/Y}|_{D_+(x_i)}=\bigoplus_{j\neq i}\mathcal{O}_{D_+(x_i)}\mathrm{d}\frac{x_j}{x_i} \]
+Setze nun $\varphi_i(\mathrm{d}\frac{x_j}{x_i})=\frac{1}{x_i^2}(x_ie_j-x_je_i)$.
+$\varphi_i$ ist ein Isomorphismus.
+Die $\varphi_i$ lassen sich nun zu einem Isomorphismus $\varphi:\Omega_{X/Y}\to\widetilde{M}$ verkleben.\qed
+
 \printindex

### 2015-05-16 Algebraische Geometrie/AlgebraischeGeometrie.tex
Ende Kapitel 2

@@
 \item $X$ heißt \textit{vollständig}\index{Kurve!vollständig}, wenn $X\to k$ eigentlich ist.
-\item $X$ heißt \textit{nicht-singulär}\index{Kurve!nicht-singulär}, falls alle lokalen Ringe von $X$ regulär sind.
+\item $X$ heißt \textit{nicht-singulär}\index{nicht-singulär}, falls alle lokalen Ringe von $X$ regulär sind.
 \end{itemize}
@@
 \paragraph{Satz 8.21.}\label{8.21} Sei $f:X\to Y$ ein Morphismus und $Z$ ein abgeschlossenes Unterschema von $X$ zum Ideal $\mathcal{J}$. Dann gibt es eine exakte Sequenz von Garben auf $Z$:
-\[\mathcal{J}/\mathcal{J}^2 \longrightarrow \Omega_{X/Y}\otimes_{\mathcal{O}_X}\mathcal{O}_Z\longrightarrow\Omega_{Z/Y}\longrightarrow 0 \]
+\[\mathcal{J}/\mathcal{J}^2 \stackrel{\delta}{\longrightarrow} \Omega_{X/Y}\otimes_{\mathcal{O}_X}\mathcal{O}_Z\longrightarrow\Omega_{Z/Y}\longrightarrow 0 \]
 
@@
 
+\subsection{Nicht-singuläre Varietäten}
+
+\paragraph{Definition 8.24.}\label{8.24} Eine Varietät $X$ über einem algebraisch abgeschlossenen Körper $k$, d.h.
+ein integres, separiertes Schema von endlichem Typ, heißt \textit{nicht-singulär}\index{nicht-singulär}, wenn alle lokalen Ringe regulär sind.
+
+\paragraph{Satz 8.25.}\label{8.25} Sei $X$ ein irreduzibles, separiertes Schema von endlichem Typ über einem algebraisch abgeschlossenen Körper $k$.
+Dann ist $\Omega_{X/k}$ genau dann lokal frei vom Rang $n=\dim(X)$, wenn $X$ eine nicht-singuläre Varietät über $k$ ist.
+
+\iffalse
+\paragraph{Beweis.} Sei $x\in X$ ein abgeschlossener Punkt.
+Dann ist $\dim(\mathcal{O}_{X,x})=n$ und $\kappa(x)=k$.
+Es ist $B=\mathcal{O}_{X,x}$ eine endlich erzeugte $k$"~Algebra und $\Omega_{B/k}=(\Omega_{X/k})_x$.
+Nach \hyperref[8.15]{Satz 8.15} ist $(\Omega_{X/k})_x$ genau dann frei vom Rang $n$, wenn $\mathcal{O}_{X,x}$ regulär ist.
+Somit haben wir mit \hyperref[8.26]{Lemma 8.26} folgende Äquivalenzen:
+\begin{align*}
+& \text{Alle lokalen Ringe sind regulär}\\
+\iff & \text{Alle lokalen Ringe von abgeschlossenen Punkten sind regulär}\\
+\iff & (\Omega_{X/k})_x \text{ ist frei vom Rang $n$ für alle abgeschlossene $x\in X$}\\
+\iff & \Omega_{X/k}\text{ ist lokal frei vom Rang $n$}\qedhere
+\end{align*}
+\fi
+
+\paragraph{Lemma 8.26.}\label{8.26} Sei $X$ ein noethersches Schema und $\mathcal{F}$ eine kohärente Garbe.
+Dann ist $\mathcal{F}$ genau dann lokal frei, wenn $\mathcal{F}_x$ freier $\mathcal{O}_{X,x}$"~Modul für alle $x\in X$ ist.
+
+\paragraph{Korollar 8.27.}\label{8.27} Sei $X$ eine nicht-singuläre Varietät über $k$ und $Y\subset X$ ein irreduzibles, abgeschlossenes Unterschema zur Idealgarbe $\mathcal{J}$.
+Dann ist $Y$ genau dann nicht-singulär, wenn folgende Bedingungen erfüllt sind:
+\begin{enumerate}[(i)]
+\item $\Omega_{Y/k}$ ist lokal frei.
+\item In der exakten Sequenz aus \hyperref[8.21]{Satz 8.21} ist $\delta$ injektiv:
+\[0\longrightarrow\mathcal{J}/\mathcal{J}^2\stackrel{\delta}{\longrightarrow}\Omega_{X/k}\otimes_{\mathcal{O}_X}\mathcal{O}_Y\longrightarrow\Omega_{Y/k}\longrightarrow 0 \]
+\end{enumerate}
+
+\chapter{Kohomologie}
+
+Wir verwenden folgende Bezeichnungen für abelsche Kategorien:
+\begin{itemize}
+\item $\mathbf{Ab}$, die Kategorie der abelschen Gruppen
+\item $\mathbf{Mod}(A)$, die Kategorie der Moduln über einem kommutativen Ring $A$ mit Eins
+\item $\mathbf{Ab}(X)$, die Kategorie der Garben abelscher Gruppen über einem topologischen Raum $X$
+\item $\mathbf{Mod}(X)$, die Kategorie der $\mathcal{O}_X$"~Moduln auf einem geringten Raum $(X,\mathcal{O}_X)$
+\item $\mathbf{Qcoh}(X)$, die Kategorie der quasikohärenten $\mathcal{O}_X$"~Moduln auf einem Schema $X$
+\item $\mathbf{Coh}(X)$, die Kategorie der kohärenten $\mathcal{O}_X$"~Modulgarben auf einem noetherschen Schema $X$
+\end{itemize}
+
 \printindex

### 2015-05-19 Algebraische Geometrie/AlgebraischeGeometrie.tex
19.05.2015

@@
 \paragraph{Definition.} Sei $X$ ein Schema.
-Ein quasikohärenter $\mathcal{O}_X$"~Modul $\mathcal{S}$ habe die Eigenschaft $(\star)$, wenn $X$ noethersch ist und wenn er eine Struktur einer graduierten $\mathcal{O}_X$"~Algebra trägt, d.h.
+Ein quasikohärenter $\mathcal{O}_X$"~Modul $\mathcal{S}$ habe die Eigenschaft $(\dagger)$, wenn $X$ noethersch ist und wenn er eine Struktur einer graduierten $\mathcal{O}_X$"~Algebra trägt, d.h.
 $\mathcal{S}=\bigoplus_{d\geq 0}\mathcal{S}_d$ mit homogenen Teilen $\mathcal{S}_d$, so dass:
@@
 
-\paragraph{Definition 7.9.}\label{7.9} Sei $X$ ein Schema und $\mathcal{S}$ ein graduierter $\mathcal{O}_X$"~Modul mit $(\star)$.
+\paragraph{Definition 7.9.}\label{7.9} Sei $X$ ein Schema und $\mathcal{S}$ ein graduierter $\mathcal{O}_X$"~Modul mit $(\dagger)$.
 Sei $U=\operatorname{Spec}(A)\subset_\text{o}X$, $\mathcal{S}(U)=\Gamma(U,\mathcal{S}|_U)$ die graduierte $A$"~Algebra und $\pi_U:\operatorname{Proj}\mathcal{S}(U)\to U$ der kanonische Morphismus.
@@
 
-\paragraph{Lemma 7.11.}\label{7.11} Sei $\mathcal{S}$ eine Garbe graduierter Algebren mit $(\star)$ auf $X$.
+\paragraph{Lemma 7.11.}\label{7.11} Sei $\mathcal{S}$ eine Garbe graduierter Algebren mit $(\dagger)$ auf $X$.
 Sei $\mathcal{L}$ eine invertierbare Garbe auf $X$ und $\mathcal{S}'$ die folgende Garbe graduierter Algebren:
 \[\mathcal{S}_d' =\mathcal{S}_1\otimes\mathcal{L}^d,\quad d\geq 0 \]
-Dann erfüllt $\mathcal{S}'$ ebenfalls $(\star)$ und es gibt einen natürlichen Isomorphismus:
+Dann erfüllt $\mathcal{S}'$ ebenfalls $(\dagger)$ und es gibt einen natürlichen Isomorphismus:
 \[\begin{tikzcd}
@@
 
-\paragraph{Satz 7.12.}\label{7.12} Sei $X$ und $\mathcal{S}$ mit $(\star)$. Sei $P=\operatorname{\mathbf{Proj}}(\mathcal{S})$ mit $\pi:P\to X$. Dann gilt:
+\paragraph{Satz 7.12.}\label{7.12} Sei $X$ und $\mathcal{S}$ mit $(\dagger)$. Sei $P=\operatorname{\mathbf{Proj}}(\mathcal{S})$ mit $\pi:P\to X$. Dann gilt:
 \begin{enumerate}[(i)]
@@
 Sei $\mathcal{S}=S(\mathcal{E})$ die symmetrische Algebra von $\mathcal{E}$ und $\mathbf{P}(\mathcal{E})=\operatorname{\mathbf{Proj}}(\mathcal{S})$.
-Dann ist $\mathcal{S}$ eine Garbe von graduierten $\mathcal{O}_X$"~Algebren, die $(\star)$ erfüllt.
+Dann ist $\mathcal{S}$ eine Garbe von graduierten $\mathcal{O}_X$"~Algebren, die $(\dagger)$ erfüllt.
 $\pi:\mathbf{P}(\mathcal{E})\to X$ heißt \textit{projektives Vektorbündel}\index{Vektorbündel!projektiv} über $X$.
@@
 Sei $\mathcal{S}=\bigoplus_{d\geq 0}\mathcal{J}^d$.
-Dann erfüllen $X$ und $\mathcal{S}$ die Bedingung $(\star)$.
+Dann erfüllen $X$ und $\mathcal{S}$ die Bedingung $(\dagger)$.
 $\widetilde{X}=\mathbf{Proj}(\mathcal{S})$ heißt die \textit{Aufblasung} von $X$ bzgl.
@@
 \paragraph{Beispiel 8.8.}\label{8.8} Sei $B$ eine $A$"~Algebra und $C=B[x_1,\ldots,x_n]$.
-Sei $M$ ein beliebiger $C$"~Modul und $D\in\operatorname{Der}_A(B,M)$.
+Sei $M$ ein beliebiger $C$"~Mo\-dul und $D\in\operatorname{Der}_A(B,M)$.
 Sei $\widetilde{D}:C\to M$ die einem Polynom $F$, den Polynom $F^D$ zuordnet, wobei $F^D$ das Polynom ist, das man durch Anwenden von $D$ auf die Koeffizienten von $F$ erhält.
@@
 
+\section{Kohomologie von Garben}
+
+\paragraph{Satz 9.15.}\label{9.15} Sei $A$ ein Ring. Dann ist jeder $A$"~Modul isomorph zu einem Untermodul eines injektiven $A$"~Moduls.
+
+\paragraph{Satz 9.16.}\label{9.16} Sei $(X,\mathcal{O}_X)$ ein geringter Raum. Dann hat die Kategorie $\mathbf{Mod}(\mathcal{O}_X)$ genügend viele Injektive.
+
+\paragraph{Beweis.} Sei $\mathcal{F}\in\mathbf{Mod}(\mathcal{O}_X)$.
+Dann ist $\mathcal{F}_x\in\mathbf{Mod}(\mathcal{O}_{X,x})$.
+Nach \hyperref[9.15]{Satz 9.15} gibt es ein Monomorphismus $\mathcal{F}_x\to I_x$ mit einem injektiven $\mathcal{O}_{X,x}$"~Modul $I_x$.
+Sei $j:\{x\}\hookrightarrow X$ die kanonische Inklusion und betrachte die $I_x$ als Garben auf $\{x\}$.
+Setze:
+\[\mathcal{J}=\prod_{x\in X}j_\ast I_x \]
+Dann gilt:
+\[\operatorname{Hom}_{\mathcal{O}_X}(\mathcal{F},\mathcal{J}) = \prod_{x\in X}\operatorname{Hom}_{\mathcal{O}_X}(\mathcal{F},j_\ast I_x) = \prod_{x\in X}\operatorname{Hom}_{\mathcal{O}_{X,x}}(\mathcal{F}_x,I_x) \]
+Die $\mathcal{F}_x\hookrightarrow I_x$ geben ein Monomorphismus $\mathcal{F}\hookrightarrow \mathcal{J}$. Ferner gilt:
+\[\operatorname{Hom}_{\mathcal{O}_X}(-,\mathcal{J})=\prod_{x\in X}\operatorname{Hom}_{\mathcal{O}_{X,x}}((-)_x, I_x) \]
+Da Halmbildung exakt und $I_x$ injektive Objekte sind, ist dies ein exakter Funktor. Somit ist $\mathcal{J}$ ein injektiver $\mathcal{O}_X$"~Modul.\qed
+
+\paragraph{Korollar 9.17.}\label{9.17} Sei $X$ ein topologischer Raum. Dann besitzt $\mathbf{Ab}(X)$ genügend viele Injektive.
+
+\paragraph{Beweis.} Sei $\mathcal{O}_X$ die konstante Garbe $\mathbb{Z}$ auf $X$. Dann ist $(X,\mathcal{O}_X)$ ein geringter Raum und es gilt $\mathbf{Mod}(\mathcal{O}_X)=\mathbf{Ab}(X)$.\qed
+
+\paragraph{Definition 9.18.}\label{9.18} Sei $X$ ein topologischer Raum und $\Gamma(X,-):\mathbf{Ab}(X)\to \mathbf{Ab}$ der globale Schnittfunktor. Wir definieren:
+\[\mathrm{H}^i(X,-)= \mathrm{R}^i\Gamma(X,-) \]
+$\mathrm{H}^i(X,\mathcal{F})$ heißt \textit{$i$-te Kohomologiegruppe}\index{Kohomologiegruppe} von der Garbe $\mathcal{F}$.
+Haben $X$ und $\mathcal{F}$ Zusatzstrukturen, z.B.
+$X$ Schema oder $\mathcal{F}\in \mathbf{Qcoh}(X)$, so wird die Kohomologie trotzdem immer im obigen Sinne verstanden, d.h.
+$\mathcal{F}\in\mathbf{Ab}(X)$.
+
+Sei $(X,\mathcal{O}_X)$ ein geringter Raum und $\mathcal{F}$ ein $\mathcal{O}_X$"~Modul.
+$\mathcal{F}$ heißt\textit{welk}\index{$\mathcal{O}_X$-Modul!welk}, falls für alle offenen $V\subset U$ die Restriktionsabbildung $\mathcal{F}(U)\to\mathcal{F}(V)$ surjektiv ist.
+
+\paragraph{Lemma 9.19.}\label{9.19} Ist $(X,\mathcal{O}_X)$ ein geringter Raum, so ist jeder injektiver $\mathcal{O}_X$"~Modul $\mathcal{J}$ welk.
+
+\paragraph{Beweis.} Seien $i:U\hookrightarrow X$ und $j:V\hookrightarrow X$ die natürlichen Inklusionen.
+Betrachte die Inklusion $j_!\mathcal{O}_X|_V\hookrightarrow i_!\mathcal{O}_X|_U$ von $\mathcal{O}_X$"~Moduln.
+Es folgt die Surjektivität von:
+\[\mathcal{J}(U) = \operatorname{Hom}(i_!\mathcal{O}_X|_U,\mathcal{J})\to \operatorname{Hom}(j_!\mathcal{O}_X|_V,\mathcal{J})=\mathcal{J}(V)\qedhere \]
+
+\paragraph{Satz 9.20.}\label{9.20} Sei $X$ ein topologischer Raum und $\mathcal{F}$ eine welke Garbe auf $X$. Dann ist $\mathrm{H}^i(X,\mathcal{F})=0$ für alle $i>0$.
+
 \printindex

### 2015-05-19 Algebraische Geometrie/AlgebraischeGeometrie.tex
Fehler fixen

@@
 \[\Gamma_\ast(\mathcal{F})=\bigoplus_{n\in\mathbb{Z}}\Gamma(X,\mathcal{F}(n)) \]
-mit der folgenden $S$"~Wirkung: Für $s\in S_d\cong\Gamma(X,\mathcal{O}_X(d))$ und $t\in\Gamma(X,\mathcal{F}(n))$ setze $st= s\otimes t\in\Gamma(X,\mathcal{F}(n)\otimes_{\mathcal{O}_X}\mathcal{O}_X(d))=\Gamma(X,\mathcal{F}(n+d))$.
+mit der folgenden $S$"~Wirkung: Ein $s\in S_d$ induziert ein $s\in \Gamma(X,\mathcal{O}_X(d))$.
+Für $t\in\Gamma(X,\mathcal{F}(n))$ setze $st= s\otimes t\in\Gamma(X,\mathcal{F}(n)\otimes_{\mathcal{O}_X}\mathcal{O}_X(d))=\Gamma(X,\mathcal{F}(n+d))$.
 

### 2015-05-23 Algebraische Geometrie/AlgebraischeGeometrie.tex
21.05.2015

@@
 
+\paragraph{Lemma 9.21.}\label{9.21} Sei $X$ ein topologischer Raum. Sei $0\to\mathcal{G}'\to\mathcal{G}\to\mathcal{G}''\to 0$ eine exakte Sequenz von Garben.
+\begin{enumerate}[(i)]
+\item Ist $\mathcal{G}'$ eine welke Garbe, so ist für alle $U\subset_\text{o}X$ die folgende Sequenz exakt:
+\[0\longrightarrow \mathcal{G}'\longrightarrow \mathcal{G}\longrightarrow \mathcal{G}''\longrightarrow 0 \]
+\item Sind $\mathcal{G'}$ und $\mathcal{G}$ welk, so auch $\mathcal{G}''$.
+\end{enumerate}
+
+\paragraph{Beweis.}
+\begin{enumerate}[(i)]
+\item Sei o.B.d.A. $U=X$ und sei $s''\in\mathcal{G}''(X)$. Setze:
+\[E=\{(U,s)\mid U\subset_\text{o}X,\ s\in\mathcal{G}(U),\ s\mapsto s''|_U \} \]
+$E$ ist partiell geordnet bzgl. Inklusion. Jede Kette in $E$ besitzt eine obere Schranke in $E$. Nach Zorns Lemma gibt es ein maximales Element $(U,s)\in E$.
+
+Gibt es ein $x\in X\setminus U$, so existiert eine offene Umgebung $V$ von $x$ und $t\in\mathcal{G}(V)$ mit $t\mapsto s''|_V$, da $\mathcal{G}\to\mathcal{G}''$ surjektiv ist.
+Da $\mathcal{G}$ welk ist, so setzt sich $s-t\in\mathcal{G}(U\cap V)$ zu einem Schnitt auf $\mathcal{G}(V)$ fort.
+Ersetzen wir $t$ durch $s-t\in\mathcal{G}(V)$, können wir o.B.d.A.
+$s=t$ auf $V$ annehmen.
+Die Existenz von $(V,s)$ ist ein Widerspruch zur Maximalität von $(U,s)$ Somit folgt $U=X$.
+\item Für $V\subset U$ offen in $X$ kommutiert das folgende Diagramm mit exakten Zeilen:
+\[\begin{tikzcd}
+0 \rar & \mathcal{G}'(U)\rar\dar[two heads] & \mathcal{G}(U)\rar\dar[two heads] & \mathcal{G}''(U)\rar\dar & 0\\
+0 \rar & \mathcal{G}'(V)\rar & \mathcal{G}(V)\rar & \mathcal{G}''(V)\rar & 0
+\end{tikzcd} \]
+Es folgt die Surjektivität von $\mathcal{G}''(U)\to\mathcal{G}''(V)$.\qed
+\end{enumerate}
+
+\paragraph{Beweis von \hyperref[9.20]{Satz 9.20}.} Sei $0\to\mathcal{F}\to\mathcal{J}\to\mathcal{G}\to 0$ exakt mit $\mathcal{J}$ injektiv.
+Nun sind $\mathcal{F}$ und $\mathcal{J}$ welk.
+Nach \hyperref[9.21]{Lemma 9.21} ist auch $\mathcal{G}$ welk und wir haben eine exakte Folge:
+\[0\longrightarrow\mathcal{F}(X)\longrightarrow\mathcal{J}(X)\longrightarrow\mathcal{G}(X)\longrightarrow 0 \]
+Da $\mathrm{H}^i(X,\mathcal{J})=0$ für alle $i>0$, gilt $\mathrm{H}^1(X,\mathcal{F})=0$ und $\mathrm{H}^i(X,\mathcal{F})=\mathrm{H}^{i-1}(X,\mathcal{G})$ für alle $i\geq 2$.
+Somit folgt die Aussage per Induktion nach $i$.\qed
+
+\paragraph{Satz 9.22.}\label{9.22} Sei $(X,\mathcal{O}_X)$ ein geringter Raum. Dann sind die derivierten Funktoren von
+\[\Gamma(X,-): \mathbf{Mod}(X)\to\mathbf{Ab} \]
+gerade die Kohomologie-Funktoren.
+
+\paragraph{Beweis.} $\mathrm{R}^i\Gamma(X,-)$ wird über eine injektive Auflösung in $\mathbf{Mod}(X)$ berechnet.
+Nun sind injektive Objekte in $\mathbf{Mod}(X)$ nach \hyperref[9.19]{Lemma 9.19} welk und nach \hyperref[9.20]{Satz 9.20} insbesondere azyklisch bzgl.
+$\Gamma(X,-)$.Somit ist $\mathrm{R}^i\Gamma(X,-)=\mathrm{H}^i(X,-)$.\qed
+
+\paragraph{Bemerkung 9.23.}\label{9.23} Sei $(X,\mathcal{O}_X)$ ein geringter Raum und $\mathcal{F}$ ein $\mathcal{O}_X$"~Modul.
+Da die Kohomologie durch Auflösungen in der Kategorie $\mathbf{Mod}(X)$ berechnet werden kann, haben alle Kohomologiegruppen von $\mathcal{F}$ eine $A$"~Modulstruktur mit $A=\Gamma(X,\mathcal{O}_X)$.
+
+\paragraph{Lemma 9.24.}\label{9.24} Sei $X$ ein noetherscher topologischer Raum und $\mathcal{F}=\varinjlim_\alpha\mathcal{F}_\alpha$. Dann gilt:
+\begin{enumerate}[(i)]
+\item $\mathcal{F}(X)=\varinjlim_\alpha\mathcal{F}_\alpha(X)$
+\item Sind alle $\mathcal{F}_\alpha$ welk, so auch $\mathcal{F}$.
+\end{enumerate}
+
+\paragraph{Beweis.}\begin{enumerate}
+\item[(ii)] Für alle $\alpha$ und $V\subset U$ offen in $X$ ist $\mathcal{F}_\alpha(U)\to\mathcal{F}_\alpha(V)$ surjektiv. Da $\varinjlim$ exakt ist, folgt mit (i) die Surjektivität von:
+\[\mathcal{F}(U)=\varinjlim_\alpha\mathcal{F}_\alpha(U)\to\varinjlim\mathcal{F}_\alpha(V) = \mathcal{F}(V) \]
+\item[(i)] Der Funktor $\Gamma(X,-)$ vertauscht mit $\varinjlim$, siehe z.B. Godement, Topologie algébrique et théorie des faisceaux, p. 169.\qed
+\end{enumerate}
+
+\paragraph{Satz 9.25.}\label{9.25} Sei $X$ ein noetherscher topologischer Raum und $(\mathcal{F}_\alpha)_{\alpha\in A}$ ein induktives System in $\mathbf{Ab}(X)$.
+Dann gibt es einen natürlichen Isomorphismus:
+\[\varinjlim\mathrm{H}^i(X,\mathcal{F}_\alpha) \stackrel{\sim}{\to} \mathrm{H}^i(X,\varinjlim\mathcal{F}_\alpha) \]
+
+\paragraph{Beweis.} Für alle $\alpha$ haben wir natürliche Abbildungen $\mathcal{F}_\alpha\to\varinjlim\mathcal{F}_\alpha$.
+Wir erhalten $\mathrm{H}^i(X,\mathcal{F}_\alpha)\to\mathrm{H}^i(X,\varinjlim\mathcal{F}_\alpha)$ und daher eine Abbildung:
+\[\varinjlim\mathrm{H}^i(X,\mathcal{F}_\alpha)\to\mathrm{H}^i(X,\varinjlim\mathcal{F}_\alpha) \]
+Dies ist nach \hyperref[9.24]{Lemma 9.24 (i)} für $i=0$ ein Isomorphismus.
+Betrachte die abelsche Kategorie $\mathbf{Ind}_A(\mathbf{Ab}(X))$ der direkten Limiten von Objekten aus $\mathbf{Ab}(X)$ indiziert durch $A$.
+Da $\varinjlim$ exakt ist, haben wir eine natürliche Transformation von $\delta$-Funktoren $\mathbf{Ind}_A(\mathbf{Ab}(X))\to\mathbf{Ab}$
+\[\varinjlim\mathrm{H}^i(X,-)\to\mathrm{H}^i(X,\varinjlim -) \]
+die für $i=0$ übereinstimmt. Jetzt genügt es zu zeigen, dass beide Funktoren auslöschbar sind. Dann sind beide universell und insbesondere gleich.
+
+Sei $(\mathcal{F}_\alpha)\in\mathbf{Ind}_A(\mathbf{Ab}(X))$. Zu $\alpha$ definieren wir die Garbe $\mathcal{G}_\alpha$ wie folgt:
+\[U\mapsto \Big\{s:U\to\coprod_{P\in U}(\mathcal{F}_\alpha)_P\mid s(P)\in (\mathcal{F}_\alpha)_P \Big\} \]
+Dann ist $\mathcal{G}_\alpha$ welk und $\mathcal{F}_\alpha\hookrightarrow\mathcal{G}_\alpha$.
+Nach Konstruktion bilden die $\mathcal{G}_\alpha$ ein induktives System und wir erhalten einen Monomorphismus $u:(\mathcal{F}_\alpha)\hookrightarrow (\mathcal{G}_\alpha)$ in $\mathbf{Ind}_A(\mathbf{Ab}(X))$.
+Ferner gilt $\mathrm{H}^i(X,\mathcal{G}_\alpha)=0$ für alle $i>0$ nach \hyperref[9.20]{Satz 9.20}, also $\varinjlim\mathrm{H}^i(X,\mathcal{G}_\alpha)=0$ für alle $i>0$.
+Somit ist $\varinjlim\mathrm{H}^i(X,-)$ auslöschbar.
+
+Da $\varinjlim\mathcal{G}_\alpha$ nach \hyperref[9.24]{Satz 9.24 (ii)} welk ist, ist auch $\mathrm{H}^i(X,\varinjlim\mathcal{G}_\alpha)=0$ für alle $i>0$, d.h.
+auch $\mathrm{H}^i(X,\varinjlim -)$ ist auslöschbar.\qed
+
+\paragraph{Lemma 9.26.}\label{9.26} Sei $Y\subset X$ abgeschlossen mit natürlicher Inklusion $i:Y\hookrightarrow X$ und $\mathcal{F}\in\mathbf{Ab}(Y)$. Dann gilt:
+\[\mathrm{H}^i(Y,\mathcal{F})=\mathrm{H}^i(X, i_\ast\mathcal{F}) \]
+
+\paragraph{Beweis.} Sei $\mathcal{J}^\bullet$ eine injektive Auflösung von $\mathcal{F}$ auf $Y$; insbesondere ist $\mathcal{J}^\bullet$ eine welke Auflösung.
+Dann ist $i_\ast\mathcal{J}^\bullet$ eine welke Auflösung von $i_\ast\mathcal{F}$ auf $X$.
+Für alle $i$ gilt:
+\[\Gamma(Y,\mathcal{J}^i) = \Gamma(X,i_\ast\mathcal{J}^i) \]
+Somit ergeben sich die gleichen Kohomologiegruppen.\qed
+
+\paragraph{Satz 9.27.}\label{9.27} \textit{(Verschwindungssatz von Grothendieck)}\index{Verschwindungssatz} Sei $X$ ein noetherscher topologischer Raum der Dimension $n$.
+Dann gilt für alle abelschen Garben $\mathcal{F}$ auf $X$:
+\[\mathrm{H}^i(X,\mathcal{F})=0\quad\text{für alle }i>n \]
+
 \printindex

### 2015-05-23 Algebraische Geometrie/AlgebraischeGeometrie.tex
Anfang Beweis vom Verschwindungssatz

@@
 
+\paragraph{Beweis.} Sei $\mathcal{F}$ eine Garbe abelscher Gruppen auf $X$.
+\begin{enumerate}
+\item Schritt: \textit{(Reduktion auf irreduzible $X$)} Sei $X$ reduzibel und $Y\subset X$ eine irreduzible Komponente mit Einbettungen $i:Y\hookrightarrow X,\ j:U=X\setminus Y\hookrightarrow X$.
+Nach \hyperref[1.16]{1.16} ist die folgende Garbensequenz exakt:
+\[0\longrightarrow \mathcal{F}_U \longrightarrow\mathcal{F}\longrightarrow\mathcal{F}_Y\longrightarrow 0 \]
+wobei $\mathcal{F}_U= j_!(\mathcal{F}|_U)$ und $\mathcal{F}_Y=i_\ast(\mathcal{F}|_Y)$ ist.
+Es genügt zu zeigen, dass $\mathrm{H}^i(X,\mathcal{F}_U)=0=\mathrm{H}^i(X,\mathcal{F}_Y)$ für alle $i>n$ ist.
+Betrachte $g:\overline{U}\hookrightarrow X$.
+Nun hat $\overline{U}$ weniger irreduzible Komponenten als $X$.
+Setze $\mathcal{F}_{\overline{U}}=g^\ast\mathcal{F}_U$.
+Es folgt nach \hyperref[9.26]{Lemma 9.26}:
+\[\mathrm{H}^i(\overline{U},\mathcal{F}_{\overline{U}})=\mathrm{H}^i(X,g_\ast\mathcal{F}_{\overline{U}}) = \mathrm{H}^i(X,\mathcal{F}_U) \]
+\[\mathrm{H}^i(X,\mathcal{F}_Y)=\mathrm{H}^i(X,i_\ast\mathcal{F}|_Y)=\mathrm{H}^i(Y,\mathcal{F}|_Y) \]
+Per Induktion nach Anzahl der irreduziblen Komponenten von $X$ können wir o.B.d.A. $X$ als irreduzibel annehmen.
+\item Schritt: Sei $X$ irreduzibel und $\dim(X)=0$.
+Die einzigen offenen Mengen von $X$ sind $\varnothing$ und $X$, d.h.
+$\Gamma(X,-):\mathbf{Ab}(X)\to\mathbf{Ab}$ ist eine Äquivalenz von Kategorien.
+Insbesondere ist $\Gamma(X,-)$ exakt, d.h.
+$\mathrm{H}^i(X,\mathcal{F})=0$ für alle $i>0$.
+\item Schritt: Sei $X$ irreduzibel und $\dim(X)=n$. Sei $j:U\hookrightarrow X$ offen und $s\in\mathcal{F}(U)$. Da
+\[\mathcal{F}(U)=\operatorname{Hom}(\mathbb{Z},\mathcal{F}(U))=\operatorname{Hom}_U(\mathbb{Z}|_U,\mathcal{F}|_U)=\operatorname{Hom}_X(j_!(\mathbb{Z}|_U),\mathcal{F}) \]
+wobei $\mathbb{Z}|_U$ die konstante Garbe bezeichnet, entspricht $s$ einen Morphismus $j_!(\mathbb{Z}|_U)\to\mathcal{F}$.
+Setze $\mathcal{F}_s=\operatorname{im}(j_!(\mathbb{Z}|_U)\to\mathcal{F})$.
+$\mathcal{F}_s$ heißt die von $s$ erzeugte Untergarbe von $\mathcal{F}$.
+Sei $\alpha=(s_i)_{i\in I},\ s_i\in\mathcal{F}(U_i)$ eine beliebige Familie von Schnitten.
+Setze:
+\[\mathcal{F}_\alpha = \sum_{i\in I}\mathcal{F}_{s_i}\subset\mathcal{F}\]
+Sei $J$ die Menge aller endlichen Familien $\alpha=(s_i)$. $J$ wird mit der folgenden par\-tiel\-len Ordnung zu einem gerichteten System:
+\[\alpha'\leq\alpha\iff \alpha'\text{ ist Unterfamilie von }\alpha \]
+Für $\alpha'\leq\alpha$ gilt $\mathcal{F}_{\alpha'}\subset\mathcal{F}_\alpha$.
+Somit ist $\mathcal{F}=\varinjlim_{\alpha\in J}\mathcal{F}_\alpha$.
+Nach \hyperref[9.25]{Satz 9.25} reicht es zu zeigen, dass $\mathrm{H}^i(X,\mathcal{F}_\alpha)=0$ für alle $i>n$ und alle $\alpha\in J$.
+
+Sei nun $\alpha'\leq\alpha$.
+Betrachte die exakte Sequenz $0\to\mathcal{F}_{\alpha'}\to\mathcal{F}_\alpha\to\mathcal{G}\to 0$.
+$\mathcal{G}$ ist erzeugt von $\#\alpha - \#\alpha'$ Schnitten über geeignete offenen Mengen.
+Per Induktion nach $\#\alpha$ können wir unter Betrachtung der langen exakten Kohomologiefolge o.B.d.A.
+annehmen, dass $\mathcal{F}$ von einem Schnitt über einer geeigneten offenen Menge $U$ erzeugt wird, d.h.
+$\mathcal{F}=\mathcal{F}_s$.
+Die zu $s$ gehörige Morphismus $\mathbb{Z}_U=j_!(\mathbb{Z}|_U)\to\mathcal{F}$ ist surjektiv.
+
+Definiere die Garbe $\mathcal{R}$ durch die exakte Folge $0\to\mathcal{R}\to\mathbb{Z}_U\to\mathcal{F}\to 0$.
+Unter Betrachtung der langen exakten Kohomologiefolge genügt es zu zeigen, dass $\mathrm{H}^i(X,\mathcal{R})=0=\mathrm{H}^i(X,\mathbb{Z}_U)$ für alle $i>n$ ist.
+\end{enumerate}
+
 \printindex

### 2015-05-29 Algebraische Geometrie/AlgebraischeGeometrie.tex
Besserer Textfluss
Fix typo
Neue Dateistruktur

--- 2015-05-19 Algebraische Geometrie/AlgebraischeGeometrie.tex
@@
 
-\chapter{Varietäten}
-
-Sei $k$ ein algebraisch abgeschlossener Körper.
-
-\section{Affine Varietäten}
-
-\paragraph{Definition.} \begin{enumerate}[(i)]
-\item Die Menge aller $n$-Tupeln über $k$
-\[\mathbf{A}^n = \mathbf{A}_k^n = \{(a_1,\ldots,a_n)\mid a_i\in k\} \]
-heißt \textit{affiner $n$-dimensionaler Raum}\index{affiner Raum} über $k$.
-Ein Element $P=(a_1,\ldots,a_n)\in\mathbf{A}^n$ heißt \textit{Punkt}\index{Punkt} und die $a_i$ heißen \textit{Koordinaten}\index{Koordinate} von $P$.
-\item Der Polynomring über $k$ in $n$ Variablen bezeichnen wir mit $A=k[X_1,\ldots,X_n]$. Für $T\subset A$ definieren wir die \textit{Nullstellenmenge}\index{Nullstellenmenge} von $T$ wie folgt:
-\[Z(T)=\{P\in\mathbf{A}^n\mid \forall f\in T\colon f(P)=0\} \]
-Es gilt $Z(T)=Z(\mathfrak{a})$, wobei $\mathfrak{a}$ das von $T$ erzeugte Ideal in $A$ ist.
-\item Eine Teilmenge $Y\subset\mathbf{A}^n$ der Form $Y=Z(T)$ für ein $T\subset A$ heißt \textit{affine algebraische Menge}\index{algebraische Menge!affin}.
-Für algebraische Mengen $Y_i=Z(\mathfrak{a}_i)$ mit Ideale $\mathfrak{a}_i\subset A,\ i\in I$ gilt:
-\[Y_1\cup Y_2=Z(\mathfrak{a}_1\cap\mathfrak{a}_2),\quad \bigcap_{i\in I}Y_i = Z\Big(\sum_{i\in I}\mathfrak{a}_i\Big) \]
-Ferner gilt $\mathbf{A}^n=Z(0)$ und $\varnothing = Z(1)$.
-Wir statten $\mathbf{A}^n$ mit der sogenannten \textit{Zariski-Topologie}\index{Zariski-Topologie} aus, in dem wir eine Menge $U\subset\mathbf{A}^n$ genau dann offen nennen, wenn $\mathbf{A}^n\setminus U$ eine algebraische Menge ist.
-\item Eine \textit{affine Varietät}\index{Varietät!affin} $V$ ist eine \textit{irreduzible}\index{irreduzibel} abgeschlossene Menge in $\mathbf{A}^n$, d.h.
-aus $V=V_1\cup V_2$ mit abgeschlossenen Mengen $V_1,V_2\subset\mathbf{A}^n$ folgt $V_1=\varnothing$ oder $V_2=\varnothing$.
-
-Eine offene Teilmenge einer affinen Varietät bzgl. der induzierten Topologie nennt man \textit{quasi-affine Varietät}\index{Varietät!quasi-affin}. 
-\item Sei $Y\subset\mathbf{A}^n$ eine algebraische Menge. Dann definieren wir das Ideal:
-\[I(Y)=\{f\in A\mid \forall P\in Y\colon f(P)=0\} \]
-Der \textit{Koordinatenring}\index{Koordinatenring} von $Y$ ist definiert als $A(Y)=A/I(Y)$.
-\end{enumerate}
-
-\paragraph{Definition.} Sei $\mathfrak{a}\subset A$ ein Ideal. Das \textit{Radikal}\index{Radikal} von $\mathfrak{a}$ ist definiert als:
-\[\operatorname{Rad}(\mathfrak{a})=\{f\in A\mid\exists r>0\colon f^r\in\mathfrak{a} \} \]
-Ein Ideal $\mathfrak{a}\subset A$ heißt \textit{Radikalideal}\index{Radikalideal}, wenn $\mathfrak{a}=\operatorname{Rad}(\mathfrak{a})$ gilt.
-
-\paragraph{Satz.} Es gibt eine inklusionsumkehrende Bijektion:
-\[\{\text{Algebraische Mengen in }\mathbf{A}^n\} \to \{\text{Radikalideale in }k[X_1,\ldots,X_n]\},\ Y\mapsto I(Y) \]
-mit der Umkehrabbildung $\mathfrak{a}\mapsto Z(\mathfrak{a})$. Eine algebraische Menge $Y$ ist genau dann irreduzibel, wenn $I(Y)\subset A$ ein Primideal ist.
-
-\section{Projektive Varietäten}
-
-\paragraph{Definition.}\begin{enumerate}[(i)]
-\item Zwei Punkte $(a_0,\ldots,a_n),(b_0,\ldots,b_n)\in \mathbf{A}^{n+1}$ heißen äquivalent, wenn ein $\lambda\in k^\times$ existiert, so dass $a_i=\lambda b_i$ für alle $i$ gilt.
-Die Äquivalenzklasse von $(a_0,\ldots,a_n)$ wird mit $(a_0\: \ldots\: a_n)$ bezeichnet.
-Der \textit{projektiver $n$-dimensionaler Raum}\index{projektiver Raum} über $k$ wird definiert als:
-\[\mathbf{P}^n=\mathbf{P}_k^n = \{(a_0\: \ldots\: a_n) \mid a_i\in k\text{ nicht alle $0$} \} \]
-Ein Element $P=(a_0\: \ldots\: a_n)\in\mathbf{P}^n$ heißt \textit{Punkt}\index{Punkt} und die $a_i$ heißen \textit{homogene Koordinaten} von $P$.\index{Koordinate!homogen}
-\item Der Polynomring über $k$ in $n+1$ Variablen $S=k[X_0,\ldots,X_n]$ wird mit der folgenden Zerlegung zu einem graduierten Ring:
-\[S=\bigoplus_{d\geq 0}S_d,\quad S_d=\Big\{ \sum a_{i_0,\ldots,i_n} X_0^{i_0}\cdots X_n^{i_n}\Bigm| a_{i_0,\ldots,i_n}\in k,\ \sum_{j=0}^n i_j=d \Big\} \]
-Die Elemente in $S_d$ heißen \textit{homogene Elemente vom Grad $d$}\index{homogene Elemente}.
-Ein Ideal $\mathfrak{a}\subset S$ heißt \textit{homogenes Ideal}\index{homogenes Ideal}, wenn $\mathfrak{a}=\bigoplus_{d\geq 0}(S_d\cap\mathfrak{a})$ gilt.
-\item Sei $\mathfrak{a}\subset S$ ein homogenes Ideal. Dann setzen wir:
-\[Z(\mathfrak{a}) = \{P\in\mathbf{P}^n\mid \forall f\in\mathfrak{a}\text{ homogen}\colon f(P)=0 \} \]
-Diese ist wohldefiniert, da $f(\lambda a_0,\ldots,\lambda a_n)=\lambda^d f(a_0,\ldots,a_n)$ für $f\in S_d$.
-Eine Menge $Y\subset\mathbf{P}^n$ heißt \textit{projektive algebraische Menge}\index{algebraische Menge!projektiv}, wenn $Y=Z(\mathfrak{a})$ für ein homogenes Ideal $\mathfrak{a}\subset S$ gilt.
-
-Analog wie im affinen Fall, können wir auch $\mathbf{P}^n$ mit der \textit{Zariski-Topologie} ausstatten, d.h.
-eine Menge $U\subset\mathbf{P}^n$ ist genau dann offen, wenn $\mathbf{P}^n\setminus U$ eine projektive algebraische Menge ist.\index{Zariski-Topologie}
-\item Eine \textit{projektive Varietät}\index{Varietät!projektiv} $V$ ist eine irreduzible abgeschlossene Menge in $\mathbf{P}^n$.
-Eine offene Teilmenge einer projektiven Varietät bzgl.
-der induzierten Topologie nennt man \textit{quasi-projektive Varietät}\index{Varietät!quasi-projektiv}.
-\item Sei $Y\subset\mathbf{P}^n$ eine algebraische Menge. Dann setzen wir $I(Y)$ als das Ideal in $S$, das von der folgenden Menge erzeugt wird:
-\[\{f\in S \text{ homogen}\mid \forall P\in Y\colon f(P)=0\} \]
-$I(Y)$ ist ein homogenes Ideal in $S$. Der \textit{homogene Koordinatenring}\index{Koordinatenring!homogen} von $Y$ ist definiert als $S(Y)=S/I(Y)$.
-\end{enumerate}
-
-\paragraph{Satz.} Wir haben eine inklusionsumkehrende Bijektion:
-\[\{\text{Algebraische Mengen in }\mathbf{P}^n\}\to\{\text{Radikalideale in }k[X_0,\ldots,X_n]\},\ Y\mapsto I(Y) \]
-mit der Umkehrabbildung $\mathfrak{a}\mapsto Z(\mathfrak{a})$.
-
-\paragraph{Satz.} Sei $Y$ eine (quasi-)projektive Varietät. Dann wird $Y$ von offenen Mengen der Form $Y\cap U_i,\ i=0,\ldots,n$ überdeckt mit:
-\[U_i=\{(a_0\: \ldots\: a_n)\in \mathbf{P}^n\mid a_i\neq 0\} \]
-Die Abbildungen $\varphi_i: U_i\to\mathbf{A}^n,\ (a_0\: \ldots\: a_n)\mapsto \big(\frac{a_0}{a_i},\ldots,\widehat{\frac{a_i}{a_i}},\ldots,\frac{a_n}{a_i}\big)$ sind wohldefiniert und Ho\-möo\-mor\-phis\-men, d.h.
-die $Y\cap U_i$ sind (quasi-)affine Varietäten.
-
-\section{Morphismen}
-
-\paragraph{Definition.} \begin{enumerate}[(i)]
-\item Sei $Y\subset\mathbf{A}^n$ eine quasi-affine Varietät.
-Eine Abbildung $f:Y\to k$ heißt \textit{reguläre Funktion} in $P\in Y$, wenn eine offene Umgebung $U\subset Y$ mit $P\in U$ existiert, so dass $f=\frac{g}{h}$ auf $U$ für gewisse $g,h\in A$ gilt.
-
-Sei $Y\subset\mathbf{P}^n$ eine quasi-projektive Varietät.
-Eine Abbildung $f:Y\to k$ heißt \textit{reguläre Funktion}\index{reguläre Funktion} in $P\in Y$, wenn eine offene Umgebung $U\subset Y$ mit $P\in U$ existiert, so dass $f=\frac{g}{h}$ auf $U$ für gewisse homogene Polynome $g,h\in S$ vom gleichen Grad.
-
-Identifizieren wir $k\cong \mathbf{A}^1$ so ist eine reguläre Funktion notwendigerweise stetig.
-\item Eine stetige Abbildung $\varphi:X\to Y$ zwischen zwei (quasi-projektive) Varietäten heißt \textit{Morphismus}\index{Morphismus!Varietäten}, wenn für jede offene Menge $V\subset Y$ und reguläre Funktion $f:V\to k$ auch $f\circ\varphi:\varphi^{-1}(V)\to k$ regulär ist.
-
-Damit erhält man die Kategorie $\textbf{Var$(k)$}$ aller Varietäten auf $k$.
-\end{enumerate}
-
-%\paragraph{Bemerkung.} Sei $Y$ eine Varietät und $f,g:Y\to k$ reguläre Funktionen.
-Existiert eine nichtleere offene Menge $U\subset Y$ mit $f=g$ auf $U$, so folgt $f=g$ auf ganz $Y$: Sei dazu o.B.d.A.
-$U$ so klein, dass $f$ und $g$ als Brüche homogener Polynome vom gleichen Grad dargestellt werden kann.
-Dann ist die Nullstellenmenge $\{P\in Y\mid (f-g)(P)=0\}$ abgeschlossen und dicht in $Y$.
-
-\paragraph{Definition.} Sei $Y$ eine Varietät und $P\in Y$ ein Punkt.
-\begin{enumerate}[(i)]
-\item Wir bezeichnen den Ring aller regulären Funktionen auf $Y$ mit $\mathcal{O}(Y)$.
-\item $\mathcal{O}_{P,Y}=\{\langle U,f\rangle\mid P\in U\subset_\text{o} Y,\ f\text{ ist auf $U$ regulär}\}$ heißt der Ring der \textit{Keime} regulärer Funktionen auf $Y$ in $P$.\index{Keim} Wir identifizieren zwei Keime $\langle U,f\rangle=\langle V,g\rangle$, wenn $f=g$ auf $U\cap V$ gilt.
-
-$\mathcal{O}_{P,Y}$ ist ein lokaler Ring, dessen Maximalideal wir mit $\mathfrak{m}_P$ bezeichnen.
-\item $K(Y)=\{\langle U,f\rangle\mid\varnothing\neq U\subset_\text{o}Y,\ f\text{ ist auf $U$ regulär}\}$ heißt der \textit{Funktionenkörper}\index{Funktionenkörper}  von $Y$.
-Die Elemente von $K(Y)$ heißen \textit{rationale Funktionen} auf $Y$\index{rationale Funktion}.
-\end{enumerate}
-Es gilt $\mathcal{O}(Y)\subset\mathcal{O}_{P,Y}\subset K(Y)$.
-
-\paragraph{Theorem.} Sei $Y\subset\mathbf{A}^n$ eine affine Varietät. Dann gilt:
-\begin{enumerate}[(i)]
-\item $\mathcal{O}(Y)\cong A(Y)$
-\item Die Abbildung $Y\to \{\text{Maximale Ideale in }A(Y)\},\ P\mapsto \mathfrak{m}_P\subset\mathcal{O}_{P,Y}$ ist eine Bijektion.
-\item $\mathcal{O}_{P,Y}\cong A(Y)_{\mathfrak{m}_P}$ und $\dim\mathcal{O}_{P,Y}=\dim Y$.
-\item $K(Y)\cong\operatorname{Quot}(A(Y))$
-\end{enumerate}
-
-\paragraph{Theorem.} Sei $Y\subset\mathbf{P}^n$ eine projektive Varietät. Dann gilt:
-\begin{enumerate}[(i)]
-\item $\mathcal{O}(Y)=k$
-\item $\mathcal{O}_{P,Y}\cong S(Y)_{(\mathfrak{m}_P)}$
-\item $K(Y)\cong S(Y)_{((0))}$
-\end{enumerate}
-
-\paragraph{Theorem.} Sei $X$ eine beliebige Varietät und $Y$ eine affine Varietät. Dann haben wir eine Bijektion:
-\[\operatorname{Mor}_{\textbf{Var}}(X,Y) \to \operatorname{Hom}_{\textbf{$k$-Alg}}(A(Y),\mathcal{O}(X)),\ f\mapsto(\varphi\mapsto \varphi\circ f) \]
-Ist $X$ ebenfalls affin, so gilt $X\cong Y$, genau dann wenn $A(X)\cong A(Y)$.
-Der Funktor $\textbf{affine Var$(k)$}\to\textbf{nullteilerfreie $k$-Alg},\ X\mapsto A(X)$ ist eine pfeilumkehrende Äquivalenz von Kategorien.
-
-\chapter{Schemata}
-\section{Garben}
-
-\paragraph{Definition 1.1.}\label{1.1} Sei $X$ ein topologischer Raum.
-\begin{enumerate}[(i)]
-\item Die Menge aller offenen Teilmengen in $X$ bilden zusammen mit den natürlichen Inklusionen eine Kategorie $\textbf{Top$(X)$}$.
-\item Eine \textit{Prägarbe}\index{Prägarbe} $F$ abelscher Gruppen ist nichts anderes als ein kontravarianter Funk\-tor $F:\textbf{Top$(X)$}\to\textbf{Ab}$ mit $F(\varnothing)=0$.
-\end{enumerate}
-
-\paragraph{Bemerkung.}\begin{enumerate}
-\item Eine Prägarbe besteht also aus abelschen Gruppen $F(U),\ U\subset_\text{o}X$ und Homomorphismen abelscher Gruppen $\operatorname{res}_V^U:F(U)\to F(V)$ für alle offenen $V\subset U$, so dass $\operatorname{res}_U^U=\operatorname{id}_{F(U)}$.
-Für offene Mengen $W\subset V\subset U$ gelte ferner $\operatorname{res}_W^U=\operatorname{res}_W^V\circ\operatorname{res}_V^U$.
-\item Ebenso können Prägarben in eine beliebige Kategorie gebildet werden, z.B. $\textbf{Ringe}$ und $\textbf{Mengen}$.
-\item Die Elemente von $F(U)$ heißen \textit{Schnitte} von $F$ über $U$.\index{Schnitt} Manchmal schreiben wir auch $\Gamma(U,F)=F(U)$.
-Die $\operatorname{res}_V^U$ heißen \textit{Restriktionsabbildungen}\index{Restriktionsabbildung}.
-Wir schreiben auch $\operatorname{res}_V^U(s)=s|_V$.
-\end{enumerate}
-
-\paragraph{Definition 1.2.}\label{1.2} Eine Prägarbe $F$ auf einem topologischen Raum $X$ heißt \textit{Garbe}\index{Garbe}, falls die folgenden Diagramme exakt sind:
-\[\begin{tikzcd}
-0\ar[r] & F(U)\ar[r, "\operatorname{res}"] & \displaystyle\prod_i F(U_i)\ar[r, "\operatorname{res}", shift left] \ar[r, shift right] & \displaystyle\prod_{i,j}F(U_i\cap U_j)
-\end{tikzcd} \]
-für alle $U\subset_\text{o}X$ und jede offene Überdeckung $U=\bigcup_i U_i$, d.h:
-\begin{enumerate}[(i)]
-\item $s\mapsto (\operatorname{res}^U_{U_i}(s))_i$ ist injektiv, d.h. aus $s|_{U_i}=0$ für alle $i$ folgt $s=0$.
-\item Sei $s_i\in F(U_i)$ für alle $i$ gegeben, so dass $s_i|_{U_i\cap U_j}=s_j|_{U_i\cap U_j}$ für alle $i,j$. Dann gibt es ein $s\in F(U)$ mit $s|_{U_i}=s_i$ für alle $i$.
-\end{enumerate}
-
-\paragraph{Definition 1.3.}\label{1.3} \begin{enumerate}[(i)]
-\item Ein \textit{Morphismus}\index{Morphismus!Prägarben} $\varphi:F\to G$ von Prägarben auf $X$ ist ein Morphismus kontravarianter Funktoren, d.h.
-eine Kollektion von Morphismen $(\varphi(U))_{U\subset_\text{o}X}$, so dass für alle offenen Mengen $V\subset U$ folgendes Diagramm kommutativ ist:
-\[\begin{tikzcd}
-F(U)\ar[rr, "\varphi(U)"]\ar[d, "\operatorname{res}^U_V"'] && G(U)\ar[d, "\operatorname{res}^U_V"]\\
-F(V)\ar[rr, "\varphi(V)"'] && G(V)
-\end{tikzcd} \]
-\item Ein \textit{Morphismus}\index{Morphismus!Garben} von Garben ist ein Morphismus von Prägarben. Die (Prä-)Garben bilden eine Kategorie.
-\end{enumerate}
-
-\paragraph{Beispiel 1.4.}\label{1.4} \begin{enumerate}
-\item Sei $X$ eine Varietät über $k$.
-Betrachte den Funktor $\mathcal{O}:\textbf{Top$(X)$}\to\textbf{komm Ringe}$ mit den gewöhnlichen Restriktionsabbildungen $\operatorname{res}^U_V:\mathcal{O}(U)\to\mathcal{O}(V)$ ist offensichtlich eine Prägarbe von Ringen.
-Da ferner reguläre Funktionen $0$ ist, wenn sie lokal $0$ ist, und eine lokal reguläre Funktion auch global regulär ist, ist $\mathcal{O}$ auch eine Garbe.
-\item Sei $X$ ein topologischer Raum und $A$ eine abelsche Gruppe.
-Die \textit{konstante Garbe}\index{Garbe!konstant} $\mathcal{A}$ auf $X$ ist folgendermaßen definiert: Wir statten $A$ mit der diskreten Topologie aus.
-Für jedes $U\subset_\text{o}X$ setze:
-\[\mathcal{A}(U)=\{f:U\to A\mid f\text{ stetig}\} \]
-Ist $U$ zusammenhängend, so gilt $\mathcal{A}(U)\stackrel{\sim}{\to} A,\ f\mapsto f(x)$, wobei $x\in U$ beliebig.
-\end{enumerate}
-
-\paragraph{Definition 1.5.}\label{1.5} Sei $F$ eine Prägarbe auf $X$ und $P\in X$. Der \textit{Halm}\index{Halm} $F_P$ von $F$ in $P$ ist definiert als:
-\[F_P =\varinjlim_{\substack{U\subset_\text{o}X\\ P\in U}} F(U)=\coprod_{\substack{U\subset_\text{o}X\\ P\in U}} F(U)\Big/{\sim} \]
-wobei zwei Elemente $s\in F(U),\ t\in F(V)$ genau dann äquivalent $s\sim t$ sind, wenn es ein $\varnothing \neq W\subset_\text{o}X$ mit $W\subset U\cap V$ existiert, so dass $s|_W=t|_W$ gilt.
-Die Elemente eines Halms heißen \textit{Keime}\index{Keim} der Schnitte von $F$ in $P$.
-
-\paragraph{Beispiel.} Sei $X$ eine Varietät, $P\in X$ ein Punkt und $\mathcal{O}$ die Garbe der regulären Funktionen. Dann ist der Halm in $P$ gerade der lokale Ring $\mathcal{O}_{P,X}$.
-
-\paragraph{Bemerkung.} Ein Morphismus $\phi:F\to G$ von Prägarben induziert für alle $P\in X$ ein Gruppenhomomorphismus $\phi_P:F_P\to G_P$.
-
-\paragraph{Satz 1.6.}\label{1.6} Sei $\phi:F\to G$ ein Morphismus von Garben auf einem topologischen Raum $X$. Dann gilt:
-\[\phi:F\to G\text{ ist Isomorphismus}\iff \phi_P:F_P\to G_P\text{ ist Isomorphismus für alle }P\in X \]
-Für Prägarben gilt dieser Satz im Allgemeinen nicht.
-
-\paragraph{Beweis.} Ist $\phi$ ein Isomorphismus, so auch alle $\phi_P,\ P\in X$.
-Sei umgekehrt $\phi_P$ Isomorphismen für alle $P\in X$.
-Es genügt zu zeigen, dass $\phi(U):F(U)\to G(U)$ für alle $U\subset_\text{o}X$ ein Isomorphismus ist.
-Sei $U\subset_\text{o}X$ und setze $\varphi=\phi(U)$.
-\begin{itemize}
-\item \textit{Injektivität von $\varphi$:} Sei $s\in F(U)$ mit $0=\varphi(s)\in G(U)$.
-Dann gilt für das Bild $\varphi(s)_P$ von $\varphi(s)$ im Halm $0=\varphi(s)_P\in F_P$.
-Wegen $\varphi(s)_P=\phi_P(s_P)$ für das Bild $s_P\in F_P$ von $s$, folgt wegen der Injektivität von $\phi_P$ nun $s_P=0$ für alle $P\in U$.
-
-Per Definition gibt es für jedes $P\in U$ eine offene Umgebung $W_P\subset_\text{o}X$ von $P$ mit $W_P\subset U$, so dass $s|_{W_P}=0$ gilt.
-Dann bilden die $W_P$ eine offene Überdeckung von $U=\bigcup_{P\in U}W_P$.
-Da $F$ eine Garbe ist, folgt $s=0$.
-
-Wir haben gezeigt, dass $\phi(U)$ für alle $U\subset_\text{o}X$ injektiv ist, genau dann wenn $\phi_P$ für alle $P\in X$ injektiv ist.
-\item \textit{Surjektivität von $\varphi$:} Sei $t\in G(U)$ ein Schnitt und $t_P\in G_P$ sein Keim in $P$.
-Da $\phi_P$ surjektiv ist, existiert ein $s_P\in F_P$ mit $\phi_P(s_P)=t_P$.
-Sei $s_P$ durch den Schnitt $s(P)\in F(V_P)$ mit $V_P\subset_\text{o}U,\ P\in V_P$ repräsentiert.
-Dann sind $\phi(V_P)(s(P))$ und $t|_{V_P}$ zwei Elemente aus $G(V_P)$ mit demselben Keim.
-Durch das Verkleinern von $V_P$ folgt $\phi(V_P)(s(P))=t|_{V_P}$ in $G(V_P)$.
-
-Dann bilden die $V_P$ eine offene Überdeckung von $U=\bigcup_{P\in U}V_P$.
-Es gilt außerdem $s(P)|_{V_P\cap V_Q}=s(Q)|_{V_P\cap V_Q}$ für alle $P,Q\in U$, denn beide Elemente sind Schnitte aus $F(V_P\cap V_Q)$, die durch $\phi(V_P\cap V_Q)$ auf $t|_{V_P\cap V_Q}$ abgebildet werden, und $\phi(V_P\cap V_Q)$ aus dem ersten Teil injektiv ist.
-
-Da $F$ eine Garbe ist, existiert ein $s\in F(U)$ mit $s|_{V_P}=s(P)$ für alle $P\in U$.
-Schließlich gilt $\varphi(s)|_{V_P}=t|_{V_P}$ für alle $P\in U$, d.h.
-$(\varphi(s)-t)|_{V_P}=0$.
-Da $G$ eine Garbe ist, folgt $\varphi(s)=t$.\qed
-\end{itemize}
-
-\paragraph{Definition 1.7.}\label{1.7} Sei $\varphi:F\to G$ ein Morphismus von Prägarben. Die Prägarben
-\[U\mapsto \ker\varphi(U),\quad U\mapsto\operatorname{coker}\varphi(U),\quad U\mapsto\operatorname{im}\varphi(U) \]
-heißen \textit{Prägarbenkern}, \textit{-kokern} und \textit{-bild} von $\varphi$.\index{Kern}\index{Kokern}\index{Bild} Sind $F$ und $G$ Garben, so sind Kokern und Bild nicht notwendig Garben.
-
-\paragraph{Satz \& Definition 1.8.}\label{1.8} Sei $F$ eine Prägarbe. Dann existiert eine Garbe $F^+$ und ein Morphismus von Prägarben $\theta:F\to F^+$ mit folgender Universaleigenschaft:
-
-Sei $G$ eine Garbe und $\phi:F\to G$ ein Morphismus von Prägarben. Dann existiert ein eindeutig bestimmter Morphismus $\psi:F^+\to G$, so dass das folgende Diagramm kommutiert:
-\[\begin{tikzcd} 
-F\ar[rr, "\theta"]\ar[d, "\phi"'] && F^+\ar[dll, "\psi", dashed]\\
-G
-\end{tikzcd} \]
-$F^+$ ist somit eindeutig bestimmt und heißt die zu $F$ \textit{assoziierte Garbe}\index{assoziierte Garbe!Prägarbe}.
-
-\paragraph{Beweis.} Für jede offene Menge $U\subset X$ setze $F^+(U)$ als die Menge aller Abbildungen $s:U\to\coprod_{P\in U}F_P$, so dass:
-\begin{enumerate}[(i)]
-\item Für alle $P\in U$ gilt $s(P)\in F_P$.
-\item Für alle $P\in U$ gibt es eine offene Umgebung $V$ von $P$ mit $V\subset U$ und ein Element $t\in F(V)$, so dass für alle $Q\in V$ der Keim $t_Q$ von $t$ in $Q$ gleich $s(Q)$ ist.
-\end{enumerate}
-Somit wird $F^+$ zu einer Garbe bzgl.
-der natürlichen Restriktionsabbildungen und besitzt die verlangte Universaleigenschaft.
-Für jeden Punkt $P\in X$ gilt $F^+_P=F_P$.
-Ist $F$ eine Garbe, so ist $F^+\cong F$ via $\theta$.\qed
-
-\paragraph{Definition 1.9.}\label{1.9} 
-\begin{enumerate}[(i)]
-\item Eine \textit{Untergarbe}\index{Garbe!Untergarbe} von $F$ ist eine Garbe $F'$ derart, dass:
-\begin{enumerate}[(a)]
-\item $F'(U)\subset F(U)$ ist eine Untergruppe für alle $U\subset_\text{o}X$.
-\item Für offene Mengen $V\subset U$ gilt $\operatorname{res}'{}^U_V=\operatorname{res}^U_V|_{F'(U)}$.
-\end{enumerate}
-Insbesondere ist $F'_P\subset F_P$ eine Untergruppe.
-\item Der \textit{Kern}\index{Kern} von $\varphi$ ist die Prägarbe $\ker(\varphi)$, die bereits eine Garbe ist. \textit{Grund:}
-
-Sei $U\subset_\text{o}X$ und $U=\bigcup U_i$ eine offene Überdeckung.
-Sei $s\in\ker\varphi(U)$ mit $s|_{U_i}=0$ für alle $i$.
-Da $F$ eine Garbe ist und $s\in F(U)$, folgt $s=0$.
-Sei nun $s_i\in \ker\varphi(U_i)$ für alle $i$ gegeben mit $s_i|_{U_i\cap U_j}=s_j|_{U_i\cap U_j}$ für alle $i,j$.
-Da $F$ eine Garbe ist, existiert ein $s\in F(U)$ mit $s|_{U_i}=s_i$ für alle $i$.
-Zu zeigen ist noch $s\in \ker\varphi(U)$.
-Es gilt für alle $i$:
-\[0=\varphi(U_i)(s_i)=\varphi(U_i)(s|_{U_i}) = \varphi(U)(s)|_{U_i}\in G(U_i) \]
-Da nun auch $G$ eine Garbe ist, folgt $\varphi(U)(s)=0$.
-\item $\varphi$ heißt \textit{injektiv}\index{Injektivität}, falls $\ker(\varphi)=0$.
-
-Mit anderen Worten: $\varphi$ ist genau dann injektiv, wenn $\varphi(U):F(U)\to G(U)$ für alle $U\subset_\text{o}X$ injektiv ist.
-\item Das \textit{Bild}\index{Bild} $\operatorname{im}(\varphi)$ von $\varphi$ ist die assoziierte Garbe des Prägarbenbilds von $\varphi$.
-
-Nach der Universaleigenschaft gibt es einen natürlichen Morphismus $\psi:\operatorname{im}(\varphi)\to G$.
-Dieser ist injektiv, da $(\operatorname{im}\varphi)_P:\operatorname{im}(\varphi_P) \to G_P$ für alle $P\in X$ injektiv ist.
-\item $\varphi$ heißt \textit{surjektiv}\index{Surjektivität}, wenn $\operatorname{im}(\varphi)=G$.
-\item Eine Garbensequenz
-\[\begin{tikzcd}
-\cdots \ar[r] & F^i\ar[r, "\varphi^i"] & F^{i+1}\ar[r, "\varphi^{i+1}"] & F^{i+2}\ar[r] & \cdots 
-\end{tikzcd} \]
-heißt \textit{exakt}\index{Exaktheit}, falls $\ker(\varphi^{i+1})=\operatorname{im}(\varphi^i)$ für alle $i$ gilt.
-\item Sei $F'$ eine Untergarbe von $F$.
-Die \textit{Quotientengarbe}\index{Quotientengarbe} $F/F'$ ist die assoziierte Garbe zur Prägarbe $U\mapsto F(U)/F'(U)$.
-Offensichtlich gilt $(F/F')_P=F_P/F'_P$ für alle $P\in X$.
-\item Der \textit{Kokern}\index{Kokern} von $\varphi$ ist die assoziierte Garbe zum Prägarbenkokern von $\varphi$.
-\end{enumerate}
-
-\paragraph{Regeln 1.10.}\label{1.10} Seien $F,G$ Garben auf $X$ und $\varphi:F\to G$ ein Morphismus von Garben.
-\begin{enumerate}[(i)]
-\item $\varphi$ ist genau dann injektiv, wenn $0\to F\to G$ exakt ist und genau dann surjektiv, wenn $F\to G\to 0$ exakt ist.
-\item Eine Garbensequenz $\cdots \to F^i\to F^{i+1}\to F^{i+2}\to\cdots$ ist genau dann exakt, wenn ihre entsprechenden Halmsequenzen in allen Punkten $P\in X$ exakt ist. \textit{Grund:}
-\[(\operatorname{im} \varphi^i)_P=\operatorname{im}(\varphi^i_P),\quad (\ker\varphi^{i+1})_P=\ker(\varphi^{i+1}_P) \]
-Insbesondere ist ein Garbenmorphismus genau dann injektiv bzw. surjektiv falls alle Halmabbildungen injektiv bzw. surjektiv sind.
-\item $\varphi$ ist genau dann surjektiv, wenn für alle $U\subset_\text{o}X$ und $s\in G(U)$ eine Überdeckung $U=\bigcup_{i\in I} U_i$ mit Schnitten $t_i\in F(U_i),\ i\in I$ existieren, so dass $\varphi(t_i)=s|_{U_i}$.
-
-Ist $\varphi$ surjektiv, so muss im Allgemeinen $\varphi(U):F(U)\to G(U)$ nicht surjektiv sein.
-\item Sei $0\to F'\to F\to F''$ eine exakte Garbensequenz und $U\subset_\text{o}X$. Dann ist auch die folgende Sequenz exakt:
-\[\begin{tikzcd}[cramped, sep=small]
-0 \ar[r] & \Gamma(U,F') \ar[r]& \Gamma(U,F) \ar[r]& \Gamma(U,F'')
-\end{tikzcd} \]
-Der Funktor $\Gamma(U,-)$ ist linksexakt, aber nicht exakt.
-\item Sei $\varphi:F\to G$ ein injektiver Morphismus von Prägarben. Dann ist der induzierte Morphismus $\varphi^+:F^+\to G^+$ der assoziierten Garben auch injektiv. Der Funktor $-^+$ ist sogar exakt.
-\end{enumerate}
-
-\paragraph{Regeln 1.11.}\label{1.11} \begin{enumerate}[(i)]
-\item Sei $F'$ eine Untergarbe von $F$. Dann ist die Sequenz $0\to F'\to F\to F/F'\to 0$ exakt, da sie halmweise exakt ist.
-\item Sei $\varphi:F\to G$ ein Garbenmorphismus. Dann gilt:
-\[\operatorname{im}(\varphi)\cong F/\ker(\varphi),\quad \operatorname{coker}(\varphi)\cong G/ \operatorname{im}(\varphi) \]
-d.h. die Folgen $0\to\ker(\varphi)\to F\to\operatorname{im}(\varphi)\to 0$ und $0\to\operatorname{im}(\varphi)\to G\to\operatorname{coker}(\varphi)\to 0$ sind exakt.
-\end{enumerate}
-
-\paragraph{Definition 1.12.}\label{1.12} \begin{enumerate}[(i)]
-\item Seien $F$ und $G$ Garben auf $X$. Die \textit{Summe}\index{Summe} $F\oplus G$ von $F$ und $G$ ist die Garbe $U\mapsto F(U)\oplus G(U)$.
-\item Sei $(F_i,\varphi_{i,j})$ ein direktes System von Garben auf $X$.
-Der \textit{direkte Limes}\index{Limes!direkt} $(\varinjlim F_i,\varphi_i)$ ist die assoziierte Garbe zur Prägarbe $U\mapsto \varinjlim F_i(U)$.
-
-Der direkte Limes besitzt die übliche Universaleigenschaft: Sei $G$ eine Garbe und $\psi_i:F_i\to G$ Morphismen mit $\varphi_k\varphi_{ik}=\psi_i$ für alle $i\leq k$.
-Dann existiert ein eindeutig bestimmter Morphismus $\psi:\varinjlim F_i\to G$, so dass das folgende Diagramm kommutiert:
-\[\begin{tikzcd}
-F_i\ar[rr, "\varphi_i"]\ar[d, "\psi_i"'] & & \varinjlim F_i\ar[lld, "\psi", dashed]\\
-G
-\end{tikzcd} \]
-\item Ebenso wird der \textit{projektive Limes}\index{Limes!projektiv} definiert, wobei alle Pfeile umgedreht werden. Es ist außerdem $U\mapsto\varprojlim F_i(U)$ bereits eine Garbe.
-\item Sei $F$ eine Garbe auf $X$ und $s\in F(U)$ ein Schnitt über $U\subset_\text{o}X$. Dann heißt
-\[\operatorname{Supp}(s)=\{P\in U\mid s_P\neq 0\} \]
-wobei $s_P\in F_P$ der Keim von $s$ in $P$ bezeichnet, der \textit{Support}\index{Support} von $s$. $\operatorname{Supp}(s)$ ist abgeschlossen in $U$.
-\[\operatorname{Supp}(F)=\{P\in X\mid F_P\neq 0\} \]
-heißt \textit{Support} von $F$. Dieser ist nicht notwendigerweise abgeschlossen.
-\item Seien $F,G$ Garben abelscher Gruppen auf $X$.
-Für ein $U\subset_\text{o}X$ sei $F|_U$ die Einschränkung von $F$ auf $U$, d.h.
-$F|_U(V)=F(V)$ für alle $V\subset_\text{o}U$.
-Dann ist die Menge $\operatorname{Hom}(F|_U,G|_U)$ der Morphismen von $F|_U$ nach $G|_U$ eine abelsche Gruppe.
-\[U\mapsto\operatorname{Hom}(F|_U,G|_U)\]
-definiert eine Garbe und wird die \textit{Hom-Garbe}\index{Hom-Garbe} genannt. Sie wird mit $\operatorname{\mathcal{H}om}(F,G)$ bezeichnet.
-\end{enumerate}
-
-\paragraph{Definition 1.13.}\label{1.13} Sei $f:X\to Y$ eine stetige Abbildung topologischer Räume, $F$ eine Garbe auf $X$ und $G$ eine Garbe auf $Y$.
-\begin{enumerate}[(i)]
-\item Die \textit{direkte Bildgarbe}\index{direkte Bildgarbe} $f_\ast F$ von $F$ auf $Y$ ist die Garbe
-\[V\mapsto (f_\ast F)(V)=F(f^{-1}(V)) \]
-\item Die \textit{Urbildgarbe}\index{Urbildgarbe} $f^{-1}G$ von $G$ auf $X$ ist die assoziierte Garbe zur Prägarbe
-\[U\mapsto (f^{-1}G)(U)= \varinjlim_{\substack{V\subset_\text{o}Y\\ f(U)\subset V}}G(V) \]
-\end{enumerate}
-
-\paragraph{Regeln 1.14.}\label{1.14} Seien $X,Y$ topologische Räume.
-\begin{enumerate}[(i)]
-\item Sei $Z\subset X$ ein Teilraum mit der Inklusionsabbildung $i:Z\hookrightarrow X$ und $F$ eine Garbe auf $X$. Dann gilt:
-\[i^{-1}F = F|_Z \]
-Offensichtlich gilt $(F|_Z)_P=F_P$ für $P\in Z$.
-\item Seien $\mathbf{Ab}(X)$ und $\mathbf{Ab}(Y)$ die Kategorien der Garben auf $X$ bzw. $Y$. Dann sind $f_\ast:\mathbf{Ab}(X)\to\mathbf{Ab}(Y)$ und $f^{-1}:\mathbf{Ab}(Y)\to\mathbf{Ab}(X)$ Funktoren.
-\item Sei $f:X\to Y$ stetig. Dann sind die \textit{Adjunktionsabbildungen}\index{Adjunktionsabbildung} 
-\[\operatorname{ad}:f^{-1}f_\ast F\to F,\quad\operatorname{ad} :G\to f_\ast f^{-1}G \]
-für Garben $F$ auf $X$ bzw. $G$ auf $Y$ Garbenmorphismen und wie folgt definiert:
-\begin{itemize}
-\item Sei $U\subset_\text{o}X$ und $s\in (f^{-1}f_\ast F)(U)$ ein Schnitt, das durch $s'\in (f_\ast F)(V)$ mit $f(U)\subset V\subset_\text{o}Y$ repräsentiert wird, d.h.
-$s'\in F(f^{-1}(V))$ mit $U\subset f^{-1}(V)$.
-Dann setzen wir:
-\[(f^{-1}f_\ast F)(U)\to F(U),\ s\mapsto \operatorname{res}_U^{f^{-1}(V)}s'\in F(U)\]
-\item Sei $V\subset_\text{o}Y$. Es besteht $(f_\ast f^{-1}G)(V)=(f^{-1}G)(f^{-1}(V))$ aus Abbildungen der Form $f^{-1}(V)\to\coprod_{P\in f^{-1}(V)}(f^{-1}G)_P$. Setze nun:
-\[G(V)\to (f_\ast f^{-1}G)(V),\ s\mapsto (s\circ f: f^{-1}(V)\to\coprod  f^{-1}(G)_P,\ P\mapsto s_{f(P)}) \]
-\end{itemize}
-Es existiert eine natürliche Bijektion:
-\[\operatorname{Hom}_X(f^{-1}G,F)\cong\operatorname{Hom}_Y(G,f_\ast F) \]
-in dem wir ein $\varphi:f^{-1}(G)\to F$ auf $\psi: G\stackrel{\operatorname{ad}}{\longrightarrow} f_\ast f^{-1}G\stackrel{f_\ast(\varphi)}{\longrightarrow} f_\ast F$ schicken und ein $\psi:G\to f_\ast F$ auf $\varphi:f^{-1}G \stackrel{f^{-1}(\psi)}{\longrightarrow} f^{-1}f_\ast F\stackrel{\operatorname{ad}}{\longrightarrow}F$ schicken.
-Somit ist $f^{-1}$ linksadjungiert zu $f_\ast$.
-\end{enumerate}
-
-\paragraph{Definition 1.15.}\label{1.15} Sei $X$ ein topologischer Raum mit $P\in X$ und $A$ eine abelsche Gruppe.
-Sei $\mathcal{A}$ die konstante Garbe auf $\overline{\{P\}}$ und $i:\overline{\{P\}}\hookrightarrow X$ die natürliche Inklusion.
-Dann heißt die Garbe $i_\ast \mathcal{A}$ \textit{Wolkenkratzergarbe}.
-Es gilt:
-\[(i_\ast\mathcal{A})(U) = \begin{cases}
-A, &\text{wenn } P\in U\\
-0, &\text{sonst}
-\end{cases},\quad (i_\ast \mathcal{A})_Q = \begin{cases}
-A,&\text{wenn } Q\in\overline{\{P\}}\\
-0,&\text{sonst}
-\end{cases} \]
-
-\paragraph{Definition 1.16.}\label{1.16} Sei $X$ ein topologischer Raum, $Z\subset X$ abgeschlossen und $U=X\setminus Z$.
-Seien $j:U\hookrightarrow X$, $i:Z\hookrightarrow X$ die Inklusionsabbildungen.
-Ist $F$ eine Garbe auf $Z$, so gilt:
-\[(i_\ast F)_P=\begin{cases}
-F_P, &\text{wenn }P\in Z\\
-0, &\text{sonst}
-\end{cases} \]
-Sei $F$ eine Garbe auf $U$ und $j_!(F)$ die Garbe auf $X$, die zur Prägarbe
-\[V\mapsto\begin{cases}
-F(V),&\text{wenn }V\subset U\\
-0,&\text{sonst}
-\end{cases} \]
-assoziiert ist. Sie heißt die \textit{außerhalb $U$ durch Null fortgesetzte Garbe}\index{Garbe!Fortsetzung} von $F$. Es gilt:
-\[(j_!F)_P=\begin{cases}
-F_P,&\text{wenn }P\in U\\
-0,&\text{sonst}
-\end{cases} \]
-Sei $F$ eine Garbe auf $X$. Dann ist die folgende Sequenz exakt:
-\[\begin{tikzcd}[cramped, sep=small]
-0\ar[r] & j_!(F|_U)\ar[r] & F \ar[r]& i_\ast(F|_Z)\ar[r] & 0
-\end{tikzcd} \]
-
-\section{Schemata}
-
-Sei $A$ ein kommutativer Ring mit Eins und $\operatorname{Spec}(A)$ die Menge aller Primideale von $A$. Für ein Ideal $\mathfrak{a}\subset A$, setzen wir:
-\[V(\mathfrak{a}) = \{\mathfrak{p}\in\operatorname{Spec}(A)\mid \mathfrak{a}\subset\mathfrak{p}\} \]
-
-\paragraph{Lemma 2.1.}\label{2.1} \begin{enumerate}[(i)]
-\item Sind $\mathfrak{a},\mathfrak{b}$ zwei Ideale von $A$, so gilt $V(\mathfrak{ab})=V(\mathfrak{a})\cup V(\mathfrak{b})$.
-\item Sind $\mathfrak{a}_i\subset A,\ i\in I$ Ideale, so gilt $V\big(\sum\mathfrak{a}_i\big)=\bigcap V(\mathfrak{a}_i)$.
-\item Sind $\mathfrak{a},\mathfrak{b}\subset A$ Ideale, gilt: $V(\mathfrak{a})\subset V(\mathfrak{b}) \iff \operatorname{Rad}(\mathfrak{a})\supset\operatorname{Rad}(\mathfrak{b})$
-\end{enumerate}
-Wegen $V(A)=\varnothing$ und $V(0)=\operatorname{Spec}(A)$ sehen wir, dass wir Teilmengen der Form $V(\mathfrak{a})$ zu abgeschlossene Mengen in $\operatorname{Spec}(A)$ erklären können.
-Somit erhalten wir die \textit{Zariski-Topologie}\index{Zariski-Topologie} auf $\operatorname{Spec}(A)$.
-
-Setzen wir $D(f)=\operatorname{Spec}(A)\setminus V(f)$ für $f\in A$, so bilden diese offene Mengen eine Basis der Topologie auf $\operatorname{Spec}$.
-
-\paragraph{Definition 2.2.}\label{2.2} Wir definieren eine Ringgarbe $\mathcal{O}$ auf $\operatorname{Spec}(A)$ wie folgt: Sei $U\subset_\text{o}\operatorname{Spec}(A)$.
-Setze $\mathcal{O}(U)$ als die Menge aller Abbildungen $s:U\to\coprod_{\mathfrak{p}\in U}A_\mathfrak{p}$, so dass:
-\begin{enumerate}[(i)]
-\item Für alle $\mathfrak{p}\in U$ gilt $s(\mathfrak{p})\in A_\mathfrak{p}$.
-\item Für alle $\mathfrak{p}\in U$ gibt es eine offene Umgebung $V$ von $\mathfrak{p}$ mit $V\subset U$ und Elemente $a,f\in A$, so dass für alle $\mathfrak{q}\in V$ stets $f\not\in\mathfrak{q}$ und $s(\mathfrak{q})=\frac{a}{f}$ in $A_\mathfrak{q}$ gilt.
-\end{enumerate}
-Offensichtlich ist mit $s,t\in\mathcal{O}(U)$ auch $s+t,st\in\mathcal{O}(U)$.
-Ferner ist für $V\subset U$ offen $\operatorname{res}^U_V:\mathcal{O}(U)\to\mathcal{O}(V)$ ein Ringhomomorphismus.
-$\mathcal{O}$ heißt \textit{Strukturgarbe}\index{Strukturgarbe}.
-Das \textit{Spektrum}\index{Spektrum} von $A$ ist das Paar $(\operatorname{Spec}A,\mathcal{O})$.
-
-\paragraph{Satz 2.3.}\label{2.3} Sei $A$ ein Ring und $(\operatorname{Spec} A,\mathcal{O})$ sein Spektrum. Dann gilt:
-\begin{enumerate}[(i)]
-\item Der Halm $\mathcal{O}_\mathfrak{p}$ ist isomorph zu $A_\mathfrak{p}$ für alle $\mathfrak{p}\in\operatorname{Spec}(A)$.
-\item $\mathcal{O}(D(f))\cong A_f$ für alle $f\in A$
-\item $\Gamma(\operatorname{Spec}(A),\mathcal{O})=A$
-\end{enumerate}
-
-\paragraph{Beweis.} (iii) folgt aus (ii) mit $f=1$.
-\begin{enumerate}[(i)]
-\item Die Abbildungen $\mathcal{O}(U)\to A_\mathfrak{p},\ s\mapsto s(\mathfrak{p})$ mit $\mathfrak{p}\in U\subset_\text{o}\operatorname{Spec}(A)$ sind kompatibel und induzieren einen Homomorphismus $\varphi:\mathcal{O}_\mathfrak{p}\to A_\mathfrak{p}$.
-\begin{itemize}
-\item \textit{Surjektivität:} Sei $\frac{a}{f}\in A_\mathfrak{p}$ mit $a,f\in A,\ f\not\in\mathfrak{p}$.
-Dann ist $D(f)$ eine offene Umgebung von $\mathfrak{p}$ und es gibt ein $s\in\mathcal{O}(D(f))$ mit $s(\mathfrak{p})=\frac{a}{f}$.
-\item \textit{Injektivität:} Sei $U$ eine Umgebung von $\mathfrak{p}$ und $s,t\in\mathcal{O}(U)$ mit $s(\mathfrak{p})=t(\mathfrak{p})$.
-Verkleinern wir $U$ wenn nötig, so können wir o.B.d.A.
-annehmen, dass:
-\[s=\frac{a}{f},\quad t=\frac{b}{g}\quad \text{ für gewisse } a,b,g,f\in A,\ g,f\not\in\mathfrak{p} \]
-Wegen $\frac{a}{f}=\frac{b}{g}$ in $A_\mathfrak{p}$, gibt es ein $h\not\in\mathfrak{p}$, so dass $h(ga-bf)=0$ in $A$.
-Insbesondere ist $\frac{a}{f}=\frac{b}{g}$ in $A_\mathfrak{q}$ für alle $\mathfrak{q}\in\operatorname{Spec}(A)$ mit $g,f,h\not\in\mathfrak{q}$.
-Somit ist $s=t$ auf der offenen Umgebung $D(f)\cap D(g)\cap D(h)$ von $\mathfrak{p}$ und haben daher denselben Keim.
-\end{itemize}
-\item Sei $f\in A$ und $\mathfrak{p}\in D(f)$, d.h. $(f)\subset A\setminus\mathfrak{p}$. Betrachte die kanonische Abbildung $\lambda_\mathfrak{p}:A_f\to A_\mathfrak{p}$. Setze:
-\[\psi: A_f \to\mathcal{O}(D(f)),\ \frac{a}{f^n}\mapsto \Big(\mathfrak{p}\mapsto \lambda_\mathfrak{p}\Big(\frac{a}{f^n}\Big)\Big) \]
-\begin{itemize}
-\item \textit{Injektivität:} Sei $\psi\big(\frac{a}{f^n}\big)=\psi \big(\frac{b}{f^m}\big)$ und $\mathfrak{p}\in D(f)$.
-Dann ist $\frac{a}{f^n}=\frac{b}{f^m}$ in $A_\mathfrak{p}$, d.h.
-es gibt ein $h\not\in\mathfrak{p}$ mit $h(f^ma-f^nb)=0$.
-Setze $\mathfrak{a}=\operatorname{Ann}(f^ma-f^nb)$.
-Dann ist $\mathfrak{a}\not\subset\mathfrak{p}$, da $h\in\mathfrak{a}$, also folgt $\mathfrak{p}\not\in V(\mathfrak{a})$.
-Wir haben also $V(\mathfrak{a})\subset V(f)$ gezeigt.
-Nach \hyperref[2.1]{Lemma 2.1 (iii)} folgt $f\in\operatorname{Rad}(\mathfrak{a})$, d.h.
-$f^e\in\mathfrak{a}$ für ein $e>0$.
-Per Definition gilt $f^e(f^na-f^mb)=0$, also $\frac{a}{f^n}=\frac{b}{f^m}$ in $A_f$.
-\item \textit{Surjektivität:} Sei $s\in\mathcal{O}(D(f))$.
-Nach Definition von $\mathcal{O}$ ist $D(f)=\bigcup V_i$ mit $s=\frac{a_i}{g_i}$ auf $V_i$ für gewisse $a_i,g_i\in A,\ g_i\not\in\mathfrak{p}$ für alle $\mathfrak{p}\in V_i$.
-Insbesondere gilt $V_i\subset D(g_i)$.
-Da die $D(h)$ eine Basis der Topologie bilden, können wir o.B.d.A.
-$V_i=D(h_i)$ annehmen, also $D(h_i)\subset D(g_i)$.
-Es folgt $V(h_i)\supset V(g_i)$ und nach \hyperref[2.1]{Lemma 2.1 (iii)} auch $\operatorname{Rad}(h_i)\subset\operatorname{Rad}(g_i)$.
-Wähle ein $n$, so dass $h_i^n\in (g_i)$ für alle $i$, d.h.
-$h_i^n=c_ig_i$ für ein $c_i\in A$.
-Es folgt:
-\[\frac{a_i}{g_i}=\frac{c_ia_i}{h_i^n}\]
-Ersetzt man $a_i$ durch $c_ia_i$ und $h_i$ durch $h_i^n$, so können wir o.B.d.A. $D(f)\subset\bigcup D(h_i)$ und $s=\frac{a_i}{h_i}$ auf $D(h_i)$ annehmen.
-
-Wir zeigen nun, dass $D(f)$ durch endlich viele $D(h_i)$ überdeckt werden kann. Wir haben mit \hyperref[2.1]{Lemma 2.1} Äquivalenzen:
-\begin{align*}
-D(f)\subset \bigcup_{i}D(h_i) &\iff V(f)\supset\bigcap_i V(h_i) = V\Big(\sum_i (h_i)\Big)\\
-&\iff f\in\operatorname{Rad}\Big(\sum_i (h_i)\Big)\\
-&\iff \exists n\in\mathbb{N}\colon f^n\in\sum_i (h_i)
-\end{align*}
-Daher ist $f^n$ eine endliche Summe der Form $f^n=\sum_{i=1}^r b_ih_i$ für gewisse $b_i\in A$, d.h. $D(f)\subset D(h_1)\cup\cdots\cup D(h_r)$.
-
-Nun gilt: \[D(h_i)\cap D(h_j)=\operatorname{Spec}(A)\setminus (V(h_i)\cup V(h_j))=\operatorname{Spec}(A)\setminus V(h_ih_j)=D(h_ih_j)\]
-Auf $D(h_ih_j)$ wird $s$ repräsentiert durch $\frac{a_i}{h_i}$ und $\frac{a_j}{h_j}$ in $A_{h_ih_j}$.
-Wenden wir die Injektivität von $\psi$ auf $D(h_ih_j)$ an, so erhalten wir $\frac{a_i}{h_i}=\frac{a_j}{h_j}$ in $A_{h_ih_j}$.
-Es folgt $(h_ih_j)^n(h_ja_i-h_ia_j)=0$ für ein $m$.
-Sei $m$ so groß, dass dies für alle endlich vielen $i,j$ gilt, also gilt für alle $i,j$:
-\[h_j^{m+1}(h_i^ma_i)-h_i^{m+1}(h_j^ma_j)=0 \]
-Ersetzen wir nun $h_i$ durch $h_i^{m+1}$ und $a_i$ durch $a_ih_i^m$, so wird $s$ auf $D(h_i)$ immer noch durch $\frac{a_i}{h_i}$ repräsentiert und es gilt $h_ja_i=h_ia_j$ für alle $i,j$.
-
-Schreibe nun $f^n=\sum b_ih_i$ für ein $n$ und setze $a=\sum b_ia_i$. Es folgt für alle $j$:
-\[h_ja = \sum_i b_ia_ih_j=\sum _i b_ih_ia_j=f^na_j \]
-Also gilt $\frac{a}{f^n}=\frac{a_j}{h_j}$ auf $D(h_j)$ für alle $j$, d.h. $\psi\big(\frac{a}{f^n}\big)=s$.\qed
-\end{itemize}
-\end{enumerate}
-
-\paragraph{Definition 2.4.}\label{2.4} \begin{enumerate}[(i)]
-\item Ein \textit{geringter Raum}\index{geringter Raum} $(X,\mathcal{O}_X)$ besteht aus einem topologischen Raum $X$ und einer Ringgarbe $\mathcal{O}_X$ auf $X$.
-Ein \textit{Morphismus von geringten Räumen}\index{Morphismus!geringte Räume} ist ein Paar
-\[(f,f^\sharp):(X,\mathcal{O}_X)\to (Y,\mathcal{O}_Y) \]
-wobei $f:X\to Y$ eine stetige Abbildung und $f^\sharp:\mathcal{O}_Y\to f_\ast\mathcal{O}_X$ ein Morphismus von Ringgarben auf $Y$ ist.
-\item Ein geringter Raum $(X,\mathcal{O}_X)$ heißt \textit{lokal}\index{geringter Raum!lokal}, falls für alle $P\in X$ der Halm $\mathcal{O}_{X,P}$ ein lokaler Ring ist.
-Ein \textit{Morphismus von lokal geringten Räumen}\index{Morphismus!lokal geringte Räume} ist ein Morphismus $(f,f^\sharp)$ von geringten Räumen derart, dass für alle $P\in X$ die induzierte Abbildung $f_P^\sharp:\mathcal{O}_{Y,f(P)}\to\mathcal{O}_{X,P}$ lokale Homomorphismen sind.
-\end{enumerate}
-
-\paragraph{Bemerkung 2.5.}\label{2.5}\begin{itemize}
-\item Die (lokal) geringte Räume bilden eine Kategorie.
-\item Ein Morphismus $(f,f^\sharp)$ von (lokal) geringten Räumen ist genau dann ein Isomorphismus, wenn $f$ ein Homöomorphismus ist und $f^\sharp$ ein Garbenisomorphismus ist.
-\end{itemize}
-
-\paragraph{Satz 2.6.}\label{2.6} Seien $A,B$ Ringe.
-\begin{enumerate}[(i)]
-\item $(\operatorname{Spec}A,\mathcal{O})$ ist ein lokal geringter Raum.
-\item Sei $\varphi:A\to B$ ein Ringhomomorphismus. Dann induziert $\varphi$ einen natürlichen Morphismus von lokal geringten Räumen:
-\[(f,f^\sharp):(\operatorname{Spec}B,\mathcal{O}_B)\to (\operatorname{Spec}A,\mathcal{O}_A),\ f(\mathfrak{p})=\varphi^{-1}(\mathfrak{p}) \]
-\item Jeder Morphismus $(f,f^\sharp):(\operatorname{Spec}B,\mathcal{O}_B)\to(\operatorname{Spec}A,\mathcal{O}_A)$ von lokal geringten Räu\-men ist induziert von einem Ringhomomorphismus $\varphi:A\to B$.
-\end{enumerate}
-
-\paragraph{Beweis.} \begin{enumerate}[(i)]
-\item folgt aus \hyperref[2.3]{Satz 2.3 (i)}.
-\item Definiere $f$ durch $f(\mathfrak{p})=\varphi^{-1}(\mathfrak{p})$ für alle $\mathfrak{p}\in\operatorname{Spec}(B)$.
-Sei $\mathfrak{a}\subset A$ ein Ideal.
-Dann ist $f^{-1}(V(\mathfrak{a}))=V(\varphi(\mathfrak{a}))$.
-Daher ist $f$ stetig.
-Sei $\mathfrak{p}\in\operatorname{Spec}(B)$.
-Dann liefert $\varphi$ einen lokalen Homomorphismus $\varphi_\mathfrak{p}:A_{\varphi^{-1}(\mathfrak{p})}\to B_\mathfrak{p}$.
-Das liefert für $V\subset_\text{o}\operatorname{Spec}(A)$ einen Ringhomomorphismus:
-\[f^\sharp(V):\mathcal{O}_A(V)\to\mathcal{O}_B(f^{-1}(V)) \]
-indem man eine Abbildung $s:V\to\coprod_{\mathfrak{q}\in V} A_\mathfrak{q}$ auf die folgende Abbildung $f^\sharp(s): f^{-1}(V)\to\coprod_{\mathfrak{p}\in f^{-1}(V)}B_\mathfrak{p}$ schickt:
-\[ \begin{tikzcd}[row sep=.1em]
-f^{-1}(V) \ar[r, "f"] & V\ar[r, "s"] & A_{\varphi^{-1}(\mathfrak{p})}\ar[r, "\varphi_\mathfrak{p}"] & B_\mathfrak{p}\\
-\mathfrak{p}\ar[r, mapsto] & f(\mathfrak{p}) = \varphi^{-1}(\mathfrak{p})\ar[r, mapsto] & s(\varphi^{-1}(\mathfrak{p}))\ar[r, mapsto] & \varphi_\mathfrak{p}(s(\varphi^{-1}(\mathfrak{p})))
-\end{tikzcd} \]
-Die $f^\sharp(V)$ gibt uns einen Garbenmorphismus $f^\sharp:\mathcal{O}_A\to f_\ast \mathcal{O}_B$.
-Die durch $f^\sharp$ induzierte Abbildungen auf den Halmen sind gerade die $\varphi_\mathfrak{p}$.
-Somit ist $(f,f^\sharp)$ ein Morphismus lokal geringten Räumen.
-\item Sei $(f,f^\sharp):(\operatorname{Spec}B,\mathcal{O}_B)\to(\operatorname{Spec}A,\mathcal{O}_A)$ ein Morphismus von lokal geringten Räu\-men. $f^\sharp$ induziert einen Ringhomomorphismus:
-\[\varphi: A=\Gamma(\operatorname{Spec}A,\mathcal{O}_A)\to\Gamma(\operatorname{Spec}B,\mathcal{O}_B) =B \]
-Sei $\mathfrak{p}\in\operatorname{Spec}(B)$. Dann haben wir induzierte lokale Homomorphismen mit kommutativem Diagramm:
-\[\begin{tikzcd}
-A_{f(\mathfrak{p})} = \mathcal{O}_{A,f(\mathfrak{p})}  \ar[r, "f^\sharp_\mathfrak{p}"] & \mathcal{O}_{B,\mathfrak{p}} = B_\mathfrak{p}\\
-A \ar[r, "\varphi"']\ar[u] & B\ar[u]
-\end{tikzcd} \]
-Da die $f^\sharp_\mathfrak{p}$ lokale Homomorphismen sind, folgt $f(\mathfrak{p})=\varphi^{-1}(\mathfrak{p})$. Somit ist $f^\sharp$ von dem Ringhomomorphismus $\varphi$ induziert.\qed
-\end{enumerate}
-
-\paragraph{Definition 2.7.}\label{2.7} \begin{enumerate}[(i)]
-\item Ein \textit{affines Schema}\index{Schema!affin} ist ein lokal geringter Raum $(X,\mathcal{O}_X)$, der als lokal geringter Raum isomorph zum Spektrum eines Rings ist.
-\item Ein \textit{Schema}\index{Schema} ist ein lokal geringter Raum $(X,\mathcal{O}_X)$ derart, dass jeder Punkt $P\in X$ eine offene Umgebung $U\subset X$ besitzt, so dass $(U,\mathcal{O}_X|_U)$ ein affines Schema ist.
-\end{enumerate}
-$\mathcal{O}_X$ heißt \textit{Strukturgarbe}\index{Strukturgarbe}. Ein \textit{Morphismus}\index{Morphismus!Schemata} von Schemata ist ein Morphismus von lokal geringten Räumen.
-
-\paragraph{Beispiel 2.8.}\label{2.8}\begin{enumerate}
-\item Sei $k$ ein Körper. $\operatorname{Spec}(k)$ ist ein affines Schema, dessen topologischer Raum aus einem Punkt besteht.
-\end{enumerate}
-
-\paragraph{Definition 2.9.}\label{2.9} Sei $K$ ein Körper.
-Eine \textit{diskrete Bewertung}\index{Bewertung!diskret} von $K$ ist eine Abbildung $v:K\to\mathbb{Z}\cup\{\infty\}$, so dass für alle $x,y\in K$ gilt:
-\begin{enumerate}[(i)]
-\item $v(xy)=v(x)+v(y)$
-\item $v(x+y)\geq \min\{v(x),v(y)\}$
-\item $v(x)=\infty\iff x=0$
-\end{enumerate}
-$R=\{x\in K\mid v(x)\geq 0\}$ definiert einen Teilring von $K$ und heißt \textit{diskreter Be\-wer\-tungs\-ring}\index{Bewertungsring!diskret} von $v$.
-$R$ ist ein lokaler Hauptidealring mit Maximalideal $\mathfrak{m}=\{x\in K\mid v(x) >0\}$.
-$R/\mathfrak{m}$ heißt \textit{Restklassenkörper}\index{Restklassenkörper} von $v$.
-Ein \textit{diskreter Bewertungsring}\index{Bewertungsring!diskret} $A$ ist ein nullteilerfreier Ring, der diskreter Bewertungsring für eine Bewertung seines Quotientenkörpers ist.
-
-\paragraph{Beispiel 2.8}\begin{enumerate}
-\item[2.] Sei $R$ ein diskreter Bewertungsring. Es ist $T=\operatorname{Spec}(R)$ ein affines Schema, bestehend aus zwei Punkten:
-\begin{itemize}
-\item Der Punkt $t_0=\mathfrak{m}\in\operatorname{Spec}(R)$ ist abgeschlossen, da $V(\mathfrak{m})=\{\mathfrak{m}\}$ und besitzt $R=R_{t_0}$ als lokalen Ring.
-\item Der Punkt $t_1=(0)\in\operatorname{Spec}(R)$ ist offen und dicht in $\operatorname{Spec}(R)$, da $V(0)=\operatorname{Spec}(R)$. $t_1$ besitzt $K=\operatorname{Quot}(R)=R_{t_1}$ als lokalen Ring.
-\end{itemize}
-\[\begin{tikzcd}[row sep=.1em]
-\operatorname{Spec}(K)\ar[r] & \operatorname{Spec}(R) &\ar[l] \operatorname{Spec}(R/\mathfrak{m})\\
-(0)\ar[r, mapsto] & t_1\quad t_0 & \ar[l, mapsto] (0)
-\end{tikzcd} \]
-\item[3.] Sei $k$ ein Körper.
-Die \textit{affine Gerade}\index{affine Gerade} $\mathbf{A}_k^1$ über $k$ ist $\operatorname{Spec}k[X]$.
-Sei $\xi$ das Nullideal in $\operatorname{Spec}k[X]$.
-Dann ist $\overline{\{\xi\}}=\mathbf{A}_k^1$.
-Ein solcher Punkt heißt \textit{generischer Punkt}\index{generischer Punkt}.
-Alle anderen Punkte sind abgeschlossen, da diese den maximalen Idealen in $k[X]$ entsprechen.
-Es besteht eine Bijektion zwischen den irreduziblen, nicht-konstanten, normierten Polynomen aus $k[X]$ und den abgeschlossenen Punkten von $\mathbf{A}_k^1$.
-
-Ist $k$ algebraisch abgeschlossen, so besteht eine Bijektion zwischen den Elementen aus $k$ und den abgeschlossenen Punkten von $\mathbf{A}_k^1$.
-\item[4.] Allgemeiner definieren wir den \textit{affinen $n$-dimensionalen Raum}\index{affiner Raum} über $k$ als: \[\mathbf{A}_k^n=\operatorname{Spec}k[X_1,\ldots,X_n]\]
-\item[5.] Sei $k$ algebraisch abgeschlossen.
-Dann entsprechen die abgeschlossenen Punkte von $\mathbf{A}_k^n$ nach dem hilbertschen Nullstellensatz bijektiv den $n$-Tupeln von Elementen aus $k$.
-Ferner gibt es einen generischen Punkt $\xi$, der dem Nullideal in $k[X_1,\ldots,X_n]$ entspricht, d.h.
-$\overline{\{\xi\}}=\mathbf{A}_k^n$.
-\end{enumerate}
-
-\paragraph{Definition 2.10.}\label{2.10} \textit{(Offene Unterschemata)}\index{Schema!Unterschema!offen} Sei $(X,\mathcal{O}_X)$ ein Schema und $U\subset X$ offen.
-Dann ist $(U,\mathcal{O}_X|_U)$ ein Schema.
-Diese Aussage ist nichttrivial und wir werden sie \hyperref[2.10-beweis]{später} zeigen.
-Die offene Menge $U$ besitzt die \textit{induzierte Unterschemastruktur}.
-
-\paragraph{Definition 2.11.}\label{2.11} \textit{(Verkleben von Schemata)}\index{Verklebung!Schema} Sei $\{X_i\}$ eine Familie von Schemata und $U_{ij}\subset X_i,\ i\neq j$ offene Teilmengen mit induzierter Struktur.
-Ferner haben wir für  $i\neq j$ Isomorphismen von Schemata:
-\[\varphi_{ij}: (U_{ij},\mathcal{O}_{X_i}|_{U_{ij}}) \stackrel{\sim}{\to} (U_{ji},\mathcal{O}_{X_j}|_{U_{ji}}) \]
-mit $\varphi_{ij}=\varphi_{ji}^{-1}$ und $\varphi_{ij}(U_{ij}\cap U_{ik})=U_{ji}\cap U_{jk}$ und $\varphi_{jk}=\varphi_{ik}\circ\varphi_{ij}$ auf $U_{ij}\cap U_{ik}$ für alle paar\-wei\-se verschiedene $i,j,k$.
-Wir erhalten ein Schema $X$ durch \textit{Verkleben} der $X_i$ längst $U_{ij}$ bzgl.
-$\varphi_{ij}$:
-\[X=\dot{\bigcup_i}\ X_i\Big/{\sim},\quad x_i\sim\varphi_{ij}(x_i)\text{ für alle }x_i\in U_{ij},\ i\neq j \]
-$X$ besitze die Quotiententopologie.
-Es existiert für jedes $j$ ein Morphismus $\psi_j:X_j\to X$ von Schemata, das ein Isomorphismus auf einem offenen Unterschema in $X$ induziert mit $X=\bigcup\psi_j(X_j)$ und $\psi_i(U_{ij})=\psi_i(X_i)\cap \psi_j(X_j)$ und $\psi_i=\psi_j\circ\varphi_{ij}$ auf $U_{ij}$ für alle $i\neq j$.
-Die Strukturgarbe auf $X$ ist folgendermaßen gegeben: Sei $V\subset_\text{o} X$.
-Setze:
-\[\mathcal{O}_X(V)=\Big\{ (s_i)\in\prod_i \mathcal{O}_{X_i}(\psi_i^{-1}(V))\Bigm| \varphi_{ij}(s_i|_{\psi_i^{-1}(V)\cap U_{ij}}) = s_j|_{\psi_j^{-1}(V)\cap U_{ji}} \Big\} \]
-Somit ist $(X,\mathcal{O}_X)$ ein lokal geringter Raum. Da alle $X_i$ Schemata sind, besitzt jeder Punkt von $X$ eine affine Umgebung. Also ist $X$ ein Schema.
-
-\paragraph{Beispiel 2.12.}\label{2.12} Sei $k$ ein Körper, $X_1=X_2=\mathbf{A}_k^1$ und $U_1=U_2=\mathbf{A}_k^1\setminus\{P\}$ mit einem abgeschlossenen Punkt $P$.
-Ist $\varphi:U_1\to U_2$ die Identität, so ist die Verklebung $X$ von $X_1$ und $X_2$ längst $\varphi$ die affine Gerade, wobei der Punkt $P$ verdoppelt wurde.
-$X$ ist selbst nicht mehr affin.
-
-\paragraph{Satz 2.13.}\label{2.13} Sei $A$ ein Ring und $(X,\mathcal{O}_X)$ ein Schema. Dann ist die Abbildung bijektiv:
-\[\alpha:\operatorname{Hom}_\mathbf{Sch}(X,\operatorname{Spec}A)\to \operatorname{Hom}_\textbf{Ringe}(A,\Gamma(X,\mathcal{O}_X)) \]
-wobei wir $(f:X\to\operatorname{Spec}A ,\ f^\sharp:\mathcal{O}_A\to f_\ast\mathcal{O}_X)$ durch das Nehmen der globalen Schnitte auf den Ringhomomorphismus $A=\Gamma(\operatorname{Spec}A,\mathcal{O}_A)\to\Gamma(X,\mathcal{O}_X)$ schicken.
-
-\paragraph{Beweis.} Sei $X=\bigcup_\nu X_\nu$ eine affine Überdeckung.
-Ein Morphismus $(f,f^\sharp)$ ist eindeutig durch seine Einschränkungen $(f_\nu,f_\nu^\sharp)$ auf $X_\nu$ bestimmt.
-Diese sind nach \hyperref[2.6]{Satz 2.6} wiederum eindeutig bestimmt durch $\alpha(f_\nu)$.
-Ferner kommutiert das folgende Diagramm:
-\[\begin{tikzcd}
-A\ar[d, "\alpha(f)"']\ar[rr, "\alpha(f_\nu)"] && \Gamma(X_\nu,\mathcal{O}_{X_\nu})\\
-\Gamma(X,\mathcal{O}_X)\ar[rru, "\operatorname{res}"']
-\end{tikzcd} \]
-weshalb $\alpha(f)$ schon alle $\alpha(f_\nu)$ eindeutig bestimmt.
-Somit ist $\alpha$ injektiv.
-Sei nun ein Ringhomomorphismus $h:A\to\Gamma(X,\mathcal{O}_X)$ gegeben und $h_\nu:A\to\Gamma(X,\mathcal{O}_X)\to\Gamma(X_\nu,\mathcal{O}_X)$.
-Nach \hyperref[2.6]{Satz 2.6 (iii)} existiert ein $f_\nu:X_\nu\to\operatorname{Spec}(A)$ mit $\alpha(f_\nu)=h_\nu$.
-Für alle $\nu,\mu$ ist das folgende Diagramm kommutativ:
-\[\begin{tikzcd}
-& \Gamma(X_\nu,\mathcal{O}_X)\ar[dr, "\operatorname{res}"] &\\
-A\ar[ur, "h_\nu"] \ar[dr, "h_\mu"'] & & \Gamma(X_\nu\cap X_\mu,\mathcal{O}_X)\\
-& \Gamma(X_\mu,\mathcal{O}_X)\ar[ur, "\operatorname{res}"'] &
-\end{tikzcd} \]
-Aus der Injektivität von $\alpha$ folgt $f_\nu=f_\mu$ auf $X_\nu\cap X_\mu$. Kleben wir die $f_\nu$ nun zusammen, so erhalten wir ein $f:X\to\operatorname{Spec}(A)$ mit $\alpha(f)=h$.\qed
-
-\paragraph{Korollar 2.14.}\label{2.14} Es gibt eine pfeilumkehrende Kategorienäquivalenz zwischen der Kategorie der affinen Schemata und der Kategorie der kommutativen Ringe mit $1$.
-
-\paragraph{Korollar 2.15.}\label{2.15} $\operatorname{Spec}(\mathbb{Z})$ ist Endobjekt in der Kategorie der Schemata, d.h.
-für jedes Schema $X$ existiert ein eindeutiger Morphismus $f:X\to\operatorname{Spec}(\mathbb{Z})$.
-
-\paragraph{Satz 2.16.}\label{2.16} Sei $A$ ein Ring, $X=\operatorname{Spec}(A)$ und $f\in A$. Dann gilt:
-\[(D(f),\mathcal{O}_X|_{D(f)})\cong \operatorname{Spec}(A_f) \]
-
-\paragraph{Beweis.} Die natürliche Abbildung $\varphi:A\to A_f$ induziert einen Homöomorphismus:
-\[\psi:\operatorname{Spec}(A_f) \to \{\mathfrak{p}\in\operatorname{Spec}(A)\mid f\not\in\mathfrak{p} \}=D(f),\ \mathfrak{p}\mapsto\varphi^{-1}(\mathfrak{p})\qedhere \]
-
-\paragraph{Beweis zu \hyperref[2.10]{Definition 2.10}.}\label{2.10-beweis} Sei $(X,\mathcal{O}_X)$ ein Schema und $U\subset X$ offen.
-Dann ist $(U,\mathcal{O}_X|_U)$ ein lokal geringter Raum.
-Sei $P\in U$.
-Wir zeigen, dass eine Umgebung $V\subset_\text{o} U$ von $P$ existiert, so dass $(V,\mathcal{O}_X|_V)$ affin ist.
-Da $(X,\mathcal{O}_X)$ ein Schema ist, existiert ein $V'\subset_\text{o}X,\ P\in V'$ mit $(V',\mathcal{O}_X|_{V'})$ affin.
-Sei also $V'=\operatorname{Spec}(A)$ für einen Ring $A$.
-Da die $D(f),\ f\in A$ eine Basis der Topologie auf $X$ bilden, existiert ein $f\in A$, so dass $P\in D(f)\subset V'\cap U$.
-Wegen \hyperref[2.16]{Satz 2.16} ist $(D(f),\mathcal{O}_X|_{D(f)})\cong\operatorname{Spec}(A_f)$ affin.\qed
-
-\paragraph{Satz 2.17.}\label{2.17} Sei $X$ ein Schema und $Z\subset X$ irreduzibel und abgeschlossen.
-Dann existiert genau ein Punkt $\xi\in Z$ derart, dass $\overline{\{\xi\}}=Z$.
-$\xi$ heißt \textit{generischer Punkt}\index{generischer Punkt} von $Z$.
-
-\paragraph{Beweis.}\begin{itemize}
-\item \textit{Existenz:} Sei $U\subset_\text{o}X$ affin mit $Z\cap U\neq\varnothing$. Da $Z\cap U$ abgeschlossen in $U$ ist, gibt es ein Radikalideal $\mathfrak{p}$ mit $Z\cap U=V(\mathfrak{p})$.
-
-Nun ist $Z\cap U$ irreduzibel, da für offene Mengen $V_1,V_2\subset_\text{o} Z\cap U$ stets $V_1,V_2\subset_\text{o}Z$ gilt und aus der Irreduzibilität von $Z$ stets $V_1\cap V_2\neq\varnothing$ folgt.
-
-Ferner ist $\mathfrak{p}$ ein Primideal, denn ist $\mathfrak{ab}\subset\mathfrak{p}$ für gewisse Ideale $\mathfrak{a}$ und $\mathfrak{b}$, so folgt $\operatorname{Rad}(\mathfrak{ab})\subset\operatorname{Rad}(\mathfrak{p})=\mathfrak{p}$.
-Dies ist äquivalent zu $V(\mathfrak{p})\subset V(\mathfrak{ab})=V(\mathfrak{a})\cup V(\mathfrak{b})$, daher:
-\[Z\cap U =V(\mathfrak{p})=(V(\mathfrak{p})\cap V(\mathfrak{a}))\cup (V(\mathfrak{p})\cap V(\mathfrak{b})) \]
-Da $Z\cap U$ irreduzibel ist, folgt o.B.d.A.
-$V(\mathfrak{p})=V(\mathfrak{p})\cap V(\mathfrak{a})\subset V(\mathfrak{a})$, d.h.
-$\mathfrak{a}\subset\operatorname{Rad}(\mathfrak{a})\subset\operatorname{Rad}(\mathfrak{p})=\mathfrak{p}$.
-
-$\mathfrak{p}$ ist nun generischer Punkt von $Z\cap U$, da:
-\[\overline{\{\mathfrak{p}\}}=\bigcap_{\mathfrak{p}\in V(\mathfrak{a})}V(\mathfrak{a})=\bigcap_{\mathfrak{a}\subset\mathfrak{p}} V(\mathfrak{a})\subset V(\mathfrak{p}) \]
-Die andere Inklusion folgt aus $\mathfrak{a}\subset\mathfrak{p}$.
-Da $\mathfrak{p}$ prim ist, folgt $\operatorname{Rad}(\mathfrak{a})\subset\mathfrak{p}$ und somit $V(\mathfrak{p})\subset V(\mathfrak{a})$.
-Somit ist $Z\cap U=\overline{\{\mathfrak{p}\}}$.
-
-Bezeichne mit $\overline{\{\mathfrak{p}\}}^X$ den topologischen Abschluss in $X$.
-Dann ist $Z\cap U\subset\overline{\{\mathfrak{p}\}}^X$.
-Da $Z$ irreduzibel ist, gilt $Z=\overline{Z\cap U}\subset\overline{\{\mathfrak{p}\}}^X$, also $Z=\overline{\{\mathfrak{p}\}}^X$.
-\item \textit{Eindeutigkeit:} Sei $\overline{\{\xi\}}^X=Z=\overline{\{\xi'\}}^X$.
-Sei $U\subset_\text{o}X$ eine affine Umgebung von $\xi\in U$.
-Dann ist $\xi'\in U$, da aus $\xi'\in Z\setminus U$ der Widerspruch $\xi\in Z=\overline{\{\xi'\}}^X\subset X\setminus U$ folgt.
-Wähle nun Radikalideale $\mathfrak{p}'$ und $\mathfrak{p}$ mit:
-\[V(\mathfrak{p}')= \overline{\{\xi'\}}^U=U\cap \overline{\{\xi'\}}^X=U\cap\overline{\{\xi\}}^X=\overline{\{\xi\}}^U=V(\mathfrak{p}) \]
-Es folgt $\xi'=\mathfrak{p}'=\operatorname{Rad}(\mathfrak{p}')=\operatorname{Rad}(\mathfrak{p})=\mathfrak{p}=\xi$.\qed
-\end{itemize}
-
-\paragraph{Satz 2.18.}\label{2.18} Sei $X$ ein Schema und $K$ ein Körper. Dann gibt es eine Bijektion:
-\[\operatorname{Hom}_\mathbf{Sch}(\operatorname{Spec}K,X)\stackrel{\sim}{\to} \{(x,i)\mid x\in X,\ i:\kappa(x)\hookrightarrow K\text{ Ringhomomorphismus} \} \]
-wobei $\kappa(x)=\mathcal{O}_{X,x}/\mathfrak{m}_x$ der Restklassenkörper von $\mathcal{O}_{X,x}$ bezeichnet. Die Elemente heißen \textit{$K$-wertige Punkte} von $X$.\index{Punkt!$K$-wertig}
-
-\paragraph{Beweis.} Sei ein Morphismus $f:\operatorname{Spec}(K)\to X$ gegeben.
-Setze $x=f((0))\in X$.
-$f^\sharp$ induziert einen lokalen Homomorphismus $\mathcal{O}_{X,x}\to\mathcal{O}_{K,(0)}=K$.
-$f^\sharp$ faktorisiert daher über $\mathcal{O}_{X,x}/\mathfrak{m}_x=\kappa(x)$ und induziert einen Homomorphismus $i:\kappa(x)\hookrightarrow K$.
-
-Sei nun umgekehrt ein $x\in X$ und $i:\kappa(x)\hookrightarrow K$ gegeben. $i$ definiert nach \hyperref[2.6]{Satz 2.6} ein Schemamorphismus:
-\[f:\operatorname{Spec}(K)\to\operatorname{Spec}\kappa(x)\to\operatorname{Spec}(\mathcal{O}_{X,x})\stackrel{\psi}{\to} X \]
-wobei $\psi$ die folgende kanonische Abbildung ist:
-
-Sei $U\subset_\text{o}X$ eine affine Umgebung von $x$ mit $U=\operatorname{Spec}(A)$.
-Nach \hyperref[2.3]{Satz 2.3 (i)} gilt $\mathcal{O}_{X,x}=\mathcal{O}_{U,x}=A_x$.
-Die kanonische Abbildung $A\to A_x$ induziert $\psi:\operatorname{Spec}(\mathcal{O}_{X,x})\to U\hookrightarrow X$.
-Diese Abbildung ist unabhängig von $U$:
-
-Sei $U'\subset_\text{o}X$ eine weitere affine Umgebung von $x$.
-Dann existiert eine affine Umgebung $U''\subset_\text{o}U\cap U'$ mit $x\in U''$, also können wir o.B.d.A.
-$\operatorname{Spec}(A)= U\subset U'=\operatorname{Spec}(A')$ annehmen.
-Es existiert ein kanonischer Homomorphismus $A'\to A$ derart, dass das folgende Diagramm kommutiert:
-\[\begin{tikzcd}
-A'\ar[rd]\ar[r] & A\ar[d]\\
-& A'_x= \mathcal{O}_{X,x}= A_x &
-\end{tikzcd} \]
-Somit kommutiert:
-\[\begin{tikzcd}
-X & U'\ar[l] & U\ar[l]\\
- && \operatorname{Spec}(\mathcal{O}_{X,x})\ar[lu]\ar[u]
-\end{tikzcd}\qedhere\]
-
-\paragraph{Satz 2.19.}\label{2.19} \begin{enumerate}[(i)]
-\item Sei $A$ ein Ring und $f\in A$. Dann gilt:
-\[D(f)=\varnothing\iff f\text{ nilpotent} \]
-\item Sei $\varphi:A\to B$ ein Ringhomomorphismus und $f: Y=\operatorname{Spec}(B)\to\operatorname{Spec}(A)=X$ der durch $\varphi$ induzierte Morphismus affiner Schemata. Dann gilt:
-\begin{enumerate}[(a)]
-\item $\varphi$ ist genau dann injektiv, wenn $f^\sharp:\mathcal{O}_X\to f_\ast\mathcal{O}_Y$ injektiv ist. In diesem Fall ist $f$ \textit{dominant}\index{dominant}, d.h. $f(Y)\subset X$ ist dicht.
-\item $\varphi$ ist genau dann surjektiv, wenn $f^\sharp$ surjektiv ist und $f$ ein Homöomorphismus auf eine abgeschlossene Teilmenge von $X$ ist.
-\end{enumerate}
-\end{enumerate}
-
-\paragraph{Beweis.} Übung.\qed
-
-% \paragraph{Beispiel.} Das Spektrum $T=\operatorname{Spec}( k[X,Y]/(X^2))$ entspricht der $Y$-Achse, also besitzt der Ring $\mathcal{O}_T(T)= k[X,Y]/(X^2)$ nilpotente Elemente.
-
-\paragraph{Definition 2.20.}\label{2.20} Ein Schema $(X,\mathcal{O}_X)$ heißt \textit{reduziert}\index{Schema!reduziert}, falls $\mathcal{O}_X(U)$ für alle $U\subset_\text{o}X$ reduziert sind, d.h.
-keine nilpotente Elemente besitzt.
-
-\paragraph{Regeln 2.21.}\label{2.21} Sei $(X,\mathcal{O}_X)$ ein Schema.
-\begin{enumerate}[(i)]
-\item $(X, \mathcal{O}_X)$ ist genau dann reduziert, wenn $\mathcal{O}_{X,P}$ für alle $P\in X$ keine nilpotente Elemente besitzt.
-\item Sei $\mathcal{O}_X^\text{red}$ die assoziierte Garbe zur folgenden Prägarbe:
-\[ U \mapsto \mathcal{O}_X(U)/\mathfrak{N}_U \]
-wobei $\mathfrak{N}_U$ das Nilradikal von $\mathcal{O}_X(U)$ bezeichnet.
-Dann ist $(X,\mathcal{O}_X^\text{red})$ ein Schema, das zu $X$ assoziierte \textit{reduzierte Schema}\index{Schema!reduziert} $X_\text{red}$.
-Es gibt einen Morphismus $f:X_\text{red}\to X$ mit dem Homöomorphismus $\operatorname{id}$ auf den unterliegenden topologischen Räumen und $f^\sharp:\mathcal{O}_X\to f_\ast\mathcal{O}_X^\text{red}$ gegeben durch:
-\[\mathcal{O}_X(U)\to\mathcal{O}_X^\text{red}(U),\ s\mapsto \Big(U\stackrel{s}{\to}\coprod_{P\in U}\mathcal{O}_{X,P} \to\coprod_{P\in U}\mathcal{O}_{X,P}^\text{red}\Big) \]
-\item Sei $f:X\to Y$ ein Morphismus von Schemata mit $X$ reduziert.
-Setze $g=f$ auf den unterliegenden topologischen Räumen.
-Da $X$ reduziert ist, faktorisiert $f^\sharp:\mathcal{O}_Y\to f_\ast\mathcal{O}_X$ über $\mathcal{O}_Y^\text{red}$.
-$f^\sharp$ induziert $g^\sharp:\mathcal{O}_Y^\text{red}\to g_\ast\mathcal{O}_X=f_\ast\mathcal{O}_X$.
-Es gibt also einen eindeutig bestimmten Morphismus $g:X\to Y_\text{red}$ mit kommutativem Diagramm:
-\[\begin{tikzcd}
-X\ar[rr, "f"]\ar[drr, "g"'] && Y\\
-&& Y_\text{red}\ar[u]
-\end{tikzcd} \]
-\end{enumerate}
-
-\paragraph{} Sei $S=\bigoplus_{d\geq 0} S_d$ ein graduierter Ring und $S_+=\bigoplus_{d>0}S_d$.
-Ein Primideal $\mathfrak{p}\subset S$ ist genau dann homogen, wenn aus $fg\in\mathfrak{p}$ für gewisse homogene Elemente $f,g\in S$ stets $f\in\mathfrak{p}$ oder $g\in\mathfrak{p}$ folgt.
-Für ein homogenes Ideal $\mathfrak{a}\subset S$ setze:
-\[\operatorname{Proj}(S)=\{\mathfrak{p}\subset S\text{ homogenes Primideal}\mid S_+\not\subset\mathfrak{p}\},\quad V(\mathfrak{a})=\{\mathfrak{p}\in\operatorname{Proj}(S)\mid\mathfrak{a}\subset\mathfrak{p}\} \]
-
-\paragraph{Lemma 2.22.}\label{2.22}\begin{enumerate}[(i)]
-\item Sind $\mathfrak{a},\mathfrak{b}\subset S$ homogene Ideale, so gilt $V(\mathfrak{ab})=V(\mathfrak{a})\cup V(\mathfrak{b})$.
-\item Ist $(\mathfrak{a}_i)_i$ eine Familie homogener Ideale in $S$, so folgt $V\big(\sum\mathfrak{a}_i\big)=\bigcap V(\mathfrak{a}_i)$.
-\end{enumerate}
-Damit wird auf $\operatorname{Proj}(S)$ eine Topologie definiert. Die abgeschlossenen Mengen sind genau die Mengen der Form $V(\mathfrak{a})$ für ein homogenes Ideal $\mathfrak{a}\subset S$.
-
-\paragraph{Beweis.} Wie in \hyperref[2.1]{Lemma 2.1} unter Beachtung, dass homogene Ideale von homogene Elemente erzeugt werden.\qed
-
-\paragraph{Definition.} Sei $T\subset S$ multiplikativ abgeschlossen, die aus homogenen Elementen besteht. Dann wird $T^{-1}S = \bigoplus_{i\geq 0} (T^{-1}S)_i$ zu einem graduierten Ring:
-\[(T^{-1}S)_i=\Big\{\frac{s}{t}\in T^{-1}S\Bigm| s\in S\text{ homogen},\ t\in T,\ \deg(s)-\deg(t)=i \Big\} \]
-Ist $\mathfrak{p}\subset S$ ein homogenes Primideal und $f\in S$ ein homogenes Element, so ist die \textit{homogene Lokalisierung}\index{homogene Lokalisierung} bzgl.
-$\mathfrak{p}$ bzw.
-$f$ definiert als:
-\[S_{(\mathfrak{p})} = (S_\mathfrak{p})_0,\quad S_{(f)} = (S_f)_0 \]
-
-\paragraph{Definition.} Wir definieren eine Ringgarbe $\mathcal{O}$ auf $\operatorname{Proj}(S)$ wie folgt: Sei $U\subset_\text{o}\operatorname{Proj}(S)$ und setze $\mathcal{O}(U)$ als die Menge aller Abbildungen $s:U\to\coprod_{\mathfrak{p}\in U}S_{(\mathfrak{p})}$, so dass:
-\begin{enumerate}[(i)]
-\item Für alle $\mathfrak{p}\in U$ gilt $s(\mathfrak{p})\in S_{(\mathfrak{p})}$.
-\item Für alle $\mathfrak{p}\in U$ existiert eine offene Umgebung $V$ von $\mathfrak{p}$ mit $V\subset U$ und homogene Elemente $a,f\in S$ mit $\deg(a)=\deg(f)$ derart, dass für alle $\mathfrak{q}\in V$ stets $f\not\in\mathfrak{q}$ und $s(\mathfrak{q})=\frac{a}{f}$ in $S_{(\mathfrak{q})}$ gilt.
-\end{enumerate}
-
-\paragraph{Satz 2.23.}\label{2.23} Sei $S$ ein graduierter Ring. Dann gilt:
-\begin{enumerate}[(i)]
-\item $\mathcal{O}_\mathfrak{p}\cong S_{(\mathfrak{p})}$ für alle $\mathfrak{p}\in\operatorname{Proj}(S)$.
-\item Für ein homogenes $f\in S_+$ setze $D_+(f)=\{\mathfrak{p}\in\operatorname{Proj}(S)\mid f\not\in\mathfrak{p}\}$. Dann ist $D_+(f)$ offen in $\operatorname{Proj}(S)$ und es gilt:
-\[\operatorname{Proj}(S) =\bigcup_{f\in S_+\text{ homogen}}D_+(f) \]
-Es gibt einen Isomorphismus lokal geringter Räume $(D_+(f),\mathcal{O}|_{D_+(f)})\cong\operatorname{Spec}S_{(f)}$.
-\item $(\operatorname{Proj}S,\mathcal{O})$ ist ein Schema.
-\end{enumerate}
-
-\paragraph{Beweis.}\begin{enumerate}[(i)]
-\item Die Abbildung $\mathcal{O}_\mathfrak{p}\to S_{(\mathfrak{p})},\ s_\mathfrak{p}\mapsto s(\mathfrak{p})$, wobei $s$ ein Repräsentant von $s_\mathfrak{p}$ ist, ist ein Isomorphismus.
-Beweis analog wie \hyperref[2.3]{Satz 2.3 (i)}.
-\item Da $D_+(f)=\operatorname{Proj}(S)\setminus V(f)$, ist $D_+(f)$ offen.
-Sei $\mathfrak{p}\in\operatorname{Proj}(S)$, d.h.
-$\mathfrak{p}\subset S$ ist ein homogenes Primideal mit $S_+\not\subset\mathfrak{p}$.
-Sei $f\in S_+\setminus\mathfrak{p}$.
-Dann ist $\mathfrak{p}\not\in V(f)$, also $\mathfrak{p}\in D_+(f)$.
-Daher ist $\operatorname{Proj}(S)=\bigcup D_+(f)$.
-
-Sei $f\in S_+$.
-Wir definieren ein Morphismus lokal geringter Räume $(\phi,\phi^\sharp): D_+(f)\to\operatorname{Spec}S_{(f)}$ wie folgt: Sei $S\rightarrow S_f$ der natürliche Homomorphismus.
-Sei $\mathfrak{a}\subset S$ ein homogenes Ideal und setze $\phi(\mathfrak{a})=\mathfrak{a}S_f\cap S_{(f)}$.
-Beachte, dass $S_{(f)}=(S_f)_0\subset S_f$ ein Teilring ist.
-Für $\mathfrak{p}\in D_+(f)$ ist $\phi(\mathfrak{p})\in\operatorname{Spec}S_{(f)}$, siehe \hyperref[2.16]{Satz 2.16}, und $\phi$ ist bijektiv.
-Sei $\mathfrak{a}\subset S$ ein homogenes Ideal.
-Dann ist $\mathfrak{p}\supset\mathfrak{a}$ genau dann, wenn $\phi(\mathfrak{p})\supset\phi(\mathfrak{a})$.
-Daher ist $\phi$ ein Homöomorphismus.
-
-$\phi^\sharp:\mathcal{O}_{\operatorname{Spec} S_{(f)}}\to\phi_\ast \big(\mathcal{O}_{\operatorname{Proj}(S)}|_{D_+(f)}\big)$ wird wie folgt definiert:
-\[\mathcal{O}_{\operatorname{Spec}S_{(f)}}(U)\to\mathcal{O}_{\operatorname{Proj}(S)}(\phi^{-1}(U)),\ s\mapsto \Big(\phi^{-1}(U)\stackrel{s\circ\phi}{\longrightarrow}\coprod (S_{(f)})_{\phi(\mathfrak{p})} \cong\coprod S_{(\mathfrak{p})}\Big) \]
-Dieses ist ein Isomorphismus.
-\item folgt aus (i) und (ii).\qed
-\end{enumerate}
-
-\paragraph{Beispiel 2.24.}\label{2.24} Sei $A$ ein Ring. Dann heißt
-\[\mathbf{P}_A^n = \operatorname{Proj} A[X_0,\ldots,X_n] \]
-der \textit{$n$-dimensionaler projektiver Raum}\index{projektiver Raum} über $A$.
-Ist speziell $A=k$ ein algebraisch abgeschlossener Körper, so ist $\mathbf{P}_k^n$ ein Schema.
-Dessen Teilraum aller abgeschlossenen Punkte ist homöomorph zur projektiven $n$-dimensionalen Varietät.
-
-\paragraph{Definition 2.25.} \label{2.25} Sei $S$ ein beliebiges Schema.
-Ein \textit{Schema über $S$}\index{Schema!über $S$} ist ein Schema $X$ zusammen mit einem Morphismus $X\to S$, der sogenannte \textit{Strukturmorphismus}\index{Strukturmorphismus}.
-Ein Morphismus zweier Schemata $X$ und $Y$ über $S$ ist ein Morphismus $f:X\to X'$, so dass das folgende Diagramm kommutiert:
-\[\begin{tikzcd}
-X\ar[rr, "f"]\ar[d] && X'\ar[lld]\\
-S
-\end{tikzcd} \]
-So ein Morphismus nennt man auch \textit{$S$-Morphismus}\index{Morphismus!$S$-Schemata}.
-Bezeichne die Kategorie aller Schemata über $S$ mit $S$-Morphismen mit $\mathbf{Sch}(S)$.
-Für einen Ring $A$ setzen wir auch $\mathbf{Sch}(A)=\mathbf{Sch}(\operatorname{Spec}A)$.
-
-\paragraph{Satz 2.26.}\label{2.26} Sei $k$ ein algebraisch abgeschlossener Körper. Dann gibt es einen natürlichen Funktor
-\[t:\mathbf{Var}(k)\to\mathbf{Sch}(k) \]
-der \textit{volltreu}\index{volltreuer Funktor} ist, d.h.
-für zwei Varietäten $V,W$ ist die durch $t$ auf den Morphismen induzierte Abbildung $\operatorname{Hom}_{\mathbf{Var}(k)}(V,W)\to \operatorname{Hom}_{\mathbf{Sch}(k)}(tV,tW)$ bijektiv.
-Für eine Varietät $V$ setze $\mathfrak{M}(V)$ als die Menge aller abgeschlossenen Punkte des Schemas $tV$ mit der Teilraumtopologie.
-Es gibt einen Homöomorphismus topologischer Räume $V\cong\mathfrak{M}(V)$.
-Die Garbe der regulären Funktionen ist via diesen Homöomorphismus isomorph zu $\mathcal{O}_{tV}|_{\mathfrak{M}(V)}$.
-
-\paragraph{Beweis.} Siehe z.B. Hartshorne Kapitel II, Proposition 2.6 oder Mumford, Theorem 2 auf Seite 168.
-
-\section{Erste Eigenschaften von Schemata}
-
-\paragraph{Definition 3.1.}\label{3.1} \begin{enumerate}[(i)]
-\item Ein Schema heißt \textit{zusammenhängend}\index{Schema!zusammenhängend}, falls es als topologischer Raum zu\-sam\-men\-hän\-gend ist.
-\item Ein Schema heißt \textit{irreduzibel}\index{Schema!irreduzibel}, falls es als topologischer Raum irreduzibel ist.
-\item Ein Schema heißt \textit{reduziert}\index{Schema!reduziert} falls für alle $U\subset_\text{o}X$ der Ring $\mathcal{O}_X(U)$ reduziert ist.
-\item Ein Schema heißt \textit{integer}\index{Schema!integer}, falls für alle $U\subset_\text{o}X$ der Ring $\mathcal{O}_X(U)$ nullteilerfrei ist.
-\end{enumerate}
-
-\paragraph{Beispiel 3.2.}\label{3.2} Sei $X=\operatorname{Spec}(A)$ ein affines Schema. Dann gilt:
-\begin{enumerate}[(i)]
-\item $X$ ist irreduzibel $\iff \mathfrak{N}(A)$ ist ein Primideal
-\item $X$ ist reduziert $\iff A$ ist reduziert $\iff\mathfrak{N}(A)=0$
-\item $X$ ist integer $\iff A$ ist nullteilerfrei
-\end{enumerate}
-
-\paragraph{Beweis.} (i) und (ii) sind klar.
-Ist $X$ integer, so ist $A=\mathcal{O}_X(X)$ nullteilerfrei.
-Sei nun umgekehrt $A$ nullteilerfrei, d.h.
-$\mathfrak{N}(A)=0$ ist ein Primideal.
-Nach (i) und (ii) ist $X$ irreduzibel und reduziert.
-Daher folgt die Aussage aus dem nächsten Satz.\qed
-
-\paragraph{Satz 3.3.}\label{3.3} Ein Schema $X$ ist genau dann integer, wenn $X$ reduziert und irreduzibel ist.
-
-\paragraph{Beweis.} Sei $X$ integer. Dann ist $X$ offensichtlich reduziert. Wäre $X$ nicht irreduzibel, so gäbe es $\varnothing\neq U_1,U_2\subset_\text{o}X$ mit $U_1\cap U_2=\varnothing$. Dann ist:
-\[\mathcal{O}_X(U_1\cup U_2)=\mathcal{O}_X(U_1)\times\mathcal{O}_X(U_2) \]
-Somit ist $\mathcal{O}_X(U_1\cup U_2)$ nicht nullteilerfrei. Sei nun umgekehrt $X$ irreduzibel und reduziert.
-
-Sei $V\subset_\text{o}X$ affin mit $V=\operatorname{Spec}(A)$. Sei $a,b\in A$ mit $ab=0$. Es folgt:
-\[\operatorname{Spec}(A)=V(ab)=V(a)\cup V(b) \]
-Da $V$ irreduzibel ist, folgt o.B.d.A.
-$\operatorname{Spec}(A)=V(a)$.
-Da $A=\mathcal{O}_X(V)$ reduziert ist, folgt $\operatorname{Rad}(a)=(0)$, also $a=0$.
-Daher ist $\mathcal{O}_X(V)$ für jedes affine $V\subset_\text{o}X$ nullteilerfrei.
-
-Sei nun $U\subset_\text{o}X$ beliebig und $f,g\in\mathcal{O}_X(U)$ mit $fg=0$.
-Für $V\subset_\text{o}U$ affin, folgt aus $f|_V\cdot g|_V=0$ o.B.d.A.
-$f|_V=0$.
-Nun ist $U$ der Abschluss von $V$ in $U$.
-Sei $x\in U$ und $U(x)\subset_\text{o}U$ eine affine Umgebung von $x$ mit $U(x)=\operatorname{Spec}(B)$.
-Sei $f(x)=\frac{a}{h}\in B_x$ für alle $x\in U(x)$.
-Es ist $U(x)\cap V\neq\varnothing$, da $U$ irreduzibel ist.
-Wähle ein $y\in U(x)\cap V$; es folgt $0=f(y)=\frac{a}{h}\in B_y$, also gibt es ein $k\in B\setminus y$ mit $ka=0$.
-Da $B$ nullteilerfrei ist, folgt $a=0$ und $f=0$ auf $U(x)$.
-Somit folgt $f=0$.\qed
-
-\paragraph{Definition.} \begin{enumerate}[(i)]
-\item Ein Schema $X$ heißt \textit{quasikompakt}\index{quasikompakt}, wenn sein unterliegender topologischer Raum quasikompakt ist.
-\item Ein topologischer Raum $X$ heißt \textit{noethersch}\index{noethersch}, wenn jede absteigende Kette von abgeschlossenen Teilmengen in $X$ stationär wird.
-\end{enumerate}
-
-\paragraph{Bemerkung.} Sei $X$ ein noetherscher Raum.
-Nach Zorns Lemma besitzt jede nichtleere Menge $\Sigma$ von abgeschlossenen Mengen in $X$ ein minimales Element, da jede Kette in $\Sigma$ ein minimales Element besitzt.
-
-\paragraph{Satz 3.4.}\label{3.4} \begin{enumerate}[(i)]
-\item Sei $X$ ein topologischer Raum. Dann ist $X$ genau dann noethersch, wenn alle offenen Teilmengen $U\subset X$ quasikompakt sind.
-\item Sei $X$ ein affines Schema. Dann ist $X$ quasikompakt, aber nicht notwendig noethersch.
-\item Sei $A$ ein noetherscher Ring. Dann ist der $\operatorname{Spec}(A)$ unterliegender Raum noethersch.
-\end{enumerate}
-
-\paragraph{Beweis.}\begin{enumerate}[(i)]
-\item Sei $U\subset_\text{o}X$ und $U=\bigcup_{i\in I}U_i$ eine offene Überdeckung. Für eine endliche Teilmenge $J\subset I$ setze $V_J=\bigcup_{i\in J}U_i$. Dann ist $V_J\subset X$ offen und es gilt:
-\[U=\bigcup_{J\subset I\text{ endlich}}V_J \]
-Wählt man aus $\Sigma=\{X\setminus V_J\mid J\subset I\text{ endlich}\}$ ein minimales Element $X\setminus V_{J'}$.
-Dann gilt $V_{J'}\supset V_J$ für alle $J$.
-Also ist $U=V_{J'}=\bigcup_{i\in J'}U_i$ eine endliche Teilüberdeckung.
-
-Sei umgekehrt $Y_1\supset Y_2\supset\ldots$ eine Kette abgeschlossener Mengen in $X$, so ist die Menge $U=\bigcup_{j\geq 1}X\setminus Y_j$ offen in $X$.
-Wir erhalten eine endliche Teilüberdeckung $U=\bigcup_{r\geq j\geq 1}X\setminus Y_j=X\setminus Y_r$, also folgt $Y_s=Y_r$ für alle $s\geq r$.
-\item Sei $X=\operatorname{Spec}(A)=\bigcup_{i\in I}U_i$ eine offene Überdeckung.
-Wir können o.B.d.A.
-$U_i=D(f_i)$ für gewisse $f_i\in A$ annehmen.
-Sei $\mathfrak{a}=(f_i\mid i\in I)\subset A$.
-Dann gilt:
-\[X=\bigcup_{i\in I}X\setminus V(f_i) = X\setminus\bigcap_{i\in I}V(f_i)=X\setminus V(\mathfrak{a}) \]
-Es folgt $V(\mathfrak{a})=\varnothing$, also $1\in\mathfrak{a}$.
-Somit gibt es endlich viele $g_j\in A$ und $i_j\in I$ mit $1=\sum_j g_jf_{i_j}$.
-Wir erhalten die endliche Teilüberdeckung $X=\bigcup_j D(f_{i_j})$.
-\item Sei $V(\mathfrak{a}_1)\supset V(\mathfrak{a}_2)\supset\ldots$ eine Kette abgeschlossener Mengen in $\operatorname{Spec}(A)$ mit Ra\-di\-kal\-idealen $\mathfrak{a}_i\subset A$.
-Sie wird stationär, da die Kette $\mathfrak{a}_1\subset \mathfrak{a}_2\subset\ldots$ stationär wird.\qed
-\end{enumerate}
-
-\paragraph{Definition 3.5.}\label{3.5} Sei $X$ ein Schema.
-\begin{enumerate}[(i)]
-\item $X$ heißt \textit{lokal noethersch}\index{noethersch!lokal}, falls $X$ von offenen, affinen Teilmengen $\operatorname{Spec}(A_i)$ mit noetherschen Ringen $A_i$ überdeckt werden kann.
-\item $X$ heißt \textit{noethersch}\index{noethersch}, falls $X$ lokal noethersch und quasikompakt ist.
-Dies ist äquivalent dazu, dass $X$ von endlich vielen offenen, affinen Teilmengen $\operatorname{Spec}(A_i)$ mit noetherschen Ringen $A_i$ überdeckt werden kann.
-\end{enumerate}
-
-\paragraph{Bemerkung.} Ist ein Schema $X$ noethersch, so ist nach \hyperref[3.4]{Satz 3.4 (iii)} der unterliegender Raum von $X$ noethersch. Die Umkehrung gilt im Allgemeinen nicht.
-
-\paragraph{Satz 3.6.}\label{3.6} Sei $X$ ein Schema. Dann ist $X$ genau dann lokal noethersch, wenn für alle offenen, affinen Teilmengen $U=\operatorname{Spec}(A)$ stets $A$ ein noetherscher Ring ist.
-
-\paragraph{Beweis.} Die Rückrichtung ist trivial.
-Sei also $X$ lokal noethersch und $U=\operatorname{Spec}(A)$ offen in $X$.
-Wir haben eine offene affine Überdeckung $X=\bigcup_i\operatorname{Spec}(B_i)$ mit noetherschen Ringen $B_i$.
-Da die offenen Mengen $D(f)$ eine offene Basis der Topologie bilden, haben wir eine Darstellung:
-\[U=\bigcup_{i,j} D(f_{ij}),\quad f_{ij}\in B_i,\ D(f_{ij})\subset U\cap \operatorname{Spec}(B_i) \]
-mit $D(f_{ij})=\operatorname{Spec}(B_i)_{f_{ij}}$.
-Da $B_i$ noethersch sind, sind die Lokalisierungen $(B_i)_{f_{ij}}$ ebenfalls noethersche Ringe.
-Da $U$ affin und somit quasikompakt ist, kann $U$ von endlich vielen Spektren noetherscher Ringe überdeckt werden.
-
-Sei $V=\operatorname{Spec}(B)$ offen in $U$ mit noetherschen Ring $B$.
-Sei $f\in A$ und betrachte $D(f)\subset V$.
-Die natürliche Inklusion $V\hookrightarrow U$ induziert einen Ringhomomorphismus $A\to B$.
-Sei $\overline{f}$ das Bild von $f$ in $B$.
-Es gilt:
-\[\operatorname{Spec}(A_f)=D(f)=D(\overline{f})=\operatorname{Spec}(B_{\overline{f}})\]
-Es folgt $A_f\cong B_{\overline{f}}$ und $A_f$ ist noethersch.
-Wir haben nun gezeigt, dass $U$ von endlich vielen offenen Mengen der Form $D(f)=\operatorname{Spec}(A_f)$ überdeckt werden kann mit noetherschen Ringen $A_f$.
-Somit folgt die Aussage aus dem nächsten Lemma.\qed
-
-\paragraph{Lemma 1.}\label{3.6-lemma1} Sei $A$ ein Ring und $f_1,\ldots,f_r\in A$ mit $1=(f_1,\ldots,f_r)$. Sind alle $A_{f_i}$ noethersch, so ist auch $A$ noethersch.
-
-\paragraph{Lemma 2.}\label{3.6-lemma2} Sei $A$ ein Ring und $f_1,\ldots,f_r\in A$ mit $1=(f_1,\ldots,f_r)$.
-Sei $\mathfrak{a}\subset A$ ein Ideal und $\varphi_i:A\to A_{f_i}$ die Lokalisierungsabbildung.
-Dann gilt:
-\[\mathfrak{a} = \bigcap_{i=1}^r \varphi_i^{-1}(\varphi_i(\mathfrak{a}) A_{f_i}) \]
-
-\paragraph{Beweis.} Für die nichttriviale Inklusion sei $b\in A$ mit $\varphi_i(b)\in \varphi(\mathfrak{a})A_{f_i}$ für alle $i$. Schreibe:
-\[\varphi_i(b)=\frac{a_i}{f_i^{n_i}}\in A_{f_i},\quad a_i\in\mathfrak{a},\ n_i>0 \]
-Sei o.B.d.A. $n=n_1=\ldots=n_r$. Somit gibt es für alle $i$ ein $m_i\geq 0$ mit:
-\[f_i^{m_i}(f_i^nb-a_i)=0 \]
-Sei o.B.d.A.
-$m=m_1=\ldots=m_r$.
-Es folgt $f_i^{m+n}b\in\mathfrak{a}$ für alle $i$.
-Aus $1=(f_1,\ldots,f_r)$ folgt $1=(f_1^N,\ldots,f_r^N)$ für alle $N\geq 0$, insbesondere für $N=m+n$.
-Sei also $1=\sum_{i=1}^rc_if_i^N$ für gewisse $c_i\in A$.
-Dann gilt:
-\[b=\sum_{i=1}^rc_if_i^Nb\in\mathfrak{a}\qedhere \]
-
-\paragraph{Beweis von \hyperref[3.6-lemma1]{Lemma 1}.} Sei $\mathfrak{a}_1\subset\mathfrak{a}_2\subset\ldots$ eine Kette von Idealen in $A$.
-Diese induziert für alle $i$ eine Kette von Idealen in $A_{f_i}$:
-\[\varphi_i(\mathfrak{a}_1)A_{f_i}\subset\varphi_i(\mathfrak{a}_2) A_{f_i}\subset\ldots \]
-Da $A_{f_i}$ noethersch ist, wird diese Kette stationär für alle $i$.
-Es existiert also ein $s$ mit $\varphi_i(\mathfrak{a}_s)A_{f_i}=\varphi(\mathfrak{a}_{s+1})A_{f_i}=\ldots$ für alle $i$.
-Mit \hyperref[3.6-lemma2]{Lemma 2} wird auch die ursprüngliche Kette von Idealen in $A$ stationär.\qed
-
-\paragraph{Definition.} Sei $f:X\to Y$ ein Morphismus von Schemata.
-\begin{enumerate}[(i)]
-\item $f$ heißt \textit{lokal von endlichem Typ}\index{von endlichem Typ!lokal}, falls $Y$ eine offene affine Überdeckung $\bigcup_i\operatorname{Spec}(B_i)$ besitzt, so dass $f^{-1}(\operatorname{Spec}B_i)$ für alle $i$ eine offene affine Überdeckung $\bigcup_j\operatorname{Spec}(A_{ij})$ besitzt, wobei alle $A_{ij}$ endlich erzeugte $B_i$-Algebren sind.
-\item $f$ heißt \textit{von endlichem Typ}\index{von endlichem Typ}, falls $Y$ eine offene affine Überdeckung $\bigcup_i\operatorname{Spec}(B_i)$ besitzt, so dass $f^{-1}(\operatorname{Spec}B_i)$ für alle $i$ eine endliche, offene affine Überdeckung $\bigcup_j\operatorname{Spec}(A_{ij})$ besitzt, wobei alle $A_{ij}$ endlich erzeugte $B_i$-Algebren sind.
-\item $f$ heißt \textit{endlich}\index{endlich}, falls eine offene affine Überdeckung $Y=\bigcup_i\operatorname{Spec}(B_i)$ existiert, so dass $f^{-1}(\operatorname{Spec}B_i)=\operatorname{Spec}(A_i)$ für alle $i$ ist, wobei $A_i$ eine $B_i$-Algebra ist, die als $B_i$-Modul endlich erzeugt ist.
-\end{enumerate}
-
-\paragraph{Bemerkung 3.7.}\label{3.7} Sei $f:X\to Y$ ein Morphismus von Schemata.
-\begin{enumerate}[(i)]
-\item $f$ ist genau dann lokal von endlichem Typ, wenn für alle offene, affine $V=\operatorname{Spec}(B)$ in $Y$ es eine offene affine Überdeckung $f^{-1}(V)=\bigcup_j\operatorname{Spec}(A_j)$ gibt, wobei $A_j$ endlich erzeugte $B$-Algebren sind.
-\item $f$ ist genau dann von endlichem Typ, wenn für alle offene, affine $V=\operatorname{Spec}(B)$ in $Y$ es eine endliche, offene affine Überdeckung $f^{-1}(V)=\bigcup_j\operatorname{Spec}(A_j)$ gibt, wobei $A_j$ endlich erzeugte $B$-Algebren sind.
-\item $f$ ist genau dann endlich, wenn für alle offene, affine $V=\operatorname{Spec}(B)$ in $Y$ die Menge $f^{-1}(V)=\operatorname{Spec}(A)$ affin ist, wobei $A$ ein endlich erzeugter $B$-Modul ist.
-\end{enumerate}
-
-\paragraph{Beweis.} Dies werden wir später zeigen.
-
-\paragraph{Beispiel 3.8.}\label{3.8} Sei $V$ eine Varietät über einem algebraisch abgeschlossenen Körper $k$.
-Dann ist das Schema $t(V)$ in \hyperref[2.26]{Satz 2.26} ein integres, noethersches Schema von endlichem Typ über $k$.
-
-\paragraph{Beispiel 3.9.}\label{3.9} Sei $P$ ein Punkt einer Varietät und $\mathcal{O}_P$ der zugehörige Halm.
-Dann ist $\operatorname{Spec}(\mathcal{O}_P)$ ein integres, noethersches Schema, aber nicht von endlichem Typ über $k$.
-
-\paragraph{Definition 3.10.}\label{3.10} Ein \textit{offenes Unterschema}\index{Schema!Unterschema!offen} eines Schemas $X$ ist ein Schema $U$, dessen unterliegender topologischer Raum eine offene Teilmenge von $X$ ist und dessen Strukturgarbe $\mathcal{O}_U$ isomorph zu $\mathcal{O}_X|_U$ ist.
-
-Ein Schemamorphismus $f:X\to Y$ heißt \textit{offene Immersion}\index{Immersion!offen}, falls $f$ ein Isomorphismus auf ein offenes Unterschema in $Y$ induziert.
-
-\paragraph{Bemerkung.} Jede offene Teilmenge eines Schemas $X$ trägt eine eindeutig bestimmte Struktur als offenes Unterschema, siehe auch \hyperref[2.10]{Definition 2.10}.
-
-\paragraph{Definition 3.11.}\label{3.11} Ein \textit{abgeschlossenes Unterschema}\index{Schema!Unterschema!abgeschlossen} eines Schemas $X$ ist ein Schema $Y$, zusammen mit einem Morphismus $(i,i^\sharp):Y\to X$, so dass:
-\begin{enumerate}[(i)]
-\item Der $Y$ unterliegender Raum ist eine abgeschlossene Teilmenge von $X$.
-\item $i:Y\hookrightarrow X$ ist die natürliche Inklusion.
-\item $i^\sharp:\mathcal{O}_X\to i_\ast\mathcal{O}_Y$ ist surjektiv.
-\end{enumerate}
-
-Ein Schemamorphismus $f:X\to Y$ heißt \textit{abgeschlossene Immersion}\index{Immersion!abgeschlossen}, falls $f$ ein Isomorphismus auf ein abgeschlossenes Unterschema in $Y$ induziert.
-
-\paragraph{Beispiel 3.12.}\label{3.12} Sei $A$ ein Ring, $\mathfrak{a}\subset A$ ein Ideal und $X=\operatorname{Spec}(A),\ Y=\operatorname{Spec}(A/\mathfrak{a})$.
-Der Ringhomomorphismus $\varphi:A\to A/\mathfrak{a}$ induziert $f:Y\to X,\ \mathfrak{p}\mapsto\varphi^{-1}(\mathfrak{p})$.
-Dies induziert ein Morphismus von Schemata.
-$f$ ist ein Homöomorphismus von $Y$ auf $V(\mathfrak{a})$ und die Abbildung $f^\sharp:\mathcal{O}_X\to f_\ast\mathcal{O}_Y$ induziert:
-\[f_\mathfrak{p}^\sharp:\mathcal{O}_{X, f(\mathfrak{p})}=A_{\varphi^{-1}(\mathfrak{p})}\to (A/\mathfrak{a})_\mathfrak{p}=\mathcal{O}_{Y,\mathfrak{p}}\]
-$f_\mathfrak{p}^\sharp$ ist surjektiv für alle $\mathfrak{p}\in Y$, also ist nach \hyperref[1.10]{1.10 (ii)} auch $f^\sharp$ surjektiv.
-Somit erhält man für jedes $\mathfrak{a}\subset A$ auf $V(\mathfrak{a})\subset X$ eine Struktur als abgeschlossenes Unterschema in $X$.
-
-\paragraph{Bemerkung.} Ist $Y\subset X=\operatorname{Spec}(A)$ eine abgeschlossene Teilmenge, so existieren auf $Y$ viele abgeschlossene Unterschemastrukturen.
-Wir werden später sehen, das sie genau den Idealen $\mathfrak{a}\subset A$ mit $Y=V(\mathfrak{a})$ entsprechen.
-
-\paragraph{Satz 3.14.}\label{3.14} Sei $X$ ein Schema und $Y$ eine abgeschlossene Teilmenge von $X$.
-Dann besitzt $Y$ eine eindeutig bestimmte induzierte Struktur als reduziertes, abgeschlossenes Unterschema.
-
-\iffalse
-\paragraph{Beweis.}
-\begin{enumerate}
-\item Fall: Sei zunächst $X=\operatorname{Spec}(A)$ affin. Betrachte das folgende Ideal in $A$:
-\[\mathfrak{a}=\bigcap_{\mathfrak{p}\in Y}\mathfrak{p} \]
-Dann ist $V(\mathfrak{a})=Y$ und $\mathfrak{a}$ ist maximal unter allen Idealen $\mathfrak{a}'$ mit $V(\mathfrak{a}')=Y$.
-Die Unterschemastruktur $Y=V(\mathfrak{a})=\operatorname{Spec}(A/\mathfrak{a})$ auf $Y$ ist reduziert, da $A/\mathfrak{a}$ reduziert ist, wegen $\mathfrak{a}=\operatorname{Rad}(\mathfrak{a})$.
-
-Für die Eindeutigkeit sei $Y'=V(\mathfrak{a}')=\operatorname{Spec}(A/ \mathfrak{a}')$ eine weitere abgeschlossene Unterschemastruktur.
-Dann ist $\mathfrak{a}'\subset\mathfrak{a}$.
-Die kanonische Abbildung $A/\mathfrak{a}'\to A/\mathfrak{a}$ induziert ein Morphismus $Y\to Y'$.
-Da $Y$ reduziert ist, gibt es einen eindeutigen Morphismus $Y\to Y'_\text{red}$, so dass folgendes Diagramm kommutiert:
-\[\xymatrix{
-Y\ar[rr]\ar@{-->}[rrd] && Y'\\
-&& Y'_\text{red}\ar[u]
-} \]
-\end{enumerate}
-\fi
-
-\paragraph{Lemma 3.15.}\label{3.15} Sei $X$ ein topologischer Raum und $X=\bigcup_i U_i$ eine offene Überdeckung.
-Ferner sei für jedes $i$ eine Garbe $F_i$ auf $U_i$ gegeben und für alle $i,j$ seien Isomorphismen gegeben:
-\[\varphi_{ij}:F_i|_{U_i\cap U_j}\stackrel{\sim}{\to} F_j|_{U_i\cap U_j} \]
-so dass $\varphi_{ii}=\operatorname{id}$ und $\varphi_{ik}=\varphi_{jk}\varphi_{ij}$ auf $U_i\cap U_j\cap U_k$ gilt.
-Dann existiert eine eindeutig bestimmte Garbe $F$ auf $X$ und Isomorphismen $\psi_i:F|_{U_i}\stackrel{\sim}{\to}F_i$ mit $\psi_j=\varphi_{ij}\psi_i$ auf $U_i\cap U_j$.
-Wir sagen auch, dass $F$ durch \textit{Verkleben} der $F_i$ längst $\varphi_{ij}$ entsteht.\index{Verklebung!Garbe}
-
-\paragraph{Beweis.} Folgt direkt aus der Definition einer Garbe.\qed
-
-\paragraph{Definition 3.16.}\label{3.16} \begin{enumerate}[(i)]
-\item Die \textit{Dimension} $\dim(X)$ eines Schemas $X$ ist die Dimension von $X$ als topologischer Raum, d.h:\index{Dimension}
-\begin{align*}
-\dim(X)=\sup\{n\mid&\text{ es gibt eine Kette }Z_0\subsetneq Z_1\subsetneq\ldots\subsetneq Z_n\\
-&\text{ von irreduziblen, abgeschlossenen Teilmengen in }X \} 
-\end{align*}
-\item Sei $Z$ eine irreduzible, abgeschlossene Teilmenge eines Schemas $X$. Dann ist die \textit{Kodimension}\index{Kodimension} $\operatorname{codim}(Z,X)$ von $Z$ in $X$ definiert als:
-\begin{align*}
-\operatorname{codim}(Z,X)=\sup\{n\mid & \text{ es gibt eine Kette }Z=Z_0\subsetneq Z_1\subsetneq\ldots\subsetneq Z_n\\
-&\text{ von irreduziblen, abgeschlossenen Teilmengen in }X\}
-\end{align*}
-Ist $Y$ eine abgeschlossene Teilmenge von $X$, so setzen wir:
-\[\operatorname{codim}(Y,X)=\inf \{\operatorname{codim}(Z,X)\mid Z\subset Y\text{ irreduzibel, abgeschlossen} \} \]
-\end{enumerate}
-
-\paragraph{Definition 3.17.}\label{3.17} Sei $S$ ein Schema und $X,Y$ $S$-Schemata.
-Das \textit{Faserprodukt}\index{Faserprodukt} $X\times_SY$ von $X$ und $Y$ über $S$ ist ein Schema, zusammen mit Projektionsmorphismen $p_1:X\times_SY\to X$ und $p_2:X\times_SY\to Y$ derart, dass:
-\begin{enumerate}[(i)]
-\item Das folgende Diagramm kommutiert:
-\[ \begin{tikzcd}
-X\times_S Y\ar[r, "p_2"]\ar[d, "p_1"'] & Y\ar[d]\\
-X\ar[r] & S
-\end{tikzcd}\]
-\item Ist $Z$ ein $S$-Schema und $f:Z\to X,\ g:Z\to Y$ Morphismen derart, dass das folgende Diagramm kommutiert:
-\[\begin{tikzcd}
-Z\ar[rrd, bend left, "g"]\ar[ddr, "f"', bend right]\ar[rd, "\theta", dashed] &&\\
-&X\times_S Y\ar[r, "p_2"']\ar[d, "p_1"] & Y\ar[d]\\
-&X\ar[r] & S
-\end{tikzcd} \]
-so existiert einen eindeutig bestimmten Morphismus $\theta:Z\to X\times_SY$ mit $f=p_1\theta$ und $g=p_2\theta$.
-\end{enumerate}
-Wird für Schemata $X$ und $Y$ kein Bezug zu einer Basis angegeben, so ist immer das Endobjekt $S=\operatorname{Spec}(\mathbb{Z})$ gemeint, d.h. $X\times Y=X\times_{\operatorname{Spec}(\mathbb{Z})}Y$.
-
-\paragraph{Theorem 3.18.}\label{3.18} Seien $X$ und $Y$ $S$-Schemata. Dann existiert das Faserprodukt $X\times_S Y$ und ist auf Isomorphie eindeutig.
-
-\paragraph{Lemma 3.19.}\label{3.19} \textit{(Verkleben von Morphismen, vgl.
-\hyperref[2.13]{Satz 2.13})}\index{Verklebung!Morphismus} Seien $X,Y$ Schemata und $\{U_i\}$ eine offene Überdeckung von $X$.
-Ferner seien $f_i:U_i\to Y$ Morphismen gegeben, wobei $U_i$ mit der offenen Unterschemastruktur versehen ist.
-Es gelte $f_i|_{U_i\cap U_j}=f_j|_{U_i\cap U_j}$ für alle $i,j$.
-Dann gibt es einen Morphismus $f:X\to Y$ mit $f|_{U_i}=f_i$ für alle $i$.
-
-\paragraph{Beweis von \hyperref[3.18]{Theorem 3.18}.} Die Eindeutigkeit ist klar.
-\begin{enumerate}
-\item Schritt: Seien $X=\operatorname{Spec}(A),\ Y=\operatorname{Spec}(B),\ S=\operatorname{Spec}(R)$ affin.
-Somit sind $A$ und $B$ $R$-Algebren.
-Wir zeigen $X\times_S Y=\operatorname{Spec}(A\otimes_R B)$.
-Die Projektionsabbildung $p_1:\operatorname{Spec}(A\otimes_RB)\to \operatorname{Spec}(A)$ ist durch die natürliche Abbildung $\tilde{p}_1:A\to A\otimes_RB$ gegeben, analog für $p_2$.
-Offensichtlich kommutiert:
-\[\begin{tikzcd}
- A\ar[r, "\tilde{p}_1"] & A\otimes_RB\\
- R\ar[r]\ar[u] & B\ar[u, "\tilde{p}_2"']
-\end{tikzcd} \]
-Sei also $Z$ ein $S$-Schema und Morphismen $f:Z\to X,\ g:Z\to Y$ gegeben, die über $S$ gleich sind.
-Diese entsprechen Ringhomomorphismen $\tilde{f}:A\to\Gamma(Z,\mathcal{O}_Z)$ und $\tilde{g}:B\to\Gamma(Z,\mathcal{O}_Z)$ nach \hyperref[2.13]{Satz 2.13}.
-Es kommutiert:
-\[\begin{tikzcd}
-&&\Gamma(Z,\mathcal{O}_Z)\\
-A\ar[r, "\tilde{p}_2"']\ar[rru, "\tilde{f}", bend left]& A\otimes_RB\ar[ru, "\tilde{\theta}"', dashed] &\\
-R\ar[r]\ar[u] & B\ar[u, "\tilde{p}_2"]\ar[ruu, "\tilde{g}"', bend right]
-\end{tikzcd} \]
-Wegen der Universaleigenschaft des Tensorprodukts gibt es genau einen Ringhomomorphismus $\tilde{\theta}:A\otimes_RB\to\Gamma(Z,\mathcal{O}_Z)$ mit $\tilde{f}\tilde{\theta}=\tilde{p}_1$ und $\tilde{g}\tilde{\theta}=\tilde{p}_2$.
-\hyperref[2.13]{Satz 2.13} liefert ein eindeutiges $\theta:Z\to\operatorname{Spec}(A\otimes_RB)$ mit $f=p_1\theta$ und $g=p_2\theta$.
-\item Schritt: Seien $X,Y$ beliebige $S$-Schemata und $U\subset_\text{o}X$.
-Wir nehmen an, dass das Faserprodukt $X\times_SY$ mit Projektionen $p_1,p_2$ existiert.
-Wir zeigen, dass für die offene Teilmenge $p_1^{-1}(U)\subset X\times_SY$ stets $p_1^{-1}(U)=U\times_SY$ gilt.
-
-Da $p_1^{-1}(U)\subset X\times_SY$, kommutiert das Diagramm:
-\[\begin{tikzcd}
-p_1^{-1}(U)\ar[r, "p_2"]\ar[d, "p_1"'] & Y\ar[d]\\
-U\ar[r] & S
-\end{tikzcd} \]
-Sei $Z$ ein $S$-Schema und Morphismen $f:Z\to U,\ g:Z\to Y$ gegeben, so dass $(Z\stackrel{f}{\to} U\stackrel{i}{\hookrightarrow} X\to S)=(Z\stackrel{g}{\to} Y\to S)$.
-Nach der Universaleigenschaft von $X\times_SY$ existiert ein eindeutiges $\theta:Z\to X\times_SY$ mit $if=p_1\theta$ und $g=p_2\theta$.
-Insbesondere gilt $\theta(Z)\subset p_1^{-1}(U)$, also $\theta:Z\to p_1^{-1}(U)$.
-Somit erfüllt $p_1^{-1}(U)$ die Universaleigenschaft von $U\times_SY$.
-\item Schritt: Seien $X,Y$ $S$-Schemata und $\{X_i\}$ eine offene Überdeckung von $X$.
-Wir nehmen an, dass alle Faserprodukte $X_i\times_SY$ mit Projektionen $p_{1i},p_{2i}$ existieren.
-Wir zeigen, dass in diesem Fall auch $X\times_SY$ existiert.
-
-Setze $X_{ij}=X_i\cap X_j$ und $U_{ij}=p_{1i}^{-1}(X_{ij})\subset X_i\times_SY$.
-Nach Schritt 2 folgt nun $U_{ij}=X_{ij}\times_SY$.
-Wegen Eindeutigkeit existieren nun Isomorphismen $\varphi_{ij}:U_{ij}\stackrel{\sim}{\to}U_{ji}$ mit $\varphi_{ij}=\varphi_{ji}^{-1}$, $\varphi_{ij}(U_{ij}\cap U_{ik})=U_{ji}\cap U_{jk}$ und $\varphi_{ik}=\varphi_{jk}\circ\varphi_{ij}$ auf $U_{ij}\cap U_{jk}$.
-Mithilfe \hyperref[2.11]{2.11} verkleben wir die $X_i\times_SY$ via $\varphi_{ij}$ und erhalten so ein Schema $Z$ mit Morphismen $\psi_j:X_j\times_SY\to Z$, die Isomorphismen auf einem offenen Unterschema induzieren.
-Seien $p_1,p_2$ die Morphismen, die durch Verkleben der $p_{1i}$ bzw.
-$p_{2i}$ entstehen, siehe \hyperref[3.19]{Lemma 3.19}.
-Wir zeigen nun, dass $Z$ gerade das Faserprodukt $X\times_SY$ mit Projektionsmorphismen $p_1,p_2$ ist.
-
-Es gilt $Z=\bigcup_j\psi_j(X_j\times_SY)$. Also folgt die Kommutativität des zweiten Diagramms aus dem ersten:
-\[\begin{tikzcd}
-X_j\times_S Y\ar[r, "p_{1j}"]\ar[d, "p_{2j}"'] & X_j\ar[d]\\
-Y\ar[r] & S
-\end{tikzcd}\qquad\qquad \begin{tikzcd}
-Z\ar[d, "p_2"']\ar[r, "p_1"] & X\ar[d]\\
-Y\ar[r] & S
-\end{tikzcd}\]
-Sei nun $Z'$ ein weiteres $S$-Schema und $f:Z'\to X,\ g:Z'\to Y$ gegeben, die über $S$ gleich sind.
-Setze $Z_i'=f^{-1}(X_i)$ für alle $i$.
-Zu jedem $i$ existiert genau ein Morphismus ${\theta}_i:Z_i'\to X_i\times_SY\hookrightarrow Z$ mit $f|_{Z_i'}=p_{1i}\circ{\theta}_i$ und $g|_{Z_i'}=p_{2i}\circ{\theta}_i$.
-Es kommutiert:
-\[\begin{tikzcd}
-X_i\times_SY\ar[r, hook]\ar[d, "p_{1i}"'] & Z\ar[d, "p_1"]\\
-X_i\ar[r, hook] & X
-\end{tikzcd} \]
-Es gilt $Z_i'\cap Z_j'=f^{-1}(X_i\cap X_j)=f^{-1}(X_{ij})$ und daher $f|_{Z'_i\cap Z'_j}=p_{1i}\circ\theta_i|_{Z'_i\cap Z'_j}=p_{1j}\circ\theta_j|_{Z_i'\cap Z_j'}$, entsprechend für $g$.
-Wegen Eindeutigkeit folgt $\theta_i|_{Z_i'\cap Z_j'}=\theta_j|_{Z'_i\cap Z'_j}$.
-Daher können wir die $\theta_i$ zu einem Morphismus $\theta:Z'\to Z$ verkleben mit $f=p_1\theta$ und $g=p_2\theta$.
-$\theta$ ist eindeutig, da $\theta|_{Z_i'}=\theta_i$ und alle $\theta_i$ eindeutig sind.
-\item Schritt: Seien $X,Y$ $S$-Schemata und $S$ affin. Wir zeigen, dass $X\times_SY$ existiert.
-
-Seien $X=\bigcup_iX_i$ und $Y=\bigcup_jY_j$ offene affine Überdeckungen.
-Nach Schritt 1 existieren $X_i\times_S Y_j$ für alle $i,j$.
-Nach Schritt 3 existieren $X\times_SY_j$ für alle $j$.
-Wegen Symmetrie, existiert somit $ X\times_SY$ nach Schritt 3.
-\item Schritt: Seien $X,Y$ $S$-Schemata mit $S$ beliebig. Wir zeigen, dass $X\times_SY$ existiert.
-
-Seien $q:X\to S$ und $r:Y\to S$ die Strukturmorphismen und $S=\bigcup_iS_i$ eine offene affine Überdeckung.
-Setze $X_i=q^{-1}(S_i)$ und $Y_i=r^{-1}(S_i)$.
-Nach Schritt 4 existieren $X_i\times_{S_i}Y_i$.
-Wir zeigen, dass $X_i\times_{S_i}Y_i$ die Universaleigenschaft von $X_i\times_SY$ erfüllt.
-
-Seien $f:Z\to X_i$ und $g:Z\to Y$ gegeben, die über $S$ gleich sind. Dann gilt $rg(Z)=qf(Z)\subset q(X_i)\subset S_i$, also $g(Z)\subset Y_i$. Wir erhalten kommutatives Diagramm:
-\[\begin{tikzcd}
-Z\ar[ddr, "f"', bend right]\ar[rrd, "g"', bend left]\ar[rrrd, "g", bend left]\ar[rd, "\theta"', dashed]\\
-& X_i\times_{S_i}Y_i\ar[r, "p_2"']\ar[d, "p_1"] & Y_i\ar[r, hook]\ar[d] & Y\ar[dd, "r"]\\
-& X_i\ar[d, hook]\ar[r] & S_i\ar[dr, hook] &\\
-& X\ar[rr]_q & & S
-\end{tikzcd} \]
-Es existiert genau ein $\theta:Z\to X_i\times_{S_i} Y_i$ mit $f=p_1\theta$ und $g=p_2\theta$. Somit existieren auch $X_i\times_SY$. Nach Schritt 3 existiert auch $X\times_SY$.\qed
-\end{enumerate}
-
-\paragraph{Lemma 3.20.}\label{3.20} Seien $X,Y$ $S$-Schemata mit Strukturmorphismen $\xi:X\to S,\ \eta:Y\to S$ und $U,V,W$ offen in $X,Y$ bzw.
-$S$, so dass $\xi(U)\subset W$ und $\eta(V)\subset W$.
-Ferner seien $p_1,p_2$ die Projektionsmorphismen von $X\times_SY$.
-Dann gibt es einen $S$-Schemaisomorphismus:
-\[p_1^{-1}(U)\cap p_2^{-1}(V)\cong U\times_WV=U\times_SV \]
-
-\paragraph{Beweis.} $E=p_1^{-1}(U)\cap p_2^{-1}(V)$ ist offenes Unterschema von $X\times_SY$.
-Die Projektionsmorphismen von $X\times_SY$ induzieren $p_1:E\to U,\ p_2:E\to V$.
-Sei $Z$ ein $W$-Schema und Morphismen $\varphi:Z\to U,\ \psi:Z\to V$ gegeben, die über $W$ gleich sind.
-Wir erhalten Morphismen $\varphi':Z\to U\hookrightarrow X,\ \psi':Z\to V\hookrightarrow Y$, die über $S$ gleich sind.
-Es existiert einen eindeutigen Morphismus $\theta:Z\to X\times_SY$ mit $\varphi'=p_1\theta$ und $\psi'=p_2\theta$.
-Nun gilt $\theta(Z)\subset E$, da $p_1\theta(Z)=\varphi'(Z)\subset U$ und $p_2\theta(Z)=\psi'(Z)\subset V$.
-Somit ist $\theta:Z\to E$ mit $\varphi=p_1\theta,\ \psi=p_2\theta$.
-Es folgt $E\cong U\times_WV$.
-Da $W$ beliebig war, folgt auch $E\cong U\times_SV$.\qed
-
-\paragraph{Definition.} Ein \textit{Unterschema}\index{Schema!Unterschema} $Y$ eines Schemas $X$ ist ein abgeschlossenes Unterschema eines offenen Unterschemas von $X$.
-Mit anderen Worten haben wir eine abgeschlossene Immersion $\phi$ und eine offene Immersion $\psi$:
-\[Y\stackrel{\phi}{\hookrightarrow} U\stackrel{\psi}{\hookrightarrow}X \]
-Ein Morphismus $i:Y\to X$ heißt \textit{Immersion}\index{Immersion}, wenn $i$ einen Isomorphismus von $Y$ auf ein Unterschema in $X$ induziert.
-
-\paragraph{Satz 3.21.}\label{3.21} Seien $X,Y,Z$ $S$-Schemata und $W$ ein $Z$-Schema. Dann gilt:
-\begin{enumerate}[(i)]
-\item $X\times_SS\cong X$
-\item $X\times_SY\cong Y\times_SX$
-\item $(X\times_SY)\times_SZ\cong X\times_S(Y\times_SZ)$
-\item $(X\times_SZ)\times_ZW\cong X\times_SW$
-\item $(X\times_SY)\times_SZ\cong (X\times_SZ)\times_Z(Y\times_SZ)$
-\item Ist $\sigma:S\to T$ eine Immersion, so ist $X\times_SY\cong X\times_TY$.
-\end{enumerate}
-
-\paragraph{Beweis.} (i) bis (v) folgen direkt aus der Universaleigenschaft des Faserprodukts.
-Für (vi) seien $\varphi:Z\to X,\ \psi:Z\to Y$ Morphismen über $T$ und $\xi:X\to S,\ \eta:Y\to S$ die Strukturmorphismen.
-Wir erhalten folgendes kommutative Diagramm:
-\[ \begin{tikzcd}
-Z\ar[rrd, "\varphi", bend left]\ar[rd, "\theta"', dashed]\ar[ddr, "\psi"', bend right]\\
-& X\ar[r]\ar[d]\times_SY & X\ar[d, "\xi"']\ar[ddr, bend left] &\\
-& Y\ar[r, "\eta"]\ar[rrd, bend right] & S\ar[dr, "\sigma"]&\\
-&&&T
-\end{tikzcd}\]
-Wegen der Injektivität von $\sigma$, folgt aus $\sigma\xi\varphi=\sigma\eta\psi$ stets $\xi\varphi=\eta\psi$.
-Es existiert ein eindeutiges, kompatibles $\theta:Z\to X\times_SY$.
-Somit erfüllt $X\times_SY$ die Universaleigenschaft von $X\times_TY$.\qed
-
-\paragraph{Definition 3.22.}\label{3.22} Seien $f:X_1\to X_2$ und $g:Y_1\to Y_2$ $S$-Morphismen mit kommutativen Diagramm:
-\[\begin{tikzcd}
-&X_1 \ar[r, "f"]& X_2\\
-X_1\times_SY_1\ar[dr, "p_2"']\ar[ur, "p_1"]\ar[rrr, "f\times_Sg", dashed] &&& X_2\times_SY_2\ar[lu, "q_1"']\ar[ld, "q_2"]\\
-& Y_1\ar[r, "g"']&Y_2
-\end{tikzcd} \]
-Es existiert genau ein Morphismus $f\times_Sg: X_1\times_SY_1\to X_2\times_SY_2$ mit $fp_1=q_1(f\times_Sg)$ und $gp_2=q_2(f\times_Sg)$.
-
-\paragraph{Definition 3.23.}\label{3.23} Sei $f:X\to S$ ein Morphismus von Schemata. Für ein Morphismus $g:T\to S$ erhält man ein $T$-Schema $X_T=X\times_ST$ mit:
-\[\begin{tikzcd}
-X_T\ar[r]\ar[d] & T\ar[d, "g"]\\
-X\ar[r, "f"'] & S
-\end{tikzcd} \]
-Man sagt, dass $X_T$ durch \textit{Basiswechsel} von $X$ über $S$ durch $T$ erhalten bleibt.\index{Basiswechsel}
-
-\paragraph{Lemma 3.24.}\label{3.24} Sei $U\subset S$ ein offenes Unterschema. Dann gilt für den Basiswechsel $X_U$ von $f:X\to S$:
-\[X_U=X\times_SU\cong f^{-1}(U) \]
-
-\paragraph{Beweis.} Nach \hyperref[3.21]{Satz 3.21 (i)} ist die Projektion $p_1:X\times_SS\stackrel{\sim}{\to} X$ ein Isomorphismus.
-Sei in \hyperref[3.20]{Lemma 3.20} $Y=S$ und setze $f=p_2:X\to S$.
-Wir erhalten: \[f^{-1}(U)=p_1^{-1}(U)\cap p_2^{-1}(U)\cong X\times_SU\qedhere\]
-
-\paragraph{Definition.} Sei $\mathcal{P}$ eine Eigenschaft eines Morphismus' $f:X\to S$.
-Man sagt, dass $\mathcal{P}$ bei Basiswechsel \textit{erhalten} bleibt, wenn $f_T:X_T\to T$ die Eigenschaft $\mathcal{P}$ besitzt für alle Morphismen $g:T\to S$.
-
-\paragraph{Bemerkung.} Hat $f:X \to S$ eine Eigenschaft $\mathcal{P}$, die stabil unter Basiswechsel ist, so hat insbesondere $f|_{f^{-1}(U)}:f^{-1}(U)\to U$ für alle offenen $U\subset S$ die Eigenschaft $\mathcal{P}$.
-
-\paragraph{Satz 3.25.}\label{3.25}\begin{enumerate}[(i)]
-\item Die Eigenschaft (abgeschlossene, offene) Immersion zu sein bleibt bei Basiswechsel erhalten.
-\item Die Eigenschaft von endlichem Typ zu sein bleibt stabil unter Basiswechsel.
-\end{enumerate}
-
-\paragraph{Beweis.} Wird ausgelassen.
-
-\paragraph{Bemerkung.} Irreduzibilität, Reduziertheit und Integrität bleiben unter Basiswechsel nicht notwendig erhalten.
-
-\paragraph{Lemma 3.26.}\label{3.26} Sei $\psi:A\to B$ ein Ringhomomorphismus, so dass $\alpha(\psi):\operatorname{Spec}(B)\to\operatorname{Spec}(A)$ eine abgeschlossene Immersion ist.
-Sei ferner $C$ ein beliebiger Ring.
-Dann ist $\alpha(\psi\otimes\operatorname{id}_C):\operatorname{Spec}(B\otimes_AC)\to\operatorname{Spec}(A\otimes_AC)=\operatorname{Spec}(C)$ eine abgeschlossene Immersion, wobei $\alpha$ die Abbildung aus \hyperref[2.13]{Satz 2.13} bezeichnet.
-
-\paragraph{Beweis.} Da $\tilde{\psi}=\alpha(\psi)$ eine abgeschlossene Immersion ist, ist $\tilde{\psi}^\sharp:\mathcal{O}_A\to\tilde{\psi}_\ast\mathcal{O}_B$ surjektiv.
-$\tilde{\psi}$ ist Homöomorphismus auf eine abgeschlossene Teilmenge, daher ist $(\tilde{\psi}_\ast\mathcal{O}_B)_{\tilde{\psi}(\mathfrak{p})}=\mathcal{O}_{B,\mathfrak{p}}$ für alle $\mathfrak{p}\in\operatorname{Spec}(B)$.
-Nach \hyperref[3.12]{Beispiel 3.12} ist die auf den Halmen induzierte Abbildung $\tilde{\psi}^\sharp_\mathfrak{p}$ surjektiv:
-\[\tilde{\psi}^\sharp_\mathfrak{p} : A_{\psi^{-1}(\mathfrak{p})}=\mathcal{O}_{A,\tilde{\psi}(\mathfrak{p})}\to (\tilde{\psi}_\ast\mathcal{O}_B)_{\tilde{\psi}(\mathfrak{p})}=B_\mathfrak{p} \]
-Betrachte nun die exakte $A$-Modulsequenz $A\stackrel{\psi}{\to}B\to B/\psi(A)\to 0$.
-Diese induziert unter $-\otimes_AA_\mathfrak{q}$ für alle $\mathfrak{q}\in\operatorname{Spec}(A)$ die exakte Folge:
-\[A_\mathfrak{q}\to B\otimes_A A_\mathfrak{q}\to B/\psi(A)\otimes_A A_\mathfrak{q}\to 0 \]
-Sei das Bild von $\tilde{\psi}$ von der Form $V(\mathfrak{a})\cong\operatorname{Spec}(B)$ für ein Ideal $\mathfrak{a}\subset A$. Für ein Primideal $\mathfrak{q}\in\operatorname{Spec}(A)$ gilt:
-\begin{align*}
-B\otimes_A A_\mathfrak{q}\neq 0 &\iff \mathfrak{q}\in\operatorname{Supp}(B)\text{, wobei $B$ als $A$-Modul aufgefasst wird}\\
-&\iff \mathfrak{q}\supset\mathfrak{a}\\
-&\iff \mathfrak{q}\in\operatorname{im}(\tilde{\psi})\\
-&\iff\mathfrak{q}=\psi^{-1}(\mathfrak{p})\text{ für ein }\mathfrak{p}\in\operatorname{Spec}(B)\\
-&\implies B/\psi(A)\otimes_AA_\mathfrak{q}=0
-\end{align*}
-Somit ist $A_\mathfrak{q}\to B\otimes A_\mathfrak{q}$ surjektiv für alle $\mathfrak{q}\in\operatorname{Spec}(A)$.
-Es folgt die Surjektivität von $\psi$.
-Wir erhalten unter $-\otimes_AC$ die surjektive Abbildung $C\to B\otimes_AC$.
-Nach \hyperref[3.12]{Beispiel 3.12} folgt die Behauptung.\qed
-
-\paragraph{Lemma 3.27.}\label{3.27} Sei $f:X\to Y$ ein Morphismus und $Y=\bigcup_\lambda Y_\lambda$ eine offene Überdeckung.
-Dann ist $f$ genau dann eine (abgeschlossene, offene) Immersion, wenn $f_\lambda=f|_{X_\lambda}:X_\lambda \to Y_\lambda$ mit $X_\lambda=f^{-1}(Y_\lambda)$ die Eigenschaft hat.
-
-\paragraph{Beweis.} Kein Beweis.
-
-\paragraph{Bemerkung 3.28.}\label{3.28} Sind $Y_1,Y_2$ Unterschemata von $X$, so ist $Y_1\times_XY_2\to X$ ein Unterschema von $X$, d.h. eine Immersion.
-
-\paragraph{Definition 3.29.}\label{3.29} Sei $f:X\to Y$ ein $S$-Morphismus von $S$-Schemata.
-Der Morphismus $\Gamma_f=(\operatorname{id}_X,f)_S:X\to X\times_SY$ heißt \textit{$S$-Graph}\index{Graph} von $f$ und $\Gamma_f(X)$ heißt \textit{Graph} von $f$.
-Ist $f=\operatorname{id}_X$ die Identität, so heißt $\Delta_{X/S}=\Gamma_{\operatorname{id}_X}$ der \textit{Diagonalmorphismus}\index{Diagonalmorphismus}.
-
-\paragraph{Satz 3.30.}\label{3.30} Sei $f:X\to Y$ ein $S$-Morphismus von $S$-Schemata. \begin{enumerate}[(i)]
-\item Sind $Y$ und $S$ affine Schemata, so ist $\Gamma_f$ eine abgeschlossene Immersion. Insbesondere ist für $X,S$ affin $\Delta_{X/S}$ eine abgeschlossene Immersion.
-\item $\Gamma_f$ ist allgemein eine Immersion.
-\end{enumerate}
-
-\paragraph{Beweis.} \begin{enumerate}[(i)]
-\item Sei $X=\bigcup_\alpha X_\alpha$ eine offene affine Überdeckung und $X\times_SY$ das Faserprodukt.
-Nach \hyperref[3.24]{Lemma 3.24} ist $p_1^{-1}(X_\alpha)\cong X_\alpha\times_SY$ affin.
-Da $p_1\circ\Gamma_f=\operatorname{id}_X$, gilt $\Gamma_f^{-1}(p_1^{-1}(X_\alpha))=X_\alpha$.
-Da $\Gamma_f: \bigcup_\alpha\Gamma_f^{-1}(X_\alpha\times_SY) \to \bigcup_\alpha(X_\alpha\times_SY)=X\times_SY$, können wir nach \hyperref[3.27]{Lemma 3.27} o.B.d.A.
-annehmen, dass $X$ affin ist.
-
-Sei also $S=\operatorname{Spec}(R),\ X=\operatorname{Spec}(B)$ und $Y=\operatorname{Spec}(A)$.
-Sei $f$ induziert von $\varphi:A\to B$ und $\Gamma_f$ induziert von $\psi:B\otimes_RA\to A,\ b\otimes a\mapsto b\varphi(a)$.
-$\psi$ ist offensichtlich surjektiv, daher ist $\Gamma_f$ nach \hyperref[2.19]{Satz 2.19 (ii)} eine abgeschlossene Immersion.
-\item Sei zunächst $S$ affin und $Y=\bigcup_\alpha Y_\alpha$ eine offene affine Überdeckung.
-Setze $X_\alpha=f^{-1}(Y_\alpha)$, $\Gamma_{f\alpha}=\Gamma_f|_{X_\alpha}$ und $f_\alpha=f|_{X_\alpha}:X_\alpha\to Y_\alpha$.
-Es ist $\Gamma_f(X_\alpha)\subset X_\alpha\times_SY_\alpha$ nach (i) ein abgeschlossenes Unterschema und $X_\alpha\times_SY_\alpha\subset X\times_SY_\alpha$ nach \hyperref[3.25]{Satz 3.25 (i)} ein offenes Unterschema ist.
-Nach (i) ist $\Gamma_{f\alpha}:X_\alpha\to X_\alpha\times_SY_\alpha$ eine Immersion.
-Nach \hyperref[3.27]{Lemma 3.27} ist auch $\Gamma_f$ eine Immersion.
-
-Sei nun $S$ beliebig und $S=\bigcup_\lambda S_\lambda$ eine offene affine Überdeckung.
-Nach \hyperref[3.21]{Satz 3.21 (v)} ist $(X\times_SY)\times_SS_\lambda=X_\lambda\times_{S_\lambda}Y_\lambda$, wobei $X_\lambda=X\times_SS_\lambda$ und $Y_\lambda=Y\times_SS_\lambda$.
-Ferner ist $\Gamma_{f_\lambda}$ der Graph von $f_\lambda:X_\lambda\to Y_\lambda$ eine Immersion.
-Nach \hyperref[3.27]{Lemma 3.27} ist auch $\Gamma_f$ eine Immersion.\qed
-\end{enumerate}
-
-\paragraph{Definition 3.31.}\label{3.31} Sei $f:X\to Y$ ein Morphismus von Schemata und $y\in Y$ ein Punkt mit Restklassenkörper $\kappa(y)=\mathcal{O}_{Y,y}/\mathfrak{m}_{Y,y}$.
-Sei weiter $i:\operatorname{Spec}\kappa(y)\to Y$ der natürliche Morphismus gegeben durch $(y,\operatorname{id}_{\kappa(y)})$ (siehe \hyperref[2.18]{Satz 2.18}).
-Dann heißt das Faserprodukt $X_y=X\times_Y\operatorname{Spec}\kappa(y)$ die \textit{Faser}\index{Faser} von $f$ über dem Punkt $y$.
-\[\begin{tikzcd}
-X_y\ar[r, "p_2"]\ar[d, "p_1"'] & \operatorname{Spec}\kappa(y)\ar[d, "i"]\\
-X\ar[r, "f"'] & Y
-\end{tikzcd}\]
-
-\paragraph{Satz.} $X_y$ ist ein $\kappa(y)$-Schema mit $f^{-1}(y)$ als unterliegender topologischer Raum.
-
-\paragraph{Beweis.} Es ist $fp_1(X_y)=ip_2(X_y)=y$, also folgt $p_1(X_y)\subset f^{-1}(y)$ und somit $X_y=p_1^{-1}(f^{-1}(y))$.
-Sei $V\subset_\text{o}X$ affin mit $V\cap f^{-1}(y)\neq\varnothing$.
-Dann gilt:
-\[p_1^{-1}(V)=V_y\cong X\times_Y\operatorname{Spec}\kappa(y)\times_XV=X_y\times_XV \]
-Wir können somit o.B.d.A.
-$X$ und $Y$ als affin annehmen, sei also $X=\operatorname{Spec}(B),\ Y=\operatorname{Spec}(A),\ y=\mathfrak{p}$.
-Sei $\varphi:A\to B$ assoziiert zu $f$.
-Setze $\varphi':A_\mathfrak{p}\to B'=B\otimes_AA_\mathfrak{p}=\varphi(A\setminus\mathfrak{p})^{-1}B$.
-Es gibt eine Folge von Homöomorphismen:
-\begin{align*}
-f^{-1}(y) &= \{\mathfrak{q}\in\operatorname{Spec}(B)\mid\varphi^{-1}(\mathfrak{q})=\mathfrak{p} \}\\
-&\cong \{\mathfrak{q}\in\operatorname{Spec}(B')\mid\varphi'^{-1}(\mathfrak{q})=\mathfrak{p}A_\mathfrak{p} \}\\
-&\cong \{ \mathfrak{q}\in\operatorname{Spec}(B')\mid\mathfrak{q}\supset (\mathfrak{p}A_\mathfrak{p})B' \} \\
-&=\operatorname{Spec}(B'/(\mathfrak{p}A_\mathfrak{p})B')\\
-&=\operatorname{Spec}(B'\otimes_{A_\mathfrak{p}}A_\mathfrak{p}/\mathfrak{p}A_\mathfrak{p})\\
-&=\operatorname{Spec}((B\otimes_AA_\mathfrak{p})\otimes_{A_\mathfrak{p}}A_\mathfrak{p}/\mathfrak{p}A_\mathfrak{p}) =\operatorname{Spec}(B\otimes_A\kappa(\mathfrak{p}))\qedhere
-\end{align*}
-
-\paragraph{Definition 3.32.}\label{3.32} Sei $f:X\to Y$ ein Morphismus von Schemata, so dass $f(X)\subset Y$ dicht liegt, und $Y$ irreduzibel mit generischer Punkt $\xi$ mit $\xi\in f(X)$.
-Dann heißt $f^{-1}(\xi)$ \textit{generische Faser}\index{generische Faser} von $f$.
-
-\paragraph{Beispiel 3.33.}\label{3.33} Sei $k$ ein algebraisch abgeschlossener Körper und:
-\[X=\operatorname{Spec}k[X,Y,t]/(tY-X^2),\quad Y=\operatorname{Spec}k[T] \]
-Sei $f:X\to Y$ gegeben durch die kanonische Abbildung $\varphi:k[t]\to k[X,Y,t]/(tY-X^2)$.
-$X$ und $Y$ sind integre Schemata von endlichem Typ über $k$.
-Ferner ist $f$ surjektiv, da $\varphi$ injektiv ist (vgl.
-\hyperref[2.19]{Satz 2.19 (ii)}).
-Abgeschlossene Punkte von $Y$ entsprechen $k$.
-Sei $a\in Y$ ein abgeschlossener Punkt und betrachte die Faser: \[X_a=X\times_Y\operatorname{Spec}\kappa(a)=\operatorname{Spec}k[X,Y]/(aY-X^2)\]
-Im Fall $a\neq 0$ ist $X_a$ eine ebene Kurve in $\mathbf{A}_k^2$, die irreduzibel und reduziert ist. Für $a=0$ ist $X_0=\operatorname{Spec}k[X,Y]/(X^2)$ die $Y$-Achse. Diese ist nicht reduziert.
-
-\section{Separierte \& eigentliche Morphismen}
-
-\paragraph{Definition 4.1.}\label{4.1} Sei $f:X\to Y$ ein Morphismus von Schemata.
-$f$ heißt \textit{separiert}, falls $\Delta_{X/Y}:X\to X\times_YX$ eine abgeschlossene Immersion ist.
-In diesem Fall sagen wir auch, dass $X$ über $Y$ separiert ist.
-
-Ein Schema $X$ heißt \textit{separiert}\index{separiert}, falls $X\to\operatorname{Spec}(\mathbb{Z})$ separiert ist.
-
-\paragraph{Beispiel 4.2.}\label{4.2} Sei $k$ ein Körper und $X$ die affine Gerade über $k$ mit doppeltem Nullpunkt wie in \hyperref[2.12]{Beispiel 2.12}.
-Dann ist $X\times_kX$ die affine Ebene mit vierfachen Nullpunkt und $\Delta(X)$ die gewöhnliche Diagonale, die zwei Nullpunkte besitzt.
-$\Delta(X)$ ist nicht abgeschlossen, da $\overline{\Delta(X)}$ vier Nullpunkte besitzt.
-
-\paragraph{Beispiel 4.3.}\label{4.3} Ist $V$ eine Varietät über eine algebraisch abgeschlossenen Körper $k$, so ist $t(V)$ separiert über $k$. Dies werden wir später zeigen.
-
-\paragraph{Beispiel 4.4.}\label{4.4} Ist $f:X\to Y$ ein Morphismus affiner Schemata, so ist $f$ separiert nach \hyperref[3.30]{Satz 3.30 (i)}.
-
-\paragraph{Beispiel 4.5.}\label{4.5} Ist $f:X\to Y$ eine Immersion, so ist $f$ separiert, da nach \hyperref[3.21]{Satz 3.21 (vi)} $X\cong X\times_XX\cong X\times_YX$ gilt, d.h.
-$\Delta_{X/Y}:X\to X$ ist ein Isomorphismus.
-
-\paragraph{Satz 4.6.}\label{4.6} Sei $f:X\to Y$ ein Morphismus von Schemata. Dann ist $f$ genau dann separiert, wenn $\Delta(X)\subset X\times_YX$ abgeschlossen ist.
-
-\paragraph{Beweis.} Für die nicht-triviale Richtung sei $\Delta(X)\subset X\times_YX$ abgeschlossen.
-Wir müssen zeigen, dass $X\to \Delta(X)$ ein Homöomorphismus ist, und dass der Morphismus $\mathcal{O}_{X\times_YX}\to\Delta_\ast\mathcal{O}_X$ surjektiv ist.
-\begin{itemize}
-\item Sei $p_1:X\times_YX\to X$ die erste Projektion. Da $p_1\circ\Delta=\operatorname{id}_X$, induziert $\Delta$ ein Homöomorphismus auf sein Bild. 
-\item Sei $P\in X$ und $V\subset_\text{o} Y$ affin.
-Wähle $U\subset_\text{o}X$ affin mit $P\in U$ und $f(U)\subset V$.
-Dann ist $U\times_VU$ eine offene, affine Umgebung von $\Delta(P)$.
-Nach \hyperref[4.4]{Beispiel 4.4} ist $\Delta:U\to U\times_VU$ eine abgeschlossene Immersion.
-Somit ist $\mathcal{O}_{X\times_YX}|_{U\times_VU}\to \Delta_\ast\mathcal{O}_X|_U$ surjektiv.\qed
-\end{itemize}
-
-\paragraph{Definition 4.7.}\label{4.7} Ein Ring $R$ heißt \textit{Bewertungsring}\index{Bewertungsring}, wenn $R$ nullteilerfrei ist, wenn von der folgenden Form ist:
-\[R=\{x\in K^\times\mid v(x)\geq 0\} \]
-wobei $v:K^\times\to G$ eine Bewertung ist, d.h.
-$v(xy)=v(x)+v(y)$ und $v(x+y)\geq\min\{v(x),v(y)\}$, und $G$ eine total geordnete abelsche Gruppe ist.
-$R$ ist dann lokaler Ring mit Maximalideal $\mathfrak{m}=\{x\in K^\times\mid v(x)>0\}\cup\{0\}$ und $K=\operatorname{Quot}(R)$.
-
-\paragraph{Theorem 4.8.}\label{4.8} \textit{(Bewertungstheoretisches Kriterium für Separiertheit)}\index{Bewertungstheoretisches Kriterium} Sei $f:X\to Y$ ein Morphismus von Schemata mit $X$ noethersch.
-Es sind folgende Aussagen äquivalent:
-\begin{enumerate}[(i)]
-\item $f$ ist separiert.
-\item Sei $R$ ein Bewertungsring des Körpers $K=\operatorname{Quot}(R)$ und $i:U=\operatorname{Spec}(K)\hookrightarrow T=\operatorname{Spec}(R)$ die kanonische Inklusion.
-Seien ferner Morphismen $T\to Y,\ U\to X$ derart gegeben, dass das folgende Diagramm kommutativ ist:
-\[\begin{tikzcd}
-U\ar[r]\ar[d, "i"'] & X\ar[d, "f"]\\
-T\ar[r]\ar[ur, "h", dashed] & Y
-\end{tikzcd}\]
-Dann gibt es höchstens eine Abbildung $h:T\to X$, die das obige Diagramm kommutativ macht. Mit anderen Worten: Die folgende kanonische Abbildung ist für jeden Bewertungsring $R$ über $Y$ injektiv:
-\[\operatorname{Hom}_Y(\operatorname{Spec}(R),X)\to\operatorname{Hom}_Y(\operatorname{Spec}(K),X),\ h\mapsto h\circ i \]
-\end{enumerate}
-
-\paragraph{Definition 4.9.}\label{4.9}  Sei $X$ ein topologischer Raum und $x\in X$.
-Ein Punkt $x'\in X$ heißt \textit{Spezialisierung}\index{Spezialisierung} von $x$, wenn $x'\in\overline{\{x\}}$ gilt.
-Ist $x''$ Spezialisierung von $x'$ und $x'$ Spezialisierung von $x$, so ist $x''$ auch Spezialisierung von $x$.
-
-Eine Teilmenge $Z\subset X$ heißt \textit{stabil unter Spezialisierungen}, falls mit $x\in Z$ auch jede Spezialisierung in $Z$ ist.
-Abgeschlossene Mengen sind stabil unter Spezialisierungen.
-Die Umkehrung gilt im Allgemeinen nicht.
-
-Ein Morphismus von Schemata $f$ heißt \textit{quasikompakt}\index{quasikompakt}, falls es eine offene affine Ü\-ber\-deckung $Y=\bigcup_\alpha Y_\alpha$ existiert, so dass alle $f^{-1}(Y_\alpha)$ quasikompakt sind.
-Es gilt:
-\[f\text{ quasikompakt}\iff\forall V\subset_\text{o}Y\text{ affin}\colon f^{-1}(V)\text{ quasikompakt} \]
-
-\paragraph{Satz 4.10.}\label{4.10} Sei $f:X\to Y$ ein quasikompakter Morphismus von Schemata. Dann sind folgende Aussagen äquivalent:
-\begin{enumerate}[(i)]
-\item $f$ ist abgeschlossen.
-\item Ist $x\in X$ und $y'$ eine Spezialisierung von $y=f(x)$, so gibt es eine Spezialisierung $x'$ von $x$ mit $f(x')=y'$.
-Mit anderen Worten: $f(\overline{\{x\}})$ ist stabil unter Spezialisierungen für alle $x\in X$.
-\end{enumerate}
-
-\paragraph{Beweis.} (i)$\implies$(ii) ist trivial.
-Sei also $X'\subset X$ abgeschlossen und setze $Y'=\overline{f(X')}$.
-Wir zeigen $Y'=f(X')$.
-Wir versehen $X'$ und $Y'$ mit der eindeutig bestimmten, reduzierte abgeschlossene Unterschemastruktur in $X$ bzw.
-$Y$.
-Seien $i:X'\hookrightarrow X$ und $j:Y'\hookrightarrow Y$ die natürlichen Inklusionen.
-Das Urbildschema $(f\circ i)^{-1}(Y')$ ist reduziert, besitzt $X'$ als unterliegender topologischer Raum und ist als Schema gleich $X'$, da die reduzierte Unterschemastrukturen eindeutig sind.
-Somit existiert ein eindeutig bestimmter Morphismus $f':X'\to Y'$:
-\[\begin{tikzcd}
-X'\ar[r, "f'"]\ar[d, "i"'] & Y'\ar[d, "j"]\\
-X\ar[r, "f"'] & Y
-\end{tikzcd} \]
-$f'$ erfüllt ebenfalls die Voraussetzung (ii).
-Ferner ist $f'$ quasikompakt.
-$i$ ist als abgeschlossene Immersion quasikompakt, also auch $f\circ i=j\circ f':X'\to Y$.
-Unter Basiswechsel sehen wir, dass $X'\times_YY'\to Y'$ quasikompakt ist.
-Da $j$ eine Immersion ist, ist $X'\cong X'\times_{Y'}Y' = X'\times_YY'$, also ist $X'\to Y'$ quasikompakt.
-
-Wir müssen noch zeigen: Ist $f:X\to Y$ ein quasikompakter, dominanter Morphismus reduzierter Schema, welcher (ii) erfüllt, so ist $f$ surjektiv.
-Sei dazu $y'\in Y$ gegeben und $y$ ein generischer Punkt der irreduziblen Komponente von $Y$, in der $y'$ liegt.
-Somit ist $y'$ eine Spezialisierung von $y$.
-Gilt nun $f^{-1}(y)\neq\varnothing$, so gibt es ein $x\in X$ mit $f(x)=y$.
-Wegen (ii) gibt es eine Spezialisierung $x'$ von $x$ mit $f(x')=y'$ und wir sind fertig.
-Somit ist noch das nächste Lemma zu zeigen.\qed
-
-\paragraph{Lemma 4.12.}\label{4.12} Für einen quasikompakten Morphismus $f:X\to Y$ ist äquivalent:
-\begin{enumerate}[(i)]
-\item $f$ ist dominant.
-\item Für die generischen Punkte $y$ von $Y$ der irreduziblen Komponenten gilt $f^{-1}(y)\neq\varnothing$.
-\end{enumerate}
-
-\iffalse
-\paragraph{Beweis.} (ii)$\implies$(i) ist trivial.
-Für die andere Richtung ersetzen wir $f$ durch $f_\text{red}$ und nehmen $X,Y$ als reduziert an.
-Sei $y\in Y$ ein generischer Punkt einer irreduziblen Komponente von $Y$ und $U\subset_\text{o}Y$ eine affine Umgebung von $y$.
-$U$ ist insbesondere quasikompakt, also auch $f^{-1}(U)$, d.h.
-$f^{-1}(U)=\bigcup_iV_i$ hat eine endliche Überdeckung durch offene, affine $V_i$.
-Da $f$ nach Voraussetzung dominant ist, folgt $U=\overline{ff^{-1}(U)}=\bigcup_i\overline{f(V_i)}$.
-Daher gibt es ein $i$ mit $y\in\overline{f(V_i)}$.
-$\overline{f(V_i)}\subset U$ wird mit der reduzierten, abgeschlossenen Unterschemastruktur versehen.
-Da alle $V_i$ reduziert sind, faktorisiert $f|_{V_i}:V_i\to U$ eindeutig in der Form $V_i\stackrel{g}{\to} \overline{f(V_i)}\hookrightarrow U$.
-
-Somit genügt es zu zeigen, dass $g^{-1}(y)\neq \varnothing$, da $g^{-1}(y)=f^{-1}(y)\cap V_i$.
-$y$ ist generischer Punkt einer irreduziblen Komponente von $Y$, daher gilt $\mathcal{O}_{U,y}=\mathcal{O}_{Y,y}$.
-\fi
-
-\paragraph{Korollar 4.13.}\label{4.13} Eine quasikompakte Immersion $f:X\to Y$ ist genau dann eine abgeschlossene Immersion, wenn $f(X)$ in $Y$ stabil unter Spezialisierungen ist.
-
-\paragraph{Beweis.} $f$ faktorisiert in eindeutiger Weise in der Form $X\stackrel{\sim}{\to}Z\stackrel{i}{\hookrightarrow} Y$, wobei $i$ die kanonische Inklusion des Unterschema $Z\subset Y$ ist.
-Es ist $i$ quasikompakt.
-Ist $f(X)=Z$ stabil unter Spezialisierungen, ist (ii) in \hyperref[4.10]{Satz 4.10} erfüllt, daher ist $i:Z\to Y$ abgeschlossen und $Z$ ein abgeschlossenes Unterschema in $Y$.\qed
-
-\paragraph{Lemma 4.14.}\label{4.14} Sei $R$ ein Bewertungsring eines Körpers $K$ und $T=\operatorname{Spec}(R),\ U=\operatorname{Spec}(K)$. Sei $X$ ein Schema. Dann gilt:
-\begin{enumerate}[(i)]
-\item $\operatorname{Hom}(U,X)$ ist die Menge aller Paare $(x,i)$ mit $x\in X$ und Körperhomomorphismus $i:\kappa(x)\hookrightarrow K$, wobei $\kappa(x)$ der Restklassenkörper in $x$ bezeichnet.
-\item $\operatorname{Hom}(T,X)$ ist die Menge aller Tripel $(x_0,x_1,i)$ mit $x_0,x_1\in X,\ x_0\in\overline{\{x_1\}}$ und Körperhomomorphismus $\kappa(x_1)\hookrightarrow K$, so dass $R$ über $\mathcal{O}_{\overline{\{x_1\}},x_0}$ \textit{dominiert}\index{dominiert}, wobei $\overline{\{x_1\}}$ mit der reduzierten Unterschemastruktur ausgestattet ist.
-\end{enumerate}
-
-\paragraph{Definition 4.15.}\label{4.15} Seien $(A,\mathfrak{m}_A),(B,\mathfrak{m}_B)$ lokale Ringe, die in einem Körper $K$ eingebettet sind.
-Wir sagen $B$ \textit{dominiert}\index{dominiert} $A$, wenn $A\subset B$ und $\mathfrak{m}_B\cap A=\mathfrak{m}_A$ gilt.
-Mit anderen Worten, wenn die natürliche Inklusion $A\hookrightarrow B$ ein lokaler Homomorphismus ist.
-
-\paragraph{Beweis von \hyperref[4.14]{Lemma 4.14}.} (i) ist gerade \hyperref[2.18]{Satz 2.18}.
-Setze $t_0=\mathfrak{m}_R\in T$ und $t_1=(0)\in T$.
-Für (ii) sei $f:T\to X$ ein Morphismus.
-Setze $x_i=f(t_i)$ und $Z=\overline{\{x_1\}}$ mit der reduzierten Unterschemastruktur.
-Dann gilt $f^{-1}(Z)=T$ als Mengen.
-Nach \hyperref[2.21]{Regeln~2.21~(iii)} faktorisiert $f$ in der Form:
-\[\begin{tikzcd}
-T\ar[rr, "f"]\ar[rrd] &&X\\
-&& Z\ar[u]
-\end{tikzcd} \]
-Beachte, dass für $U\subset_\text{o} Z$ mit $x_0\in U$ auch $x_1\in U$ gilt.
-Das induziert $\mathcal{O}_{Z,x_0}\to\mathcal{O}_{Z,x_1}$, analog haben wir $\mathcal{O}_{T,t_0}\to\mathcal{O}_{T,t_1}$.
-Wir erhalten folgendes kommutative Diagramm:
-\[\begin{tikzcd}
-\mathcal{O}_{Z,x_1} \ar[r]& \mathcal{O}_{T,t_1}=K\\
-\mathcal{O}_{Z,x_0} \ar[r]\ar[u]& \mathcal{O}_{T,t_0}=R\ar[u, hook]\\
-\mathfrak{m}_{Z,x_0}\ar[r]\ar[u, hook] & \mathfrak{m}_{T,t_0}\ar[u, hook]
-\end{tikzcd}\]
-Da $Z$ reduziert und irreduzibel ist, ist $Z$ integer. Daher ist $\mathcal{O}_{Z,x_1}=\kappa(x_1)$, alle Abbildungen sind injektiv und wir sehen, dass $R$ über $\mathcal{O}_{Z,x_0}$ dominiert.
-
-Sei umgekehrt $x_0,x_1\in X$ mit $x_0\in\overline{\{x_1\}}= Z$ und $\mathcal{O}_{Z,x_0}\hookrightarrow R$ ein lokaler Homomorphismus.
-Das induziert $T=\operatorname{Spec}(R)\to\operatorname{Spec}(\mathcal{O}_{Z,x_0})\to Z\to X$.
-Diese beiden Konstruktionen sind invers zueinander.\qed
-
-\paragraph{Beweis zu \hyperref[4.8]{Theorem 4.8}.} Sei $f:X\to Y$ ein Morphismus von Schemata mit $X$ noethersch.
-\begin{itemize}
-\item Sei zunächst $f$ separiert und $R$ ein Bewertungsring des Körpers $K=\operatorname{Quot}(R)$ mit Inklusion $i:U=\operatorname{Spec}(K)\hookrightarrow\operatorname{Spec}(R)=T$.
-Seien $T\to Y,\ U\to X$ gegeben und $h_1,h_2:T\to X$ mit kommutativem Diagramm:
-\[\begin{tikzcd}
-U\ar[d, "i"']\ar[r] & X\ar[d, "f"]\\
-T\ar[r]\ar[ur, "h_1", shift left]\ar[ur, "h_2"', shift right] & Y
-\end{tikzcd} \]
-Setze $t_0=\mathfrak{m}_R\in T$ und $t_1=(0)\in T$.
-Aus der Kommutativität $h_1i=h_2i$ folgt insbesondere $h_1(t_1)=h_2(t_1)$.
-Setze $h''=(h_1,h_2)_Y:T\to X\times_Y X$.
-Betrachte nun folgendes kommutative Diagramm:
-\[\begin{tikzcd}
-U\ar[r]\ar[d, "i"]\ar[rdd, "h_2i"', bend right=90]\ar[rrd, "h_1i", bend left=50] & X\ar[d, "\Delta"] &\\
-T\ar[r, "h''"'] & X\times_Y X\ar[r, "p_1"]\ar[d, "p_2"'] & X\ar[d, "f"]\\
-& X\ar[r, "f"'] & Y
-\end{tikzcd}\]
-Es gilt $\{h''(t_1)\}= h''i(U)\subset\Delta(X)$.
-Da $f$ separiert ist, ist $\Delta(X)$ abgeschlossen und es folgt $h''(t_0)\in h''(\overline{\{t_1\}})\subset\overline{\{h''(t_1)\}}\subset\Delta(X)$.
-Es folgt:
-\[h_1(t_0)=p_1h''(t_0)=p_2h''(t_0)=h_2(t_0) \]
-Aus der Kommutativität folgt, dass $h_1^\sharp$ und $h_2^\sharp$ dieselbe Abbildung $\kappa(x_1)\hookrightarrow K$ mit $x_1=h_1(t_1)=h_2(t_1)$ induzieren.
-Mit \hyperref[4.14]{Lemma 4.14 (ii)} folgt $h_1=h_2$.
-\item Für die andere Richtung genügt es nach \hyperref[4.6]{Satz 4.6} zu zeigen, dass $\Delta(X)\subset X\times_YX$ abgeschlossen ist.
-Nach \hyperref[3.30]{Satz 3.30 (ii)} ist $\Delta$ eine Immersion.
-Da $X$ noethersch ist, ist $\Delta$ quasikompakt.
-Nach \hyperref[4.13]{Korollar 4.13} genügt es zu zeigen, dass $\Delta(X)$ stabil unter Spezialisierungen ist.
-
-Sei also $\xi_1\in\Delta(X)$ und $\xi_0\in\overline{\{\xi_1\}}$.
-Sei $K=\kappa(\xi_1)$ und $\mathcal{O}=\mathcal{O}_{\overline{\{\xi_1\}},\xi_0}$, wobei $\overline{\{\xi_1\}}$ mit der reduzierten Unterschemastruktur versehen ist, d.h.
-$\mathcal{O}$ ist ein lokaler Ring in $K$.
-Mit \hyperref[4.16]{Satz 4.16} und dem Lemma von Zorn sehen wir, dass für jeden lokalen Ring $\mathcal{O}$ in $K$ einen Bewertungsring $R$ gibt, der $\mathcal{O}$ dominiert.
-Sei $R$ ein solcher Bewertungsring.
-Nach \hyperref[4.14]{4.14 (ii)} haben wir einen Morphismus $h:T=\operatorname{Spec}(R)\to X\times_YX$ mit $t_i\mapsto\xi_i$, wobei $t_0=\mathfrak{m}_R\in T,\ t_1=(0)\in T$.
-Betrachte das kommutative Diagramm:
-\[\begin{tikzcd}
-\operatorname{Spec}(K)\ar[r, "i"] & T\ar[dr, "h"]\ar[rrd, "p_1h", bend left]\ar[ddr, "p_2h"', bend right]\\
-& & X\times_YX\ar[r, "p_1"']\ar[d, "p_2"] & X\ar[d, "f"]\\
-&& X\ar[r, "f"'] & Y
-\end{tikzcd}\]
-Ist $x\in X$, so ist $\kappa(x)\cong\kappa(\Delta(x))$, wegen:
-\[\begin{tikzcd}
-X\ar[rr, "\Delta"]\ar[rrd, "\operatorname{id}"'] & & X\times_Y X\ar[d, "p_i"]\\
-&& X
-\end{tikzcd} \]
-Betrachte nun das folgende Diagramm:
-\[\begin{tikzcd}
-\operatorname{Spec}(K)\ar[rd, "hi"']\ar[r, "g", dashed] & X\ar[d, "\Delta"] \\
-& X\times_YX 
-\end{tikzcd}\]
-Wähle ein $x_1\in X$ mit $\Delta(x_1)=\xi_1$ und $j:\kappa(x_1)\hookrightarrow K$ als den von $hi$ induzierten $\kappa(\Delta(x_1))\hookrightarrow K$.
-Nach \hyperref[4.14]{Lemma 4.14 (i)} erhalten wir den zu $(x_1, j)$ passenden Morphismus $g:\operatorname{Spec}(K)\to X$, das das obige Diagramm kommutativ macht.
-Also faktorisiert $\operatorname{Spec}(K)\to X\times_YX$ über $\operatorname{Spec}(K)\to X\stackrel{\Delta}{\to}X\times_YX$, d.h.
-$p_1hi=p_2hi$.
-Nach Voraussetzung folgt $p_1h=p_2h$, d.h.
-auch $T\to X\times_YX$ faktorisiert über $T\to X\stackrel{\Delta}{\to}X\times_YX$.
-Somit folgt $\xi_0\in \Delta(X)$.\qed
-\end{itemize}
-
-\paragraph{Satz 4.16.}\label{4.16} Sei $K$ ein Körper und $R\subset K$ ein lokaler Ring.
-Dann ist $R$ genau dann ein Bewertungsring, wenn für jeden lokalen Ring $R\subset S\subset K$ mit lokalen Homomorphismus $R\hookrightarrow S$ stets $R=S$ folgt.
-
-\paragraph{Beweis.} Siehe z.B. Bourbaki, Algebra VI §13.
-
-\paragraph{Satz 4.17.}\label{4.17} Alle involvierten Schemata seine noethersch. Dann gilt:
-\begin{enumerate}[(i)]
-\item Offene und abgeschlossene Immersionen sind separiert.
-\item Die Komposition separierter Morphismen ist separiert.
-\item Separiertheit ist stabil unter Basiswechsel.
-\item Sind $f:X\to Y,\ f':X'\to Y'$ separierte Morphismen von $S$-Schemata, so ist auch $f\times_Sf':X\times_SX'\to Y\times_SY'$ separiert.
-\item Ist $f\circ f$ separiert, so ist auch $f$ separiert.
-\item $f:X\to Y$ ist genau dann separiert, wenn es eine offene Überdeckung $Y=\bigcup_\alpha Y_\alpha$ gibt, so dass alle $f^{-1}(Y_\alpha)\to Y_\alpha$ separiert sind.
-\end{enumerate}
-
-\paragraph{Beweis.} Folgt alles aus \hyperref[4.8]{Theorem 4.8}.\qed
-
-\paragraph{Definition 4.18.}\label{4.18} Ein Morphismus von Schemata $f:X\to Y$ heißt \textit{eigentlich}\index{eigentlich}, wenn gilt:
-\begin{enumerate}[(i)]
-\item $f$ ist von endlichem Typ.
-\item $f$ ist separiert.
-\item $f$ ist \textit{universell abgeschlossen}\index{universell abgeschlossen}, d.h. für jeden Morphismus $Y'\to Y$ ist der Morphismus $f':X'=X\times_YY'\to Y'$ abgeschlossen.
-\end{enumerate}
-In diesem Fall heißt $X$ auch \textit{eigentlich über} $Y$.
-
-\paragraph{Beispiel 4.19.}\label{4.19} Sei $k$ ein Körper und $X=\mathbf{A}_k^1$ die affine Gerade.
-Dann ist $X$ separiert und von endlichem Typ.
-Basiserweiterung mit $X\to k$ ergibt $X\times_kX\to X$, die Projektionsabbildung $\mathbf{A}_k^2\to\mathbf{A}_k^1$.
-Sei $Y=\operatorname{Spec}(k[X,Y]/(XY-1))$ die hyperbolische Kurve in $\mathbf{A}_k^2$.
-$Y\subset\mathbf{A}_k^2$.
-Diese ist abgeschlossen in $\mathbf{A}_k^2$ und projeziert sich in $\mathbf{A}_k^1$ auf $\mathbf{A}_k^1\setminus\{0\}$, die in $\mathbf{A}_k^1$ nicht abgeschlossen ist.
-Daher ist die Projektionsabbildung nicht abgeschlossen.
-In diesem Beispiel fehlt der unendliche Punkt von $Y$.
-Wir werden später zeigen, dass $X$ eigentlich über $k$ ist, wenn $X$ eine sogenannte projektive Varietät.
-
-\paragraph{Theorem 4.20.}\label{4.20} \textit{(Bewertungstheoretisches Kriterium für Eigentlichkeit)}\index{Bewertungstheoretisches Kriterium} Sei $f:X\to Y$ ein Morphismus von endlichem Typ mit $X$ noethersch.
-Dann sind äquivalent:
-\begin{enumerate}[(i)]
-\item $f$ ist eigentlich.
-\item Sei $R$ ein Bewertungsring des Körpers $K=\operatorname{Quot}(R)$ und $i:U=\operatorname{Spec}(K)\hookrightarrow T=\operatorname{Spec}(R)$ die kanonische Inklusion.
-Seien ferner Morphismen $T\to Y$ und $U\to X$ derart gegeben, dass das folgende Diagramm kommutativ ist:
-\[\begin{tikzcd}
-U\ar[r, "u"]\ar[d, "i"'] & X\ar[d, "f"]\\
-T\ar[r, "t"']\ar[ru, "h", dashed] & Y
-\end{tikzcd}\]
-Dann existiert ein eindeutig bestimmter Morphismus $h:T\to X$, der das Diagramm kommutativ macht. Mit anderen worten ist die folgende Abbildung bijektiv:
-\[\operatorname{Hom}_Y(T,X)\to\operatorname{Hom}_Y(U,X),\ h\mapsto h\circ i \]
-\end{enumerate}
-
-\paragraph{Lemma 4.21.}\label{4.21} \begin{enumerate}[(i)]
-\item Abgeschlossene Immersinoen sind von endlichem Typ. Quasikompakte offene Immersionen sind von endlichem Typ.
-\item Kompositum zweier Morphismen von endlichem Typ ist von endlichem Typ.
-\item Ein Morphismus $f:X\to Y$ ist genau dann von endlichem Typ, wenn $f$ lokal von endlichem Typ und quasikompakt ist.
-\item Seien $X\stackrel{f}{\to}Y\stackrel{g}{\to}Z$ Morphismen mit $f$ quasikompakt und $g\circ f$ von endlichem Typ. Dann ist $f$ von endlichem Typ.
-\end{enumerate}
-
-\paragraph{Beweis von \hyperref[4.20]{Theorem 4.20}.}\begin{itemize}
-\item Sei $f$ eigentlich.
-Nach Definition ist $f$ separiert, somit ist ein Morphismus $h:T\to X$ wie oben eindeutig bestimmt nach \hyperref[4.8]{Theorem 4.8}.
-Es bleibt die Existenz zu zeigen.
-Betrachte die Basiserweiterung $X_T=X\times_YT$ von $X$ mit $T\to Y$:
-\[\begin{tikzcd}
-U \ar[dr, "\theta"', dashed]\ar[rrd, "u", bend left]\ar[ddr, "i"', bend right] &&\\
-& X_T\ar[r, "p_1"']\ar[d, "f'"] & X\ar[d, "f"]\\
-& T\ar[r, "t"'] & Y
-\end{tikzcd} \]
-Wir erhalten ein $\theta:U\to X_T$, so dass das obige Diagramm kommutiert.
-Sei $\xi_1\in X_T$ das Bild des einzigen Punktes $t_1\in U$ und $Z=\overline{\{\xi_1\}}\subset X_T$ mit der reduzierten Unterschemastruktur.
-Da $f$ universell abgeschlossen ist, ist $f'$ abgeschlossen, somit ist $f'(Z)\subset T$ abgeschlossen.
-Da $f'(\xi_1)=t_1$ und $t_1$ der generische Punkt von $T$ ist, gilt $f'(Z)=T$.
-Es existiert also ein $\xi_0\in Z$ mit $f'(\xi_0)=t_0$, wobei $t_0$ der abgeschlossene Punkt in $T$ ist.
-
-Betrachte den zu $f'$ gehörigen lokalen Homomorphismus $R\hookrightarrow\mathcal{O}_{Z,\xi_0}$.
-Da $\xi_1$ der generische Punkt von $Z$ ist, gilt $\kappa(\xi_1)=\mathcal{O}_{Z,\xi_1}$ und nach \hyperref[4.14]{Lemma 4.14 (i)} erhalten wir $\kappa(\xi_1)\subset K$, der durch $U\to Z$ induziert wird.
-Nach \hyperref[4.16]{Satz 4.16} ist $R$ als Bewertungsring maximal unter allen lokalen Ringen in $K$ bzgl.
-Dominanz.
-Ferner gilt $\mathcal{O}_{Z,\xi_0}\subset\mathcal{O}_{Z,\xi_1}=\kappa(\xi_1)\subset K$ und $\mathcal{O}_{Z,\xi_0}\subsetneq K$, da $\xi_0\neq\xi_1$.
-Wegen $f'(\xi_0)=t_0$ dominiert $\mathcal{O}_{Z,\xi_0}$ den Ring $R$, d.h.
-$R\cong\mathcal{O}_{Z,\xi_0}$, insbesondere dominiert $R$ über $\mathcal{O}_{Z,\xi_0}$.
-Nach \hyperref[4.14]{Lemma 4.14 (ii)} erhalten wir Morphismus $h':T\cong\operatorname{Spec}(\mathcal{O}_{Z,\xi_0})\to X_T,\ t_i\mapsto \xi_i$.
-Wir erhalten $h:T\stackrel{h'}{\to}X_T\stackrel{p_1}{\to}X$ mit $fh=fp_1h'=tf'h'=t$ und $hi=p_1h'i=p_1h'f'\theta = p_1\theta = u$.
-\item Es gelte (ii).
-$f$ ist nach Voraussetzung von endlichem Typ und separiert nach \hyperref[4.8]{Theorem 4.8}.
-Sei also $Y'\to Y$ ein Morphismus und $f':X'=X\times_YY'\to Y'$ die Basiserweiterung von $f$.
-Wir zeigen, dass $f'$ abgeschlossen ist.
-Sei $Z\subset X'$ abgeschlossen mit der reduzierten Unterschemastruktur.
-Betrachte das kommutative Diagramm:
-\[ \begin{tikzcd}
-Z\ar[r, hook]\ar[dr] & X'\ar[r]\ar[d, "f'"] & X\ar[d, "f"]\\
-& Y'\ar[r] & Y
-\end{tikzcd} \]
-Da $f$ von endlichem Typ ist, ist nach \hyperref[3.25]{Satz 3.25 (ii)} auch $f'$ von endlichem Typ.
-Es folgt, dass $f'|_Z:Z\to Y'$ von endlichem Typ ist und somit quasikompakt.
-$f'|_Z$ faktorisiert über $Z\to f'(Z)\hookrightarrow Y'$.
-Wir sehen, dass $f'(Z)\hookrightarrow Y'$ quasikompakt ist.
-Nach \hyperref[4.13]{Korollar 4.13} genügt es zu zeigen, dass $f'(Z)$ stabil unter Spezialisierungen ist.
-
-Sei also $z_1\in Z$ und $y_1=f'(z_1),\ y_0\in\overline{\{y_1\}}$.
-Wir versehen $\overline{\{y_1\}}$ mit der reduzierten Unterschemastruktur.
-Sei $\mathcal{O}=\mathcal{O}_{\overline{\{y_1\}},y_0 }$.
-Dann ist $\operatorname{Quot}(\mathcal{O})=\kappa(y_1)\hookrightarrow \kappa(z_1)=:K$.
-Sei $R$ ein Bewertungsring von $K$, der $\mathcal{O}$ dominiert.
-Nach \hyperref[4.14]{Lemma 4.14} haben wir Morphismen:
-\[U=\operatorname{Spec}(K)\to Z,\ t_1\mapsto z_1 \]
-\[T=\operatorname{Spec}(R)\to Y',\ t_i\mapsto y_i \]
-Betrachte folgendes kommutative Diagramm:
-\[\begin{tikzcd}
-U\ar[r] \ar[d, "i"'] & Z\ar[r, hook] & X'\ar[r]& X\ar[d, "f"]\\
-T\ar[rr]\ar[rrru, dashed]\ar[rru, dashed] & & Y'\ar[r]\ar[from=u, "f'", crossing over]\ar[from=ul, crossing over] & Y
-\end{tikzcd} \]
-Nach Voraussetzung gibt es ein Morphismus $h:T\to X$, der das Diagramm kommutativ macht.
-Aus der Universaleigenschaft des Faserprodukts $X'$ erhalten wir ein Morphismus $h':T\to X'$, der $h:T\to X$ liftet.
-Nach Voraussetzung ist $Z$ abgeschlossen und da der generische Punkt $t_1\in T$ auf $z_1\in X'$ abgebildet wird, faktorisiert $h':T\to X'$ über $h':T\to Z\to X'$.
-Setze $z_0=h'(t_0)\in Z$.
-Dann ist $f'(z_0)=y_0$, also $y_0\in f'(Z)$.\qed
-\end{itemize}
-
-\paragraph{Korollar 4.22.}\label{4.22} Alle vorkommenden Schemata seien noethersch. \begin{enumerate}[(i)]
-\item Abgeschlossene Immersionen sind eigentlich.
-\item Kompositum zweier eigentlicher Morphismen ist eigentlich.
-\item Eigentliche Morphismen sind stabil unter Basiserweiterung.
-\item Seien $f:X\to Y$ und $f':X'\to Y'$ eigentliche Morphismen von $S$-Schemata. Dann ist $f\times f':X\times_SX'\to Y\times_SY'$ eigentlich.
-\item Seien $f:X\to Y,\ g:Y\to Z$ Morphismen. Ist $g\circ f$ eigentlich und $g$ separabel, so ist $f$ eigentlich.
-\item Ein Morphismus $f:X\to Y$ ist genau dann eigentlich, wenn es eine offene Ü\-ber\-deckung $Y=\bigcup_\alpha Y_\alpha$ gibt, so dass $f^{-1}(Y_\alpha)\to Y_\alpha$ für alle $\alpha$ eigentlich ist.
-\end{enumerate}
-
-\paragraph{Beweis.} Wir zeigen nur (v).
-Da $X$ noethersch ist, ist $f$ quasikompakt.
-Nach \hyperref[4.21]{Lemma 4.21} ist $f$ von endlichem Typ.
-Nach \hyperref[4.17]{Satz 4.17} ist $f$ separiert.
-Sei nun $R$ ein Bewertungsring und seien Morphismen $U\to X,\ T\to Y$ gegeben mit kommutativem Diagramm:
-\[\begin{tikzcd}
-U \ar[r, "u"]\ar[d, "i"']& X\ar[d, "f"]\\
-T\ar[r, "t"]\ar[dr, "t'"'] & Y\ar[d, "g"]\\
-& Z
-\end{tikzcd} \]
-Wir wollen ein $h:T\to X$ konstruieren mit $fh=t$ und $hi=u$.
-Setze $t'=gt$.
-Da $gf$ eigentlich ist, gibt es genau ein $h:T\to X$ mit $gfh=t'$ und $hi=u$.
-Sei $t''=fh$.
-Nun ist $g$ separabel und $ti=fu=fhi=t''i$ und $gt=t'=gfh=gt''$.
-Nach \hyperref[4.8]{Theorem 4.8} folgt $t=t''$ und somit $fh=t$.\qed
-
-\paragraph{} Ist $\varphi:A\to B$ ein Ringhomomorphismus und $\operatorname{Spec}(B)\to\operatorname{Spec}(A)$ der zugehörige Morphismus, so gilt:
-\[\mathbf{P}_B^n\cong\mathbf{P}_A^n\times_{\operatorname{Spec}(A)}\operatorname{Spec}(B) \]
-Insbesondere gilt $\mathbf{P}_A^n\cong\mathbf{P}_\mathbb{Z}^n\times_\mathbb{Z}A$.
-
-\paragraph{Definition 4.23.}\label{4.23} \begin{enumerate}[(i)]
-\item Sei $Y$ ein Schema. Dann heißt $\mathbf{P}_Y^n=\mathbf{P}_\mathbb{Z}^n\times_\mathbb{Z}Y$ der \textit{$n$-dimensionale projektive Raum über $Y$}\index{projektiver Raum}.
-\item Ein Morphismus $f:X\to Y$ von Schemata heißt \textit{projektiv}, wenn er eine Faktorisierung $f:X\stackrel{i}{\hookrightarrow} \mathbf{P}_Y^n\stackrel{p_2}{\to} Y$ besitzt, wobei $i$ eine abgeschlossene Immersion ist.
-\item Ein Morphismus $f:X\to Y$ von Schemata heißt \textit{quasiprojektiv}\index{quasiprojektiv}, falls er eine Faktorisierung der Form $f:X\stackrel{i}{\hookrightarrow} X'\stackrel{p}{\to}Y$ besitzt, wobei $i$ eine offene Immersion und $p$ projektiv ist.
-\end{enumerate}
-
-\paragraph{Theorem 4.24.}\label{4.24} Ein projektiver Morphismus von noetherschen Schemata ist eigentlich.
-Ein quasiprojektiver Morphismus von noetherschen Schemata ist von endlichem Typ und separiert.
-
-\paragraph{Beweis.} Es reicht zu zeigen, dass $\mathbf{P}_\mathbb{Z}^n$ eigentlich über $\mathbb{Z}$ ist, da der Basiswechsel nach \hyperref[4.22]{Korollar 4.22 (iii)} eigentlich über $Y$ ist.
-Ist $f:X\hookrightarrow \mathbf{P}^n_Y \to Y$ projektiv, so ist er als Verkettung von eigentlichen Morphismen wieder eigentlich, siehe \hyperref[4.22]{Korollar 4.22}.
-Ist $f:X\hookrightarrow X'\to Y$ quasiprojektiv, so ist $f$ als Verkettung separierter Morphismen separiert, siehe \hyperref[4.17]{Korollar 4.17}.
-Da $X$ noethersch ist, ist $X\hookrightarrow X'$ eine quasikompakte offene Immersion, also nach \hyperref[4.21]{Lemma 4.21 (i)} vom endlichem Typ.
-
-Sei also $X=\mathbf{P}_\mathbb{Z}^n$ und $X=\bigcup V_i$ eine offene affine Überdeckung mit $V_i=D_+(x_i)=\operatorname{Spec}\mathbb{Z}\big[\frac{x_0}{x_i},\ldots,\frac{x_n}{x_i} \big]$.
-Es ist also $X$ von endlichem Typ.
-Sei nun $R$ ein Bewertungsring mit Quotientenkörper $K=\operatorname{Quot}(R)$ und $U\to X,\ T\to\operatorname{Spec}(\mathbb{Z})$ Morphismen mit kommutativem Diagramm:
-\[\begin{tikzcd}
-U\ar[r, "u"]\ar[d, "i"'] & X\ar[d]\\
-T\ar[r, "t"'] & \operatorname{Spec}(\mathbb{Z})
-\end{tikzcd}\]
-Wir zeigen nun per Induktion, dass genau ein Morphismus $h:T\to X$ existiert, der das obige Diagramm kommutativ ergänzt.
-Für $n=0$ ist $X=\operatorname{Spec}(\mathbb{Z})$ und die Aussage ist klar.
-Sei $n\geq 1$ und $\xi_1$ das Bild des einzigen Punktes aus $U$ in $X$.
-Ist $\xi_1\in X\setminus V_i$ für ein $i$, so folgt die Behauptung nach Induktionsvoraussetzung, da die Hyperebene $X\setminus V_i$ isomorph zu $\mathbf{P}_\mathbb{Z}^{n-1}$ ist.
-
-Sei also $\xi_1\in\bigcap V_i$, d.h.
-alle Funktionen der Form $\frac{x_i}{x_j}$ sind invertierbar in $\mathcal{O}_{\xi_1}$.
-Der Morphismus $U\to X$ liefert Inklusion $\kappa(\xi_1)\hookrightarrow K$.
-Sei weiter $f_{ij}\in K^\times$ das Bild von $\frac{x_i}{x_j}$ unter $\mathcal{O}_{\xi_1}\to \kappa(\xi_1)\hookrightarrow K$.
-Dann folgt $f_{ik}=f_{ij}f_{jk}$ für alle $i,j,k$.
-Sei $v:K\to G$ die zu $R$ gehörige Bewertung und setze $g_i=v(f_{i0})$ für alle $i$.
-Sei $k$ derart, dass $g_k$ minimal unter $g_0,\ldots,g_n$ ist.
-Es gilt für alle $i$:
-\[v(f_{ik})=v(f_{i0})-v(f_{k0})=g_i-g_k\geq 0 \]
-Es folgt $f_{ik}\in R$ für alle $i$. Es gibt Homomorphismen:
-\[\begin{tikzcd}
-\mathbb{Z}\big[\frac{x_0}{x_k},\ldots,\frac{x_n}{x_k}\big]\ar[rr, "\varphi", "\frac{x_i}{x_k}\mapsto f_{ik}"'] \ar[d]&& R\ar[d]\\
-\mathcal{O}_{\xi_1}\ar[r] & \kappa(\xi_1)\ar[r] & K
-\end{tikzcd}\]
-Wir erhalten den zu $\varphi$ gehörige Morphismus $T\to V_k$ und somit $h:T\to V_k\hookrightarrow X$.
-Offensichtlich ist $T\to X\to\operatorname{Spec}(\mathbb{Z})$ der Morphismus $t$ unt $U\to T\to X$ der Morphismus $u$.
-Ferner ist $h$ eindeutig nach Konstruktion.\qed
-
-\paragraph{Satz 4.25.}\label{4.25} Sei $k$ ein algebraisch abgeschlossener Körper.
-Das Bild des Funktors $t:\mathbf{Var}(k)\to \mathbf{Sch}(k)$ ist die Menge aller quasiprojektiven, integren Schemata über $k$.
-Insbesondere ist $t(V)$ integer, separiert und von endlichem Typ für jede Varietät $V$.
-
-\paragraph{Beweis.} Ohne Beweis.
-
-\paragraph{Satz 4.26.}\label{4.26} Sei $X$ ein Schema von endlichem Typ über einem Körper $k$. Dann ist die Menge der abgeschlossenen Punkte in $X$ dicht in $X$.
-
-\iffalse
-\paragraph{Beweis.} Es genügt die Aussage für affine Schemata zu zeigen.
-Sei $X=\operatorname{Spec}(A)$ mit einer endlich erzeugten $k$-Algebra $A$ und $\mathfrak{M}$ die Menge der abgeschlossenen Punkte von $\operatorname{Spec}(A)$.
-Sei $f\in A$ mit $D(f)\neq\varnothing$.
-Wir zeigen $\mathfrak{M}\cap D(f)\neq\varnothing$.
-Es ist $D(f)\cong\operatorname{Spec}(A_f)$.
-Es existiert ein abgeschlossener Punkt $y\in \operatorname{Spec}(A_f)$, d.h.
-ein Maximalideal in $A_f$.
-Nun ist $A_f=A\big[\frac{1}{f}\big]$ endlich erzeugt über $k$, also $\kappa(y)=A_f/y$ eine algebraische Erweiterung von $k$ nach dem Hilbertschen Nullstellensatz.
-Sei $x$ der zu $y$ gehörige Punkt in $D(f)$.
-\fi
-
-\paragraph{Definition 4.27.}\label{4.27} Eine \textit{(abstrakte) Varietät}\index{Varietät!abstrakt} ist ein integres, separiertes Schema $X$ von endlichem Typ über einem algebraisch abgeschlossenen Körper $k$.
-Ist $X$ über $k$ eigentlich, so heißt $X$ \textit{vollständig}\index{Varietät!eigentlich}.
-Quasiprojektive abstrakte Varietäten entsprechen den klassischen Varietäten.
-
-\paragraph{Bemerkung 4.28.}\label{4.28}\begin{enumerate}[(i)]
-\item Eine projektive, abstrakte Varietät ist vollständig nach \hyperref[4.24]{Theorem 4.24}.
-\item Es gibt vollständige Varietäten, die nicht projektiv sind, d.h. die Klasse der abstrakten Varietäten ist größer als die Klasse der klassischen Varietäten.
-\item Jede vollständige abstrakte Varietät der Dimension $1$, d.h. eine vollständige Kurve, ist projektiv.
-\item Jede Varietät kann als offene Menge in eine vollständige Varietät eingebettet werden.
-\end{enumerate}
-
-\section{Modulgarben}
-
-\paragraph{Definition 5.1.}\label{5.1} Sei $(X,\mathcal{O}_X)$ ein geringter Raum.
-\begin{enumerate}[(i)]
-\item Eine \textit{Garbe von $\mathcal{O}_X$"~Moduln} bzw.
-\textit{$\mathcal{O}_X$"~Modul}\index{$\mathcal{O}_X$-Modul} ist eine Garbe $\mathcal{F}$ auf $X$ derart, dass $\mathcal{F}(U)$ ein $\mathcal{O}_X(U)$"~Modul für alle $U\subset_\text{o}X$ ist und alle $\operatorname{res}:\mathcal{F}(U)\to\mathcal{F}(V),\ V\subset U$ verträglich mit den Modulstrukturen via $\operatorname{res}:\mathcal{O}_X(U)\to\mathcal{O}_X(V)$ ist.
-\item Ein \textit{Morphismus}\index{Morphismus!$\mathcal{O}_X$-Moduln} $\mathcal{F}\to\mathcal{G}$ von $\mathcal{O}_X$-Moduln ist ein Garbenmorphismus, so dass $\mathcal{F}(U)\to\mathcal{G}(U)$ ein $\mathcal{O}_X(U)$"~Modulhomomorphismus für alle $U\subset_\text{o}X$ ist.
-\item Eine Sequenz von $\mathcal{O}_X$"~Moduln heißt \textit{exakt}\index{Exaktheit}, wenn sie exakt als Garbensequenz abelscher Garben ist.
-\item Sind $\mathcal{F},\mathcal{G}$ $\mathcal{O}_X$"~Moduln, so bezeichnet $\operatorname{Hom}_{\mathcal{O}_X}(\mathcal{F},\mathcal{G})$ die Gruppe der Morphismen von $\mathcal{F}$ nach $\mathcal{G}$.
-\item Sind $\mathcal{F},\mathcal{G}$ $\mathcal{O}_X$"~Moduln, so heißt die Garbe $U\mapsto \operatorname{Hom}_{\mathcal{O}_X|_U}(\mathcal{F}|_U,\mathcal{G}|_U)$ die \textit{Hom-Garbe}\index{Hom-Garbe} und wird mit $\operatorname{\mathcal{H}om}(\mathcal{F},\mathcal{G})$ bezeichnet.
-Diese ist ein $\mathcal{O}_X$"~Modul.
-\item Seien $\mathcal{F},\mathcal{G}$ $\mathcal{O}_X$"~Moduln.
-Dann heißt die zur Prägarbe $U\mapsto\mathcal{F}(U)\otimes_{\mathcal{O}_X(U)}\mathcal{G}(U)$ assoziierte Garbe das \textit{Tensorprodukt}\index{Tensorprodukt} von $\mathcal{F}$ und $\mathcal{G}$.
-Diese wird  mit $\mathcal{F}\otimes_{\mathcal{O}_X}\mathcal{G}$ bezeichnet.
-\item Ein $\mathcal{O}_X$"~Modul $\mathcal{F}$ heißt \textit{frei}\index{$\mathcal{O}_X$-Modul!frei}, wenn $\mathcal{F}$ isomorph zu einer direkten Summe von Exemplaren von $\mathcal{O}_X$ ist.
-\item Ein $\mathcal{O}_X$"~Modul $\mathcal{F}$ heißt \textit{lokal frei}\index{$\mathcal{O}_X$-Modul!lokal frei}, wenn $X$ durch offene Mengen $U$ überdeckt werden kann, so dass $\mathcal{F}|_U$ freier $\mathcal{O}_X|_U$"~Modul ist.
-
-Der \textit{Rang} $r$\index{Rang} von $\mathcal{F}$ auf $U$ ist gerade die Anzahl der Kopien von $\mathcal{O}_X|_U$.
-Wir schreiben $\operatorname{rang}(\mathcal{F}|_U)=r$.
-Ist $X$ zusammenhängend, so ist dieser Rang überall gleich.
-\item Ein lokal freier $\mathcal{O}_X$"~Modul vom Rang $1$ heißt \textit{invertierbare Garbe}\index{Garbe!invertierbar}.
-\item Eine \textit{Idealgarbe} auf $X$\index{Idealgarbe} ist ein $\mathcal{O}_X$"~Modul $\mathcal{J}$, der Untergarbe von $\mathcal{O}_X$ ist, d.h.
-$\mathcal{J}(U)$ ist ein Ideal von $\mathcal{O}_X(U)$ für $U\subset_\text{o}X$.
-\item Sei $f:(X,\mathcal{O}_X)\to (Y,\mathcal{O}_Y)$ ein Morphismus von geringten Räumen und $\mathcal{F}$ ein $\mathcal{O}_X$"~Modul.
-Dann ist $f_\ast\mathcal{F}$ ein $f_\ast\mathcal{O}_X$"~Modul und somit auch ein $\mathcal{O}_Y$"~Modul unter dem Garbenmorphismus $f^\sharp:\mathcal{O}_Y\to f_\ast\mathcal{O}_X$.
-$f_\ast\mathcal{F}$ heißt \textit{direktes Bild} von $\mathcal{F}$ unter $f$.\index{direktes Bild}
-
-Sei $\mathcal{G}$ ein $\mathcal{O}_Y$"~Modul.
-Dann ist $f^{-1}\mathcal{G}$ ein $f^{-1}\mathcal{O}_Y$"~Modul.
-Betrachte das Bild $\theta$ von $f^\sharp$ unter dem Adjunktionsisomorphismus $\operatorname{Hom}_Y(\mathcal{O}_Y,f_\ast\mathcal{O}_X)\to\operatorname{Hom}_X(f^{-1}\mathcal{O}_Y,\mathcal{O}_X)$.
-Durch $\theta$ wird $\mathcal{O}_X$ zu einem $f^{-1}\mathcal{O}_Y$"~Modul.
-Wir definieren:
-\[f^\ast\mathcal{G}= f^{-1}\mathcal{G}\otimes_{f^{-1}\mathcal{O}_Y}\mathcal{O}_X\]
-Somit ist $f^\ast\mathcal{G}$ ein $\mathcal{O}_X$"~Modul, das \textit{Urbild}\index{Urbild} von $\mathcal{G}$ unter $f$.
-
-Für jeden $\mathcal{O}_X$"~Modul $\mathcal{F}$ und $\mathcal{O}_Y$"~Modul $\mathcal{G}$ gilt:
-\[\operatorname{Hom}_{\mathcal{O}_X}(f^\ast\mathcal{G},\mathcal{F})\cong\operatorname{Hom}_{\mathcal{O}_Y}(\mathcal{G},f_\ast\mathcal{F}) \]
-Somit ist $f^\ast$ linksadjungiert zu $f_\ast$.
-\end{enumerate}
-
-\paragraph{Bemerkung.}\begin{enumerate}[(i)]
-\item Kern, Bild und Kokern eines Morphismus von $\mathcal{O}_X$"~Moduln ist wieder ein $\mathcal{O}_X$"~Modul.
-\item Sind $\mathcal{F},\mathcal{F}'$ $\mathcal{O}_X$"~Moduln und $\mathcal{F}'$ eine Untergarbe von $\mathcal{F}$, so ist $\mathcal{F}/\mathcal{F}'$ ein $\mathcal{O}_X$-Modul.
-\item Direkte Summen, direkte Produkte und projektive Limiten von $\mathcal{O}_X$"~Moduln sind wieder $\mathcal{O}_X$"~Moduln.
-\end{enumerate}
-
-\paragraph{Definition.} Sei $A$ ein Ring und $M$ ein $A$-Modul.
-Wir definieren die \textit{zu $M$ assoziierte Garbe}\index{assoziierte Garbe!Modul} $\widetilde{M}$ auf $\operatorname{Spec}(A)$ wie folgt: Für $U\subset_\text{o}\operatorname{Spec}(A)$ setzen wir $\widetilde{M}(U)$ als die Menge aller Abbildungen $s:U\to\coprod_{\mathfrak{p}\in U}M_\mathfrak{p}$, so dass:
-\begin{enumerate}[(i)]
-\item Für alle $\mathfrak{p}\in U$ gilt $s(\mathfrak{p})\in M_\mathfrak{p}$.
-\item Für alle $\mathfrak{p}\in U$ gibt es eine offene Umgebung $V$ von $\mathfrak{p}$ mit $V\subset U$ und Elemente $m\in M,\ f\in A$, so dass für alle $\mathfrak{q}\in V$ stets $f\not\in\mathfrak{q}$ und $s(\mathfrak{q})=\frac{m}{f}\in M_\mathfrak{q}$ gilt.
-\end{enumerate}
-Dann ist $\widetilde{M}$ mit den gewöhnlichen Restriktionsabbildungen eine Garbe.
-
-\paragraph{Satz 5.2.}\label{5.2} Sei $A$ ein Ring und $M$ ein $A$-Modul mit der assoziierten Garbe $\widetilde{M}$ auf $X=\operatorname{Spec}(A)$. Dann gilt:\begin{enumerate}[(i)]
-\item $\widetilde{M}$ ist ein $\mathcal{O}_X$"~Modul.
-\item Für jedes $\mathfrak{p}\in X$ gilt für den Halm von $\widetilde{M}$ in $\mathfrak{p}$ stets $\widetilde{M}_\mathfrak{p}\cong M_\mathfrak{p}$.
-\item Für alle $f\in A$ gibt es einen $A_f$-Modulisomorphismus $\widetilde{M}(D(f))\cong M_f$.
-\item Insbesondere gilt $\Gamma(X,\widetilde{M})\cong M$.
-\end{enumerate}
-
-\paragraph{Beweis.} (i) ist klar. (iv) folgt aus (iii) mit $f=1$. (ii) und (iii) gehen analog zu \hyperref[2.3]{2.3}.\qed
-
-\paragraph{Satz 5.3.}\label{5.3} Sei $A$ ein Ring und $X=\operatorname{Spec}(A)$.
-Ferner sei $\varphi:A\to B$ ein Ringhomomorphismus und $f:\operatorname{Spec}(B)\to \operatorname{Spec}(A)$ der entsprechende Morphismus.
-Dann gilt:
-\begin{enumerate}[(i)]
-\item Die Abbildung $M\mapsto\widetilde{M}$ liefert einen exakten, volltreuen Funktor von der Kategorie der $A$"~Moduln in die Kategorie der $\mathcal{O}_X$"~Moduln.
-\item Seien $M,N$ $A$"~Moduln. Dann gilt $\widetilde{M\otimes_AN}\cong \widetilde{M}\otimes_{\mathcal{O}_X}\widetilde{N}$.
-\item Sei $(M_i)_i$ eine Familie von $A$-Moduln. Dann gilt $\widetilde{\bigoplus_i M_i}\cong\bigoplus_i\widetilde{M_i}$.
-\item Sei $N$ ein $B$-Modul. Dann gilt $f_\ast\widetilde{N}\cong \widetilde{_AN}$, wobei $_AN$ den Modul $N$ als $A$"~Modul via $\varphi$ bezeichnet.
-\item Sei $M$ ein $A$"~Modul. Dann gilt $f^\ast\widetilde{M}\cong \widetilde{M\otimes_A B}$.
-\end{enumerate}
-
-\paragraph{Beweis.} Für (i) zeigen wir: \begin{itemize}
-\item \textit{Funktorialität:} Sei $\psi:M\to N$ ein $A$"~Modulhomomorphismus.
-Dieser induziert für alle $f\in A$ einen $A_f$"~Modulhomomorphismus $\psi_f:M_f\to N_f$.
-Ist $D(f)\supset D(g)$, so erhalten wir das folgende kommutative Diagramm:
-\[\begin{tikzcd}
-M_f\ar[rr, "\psi_f"]\ar[d] && N_f\ar[d]\\
-M_g\ar[rr, "\psi_g"'] && N_g
-\end{tikzcd} \]
-Da $D(f),\ f\in A$ eine Basis der Topologie bilden, induzieren $\psi_f,\ f\in A$ einen $\widetilde{A}$"~Homomorphismus $\widetilde{\psi}:\widetilde{M}\to\widetilde{N}$ mit $\widetilde{\psi}|_{D(f)}=\psi_f$.
-\item \textit{Volltreu:} Die Umkehrabbildung zu $\operatorname{Hom}_A(M,N)\to\operatorname{Hom}_{\widetilde{A}}(\widetilde{M},\widetilde{N})$ ist gegeben durch das Bilden der globalen Schnitte: 
-\[\operatorname{Hom}_{\widetilde{A}}(\widetilde{M},\widetilde{N})\to\operatorname{Hom}_A(\Gamma(X,\widetilde{M}),\Gamma(X,\widetilde{N}))=\operatorname{Hom}_A(M,N)\]
-\item \textit{Exaktheit:} Sei $0\to M'\to M\to M''\to 0$ eine kurze exakte Folge von $A$"~Moduln.
-Da Lokalisierungen exakt sind, ist $0\to M'_\mathfrak{p}\to M_\mathfrak{p}\to M''_\mathfrak{p}\to 0$ exakt.
-Nach \hyperref[5.2]{Satz 5.2 (ii)} ist $\widetilde{M}_\mathfrak{p}=M_\mathfrak{p}$.
-Da jede Halmsequenz exakt ist, folgt nach \hyperref[1.10]{1.10} die Exaktheit von $0\to\widetilde{M'}\to\widetilde{M}\to\widetilde{M''}\to 0$.
-\end{itemize}
-(ii) und (iii) folgen, da direkte Summen und Tensorprodukte mit Lokalisierungen kommutieren. Für (iv) sei $g\in A$. Es gilt:
-\[\Gamma(D(g),f_\ast\widetilde{N}) = \Gamma(f^{-1}(D(g)),\widetilde{N}) = \Gamma(D(\varphi(g)),\widetilde{N}) \cong N_{\varphi(g)} = N_g\cong \Gamma(D(g),\widetilde{N})\]
-Für (v) sei $N$ ein $B$-Modul. Dann gilt:
-\begin{align*}
-\operatorname{Hom}_{\widetilde{B}}(f^\ast\widetilde{M},\widetilde{N})&=\operatorname{Hom}_{\widetilde{A}}(\widetilde{M},f_\ast \widetilde{N}) \stackrel{\text{(iv)}}{=}\operatorname{Hom}_{\widetilde{A}}(\widetilde{M}, \widetilde{_AN})\\
-&= \operatorname{Hom}_A(M, {_AN}) \stackrel{\star}{\cong}\operatorname{Hom}_B(M\otimes_A B,N)\cong\operatorname{Hom}_{\widetilde{B}}(\widetilde{M\otimes_A B},\widetilde{N})
-\end{align*}
-wobei $\star$ durch die Abbildung $\eta\mapsto (m\otimes b\mapsto \eta(m)b)$ gegeben ist.\qed
-
-\paragraph{Definition 5.4.}\label{5.4} \begin{enumerate}[(i)]
-\item Sei $(X,\mathcal{O}_X)$ ein Schema.
-Ein $\mathcal{O}_X$"~Modul $\mathcal{F}$ heißt \textit{quasikohärent}\index{$\mathcal{O}_X$-Modul!quasikohärent}, falls es eine offene affine Überdeckung $U_i=\operatorname{Spec}(A_i),\ i\in I$ von $X$ gibt, so dass für jedes $i$ ein $A_i$"~Modul $M_i$ existiert mit $\mathcal{F}|_{U_i}\cong\widetilde{M_i}$.
-\item $\mathcal{F}$ heißt \textit{kohärent}\index{kohärent}, falls $\mathcal{F}$ quasikohärent ist und alle vorkommenden $M_i$ in (i) endlich erzeugte $A_i$"~Moduln sind.
-\end{enumerate}
-
-\paragraph{Beispiel 5.5.}\label{5.5} Für jedes Schema $X$ ist $\mathcal{O}_X$ kohärent, da $\mathcal{O}_X|_{\operatorname{Spec}(A)}=\widetilde{A}$.
-
-\paragraph{Beispiel 5.6.}\label{5.6} Sei $X=\operatorname{Spec}(A)$ affin und $Y\subset X$ ein abgeschlossenes Unterschema, das durch das Ideal $\mathfrak{a}\subset A$ definiert ist.
-Sei $i:Y\hookrightarrow X$ die natürliche Inklusion.
-Es ist $\mathcal{O}_Y\cong\widetilde{A/\mathfrak{a}}$ und somit $i_\ast\mathcal{O}_Y=\widetilde{A/\mathfrak{a}}$, wobei hier $A/\mathfrak{a}$ als $A$"~Modul aufgefasst wird.
-Somit ist $i_\ast\mathcal{O}_Y$ ein kohärenter $\mathcal{O}_X$"~Modul.
-
-\paragraph{Beispiel 5.7.}\label{5.7} Sei $X=\operatorname{Spec}(A)$ affin und $U\subsetneq_\text{o}X$ mit der natürlichen Inklusion $j:U\hookrightarrow X$.
-Betrachte die Garbe $j_!\mathcal{O}_U$, die außerhalb $U$ durch Null fortgesetzte Garbe von $\mathcal{O}_U$.
-$j_!\mathcal{O}_U$ ist nicht quasikohärent:
-
-Sei $X$ irreduzibel und $V=\operatorname{Spec}(A)\subset_\text{o}X$ mit $V\subsetneq U$.
-Wäre $j_!\mathcal{O}_U|_V\cong\widetilde{M}$ für einen $A$"~Modul $M$., so ist $(j_!\mathcal{O}_U|_V)(V)=M$, aber $(j_!\mathcal{O}_U|_V)(V)=0$ und $j_!\mathcal{O}_U|_V\neq 0$.
-
-\paragraph{Lemma 5.8.}\label{5.8} Sei $X=\operatorname{Spec}(A)$ ein affines Schema, $f\in A$ und $\mathcal{F}$ ein quasikohärenter $\mathcal{O}_X$"~Modul.
-\begin{enumerate}[(i)]
-\item Sei $s\in\Gamma(X,\mathcal{F})$ mit $s|_{D(f)}=0$. Dann existiert ein $n>0$ mit $f^ns=0$.
-\item Sei $t\in\Gamma(D(f),\mathcal{F})$. Dann existiert ein $n>0$ und $t'\in\mathcal{F}(X)$ mit $f^nt=t'|_{D(f)}$.
-\end{enumerate}
-
-\paragraph{Beweis.} Wir zeigen zunächst, dass es eine Überdeckung der Form $X=\bigcup_{i=1}^m D(g_i)$ gibt, so dass $\mathcal{F}|_{D(g_i)}\cong\widetilde{M_i}$ für einen $A_{g_i}$"~Modul $M_i$.
-
-Da $\mathcal{F}$ quasikohärent ist, existiert eine offene affine Überdeckung aus Mengen der Form $V=\operatorname{Spec}(B)$ mit $\mathcal{F}|_V=\widetilde{M}$ für einen $B$"~Modul $M$.
-Wir schreiben $V=\bigcup_{\text{gewisse }g\in A}D(g)$.
-Die natürlichen Morphismen $D(g)\hookrightarrow V$ liefern Ringhomomorphismen $B\to A_g$.
-Nach \hyperref[5.3]{Satz 5.3 (v)} ist $\mathcal{F}|_{D(g)}\cong \widetilde{M\otimes_B A_g}$.
-Da $X$ affin und somit quasikompakt ist, kann $X$ durch solche Mengen endlich überdeckt werden.
-\begin{enumerate}[(i)]
-\item Setze $s_i$ als das Bild von $s|_{D(g_i)}$ unter $\Gamma(D(g_i),\mathcal{F})=\Gamma(D(g_i),\widetilde{M_i})\cong M_i$.
-Wegen $D(fg_i)=D(f)\cap D(g_i)$ folgt nach \hyperref[5.2]{Satz 5.2 (iii)} $\Gamma(D(fg_i),\mathcal{F})=(M_i)_f$.
-Also ist $s_i=0$ in $(M_i)_f$.
-Nach Definition gibt es ein $n_i\in\mathbb{N}$ mit $f^{n_i}s_i=0$ in $M_i$.
-Sei $n$ das Maximum aller $n_i,\ i=1,\ldots,m$.
-Dann folgt $f^ns=0$ aus der ersten Garbeneigenschaft.
-\item Betrachte die Einschränkungen $t\in\Gamma(D(fg_i),\mathcal{F})=(M_i)_f$.
-Für alle $i$ gibt es ein $n_i\in\mathbb{N}$, so dass $f^{n_i}t=t_i|_{D(fg_i)}$ für ein $t_i\in \Gamma(D(g_i),\mathcal{F})=M_i$.
-Sei $n$ das Maximum aller $n_i,\ i=1,\ldots,m$.
-Dann gibt es für alle $i$ ein $t_i\in\Gamma(D(g_i),\mathcal{F})$ mit $f^n t = t_i|_{D(fg_i)}$.
-
-Auf $D(g_i)\cap D(g_j)=D(g_ig_j)$ haben wir Schnitte $t_i,t_j$ konstruiert, die auf $D(fg_ig_j)$ übereinstimmen.
-Nach (i) gibt es ein $m_{ij}$, so dass $f^{m_{ij}}(t_i-t_j)=0$ auf $D(g_ig_j)$ gilt.
-Sei $m$ das Maximum aller $m_{ij}$, so dass $f^m(t_i-t_j)=0$ auf $D(g_ig_j)$ für alle $i,j$ gilt.
-Die lokalen Schnitte $f^mt_i$ in $\Gamma(D(g_i),\mathcal{F})$ verkleben sich somit zu einem globalen Schnitt $t'$ von $\mathcal{F}$ zusammen mit $t'|_{D(f)}=f^{n+m}t$.\qed
-\end{enumerate}
-
-\paragraph{Satz 5.9.}\label{5.9} Sei $X$ ein Schema und $\mathcal{F}$ ein $\mathcal{O}_X$"~Modul.
-Dann ist $\mathcal{F}$ genau dann quasikohärent, wenn für alle affinen $U=\operatorname{Spec}(A)\subset_\text{o}X$ ein $A$"~Modul $M$ existiert mit $\mathcal{F}|_U\cong\widetilde{M}$.
-
-Ist $X$ noethersch, so ist $\mathcal{F}$ genau dann kohärent, wenn für alle affinen $U=\operatorname{Spec}(A)\subset_\text{o}X$ ein endlich erzeugter $A$"~Modul $M$ existiert mit $\mathcal{F}|_U\cong\widetilde{M}$.
-
-\paragraph{Lemma.} Sei $X=\operatorname{Spec}(A)$ affin, $M$ ein $A$"~Modul und $\mathcal{F}$ ein $\mathcal{O}_X$"~Modul. Dann ist
-\[\operatorname{Hom}_A(M,\Gamma(X,\mathcal{F}))\to\operatorname{Hom}_{\widetilde{A}}(\widetilde{M},\mathcal{F}),\ \varphi\mapsto\widetilde{\varphi} \]
-ein Isomorphismus mit Umkehrabbildung $\theta\mapsto\Gamma(X,\theta)$.
-
-\paragraph{Beweis von \hyperref[5.9]{Satz 5.9}.} Die Rückrichtungen der beiden Aussagen sind trivial.
-Sei $U\subset_\text{o}X$ affin.
-Nach dem Beweis von \hyperref[5.8]{Lemma 5.8} gibt es eine Basis der Topologie von $U$, bestehend aus affinen Teilmengen $V_i$ derart, dass $\mathcal{F}|_{V_i}\cong\widetilde{N_i}$ mit einem Modul $N_i$ ist.
-Somit ist $\mathcal{F}|_U$ quasikohärent.
-Wir können somit o.B.d.A.
-$X=U=\operatorname{Spec}(A)$ als affin annehmen.
-
-Setze $M=\Gamma(X,\mathcal{F})$ und $\alpha:\widetilde{M}\to\mathcal{F}$ als das Bild von $\operatorname{id}_M$ unter der Abbildung im vorherigen Lemma.
-Wie im Beweis von \hyperref[5.8]{Lemma 5.8} gezeigt, gibt es eine Überdeckung der Form $X=\bigcup_{i=1}^mD(g_i)$ mit $\mathcal{F}|_{D(g_i)}\cong\widetilde{M_i}$ für einen $A_{g_i}$"~Modul $M_i$.
-Es gilt $M_i=\mathcal{F}(D(g_i))\cong M_{g_i}$ und wir haben ein kommutatives Diagramm:
-\[\begin{tikzcd}
-\mathcal{F}(X)_{g_i}\ar[rr, "\cong"] && \mathcal{F}(D(g_i))\\
-\mathcal{F}(X)\ar[u]\ar[urr, "\operatorname{res}"']
-\end{tikzcd} \]
-Somit ist $\alpha|_{D(g_i)}$ ein Isomorphismus für alle $i$. Da die $D(g_i)$ ganz $X$ überdecken, ist $\alpha$ ein Isomorphismus.
-
-Sei nun $X$ zusätzlich noethersch.
-Dann sind die $A_{g_i}$"~Moduln $M_{g_i}$ endlich erzeugt.
-Wir zeigen, dass $M$ endlich erzeugt ist.
-Da $A$ noethersch ist, sind alle $A_{g_i}$ noethersch.
-Also sind die endlich erzeugten $A_{g_i}$"~Moduln $M_{g_i}$ noethersch.
-Analog zu \hyperref[3.6]{Satz 3.6} folgt $M$ noethersch.
-Insbesondere ist $M$ endlich erzeugt.
-\qed
-
-\paragraph{Korollar 5.10.}\label{5.10} Sei $A$ ein Ring und $X=\operatorname{Spec}(A)$. Dann ist
-\[\{\text{Kategorie der $A$-Moduln}\}\to\{\text{Kategorie der quasikohärenten $\mathcal{O}_X$-Moduln}\},\ M\mapsto\widetilde{M} \]
-ist eine Kategorienäquivalenz mit der Umkehrabbildung $\mathcal{F}\mapsto\Gamma(X,\mathcal{F})$.
-Ist $A$ noethersch, so geben dieselben Funktoren eine Kategorienäquivalenz zwischen den endlich erzeugten $A$-Moduln und den kohärenten $\mathcal{O}_X$"~Moduln.
-
-\paragraph{Beweis.} Sei $\mathcal{F}$ ein quasikohärenter $\mathcal{O}_X$"~Modul.
-Nach \hyperref[5.9]{Satz 5.9} existiert ein $A$"~Modul $M$ mit $\mathcal{F}\cong\widetilde{M}$.
-Nach \hyperref[5.2]{Satz 5.2 (iv)} ist $\Gamma(X,\mathcal{F})=M$, also:
-\[M\mapsto\widetilde{M}\mapsto \Gamma(X,\widetilde{M})=M,\quad\mathcal{F}=\widetilde{M}\mapsto \Gamma(X,\widetilde{M})=M\mapsto\widetilde{M}\qedhere \]
-
-\paragraph{Satz 5.11.}\label{5.11} Sei $X=\operatorname{Spec}(A)$ ein affines Schema und sei $0\to\mathcal{F}'\to\mathcal{F}\to\mathcal{F}''\to 0$ eine exakte Sequenz von $\mathcal{O}_X$"~Moduln, wobei $\mathcal{F}'$ quasikohärent ist.
-Dann ist die Sequenz der globalen Schnitte ebenfalls exakt:
-\[0\to\Gamma(X,\mathcal{F}')\to\Gamma(X,\mathcal{F})\to\Gamma(X,\mathcal{F}'')\to 0 \]
-
-\paragraph{Beweis.} Da $\Gamma(X,-)$ linksexakt ist, bleibt nur die Surjektivität von $\Gamma(X,\mathcal{F})\to\Gamma(X,\mathcal{F}'')$ zu zeigen.
-Sei $s\in\Gamma(X,\mathcal{F}'')$ gegeben.
-Für $x\in X$ gibt es ein $f\in A$ mit $x\in D(f) \subset X$, so dass $s|_{D(f)}$ sich zu einem $t\in\mathcal{F}(D(f))$ liftet.
-Wir zeigen nun, dass es ein $r>0$ existiert, so dass sich $f^rs$ zu einem $t''\in\mathcal{F}(X)$ liftet.
-
-Sei $X=\bigcup_i D(g_i)$ eine endliche offene Überdeckung, so dass sich $s|_{D(g_i)}$ zu einem $t_i\in\mathcal{F}(D(g_i))$ liftet.
-Auf $D(f)\cap D(g_i)=D(fg_i)$ liften $t_i,t\in\mathcal{F}(D(fg_i))$ beide $s$.
-Aus der Linksexaktheit von $\Gamma(D(fg_i), -)$ folgt $t-t_i\in\mathcal{F}'(D(fg_i))$.
-Da $\mathcal{F}'$ quasikohärent ist, folgt aus \hyperref[5.8]{Lemma 5.8 (ii)} die Existenz eines $n>0$ und $u_i\in\mathcal{F}'(D(g_i))$ mit $u_i|_{D(fg_i)}=f^n(t-t_i)$.
-Wir können $n$ unabhängig von $i$ wählen.
-Setze $t_i'=f^nt_i+u_i\in\mathcal{F}(D(g_i))$.
-Dann ist $t'_i$ ein Lift von $f^ns|_{D(g_i)}$ und $\star\ t_i'=f^nt$ auf $\mathcal{F}(D(fg_i))$.
-Auf $D(g_ig_j)$ liften $t_i'$ und $t_j'$ beide $f^ns$, also $t_i'-t_j'\in\mathcal{F}'(D(g_ig_j))$ und daher $t'_i=t'_j$ auf $\mathcal{F}(D(fg_ig_j))$ wegen $\star$.
-Nach \hyperref[5.8]{Lemma 5.8 (i)} existiert ein $m>0$, so dass $f^m(t_i'-t_j')=0$ auf $\mathcal{F}(D(g_ig_j))$.
-Wir können $m$ unabhängig von $i,j$ wählen.
-Somit verkleben sich die $f^mt_i'$ zu einem $t''\in\mathcal{F}(X)$ zusammen und $t''$ ist ein Lift von $f^{n+m}s$.
-
-Sei nun $X=\bigcup_i D(f_i)$ eine endliche offene Überdeckung, so dass sich $s|_{D(f_i)}$  zu einem Schnitt aus $\mathcal{F}(D(f_i))$ liften lässt.
-Für alle $i$ existiert ein $n$ und ein Lift $t_i\in\Gamma(X,\mathcal{F})$ von $f_i^ns$.
-Wir können o.B.d.A.
-$n$ unabhängig von $i$ annehmen.
-Da $X=\bigcup_i D(f_i)$, gilt $(f_1^n,\ldots,f_k^n)=A$, also $1=\sum_{i=1}^k a_if_i^n$ für gewisse $a_i\in A$.
-Setze $t=\sum_{i=1}^k a_it_i\in\mathcal{F}(X)$.
-Dann ist $t$ ein Lift von $\sum_{i=1}^k a_if_i^ns=s\in\mathcal{F}''(X)$.\qed
-
-\paragraph{Satz 5.12.}\label{5.12} Sei $X$ ein Schema.
-\begin{enumerate}[(i)]
-\item Kern, Kokern und Bild eines Morphismus von quasikohärenten Garben ist wieder quasikohärent.
-\item Ist $0\to\mathcal{F}'\to\mathcal{F}\to\mathcal{F}''\to 0$ eine kurze exakte Sequenz von $\mathcal{O}_X$"~Moduln und $\mathcal{F}',\mathcal{F}''$ quasikohärent, so ist $\mathcal{F}$ quasikohärent.
-\item Ist $X$ noethersch, so gilt (ii) auch für kohärente Garben.
-\end{enumerate}
-
-\paragraph{Beweis.} Da Quasikohärenz bzw.
-Kohärenz eine lokale Eigenschaft ist, können wir ohne Einschränkung $X=\operatorname{Spec}(A)$ als affin annehmen.
-Nach \hyperref[5.10]{Korollar 5.10} gelten (i) und (ii) für Modulgarben der Form $\widetilde{M}$.
-Da $M\mapsto \widetilde{M}$ nach \hyperref[5.3]{5.3 (i)} ein exakter, volltreuer Funktor ist, folgt die Aussage über Kern, Kokern und Bild.
-
-Sei nun $0\to\mathcal{F}'\to\mathcal{F}\to\mathcal{F}''\to 0$ eine kurze exakte Sequenz von $\mathcal{O}_X$"~Moduln mit $\mathcal{F}',\mathcal{F}''$ quasikohärent.
-Nach \hyperref[5.11]{Satz 5.11} ist die folgende Folge exakt:
-\[0\to\Gamma(X,\mathcal{F}')\to\Gamma(X,\mathcal{F})\to\Gamma(X,\mathcal{F}'')\to 0 \]
-Da $M\mapsto\widetilde{M}$ ein exakter Funktor ist, folgt die Exaktheit von:
-\[\begin{tikzcd}
-0  \ar[r]& \widetilde{\Gamma(X,\mathcal{F}')} \ar[d] \ar[r]& \widetilde{\Gamma(X,\mathcal{F})} \ar[d] \ar[r]& \widetilde{\Gamma(X,\mathcal{F}'')} \ar[d] \ar[r]& 0\\
-0 \ar[r] & \mathcal{F}'  \ar[r]& \mathcal{F}  \ar[r]&  \ar[r]\mathcal{F}''  \ar[r]& 0
-\end{tikzcd} \]
-Die vertikalen Pfeile links und rechts sind nach \hyperref[5.10]{Korollar 5.10} Isomorphismen.
-Nach dem 5er Lemma ist somit auch der mittlere Pfeil ein Isomorphismus und $\mathcal{F}$ ist quasikohärent.
-
-Sei nun $X$ noethersch und $\mathcal{F}',\mathcal{F}''$ kohärent.
-Dann sind $\Gamma(X,\mathcal{F}'),\Gamma(X,\mathcal{F}'')$ endlich erzeugt und somit auch $\Gamma(X,\mathcal{F})$.
-Es folgt die Kohärenz von $\mathcal{F}$.\qed
-
-\paragraph{Satz 5.13.}\label{5.13} Sei $f:X\to Y$ ein Morphismus von Schemata.
-\begin{enumerate}[(i)]
-\item Sei $\mathcal{G}$ ein quasikohärenter $\mathcal{O}_Y$"~Modul. Dann ist $f^\ast\mathcal{G}$ quasikohärenter $\mathcal{O}_X$"~Modul.
-\item Seien $X,Y$ noethersch und $\mathcal{G}$ kohärenter $\mathcal{O}_Y$"~Modul. Dann ist $f^\ast\mathcal{G}$ kohärent.
-\item Sei $X$ noethersch oder $f$ quasikompakt und separiert. Ist $\mathcal{F}$ ein quasikohärenter $\mathcal{O}_X$"~Modul, so ist $f_\ast\mathcal{F}$ quasikohärenter $\mathcal{O}_Y$"~Modul.
-\end{enumerate}
-
-\paragraph{Lemma 5.14.}\label{5.14} Sei $Y$ ein affines Schema, $f:X\to Y$ ein separierter Morphismus und $U,V\subset_\text{o}X$ affin.
-Dann ist $U\cap V$ ein abgeschlossenes Unterschema eines affinen Schemas.
-Wir werden in \hyperref[5.18]{Korollar 5.18} sehen, dass $U\cap V$ sogar affin ist.
-Insbesondere ist $U\cap V$ quasikompakt.
-
-\paragraph{Beweis.} Betrachte das kartesische Diagramm:
-\[\begin{tikzcd}
-X\times_Y X\ar[r, "p_2"]\ar[d, "p_1"'] & X\ar[d, "f"]\\
-X\ar[r, "f"'] & Y
-\end{tikzcd} \]
-Nach \hyperref[3.20]{Lemma~3.20} ist $p_1^{-1}(U)\cap p_2^{-1}(V)=U\times_YV$ affin.
-Es gilt: \[U\cap V\cong \Delta_{X/Y}(X)\cap U\times_YV\subset U\times_YV\] Nun ist $\Delta_{X/Y}(X)$ abgeschlossen in $X\times_YX$, d.h.
-$U\cap V$ ist ein abgeschlossenes Un\-ter\-sche\-ma in $U\times_YV$.\qed
-
-\paragraph{Beweis von \hyperref[5.13]{Satz 5.13}.} Für (i) und (ii) ist die Aussage lokal in $X$ und in $Y$.
-Daher können wir o.B.d.A.
-$X=\operatorname{Spec}(B)$ und $Y=\operatorname{Spec}(A)$ als affin annehmen.
-Dann folgt die Behauptung aus der Kategorienäquivalenz \hyperref[5.10]{Korollar 5.10} und \hyperref[5.3]{Satz 5.3 (v)} $f^\ast\widetilde{M} = \widetilde{M\otimes_AB}$.
-
-Für (iii) können wir nur $Y$ als affin annehmen.
-Sei $X=\bigcup_i U_i$ eine endliche, offene affine Überdeckung, da in beiden Fällen $X$ quasikompakt ist.
-Setze $U_{ij}=U_i\cap U_j$.
-In beiden Fällen sind $U_{ij}$ quasikompakt, siehe \hyperref[5.14]{Lemma 5.14} und \hyperref[3.5]{3.5}, \hyperref[3.4]{3.4}.
-Sei also $U_{ij}=\bigcup_k U_{ijk}$ eine endliche, offene affine Überdeckung.
-Nach der Garbeneigenschaft haben wir die folgende exakte Sequenz:
-\[0\to f_\ast\mathcal{F} \to\bigoplus_if_\ast(\mathcal{F}|_{U_i}) \to\bigoplus_{i,j,k}f_\ast(\mathcal{F}|_{U_{ijk}}) \]
-Wegen \hyperref[5.3]{Satz 5.3 (iv)} sind $f_\ast(\mathcal{F}|_{U_i})$ und $f_\ast(\mathcal{F}|_{U_{ijk}})$ quasikohärent.
-Nach \hyperref[5.12]{Satz 5.12 (ii)} sind $\bigoplus_i f_\ast(\mathcal{F}|_{U_i})$ und $\bigoplus_{i,j,k}f_\ast(\mathcal{F}|_{U_{ijk}})$ quasikohärent.
-Nach \hyperref[5.12]{5.12 (i)} ist daher auch $f_\ast\mathcal{F}$ quasikohärent.\qed
-
-\paragraph{Bemerkung.} Sind $X,Y$ noethersch und $\mathcal{F}$  kohärent, so muss $f_\ast\mathcal{F}$ nicht notwendigerweise kohärent sein.
-
-\paragraph{Definition 5.16.}\label{5.16} Sei $Y$ ein abgeschlossenes Unterschema von $X$ und $i:Y\hookrightarrow X$ der Inklusionsmorphismus.
-Dann ist die \textit{zu $Y$ gehörige Idealgarbe} auf $X$.\index{Idealgarbe} wie folgt definiert:
-\[\mathcal{J}_Y=\ker(i^\sharp:\mathcal{O}_X\to i_\ast\mathcal{O}_Y) \]
-
-\paragraph{Satz 5.17.}\label{5.17} Sei $X$ ein Schema. Dann gilt:
-\begin{enumerate}[(i)]
-\item Ist $Y$ ein abgeschlossenes Unterschema von $X$, so ist $\mathcal{J}_Y$ eine quasikohärente Idealgarbe auf $X$.
-\item Ist $X$ zusätzlich noethersch, so ist $\mathcal{J}_Y$ kohärent.
-\item Jede quasikohärente Idealgarbe auf $X$ bestimmt in eindeutiger Weise ein abgeschlossenes Unterschema.
-\end{enumerate}
-
-\paragraph{Beweis.}\begin{enumerate}[(i)]
-\item Sei $Y\subset X$ abgeschlossen.
-Dann ist $i:Y\hookrightarrow X$ ein quasikompakter Morphismus.
-Wegen \hyperref[4.17]{Satz 4.17 (i)} ist $i$ separiert.
-Nach \hyperref[5.13]{Satz 5.13 (iii)} ist $i_\ast\mathcal{O}_Y$ quasikohärent, also $\mathcal{J}_Y$ quasikohärent nach \hyperref[5.12]{Satz 5.12 (i)}.
-\item Ist $X$ noethersch und $U\subset_\text{o}X$ affin mit $U=\operatorname{Spec}(A)$, so ist auch $A$ noethersch nach \hyperref[3.6]{Satz 3.6}.
-Daher ist $I=\Gamma(U,\mathcal{J}_Y|_U)$ ein endlich erzeugtes Ideal in $A$.
-Nach \hyperref[5.9]{Satz 5.9} ist $\mathcal{J}_Y$ kohärent.
-\item Sei $\mathcal{J}$ eine quasikohärente Idealgarbe auf $X$. Setze:
-\[Y=\operatorname{Supp}(\mathcal{O}_X/\mathcal{J})=\{x\in X\mid (\mathcal{O}_X/\mathcal{J})_x\neq 0 \}\subset X \]
-Wir zeigen, dass $(Y,\mathcal{O}_X/\mathcal{J})$ ein abgeschlossenes Unterschema von $X$ ist.
-Dies ist eine lokale Frage, sei o.B.d.A.
-$X=\operatorname{Spec}(A)$ affin.
-Da $\mathcal{J}$ quasikohärent ist, folgt $\mathcal{J}=\widetilde{\mathfrak{a}}$ für ein Ideal $\mathfrak{a}\subset A$.
-Es gilt:
-\begin{align*}
-Y&=\operatorname{Supp}(\mathcal{O}_X/\mathcal{J})=\operatorname{Supp}(\widetilde{A/\mathfrak{a}})\\
-&= \{\mathfrak{p}\in X\mid (A/\mathfrak{a})_\mathfrak{p}\neq 0\}\\
-&=\{\mathfrak{p}\in X\mid\mathfrak{a}\subset\mathfrak{p}\}=\operatorname{Spec}(A/\mathfrak{a})
-\end{align*}
-Die Eindeutigkeit ist klar.\qed
-\end{enumerate}
-
-\paragraph{Korollar 5.18.}\label{5.18} Sei $X=\operatorname{Spec}(A)$ affin. Dann haben wir eine Bijektion:
-\[\{\mathfrak{a}\mid \mathfrak{a}\subset A \text{ Ideal}\}\to \{Y\mid Y\subset X \text{ abgeschlossenes Unterschema}\},\ \mathfrak{a}\mapsto \operatorname{Spec}(A/\mathfrak{a}) \]
-Insbesondere ist jedes abgeschlossenes Unterschema eines affinen Schemas wieder affin.
-
-\paragraph{Definition 5.19.}\label{5.19} Sei $S=\bigoplus_d S_d$ ein graduierter Ring.
-Ein $S$"~Modul heißt \textit{graduierter $S$"~Modul}\index{Modul!graduiert}, falls $M=\bigoplus_d M_d$ mit $S_d\cdot M_e\subset S_{d+e}$ gilt.
-Sei $\ell\in\mathbb{Z}$.
-Dann definieren wir den \textit{getwisteten $S$"~Modul}\index{Modul!getwistet} $M(\ell)$ von $M$ durch:
-\[M(\ell)_d = M_{d+\ell} \]
-
-\paragraph{Definition 5.20.}\label{5.20} Sei $S$ ein graduierter Ring und $M$ ein graduierter $S$"~Modul.
-Die zu $M$ assoziierte Garbe $\widetilde{M}$ auf $\operatorname{Proj}(S)$ ist wie folgt definiert: Sei $U\subset_\text{o}\operatorname{Proj}(S)$ und setze $\widetilde{M}(U)$ als die Menge aller Abbildungen $s:U\to\coprod_{\mathfrak{p}\in U}M_{(\mathfrak{p})}$, so dass:
-\begin{enumerate}[(i)]
-\item Für alle $\mathfrak{p}\in U$ gilt $s(\mathfrak{p})\in M_{(\mathfrak{p})}$.
-\item Für alle $\mathfrak{p}\in U$ existiert eine offene Umgebung $V$ von $\mathfrak{p}$ mit $V\subset U$ und homogene Elemente $m\in M,\ f\in S$ mit $\deg(m)=\deg(f)$ derart, dass für alle $\mathfrak{q}\in V$ stets $f\not\in\mathfrak{q}$ und $s(\mathfrak{q})=\frac{m}{f}$ in $M_{(\mathfrak{q})}$ gilt.
-\end{enumerate}
-$\widetilde{M}$ wird zu einer Garbe mit den gewöhnlichen Restriktionsabbildungen.
-
-\paragraph{Satz 5.21.}\label{5.21} Sei $S$ ein graduierter Ring, $M$ ein graduierter $S$"~Modul und $X=\operatorname{Proj}(S)$. Dann gilt:
-\begin{enumerate}[(i)]
-\item $(\widetilde{M})_\mathfrak{p}=M_{(\mathfrak{p})}$ für alle $\mathfrak{p}\in X$.
-\item Für alle homogene Elemente $f\in S_+$ ist $\widetilde{M}|_{D_+(f)}\cong\widetilde{M_{(f)}}$ bzgl. des Isomorphismus' $D_+(f)\cong\operatorname{Spec}S_{(f)}$.
-\item $\widetilde{M}$ ist ein quasikohärenter $\mathcal{O}_X$"~Modul. Ist $X$ noethersch und $M$ endlich erzeugt, so ist $\widetilde{M}$ kohärent.
-\end{enumerate}
-
-\paragraph{Beweis.} (i) und (ii) sind analog zu \hyperref[2.23]{Satz 2.23}. (iii) folgt aus (ii).\qed
-
-\paragraph{Definition 5.22.}\label{5.22} Sei $S$ ein graduierter Ring, $X=\operatorname{Proj}(S)$ und $\mathcal{F}$ ein $\mathcal{O}_X$"~Modul.
-Für $n\in\mathbb{Z}$ definieren wir die \textit{getwistete Garbe}\index{Garbe!getwistet} $\mathcal{F}(n)$ von $\mathcal{F}$ wie folgt:
-\[\mathcal{F}(n)=\mathcal{F}\otimes_{\mathcal{O}_X}\mathcal{O}_X(n),\quad \mathcal{O}_X(n)=\widetilde{S(n)} \]
-
-\paragraph{Satz 5.23.}\label{5.23} Sei $S$ ein graduierter Ring und $X=\operatorname{Proj}(S)$, wobei $S$ als $S_0$"~Algebra von $S_1$ erzeugt wird. Dann gilt:
-\begin{enumerate}[(i)]
-\item $\mathcal{O}_X(n)$ ist eine invertierbare Garbe auf $X$.
-\item Sind $M,N$ graduierte $S$"~Moduln, so ist $\widetilde{M\otimes_SN}\cong\widetilde{M}\otimes_{\mathcal{O}_X}\widetilde{N}$.
-Insbesondere gilt $\widetilde{M(n)}\cong\widetilde{M}(n)$ und $\mathcal{O}_X(n+m)\cong\mathcal{O}_X(n)\otimes\mathcal{O}_X(m)$.
-\item Sei $T$ ein weiterer graduierter Ring, der von $T_1$ als $T_0$"~Algebra erzeugt wird und $\varphi:S\to T$ ein Homomorphismus graduierter Ringe.
-Sei $U\subset_\text{o} Y=\operatorname{Proj}(T)$ und $f:U\to X$ der durch $\varphi$ induzierte Morphismus.
-Dann gilt für jeder graduierte $S$"~Modul $M$ und jeder graduierte $T$"~Modul $N$:
-\[f^\ast(\widetilde{M})\cong\widetilde{M\otimes_ST}|_U,\quad f_\ast(\widetilde{N}|_U)\cong (\widetilde{_SN}) \]
-Insbesondere gilt $f^\ast(\mathcal{O}_X(n))\cong\mathcal{O}_Y(n)|_U$ und $f_\ast(\mathcal{O}_X(n)|_U)=(f_\ast\mathcal{O}_U)(n)$.
-\end{enumerate}
-
-\paragraph{Beweis.}\begin{enumerate}[(i)]
-\item Sei $f\in S_1$ und betrachte $\mathcal{O}_X(n)|_{D_+(f)}\cong\widetilde{S(n)_{(f)}}$ auf $\operatorname{Spec}S_{(f)}$.
-Es ist $S(n)_{(f)}$ freier $S_{(f)}$"~Modul vom Rang $1$ via dem Isomorphismus $(S_f)_0\to (S_f)_n,\ s\mapsto f^ns$ für alle $n$.
-Da $S$ von $S_1$ als $S_0$"~Algebra erzeugt wird, gilt $X=\bigcup_{f\in S_1}D_+(f)$.
-Daher ist $\mathcal{O}_X(n)$ invertierbar.
-\item Sei $f\in S_1$.
-Es gilt $(M\otimes_SN)_{(f)}=M_{(f)}\otimes_{S_{(f)}}N_{(f)}$.
-Da $S$ von $S_1$ erzeugt wird, folgt $\widetilde{M\otimes_SN}\cong\widetilde{M}\otimes_{\mathcal{O}_X}\widetilde{N}$.
-\item Analog wie im affinen Fall.\qed
-\end{enumerate}
-
-\paragraph{Definition 5.24.}\label{5.24} Sei $S$ ein graduierter Ring, $X=\operatorname{Proj}(S)$ und $\mathcal{F}$ ein $\mathcal{O}_X$"~Modul.
-Der zu $\mathcal{F}$ \textit{assoziierte graduierte $S$"~Modul}\index{Modul!assoziiert} $\Gamma_\ast(\mathcal{F})$ ist definiert als die Gruppe
-\[\Gamma_\ast(\mathcal{F})=\bigoplus_{n\in\mathbb{Z}}\Gamma(X,\mathcal{F}(n)) \]
-mit der folgenden $S$"~Wirkung: Ein $s\in S_d$ induziert ein $s\in \Gamma(X,\mathcal{O}_X(d))$.
-Für $t\in\Gamma(X,\mathcal{F}(n))$ setze $st= s\otimes t\in\Gamma(X,\mathcal{F}(n)\otimes_{\mathcal{O}_X}\mathcal{O}_X(d))=\Gamma(X,\mathcal{F}(n+d))$.
-
-\paragraph{Satz 5.25.}\label{5.25} Sei $A$ ein Ring und $X=\mathbf{P}_A^r$ mit $r\geq 1$ und $S=A[X_0,\ldots,X_r]$. Dann gilt:
-\[\Gamma_\ast(\mathcal{O}_X)= S \]
-
-\paragraph{Beweis.} Sei $X=\bigcup_{i=0}^r D_+(X_i)$.
-Ein $t\in\Gamma(X,\mathcal{O}_X(n))$ entspricht eine Familie $t_i\in\Gamma(D_+(X_i),\mathcal{O}_X(n)),\ i=1,\ldots,r$ mit $t_i=t_j$ auf $D_+(X_iX_j)$.
-$t_i$ ist ein homogenes Element $s_i\in S_{X_i}$ vom Grad $n$ und $t_i|_{D_+(X_iX_j)}$ entspricht dem Bild von $s_i$ in $S_{X_iX_j}$.
-Es folgt:
-\[\Gamma_\ast(\mathcal{O}_X)=\Big\{ (t_0,\ldots,t_r)\in\prod_{i=0}^r S_{X_i}\mid t_i=t_j\text{ auf } S_{X_iX_j}\text{ für alle }i,j \Big\} \]
-Da keine $X_i$ Nullteiler sind, haben wir Inklusionen $S\hookrightarrow S_{X_i}\hookrightarrow S_{X_iX_j}\hookrightarrow S_{X_0\cdots X_r}$ und:
-\[\Gamma_\ast(\mathcal{O}_X)=\bigcap_{i=0}^r S_{X_i}\subset S_{X_0\cdots X_r} \]
-Jedes homogene $t\in S_{X_0\cdots X_r}$ lässt sich eindeutig in der folgenden Form schreiben:
-\[t=X_0^{i_0}\cdots X_r^{i_r}f,\quad i_j\in\mathbb{Z}\]
-wobei $f\in S$ ein homogenes Element ist, das durch kein $X_i$ teilbar ist. Es ist $t$ genau dann in $S_{X_i}$, wenn $i_j\geq 0$ für alle $j\neq i$ gilt. Also ist $\bigcap S_{X_i}=S$.\qed
-
-\paragraph{Lemma 5.26.}\label{5.26} Sei $X$ ein Schema, $\mathcal{L}$ eine invertierbare Garbe auf $X$ und $f\in\Gamma(X,\mathcal{L})$.
-Setze $X_f=\{x\in X\mid f_x\not\in\mathfrak{m}_x\mathcal{L}_x\}\subset_\text{o}X$ und sei $\mathcal{F}$ eine quasikohärente Garbe auf $X$.
-\begin{enumerate}[(i)]
-\item Sei $X$ quasikompakt und $s\in\Gamma(X,\mathcal{F})$ mit $s|_{X_f}=0$. Dann gibt es ein $n>0$, so dass $f^ns=0\in\Gamma(X,\mathcal{F}\otimes\mathcal{L}^{\otimes n})$.
-\item Sei $X=\bigcup U_i$ eine endliche, offene affine Überdeckung, so dass $\mathcal{L}|_{U_i}$ für alle $i$ frei und $U_i\cap U_j$ für alle $i,j$ quasikompakt sind.
-Zu $t\in\Gamma(X_f,\mathcal{F})$ gibt es ein $n>0$, so dass sich $f^nt\in\Gamma(X_f,\mathcal{F}\otimes\mathcal{L}^{\otimes n})$ zu einem globalen Schnitt auf ganz $X$ fortsetzen lässt.
-\end{enumerate}
-
-\paragraph{Bemerkung 5.27.}\label{5.27} Voraussetzungen in \hyperref[5.26]{Lemma 5.26 (i) und (ii)} sind erfüllt, wenn $X$ noethersch ist, oder wenn $X$ quasikompakt und separiert ist.
-
-\paragraph{Beweis von \hyperref[5.26]{Lemma 5.26}.} \begin{enumerate}[(i)]
-\item Sei $X=\bigcup U_i$ eine endliche, offene affine Überdeckung mit $\mathcal{L}|_{U_i}$ frei.
-Betrachte $U=U_i$ und sei $\psi:\mathcal{L}|_U\stackrel{\sim}{\to}\mathcal{O}_U$ ein Isomorphismus.
-Da $\mathcal{F}$ quasikohärent ist, folgt $\mathcal{F}|_U\cong\widetilde{M}$ für ein $A$"~Modul $M$, wobei $U=\operatorname{Spec}(A)$.
-Für ein $s\in\Gamma(X,\mathcal{F})$ ist $s|_U\in M$.
-Setze $g:=\psi(f|_U)\in A$.
-Es ist $X_f\cap U=D(g)$ und $s|_{X_f}=0$.
-Nach \hyperref[5.8]{Lemma 5.8 (i)} gibt es ein $n>0$ mit $g^ns=0\in M$.
-Der Isomorphismus
-\[\operatorname{id}\otimes \psi^{\otimes n}:\mathcal{F}\otimes \mathcal{L}^{\otimes n}|_U\to\mathcal{F}|_U \]
-liefert $0=f^ns=\Gamma(U, \mathcal{F}\otimes\mathcal{L}^{\otimes n})$ für alle $U=U_i$. Wählt man $n$ so groß, dass die obige Aussage für alle $U_i$ gilt, so folgt $f^ns=0$ auf $X$.
-\item Analog zu (i) mit \hyperref[5.8]{Lemma 5.8 (ii)}.\qed
-\end{enumerate}
-
-\paragraph{Satz 5.28.}\label{5.28} Sei $S$ ein graduierter Ring, der durch $S_1$ als $S_0$"~Algebra endlich erzeugt wird.
-Sei $X=\operatorname{Proj}(S)$ und $\mathcal{F}$ ein quasikohärenter $\mathcal{O}_X$"~Modul.
-Dann gibt es einen natürlichen Isomorphismus:
-\[\beta:\widetilde{\Gamma_\ast(\mathcal{F})}\stackrel{\sim}{\to}\mathcal{F} \]
-
-\paragraph{Beweis.} Es ist $X=\bigcup_{f\in S_1}D_+(f)$ eine endliche Vereinigung. Für $f\in S_1$ definiere:
-\[\beta_f:\widetilde{\Gamma_\ast(\mathcal{F})_{(f)}}=\widetilde{\Gamma_\ast(\mathcal{F})}|_{D_+(f)}\to\mathcal{F}|_{D_+(f)} \]
-durch $\bigoplus_d\Gamma(D_+(f),\mathcal{F}(d))\to\Gamma(D_+(f),\mathcal{F})$, das induziert wird durch:
-\[\Gamma(D_+(f),\mathcal{F}(d))\to\Gamma(D_+(f),\mathcal{F}(d)\otimes \mathcal{O}_X(-d))= \Gamma(D_+(f),\mathcal{F}),\ \frac{m}{f^d}\mapsto m\otimes f^{-d} \]
-Wir erhalten eine Abbildung $\beta:\widetilde{\Gamma_\ast(\mathcal{F})}\to\mathcal{F}$.
-Wir zeigen nun, dass alle $\beta_f$ Isomorphismen sind.
-Es genügt zu zeigen, dass $\Gamma_\ast(\mathcal{F})_{(f)}\to\Gamma(D_+(f),\mathcal{F})$ Isomorphismen sind.
-\hyperref[5.26]{Lemma 5.26 (i)} liefert die Injektivität und \hyperref[5.26]{Lemma 5.26 (ii)} die Surjektivität.\qed
-
-\paragraph{Korollar 5.29.}\label{5.29} Sei $A$ ein Ring. Dann gilt:
-\begin{enumerate}[(i)]
-\item Ist $Y\hookrightarrow\mathbf{P}_A^r$ ein abgeschlossenes Unterschema, so existiert ein homogenes Ideal $I\subset S=A[X_0,\ldots,X_r]$, so dass $Y=\operatorname{Proj}(S/I)\hookrightarrow\operatorname{Proj}(S)=X$.
-\item Sei $Y$ ein Schema über $\operatorname{Spec}(A)$.
-Dann ist $Y$ genau dann projektiv, wenn $Y\cong\operatorname{Proj}(S)$ für einen graduierten Ring $S$, der von $S_1$ als $S_0=A$"~Algebra endlich erzeugt wird.
-\end{enumerate}
-
-\paragraph{Beweis.} \begin{enumerate}[(i)]
-\item Sei $\mathcal{J}_Y\subset\mathcal{O}_X$ die Idealgarbe von $Y$ auf $X=\mathbf{P}_A^r$.
-Da $\mathcal{J}_Y(d)\subset\mathcal{O}_X(d)$, folgt $\Gamma_\ast(\mathcal{J}_Y)\subset\Gamma_\ast(\mathcal{O}_X)$.
-Nach \hyperref[5.25]{Satz 5.25} ist $\Gamma_\ast(\mathcal{O}_X)=S$, d.h.
-$I=\Gamma_\ast(\mathcal{J}_Y)$ ist ein homogenes Ideal in $S$.
-Setze $Y'=\operatorname{Proj}(S/I)$.
-$Y'$ ist ein abgeschlossenes Unterschema von $X$ mit Idealgarbe $\mathcal{J}_{Y'}=\widetilde{I}$.
-Da $\mathcal{J}_Y$ nach \hyperref[5.17]{Satz 5.17 (i)} quasikohärent ist, folgt $\mathcal{J}_Y\cong\widetilde{\Gamma_\ast(\mathcal{J}_Y)}$ nach \hyperref[5.28]{Satz 5.28}.
-Nun gilt:
-\[\mathcal{J}_Y\cong\widetilde{\Gamma_\ast(\mathcal{J}_Y)}=\widetilde{I}=\mathcal{J}_{Y'} \]
-Nach \hyperref[5.17]{Satz 5.17} folgt $Y=Y'$, also ist $Y$ das abgeschlossene Unterschema, das durch $I$ definiert ist.
-\item Es gilt:
-\begin{align*}
-Y\text{ projektiv} &\iff Y\text{ ist abgeschlossenes Unterschema von }\mathbf{P}_A^r\text{ für ein }r\\
-&\iff Y\cong\operatorname{Proj}(S'/I)\text{ für ein homogenes Ideal }I\subset S' = A[X_0,\ldots,X_r]
-\end{align*}
-Wir zeigen zunächst, dass $I$ und $I'=\bigoplus_{d\geq d_0}I_d$ dasselbe abgeschlossene Unterschema bestimmen.
-Dafür zeigen wir $\mathfrak{p}\supset I$, wenn $\mathfrak{p}\supset I'$ für alle $\mathfrak{p}\in\operatorname{Proj}(S')$.
-Wegen $\mathfrak{p}\not\supset S_+$ gibt es ein $x_i\not\in\mathfrak{p}$.
-Sei nun $x\in I_r,\ r<d_0$, dann ist $x_i^{d_0-r}x\in I_{d_0}\in\mathfrak{p}$.
-Da $\mathfrak{p}$ ein Primideal ist, folgt $x\in\mathfrak{p}$.
-
-Somit können wir o.B.d.A. $I\subset S'_+$ annehmen. Also ist $A=(S'/I)_0$ und $S=S'/I$ wird als $A$"~Algebra von $S_1$ endlich erzeugt.
-
-Umgekehrt ist jeder graduierte Ring $S$, der von $S_1$ als $S_0=A$"~Algebra endlich erzeugt ist, Quotient des Polynomrings und $\operatorname{Proj}(S)$ ist projektiv.\qed
-\end{enumerate}
-
-\paragraph{Definition 5.30.}\label{5.30} Sei $Y$ ein Schema.
-Der kanonische Morphismus $g:\mathbf{P}_Y^r=\mathbf{P}_\mathbb{Z}^r\times Y\to\mathbf{P}_\mathbb{Z}^r$ definiert die \textit{getwistete Garbe} $\mathcal{O}(1)$ auf $\mathbf{P}_Y^r$ durch:
-\[\mathcal{O}(1)=g^\ast\mathcal{O}(1) \]
-
-\paragraph{Bemerkung.} Ist $Y=\operatorname{Spec}(A)$ affin, so ist $\mathcal{O}(1)$ die bereits in \hyperref[5.22]{Definition 5.22} definierte Garbe auf $\mathbf{P}_A^r$.
-
-\paragraph{Definition 5.31.}\label{5.31} Sei $X$ ein Schema über $Y$.
-Eine invertierbare Garbe $\mathcal{L}$ auf $X$ heißt \textit{sehr ampel}\index{$\mathcal{O}_X$-Modul!sehr ampel} bzgl.
-$Y$, wenn es eine Immersion $i:X\hookrightarrow\mathbf{P}_Y^r$ für ein $r$ gibt, so dass $i^\ast(\mathcal{O}(1))\cong\mathcal{L}$.
-
-\paragraph{Satz 5.32.}\label{5.32} Sei $Y$ ein noethersches Schema und $X$ ein Schema über $Y$. Dann ist $X$ genau dann projektiv über $Y$, wenn:
-\begin{enumerate}[(i)]
-\item $X$ ist eigentlich über $Y$.
-\item Es gibt eine sehr ample Garbe auf $X$ bzgl. $Y$.
-\end{enumerate}
-
-\paragraph{Beweis.} Sei $X$ projektiv.
-Dann folgt (i) aus \hyperref[4.24]{Theorem~4.24} und es gibt eine abgeschlossene Immersion $i:X\to\mathbf{P}_Y^r$ für ein $r$, so dass $i^\ast\mathcal{O}(1)$ sehr ampel ist.
-
-Sei umgekehrt $i:X\hookrightarrow\mathbf{P}_Y^r$ eine Immersion und $\mathcal{L}\cong i^\ast(\mathcal{O}(1))$ eine sehr ample Garbe auf $X$ bzgl. $Y$. Betrachte das kartesische Quadrat:
-\[\begin{tikzcd}
-\mathbf{P}^r_Y = Y\times \mathbf{P}_\mathbb{Z}^r \ar[r]\ar[d]& Y\ar[d]\\
-\mathbf{P}_\mathbb{Z}^r\ar[r] & \mathbb{Z}
-\end{tikzcd} \]
-$\mathbf{P}_\mathbb{Z}^r\to\mathbb{Z}$ ist separiert, da projektiv, also ist auch der Basiswechsel $\mathbf{P}_Y^r\to Y$ separiert.
-Ferner ist $X\to Y$ eigentlich, nach \hyperref[4.22]{Korollar 4.22 (v)} ist auch $X\hookrightarrow \mathbf{P}_Y^r$ eigentlich, also insbesondere abgeschlossen.\qed
-
-\paragraph{Definition 5.33.}\label{5.33} Sei $X$ ein Schema und $\mathcal{F}$ ein $\mathcal{O}_X$"~Modul.
-$\mathcal{F}$ heißt \textit{von globalen Schnitten erzeugt}\index{$\mathcal{O}_X$-Modul!von globalen Schnitten erzeugt}, wenn es eine Familie von globalen Schnitten $s_i\in\Gamma(X,\mathcal{F}),\ i\in I$ gibt, so dass für alle $x\in X$ die Bilder der $s_i$ den Halm $\mathcal{F}_x$ als $\mathcal{O}_X$"~Modul erzeugen.
-Dies ist äquivalent zu: Es gibt einen surjektiven Garbenmorphismus $\bigoplus_{i\in I}\mathcal{O}_X\to \mathcal{F}$.
-
-\paragraph{Beispiel 5.34.}\label{5.34} Sei $X=\operatorname{Spec}(A)$ und $\mathcal{F}=\widetilde{M}$ für ein $A$"~Modul $M$.
-Dann wird $\mathcal{F}$ von globalen Schnitten erzeugt; jedes Erzeugendensystem von $M$ als $A$"~Modul liefern solche Schnitte.
-Die Surjektion $A^{(I)}\to M$ induziert surjektives $\mathcal{O}_X^{(I)}\to\mathcal{F}$.
-
-\paragraph{Beispiel 5.35.}\label{5.35} Sei $X=\operatorname{Proj}(S)$ mit einem graduierten Ring $S$, der von $S_1$ als $S_0$"~Algebra erzeugt wird.
-Dann liefern die Elemente aus $S_1$ globale Schnitte von $\mathcal{O}_X(1)$ und erzeugen diesen quasikohärenten Modul.
-
-\paragraph{Lemma 5.36.}\label{5.36} \begin{enumerate}[(i)]
-\item Abgeschlossene Immersionen sind endliche Morphismen.
-\item Sei $f:X\to Y$ ein endlicher Morphismues noetherscher Schemata und $\mathcal{F}$ ein kohärenter $\mathcal{O}_X$"~Modul. Dann ist $f_\ast\mathcal{F}$ kohärent.
-\end{enumerate}
-
-\paragraph{Beweis.}\begin{enumerate}[(i)]
-\item Sei $f:X\to Y$ eine abgeschlossene Immersion.
-Sei $V=\operatorname{Spec}(B)\subset_\text{o}Y$.
-Es ist $f^{-1}(V)=X\times_YV\to V$ als Basiswechsel von $f$ eine abgeschlossene Immersion.
-Somit ist $f^{-1}(V)\cong\operatorname{Spec}(B/I)$ für ein Ideal $I\subset B$ affin und ferner $B/I$ ein endlich erzeugter $B$"~Modul.
-\item Nach \hyperref[5.13]{Satz 5.13 (iii)} ist $f_\ast\mathcal{F}$ quasikohärent.
-Sei $V=\operatorname{Spec}(B)\subset_\text{o}Y$.
-Da $f$ endlich ist, ist $f^{-1}(V)=\operatorname{Spec}(A)$ affin und es gilt $\mathcal{F}|_{f^{-1}(V)}=\widetilde{N}$ für ein endlich erzeugter $A$"~Modul $N$.
-Nach \hyperref[5.3]{Satz 5.3 (iv)} gilt:
-\[f_\ast\mathcal{F}|_V = f_\ast\widetilde{N} = \widetilde{_BN} \]
-Da $f$ endlich ist, ist $A$ ein endlich erzeugter $B$"~Modul und somit auch $_BN$.\qed
-\end{enumerate}
-
-\paragraph{Satz 5.37.}\label{5.37} \textit{(Serre)} Sei $X$ ein projektives Schema über $\operatorname{Spec}(A)$ mit $A$ noethersch.
-Sei $\mathcal{O}_X(1)$ eine sehr ample Garbe auf $X$ und $\mathcal{F}$ ein kohärenter $\mathcal{O}_X$"~Modul.
-Dann existiert ein $n_0\in\mathbb{Z}$, so dass für alle $n\geq n_0$ der getwistete $\mathcal{O}_X$"~Modul $\mathcal{F}(n)$ von endlich vielen globalen Schnitten erzeugt wird.
-
-\paragraph{Beweis.} Sei $i:X\hookrightarrow \mathbf{P}_A^r$ eine abgeschlossene Immersion mit $\mathcal{O}_X(1)\cong i^\ast(\mathcal{O}(1))$.
-Nach \hyperref[5.36]{Lemma 5.36} ist $i_\ast\mathcal{F}$ kohärent auf $\mathbf{P}_A^r$ und nach \hyperref[5.23]{Satz 5.23 (iii)} gilt $(i_\ast\mathcal{F})(n)=i_\ast(\mathcal{F}(n))$.
-Wird nun $i_\ast\mathcal{F}(n)$ von endlich vielen globalen Schnitten erzeugt, so ist dies auch für $\mathcal{F}(n)$ der Fall, betrachte dafür:
-\[\Gamma(\mathbf{P}_A^r, i_\ast\mathcal{F}(n)) = \mathcal{F}(n)(i^{-1}(\mathbf{P}_A^r))=\Gamma(X,\mathcal{F}(n)) \]
-\[(i_\ast\mathcal{F}(n))_x = \begin{cases}
-\mathcal{F}(n)_x, & x\in i(X)\\
-0, &\text{sonst}
-\end{cases} \]
-Sei also o.B.d.A.
-$X=\mathbf{P}_A^r=\operatorname{Proj}A[X_0,\ldots,X_r]$.
-Es ist $X=\bigcup_{i=0}^r D_+(X_i)$.
-Für jedes $i$ ist $\mathcal{F}|_{D_+(X_i)}\cong\widetilde{M_i}$ für ein endlich erzeugter Modul $M_i$ über $B_i=A\big[\frac{X_0}{X_i},\ldots,\widehat{\frac{X_i}{X_i}},\ldots,\frac{X_r}{X_i}\big]$.
-Sei $(s_{ij})_j$ ein Erzeugendensystem von $M_i$.
-Wegen \hyperref[5.26]{Lemma 5.26} existiert ein $n_0>0$, so dass für alle $n\geq n_0$ die Schnitte $X_i^ns_{ij}$ sich zu globalen Schnitten $t_{ij}$ von $\mathcal{F}(n)$ liften lassen.
-Wir können $n_0$ unabhängig von $i$ und $j$ wählen.
-Sei $\mathcal{F}(n)|_{D_+(X_i)}\cong\widetilde{M_i'}$ für ein $B_i$"~Modul $\widetilde{M_i'}$.
-Die Abbildungen $X_i^n:\mathcal{F}\to\mathcal{F}(n)$ induzieren Isomorphismen $M_i\to M_i'$.
-Da $\{X_i^ns_{ij}\mid j\}$ ganz $M_i'$ erzeugen, erzeugen $t_{ij}\in\Gamma(X,\mathcal{F}(n))$ als globale Schnitte ganz $\mathcal{F}(n)$.\qed
-
-\paragraph{Korollar 5.38.}\label{5.38} Sei $X$ ein projektives Schema über $\operatorname{Spec}(A)$ mit $A$ noethersch.
-Dann gibt es für jeden kohärenten $\mathcal{O}_X$"~Modul $\mathcal{F}$ eine Surjektion $\mathcal{O}_X(n)^N\to\mathcal{F}$ mit $n,N\in\mathbb{Z}$.
-
-\paragraph{Beweis.} Nach \hyperref[5.37]{Satz 5.37} gibt es eine Surjektion $\mathcal{O}_X^N\to\mathcal{F}(n)$. Tensorieren mit $\mathcal{O}_X(-n)$ gibt die Behauptung.\qed
-
-\paragraph{Satz 5.39.}\label{5.39} Sei $k$ ein Körper, $A$ eine endlich erzeugte $k$"~Algebra und $X$ projektives Schema über $\operatorname{Spec}(A)$.
-Ferner sei $\mathcal{F}$ ein kohärenter $\mathcal{O}_X$"~Modul.
-Dann ist $\Gamma(X,\mathcal{F})$ ein endlich erzeugter $A$"~Modul.
-
-\paragraph{Beweis.} Wir werden diesen Satz später kohomologisch beweisen.
-
-\paragraph{Korollar 5.40.}\label{5.40} Sei $f:X\to Y$ ein projektiver Morphismus von Schemata von endlichem Typ über einem Körper $k$.
-Ist $\mathcal{F}$ kohärent auf $X$, so ist auch $f_\ast\mathcal{F}$ kohärent auf $Y$.
-Insbesondere ist für $A=k$ der Modul $\Gamma(X,\mathcal{F})$ ein endlich dimensionierter $k$"~Vektorraum.
-
-\paragraph{Beweis.} Sei o.B.d.A.
-$Y=\operatorname{Spec}(A)$ affin, wobei $A$ eine endlich erzeugte $k$"~Algebra ist.
-Da $f$ projektiv ist, ist $f$ eigentlich und somit separiert und von endlichem Typ, also quasikompakt.
-Wegen \hyperref[5.13]{Satz 5.13 (iii)} ist $f_\ast\mathcal{F}$ quasikohärent.
-Es gilt:
-\[f_\ast\mathcal{F} = \widetilde{\Gamma(Y,f_\ast\mathcal{F})}=\widetilde{\Gamma(X,\mathcal{F})} \]
-Aber $\Gamma(X,\mathcal{F})$ ist endlich erzeugter $A$"~Modul nach \hyperref[5.39]{Satz 5.39}.\qed
-
-\section{Divisoren}
-
-\paragraph{Definition 6.1.}\label{6.1} \begin{enumerate}[(i)]
-\item Ein noetherscher lokaler Ring $(R,\mathfrak{m})$ heißt \textit{regulär}\index{regulär}, falls für $k=R/\mathfrak{m}$ gilt:
-\[\dim(R)=\dim_k (\mathfrak{m}/\mathfrak{m}^2)\]
-Ist $R$ ein noetherscher lokaler Ring, so gilt stets $\dim(R)\leq\dim_k(\mathfrak{m}/\mathfrak{m}^2)$.
-\item Ein Schema $X$ heißt \textit{regulär in Kodimension $1$}\index{regulär}, falls jeder Halm $\mathcal{O}_{X,x}$ von $X$ mit $\dim(\mathcal{O}_{X,x})=1$ regulär ist.
-\end{enumerate}
-
-\paragraph{Definition 6.3.}\label{6.3}
-Ein Ring heißt \textit{normal}\index{Ring!normal}, wenn er ganzabgeschlossen und nullteilerfrei ist.
-Ein Schema heißt \textit{normal}\index{Schema!normal}, wenn seine Halme normal sind.
-
-\paragraph{Theorem 6.4.}\label{6.4} Sei $R$ ein noetherscher, normaler Ring und $\mathfrak{p}$ ein Primideal der Hö\-he $1$.
-Dann ist $R_\mathfrak{p}$ regulär.
-Genauer: Sei $(R,\mathfrak{m})$ ein noetherscher lokaler Ring der Dimension $1$.
-Dann sind folgende Aussagen äquivalent:
-\begin{enumerate}[(i)]
-\item $R$ ist ein diskreter Bewertungsring.
-\item $R$ ist ganzabgeschlossen.
-\item $R$ ist regulär.
-\item $\mathfrak{m}$ ist ein Hauptideal.
-\end{enumerate}
-
-\paragraph{Beweis.} Siehe z.B. Matsumura: \glqq Commutative Algebra\grqq, Theorem 3.9 und Ati\-yah-Mac\-Do\-nald: \glqq Introduction to Commutative Algebra\grqq, Proposition 9.2.\qed
-
-\paragraph{Definition.} Ein Schema habe die Eigenschaft $(\star)$, wenn es noethersch, separiert, integer und regulär in Kodimension $1$ ist.
-
-\paragraph{Definition 6.5.}\label{6.5} Sei $X$ ein Schema mit $(\star)$. Dann gilt:
-\begin{enumerate}[(i)]
-\item Ein \textit{Primdivisor}\index{Primdivisor} auf $X$ ist ein abgeschlossenes, integres Unterschema der Kodimension $1$.
-\item Ein \textit{Weil-Divisor}\index{Divisor!Weil} ist ein Element der freien abelschen Gruppe $\operatorname{Div}(X)$, die von den Primdivisoren erzeugt wird.
-Wir schreiben ein Divisor als $D=\sum_i n_iY_i$ mit Primdivisoren $Y_i$ und $n_i\in\mathbb{Z}$ mit $n_i=0$ für fast alle $i$.
-Ein solcher Divisor heißt \textit{effektiv}\index{Divisor!effektiv}, falls alle $n_i\geq 0$ sind.
-\item Sei $Y$ ein Primdivisor auf $X$ und $\eta$ ein generischer Punkt in $Y$.
-Dann ist $\mathcal{O}_{X,\eta}$ nach \hyperref[6.4]{Theorem 6.4} ein diskreter Bewertungsring mit Quotientenkörper $K$, der \textit{Funktionenkörper}\index{Funktionenkörper} von $X$.
-Wir bezeichnen die zugehörige diskrete Bewertung mit  $v_Y:K^\times\to\mathbb{Z}$.
-
-Sei $f\in K^\times$.
-Ist $v_Y(f)>0$, so sagen wir, dass $f$ eine \textit{Nullstelle}\index{Nullstelle} entlang $Y$ von der Ordnung $v_Y(f)$ hat.
-Ist $v_Y(f)<0$, so sagen wir, dass $f$ ein \textit{Pol}\index{Pol} entlang $Y$ von der Ordnung $-v_Y(f)$ besitzt.
-\end{enumerate}
-
-\paragraph{Lemma 6.6.}\label{6.6} Sei $X$ ein Schema mit $(\star)$ und $f\in K^\times$. Dann ist $v_Y(f)=0$ für fast alle Primdivisoren $Y$.
-
-\paragraph{Beweis.} Sei $\varnothing \neq U = \operatorname{Spec}(A)\subset_\text{o}X$ affin mit $f|_U$ \textit{regulär}, d.h.
-$f|_U\in\mathcal{O}_X(U)$.
-Sei $Z=X\setminus U\subsetneq X$ abgeschlossen.
-Da $X$ noethersch und irreduzibel ist, ist $Z$ noethersch mit $\operatorname{codim}(Z,X)\geq 1$ und besitzt nur endlich viele irreduzible Komponenten.
-Daher enthält $Z$ höchstens endlich viele Primdivisoren von $X$, alle anderen treffen $U$.
-Es genügt also zu zeigen, dass es nur endlich viele Primdivisoren $Y$ in $U$ gibt mit $v_Y(f)\neq 0$, d.h.
-$v_Y(f)>0$.
-Es gilt mit $Y=\overline{\{ \eta\}}$:
-\[v_Y(f)>0\iff f\not\in\mathcal{O}_{U,\eta}^\times=A_\eta^\times\iff f\in\eta\iff Y=V(\eta)\subset V(f)\subset U \]
-Da $f\neq 0$, ist $V(f)\subsetneq U$ eine echte abgeschlossene Teilmenge, und enthält daher nur endlich viele irreduzible Komponenten, also Primdivisoren in $U$.\qed
-
-\paragraph{Definition 6.7.}\label{6.7} Sei $X$ ein Schema mit $(\star)$ und $f\in K^\times$. Der Divisor $\operatorname{div}(f)$ von $f$ ist definiert als:
-\[\operatorname{div}(f)=\sum_Y v_Y(f)Y \]
-wobei $Y$ über die Primdivisoren in $X$ läuft.
-Diese Summe ist nach \hyperref[6.6]{Lemma 6.6} endlich und somit wohldefiniert.
-Jeder Divisor der Form $\operatorname{div}(f)$ heißt \textit{Hauptdivisor} oder \textit{prinzipal}.\index{Hauptdivisor}\index{Divisor!prinzipal}.
-
-\paragraph{Bemerkung 6.8.}\label{6.8} Sei $f,g\in K^\times$. Dann gilt:
-\[\operatorname{div}\big(\tfrac{f}{g}\big)=\operatorname{div}(f)-\operatorname{div}(g) \]
-Somit ist $K^\times\to\operatorname{Div}(X),\ f\mapsto\operatorname{div}(f)$ ein Gruppenhomomorphismus, dessen Bild gerade die Gruppe der Hauptdivisoren in $X$ sind.
-
-\paragraph{Definition 6.9.}\label{6.9} Sei $X$ ein Schema mit $(\star)$.
-Zwei Divisoren $D,D'$ heißen \textit{linear äquivalent}\index{linear äquivalent} $D\sim D'$, wenn $D-D'$ ein Hauptdivisor ist.
-Die Gruppe der zugehörigen Äquivalenzklassen $\operatorname{Cl}(X)$ heißt \textit{Divisorenklassengruppe}\index{Divisorenklassengruppe}.
-Wir haben eine exakte Folge:
-\[K^\times\to\operatorname{Div}(X)\to\operatorname{Cl}(X)\to 0 \]
-
-\paragraph{Satz 6.10.}\label{6.10} Sei $A$ ein noetherscher, nullteilerfreier Ring. Dann gilt:
-\[A\text{ ist faktoriell}\iff\operatorname{Cl}(\operatorname{Spec}(A))=0 \]
-
-\paragraph{Beweis.} Siehe z.B. Bourbaki: Algèbre Commutative, Chapitre 7 §3 Proposition 2.\qed
-
-\paragraph{Beispiel 6.11.}\label{6.11} \begin{enumerate}
-\item Sei $X=\mathbf{A}_k^n$ für ein Körper $k$. Dann gilt $\operatorname{Cl}(X)=0$, da $k[X_1,\ldots,X_n]$ faktoriell ist.
-\item Sei $A$ ein Dedekindring. Dann ist $\operatorname{Cl}(\operatorname{Spec}(A))$ gerade die Idealklassengruppe.
-\end{enumerate}
-
-\paragraph{Satz 6.12.}\label{6.12} Sei $k$ ein Körper.
-\begin{enumerate}[(i)]
-\item Sei $X=\mathbf{A}_k^n$ und $Y\subset X$ ein abgeschlossenes Unterschema. Dann ist $Y$ genau dann ein Primdivisor, wenn:
-\[Y=V(f)\quad\text{für ein irreduzibles, nicht-konstantes }f\in k[X_1,\ldots,X_n]  \]
-\item Sei $X=\mathbf{P}_k^n$ und $Y\subset X$ ein abgeschlossenes Unterschema. Dann ist $Y$ genau dann ein Primdivisor, wenn:
-\[Y=V_+(f)\quad\text{für ein homogenes, irreduzibles }f\in k[X_0,\ldots,X_n],\ \deg(f)=r>0 \]
-\end{enumerate}
-
-\paragraph{Definition 6.13.}\label{6.13} Sei $X=\mathbf{P}^n_k$.
-Jeder Primdivisor $Y$ in $X$ hat die Form $Y=V_+(f_Y)$.
-Betrachte die Abbildung $Y\mapsto\deg(f_Y)$.
-Diese induziert ein Gruppenhomomorphismus $\operatorname{Div}(X)\to\mathbb{Z}$.
-Wir zeigen, dass dieser über $\operatorname{Cl}(X)$ faktorisiert.
-Sei $f\in K^\times$.
-Dann gilt:
-\[\deg\operatorname{div}(f)=\sum_Y v_Y(f) \deg(Y)=\sum_Y v_Y(f) \deg(f_Y) \]
-Sei $f=\frac{g}{h}$ mit homogenen $g,h$ vom selben Grad $d$.
-Sei $g=g_1^{n_1}\cdots g_r^{n_r}$ eine Zerlegung in irreduzible Elemente $g_i$ vom Grad $d_i$.
-Dann sind $Y_i=\operatorname{div}(g_i)$ nach \hyperref[6.12]{Satz 6.12 (ii)} Primdivisoren.
-Es gilt:
-\[\deg\operatorname{div}(g)=\sum_i n_i\deg(g_i) = \sum_i n_id_i = d \]
-Analog ist $\deg\operatorname{div}(h)=d$. Somit ist $\deg\operatorname{div}(f)=\deg\operatorname{div}(g)-\deg\operatorname{div}(h)=0$.\qed
-
-\paragraph{Satz 6.15.}\label{6.15} Sei $X=\mathbf{P}_k^n$. Dann gilt:
-\begin{enumerate}[(i)]
-\item Ist $D$ ein Divisor auf $X$, so ist $D\sim\deg(D)\cdot V_+(T_0)$
-\item $\deg:\operatorname{Cl}(X)\to\mathbb{Z}$ ist ein Isomorphismus.
-\end{enumerate}
-
-\paragraph{Beweis.} \begin{enumerate}[(i)]
-\item Sei $D=\sum_Y n_YY$ ein Divisor mit Primdivisoren $Y$.
-Nach \hyperref[6.12]{Satz 6.12 (ii)} ist $Y=V_+(f_Y)$ mit irreduziblen, homogenen Polynomen $f_Y$ vom Grad $r_Y$.
-Wir schreiben $f_Y=T_0^{r_Y} g_Y$, wobei $g_Y$ ein Polynom in $\frac{T_1}{T_0},\ldots,\frac{T_n}{T_0}$ ist.
-Die $g_Y$ sind rationale Funktionen auf $\mathbf{P}_k^n$ und es gilt:
-\[\operatorname{div}(g_Y) = V_+(f_Y)-r_YV_+(T_0) \implies V_+(f_Y)\sim r_YV_+(T_0) \]
-Also gilt $D=\sum_Yn_Y V_+(f_Y)\sim \big(\sum n_Y r_Y\big) V_+(T_0) = \deg(D)\cdot V_+(T_0)$.
-\item folgt aus (i) und wegen $\deg V_+(T_0)=1$.\qed
-\end{enumerate}
-
-\paragraph{Satz 6.16.}\label{6.16} Sei $X$ ein Schema mit $(\star)$, $Z\subsetneq X$ eine abgeschlossene Teilmenge und $U=X\setminus Z$. Dann gilt:
-\begin{enumerate}[(i)]
-\item Es gibt einen surjektiven Homomorphismus $\operatorname{Cl}(X)\to\operatorname{Cl}(U)$, der durch $\sum n_iY_i\mapsto \sum n_i(Y_i\cap U)$ gegeben ist, wobei wir die leeren $Y_i\cap U$ ignorieren.
-\item Ist $\operatorname{codim}(Z,X)\geq 2$, dann ist die obige Abbildung ein Isomorphismus.
-\item Ist $Z$ irreduzibel und $\operatorname{codim}(Z,X)=1$, so gibt es eine exakte Folge:
-\[\mathbb{Z}\to\operatorname{Cl}(X)\to\operatorname{Cl}(U)\to 0 \]
-wobei die erste Abbildung durch $1\mapsto Z$ gegeben ist.
-\end{enumerate}
-
-\paragraph{Beweis.}\begin{enumerate}[(i)]
-\item Ist $Y$ ein Primdivisor auf $X$, so ist $Y\cap U$ leer oder ein Primdivisor auf $U$.
-Sei $f\in K^\times$ mit $\operatorname{div}(f)=\sum n_iY_i$.
-Fassen wir $f$ als rationale Funktion auf $U$ auf, so erhalten wir $\operatorname{div}(f|_U)=\sum n_i(Y_i\cap U)$.
-Somit ist die Abbildung $\operatorname{Cl}(X)\to\operatorname{Cl}(U)$ wohldefiniert.
-
-Die Surjektivität sieht man wie folgt: Sei $Y$ ein Primdivisor auf $U$ und $\overline{Y}$ der Abschluss von $Y$ in $X$. Dann ist $\overline{Y}$ ein Primdivisor auf $X$ mit $\overline{Y}\cap U=Y$.
-\item $\operatorname{Div}(X)$ bzw. $\operatorname{Cl}(X)$ hängen nur von Teilmengen der Kodimension $1$ ab.
-\item Es ist $\ker(\operatorname{Cl}(X)\to\operatorname{Cl}(U)) = \{ [D]\in\operatorname{Cl}(X)\mid \operatorname{supp}(D)\subset Z\}=\langle[Z]\rangle$, da $Z$ irreduzibel ist.\qed
-\end{enumerate}
-
-\paragraph{Beispiel 6.17.}\label{6.17} Sei $Y$ eine irreduzible Kurve vom Grad $d$ in $\mathbf{P}_k^2$. Wir haben ein kommutatives Diagramm:
-\[\begin{tikzcd}
-\mathbb{Z} \rar\dar["\cong"'] & \operatorname{Cl}(\mathbf{P}_k^2) \rar\dar["\cong"', "\deg"]& \operatorname{Cl}(\mathbf{P}_k^2\setminus Y) \rar\dar["\cong"]& 0\\
-d\mathbb{Z} \rar[hook] & \mathbb{Z}\rar & \mathbb{Z}/d\mathbb{Z}\rar & 0
-\end{tikzcd}\]
-
-\paragraph{Definition 6.18.}\label{6.18} Sei $k$ ein algebraisch abgeschlossener Körper.
-Eine \textit{Kurve}\index{Kurve} über $k$ ist ein integres, separiertes Schema $X$ von endlichem Typ über $k$ der Dimension $1$.
-\begin{itemize}
-\item $X$ heißt \textit{vollständig}\index{Kurve!vollständig}, wenn $X\to k$ eigentlich ist.
-\item $X$ heißt \textit{nicht-singulär}\index{nicht-singulär}, falls alle lokalen Ringe von $X$ regulär sind.
-\end{itemize}
-
-\paragraph{Satz 6.19.}\label{6.19} Sei $X$ eine vollständige, nicht-singuläre Kurve über $k$ und $Y$ eine beliebige Kurve über $k$ mit einem Morphismus $f:X\to Y$. Dann gilt:
-\[f(X)=\operatorname{Pt} \quad\text{oder}\quad f(X)=Y \]
-Im zweiten Fall gilt:
-\begin{enumerate}[(i)]
-\item Die Funktionenkörpererweiterung $K(X)/K(Y)$ ist endlich.
-\item $f$ ist endlich.
-\item $Y$ ist vollständig.
-\end{enumerate}
-
-\paragraph{Lemma 6.20.}\label{6.20} Sei $f:X\to Y$ surjektiv, $g:Y\to Z$ separiert und von endlichem Typ, und $g\circ f:X\to Z$ eigentlich. Dann ist $g$ eigentlich.
-
-\paragraph{Beweis von \hyperref[6.19]{6.19}.} Da $X$ eigentlich über $k$ ist, ist $f(X)\subset Y$ ein abgeschlossenes Unterschema.
-Da $X$ irreduzibel ist, ist $f(X)$ auch irreduzibel.
-Nun ist $\dim(Y)=1$, folgt entweder $f(X)=\operatorname{Pt}$ oder $f(X)=Y$.
-Betrachte das folgende kommutative Diagramm:
-\[\begin{tikzcd}
-X \rar["f"] & f(X)\rar[hook]\ar[dr] & Y\dar\\
- & & k
-\end{tikzcd} \]
-Es ist $f(X)\hookrightarrow Y$ eine abgeschlossene Immersion, also separiert nach \hyperref[4.17]{Satz 4.17 (i)}.
-Somit ist auch $f(X)/ k$ separiert.
-Nach \hyperref[6.20]{Lemma 6.20} ist $f(X)/k$ eigentlich.
-
-Sei nun $f(X)=Y$, also ist $Y$ vollständig.
-Sei $Y=\overline{\{\eta\}}$ und $X=\overline{\{\xi\}}$.
-Da $f:X\to Y$ dominant ist, folgt $f(\xi)=\eta$ und wir erhalten eine Inklusion $K(Y)=\mathcal{O}_{Y,\eta}\to\mathcal{O}_{X,\xi}=K(X)$.
-Da beide Körper endlich erzeugt über $k$ vom Transzendenzgrad $\dim(X)=1$ sind, ist $K(X)/K(Y)$ endlich.
-
-Es bleibt zu zeigen, dass $f$ endlich ist.
-Sei $V=\operatorname{Spec}(B)\subset_\text{o}Y$ affin.
-Dann ist $B=\mathcal{O}_Y(V)\subset K(Y)\subset K(X)$.
-Es ist zu zeigen, dass $f^{-1}(V)=\operatorname{Spec}(A)$ affin ist, wobei $A$ ein endlich erzeugter $B$"~Modul ist.
-In der Tat folgt aus der Normalität von $X$ stets $f^{-1}(V)=\operatorname{Spec}(A)$, wobei $A$ der Ganzabschluss von $B$ in $K(X)$ ist.
-Alles folgt nun aus dem folgenden \hyperref[6.21]{Theorem 6.21}.\qed
-
-\paragraph{Theorem 6.21.}\label{6.21} Sei $B$ ein Integritätsring und eine endlich erzeugte Algebra über einem Körper $k$.
-Setze $K=\operatorname{Quot}(B)$ und sei $L/K$ eine endliche Erweiterung.
-Dann ist der Ganzabschluss $A$ von $B$ in $L$ ein endlich erzeugter $B$"~Modul.
-
-\paragraph{Beweis.} Siehe z.B. Zariski-Samuel: Commutative Algebra I, Chapter V, Theorem 9.\qed
-
-\paragraph{Beweis von \hyperref[6.20]{6.20}.} Es reicht zu zeigen, dass $g$ universell abgeschlossen ist. Sei $h:Y'\to Y$ ein beliebiger Morphismus und betrachte den Basiswechsel:
-\[\begin{tikzcd}
-X\times_Y Y' \ar[r, "f'"]\ar[d, "h'"'] & Y' \ar[d, "h"]\\
-X\ar[r, "f"'] & Y
-\end{tikzcd} \]
-Wir zeigen nun, dass $f'$ surjektiv ist.
-Sei $y'\in Y'$ und sei $x\in X$ mit $f(x)=h(y')=:y$.
-Behauptung: Es gibt ein $z\in X\times_YY'$ mit $h'(z)=x$ und $f'(z)=y'$.
-Seien $i_x:\{x\}\hookrightarrow X,\ i_{y'}:\{y'\}\hookrightarrow Y'$ die Inklusionen.
-Diese liefern einen $Y$"~Morphismus:
-\[\delta : \operatorname{Spec}\kappa(x)\times_{\kappa(y)}\operatorname{Spec}\kappa(y') = \operatorname{Spec}(\kappa(x)\otimes_{\kappa(y)}\kappa(y'))\hookrightarrow X\times_YY' \]
-Da $\kappa(x)\otimes_{\kappa(y)}\kappa(y')\neq 0$, gilt für jedes $z\in\operatorname{im}(\delta)$ stets $h'(z)=x$ und $f'(z)=y'$. Dies zeigt die Surjektivität von $f'$.
-
-Also ist die Surjektivität von $f$ stabil unter Basiswechsel.
-Da alle anderen Eigenschaften in den Voraussetzungen ebenfalls stabil unter Basiswechsel sind (siehe \hyperref[4.17]{4.17 (iii)} und \hyperref[4.22]{4.22}), reicht es zu zeigen, dass $g$ abgeschlossen ist.
-Sei $Y'\subset Y$ eine abgeschlossene Teilmengen.
-Dann ist wegen der Surjektivität von $f$:
-\[g(Y')=(g\circ f)\circ f^{-1}(Y') \]
-$g(Y')$ ist abgeschlossen, da $g\circ f$ abgeschlossen ist.\qed
-
-\paragraph{Definition 6.22.}\label{6.22} Sei $f:X\to Y$ ein dominanter Morphismus von Kurven.
-Dann heißt $\deg(f)=[K(X):K(Y)]$ der \textit{Grad} von $f$\index{Grad}.
-Sei $X$ eine nicht-singuläre Kurve, erfüllt also insbesondere $(\star)$.
-Ein Primdivisor von $X$ ist genau ein abgeschlossener Punkt.
-Ein Divisor $D$ ist somit von der folgenden Form:
-\[D=\sum_i n_iP_i,\quad n_i\in\mathbb{Z},\ P_i\text{ abgeschlossene Punkte} \]
-Der \textit{Grad}\index{Grad} von $D$ ist definiert als $\deg(D)=\sum n_i$.
-
-\paragraph{Definition 6.23.}\label{6.23} Sei $f:X\to Y$ ein endlicher Morphismus nicht-singulärer Kurven.
-Wir definieren ein Homomorphismus $f^\ast:\operatorname{Div}(Y)\to\operatorname{Div}(X)$ wie folgt:
-
-Sei $Q\in Y$ ein abgeschlossener Punkt und wähle ein $t\in\mathcal{O}_{Y,Q}\subset K(Y)$ mit $v_Q(t)=1$, wobei $v_Q$ die diskrete Bewertung zu $\mathcal{O}_{Y,Q}$ ist.
-$t$ heißt \textit{lokaler Parameter}\index{lokaler Parameter} an der Stelle $Q$.
-Wir setzen:
-\[f^\ast(Q) = \sum_{P\in f^{-1}(Q)}v_P(t)P \]
-Da $f$ endlich ist, ist obige Summe nach \hyperref[6.24]{Lemma 6.24} endlich.
-
-\paragraph{Lemma 6.24.}\label{6.24} \begin{enumerate}[(i)]
-\item Die Eigenschaft endlich zu sein ist stabil unter Basiswechsel.
-\item Ist $f:X\to Y$ ein endlicher Morphismus und $y\in Y$, so ist $f^{-1}(y)$ endlich.
-\end{enumerate}
-
-\paragraph{Beweis.} \begin{enumerate}[(i)]
-\item Sei $f:X\to Y$ ein endlicher Morphismus und $g:Z\to Y$ beliebig. Betrachte den folgenden Basiswechsel:
-\[\begin{tikzcd}
-X\times_YZ \dar\rar["f_Z"] & Z\dar["g"]\\
-X \rar["f"'] & Y
-\end{tikzcd} \]
-Sei $V=\operatorname{Spec}(B)\subset_\text{o}Y$ affin.
-Dann existiert ein affines $W=\operatorname{Spec}(C)\subset_\text{o}Z$ mit $g(W)\subset V$.
-Es ist $f^{-1}(V)=\operatorname{Spec}(A)$ affin, wobei $A$ ein endlich erzeugter $B$"~Modul ist.
-Es gilt:
-\begin{align*}
-f_Z^{-1}(W)= (X\times_YZ)\times_ZW &\cong X\times_YW \\
-&\cong (X\times_YV)\times_VW\\
-&\cong f^{-1}(V)\times_VW=\operatorname{Spec}(A\otimes_BC)
-\end{align*}
-und $A\otimes_BC$ ist ein endlich erzeugter $C$"~Modul.
-\item Sei $f^{-1}(y)\neq\varnothing$. Betrachte den Morphismus:
-\[f^{-1}(y)= X\times_Y\operatorname{Spec}\kappa(y)\to\operatorname{Spec}\kappa(y) \]
-Dieser ist endlich nach (i).
-Also ist $f^{-1}(y)=\operatorname{Spec}(A)$ affin, wobei $A$ ein endlich dimensionierter $\kappa(y)$"~Vektorraum ist.
-Somit ist $\dim f^{-1}(y)=\dim\kappa(y)= 0$, als topologischer Raum ist $f^{-1}(y)$ also eine endliche Menge an Punkten.\qed
-\end{enumerate}
-
-\paragraph{Bemerkung.} \begin{enumerate}[(i)]
-\item In \hyperref[6.23]{Definition 6.23} ist $f^\ast Q$ unabhängig von der Wahl des lokalen Parameters, da ein lokaler Parameter eindeutig bis auf eine Einheit in $\mathcal{O}_{Y,Q}$ bestimmt ist.
-\item $f^\ast$ \textit{respektiert lineare Äquivalenz}, d.h. für $h\in K(Y)^\times,\ \operatorname{div}(h)=\sum_Qv_Q(h)Q$ gilt:
-\[f^\ast(\operatorname{div}(h))=\sum_Qv_Q(h)\sum_{P\in f^{-1}(Q)}v_P(t)P=\sum_Pv_P(h)P=\operatorname{div}(h) \]
-da $v_P(h)=v_P(t)\cdot v_Q(h)$ ist. Somit induziert $f^\ast$ eine Abbildung $\operatorname{Cl}(Y)\to\operatorname{Cl}(X)$.\qed
-\end{enumerate}
-
-\paragraph{Satz 6.25.}\label{6.25} Sei $f:X\to Y$ ein endlicher Morphismus nicht-singulärer Kurven. Dann:
-\[\deg(f^\ast D)=\deg(f)\cdot \deg(D) \]
-
-\paragraph{Korollar 6.26.}\label{6.26} Hauptdivisoren auf einer vollständigen, nicht-singulären Kurve haben Grad $0$. Wir erhalten eine surjektive Gradabbildung:
-\[\deg:\operatorname{Cl}(X)\to\mathbb{Z} \]
-
-\paragraph{Lemma 6.27.}\label{6.27} Sei $X$ eine normale Kurve über $k$, $U\subset X$ offen, $\varphi:X\to\mathbf{P}^n$ eine rationale Abbildung.
-Dann lässt sich $\varphi$ zu einem einem Morphismus auf $X$ fortsetzen.
-
-\paragraph{Definition 6.28.}\label{6.28} Sei $(X,\mathcal{O})$ ein Schema.
-\begin{enumerate}[(i)]
-\item Sei $U=\operatorname{Spec}(A)\subset_\text{o}X$ affin und $S$ die Menge aller Nicht-Nullteiler in $A$.
-Dann heißt $K(U)=A_S$ \textit{totaler Quotientenring} von $A$.\index{totaler Quotientenring}
-\item Sei $\mathcal{K}$ die Ringgarbe, die zur Prägarbe $U\mapsto\varprojlim_{V\subset U\text{ affin}} K(V)$ assoziiert ist.
-$\mathcal{K}$ heißt \textit{Garbe der totalen Quotientenringe} von $\mathcal{O}$.
-\item $\mathcal{K}^\times:U\mapsto \mathcal{K}(U)^\times$ sei die Garbe von multiplikativen Gruppe der invertierbaren Elemente in $\mathcal{K}$.
-$\mathcal{O}^\times$ sei die Garbe der invertierbaren Elemente in $\mathcal{O}$.
-\end{enumerate}
-
-\paragraph{Definition 6.29.}\label{6.29} Sei $(X,\mathcal{O})$ ein Schema.
-Ein Element aus $\Gamma(X,\mathcal{K}^\times/\mathcal{O}^\times)$ heißt \textit{Cartier-Divisor}\index{Divisor!Cartier}.
-Ein Cartier-Divisor kann also durch ein System $(U_i,f_i)_i$ beschrieben werden, wobei $(U_i)_i$ eine offene Überdeckung von $X$ ist, und $f_i\in\Gamma(U_i,\mathcal{K}^\times)$, so dass für alle $i,j$ stets $\frac{f_i}{f_j}\in\Gamma(U_i\cap U_j,\mathcal{O}^\times)$ gilt.
-
-Ein Cartier-Divisor heißt \textit{prinzipal}\index{Divisor!prinzipal} oder \textit{Hauptdivisor}\index{Hauptdivisor}, wenn er im Bild der kanonischen Abbildung $\Gamma(X,\mathcal{K}^\times)\to\Gamma(X,\mathcal{K}^\times/\mathcal{O}^\times)$ ist.
-Zwei Cartier-Divisoren heißen \textit{linear äquivalent}\index{linear äquivalent}, falls ihre Differenz prinzipal ist.
-
-\paragraph{Satz 6.30.}\label{6.30} Sei $(X,\mathcal{O})$ ein integres, separiertes, noethersches Schema.
-Ferner sei $X$ \textit{lokal faktoriell}\index{lokal faktoriell}, d.h.
-alle lokalen Ringe sind faktoriell.
-Dann ist die Gruppe $\operatorname{Div}(X)$ von Weil-Divisoren isomorph zur Gruppe der Cartier-Divisoren $\Gamma(X,\mathcal{K}^\times/\mathcal{O}^\times)$.
-Prinzipale Weil-Divisoren entsprechen prinzipale Cartier-Divisoren.
-
-\paragraph{Beweis.} $X$ ist normal, da faktorielle Ringe insbesondere ganzabgeschlossen sind.
-Nach \hyperref[6.4]{Theorem 6.4} erfüllt $X$ $(\star)$.
-Also können wir von Weil-Divisoren sprechen.
-Da $X$ integer ist, ist $K(U)=\operatorname{Quot}(A)= K$ der Funktionenkörper von $X$ für alle $U=\operatorname{Spec}(A)\subset_\text{o}X$.
-Somit ist $\mathcal{K}$ konstante Garbe.
-
-Sei $(U_i,f_i)_i$ ein Cartier-Divisor mit einer offenen Überdeckung $(U_i)_i$ von $X$ und $f_i\in\Gamma(U_i,\mathcal{K}^\times)=K^\times$.
-Wir ordnen diesen Cartier-Divisor den folgenden Weil-Divisor zu:
-\[D=\sum_Y v_Y(f_{i_Y})Y \]
-wobei $Y$ die Primdivisoren von $X$ durchläuft und $i_Y$ ein Index mit $U_i\cap Y\neq\varnothing$. Die Summe ist endlich, da $X$ noethersch ist. $D$ ist unabhängig von Wahl der Indizes $i_Y$:
-
-Seien $i,j$ mit $U_i\cap Y\neq\varnothing$ und $U_j\cap Y\neq\varnothing$.
-Dann ist $\frac{f_i}{f_j}$ auf $U_i\cap U_j$ invertierbar, d.h.
-$\frac{f_i}{f_j}\in\Gamma(U_i\cap U_j,\mathcal{O}^\times)$.
-Es folgt $v_Y\big(\frac{f_i}{f_j}\big)=0$, also $v_Y(f_i)=v_Y(f_j)$.
-
-Sei nun umgekehrt $D=\sum_Yn_YY$ ein Weil-Divisor auf $X$ und $x\in X$. Dann induziert $D$ ein Weil-Divisor $D_x$ auf $\operatorname{Spec}(\mathcal{O}_x)$, nämlich:
-\[D_x=\sum_Y n_Y(Y\cap\operatorname{Spec}(\mathcal{O}_x)) \]
-Da $\mathcal{O}_x$ als faktoriell vorausgesetzt wird, ist nach \hyperref[6.10]{Satz 6.10} $\operatorname{Cl}(\operatorname{Spec}\mathcal{O}_x)=0$ und $D_x$ ein Hauptdivisor, d.h.
-$D_x=(f_x)$ für ein $f_x\in K^\times$.
-Fassen wir $(f_x)$ als Weil-Divisor in $X$ auf, so sehen wir, dass sich $(f_x)$ und $D$ nur bei Primdivisoren, die nicht durch $x$ gehen, unterscheiden.
-Davon gibt es nur endlich viele, deren Koeffizienten nicht verschwinden.
-Daher existiert eine offene Umgebung $U_x$ von $x$ mit $(f_x)|_{U_x}=D|_{U_x}$.
-Das System $(U_x,f_x)_{x\in X}$ liefert einen Cartier-Divisor.
-
-Geben $f,f'$ denselben Weil-Divisor auf $U\subset_\text{o}X$ offen, so ist $\frac{f}{f'}\in\Gamma(U,\mathcal{O}^\times)$, da $X$ normal ist.
-Daher ist die Konstruktion wohldefiniert.
-Die obigen Konstruktionen sind invers zueinander.\qed
-
-\paragraph{Satz 6.31.}\label{6.31} Seien $\mathcal{L},\mathcal{M}$ invertierbare Garben auf einem Schema $X$. Dann gilt:
-\begin{enumerate}[(i)]
-\item $\mathcal{L}\otimes \mathcal{M}$ ist invertierbar.
-\item $\operatorname{\mathcal{H}om}_{\mathcal{O}_X}(\mathcal{L},\mathcal{O}_X)$ ist invertierbar und wird mit $\mathcal{L}^{-1}$ bezeichnet.
-\item $\mathcal{L}\otimes\mathcal{L}^{-1}\cong\mathcal{O}_X$
-\item $\operatorname{\mathcal{H}om}_{\mathcal{O}_X}(\mathcal{L},\mathcal{L})=\mathcal{O}_X$
-\end{enumerate}
-
-\paragraph{Definition 6.32.}\label{6.32}\begin{enumerate}[(i)]
-\item Wir setzen:
-\[\mathcal{L}^{\otimes m}=\underbrace{\mathcal{L}\otimes\cdots\otimes\mathcal{L}}_{m\text{-mal}},\quad\quad \mathcal{L}^{\otimes -m}=(\mathcal{L}^{-1})^{\otimes m},\quad m\geq 0 \]
-Also gilt $\mathcal{L}^{\otimes n}\otimes\mathcal{L}^{\otimes m}=\mathcal{L}^{\otimes (n+m)}$ für alle $n,m\in\mathbb{Z}$.
-\item Die \textit{Picard-Gruppe} $\operatorname{Pic}(X)$\index{Picard-Gruppe} eines Schemas $X$ ist die abelsche Gruppe von Isomorphieklassen invertierbarer Garben auf $X$ mit der Operation $\otimes$.
-\end{enumerate}
-
-\paragraph{Definition 6.33.}\label{6.33} Sei $D$ ein Cartier-Divisor auf $X$ repräsentiert durch $(U_i,f_i)_i$. Dann ist die Untergarbe $\mathcal{L}(D)$ von $\mathcal{K}$ definiert durch:
-\[\mathcal{L}(D)|_{U_i}=\mathcal{O}_{U_i}f_i^{-1}\subset\mathcal{K}|_{U_i} \]
-Diese ist wohldefiniert, denn auf $U_i\cap U_j$ haben wir $\mathcal{L}(D)|_{U_i\cap U_j}=\mathcal{O}_{U_i\cap U_j}f_i^{-1}=\mathcal{O}_{U_i\cap U_j}f_j^{-1}$, da $\frac{f_i}{f_j}\in\Gamma(U_i\cap U_j,\mathcal{O}^\times)$.
-$\mathcal{L}(D)$ heißt die zum Divisor $D$ \textit{assoziierte Garbe}\index{assoziierte Garbe!zum Divisor}.
-
-\paragraph{Satz 6.34.}\label{6.34} Sei $X$ ein Schema.
-\begin{enumerate}[(i)]
-\item Für jeden Cartier-Divisor $D$ ist $\mathcal{L}(D)$ eine invertierbare Garbe auf $X$. Die Abbildung $D\mapsto\mathcal{L}(D)$ induziert eine Bijektion:
-\[\{\text{Cartier-Divisoren auf }X \}\to\{\text{Invertierbare Untergarben von }\mathcal{K} \} \]
-\item $\mathcal{L}(D_1-D_2)\cong\mathcal{L}(D_1)\otimes\mathcal{L}(D_2)^{-1}$ für Cartier-Divisoren $D_1,D_2$
-\item $D_1\sim D_2\iff\mathcal{L}(D_1)\cong\mathcal{L}(D_2)$
-\end{enumerate}
-
-\paragraph{Beweis.} \begin{enumerate}[(i)]
-\item Sei repräsentiert durch $(U_i,f_i)_i$ mit $f_i\in\Gamma(U_i,\mathcal{K}^\times)$.
-Dann ist $\mathcal{O}_{U_i}\to\mathcal{L}(D)|_{U_i},\ 1\mapsto f_i^{-1}$ ein Isomorphismus, also $\mathcal{L}(D)$ invertierbar.
-Sei umgekehrt $\varphi:\mathcal{L}\hookrightarrow\mathcal{K}$ eine invertierbare Untergarbe und $\{U_i\}$ eine Überdeckung von $X$ mit $\tilde{\varphi}:\mathcal{O}_{U_i}\stackrel{\sim}{\to}\mathcal{L}|_{U_i}\hookrightarrow\mathcal{K}|_{U_i}$.
-Setze $f_i=\tilde{\varphi}(U_i)(1)^{-1}$.
-Dann definiert $(U_i,f_i)_i$ ein Cartier-Divisor $D$ mit $\mathcal{L}(D)=\mathcal{L}$.
-\item Seien $D_1, D_2$ repräsentiert durch $(U_i,f_i)_i$ bzw. $(U_i,g_i)_i$. Dann wird $D_1-D_2$ repräsentiert durch $(U_i,f_ig_i^{-1})_i$. Also gilt:
-\[\mathcal{L}(D_1-D_2)|_{U_i}=\mathcal{O}_{U_i}(f_i^{-1}g_i)=\mathcal{L}(D_1)|_{U_i}\otimes \mathcal{L}(D_2)|_{U_i}^{-1}\subset\mathcal{K}|_{U_i} \]
-Somit folgt $\mathcal{L}(D_1-D_2)=\mathcal{L}(D_1)\otimes\mathcal{L}(D_2)^{-1}$.
-\item Wegen (ii) reicht es zu zeigen, dass $D$ genau dann ein Hauptdivisor ist, wenn $\mathcal{L}(D)\cong\mathcal{O}_X$.
-Sei $D$ prinzipal, d.h.
-$D=(f)$ für ein $f\in\Gamma(X,\mathcal{K}^\times)$.
-Somit ist $\mathcal{L}(D)=\mathcal{O}_Xf^{-1}\cong\mathcal{O}_X$.
-Sei umgekehrt $\varphi:\mathcal{O}_X\to\mathcal{L}(D)$ ein Isomorphismus.
-Setze $f$ als das Bild von $1$ der folgenden Komposition:
-\[\Gamma(X,\mathcal{O}_X)\stackrel{\sim}{\to} \Gamma(X,\mathcal{L}(D))\hookrightarrow\Gamma(X,\mathcal{K}^\times) \]
-Dann ist $\mathcal{L}(D)=\mathcal{L}((f^{-1}))$, also $D=(f^{-1})$.\qed
-\end{enumerate}
-
-\paragraph{Korollar 6.35.}\label{6.35} Sei $X$ ein Schema. Dann induziert die Abbildung $D\mapsto\mathcal{L}(D)$ einen injektiven Gruppenhomomorphismus:
-\[\operatorname{CaCl}(X)\hookrightarrow\operatorname{Pic}(X)\]
-wobei $\operatorname{CaCl}(X)$ die Cartier-Divisorenklassengruppe bezeichnet. Im Allgemeinen ist diese Abbildung nicht surjektiv.
-
-\paragraph{Satz 6.36.}\label{6.36} Sei $X$ ein integres Schema. Dann ist die Abbildung $\operatorname{CaCl}(X)\to\operatorname{Pic}(X)$ ein Gruppenisomorphismus.
-
-\paragraph{Beweis.} Es genügt zu zeigen, dass jede invertierbare Garbe isomorph zu einer Untergarbe von $\mathcal{K}$ ist.
-Da $X$ integer ist, ist $\mathcal{K}$ konstante Garbe mit $\mathcal{K}(U)=K$ der Funktionenkörper von $X$ für alle $U$.
-Sei $\mathcal{L}$ eine invertierbare Garbe auf $X$ und betrachte $\mathcal{L}\otimes\mathcal{K}$.
-Sei $X=\bigcup_i U_i$ eine offene Überdeckung mit $\mathcal{L}|_{U_i}\cong\mathcal{O}_{U_i}$.
-Dann gilt:
-\[(\mathcal{L}\otimes_{\mathcal{O}_X}\mathcal{K})|_{U_i}=\mathcal{L}|_{U_i}\otimes_{\mathcal{O}_{U_i}}\mathcal{K}|_{U_i}=\mathcal{K}|_{U_i} \]
-Da $X$ irreduzibel ist, folgt $\mathcal{L}\otimes\mathcal{K}\cong\mathcal{K}$.
-Die kanonische Abbildung $\mathcal{L}\to\mathcal{L}\otimes\mathcal{K}\cong\mathcal{K}$ zeigt, dass $\mathcal{L}$ isomorph zu einer Untergarbe von $\mathcal{K}$ ist.\qed
-
-\paragraph{Korollar 6.37.}\label{6.37} Sei $X$ ein noethersches, integres, separiertes und lokal faktorielles Schema. Dann existiert ein Isomorphismus:
-\[\operatorname{Cl}(X)\cong\operatorname{Pic}(X) \]
-
-\paragraph{Beweis.} Folgt aus \hyperref[6.30]{Satz 6.30} und \hyperref[6.36]{Satz 6.36}.\qed
-
-\paragraph{Korollar 6.38.}\label{6.38} Sei $k$ ein Körper und $X=\mathbf{P}_k^n$. Dann ist jede invertierbare Garbe auf $X$ isomorph zu einem $\mathcal{O}(\ell)$ für ein $\ell\in\mathbb{Z}$.
-
-\paragraph{Beweis.} Nach \hyperref[6.37]{Korollar 6.37} und \hyperref[6.15]{Satz 6.15} ist $\operatorname{Pic}(X)\cong \operatorname{Cl}(X)\cong\mathbb{Z}$.
-Ferner wird $\operatorname{Cl}(X)$ von der Hyperebene $D=V_+(T_0)$ erzeugt.
-Dann gilt $\mathcal{L}(D)T_0=\mathcal{O}(1)$, da $\mathcal{O}(1)|_{D_+(T_i)}=\mathcal{O}_{D_+(T_i)}T_i$ und $\mathcal{L}(D)|_{D_+(T_i)}=\mathcal{O}_{D_+(T_i)}\frac{T_i}{T_0}$.\qed
-
-\section{Projektive Morphismen}
-
-\subsection{Ample Garben}
-
-\paragraph{Satz 7.1.}\label{7.1} Sei $A$ ein Ring, $X$ ein Schema über $A$ und $\mathbf{P}_A^n=\operatorname{Proj}A[x_0,\ldots,x_n]$.
-\begin{enumerate}[(i)]
-\item Ist $\varphi:X\to\mathbf{P}_A^n$ ein $A$-Morphismus.
-Dann ist $\varphi^\ast(\mathcal{O}(1))$ eine invertierbare Garbe auf $X$, die von globalen Schnitten $s_i=\varphi^\ast(x_i),\ i=0,\ldots,n$ erzeugt wird, wobei $x_i\in\Gamma(\mathbf{P}_A^n,\mathcal{O}(1))$.
-\item Sei $\mathcal{L}$ eine invertierbare Garbe auf $X$ und $s_0,\ldots,s_n\in\Gamma(X,\mathcal{L})$ globale Schnitte, die $\mathcal{L}$ erzeugen.
-Dann existiert ein eindeutig bestimmter $A$"~Morphismus $\varphi:X\to\mathbf{P}_A^n$ derart, dass $\mathcal{L}\cong\varphi^\ast(\mathcal{O}(1))$ mit $s_i=\varphi^\ast(x_i)$.
-\end{enumerate}
-
-\paragraph{Beweis.} \begin{enumerate}[(i)]
-\item Nach \hyperref[5.35]{5.35} erzeugen die globalen Schnitte $x_0,\ldots,x_n$ die Garbe $\mathcal{O}(1)$.
-Ferner ist $\varphi^\ast(\mathcal{O}(1))$ invertierbar und werden von $\varphi^\ast(x_i)=s_i$ erzeugt.
-\item Setze $X_i=\{P\in X\mid (s_i)_P\not\in\mathfrak{m}_P\mathcal{L}_P\}\subset_\text{o}X$.
-Da die $s_i$ die Garbe $\mathcal{L}$ erzeugen, gilt $X=\bigcup_iX_i$.
-Sei $U_i=D_+(x_i)\cong\operatorname{Spec}A[Y_0,\ldots,\widehat{Y_i},\ldots,Y_n],\ Y_j=\frac{x_j}{x_i}$.
-Betrachte den Ringhomomorphismus $\phi:A[Y_0,\ldots,\widehat{Y_i},\ldots,Y_n]\to\Gamma(X_i,\mathcal{O}_{X_i}),\ Y_j\mapsto \frac{s_j}{s_i}$.
-Dieser ist wohldefiniert, da  $(s_i)_P\not\in \mathfrak{m}_P\mathcal{L}_P\cong\mathfrak{m}_P\mathcal{O}_{X,P}$ für alle $P\in X_i$ gilt und daher $\frac{s_j}{s_i}\in\Gamma(X_i,\mathcal{O}_{X_i})$.
-Sei $X_i\to U_i$ der zu $\phi$ gehörige $A$"~Morphismus von $A$"~Schemata.
-Verkleben ergibt ein $A$"~Morphismus $\varphi:X\to\mathbf{P}_A^n$.
-Nach Konstruktion ist $\varphi$ eindeutig bestimmt durch $\mathcal{L}\cong\varphi^\ast(\mathcal{O}(1))$ und $s_i=\varphi^\ast(x_i)$.\qed
-\end{enumerate}
-
-\paragraph{Satz 7.2.}\label{7.2} Sei $\varphi:X\to\mathbf{P}_A^n$ ein $A$"~Morphismus, der zu einer invertierbaren Garbe $Z$ auf $X$ und globale Schnitte $s_0,\ldots,x_n\in\Gamma(X,Z)$ gehört, die $\mathcal{L}$ erzeugen.
-Dann ist $\varphi$ genau dann eine abgeschlossene Immersion, wenn die folgenden Bedingungen gelten:
-\begin{enumerate}[(i)]
-\item Alle offenen Teilmengen $X_i$ im Beweis von \hyperref[7.1]{Satz 7.1 (ii)} sind affin.
-\item Für alle $i$ ist die Abbildung $A[Y_0,\ldots,\widehat{Y_i},\ldots,Y_n]\to\Gamma(X_i,\mathcal{O}_{X_i}),\ Y_j\mapsto \frac{s_j}{s_i}$ surjektiv.
-\end{enumerate}
-
-\paragraph{Beweis.} Sei $\varphi$ eine abgeschlossene Immersion.
-Dann ist $X_i=U_i\cap X$ ein abgeschlossenes Unterschema von $U_i=D_+(x_i)\subset\mathbf{P}_A^n=\operatorname{Proj}A[x_0,\ldots,x_n]$.
-Da $U_i$ affin ist, ist nach \hyperref[5.18]{Korollar 5.18} auch $X_i$ affin und wir erhalten einen surjektiven Ringhomomorphismus $A[Y_0,\ldots,\widehat{Y_i},\ldots,Y_n]\to \Gamma(X_i,\mathcal{O}_{X_i})$.
-
-Sei nun umgekehrt (i) und (ii) erfüllt.
-Dann ist $X_i$ nach (ii) ein abgeschlossenes Unterschema von $U_i$.
-Da $X=\bigcup_i X_i$ und $X_i=\varphi^{-1}(U_i)$, ist $X$ ein abgeschlossenes Unterschema von $\mathbf{P}_A^n$.\qed
-
-\paragraph{Definition 7.3.}\label{7.3} Eine invertierbare Garbe $\mathcal{L}$ auf einem noetherschen Schema $X$ heißt \textit{ampel}\index{$\mathcal{O}_X$-Modul!ampel}, falls für jede kohärente Garbe $\mathcal{F}$ auf $X$ ein $n_0>0$ existiert, so dass für alle $n\geq n_0$ der $\mathcal{O}_X$"~Modul $\mathcal{F}\otimes\mathcal{L}^n$ von globalen Schnitten erzeugt wird.
-
-\paragraph{Beispiel 7.4.}\label{7.4} Ist $X$ affin, so ist jede invertierbare Garbe ampel: Jede kohärente Garbe wird von globalen Schnitten endlich erzeugt, vgl. \hyperref[5.10]{Korollar 5.10}.
-
-\paragraph{Bemerkung 7.5.}\label{7.5} \hyperref[5.37]{Serres Theorem 5.37} besagt, dass jede sehr ample Garbe $\mathcal{L}$ auf einem Schema $X$ über einem noetherschen Ring ampel ist.
-Die Umkehrung ist im Allgemeinen falsch.
-
-\paragraph{Satz 7.6.}\label{7.6} Sei $\mathcal{L}$ eine invertierbare Garbe auf einem noetherschen Schema $X$. Dann sind folgende Aussagen äquivalent:
-\begin{enumerate}[(i)]
-\item $\mathcal{L}$ ist ampel.
-\item $\mathcal{L}^m$ ist ampel für alle $m>0$.
-\item $\mathcal{L}^m$ ist ampel für ein $m>0$.
-\end{enumerate}
-
-\paragraph{Beweis.} (i)$\implies$(ii)$\implies$(iii) ist klar.
-Sei nun $\mathcal{L}^m$ ampel für ein $m>0$ und $\mathcal{F}$ eine kohärente Garbe auf $X$.
-Nach Definition existiert ein $n_0>0$, so dass für alle $n\geq n_0$ die Garbe $\mathcal{F}\otimes(\mathcal{L}^m)^n$ von globalen Schnitten erzeugt wird.
-Betrachte die Garben $\mathcal{F}\otimes\mathcal{L}^k$ für $k=1,\ldots,m-1$.
-Diese sind kohärent, d.h.
-es gibt für jedes $k$ ein $n_k>0$, so dass für alle $n\geq n_k$ die Garbe $\mathcal{F}\otimes\mathcal{L}^k\otimes(\mathcal{L}^m)^n$ von globalen Schnitten erzeugt wird.
-Setzen wir nun $N=m\cdot\max\{n_k\mid k=1,\ldots,m-1\}$, so wird $\mathcal{F}\otimes\mathcal{L}^n$ von globalen Schnitten erzeugt für alle $n\geq N$.
-Somit ist $\mathcal{L}$ ampel.\qed
-
-\paragraph{Satz 7.7.}\label{7.7} Sei $X$ ein Schema von endlichem Typ über einem noetherschen Ring $A$ und $\mathcal{L}$ eine invertierbare Garbe auf $X$.
-Dann ist $\mathcal{L}$ genau dann ampel, wenn $\mathcal{L}^m$ sehr ampel bzgl.
-$\operatorname{Spec}(A)$ für ein $m>0$ ist.
-
-\paragraph{Lemma 7.8.}\label{7.8} Sei $X$ ein noethersches Schema, $U\subset_\text{o}X$ und $\mathcal{F}$ eine kohärente Garbe auf $U$.
-Dann existiert eine kohärente Garbe $\mathcal{F}'$ auf $X$ mit $\mathcal{F}'|_U=\mathcal{F}$.
-
-\paragraph{Beweis.} Sei o.B.d.A.
-$X$ affin.
-Sei $j:U\hookrightarrow X$ die Immersion.
-Dann ist $\mathcal{F}'=j_\ast\mathcal{F}$ nach \hyperref[5.13]{Satz 5.13 (iii)} quasikohärent.
-Ferner sind quasikohärente Garben auf einem noetherschen Schema Vereinigung seiner kohärenten Untergarben.
-Daher gilt:
-\[j_\ast\mathcal{F} =\mathcal{F}' = \bigcup_\lambda\mathcal{F}'_\lambda \]
-mit kohärenten Untergarben $\mathcal{F}'_\lambda$.
-Nach \hyperref[5.13]{Satz 5.13 (ii)} sind $j^\ast\mathcal{F}'_\lambda$ kohärent.
-Somit ist $\mathcal{F}=\bigcup_\lambda j^\ast\mathcal{F}'_\lambda$, wobei $j^\ast\mathcal{F}'_\lambda$ die kohärenten Untergarben von $\mathcal{F}$ durchläuft.
-Da $\mathcal{F}$ kohärent ist, gilt $\mathcal{F}=j^\ast\mathcal{F}'_\lambda$ für ein $\lambda$.
-Somit ist $\mathcal{F}'=\mathcal{F}'_\lambda$ kohärent.\qed
-
-\paragraph{Beweis von \hyperref[7.7]{Lemma 7.7}.} Sei zunächst $\mathcal{L}^m$ sehr ampel für ein $m>0$.
-Es existiert also eine Immersion $i:X\to\mathbf{P}_A^n$ mit $\mathcal{L}^m\cong i^\ast(\mathcal{O}(1))$.
-Sei $\overline{X}$ der Abschluss von $X$ in $\mathbf{P}_A^n$.
-Dann ist $\overline{X}$ ein projektives Schema über $\operatorname{Spec}(A)$.
-Nach \hyperref[7.5]{7.5} ist $\mathcal{O}_{\overline{X}}(1)$ ampel auf $\overline{X}$.
-Sei $\mathcal{F}$ eine kohärente Garbe auf $X$ und $\overline{\mathcal{F}}$ die kohärente Garbe in \hyperref[7.8]{Lemma 7.8} auf $\overline{X}$ mit $\overline{F}|_X=\mathcal{F}$.
-Wird $\overline{\mathcal{F}}\otimes\mathcal{O}_{\overline{X}}(\ell)$ von globalen Schnitten erzeugt, so ist dies auch für $\mathcal{F}\otimes\mathcal{O}_X(\ell)$ der Fall.
-Daher ist $\mathcal{L}^m$ ampel auf $X$ und nach \hyperref[7.6]{Satz 7.6} auch $\mathcal{L}$.
-Die andere Richtung ist schwierig.
-
-\subsection{Projektive Vektorbündel}
-
-\paragraph{Definition.} Sei $X$ ein Schema.
-Ein quasikohärenter $\mathcal{O}_X$"~Modul $\mathcal{S}$ habe die Eigenschaft $(\dagger)$, wenn $X$ noethersch ist und wenn er eine Struktur einer graduierten $\mathcal{O}_X$"~Algebra trägt, d.h.
-$\mathcal{S}=\bigoplus_{d\geq 0}\mathcal{S}_d$ mit homogenen Teilen $\mathcal{S}_d$, so dass:
-\begin{enumerate}[(i)]
-\item $\mathcal{S}_0=\mathcal{O}_X$
-\item $\mathcal{S}_1$ ist ein kohärenter $\mathcal{O}_X$"~Modul.
-\item $\mathcal{S}$ wird lokal von $\mathcal{S}_1$ als $\mathcal{O}_X$"~Modul erzeugt.
-\end{enumerate}
-
-\paragraph{Definition 7.9.}\label{7.9} Sei $X$ ein Schema und $\mathcal{S}$ ein graduierter $\mathcal{O}_X$"~Modul mit $(\dagger)$.
-Sei $U=\operatorname{Spec}(A)\subset_\text{o}X$, $\mathcal{S}(U)=\Gamma(U,\mathcal{S}|_U)$ die graduierte $A$"~Algebra und $\pi_U:\operatorname{Proj}\mathcal{S}(U)\to U$ der kanonische Morphismus.
-Für $U,V\subset_\text{o}X$ affin gilt $\pi_U^{-1}(U\cap V)\cong\pi_V^{-1}(U\cap V)$.
-Verkleben von $\pi_U$ liefert ein Schema $\operatorname{\mathbf{Proj}}(\mathcal{S})$ und ein Morphismus $\pi:\operatorname{\mathbf{Proj}}(\mathcal{S})\to X$, so dass für alle affinen $U\subset_\text{o}X$ stets $\pi^{-1}(U)\cong\operatorname{Proj}\mathcal{S}(U)$ gilt.
-Die invertierbaren Garben $\mathcal{O}(1)$ auf $\operatorname{Proj}\mathcal{S}(U)$ lassen sich zu einer invertierbaren Garbe $\mathcal{O}(1)$ auf $\operatorname{\mathbf{Proj}}(\mathcal{S})$ verkleben.
-
-\paragraph{Beispiel 7.10.}\label{7.10} Sei $\mathcal{O}_X[T_0,\ldots,T_n]=\mathcal{S}$.
-Dann ist $\operatorname{\mathbf{Proj}}(\mathcal{S})=\mathbf{P}_X^n=\mathbf{P}_\mathbb{Z}^n\times X$ mit der getwisteten Garbe $\mathcal{O}(1)$ wie in \hyperref[5.30]{Definition 5.30}.
-
-\paragraph{Lemma 7.11.}\label{7.11} Sei $\mathcal{S}$ eine Garbe graduierter Algebren mit $(\dagger)$ auf $X$.
-Sei $\mathcal{L}$ eine invertierbare Garbe auf $X$ und $\mathcal{S}'$ die folgende Garbe graduierter Algebren:
-\[\mathcal{S}_d' =\mathcal{S}_1\otimes\mathcal{L}^d,\quad d\geq 0 \]
-Dann erfüllt $\mathcal{S}'$ ebenfalls $(\dagger)$ und es gibt einen natürlichen Isomorphismus:
-\[\begin{tikzcd}
-P'=\operatorname{\mathbf{Proj}}(\mathcal{S}')\ar[d, "\pi'"']\ar[r, "\varphi", "\cong"'] & P=\operatorname{\mathbf{Proj}}(\mathcal{S})\ar[dl, "\pi"]\\
-X&
-\end{tikzcd} \]
-Es gilt $\mathcal{O}_{P'}(1)\cong \varphi^\ast\mathcal{O}_P(1)\otimes\pi'^\ast\mathcal{L}$. $\mathcal{S}'$ wird auch mit $\mathcal{S}\ast\mathcal{L}$ bezeichnet.
-
-\paragraph{Satz 7.12.}\label{7.12} Sei $X$ und $\mathcal{S}$ mit $(\dagger)$. Sei $P=\operatorname{\mathbf{Proj}}(\mathcal{S})$ mit $\pi:P\to X$. Dann gilt:
-\begin{enumerate}[(i)]
-\item $\pi$ ist eigentlicher Morphismus.
-\item Besitzt $X$ eine ample Garbe $\mathcal{L}$, so ist $\pi$ projektiv und $\mathcal{O}_P(1)\otimes\pi^\ast\mathcal{L}^n$ ist eine sehr ample invertierbare Garbe von $P$ über $X$.
-\end{enumerate}
-
-\paragraph{Definition 7.13.}\label{7.13} \begin{enumerate}[(i)]
-\item Sei $A$ ein Ring und $M$ ein $A$"~Modul. Setze:
-\[T^0(M)=A,\quad T^n(M)= M^{\otimes n},\ n\geq 1 \]
-Dann ist $T(M)=\bigoplus_{n\geq 0}T^n(M)$ eine $A$"~Algebra mit $\otimes$ als Multiplikation, die sogenannte \textit{Tensor"~Algebra}\index{Tensor-Algebra} von $M$.
-\item Die \textit{symmetrische (Tensor-)Algebra}\index{Tensor-Algebra!symmetrisch} von $M$ ist definiert als der Quotient von $T(M)$ nach dem Ideal, das von den Elementen $x\otimes y-y\otimes x,\ x,y\in M$ erzeugt wird.
-$S(M)=\bigoplus_{n\geq 0}S^n(M)$ ist dann eine kommutative, graduierte $A$"~Algebra.
-$S^n(M)$ heißt \textit{$n$-tes symmetrisches Produkt}\index{symmetrisches Produkt} von $M$.
-\item Sei $X$ ein Schema und $\mathcal{F}$ ein $\mathcal{O}_X$"~Modul.
-Dann wird die Tensor-Algebra $T(\mathcal{F})$ bzw.
-die symmetrische Algebra $S(\mathcal{F})$ von $\mathcal{F}$ als assoziierte Garbe zur Prägarbe $U\mapsto T(\mathcal{F}(U))$ bzw.
-$U\mapsto S(\mathcal{F}(U))$ definiert.
-$T(\mathcal{F})$ und $S(\mathcal{F})$ sind $\mathcal{O}_X$"~Algebren.
-\end{enumerate}
-
-\paragraph{Beispiel.} Sei $M$ ein freier $A$"~Modul vom Rang $r$. Dann gilt $S(M)\cong A[X_1,\ldots,X_r]$.
-
-\paragraph{Definition 7.14.}\label{7.14} Sei $X$ ein noethersches Schema und $\mathcal{E}$ eine lokal freie, kohärente Garbe auf $X$.
-Sei $\mathcal{S}=S(\mathcal{E})$ die symmetrische Algebra von $\mathcal{E}$ und $\mathbf{P}(\mathcal{E})=\operatorname{\mathbf{Proj}}(\mathcal{S})$.
-Dann ist $\mathcal{S}$ eine Garbe von graduierten $\mathcal{O}_X$"~Algebren, die $(\dagger)$ erfüllt.
-$\pi:\mathbf{P}(\mathcal{E})\to X$ heißt \textit{projektives Vektorbündel}\index{Vektorbündel!projektiv} über $X$.
-
-\paragraph{Bemerkung.} Ist $\mathcal{E}$ frei vom Rang $n+1$ über $U\subset_\text{o}X$, so ist $\pi^{-1}(U)\cong\mathbf{P}_U^n$.
-
-\paragraph{Satz 7.15.}\label{7.15} Sei $\pi:\mathbf{P}(\mathcal{E})\to X$ ein projektives Vektorbündel über $X$. Dann gilt mit $\mathcal{S}=S(\mathcal{E})$:
-\begin{enumerate}[(i)]
-\item Ist Rang von $\mathcal{E}$ größer gleich $2$, so gibt es einen kanonischen Isomorphismus von graduierten $\mathcal{O}_X$"~Algebren:
-\[\mathcal{S}\cong\bigoplus_{\ell\in\mathbb{Z}}\pi_\ast\mathcal{O}(\ell)\]
-Insbesondere folgt $\pi_\ast(\mathcal{O}(\ell))=0$ für $\ell<0$ und $\pi_\ast(\mathcal{O})=\mathcal{O}_X$, $\pi_\ast(\mathcal{O}(1))=\mathcal{E}$.
-\item Es gibt einen natürliche surjektiven Morphismus:
-\[\pi^\ast\mathcal{E}\to\mathcal{O}(1) \]
-\end{enumerate}
-
-\paragraph{Satz 7.16.}\label{7.16} \textit{(Universaleigenschaft des projektiven Vektorbündels)} Sei $\pi:\mathbf{P}(\mathcal{E})\to X$ ein projektives Vektorbündel von $X$ und $g:Y\to X$ ein Morphismus.
-Dann gibt es genau dann einen Morphismus $f$ mit kommutativen Diagramm
-\[\begin{tikzcd}
-Y\rar[dashed, "f"]\dar["g"'] &\mathbf{P}(\mathcal{E})\ar[dl, "\pi"]\\
-X&
-\end{tikzcd}\]
-wenn es eine invertierbare Garbe $\mathcal{L}$ auf $Y$ zusammen mit einem surjektiven Garbenmorphismus $g^\ast\mathcal{E}\to\mathcal{L}$ gibt.
-
-\subsection{Aufblasung}
-
-\paragraph{Definition 7.17.}\label{7.17} Sei $X$ ein noethersches Schema und $\mathcal{J}$ eine kohärente Idealgarbe auf $X$.
-Setze $\mathcal{J}^0=\mathcal{O}_X$ und $\mathcal{J}^d$ als das $d$-fache Produkt des Ideals $\mathcal{J}$.
-Sei $\mathcal{S}=\bigoplus_{d\geq 0}\mathcal{J}^d$.
-Dann erfüllen $X$ und $\mathcal{S}$ die Bedingung $(\dagger)$.
-$\widetilde{X}=\mathbf{Proj}(\mathcal{S})$ heißt die \textit{Aufblasung} von $X$ bzgl.
-$\mathcal{J}$.\index{Aufblasung}
-
-Ist $Y$ das abgeschlossene Unterschema von $X$, das zu $\mathcal{J}$ gehört, so sagen wir auch, dass $\widetilde{X}$ die Aufblasung von $X$ entlang $Y$ ist.
-
-\paragraph{Beispiel 7.18.}\label{7.18} Sei $X=\mathbf{A}^n_k$ und $P\in X$ der Nullpunkt, d.h.
-$X=\operatorname{Spec}(A),\ A=k[X_1,\ldots,X_n],\ I=(X_1,\ldots,X_n)$.
-Dann ist die Aufblasung von $X$ bei $P$ gegeben durch:
-\[\widetilde{X}=\operatorname{Proj}(S),\quad S=\bigoplus_{d\geq 0}I^d\]
-Betrachte die surjektive Abbildung $\psi:A[Y_1,\ldots,Y_n]\to S,\ Y_i\mapsto X_i$.
-Wir sehen, dass $\widetilde{X}$ isomorph zu dem abgeschlossenen Unterschema von $\mathbf{P}_A^{n-1}$ ist, das durch die homogenen Polynome in den $Y_i$, die $\ker(\psi)$ erzeugen, definiert ist:
-\[\ker(\psi)=\langle X_iY_j-X_jY_i\mid i,j=1,\ldots,n\rangle\]
-
-\paragraph{Definition 7.19.}\label{7.19} Sei $f:X\to Y$ ein Morphismus und $\mathcal{J}\subset\mathcal{O}_Y$ eine Idealgarbe auf $Y$.
-Wir definieren die \textit{inverse Idealgarbe}\index{Idealgarbe!invers} $\mathcal{J}'\subset\mathcal{O}_X$ auf $X$ wie folgt:
-
-$\mathcal{J}\hookrightarrow\mathcal{O}_Y\stackrel{f^\sharp}{\to}f_\ast\mathcal{O}_X$ liefert nach \hyperref[5.1]{5.1 (xi)} ein $\mathcal{O}_X$"~Modulmorphismus $f^\ast\mathcal{J}\to\mathcal{O}_X$.
-Setze $\mathcal{J}'=\operatorname{im}(f^\ast\mathcal{J}\to\mathcal{O}_X)$.
-
-Alternativ: Betrachte die Idealgarbe $f^{-1}\mathcal{J}$ auf $f^{-1}\mathcal{O}_Y$ und den Morphismus $f^{-1}\mathcal{O}_Y\to \mathcal{O}_Y$, der zu $f^\sharp$ adjungiert ist.
-Dann ist $\mathcal{J}'$ die vom Bild von $f^{-1}\mathcal{J}$ in $\mathcal{O}_X$ erzeugte Idealgarbe.
-Wir schreiben daher auch $\mathcal{J}'=f^{-1}\mathcal{J}\cdot\mathcal{O}_X$.
-
-\paragraph{Bemerkung.} Im Allgemeinen sind $f^\ast\mathcal{J}$ und $f^{-1}\mathcal{J}$ verschiedene Idealgarben.
-
-\paragraph{Satz 7.20.}\label{7.20} Sei $X$ ein noethersches Schema und $\mathcal{J}$ eine kohärente Idealgarbe auf $X$.
-Sei $\pi:\widetilde{X}\to X$ die Aufblasung von $X$ bzgl.
-$\mathcal{J}$.
-Dann gilt:
-\begin{enumerate}[(i)]
-\item Die inverse Idealgarbe $\widetilde{J}=\pi^{-1}\mathcal{J}\cdot\mathcal{O}_{\widetilde{X}}$ ist invertierbar auf $\mathcal{O}_{\widetilde{X}}$.
-\item Sei $Y$ das zu $\mathcal{J}$ gehörige abgeschlossene Unterschema und $U=X\setminus Y$. Dann ist $\pi:\pi^{-1}(U)\to U$ ein Isomorphismus.
-\end{enumerate}
-
-\paragraph{Beweis.}\begin{enumerate}[(i)]
-\item Sei $\widetilde{X}=\mathbf{Proj}(\mathcal{S}),\ \mathcal{S}=\bigoplus_{d\geq 0}\mathcal{J}^d$ und $V\subset_\text{o}X$ affin.
-Dann ist $\mathcal{O}(1)|_{\operatorname{Proj}\mathcal{S}(V)}$ die assoziierte Garbe zum graduierten $\mathcal{S}(V)$"~Modul $\mathcal{S}(V)(1)=\bigoplus_{d\geq 0}\mathcal{J}^{d+1}(V)=\mathcal{J}(V)\cdot \mathcal{S}(V)$ das von $\mathcal{J}(V)$ in $\mathcal{S}(V)$ erzeugte Ideal.
-Es folgt $\widetilde{\mathcal{J}}=\pi^{-1}\mathcal{J}\cdot\mathcal{O}_{\widetilde{X}}=\mathcal{O}_{\widetilde{X}}(1)$, also invertierbar.
-\item Es ist $\mathcal{J}|_U\cong\mathcal{O}_U$, da $0\to\mathcal{J}\to\mathcal{O}_X\stackrel{i^\sharp}{\to }i_\ast\mathcal{O}_Y\to 0$ exakt ist.
-Daher gilt $\pi^{-1}(U)=\operatorname{\mathbf{Proj}}\mathcal{O}_U[T]=U$.
-Beachte, dass für ein allgemeines $A$ stets $\operatorname{Proj}A[T]=D_+(T)=\operatorname{Spec}A[T]_{(T)}=\operatorname{Spec}(A)$ gilt.\qed
-\end{enumerate}
-
-\paragraph{Satz 7.21.}\label{7.21} \textit{(Universaleigenschaft des Aufblasens)} Sei $X$ ein noethersches Schema, $\mathcal{J}$ eine kohärente Idealgarbe auf $X$ und $\pi:\widetilde{X}\to X$ die Aufblasung von $X$ bzgl.
-$\mathcal{J}$.
-Ist $f:Z\to X$ ein Morphismus, so dass $f^{-1}\mathcal{J}\cdot\mathcal{O}_Z$ eine invertierbare Idealgarbe auf $Z$ ist, so existiert ein eindeutig bestimmter Morphismus $g:Z\to\widetilde{X}$ mit kommutativem Diagramm:
-\[\begin{tikzcd}
-Z\rar["g", dashed]\ar[rd, "f"'] & \widetilde{X}\dar["\pi"]\\
-& X
-\end{tikzcd} \]
-
-\paragraph{Lemma 7.22.}\label{7.22} Sei $X$ ein Schema, $f:\mathcal{L}\to\mathcal{M}$ ein surjektiver Morphismus invertierbarer Garben. Dann ist $f$ ein Isomorphismus.
-
-\paragraph{Beweis.} Siehe Bourbaki: Commutative Algebra II, §3.2 Korollar zu Proposition 6.\qed
-
-\paragraph{Korollar 7.23.}\label{7.23} Sei $f:Y\to X$ ein Morphismus noetherscher Schemata und $\mathcal{J}$ eine kohärente Idealgarbe auf $X$.
-Sei $\widetilde{X}$ die Aufblasung bzgl.
-$\mathcal{J}$ und $\widetilde{Y}$ die Aufblasung bzgl.
-$\mathcal{J}_Y=f^{-1}\mathcal{J}\cdot\mathcal{O}_Y$.
-Dann existiert ein eindeutig bestimmter Morphismus $\widetilde{f}$ mit kommutativem Diagramm:
-\[\begin{tikzcd}
-\widetilde{Y}\ar[rr, "\widetilde{f}"]\dar["\pi_Y"'] && \widetilde{X}\dar["\pi_X"]\\
-Y\ar[rr, "f"'] && X
-\end{tikzcd}\]
-Ist $f$ eine abgeschlossene Immersion, so auch $\widetilde{f}$.
-
-\paragraph{Beweis.} $\mathcal{J}_Y$ ist kohärent, da nach \hyperref[5.13]{Satz 5.13 (ii)} $f^\ast\mathcal{J}$ kohärent ist und nach \hyperref[5.12]{Satz 5.12 (ii)} auch $\operatorname{im}(f^\ast\mathcal{J}\to\mathcal{O}_Y)$.
-Nach \hyperref[7.20]{Satz 7.20} ist $\pi_Y^{-1}\mathcal{J}_Y\cdot\mathcal{O}_{\widetilde{Y}}$ eine invertierbare Garbe auf $\widetilde{Y}$.
-Nach \hyperref[7.21]{Satz 7.21} existiert $\widetilde{f}$ und ist eindeutig bestimmt.
-
-Sei nun $f$ eine abgeschlossene Immersion und:
-\[\widetilde{X}=\operatorname{\mathbf{Proj}}(\mathcal{S}),\quad \mathcal{S}=\bigoplus_{d\geq 0}\mathcal{J}^d,\qquad\widetilde{Y}=\mathbf{Proj}(\mathcal{S}_Y),\quad\mathcal{S}_Y=\bigoplus_{d\geq 0}\mathcal{J}_Y^d \]
-Ist $Y\hookrightarrow X$ eine abgeschlossene Immersion, so ist $\mathcal{J}\to\mathcal{J}_Y$ surjektiv und somit ist auch $\mathcal{S}\to\mathcal{S}_Y$ als Homomorphismus graduierter Ringe surjektiv.
-Daher ist $\widetilde{Y}\hookrightarrow\widetilde{X}$ eine abgeschlossene Immersion.\qed
-
-\paragraph{Definition 7.24.}\label{7.24} Ist $Y\hookrightarrow X$ eine abgeschlossene Immersion noetherscher Schemata.
-Dann heißt das abgeschlossene Unterschema $\widetilde{Y}$ von $\widetilde{X}$ auch \textit{strikte Transformation} von $Y$ bzgl.
-der Aufblasung $\pi:\widetilde{X}\to X$.\index{strikte Transformation}
-
-\section{Differentiale}
-
-Sei $A$ ein Ring, $B$ eine $A$"~Algebra und $M$ ein $B$"~Modul.
-
-\paragraph{Definition 8.1.}\label{8.1} Eine \textit{$A$"~Derivation}\index{Derivation} von $B$ nach $M$ ist eine Abbildung $\mathrm{d}:B\to M$ mit:
-\begin{enumerate}[(i)]
-\item $\mathrm{d}$ ist additiv
-\item $\mathrm{d}(b\cdot b')=b\cdot \mathrm{d}(b')+ b'\cdot \mathrm{d}(b)$
-\item $\mathrm{d}(a)=0$ für alle $a\in A$
-\end{enumerate}
-Der $B$"~Modul aller $A$"~Derivationen von $B$ nach $M$ bezeichnen wir mit $D_A(B,M)$.
-
-\paragraph{Definition 8.2.}\label{8.2} Der $B$"~Modul $\Omega_{B/A}$ der \textit{relativen Differentialformen}\index{Differentialform!relativ} von $B$ über $A$ zusammen mit einer $A$"~Derivation $\mathrm{d}:B\to\Omega_{B/A}$ ist definiert durch folgende Universaleigenschaft:
-
-Für alle $B$"~Moduln $M$ und $A$"~Derivationen $\mathrm{d}':B\to M$ gibt es genau ein $B$"~Ho\-mo\-mor\-phis\-mus $f:\Omega_{B/A}\to M$, so dass folgendes Diagramm kommutiert:
-\[\begin{tikzcd}
-B\ar[r, "\mathrm{d}"]\ar[rd, "\mathrm{d}'"'] & \Omega_{B/A}\ar[d, "f", dashed]\\
-& M
-\end{tikzcd}\]
-Die Eindeutigkeit von $\Omega_{B/A}$ ist klar. Für die Existenz betrachte den freien $B$"~Modul $F$ mit Basis $\{d_b\mid b\in B\}$. Dann gilt:
-\[\Omega_{B/A}=F / \langle d_{b+b'}-d_b-d_{b'},\ d_{bb'}-bd_{b'}-b'd_b,\ d_a\mid b,b'\in B,\ a\in A\rangle \]
-Zusammen mit der $A$"~Derivation $\mathrm{d}:B\to\Omega_{B/A},\ b\mapsto d_b$ erfüllt diese Konstruktion die Universaleigenschaft.
-
-\paragraph{Satz 8.3.}\label{8.3} Sei $B$ eine $A$"~Algebra und $f:B\otimes_A B\to B,\ b\otimes b'\mapsto bb'$.
-Betrachte $B\otimes_AB$ als $B$"~Modul via $b(b_1\otimes b_2)=(bb_1)\otimes b_2$.
-$B\otimes_A B$ wird zur $B$"~Algebra durch $(b_1\otimes b_1')(b_2\otimes b_2')=(b_1b_2)\otimes (b_1'b_2')$.
-Sei $I=\ker(f)$.
-Dann ist $I/I^2$ ein $B$"~Modul.
-Setze $\mathrm{d}:B\to I/I^2,\ \mathrm{d}b=1\otimes b-b\otimes 1\mod I^2$.
-Dann ist $\Omega_{B/A}= (I/I^2,\mathrm{d})$.
-
-\paragraph{Beweis.} Wir zeigen zunächst $I=\sum_{b\in B}B\mathrm{d}b$. Sei $\beta=\sum b_i\otimes c_i\in I$, also $\sum b_ic_i=0$. Dann gilt:
-\begin{align*}
-\beta &= \sum_i (b_i(1\otimes c_i-c_i\otimes 1) + b_ic_i\otimes 1)\\
-&= \sum_i b_i \mathrm{d}c_i + \Big(\sum_i b_ic_i\Big)\otimes 1 \in\sum_{b\in B} B\mathrm{d}b
-\end{align*}
-Wir zeigen nun, dass $\mathrm{d}$ eine $A$"~Derivation ist. $\mathrm{d}(A)=0$ und die Additivität ist klar.
-\begin{align*}
-\mathrm{d}b\mathrm{d}c &= (1\otimes b- b\otimes 1)(1\otimes c-c\otimes 1)\\
-&= bc\otimes 1 - b\otimes c- c\otimes b+1\otimes bc\\
-&=\mathrm{d}(bc) + 2bc\otimes 1- b(\mathrm{d}c+c\otimes 1) - c (\mathrm{d}b + b\otimes 1)\\
-&=\mathrm{d}(bc)-b\mathrm{d}c - c \mathrm{d}b
-\end{align*}
-Also ist $\mathrm{d}(bc)\equiv b\mathrm{d}c+c\mathrm{d}b\mod I^2$.
-Somit müssen wir noch zeigen, dass $(I/I^2,\mathrm{d})$ die Universaleigenschaft von $\Omega_{B/A}$ erfüllt.
-Sei $M$ ein $B$"~Modul und $\mathrm{D}:B\to M$ eine $A$"~Derivation.
-Zu zeigen ist die Existenz und Eindeutigkeit von einem $B$"~Homomorphismus $f:I/I^2\to M$ mit $\mathrm{D}=f\circ \mathrm{d}$.
-Da $I=\sum B\mathrm{d}b$ ist, folgt die Eindeutigkeit.
-
-Betrachte die \textit{triviale Erweiterung} $B\ast M$ von $B$ mit $M$, d.h.
-$B\oplus M$ mit der Multiplikation $(b_1,m_1)(b_2,m_2)=(b_1b_2,b_1m_2+b_2m_1)$.
-Dann ist $B\ast M$ ein Ring mit Einselement $(1,0)$.
-$B\ast M$ trägt eine $B$"~Modulstruktur via $b'(b,m)=(bb',bm)$, d.h.
-$B\ast M$ ist eine $B$"~Algebra.
-Wir haben kanonische $B$"~Homomorphismen:
-\begin{itemize}
-\item Die Einbettung $M\hookrightarrow B\ast M,\ m\mapsto (0,m)$
-\item Die Projektion $B\ast M\to B,\ (b,m)\mapsto b$
-\end{itemize}
-Betrachte den $B$"~Algebrenhomomorphismus $\varphi:B\otimes_AB\to B\ast M,\ x\otimes y\mapsto (xy,x\mathrm{D}y)$.
-Es gilt $\varphi(I)\subset M\subset B\ast M$, da $\varphi(\mathrm{d}b)=\varphi(1\otimes b- b\otimes 1)=(b,\mathrm{D}b)-(b,b\mathrm{D}1)=(0,\mathrm{D}b)$.
-Da $M^2=0$ in $B\ast M$, induziert $\varphi$ eine Abbildung $\phi: B\otimes_A B/I^2\to B\ast M$.
-Dann erfüllt $f=\phi|_{I/I^2}:I/I^2\to M$ die gewünschte Eigenschaft $f\circ\mathrm{d}=\mathrm{D}$.\qed
-
-\paragraph{Satz 8.4.}\label{8.4} Es gibt für jeden $B$"~Modul $M$ einen kanonischen $B$"~Modulisomorphismus:
-\[\operatorname{Hom}_B(\Omega_{B/A},M)\to\operatorname{Der}_A(B,M),\ f\mapsto f\circ\mathrm{d} \]
-
-\paragraph{Beweis.} Folgt direkt aus der Universaleigenschaft.\qed
-
-\paragraph{Beispiel 8.5.}\label{8.5} Sei $B=A[X_1,\ldots,X_n]$.
-Dann ist $\Omega_{B/A}=\bigoplus_{i=1}^n B\mathrm{d}X_i$ ein freier $B$"~Modul vom Rang $n$.
-Allgemein gilt: Wird $B$ als $A$"~Algebra von $y_1,\ldots,y_m$ erzeugt, so ist $\Omega_{B/A}=\sum_{i=1}^m B\mathrm{d}y_1$, da:
-\[\mathrm{d}\Big(\prod_i y_i^{n_i}\Big) = \sum_i n_i\prod_{j\neq i} y_j^{n_j}y_i^{n_i - 1}\mathrm{d}y_i \]
-Angenommen, $\sum_i b_i\mathrm{d}X_i=0$.
-Betrachte die Derivationen $\frac{\partial}{\partial X_i}\in\operatorname{Der}_A(B)$.
-Nach \hyperref[8.4]{Satz 8.4} existiert für jedes $k$ ein $f_k\in\operatorname{Hom}_B(\Omega_{B/A},B)$ mit $\frac{\partial}{\partial X_k}=f_k\circ\mathrm{d}$.
-Es gilt für alle $k$:
-\[ 0= f_k\Big(\sum_{i=1}^n b_i\mathrm{d}X_i\Big)= \sum_{i=1}^n b_i f_k(\mathrm{d}X_i) = \sum_{i=1}^n b_i \frac{\partial X_i}{\partial X_k} = \sum_{i=1}^n b_i\delta_{ik}=b_k \]
-Somit ist $\Omega_{B/A}=\bigoplus_{i=1}^nB\mathrm{d}X_i$
-
-\paragraph{Satz 8.6.}\label{8.6} \begin{enumerate}[(i)]
-\item Sei $S$ eine multiplikative Teilmenge der $A$"~Algebra $B$. Dann gilt:
-\[S^{-1}\Omega_{B/A}\cong\Omega_{S^{-1}B/A} \]
-\item Seien $A',B$ $A$"~Algebren und $B'=B\otimes_AA'$. Dann gilt:
-\[\Omega_{B'/A'}\cong\Omega_{B/A}\otimes_BB' \]
-\end{enumerate}
-
-\paragraph{Satz 8.7.}\label{8.7} Seien $\phi:A\to B,\ \psi:B\to C$ Ringhomomorphismen. Dann gilt:
-\begin{enumerate}[(i)]
-\item \textit{(Erste fundamentale exakte Sequenz)}\index{Fundamentale exakte Sequenz} Es gibt eine exakte Sequenz:
-\[\begin{tikzcd}
-\Omega_{B/A}\otimes_BC \ar[r, "v"] & \Omega_{C/A} \ar[r, "u"] & \Omega_{C/B}\rar & 0
-\end{tikzcd} \]
-\item $v$ ist injektiv und die Sequenz zerfällt, genau dann wenn für alle $C$"~Moduln $M$ und jedes $\mathrm{D}\in \operatorname{Der}_A(B,M)$ zu einer Derivation aus $\operatorname{Der}_A(C,M)$ fortgesetzt werden kann.
-\end{enumerate}
-
-\paragraph{Beweis.} Definiere $u$ und $v$ durch:
-\[v(\mathrm{d}b\otimes c)= c\mathrm{d}\psi(b),\quad u(c\mathrm{d}c') = c\mathrm{d}c',\qquad b\in B,\ c,c'\in C \]
-Äquivalent zu (i) bzw. (ii) ist die Exaktheit der $4$-Term- bzw. $5$-Termsequenzen:
-\[0\longrightarrow \operatorname{Hom}_C(\Omega_{C/B},M)\longrightarrow \operatorname{Hom}_C(\Omega_{C/A},M)\longrightarrow \operatorname{Hom}_C(\Omega_{B/A}\otimes_B C,M)\dashrightarrow 0  \]
-Da $\operatorname{Hom}_C(\Omega_{B/A}\otimes_B C,M)\cong\operatorname{Hom}_B(\Omega_{B/A},M)$, ist nach \hyperref[8.4]{Satz 8.4} ist die Exaktheit nachzuweisen von:
-\[0\longrightarrow \operatorname{Der}_B(C,M)\longrightarrow \operatorname{Der}_A(C,M)\longrightarrow \operatorname{Der}_A(B,M)\dashrightarrow 0 \]
-Die vorletzte Abbildung ist gegeben durch $(\delta:C\to M)\mapsto (\delta\circ\psi: B\to M)$.
-Ohne die $0$ rechts ist diese Folge per Definition exakt.
-Mit der $0$ rechts ist die Folge genau dann exakt, wenn die Forderung in (ii) gilt.\qed
-
-\paragraph{Beispiel 8.8.}\label{8.8} Sei $B$ eine $A$"~Algebra und $C=B[x_1,\ldots,x_n]$.
-Sei $M$ ein beliebiger $C$"~Mo\-dul und $D\in\operatorname{Der}_A(B,M)$.
-Sei $\widetilde{D}:C\to M$ die einem Polynom $F$, den Polynom $F^D$ zuordnet, wobei $F^D$ das Polynom ist, das man durch Anwenden von $D$ auf die Koeffizienten von $F$ erhält.
-Dann ist $\widetilde{D}$ eine Derivation mit $\widetilde{D}|_B=D$.
-Nach \hyperref[8.7]{Satz 8.7 (ii)} gilt:
-\[\Omega_{C/A}\cong (\Omega_{B/A}\otimes_BC)\oplus C\mathrm{d}x_1\oplus\cdots\oplus C\mathrm{d}x_n \]
-
-\paragraph{Satz 8.9.}\label{8.9}\textit{(Zweite fundamentale exakte Sequenz)}\index{Fundamentale exakte Sequenz} Sei $B$ eine $A$"~Algebra, $I$ ein Ideal in $B$ und $C=B/I$.
-Es gibt eine kanonische, exakte Folge von $C$"~Moduln:
-\[\begin{tikzcd}
-I/I^2\rar["\delta"] & \Omega_{B/A}\otimes_BC\rar["v"] & \Omega_{C/A}\rar & 0
-\end{tikzcd} \]
-wobei $v(\mathrm{d}b\otimes c)= c\mathrm{d}\overline{b}$ und $\delta$ die durch die Abbildung $I\to\Omega_{B/A}\otimes_BC,\ x\mapsto \mathrm{d}x\otimes 1$ induzierte $C$"~lineare Abbildung ist.
-
-\paragraph{Beweis.} Für $x,x'\in I$ gilt:
-\[\mathrm{d}(xx')\otimes 1 = \mathrm{d}x\otimes x' + \mathrm{d}x'\otimes x =0 \]
-Daher ist $\delta$ wohldefiniert. Nun ist für alle $C$"~Moduln $M$ die folgenden Folgen exakt:
-\[\begin{tikzcd}
-0 \rar & \operatorname{Hom}_C(\Omega_{C/A},M)\dar["\cong"] \rar & \operatorname{Hom}_C(\Omega_{B/A}\otimes_BC,M)\dar["\cong"] \rar & \operatorname{Hom}_C(I/I^2,M)\dar["\cong"]\\
-0 \rar & \operatorname{Der}_A(B/I,M) \rar & \operatorname{Der}_A(B,M) \rar["\varphi"'] & \operatorname{Hom}_B(I,M)
-\end{tikzcd} \]
-wobei $\varphi(D:B\to M) = (D|_I:I\to M)$.\qed
-
-\paragraph{Korollar 8.10.}\label{8.10} Ist $B$ eine endlich erzeugte $A$"~Algebra oder Lokalisierung einer endlich erzeugten $A$"~Algebra, so ist $\Omega_{B/A}$ endlich erzeugter $B$"~Modul.
-
-\paragraph{Beweis.} Es ist $B \cong A[x_1,\ldots,x_n]/I$ für ein Ideal $I$.
-Wegen \hyperref[8.5]{Beispiel 8.5} ist $\Omega_{A[x_1,\ldots,x_n]/A}$ ein endlich erzeugter $A[x_1,\ldots,x_n]$"~Modul.
-Nach \hyperref[8.9]{Satz 8.9} ist $\Omega_{B/A}$ ein endlich erzeugter $B$"~Modul.
-Die Aussage über die Lokalisierung folgt aus \hyperref[8.6]{Satz 8.6 (i)}.\qed
-
-\paragraph{Definition 8.11.}\label{8.11} Eine Körpererweiterung $K/k$ heißt \textit{separabel erzeugt}\index{separabel erzeugt}, falls es eine Transzendenzbasis $\{x_\lambda\}_{\lambda\in\Lambda}$ für $K/k$ gibt, so dass $K/k[x_\lambda\mid \lambda\in\Lambda]$ separabel ist.
-
-\paragraph{Satz 8.12.}\label{8.12} Sei $K/k$ eine endlich erzeugte Körpererweiterung. Dann gilt:
-\begin{enumerate}[(i)]
-\item $\dim_K\Omega_{K/k}\geq \operatorname{trdeg}(K/k)$
-\item Gleichheit in (i) gilt genau dann, wenn $K/k$ separabel erzeugt ist.
-\end{enumerate}
-
-\paragraph{Korollar 8.13.}\label{8.13} Sei $K/k$ eine endliche Erweiterung. Dann gilt:
-\[\Omega_{K/k}=0\iff K/k\text{ ist separabel} \]
-
-\paragraph{Satz 8.14.}\label{8.14} Sei $B$ ein lokaler Ring mit Maximalideal $\mathfrak{m}$ und Restklassenkörper $k$.
-Ferner gebe es einen Schnitt von $B\twoheadrightarrow k$, also ist $B$ eine $k$"~Algebra.
-Dann ist die Abbildung $\delta$ in der \hyperref[8.9]{zweiten fundamentalen Sequenz} ein Isomorphismus:
-\[\delta:\mathfrak{m}/\mathfrak{m}^2\to \Omega_{B/k}\otimes_B k \]
-
-\paragraph{Beweis.} Es ist $\operatorname{coker}(\delta)=\Omega_{k/k}=0$. Die Injektivität von $\delta$ ist äquivalent zur Surjektivität von:
-\[\delta^\ast :\operatorname{Hom}_k(\Omega_{B/k}\otimes_B k,k)\to\operatorname{Hom}_k(\mathfrak{m}/\mathfrak{m}^2,k) \]
-Nun ist $\operatorname{Hom}_k(\Omega_{B/k}\otimes_B k,k) = \operatorname{Hom}_B(\Omega_{B/k},k)=\operatorname{Der}_k(B,k)$.
-Die Abbildung $\operatorname{Der}_k(B,k)\to\operatorname{Hom}_k(\mathfrak{m}/\mathfrak{m}^2,k),\ (d:B\to k)\mapsto (d|_\mathfrak{m}:\mathfrak{m}\to k)$ kommutiert mit $\delta^\ast$.
-Beachte, dass für $d\in\operatorname{Der}_k(B,k)$ stets $d(\mathfrak{m}^2)=0\in k$ gilt.
-
-Sei nun $h\in\operatorname{Hom}_k(\mathfrak{m}/\mathfrak{m}^2,k)$ und $b\in B$.
-Wegen des Schnitts $k\to B$, hat $b$ eine eindeutige Darstellung $b=\lambda+c,\ \lambda\in k,\ c\in\mathfrak{m}$.
-Setze $d:B\to k,\ d(b) =h(\overline{c})$.
-Dann ist $d(k)=0$, $d$ additiv und es gilt:
-\begin{align*}
-d(bb') &= d(\lambda\lambda'+\lambda c'+\lambda' c+cc')\\
-&= h(\overline{\lambda c'+\lambda' c})\\
-&= \lambda h(\overline{c'}) + \lambda' h(\overline{c})\\
-&= \lambda d(b')+\lambda' d(b)= bd(b')+b'd(b)\in k
-\end{align*}
-für alle $b=\lambda+c,\ b'=\lambda'+c'\in B$. Es gilt $\delta^\ast (d)=h$.\qed
-
-\paragraph{Satz 8.15.}\label{8.15} Sei $B$ ein lokaler Ring und es gebe einen Schnitt von $B\to k$.
-Ferner sei der Restklassenkörper $k$ vollkommen.
-Sei $B$ die Lokalisierung einer endlich erzeugten $k$"~Algebra.
-Dann ist $\Omega_{B/k}$ genau dann ein freier $B$"~Modul vom Rang $\dim(B)$, wenn $B$ regulär ist.
-
-\paragraph{Beweis.} Sei $\Omega_{B/k}$ frei vom Rang $\dim(B)$. Nach \hyperref[8.14]{Satz 8.14} ist $\dim_k\mathfrak{m}/\mathfrak{m}^2=\dim(B)$, also ist $B$ per Definition regulär.
-
-Sei umgekehrt $B$ regulär von Dimension $r$. Nach \hyperref[8.14]{Satz 8.14} gilt:
-\[r =\dim_k\mathfrak{m}/\mathfrak{m}^2=\dim_k\Omega_{B/k}\otimes_B k \]
-Sei $K$ der Quotientenkörper von $B$.
-Nach \hyperref[8.6]{Satz 8.6} ist $\Omega_{B/k}\otimes_B K=\Omega_{K/k}$.
-Nun ist $K/k$ separabel erzeugt, da $k$ vollkommen (siehe z.B.
-Matsumura Chapter 10, Corollary p.
-194).
-Nach \hyperref[8.12]{Satz 8.12} gilt:
-\[\dim_K\Omega_{K/k}=\operatorname{trdeg}(K/k)=\dim(B) = r \]
-Nach \hyperref[8.10]{Korollar 8.10} ist $\Omega_{B/k}$ endlich erzeugter $B$"~Modul.\qed
-
-\paragraph{Lemma 8.16.}\label{8.16} Sei $A$ ein noetherscher, lokaler, nullteilerfreier Ring mit Restklassenkörper $k$ und Quotientenkörper $K$. Sei $M$ ein endlich erzeugter $A$"~Modul mit:
-\[\dim_k(M\otimes_Ak)=\dim_K(M\otimes_AK)=r \]
-Dann ist $M$ frei vom Rang $r$.
-
-\paragraph{Beweis.} Wegen $\dim_k(M\otimes_Ak)=r$, gibt es nach Nakayama eine Surjektion $\varphi:A^r\to M$. Setze $N=\ker(\varphi)$. Da $K$ flach über $A$ ist, haben wir eine kurze exakte Folge:
-\[0\longrightarrow N\otimes_AK\longrightarrow K^r\longrightarrow M\otimes_AK\longrightarrow 0 \]
-Da $\dim_K(M\otimes_AK)=\dim_KK^r$, folgt $N\otimes_AK=0$, also $N=0$, da $N\subset A^r$ und $A$ torsionsfrei ist.\qed
-
-\subsection{Differentialgarben}
-
-Sei $f:X\to Y$ ein Morphismus von Schemata und $\Delta:X\to X\times_YX$ die Diagonale.
-$\Delta$ ist nach \hyperref[3.30]{Satz 3.30} eine Immersion und liefert ein Isomorphismus auf $\Delta(X)$, wobei $\Delta(X)$ ein abgeschlossenes Unterschema eines offenen Unterschema $W\subset X\times_YX$ ist.
-
-\paragraph{Definition 8.17.}\label{8.17} Sei $\mathcal{J}$ die Idealgarbe von $\Delta(X)$ in $W$. Dann heißt
-\[\Omega_{X/Y}=\Delta^\ast(\mathcal{J}/\mathcal{J}^2) \]
-die Garbe der \textit{relativen Differentiale}\index{Differentialform!relativ} von $X$ über $Y$.
-
-\paragraph{Bemerkung 8.18.}\label{8.18}\begin{enumerate}[(i)]
-\item $\mathcal{J}/\mathcal{J}^2$ ist ein $\mathcal{O}_{\Delta(X)}$"~Modul und somit ist über den Isomorphismus $X\to\Delta(X)$ die Modulgarbe $\Omega_{X/Y}$ ein $\mathcal{O}_X$"~Modul.
-Wegen \hyperref[5.13]{Satz 5.13 (i)} ist $\Omega_{X/Y}$ quasikohärent.
-Ist $Y$ noethersch und $f:X\to Y$ von endlichem Typ, und somit $X$ und $X\times_YX$ noethersch, dann ist $\Omega_{X/Y}$ kohärent.
-\item Sei $U=\operatorname{Spec}(A)\subset_\text{o}Y$ und $V=\operatorname{Spec}(B)\subset_\text{o}X$ mit $f(V)\subset U$.
-Dann ist $V\times_UV$ affin offen in $X\times_YX$ mit $V\times_UV= \operatorname{Spec}(B\otimes_AB)$.
-Es ist
-\[\Delta(X)\cap V\times_UV = \operatorname{Spec}(B\otimes_AB/\ker(B\otimes_AB\to B)) \]
-ein abgeschlossenes Unterschema von $V\times_UV$. Sei $I=\ker(B\otimes_AB\to B)$, also gilt:
-\[\mathcal{J}/\mathcal{J}^2|_{\Delta(X)\cap V\times_UV} = \widetilde{I/I^2} \]
-Nach \hyperref[8.3]{Satz 8.3} folgt $\Omega_{V/U}=\widetilde{\Omega_{B/A}}$.
-Somit ist \hyperref[8.17]{Definition 8.17} verträglich mit dem affinen Fall.
-Ist $X=\bigcup V$ eine Überdeckung durch affin offenen $V$ und $Y=\bigcup U$ eine Überdeckung durch affin offenen $U$, so liefert die Verklebung von $\widetilde{\Omega_{B/A}},\ \mathrm{d}:B\to\Omega_{B/A}$ gerade $\Omega_{X/Y},\ \mathrm{d}:\mathcal{O}_X\to\Omega_{X/Y}$.
-$\mathrm{d}$ ist ein Garbenmorphismus und ist in jedem Punkt eine Derivation von lokalen Ringen.
-\end{enumerate}
-
-\paragraph{Satz 8.19.}\label{8.19} Sei $f:X\to Y$ ein Morphismus und $g:Y'\to Y$ ein weiterer Morphismus. Betrachte den Basiswechsel:
-\[\begin{tikzcd}
-X\times_YY' \rar["f'"]\dar["g'"'] & Y'\dar["g"]\\
-X\rar["f"] & Y
-\end{tikzcd} \]
-Dann gilt $\Omega_{X'/Y'}\cong (g')^\ast(\Omega_{X/Y})$.
-
-\paragraph{Beweis.} Folgt aus \hyperref[8.18]{Bemerkung 8.18 (ii)} und \hyperref[8.6]{Satz 8.6 (ii)}.\qed
-
-\paragraph{Satz 8.20.}\label{8.20} Seien $f:X\to Y$ und $g:Y\to Z$ Morphismen. Dann gibt es eine exakte Garbensequenz auf $X$:
-\[f^\ast\Omega_{Y/Z}\longrightarrow\Omega_{X/Z}\longrightarrow\Omega_{X/Y}\longrightarrow 0 \]
-
-\paragraph{Beweis.} Es reicht, die Exaktheit lokal zu überprüfen. Dies folgt aus der \hyperref[8.7]{ersten fundamentalen exakten Sequenz}.\qed
-
-\paragraph{Satz 8.21.}\label{8.21} Sei $f:X\to Y$ ein Morphismus und $Z$ ein abgeschlossenes Unterschema von $X$ zum Ideal $\mathcal{J}$. Dann gibt es eine exakte Sequenz von Garben auf $Z$:
-\[\mathcal{J}/\mathcal{J}^2 \stackrel{\delta}{\longrightarrow} \Omega_{X/Y}\otimes_{\mathcal{O}_X}\mathcal{O}_Z\longrightarrow\Omega_{Z/Y}\longrightarrow 0 \]
-
-\paragraph{Beweis.} Folgt aus der \hyperref[8.9]{zweiten fundamentalen exakten Sequenz}.\qed
-
-\paragraph{Satz 8.22.}\label{8.22} Sei $X=\mathbf{A}_Y^n$.
-Dann ist $\Omega_{X/Y}$ ein freier $\mathcal{O}_X$"~Modul vom Rang $n$, erzeugt von globalen Schnitten $\mathrm{d}x_1,\ldots,\mathrm{d}x_n$, wobei $x_1,\ldots,x_n$ die affinen Koordinaten von $\mathbf{A}_Y^n$ sind.
-
-\paragraph{Beweis.} Folgt aus \hyperref[8.5]{Beispiel 8.5}.\qed
-
-\paragraph{Satz 8.23.}\label{8.23} Sei $A$ ein Ring und $Y=\operatorname{Spec}(A)$. Sei $X=\mathbf{P}_A^n$. Dann gibt es eine exakte Sequenz von Garben auf $X$:
-\[0\longrightarrow\Omega_{X/Y}\longrightarrow \bigoplus_{i=0}^n\mathcal{O}_X(-1)\longrightarrow \mathcal{O}_X\longrightarrow 0 \]
-
-\paragraph{Beweis.} Sei $S=A[x_0,\ldots,x_n]$ der homogene Koordinatenring von $X$ und setze:
-\[E=\bigoplus_{i=0}^n S(-1)e_i\]
-als den graduierten $S$"~Modul mit Basis $e_0,\ldots,e_n$ im Grad $1$.
-Betrachte den $S$"~Mo\-dul\-ho\-mo\-mor\-phis\-mus $\varphi:E\to S,\ e_i\mapsto x_i$ im Grad $0$ und $M=\ker(\varphi)$.
-Wir erhalten eine exakte Garbensequenz:
-\[0\longrightarrow\widetilde{M}\longrightarrow\bigoplus_{i=0}^n\mathcal{O}_X(-1)\stackrel{\tilde{\varphi}}{\longrightarrow}\mathcal{O}_X\longrightarrow 0 \]
-$\varphi:E\to S$ ist in allen Graden größer gleich $1$ surjektiv, also ist $\tilde{\varphi}$ surjektiv. Wir zeigen nun $\widetilde{M}=\Omega_{X/Y}$.
-
-Lokalisierung nach $x_i$ liefert eine Surjektion von freien $S_{x_i}$"~Moduln $E_{x_i}\to S_{x_i}$.
-$M_{x_i}$ ist frei vom Rang $n$ mit Basis $\{e_j-\frac{x_j}{x_i}e_i\mid j\neq i\}$.
-Multiplikation mit $\frac{1}{x_i}$ liefert Elemente vom Grad $0$ in $M_{x_i}$.
-Also ist $\widetilde{M}|_{D_+(x_i)}$ ein freier $\mathcal{O}_{D_+(x_i)}$"~Modul, der von den globalen Schnitten $\{\frac{1}{x_i}e_j-\frac{x_j}{x_i^2}e_i\mid j\neq i \}$ erzeugt wird.
-Definiere $\varphi_i:\Omega_{X/Y}|_{D_+(x_i)}\to\widetilde{M}|_{D_+(x_i)}$ wie folgt:
-
-Es ist $D_+(x_i)=\operatorname{Spec}A[\frac{x_0}{x_i},\ldots,\widehat{\frac{x_i}{x_i}},\ldots,\frac{x_n}{x_i}]$. Nach \hyperref[8.22]{Satz 8.22} gilt:
-\[\Omega_{X/Y}|_{D_+(x_i)}=\bigoplus_{j\neq i}\mathcal{O}_{D_+(x_i)}\mathrm{d}\frac{x_j}{x_i} \]
-Setze nun $\varphi_i(\mathrm{d}\frac{x_j}{x_i})=\frac{1}{x_i^2}(x_ie_j-x_je_i)$.
-$\varphi_i$ ist ein Isomorphismus.
-Die $\varphi_i$ lassen sich nun zu einem Isomorphismus $\varphi:\Omega_{X/Y}\to\widetilde{M}$ verkleben.\qed
-
-\subsection{Nicht-singuläre Varietäten}
-
-\paragraph{Definition 8.24.}\label{8.24} Eine Varietät $X$ über einem algebraisch abgeschlossenen Körper $k$, d.h.
-ein integres, separiertes Schema von endlichem Typ, heißt \textit{nicht-singulär}\index{nicht-singulär}, wenn alle lokalen Ringe regulär sind.
-
-\paragraph{Satz 8.25.}\label{8.25} Sei $X$ ein irreduzibles, separiertes Schema von endlichem Typ über einem algebraisch abgeschlossenen Körper $k$.
-Dann ist $\Omega_{X/k}$ genau dann lokal frei vom Rang $n=\dim(X)$, wenn $X$ eine nicht-singuläre Varietät über $k$ ist.
-
-\iffalse
-\paragraph{Beweis.} Sei $x\in X$ ein abgeschlossener Punkt.
-Dann ist $\dim(\mathcal{O}_{X,x})=n$ und $\kappa(x)=k$.
-Es ist $B=\mathcal{O}_{X,x}$ eine endlich erzeugte $k$"~Algebra und $\Omega_{B/k}=(\Omega_{X/k})_x$.
-Nach \hyperref[8.15]{Satz 8.15} ist $(\Omega_{X/k})_x$ genau dann frei vom Rang $n$, wenn $\mathcal{O}_{X,x}$ regulär ist.
-Somit haben wir mit \hyperref[8.26]{Lemma 8.26} folgende Äquivalenzen:
-\begin{align*}
-& \text{Alle lokalen Ringe sind regulär}\\
-\iff & \text{Alle lokalen Ringe von abgeschlossenen Punkten sind regulär}\\
-\iff & (\Omega_{X/k})_x \text{ ist frei vom Rang $n$ für alle abgeschlossene $x\in X$}\\
-\iff & \Omega_{X/k}\text{ ist lokal frei vom Rang $n$}\qedhere
-\end{align*}
-\fi
-
-\paragraph{Lemma 8.26.}\label{8.26} Sei $X$ ein noethersches Schema und $\mathcal{F}$ eine kohärente Garbe.
-Dann ist $\mathcal{F}$ genau dann lokal frei, wenn $\mathcal{F}_x$ freier $\mathcal{O}_{X,x}$"~Modul für alle $x\in X$ ist.
-
-\paragraph{Korollar 8.27.}\label{8.27} Sei $X$ eine nicht-singuläre Varietät über $k$ und $Y\subset X$ ein irreduzibles, abgeschlossenes Unterschema zur Idealgarbe $\mathcal{J}$.
-Dann ist $Y$ genau dann nicht-singulär, wenn folgende Bedingungen erfüllt sind:
-\begin{enumerate}[(i)]
-\item $\Omega_{Y/k}$ ist lokal frei.
-\item In der exakten Sequenz aus \hyperref[8.21]{Satz 8.21} ist $\delta$ injektiv:
-\[0\longrightarrow\mathcal{J}/\mathcal{J}^2\stackrel{\delta}{\longrightarrow}\Omega_{X/k}\otimes_{\mathcal{O}_X}\mathcal{O}_Y\longrightarrow\Omega_{Y/k}\longrightarrow 0 \]
-\end{enumerate}
-
-\chapter{Kohomologie}
-
-Wir verwenden folgende Bezeichnungen für abelsche Kategorien:
-\begin{itemize}
-\item $\mathbf{Ab}$, die Kategorie der abelschen Gruppen
-\item $\mathbf{Mod}(A)$, die Kategorie der Moduln über einem kommutativen Ring $A$ mit Eins
-\item $\mathbf{Ab}(X)$, die Kategorie der Garben abelscher Gruppen über einem topologischen Raum $X$
-\item $\mathbf{Mod}(X)$, die Kategorie der $\mathcal{O}_X$"~Moduln auf einem geringten Raum $(X,\mathcal{O}_X)$
-\item $\mathbf{Qcoh}(X)$, die Kategorie der quasikohärenten $\mathcal{O}_X$"~Moduln auf einem Schema $X$
-\item $\mathbf{Coh}(X)$, die Kategorie der kohärenten $\mathcal{O}_X$"~Modulgarben auf einem noetherschen Schema $X$
-\end{itemize}
-
-\section{Kohomologie von Garben}
-
-\paragraph{Satz 9.15.}\label{9.15} Sei $A$ ein Ring. Dann ist jeder $A$"~Modul isomorph zu einem Untermodul eines injektiven $A$"~Moduls.
-
-\paragraph{Satz 9.16.}\label{9.16} Sei $(X,\mathcal{O}_X)$ ein geringter Raum. Dann hat die Kategorie $\mathbf{Mod}(\mathcal{O}_X)$ genügend viele Injektive.
-
-\paragraph{Beweis.} Sei $\mathcal{F}\in\mathbf{Mod}(\mathcal{O}_X)$.
-Dann ist $\mathcal{F}_x\in\mathbf{Mod}(\mathcal{O}_{X,x})$.
-Nach \hyperref[9.15]{Satz 9.15} gibt es ein Monomorphismus $\mathcal{F}_x\to I_x$ mit einem injektiven $\mathcal{O}_{X,x}$"~Modul $I_x$.
-Sei $j:\{x\}\hookrightarrow X$ die kanonische Inklusion und betrachte die $I_x$ als Garben auf $\{x\}$.
-Setze:
-\[\mathcal{J}=\prod_{x\in X}j_\ast I_x \]
-Dann gilt:
-\[\operatorname{Hom}_{\mathcal{O}_X}(\mathcal{F},\mathcal{J}) = \prod_{x\in X}\operatorname{Hom}_{\mathcal{O}_X}(\mathcal{F},j_\ast I_x) = \prod_{x\in X}\operatorname{Hom}_{\mathcal{O}_{X,x}}(\mathcal{F}_x,I_x) \]
-Die $\mathcal{F}_x\hookrightarrow I_x$ geben ein Monomorphismus $\mathcal{F}\hookrightarrow \mathcal{J}$. Ferner gilt:
-\[\operatorname{Hom}_{\mathcal{O}_X}(-,\mathcal{J})=\prod_{x\in X}\operatorname{Hom}_{\mathcal{O}_{X,x}}((-)_x, I_x) \]
-Da Halmbildung exakt und $I_x$ injektive Objekte sind, ist dies ein exakter Funktor. Somit ist $\mathcal{J}$ ein injektiver $\mathcal{O}_X$"~Modul.\qed
-
-\paragraph{Korollar 9.17.}\label{9.17} Sei $X$ ein topologischer Raum. Dann besitzt $\mathbf{Ab}(X)$ genügend viele Injektive.
-
-\paragraph{Beweis.} Sei $\mathcal{O}_X$ die konstante Garbe $\mathbb{Z}$ auf $X$. Dann ist $(X,\mathcal{O}_X)$ ein geringter Raum und es gilt $\mathbf{Mod}(\mathcal{O}_X)=\mathbf{Ab}(X)$.\qed
-
-\paragraph{Definition 9.18.}\label{9.18} Sei $X$ ein topologischer Raum und $\Gamma(X,-):\mathbf{Ab}(X)\to \mathbf{Ab}$ der globale Schnittfunktor. Wir definieren:
-\[\mathrm{H}^i(X,-)= \mathrm{R}^i\Gamma(X,-) \]
-$\mathrm{H}^i(X,\mathcal{F})$ heißt \textit{$i$-te Kohomologiegruppe}\index{Kohomologiegruppe} von der Garbe $\mathcal{F}$.
-Haben $X$ und $\mathcal{F}$ Zusatzstrukturen, z.B.
-$X$ Schema oder $\mathcal{F}\in \mathbf{Qcoh}(X)$, so wird die Kohomologie trotzdem immer im obigen Sinne verstanden, d.h.
-$\mathcal{F}\in\mathbf{Ab}(X)$.
-
-Sei $(X,\mathcal{O}_X)$ ein geringter Raum und $\mathcal{F}$ ein $\mathcal{O}_X$"~Modul.
-$\mathcal{F}$ heißt\textit{welk}\index{$\mathcal{O}_X$-Modul!welk}, falls für alle offenen $V\subset U$ die Restriktionsabbildung $\mathcal{F}(U)\to\mathcal{F}(V)$ surjektiv ist.
-
-\paragraph{Lemma 9.19.}\label{9.19} Ist $(X,\mathcal{O}_X)$ ein geringter Raum, so ist jeder injektiver $\mathcal{O}_X$"~Modul $\mathcal{J}$ welk.
-
-\paragraph{Beweis.} Seien $i:U\hookrightarrow X$ und $j:V\hookrightarrow X$ die natürlichen Inklusionen.
-Betrachte die Inklusion $j_!\mathcal{O}_X|_V\hookrightarrow i_!\mathcal{O}_X|_U$ von $\mathcal{O}_X$"~Moduln.
-Es folgt die Surjektivität von:
-\[\mathcal{J}(U) = \operatorname{Hom}(i_!\mathcal{O}_X|_U,\mathcal{J})\to \operatorname{Hom}(j_!\mathcal{O}_X|_V,\mathcal{J})=\mathcal{J}(V)\qedhere \]
-
-\paragraph{Satz 9.20.}\label{9.20} Sei $X$ ein topologischer Raum und $\mathcal{F}$ eine welke Garbe auf $X$. Dann ist $\mathrm{H}^i(X,\mathcal{F})=0$ für alle $i>0$.
+\input{chapter1.tex}
+\input{chapter2.tex}
+\input{chapter3.tex}
 

### 2015-05-29 Algebraische Geometrie/chapter1.tex
Neue Dateistruktur

\chapter{Varietäten}

Sei $k$ ein algebraisch abgeschlossener Körper.

\section{Affine Varietäten}

\paragraph{Definition.} \begin{enumerate}[(i)]
\item Die Menge aller $n$-Tupeln über $k$
\[\mathbf{A}^n = \mathbf{A}_k^n = \{(a_1,\ldots,a_n)\mid a_i\in k\} \]
heißt \textit{affiner $n$-dimensionaler Raum}\index{affiner Raum} über $k$. Ein Element $P=(a_1,\ldots,a_n)\in\mathbf{A}^n$ heißt \textit{Punkt}\index{Punkt} und die $a_i$ heißen \textit{Koordinaten}\index{Koordinate} von $P$.
\item Der Polynomring über $k$ in $n$ Variablen bezeichnen wir mit $A=k[X_1,\ldots,X_n]$. Für $T\subset A$ definieren wir die \textit{Nullstellenmenge}\index{Nullstellenmenge} von $T$ wie folgt:
\[Z(T)=\{P\in\mathbf{A}^n\mid \forall f\in T\colon f(P)=0\} \]
Es gilt $Z(T)=Z(\mathfrak{a})$, wobei $\mathfrak{a}$ das von $T$ erzeugte Ideal in $A$ ist.
\item Eine Teilmenge $Y\subset\mathbf{A}^n$ der Form $Y=Z(T)$ für ein $T\subset A$ heißt \textit{affine algebraische Menge}\index{algebraische Menge!affin}. Für algebraische Mengen $Y_i=Z(\mathfrak{a}_i)$ mit Ideale $\mathfrak{a}_i\subset A,\ i\in I$ gilt:
\[Y_1\cup Y_2=Z(\mathfrak{a}_1\cap\mathfrak{a}_2),\quad \bigcap_{i\in I}Y_i = Z\Big(\sum_{i\in I}\mathfrak{a}_i\Big) \]
Ferner gilt $\mathbf{A}^n=Z(0)$ und $\varnothing = Z(1)$. Wir statten $\mathbf{A}^n$ mit der sogenannten \textit{Zariski-Topologie}\index{Zariski-Topologie} aus, in dem wir eine Menge $U\subset\mathbf{A}^n$ genau dann offen nennen, wenn $\mathbf{A}^n\setminus U$ eine algebraische Menge ist.
\item Eine \textit{affine Varietät}\index{Varietät!affin} $V$ ist eine \textit{irreduzible}\index{irreduzibel} abgeschlossene Menge in $\mathbf{A}^n$, d.h. aus $V=V_1\cup V_2$ mit abgeschlossenen Mengen $V_1,V_2\subset\mathbf{A}^n$ folgt $V_1=\varnothing$ oder $V_2=\varnothing$.

Eine offene Teilmenge einer affinen Varietät bzgl. der induzierten Topologie nennt man \textit{quasi-affine Varietät}\index{Varietät!quasi-affin}.
\item Sei $Y\subset\mathbf{A}^n$ eine algebraische Menge. Dann definieren wir das Ideal:
\[I(Y)=\{f\in A\mid \forall P\in Y\colon f(P)=0\} \]
Der \textit{Koordinatenring}\index{Koordinatenring} von $Y$ ist definiert als $A(Y)=A/I(Y)$.
\end{enumerate}

\paragraph{Definition.} Sei $\mathfrak{a}\subset A$ ein Ideal. Das \textit{Radikal}\index{Radikal} von $\mathfrak{a}$ ist definiert als:
\[\operatorname{Rad}(\mathfrak{a})=\{f\in A\mid\exists r>0\colon f^r\in\mathfrak{a} \} \]
Ein Ideal $\mathfrak{a}\subset A$ heißt \textit{Radikalideal}\index{Radikalideal}, wenn $\mathfrak{a}=\operatorname{Rad}(\mathfrak{a})$ gilt.

\paragraph{Satz.} Es gibt eine inklusionsumkehrende Bijektion:
\[\{\text{Algebraische Mengen in }\mathbf{A}^n\} \to \{\text{Radikalideale in }k[X_1,\ldots,X_n]\},\ Y\mapsto I(Y) \]
mit der Umkehrabbildung $\mathfrak{a}\mapsto Z(\mathfrak{a})$. Eine algebraische Menge $Y$ ist genau dann irreduzibel, wenn $I(Y)\subset A$ ein Primideal ist.

\section{Projektive Varietäten}

\paragraph{Definition.}\begin{enumerate}[(i)]
\item Zwei Punkte $(a_0,\ldots,a_n),(b_0,\ldots,b_n)\in \mathbf{A}^{n+1}$ heißen äquivalent, wenn ein $\lambda\in k^\times$ existiert, so dass $a_i=\lambda b_i$ für alle $i$ gilt. Die Äquivalenzklasse von $(a_0,\ldots,a_n)$ wird mit $(a_0\: \ldots\: a_n)$ bezeichnet. Der \textit{projektiver $n$-dimensionaler Raum}\index{projektiver Raum} über $k$ wird definiert als:
\[\mathbf{P}^n=\mathbf{P}_k^n = \{(a_0\: \ldots\: a_n) \mid a_i\in k\text{ nicht alle $0$} \} \]
Ein Element $P=(a_0\: \ldots\: a_n)\in\mathbf{P}^n$ heißt \textit{Punkt}\index{Punkt} und die $a_i$ heißen \textit{homogene Koordinaten} von $P$.\index{Koordinate!homogen}
\item Der Polynomring über $k$ in $n+1$ Variablen $S=k[X_0,\ldots,X_n]$ wird mit der folgenden Zerlegung zu einem graduierten Ring:
\[S=\bigoplus_{d\geq 0}S_d,\quad S_d=\Big\{ \sum a_{i_0,\ldots,i_n} X_0^{i_0}\cdots X_n^{i_n}\Bigm| a_{i_0,\ldots,i_n}\in k,\ \sum_{j=0}^n i_j=d \Big\} \]
Die Elemente in $S_d$ heißen \textit{homogene Elemente vom Grad $d$}\index{homogene Elemente}. Ein Ideal $\mathfrak{a}\subset S$ heißt \textit{homogenes Ideal}\index{homogenes Ideal}, wenn $\mathfrak{a}=\bigoplus_{d\geq 0}(S_d\cap\mathfrak{a})$ gilt.
\item Sei $\mathfrak{a}\subset S$ ein homogenes Ideal. Dann setzen wir:
\[Z(\mathfrak{a}) = \{P\in\mathbf{P}^n\mid \forall f\in\mathfrak{a}\text{ homogen}\colon f(P)=0 \} \]
Diese ist wohldefiniert, da $f(\lambda a_0,\ldots,\lambda a_n)=\lambda^d f(a_0,\ldots,a_n)$ für $f\in S_d$. Eine Menge $Y\subset\mathbf{P}^n$ heißt \textit{projektive algebraische Menge}\index{algebraische Menge!projektiv}, wenn $Y=Z(\mathfrak{a})$ für ein homogenes Ideal $\mathfrak{a}\subset S$ gilt.

Analog wie im affinen Fall, können wir auch $\mathbf{P}^n$ mit der \textit{Zariski-Topologie} ausstatten, d.h. eine Menge $U\subset\mathbf{P}^n$ ist genau dann offen, wenn $\mathbf{P}^n\setminus U$ eine projektive algebraische Menge ist.\index{Zariski-Topologie}
\item Eine \textit{projektive Varietät}\index{Varietät!projektiv} $V$ ist eine irreduzible abgeschlossene Menge in $\mathbf{P}^n$. Eine offene Teilmenge einer projektiven Varietät bzgl. der induzierten Topologie nennt man \textit{quasi-projektive Varietät}\index{Varietät!quasi-projektiv}.
\item Sei $Y\subset\mathbf{P}^n$ eine algebraische Menge. Dann setzen wir $I(Y)$ als das Ideal in $S$, das von der folgenden Menge erzeugt wird:
\[\{f\in S \text{ homogen}\mid \forall P\in Y\colon f(P)=0\} \]
$I(Y)$ ist ein homogenes Ideal in $S$. Der \textit{homogene Koordinatenring}\index{Koordinatenring!homogen} von $Y$ ist definiert als $S(Y)=S/I(Y)$.
\end{enumerate}

\paragraph{Satz.} Wir haben eine inklusionsumkehrende Bijektion:
\[\{\text{Algebraische Mengen in }\mathbf{P}^n\}\to\{\text{Radikalideale in }k[X_0,\ldots,X_n]\},\ Y\mapsto I(Y) \]
mit der Umkehrabbildung $\mathfrak{a}\mapsto Z(\mathfrak{a})$.

\paragraph{Satz.} Sei $Y$ eine (quasi-)projektive Varietät. Dann wird $Y$ von offenen Mengen der Form $Y\cap U_i,\ i=0,\ldots,n$ überdeckt mit:
\[U_i=\{(a_0\: \ldots\: a_n)\in \mathbf{P}^n\mid a_i\neq 0\} \]
Die Abbildungen $\varphi_i: U_i\to\mathbf{A}^n,\ (a_0\: \ldots\: a_n)\mapsto \big(\frac{a_0}{a_i},\ldots,\widehat{\frac{a_i}{a_i}},\ldots,\frac{a_n}{a_i}\big)$ sind wohldefiniert und Ho\-möo\-mor\-phis\-men, d.h. die $Y\cap U_i$ sind (quasi-)affine Varietäten.

\section{Morphismen}

\paragraph{Definition.} \begin{enumerate}[(i)]
\item Sei $Y\subset\mathbf{A}^n$ eine quasi-affine Varietät. Eine Abbildung $f:Y\to k$ heißt \textit{reguläre Funktion} in $P\in Y$, wenn eine offene Umgebung $U\subset Y$ mit $P\in U$ existiert, so dass $f=\frac{g}{h}$ auf $U$ für gewisse $g,h\in A$ gilt.

Sei $Y\subset\mathbf{P}^n$ eine quasi-projektive Varietät. Eine Abbildung $f:Y\to k$ heißt \textit{reguläre Funktion}\index{reguläre Funktion} in $P\in Y$, wenn eine offene Umgebung $U\subset Y$ mit $P\in U$ existiert, so dass $f=\frac{g}{h}$ auf $U$ für gewisse homogene Polynome $g,h\in S$ vom gleichen Grad.

Identifizieren wir $k\cong \mathbf{A}^1$ so ist eine reguläre Funktion notwendigerweise stetig.
\item Eine stetige Abbildung $\varphi:X\to Y$ zwischen zwei (quasi-projektive) Varietäten heißt \textit{Morphismus}\index{Morphismus!Varietäten}, wenn für jede offene Menge $V\subset Y$ und reguläre Funktion $f:V\to k$ auch $f\circ\varphi:\varphi^{-1}(V)\to k$ regulär ist.

Damit erhält man die Kategorie $\textbf{Var$(k)$}$ aller Varietäten auf $k$.
\end{enumerate}

%\paragraph{Bemerkung.} Sei $Y$ eine Varietät und $f,g:Y\to k$ reguläre Funktionen. Existiert eine nichtleere offene Menge $U\subset Y$ mit $f=g$ auf $U$, so folgt $f=g$ auf ganz $Y$: Sei dazu o.B.d.A. $U$ so klein, dass $f$ und $g$ als Brüche homogener Polynome vom gleichen Grad dargestellt werden kann. Dann ist die Nullstellenmenge $\{P\in Y\mid (f-g)(P)=0\}$ abgeschlossen und dicht in $Y$.

\paragraph{Definition.} Sei $Y$ eine Varietät und $P\in Y$ ein Punkt.
\begin{enumerate}[(i)]
\item Wir bezeichnen den Ring aller regulären Funktionen auf $Y$ mit $\mathcal{O}(Y)$.
\item $\mathcal{O}_{P,Y}=\{\langle U,f\rangle\mid P\in U\subset_\text{o} Y,\ f\text{ ist auf $U$ regulär}\}$ heißt der Ring der \textit{Keime} regulärer Funktionen auf $Y$ in $P$.\index{Keim} Wir identifizieren zwei Keime $\langle U,f\rangle=\langle V,g\rangle$, wenn $f=g$ auf $U\cap V$ gilt.

$\mathcal{O}_{P,Y}$ ist ein lokaler Ring, dessen Maximalideal wir mit $\mathfrak{m}_P$ bezeichnen.
\item $K(Y)=\{\langle U,f\rangle\mid\varnothing\neq U\subset_\text{o}Y,\ f\text{ ist auf $U$ regulär}\}$ heißt der \textit{Funktionenkörper}\index{Funktionenkörper}  von $Y$. Die Elemente von $K(Y)$ heißen \textit{rationale Funktionen} auf $Y$\index{rationale Funktion}.
\end{enumerate}
Es gilt $\mathcal{O}(Y)\subset\mathcal{O}_{P,Y}\subset K(Y)$.

\paragraph{Theorem.} Sei $Y\subset\mathbf{A}^n$ eine affine Varietät. Dann gilt:
\begin{enumerate}[(i)]
\item $\mathcal{O}(Y)\cong A(Y)$
\item Die Abbildung $Y\to \{\text{Maximale Ideale in }A(Y)\},\ P\mapsto \mathfrak{m}_P\subset\mathcal{O}_{P,Y}$ ist eine Bijektion.
\item $\mathcal{O}_{P,Y}\cong A(Y)_{\mathfrak{m}_P}$ und $\dim\mathcal{O}_{P,Y}=\dim Y$.
\item $K(Y)\cong\operatorname{Quot}(A(Y))$
\end{enumerate}

\paragraph{Theorem.} Sei $Y\subset\mathbf{P}^n$ eine projektive Varietät. Dann gilt:
\begin{enumerate}[(i)]
\item $\mathcal{O}(Y)=k$
\item $\mathcal{O}_{P,Y}\cong S(Y)_{(\mathfrak{m}_P)}$
\item $K(Y)\cong S(Y)_{((0))}$
\end{enumerate}

\paragraph{Theorem.} Sei $X$ eine beliebige Varietät und $Y$ eine affine Varietät. Dann haben wir eine Bijektion:
\[\operatorname{Mor}_{\textbf{Var}}(X,Y) \to \operatorname{Hom}_{\textbf{$k$-Alg}}(A(Y),\mathcal{O}(X)),\ f\mapsto(\varphi\mapsto \varphi\circ f) \]
Ist $X$ ebenfalls affin, so gilt $X\cong Y$, genau dann wenn $A(X)\cong A(Y)$. Der Funktor $\textbf{affine Var$(k)$}\to\textbf{nullteilerfreie $k$-Alg},\ X\mapsto A(X)$ ist eine pfeilumkehrende Äquivalenz von Kategorien.

### 2015-05-29 Algebraische Geometrie/chapter2.tex
Neue Dateistruktur

\chapter{Schemata}
\section{Garben}

\paragraph{Definition 1.1.}\label{1.1} Sei $X$ ein topologischer Raum.
\begin{enumerate}[(i)]
\item Die Menge aller offenen Teilmengen in $X$ bilden zusammen mit den natürlichen Inklusionen eine Kategorie $\textbf{Top$(X)$}$.
\item Eine \textit{Prägarbe}\index{Prägarbe} $F$ abelscher Gruppen ist nichts anderes als ein kontravarianter Funk\-tor $F:\textbf{Top$(X)$}\to\textbf{Ab}$ mit $F(\varnothing)=0$.
\end{enumerate}

\paragraph{Bemerkung.}\begin{enumerate}
\item Eine Prägarbe besteht also aus abelschen Gruppen $F(U),\ U\subset_\text{o}X$ und Homomorphismen abelscher Gruppen $\operatorname{res}_V^U:F(U)\to F(V)$ für alle offenen $V\subset U$, so dass $\operatorname{res}_U^U=\operatorname{id}_{F(U)}$. Für offene Mengen $W\subset V\subset U$ gelte ferner $\operatorname{res}_W^U=\operatorname{res}_W^V\circ\operatorname{res}_V^U$.
\item Ebenso können Prägarben in eine beliebige Kategorie gebildet werden, z.B. $\textbf{Ringe}$ und $\textbf{Mengen}$.
\item Die Elemente von $F(U)$ heißen \textit{Schnitte} von $F$ über $U$.\index{Schnitt} Manchmal schreiben wir auch $\Gamma(U,F)=F(U)$. Die $\operatorname{res}_V^U$ heißen \textit{Restriktionsabbildungen}\index{Restriktionsabbildung}. Wir schreiben auch $\operatorname{res}_V^U(s)=s|_V$.
\end{enumerate}

\paragraph{Definition 1.2.}\label{1.2} Eine Prägarbe $F$ auf einem topologischen Raum $X$ heißt \textit{Garbe}\index{Garbe}, falls die folgenden Diagramme exakt sind:
\[\begin{tikzcd}
0\ar[r] & F(U)\ar[r, "\operatorname{res}"] & \displaystyle\prod_i F(U_i)\ar[r, "\operatorname{res}", shift left] \ar[r, shift right] & \displaystyle\prod_{i,j}F(U_i\cap U_j)
\end{tikzcd} \]
für alle $U\subset_\text{o}X$ und jede offene Überdeckung $U=\bigcup_i U_i$, d.h:
\begin{enumerate}[(i)]
\item $s\mapsto (\operatorname{res}^U_{U_i}(s))_i$ ist injektiv, d.h. aus $s|_{U_i}=0$ für alle $i$ folgt $s=0$.
\item Sei $s_i\in F(U_i)$ für alle $i$ gegeben, so dass $s_i|_{U_i\cap U_j}=s_j|_{U_i\cap U_j}$ für alle $i,j$. Dann gibt es ein $s\in F(U)$ mit $s|_{U_i}=s_i$ für alle $i$.
\end{enumerate}

\paragraph{Definition 1.3.}\label{1.3} \begin{enumerate}[(i)]
\item Ein \textit{Morphismus}\index{Morphismus!Prägarben} $\varphi:F\to G$ von Prägarben auf $X$ ist ein Morphismus kontravarianter Funktoren, d.h. eine Kollektion von Morphismen $(\varphi(U))_{U\subset_\text{o}X}$, so dass für alle offenen Mengen $V\subset U$ folgendes Diagramm kommutativ ist:
\[\begin{tikzcd}
F(U)\ar[rr, "\varphi(U)"]\ar[d, "\operatorname{res}^U_V"'] && G(U)\ar[d, "\operatorname{res}^U_V"]\\
F(V)\ar[rr, "\varphi(V)"'] && G(V)
\end{tikzcd} \]
\item Ein \textit{Morphismus}\index{Morphismus!Garben} von Garben ist ein Morphismus von Prägarben. Die (Prä-)Garben bilden eine Kategorie.
\end{enumerate}

\paragraph{Beispiel 1.4.}\label{1.4} \begin{enumerate}
\item Sei $X$ eine Varietät über $k$. Betrachte den Funktor $\mathcal{O}:\textbf{Top$(X)$}\to\textbf{komm Ringe}$ mit den gewöhnlichen Restriktionsabbildungen $\operatorname{res}^U_V:\mathcal{O}(U)\to\mathcal{O}(V)$ ist offensichtlich eine Prägarbe von Ringen. Da ferner reguläre Funktionen $0$ ist, wenn sie lokal $0$ ist, und eine lokal reguläre Funktion auch global regulär ist, ist $\mathcal{O}$ auch eine Garbe.
\item Sei $X$ ein topologischer Raum und $A$ eine abelsche Gruppe. Die \textit{konstante Garbe}\index{Garbe!konstant} $\mathcal{A}$ auf $X$ ist folgendermaßen definiert: Wir statten $A$ mit der diskreten Topologie aus. Für jedes $U\subset_\text{o}X$ setze:
\[\mathcal{A}(U)=\{f:U\to A\mid f\text{ stetig}\} \]
Ist $U$ zusammenhängend, so gilt $\mathcal{A}(U)\stackrel{\sim}{\to} A,\ f\mapsto f(x)$, wobei $x\in U$ beliebig.
\end{enumerate}

\paragraph{Definition 1.5.}\label{1.5} Sei $F$ eine Prägarbe auf $X$ und $P\in X$. Der \textit{Halm}\index{Halm} $F_P$ von $F$ in $P$ ist definiert als:
\[F_P =\varinjlim_{\substack{U\subset_\text{o}X\\ P\in U}} F(U)=\coprod_{\substack{U\subset_\text{o}X\\ P\in U}} F(U)\Big/{\sim} \]
wobei zwei Elemente $s\in F(U),\ t\in F(V)$ genau dann äquivalent $s\sim t$ sind, wenn es ein $\varnothing \neq W\subset_\text{o}X$ mit $W\subset U\cap V$ existiert, so dass $s|_W=t|_W$ gilt. Die Elemente eines Halms heißen \textit{Keime}\index{Keim} der Schnitte von $F$ in $P$.

\paragraph{Beispiel.} Sei $X$ eine Varietät, $P\in X$ ein Punkt und $\mathcal{O}$ die Garbe der regulären Funktionen. Dann ist der Halm in $P$ gerade der lokale Ring $\mathcal{O}_{P,X}$.

\paragraph{Bemerkung.} Ein Morphismus $\phi:F\to G$ von Prägarben induziert für alle $P\in X$ ein Gruppenhomomorphismus $\phi_P:F_P\to G_P$.

\paragraph{Satz 1.6.}\label{1.6} Sei $\phi:F\to G$ ein Morphismus von Garben auf einem topologischen Raum $X$. Dann gilt:
\[\phi:F\to G\text{ ist Isomorphismus}\iff \phi_P:F_P\to G_P\text{ ist Isomorphismus für alle }P\in X \]
Für Prägarben gilt dieser Satz im Allgemeinen nicht.

\paragraph{Beweis.} Ist $\phi$ ein Isomorphismus, so auch alle $\phi_P,\ P\in X$. Sei umgekehrt $\phi_P$ Isomorphismen für alle $P\in X$. Es genügt zu zeigen, dass $\phi(U):F(U)\to G(U)$ für alle $U\subset_\text{o}X$ ein Isomorphismus ist. Sei $U\subset_\text{o}X$ und setze $\varphi=\phi(U)$.
\begin{itemize}
\item \textit{Injektivität von $\varphi$:} Sei $s\in F(U)$ mit $0=\varphi(s)\in G(U)$. Dann gilt für das Bild $\varphi(s)_P$ von $\varphi(s)$ im Halm $0=\varphi(s)_P\in F_P$. Wegen $\varphi(s)_P=\phi_P(s_P)$ für das Bild $s_P\in F_P$ von $s$, folgt wegen der Injektivität von $\phi_P$ nun $s_P=0$ für alle $P\in U$.

Per Definition gibt es für jedes $P\in U$ eine offene Umgebung $W_P\subset_\text{o}X$ von $P$ mit $W_P\subset U$, so dass $s|_{W_P}=0$ gilt. Dann bilden die $W_P$ eine offene Überdeckung von $U=\bigcup_{P\in U}W_P$. Da $F$ eine Garbe ist, folgt $s=0$.

Wir haben gezeigt, dass $\phi(U)$ für alle $U\subset_\text{o}X$ injektiv ist, genau dann wenn $\phi_P$ für alle $P\in X$ injektiv ist.
\item \textit{Surjektivität von $\varphi$:} Sei $t\in G(U)$ ein Schnitt und $t_P\in G_P$ sein Keim in $P$. Da $\phi_P$ surjektiv ist, existiert ein $s_P\in F_P$ mit $\phi_P(s_P)=t_P$. Sei $s_P$ durch den Schnitt $s(P)\in F(V_P)$ mit $V_P\subset_\text{o}U,\ P\in V_P$ repräsentiert. Dann sind $\phi(V_P)(s(P))$ und $t|_{V_P}$ zwei Elemente aus $G(V_P)$ mit demselben Keim. Durch das Verkleinern von $V_P$ folgt $\phi(V_P)(s(P))=t|_{V_P}$ in $G(V_P)$.

Dann bilden die $V_P$ eine offene Überdeckung von $U=\bigcup_{P\in U}V_P$. Es gilt außerdem $s(P)|_{V_P\cap V_Q}=s(Q)|_{V_P\cap V_Q}$ für alle $P,Q\in U$, denn beide Elemente sind Schnitte aus $F(V_P\cap V_Q)$, die durch $\phi(V_P\cap V_Q)$ auf $t|_{V_P\cap V_Q}$ abgebildet werden, und $\phi(V_P\cap V_Q)$ aus dem ersten Teil injektiv ist.

Da $F$ eine Garbe ist, existiert ein $s\in F(U)$ mit $s|_{V_P}=s(P)$ für alle $P\in U$. Schließlich gilt $\varphi(s)|_{V_P}=t|_{V_P}$ für alle $P\in U$, d.h. $(\varphi(s)-t)|_{V_P}=0$. Da $G$ eine Garbe ist, folgt $\varphi(s)=t$.\qed
\end{itemize}

\paragraph{Definition 1.7.}\label{1.7} Sei $\varphi:F\to G$ ein Morphismus von Prägarben. Die Prägarben
\[U\mapsto \ker\varphi(U),\quad U\mapsto\operatorname{coker}\varphi(U),\quad U\mapsto\operatorname{im}\varphi(U) \]
heißen \textit{Prägarbenkern}, \textit{-kokern} und \textit{-bild} von $\varphi$.\index{Kern}\index{Kokern}\index{Bild} Sind $F$ und $G$ Garben, so sind Kokern und Bild nicht notwendig Garben.

\paragraph{Satz \& Definition 1.8.}\label{1.8} Sei $F$ eine Prägarbe. Dann existiert eine Garbe $F^+$ und ein Morphismus von Prägarben $\theta:F\to F^+$ mit folgender Universaleigenschaft:

Sei $G$ eine Garbe und $\phi:F\to G$ ein Morphismus von Prägarben. Dann existiert ein eindeutig bestimmter Morphismus $\psi:F^+\to G$, so dass das folgende Diagramm kommutiert:
\[\begin{tikzcd}
F\ar[rr, "\theta"]\ar[d, "\phi"'] && F^+\ar[dll, "\psi", dashed]\\
G
\end{tikzcd} \]
$F^+$ ist somit eindeutig bestimmt und heißt die zu $F$ \textit{assoziierte Garbe}\index{assoziierte Garbe!Prägarbe}.

\paragraph{Beweis.} Für jede offene Menge $U\subset X$ setze $F^+(U)$ als die Menge aller Abbildungen $s:U\to\coprod_{P\in U}F_P$, so dass:
\begin{enumerate}[(i)]
\item Für alle $P\in U$ gilt $s(P)\in F_P$.
\item Für alle $P\in U$ gibt es eine offene Umgebung $V$ von $P$ mit $V\subset U$ und ein Element $t\in F(V)$, so dass für alle $Q\in V$ der Keim $t_Q$ von $t$ in $Q$ gleich $s(Q)$ ist.
\end{enumerate}
Somit wird $F^+$ zu einer Garbe bzgl. der natürlichen Restriktionsabbildungen und besitzt die verlangte Universaleigenschaft. Für jeden Punkt $P\in X$ gilt $F^+_P=F_P$. Ist $F$ eine Garbe, so ist $F^+\cong F$ via $\theta$.\qed

\paragraph{Definition 1.9.}\label{1.9}
\begin{enumerate}[(i)]
\item Eine \textit{Untergarbe}\index{Garbe!Untergarbe} von $F$ ist eine Garbe $F'$ derart, dass:
\begin{enumerate}[(a)]
\item $F'(U)\subset F(U)$ ist eine Untergruppe für alle $U\subset_\text{o}X$.
\item Für offene Mengen $V\subset U$ gilt $\operatorname{res}'{}^U_V=\operatorname{res}^U_V|_{F'(U)}$.
\end{enumerate}
Insbesondere ist $F'_P\subset F_P$ eine Untergruppe.
\item Der \textit{Kern}\index{Kern} von $\varphi$ ist die Prägarbe $\ker(\varphi)$, die bereits eine Garbe ist. \textit{Grund:}

Sei $U\subset_\text{o}X$ und $U=\bigcup U_i$ eine offene Überdeckung. Sei $s\in\ker\varphi(U)$ mit $s|_{U_i}=0$ für alle $i$. Da $F$ eine Garbe ist und $s\in F(U)$, folgt $s=0$. Sei nun $s_i\in \ker\varphi(U_i)$ für alle $i$ gegeben mit $s_i|_{U_i\cap U_j}=s_j|_{U_i\cap U_j}$ für alle $i,j$. Da $F$ eine Garbe ist, existiert ein $s\in F(U)$ mit $s|_{U_i}=s_i$ für alle $i$. Zu zeigen ist noch $s\in \ker\varphi(U)$. Es gilt für alle $i$:
\[0=\varphi(U_i)(s_i)=\varphi(U_i)(s|_{U_i}) = \varphi(U)(s)|_{U_i}\in G(U_i) \]
Da nun auch $G$ eine Garbe ist, folgt $\varphi(U)(s)=0$.
\item $\varphi$ heißt \textit{injektiv}\index{Injektivität}, falls $\ker(\varphi)=0$.

Mit anderen Worten: $\varphi$ ist genau dann injektiv, wenn $\varphi(U):F(U)\to G(U)$ für alle $U\subset_\text{o}X$ injektiv ist.
\item Das \textit{Bild}\index{Bild} $\operatorname{im}(\varphi)$ von $\varphi$ ist die assoziierte Garbe des Prägarbenbilds von $\varphi$.

Nach der Universaleigenschaft gibt es einen natürlichen Morphismus $\psi:\operatorname{im}(\varphi)\to G$. Dieser ist injektiv, da $(\operatorname{im}\varphi)_P:\operatorname{im}(\varphi_P) \to G_P$ für alle $P\in X$ injektiv ist.
\item $\varphi$ heißt \textit{surjektiv}\index{Surjektivität}, wenn $\operatorname{im}(\varphi)=G$.
\item Eine Garbensequenz
\[\begin{tikzcd}
\cdots \ar[r] & F^i\ar[r, "\varphi^i"] & F^{i+1}\ar[r, "\varphi^{i+1}"] & F^{i+2}\ar[r] & \cdots
\end{tikzcd} \]
heißt \textit{exakt}\index{Exaktheit}, falls $\ker(\varphi^{i+1})=\operatorname{im}(\varphi^i)$ für alle $i$ gilt.
\item Sei $F'$ eine Untergarbe von $F$. Die \textit{Quotientengarbe}\index{Quotientengarbe} $F/F'$ ist die assoziierte Garbe zur Prägarbe $U\mapsto F(U)/F'(U)$. Offensichtlich gilt $(F/F')_P=F_P/F'_P$ für alle $P\in X$.
\item Der \textit{Kokern}\index{Kokern} von $\varphi$ ist die assoziierte Garbe zum Prägarbenkokern von $\varphi$.
\end{enumerate}

\paragraph{Regeln 1.10.}\label{1.10} Seien $F,G$ Garben auf $X$ und $\varphi:F\to G$ ein Morphismus von Garben.
\begin{enumerate}[(i)]
\item $\varphi$ ist genau dann injektiv, wenn $0\to F\to G$ exakt ist und genau dann surjektiv, wenn $F\to G\to 0$ exakt ist.
\item Eine Garbensequenz $\cdots \to F^i\to F^{i+1}\to F^{i+2}\to\cdots$ ist genau dann exakt, wenn ihre entsprechenden Halmsequenzen in allen Punkten $P\in X$ exakt ist. \textit{Grund:}
\[(\operatorname{im} \varphi^i)_P=\operatorname{im}(\varphi^i_P),\quad (\ker\varphi^{i+1})_P=\ker(\varphi^{i+1}_P) \]
Insbesondere ist ein Garbenmorphismus genau dann injektiv bzw. surjektiv falls alle Halmabbildungen injektiv bzw. surjektiv sind.
\item $\varphi$ ist genau dann surjektiv, wenn für alle $U\subset_\text{o}X$ und $s\in G(U)$ eine Überdeckung $U=\bigcup_{i\in I} U_i$ mit Schnitten $t_i\in F(U_i),\ i\in I$ existieren, so dass $\varphi(t_i)=s|_{U_i}$.

Ist $\varphi$ surjektiv, so muss im Allgemeinen $\varphi(U):F(U)\to G(U)$ nicht surjektiv sein.
\item Sei $0\to F'\to F\to F''$ eine exakte Garbensequenz und $U\subset_\text{o}X$. Dann ist auch die folgende Sequenz exakt:
\[\begin{tikzcd}[cramped, sep=small]
0 \ar[r] & \Gamma(U,F') \ar[r]& \Gamma(U,F) \ar[r]& \Gamma(U,F'')
\end{tikzcd} \]
Der Funktor $\Gamma(U,-)$ ist linksexakt, aber nicht exakt.
\item Sei $\varphi:F\to G$ ein injektiver Morphismus von Prägarben. Dann ist der induzierte Morphismus $\varphi^+:F^+\to G^+$ der assoziierten Garben auch injektiv. Der Funktor $-^+$ ist sogar exakt.
\end{enumerate}

\paragraph{Regeln 1.11.}\label{1.11} \begin{enumerate}[(i)]
\item Sei $F'$ eine Untergarbe von $F$. Dann ist die Sequenz $0\to F'\to F\to F/F'\to 0$ exakt, da sie halmweise exakt ist.
\item Sei $\varphi:F\to G$ ein Garbenmorphismus. Dann gilt:
\[\operatorname{im}(\varphi)\cong F/\ker(\varphi),\quad \operatorname{coker}(\varphi)\cong G/ \operatorname{im}(\varphi) \]
d.h. die Folgen $0\to\ker(\varphi)\to F\to\operatorname{im}(\varphi)\to 0$ und $0\to\operatorname{im}(\varphi)\to G\to\operatorname{coker}(\varphi)\to 0$ sind exakt.
\end{enumerate}

\paragraph{Definition 1.12.}\label{1.12} \begin{enumerate}[(i)]
\item Seien $F$ und $G$ Garben auf $X$. Die \textit{Summe}\index{Summe} $F\oplus G$ von $F$ und $G$ ist die Garbe $U\mapsto F(U)\oplus G(U)$.
\item Sei $(F_i,\varphi_{i,j})$ ein direktes System von Garben auf $X$. Der \textit{direkte Limes}\index{Limes!direkt} $(\varinjlim F_i,\varphi_i)$ ist die assoziierte Garbe zur Prägarbe $U\mapsto \varinjlim F_i(U)$.

Der direkte Limes besitzt die übliche Universaleigenschaft: Sei $G$ eine Garbe und $\psi_i:F_i\to G$ Morphismen mit $\varphi_k\varphi_{ik}=\psi_i$ für alle $i\leq k$. Dann existiert ein eindeutig bestimmter Morphismus $\psi:\varinjlim F_i\to G$, so dass das folgende Diagramm kommutiert:
\[\begin{tikzcd}
F_i\ar[rr, "\varphi_i"]\ar[d, "\psi_i"'] & & \varinjlim F_i\ar[lld, "\psi", dashed]\\
G
\end{tikzcd} \]
\item Ebenso wird der \textit{projektive Limes}\index{Limes!projektiv} definiert, wobei alle Pfeile umgedreht werden. Es ist außerdem $U\mapsto\varprojlim F_i(U)$ bereits eine Garbe.
\item Sei $F$ eine Garbe auf $X$ und $s\in F(U)$ ein Schnitt über $U\subset_\text{o}X$. Dann heißt
\[\operatorname{Supp}(s)=\{P\in U\mid s_P\neq 0\} \]
wobei $s_P\in F_P$ der Keim von $s$ in $P$ bezeichnet, der \textit{Support}\index{Support} von $s$. $\operatorname{Supp}(s)$ ist abgeschlossen in $U$.
\[\operatorname{Supp}(F)=\{P\in X\mid F_P\neq 0\} \]
heißt \textit{Support} von $F$. Dieser ist nicht notwendigerweise abgeschlossen.
\item Seien $F,G$ Garben abelscher Gruppen auf $X$. Für ein $U\subset_\text{o}X$ sei $F|_U$ die Einschränkung von $F$ auf $U$, d.h. $F|_U(V)=F(V)$ für alle $V\subset_\text{o}U$. Dann ist die Menge $\operatorname{Hom}(F|_U,G|_U)$ der Morphismen von $F|_U$ nach $G|_U$ eine abelsche Gruppe.
\[U\mapsto\operatorname{Hom}(F|_U,G|_U)\]
definiert eine Garbe und wird die \textit{Hom-Garbe}\index{Hom-Garbe} genannt. Sie wird mit $\operatorname{\mathcal{H}om}(F,G)$ bezeichnet.
\end{enumerate}

\paragraph{Definition 1.13.}\label{1.13} Sei $f:X\to Y$ eine stetige Abbildung topologischer Räume, $F$ eine Garbe auf $X$ und $G$ eine Garbe auf $Y$.
\begin{enumerate}[(i)]
\item Die \textit{direkte Bildgarbe}\index{direkte Bildgarbe} $f_\ast F$ von $F$ auf $Y$ ist die Garbe
\[V\mapsto (f_\ast F)(V)=F(f^{-1}(V)) \]
\item Die \textit{Urbildgarbe}\index{Urbildgarbe} $f^{-1}G$ von $G$ auf $X$ ist die assoziierte Garbe zur Prägarbe
\[U\mapsto (f^{-1}G)(U)= \varinjlim_{\substack{V\subset_\text{o}Y\\ f(U)\subset V}}G(V) \]
\end{enumerate}

\paragraph{Regeln 1.14.}\label{1.14} Seien $X,Y$ topologische Räume.
\begin{enumerate}[(i)]
\item Sei $Z\subset X$ ein Teilraum mit der Inklusionsabbildung $i:Z\hookrightarrow X$ und $F$ eine Garbe auf $X$. Dann gilt:
\[i^{-1}F = F|_Z \]
Offensichtlich gilt $(F|_Z)_P=F_P$ für $P\in Z$.
\item Seien $\mathbf{Ab}(X)$ und $\mathbf{Ab}(Y)$ die Kategorien der Garben auf $X$ bzw. $Y$. Dann sind $f_\ast:\mathbf{Ab}(X)\to\mathbf{Ab}(Y)$ und $f^{-1}:\mathbf{Ab}(Y)\to\mathbf{Ab}(X)$ Funktoren.
\item Sei $f:X\to Y$ stetig. Dann sind die \textit{Adjunktionsabbildungen}\index{Adjunktionsabbildung}
\[\operatorname{ad}:f^{-1}f_\ast F\to F,\quad\operatorname{ad} :G\to f_\ast f^{-1}G \]
für Garben $F$ auf $X$ bzw. $G$ auf $Y$ Garbenmorphismen und wie folgt definiert:
\begin{itemize}
\item Sei $U\subset_\text{o}X$ und $s\in (f^{-1}f_\ast F)(U)$ ein Schnitt, das durch $s'\in (f_\ast F)(V)$ mit $f(U)\subset V\subset_\text{o}Y$ repräsentiert wird, d.h. $s'\in F(f^{-1}(V))$ mit $U\subset f^{-1}(V)$. Dann setzen wir:
\[(f^{-1}f_\ast F)(U)\to F(U),\ s\mapsto \operatorname{res}_U^{f^{-1}(V)}s'\in F(U)\]
\item Sei $V\subset_\text{o}Y$. Es besteht $(f_\ast f^{-1}G)(V)=(f^{-1}G)(f^{-1}(V))$ aus Abbildungen der Form $f^{-1}(V)\to\coprod_{P\in f^{-1}(V)}(f^{-1}G)_P$. Setze nun:
\[G(V)\to (f_\ast f^{-1}G)(V),\ s\mapsto (s\circ f: f^{-1}(V)\to\coprod  f^{-1}(G)_P,\ P\mapsto s_{f(P)}) \]
\end{itemize}
Es existiert eine natürliche Bijektion:
\[\operatorname{Hom}_X(f^{-1}G,F)\cong\operatorname{Hom}_Y(G,f_\ast F) \]
in dem wir ein $\varphi:f^{-1}(G)\to F$ auf $\psi: G\stackrel{\operatorname{ad}}{\longrightarrow} f_\ast f^{-1}G\stackrel{f_\ast(\varphi)}{\longrightarrow} f_\ast F$ schicken und ein $\psi:G\to f_\ast F$ auf $\varphi:f^{-1}G \stackrel{f^{-1}(\psi)}{\longrightarrow} f^{-1}f_\ast F\stackrel{\operatorname{ad}}{\longrightarrow}F$ schicken. Somit ist $f^{-1}$ linksadjungiert zu $f_\ast$.
\end{enumerate}

\paragraph{Definition 1.15.}\label{1.15} Sei $X$ ein topologischer Raum mit $P\in X$ und $A$ eine abelsche Gruppe. Sei $\mathcal{A}$ die konstante Garbe auf $\overline{\{P\}}$ und $i:\overline{\{P\}}\hookrightarrow X$ die natürliche Inklusion. Dann heißt die Garbe $i_\ast \mathcal{A}$ \textit{Wolkenkratzergarbe}. Es gilt:
\[(i_\ast\mathcal{A})(U) = \begin{cases}
A, &\text{wenn } P\in U\\
0, &\text{sonst}
\end{cases},\quad (i_\ast \mathcal{A})_Q = \begin{cases}
A,&\text{wenn } Q\in\overline{\{P\}}\\
0,&\text{sonst}
\end{cases} \]

\paragraph{Definition 1.16.}\label{1.16} Sei $X$ ein topologischer Raum, $Z\subset X$ abgeschlossen und $U=X\setminus Z$. Seien $j:U\hookrightarrow X$, $i:Z\hookrightarrow X$ die Inklusionsabbildungen. Ist $F$ eine Garbe auf $Z$, so gilt:
\[(i_\ast F)_P=\begin{cases}
F_P, &\text{wenn }P\in Z\\
0, &\text{sonst}
\end{cases} \]
Sei $F$ eine Garbe auf $U$ und $j_!(F)$ die Garbe auf $X$, die zur Prägarbe
\[V\mapsto\begin{cases}
F(V),&\text{wenn }V\subset U\\
0,&\text{sonst}
\end{cases} \]
assoziiert ist. Sie heißt die \textit{außerhalb $U$ durch Null fortgesetzte Garbe}\index{Garbe!Fortsetzung} von $F$. Es gilt:
\[(j_!F)_P=\begin{cases}
F_P,&\text{wenn }P\in U\\
0,&\text{sonst}
\end{cases} \]
Sei $F$ eine Garbe auf $X$. Dann ist die folgende Sequenz exakt:
\[\begin{tikzcd}[cramped, sep=small]
0\ar[r] & j_!(F|_U)\ar[r] & F \ar[r]& i_\ast(F|_Z)\ar[r] & 0
\end{tikzcd} \]

\section{Schemata}

Sei $A$ ein kommutativer Ring mit Eins und $\operatorname{Spec}(A)$ die Menge aller Primideale von $A$. Für ein Ideal $\mathfrak{a}\subset A$, setzen wir:
\[V(\mathfrak{a}) = \{\mathfrak{p}\in\operatorname{Spec}(A)\mid \mathfrak{a}\subset\mathfrak{p}\} \]

\paragraph{Lemma 2.1.}\label{2.1} \begin{enumerate}[(i)]
\item Sind $\mathfrak{a},\mathfrak{b}$ zwei Ideale von $A$, so gilt $V(\mathfrak{ab})=V(\mathfrak{a})\cup V(\mathfrak{b})$.
\item Sind $\mathfrak{a}_i\subset A,\ i\in I$ Ideale, so gilt $V\big(\sum\mathfrak{a}_i\big)=\bigcap V(\mathfrak{a}_i)$.
\item Sind $\mathfrak{a},\mathfrak{b}\subset A$ Ideale, gilt: $V(\mathfrak{a})\subset V(\mathfrak{b}) \iff \operatorname{Rad}(\mathfrak{a})\supset\operatorname{Rad}(\mathfrak{b})$
\end{enumerate}
Wegen $V(A)=\varnothing$ und $V(0)=\operatorname{Spec}(A)$ sehen wir, dass wir Teilmengen der Form $V(\mathfrak{a})$ zu abgeschlossene Mengen in $\operatorname{Spec}(A)$ erklären können. Somit erhalten wir die \textit{Zariski-Topologie}\index{Zariski-Topologie} auf $\operatorname{Spec}(A)$.

Setzen wir $D(f)=\operatorname{Spec}(A)\setminus V(f)$ für $f\in A$, so bilden diese offene Mengen eine Basis der Topologie auf $\operatorname{Spec}$.

\paragraph{Definition 2.2.}\label{2.2} Wir definieren eine Ringgarbe $\mathcal{O}$ auf $\operatorname{Spec}(A)$ wie folgt: Sei $U\subset_\text{o}\operatorname{Spec}(A)$. Setze $\mathcal{O}(U)$ als die Menge aller Abbildungen $s:U\to\coprod_{\mathfrak{p}\in U}A_\mathfrak{p}$, so dass:
\begin{enumerate}[(i)]
\item Für alle $\mathfrak{p}\in U$ gilt $s(\mathfrak{p})\in A_\mathfrak{p}$.
\item Für alle $\mathfrak{p}\in U$ gibt es eine offene Umgebung $V$ von $\mathfrak{p}$ mit $V\subset U$ und Elemente $a,f\in A$, so dass für alle $\mathfrak{q}\in V$ stets $f\not\in\mathfrak{q}$ und $s(\mathfrak{q})=\frac{a}{f}$ in $A_\mathfrak{q}$ gilt.
\end{enumerate}
Offensichtlich ist mit $s,t\in\mathcal{O}(U)$ auch $s+t,st\in\mathcal{O}(U)$. Ferner ist für $V\subset U$ offen $\operatorname{res}^U_V:\mathcal{O}(U)\to\mathcal{O}(V)$ ein Ringhomomorphismus. $\mathcal{O}$ heißt \textit{Strukturgarbe}\index{Strukturgarbe}. Das \textit{Spektrum}\index{Spektrum} von $A$ ist das Paar $(\operatorname{Spec}A,\mathcal{O})$.

\paragraph{Satz 2.3.}\label{2.3} Sei $A$ ein Ring und $(\operatorname{Spec} A,\mathcal{O})$ sein Spektrum. Dann gilt:
\begin{enumerate}[(i)]
\item Der Halm $\mathcal{O}_\mathfrak{p}$ ist isomorph zu $A_\mathfrak{p}$ für alle $\mathfrak{p}\in\operatorname{Spec}(A)$.
\item $\mathcal{O}(D(f))\cong A_f$ für alle $f\in A$
\item $\Gamma(\operatorname{Spec}(A),\mathcal{O})=A$
\end{enumerate}

\paragraph{Beweis.} (iii) folgt aus (ii) mit $f=1$.
\begin{enumerate}[(i)]
\item Die Abbildungen $\mathcal{O}(U)\to A_\mathfrak{p},\ s\mapsto s(\mathfrak{p})$ mit $\mathfrak{p}\in U\subset_\text{o}\operatorname{Spec}(A)$ sind kompatibel und induzieren einen Homomorphismus $\varphi:\mathcal{O}_\mathfrak{p}\to A_\mathfrak{p}$.
\begin{itemize}
\item \textit{Surjektivität:} Sei $\frac{a}{f}\in A_\mathfrak{p}$ mit $a,f\in A,\ f\not\in\mathfrak{p}$. Dann ist $D(f)$ eine offene Umgebung von $\mathfrak{p}$ und es gibt ein $s\in\mathcal{O}(D(f))$ mit $s(\mathfrak{p})=\frac{a}{f}$.
\item \textit{Injektivität:} Sei $U$ eine Umgebung von $\mathfrak{p}$ und $s,t\in\mathcal{O}(U)$ mit $s(\mathfrak{p})=t(\mathfrak{p})$. Verkleinern wir $U$ wenn nötig, so können wir o.B.d.A. annehmen, dass:
\[s=\frac{a}{f},\quad t=\frac{b}{g}\quad \text{ für gewisse } a,b,g,f\in A,\ g,f\not\in\mathfrak{p} \]
Wegen $\frac{a}{f}=\frac{b}{g}$ in $A_\mathfrak{p}$, gibt es ein $h\not\in\mathfrak{p}$, so dass $h(ga-bf)=0$ in $A$. Insbesondere ist $\frac{a}{f}=\frac{b}{g}$ in $A_\mathfrak{q}$ für alle $\mathfrak{q}\in\operatorname{Spec}(A)$ mit $g,f,h\not\in\mathfrak{q}$. Somit ist $s=t$ auf der offenen Umgebung $D(f)\cap D(g)\cap D(h)$ von $\mathfrak{p}$ und haben daher denselben Keim.
\end{itemize}
\item Sei $f\in A$ und $\mathfrak{p}\in D(f)$, d.h. $(f)\subset A\setminus\mathfrak{p}$. Betrachte die kanonische Abbildung $\lambda_\mathfrak{p}:A_f\to A_\mathfrak{p}$. Setze:
\[\psi: A_f \to\mathcal{O}(D(f)),\ \frac{a}{f^n}\mapsto \Big(\mathfrak{p}\mapsto \lambda_\mathfrak{p}\Big(\frac{a}{f^n}\Big)\Big) \]
\begin{itemize}
\item \textit{Injektivität:} Sei $\psi\big(\frac{a}{f^n}\big)=\psi \big(\frac{b}{f^m}\big)$ und $\mathfrak{p}\in D(f)$. Dann ist $\frac{a}{f^n}=\frac{b}{f^m}$ in $A_\mathfrak{p}$, d.h. es gibt ein $h\not\in\mathfrak{p}$ mit $h(f^ma-f^nb)=0$. Setze $\mathfrak{a}=\operatorname{Ann}(f^ma-f^nb)$. Dann ist $\mathfrak{a}\not\subset\mathfrak{p}$, da $h\in\mathfrak{a}$, also folgt $\mathfrak{p}\not\in V(\mathfrak{a})$. Wir haben also $V(\mathfrak{a})\subset V(f)$ gezeigt. Nach \hyperref[2.1]{Lemma 2.1 (iii)} folgt $f\in\operatorname{Rad}(\mathfrak{a})$, d.h. $f^e\in\mathfrak{a}$ für ein $e>0$. Per Definition gilt $f^e(f^na-f^mb)=0$, also $\frac{a}{f^n}=\frac{b}{f^m}$ in $A_f$.
\item \textit{Surjektivität:} Sei $s\in\mathcal{O}(D(f))$. Nach Definition von $\mathcal{O}$ ist $D(f)=\bigcup V_i$ mit $s=\frac{a_i}{g_i}$ auf $V_i$ für gewisse $a_i,g_i\in A,\ g_i\not\in\mathfrak{p}$ für alle $\mathfrak{p}\in V_i$. Insbesondere gilt $V_i\subset D(g_i)$. Da die $D(h)$ eine Basis der Topologie bilden, können wir o.B.d.A. $V_i=D(h_i)$ annehmen, also $D(h_i)\subset D(g_i)$. Es folgt $V(h_i)\supset V(g_i)$ und nach \hyperref[2.1]{Lemma 2.1 (iii)} auch $\operatorname{Rad}(h_i)\subset\operatorname{Rad}(g_i)$. Wähle ein $n$, so dass $h_i^n\in (g_i)$ für alle $i$, d.h. $h_i^n=c_ig_i$ für ein $c_i\in A$. Es folgt:
\[\frac{a_i}{g_i}=\frac{c_ia_i}{h_i^n}\]
Ersetzt man $a_i$ durch $c_ia_i$ und $h_i$ durch $h_i^n$, so können wir o.B.d.A. $D(f)\subset\bigcup D(h_i)$ und $s=\frac{a_i}{h_i}$ auf $D(h_i)$ annehmen.

Wir zeigen nun, dass $D(f)$ durch endlich viele $D(h_i)$ überdeckt werden kann. Wir haben mit \hyperref[2.1]{Lemma 2.1} Äquivalenzen:
\begin{align*}
D(f)\subset \bigcup_{i}D(h_i) &\iff V(f)\supset\bigcap_i V(h_i) = V\Big(\sum_i (h_i)\Big)\\
&\iff f\in\operatorname{Rad}\Big(\sum_i (h_i)\Big)\\
&\iff \exists n\in\mathbb{N}\colon f^n\in\sum_i (h_i)
\end{align*}
Daher ist $f^n$ eine endliche Summe der Form $f^n=\sum_{i=1}^r b_ih_i$ für gewisse $b_i\in A$, d.h. $D(f)\subset D(h_1)\cup\cdots\cup D(h_r)$.

Nun gilt: \[D(h_i)\cap D(h_j)=\operatorname{Spec}(A)\setminus (V(h_i)\cup V(h_j))=\operatorname{Spec}(A)\setminus V(h_ih_j)=D(h_ih_j)\]
Auf $D(h_ih_j)$ wird $s$ repräsentiert durch $\frac{a_i}{h_i}$ und $\frac{a_j}{h_j}$ in $A_{h_ih_j}$. Wenden wir die Injektivität von $\psi$ auf $D(h_ih_j)$ an, so erhalten wir $\frac{a_i}{h_i}=\frac{a_j}{h_j}$ in $A_{h_ih_j}$. Es folgt $(h_ih_j)^n(h_ja_i-h_ia_j)=0$ für ein $m$. Sei $m$ so groß, dass dies für alle endlich vielen $i,j$ gilt, also gilt für alle $i,j$:
\[h_j^{m+1}(h_i^ma_i)-h_i^{m+1}(h_j^ma_j)=0 \]
Ersetzen wir nun $h_i$ durch $h_i^{m+1}$ und $a_i$ durch $a_ih_i^m$, so wird $s$ auf $D(h_i)$ immer noch durch $\frac{a_i}{h_i}$ repräsentiert und es gilt $h_ja_i=h_ia_j$ für alle $i,j$.

Schreibe nun $f^n=\sum b_ih_i$ für ein $n$ und setze $a=\sum b_ia_i$. Es folgt für alle $j$:
\[h_ja = \sum_i b_ia_ih_j=\sum _i b_ih_ia_j=f^na_j \]
Also gilt $\frac{a}{f^n}=\frac{a_j}{h_j}$ auf $D(h_j)$ für alle $j$, d.h. $\psi\big(\frac{a}{f^n}\big)=s$.\qed
\end{itemize}
\end{enumerate}

\paragraph{Definition 2.4.}\label{2.4} \begin{enumerate}[(i)]
\item Ein \textit{geringter Raum}\index{geringter Raum} $(X,\mathcal{O}_X)$ besteht aus einem topologischen Raum $X$ und einer Ringgarbe $\mathcal{O}_X$ auf $X$. Ein \textit{Morphismus von geringten Räumen}\index{Morphismus!geringte Räume} ist ein Paar
\[(f,f^\sharp):(X,\mathcal{O}_X)\to (Y,\mathcal{O}_Y) \]
wobei $f:X\to Y$ eine stetige Abbildung und $f^\sharp:\mathcal{O}_Y\to f_\ast\mathcal{O}_X$ ein Morphismus von Ringgarben auf $Y$ ist.
\item Ein geringter Raum $(X,\mathcal{O}_X)$ heißt \textit{lokal}\index{geringter Raum!lokal}, falls für alle $P\in X$ der Halm $\mathcal{O}_{X,P}$ ein lokaler Ring ist. Ein \textit{Morphismus von lokal geringten Räumen}\index{Morphismus!lokal geringte Räume} ist ein Morphismus $(f,f^\sharp)$ von geringten Räumen derart, dass für alle $P\in X$ die induzierte Abbildung $f_P^\sharp:\mathcal{O}_{Y,f(P)}\to\mathcal{O}_{X,P}$ lokale Homomorphismen sind.
\end{enumerate}

\paragraph{Bemerkung 2.5.}\label{2.5}\begin{itemize}
\item Die (lokal) geringte Räume bilden eine Kategorie.
\item Ein Morphismus $(f,f^\sharp)$ von (lokal) geringten Räumen ist genau dann ein Isomorphismus, wenn $f$ ein Homöomorphismus ist und $f^\sharp$ ein Garbenisomorphismus ist.
\end{itemize}

\paragraph{Satz 2.6.}\label{2.6} Seien $A,B$ Ringe.
\begin{enumerate}[(i)]
\item $(\operatorname{Spec}A,\mathcal{O})$ ist ein lokal geringter Raum.
\item Sei $\varphi:A\to B$ ein Ringhomomorphismus. Dann induziert $\varphi$ einen natürlichen Morphismus von lokal geringten Räumen:
\[(f,f^\sharp):(\operatorname{Spec}B,\mathcal{O}_B)\to (\operatorname{Spec}A,\mathcal{O}_A),\ f(\mathfrak{p})=\varphi^{-1}(\mathfrak{p}) \]
\item Jeder Morphismus $(f,f^\sharp):(\operatorname{Spec}B,\mathcal{O}_B)\to(\operatorname{Spec}A,\mathcal{O}_A)$ von lokal geringten Räu\-men ist induziert von einem Ringhomomorphismus $\varphi:A\to B$.
\end{enumerate}

\paragraph{Beweis.} \begin{enumerate}[(i)]
\item folgt aus \hyperref[2.3]{Satz 2.3 (i)}.
\item Definiere $f$ durch $f(\mathfrak{p})=\varphi^{-1}(\mathfrak{p})$ für alle $\mathfrak{p}\in\operatorname{Spec}(B)$. Sei $\mathfrak{a}\subset A$ ein Ideal. Dann ist $f^{-1}(V(\mathfrak{a}))=V(\varphi(\mathfrak{a}))$. Daher ist $f$ stetig. Sei $\mathfrak{p}\in\operatorname{Spec}(B)$. Dann liefert $\varphi$ einen lokalen Homomorphismus $\varphi_\mathfrak{p}:A_{\varphi^{-1}(\mathfrak{p})}\to B_\mathfrak{p}$. Das liefert für $V\subset_\text{o}\operatorname{Spec}(A)$ einen Ringhomomorphismus:
\[f^\sharp(V):\mathcal{O}_A(V)\to\mathcal{O}_B(f^{-1}(V)) \]
indem man eine Abbildung $s:V\to\coprod_{\mathfrak{q}\in V} A_\mathfrak{q}$ auf die folgende Abbildung $f^\sharp(s): f^{-1}(V)\to\coprod_{\mathfrak{p}\in f^{-1}(V)}B_\mathfrak{p}$ schickt:
\[ \begin{tikzcd}[row sep=.1em]
f^{-1}(V) \ar[r, "f"] & V\ar[r, "s"] & A_{\varphi^{-1}(\mathfrak{p})}\ar[r, "\varphi_\mathfrak{p}"] & B_\mathfrak{p}\\
\mathfrak{p}\ar[r, mapsto] & f(\mathfrak{p}) = \varphi^{-1}(\mathfrak{p})\ar[r, mapsto] & s(\varphi^{-1}(\mathfrak{p}))\ar[r, mapsto] & \varphi_\mathfrak{p}(s(\varphi^{-1}(\mathfrak{p})))
\end{tikzcd} \]
Die $f^\sharp(V)$ gibt uns einen Garbenmorphismus $f^\sharp:\mathcal{O}_A\to f_\ast \mathcal{O}_B$. Die durch $f^\sharp$ induzierte Abbildungen auf den Halmen sind gerade die $\varphi_\mathfrak{p}$. Somit ist $(f,f^\sharp)$ ein Morphismus lokal geringten Räumen.
\item Sei $(f,f^\sharp):(\operatorname{Spec}B,\mathcal{O}_B)\to(\operatorname{Spec}A,\mathcal{O}_A)$ ein Morphismus von lokal geringten Räu\-men. $f^\sharp$ induziert einen Ringhomomorphismus:
\[\varphi: A=\Gamma(\operatorname{Spec}A,\mathcal{O}_A)\to\Gamma(\operatorname{Spec}B,\mathcal{O}_B) =B \]
Sei $\mathfrak{p}\in\operatorname{Spec}(B)$. Dann haben wir induzierte lokale Homomorphismen mit kommutativem Diagramm:
\[\begin{tikzcd}
A_{f(\mathfrak{p})} = \mathcal{O}_{A,f(\mathfrak{p})}  \ar[r, "f^\sharp_\mathfrak{p}"] & \mathcal{O}_{B,\mathfrak{p}} = B_\mathfrak{p}\\
A \ar[r, "\varphi"']\ar[u] & B\ar[u]
\end{tikzcd} \]
Da die $f^\sharp_\mathfrak{p}$ lokale Homomorphismen sind, folgt $f(\mathfrak{p})=\varphi^{-1}(\mathfrak{p})$. Somit ist $f^\sharp$ von dem Ringhomomorphismus $\varphi$ induziert.\qed
\end{enumerate}

\paragraph{Definition 2.7.}\label{2.7} \begin{enumerate}[(i)]
\item Ein \textit{affines Schema}\index{Schema!affin} ist ein lokal geringter Raum $(X,\mathcal{O}_X)$, der als lokal geringter Raum isomorph zum Spektrum eines Rings ist.
\item Ein \textit{Schema}\index{Schema} ist ein lokal geringter Raum $(X,\mathcal{O}_X)$ derart, dass jeder Punkt $P\in X$ eine offene Umgebung $U\subset X$ besitzt, so dass $(U,\mathcal{O}_X|_U)$ ein affines Schema ist.
\end{enumerate}
$\mathcal{O}_X$ heißt \textit{Strukturgarbe}\index{Strukturgarbe}. Ein \textit{Morphismus}\index{Morphismus!Schemata} von Schemata ist ein Morphismus von lokal geringten Räumen.

\paragraph{Beispiel 2.8.}\label{2.8}\begin{enumerate}
\item Sei $k$ ein Körper. $\operatorname{Spec}(k)$ ist ein affines Schema, dessen topologischer Raum aus einem Punkt besteht.
\end{enumerate}

\paragraph{Definition 2.9.}\label{2.9} Sei $K$ ein Körper. Eine \textit{diskrete Bewertung}\index{Bewertung!diskret} von $K$ ist eine Abbildung $v:K\to\mathbb{Z}\cup\{\infty\}$, so dass für alle $x,y\in K$ gilt:
\begin{enumerate}[(i)]
\item $v(xy)=v(x)+v(y)$
\item $v(x+y)\geq \min\{v(x),v(y)\}$
\item $v(x)=\infty\iff x=0$
\end{enumerate}
$R=\{x\in K\mid v(x)\geq 0\}$ definiert einen Teilring von $K$ und heißt \textit{diskreter Be\-wer\-tungs\-ring}\index{Bewertungsring!diskret} von $v$. $R$ ist ein lokaler Hauptidealring mit Maximalideal $\mathfrak{m}=\{x\in K\mid v(x) >0\}$. $R/\mathfrak{m}$ heißt \textit{Restklassenkörper}\index{Restklassenkörper} von $v$. Ein \textit{diskreter Bewertungsring}\index{Bewertungsring!diskret} $A$ ist ein nullteilerfreier Ring, der diskreter Bewertungsring für eine Bewertung seines Quotientenkörpers ist.

\paragraph{Beispiel 2.8}\begin{enumerate}
\item[2.] Sei $R$ ein diskreter Bewertungsring. Es ist $T=\operatorname{Spec}(R)$ ein affines Schema, bestehend aus zwei Punkten:
\begin{itemize}
\item Der Punkt $t_0=\mathfrak{m}\in\operatorname{Spec}(R)$ ist abgeschlossen, da $V(\mathfrak{m})=\{\mathfrak{m}\}$ und besitzt $R=R_{t_0}$ als lokalen Ring.
\item Der Punkt $t_1=(0)\in\operatorname{Spec}(R)$ ist offen und dicht in $\operatorname{Spec}(R)$, da $V(0)=\operatorname{Spec}(R)$. $t_1$ besitzt $K=\operatorname{Quot}(R)=R_{t_1}$ als lokalen Ring.
\end{itemize}
\[\begin{tikzcd}[row sep=.1em]
\operatorname{Spec}(K)\ar[r] & \operatorname{Spec}(R) &\ar[l] \operatorname{Spec}(R/\mathfrak{m})\\
(0)\ar[r, mapsto] & t_1\quad t_0 & \ar[l, mapsto] (0)
\end{tikzcd} \]
\item[3.] Sei $k$ ein Körper. Die \textit{affine Gerade}\index{affine Gerade} $\mathbf{A}_k^1$ über $k$ ist $\operatorname{Spec}k[X]$. Sei $\xi$ das Nullideal in $\operatorname{Spec}k[X]$. Dann ist $\overline{\{\xi\}}=\mathbf{A}_k^1$. Ein solcher Punkt heißt \textit{generischer Punkt}\index{generischer Punkt}. Alle anderen Punkte sind abgeschlossen, da diese den maximalen Idealen in $k[X]$ entsprechen. Es besteht eine Bijektion zwischen den irreduziblen, nicht-konstanten, normierten Polynomen aus $k[X]$ und den abgeschlossenen Punkten von $\mathbf{A}_k^1$.

Ist $k$ algebraisch abgeschlossen, so besteht eine Bijektion zwischen den Elementen aus $k$ und den abgeschlossenen Punkten von $\mathbf{A}_k^1$.
\item[4.] Allgemeiner definieren wir den \textit{affinen $n$-dimensionalen Raum}\index{affiner Raum} über $k$ als: \[\mathbf{A}_k^n=\operatorname{Spec}k[X_1,\ldots,X_n]\]
\item[5.] Sei $k$ algebraisch abgeschlossen. Dann entsprechen die abgeschlossenen Punkte von $\mathbf{A}_k^n$ nach dem hilbertschen Nullstellensatz bijektiv den $n$-Tupeln von Elementen aus $k$. Ferner gibt es einen generischen Punkt $\xi$, der dem Nullideal in $k[X_1,\ldots,X_n]$ entspricht, d.h. $\overline{\{\xi\}}=\mathbf{A}_k^n$.
\end{enumerate}

\paragraph{Definition 2.10.}\label{2.10} \textit{(Offene Unterschemata)}\index{Schema!Unterschema!offen} Sei $(X,\mathcal{O}_X)$ ein Schema und $U\subset X$ offen. Dann ist $(U,\mathcal{O}_X|_U)$ ein Schema. Diese Aussage ist nichttrivial und wir werden sie \hyperref[2.10-beweis]{später} zeigen. Die offene Menge $U$ besitzt die \textit{induzierte Unterschemastruktur}.

\paragraph{Definition 2.11.}\label{2.11} \textit{(Verkleben von Schemata)}\index{Verklebung!Schema} Sei $\{X_i\}$ eine Familie von Schemata und $U_{ij}\subset X_i,\ i\neq j$ offene Teilmengen mit induzierter Struktur. Ferner haben wir für  $i\neq j$ Isomorphismen von Schemata:
\[\varphi_{ij}: (U_{ij},\mathcal{O}_{X_i}|_{U_{ij}}) \stackrel{\sim}{\to} (U_{ji},\mathcal{O}_{X_j}|_{U_{ji}}) \]
mit $\varphi_{ij}=\varphi_{ji}^{-1}$ und $\varphi_{ij}(U_{ij}\cap U_{ik})=U_{ji}\cap U_{jk}$ und $\varphi_{jk}=\varphi_{ik}\circ\varphi_{ij}$ auf $U_{ij}\cap U_{ik}$ für alle paar\-wei\-se verschiedene $i,j,k$. Wir erhalten ein Schema $X$ durch \textit{Verkleben} der $X_i$ längst $U_{ij}$ bzgl. $\varphi_{ij}$:
\[X=\dot{\bigcup_i}\ X_i\Big/{\sim},\quad x_i\sim\varphi_{ij}(x_i)\text{ für alle }x_i\in U_{ij},\ i\neq j \]
$X$ besitze die Quotiententopologie. Es existiert für jedes $j$ ein Morphismus $\psi_j:X_j\to X$ von Schemata, das ein Isomorphismus auf einem offenen Unterschema in $X$ induziert mit $X=\bigcup\psi_j(X_j)$ und $\psi_i(U_{ij})=\psi_i(X_i)\cap \psi_j(X_j)$ und $\psi_i=\psi_j\circ\varphi_{ij}$ auf $U_{ij}$ für alle $i\neq j$. Die Strukturgarbe auf $X$ ist folgendermaßen gegeben: Sei $V\subset_\text{o} X$. Setze:
\[\mathcal{O}_X(V)=\Big\{ (s_i)\in\prod_i \mathcal{O}_{X_i}(\psi_i^{-1}(V))\Bigm| \varphi_{ij}(s_i|_{\psi_i^{-1}(V)\cap U_{ij}}) = s_j|_{\psi_j^{-1}(V)\cap U_{ji}} \Big\} \]
Somit ist $(X,\mathcal{O}_X)$ ein lokal geringter Raum. Da alle $X_i$ Schemata sind, besitzt jeder Punkt von $X$ eine affine Umgebung. Also ist $X$ ein Schema.

\paragraph{Beispiel 2.12.}\label{2.12} Sei $k$ ein Körper, $X_1=X_2=\mathbf{A}_k^1$ und $U_1=U_2=\mathbf{A}_k^1\setminus\{P\}$ mit einem abgeschlossenen Punkt $P$. Ist $\varphi:U_1\to U_2$ die Identität, so ist die Verklebung $X$ von $X_1$ und $X_2$ längst $\varphi$ die affine Gerade, wobei der Punkt $P$ verdoppelt wurde. $X$ ist selbst nicht mehr affin.

\paragraph{Satz 2.13.}\label{2.13} Sei $A$ ein Ring und $(X,\mathcal{O}_X)$ ein Schema. Dann ist die Abbildung bijektiv:
\[\alpha:\operatorname{Hom}_\mathbf{Sch}(X,\operatorname{Spec}A)\to \operatorname{Hom}_\textbf{Ringe}(A,\Gamma(X,\mathcal{O}_X)) \]
wobei wir $(f:X\to\operatorname{Spec}A ,\ f^\sharp:\mathcal{O}_A\to f_\ast\mathcal{O}_X)$ durch das Nehmen der globalen Schnitte auf den Ringhomomorphismus $A=\Gamma(\operatorname{Spec}A,\mathcal{O}_A)\to\Gamma(X,\mathcal{O}_X)$ schicken.

\paragraph{Beweis.} Sei $X=\bigcup_\nu X_\nu$ eine affine Überdeckung. Ein Morphismus $(f,f^\sharp)$ ist eindeutig durch seine Einschränkungen $(f_\nu,f_\nu^\sharp)$ auf $X_\nu$ bestimmt. Diese sind nach \hyperref[2.6]{Satz 2.6} wiederum eindeutig bestimmt durch $\alpha(f_\nu)$. Ferner kommutiert das folgende Diagramm:
\[\begin{tikzcd}
A\ar[d, "\alpha(f)"']\ar[rr, "\alpha(f_\nu)"] && \Gamma(X_\nu,\mathcal{O}_{X_\nu})\\
\Gamma(X,\mathcal{O}_X)\ar[rru, "\operatorname{res}"']
\end{tikzcd} \]
weshalb $\alpha(f)$ schon alle $\alpha(f_\nu)$ eindeutig bestimmt. Somit ist $\alpha$ injektiv. Sei nun ein Ringhomomorphismus $h:A\to\Gamma(X,\mathcal{O}_X)$ gegeben und $h_\nu:A\to\Gamma(X,\mathcal{O}_X)\to\Gamma(X_\nu,\mathcal{O}_X)$. Nach \hyperref[2.6]{Satz 2.6 (iii)} existiert ein $f_\nu:X_\nu\to\operatorname{Spec}(A)$ mit $\alpha(f_\nu)=h_\nu$. Für alle $\nu,\mu$ ist das folgende Diagramm kommutativ:
\[\begin{tikzcd}
& \Gamma(X_\nu,\mathcal{O}_X)\ar[dr, "\operatorname{res}"] &\\
A\ar[ur, "h_\nu"] \ar[dr, "h_\mu"'] & & \Gamma(X_\nu\cap X_\mu,\mathcal{O}_X)\\
& \Gamma(X_\mu,\mathcal{O}_X)\ar[ur, "\operatorname{res}"'] &
\end{tikzcd} \]
Aus der Injektivität von $\alpha$ folgt $f_\nu=f_\mu$ auf $X_\nu\cap X_\mu$. Kleben wir die $f_\nu$ nun zusammen, so erhalten wir ein $f:X\to\operatorname{Spec}(A)$ mit $\alpha(f)=h$.\qed

\paragraph{Korollar 2.14.}\label{2.14} Es gibt eine pfeilumkehrende Kategorienäquivalenz zwischen der Kategorie der affinen Schemata und der Kategorie der kommutativen Ringe mit $1$.

\paragraph{Korollar 2.15.}\label{2.15} $\operatorname{Spec}(\mathbb{Z})$ ist Endobjekt in der Kategorie der Schemata, d.h. für jedes Schema $X$ existiert ein eindeutiger Morphismus $f:X\to\operatorname{Spec}(\mathbb{Z})$.

\paragraph{Satz 2.16.}\label{2.16} Sei $A$ ein Ring, $X=\operatorname{Spec}(A)$ und $f\in A$. Dann gilt:
\[(D(f),\mathcal{O}_X|_{D(f)})\cong \operatorname{Spec}(A_f) \]

\paragraph{Beweis.} Die natürliche Abbildung $\varphi:A\to A_f$ induziert einen Homöomorphismus:
\[\psi:\operatorname{Spec}(A_f) \to \{\mathfrak{p}\in\operatorname{Spec}(A)\mid f\not\in\mathfrak{p} \}=D(f),\ \mathfrak{p}\mapsto\varphi^{-1}(\mathfrak{p})\qedhere \]

\paragraph{Beweis zu \hyperref[2.10]{Definition 2.10}.}\label{2.10-beweis} Sei $(X,\mathcal{O}_X)$ ein Schema und $U\subset X$ offen. Dann ist $(U,\mathcal{O}_X|_U)$ ein lokal geringter Raum. Sei $P\in U$. Wir zeigen, dass eine Umgebung $V\subset_\text{o} U$ von $P$ existiert, so dass $(V,\mathcal{O}_X|_V)$ affin ist. Da $(X,\mathcal{O}_X)$ ein Schema ist, existiert ein $V'\subset_\text{o}X,\ P\in V'$ mit $(V',\mathcal{O}_X|_{V'})$ affin. Sei also $V'=\operatorname{Spec}(A)$ für einen Ring $A$. Da die $D(f),\ f\in A$ eine Basis der Topologie auf $X$ bilden, existiert ein $f\in A$, so dass $P\in D(f)\subset V'\cap U$. Wegen \hyperref[2.16]{Satz 2.16} ist $(D(f),\mathcal{O}_X|_{D(f)})\cong\operatorname{Spec}(A_f)$ affin.\qed

\paragraph{Satz 2.17.}\label{2.17} Sei $X$ ein Schema und $Z\subset X$ irreduzibel und abgeschlossen. Dann existiert genau ein Punkt $\xi\in Z$ derart, dass $\overline{\{\xi\}}=Z$. $\xi$ heißt \textit{generischer Punkt}\index{generischer Punkt} von $Z$.

\paragraph{Beweis.}\begin{itemize}
\item \textit{Existenz:} Sei $U\subset_\text{o}X$ affin mit $Z\cap U\neq\varnothing$. Da $Z\cap U$ abgeschlossen in $U$ ist, gibt es ein Radikalideal $\mathfrak{p}$ mit $Z\cap U=V(\mathfrak{p})$.

Nun ist $Z\cap U$ irreduzibel, da für offene Mengen $V_1,V_2\subset_\text{o} Z\cap U$ stets $V_1,V_2\subset_\text{o}Z$ gilt und aus der Irreduzibilität von $Z$ stets $V_1\cap V_2\neq\varnothing$ folgt.

Ferner ist $\mathfrak{p}$ ein Primideal, denn ist $\mathfrak{ab}\subset\mathfrak{p}$ für gewisse Ideale $\mathfrak{a}$ und $\mathfrak{b}$, so folgt $\operatorname{Rad}(\mathfrak{ab})\subset\operatorname{Rad}(\mathfrak{p})=\mathfrak{p}$. Dies ist äquivalent zu $V(\mathfrak{p})\subset V(\mathfrak{ab})=V(\mathfrak{a})\cup V(\mathfrak{b})$, daher:
\[Z\cap U =V(\mathfrak{p})=(V(\mathfrak{p})\cap V(\mathfrak{a}))\cup (V(\mathfrak{p})\cap V(\mathfrak{b})) \]
Da $Z\cap U$ irreduzibel ist, folgt o.B.d.A. $V(\mathfrak{p})=V(\mathfrak{p})\cap V(\mathfrak{a})\subset V(\mathfrak{a})$, d.h. $\mathfrak{a}\subset\operatorname{Rad}(\mathfrak{a})\subset\operatorname{Rad}(\mathfrak{p})=\mathfrak{p}$.

$\mathfrak{p}$ ist nun generischer Punkt von $Z\cap U$, da:
\[\overline{\{\mathfrak{p}\}}=\bigcap_{\mathfrak{p}\in V(\mathfrak{a})}V(\mathfrak{a})=\bigcap_{\mathfrak{a}\subset\mathfrak{p}} V(\mathfrak{a})\subset V(\mathfrak{p}) \]
Die andere Inklusion folgt aus $\mathfrak{a}\subset\mathfrak{p}$. Da $\mathfrak{p}$ prim ist, folgt $\operatorname{Rad}(\mathfrak{a})\subset\mathfrak{p}$ und somit $V(\mathfrak{p})\subset V(\mathfrak{a})$. Somit ist $Z\cap U=\overline{\{\mathfrak{p}\}}$.

Bezeichne mit $\overline{\{\mathfrak{p}\}}^X$ den topologischen Abschluss in $X$. Dann ist $Z\cap U\subset\overline{\{\mathfrak{p}\}}^X$. Da $Z$ irreduzibel ist, gilt $Z=\overline{Z\cap U}\subset\overline{\{\mathfrak{p}\}}^X$, also $Z=\overline{\{\mathfrak{p}\}}^X$.
\item \textit{Eindeutigkeit:} Sei $\overline{\{\xi\}}^X=Z=\overline{\{\xi'\}}^X$. Sei $U\subset_\text{o}X$ eine affine Umgebung von $\xi\in U$. Dann ist $\xi'\in U$, da aus $\xi'\in Z\setminus U$ der Widerspruch $\xi\in Z=\overline{\{\xi'\}}^X\subset X\setminus U$ folgt. Wähle nun Radikalideale $\mathfrak{p}'$ und $\mathfrak{p}$ mit:
\[V(\mathfrak{p}')= \overline{\{\xi'\}}^U=U\cap \overline{\{\xi'\}}^X=U\cap\overline{\{\xi\}}^X=\overline{\{\xi\}}^U=V(\mathfrak{p}) \]
Es folgt $\xi'=\mathfrak{p}'=\operatorname{Rad}(\mathfrak{p}')=\operatorname{Rad}(\mathfrak{p})=\mathfrak{p}=\xi$.\qed
\end{itemize}

\paragraph{Satz 2.18.}\label{2.18} Sei $X$ ein Schema und $K$ ein Körper. Dann gibt es eine Bijektion:
\[\operatorname{Hom}_\mathbf{Sch}(\operatorname{Spec}K,X)\stackrel{\sim}{\to} \{(x,i)\mid x\in X,\ i:\kappa(x)\hookrightarrow K\text{ Ringhomomorphismus} \} \]
wobei $\kappa(x)=\mathcal{O}_{X,x}/\mathfrak{m}_x$ der Restklassenkörper von $\mathcal{O}_{X,x}$ bezeichnet. Die Elemente heißen \textit{$K$-wertige Punkte} von $X$.\index{Punkt!$K$-wertig}

\paragraph{Beweis.} Sei ein Morphismus $f:\operatorname{Spec}(K)\to X$ gegeben. Setze $x=f((0))\in X$. $f^\sharp$ induziert einen lokalen Homomorphismus $\mathcal{O}_{X,x}\to\mathcal{O}_{K,(0)}=K$. $f^\sharp$ faktorisiert daher über $\mathcal{O}_{X,x}/\mathfrak{m}_x=\kappa(x)$ und induziert einen Homomorphismus $i:\kappa(x)\hookrightarrow K$.

Sei nun umgekehrt ein $x\in X$ und $i:\kappa(x)\hookrightarrow K$ gegeben. $i$ definiert nach \hyperref[2.6]{Satz 2.6} ein Schemamorphismus:
\[f:\operatorname{Spec}(K)\to\operatorname{Spec}\kappa(x)\to\operatorname{Spec}(\mathcal{O}_{X,x})\stackrel{\psi}{\to} X \]
wobei $\psi$ die folgende kanonische Abbildung ist:

Sei $U\subset_\text{o}X$ eine affine Umgebung von $x$ mit $U=\operatorname{Spec}(A)$. Nach \hyperref[2.3]{Satz 2.3 (i)} gilt $\mathcal{O}_{X,x}=\mathcal{O}_{U,x}=A_x$. Die kanonische Abbildung $A\to A_x$ induziert $\psi:\operatorname{Spec}(\mathcal{O}_{X,x})\to U\hookrightarrow X$. Diese Abbildung ist unabhängig von $U$:

Sei $U'\subset_\text{o}X$ eine weitere affine Umgebung von $x$. Dann existiert eine affine Umgebung $U''\subset_\text{o}U\cap U'$ mit $x\in U''$, also können wir o.B.d.A. $\operatorname{Spec}(A)= U\subset U'=\operatorname{Spec}(A')$ annehmen. Es existiert ein kanonischer Homomorphismus $A'\to A$ derart, dass das folgende Diagramm kommutiert:
\[\begin{tikzcd}
A'\ar[rd]\ar[r] & A\ar[d]\\
& A'_x= \mathcal{O}_{X,x}= A_x &
\end{tikzcd} \]
Somit kommutiert:
\[\begin{tikzcd}
X & U'\ar[l] & U\ar[l]\\
 && \operatorname{Spec}(\mathcal{O}_{X,x})\ar[lu]\ar[u]
\end{tikzcd}\qedhere\]

\paragraph{Satz 2.19.}\label{2.19} \begin{enumerate}[(i)]
\item Sei $A$ ein Ring und $f\in A$. Dann gilt:
\[D(f)=\varnothing\iff f\text{ nilpotent} \]
\item Sei $\varphi:A\to B$ ein Ringhomomorphismus und $f: Y=\operatorname{Spec}(B)\to\operatorname{Spec}(A)=X$ der durch $\varphi$ induzierte Morphismus affiner Schemata. Dann gilt:
\begin{enumerate}[(a)]
\item $\varphi$ ist genau dann injektiv, wenn $f^\sharp:\mathcal{O}_X\to f_\ast\mathcal{O}_Y$ injektiv ist. In diesem Fall ist $f$ \textit{dominant}\index{dominant}, d.h. $f(Y)\subset X$ ist dicht.
\item $\varphi$ ist genau dann surjektiv, wenn $f^\sharp$ surjektiv ist und $f$ ein Homöomorphismus auf eine abgeschlossene Teilmenge von $X$ ist.
\end{enumerate}
\end{enumerate}

\paragraph{Beweis.} Übung.\qed

% \paragraph{Beispiel.} Das Spektrum $T=\operatorname{Spec}( k[X,Y]/(X^2))$ entspricht der $Y$-Achse, also besitzt der Ring $\mathcal{O}_T(T)= k[X,Y]/(X^2)$ nilpotente Elemente.

\paragraph{Definition 2.20.}\label{2.20} Ein Schema $(X,\mathcal{O}_X)$ heißt \textit{reduziert}\index{Schema!reduziert}, falls $\mathcal{O}_X(U)$ für alle $U\subset_\text{o}X$ reduziert sind, d.h. keine nilpotente Elemente besitzt.

\paragraph{Regeln 2.21.}\label{2.21} Sei $(X,\mathcal{O}_X)$ ein Schema.
\begin{enumerate}[(i)]
\item $(X, \mathcal{O}_X)$ ist genau dann reduziert, wenn $\mathcal{O}_{X,P}$ für alle $P\in X$ keine nilpotente Elemente besitzt.
\item Sei $\mathcal{O}_X^\text{red}$ die assoziierte Garbe zur folgenden Prägarbe:
\[ U \mapsto \mathcal{O}_X(U)/\mathfrak{N}_U \]
wobei $\mathfrak{N}_U$ das Nilradikal von $\mathcal{O}_X(U)$ bezeichnet. Dann ist $(X,\mathcal{O}_X^\text{red})$ ein Schema, das zu $X$ assoziierte \textit{reduzierte Schema}\index{Schema!reduziert} $X_\text{red}$. Es gibt einen Morphismus $f:X_\text{red}\to X$ mit dem Homöomorphismus $\operatorname{id}$ auf den unterliegenden topologischen Räumen und $f^\sharp:\mathcal{O}_X\to f_\ast\mathcal{O}_X^\text{red}$ gegeben durch:
\[\mathcal{O}_X(U)\to\mathcal{O}_X^\text{red}(U),\ s\mapsto \Big(U\stackrel{s}{\to}\coprod_{P\in U}\mathcal{O}_{X,P} \to\coprod_{P\in U}\mathcal{O}_{X,P}^\text{red}\Big) \]
\item Sei $f:X\to Y$ ein Morphismus von Schemata mit $X$ reduziert. Setze $g=f$ auf den unterliegenden topologischen Räumen. Da $X$ reduziert ist, faktorisiert $f^\sharp:\mathcal{O}_Y\to f_\ast\mathcal{O}_X$ über $\mathcal{O}_Y^\text{red}$. $f^\sharp$ induziert $g^\sharp:\mathcal{O}_Y^\text{red}\to g_\ast\mathcal{O}_X=f_\ast\mathcal{O}_X$. Es gibt also einen eindeutig bestimmten Morphismus $g:X\to Y_\text{red}$ mit kommutativem Diagramm:
\[\begin{tikzcd}
X\ar[rr, "f"]\ar[drr, "g"'] && Y\\
&& Y_\text{red}\ar[u]
\end{tikzcd} \]
\end{enumerate}

\paragraph{} Sei $S=\bigoplus_{d\geq 0} S_d$ ein graduierter Ring und $S_+=\bigoplus_{d>0}S_d$. Ein Primideal $\mathfrak{p}\subset S$ ist genau dann homogen, wenn aus $fg\in\mathfrak{p}$ für gewisse homogene Elemente $f,g\in S$ stets $f\in\mathfrak{p}$ oder $g\in\mathfrak{p}$ folgt. Für ein homogenes Ideal $\mathfrak{a}\subset S$ setze:
\[\operatorname{Proj}(S)=\{\mathfrak{p}\subset S\text{ homogenes Primideal}\mid S_+\not\subset\mathfrak{p}\},\quad V(\mathfrak{a})=\{\mathfrak{p}\in\operatorname{Proj}(S)\mid\mathfrak{a}\subset\mathfrak{p}\} \]

\paragraph{Lemma 2.22.}\label{2.22}\begin{enumerate}[(i)]
\item Sind $\mathfrak{a},\mathfrak{b}\subset S$ homogene Ideale, so gilt $V(\mathfrak{ab})=V(\mathfrak{a})\cup V(\mathfrak{b})$.
\item Ist $(\mathfrak{a}_i)_i$ eine Familie homogener Ideale in $S$, so folgt $V\big(\sum\mathfrak{a}_i\big)=\bigcap V(\mathfrak{a}_i)$.
\end{enumerate}
Damit wird auf $\operatorname{Proj}(S)$ eine Topologie definiert. Die abgeschlossenen Mengen sind genau die Mengen der Form $V(\mathfrak{a})$ für ein homogenes Ideal $\mathfrak{a}\subset S$.

\paragraph{Beweis.} Wie in \hyperref[2.1]{Lemma 2.1} unter Beachtung, dass homogene Ideale von homogene Elemente erzeugt werden.\qed

\paragraph{Definition.} Sei $T\subset S$ multiplikativ abgeschlossen, die aus homogenen Elementen besteht. Dann wird $T^{-1}S = \bigoplus_{i\geq 0} (T^{-1}S)_i$ zu einem graduierten Ring:
\[(T^{-1}S)_i=\Big\{\frac{s}{t}\in T^{-1}S\Bigm| s\in S\text{ homogen},\ t\in T,\ \deg(s)-\deg(t)=i \Big\} \]
Ist $\mathfrak{p}\subset S$ ein homogenes Primideal und $f\in S$ ein homogenes Element, so ist die \textit{homogene Lokalisierung}\index{homogene Lokalisierung} bzgl. $\mathfrak{p}$ bzw. $f$ definiert als:
\[S_{(\mathfrak{p})} = (S_\mathfrak{p})_0,\quad S_{(f)} = (S_f)_0 \]

\paragraph{Definition.} Wir definieren eine Ringgarbe $\mathcal{O}$ auf $\operatorname{Proj}(S)$ wie folgt: Sei $U\subset_\text{o}\operatorname{Proj}(S)$ und setze $\mathcal{O}(U)$ als die Menge aller Abbildungen $s:U\to\coprod_{\mathfrak{p}\in U}S_{(\mathfrak{p})}$, so dass:
\begin{enumerate}[(i)]
\item Für alle $\mathfrak{p}\in U$ gilt $s(\mathfrak{p})\in S_{(\mathfrak{p})}$.
\item Für alle $\mathfrak{p}\in U$ existiert eine offene Umgebung $V$ von $\mathfrak{p}$ mit $V\subset U$ und homogene Elemente $a,f\in S$ mit $\deg(a)=\deg(f)$ derart, dass für alle $\mathfrak{q}\in V$ stets $f\not\in\mathfrak{q}$ und $s(\mathfrak{q})=\frac{a}{f}$ in $S_{(\mathfrak{q})}$ gilt.
\end{enumerate}

\paragraph{Satz 2.23.}\label{2.23} Sei $S$ ein graduierter Ring. Dann gilt:
\begin{enumerate}[(i)]
\item $\mathcal{O}_\mathfrak{p}\cong S_{(\mathfrak{p})}$ für alle $\mathfrak{p}\in\operatorname{Proj}(S)$.
\item Für ein homogenes $f\in S_+$ setze $D_+(f)=\{\mathfrak{p}\in\operatorname{Proj}(S)\mid f\not\in\mathfrak{p}\}$. Dann ist $D_+(f)$ offen in $\operatorname{Proj}(S)$ und es gilt:
\[\operatorname{Proj}(S) =\bigcup_{f\in S_+\text{ homogen}}D_+(f) \]
Es gibt einen Isomorphismus lokal geringter Räume $(D_+(f),\mathcal{O}|_{D_+(f)})\cong\operatorname{Spec}S_{(f)}$.
\item $(\operatorname{Proj}S,\mathcal{O})$ ist ein Schema.
\end{enumerate}

\paragraph{Beweis.}\begin{enumerate}[(i)]
\item Die Abbildung $\mathcal{O}_\mathfrak{p}\to S_{(\mathfrak{p})},\ s_\mathfrak{p}\mapsto s(\mathfrak{p})$, wobei $s$ ein Repräsentant von $s_\mathfrak{p}$ ist, ist ein Isomorphismus. Beweis analog wie \hyperref[2.3]{Satz 2.3 (i)}.
\item Da $D_+(f)=\operatorname{Proj}(S)\setminus V(f)$, ist $D_+(f)$ offen. Sei $\mathfrak{p}\in\operatorname{Proj}(S)$, d.h. $\mathfrak{p}\subset S$ ist ein homogenes Primideal mit $S_+\not\subset\mathfrak{p}$. Sei $f\in S_+\setminus\mathfrak{p}$. Dann ist $\mathfrak{p}\not\in V(f)$, also $\mathfrak{p}\in D_+(f)$. Daher ist $\operatorname{Proj}(S)=\bigcup D_+(f)$.

Sei $f\in S_+$. Wir definieren ein Morphismus lokal geringter Räume $(\phi,\phi^\sharp): D_+(f)\to\operatorname{Spec}S_{(f)}$ wie folgt: Sei $S\rightarrow S_f$ der natürliche Homomorphismus. Sei $\mathfrak{a}\subset S$ ein homogenes Ideal und setze $\phi(\mathfrak{a})=\mathfrak{a}S_f\cap S_{(f)}$. Beachte, dass $S_{(f)}=(S_f)_0\subset S_f$ ein Teilring ist. Für $\mathfrak{p}\in D_+(f)$ ist $\phi(\mathfrak{p})\in\operatorname{Spec}S_{(f)}$, siehe \hyperref[2.16]{Satz 2.16}, und $\phi$ ist bijektiv. Sei $\mathfrak{a}\subset S$ ein homogenes Ideal. Dann ist $\mathfrak{p}\supset\mathfrak{a}$ genau dann, wenn $\phi(\mathfrak{p})\supset\phi(\mathfrak{a})$. Daher ist $\phi$ ein Homöomorphismus.

$\phi^\sharp:\mathcal{O}_{\operatorname{Spec} S_{(f)}}\to\phi_\ast \big(\mathcal{O}_{\operatorname{Proj}(S)}|_{D_+(f)}\big)$ wird wie folgt definiert:
\[\mathcal{O}_{\operatorname{Spec}S_{(f)}}(U)\to\mathcal{O}_{\operatorname{Proj}(S)}(\phi^{-1}(U)),\ s\mapsto \Big(\phi^{-1}(U)\stackrel{s\circ\phi}{\longrightarrow}\coprod (S_{(f)})_{\phi(\mathfrak{p})} \cong\coprod S_{(\mathfrak{p})}\Big) \]
Dieses ist ein Isomorphismus.
\item folgt aus (i) und (ii).\qed
\end{enumerate}

\paragraph{Beispiel 2.24.}\label{2.24} Sei $A$ ein Ring. Dann heißt
\[\mathbf{P}_A^n = \operatorname{Proj} A[X_0,\ldots,X_n] \]
der \textit{$n$-dimensionaler projektiver Raum}\index{projektiver Raum} über $A$. Ist speziell $A=k$ ein algebraisch abgeschlossener Körper, so ist $\mathbf{P}_k^n$ ein Schema. Dessen Teilraum aller abgeschlossenen Punkte ist homöomorph zur projektiven $n$-dimensionalen Varietät.

\paragraph{Definition 2.25.} \label{2.25} Sei $S$ ein beliebiges Schema. Ein \textit{Schema über $S$}\index{Schema!über $S$} ist ein Schema $X$ zusammen mit einem Morphismus $X\to S$, der sogenannte \textit{Strukturmorphismus}\index{Strukturmorphismus}. Ein Morphismus zweier Schemata $X$ und $Y$ über $S$ ist ein Morphismus $f:X\to X'$, so dass das folgende Diagramm kommutiert:
\[\begin{tikzcd}
X\ar[rr, "f"]\ar[d] && X'\ar[lld]\\
S
\end{tikzcd} \]
So ein Morphismus nennt man auch \textit{$S$-Morphismus}\index{Morphismus!$S$-Schemata}. Bezeichne die Kategorie aller Schemata über $S$ mit $S$-Morphismen mit $\mathbf{Sch}(S)$. Für einen Ring $A$ setzen wir auch $\mathbf{Sch}(A)=\mathbf{Sch}(\operatorname{Spec}A)$.

\paragraph{Satz 2.26.}\label{2.26} Sei $k$ ein algebraisch abgeschlossener Körper. Dann gibt es einen natürlichen Funktor
\[t:\mathbf{Var}(k)\to\mathbf{Sch}(k) \]
der \textit{volltreu}\index{volltreuer Funktor} ist, d.h. für zwei Varietäten $V,W$ ist die durch $t$ auf den Morphismen induzierte Abbildung $\operatorname{Hom}_{\mathbf{Var}(k)}(V,W)\to \operatorname{Hom}_{\mathbf{Sch}(k)}(tV,tW)$ bijektiv. Für eine Varietät $V$ setze $\mathfrak{M}(V)$ als die Menge aller abgeschlossenen Punkte des Schemas $tV$ mit der Teilraumtopologie. Es gibt einen Homöomorphismus topologischer Räume $V\cong\mathfrak{M}(V)$. Die Garbe der regulären Funktionen ist via diesen Homöomorphismus isomorph zu $\mathcal{O}_{tV}|_{\mathfrak{M}(V)}$.

\paragraph{Beweis.} Siehe z.B. Hartshorne Kapitel II, Proposition 2.6 oder Mumford, Theorem 2 auf Seite 168.

\section{Erste Eigenschaften von Schemata}

\paragraph{Definition 3.1.}\label{3.1} \begin{enumerate}[(i)]
\item Ein Schema heißt \textit{zusammenhängend}\index{Schema!zusammenhängend}, falls es als topologischer Raum zu\-sam\-men\-hän\-gend ist.
\item Ein Schema heißt \textit{irreduzibel}\index{Schema!irreduzibel}, falls es als topologischer Raum irreduzibel ist.
\item Ein Schema heißt \textit{reduziert}\index{Schema!reduziert} falls für alle $U\subset_\text{o}X$ der Ring $\mathcal{O}_X(U)$ reduziert ist.
\item Ein Schema heißt \textit{integer}\index{Schema!integer}, falls für alle $U\subset_\text{o}X$ der Ring $\mathcal{O}_X(U)$ nullteilerfrei ist.
\end{enumerate}

\paragraph{Beispiel 3.2.}\label{3.2} Sei $X=\operatorname{Spec}(A)$ ein affines Schema. Dann gilt:
\begin{enumerate}[(i)]
\item $X$ ist irreduzibel $\iff \mathfrak{N}(A)$ ist ein Primideal
\item $X$ ist reduziert $\iff A$ ist reduziert $\iff\mathfrak{N}(A)=0$
\item $X$ ist integer $\iff A$ ist nullteilerfrei
\end{enumerate}

\paragraph{Beweis.} (i) und (ii) sind klar. Ist $X$ integer, so ist $A=\mathcal{O}_X(X)$ nullteilerfrei. Sei nun umgekehrt $A$ nullteilerfrei, d.h. $\mathfrak{N}(A)=0$ ist ein Primideal. Nach (i) und (ii) ist $X$ irreduzibel und reduziert. Daher folgt die Aussage aus dem nächsten Satz.\qed

\paragraph{Satz 3.3.}\label{3.3} Ein Schema $X$ ist genau dann integer, wenn $X$ reduziert und irreduzibel ist.

\paragraph{Beweis.} Sei $X$ integer. Dann ist $X$ offensichtlich reduziert. Wäre $X$ nicht irreduzibel, so gäbe es $\varnothing\neq U_1,U_2\subset_\text{o}X$ mit $U_1\cap U_2=\varnothing$. Dann ist:
\[\mathcal{O}_X(U_1\cup U_2)=\mathcal{O}_X(U_1)\times\mathcal{O}_X(U_2) \]
Somit ist $\mathcal{O}_X(U_1\cup U_2)$ nicht nullteilerfrei. Sei nun umgekehrt $X$ irreduzibel und reduziert.

Sei $V\subset_\text{o}X$ affin mit $V=\operatorname{Spec}(A)$. Sei $a,b\in A$ mit $ab=0$. Es folgt:
\[\operatorname{Spec}(A)=V(ab)=V(a)\cup V(b) \]
Da $V$ irreduzibel ist, folgt o.B.d.A. $\operatorname{Spec}(A)=V(a)$. Da $A=\mathcal{O}_X(V)$ reduziert ist, folgt $\operatorname{Rad}(a)=(0)$, also $a=0$. Daher ist $\mathcal{O}_X(V)$ für jedes affine $V\subset_\text{o}X$ nullteilerfrei.

Sei nun $U\subset_\text{o}X$ beliebig und $f,g\in\mathcal{O}_X(U)$ mit $fg=0$. Für $V\subset_\text{o}U$ affin, folgt aus $f|_V\cdot g|_V=0$ o.B.d.A. $f|_V=0$. Nun ist $U$ der Abschluss von $V$ in $U$. Sei $x\in U$ und $U(x)\subset_\text{o}U$ eine affine Umgebung von $x$ mit $U(x)=\operatorname{Spec}(B)$. Sei $f(x)=\frac{a}{h}\in B_x$ für alle $x\in U(x)$. Es ist $U(x)\cap V\neq\varnothing$, da $U$ irreduzibel ist. Wähle ein $y\in U(x)\cap V$; es folgt $0=f(y)=\frac{a}{h}\in B_y$, also gibt es ein $k\in B\setminus y$ mit $ka=0$. Da $B$ nullteilerfrei ist, folgt $a=0$ und $f=0$ auf $U(x)$. Somit folgt $f=0$.\qed

\paragraph{Definition.} \begin{enumerate}[(i)]
\item Ein Schema $X$ heißt \textit{quasikompakt}\index{quasikompakt}, wenn sein unterliegender topologischer Raum quasikompakt ist.
\item Ein topologischer Raum $X$ heißt \textit{noethersch}\index{noethersch}, wenn jede absteigende Kette von abgeschlossenen Teilmengen in $X$ stationär wird.
\end{enumerate}

\paragraph{Bemerkung.} Sei $X$ ein noetherscher Raum. Nach Zorns Lemma besitzt jede nichtleere Menge $\Sigma$ von abgeschlossenen Mengen in $X$ ein minimales Element, da jede Kette in $\Sigma$ ein minimales Element besitzt.

\paragraph{Satz 3.4.}\label{3.4} \begin{enumerate}[(i)]
\item Sei $X$ ein topologischer Raum. Dann ist $X$ genau dann noethersch, wenn alle offenen Teilmengen $U\subset X$ quasikompakt sind.
\item Sei $X$ ein affines Schema. Dann ist $X$ quasikompakt, aber nicht notwendig noethersch.
\item Sei $A$ ein noetherscher Ring. Dann ist der $\operatorname{Spec}(A)$ unterliegender Raum noethersch.
\end{enumerate}

\paragraph{Beweis.}\begin{enumerate}[(i)]
\item Sei $U\subset_\text{o}X$ und $U=\bigcup_{i\in I}U_i$ eine offene Überdeckung. Für eine endliche Teilmenge $J\subset I$ setze $V_J=\bigcup_{i\in J}U_i$. Dann ist $V_J\subset X$ offen und es gilt:
\[U=\bigcup_{J\subset I\text{ endlich}}V_J \]
Wählt man aus $\Sigma=\{X\setminus V_J\mid J\subset I\text{ endlich}\}$ ein minimales Element $X\setminus V_{J'}$. Dann gilt $V_{J'}\supset V_J$ für alle $J$. Also ist $U=V_{J'}=\bigcup_{i\in J'}U_i$ eine endliche Teilüberdeckung.

Sei umgekehrt $Y_1\supset Y_2\supset\ldots$ eine Kette abgeschlossener Mengen in $X$, so ist die Menge $U=\bigcup_{j\geq 1}X\setminus Y_j$ offen in $X$. Wir erhalten eine endliche Teilüberdeckung $U=\bigcup_{r\geq j\geq 1}X\setminus Y_j=X\setminus Y_r$, also folgt $Y_s=Y_r$ für alle $s\geq r$.
\item Sei $X=\operatorname{Spec}(A)=\bigcup_{i\in I}U_i$ eine offene Überdeckung. Wir können o.B.d.A. $U_i=D(f_i)$ für gewisse $f_i\in A$ annehmen. Sei $\mathfrak{a}=(f_i\mid i\in I)\subset A$. Dann gilt:
\[X=\bigcup_{i\in I}X\setminus V(f_i) = X\setminus\bigcap_{i\in I}V(f_i)=X\setminus V(\mathfrak{a}) \]
Es folgt $V(\mathfrak{a})=\varnothing$, also $1\in\mathfrak{a}$. Somit gibt es endlich viele $g_j\in A$ und $i_j\in I$ mit $1=\sum_j g_jf_{i_j}$. Wir erhalten die endliche Teilüberdeckung $X=\bigcup_j D(f_{i_j})$.
\item Sei $V(\mathfrak{a}_1)\supset V(\mathfrak{a}_2)\supset\ldots$ eine Kette abgeschlossener Mengen in $\operatorname{Spec}(A)$ mit Ra\-di\-kal\-idealen $\mathfrak{a}_i\subset A$. Sie wird stationär, da die Kette $\mathfrak{a}_1\subset \mathfrak{a}_2\subset\ldots$ stationär wird.\qed
\end{enumerate}

\paragraph{Definition 3.5.}\label{3.5} Sei $X$ ein Schema.
\begin{enumerate}[(i)]
\item $X$ heißt \textit{lokal noethersch}\index{noethersch!lokal}, falls $X$ von offenen, affinen Teilmengen $\operatorname{Spec}(A_i)$ mit noetherschen Ringen $A_i$ überdeckt werden kann.
\item $X$ heißt \textit{noethersch}\index{noethersch}, falls $X$ lokal noethersch und quasikompakt ist. Dies ist äquivalent dazu, dass $X$ von endlich vielen offenen, affinen Teilmengen $\operatorname{Spec}(A_i)$ mit noetherschen Ringen $A_i$ überdeckt werden kann.
\end{enumerate}

\paragraph{Bemerkung.} Ist ein Schema $X$ noethersch, so ist nach \hyperref[3.4]{Satz 3.4 (iii)} der unterliegender Raum von $X$ noethersch. Die Umkehrung gilt im Allgemeinen nicht.

\paragraph{Satz 3.6.}\label{3.6} Sei $X$ ein Schema. Dann ist $X$ genau dann lokal noethersch, wenn für alle offenen, affinen Teilmengen $U=\operatorname{Spec}(A)$ stets $A$ ein noetherscher Ring ist.

\paragraph{Beweis.} Die Rückrichtung ist trivial. Sei also $X$ lokal noethersch und $U=\operatorname{Spec}(A)$ offen in $X$. Wir haben eine offene affine Überdeckung $X=\bigcup_i\operatorname{Spec}(B_i)$ mit noetherschen Ringen $B_i$. Da die offenen Mengen $D(f)$ eine offene Basis der Topologie bilden, haben wir eine Darstellung:
\[U=\bigcup_{i,j} D(f_{ij}),\quad f_{ij}\in B_i,\ D(f_{ij})\subset U\cap \operatorname{Spec}(B_i) \]
mit $D(f_{ij})=\operatorname{Spec}(B_i)_{f_{ij}}$. Da $B_i$ noethersch sind, sind die Lokalisierungen $(B_i)_{f_{ij}}$ ebenfalls noethersche Ringe. Da $U$ affin und somit quasikompakt ist, kann $U$ von endlich vielen Spektren noetherscher Ringe überdeckt werden.

Sei $V=\operatorname{Spec}(B)$ offen in $U$ mit noetherschen Ring $B$. Sei $f\in A$ und betrachte $D(f)\subset V$. Die natürliche Inklusion $V\hookrightarrow U$ induziert einen Ringhomomorphismus $A\to B$. Sei $\overline{f}$ das Bild von $f$ in $B$. Es gilt:
\[\operatorname{Spec}(A_f)=D(f)=D(\overline{f})=\operatorname{Spec}(B_{\overline{f}})\]
Es folgt $A_f\cong B_{\overline{f}}$ und $A_f$ ist noethersch. Wir haben nun gezeigt, dass $U$ von endlich vielen offenen Mengen der Form $D(f)=\operatorname{Spec}(A_f)$ überdeckt werden kann mit noetherschen Ringen $A_f$. Somit folgt die Aussage aus dem nächsten Lemma.\qed

\paragraph{Lemma 1.}\label{3.6-lemma1} Sei $A$ ein Ring und $f_1,\ldots,f_r\in A$ mit $1=(f_1,\ldots,f_r)$. Sind alle $A_{f_i}$ noethersch, so ist auch $A$ noethersch.

\paragraph{Lemma 2.}\label{3.6-lemma2} Sei $A$ ein Ring und $f_1,\ldots,f_r\in A$ mit $1=(f_1,\ldots,f_r)$. Sei $\mathfrak{a}\subset A$ ein Ideal und $\varphi_i:A\to A_{f_i}$ die Lokalisierungsabbildung. Dann gilt:
\[\mathfrak{a} = \bigcap_{i=1}^r \varphi_i^{-1}(\varphi_i(\mathfrak{a}) A_{f_i}) \]

\paragraph{Beweis.} Für die nichttriviale Inklusion sei $b\in A$ mit $\varphi_i(b)\in \varphi(\mathfrak{a})A_{f_i}$ für alle $i$. Schreibe:
\[\varphi_i(b)=\frac{a_i}{f_i^{n_i}}\in A_{f_i},\quad a_i\in\mathfrak{a},\ n_i>0 \]
Sei o.B.d.A. $n=n_1=\ldots=n_r$. Somit gibt es für alle $i$ ein $m_i\geq 0$ mit:
\[f_i^{m_i}(f_i^nb-a_i)=0 \]
Sei o.B.d.A. $m=m_1=\ldots=m_r$. Es folgt $f_i^{m+n}b\in\mathfrak{a}$ für alle $i$. Aus $1=(f_1,\ldots,f_r)$ folgt $1=(f_1^N,\ldots,f_r^N)$ für alle $N\geq 0$, insbesondere für $N=m+n$. Sei also $1=\sum_{i=1}^rc_if_i^N$ für gewisse $c_i\in A$. Dann gilt:
\[b=\sum_{i=1}^rc_if_i^Nb\in\mathfrak{a}\qedhere \]

\paragraph{Beweis von \hyperref[3.6-lemma1]{Lemma 1}.} Sei $\mathfrak{a}_1\subset\mathfrak{a}_2\subset\ldots$ eine Kette von Idealen in $A$. Diese induziert für alle $i$ eine Kette von Idealen in $A_{f_i}$:
\[\varphi_i(\mathfrak{a}_1)A_{f_i}\subset\varphi_i(\mathfrak{a}_2) A_{f_i}\subset\ldots \]
Da $A_{f_i}$ noethersch ist, wird diese Kette stationär für alle $i$. Es existiert also ein $s$ mit $\varphi_i(\mathfrak{a}_s)A_{f_i}=\varphi(\mathfrak{a}_{s+1})A_{f_i}=\ldots$ für alle $i$. Mit \hyperref[3.6-lemma2]{Lemma 2} wird auch die ursprüngliche Kette von Idealen in $A$ stationär.\qed

\paragraph{Definition.} Sei $f:X\to Y$ ein Morphismus von Schemata.
\begin{enumerate}[(i)]
\item $f$ heißt \textit{lokal von endlichem Typ}\index{von endlichem Typ!lokal}, falls $Y$ eine offene affine Überdeckung $\bigcup_i\operatorname{Spec}(B_i)$ besitzt, so dass $f^{-1}(\operatorname{Spec}B_i)$ für alle $i$ eine offene affine Überdeckung $\bigcup_j\operatorname{Spec}(A_{ij})$ besitzt, wobei alle $A_{ij}$ endlich erzeugte $B_i$-Algebren sind.
\item $f$ heißt \textit{von endlichem Typ}\index{von endlichem Typ}, falls $Y$ eine offene affine Überdeckung $\bigcup_i\operatorname{Spec}(B_i)$ besitzt, so dass $f^{-1}(\operatorname{Spec}B_i)$ für alle $i$ eine endliche, offene affine Überdeckung $\bigcup_j\operatorname{Spec}(A_{ij})$ besitzt, wobei alle $A_{ij}$ endlich erzeugte $B_i$-Algebren sind.
\item $f$ heißt \textit{endlich}\index{endlich}, falls eine offene affine Überdeckung $Y=\bigcup_i\operatorname{Spec}(B_i)$ existiert, so dass $f^{-1}(\operatorname{Spec}B_i)=\operatorname{Spec}(A_i)$ für alle $i$ ist, wobei $A_i$ eine $B_i$-Algebra ist, die als $B_i$-Modul endlich erzeugt ist.
\end{enumerate}

\paragraph{Bemerkung 3.7.}\label{3.7} Sei $f:X\to Y$ ein Morphismus von Schemata.
\begin{enumerate}[(i)]
\item $f$ ist genau dann lokal von endlichem Typ, wenn für alle offene, affine $V=\operatorname{Spec}(B)$ in $Y$ es eine offene affine Überdeckung $f^{-1}(V)=\bigcup_j\operatorname{Spec}(A_j)$ gibt, wobei $A_j$ endlich erzeugte $B$-Algebren sind.
\item $f$ ist genau dann von endlichem Typ, wenn für alle offene, affine $V=\operatorname{Spec}(B)$ in $Y$ es eine endliche, offene affine Überdeckung $f^{-1}(V)=\bigcup_j\operatorname{Spec}(A_j)$ gibt, wobei $A_j$ endlich erzeugte $B$-Algebren sind.
\item $f$ ist genau dann endlich, wenn für alle offene, affine $V=\operatorname{Spec}(B)$ in $Y$ die Menge $f^{-1}(V)=\operatorname{Spec}(A)$ affin ist, wobei $A$ ein endlich erzeugter $B$-Modul ist.
\end{enumerate}

\paragraph{Beweis.} Dies werden wir später zeigen.

\paragraph{Beispiel 3.8.}\label{3.8} Sei $V$ eine Varietät über einem algebraisch abgeschlossenen Körper $k$. Dann ist das Schema $t(V)$ in \hyperref[2.26]{Satz 2.26} ein integres, noethersches Schema von endlichem Typ über $k$.

\paragraph{Beispiel 3.9.}\label{3.9} Sei $P$ ein Punkt einer Varietät und $\mathcal{O}_P$ der zugehörige Halm. Dann ist $\operatorname{Spec}(\mathcal{O}_P)$ ein integres, noethersches Schema, aber nicht von endlichem Typ über $k$.

\paragraph{Definition 3.10.}\label{3.10} Ein \textit{offenes Unterschema}\index{Schema!Unterschema!offen} eines Schemas $X$ ist ein Schema $U$, dessen unterliegender topologischer Raum eine offene Teilmenge von $X$ ist und dessen Strukturgarbe $\mathcal{O}_U$ isomorph zu $\mathcal{O}_X|_U$ ist.

Ein Schemamorphismus $f:X\to Y$ heißt \textit{offene Immersion}\index{Immersion!offen}, falls $f$ ein Isomorphismus auf ein offenes Unterschema in $Y$ induziert.

\paragraph{Bemerkung.} Jede offene Teilmenge eines Schemas $X$ trägt eine eindeutig bestimmte Struktur als offenes Unterschema, siehe auch \hyperref[2.10]{Definition 2.10}.

\paragraph{Definition 3.11.}\label{3.11} Ein \textit{abgeschlossenes Unterschema}\index{Schema!Unterschema!abgeschlossen} eines Schemas $X$ ist ein Schema $Y$, zusammen mit einem Morphismus $(i,i^\sharp):Y\to X$, so dass:
\begin{enumerate}[(i)]
\item Der $Y$ unterliegender Raum ist eine abgeschlossene Teilmenge von $X$.
\item $i:Y\hookrightarrow X$ ist die natürliche Inklusion.
\item $i^\sharp:\mathcal{O}_X\to i_\ast\mathcal{O}_Y$ ist surjektiv.
\end{enumerate}

Ein Schemamorphismus $f:X\to Y$ heißt \textit{abgeschlossene Immersion}\index{Immersion!abgeschlossen}, falls $f$ ein Isomorphismus auf ein abgeschlossenes Unterschema in $Y$ induziert.

\paragraph{Beispiel 3.12.}\label{3.12} Sei $A$ ein Ring, $\mathfrak{a}\subset A$ ein Ideal und $X=\operatorname{Spec}(A),\ Y=\operatorname{Spec}(A/\mathfrak{a})$. Der Ringhomomorphismus $\varphi:A\to A/\mathfrak{a}$ induziert $f:Y\to X,\ \mathfrak{p}\mapsto\varphi^{-1}(\mathfrak{p})$. Dies induziert ein Morphismus von Schemata. $f$ ist ein Homöomorphismus von $Y$ auf $V(\mathfrak{a})$ und die Abbildung $f^\sharp:\mathcal{O}_X\to f_\ast\mathcal{O}_Y$ induziert:
\[f_\mathfrak{p}^\sharp:\mathcal{O}_{X, f(\mathfrak{p})}=A_{\varphi^{-1}(\mathfrak{p})}\to (A/\mathfrak{a})_\mathfrak{p}=\mathcal{O}_{Y,\mathfrak{p}}\]
$f_\mathfrak{p}^\sharp$ ist surjektiv für alle $\mathfrak{p}\in Y$, also ist nach \hyperref[1.10]{1.10 (ii)} auch $f^\sharp$ surjektiv. Somit erhält man für jedes $\mathfrak{a}\subset A$ auf $V(\mathfrak{a})\subset X$ eine Struktur als abgeschlossenes Unterschema in $X$.

\paragraph{Bemerkung.} Ist $Y\subset X=\operatorname{Spec}(A)$ eine abgeschlossene Teilmenge, so existieren auf $Y$ viele abgeschlossene Unterschemastrukturen. Wir werden später sehen, das sie genau den Idealen $\mathfrak{a}\subset A$ mit $Y=V(\mathfrak{a})$ entsprechen.

\paragraph{Satz 3.14.}\label{3.14} Sei $X$ ein Schema und $Y$ eine abgeschlossene Teilmenge von $X$. Dann besitzt $Y$ eine eindeutig bestimmte induzierte Struktur als reduziertes, abgeschlossenes Unterschema.

\iffalse
\paragraph{Beweis.}
\begin{enumerate}
\item Fall: Sei zunächst $X=\operatorname{Spec}(A)$ affin. Betrachte das folgende Ideal in $A$:
\[\mathfrak{a}=\bigcap_{\mathfrak{p}\in Y}\mathfrak{p} \]
Dann ist $V(\mathfrak{a})=Y$ und $\mathfrak{a}$ ist maximal unter allen Idealen $\mathfrak{a}'$ mit $V(\mathfrak{a}')=Y$. Die Unterschemastruktur $Y=V(\mathfrak{a})=\operatorname{Spec}(A/\mathfrak{a})$ auf $Y$ ist reduziert, da $A/\mathfrak{a}$ reduziert ist, wegen $\mathfrak{a}=\operatorname{Rad}(\mathfrak{a})$.

Für die Eindeutigkeit sei $Y'=V(\mathfrak{a}')=\operatorname{Spec}(A/ \mathfrak{a}')$ eine weitere abgeschlossene Unterschemastruktur. Dann ist $\mathfrak{a}'\subset\mathfrak{a}$. Die kanonische Abbildung $A/\mathfrak{a}'\to A/\mathfrak{a}$ induziert ein Morphismus $Y\to Y'$. Da $Y$ reduziert ist, gibt es einen eindeutigen Morphismus $Y\to Y'_\text{red}$, so dass folgendes Diagramm kommutiert:
\[\xymatrix{
Y\ar[rr]\ar@{-->}[rrd] && Y'\\
&& Y'_\text{red}\ar[u]
} \]
\end{enumerate}
\fi

\paragraph{Lemma 3.15.}\label{3.15} Sei $X$ ein topologischer Raum und $X=\bigcup_i U_i$ eine offene Überdeckung. Ferner sei für jedes $i$ eine Garbe $F_i$ auf $U_i$ gegeben und für alle $i,j$ seien Isomorphismen gegeben:
\[\varphi_{ij}:F_i|_{U_i\cap U_j}\stackrel{\sim}{\to} F_j|_{U_i\cap U_j} \]
so dass $\varphi_{ii}=\operatorname{id}$ und $\varphi_{ik}=\varphi_{jk}\varphi_{ij}$ auf $U_i\cap U_j\cap U_k$ gilt. Dann existiert eine eindeutig bestimmte Garbe $F$ auf $X$ und Isomorphismen $\psi_i:F|_{U_i}\stackrel{\sim}{\to}F_i$ mit $\psi_j=\varphi_{ij}\psi_i$ auf $U_i\cap U_j$. Wir sagen auch, dass $F$ durch \textit{Verkleben} der $F_i$ längst $\varphi_{ij}$ entsteht.\index{Verklebung!Garbe}

\paragraph{Beweis.} Folgt direkt aus der Definition einer Garbe.\qed

\paragraph{Definition 3.16.}\label{3.16} \begin{enumerate}[(i)]
\item Die \textit{Dimension} $\dim(X)$ eines Schemas $X$ ist die Dimension von $X$ als topologischer Raum, d.h:\index{Dimension}
\begin{align*}
\dim(X)=\sup\{n\mid&\text{ es gibt eine Kette }Z_0\subsetneq Z_1\subsetneq\ldots\subsetneq Z_n\\
&\text{ von irreduziblen, abgeschlossenen Teilmengen in }X \}
\end{align*}
\item Sei $Z$ eine irreduzible, abgeschlossene Teilmenge eines Schemas $X$. Dann ist die \textit{Kodimension}\index{Kodimension} $\operatorname{codim}(Z,X)$ von $Z$ in $X$ definiert als:
\begin{align*}
\operatorname{codim}(Z,X)=\sup\{n\mid & \text{ es gibt eine Kette }Z=Z_0\subsetneq Z_1\subsetneq\ldots\subsetneq Z_n\\
&\text{ von irreduziblen, abgeschlossenen Teilmengen in }X\}
\end{align*}
Ist $Y$ eine abgeschlossene Teilmenge von $X$, so setzen wir:
\[\operatorname{codim}(Y,X)=\inf \{\operatorname{codim}(Z,X)\mid Z\subset Y\text{ irreduzibel, abgeschlossen} \} \]
\end{enumerate}

\paragraph{Definition 3.17.}\label{3.17} Sei $S$ ein Schema und $X,Y$ $S$-Schemata. Das \textit{Faserprodukt}\index{Faserprodukt} $X\times_SY$ von $X$ und $Y$ über $S$ ist ein Schema, zusammen mit Projektionsmorphismen $p_1:X\times_SY\to X$ und $p_2:X\times_SY\to Y$ derart, dass:
\begin{enumerate}[(i)]
\item Das folgende Diagramm kommutiert:
\[ \begin{tikzcd}
X\times_S Y\ar[r, "p_2"]\ar[d, "p_1"'] & Y\ar[d]\\
X\ar[r] & S
\end{tikzcd}\]
\item Ist $Z$ ein $S$-Schema und $f:Z\to X,\ g:Z\to Y$ Morphismen derart, dass das folgende Diagramm kommutiert:
\[\begin{tikzcd}
Z\ar[rrd, bend left, "g"]\ar[ddr, "f"', bend right]\ar[rd, "\theta", dashed] &&\\
&X\times_S Y\ar[r, "p_2"']\ar[d, "p_1"] & Y\ar[d]\\
&X\ar[r] & S
\end{tikzcd} \]
so existiert einen eindeutig bestimmten Morphismus $\theta:Z\to X\times_SY$ mit $f=p_1\theta$ und $g=p_2\theta$.
\end{enumerate}
Wird für Schemata $X$ und $Y$ kein Bezug zu einer Basis angegeben, so ist immer das Endobjekt $S=\operatorname{Spec}(\mathbb{Z})$ gemeint, d.h. $X\times Y=X\times_{\operatorname{Spec}(\mathbb{Z})}Y$.

\paragraph{Theorem 3.18.}\label{3.18} Seien $X$ und $Y$ $S$-Schemata. Dann existiert das Faserprodukt $X\times_S Y$ und ist auf Isomorphie eindeutig.

\paragraph{Lemma 3.19.}\label{3.19} \textit{(Verkleben von Morphismen, vgl. \hyperref[2.13]{Satz 2.13})}\index{Verklebung!Morphismus} Seien $X,Y$ Schemata und $\{U_i\}$ eine offene Überdeckung von $X$. Ferner seien $f_i:U_i\to Y$ Morphismen gegeben, wobei $U_i$ mit der offenen Unterschemastruktur versehen ist. Es gelte $f_i|_{U_i\cap U_j}=f_j|_{U_i\cap U_j}$ für alle $i,j$. Dann gibt es einen Morphismus $f:X\to Y$ mit $f|_{U_i}=f_i$ für alle $i$.

\paragraph{Beweis von \hyperref[3.18]{Theorem 3.18}.} Die Eindeutigkeit ist klar.
\begin{enumerate}
\item Schritt: Seien $X=\operatorname{Spec}(A),\ Y=\operatorname{Spec}(B),\ S=\operatorname{Spec}(R)$ affin. Somit sind $A$ und $B$ $R$-Algebren. Wir zeigen $X\times_S Y=\operatorname{Spec}(A\otimes_R B)$. Die Projektionsabbildung $p_1:\operatorname{Spec}(A\otimes_RB)\to \operatorname{Spec}(A)$ ist durch die natürliche Abbildung $\tilde{p}_1:A\to A\otimes_RB$ gegeben, analog für $p_2$. Offensichtlich kommutiert:
\[\begin{tikzcd}
 A\ar[r, "\tilde{p}_1"] & A\otimes_RB\\
 R\ar[r]\ar[u] & B\ar[u, "\tilde{p}_2"']
\end{tikzcd} \]
Sei also $Z$ ein $S$-Schema und Morphismen $f:Z\to X,\ g:Z\to Y$ gegeben, die über $S$ gleich sind. Diese entsprechen Ringhomomorphismen $\tilde{f}:A\to\Gamma(Z,\mathcal{O}_Z)$ und $\tilde{g}:B\to\Gamma(Z,\mathcal{O}_Z)$ nach \hyperref[2.13]{Satz 2.13}. Es kommutiert:
\[\begin{tikzcd}
&&\Gamma(Z,\mathcal{O}_Z)\\
A\ar[r, "\tilde{p}_2"']\ar[rru, "\tilde{f}", bend left]& A\otimes_RB\ar[ru, "\tilde{\theta}"', dashed] &\\
R\ar[r]\ar[u] & B\ar[u, "\tilde{p}_2"]\ar[ruu, "\tilde{g}"', bend right]
\end{tikzcd} \]
Wegen der Universaleigenschaft des Tensorprodukts gibt es genau einen Ringhomomorphismus $\tilde{\theta}:A\otimes_RB\to\Gamma(Z,\mathcal{O}_Z)$ mit $\tilde{f}\tilde{\theta}=\tilde{p}_1$ und $\tilde{g}\tilde{\theta}=\tilde{p}_2$. \hyperref[2.13]{Satz 2.13} liefert ein eindeutiges $\theta:Z\to\operatorname{Spec}(A\otimes_RB)$ mit $f=p_1\theta$ und $g=p_2\theta$.
\item Schritt: Seien $X,Y$ beliebige $S$-Schemata und $U\subset_\text{o}X$. Wir nehmen an, dass das Faserprodukt $X\times_SY$ mit Projektionen $p_1,p_2$ existiert. Wir zeigen, dass für die offene Teilmenge $p_1^{-1}(U)\subset X\times_SY$ stets $p_1^{-1}(U)=U\times_SY$ gilt.

Da $p_1^{-1}(U)\subset X\times_SY$, kommutiert das Diagramm:
\[\begin{tikzcd}
p_1^{-1}(U)\ar[r, "p_2"]\ar[d, "p_1"'] & Y\ar[d]\\
U\ar[r] & S
\end{tikzcd} \]
Sei $Z$ ein $S$-Schema und Morphismen $f:Z\to U,\ g:Z\to Y$ gegeben, so dass $(Z\stackrel{f}{\to} U\stackrel{i}{\hookrightarrow} X\to S)=(Z\stackrel{g}{\to} Y\to S)$. Nach der Universaleigenschaft von $X\times_SY$ existiert ein eindeutiges $\theta:Z\to X\times_SY$ mit $if=p_1\theta$ und $g=p_2\theta$. Insbesondere gilt $\theta(Z)\subset p_1^{-1}(U)$, also $\theta:Z\to p_1^{-1}(U)$. Somit erfüllt $p_1^{-1}(U)$ die Universaleigenschaft von $U\times_SY$.
\item Schritt: Seien $X,Y$ $S$-Schemata und $\{X_i\}$ eine offene Überdeckung von $X$. Wir nehmen an, dass alle Faserprodukte $X_i\times_SY$ mit Projektionen $p_{1i},p_{2i}$ existieren. Wir zeigen, dass in diesem Fall auch $X\times_SY$ existiert.

Setze $X_{ij}=X_i\cap X_j$ und $U_{ij}=p_{1i}^{-1}(X_{ij})\subset X_i\times_SY$. Nach Schritt 2 folgt nun $U_{ij}=X_{ij}\times_SY$. Wegen Eindeutigkeit existieren nun Isomorphismen $\varphi_{ij}:U_{ij}\stackrel{\sim}{\to}U_{ji}$ mit $\varphi_{ij}=\varphi_{ji}^{-1}$, $\varphi_{ij}(U_{ij}\cap U_{ik})=U_{ji}\cap U_{jk}$ und $\varphi_{ik}=\varphi_{jk}\circ\varphi_{ij}$ auf $U_{ij}\cap U_{jk}$. Mithilfe \hyperref[2.11]{2.11} verkleben wir die $X_i\times_SY$ via $\varphi_{ij}$ und erhalten so ein Schema $Z$ mit Morphismen $\psi_j:X_j\times_SY\to Z$, die Isomorphismen auf einem offenen Unterschema induzieren. Seien $p_1,p_2$ die Morphismen, die durch Verkleben der $p_{1i}$ bzw. $p_{2i}$ entstehen, siehe \hyperref[3.19]{Lemma 3.19}. Wir zeigen nun, dass $Z$ gerade das Faserprodukt $X\times_SY$ mit Projektionsmorphismen $p_1,p_2$ ist.

Es gilt $Z=\bigcup_j\psi_j(X_j\times_SY)$. Also folgt die Kommutativität des zweiten Diagramms aus dem ersten:
\[\begin{tikzcd}
X_j\times_S Y\ar[r, "p_{1j}"]\ar[d, "p_{2j}"'] & X_j\ar[d]\\
Y\ar[r] & S
\end{tikzcd}\qquad\qquad \begin{tikzcd}
Z\ar[d, "p_2"']\ar[r, "p_1"] & X\ar[d]\\
Y\ar[r] & S
\end{tikzcd}\]
Sei nun $Z'$ ein weiteres $S$-Schema und $f:Z'\to X,\ g:Z'\to Y$ gegeben, die über $S$ gleich sind. Setze $Z_i'=f^{-1}(X_i)$ für alle $i$. Zu jedem $i$ existiert genau ein Morphismus ${\theta}_i:Z_i'\to X_i\times_SY\hookrightarrow Z$ mit $f|_{Z_i'}=p_{1i}\circ{\theta}_i$ und $g|_{Z_i'}=p_{2i}\circ{\theta}_i$. Es kommutiert:
\[\begin{tikzcd}
X_i\times_SY\ar[r, hook]\ar[d, "p_{1i}"'] & Z\ar[d, "p_1"]\\
X_i\ar[r, hook] & X
\end{tikzcd} \]
Es gilt $Z_i'\cap Z_j'=f^{-1}(X_i\cap X_j)=f^{-1}(X_{ij})$ und daher $f|_{Z'_i\cap Z'_j}=p_{1i}\circ\theta_i|_{Z'_i\cap Z'_j}=p_{1j}\circ\theta_j|_{Z_i'\cap Z_j'}$, entsprechend für $g$. Wegen Eindeutigkeit folgt $\theta_i|_{Z_i'\cap Z_j'}=\theta_j|_{Z'_i\cap Z'_j}$. Daher können wir die $\theta_i$ zu einem Morphismus $\theta:Z'\to Z$ verkleben mit $f=p_1\theta$ und $g=p_2\theta$. $\theta$ ist eindeutig, da $\theta|_{Z_i'}=\theta_i$ und alle $\theta_i$ eindeutig sind.
\item Schritt: Seien $X,Y$ $S$-Schemata und $S$ affin. Wir zeigen, dass $X\times_SY$ existiert.

Seien $X=\bigcup_iX_i$ und $Y=\bigcup_jY_j$ offene affine Überdeckungen. Nach Schritt 1 existieren $X_i\times_S Y_j$ für alle $i,j$. Nach Schritt 3 existieren $X\times_SY_j$ für alle $j$. Wegen Symmetrie, existiert somit $ X\times_SY$ nach Schritt 3.
\item Schritt: Seien $X,Y$ $S$-Schemata mit $S$ beliebig. Wir zeigen, dass $X\times_SY$ existiert.

Seien $q:X\to S$ und $r:Y\to S$ die Strukturmorphismen und $S=\bigcup_iS_i$ eine offene affine Überdeckung. Setze $X_i=q^{-1}(S_i)$ und $Y_i=r^{-1}(S_i)$. Nach Schritt 4 existieren $X_i\times_{S_i}Y_i$. Wir zeigen, dass $X_i\times_{S_i}Y_i$ die Universaleigenschaft von $X_i\times_SY$ erfüllt.

Seien $f:Z\to X_i$ und $g:Z\to Y$ gegeben, die über $S$ gleich sind. Dann gilt $rg(Z)=qf(Z)\subset q(X_i)\subset S_i$, also $g(Z)\subset Y_i$. Wir erhalten kommutatives Diagramm:
\[\begin{tikzcd}
Z\ar[ddr, "f"', bend right]\ar[rrd, "g"', bend left]\ar[rrrd, "g", bend left]\ar[rd, "\theta"', dashed]\\
& X_i\times_{S_i}Y_i\ar[r, "p_2"']\ar[d, "p_1"] & Y_i\ar[r, hook]\ar[d] & Y\ar[dd, "r"]\\
& X_i\ar[d, hook]\ar[r] & S_i\ar[dr, hook] &\\
& X\ar[rr]_q & & S
\end{tikzcd} \]
Es existiert genau ein $\theta:Z\to X_i\times_{S_i} Y_i$ mit $f=p_1\theta$ und $g=p_2\theta$. Somit existieren auch $X_i\times_SY$. Nach Schritt 3 existiert auch $X\times_SY$.\qed
\end{enumerate}

\paragraph{Lemma 3.20.}\label{3.20} Seien $X,Y$ $S$-Schemata mit Strukturmorphismen $\xi:X\to S,\ \eta:Y\to S$ und $U,V,W$ offen in $X,Y$ bzw. $S$, so dass $\xi(U)\subset W$ und $\eta(V)\subset W$. Ferner seien $p_1,p_2$ die Projektionsmorphismen von $X\times_SY$. Dann gibt es einen $S$-Schemaisomorphismus:
\[p_1^{-1}(U)\cap p_2^{-1}(V)\cong U\times_WV=U\times_SV \]

\paragraph{Beweis.} $E=p_1^{-1}(U)\cap p_2^{-1}(V)$ ist offenes Unterschema von $X\times_SY$. Die Projektionsmorphismen von $X\times_SY$ induzieren $p_1:E\to U,\ p_2:E\to V$. Sei $Z$ ein $W$-Schema und Morphismen $\varphi:Z\to U,\ \psi:Z\to V$ gegeben, die über $W$ gleich sind. Wir erhalten Morphismen $\varphi':Z\to U\hookrightarrow X,\ \psi':Z\to V\hookrightarrow Y$, die über $S$ gleich sind. Es existiert einen eindeutigen Morphismus $\theta:Z\to X\times_SY$ mit $\varphi'=p_1\theta$ und $\psi'=p_2\theta$. Nun gilt $\theta(Z)\subset E$, da $p_1\theta(Z)=\varphi'(Z)\subset U$ und $p_2\theta(Z)=\psi'(Z)\subset V$. Somit ist $\theta:Z\to E$ mit $\varphi=p_1\theta,\ \psi=p_2\theta$. Es folgt $E\cong U\times_WV$. Da $W$ beliebig war, folgt auch $E\cong U\times_SV$.\qed

\paragraph{Definition.} Ein \textit{Unterschema}\index{Schema!Unterschema} $Y$ eines Schemas $X$ ist ein abgeschlossenes Unterschema eines offenen Unterschemas von $X$. Mit anderen Worten haben wir eine abgeschlossene Immersion $\phi$ und eine offene Immersion $\psi$:
\[Y\stackrel{\phi}{\hookrightarrow} U\stackrel{\psi}{\hookrightarrow}X \]
Ein Morphismus $i:Y\to X$ heißt \textit{Immersion}\index{Immersion}, wenn $i$ einen Isomorphismus von $Y$ auf ein Unterschema in $X$ induziert.

\paragraph{Satz 3.21.}\label{3.21} Seien $X,Y,Z$ $S$-Schemata und $W$ ein $Z$-Schema. Dann gilt:
\begin{enumerate}[(i)]
\item $X\times_SS\cong X$
\item $X\times_SY\cong Y\times_SX$
\item $(X\times_SY)\times_SZ\cong X\times_S(Y\times_SZ)$
\item $(X\times_SZ)\times_ZW\cong X\times_SW$
\item $(X\times_SY)\times_SZ\cong (X\times_SZ)\times_Z(Y\times_SZ)$
\item Ist $\sigma:S\to T$ eine Immersion, so ist $X\times_SY\cong X\times_TY$.
\end{enumerate}

\paragraph{Beweis.} (i) bis (v) folgen direkt aus der Universaleigenschaft des Faserprodukts. Für (vi) seien $\varphi:Z\to X,\ \psi:Z\to Y$ Morphismen über $T$ und $\xi:X\to S,\ \eta:Y\to S$ die Strukturmorphismen. Wir erhalten folgendes kommutative Diagramm:
\[ \begin{tikzcd}
Z\ar[rrd, "\varphi", bend left]\ar[rd, "\theta"', dashed]\ar[ddr, "\psi"', bend right]\\
& X\ar[r]\ar[d]\times_SY & X\ar[d, "\xi"']\ar[ddr, bend left] &\\
& Y\ar[r, "\eta"]\ar[rrd, bend right] & S\ar[dr, "\sigma"]&\\
&&&T
\end{tikzcd}\]
Wegen der Injektivität von $\sigma$, folgt aus $\sigma\xi\varphi=\sigma\eta\psi$ stets $\xi\varphi=\eta\psi$. Es existiert ein eindeutiges, kompatibles $\theta:Z\to X\times_SY$. Somit erfüllt $X\times_SY$ die Universaleigenschaft von $X\times_TY$.\qed

\paragraph{Definition 3.22.}\label{3.22} Seien $f:X_1\to X_2$ und $g:Y_1\to Y_2$ $S$-Morphismen mit kommutativen Diagramm:
\[\begin{tikzcd}
&X_1 \ar[r, "f"]& X_2\\
X_1\times_SY_1\ar[dr, "p_2"']\ar[ur, "p_1"]\ar[rrr, "f\times_Sg", dashed] &&& X_2\times_SY_2\ar[lu, "q_1"']\ar[ld, "q_2"]\\
& Y_1\ar[r, "g"']&Y_2
\end{tikzcd} \]
Es existiert genau ein Morphismus $f\times_Sg: X_1\times_SY_1\to X_2\times_SY_2$ mit $fp_1=q_1(f\times_Sg)$ und $gp_2=q_2(f\times_Sg)$.

\paragraph{Definition 3.23.}\label{3.23} Sei $f:X\to S$ ein Morphismus von Schemata. Für ein Morphismus $g:T\to S$ erhält man ein $T$-Schema $X_T=X\times_ST$ mit:
\[\begin{tikzcd}
X_T\ar[r]\ar[d] & T\ar[d, "g"]\\
X\ar[r, "f"'] & S
\end{tikzcd} \]
Man sagt, dass $X_T$ durch \textit{Basiswechsel} von $X$ über $S$ durch $T$ erhalten bleibt.\index{Basiswechsel}

\paragraph{Lemma 3.24.}\label{3.24} Sei $U\subset S$ ein offenes Unterschema. Dann gilt für den Basiswechsel $X_U$ von $f:X\to S$:
\[X_U=X\times_SU\cong f^{-1}(U) \]

\paragraph{Beweis.} Nach \hyperref[3.21]{Satz 3.21 (i)} ist die Projektion $p_1:X\times_SS\stackrel{\sim}{\to} X$ ein Isomorphismus. Sei in \hyperref[3.20]{Lemma 3.20} $Y=S$ und setze $f=p_2:X\to S$. Wir erhalten: \[f^{-1}(U)=p_1^{-1}(U)\cap p_2^{-1}(U)\cong X\times_SU\qedhere\]

\paragraph{Definition.} Sei $\mathcal{P}$ eine Eigenschaft eines Morphismus' $f:X\to S$. Man sagt, dass $\mathcal{P}$ bei Basiswechsel \textit{erhalten} bleibt, wenn $f_T:X_T\to T$ die Eigenschaft $\mathcal{P}$ besitzt für alle Morphismen $g:T\to S$.

\paragraph{Bemerkung.} Hat $f:X \to S$ eine Eigenschaft $\mathcal{P}$, die stabil unter Basiswechsel ist, so hat insbesondere $f|_{f^{-1}(U)}:f^{-1}(U)\to U$ für alle offenen $U\subset S$ die Eigenschaft $\mathcal{P}$.

\paragraph{Satz 3.25.}\label{3.25}\begin{enumerate}[(i)]
\item Die Eigenschaft (abgeschlossene, offene) Immersion zu sein bleibt bei Basiswechsel erhalten.
\item Die Eigenschaft von endlichem Typ zu sein bleibt stabil unter Basiswechsel.
\end{enumerate}

\paragraph{Beweis.} Wird ausgelassen.

\paragraph{Bemerkung.} Irreduzibilität, Reduziertheit und Integrität bleiben unter Basiswechsel nicht notwendig erhalten.

\paragraph{Lemma 3.26.}\label{3.26} Sei $\psi:A\to B$ ein Ringhomomorphismus, so dass $\alpha(\psi):\operatorname{Spec}(B)\to\operatorname{Spec}(A)$ eine abgeschlossene Immersion ist. Sei ferner $C$ ein beliebiger Ring. Dann ist $\alpha(\psi\otimes\operatorname{id}_C):\operatorname{Spec}(B\otimes_AC)\to\operatorname{Spec}(A\otimes_AC)=\operatorname{Spec}(C)$ eine abgeschlossene Immersion, wobei $\alpha$ die Abbildung aus \hyperref[2.13]{Satz 2.13} bezeichnet.

\paragraph{Beweis.} Da $\tilde{\psi}=\alpha(\psi)$ eine abgeschlossene Immersion ist, ist $\tilde{\psi}^\sharp:\mathcal{O}_A\to\tilde{\psi}_\ast\mathcal{O}_B$ surjektiv. $\tilde{\psi}$ ist Homöomorphismus auf eine abgeschlossene Teilmenge, daher ist $(\tilde{\psi}_\ast\mathcal{O}_B)_{\tilde{\psi}(\mathfrak{p})}=\mathcal{O}_{B,\mathfrak{p}}$ für alle $\mathfrak{p}\in\operatorname{Spec}(B)$. Nach \hyperref[3.12]{Beispiel 3.12} ist die auf den Halmen induzierte Abbildung $\tilde{\psi}^\sharp_\mathfrak{p}$ surjektiv:
\[\tilde{\psi}^\sharp_\mathfrak{p} : A_{\psi^{-1}(\mathfrak{p})}=\mathcal{O}_{A,\tilde{\psi}(\mathfrak{p})}\to (\tilde{\psi}_\ast\mathcal{O}_B)_{\tilde{\psi}(\mathfrak{p})}=B_\mathfrak{p} \]
Betrachte nun die exakte $A$-Modulsequenz $A\stackrel{\psi}{\to}B\to B/\psi(A)\to 0$. Diese induziert unter $-\otimes_AA_\mathfrak{q}$ für alle $\mathfrak{q}\in\operatorname{Spec}(A)$ die exakte Folge:
\[A_\mathfrak{q}\to B\otimes_A A_\mathfrak{q}\to B/\psi(A)\otimes_A A_\mathfrak{q}\to 0 \]
Sei das Bild von $\tilde{\psi}$ von der Form $V(\mathfrak{a})\cong\operatorname{Spec}(B)$ für ein Ideal $\mathfrak{a}\subset A$. Für ein Primideal $\mathfrak{q}\in\operatorname{Spec}(A)$ gilt:
\begin{align*}
B\otimes_A A_\mathfrak{q}\neq 0 &\iff \mathfrak{q}\in\operatorname{Supp}(B)\text{, wobei $B$ als $A$-Modul aufgefasst wird}\\
&\iff \mathfrak{q}\supset\mathfrak{a}\\
&\iff \mathfrak{q}\in\operatorname{im}(\tilde{\psi})\\
&\iff\mathfrak{q}=\psi^{-1}(\mathfrak{p})\text{ für ein }\mathfrak{p}\in\operatorname{Spec}(B)\\
&\implies B/\psi(A)\otimes_AA_\mathfrak{q}=0
\end{align*}
Somit ist $A_\mathfrak{q}\to B\otimes A_\mathfrak{q}$ surjektiv für alle $\mathfrak{q}\in\operatorname{Spec}(A)$. Es folgt die Surjektivität von $\psi$. Wir erhalten unter $-\otimes_AC$ die surjektive Abbildung $C\to B\otimes_AC$. Nach \hyperref[3.12]{Beispiel 3.12} folgt die Behauptung.\qed

\paragraph{Lemma 3.27.}\label{3.27} Sei $f:X\to Y$ ein Morphismus und $Y=\bigcup_\lambda Y_\lambda$ eine offene Überdeckung. Dann ist $f$ genau dann eine (abgeschlossene, offene) Immersion, wenn $f_\lambda=f|_{X_\lambda}:X_\lambda \to Y_\lambda$ mit $X_\lambda=f^{-1}(Y_\lambda)$ die Eigenschaft hat.

\paragraph{Beweis.} Kein Beweis.

\paragraph{Bemerkung 3.28.}\label{3.28} Sind $Y_1,Y_2$ Unterschemata von $X$, so ist $Y_1\times_XY_2\to X$ ein Unterschema von $X$, d.h. eine Immersion.

\paragraph{Definition 3.29.}\label{3.29} Sei $f:X\to Y$ ein $S$-Morphismus von $S$-Schemata. Der Morphismus $\Gamma_f=(\operatorname{id}_X,f)_S:X\to X\times_SY$ heißt \textit{$S$-Graph}\index{Graph} von $f$ und $\Gamma_f(X)$ heißt \textit{Graph} von $f$. Ist $f=\operatorname{id}_X$ die Identität, so heißt $\Delta_{X/S}=\Gamma_{\operatorname{id}_X}$ der \textit{Diagonalmorphismus}\index{Diagonalmorphismus}.

\paragraph{Satz 3.30.}\label{3.30} Sei $f:X\to Y$ ein $S$-Morphismus von $S$-Schemata. \begin{enumerate}[(i)]
\item Sind $Y$ und $S$ affine Schemata, so ist $\Gamma_f$ eine abgeschlossene Immersion. Insbesondere ist für $X,S$ affin $\Delta_{X/S}$ eine abgeschlossene Immersion.
\item $\Gamma_f$ ist allgemein eine Immersion.
\end{enumerate}

\paragraph{Beweis.} \begin{enumerate}[(i)]
\item Sei $X=\bigcup_\alpha X_\alpha$ eine offene affine Überdeckung und $X\times_SY$ das Faserprodukt. Nach \hyperref[3.24]{Lemma 3.24} ist $p_1^{-1}(X_\alpha)\cong X_\alpha\times_SY$ affin. Da $p_1\circ\Gamma_f=\operatorname{id}_X$, gilt $\Gamma_f^{-1}(p_1^{-1}(X_\alpha))=X_\alpha$. Da $\Gamma_f: \bigcup_\alpha\Gamma_f^{-1}(X_\alpha\times_SY) \to \bigcup_\alpha(X_\alpha\times_SY)=X\times_SY$, können wir nach \hyperref[3.27]{Lemma 3.27} o.B.d.A. annehmen, dass $X$ affin ist.

Sei also $S=\operatorname{Spec}(R),\ X=\operatorname{Spec}(B)$ und $Y=\operatorname{Spec}(A)$. Sei $f$ induziert von $\varphi:A\to B$ und $\Gamma_f$ induziert von $\psi:B\otimes_RA\to A,\ b\otimes a\mapsto b\varphi(a)$. $\psi$ ist offensichtlich surjektiv, daher ist $\Gamma_f$ nach \hyperref[2.19]{Satz 2.19 (ii)} eine abgeschlossene Immersion.
\item Sei zunächst $S$ affin und $Y=\bigcup_\alpha Y_\alpha$ eine offene affine Überdeckung. Setze $X_\alpha=f^{-1}(Y_\alpha)$, $\Gamma_{f\alpha}=\Gamma_f|_{X_\alpha}$ und $f_\alpha=f|_{X_\alpha}:X_\alpha\to Y_\alpha$. Es ist $\Gamma_f(X_\alpha)\subset X_\alpha\times_SY_\alpha$ nach (i) ein abgeschlossenes Unterschema und $X_\alpha\times_SY_\alpha\subset X\times_SY_\alpha$ nach \hyperref[3.25]{Satz 3.25 (i)} ein offenes Unterschema ist. Nach (i) ist $\Gamma_{f\alpha}:X_\alpha\to X_\alpha\times_SY_\alpha$ eine Immersion. Nach \hyperref[3.27]{Lemma 3.27} ist auch $\Gamma_f$ eine Immersion.

Sei nun $S$ beliebig und $S=\bigcup_\lambda S_\lambda$ eine offene affine Überdeckung. Nach \hyperref[3.21]{Satz 3.21 (v)} ist $(X\times_SY)\times_SS_\lambda=X_\lambda\times_{S_\lambda}Y_\lambda$, wobei $X_\lambda=X\times_SS_\lambda$ und $Y_\lambda=Y\times_SS_\lambda$. Ferner ist $\Gamma_{f_\lambda}$ der Graph von $f_\lambda:X_\lambda\to Y_\lambda$ eine Immersion. Nach \hyperref[3.27]{Lemma 3.27} ist auch $\Gamma_f$ eine Immersion.\qed
\end{enumerate}

\paragraph{Definition 3.31.}\label{3.31} Sei $f:X\to Y$ ein Morphismus von Schemata und $y\in Y$ ein Punkt mit Restklassenkörper $\kappa(y)=\mathcal{O}_{Y,y}/\mathfrak{m}_{Y,y}$. Sei weiter $i:\operatorname{Spec}\kappa(y)\to Y$ der natürliche Morphismus gegeben durch $(y,\operatorname{id}_{\kappa(y)})$ (siehe \hyperref[2.18]{Satz 2.18}). Dann heißt das Faserprodukt $X_y=X\times_Y\operatorname{Spec}\kappa(y)$ die \textit{Faser}\index{Faser} von $f$ über dem Punkt $y$.
\[\begin{tikzcd}
X_y\ar[r, "p_2"]\ar[d, "p_1"'] & \operatorname{Spec}\kappa(y)\ar[d, "i"]\\
X\ar[r, "f"'] & Y
\end{tikzcd}\]

\paragraph{Satz.} $X_y$ ist ein $\kappa(y)$-Schema mit $f^{-1}(y)$ als unterliegender topologischer Raum.

\paragraph{Beweis.} Es ist $fp_1(X_y)=ip_2(X_y)=y$, also folgt $p_1(X_y)\subset f^{-1}(y)$ und somit $X_y=p_1^{-1}(f^{-1}(y))$. Sei $V\subset_\text{o}X$ affin mit $V\cap f^{-1}(y)\neq\varnothing$. Dann gilt:
\[p_1^{-1}(V)=V_y\cong X\times_Y\operatorname{Spec}\kappa(y)\times_XV=X_y\times_XV \]
Wir können somit o.B.d.A. $X$ und $Y$ als affin annehmen, sei also $X=\operatorname{Spec}(B),\ Y=\operatorname{Spec}(A),\ y=\mathfrak{p}$. Sei $\varphi:A\to B$ assoziiert zu $f$. Setze $\varphi':A_\mathfrak{p}\to B'=B\otimes_AA_\mathfrak{p}=\varphi(A\setminus\mathfrak{p})^{-1}B$. Es gibt eine Folge von Homöomorphismen:
\begin{align*}
f^{-1}(y) &= \{\mathfrak{q}\in\operatorname{Spec}(B)\mid\varphi^{-1}(\mathfrak{q})=\mathfrak{p} \}\\
&\cong \{\mathfrak{q}\in\operatorname{Spec}(B')\mid\varphi'^{-1}(\mathfrak{q})=\mathfrak{p}A_\mathfrak{p} \}\\
&\cong \{ \mathfrak{q}\in\operatorname{Spec}(B')\mid\mathfrak{q}\supset (\mathfrak{p}A_\mathfrak{p})B' \} \\
&=\operatorname{Spec}(B'/(\mathfrak{p}A_\mathfrak{p})B')\\
&=\operatorname{Spec}(B'\otimes_{A_\mathfrak{p}}A_\mathfrak{p}/\mathfrak{p}A_\mathfrak{p})\\
&=\operatorname{Spec}((B\otimes_AA_\mathfrak{p})\otimes_{A_\mathfrak{p}}A_\mathfrak{p}/\mathfrak{p}A_\mathfrak{p}) =\operatorname{Spec}(B\otimes_A\kappa(\mathfrak{p}))\qedhere
\end{align*}

\paragraph{Definition 3.32.}\label{3.32} Sei $f:X\to Y$ ein Morphismus von Schemata, so dass $f(X)\subset Y$ dicht liegt, und $Y$ irreduzibel mit generischer Punkt $\xi$ mit $\xi\in f(X)$. Dann heißt $f^{-1}(\xi)$ \textit{generische Faser}\index{generische Faser} von $f$.

\paragraph{Beispiel 3.33.}\label{3.33} Sei $k$ ein algebraisch abgeschlossener Körper und:
\[X=\operatorname{Spec}k[X,Y,t]/(tY-X^2),\quad Y=\operatorname{Spec}k[T] \]
Sei $f:X\to Y$ gegeben durch die kanonische Abbildung $\varphi:k[t]\to k[X,Y,t]/(tY-X^2)$. $X$ und $Y$ sind integre Schemata von endlichem Typ über $k$. Ferner ist $f$ surjektiv, da $\varphi$ injektiv ist (vgl. \hyperref[2.19]{Satz 2.19 (ii)}). Abgeschlossene Punkte von $Y$ entsprechen $k$. Sei $a\in Y$ ein abgeschlossener Punkt und betrachte die Faser: \[X_a=X\times_Y\operatorname{Spec}\kappa(a)=\operatorname{Spec}k[X,Y]/(aY-X^2)\]
Im Fall $a\neq 0$ ist $X_a$ eine ebene Kurve in $\mathbf{A}_k^2$, die irreduzibel und reduziert ist. Für $a=0$ ist $X_0=\operatorname{Spec}k[X,Y]/(X^2)$ die $Y$-Achse. Diese ist nicht reduziert.

\section{Separierte \& eigentliche Morphismen}

\paragraph{Definition 4.1.}\label{4.1} Sei $f:X\to Y$ ein Morphismus von Schemata. $f$ heißt \textit{separiert}, falls $\Delta_{X/Y}:X\to X\times_YX$ eine abgeschlossene Immersion ist. In diesem Fall sagen wir auch, dass $X$ über $Y$ separiert ist.

Ein Schema $X$ heißt \textit{separiert}\index{separiert}, falls $X\to\operatorname{Spec}(\mathbb{Z})$ separiert ist.

\paragraph{Beispiel 4.2.}\label{4.2} Sei $k$ ein Körper und $X$ die affine Gerade über $k$ mit doppeltem Nullpunkt wie in \hyperref[2.12]{Beispiel 2.12}. Dann ist $X\times_kX$ die affine Ebene mit vierfachen Nullpunkt und $\Delta(X)$ die gewöhnliche Diagonale, die zwei Nullpunkte besitzt. $\Delta(X)$ ist nicht abgeschlossen, da $\overline{\Delta(X)}$ vier Nullpunkte besitzt.

\paragraph{Beispiel 4.3.}\label{4.3} Ist $V$ eine Varietät über eine algebraisch abgeschlossenen Körper $k$, so ist $t(V)$ separiert über $k$. Dies werden wir später zeigen.

\paragraph{Beispiel 4.4.}\label{4.4} Ist $f:X\to Y$ ein Morphismus affiner Schemata, so ist $f$ separiert nach \hyperref[3.30]{Satz 3.30 (i)}.

\paragraph{Beispiel 4.5.}\label{4.5} Ist $f:X\to Y$ eine Immersion, so ist $f$ separiert, da nach \hyperref[3.21]{Satz 3.21 (vi)} $X\cong X\times_XX\cong X\times_YX$ gilt, d.h. $\Delta_{X/Y}:X\to X$ ist ein Isomorphismus.

\paragraph{Satz 4.6.}\label{4.6} Sei $f:X\to Y$ ein Morphismus von Schemata. Dann ist $f$ genau dann separiert, wenn $\Delta(X)\subset X\times_YX$ abgeschlossen ist.

\paragraph{Beweis.} Für die nicht-triviale Richtung sei $\Delta(X)\subset X\times_YX$ abgeschlossen. Wir müssen zeigen, dass $X\to \Delta(X)$ ein Homöomorphismus ist, und dass der Morphismus $\mathcal{O}_{X\times_YX}\to\Delta_\ast\mathcal{O}_X$ surjektiv ist.
\begin{itemize}
\item Sei $p_1:X\times_YX\to X$ die erste Projektion. Da $p_1\circ\Delta=\operatorname{id}_X$, induziert $\Delta$ ein Homöomorphismus auf sein Bild.
\item Sei $P\in X$ und $V\subset_\text{o} Y$ affin. Wähle $U\subset_\text{o}X$ affin mit $P\in U$ und $f(U)\subset V$. Dann ist $U\times_VU$ eine offene, affine Umgebung von $\Delta(P)$. Nach \hyperref[4.4]{Beispiel 4.4} ist $\Delta:U\to U\times_VU$ eine abgeschlossene Immersion. Somit ist $\mathcal{O}_{X\times_YX}|_{U\times_VU}\to \Delta_\ast\mathcal{O}_X|_U$ surjektiv.\qed
\end{itemize}

\paragraph{Definition 4.7.}\label{4.7} Ein Ring $R$ heißt \textit{Bewertungsring}\index{Bewertungsring}, wenn $R$ nullteilerfrei ist, wenn von der folgenden Form ist:
\[R=\{x\in K^\times\mid v(x)\geq 0\} \]
wobei $v:K^\times\to G$ eine Bewertung ist, d.h. $v(xy)=v(x)+v(y)$ und $v(x+y)\geq\min\{v(x),v(y)\}$, und $G$ eine total geordnete abelsche Gruppe ist. $R$ ist dann lokaler Ring mit Maximalideal $\mathfrak{m}=\{x\in K^\times\mid v(x)>0\}\cup\{0\}$ und $K=\operatorname{Quot}(R)$.

\paragraph{Theorem 4.8.}\label{4.8} \textit{(Bewertungstheoretisches Kriterium für Separiertheit)}\index{Bewertungstheoretisches Kriterium} Sei $f:X\to Y$ ein Morphismus von Schemata mit $X$ noethersch. Es sind folgende Aussagen äquivalent:
\begin{enumerate}[(i)]
\item $f$ ist separiert.
\item Sei $R$ ein Bewertungsring des Körpers $K=\operatorname{Quot}(R)$ und $i:U=\operatorname{Spec}(K)\hookrightarrow T=\operatorname{Spec}(R)$ die kanonische Inklusion. Seien ferner Morphismen $T\to Y,\ U\to X$ derart gegeben, dass das folgende Diagramm kommutativ ist:
\[\begin{tikzcd}
U\ar[r]\ar[d, "i"'] & X\ar[d, "f"]\\
T\ar[r]\ar[ur, "h", dashed] & Y
\end{tikzcd}\]
Dann gibt es höchstens eine Abbildung $h:T\to X$, die das obige Diagramm kommutativ macht. Mit anderen Worten: Die folgende kanonische Abbildung ist für jeden Bewertungsring $R$ über $Y$ injektiv:
\[\operatorname{Hom}_Y(\operatorname{Spec}(R),X)\to\operatorname{Hom}_Y(\operatorname{Spec}(K),X),\ h\mapsto h\circ i \]
\end{enumerate}

\paragraph{Definition 4.9.}\label{4.9}  Sei $X$ ein topologischer Raum und $x\in X$. Ein Punkt $x'\in X$ heißt \textit{Spezialisierung}\index{Spezialisierung} von $x$, wenn $x'\in\overline{\{x\}}$ gilt. Ist $x''$ Spezialisierung von $x'$ und $x'$ Spezialisierung von $x$, so ist $x''$ auch Spezialisierung von $x$.

Eine Teilmenge $Z\subset X$ heißt \textit{stabil unter Spezialisierungen}, falls mit $x\in Z$ auch jede Spezialisierung in $Z$ ist. Abgeschlossene Mengen sind stabil unter Spezialisierungen. Die Umkehrung gilt im Allgemeinen nicht.

Ein Morphismus von Schemata $f$ heißt \textit{quasikompakt}\index{quasikompakt}, falls es eine offene affine Ü\-ber\-deckung $Y=\bigcup_\alpha Y_\alpha$ existiert, so dass alle $f^{-1}(Y_\alpha)$ quasikompakt sind. Es gilt:
\[f\text{ quasikompakt}\iff\forall V\subset_\text{o}Y\text{ affin}\colon f^{-1}(V)\text{ quasikompakt} \]

\paragraph{Satz 4.10.}\label{4.10} Sei $f:X\to Y$ ein quasikompakter Morphismus von Schemata. Dann sind folgende Aussagen äquivalent:
\begin{enumerate}[(i)]
\item $f$ ist abgeschlossen.
\item Ist $x\in X$ und $y'$ eine Spezialisierung von $y=f(x)$, so gibt es eine Spezialisierung $x'$ von $x$ mit $f(x')=y'$. Mit anderen Worten: $f(\overline{\{x\}})$ ist stabil unter Spezialisierungen für alle $x\in X$.
\end{enumerate}

\paragraph{Beweis.} (i)$\implies$(ii) ist trivial. Sei also $X'\subset X$ abgeschlossen und setze $Y'=\overline{f(X')}$. Wir zeigen $Y'=f(X')$. Wir versehen $X'$ und $Y'$ mit der eindeutig bestimmten, reduzierte abgeschlossene Unterschemastruktur in $X$ bzw. $Y$. Seien $i:X'\hookrightarrow X$ und $j:Y'\hookrightarrow Y$ die natürlichen Inklusionen. Das Urbildschema $(f\circ i)^{-1}(Y')$ ist reduziert, besitzt $X'$ als unterliegender topologischer Raum und ist als Schema gleich $X'$, da die reduzierte Unterschemastrukturen eindeutig sind. Somit existiert ein eindeutig bestimmter Morphismus $f':X'\to Y'$:
\[\begin{tikzcd}
X'\ar[r, "f'"]\ar[d, "i"'] & Y'\ar[d, "j"]\\
X\ar[r, "f"'] & Y
\end{tikzcd} \]
$f'$ erfüllt ebenfalls die Voraussetzung (ii). Ferner ist $f'$ quasikompakt. $i$ ist als abgeschlossene Immersion quasikompakt, also auch $f\circ i=j\circ f':X'\to Y$. Unter Basiswechsel sehen wir, dass $X'\times_YY'\to Y'$ quasikompakt ist. Da $j$ eine Immersion ist, ist $X'\cong X'\times_{Y'}Y' = X'\times_YY'$, also ist $X'\to Y'$ quasikompakt.

Wir müssen noch zeigen: Ist $f:X\to Y$ ein quasikompakter, dominanter Morphismus reduzierter Schema, welcher (ii) erfüllt, so ist $f$ surjektiv. Sei dazu $y'\in Y$ gegeben und $y$ ein generischer Punkt der irreduziblen Komponente von $Y$, in der $y'$ liegt. Somit ist $y'$ eine Spezialisierung von $y$. Gilt nun $f^{-1}(y)\neq\varnothing$, so gibt es ein $x\in X$ mit $f(x)=y$. Wegen (ii) gibt es eine Spezialisierung $x'$ von $x$ mit $f(x')=y'$ und wir sind fertig. Somit ist noch das nächste Lemma zu zeigen.\qed

\paragraph{Lemma 4.12.}\label{4.12} Für einen quasikompakten Morphismus $f:X\to Y$ ist äquivalent:
\begin{enumerate}[(i)]
\item $f$ ist dominant.
\item Für die generischen Punkte $y$ von $Y$ der irreduziblen Komponenten gilt $f^{-1}(y)\neq\varnothing$.
\end{enumerate}

\iffalse
\paragraph{Beweis.} (ii)$\implies$(i) ist trivial. Für die andere Richtung ersetzen wir $f$ durch $f_\text{red}$ und nehmen $X,Y$ als reduziert an. Sei $y\in Y$ ein generischer Punkt einer irreduziblen Komponente von $Y$ und $U\subset_\text{o}Y$ eine affine Umgebung von $y$. $U$ ist insbesondere quasikompakt, also auch $f^{-1}(U)$, d.h. $f^{-1}(U)=\bigcup_iV_i$ hat eine endliche Überdeckung durch offene, affine $V_i$. Da $f$ nach Voraussetzung dominant ist, folgt $U=\overline{ff^{-1}(U)}=\bigcup_i\overline{f(V_i)}$. Daher gibt es ein $i$ mit $y\in\overline{f(V_i)}$. $\overline{f(V_i)}\subset U$ wird mit der reduzierten, abgeschlossenen Unterschemastruktur versehen. Da alle $V_i$ reduziert sind, faktorisiert $f|_{V_i}:V_i\to U$ eindeutig in der Form $V_i\stackrel{g}{\to} \overline{f(V_i)}\hookrightarrow U$.

Somit genügt es zu zeigen, dass $g^{-1}(y)\neq \varnothing$, da $g^{-1}(y)=f^{-1}(y)\cap V_i$. $y$ ist generischer Punkt einer irreduziblen Komponente von $Y$, daher gilt $\mathcal{O}_{U,y}=\mathcal{O}_{Y,y}$.
\fi

\paragraph{Korollar 4.13.}\label{4.13} Eine quasikompakte Immersion $f:X\to Y$ ist genau dann eine abgeschlossene Immersion, wenn $f(X)$ in $Y$ stabil unter Spezialisierungen ist.

\paragraph{Beweis.} $f$ faktorisiert in eindeutiger Weise in der Form $X\stackrel{\sim}{\to}Z\stackrel{i}{\hookrightarrow} Y$, wobei $i$ die kanonische Inklusion des Unterschema $Z\subset Y$ ist. Es ist $i$ quasikompakt. Ist $f(X)=Z$ stabil unter Spezialisierungen, ist (ii) in \hyperref[4.10]{Satz 4.10} erfüllt, daher ist $i:Z\to Y$ abgeschlossen und $Z$ ein abgeschlossenes Unterschema in $Y$.\qed

\paragraph{Lemma 4.14.}\label{4.14} Sei $R$ ein Bewertungsring eines Körpers $K$ und $T=\operatorname{Spec}(R),\ U=\operatorname{Spec}(K)$. Sei $X$ ein Schema. Dann gilt:
\begin{enumerate}[(i)]
\item $\operatorname{Hom}(U,X)$ ist die Menge aller Paare $(x,i)$ mit $x\in X$ und Körperhomomorphismus $i:\kappa(x)\hookrightarrow K$, wobei $\kappa(x)$ der Restklassenkörper in $x$ bezeichnet.
\item $\operatorname{Hom}(T,X)$ ist die Menge aller Tripel $(x_0,x_1,i)$ mit $x_0,x_1\in X,\ x_0\in\overline{\{x_1\}}$ und Körperhomomorphismus $\kappa(x_1)\hookrightarrow K$, so dass $R$ über $\mathcal{O}_{\overline{\{x_1\}},x_0}$ \textit{dominiert}\index{dominiert}, wobei $\overline{\{x_1\}}$ mit der reduzierten Unterschemastruktur ausgestattet ist.
\end{enumerate}

\paragraph{Definition 4.15.}\label{4.15} Seien $(A,\mathfrak{m}_A),(B,\mathfrak{m}_B)$ lokale Ringe, die in einem Körper $K$ eingebettet sind. Wir sagen $B$ \textit{dominiert}\index{dominiert} $A$, wenn $A\subset B$ und $\mathfrak{m}_B\cap A=\mathfrak{m}_A$ gilt. Mit anderen Worten, wenn die natürliche Inklusion $A\hookrightarrow B$ ein lokaler Homomorphismus ist.

\paragraph{Beweis von \hyperref[4.14]{Lemma 4.14}.} (i) ist gerade \hyperref[2.18]{Satz 2.18}. Setze $t_0=\mathfrak{m}_R\in T$ und $t_1=(0)\in T$. Für (ii) sei $f:T\to X$ ein Morphismus. Setze $x_i=f(t_i)$ und $Z=\overline{\{x_1\}}$ mit der reduzierten Unterschemastruktur. Dann gilt $f^{-1}(Z)=T$ als Mengen. Nach \hyperref[2.21]{Regeln~2.21~(iii)} faktorisiert $f$ in der Form:
\[\begin{tikzcd}
T\ar[rr, "f"]\ar[rrd] &&X\\
&& Z\ar[u]
\end{tikzcd} \]
Beachte, dass für $U\subset_\text{o} Z$ mit $x_0\in U$ auch $x_1\in U$ gilt. Das induziert $\mathcal{O}_{Z,x_0}\to\mathcal{O}_{Z,x_1}$, analog haben wir $\mathcal{O}_{T,t_0}\to\mathcal{O}_{T,t_1}$. Wir erhalten folgendes kommutative Diagramm:
\[\begin{tikzcd}
\mathcal{O}_{Z,x_1} \ar[r]& \mathcal{O}_{T,t_1}=K\\
\mathcal{O}_{Z,x_0} \ar[r]\ar[u]& \mathcal{O}_{T,t_0}=R\ar[u, hook]\\
\mathfrak{m}_{Z,x_0}\ar[r]\ar[u, hook] & \mathfrak{m}_{T,t_0}\ar[u, hook]
\end{tikzcd}\]
Da $Z$ reduziert und irreduzibel ist, ist $Z$ integer. Daher ist $\mathcal{O}_{Z,x_1}=\kappa(x_1)$, alle Abbildungen sind injektiv und wir sehen, dass $R$ über $\mathcal{O}_{Z,x_0}$ dominiert.

Sei umgekehrt $x_0,x_1\in X$ mit $x_0\in\overline{\{x_1\}}= Z$ und $\mathcal{O}_{Z,x_0}\hookrightarrow R$ ein lokaler Homomorphismus. Das induziert $T=\operatorname{Spec}(R)\to\operatorname{Spec}(\mathcal{O}_{Z,x_0})\to Z\to X$. Diese beiden Konstruktionen sind invers zueinander.\qed

\paragraph{Beweis zu \hyperref[4.8]{Theorem 4.8}.} Sei $f:X\to Y$ ein Morphismus von Schemata mit $X$ noethersch.
\begin{itemize}
\item Sei zunächst $f$ separiert und $R$ ein Bewertungsring des Körpers $K=\operatorname{Quot}(R)$ mit Inklusion $i:U=\operatorname{Spec}(K)\hookrightarrow\operatorname{Spec}(R)=T$. Seien $T\to Y,\ U\to X$ gegeben und $h_1,h_2:T\to X$ mit kommutativem Diagramm:
\[\begin{tikzcd}
U\ar[d, "i"']\ar[r] & X\ar[d, "f"]\\
T\ar[r]\ar[ur, "h_1", shift left]\ar[ur, "h_2"', shift right] & Y
\end{tikzcd} \]
Setze $t_0=\mathfrak{m}_R\in T$ und $t_1=(0)\in T$. Aus der Kommutativität $h_1i=h_2i$ folgt insbesondere $h_1(t_1)=h_2(t_1)$. Setze $h''=(h_1,h_2)_Y:T\to X\times_Y X$. Betrachte nun folgendes kommutative Diagramm:
\[\begin{tikzcd}
U\ar[r]\ar[d, "i"]\ar[rdd, "h_2i"', bend right=90]\ar[rrd, "h_1i", bend left=50] & X\ar[d, "\Delta"] &\\
T\ar[r, "h''"'] & X\times_Y X\ar[r, "p_1"]\ar[d, "p_2"'] & X\ar[d, "f"]\\
& X\ar[r, "f"'] & Y
\end{tikzcd}\]
Es gilt $\{h''(t_1)\}= h''i(U)\subset\Delta(X)$. Da $f$ separiert ist, ist $\Delta(X)$ abgeschlossen und es folgt $h''(t_0)\in h''(\overline{\{t_1\}})\subset\overline{\{h''(t_1)\}}\subset\Delta(X)$. Es folgt:
\[h_1(t_0)=p_1h''(t_0)=p_2h''(t_0)=h_2(t_0) \]
Aus der Kommutativität folgt, dass $h_1^\sharp$ und $h_2^\sharp$ dieselbe Abbildung $\kappa(x_1)\hookrightarrow K$ mit $x_1=h_1(t_1)=h_2(t_1)$ induzieren. Mit \hyperref[4.14]{Lemma 4.14 (ii)} folgt $h_1=h_2$.
\item Für die andere Richtung genügt es nach \hyperref[4.6]{Satz 4.6} zu zeigen, dass $\Delta(X)\subset X\times_YX$ abgeschlossen ist. Nach \hyperref[3.30]{Satz 3.30 (ii)} ist $\Delta$ eine Immersion. Da $X$ noethersch ist, ist $\Delta$ quasikompakt. Nach \hyperref[4.13]{Korollar 4.13} genügt es zu zeigen, dass $\Delta(X)$ stabil unter Spezialisierungen ist.

Sei also $\xi_1\in\Delta(X)$ und $\xi_0\in\overline{\{\xi_1\}}$. Sei $K=\kappa(\xi_1)$ und $\mathcal{O}=\mathcal{O}_{\overline{\{\xi_1\}},\xi_0}$, wobei $\overline{\{\xi_1\}}$ mit der reduzierten Unterschemastruktur versehen ist, d.h. $\mathcal{O}$ ist ein lokaler Ring in $K$. Mit \hyperref[4.16]{Satz 4.16} und dem Lemma von Zorn sehen wir, dass für jeden lokalen Ring $\mathcal{O}$ in $K$ einen Bewertungsring $R$ gibt, der $\mathcal{O}$ dominiert. Sei $R$ ein solcher Bewertungsring. Nach \hyperref[4.14]{4.14 (ii)} haben wir einen Morphismus $h:T=\operatorname{Spec}(R)\to X\times_YX$ mit $t_i\mapsto\xi_i$, wobei $t_0=\mathfrak{m}_R\in T,\ t_1=(0)\in T$. Betrachte das kommutative Diagramm:
\[\begin{tikzcd}
\operatorname{Spec}(K)\ar[r, "i"] & T\ar[dr, "h"]\ar[rrd, "p_1h", bend left]\ar[ddr, "p_2h"', bend right]\\
& & X\times_YX\ar[r, "p_1"']\ar[d, "p_2"] & X\ar[d, "f"]\\
&& X\ar[r, "f"'] & Y
\end{tikzcd}\]
Ist $x\in X$, so ist $\kappa(x)\cong\kappa(\Delta(x))$, wegen:
\[\begin{tikzcd}
X\ar[rr, "\Delta"]\ar[rrd, "\operatorname{id}"'] & & X\times_Y X\ar[d, "p_i"]\\
&& X
\end{tikzcd} \]
Betrachte nun das folgende Diagramm:
\[\begin{tikzcd}
\operatorname{Spec}(K)\ar[rd, "hi"']\ar[r, "g", dashed] & X\ar[d, "\Delta"] \\
& X\times_YX
\end{tikzcd}\]
Wähle ein $x_1\in X$ mit $\Delta(x_1)=\xi_1$ und $j:\kappa(x_1)\hookrightarrow K$ als den von $hi$ induzierten $\kappa(\Delta(x_1))\hookrightarrow K$. Nach \hyperref[4.14]{Lemma 4.14 (i)} erhalten wir den zu $(x_1, j)$ passenden Morphismus $g:\operatorname{Spec}(K)\to X$, das das obige Diagramm kommutativ macht. Also faktorisiert $\operatorname{Spec}(K)\to X\times_YX$ über $\operatorname{Spec}(K)\to X\stackrel{\Delta}{\to}X\times_YX$, d.h. $p_1hi=p_2hi$. Nach Voraussetzung folgt $p_1h=p_2h$, d.h. auch $T\to X\times_YX$ faktorisiert über $T\to X\stackrel{\Delta}{\to}X\times_YX$. Somit folgt $\xi_0\in \Delta(X)$.\qed
\end{itemize}

\paragraph{Satz 4.16.}\label{4.16} Sei $K$ ein Körper und $R\subset K$ ein lokaler Ring. Dann ist $R$ genau dann ein Bewertungsring, wenn für jeden lokalen Ring $R\subset S\subset K$ mit lokalen Homomorphismus $R\hookrightarrow S$ stets $R=S$ folgt.

\paragraph{Beweis.} Siehe z.B. Bourbaki, Algebra VI §13.

\paragraph{Satz 4.17.}\label{4.17} Alle involvierten Schemata seine noethersch. Dann gilt:
\begin{enumerate}[(i)]
\item Offene und abgeschlossene Immersionen sind separiert.
\item Die Komposition separierter Morphismen ist separiert.
\item Separiertheit ist stabil unter Basiswechsel.
\item Sind $f:X\to Y,\ f':X'\to Y'$ separierte Morphismen von $S$-Schemata, so ist auch $f\times_Sf':X\times_SX'\to Y\times_SY'$ separiert.
\item Ist $f\circ f$ separiert, so ist auch $f$ separiert.
\item $f:X\to Y$ ist genau dann separiert, wenn es eine offene Überdeckung $Y=\bigcup_\alpha Y_\alpha$ gibt, so dass alle $f^{-1}(Y_\alpha)\to Y_\alpha$ separiert sind.
\end{enumerate}

\paragraph{Beweis.} Folgt alles aus \hyperref[4.8]{Theorem 4.8}.\qed

\paragraph{Definition 4.18.}\label{4.18} Ein Morphismus von Schemata $f:X\to Y$ heißt \textit{eigentlich}\index{eigentlich}, wenn gilt:
\begin{enumerate}[(i)]
\item $f$ ist von endlichem Typ.
\item $f$ ist separiert.
\item $f$ ist \textit{universell abgeschlossen}\index{universell abgeschlossen}, d.h. für jeden Morphismus $Y'\to Y$ ist der Morphismus $f':X'=X\times_YY'\to Y'$ abgeschlossen.
\end{enumerate}
In diesem Fall heißt $X$ auch \textit{eigentlich über} $Y$.

\paragraph{Beispiel 4.19.}\label{4.19} Sei $k$ ein Körper und $X=\mathbf{A}_k^1$ die affine Gerade. Dann ist $X$ separiert und von endlichem Typ. Basiserweiterung mit $X\to k$ ergibt $X\times_kX\to X$, die Projektionsabbildung $\mathbf{A}_k^2\to\mathbf{A}_k^1$. Sei $Y=\operatorname{Spec}(k[X,Y]/(XY-1))$ die hyperbolische Kurve in $\mathbf{A}_k^2$. $Y\subset\mathbf{A}_k^2$. Diese ist abgeschlossen in $\mathbf{A}_k^2$ und projeziert sich in $\mathbf{A}_k^1$ auf $\mathbf{A}_k^1\setminus\{0\}$, die in $\mathbf{A}_k^1$ nicht abgeschlossen ist. Daher ist die Projektionsabbildung nicht abgeschlossen. In diesem Beispiel fehlt der unendliche Punkt von $Y$. Wir werden später zeigen, dass $X$ eigentlich über $k$ ist, wenn $X$ eine sogenannte projektive Varietät.

\paragraph{Theorem 4.20.}\label{4.20} \textit{(Bewertungstheoretisches Kriterium für Eigentlichkeit)}\index{Bewertungstheoretisches Kriterium} Sei $f:X\to Y$ ein Morphismus von endlichem Typ mit $X$ noethersch. Dann sind äquivalent:
\begin{enumerate}[(i)]
\item $f$ ist eigentlich.
\item Sei $R$ ein Bewertungsring des Körpers $K=\operatorname{Quot}(R)$ und $i:U=\operatorname{Spec}(K)\hookrightarrow T=\operatorname{Spec}(R)$ die kanonische Inklusion. Seien ferner Morphismen $T\to Y$ und $U\to X$ derart gegeben, dass das folgende Diagramm kommutativ ist:
\[\begin{tikzcd}
U\ar[r, "u"]\ar[d, "i"'] & X\ar[d, "f"]\\
T\ar[r, "t"']\ar[ru, "h", dashed] & Y
\end{tikzcd}\]
Dann existiert ein eindeutig bestimmter Morphismus $h:T\to X$, der das Diagramm kommutativ macht. Mit anderen worten ist die folgende Abbildung bijektiv:
\[\operatorname{Hom}_Y(T,X)\to\operatorname{Hom}_Y(U,X),\ h\mapsto h\circ i \]
\end{enumerate}

\paragraph{Lemma 4.21.}\label{4.21} \begin{enumerate}[(i)]
\item Abgeschlossene Immersinoen sind von endlichem Typ. Quasikompakte offene Immersionen sind von endlichem Typ.
\item Kompositum zweier Morphismen von endlichem Typ ist von endlichem Typ.
\item Ein Morphismus $f:X\to Y$ ist genau dann von endlichem Typ, wenn $f$ lokal von endlichem Typ und quasikompakt ist.
\item Seien $X\stackrel{f}{\to}Y\stackrel{g}{\to}Z$ Morphismen mit $f$ quasikompakt und $g\circ f$ von endlichem Typ. Dann ist $f$ von endlichem Typ.
\end{enumerate}

\paragraph{Beweis von \hyperref[4.20]{Theorem 4.20}.}\begin{itemize}
\item Sei $f$ eigentlich. Nach Definition ist $f$ separiert, somit ist ein Morphismus $h:T\to X$ wie oben eindeutig bestimmt nach \hyperref[4.8]{Theorem 4.8}. Es bleibt die Existenz zu zeigen. Betrachte die Basiserweiterung $X_T=X\times_YT$ von $X$ mit $T\to Y$:
\[\begin{tikzcd}
U \ar[dr, "\theta"', dashed]\ar[rrd, "u", bend left]\ar[ddr, "i"', bend right] &&\\
& X_T\ar[r, "p_1"']\ar[d, "f'"] & X\ar[d, "f"]\\
& T\ar[r, "t"'] & Y
\end{tikzcd} \]
Wir erhalten ein $\theta:U\to X_T$, so dass das obige Diagramm kommutiert. Sei $\xi_1\in X_T$ das Bild des einzigen Punktes $t_1\in U$ und $Z=\overline{\{\xi_1\}}\subset X_T$ mit der reduzierten Unterschemastruktur. Da $f$ universell abgeschlossen ist, ist $f'$ abgeschlossen, somit ist $f'(Z)\subset T$ abgeschlossen. Da $f'(\xi_1)=t_1$ und $t_1$ der generische Punkt von $T$ ist, gilt $f'(Z)=T$. Es existiert also ein $\xi_0\in Z$ mit $f'(\xi_0)=t_0$, wobei $t_0$ der abgeschlossene Punkt in $T$ ist.

Betrachte den zu $f'$ gehörigen lokalen Homomorphismus $R\hookrightarrow\mathcal{O}_{Z,\xi_0}$. Da $\xi_1$ der generische Punkt von $Z$ ist, gilt $\kappa(\xi_1)=\mathcal{O}_{Z,\xi_1}$ und nach \hyperref[4.14]{Lemma 4.14 (i)} erhalten wir $\kappa(\xi_1)\subset K$, der durch $U\to Z$ induziert wird. Nach \hyperref[4.16]{Satz 4.16} ist $R$ als Bewertungsring maximal unter allen lokalen Ringen in $K$ bzgl. Dominanz. Ferner gilt $\mathcal{O}_{Z,\xi_0}\subset\mathcal{O}_{Z,\xi_1}=\kappa(\xi_1)\subset K$ und $\mathcal{O}_{Z,\xi_0}\subsetneq K$, da $\xi_0\neq\xi_1$. Wegen $f'(\xi_0)=t_0$ dominiert $\mathcal{O}_{Z,\xi_0}$ den Ring $R$, d.h. $R\cong\mathcal{O}_{Z,\xi_0}$, insbesondere dominiert $R$ über $\mathcal{O}_{Z,\xi_0}$. Nach \hyperref[4.14]{Lemma 4.14 (ii)} erhalten wir Morphismus $h':T\cong\operatorname{Spec}(\mathcal{O}_{Z,\xi_0})\to X_T,\ t_i\mapsto \xi_i$. Wir erhalten $h:T\stackrel{h'}{\to}X_T\stackrel{p_1}{\to}X$ mit $fh=fp_1h'=tf'h'=t$ und $hi=p_1h'i=p_1h'f'\theta = p_1\theta = u$.
\item Es gelte (ii). $f$ ist nach Voraussetzung von endlichem Typ und separiert nach \hyperref[4.8]{Theorem 4.8}. Sei also $Y'\to Y$ ein Morphismus und $f':X'=X\times_YY'\to Y'$ die Basiserweiterung von $f$. Wir zeigen, dass $f'$ abgeschlossen ist. Sei $Z\subset X'$ abgeschlossen mit der reduzierten Unterschemastruktur. Betrachte das kommutative Diagramm:
\[ \begin{tikzcd}
Z\ar[r, hook]\ar[dr] & X'\ar[r]\ar[d, "f'"] & X\ar[d, "f"]\\
& Y'\ar[r] & Y
\end{tikzcd} \]
Da $f$ von endlichem Typ ist, ist nach \hyperref[3.25]{Satz 3.25 (ii)} auch $f'$ von endlichem Typ. Es folgt, dass $f'|_Z:Z\to Y'$ von endlichem Typ ist und somit quasikompakt. $f'|_Z$ faktorisiert über $Z\to f'(Z)\hookrightarrow Y'$. Wir sehen, dass $f'(Z)\hookrightarrow Y'$ quasikompakt ist. Nach \hyperref[4.13]{Korollar 4.13} genügt es zu zeigen, dass $f'(Z)$ stabil unter Spezialisierungen ist.

Sei also $z_1\in Z$ und $y_1=f'(z_1),\ y_0\in\overline{\{y_1\}}$. Wir versehen $\overline{\{y_1\}}$ mit der reduzierten Unterschemastruktur. Sei $\mathcal{O}=\mathcal{O}_{\overline{\{y_1\}},y_0 }$. Dann ist $\operatorname{Quot}(\mathcal{O})=\kappa(y_1)\hookrightarrow \kappa(z_1)=:K$.  Sei $R$ ein Bewertungsring von $K$, der $\mathcal{O}$ dominiert. Nach \hyperref[4.14]{Lemma 4.14} haben wir Morphismen:
\[U=\operatorname{Spec}(K)\to Z,\ t_1\mapsto z_1 \]
\[T=\operatorname{Spec}(R)\to Y',\ t_i\mapsto y_i \]
Betrachte folgendes kommutative Diagramm:
\[\begin{tikzcd}
U\ar[r] \ar[d, "i"'] & Z\ar[r, hook] & X'\ar[r]& X\ar[d, "f"]\\
T\ar[rr]\ar[rrru, dashed]\ar[rru, dashed] & & Y'\ar[r]\ar[from=u, "f'", crossing over]\ar[from=ul, crossing over] & Y
\end{tikzcd} \]
Nach Voraussetzung gibt es ein Morphismus $h:T\to X$, der das Diagramm kommutativ macht. Aus der Universaleigenschaft des Faserprodukts $X'$ erhalten wir ein Morphismus $h':T\to X'$, der $h:T\to X$ liftet. Nach Voraussetzung ist $Z$ abgeschlossen und da der generische Punkt $t_1\in T$ auf $z_1\in X'$ abgebildet wird, faktorisiert $h':T\to X'$ über $h':T\to Z\to X'$. Setze $z_0=h'(t_0)\in Z$. Dann ist $f'(z_0)=y_0$, also $y_0\in f'(Z)$.\qed
\end{itemize}

\paragraph{Korollar 4.22.}\label{4.22} Alle vorkommenden Schemata seien noethersch. \begin{enumerate}[(i)]
\item Abgeschlossene Immersionen sind eigentlich.
\item Kompositum zweier eigentlicher Morphismen ist eigentlich.
\item Eigentliche Morphismen sind stabil unter Basiserweiterung.
\item Seien $f:X\to Y$ und $f':X'\to Y'$ eigentliche Morphismen von $S$-Schemata. Dann ist $f\times f':X\times_SX'\to Y\times_SY'$ eigentlich.
\item Seien $f:X\to Y,\ g:Y\to Z$ Morphismen. Ist $g\circ f$ eigentlich und $g$ separabel, so ist $f$ eigentlich.
\item Ein Morphismus $f:X\to Y$ ist genau dann eigentlich, wenn es eine offene Ü\-ber\-deckung $Y=\bigcup_\alpha Y_\alpha$ gibt, so dass $f^{-1}(Y_\alpha)\to Y_\alpha$ für alle $\alpha$ eigentlich ist.
\end{enumerate}

\paragraph{Beweis.} Wir zeigen nur (v). Da $X$ noethersch ist, ist $f$ quasikompakt. Nach \hyperref[4.21]{Lemma 4.21} ist $f$ von endlichem Typ. Nach \hyperref[4.17]{Satz 4.17} ist $f$ separiert. Sei nun $R$ ein Bewertungsring und seien Morphismen $U\to X,\ T\to Y$ gegeben mit kommutativem Diagramm:
\[\begin{tikzcd}
U \ar[r, "u"]\ar[d, "i"']& X\ar[d, "f"]\\
T\ar[r, "t"]\ar[dr, "t'"'] & Y\ar[d, "g"]\\
& Z
\end{tikzcd} \]
Wir wollen ein $h:T\to X$ konstruieren mit $fh=t$ und $hi=u$. Setze $t'=gt$. Da $gf$ eigentlich ist, gibt es genau ein $h:T\to X$ mit $gfh=t'$ und $hi=u$. Sei $t''=fh$. Nun ist $g$ separabel und $ti=fu=fhi=t''i$ und $gt=t'=gfh=gt''$. Nach \hyperref[4.8]{Theorem 4.8} folgt $t=t''$ und somit $fh=t$.\qed

\paragraph{} Ist $\varphi:A\to B$ ein Ringhomomorphismus und $\operatorname{Spec}(B)\to\operatorname{Spec}(A)$ der zugehörige Morphismus, so gilt:
\[\mathbf{P}_B^n\cong\mathbf{P}_A^n\times_{\operatorname{Spec}(A)}\operatorname{Spec}(B) \]
Insbesondere gilt $\mathbf{P}_A^n\cong\mathbf{P}_\mathbb{Z}^n\times_\mathbb{Z}A$.

\paragraph{Definition 4.23.}\label{4.23} \begin{enumerate}[(i)]
\item Sei $Y$ ein Schema. Dann heißt $\mathbf{P}_Y^n=\mathbf{P}_\mathbb{Z}^n\times_\mathbb{Z}Y$ der \textit{$n$-dimensionale projektive Raum über $Y$}\index{projektiver Raum}.
\item Ein Morphismus $f:X\to Y$ von Schemata heißt \textit{projektiv}, wenn er eine Faktorisierung $f:X\stackrel{i}{\hookrightarrow} \mathbf{P}_Y^n\stackrel{p_2}{\to} Y$ besitzt, wobei $i$ eine abgeschlossene Immersion ist.
\item Ein Morphismus $f:X\to Y$ von Schemata heißt \textit{quasiprojektiv}\index{quasiprojektiv}, falls er eine Faktorisierung der Form $f:X\stackrel{i}{\hookrightarrow} X'\stackrel{p}{\to}Y$ besitzt, wobei $i$ eine offene Immersion und $p$ projektiv ist.
\end{enumerate}

\paragraph{Theorem 4.24.}\label{4.24} Ein projektiver Morphismus von noetherschen Schemata ist eigentlich. Ein quasiprojektiver Morphismus von noetherschen Schemata ist von endlichem Typ und separiert.

\paragraph{Beweis.} Es reicht zu zeigen, dass $\mathbf{P}_\mathbb{Z}^n$ eigentlich über $\mathbb{Z}$ ist, da der Basiswechsel nach \hyperref[4.22]{Korollar 4.22 (iii)} eigentlich über $Y$ ist. Ist $f:X\hookrightarrow \mathbf{P}^n_Y \to Y$ projektiv, so ist er als Verkettung von eigentlichen Morphismen wieder eigentlich, siehe \hyperref[4.22]{Korollar 4.22}. Ist $f:X\hookrightarrow X'\to Y$ quasiprojektiv, so ist $f$ als Verkettung separierter Morphismen separiert, siehe \hyperref[4.17]{Korollar 4.17}. Da $X$ noethersch ist, ist $X\hookrightarrow X'$ eine quasikompakte offene Immersion, also nach \hyperref[4.21]{Lemma 4.21 (i)} vom endlichem Typ.

Sei also $X=\mathbf{P}_\mathbb{Z}^n$ und $X=\bigcup V_i$ eine offene affine Überdeckung mit $V_i=D_+(x_i)=\operatorname{Spec}\mathbb{Z}\big[\frac{x_0}{x_i},\ldots,\frac{x_n}{x_i} \big]$. Es ist also $X$ von endlichem Typ. Sei nun $R$ ein Bewertungsring mit Quotientenkörper $K=\operatorname{Quot}(R)$ und $U\to X,\ T\to\operatorname{Spec}(\mathbb{Z})$ Morphismen mit kommutativem Diagramm:
\[\begin{tikzcd}
U\ar[r, "u"]\ar[d, "i"'] & X\ar[d]\\
T\ar[r, "t"'] & \operatorname{Spec}(\mathbb{Z})
\end{tikzcd}\]
Wir zeigen nun per Induktion, dass genau ein Morphismus $h:T\to X$ existiert, der das obige Diagramm kommutativ ergänzt. Für $n=0$ ist $X=\operatorname{Spec}(\mathbb{Z})$ und die Aussage ist klar. Sei $n\geq 1$ und $\xi_1$ das Bild des einzigen Punktes aus $U$ in $X$. Ist $\xi_1\in X\setminus V_i$ für ein $i$, so folgt die Behauptung nach Induktionsvoraussetzung, da die Hyperebene $X\setminus V_i$ isomorph zu $\mathbf{P}_\mathbb{Z}^{n-1}$ ist.

Sei also $\xi_1\in\bigcap V_i$, d.h. alle Funktionen der Form $\frac{x_i}{x_j}$ sind invertierbar in $\mathcal{O}_{\xi_1}$. Der Morphismus $U\to X$ liefert Inklusion $\kappa(\xi_1)\hookrightarrow K$. Sei weiter $f_{ij}\in K^\times$ das Bild von $\frac{x_i}{x_j}$ unter $\mathcal{O}_{\xi_1}\to \kappa(\xi_1)\hookrightarrow K$. Dann folgt $f_{ik}=f_{ij}f_{jk}$ für alle $i,j,k$. Sei $v:K\to G$ die zu $R$ gehörige Bewertung und setze $g_i=v(f_{i0})$ für alle $i$. Sei $k$ derart, dass $g_k$ minimal unter $g_0,\ldots,g_n$ ist. Es gilt für alle $i$:
\[v(f_{ik})=v(f_{i0})-v(f_{k0})=g_i-g_k\geq 0 \]
Es folgt $f_{ik}\in R$ für alle $i$. Es gibt Homomorphismen:
\[\begin{tikzcd}
\mathbb{Z}\big[\frac{x_0}{x_k},\ldots,\frac{x_n}{x_k}\big]\ar[rr, "\varphi", "\frac{x_i}{x_k}\mapsto f_{ik}"'] \ar[d]&& R\ar[d]\\
\mathcal{O}_{\xi_1}\ar[r] & \kappa(\xi_1)\ar[r] & K
\end{tikzcd}\]
Wir erhalten den zu $\varphi$ gehörige Morphismus $T\to V_k$ und somit $h:T\to V_k\hookrightarrow X$. Offensichtlich ist $T\to X\to\operatorname{Spec}(\mathbb{Z})$ der Morphismus $t$ unt $U\to T\to X$ der Morphismus $u$. Ferner ist $h$ eindeutig nach Konstruktion.\qed

\paragraph{Satz 4.25.}\label{4.25} Sei $k$ ein algebraisch abgeschlossener Körper. Das Bild des Funktors $t:\mathbf{Var}(k)\to \mathbf{Sch}(k)$ ist die Menge aller quasiprojektiven, integren Schemata über $k$. Insbesondere ist $t(V)$ integer, separiert und von endlichem Typ für jede Varietät $V$.

\paragraph{Beweis.} Ohne Beweis.

\paragraph{Satz 4.26.}\label{4.26} Sei $X$ ein Schema von endlichem Typ über einem Körper $k$. Dann ist die Menge der abgeschlossenen Punkte in $X$ dicht in $X$.

\iffalse
\paragraph{Beweis.} Es genügt die Aussage für affine Schemata zu zeigen. Sei $X=\operatorname{Spec}(A)$ mit einer endlich erzeugten $k$-Algebra $A$ und $\mathfrak{M}$ die Menge der abgeschlossenen Punkte von $\operatorname{Spec}(A)$. Sei $f\in A$ mit $D(f)\neq\varnothing$. Wir zeigen $\mathfrak{M}\cap D(f)\neq\varnothing$. Es ist $D(f)\cong\operatorname{Spec}(A_f)$. Es existiert ein abgeschlossener Punkt $y\in \operatorname{Spec}(A_f)$, d.h. ein Maximalideal in $A_f$. Nun ist $A_f=A\big[\frac{1}{f}\big]$ endlich erzeugt über $k$, also $\kappa(y)=A_f/y$ eine algebraische Erweiterung von $k$ nach dem Hilbertschen Nullstellensatz. Sei $x$ der zu $y$ gehörige Punkt in $D(f)$.
\fi

\paragraph{Definition 4.27.}\label{4.27} Eine \textit{(abstrakte) Varietät}\index{Varietät!abstrakt} ist ein integres, separiertes Schema $X$ von endlichem Typ über einem algebraisch abgeschlossenen Körper $k$. Ist $X$ über $k$ eigentlich, so heißt $X$ \textit{vollständig}\index{Varietät!eigentlich}. Quasiprojektive abstrakte Varietäten entsprechen den klassischen Varietäten.

\paragraph{Bemerkung 4.28.}\label{4.28}\begin{enumerate}[(i)]
\item Eine projektive, abstrakte Varietät ist vollständig nach \hyperref[4.24]{Theorem 4.24}.
\item Es gibt vollständige Varietäten, die nicht projektiv sind, d.h. die Klasse der abstrakten Varietäten ist größer als die Klasse der klassischen Varietäten.
\item Jede vollständige abstrakte Varietät der Dimension $1$, d.h. eine vollständige Kurve, ist projektiv.
\item Jede Varietät kann als offene Menge in eine vollständige Varietät eingebettet werden.
\end{enumerate}

\section{Modulgarben}

\paragraph{Definition 5.1.}\label{5.1} Sei $(X,\mathcal{O}_X)$ ein geringter Raum.
\begin{enumerate}[(i)]
\item Eine \textit{Garbe von $\mathcal{O}_X$"~Moduln} bzw. \textit{$\mathcal{O}_X$"~Modul}\index{$\mathcal{O}_X$-Modul} ist eine Garbe $\mathcal{F}$ auf $X$ derart, dass $\mathcal{F}(U)$ ein $\mathcal{O}_X(U)$"~Modul für alle $U\subset_\text{o}X$ ist und alle $\operatorname{res}:\mathcal{F}(U)\to\mathcal{F}(V),\ V\subset U$ verträglich mit den Modulstrukturen via $\operatorname{res}:\mathcal{O}_X(U)\to\mathcal{O}_X(V)$ ist.
\item Ein \textit{Morphismus}\index{Morphismus!$\mathcal{O}_X$-Moduln} $\mathcal{F}\to\mathcal{G}$ von $\mathcal{O}_X$-Moduln ist ein Garbenmorphismus, so dass $\mathcal{F}(U)\to\mathcal{G}(U)$ ein $\mathcal{O}_X(U)$"~Modulhomomorphismus für alle $U\subset_\text{o}X$ ist.
\item Eine Sequenz von $\mathcal{O}_X$"~Moduln heißt \textit{exakt}\index{Exaktheit}, wenn sie exakt als Garbensequenz abelscher Garben ist.
\item Sind $\mathcal{F},\mathcal{G}$ $\mathcal{O}_X$"~Moduln, so bezeichnet $\operatorname{Hom}_{\mathcal{O}_X}(\mathcal{F},\mathcal{G})$ die Gruppe der Morphismen von $\mathcal{F}$ nach $\mathcal{G}$.
\item Sind $\mathcal{F},\mathcal{G}$ $\mathcal{O}_X$"~Moduln, so heißt die Garbe $U\mapsto \operatorname{Hom}_{\mathcal{O}_X|_U}(\mathcal{F}|_U,\mathcal{G}|_U)$ die \textit{Hom-Garbe}\index{Hom-Garbe} und wird mit $\operatorname{\mathcal{H}om}(\mathcal{F},\mathcal{G})$ bezeichnet. Diese ist ein $\mathcal{O}_X$"~Modul.
\item Seien $\mathcal{F},\mathcal{G}$ $\mathcal{O}_X$"~Moduln. Dann heißt die zur Prägarbe $U\mapsto\mathcal{F}(U)\otimes_{\mathcal{O}_X(U)}\mathcal{G}(U)$ assoziierte Garbe das \textit{Tensorprodukt}\index{Tensorprodukt} von $\mathcal{F}$ und $\mathcal{G}$. Diese wird  mit $\mathcal{F}\otimes_{\mathcal{O}_X}\mathcal{G}$ bezeichnet.
\item Ein $\mathcal{O}_X$"~Modul $\mathcal{F}$ heißt \textit{frei}\index{$\mathcal{O}_X$-Modul!frei}, wenn $\mathcal{F}$ isomorph zu einer direkten Summe von Exemplaren von $\mathcal{O}_X$ ist.
\item Ein $\mathcal{O}_X$"~Modul $\mathcal{F}$ heißt \textit{lokal frei}\index{$\mathcal{O}_X$-Modul!lokal frei}, wenn $X$ durch offene Mengen $U$ überdeckt werden kann, so dass $\mathcal{F}|_U$ freier $\mathcal{O}_X|_U$"~Modul ist.

Der \textit{Rang} $r$\index{Rang} von $\mathcal{F}$ auf $U$ ist gerade die Anzahl der Kopien von $\mathcal{O}_X|_U$. Wir schreiben $\operatorname{rang}(\mathcal{F}|_U)=r$. Ist $X$ zusammenhängend, so ist dieser Rang überall gleich.
\item Ein lokal freier $\mathcal{O}_X$"~Modul vom Rang $1$ heißt \textit{invertierbare Garbe}\index{Garbe!invertierbar}.
\item Eine \textit{Idealgarbe} auf $X$\index{Idealgarbe} ist ein $\mathcal{O}_X$"~Modul $\mathcal{J}$, der Untergarbe von $\mathcal{O}_X$ ist, d.h. $\mathcal{J}(U)$ ist ein Ideal von $\mathcal{O}_X(U)$ für $U\subset_\text{o}X$.
\item Sei $f:(X,\mathcal{O}_X)\to (Y,\mathcal{O}_Y)$ ein Morphismus von geringten Räumen und $\mathcal{F}$ ein $\mathcal{O}_X$"~Modul. Dann ist $f_\ast\mathcal{F}$ ein $f_\ast\mathcal{O}_X$"~Modul und somit auch ein $\mathcal{O}_Y$"~Modul unter dem Garbenmorphismus $f^\sharp:\mathcal{O}_Y\to f_\ast\mathcal{O}_X$. $f_\ast\mathcal{F}$ heißt \textit{direktes Bild} von $\mathcal{F}$ unter $f$.\index{direktes Bild}

Sei $\mathcal{G}$ ein $\mathcal{O}_Y$"~Modul. Dann ist $f^{-1}\mathcal{G}$ ein $f^{-1}\mathcal{O}_Y$"~Modul. Betrachte das Bild $\theta$ von $f^\sharp$ unter dem Adjunktionsisomorphismus $\operatorname{Hom}_Y(\mathcal{O}_Y,f_\ast\mathcal{O}_X)\to\operatorname{Hom}_X(f^{-1}\mathcal{O}_Y,\mathcal{O}_X)$. Durch $\theta$ wird $\mathcal{O}_X$ zu einem $f^{-1}\mathcal{O}_Y$"~Modul. Wir definieren:
\[f^\ast\mathcal{G}= f^{-1}\mathcal{G}\otimes_{f^{-1}\mathcal{O}_Y}\mathcal{O}_X\]
Somit ist $f^\ast\mathcal{G}$ ein $\mathcal{O}_X$"~Modul, das \textit{Urbild}\index{Urbild} von $\mathcal{G}$ unter $f$.

Für jeden $\mathcal{O}_X$"~Modul $\mathcal{F}$ und $\mathcal{O}_Y$"~Modul $\mathcal{G}$ gilt:
\[\operatorname{Hom}_{\mathcal{O}_X}(f^\ast\mathcal{G},\mathcal{F})\cong\operatorname{Hom}_{\mathcal{O}_Y}(\mathcal{G},f_\ast\mathcal{F}) \]
Somit ist $f^\ast$ linksadjungiert zu $f_\ast$.
\end{enumerate}

\paragraph{Bemerkung.}\begin{enumerate}[(i)]
\item Kern, Bild und Kokern eines Morphismus von $\mathcal{O}_X$"~Moduln ist wieder ein $\mathcal{O}_X$"~Modul.
\item Sind $\mathcal{F},\mathcal{F}'$ $\mathcal{O}_X$"~Moduln und $\mathcal{F}'$ eine Untergarbe von $\mathcal{F}$, so ist $\mathcal{F}/\mathcal{F}'$ ein $\mathcal{O}_X$-Modul.
\item Direkte Summen, direkte Produkte und projektive Limiten von $\mathcal{O}_X$"~Moduln sind wieder $\mathcal{O}_X$"~Moduln.
\end{enumerate}

\paragraph{Definition.} Sei $A$ ein Ring und $M$ ein $A$-Modul. Wir definieren die \textit{zu $M$ assoziierte Garbe}\index{assoziierte Garbe!Modul} $\widetilde{M}$ auf $\operatorname{Spec}(A)$ wie folgt: Für $U\subset_\text{o}\operatorname{Spec}(A)$ setzen wir $\widetilde{M}(U)$ als die Menge aller Abbildungen $s:U\to\coprod_{\mathfrak{p}\in U}M_\mathfrak{p}$, so dass:
\begin{enumerate}[(i)]
\item Für alle $\mathfrak{p}\in U$ gilt $s(\mathfrak{p})\in M_\mathfrak{p}$.
\item Für alle $\mathfrak{p}\in U$ gibt es eine offene Umgebung $V$ von $\mathfrak{p}$ mit $V\subset U$ und Elemente $m\in M,\ f\in A$, so dass für alle $\mathfrak{q}\in V$ stets $f\not\in\mathfrak{q}$ und $s(\mathfrak{q})=\frac{m}{f}\in M_\mathfrak{q}$ gilt.
\end{enumerate}
Dann ist $\widetilde{M}$ mit den gewöhnlichen Restriktionsabbildungen eine Garbe.

\paragraph{Satz 5.2.}\label{5.2} Sei $A$ ein Ring und $M$ ein $A$-Modul mit der assoziierten Garbe $\widetilde{M}$ auf $X=\operatorname{Spec}(A)$. Dann gilt:\begin{enumerate}[(i)]
\item $\widetilde{M}$ ist ein $\mathcal{O}_X$"~Modul.
\item Für jedes $\mathfrak{p}\in X$ gilt für den Halm von $\widetilde{M}$ in $\mathfrak{p}$ stets $\widetilde{M}_\mathfrak{p}\cong M_\mathfrak{p}$.
\item Für alle $f\in A$ gibt es einen $A_f$-Modulisomorphismus $\widetilde{M}(D(f))\cong M_f$.
\item Insbesondere gilt $\Gamma(X,\widetilde{M})\cong M$.
\end{enumerate}

\paragraph{Beweis.} (i) ist klar. (iv) folgt aus (iii) mit $f=1$. (ii) und (iii) gehen analog zu \hyperref[2.3]{2.3}.\qed

\paragraph{Satz 5.3.}\label{5.3} Sei $A$ ein Ring und $X=\operatorname{Spec}(A)$. Ferner sei $\varphi:A\to B$ ein Ringhomomorphismus und $f:\operatorname{Spec}(B)\to \operatorname{Spec}(A)$ der entsprechende Morphismus. Dann gilt:
\begin{enumerate}[(i)]
\item Die Abbildung $M\mapsto\widetilde{M}$ liefert einen exakten, volltreuen Funktor von der Kategorie der $A$"~Moduln in die Kategorie der $\mathcal{O}_X$"~Moduln.
\item Seien $M,N$ $A$"~Moduln. Dann gilt $\widetilde{M\otimes_AN}\cong \widetilde{M}\otimes_{\mathcal{O}_X}\widetilde{N}$.
\item Sei $(M_i)_i$ eine Familie von $A$-Moduln. Dann gilt $\widetilde{\bigoplus_i M_i}\cong\bigoplus_i\widetilde{M_i}$.
\item Sei $N$ ein $B$-Modul. Dann gilt $f_\ast\widetilde{N}\cong \widetilde{_AN}$, wobei $_AN$ den Modul $N$ als $A$"~Modul via $\varphi$ bezeichnet.
\item Sei $M$ ein $A$"~Modul. Dann gilt $f^\ast\widetilde{M}\cong \widetilde{M\otimes_A B}$.
\end{enumerate}

\paragraph{Beweis.} Für (i) zeigen wir: \begin{itemize}
\item \textit{Funktorialität:} Sei $\psi:M\to N$ ein $A$"~Modulhomomorphismus. Dieser induziert für alle $f\in A$ einen $A_f$"~Modulhomomorphismus $\psi_f:M_f\to N_f$. Ist $D(f)\supset D(g)$, so erhalten wir das folgende kommutative Diagramm:
\[\begin{tikzcd}
M_f\ar[rr, "\psi_f"]\ar[d] && N_f\ar[d]\\
M_g\ar[rr, "\psi_g"'] && N_g
\end{tikzcd} \]
Da $D(f),\ f\in A$ eine Basis der Topologie bilden, induzieren $\psi_f,\ f\in A$ einen $\widetilde{A}$"~Homomorphismus $\widetilde{\psi}:\widetilde{M}\to\widetilde{N}$ mit $\widetilde{\psi}|_{D(f)}=\psi_f$.
\item \textit{Volltreu:} Die Umkehrabbildung zu $\operatorname{Hom}_A(M,N)\to\operatorname{Hom}_{\widetilde{A}}(\widetilde{M},\widetilde{N})$ ist gegeben durch das Bilden der globalen Schnitte:
\[\operatorname{Hom}_{\widetilde{A}}(\widetilde{M},\widetilde{N})\to\operatorname{Hom}_A(\Gamma(X,\widetilde{M}),\Gamma(X,\widetilde{N}))=\operatorname{Hom}_A(M,N)\]
\item \textit{Exaktheit:} Sei $0\to M'\to M\to M''\to 0$ eine kurze exakte Folge von $A$"~Moduln. Da Lokalisierungen exakt sind, ist $0\to M'_\mathfrak{p}\to M_\mathfrak{p}\to M''_\mathfrak{p}\to 0$ exakt. Nach \hyperref[5.2]{Satz 5.2 (ii)} ist $\widetilde{M}_\mathfrak{p}=M_\mathfrak{p}$. Da jede Halmsequenz exakt ist, folgt nach \hyperref[1.10]{1.10} die Exaktheit von $0\to\widetilde{M'}\to\widetilde{M}\to\widetilde{M''}\to 0$.
\end{itemize}
(ii) und (iii) folgen, da direkte Summen und Tensorprodukte mit Lokalisierungen kommutieren. Für (iv) sei $g\in A$. Es gilt:
\[\Gamma(D(g),f_\ast\widetilde{N}) = \Gamma(f^{-1}(D(g)),\widetilde{N}) = \Gamma(D(\varphi(g)),\widetilde{N}) \cong N_{\varphi(g)} = N_g\cong \Gamma(D(g),\widetilde{N})\]
Für (v) sei $N$ ein $B$-Modul. Dann gilt:
\begin{align*}
\operatorname{Hom}_{\widetilde{B}}(f^\ast\widetilde{M},\widetilde{N})&=\operatorname{Hom}_{\widetilde{A}}(\widetilde{M},f_\ast \widetilde{N}) \stackrel{\text{(iv)}}{=}\operatorname{Hom}_{\widetilde{A}}(\widetilde{M}, \widetilde{_AN})\\
&= \operatorname{Hom}_A(M, {_AN}) \stackrel{\star}{\cong}\operatorname{Hom}_B(M\otimes_A B,N)\cong\operatorname{Hom}_{\widetilde{B}}(\widetilde{M\otimes_A B},\widetilde{N})
\end{align*}
wobei $\star$ durch die Abbildung $\eta\mapsto (m\otimes b\mapsto \eta(m)b)$ gegeben ist.\qed

\paragraph{Definition 5.4.}\label{5.4} \begin{enumerate}[(i)]
\item Sei $(X,\mathcal{O}_X)$ ein Schema. Ein $\mathcal{O}_X$"~Modul $\mathcal{F}$ heißt \textit{quasikohärent}\index{$\mathcal{O}_X$-Modul!quasikohärent}, falls es eine offene affine Überdeckung $U_i=\operatorname{Spec}(A_i),\ i\in I$ von $X$ gibt, so dass für jedes $i$ ein $A_i$"~Modul $M_i$ existiert mit $\mathcal{F}|_{U_i}\cong\widetilde{M_i}$.
\item $\mathcal{F}$ heißt \textit{kohärent}\index{kohärent}, falls $\mathcal{F}$ quasikohärent ist und alle vorkommenden $M_i$ in (i) endlich erzeugte $A_i$"~Moduln sind.
\end{enumerate}

\paragraph{Beispiel 5.5.}\label{5.5} Für jedes Schema $X$ ist $\mathcal{O}_X$ kohärent, da $\mathcal{O}_X|_{\operatorname{Spec}(A)}=\widetilde{A}$.

\paragraph{Beispiel 5.6.}\label{5.6} Sei $X=\operatorname{Spec}(A)$ affin und $Y\subset X$ ein abgeschlossenes Unterschema, das durch das Ideal $\mathfrak{a}\subset A$ definiert ist. Sei $i:Y\hookrightarrow X$ die natürliche Inklusion. Es ist $\mathcal{O}_Y\cong\widetilde{A/\mathfrak{a}}$ und somit $i_\ast\mathcal{O}_Y=\widetilde{A/\mathfrak{a}}$, wobei hier $A/\mathfrak{a}$ als $A$"~Modul aufgefasst wird. Somit ist $i_\ast\mathcal{O}_Y$ ein kohärenter $\mathcal{O}_X$"~Modul.

\paragraph{Beispiel 5.7.}\label{5.7} Sei $X=\operatorname{Spec}(A)$ affin und $U\subsetneq_\text{o}X$ mit der natürlichen Inklusion $j:U\hookrightarrow X$. Betrachte die Garbe $j_!\mathcal{O}_U$, die außerhalb $U$ durch Null fortgesetzte Garbe von $\mathcal{O}_U$. $j_!\mathcal{O}_U$ ist nicht quasikohärent:

Sei $X$ irreduzibel und $V=\operatorname{Spec}(A)\subset_\text{o}X$ mit $V\subsetneq U$. Wäre $j_!\mathcal{O}_U|_V\cong\widetilde{M}$ für einen $A$"~Modul $M$., so ist $(j_!\mathcal{O}_U|_V)(V)=M$, aber $(j_!\mathcal{O}_U|_V)(V)=0$ und $j_!\mathcal{O}_U|_V\neq 0$.

\paragraph{Lemma 5.8.}\label{5.8} Sei $X=\operatorname{Spec}(A)$ ein affines Schema, $f\in A$ und $\mathcal{F}$ ein quasikohärenter $\mathcal{O}_X$"~Modul.
\begin{enumerate}[(i)]
\item Sei $s\in\Gamma(X,\mathcal{F})$ mit $s|_{D(f)}=0$. Dann existiert ein $n>0$ mit $f^ns=0$.
\item Sei $t\in\Gamma(D(f),\mathcal{F})$. Dann existiert ein $n>0$ und $t'\in\mathcal{F}(X)$ mit $f^nt=t'|_{D(f)}$.
\end{enumerate}

\paragraph{Beweis.} Wir zeigen zunächst, dass es eine Überdeckung der Form $X=\bigcup_{i=1}^m D(g_i)$ gibt, so dass $\mathcal{F}|_{D(g_i)}\cong\widetilde{M_i}$ für einen $A_{g_i}$"~Modul $M_i$.

Da $\mathcal{F}$ quasikohärent ist, existiert eine offene affine Überdeckung aus Mengen der Form $V=\operatorname{Spec}(B)$ mit $\mathcal{F}|_V=\widetilde{M}$ für einen $B$"~Modul $M$. Wir schreiben $V=\bigcup_{\text{gewisse }g\in A}D(g)$. Die natürlichen Morphismen $D(g)\hookrightarrow V$ liefern Ringhomomorphismen $B\to A_g$. Nach \hyperref[5.3]{Satz 5.3 (v)} ist $\mathcal{F}|_{D(g)}\cong \widetilde{M\otimes_B A_g}$. Da $X$ affin und somit quasikompakt ist, kann $X$ durch solche Mengen endlich überdeckt werden.
\begin{enumerate}[(i)]
\item Setze $s_i$ als das Bild von $s|_{D(g_i)}$ unter $\Gamma(D(g_i),\mathcal{F})=\Gamma(D(g_i),\widetilde{M_i})\cong M_i$. Wegen $D(fg_i)=D(f)\cap D(g_i)$ folgt nach \hyperref[5.2]{Satz 5.2 (iii)} $\Gamma(D(fg_i),\mathcal{F})=(M_i)_f$. Also ist $s_i=0$ in $(M_i)_f$. Nach Definition gibt es ein $n_i\in\mathbb{N}$ mit $f^{n_i}s_i=0$ in $M_i$. Sei $n$ das Maximum aller $n_i,\ i=1,\ldots,m$. Dann folgt $f^ns=0$ aus der ersten Garbeneigenschaft.
\item Betrachte die Einschränkungen $t\in\Gamma(D(fg_i),\mathcal{F})=(M_i)_f$. Für alle $i$ gibt es ein $n_i\in\mathbb{N}$, so dass $f^{n_i}t=t_i|_{D(fg_i)}$ für ein $t_i\in \Gamma(D(g_i),\mathcal{F})=M_i$. Sei $n$ das Maximum aller $n_i,\ i=1,\ldots,m$. Dann gibt es für alle $i$ ein $t_i\in\Gamma(D(g_i),\mathcal{F})$ mit $f^n t = t_i|_{D(fg_i)}$.

Auf $D(g_i)\cap D(g_j)=D(g_ig_j)$ haben wir Schnitte $t_i,t_j$ konstruiert, die auf $D(fg_ig_j)$ übereinstimmen. Nach (i) gibt es ein $m_{ij}$, so dass $f^{m_{ij}}(t_i-t_j)=0$ auf $D(g_ig_j)$ gilt. Sei $m$ das Maximum aller $m_{ij}$, so dass $f^m(t_i-t_j)=0$ auf $D(g_ig_j)$ für alle $i,j$ gilt. Die lokalen Schnitte $f^mt_i$ in $\Gamma(D(g_i),\mathcal{F})$ verkleben sich somit zu einem globalen Schnitt $t'$ von $\mathcal{F}$ zusammen mit $t'|_{D(f)}=f^{n+m}t$.\qed
\end{enumerate}

\paragraph{Satz 5.9.}\label{5.9} Sei $X$ ein Schema und $\mathcal{F}$ ein $\mathcal{O}_X$"~Modul. Dann ist $\mathcal{F}$ genau dann quasikohärent, wenn für alle affinen $U=\operatorname{Spec}(A)\subset_\text{o}X$ ein $A$"~Modul $M$ existiert mit $\mathcal{F}|_U\cong\widetilde{M}$.

Ist $X$ noethersch, so ist $\mathcal{F}$ genau dann kohärent, wenn für alle affinen $U=\operatorname{Spec}(A)\subset_\text{o}X$ ein endlich erzeugter $A$"~Modul $M$ existiert mit $\mathcal{F}|_U\cong\widetilde{M}$.

\paragraph{Lemma.} Sei $X=\operatorname{Spec}(A)$ affin, $M$ ein $A$"~Modul und $\mathcal{F}$ ein $\mathcal{O}_X$"~Modul. Dann ist
\[\operatorname{Hom}_A(M,\Gamma(X,\mathcal{F}))\to\operatorname{Hom}_{\widetilde{A}}(\widetilde{M},\mathcal{F}),\ \varphi\mapsto\widetilde{\varphi} \]
ein Isomorphismus mit Umkehrabbildung $\theta\mapsto\Gamma(X,\theta)$.

\paragraph{Beweis von \hyperref[5.9]{Satz 5.9}.} Die Rückrichtungen der beiden Aussagen sind trivial. Sei $U\subset_\text{o}X$ affin. Nach dem Beweis von \hyperref[5.8]{Lemma 5.8} gibt es eine Basis der Topologie von $U$, bestehend aus affinen Teilmengen $V_i$ derart, dass $\mathcal{F}|_{V_i}\cong\widetilde{N_i}$ mit einem Modul $N_i$ ist. Somit ist $\mathcal{F}|_U$ quasikohärent. Wir können somit o.B.d.A. $X=U=\operatorname{Spec}(A)$ als affin annehmen.

Setze $M=\Gamma(X,\mathcal{F})$ und $\alpha:\widetilde{M}\to\mathcal{F}$ als das Bild von $\operatorname{id}_M$ unter der Abbildung im vorherigen Lemma. Wie im Beweis von \hyperref[5.8]{Lemma 5.8} gezeigt, gibt es eine Überdeckung der Form $X=\bigcup_{i=1}^mD(g_i)$ mit $\mathcal{F}|_{D(g_i)}\cong\widetilde{M_i}$ für einen $A_{g_i}$"~Modul $M_i$. Es gilt $M_i=\mathcal{F}(D(g_i))\cong M_{g_i}$ und wir haben ein kommutatives Diagramm:
\[\begin{tikzcd}
\mathcal{F}(X)_{g_i}\ar[rr, "\cong"] && \mathcal{F}(D(g_i))\\
\mathcal{F}(X)\ar[u]\ar[urr, "\operatorname{res}"']
\end{tikzcd} \]
Somit ist $\alpha|_{D(g_i)}$ ein Isomorphismus für alle $i$. Da die $D(g_i)$ ganz $X$ überdecken, ist $\alpha$ ein Isomorphismus.

Sei nun $X$ zusätzlich noethersch. Dann sind die $A_{g_i}$"~Moduln $M_{g_i}$ endlich erzeugt. Wir zeigen, dass $M$ endlich erzeugt ist. Da $A$ noethersch ist, sind alle $A_{g_i}$ noethersch. Also sind die endlich erzeugten $A_{g_i}$"~Moduln $M_{g_i}$ noethersch. Analog zu \hyperref[3.6]{Satz 3.6} folgt $M$ noethersch. Insbesondere ist $M$ endlich erzeugt. \qed

\paragraph{Korollar 5.10.}\label{5.10} Sei $A$ ein Ring und $X=\operatorname{Spec}(A)$. Dann ist
\[\{\text{Kategorie der $A$-Moduln}\}\to\{\text{Kategorie der quasikohärenten $\mathcal{O}_X$-Moduln}\},\ M\mapsto\widetilde{M} \]
ist eine Kategorienäquivalenz mit der Umkehrabbildung $\mathcal{F}\mapsto\Gamma(X,\mathcal{F})$. Ist $A$ noethersch, so geben dieselben Funktoren eine Kategorienäquivalenz zwischen den endlich erzeugten $A$-Moduln und den kohärenten $\mathcal{O}_X$"~Moduln.

\paragraph{Beweis.} Sei $\mathcal{F}$ ein quasikohärenter $\mathcal{O}_X$"~Modul. Nach \hyperref[5.9]{Satz 5.9} existiert ein $A$"~Modul $M$ mit $\mathcal{F}\cong\widetilde{M}$. Nach \hyperref[5.2]{Satz 5.2 (iv)} ist $\Gamma(X,\mathcal{F})=M$, also:
\[M\mapsto\widetilde{M}\mapsto \Gamma(X,\widetilde{M})=M,\quad\mathcal{F}=\widetilde{M}\mapsto \Gamma(X,\widetilde{M})=M\mapsto\widetilde{M}\qedhere \]

\paragraph{Satz 5.11.}\label{5.11} Sei $X=\operatorname{Spec}(A)$ ein affines Schema und sei $0\to\mathcal{F}'\to\mathcal{F}\to\mathcal{F}''\to 0$ eine exakte Sequenz von $\mathcal{O}_X$"~Moduln, wobei $\mathcal{F}'$ quasikohärent ist. Dann ist die Sequenz der globalen Schnitte ebenfalls exakt:
\[0\to\Gamma(X,\mathcal{F}')\to\Gamma(X,\mathcal{F})\to\Gamma(X,\mathcal{F}'')\to 0 \]

\paragraph{Beweis.} Da $\Gamma(X,-)$ linksexakt ist, bleibt nur die Surjektivität von $\Gamma(X,\mathcal{F})\to\Gamma(X,\mathcal{F}'')$ zu zeigen. Sei $s\in\Gamma(X,\mathcal{F}'')$ gegeben. Für $x\in X$ gibt es ein $f\in A$ mit $x\in D(f) \subset X$, so dass $s|_{D(f)}$ sich zu einem $t\in\mathcal{F}(D(f))$ liftet. Wir zeigen nun, dass es ein $r>0$ existiert, so dass sich $f^rs$ zu einem $t''\in\mathcal{F}(X)$ liftet.

Sei $X=\bigcup_i D(g_i)$ eine endliche offene Überdeckung, so dass sich $s|_{D(g_i)}$ zu einem $t_i\in\mathcal{F}(D(g_i))$ liftet. Auf $D(f)\cap D(g_i)=D(fg_i)$ liften $t_i,t\in\mathcal{F}(D(fg_i))$ beide $s$. Aus der Linksexaktheit von $\Gamma(D(fg_i), -)$ folgt $t-t_i\in\mathcal{F}'(D(fg_i))$. Da $\mathcal{F}'$ quasikohärent ist, folgt aus \hyperref[5.8]{Lemma 5.8 (ii)} die Existenz eines $n>0$ und $u_i\in\mathcal{F}'(D(g_i))$ mit $u_i|_{D(fg_i)}=f^n(t-t_i)$. Wir können $n$ unabhängig von $i$ wählen. Setze $t_i'=f^nt_i+u_i\in\mathcal{F}(D(g_i))$. Dann ist $t'_i$ ein Lift von $f^ns|_{D(g_i)}$ und $\star\ t_i'=f^nt$ auf $\mathcal{F}(D(fg_i))$. Auf $D(g_ig_j)$ liften $t_i'$ und $t_j'$ beide $f^ns$, also $t_i'-t_j'\in\mathcal{F}'(D(g_ig_j))$ und daher $t'_i=t'_j$ auf $\mathcal{F}(D(fg_ig_j))$ wegen $\star$. Nach \hyperref[5.8]{Lemma 5.8 (i)} existiert ein $m>0$, so dass $f^m(t_i'-t_j')=0$ auf $\mathcal{F}(D(g_ig_j))$. Wir können $m$ unabhängig von $i,j$ wählen. Somit verkleben sich die $f^mt_i'$ zu einem $t''\in\mathcal{F}(X)$ zusammen und $t''$ ist ein Lift von $f^{n+m}s$.

Sei nun $X=\bigcup_i D(f_i)$ eine endliche offene Überdeckung, so dass sich $s|_{D(f_i)}$  zu einem Schnitt aus $\mathcal{F}(D(f_i))$ liften lässt. Für alle $i$ existiert ein $n$ und ein Lift $t_i\in\Gamma(X,\mathcal{F})$ von $f_i^ns$. Wir können o.B.d.A. $n$ unabhängig von $i$ annehmen. Da $X=\bigcup_i D(f_i)$, gilt $(f_1^n,\ldots,f_k^n)=A$, also $1=\sum_{i=1}^k a_if_i^n$ für gewisse $a_i\in A$. Setze $t=\sum_{i=1}^k a_it_i\in\mathcal{F}(X)$. Dann ist $t$ ein Lift von $\sum_{i=1}^k a_if_i^ns=s\in\mathcal{F}''(X)$.\qed

\paragraph{Satz 5.12.}\label{5.12} Sei $X$ ein Schema.
\begin{enumerate}[(i)]
\item Kern, Kokern und Bild eines Morphismus von quasikohärenten Garben ist wieder quasikohärent.
\item Ist $0\to\mathcal{F}'\to\mathcal{F}\to\mathcal{F}''\to 0$ eine kurze exakte Sequenz von $\mathcal{O}_X$"~Moduln und $\mathcal{F}',\mathcal{F}''$ quasikohärent, so ist $\mathcal{F}$ quasikohärent.
\item Ist $X$ noethersch, so gilt (ii) auch für kohärente Garben.
\end{enumerate}

\paragraph{Beweis.} Da Quasikohärenz bzw. Kohärenz eine lokale Eigenschaft ist, können wir ohne Einschränkung $X=\operatorname{Spec}(A)$ als affin annehmen. Nach \hyperref[5.10]{Korollar 5.10} gelten (i) und (ii) für Modulgarben der Form $\widetilde{M}$. Da $M\mapsto \widetilde{M}$ nach \hyperref[5.3]{5.3 (i)} ein exakter, volltreuer Funktor ist, folgt die Aussage über Kern, Kokern und Bild.

Sei nun $0\to\mathcal{F}'\to\mathcal{F}\to\mathcal{F}''\to 0$ eine kurze exakte Sequenz von $\mathcal{O}_X$"~Moduln mit $\mathcal{F}',\mathcal{F}''$ quasikohärent. Nach \hyperref[5.11]{Satz 5.11} ist die folgende Folge exakt:
\[0\to\Gamma(X,\mathcal{F}')\to\Gamma(X,\mathcal{F})\to\Gamma(X,\mathcal{F}'')\to 0 \]
Da $M\mapsto\widetilde{M}$ ein exakter Funktor ist, folgt die Exaktheit von:
\[\begin{tikzcd}
0  \ar[r]& \widetilde{\Gamma(X,\mathcal{F}')} \ar[d] \ar[r]& \widetilde{\Gamma(X,\mathcal{F})} \ar[d] \ar[r]& \widetilde{\Gamma(X,\mathcal{F}'')} \ar[d] \ar[r]& 0\\
0 \ar[r] & \mathcal{F}'  \ar[r]& \mathcal{F}  \ar[r]&  \ar[r]\mathcal{F}''  \ar[r]& 0
\end{tikzcd} \]
Die vertikalen Pfeile links und rechts sind nach \hyperref[5.10]{Korollar 5.10} Isomorphismen. Nach dem 5er Lemma ist somit auch der mittlere Pfeil ein Isomorphismus und $\mathcal{F}$ ist quasikohärent.

Sei nun $X$ noethersch und $\mathcal{F}',\mathcal{F}''$ kohärent. Dann sind $\Gamma(X,\mathcal{F}'),\Gamma(X,\mathcal{F}'')$ endlich erzeugt und somit auch $\Gamma(X,\mathcal{F})$. Es folgt die Kohärenz von $\mathcal{F}$.\qed

\paragraph{Satz 5.13.}\label{5.13} Sei $f:X\to Y$ ein Morphismus von Schemata.
\begin{enumerate}[(i)]
\item Sei $\mathcal{G}$ ein quasikohärenter $\mathcal{O}_Y$"~Modul. Dann ist $f^\ast\mathcal{G}$ quasikohärenter $\mathcal{O}_X$"~Modul.
\item Seien $X,Y$ noethersch und $\mathcal{G}$ kohärenter $\mathcal{O}_Y$"~Modul. Dann ist $f^\ast\mathcal{G}$ kohärent.
\item Sei $X$ noethersch oder $f$ quasikompakt und separiert. Ist $\mathcal{F}$ ein quasikohärenter $\mathcal{O}_X$"~Modul, so ist $f_\ast\mathcal{F}$ quasikohärenter $\mathcal{O}_Y$"~Modul.
\end{enumerate}

\paragraph{Lemma 5.14.}\label{5.14} Sei $Y$ ein affines Schema, $f:X\to Y$ ein separierter Morphismus und $U,V\subset_\text{o}X$ affin. Dann ist $U\cap V$ ein abgeschlossenes Unterschema eines affinen Schemas. Wir werden in \hyperref[5.18]{Korollar 5.18} sehen, dass $U\cap V$ sogar affin ist. Insbesondere ist $U\cap V$ quasikompakt.

\paragraph{Beweis.} Betrachte das kartesische Diagramm:
\[\begin{tikzcd}
X\times_Y X\ar[r, "p_2"]\ar[d, "p_1"'] & X\ar[d, "f"]\\
X\ar[r, "f"'] & Y
\end{tikzcd} \]
Nach \hyperref[3.20]{Lemma~3.20} ist $p_1^{-1}(U)\cap p_2^{-1}(V)=U\times_YV$ affin. Es gilt: \[U\cap V\cong \Delta_{X/Y}(X)\cap U\times_YV\subset U\times_YV\] Nun ist $\Delta_{X/Y}(X)$ abgeschlossen in $X\times_YX$, d.h. $U\cap V$ ist ein abgeschlossenes Un\-ter\-sche\-ma in $U\times_YV$.\qed

\paragraph{Beweis von \hyperref[5.13]{Satz 5.13}.} Für (i) und (ii) ist die Aussage lokal in $X$ und in $Y$. Daher können wir o.B.d.A. $X=\operatorname{Spec}(B)$ und $Y=\operatorname{Spec}(A)$ als affin annehmen. Dann folgt die Behauptung aus der Kategorienäquivalenz \hyperref[5.10]{Korollar 5.10} und \hyperref[5.3]{Satz 5.3 (v)} $f^\ast\widetilde{M} = \widetilde{M\otimes_AB}$.

Für (iii) können wir nur $Y$ als affin annehmen. Sei $X=\bigcup_i U_i$ eine endliche, offene affine Überdeckung, da in beiden Fällen $X$ quasikompakt ist. Setze $U_{ij}=U_i\cap U_j$. In beiden Fällen sind $U_{ij}$ quasikompakt, siehe \hyperref[5.14]{Lemma 5.14} und \hyperref[3.5]{3.5}, \hyperref[3.4]{3.4}. Sei also $U_{ij}=\bigcup_k U_{ijk}$ eine endliche, offene affine Überdeckung. Nach der Garbeneigenschaft haben wir die folgende exakte Sequenz:
\[0\to f_\ast\mathcal{F} \to\bigoplus_if_\ast(\mathcal{F}|_{U_i}) \to\bigoplus_{i,j,k}f_\ast(\mathcal{F}|_{U_{ijk}}) \]
Wegen \hyperref[5.3]{Satz 5.3 (iv)} sind $f_\ast(\mathcal{F}|_{U_i})$ und $f_\ast(\mathcal{F}|_{U_{ijk}})$ quasikohärent. Nach \hyperref[5.12]{Satz 5.12 (ii)} sind $\bigoplus_i f_\ast(\mathcal{F}|_{U_i})$ und $\bigoplus_{i,j,k}f_\ast(\mathcal{F}|_{U_{ijk}})$ quasikohärent. Nach \hyperref[5.12]{5.12 (i)} ist daher auch $f_\ast\mathcal{F}$ quasikohärent.\qed

\paragraph{Bemerkung.} Sind $X,Y$ noethersch und $\mathcal{F}$  kohärent, so muss $f_\ast\mathcal{F}$ nicht notwendigerweise kohärent sein.

\paragraph{Definition 5.16.}\label{5.16} Sei $Y$ ein abgeschlossenes Unterschema von $X$ und $i:Y\hookrightarrow X$ der Inklusionsmorphismus. Dann ist die \textit{zu $Y$ gehörige Idealgarbe} auf $X$.\index{Idealgarbe} wie folgt definiert:
\[\mathcal{J}_Y=\ker(i^\sharp:\mathcal{O}_X\to i_\ast\mathcal{O}_Y) \]

\paragraph{Satz 5.17.}\label{5.17} Sei $X$ ein Schema. Dann gilt:
\begin{enumerate}[(i)]
\item Ist $Y$ ein abgeschlossenes Unterschema von $X$, so ist $\mathcal{J}_Y$ eine quasikohärente Idealgarbe auf $X$.
\item Ist $X$ zusätzlich noethersch, so ist $\mathcal{J}_Y$ kohärent.
\item Jede quasikohärente Idealgarbe auf $X$ bestimmt in eindeutiger Weise ein abgeschlossenes Unterschema.
\end{enumerate}

\paragraph{Beweis.}\begin{enumerate}[(i)]
\item Sei $Y\subset X$ abgeschlossen. Dann ist $i:Y\hookrightarrow X$ ein quasikompakter Morphismus. Wegen \hyperref[4.17]{Satz 4.17 (i)} ist $i$ separiert. Nach \hyperref[5.13]{Satz 5.13 (iii)} ist $i_\ast\mathcal{O}_Y$ quasikohärent, also $\mathcal{J}_Y$ quasikohärent nach \hyperref[5.12]{Satz 5.12 (i)}.
\item Ist $X$ noethersch und $U\subset_\text{o}X$ affin mit $U=\operatorname{Spec}(A)$, so ist auch $A$ noethersch nach \hyperref[3.6]{Satz 3.6}. Daher ist $I=\Gamma(U,\mathcal{J}_Y|_U)$ ein endlich erzeugtes Ideal in $A$. Nach \hyperref[5.9]{Satz 5.9} ist $\mathcal{J}_Y$ kohärent.
\item Sei $\mathcal{J}$ eine quasikohärente Idealgarbe auf $X$. Setze:
\[Y=\operatorname{Supp}(\mathcal{O}_X/\mathcal{J})=\{x\in X\mid (\mathcal{O}_X/\mathcal{J})_x\neq 0 \}\subset X \]
Wir zeigen, dass $(Y,\mathcal{O}_X/\mathcal{J})$ ein abgeschlossenes Unterschema von $X$ ist. Dies ist eine lokale Frage, sei o.B.d.A. $X=\operatorname{Spec}(A)$ affin. Da $\mathcal{J}$ quasikohärent ist, folgt $\mathcal{J}=\widetilde{\mathfrak{a}}$ für ein Ideal $\mathfrak{a}\subset A$. Es gilt:
\begin{align*}
Y&=\operatorname{Supp}(\mathcal{O}_X/\mathcal{J})=\operatorname{Supp}(\widetilde{A/\mathfrak{a}})\\
&= \{\mathfrak{p}\in X\mid (A/\mathfrak{a})_\mathfrak{p}\neq 0\}\\
&=\{\mathfrak{p}\in X\mid\mathfrak{a}\subset\mathfrak{p}\}=\operatorname{Spec}(A/\mathfrak{a})
\end{align*}
Die Eindeutigkeit ist klar.\qed
\end{enumerate}

\paragraph{Korollar 5.18.}\label{5.18} Sei $X=\operatorname{Spec}(A)$ affin. Dann haben wir eine Bijektion:
\[\{\mathfrak{a}\mid \mathfrak{a}\subset A \text{ Ideal}\}\to \{Y\mid Y\subset X \text{ abgeschlossenes Unterschema}\},\ \mathfrak{a}\mapsto \operatorname{Spec}(A/\mathfrak{a}) \]
Insbesondere ist jedes abgeschlossenes Unterschema eines affinen Schemas wieder affin.

\paragraph{Definition 5.19.}\label{5.19} Sei $S=\bigoplus_d S_d$ ein graduierter Ring. Ein $S$"~Modul heißt \textit{graduierter $S$"~Modul}\index{Modul!graduiert}, falls $M=\bigoplus_d M_d$ mit $S_d\cdot M_e\subset S_{d+e}$ gilt. Sei $\ell\in\mathbb{Z}$. Dann definieren wir den \textit{getwisteten $S$"~Modul}\index{Modul!getwistet} $M(\ell)$ von $M$ durch:
\[M(\ell)_d = M_{d+\ell} \]

\paragraph{Definition 5.20.}\label{5.20} Sei $S$ ein graduierter Ring und $M$ ein graduierter $S$"~Modul. Die zu $M$ assoziierte Garbe $\widetilde{M}$ auf $\operatorname{Proj}(S)$ ist wie folgt definiert: Sei $U\subset_\text{o}\operatorname{Proj}(S)$ und setze $\widetilde{M}(U)$ als die Menge aller Abbildungen $s:U\to\coprod_{\mathfrak{p}\in U}M_{(\mathfrak{p})}$, so dass:
\begin{enumerate}[(i)]
\item Für alle $\mathfrak{p}\in U$ gilt $s(\mathfrak{p})\in M_{(\mathfrak{p})}$.
\item Für alle $\mathfrak{p}\in U$ existiert eine offene Umgebung $V$ von $\mathfrak{p}$ mit $V\subset U$ und homogene Elemente $m\in M,\ f\in S$ mit $\deg(m)=\deg(f)$ derart, dass für alle $\mathfrak{q}\in V$ stets $f\not\in\mathfrak{q}$ und $s(\mathfrak{q})=\frac{m}{f}$ in $M_{(\mathfrak{q})}$ gilt.
\end{enumerate}
$\widetilde{M}$ wird zu einer Garbe mit den gewöhnlichen Restriktionsabbildungen.

\paragraph{Satz 5.21.}\label{5.21} Sei $S$ ein graduierter Ring, $M$ ein graduierter $S$"~Modul und $X=\operatorname{Proj}(S)$. Dann gilt:
\begin{enumerate}[(i)]
\item $(\widetilde{M})_\mathfrak{p}=M_{(\mathfrak{p})}$ für alle $\mathfrak{p}\in X$.
\item Für alle homogene Elemente $f\in S_+$ ist $\widetilde{M}|_{D_+(f)}\cong\widetilde{M_{(f)}}$ bzgl. des Isomorphismus' $D_+(f)\cong\operatorname{Spec}S_{(f)}$.
\item $\widetilde{M}$ ist ein quasikohärenter $\mathcal{O}_X$"~Modul. Ist $X$ noethersch und $M$ endlich erzeugt, so ist $\widetilde{M}$ kohärent.
\end{enumerate}

\paragraph{Beweis.} (i) und (ii) sind analog zu \hyperref[2.23]{Satz 2.23}. (iii) folgt aus (ii).\qed

\paragraph{Definition 5.22.}\label{5.22} Sei $S$ ein graduierter Ring, $X=\operatorname{Proj}(S)$ und $\mathcal{F}$ ein $\mathcal{O}_X$"~Modul. Für $n\in\mathbb{Z}$ definieren wir die \textit{getwistete Garbe}\index{Garbe!getwistet} $\mathcal{F}(n)$ von $\mathcal{F}$ wie folgt:
\[\mathcal{F}(n)=\mathcal{F}\otimes_{\mathcal{O}_X}\mathcal{O}_X(n),\quad \mathcal{O}_X(n)=\widetilde{S(n)} \]

\paragraph{Satz 5.23.}\label{5.23} Sei $S$ ein graduierter Ring und $X=\operatorname{Proj}(S)$, wobei $S$ als $S_0$"~Algebra von $S_1$ erzeugt wird. Dann gilt:
\begin{enumerate}[(i)]
\item $\mathcal{O}_X(n)$ ist eine invertierbare Garbe auf $X$.
\item Sind $M,N$ graduierte $S$"~Moduln, so ist $\widetilde{M\otimes_SN}\cong\widetilde{M}\otimes_{\mathcal{O}_X}\widetilde{N}$. Insbesondere gilt $\widetilde{M(n)}\cong\widetilde{M}(n)$ und $\mathcal{O}_X(n+m)\cong\mathcal{O}_X(n)\otimes\mathcal{O}_X(m)$.
\item Sei $T$ ein weiterer graduierter Ring, der von $T_1$ als $T_0$"~Algebra erzeugt wird und $\varphi:S\to T$ ein Homomorphismus graduierter Ringe. Sei $U\subset_\text{o} Y=\operatorname{Proj}(T)$ und $f:U\to X$ der durch $\varphi$ induzierte Morphismus. Dann gilt für jeder graduierte $S$"~Modul $M$ und jeder graduierte $T$"~Modul $N$:
\[f^\ast(\widetilde{M})\cong\widetilde{M\otimes_ST}|_U,\quad f_\ast(\widetilde{N}|_U)\cong (\widetilde{_SN}) \]
Insbesondere gilt $f^\ast(\mathcal{O}_X(n))\cong\mathcal{O}_Y(n)|_U$ und $f_\ast(\mathcal{O}_X(n)|_U)=(f_\ast\mathcal{O}_U)(n)$.
\end{enumerate}

\paragraph{Beweis.}\begin{enumerate}[(i)]
\item Sei $f\in S_1$ und betrachte $\mathcal{O}_X(n)|_{D_+(f)}\cong\widetilde{S(n)_{(f)}}$ auf $\operatorname{Spec}S_{(f)}$. Es ist $S(n)_{(f)}$ freier $S_{(f)}$"~Modul vom Rang $1$ via dem Isomorphismus $(S_f)_0\to (S_f)_n,\ s\mapsto f^ns$ für alle $n$. Da $S$ von $S_1$ als $S_0$"~Algebra erzeugt wird, gilt $X=\bigcup_{f\in S_1}D_+(f)$. Daher ist $\mathcal{O}_X(n)$ invertierbar.
\item Sei $f\in S_1$. Es gilt $(M\otimes_SN)_{(f)}=M_{(f)}\otimes_{S_{(f)}}N_{(f)}$. Da $S$ von $S_1$ erzeugt wird, folgt $\widetilde{M\otimes_SN}\cong\widetilde{M}\otimes_{\mathcal{O}_X}\widetilde{N}$.
\item Analog wie im affinen Fall.\qed
\end{enumerate}

\paragraph{Definition 5.24.}\label{5.24} Sei $S$ ein graduierter Ring, $X=\operatorname{Proj}(S)$ und $\mathcal{F}$ ein $\mathcal{O}_X$"~Modul. Der zu $\mathcal{F}$ \textit{assoziierte graduierte $S$"~Modul}\index{Modul!assoziiert} $\Gamma_\ast(\mathcal{F})$ ist definiert als die Gruppe
\[\Gamma_\ast(\mathcal{F})=\bigoplus_{n\in\mathbb{Z}}\Gamma(X,\mathcal{F}(n)) \]
mit der folgenden $S$"~Wirkung: Ein $s\in S_d$ induziert ein $s\in \Gamma(X,\mathcal{O}_X(d))$. Für $t\in\Gamma(X,\mathcal{F}(n))$ setze $st= s\otimes t\in\Gamma(X,\mathcal{F}(n)\otimes_{\mathcal{O}_X}\mathcal{O}_X(d))=\Gamma(X,\mathcal{F}(n+d))$.

\paragraph{Satz 5.25.}\label{5.25} Sei $A$ ein Ring und $X=\mathbf{P}_A^r$ mit $r\geq 1$ und $S=A[X_0,\ldots,X_r]$. Dann gilt:
\[\Gamma_\ast(\mathcal{O}_X)= S \]

\paragraph{Beweis.} Sei $X=\bigcup_{i=0}^r D_+(X_i)$. Ein $t\in\Gamma(X,\mathcal{O}_X(n))$ entspricht eine Familie $t_i\in\Gamma(D_+(X_i),\mathcal{O}_X(n)),\ i=1,\ldots,r$ mit $t_i=t_j$ auf $D_+(X_iX_j)$. $t_i$ ist ein homogenes Element $s_i\in S_{X_i}$ vom Grad $n$ und $t_i|_{D_+(X_iX_j)}$ entspricht dem Bild von $s_i$ in $S_{X_iX_j}$. Es folgt:
\[\Gamma_\ast(\mathcal{O}_X)=\Big\{ (t_0,\ldots,t_r)\in\prod_{i=0}^r S_{X_i}\mid t_i=t_j\text{ auf } S_{X_iX_j}\text{ für alle }i,j \Big\} \]
Da keine $X_i$ Nullteiler sind, haben wir Inklusionen $S\hookrightarrow S_{X_i}\hookrightarrow S_{X_iX_j}\hookrightarrow S_{X_0\cdots X_r}$ und:
\[\Gamma_\ast(\mathcal{O}_X)=\bigcap_{i=0}^r S_{X_i}\subset S_{X_0\cdots X_r} \]
Jedes homogene $t\in S_{X_0\cdots X_r}$ lässt sich eindeutig in der folgenden Form schreiben:
\[t=X_0^{i_0}\cdots X_r^{i_r}f,\quad i_j\in\mathbb{Z}\]
wobei $f\in S$ ein homogenes Element ist, das durch kein $X_i$ teilbar ist. Es ist $t$ genau dann in $S_{X_i}$, wenn $i_j\geq 0$ für alle $j\neq i$ gilt. Also ist $\bigcap S_{X_i}=S$.\qed

\paragraph{Lemma 5.26.}\label{5.26} Sei $X$ ein Schema, $\mathcal{L}$ eine invertierbare Garbe auf $X$ und $f\in\Gamma(X,\mathcal{L})$. Setze $X_f=\{x\in X\mid f_x\not\in\mathfrak{m}_x\mathcal{L}_x\}\subset_\text{o}X$ und sei $\mathcal{F}$ eine quasikohärente Garbe auf $X$.
\begin{enumerate}[(i)]
\item Sei $X$ quasikompakt und $s\in\Gamma(X,\mathcal{F})$ mit $s|_{X_f}=0$. Dann gibt es ein $n>0$, so dass $f^ns=0\in\Gamma(X,\mathcal{F}\otimes\mathcal{L}^{\otimes n})$.
\item Sei $X=\bigcup U_i$ eine endliche, offene affine Überdeckung, so dass $\mathcal{L}|_{U_i}$ für alle $i$ frei und $U_i\cap U_j$ für alle $i,j$ quasikompakt sind. Zu $t\in\Gamma(X_f,\mathcal{F})$ gibt es ein $n>0$, so dass sich $f^nt\in\Gamma(X_f,\mathcal{F}\otimes\mathcal{L}^{\otimes n})$ zu einem globalen Schnitt auf ganz $X$ fortsetzen lässt.
\end{enumerate}

\paragraph{Bemerkung 5.27.}\label{5.27} Voraussetzungen in \hyperref[5.26]{Lemma 5.26 (i) und (ii)} sind erfüllt, wenn $X$ noethersch ist, oder wenn $X$ quasikompakt und separiert ist.

\paragraph{Beweis von \hyperref[5.26]{Lemma 5.26}.} \begin{enumerate}[(i)]
\item Sei $X=\bigcup U_i$ eine endliche, offene affine Überdeckung mit $\mathcal{L}|_{U_i}$ frei. Betrachte $U=U_i$ und sei $\psi:\mathcal{L}|_U\stackrel{\sim}{\to}\mathcal{O}_U$ ein Isomorphismus. Da $\mathcal{F}$ quasikohärent ist, folgt $\mathcal{F}|_U\cong\widetilde{M}$ für ein $A$"~Modul $M$, wobei $U=\operatorname{Spec}(A)$. Für ein $s\in\Gamma(X,\mathcal{F})$ ist $s|_U\in M$. Setze $g:=\psi(f|_U)\in A$. Es ist $X_f\cap U=D(g)$ und $s|_{X_f}=0$. Nach \hyperref[5.8]{Lemma 5.8 (i)} gibt es ein $n>0$ mit $g^ns=0\in M$. Der Isomorphismus
\[\operatorname{id}\otimes \psi^{\otimes n}:\mathcal{F}\otimes \mathcal{L}^{\otimes n}|_U\to\mathcal{F}|_U \]
liefert $0=f^ns=\Gamma(U, \mathcal{F}\otimes\mathcal{L}^{\otimes n})$ für alle $U=U_i$. Wählt man $n$ so groß, dass die obige Aussage für alle $U_i$ gilt, so folgt $f^ns=0$ auf $X$.
\item Analog zu (i) mit \hyperref[5.8]{Lemma 5.8 (ii)}.\qed
\end{enumerate}

\paragraph{Satz 5.28.}\label{5.28} Sei $S$ ein graduierter Ring, der durch $S_1$ als $S_0$"~Algebra endlich erzeugt wird. Sei $X=\operatorname{Proj}(S)$ und $\mathcal{F}$ ein quasikohärenter $\mathcal{O}_X$"~Modul. Dann gibt es einen natürlichen Isomorphismus:
\[\beta:\widetilde{\Gamma_\ast(\mathcal{F})}\stackrel{\sim}{\to}\mathcal{F} \]

\paragraph{Beweis.} Es ist $X=\bigcup_{f\in S_1}D_+(f)$ eine endliche Vereinigung. Für $f\in S_1$ definiere:
\[\beta_f:\widetilde{\Gamma_\ast(\mathcal{F})_{(f)}}=\widetilde{\Gamma_\ast(\mathcal{F})}|_{D_+(f)}\to\mathcal{F}|_{D_+(f)} \]
durch $\bigoplus_d\Gamma(D_+(f),\mathcal{F}(d))\to\Gamma(D_+(f),\mathcal{F})$, das induziert wird durch:
\[\Gamma(D_+(f),\mathcal{F}(d))\to\Gamma(D_+(f),\mathcal{F}(d)\otimes \mathcal{O}_X(-d))= \Gamma(D_+(f),\mathcal{F}),\ \frac{m}{f^d}\mapsto m\otimes f^{-d} \]
Wir erhalten eine Abbildung $\beta:\widetilde{\Gamma_\ast(\mathcal{F})}\to\mathcal{F}$. Wir zeigen nun, dass alle $\beta_f$ Isomorphismen sind. Es genügt zu zeigen, dass $\Gamma_\ast(\mathcal{F})_{(f)}\to\Gamma(D_+(f),\mathcal{F})$ Isomorphismen sind. \hyperref[5.26]{Lemma 5.26 (i)} liefert die Injektivität und \hyperref[5.26]{Lemma 5.26 (ii)} die Surjektivität.\qed

\paragraph{Korollar 5.29.}\label{5.29} Sei $A$ ein Ring. Dann gilt:
\begin{enumerate}[(i)]
\item Ist $Y\hookrightarrow\mathbf{P}_A^r$ ein abgeschlossenes Unterschema, so existiert ein homogenes Ideal $I\subset S=A[X_0,\ldots,X_r]$, so dass $Y=\operatorname{Proj}(S/I)\hookrightarrow\operatorname{Proj}(S)=X$.
\item Sei $Y$ ein Schema über $\operatorname{Spec}(A)$. Dann ist $Y$ genau dann projektiv, wenn $Y\cong\operatorname{Proj}(S)$ für einen graduierten Ring $S$, der von $S_1$ als $S_0=A$"~Algebra endlich erzeugt wird.
\end{enumerate}

\paragraph{Beweis.} \begin{enumerate}[(i)]
\item Sei $\mathcal{J}_Y\subset\mathcal{O}_X$ die Idealgarbe von $Y$ auf $X=\mathbf{P}_A^r$. Da $\mathcal{J}_Y(d)\subset\mathcal{O}_X(d)$, folgt $\Gamma_\ast(\mathcal{J}_Y)\subset\Gamma_\ast(\mathcal{O}_X)$. Nach \hyperref[5.25]{Satz 5.25} ist $\Gamma_\ast(\mathcal{O}_X)=S$, d.h. $I=\Gamma_\ast(\mathcal{J}_Y)$ ist ein homogenes Ideal in $S$. Setze $Y'=\operatorname{Proj}(S/I)$. $Y'$ ist ein abgeschlossenes Unterschema von $X$ mit Idealgarbe $\mathcal{J}_{Y'}=\widetilde{I}$. Da $\mathcal{J}_Y$ nach \hyperref[5.17]{Satz 5.17 (i)} quasikohärent ist, folgt $\mathcal{J}_Y\cong\widetilde{\Gamma_\ast(\mathcal{J}_Y)}$ nach \hyperref[5.28]{Satz 5.28}. Nun gilt:
\[\mathcal{J}_Y\cong\widetilde{\Gamma_\ast(\mathcal{J}_Y)}=\widetilde{I}=\mathcal{J}_{Y'} \]
Nach \hyperref[5.17]{Satz 5.17} folgt $Y=Y'$, also ist $Y$ das abgeschlossene Unterschema, das durch $I$ definiert ist.
\item Es gilt:
\begin{align*}
Y\text{ projektiv} &\iff Y\text{ ist abgeschlossenes Unterschema von }\mathbf{P}_A^r\text{ für ein }r\\
&\iff Y\cong\operatorname{Proj}(S'/I)\text{ für ein homogenes Ideal }I\subset S' = A[X_0,\ldots,X_r]
\end{align*}
Wir zeigen zunächst, dass $I$ und $I'=\bigoplus_{d\geq d_0}I_d$ dasselbe abgeschlossene Unterschema bestimmen. Dafür zeigen wir $\mathfrak{p}\supset I$, wenn $\mathfrak{p}\supset I'$ für alle $\mathfrak{p}\in\operatorname{Proj}(S')$. Wegen $\mathfrak{p}\not\supset S_+$ gibt es ein $x_i\not\in\mathfrak{p}$. Sei nun $x\in I_r,\ r<d_0$, dann ist $x_i^{d_0-r}x\in I_{d_0}\in\mathfrak{p}$. Da $\mathfrak{p}$ ein Primideal ist, folgt $x\in\mathfrak{p}$.

Somit können wir o.B.d.A. $I\subset S'_+$ annehmen. Also ist $A=(S'/I)_0$ und $S=S'/I$ wird als $A$"~Algebra von $S_1$ endlich erzeugt.

Umgekehrt ist jeder graduierte Ring $S$, der von $S_1$ als $S_0=A$"~Algebra endlich erzeugt ist, Quotient des Polynomrings und $\operatorname{Proj}(S)$ ist projektiv.\qed
\end{enumerate}

\paragraph{Definition 5.30.}\label{5.30} Sei $Y$ ein Schema. Der kanonische Morphismus $g:\mathbf{P}_Y^r=\mathbf{P}_\mathbb{Z}^r\times Y\to\mathbf{P}_\mathbb{Z}^r$ definiert die \textit{getwistete Garbe} $\mathcal{O}(1)$ auf $\mathbf{P}_Y^r$ durch:
\[\mathcal{O}(1)=g^\ast\mathcal{O}(1) \]

\paragraph{Bemerkung.} Ist $Y=\operatorname{Spec}(A)$ affin, so ist $\mathcal{O}(1)$ die bereits in \hyperref[5.22]{Definition 5.22} definierte Garbe auf $\mathbf{P}_A^r$.

\paragraph{Definition 5.31.}\label{5.31} Sei $X$ ein Schema über $Y$. Eine invertierbare Garbe $\mathcal{L}$ auf $X$ heißt \textit{sehr ampel}\index{$\mathcal{O}_X$-Modul!sehr ampel} bzgl. $Y$, wenn es eine Immersion $i:X\hookrightarrow\mathbf{P}_Y^r$ für ein $r$ gibt, so dass $i^\ast(\mathcal{O}(1))\cong\mathcal{L}$.

\paragraph{Satz 5.32.}\label{5.32} Sei $Y$ ein noethersches Schema und $X$ ein Schema über $Y$. Dann ist $X$ genau dann projektiv über $Y$, wenn:
\begin{enumerate}[(i)]
\item $X$ ist eigentlich über $Y$.
\item Es gibt eine sehr ample Garbe auf $X$ bzgl. $Y$.
\end{enumerate}

\paragraph{Beweis.} Sei $X$ projektiv. Dann folgt (i) aus \hyperref[4.24]{Theorem~4.24} und es gibt eine abgeschlossene Immersion $i:X\to\mathbf{P}_Y^r$ für ein $r$, so dass $i^\ast\mathcal{O}(1)$ sehr ampel ist.

Sei umgekehrt $i:X\hookrightarrow\mathbf{P}_Y^r$ eine Immersion und $\mathcal{L}\cong i^\ast(\mathcal{O}(1))$ eine sehr ample Garbe auf $X$ bzgl. $Y$. Betrachte das kartesische Quadrat:
\[\begin{tikzcd}
\mathbf{P}^r_Y = Y\times \mathbf{P}_\mathbb{Z}^r \ar[r]\ar[d]& Y\ar[d]\\
\mathbf{P}_\mathbb{Z}^r\ar[r] & \mathbb{Z}
\end{tikzcd} \]
$\mathbf{P}_\mathbb{Z}^r\to\mathbb{Z}$ ist separiert, da projektiv, also ist auch der Basiswechsel $\mathbf{P}_Y^r\to Y$ separiert. Ferner ist $X\to Y$ eigentlich, nach \hyperref[4.22]{Korollar 4.22 (v)} ist auch $X\hookrightarrow \mathbf{P}_Y^r$ eigentlich, also insbesondere abgeschlossen.\qed

\paragraph{Definition 5.33.}\label{5.33} Sei $X$ ein Schema und $\mathcal{F}$ ein $\mathcal{O}_X$"~Modul. $\mathcal{F}$ heißt \textit{von globalen Schnitten erzeugt}\index{$\mathcal{O}_X$-Modul!von globalen Schnitten erzeugt}, wenn es eine Familie von globalen Schnitten $s_i\in\Gamma(X,\mathcal{F}),\ i\in I$ gibt, so dass für alle $x\in X$ die Bilder der $s_i$ den Halm $\mathcal{F}_x$ als $\mathcal{O}_X$"~Modul erzeugen. Dies ist äquivalent zu: Es gibt einen surjektiven Garbenmorphismus $\bigoplus_{i\in I}\mathcal{O}_X\to \mathcal{F}$.

\paragraph{Beispiel 5.34.}\label{5.34} Sei $X=\operatorname{Spec}(A)$ und $\mathcal{F}=\widetilde{M}$ für ein $A$"~Modul $M$. Dann wird $\mathcal{F}$ von globalen Schnitten erzeugt; jedes Erzeugendensystem von $M$ als $A$"~Modul liefern solche Schnitte. Die Surjektion $A^{(I)}\to M$ induziert surjektives $\mathcal{O}_X^{(I)}\to\mathcal{F}$.

\paragraph{Beispiel 5.35.}\label{5.35} Sei $X=\operatorname{Proj}(S)$ mit einem graduierten Ring $S$, der von $S_1$ als $S_0$"~Algebra erzeugt wird. Dann liefern die Elemente aus $S_1$ globale Schnitte von $\mathcal{O}_X(1)$ und erzeugen diesen quasikohärenten Modul.

\paragraph{Lemma 5.36.}\label{5.36} \begin{enumerate}[(i)]
\item Abgeschlossene Immersionen sind endliche Morphismen.
\item Sei $f:X\to Y$ ein endlicher Morphismues noetherscher Schemata und $\mathcal{F}$ ein kohärenter $\mathcal{O}_X$"~Modul. Dann ist $f_\ast\mathcal{F}$ kohärent.
\end{enumerate}

\paragraph{Beweis.}\begin{enumerate}[(i)]
\item Sei $f:X\to Y$ eine abgeschlossene Immersion. Sei $V=\operatorname{Spec}(B)\subset_\text{o}Y$. Es ist $f^{-1}(V)=X\times_YV\to V$ als Basiswechsel von $f$ eine abgeschlossene Immersion. Somit ist $f^{-1}(V)\cong\operatorname{Spec}(B/I)$ für ein Ideal $I\subset B$ affin und ferner $B/I$ ein endlich erzeugter $B$"~Modul.
\item Nach \hyperref[5.13]{Satz 5.13 (iii)} ist $f_\ast\mathcal{F}$ quasikohärent. Sei $V=\operatorname{Spec}(B)\subset_\text{o}Y$. Da $f$ endlich ist, ist $f^{-1}(V)=\operatorname{Spec}(A)$ affin und es gilt $\mathcal{F}|_{f^{-1}(V)}=\widetilde{N}$ für ein endlich erzeugter $A$"~Modul $N$. Nach \hyperref[5.3]{Satz 5.3 (iv)} gilt:
\[f_\ast\mathcal{F}|_V = f_\ast\widetilde{N} = \widetilde{_BN} \]
Da $f$ endlich ist, ist $A$ ein endlich erzeugter $B$"~Modul und somit auch $_BN$.\qed
\end{enumerate}

\paragraph{Satz 5.37.}\label{5.37} \textit{(Serre)} Sei $X$ ein projektives Schema über $\operatorname{Spec}(A)$ mit $A$ noethersch. Sei $\mathcal{O}_X(1)$ eine sehr ample Garbe auf $X$ und $\mathcal{F}$ ein kohärenter $\mathcal{O}_X$"~Modul. Dann existiert ein $n_0\in\mathbb{Z}$, so dass für alle $n\geq n_0$ der getwistete $\mathcal{O}_X$"~Modul $\mathcal{F}(n)$ von endlich vielen globalen Schnitten erzeugt wird.

\paragraph{Beweis.} Sei $i:X\hookrightarrow \mathbf{P}_A^r$ eine abgeschlossene Immersion mit $\mathcal{O}_X(1)\cong i^\ast(\mathcal{O}(1))$. Nach \hyperref[5.36]{Lemma 5.36} ist $i_\ast\mathcal{F}$ kohärent auf $\mathbf{P}_A^r$ und nach \hyperref[5.23]{Satz 5.23 (iii)} gilt $(i_\ast\mathcal{F})(n)=i_\ast(\mathcal{F}(n))$. Wird nun $i_\ast\mathcal{F}(n)$ von endlich vielen globalen Schnitten erzeugt, so ist dies auch für $\mathcal{F}(n)$ der Fall, betrachte dafür:
\[\Gamma(\mathbf{P}_A^r, i_\ast\mathcal{F}(n)) = \mathcal{F}(n)(i^{-1}(\mathbf{P}_A^r))=\Gamma(X,\mathcal{F}(n)) \]
\[(i_\ast\mathcal{F}(n))_x = \begin{cases}
\mathcal{F}(n)_x, & x\in i(X)\\
0, &\text{sonst}
\end{cases} \]
Sei also o.B.d.A. $X=\mathbf{P}_A^r=\operatorname{Proj}A[X_0,\ldots,X_r]$. Es ist $X=\bigcup_{i=0}^r D_+(X_i)$. Für jedes $i$ ist $\mathcal{F}|_{D_+(X_i)}\cong\widetilde{M_i}$ für ein endlich erzeugter Modul $M_i$ über $B_i=A\big[\frac{X_0}{X_i},\ldots,\widehat{\frac{X_i}{X_i}},\ldots,\frac{X_r}{X_i}\big]$. Sei $(s_{ij})_j$ ein Erzeugendensystem von $M_i$. Wegen \hyperref[5.26]{Lemma 5.26} existiert ein $n_0>0$, so dass für alle $n\geq n_0$ die Schnitte $X_i^ns_{ij}$ sich zu globalen Schnitten $t_{ij}$ von $\mathcal{F}(n)$ liften lassen. Wir können $n_0$ unabhängig von $i$ und $j$ wählen. Sei $\mathcal{F}(n)|_{D_+(X_i)}\cong\widetilde{M_i'}$ für ein $B_i$"~Modul $\widetilde{M_i'}$. Die Abbildungen $X_i^n:\mathcal{F}\to\mathcal{F}(n)$ induzieren Isomorphismen $M_i\to M_i'$. Da $\{X_i^ns_{ij}\mid j\}$ ganz $M_i'$ erzeugen, erzeugen $t_{ij}\in\Gamma(X,\mathcal{F}(n))$ als globale Schnitte ganz $\mathcal{F}(n)$.\qed

\paragraph{Korollar 5.38.}\label{5.38} Sei $X$ ein projektives Schema über $\operatorname{Spec}(A)$ mit $A$ noethersch. Dann gibt es für jeden kohärenten $\mathcal{O}_X$"~Modul $\mathcal{F}$ eine Surjektion $\mathcal{O}_X(n)^N\to\mathcal{F}$ mit $n,N\in\mathbb{Z}$.

\paragraph{Beweis.} Nach \hyperref[5.37]{Satz 5.37} gibt es eine Surjektion $\mathcal{O}_X^N\to\mathcal{F}(n)$. Tensorieren mit $\mathcal{O}_X(-n)$ gibt die Behauptung.\qed

\paragraph{Satz 5.39.}\label{5.39} Sei $k$ ein Körper, $A$ eine endlich erzeugte $k$"~Algebra und $X$ projektives Schema über $\operatorname{Spec}(A)$. Ferner sei $\mathcal{F}$ ein kohärenter $\mathcal{O}_X$"~Modul. Dann ist $\Gamma(X,\mathcal{F})$ ein endlich erzeugter $A$"~Modul.

\paragraph{Beweis.} Wir werden diesen Satz später kohomologisch beweisen.

\paragraph{Korollar 5.40.}\label{5.40} Sei $f:X\to Y$ ein projektiver Morphismus von Schemata von endlichem Typ über einem Körper $k$. Ist $\mathcal{F}$ kohärent auf $X$, so ist auch $f_\ast\mathcal{F}$ kohärent auf $Y$. Insbesondere ist für $A=k$ der Modul $\Gamma(X,\mathcal{F})$ ein endlich dimensionierter $k$"~Vektorraum.

\paragraph{Beweis.} Sei o.B.d.A. $Y=\operatorname{Spec}(A)$ affin, wobei $A$ eine endlich erzeugte $k$"~Algebra ist. Da $f$ projektiv ist, ist $f$ eigentlich und somit separiert und von endlichem Typ, also quasikompakt. Wegen \hyperref[5.13]{Satz 5.13 (iii)} ist $f_\ast\mathcal{F}$ quasikohärent. Es gilt:
\[f_\ast\mathcal{F} = \widetilde{\Gamma(Y,f_\ast\mathcal{F})}=\widetilde{\Gamma(X,\mathcal{F})} \]
Aber $\Gamma(X,\mathcal{F})$ ist endlich erzeugter $A$"~Modul nach \hyperref[5.39]{Satz 5.39}.\qed

\section{Divisoren}

\paragraph{Definition 6.1.}\label{6.1} \begin{enumerate}[(i)]
\item Ein noetherscher lokaler Ring $(R,\mathfrak{m})$ heißt \textit{regulär}\index{regulär}, falls für $k=R/\mathfrak{m}$ gilt:
\[\dim(R)=\dim_k (\mathfrak{m}/\mathfrak{m}^2)\]
Ist $R$ ein noetherscher lokaler Ring, so gilt stets $\dim(R)\leq\dim_k(\mathfrak{m}/\mathfrak{m}^2)$.
\item Ein Schema $X$ heißt \textit{regulär in Kodimension $1$}\index{regulär}, falls jeder Halm $\mathcal{O}_{X,x}$ von $X$ mit $\dim(\mathcal{O}_{X,x})=1$ regulär ist.
\end{enumerate}

\paragraph{Definition 6.3.}\label{6.3}
Ein Ring heißt \textit{normal}\index{Ring!normal}, wenn er ganzabgeschlossen und nullteilerfrei ist.
Ein Schema heißt \textit{normal}\index{Schema!normal}, wenn seine Halme normal sind.

\paragraph{Theorem 6.4.}\label{6.4} Sei $R$ ein noetherscher, normaler Ring und $\mathfrak{p}$ ein Primideal der Hö\-he $1$. Dann ist $R_\mathfrak{p}$ regulär. Genauer: Sei $(R,\mathfrak{m})$ ein noetherscher lokaler Ring der Dimension $1$. Dann sind folgende Aussagen äquivalent:
\begin{enumerate}[(i)]
\item $R$ ist ein diskreter Bewertungsring.
\item $R$ ist ganzabgeschlossen.
\item $R$ ist regulär.
\item $\mathfrak{m}$ ist ein Hauptideal.
\end{enumerate}

\paragraph{Beweis.} Siehe z.B. Matsumura: \glqq Commutative Algebra\grqq, Theorem 3.9 und Ati\-yah-Mac\-Do\-nald: \glqq Introduction to Commutative Algebra\grqq, Proposition 9.2.\qed

\paragraph{Definition.} Ein Schema habe die Eigenschaft $(\star)$, wenn es noethersch, separiert, integer und regulär in Kodimension $1$ ist.

\paragraph{Definition 6.5.}\label{6.5} Sei $X$ ein Schema mit $(\star)$. Dann gilt:
\begin{enumerate}[(i)]
\item Ein \textit{Primdivisor}\index{Primdivisor} auf $X$ ist ein abgeschlossenes, integres Unterschema der Kodimension $1$.
\item Ein \textit{Weil-Divisor}\index{Divisor!Weil} ist ein Element der freien abelschen Gruppe $\operatorname{Div}(X)$, die von den Primdivisoren erzeugt wird. Wir schreiben ein Divisor als $D=\sum_i n_iY_i$ mit Primdivisoren $Y_i$ und $n_i\in\mathbb{Z}$ mit $n_i=0$ für fast alle $i$. Ein solcher Divisor heißt \textit{effektiv}\index{Divisor!effektiv}, falls alle $n_i\geq 0$ sind.
\item Sei $Y$ ein Primdivisor auf $X$ und $\eta$ ein generischer Punkt in $Y$. Dann ist $\mathcal{O}_{X,\eta}$ nach \hyperref[6.4]{Theorem 6.4} ein diskreter Bewertungsring mit Quotientenkörper $K$, der \textit{Funktionenkörper}\index{Funktionenkörper} von $X$. Wir bezeichnen die zugehörige diskrete Bewertung mit  $v_Y:K^\times\to\mathbb{Z}$.

Sei $f\in K^\times$. Ist $v_Y(f)>0$, so sagen wir, dass $f$ eine \textit{Nullstelle}\index{Nullstelle} entlang $Y$ von der Ordnung $v_Y(f)$ hat. Ist $v_Y(f)<0$, so sagen wir, dass $f$ ein \textit{Pol}\index{Pol} entlang $Y$ von der Ordnung $-v_Y(f)$ besitzt.
\end{enumerate}

\paragraph{Lemma 6.6.}\label{6.6} Sei $X$ ein Schema mit $(\star)$ und $f\in K^\times$. Dann ist $v_Y(f)=0$ für fast alle Primdivisoren $Y$.

\paragraph{Beweis.} Sei $\varnothing \neq U = \operatorname{Spec}(A)\subset_\text{o}X$ affin mit $f|_U$ \textit{regulär}, d.h. $f|_U\in\mathcal{O}_X(U)$. Sei $Z=X\setminus U\subsetneq X$ abgeschlossen. Da $X$ noethersch und irreduzibel ist, ist $Z$ noethersch mit $\operatorname{codim}(Z,X)\geq 1$ und besitzt nur endlich viele irreduzible Komponenten. Daher enthält $Z$ höchstens endlich viele Primdivisoren von $X$, alle anderen treffen $U$. Es genügt also zu zeigen, dass es nur endlich viele Primdivisoren $Y$ in $U$ gibt mit $v_Y(f)\neq 0$, d.h. $v_Y(f)>0$. Es gilt mit $Y=\overline{\{ \eta\}}$:
\[v_Y(f)>0\iff f\not\in\mathcal{O}_{U,\eta}^\times=A_\eta^\times\iff f\in\eta\iff Y=V(\eta)\subset V(f)\subset U \]
Da $f\neq 0$, ist $V(f)\subsetneq U$ eine echte abgeschlossene Teilmenge, und enthält daher nur endlich viele irreduzible Komponenten, also Primdivisoren in $U$.\qed

\paragraph{Definition 6.7.}\label{6.7} Sei $X$ ein Schema mit $(\star)$ und $f\in K^\times$. Der Divisor $\operatorname{div}(f)$ von $f$ ist definiert als:
\[\operatorname{div}(f)=\sum_Y v_Y(f)Y \]
wobei $Y$ über die Primdivisoren in $X$ läuft. Diese Summe ist nach \hyperref[6.6]{Lemma 6.6} endlich und somit wohldefiniert. Jeder Divisor der Form $\operatorname{div}(f)$ heißt \textit{Hauptdivisor} oder \textit{prinzipal}.\index{Hauptdivisor}\index{Divisor!prinzipal}.

\paragraph{Bemerkung 6.8.}\label{6.8} Sei $f,g\in K^\times$. Dann gilt:
\[\operatorname{div}\big(\tfrac{f}{g}\big)=\operatorname{div}(f)-\operatorname{div}(g) \]
Somit ist $K^\times\to\operatorname{Div}(X),\ f\mapsto\operatorname{div}(f)$ ein Gruppenhomomorphismus, dessen Bild gerade die Gruppe der Hauptdivisoren in $X$ sind.

\paragraph{Definition 6.9.}\label{6.9} Sei $X$ ein Schema mit $(\star)$. Zwei Divisoren $D,D'$ heißen \textit{linear äquivalent}\index{linear äquivalent} $D\sim D'$, wenn $D-D'$ ein Hauptdivisor ist. Die Gruppe der zugehörigen Äquivalenzklassen $\operatorname{Cl}(X)$ heißt \textit{Divisorenklassengruppe}\index{Divisorenklassengruppe}. Wir haben eine exakte Folge:
\[K^\times\to\operatorname{Div}(X)\to\operatorname{Cl}(X)\to 0 \]

\paragraph{Satz 6.10.}\label{6.10} Sei $A$ ein noetherscher, nullteilerfreier Ring. Dann gilt:
\[A\text{ ist faktoriell}\iff\operatorname{Cl}(\operatorname{Spec}(A))=0 \]

\paragraph{Beweis.} Siehe z.B. Bourbaki: Algèbre Commutative, Chapitre 7 §3 Proposition 2.\qed

\paragraph{Beispiel 6.11.}\label{6.11} \begin{enumerate}
\item Sei $X=\mathbf{A}_k^n$ für ein Körper $k$. Dann gilt $\operatorname{Cl}(X)=0$, da $k[X_1,\ldots,X_n]$ faktoriell ist.
\item Sei $A$ ein Dedekindring. Dann ist $\operatorname{Cl}(\operatorname{Spec}(A))$ gerade die Idealklassengruppe.
\end{enumerate}

\paragraph{Satz 6.12.}\label{6.12} Sei $k$ ein Körper.
\begin{enumerate}[(i)]
\item Sei $X=\mathbf{A}_k^n$ und $Y\subset X$ ein abgeschlossenes Unterschema. Dann ist $Y$ genau dann ein Primdivisor, wenn:
\[Y=V(f)\quad\text{für ein irreduzibles, nicht-konstantes }f\in k[X_1,\ldots,X_n]  \]
\item Sei $X=\mathbf{P}_k^n$ und $Y\subset X$ ein abgeschlossenes Unterschema. Dann ist $Y$ genau dann ein Primdivisor, wenn:
\[Y=V_+(f)\quad\text{für ein homogenes, irreduzibles }f\in k[X_0,\ldots,X_n],\ \deg(f)=r>0 \]
\end{enumerate}

\paragraph{Definition 6.13.}\label{6.13} Sei $X=\mathbf{P}^n_k$. Jeder Primdivisor $Y$ in $X$ hat die Form $Y=V_+(f_Y)$. Betrachte die Abbildung $Y\mapsto\deg(f_Y)$. Diese induziert ein Gruppenhomomorphismus $\operatorname{Div}(X)\to\mathbb{Z}$. Wir zeigen, dass dieser über $\operatorname{Cl}(X)$ faktorisiert. Sei $f\in K^\times$. Dann gilt:
\[\deg\operatorname{div}(f)=\sum_Y v_Y(f) \deg(Y)=\sum_Y v_Y(f) \deg(f_Y) \]
Sei $f=\frac{g}{h}$ mit homogenen $g,h$ vom selben Grad $d$. Sei $g=g_1^{n_1}\cdots g_r^{n_r}$ eine Zerlegung in irreduzible Elemente $g_i$ vom Grad $d_i$. Dann sind $Y_i=\operatorname{div}(g_i)$ nach \hyperref[6.12]{Satz 6.12 (ii)} Primdivisoren. Es gilt:
\[\deg\operatorname{div}(g)=\sum_i n_i\deg(g_i) = \sum_i n_id_i = d \]
Analog ist $\deg\operatorname{div}(h)=d$. Somit ist $\deg\operatorname{div}(f)=\deg\operatorname{div}(g)-\deg\operatorname{div}(h)=0$.\qed

\paragraph{Satz 6.15.}\label{6.15} Sei $X=\mathbf{P}_k^n$. Dann gilt:
\begin{enumerate}[(i)]
\item Ist $D$ ein Divisor auf $X$, so ist $D\sim\deg(D)\cdot V_+(T_0)$
\item $\deg:\operatorname{Cl}(X)\to\mathbb{Z}$ ist ein Isomorphismus.
\end{enumerate}

\paragraph{Beweis.} \begin{enumerate}[(i)]
\item Sei $D=\sum_Y n_YY$ ein Divisor mit Primdivisoren $Y$. Nach \hyperref[6.12]{Satz 6.12 (ii)} ist $Y=V_+(f_Y)$ mit irreduziblen, homogenen Polynomen $f_Y$ vom Grad $r_Y$. Wir schreiben $f_Y=T_0^{r_Y} g_Y$, wobei $g_Y$ ein Polynom in $\frac{T_1}{T_0},\ldots,\frac{T_n}{T_0}$ ist. Die $g_Y$ sind rationale Funktionen auf $\mathbf{P}_k^n$ und es gilt:
\[\operatorname{div}(g_Y) = V_+(f_Y)-r_YV_+(T_0) \implies V_+(f_Y)\sim r_YV_+(T_0) \]
Also gilt $D=\sum_Yn_Y V_+(f_Y)\sim \big(\sum n_Y r_Y\big) V_+(T_0) = \deg(D)\cdot V_+(T_0)$.
\item folgt aus (i) und wegen $\deg V_+(T_0)=1$.\qed
\end{enumerate}

\paragraph{Satz 6.16.}\label{6.16} Sei $X$ ein Schema mit $(\star)$, $Z\subsetneq X$ eine abgeschlossene Teilmenge und $U=X\setminus Z$. Dann gilt:
\begin{enumerate}[(i)]
\item Es gibt einen surjektiven Homomorphismus $\operatorname{Cl}(X)\to\operatorname{Cl}(U)$, der durch $\sum n_iY_i\mapsto \sum n_i(Y_i\cap U)$ gegeben ist, wobei wir die leeren $Y_i\cap U$ ignorieren.
\item Ist $\operatorname{codim}(Z,X)\geq 2$, dann ist die obige Abbildung ein Isomorphismus.
\item Ist $Z$ irreduzibel und $\operatorname{codim}(Z,X)=1$, so gibt es eine exakte Folge:
\[\mathbb{Z}\to\operatorname{Cl}(X)\to\operatorname{Cl}(U)\to 0 \]
wobei die erste Abbildung durch $1\mapsto Z$ gegeben ist.
\end{enumerate}

\paragraph{Beweis.}\begin{enumerate}[(i)]
\item Ist $Y$ ein Primdivisor auf $X$, so ist $Y\cap U$ leer oder ein Primdivisor auf $U$. Sei $f\in K^\times$ mit $\operatorname{div}(f)=\sum n_iY_i$. Fassen wir $f$ als rationale Funktion auf $U$ auf, so erhalten wir $\operatorname{div}(f|_U)=\sum n_i(Y_i\cap U)$. Somit ist die Abbildung $\operatorname{Cl}(X)\to\operatorname{Cl}(U)$ wohldefiniert.

Die Surjektivität sieht man wie folgt: Sei $Y$ ein Primdivisor auf $U$ und $\overline{Y}$ der Abschluss von $Y$ in $X$. Dann ist $\overline{Y}$ ein Primdivisor auf $X$ mit $\overline{Y}\cap U=Y$.
\item $\operatorname{Div}(X)$ bzw. $\operatorname{Cl}(X)$ hängen nur von Teilmengen der Kodimension $1$ ab.
\item Es ist $\ker(\operatorname{Cl}(X)\to\operatorname{Cl}(U)) = \{ [D]\in\operatorname{Cl}(X)\mid \operatorname{supp}(D)\subset Z\}=\langle[Z]\rangle$, da $Z$ irreduzibel ist.\qed
\end{enumerate}

\paragraph{Beispiel 6.17.}\label{6.17} Sei $Y$ eine irreduzible Kurve vom Grad $d$ in $\mathbf{P}_k^2$. Wir haben ein kommutatives Diagramm:
\[\begin{tikzcd}
\mathbb{Z} \rar\dar["\cong"'] & \operatorname{Cl}(\mathbf{P}_k^2) \rar\dar["\cong"', "\deg"]& \operatorname{Cl}(\mathbf{P}_k^2\setminus Y) \rar\dar["\cong"]& 0\\
d\mathbb{Z} \rar[hook] & \mathbb{Z}\rar & \mathbb{Z}/d\mathbb{Z}\rar & 0
\end{tikzcd}\]

\paragraph{Definition 6.18.}\label{6.18} Sei $k$ ein algebraisch abgeschlossener Körper. Eine \textit{Kurve}\index{Kurve} über $k$ ist ein integres, separiertes Schema $X$ von endlichem Typ über $k$ der Dimension $1$.
\begin{itemize}
\item $X$ heißt \textit{vollständig}\index{Kurve!vollständig}, wenn $X\to k$ eigentlich ist.
\item $X$ heißt \textit{nicht-singulär}\index{nicht-singulär}, falls alle lokalen Ringe von $X$ regulär sind.
\end{itemize}

\paragraph{Satz 6.19.}\label{6.19} Sei $X$ eine vollständige, nicht-singuläre Kurve über $k$ und $Y$ eine beliebige Kurve über $k$ mit einem Morphismus $f:X\to Y$. Dann gilt:
\[f(X)=\operatorname{Pt} \quad\text{oder}\quad f(X)=Y \]
Im zweiten Fall gilt:
\begin{enumerate}[(i)]
\item Die Funktionenkörpererweiterung $K(X)/K(Y)$ ist endlich.
\item $f$ ist endlich.
\item $Y$ ist vollständig.
\end{enumerate}

\paragraph{Lemma 6.20.}\label{6.20} Sei $f:X\to Y$ surjektiv, $g:Y\to Z$ separiert und von endlichem Typ, und $g\circ f:X\to Z$ eigentlich. Dann ist $g$ eigentlich.

\paragraph{Beweis von \hyperref[6.19]{6.19}.} Da $X$ eigentlich über $k$ ist, ist $f(X)\subset Y$ ein abgeschlossenes Unterschema. Da $X$ irreduzibel ist, ist $f(X)$ auch irreduzibel. Nun ist $\dim(Y)=1$, folgt entweder $f(X)=\operatorname{Pt}$ oder $f(X)=Y$. Betrachte das folgende kommutative Diagramm:
\[\begin{tikzcd}
X \rar["f"] & f(X)\rar[hook]\ar[dr] & Y\dar\\
 & & k
\end{tikzcd} \]
Es ist $f(X)\hookrightarrow Y$ eine abgeschlossene Immersion, also separiert nach \hyperref[4.17]{Satz 4.17 (i)}. Somit ist auch $f(X)/ k$ separiert. Nach \hyperref[6.20]{Lemma 6.20} ist $f(X)/k$ eigentlich.

Sei nun $f(X)=Y$, also ist $Y$ vollständig. Sei $Y=\overline{\{\eta\}}$ und $X=\overline{\{\xi\}}$. Da $f:X\to Y$ dominant ist, folgt $f(\xi)=\eta$ und wir erhalten eine Inklusion $K(Y)=\mathcal{O}_{Y,\eta}\to\mathcal{O}_{X,\xi}=K(X)$. Da beide Körper endlich erzeugt über $k$ vom Transzendenzgrad $\dim(X)=1$ sind, ist $K(X)/K(Y)$ endlich.

Es bleibt zu zeigen, dass $f$ endlich ist. Sei $V=\operatorname{Spec}(B)\subset_\text{o}Y$ affin. Dann ist $B=\mathcal{O}_Y(V)\subset K(Y)\subset K(X)$. Es ist zu zeigen, dass $f^{-1}(V)=\operatorname{Spec}(A)$ affin ist, wobei $A$ ein endlich erzeugter $B$"~Modul ist. In der Tat folgt aus der Normalität von $X$ stets $f^{-1}(V)=\operatorname{Spec}(A)$, wobei $A$ der Ganzabschluss von $B$ in $K(X)$ ist. Alles folgt nun aus dem folgenden \hyperref[6.21]{Theorem 6.21}.\qed

\paragraph{Theorem 6.21.}\label{6.21} Sei $B$ ein Integritätsring und eine endlich erzeugte Algebra über einem Körper $k$. Setze $K=\operatorname{Quot}(B)$ und sei $L/K$ eine endliche Erweiterung. Dann ist der Ganzabschluss $A$ von $B$ in $L$ ein endlich erzeugter $B$"~Modul.

\paragraph{Beweis.} Siehe z.B. Zariski-Samuel: Commutative Algebra I, Chapter V, Theorem 9.\qed

\paragraph{Beweis von \hyperref[6.20]{6.20}.} Es reicht zu zeigen, dass $g$ universell abgeschlossen ist. Sei $h:Y'\to Y$ ein beliebiger Morphismus und betrachte den Basiswechsel:
\[\begin{tikzcd}
X\times_Y Y' \ar[r, "f'"]\ar[d, "h'"'] & Y' \ar[d, "h"]\\
X\ar[r, "f"'] & Y
\end{tikzcd} \]
Wir zeigen nun, dass $f'$ surjektiv ist. Sei $y'\in Y'$ und sei $x\in X$ mit $f(x)=h(y')=:y$. Behauptung: Es gibt ein $z\in X\times_YY'$ mit $h'(z)=x$ und $f'(z)=y'$. Seien $i_x:\{x\}\hookrightarrow X,\ i_{y'}:\{y'\}\hookrightarrow Y'$ die Inklusionen. Diese liefern einen $Y$"~Morphismus:
\[\delta : \operatorname{Spec}\kappa(x)\times_{\kappa(y)}\operatorname{Spec}\kappa(y') = \operatorname{Spec}(\kappa(x)\otimes_{\kappa(y)}\kappa(y'))\hookrightarrow X\times_YY' \]
Da $\kappa(x)\otimes_{\kappa(y)}\kappa(y')\neq 0$, gilt für jedes $z\in\operatorname{im}(\delta)$ stets $h'(z)=x$ und $f'(z)=y'$. Dies zeigt die Surjektivität von $f'$.

Also ist die Surjektivität von $f$ stabil unter Basiswechsel. Da alle anderen Eigenschaften in den Voraussetzungen ebenfalls stabil unter Basiswechsel sind (siehe \hyperref[4.17]{4.17 (iii)} und \hyperref[4.22]{4.22}), reicht es zu zeigen, dass $g$ abgeschlossen ist. Sei $Y'\subset Y$ eine abgeschlossene Teilmengen. Dann ist wegen der Surjektivität von $f$:
\[g(Y')=(g\circ f)\circ f^{-1}(Y') \]
$g(Y')$ ist abgeschlossen, da $g\circ f$ abgeschlossen ist.\qed

\paragraph{Definition 6.22.}\label{6.22} Sei $f:X\to Y$ ein dominanter Morphismus von Kurven. Dann heißt $\deg(f)=[K(X):K(Y)]$ der \textit{Grad} von $f$\index{Grad}. Sei $X$ eine nicht-singuläre Kurve, erfüllt also insbesondere $(\star)$. Ein Primdivisor von $X$ ist genau ein abgeschlossener Punkt. Ein Divisor $D$ ist somit von der folgenden Form:
\[D=\sum_i n_iP_i,\quad n_i\in\mathbb{Z},\ P_i\text{ abgeschlossene Punkte} \]
Der \textit{Grad}\index{Grad} von $D$ ist definiert als $\deg(D)=\sum n_i$.

\paragraph{Definition 6.23.}\label{6.23} Sei $f:X\to Y$ ein endlicher Morphismus nicht-singulärer Kurven. Wir definieren ein Homomorphismus $f^\ast:\operatorname{Div}(Y)\to\operatorname{Div}(X)$ wie folgt:

Sei $Q\in Y$ ein abgeschlossener Punkt und wähle ein $t\in\mathcal{O}_{Y,Q}\subset K(Y)$ mit $v_Q(t)=1$, wobei $v_Q$ die diskrete Bewertung zu $\mathcal{O}_{Y,Q}$ ist. $t$ heißt \textit{lokaler Parameter}\index{lokaler Parameter} an der Stelle $Q$. Wir setzen:
\[f^\ast(Q) = \sum_{P\in f^{-1}(Q)}v_P(t)P \]
Da $f$ endlich ist, ist obige Summe nach \hyperref[6.24]{Lemma 6.24} endlich.

\paragraph{Lemma 6.24.}\label{6.24} \begin{enumerate}[(i)]
\item Die Eigenschaft endlich zu sein ist stabil unter Basiswechsel.
\item Ist $f:X\to Y$ ein endlicher Morphismus und $y\in Y$, so ist $f^{-1}(y)$ endlich.
\end{enumerate}

\paragraph{Beweis.} \begin{enumerate}[(i)]
\item Sei $f:X\to Y$ ein endlicher Morphismus und $g:Z\to Y$ beliebig. Betrachte den folgenden Basiswechsel:
\[\begin{tikzcd}
X\times_YZ \dar\rar["f_Z"] & Z\dar["g"]\\
X \rar["f"'] & Y
\end{tikzcd} \]
Sei $V=\operatorname{Spec}(B)\subset_\text{o}Y$ affin. Dann existiert ein affines $W=\operatorname{Spec}(C)\subset_\text{o}Z$ mit $g(W)\subset V$. Es ist $f^{-1}(V)=\operatorname{Spec}(A)$ affin, wobei $A$ ein endlich erzeugter $B$"~Modul ist. Es gilt:
\begin{align*}
f_Z^{-1}(W)= (X\times_YZ)\times_ZW &\cong X\times_YW \\
&\cong (X\times_YV)\times_VW\\
&\cong f^{-1}(V)\times_VW=\operatorname{Spec}(A\otimes_BC)
\end{align*}
und $A\otimes_BC$ ist ein endlich erzeugter $C$"~Modul.
\item Sei $f^{-1}(y)\neq\varnothing$. Betrachte den Morphismus:
\[f^{-1}(y)= X\times_Y\operatorname{Spec}\kappa(y)\to\operatorname{Spec}\kappa(y) \]
Dieser ist endlich nach (i). Also ist $f^{-1}(y)=\operatorname{Spec}(A)$ affin, wobei $A$ ein endlich dimensionierter $\kappa(y)$"~Vektorraum ist. Somit ist $\dim f^{-1}(y)=\dim\kappa(y)= 0$, als topologischer Raum ist $f^{-1}(y)$ also eine endliche Menge an Punkten.\qed
\end{enumerate}

\paragraph{Bemerkung.} \begin{enumerate}[(i)]
\item In \hyperref[6.23]{Definition 6.23} ist $f^\ast Q$ unabhängig von der Wahl des lokalen Parameters, da ein lokaler Parameter eindeutig bis auf eine Einheit in $\mathcal{O}_{Y,Q}$ bestimmt ist.
\item $f^\ast$ \textit{respektiert lineare Äquivalenz}, d.h. für $h\in K(Y)^\times,\ \operatorname{div}(h)=\sum_Qv_Q(h)Q$ gilt:
\[f^\ast(\operatorname{div}(h))=\sum_Qv_Q(h)\sum_{P\in f^{-1}(Q)}v_P(t)P=\sum_Pv_P(h)P=\operatorname{div}(h) \]
da $v_P(h)=v_P(t)\cdot v_Q(h)$ ist. Somit induziert $f^\ast$ eine Abbildung $\operatorname{Cl}(Y)\to\operatorname{Cl}(X)$.\qed
\end{enumerate}

\paragraph{Satz 6.25.}\label{6.25} Sei $f:X\to Y$ ein endlicher Morphismus nicht-singulärer Kurven. Dann:
\[\deg(f^\ast D)=\deg(f)\cdot \deg(D) \]

\paragraph{Korollar 6.26.}\label{6.26} Hauptdivisoren auf einer vollständigen, nicht-singulären Kurve haben Grad $0$. Wir erhalten eine surjektive Gradabbildung:
\[\deg:\operatorname{Cl}(X)\to\mathbb{Z} \]

\paragraph{Lemma 6.27.}\label{6.27} Sei $X$ eine normale Kurve über $k$, $U\subset X$ offen, $\varphi:X\to\mathbf{P}^n$ eine rationale Abbildung. Dann lässt sich $\varphi$ zu einem einem Morphismus auf $X$ fortsetzen.

\paragraph{Definition 6.28.}\label{6.28} Sei $(X,\mathcal{O})$ ein Schema.
\begin{enumerate}[(i)]
\item Sei $U=\operatorname{Spec}(A)\subset_\text{o}X$ affin und $S$ die Menge aller Nicht-Nullteiler in $A$. Dann heißt $K(U)=A_S$ \textit{totaler Quotientenring} von $A$.\index{totaler Quotientenring}
\item Sei $\mathcal{K}$ die Ringgarbe, die zur Prägarbe $U\mapsto\varprojlim_{V\subset U\text{ affin}} K(V)$ assoziiert ist. $\mathcal{K}$ heißt \textit{Garbe der totalen Quotientenringe} von $\mathcal{O}$.
\item $\mathcal{K}^\times:U\mapsto \mathcal{K}(U)^\times$ sei die Garbe von multiplikativen Gruppe der invertierbaren Elemente in $\mathcal{K}$. $\mathcal{O}^\times$ sei die Garbe der invertierbaren Elemente in $\mathcal{O}$.
\end{enumerate}

\paragraph{Definition 6.29.}\label{6.29} Sei $(X,\mathcal{O})$ ein Schema. Ein Element aus $\Gamma(X,\mathcal{K}^\times/\mathcal{O}^\times)$ heißt \textit{Cartier-Divisor}\index{Divisor!Cartier}. Ein Cartier-Divisor kann also durch ein System $(U_i,f_i)_i$ beschrieben werden, wobei $(U_i)_i$ eine offene Überdeckung von $X$ ist, und $f_i\in\Gamma(U_i,\mathcal{K}^\times)$, so dass für alle $i,j$ stets $\frac{f_i}{f_j}\in\Gamma(U_i\cap U_j,\mathcal{O}^\times)$ gilt.

Ein Cartier-Divisor heißt \textit{prinzipal}\index{Divisor!prinzipal} oder \textit{Hauptdivisor}\index{Hauptdivisor}, wenn er im Bild der kanonischen Abbildung $\Gamma(X,\mathcal{K}^\times)\to\Gamma(X,\mathcal{K}^\times/\mathcal{O}^\times)$ ist. Zwei Cartier-Divisoren heißen \textit{linear äquivalent}\index{linear äquivalent}, falls ihre Differenz prinzipal ist.

\paragraph{Satz 6.30.}\label{6.30} Sei $(X,\mathcal{O})$ ein integres, separiertes, noethersches Schema. Ferner sei $X$ \textit{lokal faktoriell}\index{lokal faktoriell}, d.h. alle lokalen Ringe sind faktoriell. Dann ist die Gruppe $\operatorname{Div}(X)$ von Weil-Divisoren isomorph zur Gruppe der Cartier-Divisoren $\Gamma(X,\mathcal{K}^\times/\mathcal{O}^\times)$. Prinzipale Weil-Divisoren entsprechen prinzipale Cartier-Divisoren.

\paragraph{Beweis.} $X$ ist normal, da faktorielle Ringe insbesondere ganzabgeschlossen sind. Nach \hyperref[6.4]{Theorem 6.4} erfüllt $X$ $(\star)$. Also können wir von Weil-Divisoren sprechen. Da $X$ integer ist, ist $K(U)=\operatorname{Quot}(A)= K$ der Funktionenkörper von $X$ für alle $U=\operatorname{Spec}(A)\subset_\text{o}X$. Somit ist $\mathcal{K}$ konstante Garbe.

Sei $(U_i,f_i)_i$ ein Cartier-Divisor mit einer offenen Überdeckung $(U_i)_i$ von $X$ und $f_i\in\Gamma(U_i,\mathcal{K}^\times)=K^\times$. Wir ordnen diesen Cartier-Divisor den folgenden Weil-Divisor zu:
\[D=\sum_Y v_Y(f_{i_Y})Y \]
wobei $Y$ die Primdivisoren von $X$ durchläuft und $i_Y$ ein Index mit $U_i\cap Y\neq\varnothing$. Die Summe ist endlich, da $X$ noethersch ist. $D$ ist unabhängig von Wahl der Indizes $i_Y$:

Seien $i,j$ mit $U_i\cap Y\neq\varnothing$ und $U_j\cap Y\neq\varnothing$. Dann ist $\frac{f_i}{f_j}$ auf $U_i\cap U_j$ invertierbar, d.h. $\frac{f_i}{f_j}\in\Gamma(U_i\cap U_j,\mathcal{O}^\times)$. Es folgt $v_Y\big(\frac{f_i}{f_j}\big)=0$, also $v_Y(f_i)=v_Y(f_j)$.

Sei nun umgekehrt $D=\sum_Yn_YY$ ein Weil-Divisor auf $X$ und $x\in X$. Dann induziert $D$ ein Weil-Divisor $D_x$ auf $\operatorname{Spec}(\mathcal{O}_x)$, nämlich:
\[D_x=\sum_Y n_Y(Y\cap\operatorname{Spec}(\mathcal{O}_x)) \]
Da $\mathcal{O}_x$ als faktoriell vorausgesetzt wird, ist nach \hyperref[6.10]{Satz 6.10} $\operatorname{Cl}(\operatorname{Spec}\mathcal{O}_x)=0$ und $D_x$ ein Hauptdivisor, d.h. $D_x=(f_x)$ für ein $f_x\in K^\times$. Fassen wir $(f_x)$ als Weil-Divisor in $X$ auf, so sehen wir, dass sich $(f_x)$ und $D$ nur bei Primdivisoren, die nicht durch $x$ gehen, unterscheiden. Davon gibt es nur endlich viele, deren Koeffizienten nicht verschwinden. Daher existiert eine offene Umgebung $U_x$ von $x$ mit $(f_x)|_{U_x}=D|_{U_x}$. Das System $(U_x,f_x)_{x\in X}$ liefert einen Cartier-Divisor.

Geben $f,f'$ denselben Weil-Divisor auf $U\subset_\text{o}X$ offen, so ist $\frac{f}{f'}\in\Gamma(U,\mathcal{O}^\times)$, da $X$ normal ist. Daher ist die Konstruktion wohldefiniert. Die obigen Konstruktionen sind invers zueinander.\qed

\paragraph{Satz 6.31.}\label{6.31} Seien $\mathcal{L},\mathcal{M}$ invertierbare Garben auf einem Schema $X$. Dann gilt:
\begin{enumerate}[(i)]
\item $\mathcal{L}\otimes \mathcal{M}$ ist invertierbar.
\item $\operatorname{\mathcal{H}om}_{\mathcal{O}_X}(\mathcal{L},\mathcal{O}_X)$ ist invertierbar und wird mit $\mathcal{L}^{-1}$ bezeichnet.
\item $\mathcal{L}\otimes\mathcal{L}^{-1}\cong\mathcal{O}_X$
\item $\operatorname{\mathcal{H}om}_{\mathcal{O}_X}(\mathcal{L},\mathcal{L})=\mathcal{O}_X$
\end{enumerate}

\paragraph{Definition 6.32.}\label{6.32}\begin{enumerate}[(i)]
\item Wir setzen:
\[\mathcal{L}^{\otimes m}=\underbrace{\mathcal{L}\otimes\cdots\otimes\mathcal{L}}_{m\text{-mal}},\quad\quad \mathcal{L}^{\otimes -m}=(\mathcal{L}^{-1})^{\otimes m},\quad m\geq 0 \]
Also gilt $\mathcal{L}^{\otimes n}\otimes\mathcal{L}^{\otimes m}=\mathcal{L}^{\otimes (n+m)}$ für alle $n,m\in\mathbb{Z}$.
\item Die \textit{Picard-Gruppe} $\operatorname{Pic}(X)$\index{Picard-Gruppe} eines Schemas $X$ ist die abelsche Gruppe von Isomorphieklassen invertierbarer Garben auf $X$ mit der Operation $\otimes$.
\end{enumerate}

\paragraph{Definition 6.33.}\label{6.33} Sei $D$ ein Cartier-Divisor auf $X$ repräsentiert durch $(U_i,f_i)_i$. Dann ist die Untergarbe $\mathcal{L}(D)$ von $\mathcal{K}$ definiert durch:
\[\mathcal{L}(D)|_{U_i}=\mathcal{O}_{U_i}f_i^{-1}\subset\mathcal{K}|_{U_i} \]
Diese ist wohldefiniert, denn auf $U_i\cap U_j$ haben wir $\mathcal{L}(D)|_{U_i\cap U_j}=\mathcal{O}_{U_i\cap U_j}f_i^{-1}=\mathcal{O}_{U_i\cap U_j}f_j^{-1}$, da $\frac{f_i}{f_j}\in\Gamma(U_i\cap U_j,\mathcal{O}^\times)$. $\mathcal{L}(D)$ heißt die zum Divisor $D$ \textit{assoziierte Garbe}\index{assoziierte Garbe!zum Divisor}.

\paragraph{Satz 6.34.}\label{6.34} Sei $X$ ein Schema.
\begin{enumerate}[(i)]
\item Für jeden Cartier-Divisor $D$ ist $\mathcal{L}(D)$ eine invertierbare Garbe auf $X$. Die Abbildung $D\mapsto\mathcal{L}(D)$ induziert eine Bijektion:
\[\{\text{Cartier-Divisoren auf }X \}\to\{\text{Invertierbare Untergarben von }\mathcal{K} \} \]
\item $\mathcal{L}(D_1-D_2)\cong\mathcal{L}(D_1)\otimes\mathcal{L}(D_2)^{-1}$ für Cartier-Divisoren $D_1,D_2$
\item $D_1\sim D_2\iff\mathcal{L}(D_1)\cong\mathcal{L}(D_2)$
\end{enumerate}

\paragraph{Beweis.} \begin{enumerate}[(i)]
\item Sei repräsentiert durch $(U_i,f_i)_i$ mit $f_i\in\Gamma(U_i,\mathcal{K}^\times)$. Dann ist $\mathcal{O}_{U_i}\to\mathcal{L}(D)|_{U_i},\ 1\mapsto f_i^{-1}$ ein Isomorphismus, also $\mathcal{L}(D)$ invertierbar. Sei umgekehrt $\varphi:\mathcal{L}\hookrightarrow\mathcal{K}$ eine invertierbare Untergarbe und $\{U_i\}$ eine Überdeckung von $X$ mit $\tilde{\varphi}:\mathcal{O}_{U_i}\stackrel{\sim}{\to}\mathcal{L}|_{U_i}\hookrightarrow\mathcal{K}|_{U_i}$. Setze $f_i=\tilde{\varphi}(U_i)(1)^{-1}$. Dann definiert $(U_i,f_i)_i$ ein Cartier-Divisor $D$ mit $\mathcal{L}(D)=\mathcal{L}$.
\item Seien $D_1, D_2$ repräsentiert durch $(U_i,f_i)_i$ bzw. $(U_i,g_i)_i$. Dann wird $D_1-D_2$ repräsentiert durch $(U_i,f_ig_i^{-1})_i$. Also gilt:
\[\mathcal{L}(D_1-D_2)|_{U_i}=\mathcal{O}_{U_i}(f_i^{-1}g_i)=\mathcal{L}(D_1)|_{U_i}\otimes \mathcal{L}(D_2)|_{U_i}^{-1}\subset\mathcal{K}|_{U_i} \]
Somit folgt $\mathcal{L}(D_1-D_2)=\mathcal{L}(D_1)\otimes\mathcal{L}(D_2)^{-1}$.
\item Wegen (ii) reicht es zu zeigen, dass $D$ genau dann ein Hauptdivisor ist, wenn $\mathcal{L}(D)\cong\mathcal{O}_X$. Sei $D$ prinzipal, d.h. $D=(f)$ für ein $f\in\Gamma(X,\mathcal{K}^\times)$. Somit ist $\mathcal{L}(D)=\mathcal{O}_Xf^{-1}\cong\mathcal{O}_X$. Sei umgekehrt $\varphi:\mathcal{O}_X\to\mathcal{L}(D)$ ein Isomorphismus. Setze $f$ als das Bild von $1$ der folgenden Komposition:
\[\Gamma(X,\mathcal{O}_X)\stackrel{\sim}{\to} \Gamma(X,\mathcal{L}(D))\hookrightarrow\Gamma(X,\mathcal{K}^\times) \]
Dann ist $\mathcal{L}(D)=\mathcal{L}((f^{-1}))$, also $D=(f^{-1})$.\qed
\end{enumerate}

\paragraph{Korollar 6.35.}\label{6.35} Sei $X$ ein Schema. Dann induziert die Abbildung $D\mapsto\mathcal{L}(D)$ einen injektiven Gruppenhomomorphismus:
\[\operatorname{CaCl}(X)\hookrightarrow\operatorname{Pic}(X)\]
wobei $\operatorname{CaCl}(X)$ die Cartier-Divisorenklassengruppe bezeichnet. Im Allgemeinen ist diese Abbildung nicht surjektiv.

\paragraph{Satz 6.36.}\label{6.36} Sei $X$ ein integres Schema. Dann ist die Abbildung $\operatorname{CaCl}(X)\to\operatorname{Pic}(X)$ ein Gruppenisomorphismus.

\paragraph{Beweis.} Es genügt zu zeigen, dass jede invertierbare Garbe isomorph zu einer Untergarbe von $\mathcal{K}$ ist. Da $X$ integer ist, ist $\mathcal{K}$ konstante Garbe mit $\mathcal{K}(U)=K$ der Funktionenkörper von $X$ für alle $U$. Sei $\mathcal{L}$ eine invertierbare Garbe auf $X$ und betrachte $\mathcal{L}\otimes\mathcal{K}$. Sei $X=\bigcup_i U_i$ eine offene Überdeckung mit $\mathcal{L}|_{U_i}\cong\mathcal{O}_{U_i}$. Dann gilt:
\[(\mathcal{L}\otimes_{\mathcal{O}_X}\mathcal{K})|_{U_i}=\mathcal{L}|_{U_i}\otimes_{\mathcal{O}_{U_i}}\mathcal{K}|_{U_i}=\mathcal{K}|_{U_i} \]
Da $X$ irreduzibel ist, folgt $\mathcal{L}\otimes\mathcal{K}\cong\mathcal{K}$. Die kanonische Abbildung $\mathcal{L}\to\mathcal{L}\otimes\mathcal{K}\cong\mathcal{K}$ zeigt, dass $\mathcal{L}$ isomorph zu einer Untergarbe von $\mathcal{K}$ ist.\qed

\paragraph{Korollar 6.37.}\label{6.37} Sei $X$ ein noethersches, integres, separiertes und lokal faktorielles Schema. Dann existiert ein Isomorphismus:
\[\operatorname{Cl}(X)\cong\operatorname{Pic}(X) \]

\paragraph{Beweis.} Folgt aus \hyperref[6.30]{Satz 6.30} und \hyperref[6.36]{Satz 6.36}.\qed

\paragraph{Korollar 6.38.}\label{6.38} Sei $k$ ein Körper und $X=\mathbf{P}_k^n$. Dann ist jede invertierbare Garbe auf $X$ isomorph zu einem $\mathcal{O}(\ell)$ für ein $\ell\in\mathbb{Z}$.

\paragraph{Beweis.} Nach \hyperref[6.37]{Korollar 6.37} und \hyperref[6.15]{Satz 6.15} ist $\operatorname{Pic}(X)\cong \operatorname{Cl}(X)\cong\mathbb{Z}$. Ferner wird $\operatorname{Cl}(X)$ von der Hyperebene $D=V_+(T_0)$ erzeugt. Dann gilt $\mathcal{L}(D)T_0=\mathcal{O}(1)$, da $\mathcal{O}(1)|_{D_+(T_i)}=\mathcal{O}_{D_+(T_i)}T_i$ und $\mathcal{L}(D)|_{D_+(T_i)}=\mathcal{O}_{D_+(T_i)}\frac{T_i}{T_0}$.\qed

\section{Projektive Morphismen}

\subsection{Ample Garben}

\paragraph{Satz 7.1.}\label{7.1} Sei $A$ ein Ring, $X$ ein Schema über $A$ und $\mathbf{P}_A^n=\operatorname{Proj}A[x_0,\ldots,x_n]$.
\begin{enumerate}[(i)]
\item Ist $\varphi:X\to\mathbf{P}_A^n$ ein $A$-Morphismus. Dann ist $\varphi^\ast(\mathcal{O}(1))$ eine invertierbare Garbe auf $X$, die von globalen Schnitten $s_i=\varphi^\ast(x_i),\ i=0,\ldots,n$ erzeugt wird, wobei $x_i\in\Gamma(\mathbf{P}_A^n,\mathcal{O}(1))$.
\item Sei $\mathcal{L}$ eine invertierbare Garbe auf $X$ und $s_0,\ldots,s_n\in\Gamma(X,\mathcal{L})$ globale Schnitte, die $\mathcal{L}$ erzeugen. Dann existiert ein eindeutig bestimmter $A$"~Morphismus $\varphi:X\to\mathbf{P}_A^n$ derart, dass $\mathcal{L}\cong\varphi^\ast(\mathcal{O}(1))$ mit $s_i=\varphi^\ast(x_i)$.
\end{enumerate}

\paragraph{Beweis.} \begin{enumerate}[(i)]
\item Nach \hyperref[5.35]{5.35} erzeugen die globalen Schnitte $x_0,\ldots,x_n$ die Garbe $\mathcal{O}(1)$. Ferner ist $\varphi^\ast(\mathcal{O}(1))$ invertierbar und werden von $\varphi^\ast(x_i)=s_i$ erzeugt.
\item Setze $X_i=\{P\in X\mid (s_i)_P\not\in\mathfrak{m}_P\mathcal{L}_P\}\subset_\text{o}X$. Da die $s_i$ die Garbe $\mathcal{L}$ erzeugen, gilt $X=\bigcup_iX_i$. Sei $U_i=D_+(x_i)\cong\operatorname{Spec}A[Y_0,\ldots,\widehat{Y_i},\ldots,Y_n],\ Y_j=\frac{x_j}{x_i}$. Betrachte den Ringhomomorphismus $\phi:A[Y_0,\ldots,\widehat{Y_i},\ldots,Y_n]\to\Gamma(X_i,\mathcal{O}_{X_i}),\ Y_j\mapsto \frac{s_j}{s_i}$. Dieser ist wohldefiniert, da  $(s_i)_P\not\in \mathfrak{m}_P\mathcal{L}_P\cong\mathfrak{m}_P\mathcal{O}_{X,P}$ für alle $P\in X_i$ gilt und daher $\frac{s_j}{s_i}\in\Gamma(X_i,\mathcal{O}_{X_i})$. Sei $X_i\to U_i$ der zu $\phi$ gehörige $A$"~Morphismus von $A$"~Schemata. Verkleben ergibt ein $A$"~Morphismus $\varphi:X\to\mathbf{P}_A^n$. Nach Konstruktion ist $\varphi$ eindeutig bestimmt durch $\mathcal{L}\cong\varphi^\ast(\mathcal{O}(1))$ und $s_i=\varphi^\ast(x_i)$.\qed
\end{enumerate}

\paragraph{Satz 7.2.}\label{7.2} Sei $\varphi:X\to\mathbf{P}_A^n$ ein $A$"~Morphismus, der zu einer invertierbaren Garbe $Z$ auf $X$ und globale Schnitte $s_0,\ldots,x_n\in\Gamma(X,Z)$ gehört, die $\mathcal{L}$ erzeugen. Dann ist $\varphi$ genau dann eine abgeschlossene Immersion, wenn die folgenden Bedingungen gelten:
\begin{enumerate}[(i)]
\item Alle offenen Teilmengen $X_i$ im Beweis von \hyperref[7.1]{Satz 7.1 (ii)} sind affin.
\item Für alle $i$ ist die Abbildung $A[Y_0,\ldots,\widehat{Y_i},\ldots,Y_n]\to\Gamma(X_i,\mathcal{O}_{X_i}),\ Y_j\mapsto \frac{s_j}{s_i}$ surjektiv.
\end{enumerate}

\paragraph{Beweis.} Sei $\varphi$ eine abgeschlossene Immersion. Dann ist $X_i=U_i\cap X$ ein abgeschlossenes Unterschema von $U_i=D_+(x_i)\subset\mathbf{P}_A^n=\operatorname{Proj}A[x_0,\ldots,x_n]$. Da $U_i$ affin ist, ist nach \hyperref[5.18]{Korollar 5.18} auch $X_i$ affin und wir erhalten einen surjektiven Ringhomomorphismus $A[Y_0,\ldots,\widehat{Y_i},\ldots,Y_n]\to \Gamma(X_i,\mathcal{O}_{X_i})$.

Sei nun umgekehrt (i) und (ii) erfüllt. Dann ist $X_i$ nach (ii) ein abgeschlossenes Unterschema von $U_i$. Da $X=\bigcup_i X_i$ und $X_i=\varphi^{-1}(U_i)$, ist $X$ ein abgeschlossenes Unterschema von $\mathbf{P}_A^n$.\qed

\paragraph{Definition 7.3.}\label{7.3} Eine invertierbare Garbe $\mathcal{L}$ auf einem noetherschen Schema $X$ heißt \textit{ampel}\index{$\mathcal{O}_X$-Modul!ampel}, falls für jede kohärente Garbe $\mathcal{F}$ auf $X$ ein $n_0>0$ existiert, so dass für alle $n\geq n_0$ der $\mathcal{O}_X$"~Modul $\mathcal{F}\otimes\mathcal{L}^n$ von globalen Schnitten erzeugt wird.

\paragraph{Beispiel 7.4.}\label{7.4} Ist $X$ affin, so ist jede invertierbare Garbe ampel: Jede kohärente Garbe wird von globalen Schnitten endlich erzeugt, vgl. \hyperref[5.10]{Korollar 5.10}.

\paragraph{Bemerkung 7.5.}\label{7.5} \hyperref[5.37]{Serres Theorem 5.37} besagt, dass jede sehr ample Garbe $\mathcal{L}$ auf einem Schema $X$ über einem noetherschen Ring ampel ist. Die Umkehrung ist im Allgemeinen falsch.

\paragraph{Satz 7.6.}\label{7.6} Sei $\mathcal{L}$ eine invertierbare Garbe auf einem noetherschen Schema $X$. Dann sind folgende Aussagen äquivalent:
\begin{enumerate}[(i)]
\item $\mathcal{L}$ ist ampel.
\item $\mathcal{L}^m$ ist ampel für alle $m>0$.
\item $\mathcal{L}^m$ ist ampel für ein $m>0$.
\end{enumerate}

\paragraph{Beweis.} (i)$\implies$(ii)$\implies$(iii) ist klar. Sei nun $\mathcal{L}^m$ ampel für ein $m>0$ und $\mathcal{F}$ eine kohärente Garbe auf $X$. Nach Definition existiert ein $n_0>0$, so dass für alle $n\geq n_0$ die Garbe $\mathcal{F}\otimes(\mathcal{L}^m)^n$ von globalen Schnitten erzeugt wird. Betrachte die Garben $\mathcal{F}\otimes\mathcal{L}^k$ für $k=1,\ldots,m-1$. Diese sind kohärent, d.h. es gibt für jedes $k$ ein $n_k>0$, so dass für alle $n\geq n_k$ die Garbe $\mathcal{F}\otimes\mathcal{L}^k\otimes(\mathcal{L}^m)^n$ von globalen Schnitten erzeugt wird. Setzen wir nun $N=m\cdot\max\{n_k\mid k=1,\ldots,m-1\}$, so wird $\mathcal{F}\otimes\mathcal{L}^n$ von globalen Schnitten erzeugt für alle $n\geq N$. Somit ist $\mathcal{L}$ ampel.\qed

\paragraph{Satz 7.7.}\label{7.7} Sei $X$ ein Schema von endlichem Typ über einem noetherschen Ring $A$ und $\mathcal{L}$ eine invertierbare Garbe auf $X$. Dann ist $\mathcal{L}$ genau dann ampel, wenn $\mathcal{L}^m$ sehr ampel bzgl. $\operatorname{Spec}(A)$ für ein $m>0$ ist.

\paragraph{Lemma 7.8.}\label{7.8} Sei $X$ ein noethersches Schema, $U\subset_\text{o}X$ und $\mathcal{F}$ eine kohärente Garbe auf $U$. Dann existiert eine kohärente Garbe $\mathcal{F}'$ auf $X$ mit $\mathcal{F}'|_U=\mathcal{F}$.

\paragraph{Beweis.} Sei o.B.d.A. $X$ affin. Sei $j:U\hookrightarrow X$ die Immersion. Dann ist $\mathcal{F}'=j_\ast\mathcal{F}$ nach \hyperref[5.13]{Satz 5.13 (iii)} quasikohärent. Ferner sind quasikohärente Garben auf einem noetherschen Schema Vereinigung seiner kohärenten Untergarben. Daher gilt:
\[j_\ast\mathcal{F} =\mathcal{F}' = \bigcup_\lambda\mathcal{F}'_\lambda \]
mit kohärenten Untergarben $\mathcal{F}'_\lambda$. Nach \hyperref[5.13]{Satz 5.13 (ii)} sind $j^\ast\mathcal{F}'_\lambda$ kohärent. Somit ist $\mathcal{F}=\bigcup_\lambda j^\ast\mathcal{F}'_\lambda$, wobei $j^\ast\mathcal{F}'_\lambda$ die kohärenten Untergarben von $\mathcal{F}$ durchläuft. Da $\mathcal{F}$ kohärent ist, gilt $\mathcal{F}=j^\ast\mathcal{F}'_\lambda$ für ein $\lambda$. Somit ist $\mathcal{F}'=\mathcal{F}'_\lambda$ kohärent.\qed

\paragraph{Beweis von \hyperref[7.7]{Lemma 7.7}.} Sei zunächst $\mathcal{L}^m$ sehr ampel für ein $m>0$. Es existiert also eine Immersion $i:X\to\mathbf{P}_A^n$ mit $\mathcal{L}^m\cong i^\ast(\mathcal{O}(1))$. Sei $\overline{X}$ der Abschluss von $X$ in $\mathbf{P}_A^n$. Dann ist $\overline{X}$ ein projektives Schema über $\operatorname{Spec}(A)$. Nach \hyperref[7.5]{7.5} ist $\mathcal{O}_{\overline{X}}(1)$ ampel auf $\overline{X}$. Sei $\mathcal{F}$ eine kohärente Garbe auf $X$ und $\overline{\mathcal{F}}$ die kohärente Garbe in \hyperref[7.8]{Lemma 7.8} auf $\overline{X}$ mit $\overline{F}|_X=\mathcal{F}$. Wird $\overline{\mathcal{F}}\otimes\mathcal{O}_{\overline{X}}(\ell)$ von globalen Schnitten erzeugt, so ist dies auch für $\mathcal{F}\otimes\mathcal{O}_X(\ell)$ der Fall. Daher ist $\mathcal{L}^m$ ampel auf $X$ und nach \hyperref[7.6]{Satz 7.6} auch $\mathcal{L}$. Die andere Richtung ist schwierig.

\subsection{Projektive Vektorbündel}

\paragraph{Definition.} Sei $X$ ein Schema. Ein quasikohärenter $\mathcal{O}_X$"~Modul $\mathcal{S}$ habe die Eigenschaft $(\dagger)$, wenn $X$ noethersch ist und wenn er eine Struktur einer graduierten $\mathcal{O}_X$"~Algebra trägt, d.h. $\mathcal{S}=\bigoplus_{d\geq 0}\mathcal{S}_d$ mit homogenen Teilen $\mathcal{S}_d$, so dass:
\begin{enumerate}[(i)]
\item $\mathcal{S}_0=\mathcal{O}_X$
\item $\mathcal{S}_1$ ist ein kohärenter $\mathcal{O}_X$"~Modul.
\item $\mathcal{S}$ wird lokal von $\mathcal{S}_1$ als $\mathcal{O}_X$"~Modul erzeugt.
\end{enumerate}

\paragraph{Definition 7.9.}\label{7.9} Sei $X$ ein Schema und $\mathcal{S}$ ein graduierter $\mathcal{O}_X$"~Modul mit $(\dagger)$. Sei $U=\operatorname{Spec}(A)\subset_\text{o}X$, $\mathcal{S}(U)=\Gamma(U,\mathcal{S}|_U)$ die graduierte $A$"~Algebra und $\pi_U:\operatorname{Proj}\mathcal{S}(U)\to U$ der kanonische Morphismus. Für $U,V\subset_\text{o}X$ affin gilt $\pi_U^{-1}(U\cap V)\cong\pi_V^{-1}(U\cap V)$. Verkleben von $\pi_U$ liefert ein Schema $\operatorname{\mathbf{Proj}}(\mathcal{S})$ und ein Morphismus $\pi:\operatorname{\mathbf{Proj}}(\mathcal{S})\to X$, so dass für alle affinen $U\subset_\text{o}X$ stets $\pi^{-1}(U)\cong\operatorname{Proj}\mathcal{S}(U)$ gilt. Die invertierbaren Garben $\mathcal{O}(1)$ auf $\operatorname{Proj}\mathcal{S}(U)$ lassen sich zu einer invertierbaren Garbe $\mathcal{O}(1)$ auf $\operatorname{\mathbf{Proj}}(\mathcal{S})$ verkleben.

\paragraph{Beispiel 7.10.}\label{7.10} Sei $\mathcal{O}_X[T_0,\ldots,T_n]=\mathcal{S}$. Dann ist $\operatorname{\mathbf{Proj}}(\mathcal{S})=\mathbf{P}_X^n=\mathbf{P}_\mathbb{Z}^n\times X$ mit der getwisteten Garbe $\mathcal{O}(1)$ wie in \hyperref[5.30]{Definition 5.30}.

\paragraph{Lemma 7.11.}\label{7.11} Sei $\mathcal{S}$ eine Garbe graduierter Algebren mit $(\dagger)$ auf $X$. Sei $\mathcal{L}$ eine invertierbare Garbe auf $X$ und $\mathcal{S}'$ die folgende Garbe graduierter Algebren:
\[\mathcal{S}_d' =\mathcal{S}_1\otimes\mathcal{L}^d,\quad d\geq 0 \]
Dann erfüllt $\mathcal{S}'$ ebenfalls $(\dagger)$ und es gibt einen natürlichen Isomorphismus:
\[\begin{tikzcd}
P'=\operatorname{\mathbf{Proj}}(\mathcal{S}')\ar[d, "\pi'"']\ar[r, "\varphi", "\cong"'] & P=\operatorname{\mathbf{Proj}}(\mathcal{S})\ar[dl, "\pi"]\\
X&
\end{tikzcd} \]
Es gilt $\mathcal{O}_{P'}(1)\cong \varphi^\ast\mathcal{O}_P(1)\otimes\pi'^\ast\mathcal{L}$. $\mathcal{S}'$ wird auch mit $\mathcal{S}\ast\mathcal{L}$ bezeichnet.

\paragraph{Satz 7.12.}\label{7.12} Sei $X$ und $\mathcal{S}$ mit $(\dagger)$. Sei $P=\operatorname{\mathbf{Proj}}(\mathcal{S})$ mit $\pi:P\to X$. Dann gilt:
\begin{enumerate}[(i)]
\item $\pi$ ist eigentlicher Morphismus.
\item Besitzt $X$ eine ample Garbe $\mathcal{L}$, so ist $\pi$ projektiv und $\mathcal{O}_P(1)\otimes\pi^\ast\mathcal{L}^n$ ist eine sehr ample invertierbare Garbe von $P$ über $X$.
\end{enumerate}

\paragraph{Definition 7.13.}\label{7.13} \begin{enumerate}[(i)]
\item Sei $A$ ein Ring und $M$ ein $A$"~Modul. Setze:
\[T^0(M)=A,\quad T^n(M)= M^{\otimes n},\ n\geq 1 \]
Dann ist $T(M)=\bigoplus_{n\geq 0}T^n(M)$ eine $A$"~Algebra mit $\otimes$ als Multiplikation, die sogenannte \textit{Tensor"~Algebra}\index{Tensor-Algebra} von $M$.
\item Die \textit{symmetrische (Tensor-)Algebra}\index{Tensor-Algebra!symmetrisch} von $M$ ist definiert als der Quotient von $T(M)$ nach dem Ideal, das von den Elementen $x\otimes y-y\otimes x,\ x,y\in M$ erzeugt wird. $S(M)=\bigoplus_{n\geq 0}S^n(M)$ ist dann eine kommutative, graduierte $A$"~Algebra. $S^n(M)$ heißt \textit{$n$-tes symmetrisches Produkt}\index{symmetrisches Produkt} von $M$.
\item Sei $X$ ein Schema und $\mathcal{F}$ ein $\mathcal{O}_X$"~Modul. Dann wird die Tensor-Algebra $T(\mathcal{F})$ bzw. die symmetrische Algebra $S(\mathcal{F})$ von $\mathcal{F}$ als assoziierte Garbe zur Prägarbe $U\mapsto T(\mathcal{F}(U))$ bzw. $U\mapsto S(\mathcal{F}(U))$ definiert. $T(\mathcal{F})$ und $S(\mathcal{F})$ sind $\mathcal{O}_X$"~Algebren.
\end{enumerate}

\paragraph{Beispiel.} Sei $M$ ein freier $A$"~Modul vom Rang $r$. Dann gilt $S(M)\cong A[X_1,\ldots,X_r]$.

\paragraph{Definition 7.14.}\label{7.14} Sei $X$ ein noethersches Schema und $\mathcal{E}$ eine lokal freie, kohärente Garbe auf $X$. Sei $\mathcal{S}=S(\mathcal{E})$ die symmetrische Algebra von $\mathcal{E}$ und $\mathbf{P}(\mathcal{E})=\operatorname{\mathbf{Proj}}(\mathcal{S})$. Dann ist $\mathcal{S}$ eine Garbe von graduierten $\mathcal{O}_X$"~Algebren, die $(\dagger)$ erfüllt. $\pi:\mathbf{P}(\mathcal{E})\to X$ heißt \textit{projektives Vektorbündel}\index{Vektorbündel!projektiv} über $X$.

\paragraph{Bemerkung.} Ist $\mathcal{E}$ frei vom Rang $n+1$ über $U\subset_\text{o}X$, so ist $\pi^{-1}(U)\cong\mathbf{P}_U^n$.

\paragraph{Satz 7.15.}\label{7.15} Sei $\pi:\mathbf{P}(\mathcal{E})\to X$ ein projektives Vektorbündel über $X$. Dann gilt mit $\mathcal{S}=S(\mathcal{E})$:
\begin{enumerate}[(i)]
\item Ist Rang von $\mathcal{E}$ größer gleich $2$, so gibt es einen kanonischen Isomorphismus von graduierten $\mathcal{O}_X$"~Algebren:
\[\mathcal{S}\cong\bigoplus_{\ell\in\mathbb{Z}}\pi_\ast\mathcal{O}(\ell)\]
Insbesondere folgt $\pi_\ast(\mathcal{O}(\ell))=0$ für $\ell<0$ und $\pi_\ast(\mathcal{O})=\mathcal{O}_X$, $\pi_\ast(\mathcal{O}(1))=\mathcal{E}$.
\item Es gibt einen natürliche surjektiven Morphismus:
\[\pi^\ast\mathcal{E}\to\mathcal{O}(1) \]
\end{enumerate}

\paragraph{Satz 7.16.}\label{7.16} \textit{(Universaleigenschaft des projektiven Vektorbündels)} Sei $\pi:\mathbf{P}(\mathcal{E})\to X$ ein projektives Vektorbündel von $X$ und $g:Y\to X$ ein Morphismus. Dann gibt es genau dann einen Morphismus $f$ mit kommutativen Diagramm
\[\begin{tikzcd}
Y\rar[dashed, "f"]\dar["g"'] &\mathbf{P}(\mathcal{E})\ar[dl, "\pi"]\\
X&
\end{tikzcd}\]
wenn es eine invertierbare Garbe $\mathcal{L}$ auf $Y$ zusammen mit einem surjektiven Garbenmorphismus $g^\ast\mathcal{E}\to\mathcal{L}$ gibt.

\subsection{Aufblasung}

\paragraph{Definition 7.17.}\label{7.17} Sei $X$ ein noethersches Schema und $\mathcal{J}$ eine kohärente Idealgarbe auf $X$. Setze $\mathcal{J}^0=\mathcal{O}_X$ und $\mathcal{J}^d$ als das $d$-fache Produkt des Ideals $\mathcal{J}$. Sei $\mathcal{S}=\bigoplus_{d\geq 0}\mathcal{J}^d$. Dann erfüllen $X$ und $\mathcal{S}$ die Bedingung $(\dagger)$. $\widetilde{X}=\mathbf{Proj}(\mathcal{S})$ heißt die \textit{Aufblasung} von $X$ bzgl. $\mathcal{J}$.\index{Aufblasung}

Ist $Y$ das abgeschlossene Unterschema von $X$, das zu $\mathcal{J}$ gehört, so sagen wir auch, dass $\widetilde{X}$ die Aufblasung von $X$ entlang $Y$ ist.

\paragraph{Beispiel 7.18.}\label{7.18} Sei $X=\mathbf{A}^n_k$ und $P\in X$ der Nullpunkt, d.h. $X=\operatorname{Spec}(A),\ A=k[X_1,\ldots,X_n],\ I=(X_1,\ldots,X_n)$. Dann ist die Aufblasung von $X$ bei $P$ gegeben durch:
\[\widetilde{X}=\operatorname{Proj}(S),\quad S=\bigoplus_{d\geq 0}I^d\]
Betrachte die surjektive Abbildung $\psi:A[Y_1,\ldots,Y_n]\to S,\ Y_i\mapsto X_i$. Wir sehen, dass $\widetilde{X}$ isomorph zu dem abgeschlossenen Unterschema von $\mathbf{P}_A^{n-1}$ ist, das durch die homogenen Polynome in den $Y_i$, die $\ker(\psi)$ erzeugen, definiert ist:
\[\ker(\psi)=\langle X_iY_j-X_jY_i\mid i,j=1,\ldots,n\rangle\]

\paragraph{Definition 7.19.}\label{7.19} Sei $f:X\to Y$ ein Morphismus und $\mathcal{J}\subset\mathcal{O}_Y$ eine Idealgarbe auf $Y$. Wir definieren die \textit{inverse Idealgarbe}\index{Idealgarbe!invers} $\mathcal{J}'\subset\mathcal{O}_X$ auf $X$ wie folgt:

$\mathcal{J}\hookrightarrow\mathcal{O}_Y\stackrel{f^\sharp}{\to}f_\ast\mathcal{O}_X$ liefert nach \hyperref[5.1]{5.1 (xi)} ein $\mathcal{O}_X$"~Modulmorphismus $f^\ast\mathcal{J}\to\mathcal{O}_X$. Setze $\mathcal{J}'=\operatorname{im}(f^\ast\mathcal{J}\to\mathcal{O}_X)$.

Alternativ: Betrachte die Idealgarbe $f^{-1}\mathcal{J}$ auf $f^{-1}\mathcal{O}_Y$ und den Morphismus $f^{-1}\mathcal{O}_Y\to \mathcal{O}_Y$, der zu $f^\sharp$ adjungiert ist. Dann ist $\mathcal{J}'$ die vom Bild von $f^{-1}\mathcal{J}$ in $\mathcal{O}_X$ erzeugte Idealgarbe. Wir schreiben daher auch $\mathcal{J}'=f^{-1}\mathcal{J}\cdot\mathcal{O}_X$.

\paragraph{Bemerkung.} Im Allgemeinen sind $f^\ast\mathcal{J}$ und $f^{-1}\mathcal{J}$ verschiedene Idealgarben.

\paragraph{Satz 7.20.}\label{7.20} Sei $X$ ein noethersches Schema und $\mathcal{J}$ eine kohärente Idealgarbe auf $X$. Sei $\pi:\widetilde{X}\to X$ die Aufblasung von $X$ bzgl. $\mathcal{J}$. Dann gilt:
\begin{enumerate}[(i)]
\item Die inverse Idealgarbe $\widetilde{J}=\pi^{-1}\mathcal{J}\cdot\mathcal{O}_{\widetilde{X}}$ ist invertierbar auf $\mathcal{O}_{\widetilde{X}}$.
\item Sei $Y$ das zu $\mathcal{J}$ gehörige abgeschlossene Unterschema und $U=X\setminus Y$. Dann ist $\pi:\pi^{-1}(U)\to U$ ein Isomorphismus.
\end{enumerate}

\paragraph{Beweis.}\begin{enumerate}[(i)]
\item Sei $\widetilde{X}=\mathbf{Proj}(\mathcal{S}),\ \mathcal{S}=\bigoplus_{d\geq 0}\mathcal{J}^d$ und $V\subset_\text{o}X$ affin. Dann ist $\mathcal{O}(1)|_{\operatorname{Proj}\mathcal{S}(V)}$ die assoziierte Garbe zum graduierten $\mathcal{S}(V)$"~Modul $\mathcal{S}(V)(1)=\bigoplus_{d\geq 0}\mathcal{J}^{d+1}(V)=\mathcal{J}(V)\cdot \mathcal{S}(V)$ das von $\mathcal{J}(V)$ in $\mathcal{S}(V)$ erzeugte Ideal. Es folgt $\widetilde{\mathcal{J}}=\pi^{-1}\mathcal{J}\cdot\mathcal{O}_{\widetilde{X}}=\mathcal{O}_{\widetilde{X}}(1)$, also invertierbar.
\item Es ist $\mathcal{J}|_U\cong\mathcal{O}_U$, da $0\to\mathcal{J}\to\mathcal{O}_X\stackrel{i^\sharp}{\to }i_\ast\mathcal{O}_Y\to 0$ exakt ist. Daher gilt $\pi^{-1}(U)=\operatorname{\mathbf{Proj}}\mathcal{O}_U[T]=U$. Beachte, dass für ein allgemeines $A$ stets $\operatorname{Proj}A[T]=D_+(T)=\operatorname{Spec}A[T]_{(T)}=\operatorname{Spec}(A)$ gilt.\qed
\end{enumerate}

\paragraph{Satz 7.21.}\label{7.21} \textit{(Universaleigenschaft des Aufblasens)} Sei $X$ ein noethersches Schema, $\mathcal{J}$ eine kohärente Idealgarbe auf $X$ und $\pi:\widetilde{X}\to X$ die Aufblasung von $X$ bzgl. $\mathcal{J}$. Ist $f:Z\to X$ ein Morphismus, so dass $f^{-1}\mathcal{J}\cdot\mathcal{O}_Z$ eine invertierbare Idealgarbe auf $Z$ ist, so existiert ein eindeutig bestimmter Morphismus $g:Z\to\widetilde{X}$ mit kommutativem Diagramm:
\[\begin{tikzcd}
Z\rar["g", dashed]\ar[rd, "f"'] & \widetilde{X}\dar["\pi"]\\
& X
\end{tikzcd} \]

\paragraph{Lemma 7.22.}\label{7.22} Sei $X$ ein Schema, $f:\mathcal{L}\to\mathcal{M}$ ein surjektiver Morphismus invertierbarer Garben. Dann ist $f$ ein Isomorphismus.

\paragraph{Beweis.} Siehe Bourbaki: Commutative Algebra II, §3.2 Korollar zu Proposition 6.\qed

\paragraph{Korollar 7.23.}\label{7.23} Sei $f:Y\to X$ ein Morphismus noetherscher Schemata und $\mathcal{J}$ eine kohärente Idealgarbe auf $X$. Sei $\widetilde{X}$ die Aufblasung bzgl. $\mathcal{J}$ und $\widetilde{Y}$ die Aufblasung bzgl. $\mathcal{J}_Y=f^{-1}\mathcal{J}\cdot\mathcal{O}_Y$. Dann existiert ein eindeutig bestimmter Morphismus $\widetilde{f}$ mit kommutativem Diagramm:
\[\begin{tikzcd}
\widetilde{Y}\ar[rr, "\widetilde{f}"]\dar["\pi_Y"'] && \widetilde{X}\dar["\pi_X"]\\
Y\ar[rr, "f"'] && X
\end{tikzcd}\]
Ist $f$ eine abgeschlossene Immersion, so auch $\widetilde{f}$.

\paragraph{Beweis.} $\mathcal{J}_Y$ ist kohärent, da nach \hyperref[5.13]{Satz 5.13 (ii)} $f^\ast\mathcal{J}$ kohärent ist und nach \hyperref[5.12]{Satz 5.12 (ii)} auch $\operatorname{im}(f^\ast\mathcal{J}\to\mathcal{O}_Y)$. Nach \hyperref[7.20]{Satz 7.20} ist $\pi_Y^{-1}\mathcal{J}_Y\cdot\mathcal{O}_{\widetilde{Y}}$ eine invertierbare Garbe auf $\widetilde{Y}$. Nach \hyperref[7.21]{Satz 7.21} existiert $\widetilde{f}$ und ist eindeutig bestimmt.

Sei nun $f$ eine abgeschlossene Immersion und:
\[\widetilde{X}=\operatorname{\mathbf{Proj}}(\mathcal{S}),\quad \mathcal{S}=\bigoplus_{d\geq 0}\mathcal{J}^d,\qquad\widetilde{Y}=\mathbf{Proj}(\mathcal{S}_Y),\quad\mathcal{S}_Y=\bigoplus_{d\geq 0}\mathcal{J}_Y^d \]
Ist $Y\hookrightarrow X$ eine abgeschlossene Immersion, so ist $\mathcal{J}\to\mathcal{J}_Y$ surjektiv und somit ist auch $\mathcal{S}\to\mathcal{S}_Y$ als Homomorphismus graduierter Ringe surjektiv. Daher ist $\widetilde{Y}\hookrightarrow\widetilde{X}$ eine abgeschlossene Immersion.\qed

\paragraph{Definition 7.24.}\label{7.24} Ist $Y\hookrightarrow X$ eine abgeschlossene Immersion noetherscher Schemata. Dann heißt das abgeschlossene Unterschema $\widetilde{Y}$ von $\widetilde{X}$ auch \textit{strikte Transformation} von $Y$ bzgl. der Aufblasung $\pi:\widetilde{X}\to X$.\index{strikte Transformation}

\section{Differentiale}

Sei $A$ ein Ring, $B$ eine $A$"~Algebra und $M$ ein $B$"~Modul.

\paragraph{Definition 8.1.}\label{8.1} Eine \textit{$A$"~Derivation}\index{Derivation} von $B$ nach $M$ ist eine Abbildung $\mathrm{d}:B\to M$ mit:
\begin{enumerate}[(i)]
\item $\mathrm{d}$ ist additiv
\item $\mathrm{d}(b\cdot b')=b\cdot \mathrm{d}(b')+ b'\cdot \mathrm{d}(b)$
\item $\mathrm{d}(a)=0$ für alle $a\in A$
\end{enumerate}
Der $B$"~Modul aller $A$"~Derivationen von $B$ nach $M$ bezeichnen wir mit $D_A(B,M)$.

\paragraph{Definition 8.2.}\label{8.2} Der $B$"~Modul $\Omega_{B/A}$ der \textit{relativen Differentialformen}\index{Differentialform!relativ} von $B$ über $A$ zusammen mit einer $A$"~Derivation $\mathrm{d}:B\to\Omega_{B/A}$ ist definiert durch folgende Universaleigenschaft:

Für alle $B$"~Moduln $M$ und $A$"~Derivationen $\mathrm{d}':B\to M$ gibt es genau ein $B$"~Ho\-mo\-mor\-phis\-mus $f:\Omega_{B/A}\to M$, so dass folgendes Diagramm kommutiert:
\[\begin{tikzcd}
B\ar[r, "\mathrm{d}"]\ar[rd, "\mathrm{d}'"'] & \Omega_{B/A}\ar[d, "f", dashed]\\
& M
\end{tikzcd}\]
Die Eindeutigkeit von $\Omega_{B/A}$ ist klar. Für die Existenz betrachte den freien $B$"~Modul $F$ mit Basis $\{d_b\mid b\in B\}$. Dann gilt:
\[\Omega_{B/A}=F / \langle d_{b+b'}-d_b-d_{b'},\ d_{bb'}-bd_{b'}-b'd_b,\ d_a\mid b,b'\in B,\ a\in A\rangle \]
Zusammen mit der $A$"~Derivation $\mathrm{d}:B\to\Omega_{B/A},\ b\mapsto d_b$ erfüllt diese Konstruktion die Universaleigenschaft.

\paragraph{Satz 8.3.}\label{8.3} Sei $B$ eine $A$"~Algebra und $f:B\otimes_A B\to B,\ b\otimes b'\mapsto bb'$. Betrachte $B\otimes_AB$ als $B$"~Modul via $b(b_1\otimes b_2)=(bb_1)\otimes b_2$. $B\otimes_A B$ wird zur $B$"~Algebra durch $(b_1\otimes b_1')(b_2\otimes b_2')=(b_1b_2)\otimes (b_1'b_2')$. Sei $I=\ker(f)$. Dann ist $I/I^2$ ein $B$"~Modul. Setze $\mathrm{d}:B\to I/I^2,\ \mathrm{d}b=1\otimes b-b\otimes 1\mod I^2$. Dann ist $\Omega_{B/A}= (I/I^2,\mathrm{d})$.

\paragraph{Beweis.} Wir zeigen zunächst $I=\sum_{b\in B}B\mathrm{d}b$. Sei $\beta=\sum b_i\otimes c_i\in I$, also $\sum b_ic_i=0$. Dann gilt:
\begin{align*}
\beta &= \sum_i (b_i(1\otimes c_i-c_i\otimes 1) + b_ic_i\otimes 1)\\
&= \sum_i b_i \mathrm{d}c_i + \Big(\sum_i b_ic_i\Big)\otimes 1 \in\sum_{b\in B} B\mathrm{d}b
\end{align*}
Wir zeigen nun, dass $\mathrm{d}$ eine $A$"~Derivation ist. $\mathrm{d}(A)=0$ und die Additivität ist klar.
\begin{align*}
\mathrm{d}b\mathrm{d}c &= (1\otimes b- b\otimes 1)(1\otimes c-c\otimes 1)\\
&= bc\otimes 1 - b\otimes c- c\otimes b+1\otimes bc\\
&=\mathrm{d}(bc) + 2bc\otimes 1- b(\mathrm{d}c+c\otimes 1) - c (\mathrm{d}b + b\otimes 1)\\
&=\mathrm{d}(bc)-b\mathrm{d}c - c \mathrm{d}b
\end{align*}
Also ist $\mathrm{d}(bc)\equiv b\mathrm{d}c+c\mathrm{d}b\mod I^2$. Somit müssen wir noch zeigen, dass $(I/I^2,\mathrm{d})$ die Universaleigenschaft von $\Omega_{B/A}$ erfüllt. Sei $M$ ein $B$"~Modul und $\mathrm{D}:B\to M$ eine $A$"~Derivation. Zu zeigen ist die Existenz und Eindeutigkeit von einem $B$"~Homomorphismus $f:I/I^2\to M$ mit $\mathrm{D}=f\circ \mathrm{d}$. Da $I=\sum B\mathrm{d}b$ ist, folgt die Eindeutigkeit.

Betrachte die \textit{triviale Erweiterung} $B\ast M$ von $B$ mit $M$, d.h. $B\oplus M$ mit der Multiplikation $(b_1,m_1)(b_2,m_2)=(b_1b_2,b_1m_2+b_2m_1)$. Dann ist $B\ast M$ ein Ring mit Einselement $(1,0)$. $B\ast M$ trägt eine $B$"~Modulstruktur via $b'(b,m)=(bb',bm)$, d.h. $B\ast M$ ist eine $B$"~Algebra. Wir haben kanonische $B$"~Homomorphismen:
\begin{itemize}
\item Die Einbettung $M\hookrightarrow B\ast M,\ m\mapsto (0,m)$
\item Die Projektion $B\ast M\to B,\ (b,m)\mapsto b$
\end{itemize}
Betrachte den $B$"~Algebrenhomomorphismus $\varphi:B\otimes_AB\to B\ast M,\ x\otimes y\mapsto (xy,x\mathrm{D}y)$. Es gilt $\varphi(I)\subset M\subset B\ast M$, da $\varphi(\mathrm{d}b)=\varphi(1\otimes b- b\otimes 1)=(b,\mathrm{D}b)-(b,b\mathrm{D}1)=(0,\mathrm{D}b)$. Da $M^2=0$ in $B\ast M$, induziert $\varphi$ eine Abbildung $\phi: B\otimes_A B/I^2\to B\ast M$. Dann erfüllt $f=\phi|_{I/I^2}:I/I^2\to M$ die gewünschte Eigenschaft $f\circ\mathrm{d}=\mathrm{D}$.\qed

\paragraph{Satz 8.4.}\label{8.4} Es gibt für jeden $B$"~Modul $M$ einen kanonischen $B$"~Modulisomorphismus:
\[\operatorname{Hom}_B(\Omega_{B/A},M)\to\operatorname{Der}_A(B,M),\ f\mapsto f\circ\mathrm{d} \]

\paragraph{Beweis.} Folgt direkt aus der Universaleigenschaft.\qed

\paragraph{Beispiel 8.5.}\label{8.5} Sei $B=A[X_1,\ldots,X_n]$. Dann ist $\Omega_{B/A}=\bigoplus_{i=1}^n B\mathrm{d}X_i$ ein freier $B$"~Modul vom Rang $n$. Allgemein gilt: Wird $B$ als $A$"~Algebra von $y_1,\ldots,y_m$ erzeugt, so ist $\Omega_{B/A}=\sum_{i=1}^m B\mathrm{d}y_1$, da:
\[\mathrm{d}\Big(\prod_i y_i^{n_i}\Big) = \sum_i n_i\prod_{j\neq i} y_j^{n_j}y_i^{n_i - 1}\mathrm{d}y_i \]
Angenommen, $\sum_i b_i\mathrm{d}X_i=0$. Betrachte die Derivationen $\frac{\partial}{\partial X_i}\in\operatorname{Der}_A(B)$. Nach \hyperref[8.4]{Satz 8.4} existiert für jedes $k$ ein $f_k\in\operatorname{Hom}_B(\Omega_{B/A},B)$ mit $\frac{\partial}{\partial X_k}=f_k\circ\mathrm{d}$. Es gilt für alle $k$:
\[ 0= f_k\Big(\sum_{i=1}^n b_i\mathrm{d}X_i\Big)= \sum_{i=1}^n b_i f_k(\mathrm{d}X_i) = \sum_{i=1}^n b_i \frac{\partial X_i}{\partial X_k} = \sum_{i=1}^n b_i\delta_{ik}=b_k \]
Somit ist $\Omega_{B/A}=\bigoplus_{i=1}^nB\mathrm{d}X_i$

\paragraph{Satz 8.6.}\label{8.6} \begin{enumerate}[(i)]
\item Sei $S$ eine multiplikative Teilmenge der $A$"~Algebra $B$. Dann gilt:
\[S^{-1}\Omega_{B/A}\cong\Omega_{S^{-1}B/A} \]
\item Seien $A',B$ $A$"~Algebren und $B'=B\otimes_AA'$. Dann gilt:
\[\Omega_{B'/A'}\cong\Omega_{B/A}\otimes_BB' \]
\end{enumerate}

\paragraph{Satz 8.7.}\label{8.7} Seien $\phi:A\to B,\ \psi:B\to C$ Ringhomomorphismen. Dann gilt:
\begin{enumerate}[(i)]
\item \textit{(Erste fundamentale exakte Sequenz)}\index{Fundamentale exakte Sequenz} Es gibt eine exakte Sequenz:
\[\begin{tikzcd}
\Omega_{B/A}\otimes_BC \ar[r, "v"] & \Omega_{C/A} \ar[r, "u"] & \Omega_{C/B}\rar & 0
\end{tikzcd} \]
\item $v$ ist injektiv und die Sequenz zerfällt, genau dann wenn für alle $C$"~Moduln $M$ und jedes $\mathrm{D}\in \operatorname{Der}_A(B,M)$ zu einer Derivation aus $\operatorname{Der}_A(C,M)$ fortgesetzt werden kann.
\end{enumerate}

\paragraph{Beweis.} Definiere $u$ und $v$ durch:
\[v(\mathrm{d}b\otimes c)= c\mathrm{d}\psi(b),\quad u(c\mathrm{d}c') = c\mathrm{d}c',\qquad b\in B,\ c,c'\in C \]
Äquivalent zu (i) bzw. (ii) ist die Exaktheit der $4$-Term- bzw. $5$-Termsequenzen:
\[0\longrightarrow \operatorname{Hom}_C(\Omega_{C/B},M)\longrightarrow \operatorname{Hom}_C(\Omega_{C/A},M)\longrightarrow \operatorname{Hom}_C(\Omega_{B/A}\otimes_B C,M)\dashrightarrow 0  \]
Da $\operatorname{Hom}_C(\Omega_{B/A}\otimes_B C,M)\cong\operatorname{Hom}_B(\Omega_{B/A},M)$, ist nach \hyperref[8.4]{Satz 8.4} ist die Exaktheit nachzuweisen von:
\[0\longrightarrow \operatorname{Der}_B(C,M)\longrightarrow \operatorname{Der}_A(C,M)\longrightarrow \operatorname{Der}_A(B,M)\dashrightarrow 0 \]
Die vorletzte Abbildung ist gegeben durch $(\delta:C\to M)\mapsto (\delta\circ\psi: B\to M)$. Ohne die $0$ rechts ist diese Folge per Definition exakt. Mit der $0$ rechts ist die Folge genau dann exakt, wenn die Forderung in (ii) gilt.\qed

\paragraph{Beispiel 8.8.}\label{8.8} Sei $B$ eine $A$"~Algebra und $C=B[x_1,\ldots,x_n]$. Sei $M$ ein beliebiger $C$"~Mo\-dul und $D\in\operatorname{Der}_A(B,M)$. Sei $\widetilde{D}:C\to M$ die einem Polynom $F$, den Polynom $F^D$ zuordnet, wobei $F^D$ das Polynom ist, das man durch Anwenden von $D$ auf die Koeffizienten von $F$ erhält. Dann ist $\widetilde{D}$ eine Derivation mit $\widetilde{D}|_B=D$. Nach \hyperref[8.7]{Satz 8.7 (ii)} gilt:
\[\Omega_{C/A}\cong (\Omega_{B/A}\otimes_BC)\oplus C\mathrm{d}x_1\oplus\cdots\oplus C\mathrm{d}x_n \]

\paragraph{Satz 8.9.}\label{8.9}\textit{(Zweite fundamentale exakte Sequenz)}\index{Fundamentale exakte Sequenz} Sei $B$ eine $A$"~Algebra, $I$ ein Ideal in $B$ und $C=B/I$. Es gibt eine kanonische, exakte Folge von $C$"~Moduln:
\[\begin{tikzcd}
I/I^2\rar["\delta"] & \Omega_{B/A}\otimes_BC\rar["v"] & \Omega_{C/A}\rar & 0
\end{tikzcd} \]
wobei $v(\mathrm{d}b\otimes c)= c\mathrm{d}\overline{b}$ und $\delta$ die durch die Abbildung $I\to\Omega_{B/A}\otimes_BC,\ x\mapsto \mathrm{d}x\otimes 1$ induzierte $C$"~lineare Abbildung ist.

\paragraph{Beweis.} Für $x,x'\in I$ gilt:
\[\mathrm{d}(xx')\otimes 1 = \mathrm{d}x\otimes x' + \mathrm{d}x'\otimes x =0 \]
Daher ist $\delta$ wohldefiniert. Nun ist für alle $C$"~Moduln $M$ die folgenden Folgen exakt:
\[\begin{tikzcd}
0 \rar & \operatorname{Hom}_C(\Omega_{C/A},M)\dar["\cong"] \rar & \operatorname{Hom}_C(\Omega_{B/A}\otimes_BC,M)\dar["\cong"] \rar & \operatorname{Hom}_C(I/I^2,M)\dar["\cong"]\\
0 \rar & \operatorname{Der}_A(B/I,M) \rar & \operatorname{Der}_A(B,M) \rar["\varphi"'] & \operatorname{Hom}_B(I,M)
\end{tikzcd} \]
wobei $\varphi(D:B\to M) = (D|_I:I\to M)$.\qed

\paragraph{Korollar 8.10.}\label{8.10} Ist $B$ eine endlich erzeugte $A$"~Algebra oder Lokalisierung einer endlich erzeugten $A$"~Algebra, so ist $\Omega_{B/A}$ endlich erzeugter $B$"~Modul.

\paragraph{Beweis.} Es ist $B \cong A[x_1,\ldots,x_n]/I$ für ein Ideal $I$. Wegen \hyperref[8.5]{Beispiel 8.5} ist $\Omega_{A[x_1,\ldots,x_n]/A}$ ein endlich erzeugter $A[x_1,\ldots,x_n]$"~Modul. Nach \hyperref[8.9]{Satz 8.9} ist $\Omega_{B/A}$ ein endlich erzeugter $B$"~Modul. Die Aussage über die Lokalisierung folgt aus \hyperref[8.6]{Satz 8.6 (i)}.\qed

\paragraph{Definition 8.11.}\label{8.11} Eine Körpererweiterung $K/k$ heißt \textit{separabel erzeugt}\index{separabel erzeugt}, falls es eine Transzendenzbasis $\{x_\lambda\}_{\lambda\in\Lambda}$ für $K/k$ gibt, so dass $K/k[x_\lambda\mid \lambda\in\Lambda]$ separabel ist.

\paragraph{Satz 8.12.}\label{8.12} Sei $K/k$ eine endlich erzeugte Körpererweiterung. Dann gilt:
\begin{enumerate}[(i)]
\item $\dim_K\Omega_{K/k}\geq \operatorname{trdeg}(K/k)$
\item Gleichheit in (i) gilt genau dann, wenn $K/k$ separabel erzeugt ist.
\end{enumerate}

\paragraph{Korollar 8.13.}\label{8.13} Sei $K/k$ eine endliche Erweiterung. Dann gilt:
\[\Omega_{K/k}=0\iff K/k\text{ ist separabel} \]

\paragraph{Satz 8.14.}\label{8.14} Sei $B$ ein lokaler Ring mit Maximalideal $\mathfrak{m}$ und Restklassenkörper $k$. Ferner gebe es einen Schnitt von $B\twoheadrightarrow k$, also ist $B$ eine $k$"~Algebra. Dann ist die Abbildung $\delta$ in der \hyperref[8.9]{zweiten fundamentalen Sequenz} ein Isomorphismus:
\[\delta:\mathfrak{m}/\mathfrak{m}^2\to \Omega_{B/k}\otimes_B k \]

\paragraph{Beweis.} Es ist $\operatorname{coker}(\delta)=\Omega_{k/k}=0$. Die Injektivität von $\delta$ ist äquivalent zur Surjektivität von:
\[\delta^\ast :\operatorname{Hom}_k(\Omega_{B/k}\otimes_B k,k)\to\operatorname{Hom}_k(\mathfrak{m}/\mathfrak{m}^2,k) \]
Nun ist $\operatorname{Hom}_k(\Omega_{B/k}\otimes_B k,k) = \operatorname{Hom}_B(\Omega_{B/k},k)=\operatorname{Der}_k(B,k)$. Die Abbildung $\operatorname{Der}_k(B,k)\to\operatorname{Hom}_k(\mathfrak{m}/\mathfrak{m}^2,k),\ (d:B\to k)\mapsto (d|_\mathfrak{m}:\mathfrak{m}\to k)$ kommutiert mit $\delta^\ast$. Beachte, dass für $d\in\operatorname{Der}_k(B,k)$ stets $d(\mathfrak{m}^2)=0\in k$ gilt.

Sei nun $h\in\operatorname{Hom}_k(\mathfrak{m}/\mathfrak{m}^2,k)$ und $b\in B$. Wegen des Schnitts $k\to B$, hat $b$ eine eindeutige Darstellung $b=\lambda+c,\ \lambda\in k,\ c\in\mathfrak{m}$. Setze $d:B\to k,\ d(b) =h(\overline{c})$. Dann ist $d(k)=0$, $d$ additiv und es gilt:
\begin{align*}
d(bb') &= d(\lambda\lambda'+\lambda c'+\lambda' c+cc')\\
&= h(\overline{\lambda c'+\lambda' c})\\
&= \lambda h(\overline{c'}) + \lambda' h(\overline{c})\\
&= \lambda d(b')+\lambda' d(b)= bd(b')+b'd(b)\in k
\end{align*}
für alle $b=\lambda+c,\ b'=\lambda'+c'\in B$. Es gilt $\delta^\ast (d)=h$.\qed

\paragraph{Satz 8.15.}\label{8.15} Sei $B$ ein lokaler Ring und es gebe einen Schnitt von $B\to k$. Ferner sei der Restklassenkörper $k$ vollkommen. Sei $B$ die Lokalisierung einer endlich erzeugten $k$"~Algebra. Dann ist $\Omega_{B/k}$ genau dann ein freier $B$"~Modul vom Rang $\dim(B)$, wenn $B$ regulär ist.

\paragraph{Beweis.} Sei $\Omega_{B/k}$ frei vom Rang $\dim(B)$. Nach \hyperref[8.14]{Satz 8.14} ist $\dim_k\mathfrak{m}/\mathfrak{m}^2=\dim(B)$, also ist $B$ per Definition regulär.

Sei umgekehrt $B$ regulär von Dimension $r$. Nach \hyperref[8.14]{Satz 8.14} gilt:
\[r =\dim_k\mathfrak{m}/\mathfrak{m}^2=\dim_k\Omega_{B/k}\otimes_B k \]
Sei $K$ der Quotientenkörper von $B$. Nach \hyperref[8.6]{Satz 8.6} ist $\Omega_{B/k}\otimes_B K=\Omega_{K/k}$. Nun ist $K/k$ separabel erzeugt, da $k$ vollkommen (siehe z.B. Matsumura Chapter 10, Corollary p. 194). Nach \hyperref[8.12]{Satz 8.12} gilt:
\[\dim_K\Omega_{K/k}=\operatorname{trdeg}(K/k)=\dim(B) = r \]
Nach \hyperref[8.10]{Korollar 8.10} ist $\Omega_{B/k}$ endlich erzeugter $B$"~Modul.\qed

\paragraph{Lemma 8.16.}\label{8.16} Sei $A$ ein noetherscher, lokaler, nullteilerfreier Ring mit Restklassenkörper $k$ und Quotientenkörper $K$. Sei $M$ ein endlich erzeugter $A$"~Modul mit:
\[\dim_k(M\otimes_Ak)=\dim_K(M\otimes_AK)=r \]
Dann ist $M$ frei vom Rang $r$.

\paragraph{Beweis.} Wegen $\dim_k(M\otimes_Ak)=r$, gibt es nach Nakayama eine Surjektion $\varphi:A^r\to M$. Setze $N=\ker(\varphi)$. Da $K$ flach über $A$ ist, haben wir eine kurze exakte Folge:
\[0\longrightarrow N\otimes_AK\longrightarrow K^r\longrightarrow M\otimes_AK\longrightarrow 0 \]
Da $\dim_K(M\otimes_AK)=\dim_KK^r$, folgt $N\otimes_AK=0$, also $N=0$, da $N\subset A^r$ und $A$ torsionsfrei ist.\qed

\subsection{Differentialgarben}

Sei $f:X\to Y$ ein Morphismus von Schemata und $\Delta:X\to X\times_YX$ die Diagonale. $\Delta$ ist nach \hyperref[3.30]{Satz 3.30} eine Immersion und liefert ein Isomorphismus auf $\Delta(X)$, wobei $\Delta(X)$ ein abgeschlossenes Unterschema eines offenen Unterschema $W\subset X\times_YX$ ist.

\paragraph{Definition 8.17.}\label{8.17} Sei $\mathcal{J}$ die Idealgarbe von $\Delta(X)$ in $W$. Dann heißt
\[\Omega_{X/Y}=\Delta^\ast(\mathcal{J}/\mathcal{J}^2) \]
die Garbe der \textit{relativen Differentiale}\index{Differentialform!relativ} von $X$ über $Y$.

\paragraph{Bemerkung 8.18.}\label{8.18}\begin{enumerate}[(i)]
\item $\mathcal{J}/\mathcal{J}^2$ ist ein $\mathcal{O}_{\Delta(X)}$"~Modul und somit ist über den Isomorphismus $X\to\Delta(X)$ die Modulgarbe $\Omega_{X/Y}$ ein $\mathcal{O}_X$"~Modul. Wegen \hyperref[5.13]{Satz 5.13 (i)} ist $\Omega_{X/Y}$ quasikohärent. Ist $Y$ noethersch und $f:X\to Y$ von endlichem Typ, und somit $X$ und $X\times_YX$ noethersch, dann ist $\Omega_{X/Y}$ kohärent.
\item Sei $U=\operatorname{Spec}(A)\subset_\text{o}Y$ und $V=\operatorname{Spec}(B)\subset_\text{o}X$ mit $f(V)\subset U$. Dann ist $V\times_UV$ affin offen in $X\times_YX$ mit $V\times_UV= \operatorname{Spec}(B\otimes_AB)$. Es ist
\[\Delta(X)\cap V\times_UV = \operatorname{Spec}(B\otimes_AB/\ker(B\otimes_AB\to B)) \]
ein abgeschlossenes Unterschema von $V\times_UV$. Sei $I=\ker(B\otimes_AB\to B)$, also gilt:
\[\mathcal{J}/\mathcal{J}^2|_{\Delta(X)\cap V\times_UV} = \widetilde{I/I^2} \]
Nach \hyperref[8.3]{Satz 8.3} folgt $\Omega_{V/U}=\widetilde{\Omega_{B/A}}$. Somit ist \hyperref[8.17]{Definition 8.17} verträglich mit dem affinen Fall. Ist $X=\bigcup V$ eine Überdeckung durch affin offenen $V$ und $Y=\bigcup U$ eine Überdeckung durch affin offenen $U$, so liefert die Verklebung von $\widetilde{\Omega_{B/A}},\ \mathrm{d}:B\to\Omega_{B/A}$ gerade $\Omega_{X/Y},\ \mathrm{d}:\mathcal{O}_X\to\Omega_{X/Y}$. $\mathrm{d}$ ist ein Garbenmorphismus und ist in jedem Punkt eine Derivation von lokalen Ringen.
\end{enumerate}

\paragraph{Satz 8.19.}\label{8.19} Sei $f:X\to Y$ ein Morphismus und $g:Y'\to Y$ ein weiterer Morphismus. Betrachte den Basiswechsel:
\[\begin{tikzcd}
X\times_YY' \rar["f'"]\dar["g'"'] & Y'\dar["g"]\\
X\rar["f"] & Y
\end{tikzcd} \]
Dann gilt $\Omega_{X'/Y'}\cong (g')^\ast(\Omega_{X/Y})$.

\paragraph{Beweis.} Folgt aus \hyperref[8.18]{Bemerkung 8.18 (ii)} und \hyperref[8.6]{Satz 8.6 (ii)}.\qed

\paragraph{Satz 8.20.}\label{8.20} Seien $f:X\to Y$ und $g:Y\to Z$ Morphismen. Dann gibt es eine exakte Garbensequenz auf $X$:
\[f^\ast\Omega_{Y/Z}\longrightarrow\Omega_{X/Z}\longrightarrow\Omega_{X/Y}\longrightarrow 0 \]

\paragraph{Beweis.} Es reicht, die Exaktheit lokal zu überprüfen. Dies folgt aus der \hyperref[8.7]{ersten fundamentalen exakten Sequenz}.\qed

\paragraph{Satz 8.21.}\label{8.21} Sei $f:X\to Y$ ein Morphismus und $Z$ ein abgeschlossenes Unterschema von $X$ zum Ideal $\mathcal{J}$. Dann gibt es eine exakte Sequenz von Garben auf $Z$:
\[\mathcal{J}/\mathcal{J}^2 \stackrel{\delta}{\longrightarrow} \Omega_{X/Y}\otimes_{\mathcal{O}_X}\mathcal{O}_Z\longrightarrow\Omega_{Z/Y}\longrightarrow 0 \]

\paragraph{Beweis.} Folgt aus der \hyperref[8.9]{zweiten fundamentalen exakten Sequenz}.\qed

\paragraph{Satz 8.22.}\label{8.22} Sei $X=\mathbf{A}_Y^n$. Dann ist $\Omega_{X/Y}$ ein freier $\mathcal{O}_X$"~Modul vom Rang $n$, erzeugt von globalen Schnitten $\mathrm{d}x_1,\ldots,\mathrm{d}x_n$, wobei $x_1,\ldots,x_n$ die affinen Koordinaten von $\mathbf{A}_Y^n$ sind.

\paragraph{Beweis.} Folgt aus \hyperref[8.5]{Beispiel 8.5}.\qed

\paragraph{Satz 8.23.}\label{8.23} Sei $A$ ein Ring und $Y=\operatorname{Spec}(A)$. Sei $X=\mathbf{P}_A^n$. Dann gibt es eine exakte Sequenz von Garben auf $X$:
\[0\longrightarrow\Omega_{X/Y}\longrightarrow \bigoplus_{i=0}^n\mathcal{O}_X(-1)\longrightarrow \mathcal{O}_X\longrightarrow 0 \]

\paragraph{Beweis.} Sei $S=A[x_0,\ldots,x_n]$ der homogene Koordinatenring von $X$ und setze:
\[E=\bigoplus_{i=0}^n S(-1)e_i\]
als den graduierten $S$"~Modul mit Basis $e_0,\ldots,e_n$ im Grad $1$. Betrachte den $S$"~Mo\-dul\-ho\-mo\-mor\-phis\-mus $\varphi:E\to S,\ e_i\mapsto x_i$ im Grad $0$ und $M=\ker(\varphi)$. Wir erhalten eine exakte Garbensequenz:
\[0\longrightarrow\widetilde{M}\longrightarrow\bigoplus_{i=0}^n\mathcal{O}_X(-1)\stackrel{\tilde{\varphi}}{\longrightarrow}\mathcal{O}_X\longrightarrow 0 \]
$\varphi:E\to S$ ist in allen Graden größer gleich $1$ surjektiv, also ist $\tilde{\varphi}$ surjektiv. Wir zeigen nun $\widetilde{M}=\Omega_{X/Y}$.

Lokalisierung nach $x_i$ liefert eine Surjektion von freien $S_{x_i}$"~Moduln $E_{x_i}\to S_{x_i}$. $M_{x_i}$ ist frei vom Rang $n$ mit Basis $\{e_j-\frac{x_j}{x_i}e_i\mid j\neq i\}$. Multiplikation mit $\frac{1}{x_i}$ liefert Elemente vom Grad $0$ in $M_{x_i}$. Also ist $\widetilde{M}|_{D_+(x_i)}$ ein freier $\mathcal{O}_{D_+(x_i)}$"~Modul, der von den globalen Schnitten $\{\frac{1}{x_i}e_j-\frac{x_j}{x_i^2}e_i\mid j\neq i \}$ erzeugt wird. Definiere $\varphi_i:\Omega_{X/Y}|_{D_+(x_i)}\to\widetilde{M}|_{D_+(x_i)}$ wie folgt:

Es ist $D_+(x_i)=\operatorname{Spec}A[\frac{x_0}{x_i},\ldots,\widehat{\frac{x_i}{x_i}},\ldots,\frac{x_n}{x_i}]$. Nach \hyperref[8.22]{Satz 8.22} gilt:
\[\Omega_{X/Y}|_{D_+(x_i)}=\bigoplus_{j\neq i}\mathcal{O}_{D_+(x_i)}\mathrm{d}\frac{x_j}{x_i} \]
Setze nun $\varphi_i(\mathrm{d}\frac{x_j}{x_i})=\frac{1}{x_i^2}(x_ie_j-x_je_i)$. $\varphi_i$ ist ein Isomorphismus. Die $\varphi_i$ lassen sich nun zu einem Isomorphismus $\varphi:\Omega_{X/Y}\to\widetilde{M}$ verkleben.\qed

\subsection{Nicht-singuläre Varietäten}

\paragraph{Definition 8.24.}\label{8.24} Eine Varietät $X$ über einem algebraisch abgeschlossenen Körper $k$, d.h. ein integres, separiertes Schema von endlichem Typ, heißt \textit{nicht-singulär}\index{nicht-singulär}, wenn alle lokalen Ringe regulär sind.

\paragraph{Satz 8.25.}\label{8.25} Sei $X$ ein irreduzibles, separiertes Schema von endlichem Typ über einem algebraisch abgeschlossenen Körper $k$. Dann ist $\Omega_{X/k}$ genau dann lokal frei vom Rang $n=\dim(X)$, wenn $X$ eine nicht-singuläre Varietät über $k$ ist.

\iffalse
\paragraph{Beweis.} Sei $x\in X$ ein abgeschlossener Punkt. Dann ist $\dim(\mathcal{O}_{X,x})=n$ und $\kappa(x)=k$. Es ist $B=\mathcal{O}_{X,x}$ eine endlich erzeugte $k$"~Algebra und $\Omega_{B/k}=(\Omega_{X/k})_x$. Nach \hyperref[8.15]{Satz 8.15} ist $(\Omega_{X/k})_x$ genau dann frei vom Rang $n$, wenn $\mathcal{O}_{X,x}$ regulär ist. Somit haben wir mit \hyperref[8.26]{Lemma 8.26} folgende Äquivalenzen:
\begin{align*}
& \text{Alle lokalen Ringe sind regulär}\\
\iff & \text{Alle lokalen Ringe von abgeschlossenen Punkten sind regulär}\\
\iff & (\Omega_{X/k})_x \text{ ist frei vom Rang $n$ für alle abgeschlossene $x\in X$}\\
\iff & \Omega_{X/k}\text{ ist lokal frei vom Rang $n$}\qedhere
\end{align*}
\fi

\paragraph{Lemma 8.26.}\label{8.26} Sei $X$ ein noethersches Schema und $\mathcal{F}$ eine kohärente Garbe. Dann ist $\mathcal{F}$ genau dann lokal frei, wenn $\mathcal{F}_x$ freier $\mathcal{O}_{X,x}$"~Modul für alle $x\in X$ ist.

\paragraph{Korollar 8.27.}\label{8.27} Sei $X$ eine nicht-singuläre Varietät über $k$ und $Y\subset X$ ein irreduzibles, abgeschlossenes Unterschema zur Idealgarbe $\mathcal{J}$. Dann ist $Y$ genau dann nicht-singulär, wenn folgende Bedingungen erfüllt sind:
\begin{enumerate}[(i)]
\item $\Omega_{Y/k}$ ist lokal frei.
\item In der exakten Sequenz aus \hyperref[8.21]{Satz 8.21} ist $\delta$ injektiv:
\[0\longrightarrow\mathcal{J}/\mathcal{J}^2\stackrel{\delta}{\longrightarrow}\Omega_{X/k}\otimes_{\mathcal{O}_X}\mathcal{O}_Y\longrightarrow\Omega_{Y/k}\longrightarrow 0 \]
\end{enumerate}

### 2015-05-29 Algebraische Geometrie/chapter3.tex
Neue Dateistruktur

\chapter{Kohomologie}

Wir verwenden folgende Bezeichnungen für abelsche Kategorien:
\begin{itemize}
\item $\mathbf{Ab}$, die Kategorie der abelschen Gruppen
\item $\mathbf{Mod}(A)$, die Kategorie der Moduln über einem kommutativen Ring $A$ mit Eins
\item $\mathbf{Ab}(X)$, die Kategorie der Garben abelscher Gruppen über einem topologischen Raum $X$
\item $\mathbf{Mod}(X)$, die Kategorie der $\mathcal{O}_X$"~Moduln auf einem geringten Raum $(X,\mathcal{O}_X)$
\item $\mathbf{Qcoh}(X)$, die Kategorie der quasikohärenten $\mathcal{O}_X$"~Moduln auf einem Schema $X$
\item $\mathbf{Coh}(X)$, die Kategorie der kohärenten $\mathcal{O}_X$"~Modulgarben auf einem noetherschen Schema $X$
\end{itemize}

\section{Kohomologie von Garben}

\paragraph{Satz 9.15.}\label{9.15} Sei $A$ ein Ring. Dann ist jeder $A$"~Modul isomorph zu einem Untermodul eines injektiven $A$"~Moduls.

\paragraph{Satz 9.16.}\label{9.16} Sei $(X,\mathcal{O}_X)$ ein geringter Raum. Dann hat die Kategorie $\mathbf{Mod}(\mathcal{O}_X)$ genügend viele Injektive.

\paragraph{Beweis.} Sei $\mathcal{F}\in\mathbf{Mod}(\mathcal{O}_X)$. Dann ist $\mathcal{F}_x\in\mathbf{Mod}(\mathcal{O}_{X,x})$. Nach \hyperref[9.15]{Satz 9.15} gibt es ein Monomorphismus $\mathcal{F}_x\to I_x$ mit einem injektiven $\mathcal{O}_{X,x}$"~Modul $I_x$. Sei $j:\{x\}\hookrightarrow X$ die kanonische Inklusion und betrachte die $I_x$ als Garben auf $\{x\}$. Setze $\mathcal{J}=\prod_{x\in X}j_\ast I_x$. Dann gilt:
\[\operatorname{Hom}_{\mathcal{O}_X}(\mathcal{F},\mathcal{J}) = \prod_{x\in X}\operatorname{Hom}_{\mathcal{O}_X}(\mathcal{F},j_\ast I_x) = \prod_{x\in X}\operatorname{Hom}_{\mathcal{O}_{X,x}}(\mathcal{F}_x,I_x) \]
Die $\mathcal{F}_x\hookrightarrow I_x$ geben ein Monomorphismus $\mathcal{F}\hookrightarrow \mathcal{J}$. Ferner gilt:
\[\operatorname{Hom}_{\mathcal{O}_X}(-,\mathcal{J})=\prod_{x\in X}\operatorname{Hom}_{\mathcal{O}_{X,x}}((-)_x, I_x) \]
Da Halmbildung exakt und $I_x$ injektive Objekte sind, ist dies ein exakter Funktor. Somit ist $\mathcal{J}$ ein injektiver $\mathcal{O}_X$"~Modul.\qed

\paragraph{Korollar 9.17.}\label{9.17} Sei $X$ ein topologischer Raum. Dann besitzt $\mathbf{Ab}(X)$ genügend viele Injektive.

\paragraph{Beweis.} Sei $\mathcal{O}_X$ die konstante Garbe $\mathbb{Z}$ auf $X$. Dann ist $(X,\mathcal{O}_X)$ ein geringter Raum und es gilt $\mathbf{Mod}(\mathcal{O}_X)=\mathbf{Ab}(X)$.\qed

\paragraph{Definition 9.18.}\label{9.18} Sei $X$ ein topologischer Raum und $\Gamma(X,-):\mathbf{Ab}(X)\to \mathbf{Ab}$ der globale Schnittfunktor. Wir definieren:
\[\mathrm{H}^i(X,-)= \mathrm{R}^i\Gamma(X,-) \]
$\mathrm{H}^i(X,\mathcal{F})$ heißt \textit{$i$-te Kohomologiegruppe}\index{Kohomologiegruppe} von der Garbe $\mathcal{F}$. Haben $X$ und $\mathcal{F}$ Zusatzstrukturen, z.B. $X$ Schema oder $\mathcal{F}\in \mathbf{Qcoh}(X)$, so wird die Kohomologie trotzdem immer im obigen Sinne verstanden, d.h. $\mathcal{F}\in\mathbf{Ab}(X)$.

Sei $(X,\mathcal{O}_X)$ ein geringter Raum und $\mathcal{F}$ ein $\mathcal{O}_X$"~Modul. $\mathcal{F}$ heißt\textit{welk}\index{$\mathcal{O}_X$-Modul!welk}, falls für alle offenen $V\subset U$ die Restriktionsabbildung $\mathcal{F}(U)\to\mathcal{F}(V)$ surjektiv ist.

\paragraph{Lemma 9.19.}\label{9.19} Ist $(X,\mathcal{O}_X)$ ein geringter Raum, so ist jeder injektiver $\mathcal{O}_X$"~Modul $\mathcal{J}$ welk.

\paragraph{Beweis.} Seien $i:U\hookrightarrow X$ und $j:V\hookrightarrow X$ die natürlichen Inklusionen. Betrachte die Inklusion $j_!\mathcal{O}_X|_V\hookrightarrow i_!\mathcal{O}_X|_U$ von $\mathcal{O}_X$"~Moduln. Es folgt die Surjektivität von:
\[\mathcal{J}(U) = \operatorname{Hom}(i_!\mathcal{O}_X|_U,\mathcal{J})\to \operatorname{Hom}(j_!\mathcal{O}_X|_V,\mathcal{J})=\mathcal{J}(V)\qedhere \]

\paragraph{Satz 9.20.}\label{9.20} Sei $X$ ein topologischer Raum und $\mathcal{F}$ eine welke Garbe auf $X$. Dann ist $\mathrm{H}^i(X,\mathcal{F})=0$ für alle $i>0$.

\paragraph{Lemma 9.21.}\label{9.21} Sei $X$ ein topologischer Raum. Sei $0\to\mathcal{G}'\to\mathcal{G}\to\mathcal{G}''\to 0$ eine exakte Sequenz von Garben.
\begin{enumerate}[(i)]
\item Ist $\mathcal{G}'$ eine welke Garbe, so ist für alle $U\subset_\text{o}X$ die folgende Sequenz exakt:
\[0\longrightarrow \mathcal{G}'(U)\longrightarrow \mathcal{G}(U)\longrightarrow \mathcal{G}''(U)\longrightarrow 0 \]
\item Sind $\mathcal{G'}$ und $\mathcal{G}$ welk, so auch $\mathcal{G}''$.
\end{enumerate}

\paragraph{Beweis.}
\begin{enumerate}[(i)]
\item Sei o.B.d.A. $U=X$ und sei $s''\in\mathcal{G}''(X)$. Setze:
\[E=\{(U,s)\mid U\subset_\text{o}X,\ s\in\mathcal{G}(U),\ s\mapsto s''|_U \} \]
$E$ ist partiell geordnet bzgl. Inklusion. Jede Kette in $E$ besitzt eine obere Schranke in $E$. Nach Zorns Lemma gibt es ein maximales Element $(U,s)\in E$.

Gibt es ein $x\in X\setminus U$, so existiert eine offene Umgebung $V$ von $x$ und $t\in\mathcal{G}(V)$ mit $t\mapsto s''|_V$, da $\mathcal{G}\to\mathcal{G}''$ surjektiv ist. Da $\mathcal{G}$ welk ist, so setzt sich $s-t\in\mathcal{G}(U\cap V)$ zu einem Schnitt auf $\mathcal{G}(V)$ fort. Ersetzen wir $t$ durch $s-t\in\mathcal{G}(V)$, können wir o.B.d.A. $s=t$ auf $V$ annehmen. Die Existenz von $(V,s)$ ist ein Widerspruch zur Maximalität von $(U,s)$ Somit folgt $U=X$.
\item Für $V\subset U$ offen in $X$ kommutiert das folgende Diagramm mit exakten Zeilen:
\[\begin{tikzcd}
0 \rar & \mathcal{G}'(U)\rar\dar[two heads] & \mathcal{G}(U)\rar\dar[two heads] & \mathcal{G}''(U)\rar\dar & 0\\
0 \rar & \mathcal{G}'(V)\rar & \mathcal{G}(V)\rar & \mathcal{G}''(V)\rar & 0
\end{tikzcd} \]
Es folgt die Surjektivität von $\mathcal{G}''(U)\to\mathcal{G}''(V)$.\qed
\end{enumerate}

\paragraph{Beweis von \hyperref[9.20]{Satz 9.20}.} Sei $0\to\mathcal{F}\to\mathcal{J}\to\mathcal{G}\to 0$ exakt mit $\mathcal{J}$ injektiv. Nun sind $\mathcal{F}$ und $\mathcal{J}$ welk. Nach \hyperref[9.21]{Lemma 9.21} ist auch $\mathcal{G}$ welk und wir haben eine exakte Folge:
\[0\longrightarrow\mathcal{F}(X)\longrightarrow\mathcal{J}(X)\longrightarrow\mathcal{G}(X)\longrightarrow 0 \]
Da $\mathrm{H}^i(X,\mathcal{J})=0$ für alle $i>0$, gilt $\mathrm{H}^1(X,\mathcal{F})=0$ und $\mathrm{H}^i(X,\mathcal{F})=\mathrm{H}^{i-1}(X,\mathcal{G})$ für alle $i\geq 2$. Somit folgt die Aussage per Induktion nach $i$.\qed

\paragraph{Satz 9.22.}\label{9.22} Sei $(X,\mathcal{O}_X)$ ein geringter Raum. Dann sind die derivierten Funktoren von
\[\Gamma(X,-): \mathbf{Mod}(X)\to\mathbf{Ab} \]
gerade die Kohomologie-Funktoren.

\paragraph{Beweis.} $\mathrm{R}^i\Gamma(X,-)$ wird über eine injektive Auflösung in $\mathbf{Mod}(X)$ berechnet. Nun sind injektive Objekte in $\mathbf{Mod}(X)$ nach \hyperref[9.19]{Lemma 9.19} welk und nach \hyperref[9.20]{Satz 9.20} insbesondere azyklisch bzgl. $\Gamma(X,-)$.Somit ist $\mathrm{R}^i\Gamma(X,-)=\mathrm{H}^i(X,-)$.\qed

\paragraph{Bemerkung 9.23.}\label{9.23} Sei $(X,\mathcal{O}_X)$ ein geringter Raum und $\mathcal{F}$ ein $\mathcal{O}_X$"~Modul. Da die Kohomologie durch Auflösungen in der Kategorie $\mathbf{Mod}(X)$ berechnet werden kann, haben alle Kohomologiegruppen von $\mathcal{F}$ eine $A$"~Modulstruktur mit $A=\Gamma(X,\mathcal{O}_X)$.

\paragraph{Lemma 9.24.}\label{9.24} Sei $X$ ein noetherscher topologischer Raum und $\mathcal{F}=\varinjlim_\alpha\mathcal{F}_\alpha$. Dann gilt:
\begin{enumerate}[(i)]
\item $\mathcal{F}(X)=\varinjlim_\alpha\mathcal{F}_\alpha(X)$
\item Sind alle $\mathcal{F}_\alpha$ welk, so auch $\mathcal{F}$.
\end{enumerate}

\paragraph{Beweis.}\begin{enumerate}
\item[(ii)] Für alle $\alpha$ und $V\subset U$ offen in $X$ ist $\mathcal{F}_\alpha(U)\to\mathcal{F}_\alpha(V)$ surjektiv. Da $\varinjlim$ exakt ist, folgt mit (i) die Surjektivität von:
\[\mathcal{F}(U)=\varinjlim_\alpha\mathcal{F}_\alpha(U)\to\varinjlim\mathcal{F}_\alpha(V) = \mathcal{F}(V) \]
\item[(i)] Der Funktor $\Gamma(X,-)$ vertauscht mit $\varinjlim$, siehe z.B. Godement, Topologie algébrique et théorie des faisceaux, p. 169.\qed
\end{enumerate}

\paragraph{Satz 9.25.}\label{9.25} Sei $X$ ein noetherscher topologischer Raum und $(\mathcal{F}_\alpha)_{\alpha\in A}$ ein induktives System in $\mathbf{Ab}(X)$. Dann gibt es einen natürlichen Isomorphismus:
\[\varinjlim\mathrm{H}^i(X,\mathcal{F}_\alpha) \stackrel{\sim}{\to} \mathrm{H}^i(X,\varinjlim\mathcal{F}_\alpha) \]

\paragraph{Beweis.} Für alle $\alpha$ haben wir natürliche Abbildungen $\mathcal{F}_\alpha\to\varinjlim\mathcal{F}_\alpha$. Wir erhalten $\mathrm{H}^i(X,\mathcal{F}_\alpha)\to\mathrm{H}^i(X,\varinjlim\mathcal{F}_\alpha)$ und daher eine Abbildung:
\[\varinjlim\mathrm{H}^i(X,\mathcal{F}_\alpha)\to\mathrm{H}^i(X,\varinjlim\mathcal{F}_\alpha) \]
Dies ist nach \hyperref[9.24]{Lemma 9.24 (i)} für $i=0$ ein Isomorphismus. Betrachte die abelsche Kategorie $\mathbf{Ind}_A(\mathbf{Ab}(X))$ der direkten Limiten von Objekten aus $\mathbf{Ab}(X)$ indiziert durch $A$. Da $\varinjlim$ exakt ist, haben wir eine natürliche Transformation von $\delta$-Funktoren $\mathbf{Ind}_A(\mathbf{Ab}(X))\to\mathbf{Ab}$
\[\varinjlim\mathrm{H}^i(X,-)\to\mathrm{H}^i(X,\varinjlim -) \]
die für $i=0$ übereinstimmt. Jetzt genügt es zu zeigen, dass beide Funktoren auslöschbar sind. Dann sind beide universell und insbesondere gleich.

Sei $(\mathcal{F}_\alpha)\in\mathbf{Ind}_A(\mathbf{Ab}(X))$. Zu $\alpha$ definieren wir die Garbe $\mathcal{G}_\alpha$ wie folgt:
\[U\mapsto \Big\{s:U\to\coprod_{P\in U}(\mathcal{F}_\alpha)_P\mid s(P)\in (\mathcal{F}_\alpha)_P \Big\} \]
Dann ist $\mathcal{G}_\alpha$ welk und $\mathcal{F}_\alpha\hookrightarrow\mathcal{G}_\alpha$. Nach Konstruktion bilden die $\mathcal{G}_\alpha$ ein induktives System und wir erhalten einen Monomorphismus $u:(\mathcal{F}_\alpha)\hookrightarrow (\mathcal{G}_\alpha)$ in $\mathbf{Ind}_A(\mathbf{Ab}(X))$. Ferner gilt $\mathrm{H}^i(X,\mathcal{G}_\alpha)=0$ für alle $i>0$ nach \hyperref[9.20]{Satz 9.20}, also $\varinjlim\mathrm{H}^i(X,\mathcal{G}_\alpha)=0$ für alle $i>0$. Somit ist $\varinjlim\mathrm{H}^i(X,-)$ auslöschbar.

Da $\varinjlim\mathcal{G}_\alpha$ nach \hyperref[9.24]{Satz 9.24 (ii)} welk ist, ist auch $\mathrm{H}^i(X,\varinjlim\mathcal{G}_\alpha)=0$ für alle $i>0$, d.h. auch $\mathrm{H}^i(X,\varinjlim -)$ ist auslöschbar.\qed

\paragraph{Lemma 9.26.}\label{9.26} Sei $Y\subset X$ abgeschlossen mit natürlicher Inklusion $i:Y\hookrightarrow X$ und $\mathcal{F}\in\mathbf{Ab}(Y)$. Dann gilt:
\[\mathrm{H}^i(Y,\mathcal{F})=\mathrm{H}^i(X, i_\ast\mathcal{F}) \]

\paragraph{Beweis.} Sei $\mathcal{J}^\bullet$ eine injektive Auflösung von $\mathcal{F}$ auf $Y$; insbesondere ist $\mathcal{J}^\bullet$ eine welke Auflösung. Dann ist $i_\ast\mathcal{J}^\bullet$ eine welke Auflösung von $i_\ast\mathcal{F}$ auf $X$. Für alle $i$ gilt:
\[\Gamma(Y,\mathcal{J}^i) = \Gamma(X,i_\ast\mathcal{J}^i) \]
Somit ergeben sich die gleichen Kohomologiegruppen.\qed

\paragraph{Satz 9.27.}\label{9.27} \textit{(Verschwindungssatz von Grothendieck)}\index{Verschwindungssatz} Sei $X$ ein noetherscher topologischer Raum der Dimension $n$. Dann gilt für alle abelschen Garben $\mathcal{F}$ auf $X$:
\[\mathrm{H}^i(X,\mathcal{F})=0\quad\text{für alle }i>n \]

\paragraph{Beweis.} Sei $\mathcal{F}$ eine Garbe abelscher Gruppen auf $X$.
\begin{enumerate}
\item Schritt: \textit{(Reduktion auf irreduzible $X$)} Sei $X$ reduzibel und $Y\subset X$ eine irreduzible Komponente mit Einbettungen $i:Y\hookrightarrow X,\ j:U=X\setminus Y\hookrightarrow X$. Nach \hyperref[1.16]{1.16} ist die folgende Garbensequenz exakt:
\[0\longrightarrow \mathcal{F}_U \longrightarrow\mathcal{F}\longrightarrow\mathcal{F}_Y\longrightarrow 0 \]
wobei $\mathcal{F}_U= j_!(\mathcal{F}|_U)$ und $\mathcal{F}_Y=i_\ast(\mathcal{F}|_Y)$ ist. Es genügt zu zeigen, dass $\mathrm{H}^i(X,\mathcal{F}_U)=0=\mathrm{H}^i(X,\mathcal{F}_Y)$ für alle $i>n$ ist. Betrachte $g:\overline{U}\hookrightarrow X$. Nun hat $\overline{U}$ weniger irreduzible Komponenten als $X$. Setze $\mathcal{F}_{\overline{U}}=g^\ast\mathcal{F}_U$. Es folgt nach \hyperref[9.26]{Lemma 9.26}:
\[\mathrm{H}^i(\overline{U},\mathcal{F}_{\overline{U}})=\mathrm{H}^i(X,g_\ast\mathcal{F}_{\overline{U}}) = \mathrm{H}^i(X,\mathcal{F}_U) \]
\[\mathrm{H}^i(X,\mathcal{F}_Y)=\mathrm{H}^i(X,i_\ast\mathcal{F}|_Y)=\mathrm{H}^i(Y,\mathcal{F}|_Y) \]
Per Induktion nach Anzahl der irreduziblen Komponenten von $X$ können wir ohne Einschränkung $X$ als irreduzibel annehmen.
\item Schritt: Sei $X$ irreduzibel und $\dim(X)=0$. Die einzigen offenen Mengen von $X$ sind $\varnothing$ und $X$, d.h. $\Gamma(X,-):\mathbf{Ab}(X)\to\mathbf{Ab}$ ist eine Äquivalenz von Kategorien. Insbesondere ist $\Gamma(X,-)$ exakt, d.h. $\mathrm{H}^i(X,\mathcal{F})=0$ für alle $i>0$.
\item Schritt: Sei $X$ irreduzibel und $\dim(X)=n$. Sei $j:U\hookrightarrow X$ offen und $s\in\mathcal{F}(U)$. Da
\[\mathcal{F}(U)=\operatorname{Hom}(\mathbb{Z},\mathcal{F}(U))=\operatorname{Hom}_U(\mathbb{Z}|_U,\mathcal{F}|_U)=\operatorname{Hom}_X(j_!(\mathbb{Z}|_U),\mathcal{F}) \]
wobei $\mathbb{Z}|_U$ die konstante Garbe bezeichnet, entspricht $s$ einen Morphismus $j_!(\mathbb{Z}|_U)\to\mathcal{F}$. Setze $\mathcal{F}_s=\operatorname{im}(j_!(\mathbb{Z}|_U)\to\mathcal{F})$. $\mathcal{F}_s$ heißt die von $s$ erzeugte Untergarbe von $\mathcal{F}$. Sei $\alpha=(s_i)_{i\in I},\ s_i\in\mathcal{F}(U_i)$ eine beliebige Familie von Schnitten. Setze:
\[\mathcal{F}_\alpha = \sum_{i\in I}\mathcal{F}_{s_i}\subset\mathcal{F}\]
Sei $J$ die Menge aller endlichen Familien $\alpha=(s_i)$. $J$ wird mit der folgenden par\-tiel\-len Ordnung zu einem gerichteten System:
\[\alpha'\leq\alpha\iff \alpha'\text{ ist Unterfamilie von }\alpha \]
Für $\alpha'\leq\alpha$ gilt $\mathcal{F}_{\alpha'}\subset\mathcal{F}_\alpha$. Somit ist $\mathcal{F}=\varinjlim_{\alpha\in J}\mathcal{F}_\alpha$. Nach \hyperref[9.25]{Satz 9.25} reicht es zu zeigen, dass $\mathrm{H}^i(X,\mathcal{F}_\alpha)=0$ für alle $i>n$ und alle $\alpha\in J$.

Sei nun $\alpha'\leq\alpha$. Betrachte die exakte Sequenz $0\to\mathcal{F}_{\alpha'}\to\mathcal{F}_\alpha\to\mathcal{G}\to 0$. $\mathcal{G}$ ist erzeugt von $\#\alpha - \#\alpha'$ Schnitten über geeignete offenen Mengen. Per Induktion nach $\#\alpha$ können wir unter Betrachtung der langen exakten Kohomologiefolge o.B.d.A. annehmen, dass $\mathcal{F}$ von einem Schnitt über einer geeigneten offenen Menge $U$ erzeugt wird, d.h. $\mathcal{F}=\mathcal{F}_s$. Die zu $s$ gehörige Morphismus $\mathbb{Z}_U=j_!(\mathbb{Z}|_U)\to\mathcal{F}$ ist surjektiv.

Definiere die Garbe $\mathcal{R}$ durch die exakte Folge $0\to\mathcal{R}\to\mathbb{Z}_U\to\mathcal{F}\to 0$. Unter Betrachtung der langen exakten Kohomologiefolge genügt es zu zeigen, dass $\mathrm{H}^i(X,\mathcal{R})=0=\mathrm{H}^i(X,\mathbb{Z}_U)$ für alle $i>n$ ist.
\item Schritt: Für $x\in U$ ist $\mathcal{R}_x$ eine Untergruppe von $\mathbb{Z}$. Ist $\mathcal{R} = 0$, so folgt $\mathcal{F}\cong\mathbb{Z}_U$. Sei nun $\mathcal{R}\neq 0$. Setze:
\[d=\min\{m>0\mid m\in\mathcal{R}_x,\ x\in U\} \]
\end{enumerate}

### 2015-05-29 Algebraische Geometrie/chapter3.tex
Lemma 9.29

@@
 Unter Betrachtung der langen exakten Kohomologiefolge genügt es zu zeigen, dass $\mathrm{H}^i(X,\mathcal{R})=0=\mathrm{H}^i(X,\mathbb{Z}_U)$ für alle $i>n$ ist.
-\item Schritt: Für $x\in U$ ist $\mathcal{R}_x$ eine Untergruppe von $\mathbb{Z}$. Ist $\mathcal{R} = 0$, so folgt $\mathcal{F}\cong\mathbb{Z}_U$. Sei nun $\mathcal{R}\neq 0$. Setze:
+\item Schritt: Wir zeigen $\mathrm{H}^i(X,\mathcal{R})=0$ für alle $i>n$ per Induktion über $n=\dim(X)$.
+Für $x\in U$ ist $\mathcal{R}_x$ eine Untergruppe von $\mathbb{Z}$.
+Ist $\mathcal{R} = 0$, so folgt $\mathcal{F}\cong\mathbb{Z}_U$.
+Sei nun $\mathcal{R}\neq 0$.
+Setze:
 \[d=\min\{m>0\mid m\in\mathcal{R}_x,\ x\in U\} \]
+Es existiert ein $\varnothing\neq V\subset_\text{o}U$ mit $\mathcal{R}|_V\cong d\cdot\mathbb{Z}|_V\subset \mathbb{Z}|_V$.
+Wir erhalten $\mathcal{R}_V=j_!(\mathcal{R}|_V)=\mathbb{Z}_V$, wobei $j:V\hookrightarrow X$ die kanonische Inklusion ist.
+Wir erhalten die exakte Sequenz:
+\[0\longrightarrow\mathbb{Z}_V\longrightarrow\mathcal{R}\longrightarrow\mathcal{R}/\mathbb{Z}_V\longrightarrow 0 \]
+Nun gilt $\operatorname{supp}(\mathcal{R}/\mathbb{Z}_V)\subset\overline{U\setminus V}\subset X$.
+Da $X$ irreduzibel ist, gilt $\dim(\overline{U\setminus V})<n$.
+Aus der Induktionsannahme folgt mit \hyperref[9.26]{Lemma 9.26} für alle $i\geq n$:
+\[\mathrm{H}^i(X,\mathcal{R}/\mathbb{Z}_V) = \mathrm{H}^i(\overline{U\setminus V},\mathcal{R}/\mathbb{Z}_V|_{\overline{U\setminus V}})=0 \]
+Nach der langen exakten Kohomologiesequenz reicht es zu zeigen, dass $\mathrm{H}^i(X,\mathbb{Z}_V)=0$ für $i>n$ gilt.
+\item Schritt: Wir zeigen $\mathrm{H}^i(X,\mathbb{Z}_V)=0$ für alle $i>n$ per Induktion über $n=\dim(X)$.
+Sei $U\subset_\text{o}X$ und $Y=X\setminus U$.
+Dann ist $0\to\mathbb{Z}_U\to\mathbb{Z}\to\mathbb{Z}_Y\to 0$ exakt nach \hyperref[1.16]{1.16}.
+Wegen $\dim(Y)<\dim(X)$ folgt nach der Induktionsannahme und \hyperref[9.26]{Lemma 9.26} für alle $i\geq n$:
+\[\mathrm{H}^i(X,\mathbb{Z}_Y)=\mathrm{H}^i(Y,\mathbb{Z}|_Y)=0 \]
+Ferner ist die konstante Garbe $\mathbb{Z}$ welk.
+Nach \hyperref[9.20]{Satz 9.20} folgt $\mathrm{H}^i(X,\mathbb{Z})=0$ für alle $i>0$.
+Die lange exakte Kohomologiesequenz liefert $\mathrm{H}^i(X,\mathbb{Z}_U)=0$ für $i>n$.\qed
 \end{enumerate}
+
+\paragraph{Satz 9.28.}\label{9.28} Sei $X=\operatorname{Spec}(A)$ noethersch. Dann ist $\mathrm{H}^i(X,\mathcal{F})=0$ für alle $i>0$ und alle quasikohärenten Garben $\mathcal{F}$.
+
+\paragraph{Bemerkung.} Dieser Satz gilt auch für nicht-noethersche Ringe $A$.
+
+\paragraph{Satz von Artin-Rees.}\index{Satz von Artin-Rees}\label{prop:artin-rees} Sei $A$ ein noetherscher Ring, $\mathfrak{a}\subset A$ ein Ideal und $M\subset N$ endlich erzeugte $A$"~Moduln.
+Dann ist die $\mathfrak{a}$-adische Topologie auf $M$ induziert von der $\mathfrak{a}$-adischen Topologie auf $N$, d.h:
+\[\forall n\geq 0,\ \exists m\geq n:\ \mathfrak{a}^nM\supset\mathfrak{a}^mN\cap M \]
+
+\paragraph{Beweis.} Siehe beispielsweise Bosch: Algebraic Geometry and Commutative Algebra, 2.3. Lemma 1.\qed
+
+\paragraph{Definition.} Sei $A$ ein Ring, $\mathfrak{a}\subset A$ ein Ideal und $M$ ein $A$"~Modul. Wir definieren:
+\[\Gamma_\mathfrak{a}(M)=\{m\in M\mid\exists n>0:\ \mathfrak{a}^nm=0 \} \]
+
+\paragraph{Lemma 9.29.}\label{9.29} Sei $A$ noethersch, $\mathfrak{a}\subset A$ ein Ideal und $I$ ein injektiver $A$"~Modul. Dann ist $\Gamma_\mathfrak{a}(I)$ ebenfalls injektiv.
+
+\paragraph{Beweis.} Setze $J=\Gamma_\mathfrak{a}(I)$.
+Wir zeigen zunächst: Für alle Ideale $\mathfrak{b}\subset A$ und alle $A$-Modulhomomorphismen $\varphi:\mathfrak{b}\to J$ gibt es einen $A$-Modulhomomorphismus $\psi:A\to J$ mit $\psi|_\mathfrak{b}=\varphi$.
+
+Sei $\mathfrak{b}\subset A$ ein notwendigerweise endlich erzeugtes Ideal und $\varphi:\mathfrak{b}\to J$ ein $A$-Modulhomomorphismus.
+Zu $x\in J$ gibt es nach Definition ein $n>0$ mit $\mathfrak{a}^nx=0$.
+Da $\mathfrak{b}$ endlich erzeugt ist, existiert ein $n$ mit:
+\[0=\mathfrak{a}^n\varphi(\mathfrak{b})=\varphi(\mathfrak{a}^n\mathfrak{b}) \]
+Nach dem \hyperref[prop:artin-rees]{Satz von Artin-Rees} gibt es ein $m\geq n$ mit $\mathfrak{a}^n\mathfrak{b}\supset\mathfrak{b}\cap\mathfrak{a}^m$, also $\varphi(\mathfrak{b}\cap\mathfrak{a}^m)=0$.
+Somit faktorisiert $\varphi$ über $\mathfrak{b}/(\mathfrak{b}\cap\mathfrak{a}^m)$:
+\[\begin{tikzcd}
+A \rar & A/\mathfrak{a}^m\ar[rrd, dashed] &&\\
+\mathfrak{b} \uar[hook]\rar & \mathfrak{b}/(\mathfrak{b}\cap\mathfrak{a}^m) \uar[hook]\rar & J\rar[hook] & I
+\end{tikzcd}\]
+Da $I$ injektiv ist, gibt es eine Fortsetzung $\psi':A/\mathfrak{a}^m\to I$ von $\varphi$, das obiges Diagramm kommutativ macht.
+Nun gilt $\mathfrak{a}^m\psi'(A/\mathfrak{a}^m)=0$, d.h.
+$\psi'(A/\mathfrak{a}^m)\subset J$.
+Somit ist $\psi:A\to A/\mathfrak{a}^m\to J$ eine Fortsetzung von $\varphi:\mathfrak{b}\to J$.
+
+Sei nun $X$ ein $A$-Modul, $X'\subset X$ ein Untermodul und $f:X'\to J$ ein Homomorphismus. Betrachte:
+\[\Sigma=\{(Y,g)\mid Y\subset X\text{ Untermodul},\ X'\subset Y,\ g:Y\to J,\ g|_{X'}=f \} \]
+Dann ist $(X',f)\in\Sigma$ und $\Sigma$ ist induktiv geordnet.
+Nach Zorn gibt es ein maximales Element $(Y,g)\in\Sigma$.
+Angenommen, $Y\neq X$.
+Dann gibt es ein $x\in X\setminus Y$.
+Betrachte das Ideal $\mathfrak{b}=\{\lambda\in A\mid\lambda x\in Y\}\subset A$ und den $A$-Modulhomomorphismus $\varphi:\mathfrak{b}\to J,\ \lambda\mapsto g(\lambda x)$.
+Diese können wir zu einem $\psi:A\to J$ fortsetzen.
+Definiere $g':\langle Y,x\rangle\to J$ durch $g'|_Y=g$ und $g'(x)=\psi(1)$.
+Diese Abbildung ist ein wohldefinierter $A$-Modulhomomorphismus, da $g'(y+\lambda x)=g'(y)+\psi(\lambda)$ für $y\in Y,\ \lambda\in A$ und ist $\lambda x\in Y$, so folgt $\lambda\in\mathfrak{b}$, d.h.
+$\psi(\lambda)=\varphi(\lambda)=g(\lambda x)=g'(\lambda x)$.
+Dies ist ein Widerspruch zur Maximalität von $(Y,g)$, daher folgt $Y=X$.\qed
+
+\paragraph{Satz 9.30.}\label{9.30} Sei $A$ ein noetherscher Ring und $I$ ein injektiver $A$-Modul. Dann ist die Garbe $\widetilde{I}$ welk über $\operatorname{Spec}(A)$.
+
+%\paragraph{Beweis.} Sei $Y=\overline{\operatorname{supp}(\widetilde{I}}$.

### 2015-05-29 Algebraische Geometrie/chapter2.tex
Consistency update

@@
 \item Sei $F$ eine Garbe auf $X$ und $s\in F(U)$ ein Schnitt über $U\subset_\text{o}X$. Dann heißt
-\[\operatorname{Supp}(s)=\{P\in U\mid s_P\neq 0\} \]
-wobei $s_P\in F_P$ der Keim von $s$ in $P$ bezeichnet, der \textit{Support}\index{Support} von $s$. $\operatorname{Supp}(s)$ ist abgeschlossen in $U$.
-\[\operatorname{Supp}(F)=\{P\in X\mid F_P\neq 0\} \]
+\[\operatorname{supp}(s)=\{P\in U\mid s_P\neq 0\} \]
+wobei $s_P\in F_P$ der Keim von $s$ in $P$ bezeichnet, der \textit{Support}\index{Support} von $s$. $\operatorname{supp}(s)$ ist abgeschlossen in $U$.
+\[\operatorname{supp}(F)=\{P\in X\mid F_P\neq 0\} \]
 heißt \textit{Support} von $F$. Dieser ist nicht notwendigerweise abgeschlossen.
@@
 Sei $\mathcal{A}$ die konstante Garbe auf $\overline{\{P\}}$ und $i:\overline{\{P\}}\hookrightarrow X$ die natürliche Inklusion.
-Dann heißt die Garbe $i_\ast \mathcal{A}$ \textit{Wolkenkratzergarbe}.
+Dann heißt die Garbe $i_\ast \mathcal{A}$ \textit{Wolkenkratzergarbe}\index{Wolkenkratzergarbe}.
 Es gilt:
@@
 \begin{align*}
-B\otimes_A A_\mathfrak{q}\neq 0 &\iff \mathfrak{q}\in\operatorname{Supp}(B)\text{, wobei $B$ als $A$-Modul aufgefasst wird}\\
+B\otimes_A A_\mathfrak{q}\neq 0 &\iff \mathfrak{q}\in\operatorname{supp}(B)\text{, wobei $B$ als $A$-Modul aufgefasst wird}\\
 &\iff \mathfrak{q}\supset\mathfrak{a}\\
@@
 \item Sei $\mathcal{J}$ eine quasikohärente Idealgarbe auf $X$. Setze:
-\[Y=\operatorname{Supp}(\mathcal{O}_X/\mathcal{J})=\{x\in X\mid (\mathcal{O}_X/\mathcal{J})_x\neq 0 \}\subset X \]
+\[Y=\operatorname{supp}(\mathcal{O}_X/\mathcal{J})=\{x\in X\mid (\mathcal{O}_X/\mathcal{J})_x\neq 0 \}\subset X \]
 Wir zeigen, dass $(Y,\mathcal{O}_X/\mathcal{J})$ ein abgeschlossenes Unterschema von $X$ ist.
@@
 \begin{align*}
-Y&=\operatorname{Supp}(\mathcal{O}_X/\mathcal{J})=\operatorname{Supp}(\widetilde{A/\mathfrak{a}})\\
+Y&=\operatorname{supp}(\mathcal{O}_X/\mathcal{J})=\operatorname{supp}(\widetilde{A/\mathfrak{a}})\\
 &= \{\mathfrak{p}\in X\mid (A/\mathfrak{a})_\mathfrak{p}\neq 0\}\\

### 2015-05-29 Algebraische Geometrie/chapter3.tex
28.05.2015

@@
 
-%\paragraph{Beweis.} Sei $Y=\overline{\operatorname{supp}(\widetilde{I}}$.
+\paragraph{Lemma 9.31.}\label{9.31} Sei $A$ ein noetherscher Ring und $I$ ein injektiver $A$-Modul. Dann ist für jedes $f\in A$ die natürliche Abbildung $\theta:I\to I_f$ surjektiv.
+
+\paragraph{Beweis.} Betrachte die folgenden Ideale in $A$:
+\[\mathfrak{b}_i=\operatorname{ann}(f^i)= \{x\in A\mid f^ix=0\} \]
+Dann gilt $\mathfrak{b}_1\subset\mathfrak{b}_2\subset\ldots\subset\mathfrak{b}_r=\mathfrak{b}_{r+1}=\ldots$ für ein $r$.
+Sei $x\in I_f$.
+Dann gibt es ein $y\in I$ und ein $n\geq 0$ mit $x=\frac{\theta(y)}{f^n}$.
+Definiere den $A$-Modulhomomorphismus:
+\[\varphi: (f^{n+r})\to I,\ f^{n+r}\mapsto f^ry \]
+Dieser ist wohldefiniert, da $\operatorname{ann}(f^{n+r})= \mathfrak{b}_{n+r}=\mathfrak{b}_r = \operatorname{ann}(f^r)\subset\operatorname{ann}(f^ry)$.
+Da $I$ injektiv ist, gibt es eine Fortsetzung $\psi:A\to I$ von $\varphi$.
+Sei $z=\psi(1)$.
+Dann gilt $f^{n+r}z=f^ry$, also $\theta(z)=\frac{\theta(y)}{f^n}=x$.\qed
+
+\paragraph{Beweis von \hyperref[9.30]{9.30}.} Sei $Y=\overline{\operatorname{supp}(\widetilde{I})}$.
+Besteht $Y$ nur aus einem abgeschlossenen Punkt, so ist $\widetilde{I}$ eine Wolkenkratzergarbe und somit welk.
+
+Sei nun $Y$ größer und die Aussage für alle kleineren $Y$ bewiesen.
+Es genügt zu zeigen, dass alle Restriktionsabbildungen der Form $\Gamma(X,\widetilde{I})\to\Gamma(U,\widetilde{I})$ surjektiv sind.
+Sei $U\subset_\text{o}X$.
+Ist $Y\cap U=\varnothing$, so ist nichts zu zeigen.
+Sei $Y\cap U\neq\varnothing$.
+Es existiert ein $f\in A$ mit $X_f=D(f)\subset U$ und $X_f\cap Y\neq\varnothing$.
+Setze $Z=X\setminus X_f$ und $\Gamma_Z(U,\widetilde{I})=\{s\in\Gamma(U,\widetilde{I})\mid\operatorname{supp}(s)\subset Z \}$.
+Betrachte:
+\[\begin{tikzcd}
+\Gamma(X,\widetilde{I})\rar & \Gamma(U,\widetilde{I})\rar & \Gamma(X_f,\widetilde{I})\\
+\Gamma_Z(X,\widetilde{I})\rar\uar[hook] & \Gamma_Z(U,\widetilde{I}) \uar[hook]&
+\end{tikzcd} \]
+Sei $s\in\Gamma(U,\widetilde{I})$ und betrachte die Restriktion $s'\in\Gamma(X_f,\widetilde{I})=I_f$.
+Nach \hyperref[9.31]{Lemma 9.31} gibt es ein $t\in I=\Gamma(X,\widetilde{I})$ mit $t|_{X_f}=s'$.
+Sein $t'$ die Restriktion von $t$ auf $U$.
+Dann ist $(s-t')|_{X_f}=0$, also $\operatorname{supp}(s-t')\subset Z$.
+Bleibt noch zu zeigen, dass $\Gamma_Z(X,\widetilde{I})\to\Gamma_Z(U,\widetilde{I})$ surjektiv ist.
+Dann existiert nämlich ein Urbild $x\in\Gamma_Z(X,\widetilde{I})$ von $s-t'$ und daher:
+\[(x+t)|_U=s-t'+t'=s\]
+Sei $J=\Gamma_Z(X,\widetilde{I})$ und $\mathfrak{a}=(f)\subset A$.
+Dann ist $J=\Gamma_\mathfrak{a}(I)$.
+Nach \hyperref[9.29]{Lemma 9.29} ist $J$ ein injektiver $A$-Modul mit $\operatorname{supp}(\widetilde{J})\subset Y\cap Z\subsetneq Y$.
+Nach Induktionsannahme ist $\widetilde{J}$ welk, also ist die Restriktionsabbildung surjektiv:
+\[\Gamma_Z(X,\widetilde{I}) = \Gamma(X,\widetilde{J})\to \Gamma(U,\widetilde{J})=\Gamma_Z(U,\widetilde{I})\qedhere \]
+
+\paragraph{Beweis von \hyperref[9.28]{9.28}.} Sei $\mathcal{F}=\widetilde{M}$ eine quasikohärente Garbe über einem noetherschen Ring $A$.
+Sei $0\to M\to I^\bullet$ eine injektive Auflösung in $\mathbf{Mod}(A)$.
+Wir erhalten eine exakte Folge $0\to\mathcal{F}\to\widetilde{I}^\bullet$.
+Nach \hyperref[9.30]{Satz 9.30} sind alle $\widetilde{I}^i$ welk und somit azyklisch für $\Gamma(X,-)$.
+Daher kann diese Auflösung zur Kohomologieberechnung benutzt werden.
+Anwendung von $\Gamma(X,-)$ gibt gerade die exakte Sequenz $0\to M\to I^\bullet$.
+Also ist $\mathrm{H}^i(X,\mathcal{F})=0$ für alle $i>0$.\qed

### 2015-06-03 Algebraische Geometrie/chapter3.tex
Čech-Kohomologie

@@
 
-\paragraph{Satz von Artin-Rees.}\index{Satz von Artin-Rees}\label{prop:artin-rees} Sei $A$ ein noetherscher Ring, $\mathfrak{a}\subset A$ ein Ideal und $M\subset N$ endlich erzeugte $A$"~Moduln.
+\paragraph{Satz von Krull.}\index{Satz von Krull}\label{prop:krull} Sei $A$ ein noetherscher Ring, $\mathfrak{a}\subset A$ ein Ideal und $M\subset N$ endlich erzeugte $A$"~Moduln.
 Dann ist die $\mathfrak{a}$-adische Topologie auf $M$ induziert von der $\mathfrak{a}$-adischen Topologie auf $N$, d.h:
@@
 \[0=\mathfrak{a}^n\varphi(\mathfrak{b})=\varphi(\mathfrak{a}^n\mathfrak{b}) \]
-Nach dem \hyperref[prop:artin-rees]{Satz von Artin-Rees} gibt es ein $m\geq n$ mit $\mathfrak{a}^n\mathfrak{b}\supset\mathfrak{b}\cap\mathfrak{a}^m$, also $\varphi(\mathfrak{b}\cap\mathfrak{a}^m)=0$.
+Nach dem \hyperref[prop:krull]{Satz von Krull} gibt es ein $m\geq n$ mit $\mathfrak{a}^n\mathfrak{b}\supset\mathfrak{b}\cap\mathfrak{a}^m$, also $\varphi(\mathfrak{b}\cap\mathfrak{a}^m)=0$.
 Somit faktorisiert $\varphi$ über $\mathfrak{b}/(\mathfrak{b}\cap\mathfrak{a}^m)$:
@@
 Also ist $\mathrm{H}^i(X,\mathcal{F})=0$ für alle $i>0$.\qed
+
+\paragraph{Korollar 9.32.}\label{9.32} Sei $X$ ein noethersches Schema und $\mathcal{F}\in\mathbf{Qcoh}(X)$. Dann kann $\mathcal{F}$ in eine welke quasikohärente Garbe eingebettet werden.
+
+\paragraph{Beweis.} Sei $X=\bigcup_i U_i$ eine endliche, offen affine Überdeckung mit $U_i=\operatorname{Spec}(A_i)$ und $\mathcal{F}|_{U_i}\cong\widetilde{M_i}$. Setze:
+\[\mathcal{G} =\bigoplus_i f_{i\ast}\widetilde{I_i} \]
+wobei $f_i:U_i\hookrightarrow X$ die kanonische Inklusion und $M_i\hookrightarrow I_i$ mit injektivem $I_i$.
+Es folgt $\mathcal{F}|_{U_i}=\widetilde{M_i} \hookrightarrow\widetilde{I_i}$.
+Betrachte den kanonischen Morphismus $\mathcal{F}\to f_{i\ast}f_i^\ast\mathcal{F}=f_{i\ast}\mathcal{F}|_{U_i}\to f_{i\ast}\widetilde{I_i}$.
+Dieser induziert einen Morphismus $\mathcal{F}\to\bigoplus_i f_{i\ast}\widetilde{I_i}=\mathcal{G}$.
+Betrachtet man die Halme, so sieht man, dass dieser Morphismus injektiv ist.
+
+Nach \hyperref[9.30]{Satz 9.30} sind $\widetilde{I_i}$ alle welk, daher sind die $f_{i\ast}\widetilde{I_i}$ welk und somit auch $\mathcal{G}$.
+Da alle $\widetilde{I_i}$ quasikohärent sind, sind auch alle $f_{i\ast}\widetilde{I_i}$ quasikohärent nach \hyperref[5.3]{Satz 5.3 (iv)} und somit ist auch $\mathcal{G}$ quasikohärent.\qed
+
+\paragraph{Theorem 9.33.}\label{9.33} \textit{(Serre)} Sei $X$ ein noethersches Schema. Dann sind folgende Aussagen äquivalent:
+\begin{enumerate}[(i)]
+\item $X$ ist affin.
+\item $\mathrm{H}^i(X,\mathcal{F})=0$ für alle quasikohärenten $\mathcal{F}$ und $i>0$.
+\item $\mathrm{H}^1(X,\mathcal{J})=0$ für alle kohärenten Idealgarben $\mathcal{J}$.
+\end{enumerate}
+
+\paragraph{Beweis.} Siehe z.B. Hartshorne: Algebraic Geometry III 3.7.\qed
+
+\section{Der Čech-Komplex}
+
+\paragraph{Definition 10-$\varepsilon$.1.}\label{9.5.1} Sei $X$ ein topologischer Raum, $\mathcal{F}$ eine Prägarbe abelscher Gruppen auf $X$, $I$ eine total geordnete Indexmenge und $\mathcal{U}=(U_i)_{i\in I}$ eine offene Überdeckung von $X$.
+Wir definieren:
+\[C^k(\mathcal{U},\mathcal{F})=\prod_{i_0<\ldots<i_k}\mathcal{F}(U_{i_0}\cap\cdots\cap U_{i_k}),\quad k\geq 0 \]
+mit Randoperatoren
+\[d: C^k(\mathcal{U},\mathcal{F})\to C^{k+1}(\mathcal{U},\mathcal{F}) \]
+wobei diese für $\alpha=(\alpha_{i_0,\ldots,i_k})_{i_0<\ldots<i_k}$ durch
+\[(d\alpha)_{i_0,\ldots,i_k} = \sum_{j=0}^{k+1} (-1)^j \cdot\alpha_{i_0,\ldots,\widehat{i_j},\ldots,i_{k+1}}|_{U_{i_0}\cap\cdots\cap U_{i_{k+1}}} \]
+definiert sind. Die Folge $C^\bullet(\mathcal{U},\mathcal{F})$ heißt \textit{Čech-Komplex} von $\mathcal{U}$ mit Koeffizienten in $\mathcal{F}$.\index{Cech-Komplex}
+
+\paragraph{Satz 10-$\varepsilon$.2.}\label{9.5.2} Es gilt für $C^\bullet(\mathcal{U},\mathcal{F})$:
+\begin{enumerate}[(i)]
+\item $d\circ d =0$
+\item Für eine Prägarbe $\mathcal{F}$ auf $X$ und eine offene Überdeckung $\mathcal{U}$ setzen wir:
+\[\check{\mathrm{H}}^k(\mathcal{U},\mathcal{F}) = \frac{\ker( C^k(\mathcal{U},\mathcal{F})\to C^{k+1}(\mathcal{U},\mathcal{F}) )}{\operatorname{im}(C^{k-1}(\mathcal{U},\mathcal{F})\to C^{k}(\mathcal{U},\mathcal{F}))},\quad k\geq 0 \]
+Dann ist $\check{\mathrm{H}}^k(\mathcal{U},-)$ ein Funktor nach $\mathbf{Ab}$.
+\item Es gilt $\check{\mathrm{H} }^0(\mathcal{U},-)=\Gamma(X,-)$.
+\end{enumerate}
+
+\paragraph{Definition 10-$\varepsilon$.3.}\label{9.5.3} $\check{\mathrm{H}}^k(\mathcal{U},\mathcal{F})$ heißt $k$-te \textit{Čech-Kohomologiegruppe}\index{Cech-Kohomologie} von $\mathcal{U}$ mit Koeffizienten in $\mathcal{F}$.
+
+\paragraph{Bemerkung 10-$\varepsilon$.4.}\label{9.5.4} Es lässt sich auch bilden:
+\[\check{\mathrm{H}}^\bullet(X,\mathcal{F}) = \varinjlim_{\mathcal{U}}\check{\mathrm{H}}^\bullet(\mathcal{U},\mathcal{F}) \]
+Es gilt stets $\check{\mathrm{H}}^1(X,\mathcal{F})\cong \mathrm{H}^1(X,\mathcal{F})$. Im Allgemeinen gilt keine Gleichheit für höhere Kohomologiegruppen.

### 2015-06-03 Algebraische Geometrie/chapter3.tex
Fix index error
Change Čech numbering

@@
 
-\paragraph{Definition 10-$\varepsilon$.1.}\label{9.5.1} Sei $X$ ein topologischer Raum, $\mathcal{F}$ eine Prägarbe abelscher Gruppen auf $X$, $I$ eine total geordnete Indexmenge und $\mathcal{U}=(U_i)_{i\in I}$ eine offene Überdeckung von $X$.
+\paragraph{Definition 10$-\varepsilon$.1.}\label{9.5.1} Sei $X$ ein topologischer Raum, $\mathcal{F}$ eine Prägarbe abelscher Gruppen auf $X$, $I$ eine total geordnete Indexmenge und $\mathcal{U}=(U_i)_{i\in I}$ eine offene Überdeckung von $X$.
 Wir definieren:
@@
 wobei diese für $\alpha=(\alpha_{i_0,\ldots,i_k})_{i_0<\ldots<i_k}$ durch
-\[(d\alpha)_{i_0,\ldots,i_k} = \sum_{j=0}^{k+1} (-1)^j \cdot\alpha_{i_0,\ldots,\widehat{i_j},\ldots,i_{k+1}}|_{U_{i_0}\cap\cdots\cap U_{i_{k+1}}} \]
+\[(d\alpha)_{i_0,\ldots,i_{k+1}} = \sum_{j=0}^{k+1} (-1)^j \cdot\alpha_{i_0,\ldots,\widehat{i_j},\ldots,i_{k+1}}|_{U_{i_0}\cap\cdots\cap U_{i_{k+1}}} \]
 definiert sind. Die Folge $C^\bullet(\mathcal{U},\mathcal{F})$ heißt \textit{Čech-Komplex} von $\mathcal{U}$ mit Koeffizienten in $\mathcal{F}$.\index{Cech-Komplex}
 
-\paragraph{Satz 10-$\varepsilon$.2.}\label{9.5.2} Es gilt für $C^\bullet(\mathcal{U},\mathcal{F})$:
+\paragraph{Satz 10$-\varepsilon$.2.}\label{9.5.2} Es gilt für $C^\bullet(\mathcal{U},\mathcal{F})$:
 \begin{enumerate}[(i)]
@@
 
-\paragraph{Definition 10-$\varepsilon$.3.}\label{9.5.3} $\check{\mathrm{H}}^k(\mathcal{U},\mathcal{F})$ heißt $k$-te \textit{Čech-Kohomologiegruppe}\index{Cech-Kohomologie} von $\mathcal{U}$ mit Koeffizienten in $\mathcal{F}$.
+\paragraph{Definition 10$-\varepsilon$.3.}\label{9.5.3} $\check{\mathrm{H}}^k(\mathcal{U},\mathcal{F})$ heißt $k$-te \textit{Čech-Kohomologiegruppe}\index{Cech-Kohomologie} von $\mathcal{U}$ mit Koeffizienten in $\mathcal{F}$.
 
-\paragraph{Bemerkung 10-$\varepsilon$.4.}\label{9.5.4} Es lässt sich auch bilden:
+\paragraph{Bemerkung 10$-\varepsilon$.4.}\label{9.5.4} Es lässt sich auch bilden:
 \[\check{\mathrm{H}}^\bullet(X,\mathcal{F}) = \varinjlim_{\mathcal{U}}\check{\mathrm{H}}^\bullet(\mathcal{U},\mathcal{F}) \]

### 2015-06-03 Algebraische Geometrie/chapter3.tex
Kohomologie des projektiven Raumes, Teil I

@@
 \[\check{\mathrm{H}}^\bullet(X,\mathcal{F}) = \varinjlim_{\mathcal{U}}\check{\mathrm{H}}^\bullet(\mathcal{U},\mathcal{F}) \]
-Es gilt stets $\check{\mathrm{H}}^1(X,\mathcal{F})\cong \mathrm{H}^1(X,\mathcal{F})$. Im Allgemeinen gilt keine Gleichheit für höhere Kohomologiegruppen.
+Es gilt stets $\check{\mathrm{H}}^1(X,\mathcal{F})\cong \mathrm{H}^1(X,\mathcal{F})$. Im Allgemeinen gibt es keine Isomorphie für höhere Kohomologiegruppen.
+
+\section{Kohomologie des projektiven Raumes}
+
+\paragraph{Satz 10.1.}\label{10.1} Sei $A$ ein noetherscher Ring, $S=A[X_0,\ldots,X_r]$ für ein $r\geq 1$ und $X=\operatorname{Proj}(S)=\mathbf{P}_A^r$. Dann gilt:
+\begin{enumerate}[(i)]
+\item $\bigoplus_{n\in\mathbb{Z}} \mathrm{H}^0(X,\mathcal{O}_X(n))\cong S$ ist ein Isomorphismus von graduierten $S$"~Moduln.
+\item $\mathrm{H}^i(X,\mathcal{O}_X(n))=0$ für $0<i<r$ und alle $n\in\mathbb{Z}$.
+\item $\mathrm{H}^r(X,\mathcal{O}_X(-r-1))\cong A$
+\item Für alle $n\in\mathbb{Z}$ ist die natürliche Paarung
+\[ \mathrm{H}^0(X,\mathcal{O}_X(n))\times\mathrm{H}^r(X,\mathcal{O}_X(-n-r-1))\to\mathrm{H}^r(X,\mathcal{O}_X(-r-1))\cong A \]
+eine perfekte Paarung freier $A$"~Moduln, d.h. sie induziert einen Isomorphismus:
+\[\mathrm{H}^0(X,\mathcal{O}_X(n))^\ast\cong \mathrm{H}^r(X,\mathcal{O}_X(-n-r-1)) \]
+Diese Paarung stimmt mit dem Cup-Produkt überein.
+\end{enumerate}
+
+\paragraph{Lemma 10.2.}\label{10.2} Sei $X$ ein separiertes, noethersches Schema, $\mathcal{U}$ eine offen affine Überdeckung und $\mathcal{F}$ eine quasikohärente Garbe auf $X$. Dann gilt:
+\[\check{\mathrm{H}}^\bullet(\mathcal{U},\mathcal{F}) = \mathrm{H}^\bullet(X,\mathcal{F}) \]
+
+\paragraph{Beweis von \hyperref[10.1]{10.1}.} Setze $\mathcal{F}=\bigoplus_{n\in\mathbb{Z}}\mathcal{O}_X(n)$.
+Da die Kohomologie mit direkten Summen vertauscht, liefert $\mathrm{H}^\bullet(X,\mathcal{F})$ gerade alle nötigen Daten.
+Sei $\mathcal{U}=(U_i)_{i=0,\ldots,r}$ mit $U_i=D_+(X_i)$.
+Nach \hyperref[10.2]{Lemma 10.2} gilt $\mathrm{H}^\bullet(X,\mathcal{F})=\check{\mathrm{H}}^\bullet(\mathcal{U},\mathcal{F})$.
+Nach \hyperref[5.21]{Satz 5.21} gilt:
+\[\mathcal{F}(U_{i_0}\cap\cdots\cap U_{i_k}) = \mathcal{F}(D_+(X_{i_0}\cdots X_{i_k})) \cong S_{X_{i_0}\cdots X_{i_k}} \]
+Somit ist:
+\[C^\bullet(\mathcal{U},\mathcal{F}) = \Big(\prod_{i_0\in I} S_{X_{i_0}}\to \prod_{i_0<i_1} S_{X_{i_0}X_{i_1}}\to\cdots\to S_{X_0\cdots X_r} \to 0\to\cdots\Big) \]
+\begin{enumerate}
+\item[(i)] folgt aus \hyperref[5.25]{Satz 5.25}.
+\item[(iii)] Es gilt:
+\[\mathrm{H}^r(X,\mathcal{F}) = \operatorname{coker}\Big(d:\prod_k S_{X_0\cdots \widehat{X_k}\cdots X_r} \to S_{X_0\cdots X_r} \Big) \]
+Wir betrachten $S_{X_0\cdots X_r}$ als freien $A$"~Modul mit Basis $\{X_0^{e_0}\cdots X_r^{e_r}\mid e_i\in\mathbb{Z}\}$.
+Damit ist das Bild von $d$ der freie Untermodul, der von den Monomen $X_0^{e_0}\cdots X_r^{e_r}$ mit $e_i\geq 0$ erzeugt wird.
+Somit ist $\mathrm{H}^r(X,\mathcal{F})$ der freie $A$"~Modul mit Basis $\{X_0^{e_0}\cdots X_r^{e_r}\mid e_i < 0\}$.
+Nun ist $\mathrm{H}^r(X,\mathcal{O}_X(-r-1))$ gerade die Elemente vom Grad $-r-1$ in $\mathrm{H}^r(X,\mathcal{F})$, d.h.
+gerade der Untermodul, der von den Monomen $\{X_0^{e_0}\cdots X_r^{e_r}\mid e_i<0,\ \sum e_i=-r-1\}$ erzeugt wird.
+Somit ist:
+\[\mathrm{H}^r(X,\mathcal{F}) = \frac{1}{X_0\cdots X_r} A\cong A \]
+\item[(iv)] Sei zunächst $n<0$.
+Dann ist $\mathrm{H}^0(X,\mathcal{O}_X(n))=0$ nach (i).
+Da $-n-r-1>-r-1$, folgt nach der Rechnung in (iii) auch $\mathrm{H}^r(X,\mathcal{O}_X(-n-r-1))=0$ und die Dualität ist trivial.
+
+Sei nun $n\geq 0$.
+Dann ist $\mathrm{H}^0(X,\mathcal{O}_X(n))$ ein freier $A$"~Modul mit Basis $\{X_0^{m_0}\cdots X_r^{m_r}\mid m_i\geq 0,\ \sum m_i=n\}$.
+$\mathrm{H}^r(X,\mathcal{O}_X(-n-r-1))$ ist freier $A$"~Modul mit Basis $\{X_0^{e_0}\cdots X_r^{e_r}\mid e_i<0,\ \sum e_i = -n-r-1\}$.
+Auf Ebene der Čech-Kohomologie ist die Paarung gegeben durch:
+\begin{align*}
+\mathrm{H}^0(X,\mathcal{O}_X(n))\times\mathrm{H}^r(X,\mathcal{O}_X(-n-r-1))&\to\mathrm{H}^r(X,\mathcal{O}_X(-r-1)),\\
+(X_0^{m_0}\cdots X_r^{m_r}, X_0^{e_0}\cdots X_r^{e_r})&\mapsto X_0^{m_0+e_0}\cdots X_r^{m_r+e_r}
+\end{align*}
+In $\mathrm{H}^r(X,\mathcal{O}_X(-r-1))$ gilt $X_0^{m_0+e_0}\cdots X_r^{m_r+e_r}\neq 0$ genau dann, wenn $m_i+e_i = -1$ für alle $i$ gilt.
+Somit ist durch $\{X_0^{-m_0-1}\cdots X_r^{-m_r-1}\in\mathrm{H}^r(X,\mathcal{O}_X(-n-r-1))\mid m_i\geq 0,\ \sum m_i=n\}$ eine Dualbasis gegeben und die Paarung ist somit perfekt.
+
+\end{enumerate}

### 2015-06-04 Algebraische Geometrie/chapter3.tex
Kohomologie des projektiven Raumes, Teil II

@@
 \end{align*}
-In $\mathrm{H}^r(X,\mathcal{O}_X(-r-1))$ gilt $X_0^{m_0+e_0}\cdots X_r^{m_r+e_r}\neq 0$ genau dann, wenn $m_i+e_i = -1$ für alle $i$ gilt.
-Somit ist durch $\{X_0^{-m_0-1}\cdots X_r^{-m_r-1}\in\mathrm{H}^r(X,\mathcal{O}_X(-n-r-1))\mid m_i\geq 0,\ \sum m_i=n\}$ eine Dualbasis gegeben und die Paarung ist somit perfekt.
+In $\mathrm{H}^r(X,\mathcal{O}_X(-r-1))$ gilt $X_0^{m_0+e_0}\cdots X_r^{m_r+e_r}\neq 0$ genau dann, wenn $m_i+e_i = -1$, d.h.
+$e_i=-m_i-1$, für alle $i$ gilt.
+Somit bilden die dualen Elemente von $X_0^{-m_0-1}\cdots X_r^{-m_r-1},\ m_i\geq 0,\ \sum m_i =n$ eine Basis von $\mathrm{H}^r(X,\mathcal{O}_X(-n-r-1))^\ast$.
+Die Paarung ist somit perfekt.
+\item[(ii)] Wir zeigen die Aussage per Induktion über $r$.
+Für $r=1$ ist die Aussage trivial.
+Wir lokalisieren $C^\bullet(\mathcal{U},\mathcal{F})$ homogen nach $X_r$ und setze $\mathcal{U}_r = (U_i\cap U_r)_{i=0,\ldots,r}$.
+Dann gilt:
+\[C^\bullet(\mathcal{U},\mathcal{F})_{(X_r)} = C^\bullet(\mathcal{U}_r,\mathcal{F}|_{U_r}) \]
+Insbesondere ist $U_r\in\mathcal{U}_r$, also folgt $\check{\mathrm{H}}^k(\mathcal{U}_r,\mathcal{F}|_{U_r})=0$ für $k>0$.
+Ferner ist die homogene Lokalisierung ein exakter Funktor und vertauscht somit mit der Kohomologie.
+Es folgt:
+\begin{align*}
+0 = \check{\mathrm{H}}^k(\mathcal{U}_r,\mathcal{F}|_{U_r}) &= \mathrm{H}^k(C^\bullet(\mathcal{U}_r,\mathcal{F}|_{U_r}))\\
+&= \mathrm{H}^k(C^\bullet(\mathcal{U},\mathcal{F})_{(X_r)})
+= \check{\mathrm{H}}^k(\mathcal{U},\mathcal{F})_{(X_r)} = \mathrm{H}^k(X,\mathcal{F})_{(X_r)}
+\end{align*}
+Also wird für alle $k>0$ jedes $\alpha\in\mathrm{H}^k(X,\mathcal{F})$ durch eine Potenz von $X_r$ annulliert.
+Wir zeigen nun, dass die $X_r$-Multiplikation $\mathrm{H}^k(X,\mathcal{F})\to\mathrm{H}^k(X,\mathcal{F})$ für $0<k<r$ injektiv ist.
+Daraus folgt die Behauptung.
 
+Betrachte die exakte Sequenz von graduierten $S$"~Moduln:
+\[0\longrightarrow S(-1)\stackrel{\cdot X_r}{\longrightarrow} S\longrightarrow S/(X_r)\longrightarrow 0  \]
+Sie wird zu einer exakten Sequenz von Garben:
+\[0\longrightarrow\mathcal{O}_X(-1)\longrightarrow\mathcal{O}_X\longrightarrow\mathcal{O}_H\longrightarrow 0 \]
+mit $H=V_+(X_r)$. Twiste diese Sequenz mit allen $\mathcal{O}_X(n)$ und bilde direkte Summe. Diese Operationen sind alle exakt, daher erhalten wir die exakte Sequenz:
+\[0\longrightarrow\mathcal{F}(-1)\stackrel{\cdot X_r}{\longrightarrow}\mathcal{F}\longrightarrow\mathcal{F}_H=\bigoplus_{n\in\mathbb{Z}}\mathcal{O}_H(n)\longrightarrow 0 \]
+Betrachte die lange exakte Kohomologiesequenz:
+\[\cdots \longrightarrow \mathrm{H}^{k-1}(X,\mathcal{F}_H)\longrightarrow\mathrm{H}^k(X,\mathcal{F}(-1))\stackrel{\cdot X_r}{\longrightarrow}\mathrm{H}^k(X,\mathcal{F})\longrightarrow\mathrm{H}^k(X,\mathcal{F}_H)\longrightarrow\cdots \]
+Nach Induktionsannahme ist $\mathrm{H}^k(H,\mathcal{F}_H)=0$ für $0<k<r-1$, da $H\cong\mathbf{P}_A^{r-1}$.
+Nach \hyperref[9.26]{Lemma 9.26} gilt $\mathrm{H}^k(H,\mathcal{F}_H)= \mathrm{H}^k(X,\mathcal{F}_H)$.
+Somit sind die $X_r$-Multiplikationen $\mathrm{H}^k(X,\mathcal{F}(-1))\to\mathrm{H}^k(X,\mathcal{F})$ für $1<k<r-1$ Isomorphismen.
+
+Für $k=1$ betrachte die exakte Sequenz $0\to S(-1)\to S\to S/(X_r)\to 0$:
+\[0\longrightarrow\mathrm{H}^0(X,\mathcal{F}(-1))\longrightarrow \mathrm{H}^0(X,\mathcal{F})\longrightarrow\mathrm{H}^0(X,\mathcal{F}_H)\longrightarrow 0 \]
+Wir erhalten die exakte Folge:
+\[0 \longrightarrow \mathrm{H}^1(X,\mathcal{F}(-1))\stackrel{\cdot X_r}{\longrightarrow}\mathrm{H}^1(X,\mathcal{F})\longrightarrow\mathrm{H}^1(X,\mathcal{F}_H)=0\longrightarrow\cdots \]
+Für $k=r-1$ betrachte die exakte Folge:
+\begin{align*}
+0=\mathrm{H}^{r-2}(X,\mathcal{F}_H)&\longrightarrow\mathrm{H}^{r-1}(X,\mathcal{F}(-1))\stackrel{\cdot X_r}{\longrightarrow}\mathrm{H}^{r-1}(X,\mathcal{F})\\
+&\longrightarrow\mathrm{H}^{r-1}(X,\mathcal{F}_H)\stackrel{\delta}{\longrightarrow}\mathrm{H}^r(X,\mathcal{F}(-1))\stackrel{\cdot X_r}{\longrightarrow}\mathrm{H}^r(X,\mathcal{F})\qedhere
+\end{align*}
 \end{enumerate}

### 2015-06-04 Algebraische Geometrie/chapter3.tex
Truncate long exact sequence
Truncate long exact sequence 2
Small fixes

@@
 Sei nun $n\geq 0$.
-Dann ist $\mathrm{H}^0(X,\mathcal{O}_X(n))$ ein freier $A$"~Modul mit Basis $\{X_0^{m_0}\cdots X_r^{m_r}\mid m_i\geq 0,\ \sum m_i=n\}$.
-$\mathrm{H}^r(X,\mathcal{O}_X(-n-r-1))$ ist freier $A$"~Modul mit Basis $\{X_0^{e_0}\cdots X_r^{e_r}\mid e_i<0,\ \sum e_i = -n-r-1\}$.
+Dann ist $\mathrm{H}^0(X,\mathcal{O}_X(n))$ ein freier $A$"~Modul mit Basis $\{X_0^{m_0}\cdots X_r^{m_r}\mid m_i\geq 0,\ \sum m_i=n\}$ und $\mathrm{H}^r(X,\mathcal{O}_X(-n-r-1))$ ein freier $A$"~Modul mit Basis $\{X_0^{e_0}\cdots X_r^{e_r}\mid e_i<0,\ \sum e_i = -n-r-1\}$.
 Auf Ebene der Čech-Kohomologie ist die Paarung gegeben durch:
@@
 Für $k=r-1$ betrachte die exakte Folge:
-\begin{align*}
-0=\mathrm{H}^{r-2}(X,\mathcal{F}_H)&\longrightarrow\mathrm{H}^{r-1}(X,\mathcal{F}(-1))\stackrel{\cdot X_r}{\longrightarrow}\mathrm{H}^{r-1}(X,\mathcal{F})\\
-&\longrightarrow\mathrm{H}^{r-1}(X,\mathcal{F}_H)\stackrel{\delta}{\longrightarrow}\mathrm{H}^r(X,\mathcal{F}(-1))\stackrel{\cdot X_r}{\longrightarrow}\mathrm{H}^r(X,\mathcal{F})\qedhere
-\end{align*}
+\[0=\mathrm{H}^{r-2}(X,\mathcal{F}_H)\longrightarrow\mathrm{H}^{r-1}(X,\mathcal{F}(-1))\stackrel{\cdot X_r}{\longrightarrow}\mathrm{H}^{r-1}(X,\mathcal{F})\qedhere\]
 \end{enumerate}

### 2015-06-04 Algebraische Geometrie/chapter3.tex
Kategorienbezeichnungen aktualisiert

@@
 \item $\mathbf{Ab}(X)$, die Kategorie der Garben abelscher Gruppen über einem topologischen Raum $X$
-\item $\mathbf{Mod}(X)$, die Kategorie der $\mathcal{O}_X$"~Moduln auf einem geringten Raum $(X,\mathcal{O}_X)$
-\item $\mathbf{Qcoh}(X)$, die Kategorie der quasikohärenten $\mathcal{O}_X$"~Moduln auf einem Schema $X$
-\item $\mathbf{Coh}(X)$, die Kategorie der kohärenten $\mathcal{O}_X$"~Modulgarben auf einem noetherschen Schema $X$
+\item $\mathbf{Mod}(X)$ oder $\mathbf{Mod}(\mathcal{O}_X)$, die Kategorie der $\mathcal{O}_X$"~Moduln auf einem geringten Raum $X$ mit Strukturgarbe $\mathcal{O}_X$.
+\item $\mathbf{Qcoh}(X)$ oder $\mathbf{Qcoh}(\mathcal{O}_X)$, die Kategorie der quasikohärenten $\mathcal{O}_X$"~Moduln auf einem Schema $X$
+\item $\mathbf{Coh}(X)$ oder $\mathbf{Coh}(\mathcal{O}_X)$, die Kategorie der kohärenten $\mathcal{O}_X$"~Modulgarben auf einem noetherschen Schema $X$
 \end{itemize}

### 2015-06-12 Algebraische Geometrie/chapter3.tex
09.06.2015

@@
 \end{enumerate}
+
+\paragraph{Satz 10.3.}\label{10.3} \textit{(Serre)} Sei $X$ ein projektives Schema über einem noetherschen Ring $A$, $\mathcal{O}_X(1)$ eine sehr ample Garbe auf $X$ über $\operatorname{Spec}(A)$ und $\mathcal{F}$ eine kohärente Garbe auf $X$.
+Dann gilt:
+\begin{enumerate}[(i)]
+\item $\mathrm{H}^i(X, \mathcal{F})$ ist endlich erzeugter $A$"~Modul für alle $i\geq 0$.
+\item Es gibt ein $n_0$, so dass $\mathrm{H}^i(X, \mathcal{F}(n))=0$ für alle $i>0$ und $n\geq n_0$.
+\end{enumerate}
+
+\paragraph{Bemerkung 10.4.}\label{10.4} Als Spezialfall von (i) gilt \hyperref[5.39]{Satz 5.39}: $\Gamma(X,\mathcal{F})$ ist endlich erzeugter $A$-Modul für kohärente Garben $\mathcal{F}$.
+
+\paragraph{Beweis.} Da $\mathcal{O}_X(1)$ sehr ampel ist, gibt es eine Immersion $i:X\hookrightarrow  \mathbf{P}_A^r$ für ein $r$ mit $\mathcal{O}_X(1)=i^\ast\mathcal{O}(1)$.
+Aus der Projektivität folgt nach \hyperref[4.24]{Theorem 4.24} die Eigentlichkeit von $X\to\operatorname{Spec}(A)$.
+Nun ist $\mathbf{P}_A^r\to\operatorname{Spec}(A)$ separiert, also folgt nach \hyperref[4.22]{Korollar 4.22 (v)} die Eigentlichkeit von $X\to\mathbf{P}_A^r$, insbesondere ist $i$ abgeschlossen.
+Da $\mathcal{F}$ kohärent ist, ist auch $i_\ast\mathcal{F}$ nach \hyperref[5.36]{Lemma 5.36 (ii)} kohärent auf $\mathbf{P}_A^r$ und nach \hyperref[9.26]{Lemma 9.26} gilt $\mathrm{H}^i(X,\mathcal{F}(n)) =\mathrm{H}^i(\mathbf{P}_A^r, i_\ast\mathcal{F}(n))$.
+Wir können also o.B.d.A.
+$X=\mathbf{P}_A^r$ annehmen.
+
+Die Aussagen (i) und (ii) gelten für Garben der Form $\mathcal{O}_X(q),\ q\in\mathbb{Z}$ nach \hyperref[10.1]{Satz 10.1}, dem \hyperref[9.27]{Verschwindungssatz} und der Tatsache:
+\[\mathrm{H}^r(X,\mathcal{O}_X(q)) \cong \operatorname{Hom}(\mathrm{H}^0(X,\mathcal{O}_X(-q-r-1)), A) =0\quad\text{für $-q-r-1<0$, d.h. q>-r-1} \]
+Es folgt (i) und (ii) für endliche direkte Summen von $\mathcal{O}_X(q),\ q\in\mathbb{Z}$.
+Sei nun $\mathcal{F}$ beliebig kohärent.
+Die allgemeine Aussage beweisen wir nun per absteigende Induktion über $i$.
+Für $i>r$ folgt alles aus dem Verschwindungssatz.
+Sei nun $i\leq r$ und die Aussage für höhere $i$ bewiesen.
+Nach \hyperref[5.38]{Korollar 5.38} gibt es eine exakte Sequenz:
+\[0\longrightarrow\mathcal{R}\longrightarrow\mathcal{O}_X(q)^N\longrightarrow\mathcal{F}\longrightarrow 0 \]
+$\mathcal{R}$ ist kohärent nach \hyperref[5.12]{Satz 5.12}. Betrachte die lange exakte Kohomologiesequenz:
+\[\cdots\longrightarrow \mathrm{H}^i(X,\mathcal{O}_X(q)^N)\longrightarrow\mathrm{H}^i(X,\mathcal{F})\longrightarrow\mathrm{H}^{i+1}(X,\mathcal{R})\longrightarrow\cdots \]
+Die äußeren beiden Gruppen sind endlich erzeugt. Da $A$ noethersch ist, folgt die endliche Erzeugbarkeit von $\mathrm{H}^i(X,\mathcal{F})$. Twisten wir die Folge um $n$ ergibt:
+\[\cdots\longrightarrow \mathrm{H}^i(X,\mathcal{O}_X(q+n)^N)\longrightarrow\mathrm{H}^i(X,\mathcal{F}(n))\longrightarrow\mathrm{H}^{i+1}(X,\mathcal{R}(n))\longrightarrow\cdots\]
+Die äußeren beiden Gruppen verschwinden für hinreichend große $n$, daher verschwindet auch $\mathcal{H}^i(X,\mathcal{F}(n))$ für $n\geq n_0$.
+Da die Kohomologiegruppen für fast alle $i$ komplett verschwinden, kann $n_0$ unabhängig von $i$ gewählt werden.\qed
+
+\paragraph{Satz 10.5.}\label{10.5} Sei $A$ ein noetherscher Ring und $X$ eigentlich über $\operatorname{Spec}(A)$. Sei $\mathcal{L}$ eine invertierbare Garbe. Dann sind äquivalent:
+\begin{enumerate}[(i)]
+\item $\mathcal{L}$ ist ampel.
+\item Für alle kohärenten Garben $\mathcal{F}$ auf $X$ gibt es ein $n_0$, so dass $\mathrm{H}^i(X,\mathcal{F}\otimes \mathcal{L}^{\otimes n})=0$ für alle $i>0$ und $n\geq n_0$.
+\end{enumerate}
+
+\paragraph{Beweis.} \begin{itemize}
+\item Sei $\mathcal{L}$ ampel auf $X$.
+Nach \hyperref[7.7]{Satz 7.7} existiert ein $m>0$, so dass $\mathcal{L}^{\otimes m}$ sehr ampel auf $X$ bzgl.
+$\operatorname{Spec}(A)$ ist.
+Da $X$ eigentlich über $A$ ist, folgt nach \hyperref[5.32]{Satz 5.32} die Projektivität von $X\to\operatorname{Spec}(A)$.
+Wenden wir \hyperref[10.3]{Satz 10.3} auf $\mathcal{F},\ \mathcal{F}\otimes\mathcal{L},\ \mathcal{F}\otimes\mathcal{L}^{\otimes 2},\ \ldots,\ \mathcal{F}\otimes\mathcal{L}^{\otimes (m-1)}$ an, so erhalten wir (ii).
+\item Sei nun (ii) erfüllt und $\mathcal{F}$ eine kohärente Garbe auf $X$.
+Wir zeigen, dass es ein $n_0$ existiert, so dass für alle $n\geq n_0$ die Garbe $\mathcal{F}\otimes\mathcal{L}^{\otimes n}$ von globalen Schnitten erzeugt wird.
+\end{itemize}
+
+\section{Ext-Gruppen}
+
+Sei $\mathcal{F}$ ein $\mathcal{O}_X$"~Modul auf einem geringten Raum $(X,\mathcal{O}_X)$. Wir haben folgende linksexakte Funktoren:
+\[\operatorname{Hom}(\mathcal{F},-):\mathbf{Mod}(X)\to\mathbf{Ab} \]
+\[\operatorname{\mathcal{H}om}_{\mathcal{O}_X}(\mathcal{F},-):\mathbf{Mod}(X)\to\mathbf{Mod}(X) \]
+
+\paragraph{Definition 11.1.}\label{11.1} Sei $(X,\mathcal{O}_X)$ ein geringter Raum und $\mathcal{F}\in\mathbf{Mod}(X)$. Für $i\geq 0$ definieren wir:
+\[\operatorname{Ext}^i(\mathcal{F},-)=\mathrm{R}^i\operatorname{Hom}(\mathcal{F},-) \]
+\[\operatorname{\mathcal{E}xt}^i(\mathcal{F},-) =\mathrm{R}^i\operatorname{\mathcal{H}om}(\mathcal{F},-) \]
+
+\paragraph{Bemerkung 11.2.}\label{11.2} Es gilt:
+\begin{enumerate}[(i)]
+\item $\operatorname{Ext}^0=\operatorname{Hom},\ \operatorname{\mathcal{E}xt}^0 =\operatorname{\mathcal{H}om}$
+\item $\operatorname{Ext}^i(\mathcal{F},\mathcal{J})=0$ und $\operatorname{\mathcal{E}xt}^i(\mathcal{F},\mathcal{J})=0$ für alle $i>0$ und injektive $\mathcal{J}$.
+\end{enumerate}
+
+\paragraph{Lemma 11.3.}\label{11.3} Sei $\mathcal{J}\in\mathbf{Mod}(X)$ injektiv und $U\subset_\text{o}X$. Dann ist $\mathcal{J}|_U\in\mathbf{Mod}(U)$ injektiv.
+
+\paragraph{Beweis.} Sei $j:U\hookrightarrow X$ die natürliche Inklusion.
+Sei $\mathcal{F}\subset\mathcal{G}$ in $\mathbf{Mod}(U)$ und $\mathcal{F}\to\mathcal{J}|_U$ gegeben.
+Es gilt $j_!\mathcal{F}\subset j_!\mathcal{G}$ und wir erhalten einen Morphismus $j_!\mathcal{F}\to j_!(\mathcal{J}|_U)\hookrightarrow\mathcal{J}$.
+Da $\mathcal{J}$ injektiv ist, existiert eine Fortsetzung $j_!\mathcal{G}\to\mathcal{J}$.
+Somit ist $\mathcal{G}=(j_!\mathcal{G})|_U\to\mathcal{J}|_U$ eine Fortsetzung von $\mathcal{F}=(j_!\mathcal{F})|_U\to\mathcal{J}|_U$.\qed
+
+\paragraph{Satz 11.4.}\label{11.4} Sei $(X,\mathcal{O}_X)$ ein geringter Raum und $U\subset_\text{o} X$. Dann gilt für $\mathcal{F},\mathcal{G}\in\mathbf{Mod}(X)$ und alle $i\geq 0$:
+\[\operatorname{\mathcal{E}xt}^i(\mathcal{F},\mathcal{G})|_U\cong \operatorname{\mathcal{E}xt}^i(\mathcal{F}|_U,\mathcal{G}|_U) \]
+
+\paragraph{Beweis.} Beide Seiten sind $\delta$-Funktoren $\mathbf{Mod}(X)\to\mathbf{Mod}(X)$ und gleich für $i=0$. Sie verschwinden beide auf Injektiven, sind somit auslöschbar, also universell.\qed
+
+\paragraph{Satz 11.5.}\label{11.5} Sei $\mathcal{G}$ ein $\mathcal{O}_X$-Modul. Dann gilt:
+\begin{enumerate}[(i)]
+\item $\operatorname{\mathcal{E}xt}^0(\mathcal{O}_X,\mathcal{G})\cong\mathcal{G}$
+\item $\operatorname{\mathcal{E}xt}^i(\mathcal{O}_X,\mathcal{G})=0$ für alle $i>0$
+\item $\operatorname{Ext}^i(\mathcal{O}_X,\mathcal{G}) \cong\mathrm{H}^i(X,\mathcal{G})$ für alle $i\geq 0$
+\end{enumerate}
+
+\paragraph{Beweis.} Es gilt $\operatorname{\mathcal{H}om}(\mathcal{O}_X,-) = \operatorname{Id}$ und $\operatorname{Hom}(\mathcal{O}_X,-)=\Gamma(X,-)$.\qed

### 2015-06-12 Algebraische Geometrie/chapter3.tex
11.06.2015, Teil 1

@@
 \paragraph{Beweis.} Es gilt $\operatorname{\mathcal{H}om}(\mathcal{O}_X,-) = \operatorname{Id}$ und $\operatorname{Hom}(\mathcal{O}_X,-)=\Gamma(X,-)$.\qed
+
+\paragraph{Satz 11.6.}\label{11.6} Sei $0\to\mathcal{F}'\to\mathcal{F}\to\mathcal{F}''\to 0$ eine exakte Folge in $\mathbf{Mod}(X)$ und $\mathcal{G}\in\mathbf{Mod}(X)$.
+Dann ist die folgende Folge exakt:
+\[0\longrightarrow\operatorname{Hom}(\mathcal{F}'',\mathcal{G})\longrightarrow\operatorname{Hom}(\mathcal{F},\mathcal{G})\longrightarrow\mathcal{Hom}(\mathcal{F}')\longrightarrow\operatorname{Ext}^1(\mathcal{F}'',\mathcal{G})\longrightarrow\operatorname{Ext}^1(\mathcal{F},\mathcal{G})\longrightarrow\cdots\]
+Analog für $\operatorname{\mathcal{H}om}$ und $\operatorname{\mathcal{E}xt}$.
+
+\paragraph{Beweis.} Sei $0\to\mathcal{G}\to\mathcal{J}^\bullet$ eine injektive Auflösung. Dann ist $\operatorname{Hom}(-,\mathcal{J}^i)$ exakt. Es folgt daher die Exaktheit von:
+\[0\longrightarrow\operatorname{Hom}(\mathcal{F}'',\mathcal{J}^\bullet)\longrightarrow\operatorname{Hom}(\mathcal{F},\mathcal{J}^\bullet)\longrightarrow\operatorname{Hom}(\mathcal{F}',\mathcal{J}^\bullet)\longrightarrow 0\]
+Die lange exakte Kohomologiesequenz liefert die Aussage. Die Aussage für $\operatorname{\mathcal{E}xt}$ folgt analog mithilfe von \hyperref[11.3]{Lemma 11.3}.\qed
+
+\paragraph{Satz 11.7.}\label{11.7} Sei $\cdots\to\mathcal{L}_1\to\mathcal{L}_0\to\mathcal{F}\to 0$ eine exakte Sequenz von $\mathcal{O}_X$-Moduln, wobei alle $\mathcal{L}_i$ lokal frei von endlichem Rang ist.
+Dann gilt für alle $\mathcal{O}_X$-Moduln $\mathcal{G}$:
+\[\operatorname{\mathcal{E}xt}^i(\mathcal{F},\mathcal{G})\cong \mathrm{H}^i(\operatorname{\mathcal{H}om}(\mathcal{L}_\bullet,\mathcal{G})) \]
+
+\paragraph{Beweis.} Beide Seiten sind $\delta$-Funktoren $\mathbf{Mod}(X)\to\mathbf{Mod}(X)$ in $\mathcal{G}$, die gleich für $i=0$ sind.
+Beide Seiten verschwinden für injektive $\mathcal{G}$.
+Somit sind beide $\delta$-Funktoren universell.\qed
+
+\paragraph{Beispiel 11.8.}\label{11.8} Sei $X$ ein quasiprojektives Schema über $\operatorname{Spec}(A)$ mit $A$ noethersch.
+Nach \hyperref[5.38]{Korollar 5.38} ist jede kohärente Garbe von $X$ ein Quotient von einer lokal freien Garbe von endlichem Rang.
+Somit besitzt jede kohärente Garbe eine lokal freie Auflösung $\mathcal{L}_\bullet\to\mathcal{F}\to 0$ von endlichem Rang.
+Somit kann $\operatorname{\mathcal{E}xt}^\bullet(\mathcal{F},-)$ durch eine lokal freie Auflösung von endlichem Rang in der ersten Variable berechnet werden.
+$\mathbf{Mod}(X)$ besitzt aber nicht genügend viele Projektive.
+
+\paragraph{Lemma 11.9.}\label{11.9} Sei $\mathcal{L}\in\mathbf{Mod}(X)$ lokal frei von endlichem Rang und $\mathcal{J}\in\mathbf{Mod}(X)$ injektiv. Dann ist $\mathcal{J}\otimes\mathcal{L}$ injektiv.
+
+\paragraph{Beweis.} Wir zeigen, dass $\operatorname{Hom}(-,\mathcal{J}\otimes\mathcal{L})$ exakter Funktor ist. Nun gilt:
+\[\operatorname{Hom}(-,\mathcal{J}\otimes\mathcal{L})=\operatorname{Hom}(-\otimes\mathcal{L}^\wedge,\mathcal{J})\]
+mit $\mathcal{L}^\wedge =\operatorname{Hom}(\mathcal{L},\mathcal{O}_X)$. Da $\mathcal{J}$ injektiv ist, folgt die Behauptung.\qed
+
+\paragraph{Satz 11.10.}\label{11.10} Sei $\mathcal{L}$ eine lokal freie Garbe von endlichem Rang und $\mathcal{L}^\wedge=\operatorname{Hom}(\mathcal{L},\mathcal{O}_X)$ sein Dual.
+Dann gilt für alle $\mathcal{F},\mathcal{G}\in\mathbf{Mod}(X)$:
+\begin{align*}
+\operatorname{Ext}^i(\mathcal{F}\otimes\mathcal{L},\mathcal{G}) &\cong \operatorname{Ext}^i(\mathcal{F},\mathcal{L}^\wedge\otimes\mathcal{G})\\
+\operatorname{\mathcal{E}xt}^i(\mathcal{F}\otimes\mathcal{L},\mathcal{G}) &\cong \operatorname{\mathcal{E}xt}^i(\mathcal{F},\mathcal{L}^\wedge\otimes\mathcal{G})
+\end{align*}
+
+\paragraph{Beweis.} Für $i=0$ sind die Aussagen klar.
+Es ist $-\otimes\mathcal{L}^\wedge$ ein exakter Funktor.
+Somit sind beide Seiten $\delta$-Funktoren in $\mathcal{G}$, die auf Injektive $\mathcal{G}$ verschwinden und daher universell.\qed
+
+\paragraph{Satz 11.11.}\label{11.11} Sei $X$ ein noethersches Schema, $\mathcal{F}$ eine kohärente Garbe auf $X$, $\mathcal{G}$ ein  $\mathcal{O}_X$-Modul und $x\in X$. Dann gilt:
+\[\operatorname{\mathcal{E}xt}^i(\mathcal{F},\mathcal{G})_x\cong \operatorname{Ext}^i_{\mathcal{O}_{X,x}}(\mathcal{F}_x,\mathcal{G}_x) = \mathrm{R}^i\operatorname{Hom}_{\mathcal{O}_{X,x}}(\mathcal{F}_x,\mathcal{G}_x) \]
+
+\paragraph{Beweis.} Die Aussage ist lokal, sei also o.B.d.A.
+$X$ affin.
+Dann besitzt $\mathcal{F}$ eine lokal freie Auflösung von endlichem Rang $\mathcal{L}_\bullet\to\mathcal{F}\to 0$ nach \hyperref[11.8]{11.8}.
+Somit ist $\mathcal{L}_{\bullet,x}\to\mathcal{F}_x\to 0$ eine freie Auflösung.
+Wir können diese Auflösung zur Berechnung der Kohomologie verwenden.
+Da $\operatorname{\mathcal{H}om}(\mathcal{L},\mathcal{G})_x=\operatorname{Hom}_{\mathcal{O}_{X,x}}(\mathcal{L}_x,\mathcal{G}_x)$ für lokal freie Garben $\mathcal{L}$ und Halmbildung exakt ist, folgt die Gleichheit der Ext-Gruppen.\qed

### 2015-06-13 Algebraische Geometrie/chapter3.tex
Ende "Ext-Gruppen"

@@
 Da $\operatorname{\mathcal{H}om}(\mathcal{L},\mathcal{G})_x=\operatorname{Hom}_{\mathcal{O}_{X,x}}(\mathcal{L}_x,\mathcal{G}_x)$ für lokal freie Garben $\mathcal{L}$ und Halmbildung exakt ist, folgt die Gleichheit der Ext-Gruppen.\qed
+
+\paragraph{Satz 11.12.}\label{11.12} Sei $X$ ein projektives Schema über einem noetherschen Ring $A$, $\mathcal{O}_X(1)$ eine sehr ample Garbe auf $X$ und $\mathcal{F},\mathcal{G}$ kohärente Garben auf $X$.
+Dann existiert ein $n_0=n_0(\mathcal{F},\mathcal{G})>0$, so dass für alle $n\geq n_0$ und alle $i\geq 0$ gilt:
+\[\operatorname{Ext}^i(\mathcal{F},\mathcal{G}(n)) =\Gamma(X,\operatorname{\mathcal{E}xt}^i(\mathcal{F},\mathcal{G}(n))) \]
+
+\paragraph{Lemma 11.13.}\label{11.13} Sei $X$ ein projektives Schema über einem noetherschen Ring $A$, $\mathcal{O}_X(1)$ eine sehr ample Garbe auf $X$ und $\mathcal{F}^1\to\mathcal{F}^2\to\cdots\to\mathcal{F}^r$ eine exakte Folge kohärenter Garben.
+Dann gibt es ein $m_0$, so dass für alle $m\geq m_0$ die folgende Sequenz exakt ist:
+\[\Gamma(X,\mathcal{F}^1(m))\longrightarrow\Gamma(X,\mathcal{F}^2(m))\longrightarrow\cdots\longrightarrow\Gamma(X,\mathcal{F}^r(m)) \]
+
+\paragraph{Beweis von \hyperref[11.12]{11.12}.} Für $i=0$ ist die Aussage klar.
+Sei zunächst $\mathcal{F}=\mathcal{O}_X$.
+Dann folgt aus \hyperref[11.5]{Satz 11.5} und \hyperref[10.3]{Satz 10.3} für alle $i>0$ und hinreichend großes $n$:
+\[\operatorname{Ext}^i(\mathcal{O}_X,\mathcal{G}(n))=\mathrm{H}^i(X,\mathcal{G}(n))=0=\Gamma(X,\operatorname{\mathcal{E}xt}^i(\mathcal{O}_X,\mathcal{G}(n))) \]
+Sei nun $\mathcal{F}$ lokal frei von endlichem Rang. Nach \hyperref[11.10]{Satz 11.10} gilt für hinreichend große $n$:
+\[\operatorname{Ext}^i(\mathcal{F},\mathcal{G}(n))=\operatorname{Ext}^i(\mathcal{O}_X,\mathcal{F}^\wedge\otimes\mathcal{G}(n))=0=\Gamma(X,\operatorname{\mathcal{E}xt}^i(\mathcal{O}_X,\mathcal{F}^\wedge\otimes\mathcal{G}(n)))=\Gamma(X,\operatorname{\mathcal{E}xt}^i(\mathcal{F},\mathcal{G}(n))) \]
+Sei nun $\mathcal{F}$ beliebig kohärent.
+Nach \hyperref[5.38]{Korollar 5.38} gibt es eine kurze exakte Folge $0\to\mathcal{R}\to\mathcal{L}\to\mathcal{F}\to 0$ mit einem lokal freien $\mathcal{L}$ von endlichem Rang.
+Wir erhalten für hinreichend große $n$ die folgende exakte Folge:
+\[0\longrightarrow\operatorname{Hom}(\mathcal{F},\mathcal{G}(n))\longrightarrow\operatorname{Hom}(\mathcal{L},\mathcal{G}(n))\longrightarrow\operatorname{Hom}(\mathcal{R},\mathcal{G}(n))\longrightarrow\operatorname{Ext}^1(\mathcal{F},\mathcal{G}(n))\longrightarrow 0 \]
+Analog für $\operatorname{\mathcal{H}om}$ und $\operatorname{\mathcal{E}xt}$:
+\[0\longrightarrow\operatorname{\mathcal{H}om}(\mathcal{F},\mathcal{G}(n))\longrightarrow\operatorname{\mathcal{H}om}(\mathcal{L},\mathcal{G}(n))\longrightarrow\operatorname{\mathcal{H}om}(\mathcal{R},\mathcal{G}(n))\longrightarrow\operatorname{\mathcal{E}xt}^1(\mathcal{F},\mathcal{G}(n))\longrightarrow 0\]
+Beachte $\operatorname{Ext}^j(\mathcal{R},\mathcal{G}(n))\cong\operatorname{Ext}^{j+1}(\mathcal{F},\mathcal{G}(n))$ und $\operatorname{\mathcal{E}xt}^j(\mathcal{R},\mathcal{G}(n))\cong\operatorname{\mathcal{E}xt}^{j+1}(\mathcal{F},\mathcal{G}(n))$ für $j\geq 1$.
+Da $\mathcal{R}$ kohärent ist, folgt die Aussage per Induktion über $j$.
+Für $j=0$ twisten wir die obige Folge genug und wenden $\Gamma(X,-)$ an.
+\hyperref[11.13]{Lemma 11.13} gibt uns ein $m$, so dass:
+\[\Gamma(X,\operatorname{\mathcal{E}xt}^1(\mathcal{F},\mathcal{G}(n+m))) =\operatorname{Ext}^1(\mathcal{F},\mathcal{G}(n+m))\qedhere \]

### 2015-06-13 Algebraische Geometrie/chapter3.tex
Konsistenz
Fix typo

@@
 Gibt es ein $x\in X\setminus U$, so existiert eine offene Umgebung $V$ von $x$ und $t\in\mathcal{G}(V)$ mit $t\mapsto s''|_V$, da $\mathcal{G}\to\mathcal{G}''$ surjektiv ist.
-Da $\mathcal{G}$ welk ist, so setzt sich $s-t\in\mathcal{G}(U\cap V)$ zu einem Schnitt auf $\mathcal{G}(V)$ fort.
-Ersetzen wir $t$ durch $s-t\in\mathcal{G}(V)$, können wir o.B.d.A.
+Da $\mathcal{G}'$ welk ist, so setzt sich $s-t\in\mathcal{G}'(U\cap V)$ zu einem Schnitt auf $\mathcal{G}'(V)$ fort.
+Ersetzen wir $t$ durch $s-t\in\mathcal{G}'(V)$, können wir o.B.d.A.
 $s=t$ auf $V$ annehmen.
@@
 
-\paragraph{Beweis von \hyperref[9.20]{Satz 9.20}.} Sei $0\to\mathcal{F}\to\mathcal{J}\to\mathcal{G}\to 0$ exakt mit $\mathcal{J}$ injektiv.
+\paragraph{Beweis von \hyperref[9.20]{9.20}.} Sei $0\to\mathcal{F}\to\mathcal{J}\to\mathcal{G}\to 0$ exakt mit $\mathcal{J}$ injektiv.
 Nun sind $\mathcal{F}$ und $\mathcal{J}$ welk.

### 2015-06-14 Algebraische Geometrie/chapter3.tex
Kleinere Fehler fixen

@@
 
-Die Aussagen (i) und (ii) gelten für Garben der Form $\mathcal{O}_X(q),\ q\in\mathbb{Z}$ nach \hyperref[10.1]{Satz 10.1}, dem \hyperref[9.27]{Verschwindungssatz} und der Tatsache:
-\[\mathrm{H}^r(X,\mathcal{O}_X(q)) \cong \operatorname{Hom}(\mathrm{H}^0(X,\mathcal{O}_X(-q-r-1)), A) =0\quad\text{für $-q-r-1<0$, d.h. q>-r-1} \]
+Die Aussagen (i) und (ii) gelten für Garben der Form $\mathcal{O}_X(q),\ q\in\mathbb{Z}$ nach \hyperref[10.1]{Satz 10.1}, dem \hyperref[9.27]{Verschwindungssatz} und der Tatsache, dass für $-q-r-1<0$, also $q>-r-1$, gilt:
+\[\mathrm{H}^r(X,\mathcal{O}_X(q)) \cong \operatorname{Hom}(\mathrm{H}^0(X,\mathcal{O}_X(-q-r-1)), A) =0 \]
 Es folgt (i) und (ii) für endliche direkte Summen von $\mathcal{O}_X(q),\ q\in\mathbb{Z}$.
@@
 Sei $\mathcal{F}$ ein $\mathcal{O}_X$"~Modul auf einem geringten Raum $(X,\mathcal{O}_X)$. Wir haben folgende linksexakte Funktoren:
-\[\operatorname{Hom}(\mathcal{F},-):\mathbf{Mod}(X)\to\mathbf{Ab} \]
-\[\operatorname{\mathcal{H}om}_{\mathcal{O}_X}(\mathcal{F},-):\mathbf{Mod}(X)\to\mathbf{Mod}(X) \]
+\[\operatorname{Hom}(\mathcal{F},-):\mathbf{Mod}(X)\to\mathbf{Ab},\quad\operatorname{\mathcal{H}om}(\mathcal{F},-):\mathbf{Mod}(X)\to\mathbf{Mod}(X) \]
 
 \paragraph{Definition 11.1.}\label{11.1} Sei $(X,\mathcal{O}_X)$ ein geringter Raum und $\mathcal{F}\in\mathbf{Mod}(X)$. Für $i\geq 0$ definieren wir:
-\[\operatorname{Ext}^i(\mathcal{F},-)=\mathrm{R}^i\operatorname{Hom}(\mathcal{F},-) \]
-\[\operatorname{\mathcal{E}xt}^i(\mathcal{F},-) =\mathrm{R}^i\operatorname{\mathcal{H}om}(\mathcal{F},-) \]
+\[\operatorname{Ext}^i(\mathcal{F},-)=\mathrm{R}^i\operatorname{Hom}(\mathcal{F},-),\quad\operatorname{\mathcal{E}xt}^i(\mathcal{F},-) =\mathrm{R}^i\operatorname{\mathcal{H}om}(\mathcal{F},-) \]
 
@@
 Dann ist die folgende Folge exakt:
-\[0\longrightarrow\operatorname{Hom}(\mathcal{F}'',\mathcal{G})\longrightarrow\operatorname{Hom}(\mathcal{F},\mathcal{G})\longrightarrow\mathcal{Hom}(\mathcal{F}')\longrightarrow\operatorname{Ext}^1(\mathcal{F}'',\mathcal{G})\longrightarrow\operatorname{Ext}^1(\mathcal{F},\mathcal{G})\longrightarrow\cdots\]
+\[0\to\operatorname{Hom}(\mathcal{F}'',\mathcal{G})\to\operatorname{Hom}(\mathcal{F},\mathcal{G})\to\operatorname{Hom}(\mathcal{F}',\mathcal{G})\to\operatorname{Ext}^1(\mathcal{F}'',\mathcal{G})\to\operatorname{Ext}^1(\mathcal{F},\mathcal{G})\to\cdots\]
 Analog für $\operatorname{\mathcal{H}om}$ und $\operatorname{\mathcal{E}xt}$.
@@
 \paragraph{Beweis.} Wir zeigen, dass $\operatorname{Hom}(-,\mathcal{J}\otimes\mathcal{L})$ exakter Funktor ist. Nun gilt:
-\[\operatorname{Hom}(-,\mathcal{J}\otimes\mathcal{L})=\operatorname{Hom}(-\otimes\mathcal{L}^\wedge,\mathcal{J})\]
-mit $\mathcal{L}^\wedge =\operatorname{Hom}(\mathcal{L},\mathcal{O}_X)$. Da $\mathcal{J}$ injektiv ist, folgt die Behauptung.\qed
+\[\operatorname{Hom}(-,\mathcal{J}\otimes\mathcal{L})=\operatorname{Hom}(-\otimes\mathcal{L}^\vee,\mathcal{J})\]
+mit $\mathcal{L}^\vee =\operatorname{Hom}(\mathcal{L},\mathcal{O}_X)$. Da $\mathcal{J}$ injektiv ist, folgt die Behauptung.\qed
 
-\paragraph{Satz 11.10.}\label{11.10} Sei $\mathcal{L}$ eine lokal freie Garbe von endlichem Rang und $\mathcal{L}^\wedge=\operatorname{Hom}(\mathcal{L},\mathcal{O}_X)$ sein Dual.
+\paragraph{Satz 11.10.}\label{11.10} Sei $\mathcal{L}$ eine lokal freie Garbe von endlichem Rang und $\mathcal{L}^\vee=\operatorname{Hom}(\mathcal{L},\mathcal{O}_X)$ sein Dual.
 Dann gilt für alle $\mathcal{F},\mathcal{G}\in\mathbf{Mod}(X)$:
 \begin{align*}
-\operatorname{Ext}^i(\mathcal{F}\otimes\mathcal{L},\mathcal{G}) &\cong \operatorname{Ext}^i(\mathcal{F},\mathcal{L}^\wedge\otimes\mathcal{G})\\
-\operatorname{\mathcal{E}xt}^i(\mathcal{F}\otimes\mathcal{L},\mathcal{G}) &\cong \operatorname{\mathcal{E}xt}^i(\mathcal{F},\mathcal{L}^\wedge\otimes\mathcal{G})
+\operatorname{Ext}^i(\mathcal{F}\otimes\mathcal{L},\mathcal{G}) &\cong \operatorname{Ext}^i(\mathcal{F},\mathcal{L}^\vee\otimes\mathcal{G})\\
+\operatorname{\mathcal{E}xt}^i(\mathcal{F}\otimes\mathcal{L},\mathcal{G}) &\cong \operatorname{\mathcal{E}xt}^i(\mathcal{F},\mathcal{L}^\vee\otimes\mathcal{G})
 \end{align*}
@@
 \paragraph{Beweis.} Für $i=0$ sind die Aussagen klar.
-Es ist $-\otimes\mathcal{L}^\wedge$ ein exakter Funktor.
+Es ist $-\otimes\mathcal{L}^\vee$ ein exakter Funktor.
 Somit sind beide Seiten $\delta$-Funktoren in $\mathcal{G}$, die auf Injektive $\mathcal{G}$ verschwinden und daher universell.\qed
@@
 \paragraph{Satz 11.11.}\label{11.11} Sei $X$ ein noethersches Schema, $\mathcal{F}$ eine kohärente Garbe auf $X$, $\mathcal{G}$ ein  $\mathcal{O}_X$-Modul und $x\in X$. Dann gilt:
-\[\operatorname{\mathcal{E}xt}^i(\mathcal{F},\mathcal{G})_x\cong \operatorname{Ext}^i_{\mathcal{O}_{X,x}}(\mathcal{F}_x,\mathcal{G}_x) = \mathrm{R}^i\operatorname{Hom}_{\mathcal{O}_{X,x}}(\mathcal{F}_x,\mathcal{G}_x) \]
+\[\operatorname{\mathcal{E}xt}^i(\mathcal{F},\mathcal{G})_x\cong \operatorname{Ext}^i_{\mathcal{O}_{X,x}}(\mathcal{F}_x,\mathcal{G}_x) = \mathrm{R}^i\operatorname{Hom}_{\mathcal{O}_{X,x}}(\mathcal{F}_x,-)(\mathcal{G}_x) \]
 
@@
 Sei nun $\mathcal{F}$ lokal frei von endlichem Rang. Nach \hyperref[11.10]{Satz 11.10} gilt für hinreichend große $n$:
-\[\operatorname{Ext}^i(\mathcal{F},\mathcal{G}(n))=\operatorname{Ext}^i(\mathcal{O}_X,\mathcal{F}^\wedge\otimes\mathcal{G}(n))=0=\Gamma(X,\operatorname{\mathcal{E}xt}^i(\mathcal{O}_X,\mathcal{F}^\wedge\otimes\mathcal{G}(n)))=\Gamma(X,\operatorname{\mathcal{E}xt}^i(\mathcal{F},\mathcal{G}(n))) \]
+\begin{align*}
+\operatorname{Ext}^i(\mathcal{F},\mathcal{G}(n))&=\operatorname{Ext}^i(\mathcal{O}_X,\mathcal{F}^\vee\otimes\mathcal{G}(n))=0\\
+&=\Gamma(X,\operatorname{\mathcal{E}xt}^i(\mathcal{O}_X,\mathcal{F}^\vee\otimes\mathcal{G}(n)))=\Gamma(X,\operatorname{\mathcal{E}xt}^i(\mathcal{F},\mathcal{G}(n)))
+\end{align*}
 Sei nun $\mathcal{F}$ beliebig kohärent.

### 2015-06-14 Algebraische Geometrie/chapter3.tex
Index hinzufügen: Ext-Gruppen
-⊗L^v ist exakter Funktor
Dual eines O_X-Moduls

@@
 
-\paragraph{Definition 11.1.}\label{11.1} Sei $(X,\mathcal{O}_X)$ ein geringter Raum und $\mathcal{F}\in\mathbf{Mod}(X)$. Für $i\geq 0$ definieren wir:
+\paragraph{Definition 11.1.}\label{11.1} Sei $(X,\mathcal{O}_X)$ ein geringter Raum und $\mathcal{F}\in\mathbf{Mod}(X)$. Für $i\geq 0$ definieren wir:\index{Ext-Gruppen}
 \[\operatorname{Ext}^i(\mathcal{F},-)=\mathrm{R}^i\operatorname{Hom}(\mathcal{F},-),\quad\operatorname{\mathcal{E}xt}^i(\mathcal{F},-) =\mathrm{R}^i\operatorname{\mathcal{H}om}(\mathcal{F},-) \]
@@
 \paragraph{Beweis.} Beide Seiten sind $\delta$-Funktoren $\mathbf{Mod}(X)\to\mathbf{Mod}(X)$ in $\mathcal{G}$, die gleich für $i=0$ sind.
-Beide Seiten verschwinden für injektive $\mathcal{G}$.
-Somit sind beide $\delta$-Funktoren universell.\qed
+Beide Seiten verschwinden für injektive $\mathcal{G}$, also sind beide $\delta$-Funktoren universell.\qed
 
@@
 \[\operatorname{Hom}(-,\mathcal{J}\otimes\mathcal{L})=\operatorname{Hom}(-\otimes\mathcal{L}^\vee,\mathcal{J})\]
-mit $\mathcal{L}^\vee =\operatorname{Hom}(\mathcal{L},\mathcal{O}_X)$. Da $\mathcal{J}$ injektiv ist, folgt die Behauptung.\qed
+mit $\mathcal{L}^\vee =\operatorname{\mathcal{H}om}(\mathcal{L},\mathcal{O}_X)$. Da $\mathcal{J}$ injektiv und $-\otimes\mathcal{L}^\vee$ exakt ist, folgt die Behauptung.\qed
 
-\paragraph{Satz 11.10.}\label{11.10} Sei $\mathcal{L}$ eine lokal freie Garbe von endlichem Rang und $\mathcal{L}^\vee=\operatorname{Hom}(\mathcal{L},\mathcal{O}_X)$ sein Dual.
+\paragraph{Satz 11.10.}\label{11.10} Sei $\mathcal{L}$ eine lokal freie Garbe von endlichem Rang und $\mathcal{L}^\vee=\operatorname{\mathcal{H}om}(\mathcal{L},\mathcal{O}_X)$ sein Dual.
 Dann gilt für alle $\mathcal{F},\mathcal{G}\in\mathbf{Mod}(X)$:

### 2015-06-19 Algebraische Geometrie/chapter3.tex
Einheitliche Bezeichnung für den Dualraum
Dualität für den projektiven Raum

@@
 eine perfekte Paarung freier $A$"~Moduln, d.h. sie induziert einen Isomorphismus:
-\[\mathrm{H}^0(X,\mathcal{O}_X(n))^\ast\cong \mathrm{H}^r(X,\mathcal{O}_X(-n-r-1)) \]
-Diese Paarung stimmt mit dem Cup-Produkt überein.
+\[\mathrm{H}^0(X,\mathcal{O}_X(n))'\cong \mathrm{H}^r(X,\mathcal{O}_X(-n-r-1)) \]
+wobei $'$ den Dualraum bezeichnet. Diese Paarung stimmt mit dem Cup-Produkt überein.
 \end{enumerate}
@@
 \[\Gamma(X,\operatorname{\mathcal{E}xt}^1(\mathcal{F},\mathcal{G}(n+m))) =\operatorname{Ext}^1(\mathcal{F},\mathcal{G}(n+m))\qedhere \]
+
+\section{Die Serre-Dualität}
+
+\paragraph{Satz 12.1.}\label{12.1} \textit{(Dualität für den projektiven Raum)}\index{Serre-Dualität} Sei $X=\mathbf{P}_k^n$ und $\omega_X=\bigwedge^n\Omega_{X/k}$ die $n$-te äußere Potenz von $\Omega_{X/k}$, die sogenannte \textit{kanonische Garbe}\index{kanonische Garbe} auf $X$.
+Dann ist $\omega_X$ eine invertierbare Garbe und es gilt:
+\begin{enumerate}[(i)]
+\item $\mathrm{H}^n(X,\omega_X)\cong k$
+\item Sei $\mathcal{F}$ kohärent auf $X$. Dann ist
+\[\operatorname{Hom}(\mathcal{F},\omega_X)\times\mathrm{H}^n(X,\mathcal{F})\to\mathrm{H}^n(X,\omega_X)\cong k \]
+eine perfekte Paarung endlich-dimensionierter $k$-Vektorräume.
+\item Für alle $i\geq 0$ gibt es eine natürliche Isomorphie von kontravarianten Funktoren $\mathbf{Coh}(X)\to\mathbf{Ab}$:
+\[\operatorname{Ext}^i(-,\omega_X)\cong\mathrm{H}^{n-i}(X,-)' \]
+wobei $'$ den Dualraum bezeichnet. Für $i=0$ ist diese gerade durh die Paarung von (ii) induziert.
+\end{enumerate}
+
+\paragraph{Beweis.} Da die Garbe der relativen Differentiale $\Omega_{X/k}$ lokal frei vom Rang $n$ ist, ist $\omega_X$ vom Rang $\binom{n}{n}=1$, also invertierbar.
+\begin{enumerate}[(i)]
+\item Aus der exakten Sequenz in \hyperref[8.23]{Satz 8.23}
+\[0\longrightarrow\Omega_{X/k}\longrightarrow\mathcal{O}_X(-1)^{n+1}\longrightarrow\mathcal{O}_X\longrightarrow 0 \]
+folgt $\omega_X=\bigwedge^n\Omega_{X/k}\cong\mathcal{O}_X(-n-1)$. Nach \hyperref[10.1]{Satz 10.1 (iii)} gilt $\mathrm{H}^n(X,\mathcal{O}_X(-n-1))\cong k$.
+\item Ein $f\in\operatorname{Hom}(\mathcal{F},\omega_X)$ liefert $\mathrm{H}^n(X,\mathcal{F})\to\mathrm{H}^n(X,\omega_X)$. Dies induziert eine Paarung:
+\[\operatorname{Hom}(\mathcal{F},\omega_X)\times\mathrm{H}^n(X,\mathcal{F})\to\mathrm{H}^n(X,\omega_X),\ (f,a)\mapsto \mathrm{H}^n(f)(a) \]
+Ist $\mathcal{F}\cong\mathcal{O}_X(q)$, so gilt $\operatorname{Hom}(\mathcal{O}_X(q),\omega_X)=\operatorname{Hom}(\mathcal{O}_X,\omega_X(-q))\cong\mathrm{H}^0(X,\omega_X(-q))$ nach \hyperref[11.5]{Satz 11.5 (iii)}.
+Betrachte das kommutative Diagramm:
+\[\begin{tikzcd}
+\mathrm{H}^0(X,\omega_X(-q)) \times \mathrm{H}^n(X,\mathcal{O}_X(q)) \rar\ar[d, shift left=4em, "\cong"]\ar[d, shift right=4em, "\cong"']& \mathrm{H}^n(X,\omega_X)\ar[d, "\cong"]\\
+\mathrm{H}^0(X,\mathcal{O}_X(-n-q-1))\times\mathrm{H}^n(X,\mathcal{O}_X(q))\rar &\mathrm{H}^n(X,\mathcal{O}_X(-n-1))
+\end{tikzcd} \]
+Daher ist die Paarung nach \hyperref[10.1]{Satz 10.1 (iv)} nicht ausgeartet. Die Aussage folgt daher auch für endliche direkte Summen von $\mathcal{O}_X(q)$.
+
+Sei nun $\mathcal{F}$ beliebig kohärent.
+Nach \hyperref[5.38]{Korollar 5.38} erhalten wir eine exakte Sequenz $\mathcal{E}_1\to\mathcal{E}_2\to\mathcal{F}\to 0$, wobei $\mathcal{E}_i$ endliche direkte Summen von $\mathcal{O}_X(q)$ ist.
+Sei $0\to\mathcal{G}'\to\mathcal{G}\to\mathcal{G}''\to 0$ eine exakte Sequenz kohärenter Garben.
+Nach dem Verschwindungssatz und unter Betrachtung der langen exakten Kohomologiefolge folgt die Exaktheit von:
+\[\mathrm{H}^n(X,\mathcal{G}')\longrightarrow\mathrm{H}^n(X,\mathcal{G})\longrightarrow\mathrm{H}^n(X,\mathcal{G}'')\longrightarrow 0\]
+Nun sind $\operatorname{Hom}(-,\omega_X)$ und $\operatorname{Hom}(\mathrm{H}^n(X,-),k)$ linksexakte kontravariante Funktoren. Also folgt die Behauptung aus dem Fünferlemma:
+\[\begin{tikzcd}
+0\rar & \operatorname{Hom}(\mathcal{F},\omega_X) \dar\rar& \operatorname{Hom}(\mathcal{E}_2,\omega_X) \rar\dar["\cong"']& \operatorname{Hom}(\mathcal{E}_1,\omega_X)\dar["\cong"]\\
+0 \rar& \mathrm{H}^n(X,\mathcal{F})'\rar & \mathrm{H}^n(X,\mathcal{E}_2)'\rar & \mathrm{H}^n(X,\mathcal{E}_1)'
+\end{tikzcd}\]
+\item Beide Seiten sind kontravariante $\delta$-Funktoren $\mathbf{Coh}(X)\to\mathbf{Ab}$.
+Für $i=0$ sind sie isomorph nach (ii).
+Also reicht es zu zeigen, dass beide $\delta$-Funktoren auslöschbar, also universell sind.
+Für $\mathcal{F}\in\mathbf{Coh}(X)$, existiert nach \hyperref[5.38]{Korollar 5.38} für alle hinreichend große $q$ eine Surjektion $\mathcal{E}\to\mathcal{F}$ mit $\mathcal{E}=\bigoplus_{i=1}^N\mathcal{O}_X(-q)$.
+Nach \hyperref[11.5]{Satz 11.5 (iii)} und \hyperref[10.3]{Satz 10.3} gilt für $i>0$ und hinreichend große $q$:
+\[\operatorname{Ext}^i(\mathcal{E},\omega_X) =\bigoplus_{i=1}^N\mathrm{H}^i(X,\omega_X(q))=0 \]
+Ferner gilt $\mathrm{H}^{n-i}(X,\mathcal{E})'=\bigoplus_{i=1}^N\mathrm{H}^{n-i}(X,\mathcal{O}_X(-q))'=0$ für $n>i>0$ nach \hyperref[10.1]{Satz 10.1}.
+Ist $i=n$, so gilt $\mathrm{H}^0(X,\mathcal{O}_X(-q))=0$ für alle $q>0$.\qed
+\end{enumerate}

### 2015-06-19 Algebraische Geometrie/chapter3.tex
Definition und Eindeutigkeit der dualisierenden Garbe

@@
 
-\paragraph{Satz 12.1.}\label{12.1} \textit{(Dualität für den projektiven Raum)}\index{Serre-Dualität} Sei $X=\mathbf{P}_k^n$ und $\omega_X=\bigwedge^n\Omega_{X/k}$ die $n$-te äußere Potenz von $\Omega_{X/k}$, die sogenannte \textit{kanonische Garbe}\index{kanonische Garbe} auf $X$.
+\paragraph{Satz 12.1.}\label{12.1} \textit{(Dualität für den projektiven Raum)}\index{Serre-Dualität} Sei $X=\mathbf{P}_k^n$ und $\omega_X=\bigwedge^n\Omega_{X/k}$ die $n$-te äußere Potenz von $\Omega_{X/k}$, die sogenannte \textit{kanonische Garbe}\index{Garbe!kanonisch} auf $X$.
 Dann ist $\omega_X$ eine invertierbare Garbe und es gilt:
@@
 \end{enumerate}
+
+\paragraph{Definition 12.2.}\label{12.2} Sei $X$ ein eigentliches Schema über einem Körper $k$ der Dimension $n$.
+Eine \textit{dualisierende Garbe}\index{Garbe!dualisierend} für $X$ ist eine kohärente Garbe $\omega_X^0$ auf $X$ zusammen mit einer \textit{Spurabbildung}\index{Spurabbildung}
+\[t:\mathrm{H}^n(X,\omega_X^0)\to k \]
+derart, dass für alle kohärenten Garben $\mathcal{F}$ auf $X$ die natürliche Paarung perfekt ist:
+\[\operatorname{Hom}(\mathcal{F},\omega_X^0) \times\mathrm{H}^n(X,\mathcal{F}) \longrightarrow\mathrm{H}^n(X,\omega_X^0)\stackrel{t}{\longrightarrow}k \]
+
+\paragraph{Satz 12.3.}\label{12.3} Sei $X$ eigentlich über einem Körper $k$.
+Seien $(\omega_X^0,t)$ und $(\omega_X^1,t')$ dualisierende Garben für $X$.
+Dann existiert einen eindeutig bestimmten Isomorphismus $\varphi:\omega_X^0\to\omega_X^1$, der mit der Spurabbildung verträglich ist, d.h.
+$t'\circ\mathrm{H}^n(\varphi)=t$.
+
+\paragraph{Beweis.} Da $\omega^1_X$ dualisierend und $\omega_X^0$ kohärent ist, folgt:
+\[\operatorname{Hom}(\mathrm{H}^n(X,\omega^0_X),k)\cong\operatorname{Hom}(\omega_X^0,\omega_X^1)\]
+Sei $\varphi$ das Bild von $t\in \operatorname{Hom}(\mathrm{H}^n(X,\omega^0_X),k)$ in $\operatorname{Hom}(\omega_X^0,\omega_X^1)$.
+Dann gilt $t'\circ\mathrm{H}^n(\varphi)=t$.
+Analog erhalten wir für $\omega_X^0$ dualisierend und $\omega_X^1$ kohärent einen Morphismus $\psi:\omega_X^1\to\omega_X^0$ mit $t\circ\mathrm{H}^n(\psi)=t'$.
+Es folgt $t\circ\mathrm{H}^n(\psi\circ\varphi)=t$.
+Da $\omega_X^0$ dualisierend ist, folgt $\psi\circ\varphi=\operatorname{id}$.
+Analog ist $\varphi\circ\psi=\operatorname{id}$.\qed

### 2015-06-19 Algebraische Geometrie/chapter3.tex
Bis 12.6

@@
 Analog ist $\varphi\circ\psi=\operatorname{id}$.\qed
+
+\paragraph{Lemma 12.4.}\label{12.4} Sei $X$ ein abgeschlossenes Unterschema von $P=\mathbf{P}_k^N$ der Kodimension $r$ und $j:X\hookrightarrow P$ die kanonische Inklusion. Dann gilt für alle $i<r$:
+\[\operatorname{\mathcal{E}xt}^i_P(j_\ast\mathcal{O}_X,\omega_P)=0 \]
+
+\paragraph{Lemma 12.5.}\label{12.5} Sei $X$ noethersch und $\mathcal{F},\mathcal{G}\in\mathbf{Coh}(X)$.
+Dann ist die Garbe $\operatorname{\mathcal{E}xt}^i(\mathcal{F},\mathcal{G})$ kohärent für alle $i\geq 0$.
+
+\paragraph{Beweis von \hyperref[12.4]{12.4}} Für alle $i$ ist nach \hyperref[12.5]{Lemma 12.5} $\mathcal{F}^i=\operatorname{\mathcal{E}xt}^i_P(j_\ast\mathcal{O}_X,\omega_P)$ kohärent auf $P$, da $j_\ast\mathcal{O}_X$ kohärent nach \hyperref[5.13]{Satz 5.13 (iii)} ist.
+Nach \hyperref[5.37]{Satz 5.37} wird $\mathcal{F}^i(q)$ von globalen Schnitten erzeugt für hinreichend große $q$.
+Somit reicht es zu zeigen, dass $\Gamma(P,\mathcal{F}^i(q))=0$ für hinreichend große $q$ und alle $i<r$.
+Nun gilt mit \hyperref[11.12]{Satz 11.12} und \hyperref[12.1]{Satz 12.1 (iii)} für hinreichend große $q$:
+\begin{align*}
+\Gamma(P,\mathcal{F}^i(q)) &\cong \operatorname{Ext}_P^i(j_\ast\mathcal{O}_X,\omega_P(q))\\
+&\cong \mathrm{H}^{N-i}(P,j_\ast\mathcal{O}_X(-q))' = \mathrm{H}^{N-i}(X,\mathcal{O}_X(-q))'
+\end{align*}
+Für $N-i>\dim(X)$ verschwinden diese nach dem Verschwindungssatz, d.h. $i<N-\dim(X)=r$.\qed
+
+\paragraph{Lemma 12.6.}\label{12.6} Sei $X$ ein abgeschlossenes Unterschema von $P=\mathbf{P}_k^N$ der Kodimension $r$ und $j:X\hookrightarrow P$ die kanonische Inklusion.
+Sei $\omega_X^0=\operatorname{\mathcal{E}xt}_P^r(j_\ast\mathcal{O}_X,\omega_P)$.
+Dann gibt es einen Isomorphismus von Funktoren $\mathbf{Mod}(X)\to\mathbf{Ab}$:
+\[\operatorname{Hom}(-,\omega_X^0)\cong\operatorname{Ext}_P^r(j_\ast-,\omega_P) \]

### 2015-06-20 Algebraische Geometrie/chapter3.tex
Beweis von 12.8

@@
 \paragraph{Lemma 12.6.}\label{12.6} Sei $X$ ein abgeschlossenes Unterschema von $P=\mathbf{P}_k^N$ der Kodimension $r$ und $j:X\hookrightarrow P$ die kanonische Inklusion.
-Sei $\omega_X^0=\operatorname{\mathcal{E}xt}_P^r(j_\ast\mathcal{O}_X,\omega_P)$.
+Sei $\omega_X^0=j^\ast\operatorname{\mathcal{E}xt}_P^r(j_\ast\mathcal{O}_X,\omega_P)$.
 Dann gibt es einen Isomorphismus von Funktoren $\mathbf{Mod}(X)\to\mathbf{Ab}$:
 \[\operatorname{Hom}(-,\omega_X^0)\cong\operatorname{Ext}_P^r(j_\ast-,\omega_P) \]
+
+\iffalse
+\paragraph{Beweis.} Sei $0\to\omega_P\to\mathcal{J}^\bullet$ eine injektive Auflösung von $\omega_P$ in $\mathbf{Mod}(P)$ und $\mathcal{F}\in\mathbf{Mod}(X)$. Dann gilt:
+\[\operatorname{Ext}_P^i(j_\ast\mathcal{F},\omega_P)=\mathrm{H}^i(\operatorname{Hom}_P(j_\ast\mathcal{F},\mathcal{J}^\bullet)) \]
+Es faktorisiert jeder Morphismus $j_\ast\mathcal{F}\to\mathcal{J}^i$ über $\mathcal{I}^i = \operatorname{\mathcal{H}om}_P(j_\ast\mathcal{O}_X,\mathcal{J}^i)$. Also gilt:
+\[\operatorname{Ext}_P^i(j_\ast\mathcal{F},\omega_P)=\mathrm{H}^i(\operatorname{Hom}_X(\mathcal{F},j^\ast\mathcal{I}^\bullet)) \]
+Nun sind die $j^\ast\mathcal{I}^i$ alle injektive $\mathcal{O}_X$"~Moduln, da $\operatorname{Hom}_X(-,j^\ast\mathcal{I}^i)=\operatorname{Hom}_P(j_\ast-,\mathcal{J}^i)$ exakt ist.
+Wegen \hyperref[12.4]{Lemma 12.4} gilt für $i<r$:
+\[\mathrm{H}^i(j^\ast\mathcal{I}^\bullet) = \mathrm{H}^i(j^\ast\operatorname{\mathcal{H}om}_P(j_\ast\mathcal{O}_X,\mathcal{J}^\bullet))=\operatorname{\mathcal{E}xt}^i_P(j_\ast\mathcal{O}_X,\omega_P)=0 \]
+Also ist der Komplex $\mathcal{I}^\bullet$ exakt bis zum $r$-ten Schritt.
+\fi
+
+\paragraph{Satz 12.7.}\label{12.7} Sei $X/k$ ein projektives Schema. Dann besitzt $X$ eine dualisierende Garbe.
+
+\iffalse
+\paragraph{Beweis.} Sei $j:X\hookrightarrow P=\mathbf{P}_k^N$ eine abgeschlossene Immersion, $n=\dim(X),\ r=\operatorname{codim}(X)$.
+Setze $\omega_X^0=j^\ast\operatorname{\mathcal{E}xt}^r_P(j_\ast\mathcal{O}_X,\omega_P)$.
+Nach \hyperref[12.6]{Lemma 12.6} gilt für alle $\mathcal{F}\in\mathbf{Mod}(X)$:
+\[\operatorname{Hom}_X(\mathcal{F},\omega_X^0) \cong\operatorname{Ext}_P^r(j_\ast\mathcal{F},\omega_P) \]
+Ist $\mathcal{F}$ kohärent, so gilt nach \hyperref[12.1]{Satz 12.1} und \hyperref[9.26]{Lemma 9.26}:
+\[\operatorname{Ext}_P^r(j_\ast\mathcal{F},\omega_P)\cong\mathrm{H}^{N-r}(P,j_\ast\mathcal{F})'\cong\mathrm{H}^n(X,\mathcal{F})' \]
+\fi
+
+\paragraph{Definition.} Sei $(A,\mathfrak{m})$ ein lokaler, noetherscher Ring und $M$ ein $A$"~Modul.
+$x_1,\ldots,x_r\in A$ heißt \textit{reguläre Sequenz}\index{reguläre Sequenz}, wenn $x_1$ kein Nullteiler von $M$ ist und für $i>1$ das Element $x_i$ kein Nullteiler von $M/(x_1,\ldots,x_{i-1})M$ ist.
+Die \textit{Tiefe}\index{Tiefe} von $M$ ist definiert als:
+\[\operatorname{depth}(M) = \sup\{r\mid \text{es existiert eine reguläre Sequenz }x_1,\ldots,x_r\in\mathfrak{m}\} \]
+$A$ heißt \textit{Cohen-Macaulay}\index{Cohen-Macaulay}, falls seine Tiefe gerade $\dim(A)$ ist.
+
+\paragraph{Satz 12.8.}\label{12.8} \textit{(Dualitätssatz für projektive Schemata)} Sei $X$ ein projektives Schema über einem algebraisch abgeschlossenen Körper $k$ der Dimension $n$.
+Sei $\omega_X^0$ die dualisierende Garbe auf $X$ und $\mathcal{O}(1)$ eine sehr ample Garbe auf $X$.
+Dann gilt:
+\begin{enumerate}[(i)]
+\item Für alle $i\geq 0$ und alle kohärenten Garben $\mathcal{F}$ auf $X$ gibt es natürliche, in $\mathcal{F}$ funktorielle Abbildungen:
+\[\theta^i:\operatorname{Ext}^i(\mathcal{F},\omega_X^0)\to\mathrm{H}^{n-i}(X,\mathcal{F})' \]
+wobei $\theta^0$ durch die Definition der dualisierende Garbe gegeben ist.
+\item Folgende Bedingungen sind äquivalent:
+\begin{enumerate}[(a)]
+\item $X$ ist \textit{Cohen-Macaulay}\index{Cohen-Macaulay}, d.h.
+alle lokalen Ringe sind Cohen-Macaulay, und \textit{äquidimensional}\index{äquidimensional}, d.h.
+alle irreduziblen Komponenten haben dieselbe Dimension.
+\item Für alle lokal freien Garben $\mathcal{F}$ auf $X$ gilt $\mathrm{H}^i(X,\mathcal{F}(-q))=0$ für alle $i<n$ und hinreichend große $q$.
+\item Alle $\theta^i,\ i\geq 0$ aus (i) sind Isomorphismen für alle $\mathcal{F}\in\mathbf{Coh}(X)$.
+\end{enumerate}
+\end{enumerate}
+
+\paragraph{Bemerkung 12.9.}\label{12.9} Ist $X/k$ nicht-singulär, so ist $X$ Cohen-Macaulay, da reguläre Ringe stets Cohen-Macaulay sind.
+
+\paragraph{Korollar 12.10.}\label{12.10} Sei $X/k$ ein projektives, äquidimensionales Cohen-Macaulay Schema der Dimension $n$ über einem algebraisch abgeschlossenen Körper $k$.
+Sei $\mathcal{F}$ eine lokal freie Garbe von endlichem Rang auf $X$.
+Dann gibt es für $i\geq 0$ natürliche Isomorphismen:
+\[\mathrm{H}^i(X,\mathcal{F})\cong\mathrm{H}^{n-i}(X,\mathcal{F}^\vee\otimes\omega_X^0)' \]
+
+\paragraph{Beweis.} Nach \hyperref[11.10]{Satz 11.10} und \hyperref[11.5]{Satz 11.5 (iii)} gilt:
+\[\mathrm{H}^i(X,\mathcal{F})\cong\operatorname{Ext}^{n-i}(\mathcal{F},\omega_X^0)'=\operatorname{Ext}^{n-i}(\mathcal{O}_X,\mathcal{F}^\vee\otimes\omega_X^0)' = \mathrm{H}^{n-i}(X,\mathcal{F}^\vee\otimes\omega_X^0)'\qedhere \]
+
+\paragraph{Korollar 12.11.}\label{12.11} Sei $X$ eine nicht-singuläre projektive Varietät der Dimension $n$ über einem abgeschlossenen Körper $k$.
+Dann ist die dualisierende Garbe $\omega_X^0$ isomorph zur kanonischen Garbe $\omega_X=\bigwedge^n\Omega_{X/k}$.
+
+\paragraph{Definition.} Sei $A$ ein Ring und $M$ ein $A$-Modul. Die \textit{projektive Dimension}\index{projektive Dimension} von $M$ ist definiert als:
+\[\operatorname{pd}_A(M)=\inf\{ m\mid\text{es gibt eine projektive Auflösung } 0\to P_m\to\cdots\to P_0\to M\to 0 \} \]
+
+\paragraph{Satz 12.12.}\label{12.12} Sei $A$ ein Ring und $M$ ein $A$-Modul. Dann gilt:
+\begin{enumerate}[(i)]
+\item $\operatorname{pd}_A(M)\leq m\iff \operatorname{Ext}^i(M,N)=0$ für $i>m$ und alle $A$-Moduln $N$.
+\item Ist $A$ regulär und $M$ endlich erzeugter $A$-Modul, so gilt:
+\[\operatorname{pd}_A(M)+\operatorname{depth}(M)=\dim(A) \]
+Es ist $\operatorname{pd}_A(M)\leq m$ genau dann, wenn $\operatorname{Ext}^i(M,A)=0$ für alle $i> m$.
+\end{enumerate}
+
+\paragraph{Beweis von \hyperref[12.8]{12.8}.}\begin{enumerate}[(i)]
+\item Sei $q$ hinreichend groß, so dass wir eine Surjektion $\mathcal{E}=\mathcal{O}_X(-q)^N\to\mathcal{F}$ erhalten.
+Nach \hyperref[11.5]{Satz 11.5 (iii)} und \hyperref[10.3]{Satz 10.3 (ii)} gilt für alle $i>0$ und hinreichend große $q$:
+\[\operatorname{Ext}^i(\mathcal{E},\omega_X^0)\cong\bigoplus_{i=1}^N\mathrm{H}^i(X,\omega_X^0(q))=0 \]
+Somit ist $\operatorname{Ext}^i(-,\omega_X^0)$ auslöschbar für $i>0$ und somit ein universeller, kontravarianter $\delta$-Funktor $\mathbf{Coh}(X)\to\mathbf{Ab}$.
+Da $\mathrm{H}^{n-i}(X,-)'$ auch kontravarianter $\delta$-Funktor ist, gibt es nach der Universaleigenschaft einen eindeutigen Morphismus $(\theta^i)_{i\geq 0}$ von $\delta$-Funktoren, wobei $\theta^0$ der vorgegebene ist.
+\item Sei (a) erfüllt.
+Sei $i:X\hookrightarrow P=\mathbf{P}_k^N$ eine abgeschlossene Immersion.
+Für alle lokal freien Garben $\mathcal{F}$ auf $X$ und jedem abgeschlossenen Punkt $x\in X$ gilt wegen Cohen-Macaulay und Äquidimensionalität:
+\[\operatorname{depth}(\mathcal{F}_x)=\dim(\mathcal{O}_{X,x})=n \]
+Sei $A=\mathcal{O}_{P,x}$. Dann ist $A$ ein regulärer lokaler Ring mit $\dim(A)=N$, da $P$ nicht-singulär ist. Nach \hyperref[12.12]{Satz 12.12 (ii)} gilt:
+\[\operatorname{pd}_A(i_\ast\mathcal{F}_x)=\dim(A) - \operatorname{depth}(\mathcal{F}_x)=N-n \]
+Aus \hyperref[11.11]{Satz 11.11} und \hyperref[12.12]{Satz 12.12 (ii)} folgt für $i>N-n$:
+\[\operatorname{\mathcal{E}xt}^i_P(i_\ast\mathcal{F},-)_x=\operatorname{Ext}^i_{\mathcal{O}_{P,x}}(i_\ast\mathcal{F}_x, -_x)=0 \]
+Für $i<n$ gilt somit $\operatorname{\mathcal{E}xt}^{N-i}_P(i_\ast\mathcal{F},-)=0$.
+Für hinreichend große $q$ folgt nach \hyperref[9.26]{Lemma 9.26}, \hyperref[12.1]{Satz 12.1} und \hyperref[11.12]{Satz 11.12}:
+\begin{align*}
+\mathrm{H}^i(X,\mathcal{F}(-q))' &= \mathrm{H}^i(P,i_\ast\mathcal{F}(-q))'\\
+&=\operatorname{Ext}_P^{N-i}(i_\ast\mathcal{F},\omega_P(q)) = \Gamma(P,\operatorname{\mathcal{E}xt}_P^{N-i}(i_\ast\mathcal{F},\omega_P(q)))=0
+\end{align*}
+Sei nun (b) erfüllt. Wie oben mit $\mathcal{F}=\mathcal{O}_X$ folgt $\operatorname{\mathcal{E}xt}^i_P(\mathcal{O}_X,\omega_P)=0$ für $i>N-n$. Da $\omega_P=\mathcal{O}_P(-N-1)$, gilt für $i>N-n$:
+\[\operatorname{\mathcal{E}xt}^i_P(\mathcal{O}_X,\omega_P)=\operatorname{\mathcal{E}xt}^i_P(\mathcal{O}_X,\mathcal{O}_P)(-N-1)=0\]
+Durch Halmbildung folgt $\operatorname{Ext}_A^i(\mathcal{O}_{X,x},A)=0$ für $A=\mathcal{O}_{P,x}$ und $i>N-n$.
+Somit gilt nach \hyperref[12.12]{Satz 12.12 (ii)} $\operatorname{pd}_A(\mathcal{O}_{X,x})\leq N-n$ und somit $\operatorname{depth}(\mathcal{O}_{X,x})\geq n=\dim(X)$.
+Da die andere Ungleichung stets gilt, folgt $\operatorname{depth}(\mathcal{O}_{X,x})=\dim(X)$ für alle abgeschlossenen Punkte $x\in X$.
+Somit ist $X$ Cohen-Macaulay, da Lokalisierungen von Cohen-Macaulay  Ringe bzgl.
+Primideale wieder Cohen-Macaulay sind.
+Es folgt (a).
+
+Sei (b) erfüllt.
+Da $\operatorname{Ext}^i(-,\omega_X^0)$ universeller $\delta$-Funktor ist, reicht es für (c) zu zeigen, dass auch $\mathrm{H}^{n-i}(X,-)'$ universeller $\delta$-Funktor ist.
+Wir zeigen die Auslöschbarkeit.
+Sei $\mathcal{F}$ kohärent und $\mathcal{E}=\mathcal{O}_X(-q)^N\to\mathcal{F}$ für hinreichend große $q$ eine Surjektion.
+Wegen (b) ist $\mathrm{H}^{n-i}(X,\mathcal{E})'=0$ für $i>0$ und hinreichend große $q$.
+
+Sei (c) erfüllt. Seien alle $\theta^i$ Isomorphismen und $\mathcal{F}$ lokal frei. Dann gilt nach \hyperref[10.3]{Satz 10.3} für $n-i>0$ und hinreichend große $q$:
+\[\mathrm{H}^i(X,\mathcal{F}(-q))\cong\operatorname{Ext}^{n-i}(\mathcal{F}(-q),\omega_X^0)'=\mathrm{H}^{n-i}(X,\mathcal{F}^\vee\otimes\omega_X^0(q))'=0\qedhere \]
+\end{enumerate}

### 2015-06-20 Algebraische Geometrie/chapter3.tex
Index Serre-Dualität hinzufügen

@@
 
-\paragraph{Satz 12.8.}\label{12.8} \textit{(Dualitätssatz für projektive Schemata)} Sei $X$ ein projektives Schema über einem algebraisch abgeschlossenen Körper $k$ der Dimension $n$.
+\paragraph{Satz 12.8.}\label{12.8} \textit{(Dualitätssatz für projektive Schemata)}\index{Serre-Dualität} Sei $X$ ein projektives Schema über einem algebraisch abgeschlossenen Körper $k$ der Dimension $n$.
 Sei $\omega_X^0$ die dualisierende Garbe auf $X$ und $\mathcal{O}(1)$ eine sehr ample Garbe auf $X$.

### 2015-06-20 Algebraische Geometrie/chapter3.tex
Anfang von "Der Satz von Riemann-Roch für Kurven"

@@
 \end{enumerate}
+
+\section{Der Satz von Riemann-Roch für Kurven}
+
+Eine Kurve ist ein eigentliches, nicht-singuläres, integres Schema $X$ der Dimension $1$ über einem algebraisch abgeschlossenen Körper $k$.
+Dann ist $X$ notwendigerweise projektiv, siehe z.B.
+Hartshorne II Proposition 6.7.
+
+\paragraph{Definition 13.1.}\label{13.1} Sei $X$ eine Kurve. Dann heißt
+\[g=g(X)=\dim_k\mathrm{H}^1(X,\mathcal{O}_X)=\dim_k\mathrm{H}^0(X,\omega_X)<\infty \]
+das \textit{Geschlecht}\index{Geschlecht} von $X$, wobei die letzte Gleichung aus der \hyperref[12.8]{Serre-Dualität} folgt.

### 2015-06-21 Algebraische Geometrie/chapter3.tex
Fix reference

@@
 Nach \hyperref[9.30]{Satz 9.30} sind $\widetilde{I_i}$ alle welk, daher sind die $f_{i\ast}\widetilde{I_i}$ welk und somit auch $\mathcal{G}$.
-Da alle $\widetilde{I_i}$ quasikohärent sind, sind auch alle $f_{i\ast}\widetilde{I_i}$ quasikohärent nach \hyperref[5.3]{Satz 5.3 (iv)} und somit ist auch $\mathcal{G}$ quasikohärent.\qed
+Da alle $\widetilde{I_i}$ quasikohärent sind, sind auch alle $f_{i\ast}\widetilde{I_i}$ quasikohärent nach \hyperref[5.13]{Satz 5.13 (iii)} und somit ist auch $\mathcal{G}$ quasikohärent.\qed
 

### 2015-06-21 Algebraische Geometrie/chapter2.tex
Kommutatives Diagramm in 8.19 verbessern
Fix
Füge fehlenden Satz 8.27 hinzu

@@
 \item Ist $0\to\mathcal{F}'\to\mathcal{F}\to\mathcal{F}''\to 0$ eine kurze exakte Sequenz von $\mathcal{O}_X$"~Moduln und $\mathcal{F}',\mathcal{F}''$ quasikohärent, so ist $\mathcal{F}$ quasikohärent.
-\item Ist $X$ noethersch, so gilt (ii) auch für kohärente Garben.
+\item Ist $X$ noethersch, so gilt (i) und (ii) auch für kohärente Garben.
 \end{enumerate}
@@
 X\times_YY' \rar["f'"]\dar["g'"'] & Y'\dar["g"]\\
-X\rar["f"] & Y
+X\rar["f"'] & Y
 \end{tikzcd} \]
@@
 
-\paragraph{Korollar 8.27.}\label{8.27} Sei $X$ eine nicht-singuläre Varietät über $k$ und $Y\subset X$ ein irreduzibles, abgeschlossenes Unterschema zur Idealgarbe $\mathcal{J}$.
+\paragraph{Korollar 8.27.}\label{8.27} Sei $X$ eine Varietät über $k$. Dann gibt es ein $U\subset_\text{o} X$, die nicht-singulär ist.
+
+\paragraph{Satz 8.28.}\label{8.28} Sei $X$ eine nicht-singuläre Varietät über $k$ und $Y\subset X$ ein irreduzibles, abgeschlossenes Unterschema zur Idealgarbe $\mathcal{J}$.
 Dann ist $Y$ genau dann nicht-singulär, wenn folgende Bedingungen erfüllt sind:

### 2015-06-26 Algebraische Geometrie/chapter3.tex
Definitionen zu Riemann-Roch

@@
 Hartshorne II Proposition 6.7.
+Da $X$ nicht-singulär ist, sind alle lokalen Ringe regulär und insbesondere faktoriell.
+Ferner stimmen die Weil-Divisoren mit den Cartier-Divisoren überein:
+\[\operatorname{Div}(X)\cong\Gamma(X,\mathcal{K}^\times/\mathcal{O}^\times_X)\]
+Wir haben einen Isomorphismus:
+\[\operatorname{Cl}(X)\cong\operatorname{Pic}(X) \]
 
@@
 das \textit{Geschlecht}\index{Geschlecht} von $X$, wobei die letzte Gleichung aus der \hyperref[12.8]{Serre-Dualität} folgt.
+
+\paragraph{Definition 13.2.}\label{13.2} \begin{enumerate}[(i)]
+\item Ein Divisor $D=\sum_i n_iP_i$ heißt \textit{effektiv}\index{Divisor!effektiv}, wenn $n_i\geq 0$ für alle $i$ gilt.
+\item Für einen Divisor $D_0$ heißt die Menge $|D_0|=\{D\in\operatorname{Div}(X)\mid D\sim D_0,\ D\text{ effektiv}\}$ \textit{vollständiges lineares System}\index{vollständiges lineares System} auf $X$.
+Sie kann evtl.
+leer sein.
+\end{enumerate}
+
+\paragraph{Bemerkung 13.3.}\label{13.3} Sei $D$ ein Divisor auf einer nicht-singulären, projektiven Varietät $X$ über einem algebraisch abgeschlossenen Körper $k$. Dann ist
+\[(\mathrm{H}^0(X,\mathcal{L}(D))\setminus\{0\})/k^\times\to |D|,\ s\mapsto (s)_0 \]
+eine Bijektion von Mengen, wobei $(s)_0$ den Nullstellendivisor von $s$ bezeichnet. Siehe z.B. Hartshorne II Proposition 7.7.
+
+\paragraph{Definition 13.4.}\label{13.4} Sei $D\in\operatorname{Div}(X)$. Wir definieren
+\[\ell(D)=\dim_k\mathrm{H}^0(X,\mathcal{L}(D)) \]
+und bezeichnen $\ell(D)-1$ als die \textit{Dimension}\index{Dimension} von $|D|$. Nach \hyperref[10.3]{Satz 10.3} ist $\ell(D)<\infty$.

### 2015-06-26 Algebraische Geometrie/chapter3.tex
Füge dualisierende Garbe für Kurven hinzu
Spezifiziere Isomorphismus Cl(X) = Pic(X)
Riemann-Roch

@@
 Wir haben einen Isomorphismus:
-\[\operatorname{Cl}(X)\cong\operatorname{Pic}(X) \]
+\[\operatorname{Cl}(X)\stackrel{\sim}{\to}\operatorname{Pic}(X),\ [D]\mapsto [\mathcal{L}(D)] \]
+Es gibt nach \hyperref[12.11]{Korollar 12.11} eine dualisierende Garbe $\omega_X=\Omega_{X/k}$. Diese ist invertierbar.
 
@@
 und bezeichnen $\ell(D)-1$ als die \textit{Dimension}\index{Dimension} von $|D|$. Nach \hyperref[10.3]{Satz 10.3} ist $\ell(D)<\infty$.
+
+\paragraph{Lemma 13.5.}\label{13.5} Sei $X$ eine Kurve und $D\in\operatorname{Div}(X)$. Dann gilt:
+\begin{enumerate}[(i)]
+\item $\ell(D)>0\implies\deg(D)\geq 0$
+\item Sind $\ell(D)>0$ und $\deg(D)=0$, so folgt $D\sim 0$, also $\mathcal{L}(D)\cong\mathcal{O}_X$.
+\end{enumerate}
+
+\paragraph{Beweis.} Ist $\ell(D)>0$, so ist $|D|\neq\varnothing$.
+Sei $D\sim D'$ mit effektivem $D'$.
+Dann ist $\deg(D)=\deg(D')\geq 0$.
+Ist zusätzlich $\deg(D)=0$, so auch $\deg(D')=0$.
+Da $D'$ effektiv ist, folgt $D'=0$.\qed
+
+\paragraph{Definition 13.6.}\label{13.6} Ein Divisor $K\in\operatorname{Div}(X)$ heißt \textit{kanonischer Divisor}\index{Divisor!kanonisch}, falls $K\sim D$ mit $\mathcal{L}(D)\cong\omega_X$.
+
+\paragraph{Definition.} Sei $X$ ein projektives Schema über einem Körper $k$ und $\mathcal{F}\in\mathbf{Coh}(X)$.
+Die \textit{Euler-Charakteristik}\index{Euler-Charakteristik} von $\mathcal{F}$ ist definiert als:
+\[\chi(\mathcal{F})=\sum_{i\geq 0}(-1)^i\dim_k\mathrm{H}^i(X,\mathcal{F}) \]
+
+\paragraph{Satz 13.7.}\label{13.7} \textit{(Riemann-Roch)} Sei $X$ eine Kurve vom Geschlecht $g$ und $D\in\operatorname{Div}(X)$. Dann gilt für einen kanonischen Divisor $K$:
+\[\ell(D)-\ell(K-D) = \deg(D)+1-g \]
+
+\paragraph{Beweis.} Der Divisor $K-D$ gehört zur invertierbaren Garbe $\omega_X\otimes\mathcal{L}(D)^\vee$. \hyperref[12.10]{Serre-Dualität} liefert nun:
+\[\mathrm{H}^0(X,\omega\otimes\mathcal{L}(D)^\vee)' \cong\mathrm{H}^1(X,\mathcal{L}(D)) \]
+Es ist somit zu zeigen $\chi(\mathcal{L}(D))=\deg(D)+1-g$. Sei zunächst $D=0$. Dann gilt:
+\[\mathrm{H}^0(X,\mathcal{L}(D))=\mathrm{H}^0(X,\mathcal{O}_X)=k \]
+Also $\dim_k\mathrm{H}^0(X,\mathcal{L}(D))=1$. Per Definition gilt $\dim_k\mathrm{H}^1(X,\mathcal{L}(D))=\dim_k\mathrm{H}^1(X,\mathcal{O}_X)=g$. Somit folgt $\chi(\mathcal{O}_X)=0+1-g$.
+
+Sei nun $D\in\operatorname{Div}(X)$ mit $\chi(\mathcal{L}(D))=\deg(D)+1-g$ und $P\in X$ ein abgeschlossener Punkt.
+Sei $\{P\}\subset X$ das zu $P$ gehörige abgeschlossene Unterschema von $X$ mit konstanter Strukturgarbe $k(P)\cong k$.
+Sei $\mathcal{J}_P$ die zughörige Idealgarbe von $\{P\}$.
+Es ist $\mathcal{J}_P=\mathcal{L}(-P)$.
+Nun ist $j_\ast k(P)$ eine Wolkenkratzergarbe auf $X$ mit Support $\{P\}$, wobei $j:\{P\}\hookrightarrow X$ die kanonische Inklusion bezeichnet.
+Wir erhalten somit eine exakte Sequenz:
+\[0\longrightarrow\mathcal{L}(-P)\longrightarrow\mathcal{O}_X\longrightarrow j_\ast k(P)\longrightarrow 0 \]
+Tensorieren mit der lokal freien Garbe $\mathcal{L}(D+P)$ ergibt die exakte Sequenz:
+\[0\longrightarrow\mathcal{L}(D)\longrightarrow\mathcal{L}(D+P)\longrightarrow j_\ast k(P)\longrightarrow 0 \]
+da $j_\ast k(P)\otimes\mathcal{L}(D+P)=j_\ast k(P)$. Es folgt:
+\[\chi(\mathcal{L}(D+P)) = \chi(\mathcal{L}(D)) + \chi(k(P))=\chi(\mathcal{L}(D))+1 = \deg(D)+1 = \deg(D+P) \]
+Da jeder Divisor von $0$ aus aufgebaut werden kann, folgt die Aussage.\qed

### 2015-06-26 Algebraische Geometrie/chapter3.tex
Korollare zum Riemann-Roch

@@
 Da jeder Divisor von $0$ aus aufgebaut werden kann, folgt die Aussage.\qed
+
+\paragraph{Korollar 13.8.}\label{13.8} Sei $X$ eine Kurve vom Geschlecht $g$ und $D\in\operatorname{Div}(X)$. Für einen kanonischen Divisor $K$ gilt:
+\begin{enumerate}[(i)]
+\item $\ell(K)=g$, $\deg(K)=2g-2$
+\item Ist $\deg(D)>2g-2$, so folgt $\ell(D)=\deg(D)+1-g$. Für eine elliptische Kurve gilt insbesondere $\ell(D)=\deg(D)$ für $\deg(D)>0$.
+\end{enumerate}
+
+\paragraph{Beweis.}\begin{enumerate}[(i)]
+\item Für $D=0$ bzw. $D=K$ folgt aus \hyperref[13.7]{Satz 13.7}:
+\[\ell(K)=-\deg(D)+g-1+\ell(D)=g,\quad \deg(K)=\ell(K)-\ell(0)+g-1=2g-2 \]
+\item Sei $D'\in\operatorname{Div}(X)$ mit $\deg(D')<0$. Nach \hyperref[13.5]{Lemma 13.5} folgt $\ell(D')\leq 0$, also $\ell(D')=0$. Wegen $\deg(K-D)=2g-2-\deg(D)<0$ folgt $\ell(K-D)=0$.\qed
+\end{enumerate}
+
+\paragraph{Beispiel 13.9.}\label{13.9} Sei $X$ eine elliptische Kurve und $P_0\in X$ ein abgeschlossener Punkt. Betrachte die Untergruppe:
+\[\operatorname{Pic}^0(X)=\{[\mathcal{L}(D)]\in\operatorname{Pic}(X)\mid\deg(D)=0 \} \]
+Dann ist die folgende Abbildung eine Bijektion:
+\[X_0\stackrel{\sim}{\to}\operatorname{Pic}^0(X),\ P\mapsto [\mathcal{L}(P-P_0)] \]
+wobei $X_0\subset X$ die Teilmenge aller abgeschlossenen Punkte von $X$ bezeichnet. Wir erhalten durch Strukturtransport eine Gruppenstruktur auf $X_0$ mit $P_0$ als neutralem Element.
+
+\paragraph{Beweis.} Sei $D\in\operatorname{Div}(X)$ mit $\deg(D)=0$. Riemann-Roch liefert:
+\[\ell(D+P_0)-\ell(K-D-P_0)=\deg(D+P_0)+1-g=1 \]
+Wegen $\deg(K)=2g-2=0$ ist $\deg(K-D-P_0)=-1$.
+Nach \hyperref[13.5]{Lemma 13.5} folgt $\ell(K-D-P_0)=0$, also $\ell(D+P_0)=1$ und die Dimension von $|D+P_0|$ ist $0$.
+Daher existiert genau ein effektiver Divisor $D'\sim D+P_0$.
+Wegen $\deg(D')=\deg(D+P_0)=1$ ist $D'=P$.\qed

### 2015-07-06 Algebraische Geometrie/chapter2.tex
Fix #5 & other minor typos

@@
 \paragraph{Beweis.} \begin{enumerate}[(i)]
-\item Nach \hyperref[5.35]{5.35} erzeugen die globalen Schnitte $x_0,\ldots,x_n$ die Garbe $\mathcal{O}(1)$.
+\item Nach \hyperref[5.25]{Satz 5.25} erzeugen die globalen Schnitte $x_0,\ldots,x_n$ die Garbe $\mathcal{O}(1)$.
 Ferner ist $\varphi^\ast(\mathcal{O}(1))$ invertierbar und werden von $\varphi^\ast(x_i)=s_i$ erzeugt.
@@
 
-\paragraph{Satz 7.2.}\label{7.2} Sei $\varphi:X\to\mathbf{P}_A^n$ ein $A$"~Morphismus, der zu einer invertierbaren Garbe $Z$ auf $X$ und globale Schnitte $s_0,\ldots,x_n\in\Gamma(X,Z)$ gehört, die $\mathcal{L}$ erzeugen.
+\paragraph{Satz 7.2.}\label{7.2} Sei $\varphi:X\to\mathbf{P}_A^n$ ein $A$"~Morphismus, der zu einer invertierbaren Garbe $\mathcal{L}$ auf $X$ und globale Schnitte $s_0,\ldots,s_n\in\Gamma(X,\mathcal{L})$ gehört, die $\mathcal{L}$ erzeugen.
 Dann ist $\varphi$ genau dann eine abgeschlossene Immersion, wenn die folgenden Bedingungen gelten:
@@
 Sei $\mathcal{L}$ eine invertierbare Garbe auf $X$ und $\mathcal{S}'$ die folgende Garbe graduierter Algebren:
-\[\mathcal{S}_d' =\mathcal{S}_1\otimes\mathcal{L}^d,\quad d\geq 0 \]
+\[\mathcal{S}_d' =\mathcal{S}_d\otimes\mathcal{L}^d,\quad d\geq 0 \]
 Dann erfüllt $\mathcal{S}'$ ebenfalls $(\dagger)$ und es gibt einen natürlichen Isomorphismus:
@@
 
-Alternativ: Betrachte die Idealgarbe $f^{-1}\mathcal{J}$ auf $f^{-1}\mathcal{O}_Y$ und den Morphismus $f^{-1}\mathcal{O}_Y\to \mathcal{O}_Y$, der zu $f^\sharp$ adjungiert ist.
+Alternativ: Betrachte die Idealgarbe $f^{-1}\mathcal{J}$ auf $f^{-1}\mathcal{O}_Y$ und den Morphismus $f^{-1}\mathcal{O}_Y\to \mathcal{O}_X$, der zu $f^\sharp$ adjungiert ist.
 Dann ist $\mathcal{J}'$ die vom Bild von $f^{-1}\mathcal{J}$ in $\mathcal{O}_X$ erzeugte Idealgarbe.
@@
 \begin{enumerate}[(i)]
-\item Die inverse Idealgarbe $\widetilde{J}=\pi^{-1}\mathcal{J}\cdot\mathcal{O}_{\widetilde{X}}$ ist invertierbar auf $\mathcal{O}_{\widetilde{X}}$.
+\item Die inverse Idealgarbe $\widetilde{\mathcal{J}}=\pi^{-1}\mathcal{J}\cdot\mathcal{O}_{\widetilde{X}}$ ist invertierbar auf $\mathcal{O}_{\widetilde{X}}$.
 \item Sei $Y$ das zu $\mathcal{J}$ gehörige abgeschlossene Unterschema und $U=X\setminus Y$. Dann ist $\pi:\pi^{-1}(U)\to U$ ein Isomorphismus.

### 2015-07-15 Master Thesis/thesis.tex
Add master thesis structure

\documentclass[11pt,a4paper,openany]{memoir}
\usepackage{fontspec}
\usepackage[english]{babel}
\usepackage{amsmath, amsfonts, amssymb, amsthm}
\usepackage{paralist}
\usepackage{tikz-cd}
\linespread{1.1}
\usepackage{hyperref}
\hypersetup{
	bookmarksopen=true,
	pdfstartview=FitH
}
\usepackage[inner=4cm,outer=3cm,top=3cm,bottom=3cm]{geometry}
\author{Yichuan Shen}
\title{Title of the Master Thesis}
\makeindex

\begin{document}

\theoremstyle{plain}
\theoremstyle{definition}
\newtheorem{theorem}{Theorem}[chapter]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{corollary}[theorem]{Corollary}
\theoremstyle{definition}
\newtheorem*{definition}{Definition}
\newtheorem*{example}{Example}
\theoremstyle{remark}
\newtheorem*{remark}{Remark}

\frontmatter
\pagenumbering{gobble}

\makeatletter
\begin{center}
\vspace*{0cm}
\begin{large}
Universität Heidelberg\\
Fakultät für Mathematik und Informatik\\
Studiengang Mathematik\\
\vspace{8mm}
\end{large}
\vfill 
\begin{large}
\textbf{MASTER THESIS}\\
\end{large}
\vspace{10mm}
\begin{huge}\@title \end{huge}\\
\vspace{10mm}
by \textsc{yichuan shen}\\
\vspace{3cm}
\vfill
Supervisor: \textsc{prof. dr. alexander schmidt}\\
\today
\end{center}
\makeatother

\clearpage
\ 
\clearpage

\vspace*{0cm}
\vfill
\paragraph{Declaration.} I declare that I carried out this master thesis independently, and only with the cited sources, literature and other professional sources. The master thesis was not used in the same or in a similar version to achieve an academic grading or is being published elsewhere.\\
\vspace{7mm}\\
\rule{7cm}{0.4pt}
\clearpage

\vspace*{0cm}
\vfill
\renewcommand{\abstractname}{Abstract}
\begin{abstract}
Lorem ipsum dolor sit amet, consectetur adipiscing elit. Curabitur interdum, dui non molestie tincidunt, lectus enim aliquet arcu, suscipit pulvinar dui tortor vitae mi. Nam cursus mi sem, ac gravida tellus congue vitae. Nam elementum eleifend lorem dignissim rhoncus. Quisque ac nulla dui. Duis rutrum eros ornare faucibus congue.
\end{abstract}

\vspace{1cm}

\renewcommand{\abstractname}{Zusammenfassung}
\begin{abstract}
Integer congue euismod justo, quis venenatis tortor. Proin sem leo, accumsan eget pulvinar vitae, tincidunt ut elit. Morbi ut purus volutpat, efficitur quam id, consequat augue. Cras ullamcorper lacus eget massa vehicula, a elementum diam lacinia. Morbi efficitur elementum malesuada. Cras a aliquam tortor, et interdum neque.
\end{abstract}
\vfill
\clearpage

\mainmatter
\pagenumbering{arabic} 
\setcounter{page}{5}
\tableofcontents

\chapter{Introduction}

\clearpage

\section{Notation}

$\begin{array}{ll}
\mathbb{N} & \text{Natural numbers} \\
\mathbb{Z} & \text{Integers}\\
\mathbb{Q} & \text{Rational numbers}\\
\mathbb{R} & \text{Real numbers}\\
\mathbb{C} & \text{Complex numbers}\\
\mathbb{Z}_p & \text{$p$-adic integers}\\
\mathbb{Q}_p & \text{$p$-adic numbers}
\end{array}$

\chapter{First chapter}

\begin{thebibliography}{99999999}

\bibitem[DDMS99]{DDMS99}
	J. D. Dixon, M. P. F. du Sautoy, A. Mann, D. Segal (1999)
	\textit{Analytic Pro-$p$ Groups}.
	Cambridge Studies in Advanced Mathematics,
	2nd Edition.

\end{thebibliography}

%\printindex
\end{document}

### 2015-07-15 Master Thesis/thesis.tex
Add main result

@@
 
-\mainmatter
 \pagenumbering{arabic} 
@@
 \chapter{Introduction}
+
+\begin{theorem}[\textit{Pop}]\label{thm:main-result}
+Let $K,K'$ be two finite field extensions of $\mathbb{Q}(T)$.
+If there is an isomorphism $\sigma:G_K\to G_{K'}$ between their absolute Galois groups, then there exists a unique field isomorphism $\varphi:\overline{K'}\to\overline{K}$ with $\sigma(g)=\varphi^{-1}g\varphi$ for all $g\in G_K$.
+In particular $\varphi$ induces a field isomorphism $K'\to K$.
+\end{theorem}
+
+\begin{theorem}[\textit{Neukirch-Uchida}]\label{thm:neukirch-uchida}
+Let $k,k'$ be two global fields.
+Then any isomorphism $G_k\to G_{k'}$ is induced by a unique isomorphism $k'^s\to k^s$.
+In particular, if $k,k'$ have isomorphic absolute Galois groups, they are isomorphic as fields.
+\end{theorem}
 

### 2015-07-15 Master Thesis/thesis.tex
Update bibliography
Update notation

@@
 Let $k,k'$ be two global fields.
-Then any isomorphism $G_k\to G_{k'}$ is induced by a unique isomorphism $k'^s\to k^s$.
+Then any isomorphism $G_k\to G_{k'}$ is induced by a unique isomorphism $k'^{\text{s}}\to k^\text{s}$.
 In particular, if $k,k'$ have isomorphic absolute Galois groups, they are isomorphic as fields.
@@
 $\begin{array}{ll}
-\mathbb{N} & \text{Natural numbers} \\
-\mathbb{Z} & \text{Integers}\\
-\mathbb{Q} & \text{Rational numbers}\\
-\mathbb{R} & \text{Real numbers}\\
-\mathbb{C} & \text{Complex numbers}\\
+\mathbb{N} & \text{natural numbers} \\
+\mathbb{Z} & \text{integers}\\
+\mathbb{Q} & \text{rational numbers}\\
+\mathbb{R} & \text{real numbers}\\
+\mathbb{C} & \text{complex numbers}\\
 \mathbb{Z}_p & \text{$p$-adic integers}\\
-\mathbb{Q}_p & \text{$p$-adic numbers}
+\mathbb{Q}_p & \text{$p$-adic numbers}\\
+\\
+\overline{K} & \text{algebraic closure of $K$}\\
+K^\text{s} & \text{separable closure of $K$}\\
+G_K & \text{absolute Galois group of $K$}
 \end{array}$
 
+\mainmatter
+\setcounter{page}{9}
 \chapter{First chapter}
 
-\begin{thebibliography}{99999999}
+\begin{thebibliography}{99999}
 
-\bibitem[DDMS99]{DDMS99}
-	J. D. Dixon, M. P. F. du Sautoy, A. Mann, D. Segal (1999)
-	\textit{Analytic Pro-$p$ Groups}.
-	Cambridge Studies in Advanced Mathematics,
-	2nd Edition.
+\bibitem[Sp96]{Sp96}
+	M. Spiess (1996)
+	\textit{An arithmetic proof of Pop's Theorem concerning Galois groups of function fields over number fields}.
+    J. reine angew. Math. \textbf{478}, 107\---126.
 

### 2015-07-15 Master Thesis/synopsis.tex
Add synopsis

\chapter{Synopsis}

\section{Some results on $\operatorname{Spv}(K)$}

\begin{definition}
Let $K$ be a field and $P$ its prime field. The \textit{(Kronecker) dimension}\index{dimension!Kronecker} $\dim(K)$ of $K$ is defined as:
\[\dim(K) = \begin{cases}
\operatorname{trdeg}(K/P), & \operatorname{char}(K)>0\\
\operatorname{trdeg}(K/P)+1, & \text{otherwise}
\end{cases} \]
\end{definition}

\begin{definition}
Let $K$ be a field of finite dimension. We denote the set of equivalence classes of non-archimedean valuations of $K$ by $\operatorname{Spv}(K)$. We often do not distinguish between a valuation and its equivalence class.

We denote the valuation ring of $v\in\operatorname{Spv}(K)$ by $(\mathcal{O}_v,\mathfrak{m}_v)$ and set $U_v=\mathcal{O}_v^\times$ and $k(v)=\mathcal{O}_v/\mathfrak{m}_v$. The \textit{rank}\index{valuation!rank} of $v$ is defined as the Krull dimension of its valuation ring $\mathcal{O}_v$. The \textit{rational rank}\index{valuation!rational rank} of $v:K\to\Gamma_v$ is defined as the rank of its valuation group $\Gamma_v$ as a $\mathbb{Z}$-module:
\[\operatorname{rrank}(v) = \dim_\mathbb{Q}\Gamma_v\otimes_\mathbb{Z}\mathbb{Q} \]
\end{definition}

\begin{lemma}
Let $v\in\operatorname{Spv}(K)$. Then we have:
\[\operatorname{rank}(v)\leq \operatorname{rrank}(v)\quad\text{and}\quad\operatorname{rrank}(v)+\dim k(v)\leq\dim(K) \]
\end{lemma}

\begin{lemma}
Let $K$ be a finitely generated field and $v\in\operatorname{Spv}(K)$ with:
\[\operatorname{rrank}(v)+\dim k(v)= \dim(K) \]
Then $\Gamma_v$ is a finitely generated $\mathbb{Z}$-module and $k(v)$ is a finitely generated field.
\end{lemma}

\begin{definition}
We define:
\begin{align*}
\operatorname{Spv}_\text{d}(K) &= \{v\in\operatorname{Spv}(K)\mid\operatorname{rank}(v)=\operatorname{rrank}(v) \} \\
\operatorname{Spv}_\text{rd}(K) &= \{v\in\operatorname{Spv}_\text{d}(K)\mid \operatorname{rrank}(v)+\dim k(v)=\dim(K) \}
\end{align*}
Valuations in $\operatorname{Spv}_\text{d}$ are called \textit{defectless}\index{valuation!defectless}, those in $\operatorname{Spv}_\text{rd}$ are called \textit{rank-defectless}\index{valuation!rank-defectless}. For $\ast\in\{\text{d},\text{rd}\}$ we set:
\[\operatorname{Spv}_\ast^{(i)}(K)=\{v\in\operatorname{Spv}_\ast(K)\mid \operatorname{rrank}(v)=i \} \]
\end{definition}

\begin{definition}
We can endow $\operatorname{Spv}(K)$ with a natural topology by declaring following sets as a basis of open sets:
\[\{v\in\operatorname{Spv}(K)\mid v(x)\geq 0\text{ for all }x\in S\} \]
where $S$ ranges over all finite subsets of $K$. 
\end{definition}

\begin{lemma}
$\operatorname{Spv}(K)$ with the topology defined above is a quasicompact $T_0$-space. Every closed subset $A\subset\operatorname{Spv}(K)$ is the closure of its closed points. In particular the set of closed points $|\operatorname{Spv}(K)|$ of $\operatorname{Spv}(K)$ is a quasicompact $T_1$-space, endowed with the induced topology.
\end{lemma}

\begin{lemma}
Let $v,w\in\operatorname{Spv}(K)$. We have:
\[\mathcal{O}_w\subset\mathcal{O}_v \iff w\in\overline{\{v\}} \]
Thus $v$ is a closed point if and only if $\dim k(v)=0$.
\end{lemma}

### 2015-07-15 Master Thesis/introduction.tex
Pop introduction

--- 2015-07-15 Master Thesis/thesis.tex
@@
-\documentclass[11pt,a4paper,openany]{memoir}
-\usepackage{fontspec}
-\usepackage[english]{babel}
-\usepackage{amsmath, amsfonts, amssymb, amsthm}
-\usepackage{paralist}
-\usepackage{tikz-cd}
-\linespread{1.1}
-\usepackage{hyperref}
-\hypersetup{
-	bookmarksopen=true,
-	pdfstartview=FitH
-}
-\usepackage[inner=4cm,outer=3cm,top=3cm,bottom=3cm]{geometry}
-\author{Yichuan Shen}
-\title{Title of the Master Thesis}
-\makeindex
-
-\begin{document}
-
-\theoremstyle{plain}
-\theoremstyle{definition}
-\newtheorem{theorem}{Theorem}[chapter]
-\newtheorem{lemma}[theorem]{Lemma}
-\newtheorem{proposition}[theorem]{Proposition}
-\newtheorem{corollary}[theorem]{Corollary}
-\theoremstyle{definition}
-\newtheorem*{definition}{Definition}
-\newtheorem*{example}{Example}
-\theoremstyle{remark}
-\newtheorem*{remark}{Remark}
-
-\frontmatter
-\pagenumbering{gobble}
-
-\makeatletter
-\begin{center}
-\vspace*{0cm}
-\begin{large}
-Universität Heidelberg\\
-Fakultät für Mathematik und Informatik\\
-Studiengang Mathematik\\
-\vspace{8mm}
-\end{large}
-\vfill 
-\begin{large}
-\textbf{MASTER THESIS}\\
-\end{large}
-\vspace{10mm}
-\begin{huge}\@title \end{huge}\\
-\vspace{10mm}
-by \textsc{yichuan shen}\\
-\vspace{3cm}
-\vfill
-Supervisor: \textsc{prof. dr. alexander schmidt}\\
-\today
-\end{center}
-\makeatother
-
-\clearpage
-\ 
-\clearpage
-
-\vspace*{0cm}
-\vfill
-\paragraph{Declaration.} I declare that I carried out this master thesis independently, and only with the cited sources, literature and other professional sources.
-The master thesis was not used in the same or in a similar version to achieve an academic grading or is being published elsewhere.\\
-\vspace{7mm}\\
-\rule{7cm}{0.4pt}
-\clearpage
-
-\vspace*{0cm}
-\vfill
-\renewcommand{\abstractname}{Abstract}
-\begin{abstract}
-Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Curabitur interdum, dui non molestie tincidunt, lectus enim aliquet arcu, suscipit pulvinar dui tortor vitae mi.
-Nam cursus mi sem, ac gravida tellus congue vitae.
-Nam elementum eleifend lorem dignissim rhoncus.
-Quisque ac nulla dui.
-Duis rutrum eros ornare faucibus congue.
-\end{abstract}
-
-\vspace{1cm}
-
-\renewcommand{\abstractname}{Zusammenfassung}
-\begin{abstract}
-Integer congue euismod justo, quis venenatis tortor.
-Proin sem leo, accumsan eget pulvinar vitae, tincidunt ut elit.
-Morbi ut purus volutpat, efficitur quam id, consequat augue.
-Cras ullamcorper lacus eget massa vehicula, a elementum diam lacinia.
-Morbi efficitur elementum malesuada.
-Cras a aliquam tortor, et interdum neque.
-\end{abstract}
-\vfill
-\clearpage
-
-\pagenumbering{arabic} 
-\setcounter{page}{5}
-\tableofcontents
-
 \chapter{Introduction}
@@
 \end{array}$
-
-\mainmatter
-\setcounter{page}{9}
-\chapter{First chapter}
-
-\begin{thebibliography}{99999}
-
-\bibitem[Sp96]{Sp96}
-	M. Spiess (1996)
-	\textit{An arithmetic proof of Pop's Theorem concerning Galois groups of function fields over number fields}.
-    J. reine angew. Math. \textbf{478}, 107\---126.
-
-\end{thebibliography}
-
-%\printindex
-\end{document}

### 2015-07-15 Master Thesis/bibliography.tex
Pop bibliography

--- 2015-07-15 Master Thesis/thesis.tex
@@
-\documentclass[11pt,a4paper,openany]{memoir}
-\usepackage{fontspec}
-\usepackage[english]{babel}
-\usepackage{amsmath, amsfonts, amssymb, amsthm}
-\usepackage{paralist}
-\usepackage{tikz-cd}
-\linespread{1.1}
-\usepackage{hyperref}
-\hypersetup{
-	bookmarksopen=true,
-	pdfstartview=FitH
-}
-\usepackage[inner=4cm,outer=3cm,top=3cm,bottom=3cm]{geometry}
-\author{Yichuan Shen}
-\title{Title of the Master Thesis}
-\makeindex
+\begin{thebibliography}{99999}
 
-\begin{document}
-
-\theoremstyle{plain}
-\theoremstyle{definition}
-\newtheorem{theorem}{Theorem}[chapter]
-\newtheorem{lemma}[theorem]{Lemma}
-\newtheorem{proposition}[theorem]{Proposition}
-\newtheorem{corollary}[theorem]{Corollary}
-\theoremstyle{definition}
-\newtheorem*{definition}{Definition}
-\newtheorem*{example}{Example}
-\theoremstyle{remark}
-\newtheorem*{remark}{Remark}
-
-\frontmatter
-\pagenumbering{gobble}
-
-\makeatletter
-\begin{center}
-\vspace*{0cm}
-\begin{large}
-Universität Heidelberg\\
-Fakultät für Mathematik und Informatik\\
-Studiengang Mathematik\\
-\vspace{8mm}
-\end{large}
-\vfill 
-\begin{large}
-\textbf{MASTER THESIS}\\
-\end{large}
-\vspace{10mm}
-\begin{huge}\@title \end{huge}\\
-\vspace{10mm}
-by \textsc{yichuan shen}\\
-\vspace{3cm}
-\vfill
-Supervisor: \textsc{prof. dr. alexander schmidt}\\
-\today
-\end{center}
-\makeatother
-
-\clearpage
-\ 
-\clearpage
-
-\vspace*{0cm}
-\vfill
-\paragraph{Declaration.} I declare that I carried out this master thesis independently, and only with the cited sources, literature and other professional sources.
-The master thesis was not used in the same or in a similar version to achieve an academic grading or is being published elsewhere.\\
-\vspace{7mm}\\
-\rule{7cm}{0.4pt}
-\clearpage
-
-\vspace*{0cm}
-\vfill
-\renewcommand{\abstractname}{Abstract}
-\begin{abstract}
-Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Curabitur interdum, dui non molestie tincidunt, lectus enim aliquet arcu, suscipit pulvinar dui tortor vitae mi.
-Nam cursus mi sem, ac gravida tellus congue vitae.
-Nam elementum eleifend lorem dignissim rhoncus.
-Quisque ac nulla dui.
-Duis rutrum eros ornare faucibus congue.
-\end{abstract}
-
-\vspace{1cm}
-
-\renewcommand{\abstractname}{Zusammenfassung}
-\begin{abstract}
-Integer congue euismod justo, quis venenatis tortor.
-Proin sem leo, accumsan eget pulvinar vitae, tincidunt ut elit.
-Morbi ut purus volutpat, efficitur quam id, consequat augue.
-Cras ullamcorper lacus eget massa vehicula, a elementum diam lacinia.
-Morbi efficitur elementum malesuada.
-Cras a aliquam tortor, et interdum neque.
-\end{abstract}
-\vfill
-\clearpage
-
-\mainmatter
-\pagenumbering{arabic} 
-\setcounter{page}{5}
-\tableofcontents
-
-\chapter{Introduction}
-
-\clearpage
-
-\section{Notation}
-
-$\begin{array}{ll}
-\mathbb{N} & \text{Natural numbers} \\
-\mathbb{Z} & \text{Integers}\\
-\mathbb{Q} & \text{Rational numbers}\\
-\mathbb{R} & \text{Real numbers}\\
-\mathbb{C} & \text{Complex numbers}\\
-\mathbb{Z}_p & \text{$p$-adic integers}\\
-\mathbb{Q}_p & \text{$p$-adic numbers}
-\end{array}$
-
-\chapter{First chapter}
-
-\begin{thebibliography}{99999999}
-
-\bibitem[DDMS99]{DDMS99}
-	J. D. Dixon, M. P. F. du Sautoy, A. Mann, D. Segal (1999)
-	\textit{Analytic Pro-$p$ Groups}.
-	Cambridge Studies in Advanced Mathematics,
-	2nd Edition.
+\bibitem[Sp96]{Sp96}
+	M. Spiess (1996)
+	\textit{An arithmetic proof of Pop's Theorem concerning Galois groups of function fields over number fields}.
+    J. reine angew. Math. \textbf{478}, 107\---126.
 
 \end{thebibliography}
-
-%\printindex
-\end{document}

### 2015-07-16 Algebraische Geometrie/chapter3.tex
Clarify proof of Riemann-Roch

@@
 
-Sei nun $D\in\operatorname{Div}(X)$ mit $\chi(\mathcal{L}(D))=\deg(D)+1-g$ und $P\in X$ ein abgeschlossener Punkt.
+Sei nun $D\in\operatorname{Div}(X)$ und $P\in X$ ein abgeschlossener Punkt.
 Sei $\{P\}\subset X$ das zu $P$ gehörige abgeschlossene Unterschema von $X$ mit konstanter Strukturgarbe $k(P)\cong k$.
@@
 da $j_\ast k(P)\otimes\mathcal{L}(D+P)=j_\ast k(P)$. Es folgt:
-\[\chi(\mathcal{L}(D+P)) = \chi(\mathcal{L}(D)) + \chi(k(P))=\chi(\mathcal{L}(D))+1 = \deg(D)+1 = \deg(D+P) \]
-Da jeder Divisor von $0$ aus aufgebaut werden kann, folgt die Aussage.\qed
+\[\chi(\mathcal{L}(D+P)) = \chi(\mathcal{L}(D)) + \chi(k(P))=\chi(\mathcal{L}(D))+1 \]
+Andererseits gilt:
+\[\deg(D) + 2 - g = \deg(D+P) + 1 - g \]
+Somit gilt $\chi(\mathcal{L}(D))=\deg(D)+1-g$ genau dann, wenn $\chi(\mathcal{L}(D+P)) = \deg(D+P)+1-g$. Da jeder Divisor von $0$ aus aufgebaut werden kann, folgt die Aussage.\qed
 

### 2015-07-29 Bachelor Thesis/thesis.tex
Add bachelor thesis

\documentclass[11pt,a4paper,openany]{memoir}
\usepackage{fontspec}
\usepackage[ngerman]{babel}
\usepackage{amsmath, amsfonts, amssymb, amsthm}
\usepackage{paralist}
\usepackage[all]{xy}
\linespread{1.1}
\usepackage{hyperref}
\hypersetup{
	bookmarksopen=true,
	pdfstartview=FitH
}
\usepackage[left=2.5cm,right=2.5cm,top=3cm,bottom=3cm]{geometry}
\author{Yichuan Shen}
\title{Die Golod-Shafarevich Ungleichung}
\makeindex
%\chapterstyle{tandh}

\makeatletter
\DeclareFontFamily{OMX}{MnSymbolE}{}
\DeclareSymbolFont{MnLargeSymbols}{OMX}{MnSymbolE}{m}{n}
\SetSymbolFont{MnLargeSymbols}{bold}{OMX}{MnSymbolE}{b}{n}
\DeclareFontShape{OMX}{MnSymbolE}{m}{n}{
    <-6>  MnSymbolE5
   <6-7>  MnSymbolE6
   <7-8>  MnSymbolE7
   <8-9>  MnSymbolE8
   <9-10> MnSymbolE9
  <10-12> MnSymbolE10
  <12->   MnSymbolE12
}{}
\DeclareFontShape{OMX}{MnSymbolE}{b}{n}{
    <-6>  MnSymbolE-Bold5
   <6-7>  MnSymbolE-Bold6
   <7-8>  MnSymbolE-Bold7
   <8-9>  MnSymbolE-Bold8
   <9-10> MnSymbolE-Bold9
  <10-12> MnSymbolE-Bold10
  <12->   MnSymbolE-Bold12
}{}

\let\llangle\@undefined
\let\rrangle\@undefined
\DeclareMathDelimiter{\llangle}{\mathopen}%
                     {MnLargeSymbols}{'164}{MnLargeSymbols}{'164}
\DeclareMathDelimiter{\rrangle}{\mathclose}%
                     {MnLargeSymbols}{'171}{MnLargeSymbols}{'171}
\makeatother

\begin{document}

\theoremstyle{plain}
\theoremstyle{definition}
\newtheorem{theorem}{Theorem}[chapter]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Satz}
\newtheorem{corollary}[theorem]{Korollar}
\theoremstyle{definition}
\newtheorem*{definition}{Definition}
\newtheorem*{example}{Beispiel}
\theoremstyle{remark}
\newtheorem*{remark}{Bemerkung}

\frontmatter
\pagenumbering{gobble}

%\begin{titlingpage}
	\begin{center}
	\vspace*{0cm}
	%\vspace{8mm}
	\begin{large}
	Universität Heidelberg\\
	%\vspace{2mm}
	Fakultät für Mathematik und Informatik\\
	%\vspace{2mm}
	Studiengang Mathematik\\
	\vspace{8mm}
	\end{large}
	\vfill 
	\begin{large}
	\textbf{BACHELORARBEIT}\\
	\end{large}
	\vspace{10mm}
	\begin{huge} Die Golod-Shafarevich Ungleichung \end{huge}\\ 
	\vspace{2mm}
	\begin{huge}für $p$"~adisch analytische pro"~$p$ Gruppen \end{huge}\\ 
	\vspace{10mm}
	von \textsc{yichuan shen}\\
	\vspace{3cm}
	\vfill
	betreut von \textsc{prof. dr. alexander schmidt}\\
	%\includegraphics[scale=.2]{seal.pdf}\\
	%\vspace{2mm}
	\today
	\end{center}
%\end{titlingpage}

\clearpage
\ 
\clearpage

\vspace*{0cm}
\vfill
\paragraph{Erklärung zur Bachelorarbeit.} Hiermit erkläre ich, dass ich die Bachelorarbeit selbst\-stän\-dig verfasst und keine anderen als die angegebenen Quellen und Hilfsmittel benutzt habe. Die Arbeit ist keinem anderen Prüfungsamt in gleicher oder vergleichbarer Form vorgelegt worden. Sie wurde bisher nicht veröffentlicht.\\
\vspace{7mm}\\
\rule{7cm}{0.4pt}
\clearpage

\vspace*{0cm}
\vfill
\renewcommand{\abstractname}{Abstract}
\begin{abstract}
In this thesis, we want to present the Golod-Shafarevich inequality and all nec\-es\-sary requirements, especially verifying the inequality for finite $p$-groups and $p$-adic analytic pro"~$p$ groups. We shall use the purely algebraic characterization of an analytic pro"~$p$ group, de\-vel\-oped by Lubotzky and Mann with powerful pro"~$p$ groups.
\end{abstract}

\vspace{1cm}

\renewcommand{\abstractname}{Zusammenfassung}
\begin{abstract}
In dieser Arbeit wollen wir die Golod-Shafarevich Ungleichung und die nötigen Voraussetzungen vorstellen, und diese für endliche $p$-Gruppen und $p$-adisch analytische pro-$p$ Gruppen nachweisen. Wir verwenden dabei die rein algebraische Charakterisierung einer analytischen pro"~$p$ Gruppe von Lubotzky und Mann über potenzreiche pro"~$p$ Gruppen.
\end{abstract}
\vfill
\clearpage

\pagenumbering{arabic} 
\setcounter{page}{5}
\tableofcontents

\chapter{Einführung}

Zu einem Zahlkörper $K=K_0$ kann man die maximal abelsche, unverzweigte Erweiterung $K_1$ von $K$ bilden. Die zugehörige Galoisgruppe von $K_1/K$ ist dann isomorph zu der Idealklassengruppe $\operatorname{Cl}_K$ von $K$. Dieses Verfahren kann man mit $K_1$ fortführen und man erhält $K_2$, usw. So entsteht der sogenannte Klassenkörperturm:
\[K=K_0\subset K_1\subset K_2\subset\ldots,\quad K_\infty:=\bigcup K_i \]

Furtwängler hat 1924 vermutet, dass es stets für jeden Zahlkörper $K$ ein $i$ gibt mit $K_i=K_{i+1}$, mit anderen Worten, dass der Klassenkörperturm über $K$ stationär wird. Wegen der Endlichkeit der Klassenzahl ist dies äquivalent dazu, dass $\operatorname{Gal}(K_\infty/K)$ endlich ist. Wäre dies der Fall, so könnte man jeden Zahlkörper in einem endlichen Erweiterungskörper $K_\infty$ einbetten, dessen Ganzheitsring ein Hauptidealring ist.

Doch 1964 konnte dies durch Golod und Shafarevich in \cite{GS64} widerlegt werden. Dazu haben sie die maximale Erweiterung $K_i^{(p)}$ in $K_i$ über $K$ mit einer $p$-Gruppe als Galoisgruppe über $K$ betrachtet, mit einer fixierten Primzahl $p$. So erhält man den $p$-Klassenkörperturm:
\[K=K_0^{(p)}\subset K_1^{(p)}\subset K_2^{(p)}\subset\ldots,\quad K_\infty^{(p)}:=\bigcup K_i^{(p)}\subset K_\infty \]
$\operatorname{Gal}(K_\infty^{(p)}/K)$ ist somit eine pro"~$p$ Gruppe. Die Golod-Shafarevich Ungleichung für endliche $p$"~Gruppen gibt uns eine hinreichende Bedingung, wann eine pro"~$p$ Gruppe $G$ unendlich ist:

\paragraph{Theorem.} \textit{(Die Golod-Shafarevich Ungleichung)} Für eine endliche $p$-Gruppe $G$ gilt:
\[ t\geq \frac{d^2}{4} \]
wobei $d$ die minimale Anzahl von Erzeugern und $t$ die minimale Anzahl der dazugehörigen Relationen von $G$ als pro"~$p$ Gruppe bezeichnet.

\paragraph{} Mithilfe dieser Ungleichung kann man somit Beispiele konstruieren, für das $\operatorname{Gal}(K_\infty^{(p)}/K)$ nicht endlich ist, und insbesondere $\operatorname{Gal}(K_\infty/K)$ nicht endlich ist. Das klassische Beispiel, das Shafarevich nutzte, ist $K=\mathbb{Q}(\sqrt{-D})$ mit $D=3\cdot 5\cdot 7\cdot 11\cdot 13\cdot 17\cdot 19$.

Lie"~Gruppen, d.h. Gruppen mit einer Mannigfaltigkeitsstruktur, deren Grup\-pen\-ope\-ra\-tio\-nen glatt sind, spielen über  $\mathbb{C}$ oder $\mathbb{R}$ vor allem in der theoretischen Physik eine große Rolle. Andererseits finden Lie-Gruppen über $\mathbb{Q}_p$, sogenannte $p$-adisch analytische Gruppen, in der arithmetischen Geometrie nützliche Anwendungen.

In dieser Arbeit wollen wir hauptsächlich (sofern nichts anderes angegeben) \cite{DDMS99} folgen und eine etwas allgemeinere Formulierung der Golod-Shafarevich Ungleichung beweisen. Diese werden wir nicht nur für endliche $p$"~Gruppen, sondern auch für $p$"~adisch analytische pro"~$p$ Gruppen nachweisen. Dabei verwenden wir die rein algebraische Charakterisierung einer analytischen pro"~$p$ Gruppe von Lubotzky und Mann \cite{LM87} über potenzreiche pro"~$p$ Gruppen.

\section{Überblick}

Im ersten Kapitel werden die Grundlagen von pro"~$p$ Gruppen behandelt. Wir werden klären, was Erzeuger und Relationen in der Kategorie der pro"~$p$ Gruppen bedeuten und im Falle einer endlich erzeugten pro"~$p$ Gruppe eine einfache Charakterisierung für die minimale Anzahl von Erzeugern herleiten.

Im zweiten Kapitel betrachten wir analytische pro"~$p$ Gruppen. Insbesondere untersuchen wir potenzreiche pro"~$p$ Gruppen und wie sich deren gruppentheoretischen Eigenschaften auf den Gruppenring über $\mathbb{F}_p$ übertragen. Diese Resultate zeigen dann, dass die Voraussetzungen der Golod-Shafarevich Ungleichung für analytische pro"~$p$ Gruppen erfüllt sind.

Im dritten Kapitel führen wir zunächst den vollständigen Gruppenring ein und untersuchen diese für endlich erzeugte, freie pro"~$p$ Gruppen über $\mathbb{F}_p$. Im zweiten Teil zeigen wir schließlich die Golod-Shafarevich Ungleichung.

\section{Danksagung}

Ich würde gerne Professor Dr. Alexander Schmidt für das Betreuen meiner Arbeit danken. Besonderen Dank gilt Dr. Malte Witte, der mir nicht nur auf Fehler und Ungenauigkeiten im Text aufmerksam gemacht hat, sondern mir auch mit viel Geduld und Verständnis meinen Problemen und Fragen geholfen hat.

\clearpage

\section{Notation}

$\begin{array}{ll}
\mathbb{N} & \text{natürliche Zahlen inklusive $0$} \\
\mathbb{Z} & \text{ganze Zahlen}\\
\mathbb{Q} & \text{rationale Zahlen}\\
\mathbb{R} & \text{reelle Zahlen}\\
\mathbb{C} & \text{komplexe Zahlen}\\
\mathbb{Z}_p & \text{$p$-adische ganze Zahlen}\\
\mathbb{Q}_p & \text{$p$-adische Zahlen}\\
\mathbb{F}_q & \text{Körper der Ordnung $q$}\\
\\
\overline{X} & \text{topologischer Abschluss von $X$}\\
\subset & \text{Teilmenge}\\
\subset_\text{o}& \text{offene Teilmenge}\\
\lhd & \text{Normalteiler}\\
\lhd_\text{o}& \text{offener Normalteiler}\\
\\
\langle X\rangle & \text{von $X$ erzeugte Gruppe}\\
X^{\{n\}} & \{x^n\mid x\in X\}\\
G^n & \langle G^{\{n\}} \rangle\text{, wobei $G$ eine multiplikative Gruppe ist} \\
\left[G:H\right] & \text{Index von $H$ in $G$}\\
AB & \{ab\mid a\in A,\ b\in B\}\\
\left[a,b\right] & a^{-1}b^{-1}ab\\
\left[A,B\right] & \langle \left[a,b\right]\mid a\in A,\ b\in B\rangle\\
\\
A^{(n)} & \text{$n$-fache direkte Summe über $A$}\\
\dim_k(V) & \text{Dimension von $V$ als $k$-Vektorraum}\\
\langle\alpha\rangle & \alpha_1+\ldots+\alpha_r\text{, wobei }\alpha=(\alpha_1,\ldots,\alpha_r)\\
\mathbf{v}^\alpha & v_1^{\alpha_1}\cdots v_r^{\alpha_r}\text{, wobei $\mathbf{v}=(v_1,\ldots,v_r)$ und $\alpha\in\mathbb{N}^r$}\\
\end{array}$

\mainmatter
\setcounter{page}{9}

\chapter{Pro"~$p$ Gruppen}

In dieser Arbeit geht es um \textit{pro"~$p$ Gruppen}. In diesem Kapitel werden die Grundlagen von \textit{proendlichen Gruppen} und pro"~$p$ Gruppen behandelt, insbesondere betrachten wir den Fall einer (topologisch) endlich erzeugten pro"~$p$ Gruppe. Ferner wird eingeführt, was eine \textit{Darstellung} einer pro"~$p$ Gruppe mittels Erzeugern und Relationen ist.

\section{Proendliche Gruppen}

\begin{definition}
Eine \textit{proendliche Gruppe}\index{proendliche Gruppe} ist eine kompakte, hausdorffsche topologische Gruppe, deren offene Normalteiler eine Umgebungsbasis der Identität bilden.
\end{definition}

\begin{remark}
Äquivalent dazu: Eine topologische Gruppe $G$ ist genau dann proendlich, wenn sie projektiver Limes endlicher, diskreter Gruppen ist. Ist $G$ proendlich, dann gibt es einen natürlichen topologischen Isomorphismus:
\[G\cong\varprojlim_{N\lhd_\text{o}G}G/N \]
Eine endliche Gruppe wird mit der diskreten Topologie zu einer proendlichen Gruppe.
\end{remark}

\iffalse
\begin{proposition}\label{1.2.i}
Sei $G$ eine proendliche Gruppe. Dann gilt:
\begin{enumerate}[(i)]
\item Jede offene Untergruppe von $G$ ist abgeschlossen, besitzt endlichen Index und enthält einen offenen Normalteiler von $G$. Der Durchschnitt aller offenen Untergruppen ist $\{1\}$. Enthält eine Untergruppe $U\subset G$ eine offene nichtleere Teilmenge, so ist $U$ offen.
\item Eine abgeschlossene Untergruppe von $G$ ist genau dann offen, wenn sie einen endlichen Index in $G$ besitzt.
\item Eine Teilmenge von $G$ ist genau dann offen, wenn sie eine Vereinigung von Mengen der Form $gN$ mit $g\in G,\ N\lhd_\text{o} G$ ist.
\item Sei $N\lhd G$ abgeschlossen. Dann ist $G/N$ proendlich und die natürliche Projektion $G\to G/N$ ist eine offene, stetige Abbildung.
\end{enumerate}
\end{proposition}

\begin{proof}
(iii) und (iv) folgen direkt aus der Definition. Für (i) und (ii) verwendet man die Kompaktheit und Hausdorff-Eigenschaft des Raumes und dass für jede Untergruppe $H\subset G$ gilt:
\[ G- H = \bigcup_{g\not\in H} gH \qedhere \]
\end{proof}
\fi

\begin{proposition}\label{1.2.iii}
Sei $G$ eine proendliche Gruppe und $X\subset G$ eine Teilmenge. Dann gilt:
\[ \overline{X} = \bigcap_{N\lhd_\text{o} G} XN \]
Ist $X$ eine Untergruppe, so gilt $\overline{X} = \bigcap\ \{K\text{ Untergruppe} \mid X\subset K\subset_\text{o}G \}$.
\end{proposition}

\begin{proof}
Sei $y\not\in\overline{X}$. Da $G\setminus\overline{X}$ offen ist, gibt es ein $N\lhd_\text{o}G$ mit $yN\cap X=\varnothing$. Dies ist äquivalent zu $y\not\in XN$. Dies zeigt $\bigcap XN\subset\overline{X}$. $X\subset\bigcap XN$ ist klar. Nun ist $XN=\bigcup_{x\in X} xN$ für jedes $N\lhd_\text{o}G$ eine endliche Vereinigung von abgeschlossenen Mengen, da $N$ von endlichem Index ist. Also ist $\bigcap XN$ abgeschlossen und $\overline{X}\subset \bigcap XN$.

Ist $X$ eine Untergruppe, so ist $XN$ eine offene Untergruppe, die $X$ enthält. Die andere Inklusion folgt dadurch, dass offene Untergruppen abgeschlossen sind.
\end{proof}

\begin{corollary}
Für jedes $N\lhd G$ ist der Abschluss $\overline{N}$ ebenfalls ein Normalteiler in $G$.
\end{corollary}

\begin{proof}
Für $K\lhd_\text{o}G$ ist $NK$ ein Normalteiler in $G$ und der Durchschnitt von Normalteilern ist wieder ein Normalteiler.
\end{proof}

\paragraph{} Der folgende Satz und seine Korollare werden als bekannt vorausgesetzt und nur der Voll\-stän\-dig\-keit halber erwähnt:

\begin{proposition}\label{1.4}
Sei $(X_\lambda)_{\lambda\in\Lambda}$ ein projektives System von nichtleeren kompakten Räumen und stetigen Ü\-ber\-gangs\-ab\-bil\-dun\-gen $(\varphi_{\mu\nu})_{\mu\leq\nu}$. Dann ist der projektive Limes $\varprojlim X_\lambda$ nichtleer und kompakt.
\end{proposition}

\iffalse
\begin{proof}
Für $\nu\leq\mu$ setze:
\[U_{\mu\nu}:=\Big\{(x_\lambda)\in\prod_{\lambda\in\Lambda}X_\lambda \mid \varphi_{\mu\nu}(x_\mu)\neq x_\nu \Big\} \]
Diese sind offene Teilmengen von $\prod X_\lambda$, da alle $\varphi_{\mu\nu}$ stetig sind. Jede endliche Vereinigung von Mengen der Form $U_{\mu\nu}$ ist echt in $\prod X_\lambda$ enthalten: Wähle dazu ein $\kappa\in \Lambda$ mit $\kappa\geq \mu,\nu$ für alle Mengen $U_{\mu\nu}$, die in der Vereinigung auftauchen. Wähle $x_\kappa\in X_\kappa$ beliebig und setze:
\[x_\lambda=\begin{cases}
\varphi_{\kappa\lambda}(x_\kappa),&\lambda\leq\kappa\\
\text{beliebiges Element in }X_\lambda,&\text{sonst}
\end{cases} \]
Dann liegt $(x_\lambda)$ nicht in der Vereinigung der $U_{\mu\nu}$. Nun ist die Vereinigung aller $U_{\mu\nu}$ das Komplement von $\varprojlim X_\lambda$ in $\prod X_\lambda$. Somit ist $\varprojlim X_\lambda$ abgeschlossen, also kompakt, da $\prod X_\lambda$ nach Tychonoff kompakt ist. Wäre nun $\varprojlim X_\lambda\neq\varnothing$, würden die $U_{\mu\nu}$ ganz $\prod X_\lambda$ überdecken. Also existiert eine endliche Teilüberdeckung, ein Widerspruch.
\end{proof}
\fi

\begin{corollary}\label{1.4cor}
Der projektive Limes von endlichen, nichtleeren Mengen ist nichtleer. 
\end{corollary}

\begin{corollary}\label{prop:compact-exact-limes}
Sei $1\to (G_i')\to (G_i) \to (G_i'')\to 1$ eine exakte Folge projektiver Systeme von kompakten topologischen Gruppen. Dann ist die folgende Folge exakt:
\[1\to\varprojlim G_i'\to\varprojlim G_i  \to \varprojlim G_i''\to 1 \]
\end{corollary}

\iffalse
\begin{proof}
Die Exaktheit von $1\to\varprojlim G_i'\to\varprojlim G_i\to\varprojlim G_i''$ ist klar. Für $(x_i)\in\varprojlim G_i''$ gilt $f^{-1}((x_i))=\varprojlim f_i^{-1}(x_i)$. Da alle $f_i^{-1}(x_i)$ nichtleer und kompakt sind, folgt $f^{-1}((x_i))\neq\varnothing$ nach Satz~\ref{1.4} und $f$ ist surjektiv.
\end{proof}

\begin{proposition}\label{1.3}
Ist $G$ eine proendliche Gruppe, so ist $G$ topologisch isomorph zu $\varprojlim_{N\lhd_\text{o} G} G/N$. Umgekehrt ist der projektive Limes endlicher, diskreter Gruppen eine proendliche Gruppe.
\end{proposition}

\begin{proof}
Sei $G$ proendlich. Der natürliche Homomorphismus $i:G\to\varprojlim G/N,\ g\mapsto (gN)_{N\lhd_\text{o}G}$ ist wegen $\bigcap_{N\lhd_\text{o}G}N=1$ injektiv. Da jede Projektion $G\to G/N$ stetig und surjektiv ist, ist $i$ stetig und surjektiv nach \ref{prop:compact-exact-limes}. Schließlich ist jeder stetige Isomorphismus zwischen kompakten Hausdorffräumen ein Homöomorphismus.

Sei nun $(G_\lambda)_{\lambda\in\Lambda}$ ein projektives System end\-li\-cher dis\-kre\-ter Gruppen. Dann ist $\prod_{\lambda\in\Lambda}G_\lambda$ als Produkt kompakter Hausdorffräume wieder ein kompakter Hausdorffraum. Per Definition der Produkttopologie enthält jede Umgebung der $1$ eine Untergruppe der folgenden Form:
\[U(S)=\prod_{\lambda\not\in S}G_\lambda\times\prod_{\lambda\in S}\{1\}\subset_\text{o}\prod_{\lambda\in\Lambda}G_\lambda \]
mit $S\subset\Lambda$ endlich. Somit ist $\prod G_\lambda$ eine proendliche Gruppe. Es ist noch zu zeigen, dass $\varprojlim G_\lambda$ in $\prod G_\lambda$ abgeschlossen ist. Sei dazu $\widehat{g}=(g_\lambda)_{\lambda\in\Lambda}\in \prod G_\lambda \setminus \varprojlim G_\lambda$. Dann existieren $\nu>\mu$ in $\Lambda$ mit $\pi_{\nu\mu}(g_\nu)\neq g_\mu$, wobei $\pi_{\nu\mu}$ die Übergangsabbildung bezeichnet. Nun ist $\widehat{g}U(\{\nu,\mu\})$ eine offene Umgebung von $\widehat{g}$ in $\prod G_\lambda$ und $\widehat{g}U(\{\nu,\mu\})\cap\varprojlim G_\lambda=\varnothing$. Also ist $\prod G_\lambda\setminus\varprojlim G_\lambda\subset_\text{o} \prod G_\lambda$.
\end{proof}
\fi

\begin{definition}
Eine Teilmenge $X\subset G$ einer topologischen Gruppe $G$ \textit{erzeugt $G$ topologisch}\index{erzeugt topologisch}, wenn: \[\overline{\langle X\rangle}=G\] $G$ heißt \textit{endlich erzeugt}\index{endlich erzeugt}, wenn $G$ von einer endlichen Menge topologisch erzeugt wird. Wir setzen:
\[d(G):=\inf\{\#X\mid G=\overline{\langle X\rangle}\} \]
Eine proendliche Gruppe $G$ heißt \textit{prozyklisch}, wenn $d(G)=1$.
\end{definition}

\iffalse
\begin{proposition}\label{1.5}
Sei $G$ eine proendliche Gruppe und $H\subset G$ eine abgeschlossene Untergruppe.
\begin{enumerate}[(i)]
\item Eine Teilmenge $X\subset H$ erzeugt $H$ genau dann topologisch, wenn das Bild von $X$ in $HN/N$ die Gruppe $HN/N$ für jedes $N\lhd_\text{o}G$ erzeugt.
\item Kann $HN/N$ für jedes $N\lhd_\text{o}G$ durch $d$ Elemente erzeugt werden, so kann $H$ durch $d$ Elemente topologisch erzeugt werden.
\end{enumerate}
\end{proposition}

\begin{proof}
\begin{enumerate}[(i)]
\item Sei zunächst $H=\overline{\langle X\rangle}$. Nach Satz~\ref{1.2.iii} gilt für alle $N\lhd_\text{o}G$:
\[ H=\bigcap_{N\lhd_\text{o}G} \langle X \rangle N \subset \langle X\rangle N\]
Sei nun umgekehrt $\langle X\rangle N/N=HN/N$ und $\varnothing\neq U\subset_\text{o}H$ eine Teilmenge. $U$ enthält eine offene Umgebung der Form $hN\cap H$ mit $h\in H,\ N\lhd_\text{o}G$. Nach Voraussetzung ist $hN=xN$ für ein $x\in\langle X\rangle\subset H$, d.h. $U\cap\langle X\rangle \neq\varnothing$. Somit liegt $\langle X\rangle$ dicht in $H$.
\item Für jedes $N\lhd_\text{o}G$ sei $Y_N$ die Menge aller $d$-Tupel mit Elementen in $G/N$, die $HN/N$ erzeugen. Alle $Y_N$ sind endlich und nichtleer. Ist $\pi_{MN}:G/M\to G/N$ die natürliche Projektion, wobei $M\subset N,\ M,N\lhd_\text{o}G$, so induziert diese eine Abbildung $\pi_{MN}: Y_M\to Y_N$. Also bilden $(Y_N)_{N\lhd_\text{o}G}$ ein projektives System. Nach Satz~\ref{1.4} ist der projektive Limes nichtleer, sei also $(X_N)\in\varprojlim Y_N$. Dann existieren $x_1,\ldots,x_d\in G$, so dass $X_N=(x_1N,\ldots,x_dN)$ für alle $N\lhd_\text{o}G$. Nach (i) ist $\{x_1,\ldots,x_d\}$ somit ein topologisches Erzeugendensystem für $H$. \qedhere
\end{enumerate}
\end{proof}
\fi

\begin{proposition}\label{1.7}
Ist $G$ eine endlich erzeugte proendliche Gruppe, dann ist jede offene Untergruppe von $G$ ebenfalls endlich erzeugt.
\end{proposition}

\begin{proof}
Sei $X$ ein endliches topologisches Erzeugendensystem von $G$ und o.B.d.A. $X^{-1}=X$. Sei $H\subset_\text{o}G$ eine Untergruppe und $T$ ein Repräsentantensystem von den Rechtsnebenklassen $H\backslash G$ mit $1\in T$. Nun ist $T$ endlich. Für jedes $x\in X$ und $t\in T$ gibt es ein $s(t,x)\in T$, so dass $Htx=Hs(t,x)$. Wir setzen:
\[Y=\{tx\cdot s(t,x)^{-1}\mid t\in T,\ x\in X \}\subset H \]
Wir zeigen, dass $Y$ nun $H$ topologisch erzeugt. Betrachte die Untergruppe $M:=\overline{\langle Y\rangle}\subset G$. Ist $a\in M,\ t\in T$ und $x\in X$, so gilt:
\[atx=atx\cdot s(t,x)^{-1} s(t,x)\in MT \]
Also ist $MTX=MT$. Da $1\in MT$ und $X=X^{-1}$, folgt $\langle X\rangle\subset MT$. Da $T$ endlich, ist $MT$ abgeschlossen, also $MT=G$. Nun gilt $M\subset H$ und $H=MT\cap H=M(T\cap H)=M$.
\end{proof}

\begin{definition}
Sei $G$ eine proendliche Gruppe. Die \textit{Frattinigruppe}\index{Frattinigruppe} $\Phi(G)$ von $G$ ist definiert durch:
\[\Phi(G) = \bigcap\ \{M\mid M\subsetneq  G\text{ maximale offene Untergruppe} \} \]
Es ist klar, dass $\Phi(G)\lhd G$ abgeschlossen ist. 
\end{definition}

\begin{proposition}\label{1.9}
Sei $G$ eine proendliche Gruppe. Dann sind äquivalent:
\begin{enumerate}[(i)]
\item $X$ erzeugt $G$ topologisch.
\item $X\cup\Phi(G)$ erzeugt $G$ topologisch.
\item $X\Phi(G)/\Phi(G)$ erzeugt $G/\Phi(G)$ topologisch.
\end{enumerate}
\end{proposition}

\begin{proof}
(i)$\implies$(ii)$\implies$(iii) ist klar. Sei (iii) erfüllt und $H\subset_\text{o}G$ eine Untergruppe, die $X$ enthält. Ist $H\neq G$, so ist $H\subset M$ für eine maximale offene Untergruppe $M\subsetneq G$ und es gilt:
\[\overline{\langle X\rangle }\Phi(G)/\Phi(G)\subset M/\Phi(G)\neq G/\Phi(G) \]
ein Widerspruch. Somit ist $H=G$ und nach Satz~\ref{1.2.iii} folgt $\overline{\langle X\rangle}=G$.
\end{proof}

\begin{remark}
Die Frattinigruppe von $G$ besteht aus Nichterzeugern, d.h. die Elemente von $\Phi(G)$ sind in jedem topologischen Erzeugendensystem von $G$ überflüssig.
\end{remark}

\begin{corollary}\label{cor:1.9}
Sei $G$ eine proendliche Gruppe. Dann gilt $d(G)=d(G/\Phi(G))$.
\end{corollary}

\section{Endliche $p$-Gruppen}

Sei $p$ eine Primzahl. Wie wir später sehen werden, sind pro"~$p$ Gruppen nichts anderes als ein projektiver Limes von endlichen $p$-Gruppen. Daher werden wir zunächst endliche $p$-Gruppen betrachten und ein paar Hilfsresultate beweisen.

\begin{definition}
Wir definieren die \textit{untere Zentralreihe} einer Gruppe $G$ wie folgt:
\[\gamma_1(G):=G,\quad \gamma_{i+1}(G):=[\gamma_i(G),G] \]
$G$ heißt \textit{nilpotent}, wenn $\gamma_k(G)=1$ für ein $k$ gilt.
\end{definition}

\begin{proposition}
Das \textit{Zentrum}\index{Zentrum} $Z(G)=\{x\in G\mid \forall g\in G:\ xg=gx \}\lhd G$ einer nichttrivialen endlichen $p$"~Gruppe $G$ ist nichttrivial.
\end{proposition}

\begin{proof}
Folgt aus der Klassengleichung.
\end{proof}

\begin{proposition}\label{0.4.ii}
Jede endliche $p$-Gruppe $G$ ist nilpotent.
\end{proposition}

\begin{proof}
Per Induktion über $\#G$. Ist $G=\{1\}$, so ist $G$ offensichtlich nilpotent. Sei nun $\#G>1$. Nach Induktionsvoraussetzung ist $G/Z(G)$ nilpotent, also gibt es ein $m$ mit $\gamma_m(G/Z(G))=1$. Es folgt somit $\gamma_{m+1}(G)=[\gamma_m(G),G]\subset [Z(G),G]=1$.
\end{proof}

\begin{proposition}\label{0.4.iii}
Sei $G$ eine endliche $p$"~Gruppe. Dann ist jede maximale Untergruppe $M\subsetneq G$ normal und vom Index $p$ in $G$.
\end{proposition}

\begin{proof}
Per Induktion über $n$, wobei $\#G=p^n$. Für $n=1$ ist die Aussage klar. Sei $M\subsetneq G$ eine maximale Untergruppe. Da $Z(G)\neq 1$, existiert ein $z\in Z(G)$ von Ordnung $p$. Ist $z\not\in M$, so gilt $M\langle z\rangle=G$ und somit $M\lhd G$ und $[G:M]=p$. Andernfalls ist $M/\langle z\rangle\subsetneq G/\langle z\rangle$ eine maximale Untergruppe und die Behauptung folgt aus der Induktionsvoraussetzung.
\end{proof}

\begin{lemma}\label{0.4.v:lemma}
Sei $G$ eine endliche $p$"~Gruppe und $1\neq N\lhd G$. Dann gilt $Z(G)\cap N\neq 1$.
\end{lemma}

\begin{proof}
Sei $k\geq 1$ so gewählt, dass $\gamma_k(G)\cap N\neq 1$ und $\gamma_{k+1}(G)\cap N=1$. Dann ist $1\neq\gamma_k(G)\cap N\subset Z(G)\cap N$.
\end{proof}

\begin{proposition}\label{0.4.v}
Sei $G$ eine endliche $p$"~Gruppe und $1\neq N\lhd G$. Dann gibt es eine maximale Untergruppe $M\subsetneq N$ mit $M\lhd G$.
\end{proposition}

\begin{proof}
Per Induktion über $\#N$. Ist $Z(G)\cap N=N$, d.h. $[N,G]=1$, so ist jede maximale Un\-ter\-grup\-pe $M\subsetneq N$ normal in $G$. Sei also $Z(G)\cap N\subsetneq N$. Da $[N,G]\lhd G$, folgt $K:= [N,G]\cap Z(G)\neq 1$ nach Lemma~\ref{0.4.v:lemma}. Da $[N,G]\subset N$, ist $K\subsetneq N$. Somit gilt $1\neq N/K\lhd G/K$ und nach Induktionsvoraussetzung gibt es eine maximale Untergruppe $M\subsetneq N$ mit $M/K\lhd G/K$, so dass $M\lhd G$ folgt.
\end{proof}

\begin{lemma}\label{1.23}
Sei $G=\langle a_1,\ldots,a_d\rangle$ eine endlich erzeugte, nilpotente Gruppe. Dann ist jedes Element in $[G,G]$ von der Form $[x_1,a_1]\cdots[x_d,a_d]$ mit $x_1,\ldots,x_d\in G$.
\end{lemma}

\begin{proof}
Sei $\gamma_c(G)\supsetneq \gamma_{c+1}(G)=1$, also $\gamma_c(G)\subset Z(G)$. Wir beweisen die Aussage per Induktion über $c$. Sei $c\geq 2$. Für $u,v\in\gamma_{c-1}(G)$ und $a,b\in G$ gilt:
\[[u,ab]=[u,b]b^{-1}[u,a]b=[u,a][u,b],\quad [uv,a]=v^{-1}[u,a]v[v,a]=[u,a][v,a] \]
Insbesondere ist $[u,a^n]=[u,a]^n=[u^n,a]$. Somit besitzt jedes $w\in\gamma_c(G)$ eine Darstellung:
\[w=[w_1,a_1]\cdots[w_d,a_d],\quad w_1,\ldots,w_d\in\gamma_{c-1}(G) \]
Sei $g\in[G,G]$. Die Induktionsvoraussetzung, auf $G/\gamma_c(G)$ angewendet, gibt uns $y_1,\ldots,y_d\in G$ und $w\in\gamma_c(G)$, so dass:
\[g=[y_1,a_1]\cdots [y_d,a_d]w \]
Schreiben wir $w=\prod[w_i,a_i]$, so folgt $g=\prod[w_iy_i,a_i]$, da $[w_i,a_i]\in Z(G)$.
\end{proof}

\section{Pro"~$p$ Gruppen}

Sei $p$ eine Primzahl. Im Gegensatz zu einer allgemeinen proendlichen Gruppe hat die Frattinigruppe einer pro"~$p$ Gruppe eine einfache Form und wir können $d(G)$ sehr einfach charakterisieren. Im Fall einer endlich erzeugten pro"~$p$ Gruppe, können wir sogar eine explizite offene Umgebungsbasis induktiv angeben.

\begin{definition}
Eine \textit{pro"~$p$ Gruppe}\index{pro-$p$ Gruppe} ist eine proendliche Gruppe, in der jeder offene Normalteiler vom Index $p^n$ für ein $n\geq 0$ ist.
\end{definition}

\begin{remark}
In einer pro"~$p$ Gruppe besitzt jede offene Untergruppe einen Index $p^n,\ n\geq 0$, da jede offene Untergruppe einen offenen Normalteiler enthält. Jede endliche $p$-Gruppe wird mit der diskreten Topologie zu einer pro"~$p$ Gruppe.
\end{remark}

\begin{proposition}\label{1.11}
Sei $G$ eine proendliche Gruppe. Ist $G$ eine pro"~$p$ Gruppe und $H\subset G$ eine abgeschlossene Untergruppe, so ist $H$ eine pro"~$p$ Gruppe. Ist $H\lhd G$ abgeschlossen, so ist auch $G/H$ eine pro"~$p$ Gruppe.
\end{proposition}

\begin{proof}
$H$ ist wieder proendlich und es gilt $[H:N]\mid [G:N]$ für alle $N\lhd_\text{o}H$. Ist $H\lhd G$, so ist $G/H$ wieder eine proendliche Gruppe und $\phi:G\to G/H$ ist stetig. Ist $N\lhd_\text{o}G/H$, so ist $\phi^{-1}(N)\lhd_\text{o}G$ und wir haben eine Bijektion $G/\phi^{-1}(N)\to (G/H)/N$.
\end{proof}

\begin{proposition}\label{1.12}
Eine topologische Gruppe $G$ ist genau dann eine pro"~$p$ Gruppe, wenn $G$ ein projektiver Limes endlicher, diskreter $p$-Gruppen ist.
\end{proposition}

\begin{proof}
Ist $G$ eine pro"~$p$ Gruppe, so ist sie insbesondere proendlich und es gilt $G\cong\varprojlim G/N$, wobei jedes $G/N$ eine endliche $p$"~Gruppe ist. Sei umgekehrt $G=\varprojlim_{\lambda\in\Lambda}G_\lambda$ mit endlichen $p$"~Gruppen $G_\lambda$. Somit enthält jede offene Untergruppe von $G$ eine Untergruppe der Form:
\[ U(S) = G\cap \Big(\prod_{\lambda\not\in S}G_\lambda\times \prod_{\lambda\in S}\{1\}\Big) \]
wobei $S\subset\Lambda$ endlich ist. Da $[G:U(S)]$ stets $\prod_{\lambda\in S}\#G_\lambda$ teilt, ist jede offene Untergruppe von $G$ vom $p$-Potenz-Index.
\end{proof}

\begin{proposition}\label{1.13}
Ist $G$ eine pro"~$p$ Gruppe, so gilt mit $G^p=\langle g^p\mid g\in G\rangle$:
\[\Phi(G)=\overline{G^p[G,G]} \]
\end{proposition}

\begin{proof}
In jeder endlichen $p$-Gruppe ist nach Satz~\ref{0.4.iii} eine echte maximale Untergruppe stets ein Normalteiler vom Index $p$. Ist also $M\subsetneq_\text{o} G$ eine maximale Untergruppe und $N\lhd_\text{o} G$ mit $N\subset M$, so ist $M/N$ eine maximale Untergruppe von $G/N$ und somit $M\lhd G$ mit $[G:M]=p$. Es folgt $G^p[G,G]\subset M$. Dies zeigt $\Phi(G)=\bigcap M\supset G^p[G,G]$. Da $\Phi(G)$ abgeschlossen ist, folgt $\Phi(G)\supset \overline{G^p[G,G]}$.

Es ist klar, dass $G^p\lhd G$ und $[G,G]\lhd G$. Daher ist $Q:=G/\overline{G^p[G,G]}$ eine pro"~$p$ Gruppe. Ist $N\lhd_\text{o}Q$, so ist $Q/N$ ein endlicher $\mathbb{F}_p$"~Vek\-tor\-raum, d.h. von der Form $Q/N\cong  \mathbb{F}_p^{(m)}$ für ein $m\geq 0$. Somit ist $\Phi(Q/N)=1$, also:
\[\Phi(G) /\overline{G^p[G,G]}=\Phi(Q)\subset\bigcap_{N\lhd_\text{o}Q}N=1\qedhere \]
\end{proof}

\begin{proposition}\label{1.14}
Sei $G$ eine pro"~$p$ Gruppe. Dann ist $G$ genau dann endlich erzeugt, wenn $\Phi(G)\lhd_\text{o}G$.
\end{proposition}

\begin{proof}
Ist $\Phi(G)\lhd_\text{o}G$, so ist $G/\Phi(G)$ endlich und besitzt daher ein endliches Erzeugendensystem. Nach Satz~\ref{1.9} ist auch $G$ endlich erzeugt.

Sei nun $G=\overline{\langle X\rangle}$ mit $d:=\#X<\infty$. Ist $N\lhd_\text{o}G$ mit $\Phi(G)\subset N$, so ist $G/N$ nach Satz~\ref{1.13} ein von $d$ Elementen erzeugter $\mathbb{F}_p$"~Vektorraum, d.h. $[G:N]\leq p^d$. Wähle nun ein $M\lhd_\text{o}G$ mit $\Phi(G)\subset M$ und maximalem Index $[G:M]$. Wir zeigen nun $M\subset N$ für jedes $N\lhd_\text{o}G$ mit $\Phi(G)\subset N$. Es gilt $\Phi(G)\subset M\cap N\lhd_\text{o}G$, daher $[G:M\cap N]\leq [G:M]$. Also ist die natürliche Surjektion $G/M\cap N\to G/M$ ein Isomorphismus. Es folgt $M=M\cap N$, d.h. $M\subset N$. Da $\Phi(G)$ normal und abgeschlossen ist, gilt mit Satz~\ref{1.2.iii}:
\[\Phi(G)=\bigcap\ \{N\mid \Phi(G)\subset N\lhd_\text{o} G\} = M \lhd_\text{o}G\qedhere \]
\end{proof}

\begin{corollary}\label{cor:1.19}
Sei $G$ eine pro"~$p$ Gruppe. Dann gilt $d(G)=\dim_{\mathbb{F}_p}(G/\Phi(G))$.
\end{corollary}

\begin{proof}
Ist $G$ eine endlich erzeugte pro"~$p$ Gruppe, so ist $G/\Phi(G)$ nach Satz~\ref{1.14} und Satz~\ref{1.13} diskret und ein $\mathbb{F}_p$-Vektorraum. Jedes Erzeugendensystem von $G/\Phi(G)$ als topologische Gruppe bildet somit ein Erzeugendensystem von $G/\Phi(G)$ als $\mathbb{F}_p$-Vektorraum und umgekehrt. Daher gilt mit Satz~\ref{cor:1.9}:
\[d(G)=d(G/\Phi(G))= \dim_{\mathbb{F}_p}(G/\Phi(G))\qedhere \] 
\end{proof}

\paragraph{} Wir wollen nun die \textit{untere $p$-Zentralreihe} einführen. Wir werden zeigen, dass diese Untergruppen in einer endlich erzeugten pro"~$p$ Gruppe sogar eine offene, abzählbare Umgebungsbasis der $1\in G$ bilden. Ein Vorteil dieser Umgebungsbasis ist, dass wir sie induktiv angeben können.

\begin{definition}
Sei $G$ eine pro"~$p$ Gruppe. Wir setzen:
\[P_1(G):=G,\quad P_{i+1}(G)=\overline{P_i(G)^p [P_i(G),G] },\ i\geq 1 \]
Nach Satz~\ref{1.13} ist $\Phi(G)=P_2(G)$.
\end{definition}

\begin{remark}
Beachte, dass $\Phi(P_i(G))\subset P_{i+1}(G)$ für alle $i$ gilt. Ferner ist klar, dass diese Gruppen Normalteiler in $G$ bilden. Wir werden später sehen, dass der Abschluss nicht nötig ist.
\end{remark}

\begin{proposition}\label{1.16}
Sei $G$ eine endlich erzeugte pro"~$p$ Gruppe. Dann ist $P_i(G)\subset G$ offen für alle $i$ und $\{P_i(G)\mid i\geq 1\}$ bildet eine offene Umgebungsbasis der $1\in G$.
\end{proposition}

\begin{proof}
Setze $G_n:=P_n(G)$. Offenbar ist $G_1=G$ endlich erzeugt und offen in $G$. Sei nun $n\geq 1$ und $G_n$ endlich erzeugt und offen in $G$. Satz~\ref{1.14} zeigt, dass $\Phi(G_n)\subset_\text{o} G_n$. Da $\Phi(G_n)\subset G_{n+1}\subset G_n$, ist $G_{n+1}\subset_\text{o}G_n\subset_\text{o}G$ und Satz~\ref{1.7} zeigt, dass $G_{n+1}$ endlich erzeugt ist.

Es genügt nun zu zeigen, dass jeder offene Normalteiler von $G$ ein $G_i$ für ein $i$ enthält. Sei $N\lhd_\text{o}G$. Dann ist $G/N$ eine endliche $p$-Gruppe. Analog zu Satz~\ref{0.4.ii} zeigt man $P_k(G/N)=1$ für ein $k$. Also reicht es zu zeigen, dass $P_n(G)N/N\subset P_n(G/N)$ für alle $n$ gilt. Für $n=1$ ist die Aussage klar. Es gelte $G_nN/N\subset P_n(G/N)$ für ein $n$. Sei $M/N=P_{n+1}(G/N)$. Dann ist $M$ abgeschlossen in $G$ und $G_n^p[G_n,G]N\subset M$, also $G_{n+1}N\subset M$. Die Behauptung folgt per vollständiger Induktion.
\end{proof}

\begin{proposition}\label{1.19}
Ist $G$ eine endlich erzeugte pro"~$p$ Gruppe, so ist $[G,G]\subset G$ abgeschlossen.
\end{proposition}

\begin{proof}
Sei $G$ topologisch von $\{a_1,\ldots,a_d\}$ erzeugt. Setze:
\[X:=\{[g_1,a_1]\cdots[g_d,a_d]\mid g_1,\ldots,g_d\in G \} \]
Dann ist $X$ als Bild der stetigen Abbildung $G^{(d)}\to G,\ (g_1,\ldots,g_d)\mapsto \prod [g_i,a_i]$ kompakt und somit abgeschlossen. Sei $N\lhd_\text{o}G$. Dann ist $G/N$ eine endliche $p$"~Gruppe mit $G/N=\langle a_1N,\ldots,a_dN\rangle$. Nach Lemma~\ref{1.23} gilt $[G/N,G/N]=XN/N$, also $[G,G]N=XN$. Also folgt:
\[[G,G]\subset\bigcap_{N\lhd_\text{o}G}XN=\overline{X}=X \]
Da $X\subset [G,G]$, ist $[G,G]=X$ abgeschlossen.
\end{proof}

\begin{corollary}\label{1.20}
Ist $G$ eine endlich erzeugte pro"~$p$ Gruppe, so gilt für alle $i$:
\[\Phi(G)=G^p[G,G],\quad P_{i+1}(G)=P_i(G)^p[P_i(G),G] \]
\end{corollary}

\begin{proof}
Setze $G_i:=P_i(G)$. Die Menge $G^{\{p\}}=\{g^p\mid g\in G\}$ ist als Bild der stetigen Abbildung $G\to G,\ g\mapsto g^p$ kompakt und somit abgeschlossen in $G$. Da $G/[G,G]$ abelsch ist, folgt $G^p[G,G]=G^{\{p\}}[G,G]$. Da $[G,G]$ nach \ref{1.19} abgeschlossen ist, sehen wir, dass $G^p[G,G]$ abgeschlossen ist, also gleich $\Phi(G)$.

Satz~\ref{1.16} und Satz~\ref{1.14} zeigen, dass $G_i$ eine endlich erzeugte pro"~$p$ Gruppe ist und daher $\Phi(G_i)\subset_\text{o}G_i$ gilt für alle $i$. Den ersten Teil auf $G_i$ angewendet ergibt:
\[\Phi(G_i)=G_i^p[G_i,G_i]\subset G_i^p [G_i,G]\]
Es folgt $G_i^p[G_i,G]\subset_\text{o}G_i\subset_\text{o}G$, also ist $G_i^p[G_i,G]$ abgeschlossen in $G$ und gleich $G_{i+1}$. Die Behauptung folgt per Induktion.
\end{proof}

\paragraph{} Wir wollen nun zeigen, dass jede Untergruppe einer endlich erzeugten pro"~$p$ Gruppe von endlichem Index offen ist. Als Konsequenz werden wir sehen, dass die Topologie einer endlich erzeugten pro"~$p$ Gruppe schon von ihrer Gruppenstruktur eindeutig bestimmt wird.

\begin{lemma}\label{1.18}
Sei $G$ eine pro"~$p$ Gruppe und $K\lhd G$ ein Normalteiler von endlichem Index. Dann ist $[G:K]$ eine $p$-Potenz.
\end{lemma}

\begin{proof}
Sei $[G:K]=m=p^r q$ mit $p\nmid q$ und setze $X:=\{g^m\mid g\in G\}$. Dann ist $X\subset G$ als Bild der stetigen Abbildung $G\to G,\ g\mapsto g^m$ abgeschlossen und $X\subset K$. Sei nun $g\in G$ und $N\lhd_\text{o}G$. Wähle ein $e$, so dass $g^{p^e}\in N$. Wir können o.B.d.A. $e\geq r$ annehmen.  Es existieren $a,b\in\mathbb{Z}$ mit $aq+bp^{e-r}=1$, also $am+bp^e=p^r$. Es folgt:
\[g^{p^r}=(g^a)^m(g^{p^e})^b\in XN \]
Da $r$ unabhängig von $N$ ist, folgt $g^{p^r}\in \bigcap_{N\lhd_\text{o}G} XN=X\subset K$. Somit ist $G/K$ eine $p$-Gruppe.
\end{proof}

\begin{lemma}
Sei $G$ eine Gruppe und $H\subset G$ eine Untergruppe von endlichem Index. Dann enthält $H$ einen Normalteiler in $G$ von endlichem Index.
\end{lemma}

\begin{proof}
Setze $n:=[G:H]$. Für jedes $g\in G$ ist die Abbildung $\sigma_g : G/H\to G/H,\ \overline{x}\mapsto \overline{gx}$ eine Bijektion. Fassen wir $\sigma_g$ als Element der symmetrischen Gruppe $\mathfrak{S}_n$ auf, so ist die Abbildung $\phi:G\to \mathfrak{S}_n,\ g\mapsto \sigma_g$ ein Homomorphismus. Für $N:=\ker(\phi)\lhd G$ gilt $N\subset H$. Ferner gilt $[G:N]=\#\operatorname{im}(\phi)\leq\#\mathfrak{S}_n = n!$.
\end{proof}

\begin{theorem}\label{1.17}
Ist $G$ eine endlich erzeugte pro"~$p$ Gruppe, so ist jede Untergruppe von endlichem Index in $G$ offen.
\end{theorem}

\begin{proof}
Sei $K\subsetneq G$ eine Untergruppe von endlichem Index. Nach vorherigem Lemma können wir $K\lhd G$ annehmen. Per Induktion nach $[G:K]$ können wir $K\subset_\text{o}M$ annehmen, wenn $M$ eine endlich erzeugte pro"~$p$ Gruppe mit $K\subset M\subsetneq G$ ist. Sei $M:=G^p[G,G]K=\Phi(G)K$ Da $\Phi(G)$ nach Satz~\ref{1.14} offen ist, ist $M\subset_\text{o}G$ und daher nach Satz~\ref{1.7} endlich erzeugt. Nach Lemma~\ref{1.18} ist $G/K$ eine endliche $p$-Gruppe. Ist also $M=G$, so folgt $\Phi(G/K)=\Phi(G)K/K=G/K$, ein Widerspruch, da nichttriviale, endliche $p$-Gruppen stets echte maximale Untergruppen besitzen. Nach der Induktionsvoraussetzung folgt $K\subset_\text{o}M\subset_\text{o}G$.
\end{proof}

\begin{corollary}\label{1.21}
Jeder abstrakte Gruppenhomomorphismus $\theta :G\to H$ von einer endlich erzeugten pro"~$p$ Gruppe $G$ in einer proendlichen Gruppe $H$ ist stetig.
\end{corollary}

\begin{proof}
Ist $K\subset_\text{o} H$ eine Untergruppe, so ist $\theta^{-1}(K)$ eine Untergruppe vom Index kleiner gleich $[H:K]$, also offen in $G$ nach Theorem~\ref{1.17}. Da solche $K$ eine offene Umgebungsbasis der $1\in H$ bilden, ist $\theta$ stetig.
\end{proof}

\begin{remark}
Somit ist jeder bijektiver Homomorphismus von einer endlich erzeugten pro"~$p$ Gruppe $G$ in eine proendliche Gruppe stets stetig und somit ein Homöomorphismus, da die Räume kompakt und hausdorffsch sind.

Insbesondere ist $G$ mit einer ggf. anderen proendlichen Topologie homöomorph zu $G$ mit ihrer ursprünglichen Topologie via der Identitätsabbildung. In diesem Sinne ist die Topologie einer endlich erzeugten pro"~$p$ Gruppe eindeutig durch ihre Gruppenstruktur bestimmt.
\end{remark}

\section{Erzeuger \& Relationen}

Die Golod-Shafarevich Ungleichung macht Aussagen über die minimale Anzahl der \textit{Erzeugern} und \textit{Relationen} von pro"~$p$ Gruppen. Was diese in der Kategorie der pro"~$p$ Gruppen sind, werden wir in diesem Abschnitt definieren.

\begin{definition}
Sei $\Gamma$ eine gewöhnliche Gruppe und $\Lambda$ die Familie aller Normalteiler vom $p$-Potenz-Index in $\Gamma$. Dann bilden die Quotienten $(\Gamma/N)_{N\in\Lambda}$ ein projektives System endlicher $p$-Gruppen. Die \textit{pro"~$p$ Vervollständigung}\index{pro-$p$ Vervollständigung} von $\Gamma$ ist definiert als:
\[\widehat{\Gamma} :=\varprojlim_{N\in\Lambda} \Gamma/N \]
\end{definition}

\begin{proposition} \textit{(Universaleigenschaft der pro"~$p$ Vervollständigung)}
Sei $\eta : \Gamma\to\widehat{\Gamma}$ die Abbildung, die ein Element $g\in\Gamma$ auf die konstante Folge $(g)_{N\in\Lambda}$ schickt. Für alle Homomorphismen $\phi:\Gamma\to G$ in eine pro"~$p$ Gruppe $G$ gibt es einen eindeutig bestimmten, stetigen Homomorphismus $\widehat{\phi}:\widehat{\Gamma}\to G$ mit $\phi = \widehat{\phi}\circ\eta$.
\end{proposition}

\begin{proof}
Sei $N\lhd_\text{o}G$ und betrachte $\varphi_N:\Gamma\stackrel{\phi}{\to} G\to G/N$. Der Kern $\ker(\varphi_N)$ ist ein Normalteiler in $\Gamma$ vom $p$-Potenz-Index, da $\varphi_N$ eine Injektion $\Gamma/\ker(\varphi_N)\to G/N$ induziert. Die Abbildungen $\widehat{\varphi}_N:\widehat{\Gamma}\to\Gamma/\ker(\varphi_N)\to G/N$ sind kompatibel, stetig und setzen sich zu einer stetigen Abbildung $\widehat{\varphi}:\widehat{\Gamma}\to G$ zusammen. $\phi=\widehat{\phi}\circ\eta$ ist klar. Ist $\widehat{\psi}$ eine weitere solche Abbildung, so stimmen $\widehat{\psi}$ und $\widehat{\phi}$ auf der dichten Teilmenge $\eta(\Gamma)\subset G$ überein. Da diese stetig sind und $G$ hausdorffsch, stimmen sie auf ganz $\widehat{\Gamma}$ überein.
\end{proof}

\begin{definition}
Sei $X$ eine endliche Menge und $F(X)$ die freie Gruppe auf $X$. Dann heißt die pro"~$p$ Vervollständigung $\widehat{F}(X) := \widehat{F(X)}$ von $F(X)$ die \textit{freie pro"~$p$ Gruppe}\index{freie pro-$p$ Gruppe} auf $X$.
\end{definition}

\begin{remark}
Für eine unendliche Menge $X$ muss die Definition einer freien pro"~$p$ Gruppe auf $X$ ein wenig modifiziert werden. Siehe z.B. \cite{Wil98}.
\end{remark}

\begin{example}
Die pro"~$p$ Vervollständigung von $\mathbb{Z}$ ist gerade $\mathbb{Z}_p$ und in diesem Fall ist die natürliche Abbildung $\eta:\mathbb{Z}\to\mathbb{Z}_p$ injektiv, da $\ker(\eta)=\bigcap_{n\in\mathbb{N}} p^n\mathbb{Z}=0$. Daher ist $\mathbb{Z}_p$ die freie pro"~$p$ Gruppe auf einem Erzeuger.
\end{example}

\begin{proposition} \textit{(Universaleigenschaft der freien pro"~$p$ Gruppe auf einer endlichen Menge $X$)} 
Jede Abbildung $\phi:X\to G$ in eine pro"~$p$ Gruppe $G$ setzt sich eindeutig zu einem stetigen Homomorphismus $\widehat{\phi} : \widehat{F}(X)\to G$ fort.
\end{proposition}

\begin{proof}
$\phi:X\to G$ setzt sich eindeutig zu einem Homomorphismus $\phi:F(X)\to G$ fort. Nach der Universaleigenschaft der pro"~$p$ Vervollständigung erhalten wir einen eindeutigen, stetigen Homomorphismus $\widehat{\phi}:\widehat{F}(X)\to G$ mit $\widehat{\phi}\circ\eta = \phi$, wobei $\eta: F(X)\to\widehat{F}(X)$ die natürliche Abbildung bezeichnet. Nach Einschränkung auf $X\subset F(X)$ erhalten wir $\widehat{\phi}\circ i=\phi$ mit $i:=\eta|_X$. Zu zeigen ist nur noch, dass $i$ eine Injektion ist.

Sei also $x,y\in X$ mit $x\neq y$. Betrachte die Abbildung $\varphi:X\to\mathbb{Z}_p$, die $x$ nach $1\in\mathbb{Z}_p$ und alle anderen Elemente nach $0\in\mathbb{Z}_p$ schickt. Nach dem erstem Teil erhalten wir einen stetigen Homomorphismus $\widehat{\varphi}:\widehat{F}(X)\to\mathbb{Z}_p$ mit $\widehat{\varphi}\circ i=\varphi$. Nach Konstruktion ist $\varphi(x)\neq\varphi(y)$, daher folgt $i(x)\neq i(y)$.
\end{proof}

\begin{remark}
Wir haben also eine Inklusion $X\subset \widehat{F}(X)$ via $i$. Da $\eta(F(X))\subset\widehat{F}(X)$ dicht liegt, folgt $\overline{\langle X\rangle}=\widehat{F}(X)$, d.h. $\widehat{F}(X)$ wird von $X$ topologisch erzeugt.
\end{remark}

\iffalse
\begin{proposition}\label{prop:freepropgroup2}
Sei $Y\subset X$ endliche Mengen und $F(Y)\subset F(X)$ die freie Gruppen auf $Y$ bzw. $X$. Setze $K$ als den kleinsten Normalteiler in $F(X)$, der $Y$ enthält und $N$ als den kleinsten Normalteiler in $\widehat{F}(X)$, der $Y$ enthält. Dann ist $F(X)/K$ eine freie Gruppe und $\widehat{F}(X)/\overline{N}$ eine freie pro"~$p$ Gruppe auf $\#X-\#Y$ Erzeugern.
\end{proposition}

\begin{proof}
Wir zeigen, dass $\widehat{F}(X)/\overline{N}$ die Universaleigenschaft von $\widehat{F}(X\setminus Y)$ erfüllt. Sei also $X\setminus Y\to G$ eine Abbildung in eine pro"~$p$ Gruppe. In dem wir Elemente aus $Y$ auf $1\in G$ schicken, erhalten wir eine Abbildung $X\to G$ und somit einen eindeutigen, stetigen Homomorphismus $\phi:\widehat{F}(X)\to G$. Offensichtlich gilt $N\subset\ker(\phi)$. Da $G$ hausdorffsch, ist $\ker(\phi)= \phi^{-1}(\{1\})$ abgeschlossen in $G$. Daher gilt $\overline{N}\subset\ker(\phi)$. $\phi$ induziert eine Abbildung $\tilde{\phi}:\widehat{F}(X)/\overline{N}\to G$. Für $F(X)/K$ analog.
\end{proof}

\begin{proposition}\label{prop:freepropgroup3}
Seien $Y\subset X$ endliche Mengen. Dann induziert die Inklusion $F(Y)\to F(X)$ auf den abstrakten freien Gruppen eine Inklusion $\widehat{F}(Y)\to\widehat{F}(X)$.
\end{proposition}

\begin{proof}
Sei $N\lhd F(Y)$ vom $p$-Potenz-Index. Wir zeigen zunächst, dass ein Normalteiler $M\lhd F(X)$ vom $p$-Potenz-Index existiert mit $N=F(Y)\cap M$. Betrachte den kleinsten Normalteiler $K\lhd F(X)$, der $X\setminus Y$ enthält und das folgende kommutative Diagramm:
\[\xymatrix{
F(X) \ar[r]^-\pi & F(X)/K\\
F(Y) \ar[u]^i \ar[ru]_\phi
} \]
wobei $\phi$ der Isomorphismus aus Satz~\ref{prop:freepropgroup2}, $\pi$ die kanonische Projektion und $i$ die natürliche Inklusion bezeichnet. Setze $M:= \pi^{-1}(\phi(N))\lhd F(X)$. Per Definition gilt: 
\[F(X)/M\cong (F(X)/K)/\phi(N)\cong F(Y)/N\]
Daher ist $M$ vom $p$-Potenz-Index und es gilt $F(Y)\cap M = i^{-1}(M) = N$. Wir erhalten für jedes $M\lhd F(X)$ vom $p$-Potenz-Index kompatible Inklusionen $F(Y)/i^{-1}(M)\hookrightarrow F(X)/M$. Nun läuft $i^{-1}(M)$ über alle Normalteiler in $F(Y)$ vom $p$-Potenz-Index. Unter $\varprojlim$ erhalten wir einen stetigen Homomorphismus $\widehat{F}(Y)\hookrightarrow\widehat{F}(X)$.
\end{proof}
\fi

\begin{definition}
Sei $G$ eine pro"~$p$ Gruppe. Für eine endliche Menge $X$ und eine beliebige Teilmenge $R\subset \widehat{F}(X)$ schreiben wir $G=\langle X;R\rangle$, wenn:
\[G \cong \widehat{F}(X)/\overline{N(R) } \]
wobei $N(R)$ den kleinsten Normalteiler in $\widehat{F}(X)$ bezeichnet, der $R$ enthält. $\langle X;R\rangle$ heißt \textit{Darstellung}\index{Darstellung} von $G$. Ist $d(G)$ endlich, so heißt die Darstellung \textit{minimal}\index{Darstellung!minimal}, wenn $\#X=d(G)$.

Ist $X$ ein endliches topologisches Erzeugendensystem von $G$, so lässt sich die natürliche Inklusion $X\hookrightarrow G$ zu einem surjektiven Homomorphismus $\pi:\widehat{F}(X)\to G$ fortsetzen. Ist $R\subset \ker(\pi)$, so induziert $\pi$ eine Abbildung $\widehat{F}(X)/\overline{N(R) }\to G$. Gilt $\overline{N(R)}=\ker(\pi)$, so ist $\langle X;R\rangle$ eine Darstellung von $G$.
\end{definition}

\begin{definition}
Für eine endlich erzeugte pro"~$p$ Gruppe $G$, definieren wir:
\[t(G):=\inf\{ \#R \mid G= \langle X;R\rangle,\ \#X = d(G) \} \]
\end{definition}

\begin{remark}
Eine endliche $p$"~Gruppe $G$ mag als pro"~$p$ Gruppe eine Darstellung auf $d(G)$ Erzeugern und $t(G)$ Relationen haben, aber die Anzahl an Relationen als abstrakte Gruppe kann deutlich größer als $t(G)$ sein.
\end{remark}

\iffalse
\begin{lemma}\label{4.33}
Sei $G$ eine pro"~$p$ Gruppe und $K\lhd_\text{o}G$. Ist $K$ endlich darstellbar, so auch $G$.
\end{lemma}

\begin{proof}
Sei zunächst $[G:K]=p$. Dann ist $G=\langle y\rangle K$ mit $y^p\in K$. Sei $\langle X;R\rangle$ eine endliche Darstellung von $K$ und $\pi:\widehat{F}(X)\to K$ der zugehörige Epimorphismus. Es existiert ein $v\in \widehat{F}(X)$ mit $y^p=\pi(v)$. Für alle $x\in X$ gibt es ein $w_x\in \widehat{F}(X)$, so dass $y^{-1}\pi(x)y=\pi(w_x)$. Nun setze $Y=X\cup\{t\}$ mit $t\not\in X$ und definiere ein Epimorphismus $\overline{\pi}:\widehat{F}(Y)\to G$ durch $\overline{\pi}(x)=\pi(x)$ für alle $x\in X$ und $\overline{\pi}(t)=y$. Nach Satz~\ref{prop:freepropgroup3} haben wir eine Inklusion $\widehat{F}(X)\subset\widehat{F}(Y)$. Setze:
\[S:=\{t^pv^{-1}\}\cup\{t^{-1}xtw_x^{-1}\mid x\in X\}\subset \widehat{F}(Y) \]
Sei $N=\overline{ N(R\cup S)}\subset \widehat{F}(Y)$ und setze $M:=\ker(\overline{\pi})$. Dann gilt $N\subset M$. Die Re\-la\-tionen $S=1$ in $\widehat{F}(Y)/N$ zeigen $\widehat{F}(X)N\lhd \widehat{F}(Y)$ und $[\widehat{F}(Y):\widehat{F}(X)N]\leq p$. Da $\overline{\pi}(\widehat{F}(Y))=G$ und $\overline{\pi}(\widehat{F}(X)N)=K$, folgt $M\subset \widehat{F}(X)N$, also $M=(M\cap \widehat{F}(X))N$. Nun gilt:
\[M\cap \widehat{F}(X)=\ker(\pi)=\overline{N(R)} \subset N \]
Somit gilt $M=N$. Also ist $\langle Y;R\cup S\rangle$ eine endliche Darstellung von $G$.

Sei nun $[G:K]=p^k$ für ein $k>1$. Da $G/K$ eine $p$"~Gruppe ist, existiert eine Reihe von Normalteilern in $G/K$:
\[ 1 = G_0/K\subset G_1/K\subset\ldots\subset G_k/K=G/K \]
mit $\#G_i/K=p^i$ und $(G_i/K)/(G_{i-1}/K)\cong\mathbb{Z}/p\mathbb{Z}$ für alle $i$. Insbesondere gilt $G_i\lhd_\text{o} G$ und $[G_i:G_{i-1}]=p$ für alle $i$. Da $K=G_0$, folgt nun, dass $G_1$ endlich darstellbar ist. Induktiv folgt, dass $G=G_k$ endlich darstellbar ist.
\end{proof}
\fi

\chapter{Analytische pro"~$p$ Gruppen}

In diesem Kapitel werden wir uns \textit{$p$-adisch analytische Gruppen} zuwenden, d.h. Gruppen, die eine Mannigfaltigkeitsstruktur über $\mathbb{Q}_p$ tragen und deren Gruppenoperationen analytische Funktionen sind. Es stellt sich heraus, dass man solche Mannigfaltigkeiten rein gruppentheoretisch beschreiben kann. Lazard zeigt dies in seinem Artikel \cite{Laz65}. Zusammen mit seinem Resultat zeigen Lubotzky und Mann in \cite{LM87} über \textit{potenzreiche pro"~$p$ Gruppen}:

\begin{theorem}\label{thm:analytische-pro-p-gruppe}
Eine pro"~$p$ Gruppe $G$ ist genau dann $p$-adisch analytisch, wenn $G$ endlich erzeugt ist und eine offene, potenzreiche pro"~$p$ Gruppe $H$ enthält, die normal in $G$ ist.
\end{theorem}

\paragraph{} Nach Satz~\ref{1.7} ist $H$ endlich erzeugt. Wir werden in dieser Arbeit hauptsächlich diese Charakterisierung einer analytischen pro"~$p$ Gruppe verwenden. Um die Struktur einer analytischen pro"~$p$ Gruppe zu verstehen, wenden wir uns daher zunächst potenzreichen Gruppen zu.

\section{Potenzreiche $p$"~Gruppen}

Wir wollen uns, genau wie bei pro"~$p$ Gruppen, zuerst auf endliche $p$"~Gruppen beschränken. Wir werden sehen, dass potenzreiche $p$"~Gruppen (und später auch potenzreiche pro"~$p$ Gruppen) viele einfache strukturelle Merkmale mit abelschen Gruppen teilen.

\begin{definition}
Sei $G$ eine endliche $p$"~Gruppe.
\begin{enumerate}[(i)]
\item $G$ heißt \textit{potenzreich}\index{potenzreiche Gruppe}, wenn $p$ ungerade und $[G,G]\subset G^p$, oder $p=2$ und $[G,G]\subset G^4$ gilt.
\item Sei $N\subset G$ eine Untergruppe. Dann heißt $N$ \textit{potenzreich eingebettet}\index{potenzreich eingebettet} in $G$, wenn $p$ ungerade und $[N,G]\subset N^p$, oder $p=2$ und $[N,G]\subset N^4$ gilt. Wir schreiben in diesem Fall $N\lhd_\text{p.e.}G$.
\end{enumerate}
\end{definition}

\begin{remark}
Ist $p$ ungerade, so folgt aus Satz~\ref{1.13} $G^p=\Phi(G)$. Ist $N\lhd_\text{p.e.}G$, so ist $N$ insbesondere potenzreich und es gilt $N\lhd G$, da $g^{-1}n g=n[n,g]\in N$ für alle $n\in N$ und $g\in G$.
\end{remark}

\begin{lemma}\label{2.2}
Sei $G$ eine endliche $p$"~Gruppe und $N,K,W\lhd G$ mit $N\subset W$. Dann gilt:
\begin{enumerate}[(i)]
\item Ist $N\lhd_\text{p.e.}G$, so gilt $NK/K\lhd_\text{p.e.}G/K$.
\item Ist $p$ ungerade und $K\subset N^p$, oder $p=2$ und $K\subset N^4$, so gilt: \[N\lhd_\text{p.e.}G \iff N/K\lhd_\text{p.e.}G/K\]
\item Ist $N\lhd_\text{p.e.}G$ und $x\in G$, so ist $\langle N,x\rangle$ potenzreich.
\item Ist $N$ nicht potenzreich in $W$ eingebettet, so gibt es ein $J\lhd G$ mit:
\begin{align*}
N^p[[N,W],W] \subset J\subsetneq N^p[N,W],\quad [N^p[N,W]:J]=p,&\quad\text{wenn $p$ ungerade}\\
\text{oder}\quad N^4[N,W]^2[[N,W],W]\subset J\subsetneq N^4[N,W],\quad [N^4[N,W]:J]=2,&\quad\text{wenn $p=2$}
\end{align*}
\end{enumerate}
\end{lemma}

\begin{proof}
(i) folgt direkt aus der Definition. Für (ii) sei $K\subset N^p$ bzw. $K\subset N^4$ und $N/K\lhd_\text{p.e.}G/K$. Dann gilt für alle $n\in N,\ g\in G$ stets $[n,g]\in N^pK=N^p$ bzw. $[n,g]\in N^4$, also $N\lhd_\text{p.e.}G$.

Für (iii) setze $H:=\langle N,x\rangle$. Nun ist $H/[N,H]$ abelsch, da alle Erzeuger miteinander kommutieren. Daher gilt $[H,H]\subset [N,H]$ und es folgt $[H,H]= [N,H]\subset[N,G]\subset N^p\subset H^p$ bzw. $[H,H]\subset H^4$. Somit ist $H$ potenzreich.

Für (iv) sei $p$ ungerade und $M:=N^p[N,W]\neq N^p$. Nun ist $N\lhd M$ und daher $N^p\lhd M$. Nach Satz~\ref{0.4.v} gibt es eine maximale Untergruppe $J/N^p\subset M/N^p$ mit $J/N^p\lhd G/N^p$, also $N^p\subset J$, $J\lhd M$, $J\lhd G$ und $[M:J]=p$ nach Satz~\ref{0.4.iii}. Schließlich ist $M/J$ einfach und nach Lemma~\ref{0.4.v:lemma} $M/J\cap Z(G/J)\neq 1$, also folgt $M/J\subset Z(G/J)$ und:
\[N^p[[N,W],W]\subset N^p[M,G] \subset J\subset M \]
Sei nun $p=2$ und $M:=N^4[N,W]\neq N^4$. Genauso wie oben finden wir ein $J\lhd G$, so dass $N^4\subset J\lhd M$ und $[M:J]=2$. Mit $M/J\subset Z(G/J)$ folgt:
\[N^4[N,W]^2[[N,W],W]\subset N^4 M^2 [M,G]\subset J\subset M\qedhere \]
\end{proof}

\begin{remark}
Um $N\lhd_\text{p.e.}W$ zu zeigen, können wir mit (ii) und (iv) ohne Einschränkung $N^p=1$ und $[[N,W],W]=1$ bzw. $N^4=[N,W]^2=[[N,W],W]=1$ annehmen. Diese Reduktionstechnik werden wir in aller Ausführlichkeit im Beweis des nächsten Satzes sehen.
\end{remark}

\begin{proposition}\label{2.3}
Sei $G$ eine endliche $p$"~Gruppe und $N\subset G$ eine Untergruppe. Ist $N\lhd_\text{p.e.}G$, so folgt auch $N^p\lhd_\text{p.e.}G$.
\end{proposition}

\begin{proof}
Sei $p$ zunächst ungerade. Nach Lemma~\ref{2.2} (ii) können wir durch Reduktion modulo $(N^p)^p$ o.B.d.A. $(N^p)^p=1$ annehmen. Ist $N^p$ nicht potenzreich in $G$ eingebettet, so gibt es nach Lemma~\ref{2.2} (iv) ein $J\lhd G$, so dass:
\[ [[N^p,G],G]\subset J\subsetneq [N^p,G],\quad [[N^p,G]:J]=p \]
Da $N^p\lhd G$, ist $J\subset[N^p,G]\subset N^p$. Nun ist $[(N/J)^p, G/J]= [N^p,G]/J$ eine Gruppe der Ordnung $p$ und damit nichttrivial. Daher ist $N^p/J$ nicht in $G/J$ potenzreich eingebettet und wir können durch Reduktion modulo $J$ zusätzlich $[[N^p,G],G]=1$ annehmen. Es folgt $[[N,G],G]\subset [N^p,G]\subset Z(G)$. Für jedes $x\in N$ und $g\in G$ gilt $[[x,g],vw]=[[x,g],w]w^{-1}\cdot [[x,g],v]w=[[x,g],v]\cdot[[x,g],w]$. Es folgt:
\begin{align*}
[x^p,g] &=\prod_{i=1}^p x^{-(p-i)}[x,g]x^{p-i}= \prod_{i=1}^p [x,g] \cdot[[x,g],x^{p-i}]\\
&=\prod_{i=1}^p [x,g]\cdot [[x,g],x]^{p-i}= [x,g]^p\cdot [[x,g],x]^\frac{p(p-1)}{2}\in [N,G]^p\subset (N^p)^p
\end{align*}
Somit ist $[N^p,G]\subset (N^p)^p$, ein Widerspruch. Sei nun $p=2$. In diesem Fall gilt $[N,G]\subset N^4$ und wir können $(N^2)^4 = [N^2,G]^2= [[N^2,G],G]=1$ annehmen. Für alle $x\in N$ und $g\in G$ gilt:
\[[x^4,g]=[x^2,g][[x^2,g],x^2][x^2,g]=[x^2,g]^2=1 \]
Somit folgt $N^4\subset Z(G)$. Da $[[x,g],x]\in [[N,G],G]\subset [N^4,G]=1$ und $[x,g]\in[N,G]\subset N^4$, folgt:
\[[x^2,g]=[x,g][[x,g],x][x,g]=[x,g]^2\in (N^4)^2=N^{\{8\}}\subset (N^2)^4\qedhere \]
\end{proof}

\paragraph{} Für eine potenzreiche $p$-Gruppe (und später potenzreiche pro"~$p$ Gruppe) haben wir eine noch einfachere Beschreibung der unteren $p$"~Zentralreihe.

\begin{lemma}
Sei $x,y\in G$. Dann gilt: 
\[(xy)^n\equiv x^ny^n[y,x]^{\frac{n(n-1)}{2}}\mod [[G,G],G] \]
\end{lemma}

\begin{proof}
Es ist $[[G,G],G]\lhd G$, da $g^{-1}[[a,b],c]g=[g^{-1}[a,b]g,g^{-1}cg]$ und $[G,G]\lhd G$. Außerdem enthält $Z(G/[[G,G],G])$ das Bild von $[G,G]$ in $G/[[G,G],G]$. Für $n=1$ ist die Aussage klar. Sei nun $n>1$ und die Aussage für alle kleineren $n$ schon bewiesen. Dann gilt:
\begin{align*}
(xy)^n = (xy)^{n-1}xy &\equiv x^{n-1}y^{n-1}[y,x]^{\frac{(n-1)(n-2)}{2}}\cdot xy\\
&\equiv x^ny^n y^{-n}x^{-1}y^{n-1}xy [y,x]^\frac{(n-1)(n-2)}{2}\\
&\equiv x^ny^ny^{-1}[y^{n-1},x] y[y,x]^\frac{(n-1)(n-2)}{2}\\
&\equiv x^ny^n[y,x]^{n-1}[y,x]^\frac{(n-1)(n-2)}{2}\mod [[G,G],G]
\end{align*}
wobei die letzte Kongruenz aus der Tatsache $[y^n,x]=\prod_{i=1}^n y^{-(n-i)}[y,x]y^{n-i}$ folgt.
\end{proof}

\begin{proposition}\label{2.4}
Sei $G$ eine potenzreiche $p$"~Gruppe und $G_i:=P_i(G)$. Dann gilt für alle $i$:
\begin{enumerate}[(i)]
\item $G_{i+1}=G_i^p=\Phi(G_i)$ und $G_i\lhd_\text{p.e.} G$.
\item $x\mapsto x^p$ induziert einen surjektiven Homomorphismus $G_i/G_{i+1}\to G_{i+1}/G_{i+2}$.
\end{enumerate}
\end{proposition}

\begin{proof}
\begin{enumerate}[(i)]
\item Da $G=G_1$ potenzreich ist, gilt $G_1\lhd_\text{p.e.}G$. Sei $G_i\lhd_\text{p.e.}G$ für ein $i\geq 1$. Dann gilt $G_{i+1}=G_i^p[G_i,G]=G_i^p$ und nach Satz~\ref{2.3} folgt $G_{i+1}\lhd_\text{p.e.}G$.
\item Wir haben gezeigt, dass alle $G_i$ potenzreich sind, $G_{i+1}=P_2(G_i)$ und $G_{i+2}=P_3(G_i)$. Nach Umbenennung können wir $i=1$ annehmen. Durch Ersetzen von $G$ mit $G/G_3$ können wir zusätzlich $G_3=1$ annehmen. Dann gilt $[G,G]\subset G_2\subset Z(G)$, also gilt nach vorherigem Lemma für alle $x,y\in G$:
\[(xy)^p=x^py^p[y,x]^{\frac{p(p-1)}{2}} \]
Ist $p$ ungerade, so gilt $p\mid \frac{p(p-1)}{2}$ und $[y,x]^\frac{p(p-1)}{2}\in G_2^p=G_3=1$. Ist $p=2$, so folgt $[G,G]\subset G^4\subset G_3=1$. In beiden Fällen gilt $(xy)^p=x^py^p$. Da $G^p=G_2$ und $(G^p)^p=G_3$, folgt die Behauptung.\qedhere
\end{enumerate}
\end{proof}

\paragraph{} Die folgenden Aussagen zeigen die strukturellen Gemeinsamkeiten mit einer abelschen Gruppe:

\begin{lemma}\label{2.5}
Ist $G=\langle a_1,\ldots,a_d\rangle$ eine potenzreiche $p$-Gruppe, so ist $G^p=\langle a_1^p,\ldots,a_d^p\rangle$.
\end{lemma}

\begin{proof}
Sei $\theta:G/G_2\to G_2/G_3$ der durch $x\mapsto x^p$ induzierter Homomorphismus. Dann ist $G_2/G_3$ erzeugt durch $\{\theta(a_1G_2),\ldots,\theta(a_dG_2) \}$, also $G_2=\langle a_1^p,\ldots,a_d^p\rangle G_3$. Da $G_3=\Phi(G_2)$ und $G_2=G^p$, folgt die Behauptung aus Satz~\ref{1.9}.
\end{proof}

\begin{proposition}\label{2.6}
Ist $G$ eine potenzreiche $p$-Gruppe, so ist jedes Element in $G^p$ eine $p$-te Potenz.
\end{proposition}

\begin{proof}
Per Induktion über $\#G$. Sei $g\in G^p$. Nach Lemma~\ref{2.4} existieren $x\in G$ und $y\in G_3$, so dass $g=x^py$. Setze $H:=\langle G^p,x\rangle$. Da $G^p=G_2\lhd_\text{p.e.}G$, folgt aus Lemma~\ref{2.2} (iii), dass $H$ potenzreich ist. Außerdem gilt $g\in H^p$, da $y\in G_3=(G^p)^p$. Ist $H\neq G$, so gibt uns die Induktionsvoraussetzung, dass $g$ eine $p$-te Potenz in $H$ ist. Ist $H=G$, so ist $G$ zyklisch, da $G=\langle G^p,x\rangle=\langle \Phi(G),x\rangle=\langle x\rangle$ und in diesem Fall ist die Behauptung trivial.
\end{proof}

\begin{theorem}\label{2.7}
Sei $G=\langle a_1,\ldots,a_d\rangle$ und $G_i:=P_i(G)$. Dann gilt für alle $i$:
\begin{enumerate}[(i)]
\item $G_{i+k}=P_{k+1}(G_i)=G_i^{p^k}$ für alle $k\geq 0$. Insbesondere gilt $G_{i+1}=\Phi(G_i)$.
\item $G_i=G^{p^{i-1}}=\{x^{p^{i-1}}\mid x\in G\}=\langle a_1^{p^{i-1}},\ldots,a_d^{p^{i-1}}\rangle\lhd_\text{p.e.}G$
\item $x\mapsto x^{p^k}$ induziert einen surjektiven Homomorphismus $G_i/G_{i+1}\to G_{i+k}/G_{i+k+1}$
\end{enumerate}
\end{theorem}

\begin{proof}
Folgt hauptsächlich per Induktion aus Lemma~\ref{2.4}. Die restlichen Aussagen folgen aus Lemma~\ref{2.5} und Satz~\ref{2.6}.
\end{proof}

\begin{corollary}\label{2.8}
Ist $G=\langle a_1,\ldots,a_d\rangle$ eine potenzreiche $p$-Gruppe, so ist $G=\langle a_1\rangle\cdots\langle a_d\rangle$ ein Produkt von zyklischen Untergruppen.
\end{corollary}

\begin{proof}
Wähle $e$ so, dass $G_e\supsetneq G_{e+1}=1$. Per Induktion über $e$ können wir annehmen, dass $G=\langle a_1\rangle\cdots\langle a_d\rangle G_e$. Nun ist $G_e=\langle a_1^{p^{e-1}},\ldots,a_d^{p^{e-1}}\rangle$ und $G_e\subset Z(G)$.
\end{proof}

\iffalse
\begin{theorem}\label{2.9}
Ist $G$ eine potenzreiche $p$-Gruppe und $H\subset G$ eine Untergruppe, so gilt: \[d(H)\leq d(G)\]
\end{theorem}

\begin{proof}
Per Induktion über $\#G$. Sei $d:=d(G)$ und setze $m:=d(G_2)$, wobei $G_i:=P_i(G)$. Nach Lemma~\ref{2.4} (i) ist $G_2$ potenzreich, d.h. nach der Induktionsvoraussetzung können wir annehmen, dass für $K:=H\cap G_2$ die Ungleichung $d(K)\leq m$ erfüllt.

Betrachte die nach Lemma~\ref{2.4} (ii) surjektive Abbildung $\pi:G/G_2\to G_2/G_3,\ x\mapsto x^p$ und es gilt $\dim_{\mathbb{F}_p}(\ker(\pi))=d-m$. Es folgt $\dim_{\mathbb{F}_p}(\ker(\pi)\cap HG_2/G_2)\leq d-m$, also:
\[\dim_{\mathbb{F}_p}(\pi (HG_2/G_2))\geq \dim_{\mathbb{F}_p}(HG_2/G_2)-(d-m)=m-(d-e) \]
wobei $e:=\dim_{\mathbb{F}_p}(HG_2/G_2)\leq d$. Sei $h_1,\ldots,h_e\in H$, so dass $HG_2=\langle h_1,\ldots,h_e\rangle G_2$. Da $\Phi(K)\subset K^p\subset G_3$ gilt, induziert $K\to KG_3/G_3$ die Surjektion $\phi:K/\Phi(K)\to KG_3/G_3$. Das Bild des $\mathbb{F}_p$"~Unterraums $V:=\langle\overline{h_1^p},\ldots,\overline{h_e^p}\rangle\subset K/\Phi(K)$ unter $\phi$ ist gerade $\pi(HG_2/G_2)$, daher ist $\dim_{\mathbb{F}_p}(V)\geq \dim_{\mathbb{F}_p}(\pi(HG_2/G_2))\geq m-(d-e)$. Da $d(K)\leq m$, können wir $y_1,\ldots,y_{d-e}\in K$ finden mit:
\[K=\langle h_1^p,\ldots,h_e^p,y_1,\ldots,y_{d-e}\rangle\Phi(K)=\langle h_1^p,\ldots,h_e^p,y_1,\ldots,y_{d-e}\rangle \]
Somit folgt $H=H\cap\langle h_1,\ldots,h_e\rangle G_2=\langle h_1,\ldots,h_e\rangle K=\langle h_1,\ldots,h_e,y_1,\ldots,y_{d-e}\rangle$ und daher $d(H)\leq d$.
\end{proof}
\fi

\section{Potenzreiche pro"~$p$ Gruppen}

Wir wollen nun die Aussagen, die wir für potenzreiche $p$"~Gruppen hergeleitet haben, ebenfalls für potenzreiche pro"~$p$ Gruppen beweisen. Dabei werden wir uns bei den meisten Aussagen auf endlich erzeugte pro"~$p$ Gruppen beschränken, da die untere $p$-Zentralreihe in diesem Fall offene Untergruppen sind.

\begin{definition}
Sei $G$ eine pro"~$p$ Gruppe.
\begin{enumerate}[(i)]
\item $G$ heißt \textit{potenzreich}\index{potenzreiche Gruppe}, wenn $p$ ungerade und $[G,G]\subset \overline{G^p}$, oder $p=2$ und $[G,G]\subset \overline{G^4}$ gilt.
\item Sei $N\subset_\text{o} G$ eine Untergruppe. Dann heißt $N$ \textit{potenzreich eingebettet}\index{potenzreich eingebettet} in $G$, wenn $p$ ungerade und $[N,G]\subset \overline{N^p}$, oder $p=2$ und $[N,G]\subset \overline{N^4}$ gilt. Wir schreiben in diesem Fall $N\lhd_\text{p.e.}G$.
\end{enumerate}
\end{definition}

\begin{proposition}\label{3.2}
Sei $G$ eine pro"~$p$ Gruppe und $N\subset_\text{o}G$ eine Untergruppe. Dann ist $N\lhd_\text{p.e.}G$, genau dann wenn $NK/K\lhd_\text{p.e.} G/K$ ist für alle $K\lhd_\text{o}G$. 
\end{proposition}

\begin{proof}
Ist $N\lhd_\text{p.e.}G$, so gilt $[N,G]\subset\overline{N^p}=\bigcap_{K\lhd_\text{o}G}N^pK$ nach Satz~\ref{1.2.iii}. Da für alle $K\lhd_\text{o}G$ stets $[N,G]\subset N^pK$ gilt, folgt $NK/K\lhd_\text{p.e.}G/K$.

Sei nun umgekehrt $NK/K\lhd_\text{p.e.}G/K$ für alle $K\lhd_\text{o}G$. Dann folgt aus $[NK/K,G/K]\subset N^pK/K$ stets $[N,G]\subset [NK,G]\subset N^pK$. Somit ist $N\lhd_\text{p.e.}G$, da:
\[[N,G]\subset\bigcap_{K\lhd_\text{o}G}N^pK=\overline{N^p}\qedhere \]
\end{proof}

\begin{remark}
Ist $G$ eine potenzreiche pro"~$p$ Gruppe, so ist $G/K$ potenzreich für alle $K\lhd_\text{o}G$.
\end{remark}

\iffalse
\begin{corollary}
Eine topologische Gruppe $G$ ist genau dann eine potenzreiche pro"~$p$ Gruppe, wenn $G$ ein projektiver Limes endlicher potenzreicher $p$-Gruppen ist, in der alle Ü\-ber\-gangs\-ab\-bil\-dun\-gen surjektiv sind.
\end{corollary}

\begin{proof}
Sei $G$ eine potenzreiche pro"~$p$ Gruppe. Dann ist $G\cong \varprojlim_{N\lhd_\text{o}G} G/N$ mit endlichen $p$-Gruppen $G/N$. Nach Satz~\ref{3.2} ist $G/N$ potenzreich. Sei umgekehrt $G=\varprojlim_{\lambda\in\Lambda}G_\lambda$. Dann ist $G$ eine pro"~$p$ Gruppe. Ist $K\lhd_\text{o}G$, so ist $G/K$ ein Quotient von gewissen $G_\lambda$, also potenzreich. Somit ist nach Satz~\ref{3.2} auch $G$ potenzreich.
\end{proof}
\fi

\begin{lemma}\label{3.4}
Sei $G$ eine endlich erzeugte, potenzreiche pro"~$p$ Gruppe. Dann gilt:
\[\Phi(G)=G^p=\{g^p\mid g\in G\}\subset_\text{o}G \]
\end{lemma}

\iffalse
\begin{proof}
Sei $g\in\overline{G^p}$. Dann ist $gN\in (G/N)^p$ für alle $N\lhd_\text{o}G$, also ist $gN$ eine $p$-te Potenz in $G/N$ nach Satz~\ref{2.6}. Also ist $g$ eine $p$-te Potenz in $G$ nach Lemma~\ref{Ex1.6}. Somit ist $\overline{G^p}\subset G^p$, also enthält $G^p=\overline{G^p}$ nur $p$-te Potenzen in $G$. Da $[G,G]\subset\overline{G^p}$, ist $G^p=\Phi(G)=P_2(G)$ und offen nach Satz~\ref{1.16}.
\end{proof}
\fi

\begin{proof}
Sei $g\in\overline{G^p}$. Für alle $N\lhd_\text{o}G$ ist $gN\in (G/N)^p$ nach Satz~\ref{2.6} eine $p$-te Potenz. Setze $X_N:=\{hN\in G/N\mid h^pN=gN \}\neq\varnothing$ für $N\lhd_\text{o}G$. Diese bilden bzgl. den natürlichen Projektionen ein projektives System. Nach Korollar~\ref{1.4cor} gibt es ein $h\in\varprojlim X_N\subset G$. Offenbar gilt $h^p=g$ und daher gilt $\Phi(G)= \overline{G^p}=G^p=\{g^p\mid g\in G\}$.
\end{proof}

\begin{corollary}\label{3.5}
Sei $G$ eine endlich erzeugte, potenzreiche pro"~$p$ Gruppe. Dann gilt für alle $i$:
\[G^{p^i}=(G^{p^{i-1}})^p=\{g^{p^i}\mid g\in G\}\subset_\text{o} G^{p^{i-1}} \]
\end{corollary}

\begin{proof}
Der Fall $i=1$ folgt aus Lemma~\ref{3.4}. Der allgemeine Fall folgt per Induktion, wenn wir $G$ mit $G^{p^{i-1}}$ ersetzen.
\end{proof}

\begin{theorem}\label{3.6}
Sei $G=\overline{\langle a_1,\ldots,a_d\rangle}$ eine endlich erzeugte, potenzreiche pro"~$p$ Gruppe. Setze $G_i:=P_i(G)$ für alle $i$. Dann gilt:
\begin{enumerate}[(i)]
\item $G_{i+k}=P_{k+1}(G_i)=G_i^{p^k}$ für alle $k\geq 0$. Insbesondere gilt $G_{i+1}=\Phi(G_i)$.
\item $G_i=G^{p^{i-1}}=\{g^{p^{i-1}}\mid g\in G\}=\overline{\langle a_1^{p^{i-1}},\ldots,a_d^{p^{i-1}}\rangle}\lhd_\text{p.e.}G$
\item $x\mapsto x^{p^k}$ induziert einen surjektiven Homomorphismus $G_i/G_{i+1}\to G_{i+k}/G_{i+k+1}$.
\end{enumerate}
\end{theorem}

\begin{proof}
Die zweite Gleichung in (ii) folgt aus Korollar~\ref{3.5}. Sei $N\lhd_\text{o}G$. Dann ist $G/N$ eine potenzreiche $p$-Gruppe. Aus Theorem~\ref{2.7} folgt $G^{p^{i-1}}N/N\lhd_\text{p.e.}G/N$. Daher gilt $G_i=G^{p^{i-1}} \lhd_\text{p.e.}G$ nach Satz~\ref{3.2}. Die restlichen Aussagen sind mit Korollar~\ref{3.5} analog zu Theorem~\ref{2.7}.
\end{proof}

\begin{proposition}\label{3.7}
Ist $G=\overline{\langle a_1,\ldots,a_d\rangle}$ eine potenzreiche pro"~$p$ Gruppe, so ist $G=\overline{\langle a_1\rangle}\cdots\overline{\langle a_d\rangle}$ ein Produkt von prozyklischen pro"~$p$ Gruppen.
\end{proposition}

\begin{proof}
Sei $A:=\overline{\langle a_1\rangle}\cdots\overline{\langle a_d\rangle}$. Als endliches Produkt von abgeschlossenen Teilmengen in $G$ ist $A\subset G$ abgeschlossen. Korollar~\ref{2.8} zeigt $AN/N=G/N$ für alle $N\lhd_\text{o}G$. Somit folgt:
\[A=\bigcap_{N\lhd_\text{o}G}AN =\bigcap_{N\lhd_\text{o}G}GN=G\qedhere\]
\end{proof}

\iffalse
\paragraph{} Ist $\Gamma$ eine abstrakte, zyklische Gruppe mit einem Erzeuger $g\in\Gamma$, so ist jedes Element in $\Gamma$ von der Form $g^\lambda$ mit $\lambda\in\mathbb{Z}$. Ähnlich kann man jedes Element einer prozyklischen pro"~$p$ Gruppe als $p$-adische Potenz eines topologischen Erzeugers darstellen:

\begin{lemma}\label{1.24}
Sei $G$ eine pro"~$p$ Gruppe, $g\in G$ und $(a_i),(b_i)$ $p$"~adisch konvergente Folgen in $\mathbb{Z}$, die den gleichen Grenzwert in $\mathbb{Z}_p$ besitzen. Dann konvergieren $(g^{a_i})$ und $(g^{b_i})$ in $G$ und besitzen den gleichen Grenzwert.
\end{lemma}

\begin{proof}
Sei $N\lhd_\text{o}G$. Dann ist $[G:N]=p^j$ für ein $j$. Für alle hinreichend große $i$ und $k$ gilt $a_i\equiv a_k\mod p^j$, also $g^{a_i}\equiv g^{a_k}\mod N$. Somit ist $(g^{a_i})$ eine Cauchy-Folge in $G$ und konvergiert daher gegen ein $g_1\in G$. Genauso konvergiert $(g^{b_i})$ gegen ein $g_2\in G$. Für hinreichend große $k$ gilt $b_k\equiv a_k\mod p^j$, $g^{b_k}\equiv g_2\mod N$ und $g^{a_k}\equiv g_1\mod N$. Daher folgt:
\[g_1^{}g_2^{-1}\equiv g^{a_k-b_k}\equiv 1\mod N \]
Da $N$ beliebig war, folgt $g_1^{}g_2^{-1}\in\bigcap_{N\lhd_\text{o}G} N=1$.
\end{proof}

\begin{definition}
Sei $G$ eine pro"~$p$ Gruppe, $g\in G$ und $\lambda\in\mathbb{Z}_p$. Dann setzen wir:
\[g^\lambda :=\lim_{n\to\infty}g^{a_n} \]
wobei $(a_n)$ eine Folge in $\mathbb{Z}$ mit Grenzwert $\lambda$ besitzt. Dies ist wohldefiniert nach Lemma~\ref{1.24}.
\end{definition}

\begin{proposition}\label{1.28}
Sei $G=\overline{\langle g\rangle}$ eine prozyklische pro"~$p$ Gruppe. Dann gilt:
\[G= \{g^\lambda\mid\lambda\in\mathbb{Z}_p \}\]
\end{proposition}

\begin{proof}
Das Bild $g^{\mathbb{Z}_p}$ der nach Korollar~\ref{1.21} stetigen Abbildung $\mathbb{Z}_p\to G,\ \lambda\mapsto g^\lambda$ enthält offenbar $\langle g\rangle$ und da jedes Element in $g^{\mathbb{Z}_p}$ der Grenzwert einer Folge in $\langle g\rangle$ ist, folgt $g^{\mathbb{Z}_p}=\overline{\langle g\rangle}$.
\end{proof}

\begin{theorem}\label{3.8}
Sei $G$ eine endlich erzeugte, potenzreiche pro"~$p$ Gruppe und $H\subset G$ eine abgeschlossene Untergruppe. Dann gilt: \[d(H)\leq d(G)\]
\end{theorem}

\begin{proof}
Nach Theorem~\ref{2.9} gilt $d(HN/N)\leq d(G/N)$ für jedes $N\lhd_\text{o}G$. Nach Satz~\ref{1.5} folgt $d(H)\leq d(G)$.
\end{proof}

\paragraph{} Nun wollen wir zeigen, dass analytische pro"~$p$ Gruppen endlich darstellbar sind. Nach Lemma~\ref{4.33} können wir die Aussage auf einen offenen Normalteiler reduzieren. Wir führen dafür \textit{gleichmäßig potenzreiche} pro"~$p$ Gruppen ein:

\begin{definition}
Eine endlich erzeugte, potenzreiche pro"~$p$ Gruppe $G$ heißt \textit{gleichmäßig potenzreich}\index{potenzreich!gleichmäßig}, wenn für alle $i$ gilt:
\[[P_i(G):P_{i+1}(G)]=[G:P_2(G)] \]
\end{definition}

\begin{lemma}\label{4.2} 
Sei $G$ eine endlich erzeugte, potenzreiche pro"~$p$ Gruppe. Dann ist $P_k(G)$ gleich\-mäßig für hinreichend große $k$.
\end{lemma}

\begin{proof}
Setze $G_i:=P_i(G)$ und $d_i:=\dim_{\mathbb{F}_p}(G_i/G_{i+1})$. Nach Theorem~\ref{3.6} (iii) gilt $d_1\geq d_2\geq d_3\geq\ldots$, also existiert ein $m$, so dass $d_k=d_m$ für alle $k\geq m$. Theorem~\ref{3.6} (ii) zeigt, dass $G_k$ potenzreich ist und Theorem~\ref{3.6} (i) zeigt $P_i(G_k)=G_{k+i-1}$ für alle $i$ und $k$.
\end{proof}

\begin{theorem}
Jede analytische pro"~$p$ Gruppe $G$ ist endlich darstellbar.
\end{theorem}

\begin{proof}
Nach Lemma~\ref{4.33} und Theorem~\ref{thm:analytische-pro-p-gruppe} können wir o.B.d.A. $G$ als endlich erzeugte, potenzreiche pro"~$p$ Gruppe annehmen. Ebenso können wir nach Lemma~\ref{4.2} annehmen, dass $G$ gleichmäßig ist. Sei $\{a_1,\ldots,a_d\}$ ein topologisches Erzeugendensystem von $G$ mit $d=d(G)$. Für alle $i,j$ gilt $[a_j,a_i]\in\overline{G^p}$, wenn $p$ ungerade ist, oder $[a_j,a_i]\in\overline{G^4}$, wenn $p=2$. Nach Theorem~\ref{3.6} (ii), Satz~\ref{3.7} und Satz~\ref{1.28} existiert für jedes $m,i,j\in\{1,\ldots,d\}$ ein $\lambda_m(i,j)\in p\mathbb{Z}_p$ für $p$ ungerade oder $\lambda_m(i,j)\in 4\mathbb{Z}_2$ für $p=2$, so dass:
\[[a_j,a_i]=\prod_{m=1}^d a_m^{\lambda_m(i,j)} \]
Setze $R:=\{[a_i,a_j] a_1^{\lambda_1(i,j)}\cdots a_d^{\lambda_d(i,j)} \mid 1\leq i<j\leq d \},\ H:=\langle a_1,\ldots,a_d ; R\rangle,\ H_i:=P_i(H)$ und $G_i:=P_i(G)$. Dann ist $R=1$ in $G$ erfüllt, also haben wir einen surjektiven Homomorphismus $\pi:H\to G$. Die Relationen $R=1$ in $H$ machen $H$ zu einer potenzreichen pro"~$p$ Gruppe. Beachte dabei, dass $[G,G]$ als abgeschlossener Normalteiler von Elementen der Form $[a_i,a_j],\ 1\leq i<j\leq d$ erzeugt werden. Mit Theorem~\ref{3.6} (iii) und Theorem~\ref{3.8} gilt für alle $i$: 
\[[H_i:H_{i+1}]\leq [H:H_2]= p^{d(H)}\leq p^d\]
Da $G$ gleichmäßig ist, folgt $[H:H_{n+1}]\leq p^{nd}=[G:G_{n+1}]$. Da $\pi$ surjektiv und $\pi(H_{n+1})\subset G_{n+1}$ gilt, ist die von $\pi$ induzierte Abbildung $H/H_{n+1}\to G/G_{n+1}$ ein Isomorphismus, also folgt $\ker(\pi)\subset H_{n+1}$ für alle $n$. Da $H$ hausdorffsch und $H_n,\ n\geq 1$ eine offene Umgebungsbasis bilden, folgt $\bigcap_{n=1}^\infty H_{n+1}=1$ und daher ist $\pi$ injektiv und somit ein topologischer Isomorphismus.
\end{proof}
\fi

\section{Der Gruppenring}

Sei $G=\overline{\langle a_1,\ldots,a_d\rangle}$ mit $d=d(G)$ eine endlich erzeugte pro"~$p$ Gruppe und $G_k:=P_k(G)$. Sei $R:=\mathbb{F}_p[G]$ der Gruppenring, $I_k:=\ker(R\to \mathbb{F}_p[G/G_k])$ und $I:=I_1$ das Augmentationsideal.

\begin{lemma}\label{lemma:augmentation-ideal}
Sei $N\lhd G$ und $J:=\ker(\mathbb{F}_p[G]\to \mathbb{F}_p[G/N])$. Dann ist $J$ als Linksideal durch $\{g-1\mid g\in N\}$ erzeugt. Ist $N=G$, so ist $J=I$ und $\{g-1\mid g\in G\setminus\{1\}\}$ eine $\mathbb{F}_p$-Basis von $I$.
\end{lemma}

\begin{proof}
Die Elemente $g-1,\ g\in N$ liegen in $J$. Sei $\alpha=\sum_{g\in G}\lambda_gg\in J,\ \lambda_g=0$ für fast alle $g$. Dann gilt für das Bild von $\alpha$ in $\mathbb{F}_p[G/N]$:
\[0=\overline{\alpha}= \sum_{g\in G} \lambda_g\overline{g} = \sum_{h\in S}\sum_{g\in N} \lambda_{gh}\overline{h}\]
wobei $S\subset G$ ein Repräsentatensystem von den Rechtsnebenklassen $N\backslash G$ bezeichnet mit $1\in S$. Es folgt $\sum_{g\in N}\lambda_{gh}=0$ für alle $h\in S$. Daher gilt: 
\[\alpha= \sum_{h\in S}\sum_{g\in N}\lambda_{gh}(gh-h) =\sum_{g\in N} \sum_{h\in S} \lambda_{gh} (g-1)h = \sum_{g\in N} (g-1)\sum_{h\in S}\lambda_{gh}h \]
Für $N=G$ ist $S=\{1\}$. Daher sehen wir, dass $\{g-1\mid g\in G \setminus\{1\}\}$ ein $\mathbb{F}_p$"~Erzeugendensystem von $I$ ist. Die lineare Unabhängigkeit ist klar.
\end{proof}

\begin{proposition}\label{lemma:unipotency}
Sei $G$ eine endliche $p$"~Gruppe der Ordnung $n$. Dann gilt $I^n=0$.
\end{proposition}

\begin{proof}
Es ist $R/I\cong\mathbb{F}_p$. Daher ist $\mathbb{F}_p\subset R$ ein Repräsentantensystem von $R/I$. Jedes $x\in R\setminus I$ ist also von der Form: 
\[x=\lambda + \sum_{g\in G} \lambda_g (g-1),\quad \lambda,\lambda_g\in \mathbb{F}_p,\ \lambda\neq 0 \]
Sei $n=p^k$, d.h. $(g-1)^{p^k}=g^{p^k}-1=0$ für alle $g\in G$. Dann gilt $x^{p^k}=\lambda^{p^k}\in R^\times$, also $R\setminus I\subset R^\times$. Angenommen, $I^n\neq 0$, insbesondere ist $I^k\neq 0$ für alle $k=0,\ldots,n-1$. Sei $k$ beliebig, aber fest, und $\{x_1,\ldots,x_m\}$ ein minimales Erzeugendensystem von $I^k$ als $R$-Linksmodul. Ist $I^k=I^{k+1}$, so hat $x_m\in I^k=I^{k+1}$ die Form:
\[x_m = i_1x_1+\ldots+i_mx_m,\quad i_1,\ldots,i_m\in I  \]
Es folgt $(1-i_m)x_m=i_1x_1+\ldots+i_{m-1}x_{m-1}$. Nun ist $1-i_m\in R\setminus I\subset R^\times$. $x_m$ wird also schon durch $x_1,\ldots,x_{m-1}$ erzeugt, ein Widerspruch. Da $k$ beliebig war, ist $I^k\neq I^{k+1}$ für alle $k=0,\ldots,n-1$ und somit $\dim_{\mathbb{F}_p}(I^k/I^{k+1})\geq 1$. Es gilt:
\[n- \dim_{\mathbb{F}_p}(I^n)= \dim_{\mathbb{F}_p}(R/I^n) = \sum_{i=0}^{n-1} \dim_{\mathbb{F}_p}(I^k/I^{k+1})\geq n  \]
Also folgt $\dim_{\mathbb{F}_p}(I^n)=0$.
\end{proof}

\begin{corollary}\label{cor:finite-group-rho}
Ist $G$ eine endliche $p$-Gruppe, so wird die Folge $c_n:=\dim_{\mathbb{F}_p}(I^n/I^{n+1})$ für hinreichend große $n$ konstant $0$. Insbesondere gilt:
\[\varrho:=\limsup_{n\to\infty}\sqrt[n]{c_n}\leq 1 \]
\end{corollary}

\begin{remark}
Wir werden sehen, dass der Konvergenzradius $\varrho^{-1}$ der Potenzreihe $\sum_{n=0}^\infty c_nX^n$ in den Voraussetzungen der Golod-Shafarevich Ungleichung auftaucht.
\end{remark}

\paragraph{} Die folgenden Aussagen werden sich später als nützlich erweisen, wenn wir den Gruppenring bzgl. der $I$-adischen Topologie vervollständigen:

\begin{lemma}\label{7.1} Sei $k\geq 1$. Dann gilt mit $m(k)=[G:G_k]$:
\[I^k\supset I_k \supset I^{m(k)}\]
\end{lemma}

\begin{proof}
Sei $n:=[G:G_k]$. Nun ist $G/G_k$ eine endliche $p$"~Gruppe der Ordnung $n$. Es folgt $(x_1-1)\cdots(x_n-1)=0$ in $\mathbb{F}_p[G/G_k]$ für alle $x_1,\ldots,x_n\in G/G_k$ nach Satz~\ref{lemma:unipotency}. Da $I_k=\ker(R\to \mathbb{F}_p[G/G_k])$, folgt $(g_1-1)\cdots(g_n-1)\in I_k$ für alle $g_1,\ldots,g_n\in G$. Somit ist $I^n\subset I_k$.

Die andere Inklusion beweisen wir per Induktion über $k$. Für $k=1$ gilt die Behauptung per Definition. Sei $k>1$ und $I^{k-1}\supset I_{k-1}$. Nach Korollar~\ref{1.20} wird $G_k$ von den Elementen der Form $x^p$ und $[x,y]$ mit $x\in G_{k-1},\ y\in G$ erzeugt. Setze $u:=x-1\in I_{k-1},\ v:=y-1\in I$. Dann gilt:
\[x^p-1=(x-1)^p=u^p,\quad [x,y]-1=x^{-1}y^{-1}(xy-yx)=x^{-1}y^{-1}(uv-vu) \]
Nach Induktionsvoraussetzung ist $u\in I^{k-1}$. Da $p\geq 2$ und $v\in I$, folgt $x^p-1\in I^k$ und $[x,y]-1\in I^k$. Wegen $xy-1=x(y-1)+(x-1)$ gilt $I_k=(G_k-1)R\subset I^k$.
\end{proof}

\begin{corollary}\label{7.2}
Es gilt $\bigcap_{n=1}^\infty I^n=0$.
\end{corollary}

\begin{proof}
Sei $c=\sum_{i=1}^n \lambda_ix_i\in R,\ c\neq 0$ mit $\lambda_i\in \mathbb{F}_p$ und paarweise verschiedene $x_i\in G$. Da $G$ hausdorffsch ist, finden wir ein $k$, dass $x_ix_j^{-1}\not\in G_k$ für alle $i\neq j$. Bezeichne die natürliche Abbildung $R\to \mathbb{F}_p[G/G_k]$ mit $\psi_k$. Dann sind $\psi_k(x_1),\ldots,\psi_k(x_n)$ paarweise verschiedene Elemente in $G/G_k$. Da es ein $i$ mit $\lambda_i\neq 0$ gibt, folgt $\psi_k(c)\neq 0$, also $c\not\in \ker(\psi_k)=I_k$. Nach Lemma~\ref{7.1} folgt $c\not\in I^m$ für ein $m$.
\end{proof}

\begin{proposition}\label{Ex1}
Die Abbildung $g\mapsto g-1$ induziert einen Isomorphismus $G/\Phi(G)\to I/I^2$.
\end{proposition}

\begin{proof}
Betrachte die wohldefinierte Abbildung $\phi:G\to I/I^2,\ g\mapsto (g-1)+I^2$. Wegen $xy-1=(x-1)+(y-1)+(x-1)(y-1)$ ist diese ein Homomorphismus. Da $I/I^2$ abelsch ist, gilt $[G,G]\subset\ker(\phi)$. Ferner gilt $g^p-1=(g-1)^p\in I^p\subset I^2$ für alle $g\in G$. Daher ist $\Phi(G)=G^p[G,G]\subset\ker(\phi)$ und wir erhalten einen Homomorphismus $\tilde{\phi}:G/\Phi(G)\to I/I^2$.

Wir wollen nun ein Inverses konstruieren. Nach Lemma~\ref{lemma:augmentation-ideal} besitzt $I$ die $\mathbb{F}_p$-Basis $g-1$, $g\in G\setminus\{1\}$. Sei $\psi:I\to G/\Phi(G)$ die durch $g-1\mapsto g\Phi(G)$ gegebene $\mathbb{F}_p$-lineare Abbildung. Wegen $(x-1)(y-1)=(xy-1)-(x-1)-(y-1)$ gilt für alle $g,h\in G$:
\[\psi((g-1)(h-1))=gh\Phi(G)\cdot  (g\Phi(G))^{-1}(h\Phi(G))^{-1}=1 \]
Also induziert $\psi$ einen Homomorphismus $\tilde{\psi}:I/I^2\to G/\Phi(G)$, die zu $\tilde{\phi}$ invers ist.
\end{proof}

\begin{corollary}\label{cor:Ex1}
Es gilt $d(G) = \dim_{\mathbb{F}_p}(I/I^2)$.
\end{corollary}

\begin{proof}
Folgt aus Korollar~\ref{cor:1.19}.
\end{proof}

\paragraph{} Wir werden nun den Rest des Kapitels ein Analogon von Korollar~\ref{cor:finite-group-rho} für analytische pro"~$p$ Gruppen beweisen. Es ist offensichtlich, dass wir dazu den Gruppenring einer endlich erzeugten, potenzreichen pro"~$p$ Gruppe über $\mathbb{F}_p$ untersuchen müssen, insbesondere wie sich die gruppentheoretischen Eigenschaften auf den Gruppenring übertragen.

\begin{definition}
Setze $\mathbf{a} :=(a_1,\ldots,a_d)$ und $\mathbf{b} :=(b_1,\ldots,b_d)$ mit $b_i:=a_i-1\in I$ für alle $i$. Für ein $d$-Tupel $\alpha=(\alpha_1,\ldots,\alpha_d)\in\mathbb{N}^d$ und ein $d$-Tupel $\mathbf{v}=(v_1,\ldots,v_d)$ setzen wir:
\[\langle\alpha\rangle :=\alpha_1+\ldots+\alpha_d,\quad \mathbf{v}^\alpha := v_1^{\alpha_1}\cdots v_d^{\alpha_d} \]
\end{definition}

\begin{lemma}\label{7.8}
Sei $\mathbf{u}:=(u_1,\ldots,u_r)\in G^{(r)}$ und $\mathbf{v}:=(v_1,\ldots,v_r)$ mit $v_i:=u_i-1\in I$. Dann gilt für alle $\beta\in\mathbb{N}^r$:
\[\mathbf{u}^\beta =\sum_{\alpha\in\mathbb{N}^r}\binom{\beta_1}{\alpha_1}\cdots\binom{\beta_r}{\alpha_r}\mathbf{v}^\alpha ,\quad \mathbf{v}^\beta=\sum_{\alpha\in\mathbb{N}^r} (-1)^{\langle \beta\rangle -\langle \alpha\rangle}\binom{\beta_1}{\alpha_1}\cdots\binom{\beta_r}{\alpha_r} \mathbf{u}^\alpha \]
Jede Summe ist endlich, da $\binom{\beta_i}{\alpha_i}=0$, wenn $\alpha_i>\beta_i$.
\end{lemma}

\begin{proof}
Dies folgt direkt aus dem binomischen Lehrsatz. Es gilt:
\begin{align*}
\mathbf{v}^\beta &= (u_1-1)^{\beta_1}\cdots (u_r-1)^{\beta_r}\\
&=\prod_{i=1}^r \Big(\sum_{\alpha_i=0}^{\beta_i} (-1)^{\beta_i-\alpha_i}\binom{\beta_i}{\alpha_i} u_i^{\alpha_i}\Big)= \sum_{\alpha\in\mathbb{N}^r}(-1)^{\langle \beta\rangle-\langle\alpha\rangle}\binom{\beta_1}{\alpha_1}\cdots\binom{\beta_r}{\alpha_r} \mathbf{u}^\alpha
\end{align*}
Analog für $\mathbf{u}^\beta = (v_1+1)^{\beta_1}\cdots (v_r+1)^{\beta_r}$.
\end{proof}

\begin{lemma}\label{7.9}
Sei $G$ potenzreich und setze $T_k:=\{ \alpha\in\mathbb{N}^d\mid \alpha_i< p^{k-1}\text{ für alle }i=1,\ldots,d \}$. Dann gilt für alle $k\geq 1$:
\[R=I_k+\sum_{\alpha\in T_k} \mathbb{F}_p \mathbf{b}^\alpha \]
\end{lemma}

\begin{proof}
$I_k$ ist der Kern der natürlichen Surjektion $\pi:R\to \mathbb{F}_p[G/G_k]$. Nun ist $G/G_k$ diskret. Nach Theorem~\ref{3.6} (ii) und Satz~\ref{3.7} kann jedes Element in $G/G_k$ in der Form $a_1^{\alpha_1}\cdots a_d^{\alpha_d}G_k$ mit $0\leq\alpha_i <p^{k-1}$ geschrieben werden. Somit ist $\{\pi(\mathbf{a}^\alpha)\mid\alpha\in T_k \}$ ein Erzeugendensystem von $\mathbb{F}_p[G/G_k]$ als $\mathbb{F}_p$"~Vektorraum. Lemma~\ref{7.8} zeigt, dass $\{\pi(\mathbf{b}^\alpha)\mid \alpha\in T_k\}$ den gleichen Vektorraum erzeugt, also ist $\pi(R)=\pi\big(\sum_{\alpha\in T_k}\mathbb{F}_p  \mathbf{b}^\alpha\big)$.
\end{proof}

\begin{lemma}\label{7.10}
Sei $G$ potenzreich und $u\in I^k,\ x\in G$. Dann gilt $ux-xu\in I^{k+2}$.
\end{lemma}

\begin{proof}
Sei $y\in G$. Da $G$ potenzreich ist, gibt es nach Lemma~\ref{3.4} ein $z\in G$, so dass:
\[yx-xy=xy([y,x]-1)=xy(z^p-1) \]
Es gilt $z^p-1=(z-1)^p\in I^p\subset I^3$, also $yx-xy\in I^3$. Da $R$ von $G$ additiv erzeugt wird, ist die Behauptung für den Fall $k=1$ gezeigt. Sei nun $k\geq 1$ und $vy-yv\in I^{k+2}$ für alle $v\in I^{k}$ und $y\in G$. Nun ist $I^{k+1}$ erzeugt von Elementen der Form $vw$ mit $v\in I^{k}$ und $w\in I$, also folgt:
\[vwx-xvw=v(wx-xw)+(vx-xv)w\in I^{k}I^3+ I^{k+2}I\subset I^{k+3}\qedhere \]
\end{proof}

\begin{proposition}\label{7.11}
Sei $G$ potenzreich und $k\geq 1$. Dann gilt:
\[ I^k = I^{k+1}+\sum_{\substack{\alpha\in\mathbb{N}^d\\\langle\alpha\rangle =k}}\mathbb{F}_p \mathbf{b}^\alpha \]
\end{proposition}

\begin{proof}
Setze $W_k:=\sum_{\langle\alpha\rangle =k}\mathbb{F}_p\mathbf{b}^\alpha$. Da $b_i=a_i-1\in I$ für alle $i$, folgt $\mathbf{b}^\alpha\in I^k$ für alle $\alpha\in\mathbb{N}^d$ mit $\langle\alpha\rangle=k$. Somit gilt $I^k\supset I^{k+1}+W_k$. Die andere Inklusion beweisen wir per Induktion über $k$. Lemma~\ref{7.9} mit $k=2$ liefert:
\[I\subset I_2+\sum_{\alpha\in T_2}\mathbb{F}_p\mathbf{b}^\alpha\subset I^2+\sum_{\langle\alpha\rangle =1}\mathbb{F}_p\mathbf{b}^\alpha+\mathbb{F}_p 1\subset I^2+W_1+\mathbb{F}_p1 \]
da $I_2\subset I^2$ nach Lemma~\ref{7.1} und $\mathbf{b}^\alpha\in I^2$ für $\langle\alpha\rangle\geq 2$ gilt. Nun ist $I\cap\mathbb{F}_p1=0$. Sei nun $k>1$ und die Aussage für kleinere $k$ schon bewiesen. Dann gilt:
\[I^k = I^{k-1} I=(I^k+W_{k-1})(I^2+W_1)\subset I^{k+1}+W_{k-1}W_1 \]
Zu zeigen ist noch $W_{k-1}W_1\subset I^{k+1}+W_k$. Nun ist $W_{k-1}W_1$ von Elementen der Form $\mathbf{b}^\alpha b_i$ mit $i=1,\ldots,d$ und $\langle\alpha\rangle =k-1$ erzeugt. Es genügt, die Aussage für diese Elemente nach\-zu\-wei\-sen. Setze $u:= b_1^{\alpha_1}\cdots b_{i-1}^{\alpha_{i-1}}$ und $v:=b_i^{\alpha_i}\cdots b_d^{\alpha_d}$. Dann gilt
\[\mathbf{b}^\alpha b_i=uvb_i=ub_iv+u(vb_i-b_iv)=\mathbf{b}^\beta+uw \]
wobei $w:=vb_i-b_iv$ und $\beta:=(\alpha_1,\ldots,\alpha_i+1,\ldots, \alpha_d)$. Da $v\in I^n$ mit $n:=\alpha_i+\ldots+\alpha_d$, folgt aus Lemma~\ref{7.10} $w\in I^{n+2}$. Da $u\in I^{\langle\alpha\rangle -n}$, folgt $uw\in I^{\langle\alpha\rangle+2}$. Zudem gilt $\langle\beta\rangle = \langle\alpha\rangle+1=k$. Daher gilt:
\[\mathbf{b}^\alpha b_i=\mathbf{b}^\beta+uw \in \mathbb{F}_p \mathbf{b}^\beta+ I^{\langle\alpha\rangle+2}\subset W_k+ I^{k+1} \qedhere \]
\end{proof}

\paragraph{} Eine Abschätzung für $\dim_{\mathbb{F}_p}(I^k/I^{k+1})$ nach oben wäre also die Anzahl aller $\alpha\in\mathbb{N}^d$ mit $\langle\alpha\rangle =k$. Diese Anzahl wollen wir im nächsten Satz berechnen:

\begin{lemma}\label{Ex7.4Ex}
Für alle $n,k\in\mathbb{N}$ gilt:
\[\sum_{j=0}^n\binom{k-2+j}{k-2} =\binom{k-1+n}{k-1} \]
\end{lemma}

\begin{proof}
Per Induktion über $n$. Für $n=0$ ist die Aussage klar. Sei die Aussage für ein $n$ bewiesen. Dann gilt für alle $k\in\mathbb{N}$:
\begin{align*}
\sum_{j=0}^{n+1}\binom{k-2+j}{k-2} &=\binom{k-1+n}{k-2} + \sum_{j=0}^n\binom{k-2+j}{k-2}\\
&=\binom{k-1+n}{k-2} + \binom{k-1+n}{k-1}=\binom{k+n}{k-1}
\end{align*}
wobei die letzte Gleichung aus der Identität $\binom{n+1}{k+1}=\binom{n}{k}+\binom{n}{k+1}$ folgt.
\end{proof}

\begin{proposition}
Sei $k\in\mathbb{N}$. Die Anzahl aller $\alpha\in\mathbb{N}^d$ mit $\langle \alpha\rangle =k$ beträgt gerade:
\[f_d(k) := \binom{k+d-1}{d-1} \]
\end{proposition}

\begin{proof}
Per Induktion über $d$. Es ist $f_1(k)=\binom{k}{0}=1$ für alle $k$, also ist die Behauptung für $d=1$ korrekt. Sei nun $d> 1$ und die Aussage für kleinere $d$ bewiesen. Sei $\alpha=(\alpha_1,\ldots,\alpha_d)\in\mathbb{N}^d$ mit $\langle\alpha\rangle=k$. Für $\alpha_1\in\mathbb{N}$ haben wir $k+1$ Möglichkeiten $\{0,\ldots,k\}$. Für $(\alpha_2,\ldots,\alpha_d)$ gibt es nach Induktionsvoraussetzung $f_{d-1}(k-\alpha_1)=\binom{k-\alpha_1+d-2}{d-2}$ Möglichkeiten. Also gibt es insgesamt folgende Anzahl von Möglichkeiten:
\[\sum_{\alpha_1=0}^k f_{d-1}(k-\alpha_1)= \sum_{\alpha_1=0}^k \binom{d-2+k-\alpha_1}{d-2}=\sum_{\alpha_1=0}^k\binom{d-2+\alpha_1}{d-2}=\binom{d-1+k}{d-1}=f_d(k) \]
wobei die vorletzte Gleichung aus dem vorherigen Lemma folgt.
\end{proof}

\begin{corollary}\label{Ex7.4}
Sei $G$ potenzreich. Dann gilt $\dim_{\mathbb{F}_p}(I^k/I^{k+1})\leq f_d(k)$.
\end{corollary}

\begin{proof}
Nach Lemma~\ref{7.11} wird $I^k/I^{k+1}$ von Elementen der Form $\mathbf{b}^\alpha$ mit $\langle\alpha\rangle =k$ erzeugt.
\end{proof}

\begin{corollary}
Ist $G$ potenzreich, so gilt $\dim_{\mathbb{F}_p}(R/I^{k+1})\leq f_{d+1}(k)$.
\end{corollary}

\begin{proof}
$\dim_{\mathbb{F}_p}(R/I^{k+1})=\sum_{i=0}^k\dim_{\mathbb{F}_p}(I^i/I^{i+1})\leq \sum_{i=0}^k\binom{i+d-1}{d-1}=\binom{d+k}{d}= f_{d+1}(k)$.
\end{proof}

\begin{proposition}\label{Ex7.5}
Sei $G$ $p$-adisch analytisch. Dann existieren Konstanten $C>0$ und $e\in\mathbb{N}$, so dass $\dim_{\mathbb{F}_p}(R/I^n)\leq Cn^e$ für alle $n\geq 0$. Insbesondere gilt für die Folge $c_n:=\dim_{\mathbb{F}_p}(I^n/I^{n+1})$:
\[\varrho:= \limsup_{n\to\infty}\sqrt[n]{c_n}\leq 1 \]
\end{proposition}

\begin{proof}
Nach Theorem~\ref{thm:analytische-pro-p-gruppe} ist $G$ endlich erzeugt und enthält eine endlich erzeugte potenzreiche pro"~$p$ Gruppe $H\lhd_\text{o}G$. Setze $I_H$ als das Augmentationsideal in $\mathbb{F}_p[H]$ und $J:=I_HR\subset I$.

Für alle $i\in I_H^n$ und $x\in R$ gilt $ix\in J^n$. Daher ist $R/J^n$ ein $\mathbb{F}_p[H]/I_H^n$"~Modul. Sei $S\subset G$ ein Repräsentantensystem von den Rechtsnebenklassen $H\backslash G$. Der surjektive Homomorphismus $\mathbb{F}_p[H]^{(S)}\cong R\to R/J^n$ induziert die folgende Surjektion:
\[(\mathbb{F}_p[H]/I_H^n)^{(S)}\to R/J^n\]
Somit ist $R/J^n$ als $\mathbb{F}_p[H]/I_H^n$"~Modul durch $\#S=[G:H]$ Elementen erzeugt. Da $J\subset I$, haben wir eine Surjektion $R/J^n\to R/I^n$. Nach vorherigem Korollar folgt:
\begin{align*}
\dim_{\mathbb{F}_p}(R/I^n)&\leq\dim_{\mathbb{F}_p}(R/J^n)\\
&\leq [G:H]\cdot\dim_{\mathbb{F}_p}(\mathbb{F}_p[H]/I_H^n)\\
&\leq [G:H]\cdot f_{e+1}(n-1)\\
&= [G:H]\cdot \frac{(n-1+e)!}{e! (n-1)!}\leq \frac{[G:H]}{e!}\cdot (n+e-1)^e
\end{align*}
wobei $e:=d(H)$. Dies zeigt die Behauptung. Da $\sqrt[n]{C}\to 1$ und $\sqrt[n]{n+1}\to 1$ für $n\to\infty$, folgt $\varrho=\limsup_{n\to\infty}\sqrt[n]{c_n} \leq \lim_{n\to\infty}\sqrt[n]{C(n+1)^e}=1$.
\end{proof}

\chapter{Die Ungleichung}

Wir werden in diesem Kapitel den \textit{vollständigen Gruppenring} $\mathbb{F}_p[[G]]$ einführen. Mit proendlichen Gruppen ist die Arbeit mit dem vollständigen Gruppenring meist einfacher. Doch können wir im Allgemeinen die Elemente von $\mathbb{F}_p[[G]]$ nicht mehr so einfach beschreiben wie im gewöhnlichen Gruppenring $\mathbb{F}_p[G]$. Für den Fall, dass $G$ eine endlich erzeugte, freie pro"~$p$ Gruppe ist, haben wir aber trotz allem eine einfache Beschreibung von $\mathbb{F}_p[[G]]$. Diese werden wir im ersten Teil dieses Kapitels präsentieren und dabei \cite{Wil98} und \cite{Koc02} folgen.

Im zweiten Teil können wir die Resultate verwenden und schließlich die \textit{Golod-Shafarevich Ungleichung} formulieren und beweisen.

\section{Der vollständige Gruppenring}

Sei $G$ eine endlich erzeugte pro"~$p$ Gruppe,  $I_k:=\ker(\mathbb{F}_p[G]\to\mathbb{F}_p[G/G_k])$ und $I:=I_1$ das Aug\-men\-ta\-tions\-i\-deal in $\mathbb{F}_p[G]$.

\begin{definition}
Der \textit{vollständige Gruppenring} von $G$ über $\mathbb{F}_p$ ist definiert als:
\[\mathbb{F}_p[[G]] := \varprojlim_{N\lhd_\text{o}G}\mathbb{F}_p[G/N] = \varprojlim_{k\geq 1}\mathbb{F}_p[G/G_k] = \varprojlim_{k\geq 1} \mathbb{F}_p[G]/I_k \]
wobei die Gleichung in der Mitte aus Satz~\ref{1.16} folgt. Also entsteht $\mathbb{F}_p[[G]]$ durch Ver\-voll\-stän\-di\-gung von $\mathbb{F}_p[G]$ bzgl. der offenen Umgebungsbasis $(I_k)_{k\geq 1}$ der $0$. Nach Lemma~\ref{7.1} induziert $(I^k)_{k\geq 1}$ dieselbe Topologie auf $\mathbb{F}_p[G]$. Daher gilt:
\[\mathbb{F}_p[[G]]\cong\varprojlim_{k\geq 1} \mathbb{F}_p[G]/I^k \]
Die Vervollständigung induziert eine offene Umgebungsbasis $(\widehat{I}^k)_{k\geq 1}$ auf $\mathbb{F}_p[[G]]$, wobei $\widehat{I}= \ker(\mathbb{F}_p[[G]]\to\mathbb{F}_p[G]/I\cong\mathbb{F}_p)$ und es gilt für alle $k$:
\[\mathbb{F}_p[G]/I^k\cong \mathbb{F}_p[[G]]/\widehat{I}^k \]
Nach Korollar~\ref{7.2} ist die natürliche Abbildung $\mathbb{F}_p[G]\to\mathbb{F}_p[[G]]$ eine Inklusion. Daher ist $\widehat{I}$ auch der topologische Abschluss von $I=(G-1)\mathbb{F}_p[G]$ in $\mathbb{F}_p[[G]]$, d.h. $\widehat{I}=(G-1)\mathbb{F}_p[[G]]$. 
\end{definition}

\begin{definition}
Ein topologischer Ring heißt \textit{proendliche $\mathbb{F}_p$"~Algebra}, wenn sie projektiver Limes endlicher, diskreter $\mathbb{F}_p$"~Algebren ist.
\end{definition}

\begin{proposition}\textit{(Universaleigenschaft des vollständigen Gruppenrings)}
Jeder stetige Homomorphismus $\phi:G\to E^\times$ von $G$ in die Einheitengruppe einer proendlichen $\mathbb{F}_p$"~Algebra $E$ setzt sich eindeutig zu einem stetigen $\mathbb{F}_p$"~Algebra-Homomorphismus $\mathbb{F}_p[[G]]\to E$ fort.
\end{proposition}

\begin{proof}
Sei $E=\varprojlim_{\lambda\in\Lambda} E_\lambda$ und $\lambda\in\Lambda$. Betrachte die Komposition $\varphi_\lambda:G\to E^\times\to E_\lambda^\times$. Dann ist $\varphi_\lambda$ stetig und $\ker(\varphi_\lambda)\lhd_\text{o}G$. $\varphi$ induziert $G/\ker(\varphi_\lambda)\to E_\lambda^\times$. Diese setzt sich offensichtlich auf einem $\mathbb{F}_p$"~Algebra-Homomorphismus $\mathbb{F}_p[G/\ker(\varphi_\lambda)]\to E_\lambda$ eindeutig fort. Die Homomorphismen $\phi_\lambda:\mathbb{F}_p[[G]]\to \mathbb{F}_p[G/\ker(\varphi_\lambda)]\to E_\lambda$ sind kompatibel und setzen sich eindeutig zu einem stetigen $\mathbb{F}_p$"~Algebra-Homomorphismus $\widehat{\phi}:\mathbb{F}_p[[G]]\to E$ zusammen. Da die natürliche Abbildung $G\hookrightarrow \mathbb{F}_p[G]\hookrightarrow\mathbb{F}_p[[G]]$ eine Inklusion ist, setzt $\widehat{\phi}$ die ursprüngliche Abbildung $\phi$ fort.
\end{proof}

\paragraph{} Das folgende Lemma lässt uns die Hauptaussagen Korollar~\ref{cor:finite-group-rho} und Satz~\ref{Ex7.5} über den gewöhnlichen Gruppenring, die wir in Kapitel 2, Abschnitt 3 hergeleitet haben, auf den voll\-stän\-di\-gen Gruppenring übertragen:

\begin{lemma}\label{lemma:nuisance}
Es gilt $I^k/I^{k+1}\cong \widehat{I}^k/\widehat{I}^{k+1}$ für alle $k$.
\end{lemma}

\begin{proof}
Wir haben ein kommutatives Diagramm mit exakten Zeilen:
\[\xymatrix{
0\ar[r] & I^k/I^{k+1} \ar[r]& \mathbb{F}_p[G]/I^{k+1} \ar[r]\ar[d]_\cong& \mathbb{F}_p[G]/I^k\ar[r]\ar[d]^\cong & 0\\
0 \ar[r] & \widehat{I}^k/\widehat{I}^{k+1} \ar[r] & \mathbb{F}_p[[G]]/\widehat{I}^{k+1} \ar[r] &\mathbb{F}_p[[G]]/\widehat{I}^k \ar[r] & 0
} \]
Es folgt die Behauptung.
\end{proof}

\paragraph{} Die Hauptaussage, die wir für die Golod-Shafarevich Ungleichung benötigen werden, ist der folgende Satz~\ref{prop:Ex2}. Dafür werden wir zeigen, dass $\mathbb{F}_p[[F]]$ für eine endlich erzeugte, freie pro"~$p$ Gruppe $F$ isomorph zu einem Potenzreihenring über nichtkommutierende Variablen ist. Diese werden wir im Nachfolgenden formal einführen.

\begin{proposition}\label{prop:Ex2}
Sei $G={F}$ die freie pro"~$p$ Gruppe auf $\{x_1,\ldots,x_d\}$. Dann haben wir einen Isomorphismus von $\mathbb{F}_p[[{F}]]$"~Rechtsmoduln:
\[\widehat{I} \cong \mathbb{F}_p[[{F}]]^{(d)}\]
\end{proposition}

\begin{definition}
Eine \textit{formale Potenzreihe} über $\mathbb{F}_p$ in nichtkommutierenden Variablen $y_1,\ldots,y_d$ ist ein Ausdruck:
\[\sum_{w\in W}\lambda_w w,\quad \lambda_w\in\mathbb{F}_p \]
wobei $W:=\{y_{i_1}\cdots y_{i_r}\mid r\in\mathbb{N},\ 1\leq i_j\leq d \}$ die Menge der Monomen in $y_1,\ldots,y_d$ bezeichnet. Der \textit{Gesamtgrad} eines Monoms $w=y_{i_1}\cdots y_{i_r}$ ist gerade die Anzahl der Faktoren und wird mit $\deg(w)=r$ bezeichnet. Die Menge aller formalen Potenzreihen über $\mathbb{F}_p$ wird auf natürlicher Weise zu einer $\mathbb{F}_p$-Algebra und wird mit $\mathbb{F}_p\llangle y_1,\ldots,y_d\rrangle$ bezeichnet. Wir haben einen Isomorphismus von $\mathbb{F}_p$"~Moduln:
\[\phi:\prod_{w\in W}\mathbb{F}_p\to\mathbb{F}_p\llangle y_1,\ldots,y_d\rrangle,\ (\lambda_w)_w\mapsto \sum_{w\in W}\lambda_w w \]
Statten wir $\mathbb{F}_p$ mit der diskreten Topologie aus, können wir die Produkttopologie von $\prod\mathbb{F}_p$ auf $\mathbb{F}_p\llangle y_1,\ldots,y_d\rrangle$ übertragen. Somit wird $\mathbb{F}_p\llangle y_1,\ldots,y_d\rrangle$ zu einem kompakten Hausdorffraum.

Für jedes $n\geq 1$ sei ${D_n}$ die Menge aller formalen Potenzreihen $\sum_{w\in W}\lambda_w w$ mit $\lambda_w=0$ für alle $w\in W$, dessen Gesamtgrad $\deg(w)<n$ ist. Es ist klar, dass alle ${D_n}$ zweiseitige Ideale in $\mathbb{F}_p\llangle y_1,\ldots,y_d\rrangle$ sind. Per Definition der Produkttopologie bilden $\{D_n\mid n\geq 1\}$ eine offene Umgebungsbasis der $0$. Daher ist die natürliche Abbildung $\eta :\mathbb{F}_p\llangle y_1,\ldots,y_d\rrangle\to\varprojlim_{n\geq 1} \mathbb{F}_p\llangle y_1,\ldots,y_d\rrangle /D_n$ injektiv. Wegen Kompaktheit und Dichtheit des Bildes, ist $\eta$ ein topologischer Isomorphismus:
\[\mathbb{F}_p\llangle y_1,\ldots,y_d\rrangle\cong\varprojlim_{n\geq 1} \mathbb{F}_p\llangle y_1,\ldots,y_d\rrangle /D_n \]
Da $\mathbb{F}_p\llangle y_1,\ldots,y_d\rrangle /D_n\cong \prod_{\substack{w\in W,\ \deg(w)<n}}\mathbb{F}_p$ alles endliche $\mathbb{F}_p$"~Algebren bzw. $p$"~Gruppen sind, ist $\mathbb{F}_p\llangle y_1,\ldots,y_d\rrangle$ eine proendliche $\mathbb{F}_p$"~Al\-ge\-bra und die additive Gruppe von $\mathbb{F}_p\llangle y_1,\ldots,y_d\rrangle$ eine pro"~$p$ Gruppe.
\end{definition}

\begin{lemma}
$V:=1+{D_1} \subset\mathbb{F}_p\llangle y_1,\ldots,y_d\rrangle^\times$ ist eine pro"~$p$ Gruppe.
\end{lemma}

\begin{proof}
Setze $P:=\mathbb{F}_p\llangle y_1,\ldots,y_d\rrangle$. Offensichtlich ist $V$ abgeschlossen bzgl. Produktbildung. Sei $x=1+y\in V$ mit $y\in {D_1}$. Da $y^n\in {D_k}$ für $n\geq k$, konvergiert $s:=\sum_{n=0}^\infty (-y)^n$ in $P$. Dann ist $s\in 1+{D_1}$ ein inverses Element zu $x$, d.h. $x\in P^\times$. Daher ist $V$ eine multiplikative Gruppe.

Für jedes $n\geq 1$, sei $V_n$ das Bild der Menge $V$ unter der Projektion $P\to P/D_n$. Es ist klar, dass $V_n$ eine multiplikative Gruppe ist. Da ${D_1}\subset_\text{o}P$, ist $V\subset_\text{o} P$ abgeschlossen und daher kompakt. Die Mengenabbildung $x\mapsto x-1$ induziert eine Bijektion $V_n\to {D_1}/D_n$. Also ist $V_n$ ist eine endliche $p$"~Gruppe, da $[D_1:D_n]\mid [P:D_n]$ eine $p$-Potenz ist. Daher ist $\varprojlim V_n$ eine pro"~$p$ Gruppe. Der Kern der natürlichen Projektion $V\to V_n$ ist gerade $1+D_n$. Da $P$ hausdorffsch ist, und $D_n,\ n\geq 1$ eine offene Umgebungsbasis von $P$ bilden, gilt $\bigcap_{n\geq 1} (1+D_n)=1$, und die natürliche Abbildung $V\to\varprojlim V_n$ ist injektiv. Da $V$ kompakt ist, ist $V\subset\varprojlim V_n$ abgeschlossen und somit eine pro"~$p$ Gruppe.
\end{proof}

\begin{theorem} \label{thm:completedgroupring-freeprop}
Sei $G={F}$ die freie pro"~$p$ Gruppe auf $\{x_1,\ldots,x_d\}$. Dann setzt sich die Abbildung $y_i\mapsto x_i-1,\ i=1,\ldots,d$ zu einem topologischen Isomorphismus fort:
\[\mathbb{F}_p\llangle y_1,\ldots,y_d\rrangle\cong\mathbb{F}_p [[{F}]] \]
\end{theorem}

\begin{proof}
Setze $P:=\mathbb{F}_p\llangle y_1,\ldots,y_d\rrangle$. Sei $P(y_1,\ldots,y_d)\in P$ eine formale Potenzreihe. Da $\widehat{I}^n,\ n\in\mathbb{N}$ eine offene Umgebungsbasis von $\mathbb{F}_p[[{F}]]$ ist, konvergiert die Potenzreihe $P(x_1-1,\ldots,x_d-1)$ in $\mathbb{F}_p[[{F}]]$. Daher ist der Einsetzungshomomorphismus wohldefiniert:
\[\psi:P\to\mathbb{F}_p[[{F}]],\ y_i\mapsto x_i-1 \]
Betrachte nun die Abbildung $x_i\mapsto y_i+1,\ i=1,\ldots,d$. Es ist $y_i+1\in V:=1+D_1$. Da $V$ nach dem vorherigen Lemma eine pro"~$p$ Gruppe ist, induziert diese einen stetigen Gruppenhomomorphismus $\varphi:{F}\to V\to P^\times$. Da $P$ eine proendliche $\mathbb{F}_p$"~Algebra ist, induziert $\varphi$ einen stetigen $\mathbb{F}_p$-Algebra-Homomorphismus $\phi:\mathbb{F}_p[[{F}]]\to P$, die offensichtlich zu $\psi$ invers ist.
\end{proof}

\begin{proof}[Beweis von Satz~\ref{prop:Ex2}.]
Wegen $xy-1=(x-1)+(y-1)+(x-1)(y-1)$, wird $\widehat{I}=(F-1)\mathbb{F}_p[[{F}]]$ als Rechtsideal in $\mathbb{F}_p[[F]]$ durch die Elemente $x_i-1,\ i=1,\ldots,d$ erzeugt. Nach Theorem~\ref{thm:completedgroupring-freeprop} können wir $\widehat{I}$ mit dem von $y_1,\ldots,y_d$ erzeugten Rechtsideal $J$ in $P:=\mathbb{F}_p\llangle y_1,\ldots,y_d\rrangle$ identifizieren. Betrachte den $P$-Modulepimorphismus:
\[ \phi: P^{(d)}\to J,\ e_i\mapsto y_i \]
wobei $e_i\in P^{(d)}$ im $i$-ten Eintrag eine $1$ besitzt und überall sonst $0$. Es genügt also nur noch zu zeigen, dass $\phi$ injektiv ist. Sei also $\sum_{i=1}^d y_ix_i=0$ mit $x_i=\sum_{w\in W}\lambda_{w,i} w\in P$. Es folgt:
\[0=\sum_{i=1}^dy_ix_i = \sum_{w\in W}\sum_{i=1}^d \lambda_{w,i} y_iw \]
Da die Variablen $y_1,\ldots,y_d$ nicht kommutieren, sind die Monome $y_iw,\ i=1,\ldots,d,\ w\in W$ paarweise verschieden. Daher folgt $\lambda_{w,i}=0$ für alle $i$ und $w$ und daher $x_i=0$ für alle $i$.
\end{proof}

\iffalse
\begin{proposition}
Sei $F$ die abstrakte freie Gruppe auf $\{x_1,\ldots,x_d\}$. Dann gilt:
\[(F-1)\mathbb{F}_p[F]=\bigoplus_{i=1}^d (x_i-1)\mathbb{F}_p[F]\]
\end{proposition}

\begin{proof}
Wegen $xy-1=(x-1)+(y-1)+(x-1)(y-1)$, wird das Ideal $I_{F}:=(F-1)\mathbb{F}_p[F]$ schon durch die Elemente $x_i-1,\ i=1,\ldots,d$ erzeugt. Betrachte nun das semidirekte Produkt $H:=\mathbb{F}_p[F]^{(d)}\rtimes F$, d.h. das kartesische Produkt mit der Gruppenoperation $(x,f)\cdot (x',f') = (x+fx',ff')$. Wir identifizieren $\mathbb{F}_p[F]^{(d)}\subset H$ durch die Inklusion $\mathbb{F}_p[F]^{(d)}\hookrightarrow H,\ x\mapsto (x,1)$. Die Abbildung $x_i\mapsto (u_i,x_i)$, wobei $u_i\in \mathbb{F}_p[F]^{(d)}$ den Eintrag $1$ in der $i$-ten Komponente und überall sonst $0$ besitzt, lässt sich zu einem Homomorphismus $\theta:F\to H$ fortsetzen. Da $I_{F}$ als $\mathbb{F}_p$-Vektorraum die Basis $f-1,\ f\in F\setminus\{1\}$ besitzt, definiert
\[\phi(f-1) := \theta(f)\cdot (0,f^{-1})\in \mathbb{F}_p[F]^{(d)} \]
eine $\mathbb{F}_p$-lineare Abbildung $\phi:I\to \mathbb{F}_p[F]^{(d)}$. $\phi$ ist ein $\mathbb{F}_p[F]$-Modulhomomorphismus, da:
\begin{align*}
\phi(x_j(x_i-1)) &= \phi(x_jx_i-1)-\phi(x_j-1)\\
&=\theta(x_jx_i)\cdot (0,x_i^{-1} x_j^{-1}) - \theta(x_j)\cdot (0,x_j^{-1})\\
&=(u_j,x_j)\cdot (u_i,x_i)\cdot (0,x_i^{-1} x_j^{-1}) - u_j\\
&= (u_j+x_ju_i,x_jx_i)\cdot (0,x_i^{-1}x_j^{-1}) - u_j \\
&=u_j+ x_ju_i-u_j=x_j\phi(x_i-1) 
\end{align*}
Mit dem $\mathbb{F}_p[F]$-Modulhomomorphismus $\psi:\mathbb{F}_p[F]^{(d)}\to I,\ u_i\mapsto x_i-1$ haben wir eine Umkehrabbildung zu $\phi$.
\end{proof}

\begin{proposition}\label{prop:Ex2}
Sei $\widehat{F}$ die freie pro"~$p$ Gruppe auf $\{x_1,\ldots,x_d\}$. Dann gilt:
\[(\widehat{F}-1)\mathbb{F}_p[[\widehat{F}]]=\bigoplus_{i=1}^d (x_i-1)\mathbb{F}_p[[\widehat{F}]]\]
\end{proposition}

\begin{proof}
Sei $\pi_n: \mathbb{F}_p[[\widehat{F}]]\to \mathbb{F}_p[\widehat{F}/\widehat{F_n}]\cong \mathbb{F}_p[F/F_n] $ die natürliche Projektion, wobei $\widehat{F_n}:=P_n(\widehat{F})$ und $F_n:=P_n(F)$. Seien $r_1,\ldots,r_d\in \mathbb{F}_p[[\widehat{F}]]$, so dass:
\[\sum_{i=1}^d(x_i-1)r_i=0 \]
Für jedes $i$ wähle ein $s_i\in \mathbb{F}_p[F/F_n]$ mit $r_i\equiv s_{i}\mod \ker(\pi_n)$. Wähle für jedes $i$ ein Urbild $y_i\in\mathbb{F}_p[F]$ von $s_i$. Dann gilt mit $I_n:=\ker(\mathbb{F}_p[F]\to\mathbb{F}_p[F/F_n])$ und $I:=I_1$:
\[ \sum_{i=1}^d(x_i-1)y_i\in I_n\subset I^n \]
Es existieren also $r_{j_1,\ldots,j_n}\in\mathbb{F}_p[F]$ für alle $j_1,\ldots,j_n=1,\ldots,d$, so dass:
\[  \sum_{i=1}^d(x_i-1)y_i = \sum_{j_1,\ldots,j_n=1,\ldots,d} (x_{j_1}-1)\cdots (x_{j_n}-1) r_{j_1,\ldots,j_n} \]
Nach dem vorherigen Satz folgt für alle $i$:
\[y_i = \sum_{j_2,\ldots,j_n=1,\ldots,d} (x_{j_2}-1)\cdots (x_{j_n}-1)r_{i,j_2,\ldots, j_n} \in I^{n-1} \]
\end{proof}
\fi

\section{Die Golod-Shafarevich Ungleichung}

Sei $G$ eine pro"~$p$ Gruppe, so dass $d:=d(G)$ und $t:=t(G)$ endlich sind und $R:=\mathbb{F}_p[[G]]$ der vollständige Gruppenring. Wir haben eine Darstellung $G=\langle X;S\rangle$ mit $\#X=d$ und $\#S=t$. Setzen wir $F:=\widehat{F}(X)$ und $N:=\overline{N(S)}\lhd F$, so erhalten wir eine kurze exakte Folge:
\[ 1\to N\to F\to G\to 1 \]

\begin{definition}
Sei $I\subset\mathbb{F}_p[G]$ das Augmentationsideal und $\widehat{I}$ sein Abschluss in $\mathbb{F}_p[[G]]$. Wir setzen:
\[ c_n := \dim_{\mathbb{F}_p}(I^n/I^{n+1}) =\dim_{\mathbb{F}_p}(\widehat{I}^n/\widehat{I}^{n+1}),\qquad \varrho:=\limsup_{n\to\infty}\sqrt[n]{c_n} \]
$\varrho^{-1}$ ist gerade der Konvergenzradius der Potenzreihe $\sum_{n=0}^\infty c_n X^n$ im archimedischen Sinne.
\end{definition}

\begin{theorem}\textit{(Die Golod-Shafarevich Ungleichung)}\label{thm:golod-shafarevich}
Sei $G$ eine endlich erzeugte pro"~$p$ Gruppe mit $d(G)>1$. Ist $\varrho\leq 1$, so gilt:
\[ t(G)\geq \frac{d(G)^2}{4} \]
Ist $d(G)\neq 2$ oder $t(G)\neq 1$, so ist diese Ungleichung strikt.
\end{theorem}

\begin{remark}
Ist $t(G)$ unendlich, so ist die Ungleichung trivialerweise erfüllt.
\end{remark}

\begin{corollary}
Für endliche $p$-Gruppen $G$ gilt:
\[t(G)\geq \frac{d(G)^2}{4} \]
\end{corollary}

\begin{proof}
Folgt aus Theorem~\ref{thm:golod-shafarevich} und Korollar~\ref{cor:finite-group-rho}. Ist $G$ zyklisch, so ist $t(G)\geq 1$, da nichttriviale freie pro"~$p$ Gruppen unendlich sind.
\end{proof}

\begin{corollary}
Für $p$-adisch analytische pro"~$p$ Gruppen $G$, die nicht prozyklisch sind, gilt:
\[t(G)\geq \frac{d(G)^2}{4}\]
\end{corollary}

\begin{proof}
Folgt aus Theorem~\ref{thm:golod-shafarevich} und Satz~\ref{Ex7.5}.
\end{proof}

\paragraph{} Bevor wir die Golod-Shafarevich Ungleichung beweisen können, benötigen wir zunächst das folgende Hilfslemma:

\begin{lemma}\label{D2}
Die kurze exakte Folge $1\to N\to F\to G\to 1$ induziert die folgende exakte Folge von $R$"~Moduln:
\[ R^{(t)}\stackrel{\alpha}{\to} R^{(d)}\to \widehat{I}\to 0 \]
mit $\alpha(R^{(t)})\subset \widehat{I}^{(d)}$.
\end{lemma}

\begin{proof}
Setze $F_n:=P_n(F)$ und $G_n:=P_n(G)$ und sei $N=\overline{N(y_1,\ldots,y_t)}\lhd F$. Wir erhalten kompatible kurze exakte Folgen:
\[ 0\to\sum_{i=1}^t(y_iF_n-1)\mathbb{F}_p[F/F_n]\to \mathbb{F}_p[F/F_n]\to\mathbb{F}_p[G/G_n]\to 0 \]
Da alle Objekte endlich sind, erhalten wir nach Korollar~\ref{prop:compact-exact-limes} unter $\varprojlim$ die kurze exakte Folge:
\[0\to K\to \mathbb{F}_p[[F]]\stackrel{\beta}{\to} \mathbb{F}_p[[G]]\to 0 \]
mit einem Ringhomomorphismus $\beta$ und $K=\sum_{i=1}^t (y_i-1)\mathbb{F}_p[[F]]$, wobei $K\subset\mathbb{F}_p[[F]]$ als endlich erzeugtes Ideal in einem kompakten Hausdorffraum kompakt und daher abgeschlossen ist.

Sei $\widehat{I}_F:=(F-1)\mathbb{F}_p[[F]]$ der Abschluss des Augmentationsideals $I_F\subset\mathbb{F}_p[F]$ in $\mathbb{F}_p[[F]]$. Dann ist $K\subset \widehat{I}_F$ und $\beta(\widehat{I}_F)=\widehat{I}$. $\beta$ induziert einen surjektiven Homomorphismus $\widehat{I}_F/\widehat{I}_F^2\to \widehat{I}/\widehat{I}^2$. Da $d(F)=d=d(G)$, gilt nach Korollar~\ref{cor:Ex1} und Lemma~\ref{lemma:nuisance} $\widehat{I}_F/\widehat{I}_F^2\cong I_F/I_F^2\cong \mathbb{F}_p^{(d)}\cong I/I^2\cong \widehat{I}/\widehat{I}^2$. Daher folgt $K=\ker(\beta)\subset \widehat{I}_F^2$. Aus der obigen kurzen exakten Folge erhalten wir somit die folgende exakte Folge:
\[ K/K^2\stackrel{\gamma}{\to}\widehat{I}_F/\widehat{I}_FK\to \widehat{I}\to 0 \]
mit $\operatorname{im}(\gamma)\subset\widehat{I}_F^2/\widehat{I}_FK$. Da $K$ ein durch $t$ Elementen erzeugter $\mathbb{F}_p[[F]]$-Rechtsmodul ist, ist $K/K^2$ ein von $t$ Elementen erzeugter $\mathbb{F}_p[[G]]$-Rechtsmodul. Wir erhalten also einen surjektiven Mo\-dul\-ho\-mo\-mor\-phis\-mus $\pi:R^{(t)}\to K/K^2$. Andererseits ist nach Satz~\ref{prop:Ex2} $\widehat{I}_F=(F-1)\mathbb{F}_p[[F]]$ ein freier $\mathbb{F}_p[[F]]$-Modul auf $d$ Erzeugern. Es folgt:
\[\widehat{I}_F/\widehat{I}_FK\cong \mathbb{F}_p[[F]]^{(d)}/\mathbb{F}_p[[F]]^{(d)}K\cong R^{(d)} \]
Zusammen erhalten wir die gewünschte exakte Folge:
\[R^{(t)}\stackrel{\alpha}{\to}R^{(d)}\to \widehat{I}\to 0 \]
mit $\alpha:=\gamma\circ\pi$. Es gilt $\alpha(R^{(t)})=\operatorname{im}(\gamma)\subset\widehat{I}^2_F/\widehat{I}_FK\cong R^{(d)}\widehat{I}=\widehat{I}^{(d)}$.
\end{proof}

\begin{proof}[Beweis von Theorem~\ref{thm:golod-shafarevich}] 
Da $\varrho\leq 1$, konvergiert die folgende Potenzreihe für $x\in (0,1)$:
\[P(x) := \sum_{n=0}^\infty c_nx^n \]
Für $n\geq 0$ setze $s_n:=\dim_{\mathbb{F}_p}(R/\widehat{I}^{n+1})=\sum_{j=0}^nc_j$ und zusätzlich $s_{-1}:=0$. Sei $n\geq 1$. Tensorieren wir die exakte Folge aus Lemma~\ref{D2} mit $R/\widehat{I}^n$, so erhalten wir:
\[ (R/\widehat{I}^n)^{(t)}\to (R/\widehat{I}^n)^{(d)} \to \widehat{I}/\widehat{I}^{n+1} \to 0 \]
Mit $\alpha$ aus Lemma~\ref{D2} und $K_n:=\alpha^{-1}( (\widehat{I}^n)^{(d)})$, erhalten wir die kurze exakte Folge:
\[ 0\to R^{(t)}/K_n \stackrel{\alpha_n}{\to} (R/\widehat{I}^n)^{(d)} \to\widehat{I}/\widehat{I}^{n+1}\to 0 \]
Sei $i\in\widehat{I}^{n-1}$ und $r\in R^{(t)}$. Dann gilt $\alpha(ir)=i\alpha(r)\in(\widehat{I}^n)^{(d)}$, da $\alpha(R^{(t)})\subset\widehat{I}^{(d)}$, d.h. $ir\in K_n$. Daher ist $R^{(t)}/K_n$ ein $R/\widehat{I}^{n-1}$"~Modul und höchstens von der $\mathbb{F}_p$"~Dimension $ts_{n-2}$. Die beiden anderen Terme haben Dimensionen $ds_{n-1}$ bzw. $s_n-1$, da $\dim_{\mathbb{F}_p}(R/\widehat{I})=1$. Es folgt:
\[ts_{n-2}-ds_{n-1}+s_n\geq 1 \]
Multiplizieren wir diese Ungleichung mit $x^{n-1}-x^n$ und addiert man die resultierenden Ungleichungen, so erhalten wir unter Beachtung von $s_n-s_{n-1}=c_n$ und $x\in (0,1)$:
\[ P(x)(tx^2-dx+1)\geq 1 \]
Da $P(x)>0$, folgt $tx^2-dx+1>0$ für alle $x\in (0,1)$. Für $x\nearrow 1$ erhalten wir $t-d+1\geq 0$, also $t\geq d-1$. Für $d=2$ folgt $t\geq 1$. Sei nun $d\geq 3$. Dann gilt $\frac{d}{2t}\leq\frac{d}{2d-2} <1$. Setzen wir $x:=\frac{d}{2t}$ in die Ungleichung $tx^2-dx+1>0$ ein, erhalten wir $-\frac{d^2}{4t}+1>0$.
\end{proof}

\iffalse
\section{Die Ungleichung für Vervollständigungen}

Sei $p$ eine Primzahl, $\Gamma$ eine Gruppe und $\Gamma=\langle X;R\rangle$ eine endliche Darstellung als abstrakte Gruppe, d.h. $\Gamma\cong F(X)/N(R)$, wobei $F(X)$ die freie Gruppe auf $X$ bezeichnet.

\begin{proposition}\label{prop:D4}
Es ist $\widehat{\Gamma}=\langle X;R\rangle$ eine Darstellung der pro"~$p$ Vervollständigung von $\Gamma$ in der Kategorie der pro"~$p$ Gruppen.
\end{proposition}

\begin{proof}
Wir zeigen, dass $\widehat{F}(X)/\overline{N(R)}$ die Universaleigenschaft der pro"~$p$ Vervollständigung von $\Gamma$ erfüllt. Sei $\Gamma \to G$ ein Homomorphismus in eine pro"~$p$ Gruppe $G$. Der Gruppenho\-mo\-mor\-phis\-mus $F(X)\to \Gamma\to G$ induziert einen eindeutigen, stetigen Homomorphismus $\pi:\widehat{F}(X)\to G$. Da $R\subset\ker(\pi)$ und $\widehat{F}(X)$ hausdorffsch, induziert $\pi$ einen eindeutigen, stetigen Homomorphismus $\widehat{F}(X)/\overline{N(R)}\to G$.
\end{proof}

\begin{definition}
Da $d(\Gamma)=d(\widehat{\Gamma})$ im Allgemeinen nicht gilt, können wir die Golod-Shafarevich Ungleichung nicht direkt für abstrakte Gruppen verwenden. Wir definieren:
\[d_p(\Gamma)=d(\Gamma/\Gamma^p[\Gamma,\Gamma]) \]
\end{definition}

\begin{proposition}\label{prop:D5}
Ist $\Gamma$ eine endlich erzeugte Gruppe, so gilt $d(\widehat{\Gamma})=d_p(\Gamma)$.
\end{proposition}

\begin{proof}
Setze $G:=\widehat{\Gamma}$. Nach Satz~\ref{prop:D4} ist $G$ topologisch endlich erzeugt. Da $\Gamma$ endlich erzeugt, ist $\Gamma/\Gamma^p[\Gamma,\Gamma]$ ein endlich dimensionaler $\mathbb{F}_p$-Vektorraum, also ist $\Gamma^p[\Gamma,\Gamma]$ vom $p$-Potenz-Index. Wir erhalten daher die Projektion $G\to\Gamma/\Gamma^p[\Gamma,\Gamma]$. Diese induziert eine Surjektion $G/\Phi(G)\to\Gamma/\Gamma^p[\Gamma,\Gamma]$. Als surjektive $\mathbb{F}_p$-lineare Abbildung zwischen endlich dimensionierte $\mathbb{F}_p$-Vektorräume ist diese ein Isomorphismus. Es folgt $d(G)=d(G/\Phi(G))=d_p(\Gamma)$.
\end{proof}

\begin{proposition}\label{prop:D6}
Sei $G=\langle X;R\rangle$ eine pro"~$p$ Gruppe mit endlicher pro"~$p$ Darstellung. Dann existiert eine Darstellung $G=\langle Y;S\rangle$ mit $\#Y=d(G)$ und $\#S=\#R-(\#X-\#Y)$.
\end{proposition}

\begin{proof}
Sei $F:=\widehat{F}(X)$, also haben wir einen surjektiven Homomorphismus $F\to G$ mit Kern $\overline{N(R)}\lhd F$. Diese induziert einen surjektiven Homomorphismus $\phi:F/\Phi(F)\to G/\Phi(G)$. Setze $K:=\ker(\phi)$. $K$ ist ein Unterraum des $\mathbb{F}_p$-Vektorraums $F/\Phi(F)$, der durch das Bild von $R$ erzeugt wird. Sei $\{r_1\Phi(F),\ldots,r_m\Phi(F)\}$ eine Basis von $K$ mit $r_1,\ldots,r_m\in R$. Dann gilt $m=d(F)-d(G)$ und $R':=\{r_1,\ldots,r_m\}\subset R$ ist Teilmenge einer freien Basis von $F$.
\end{proof}

\begin{corollary}\label{cor:D7}
Es gilt $t(\widehat{\Gamma})\leq \#R -(\#X-d_p(\Gamma))$.
\end{corollary}

\begin{theorem}
Sei $\Gamma=\langle X;R\rangle$ eine abstrakte Gruppe mit endlicher Darstellung und $p$ eine Primzahl. Erfüllt die pro"~$p$ Vervollständigung $\widehat{\Gamma}$ die Voraussetzungen von Theorem~\ref{thm:golod-shafarevich}, so gilt:
\[\#R -(\#X-d_p(\Gamma))\geq \frac{d_p(\Gamma)^2}{4} \]
\end{theorem}

\begin{proof}
Folgt direkt aus Korollar~\ref{cor:D7} und Theorem~\ref{thm:golod-shafarevich}.
\end{proof}
\fi

\begin{thebibliography}{99999999}

\bibitem[DDMS99]{DDMS99}
	J. D. Dixon, M. P. F. du Sautoy, A. Mann, D. Segal (1999)
	\textit{Analytic Pro"~$p$ Groups}.
	Cambridge Studies in Advanced Mathematics,
	2nd Edition.
\bibitem[GS64]{GS64}
	E. S. Golod, I. R. \v{S}afarevi\v{c} (1964)
	\textit{Über Klassenkörpertürme} (russisch). Izv. Akad. Nauk SSSR, Ser. Mat. \textbf{28} 261-272
\bibitem[Koc02]{Koc02}
	H. Koch (2002) \textit{Galois Theory of $p$-Extensions}. Springer-Verlag.
\bibitem[Laz65]{Laz65}
	M. Lazard (1965) \textit{Groupes analytiques $p$-adiques}. Inst. Hautes Études Scientifiques, Publ. Math. \textbf{26} 389-603
\bibitem[LM87]{LM87}
	A. Lubotzky and A. Mann (1987) \textit{Powerful $p$-Groups}. J. Algebra. \textbf{105} 484-515
\bibitem[Wil98]{Wil98}
	J. S. Wilson (1998) \textit{Profinite Groups}. Oxford University Press, Oxford.

\end{thebibliography}

%\printindex
\end{document}

### 2015-10-01 Master Thesis/synopsis.tex
Til main result

@@
 The \textit{rational rank}\index{valuation!rational rank} of $v:K\to\Gamma_v$ is defined as the rank of its valuation group $\Gamma_v$ as a $\mathbb{Z}$-module:
-\[\operatorname{rrank}(v) = \dim_\mathbb{Q}\Gamma_v\otimes_\mathbb{Z}\mathbb{Q} \]
+\[\operatorname{rrank}(v) = \dim_\mathbb{Q}(\Gamma_v\otimes_\mathbb{Z}\mathbb{Q}) \]
 \end{definition}
@@
 \end{lemma}
+
+\begin{lemma}
+Let $v\in\operatorname{Spv}(K)$. There is a canonical bijection:
+\[\{\text{Specializations of }v \} \to \operatorname{Spv}k(v),\ w\mapsto w\mod v\]
+If $w\in\operatorname{Spv}_\text{rd}^{(i)}(K)$ for some $i\geq 1$ then there exists a unique $v(w)\in\operatorname{Spv}_\text{rd}^{(i-1)}(K)$ such that $\mathcal{O}_w\subset\mathcal{O}_v$.
+$w$ is henselian if and only if $v$ and $w\mod v$ are henselian.
+\end{lemma}
+
+\paragraph{} Let $K$ be a field of characteristic $0$ and dimension $2$.
+
+\begin{lemma}
+Let $K$ be henselian with respect to two different valuations $w_1,w_2\in |\operatorname{Spv}(K)|$.
+Then $K$ is algebraically closed or $w_1,w_2\in\operatorname{Spv}_\text{rd}^{(2)}(K),\ v(w_1)=v(w_2)$ and $k(v(w_1))$ is separably closed.
+In the last case we have $\operatorname{cd}(G_K)\leq 2$.
+\end{lemma}
+
+\begin{definition}
+Let $K$ be finitely generated.
+A \textit{model} of $K$ is a regular, proper scheme $\mathfrak{X}$ over $\mathbb{Z}$ with function field $K$.
+A \textit{morphism of models} is a morphism of schemes inducing the identity on $K$.
+We denote the category of models of $K$ by $\mathbf{M}(K)$; it is filtered.
+
+For every model $\mathfrak{X}$ there is a canonical map $\pi_\mathfrak{X}:\operatorname{Spv}(K)\to\mathfrak{X}$.
+$\pi_\mathfrak{X}(v)$  is the unique point of $\mathfrak{X}$ such that $\mathcal{O}_{\mathfrak{X}, \pi_\mathfrak{X}(v)}\subset\mathcal{O}_v$ and the inclusion $\mathcal{O}_{\mathfrak{X}, \pi_\mathfrak{X}(v)}\hookrightarrow \mathcal{O}_v$ is a homomorphism of local rings.
+\end{definition}
+
+\begin{lemma}
+\begin{enumerate}[(i)]
+\item $\operatorname{Spv}(K)$ together with the maps $\pi_{\mathfrak{X}}$ is the projective limit of topological spaces: 
+\[\operatorname{Spv}(K) =\varprojlim_{\mathfrak{X}\in\mathbf{M}(K)}\mathfrak{X}\]
+In particular we have:
+\[ |\operatorname{Spv}(K)| =\varprojlim_{\mathfrak{X}\in\mathbf{M}(K)}|\mathfrak{X}| \]
+\item For all $v\in\operatorname{Spv}(K)$ we have:
+\[ \mathcal{O}_v = \varinjlim_{\mathfrak{X}\in\mathbf{M}(K)}\mathcal{O}_{\mathfrak{X},\pi_\mathfrak{X}(v)} \]
+\item For $v\in\operatorname{Spv}(K)$ let $K_v$ be the quotient field of a henselization of $\mathcal{O}_v$ and for $\mathfrak{X}\in\mathbf{M}(K),\ x\in\mathfrak{X}$ let $K_\mathfrak{X}$ be the quotient field of a henselization of $\mathcal{O}_{\mathfrak{X},x}$.
+Then:
+\[ K_v=\varinjlim_{\mathfrak{X}\in\mathbf{M}(K)}K_{\pi_\mathfrak{X}(v)} \]
+\end{enumerate}
+\end{lemma}
+
+\begin{lemma}\label{lemma:main-result}
+Let $\ell\neq 2$ be a prime. The following map, induced by the restrictions, is injective:
+\[ \mathrm{H}^3(K,\mathbb{Z}/\ell\mathbb{Z})\to\prod_{w\in |\operatorname{Spv}(K)| }\mathrm{H}^3(K_w,\mathbb{Z}/\ell\mathbb{Z}) \]
+\end{lemma}

### 2015-10-01 Master Thesis/synopsis.tex
Starting section "Local correspondence"

@@
 \end{lemma}
+
+\begin{lemma}
+Let $K$ be henselian with respect to $w\in|\operatorname{Spv}(K)|$.
+Set $k=k(w)$ and $\Gamma = \Gamma_w$.
+Then $k$ is an algebraic extension of a finite field $\mathbb{F}_p$.
+Let $\ell\neq p$ be a prime and assume $\mu_\ell\subset K$.
+Then the following conditions are equivalent:
+\begin{enumerate}[(i)]
+\item $\mathrm{H}^3(K,\mathbb{Z}/\ell\mathbb{Z})=\mathbb{Z}/\ell\mathbb{Z}$
+\item $\ell^\infty\nmid[k:\mathbb{F}_p]$ and $\dim_{\mathbb{F}_\ell}(\Gamma/\ell\Gamma) = 2$
+\end{enumerate}
+Otherwise we have $\mathrm{H}^3(K,\mathbb{Z}/\ell\mathbb{Z})=0$. If $L/K$ is an algebraic extension and (i) holds, we have:
+\[ \mathrm{H}^3(L,\mathbb{Z}/\ell\mathbb{Z}) = \begin{cases}
+\mathbb{Z}/\ell\mathbb{Z}, & \ell^\infty\nmid [L:K]\\
+0, & \text{otherwise}
+\end{cases} \]
+\end{lemma}
+
+\begin{proposition}
+Let $K_0/K$ be an algebraic extension.
+\begin{enumerate}[(i)]
+\item Suppose there exists a prime number $\ell$ with $\ell^\infty\nmid[K_0:K]$ and such that for every algebraic extension $L/K_0$ we have: 
+\[ \mathrm{H}^3(L,\mathbb{Z}/\ell\mathbb{Z}) = \begin{cases}
+\mathbb{Z}/\ell\mathbb{Z}, & \ell^\infty\nmid [L:K_0]\\
+0, & \text{otherwise}
+\end{cases} \]
+Then there exists a unique henselian $w\in|\operatorname{Spv}(K)|$.
+\item Assume that the hypothesis of (i) is satisfied for two different prime numbers. Then $w$ is defectless.
+\end{enumerate}
+\end{proposition}
+
+\section{Local correspondence}
+
+Let $K,K'$ be two finitely generated fields of characteristic $0$ and dimension $2$ and $\sigma:G_K\to G_{K'}$ be an isomorphism of profinite groups.
+For intermediate fields $L$ of $\overline{K}/K$ the corresponding intermediate field of $\overline{K'}/K'$ will be denoted by $L'$.
+In other words $\sigma G_L=G_{L'}$.
+
+\begin{lemma}
+Let $w\in\operatorname{Spv}_\text{d}^{(2)}(K)$ and let $E=K_w$ be a henselization of $w$.
+Then $E'$ is a henselization of $K'$ with respect to a unique $w'\in\operatorname{Spv}_\text{d}^{(2)}(K')$.
+The map $w\mapsto w'$ does not depend on the choice of the henselization.
+\end{lemma}
+
+\begin{lemma}
+$\sigma$ induces a bijection $\sigma:\operatorname{Spv}_\text{d}^{(1)}(K)\to\operatorname{Spv}_\text{d}^{(1)}(K')$.
+If $E=K_v\subset\overline{K}$ is a henselization of $K$ with respect to $v\in\operatorname{Spv}_\text{d}^{(1)}(K)$ then $E'$ is a henselization of $K'$ with respect to $v'=\sigma(v)$.
+\end{lemma}

### 2015-10-01 Master Thesis/synopsis.tex
Local correspondence end

@@
 If $E=K_v\subset\overline{K}$ is a henselization of $K$ with respect to $v\in\operatorname{Spv}_\text{d}^{(1)}(K)$ then $E'$ is a henselization of $K'$ with respect to $v'=\sigma(v)$.
+Furthermore, if $E^{\text{t}}$ and $E'^{\text{t}}$ denotes the inertial fields of $E$ and $E'$ respectively, then $\sigma$ induces an isomorphism:
+\[ G_{k(v)}\cong\operatorname{Gal}(E^\text{t}/E)\to\operatorname{Gal}(E'^{\text{t}}/E')\cong G_{k(v')} \]
+which is induced by a unique isomorphism $\varphi_v^\text{s}:k(v')^\text{s}\to k(v)^\text{s}$. In particular $\varphi_v=\varphi_v^\text{s}|_{k(v')}:k(v')\to k(v)$ is an isomorphism.
 \end{lemma}

### 2015-10-01 Master Thesis/synopsis.tex
Finish synopsis

@@
 
-\section{Some results on $\operatorname{Spv}(K)$}
+\section{Some Results on $\operatorname{Spv}(K)$}
 
@@
 
-\section{Local correspondence}
+\section{Local Correspondence}
 
@@
 \end{lemma}
+
+\section{Global Correspondence mod $n$}
+
+\begin{lemma}\label{4.1}
+Let $E$ be a number field and let $X$ be a smooth projective geometrically connected curve over $E$.
+Let $\pi:X'\to X$ be a connected Galois covering of degree $d$ which splits completely, i.e.
+every closed point $P\in X_0$ has exactly $d$ pre-images in $X'$.
+Then $d=1$.
+\end{lemma}
+
+\paragraph{} Let $k$ and $k'$ be the algebraic closures of $\mathbb{Q}$ in $K$ and $K'$ respectively, and let $\overline{k}$ and $\overline{k'}$ be the algebraic closure of $k$ and $k'$ in $\overline{K}$ and $\overline{K'}$ respectively.
+
+Let $X$ and $X'$ be the unique smooth projective curve over $k$ and $k'$ with function field $K$ and $K'$ respectively.
+We identify the set of closed points $X_0$ of $X$ with the set of valuations $v\in\operatorname{Spv}_\text{d}^{(1)}(K)$ with $\operatorname{char} k(v)=0$ and similarly identify $X'_0$ with the corresponding subset of $\operatorname{Spv}_\text{d}^{(1)}(K')$.
+
+\begin{lemma}\label{4.2}
+We have $(K\overline{k})'=K'\overline{k'}$, i.e. $\sigma(G_{K\overline{k}})=G_{K'\overline{k'}}$. $\sigma$ induces an isomorphism $G_k\to G_{k'}$ such that the following diagram commutes:
+\[ \begin{tikzcd}
+G_K \ar[r, "\sigma"] \ar[d, "\text{res}"'] & G_{K'}\ar[d, "\text{res}"]\\
+G_k \ar[r, "\sigma"'] & G_{k'}
+\end{tikzcd} \]
+Moreover, $\sigma:G_k\to G_{k'}$ is induced by a unique isomorphism $\varphi:\overline{k'}\to\overline{k}$.
+\end{lemma}
+
+\begin{definition}\label{4.4}
+For a finite subset $S\subset X_0$ set:
+\[ K^S=\{x\in K^\times \mid v_P(x)=0\text{ for all }P\in X_0\setminus S \} \]
+$K'^{S'}$ is defined similarly. For finite subsets $S$ of $X_0$, the corresponding subsets of $X_0'$ will be denoted by $S'$, i.e. $S'=\sigma(S)$.
+\end{definition}
+
+\begin{lemma}\label{4.5}
+\begin{enumerate}[(i)]
+\item There exists a finite subset $S\subset X_0$ such that:
+\[ \operatorname{Pic}(X\setminus S)=0=\operatorname{Pic}(X'\setminus S') \]
+Such a set $S$ is called \textit{admissible}. If $V\subset X$ is an open subscheme, then one can choose $S$ in $V_0$.
+\item Let $S$ be admissible. For all $n\in\mathbb{N}$ the following canonical maps are injective:
+\[ K^S/n \to K^\times /n,\quad K'^{S'}/n\to K'^\times /n \]
+$\varphi_n$ induces isomorphisms $\varphi_n: K'^{S'}/n\to K^S/n$.
+\end{enumerate}
+\end{lemma}
+
+\section{The Proof}
+
+\begin{definition}\label{5.1}
+An abelian group is called \textit{free-by-finite} if its torsion subgroup $A_\text{tor}$ is finite and the quotien $A/A_\text{tor}$ is free abelian.
+$A$ is free-by-finite if and only if $A$ is the direct sum of a finite group and a free abelian group.
+\end{definition}
+
+\begin{lemma}\label{5.2}
+\begin{enumerate}[(i)]
+\item Let $A\to B$ be a homomorphism of abelian groups with finite kernel and cokernel. Then $A$ is free-by-finite iff $B$ is free-by-finite.
+\item Let $0\to A'\stackrel{i}{\to}A\stackrel{p}{\to}A''\to 0$ be an exact sequence of abelian groups.
+If $A'$ and $A''$ are free-by-finite, then $A$ is free-by-finite.
+Conversely, if $A$ is free-by-finite, then $A'$ is free-by-finite.
+If additionally, $A'$ is finitely generated, then $A''$ is free-by-finite.
+\item Let $A$ be free-by-finite. Then $\bigcap_{n\in\mathbb{N}}nA=0$.
+\end{enumerate}
+\end{lemma}
+
+\begin{lemma}\label{5.3}
+Let $F$ be a number field. Then:
+\begin{enumerate}[(i)]
+\item $F^\times$ is free-by-finite.
+\item Let $F_1,\ldots,F_n$ be finite extensions of $F$.
+Then $\prod_{i=1}^{n}F_i^\times/F^\times$ is free-by-finite, where $F^\times$ is considered as a subgroup of $\prod_{i=1}^{n}F_i^\times$ via the diagonal embedding.
+\end{enumerate}
+\end{lemma}
+
+\begin{lemma}
+Let $S\subset X_0$ be admissible. Then there exists a finite subset $T\subset X_0$ with $T\cap S=\varnothing$ such that the following map is injective:
+\[ r_{S,T}:K^S \to\prod_{P\in T}k(P)^\times,\ x\mapsto(r_P(x))_{P\in T} \]
+In this case, $\operatorname{coker}(r_{S,T})$ is free-by-finite.
+\end{lemma}

### 2015-10-01 Master Thesis/synopsis.tex
Add labels

@@
 
-\begin{definition}
+\begin{definition}\label{2.1}
 Let $K$ be a field of finite dimension.
@@
 
-\begin{lemma}
+\begin{lemma}\label{2.2}
 Let $K$ be henselian with respect to two different valuations $w_1,w_2\in |\operatorname{Spv}(K)|$.
@@
 
-\begin{lemma}
+\begin{lemma}\label{2.3}
 \begin{enumerate}[(i)]
@@
 
-\begin{lemma}\label{lemma:main-result}
+\begin{lemma}\label{2.4}
 Let $\ell\neq 2$ be a prime. The following map, induced by the restrictions, is injective:
@@
 
-\begin{lemma}
+\begin{lemma}\label{2.5}
 Let $K$ be henselian with respect to $w\in|\operatorname{Spv}(K)|$.
@@
 
-\begin{lemma}
+\begin{lemma}\label{5.4}
 Let $S\subset X_0$ be admissible. Then there exists a finite subset $T\subset X_0$ with $T\cap S=\varnothing$ such that the following map is injective:

### 2015-10-01 Master Thesis/synopsis.tex
Add steps

@@
 \end{lemma}
+
+\begin{theorem}[\textit{Pop}]\label{1.1}
+Let $K,K'$ be two finitely generated fields of characteristic $0$ and dimension $2$.
+Suppose there is an isomorphism $\sigma:G_K\to G_{K'}$ between their absolute Galois groups, then there exists a unique field isomorphism $\varphi:\overline{K'}\to\overline{K}$ with $\sigma(g)=\varphi^{-1}g\varphi$ for all $g\in G_K$.
+In particular $\varphi$ induces a field isomorphism $\varphi:K'\to K$.
+\end{theorem}
+
+\begin{step}
+There exists a unique group isomorphism $\varphi:K'^\times\to K^\times$ such that for any $P\in X_0$ we have $\varphi(U_{P'})=U_P$ where $P'=\sigma(P)$ and the following diagram:
+\[ \begin{tikzcd}
+U_{P'} \ar[r, "\varphi"]\ar[d, "r_{P'}"'] & U_P \ar[d, "r_P"]\\
+k(P') \ar[r, "\varphi_P"'] & k(P)
+\end{tikzcd} \]
+\end{step}
+
+\begin{step}
+Extending $\varphi$ to a map $K'\to K$ by setting $\varphi(0)=0$ yields a field isomorphism $\varphi:K'\to K$.
+\end{step}
+
+\begin{step}
+$\varphi$ induces $\sigma:G_K\to G_{K'}$, i.e. $\sigma(g) = \varphi^{-1}g\varphi$ for all $g\in G_K$, and is uniquely determined by this property.
+\end{step}

### 2015-10-01 Master Thesis/synopsis.tex
Add labels

@@
 
-\begin{proposition}
+\begin{proposition}\label{2.6}
 Let $K_0/K$ be an algebraic extension.
@@
 
-\begin{lemma}
+\begin{lemma}\label{3.1}
 Let $w\in\operatorname{Spv}_\text{d}^{(2)}(K)$ and let $E=K_w$ be a henselization of $w$.
@@
 
-\begin{lemma}
+\begin{lemma}\label{3.2}
 $\sigma$ induces a bijection $\sigma:\operatorname{Spv}_\text{d}^{(1)}(K)\to\operatorname{Spv}_\text{d}^{(1)}(K')$.

### 2015-10-01 Master Thesis/synopsis.tex
Change model category notation

@@
 
-\paragraph{} Let $K$ be a field of characteristic $0$ and dimension $2$.
+\paragraph{} Let $K$ be a finitely generated field of characteristic $0$ and dimension $2$.
 
@@
 A \textit{morphism of models} is a morphism of schemes inducing the identity on $K$.
-We denote the category of models of $K$ by $\mathbf{M}(K)$; it is filtered.
+We denote the category of models of $K$ by $\mathfrak{M}(K)$; it is filtered.
 
@@
 \begin{enumerate}[(i)]
-\item $\operatorname{Spv}(K)$ together with the maps $\pi_{\mathfrak{X}}$ is the projective limit of topological spaces: 
-\[\operatorname{Spv}(K) =\varprojlim_{\mathfrak{X}\in\mathbf{M}(K)}\mathfrak{X}\]
+\item $\operatorname{Spv}(K)$, together with the maps $\pi_{\mathfrak{X}}$, is the projective limit of topological spaces: 
+\[\operatorname{Spv}(K) =\varprojlim_{\mathfrak{X}\in\mathfrak{M}(K)}\mathfrak{X}\]
 In particular we have:
-\[ |\operatorname{Spv}(K)| =\varprojlim_{\mathfrak{X}\in\mathbf{M}(K)}|\mathfrak{X}| \]
+\[ |\operatorname{Spv}(K)| =\varprojlim_{\mathfrak{X}\in\mathfrak{M}(K)}|\mathfrak{X}| \]
 \item For all $v\in\operatorname{Spv}(K)$ we have:
-\[ \mathcal{O}_v = \varinjlim_{\mathfrak{X}\in\mathbf{M}(K)}\mathcal{O}_{\mathfrak{X},\pi_\mathfrak{X}(v)} \]
-\item For $v\in\operatorname{Spv}(K)$ let $K_v$ be the quotient field of a henselization of $\mathcal{O}_v$ and for $\mathfrak{X}\in\mathbf{M}(K),\ x\in\mathfrak{X}$ let $K_\mathfrak{X}$ be the quotient field of a henselization of $\mathcal{O}_{\mathfrak{X},x}$.
+\[ \mathcal{O}_v = \varinjlim_{\mathfrak{X}\in\mathfrak{M}(K)}\mathcal{O}_{\mathfrak{X},\pi_\mathfrak{X}(v)} \]
+\item For $v\in\operatorname{Spv}(K)$ let $K_v$ be the quotient field of a henselization of $\mathcal{O}_v$ and for $\mathfrak{X}\in\mathfrak{M}(K),\ x\in\mathfrak{X}$ let $K_\mathfrak{X}$ be the quotient field of a henselization of $\mathcal{O}_{\mathfrak{X},x}$.
 Then:
-\[ K_v=\varinjlim_{\mathfrak{X}\in\mathbf{M}(K)}K_{\pi_\mathfrak{X}(v)} \]
+\[ K_v=\varinjlim_{\mathfrak{X}\in\mathfrak{M}(K)}K_{\pi_\mathfrak{X}(v)} \]
 \end{enumerate}

### 2015-10-01 Master Thesis/thesis.tex
Add synopsis
Print index
Pop introduction
Pop bibliography
Add tikz-babel package
Add steps
Fix declaration

--- 2015-07-15 Master Thesis/thesis.tex
@@
-\documentclass[11pt,a4paper,openany]{memoir}
+\documentclass[11pt, a4paper, openany]{memoir}
 \usepackage{fontspec}
@@
 \usepackage{amsmath, amsfonts, amssymb, amsthm}
-\usepackage{paralist}
-\usepackage{tikz-cd}
+\usepackage{paralist, tikz}
+\usetikzlibrary{cd, babel}
+\usepackage[inner=4cm, outer=3cm, top=3cm, bottom=3cm]{geometry}
+\usepackage{hyperref}
+\hypersetup{bookmarksopen=true, pdfstartview=FitH}
 \linespread{1.1}
-\usepackage{hyperref}
-\hypersetup{
-	bookmarksopen=true,
-	pdfstartview=FitH
-}
-\usepackage[inner=4cm,outer=3cm,top=3cm,bottom=3cm]{geometry}
 \author{Yichuan Shen}
@@
 \newtheorem*{remark}{Remark}
+\newtheorem{step}{Step}
 
@@
 \paragraph{Declaration.} I declare that I carried out this master thesis independently, and only with the cited sources, literature and other professional sources.
-The master thesis was not used in the same or in a similar version to achieve an academic grading or is being published elsewhere.\\
+This thesis was not used in the same or in a similar version to achieve an academic grading or is being published elsewhere.\\
 \vspace{7mm}\\
@@
 
-\mainmatter
 \pagenumbering{arabic}
@@
 
-\chapter{Introduction}
+\input{introduction.tex}
 
-\clearpage
+\mainmatter
+\setcounter{page}{9}
 
-\section{Notation}
+\input{synopsis.tex}
+\input{bibliography.tex}
 
-$\begin{array}{ll}
-\mathbb{N} & \text{Natural numbers} \\
-\mathbb{Z} & \text{Integers}\\
-\mathbb{Q} & \text{Rational numbers}\\
-\mathbb{R} & \text{Real numbers}\\
-\mathbb{C} & \text{Complex numbers}\\
-\mathbb{Z}_p & \text{$p$-adic integers}\\
-\mathbb{Q}_p & \text{$p$-adic numbers}
-\end{array}$
-
-\chapter{First chapter}
-
-\begin{thebibliography}{99999999}
-
-\bibitem[DDMS99]{DDMS99}
-	J. D. Dixon, M. P. F. du Sautoy, A. Mann, D. Segal (1999)
-	\textit{Analytic Pro-$p$ Groups}.
-	Cambridge Studies in Advanced Mathematics,
-	2nd Edition.
-
-\end{thebibliography}
-
-%\printindex
+\printindex
 \end{document}

### 2015-10-01 Master Thesis/synopsis.tex
Don't mention specializations

@@
 Let $v\in\operatorname{Spv}(K)$. There is a canonical bijection:
-\[\{\text{Specializations of }v \} \to \operatorname{Spv}k(v),\ w\mapsto w\mod v\]
+\[\{ w\in\operatorname{Spv}(K)\mid \mathcal{O}_w\subset\mathcal{O}_v \} \to \operatorname{Spv}k(v),\ w\mapsto w\mod v\]
 If $w\in\operatorname{Spv}_\text{rd}^{(i)}(K)$ for some $i\geq 1$ then there exists a unique $v(w)\in\operatorname{Spv}_\text{rd}^{(i-1)}(K)$ such that $\mathcal{O}_w\subset\mathcal{O}_v$.
@@
 \begin{step}
-There exists a unique group isomorphism $\varphi:K'^\times\to K^\times$ such that for any $P\in X_0$ we have $\varphi(U_{P'})=U_P$ where $P'=\sigma(P)$ and the following diagram:
+There exists a unique group isomorphism $\varphi:K'^\times\to K^\times$ such that for any $P\in X_0$ we have $\varphi(U_{P'})=U_P$ where $P'=\sigma(P)$ and the following diagram commutes:
 \[ \begin{tikzcd}

### 2015-10-02 Master Thesis/synopsis.tex
Add 4.3 sketch paragraph

@@
 
+\paragraph{}\label{4.3} Let $n\in\mathbb{N}$. For all intermediate fields $L$ of $\overline{K}/K$, $\sigma$ induces an isomorphism:
+\[ \varphi_n:L'^\times/n \cong\mathrm{H}^1(L', \mu_n) \to \mathrm{H}^1(L,\mu_n) \cong L^\times/n \]
+Obviously, for intermediate fields $L\subset M$ of $\overline{K}/K$ the following diagram commutes:
+\[ \begin{tikzcd}
+L'^\times/n \ar[r, "\varphi_n"]\dar & L^\times/n \dar\\
+M'^\times/n \ar[r, "\varphi_n"'] & M^\times/n
+\end{tikzcd} \]
+Let $P\in X_0$ and $v_P:K\to\mathbb{Z}\cup\{\infty\}$ be the corresponding discrete normalized valuation and let $U_P$ be the group of units of the valuation ring $\mathcal{O}_P$.
+We denote the residue map $U_P\to k(P)^\times,\ x\mapsto x\mod P$ by $r_P$.
+For $n\in\mathbb{N}$ the following sequence is exact:
+\[ 0\to U_P/n\to K^\times/n\to\mathbb{Z}/n\mathbb{Z}\to 0 \]
+
+\begin{lemma}\label{4.3}
+\begin{enumerate}[(i)]
+\item $\varphi_n$ induces an isomorphism $\varphi_n: U_{P'}/n\to U_P/n$.
+\item The following diagram commutes:
+\[ \begin{tikzcd}
+U_{P'}/n \rar["\varphi_n"] \dar["r_{P'}/n"'] & U_P/n \dar["r_P/n"]\\
+k(P')^\times/n \rar["\varphi_n"'] & k(P)^\times/n
+\end{tikzcd} \]
+\end{enumerate}
+\end{lemma}
+
 \begin{definition}\label{4.4}

### 2015-10-12 Master Thesis/synopsis.tex
Fix multiple identical labels
Clarify definition of admissible pair
Fix

@@
 \begin{lemma}\label{3.2}
-$\sigma$ induces a bijection $\sigma:\operatorname{Spv}_\text{d}^{(1)}(K)\to\operatorname{Spv}_\text{d}^{(1)}(K')$.
-If $E=K_v\subset\overline{K}$ is a henselization of $K$ with respect to $v\in\operatorname{Spv}_\text{d}^{(1)}(K)$ then $E'$ is a henselization of $K'$ with respect to $v'=\sigma(v)$.
-Furthermore, if $E^{\text{t}}$ and $E'^{\text{t}}$ denotes the inertial fields of $E$ and $E'$ respectively, then $\sigma$ induces an isomorphism:
-\[ G_{k(v)}\cong\operatorname{Gal}(E^\text{t}/E)\to\operatorname{Gal}(E'^{\text{t}}/E')\cong G_{k(v')} \]
+$\sigma$ induces a bijection $\sigma:\operatorname{Spv}_\text{d}^{(1)}(K) \to\operatorname{Spv}_\text{d}^{(1)}(K')$.
+If $E=K_v\subset\overline{K}$ is a henselization of $K$ with respect to $v\in\operatorname{Spv}_\text{d}^{(1)}(K)$, then $E'$ is a henselization of $K'$ with respect to $v'=\sigma(v)$.
+Furthermore, if $E^{\text{t}}$ and $E'^{\text{t}}$ denote the inertial fields of $E$ and $E'$ respectively, then $\sigma$ induces an isomorphism:
+\[ G_{k(v)}\cong\operatorname{Gal}(E^\text{t}/E) \to\operatorname{Gal}(E'^{\text{t}}/E')\cong G_{k(v')} \]
 which is induced by a unique isomorphism $\varphi_v^\text{s}:k(v')^\text{s}\to k(v)^\text{s}$. In particular $\varphi_v=\varphi_v^\text{s}|_{k(v')}:k(v')\to k(v)$ is an isomorphism.
@@
 
-\paragraph{}\label{4.3} Let $n\in\mathbb{N}$. For all intermediate fields $L$ of $\overline{K}/K$, $\sigma$ induces an isomorphism:
+\paragraph{} Let $n\in\mathbb{N}$. For all intermediate fields $L$ of $\overline{K}/K$, $\sigma$ induces an isomorphism:
 \[ \varphi_n:L'^\times/n \cong\mathrm{H}^1(L', \mu_n) \to \mathrm{H}^1(L,\mu_n) \cong L^\times/n \]
@@
 Let $P\in X_0$ and $v_P:K\to\mathbb{Z}\cup\{\infty\}$ be the corresponding discrete normalized valuation and let $U_P$ be the group of units of the valuation ring $\mathcal{O}_P$.
-We denote the residue map $U_P\to k(P)^\times,\ x\mapsto x\mod P$ by $r_P$.
+We denote the natural projection $U_P\to k(P)^\times,\ x\mapsto x\mod P$ by $r_P$.
 For $n\in\mathbb{N}$ the following sequence is exact:
@@
 \[ K^S=\{x\in K^\times \mid v_P(x)=0\text{ for all }P\in X_0\setminus S \} \]
-$K'^{S'}$ is defined similarly. For finite subsets $S$ of $X_0$, the corresponding subsets of $X_0'$ will be denoted by $S'$, i.e. $S'=\sigma(S)$.
+$K'^{S'}$ is defined similarly. For finite subsets $S$ of $X_0$, the corresponding subsets of $X_0'$ will be denoted by $S'$, i.e. $S'=\{\sigma P\mid P\in S \} $.
 \end{definition}
@@
 
+\begin{definition}
+For each admissible $S\subset X_0$ we can find a finite subset $T\subset X_0\setminus S$ such that both $r_{S,T}$ and $r_{S',T'}$ are injective.
+Such a $T$ will be called $S$-\textit{admissible} and $(S,T)$ is called \textit{admissible pair}.
+\end{definition}
+
 \begin{theorem}[\textit{Pop}]\label{1.1}
@@
 U_{P'} \ar[r, "\varphi"]\ar[d, "r_{P'}"'] & U_P \ar[d, "r_P"]\\
-k(P') \ar[r, "\varphi_P"'] & k(P)
+k(P')^\times \ar[r, "\varphi_P"'] & k(P)^\times
 \end{tikzcd} \]

### 2015-10-13 Master Thesis/synopsis.tex
Add Etale Kohomologie

@@
 
+\begin{proof}
+Since any rational function on $X$ is uniquely determined by its values at those points where it is defined, $\varphi$ is unique.
+Furthermore, the following maps are injective for all finite subsets $S\subset X_0$:
+\[ r_S:K^S \to\prod_{P\in X_0\setminus S} k(P)^\times,\ x\mapsto (r_P(x))_P \]
+For existence, let $(S,T)$ be an admissible pair. First, we show that there is a unique homomorphism $\varphi^{S,T}$, making the following diagram commutative:
+\[ (\star) \qquad \begin{tikzcd}
+K'^{S'} \rar["\varphi^{S,T}", dashed] \dar["r_{S',T'}"'] & K^S \dar["r_{S,T}"]\\
+\displaystyle \prod_{P\in T}k(\sigma P)^\times \rar["\prod\varphi_P"'] & \displaystyle \prod_{P\in T}k(P)^\times
+\end{tikzcd} \phantom{\qquad (\star)} \]
+The vertical maps are injective by definition.
+Thus, it suffices to show that the composition $\varrho:K'^{S'}\to\prod k(\sigma P)^\times\to \prod k(P)^\times\to\operatorname{coker}(r_{S,T})$ is trivial.
+Consider the commutative diagram which results from tensoring $\mathbb{Z}/n\mathbb{Z}$ with $(\star)$:
+\[ \begin{tikzcd}
+K'^{S'}/n \rar[dashed]\dar & K^S/n\dar\\
+\displaystyle \prod_{P\in T}k(\sigma P)^\times/n \rar & \displaystyle \prod_{P\in T}k(P)^\times/n
+\end{tikzcd} \]
+According to Lemma~\ref{4.3} and Lemma~\ref{4.5} this diagram can now be completed by $\varphi_n: K'^{S'}/n\to K^S/n$.
+This implies $\varrho/n=0$, i.e.
+$\operatorname{im}(\varrho)\subset n\operatorname{coker}(r_{S,T})$ for every $n\in\mathbb{N}$.
+By Lemma~\ref{5.2} (iii) and Lemma~\ref{5.4} it follows $\varrho = 0$.
+
+Let $T_1, T_2\subset X_0\setminus S$ be two $S$-admissible subsets and suppose $T_1\subset T_2$. We have:
+\[ r_{S,T_1}\circ\varphi^{S,T_2}=\Big(\prod_{P\in T_1}\varphi_P\Big)\circ r_{S',T_1'} = r_{S,T_1}\circ\varphi^{S,T_1} \]
+Since $r_{S,T_1}$ is injective we get $\varphi^{S,T_2}=\varphi^{S,T_1}$.
+Hence, $\varphi^{S,T}$ is actually independent of $T$, thus we can omit the $T$ from its notation.
+By symmetry, $\varphi^S$ is an isomorphism.
+
+Now, if $T$ ranges over all $S$-admissible subsets of $X_0$, then $(\star)$ shows:
+\[ r_S\circ\varphi^S = \Big(\prod_{P\in X_0\setminus S}\varphi_P \Big)\circ r_{S'} \]
+In other words, we can replace $T$ by $X_0\setminus S$ in $(\star)$. If $S_1\subset S_2$ are two admissible subsets, then:
+\[ r_{S_2}\circ\varphi^{S_2}|_{K'^{S'_1}} = \Big(\prod_{P\in X_0\setminus S_2}\varphi_P \Big)\circ r_{S'_2}|_{K'^{S'_1}} = r_{S_2}\circ\varphi^{S_1} \]
+Hence, $\varphi^{S_1}=\varphi^{S_2}|_{K'^{S'_1}}$.
+Thus, we can combine the isomorphisms $\varphi^S$, where $S$ ranges over the admissible subsets of $X_0$, into an isomorphism $\varphi:K'^\times \to K^\times$.
+$\varphi$ has the desired properties.
+\end{proof}
+
 \begin{step}

### 2015-10-16 Master Thesis/synopsis.tex
Fix references not working

@@
 
-\begin{lemma}\label{2.3}
+\begin{lemma}\phantomsection\label{2.3}
 \begin{enumerate}[(i)]
@@
 
-\begin{lemma}\label{4.3}
+\begin{lemma}\phantomsection\label{4.3}
 \begin{enumerate}[(i)]
@@
 
-\begin{lemma}\label{4.5}
+\begin{lemma}\phantomsection\label{4.5}
 \begin{enumerate}[(i)]
@@
 
-\begin{lemma}\label{5.2}
+\begin{lemma}\phantomsection\label{5.2}
 \begin{enumerate}[(i)]
@@
 Hence, $\varphi^{S_1}=\varphi^{S_2}|_{K'^{S'_1}}$.
-Thus, we can combine the isomorphisms $\varphi^S$, where $S$ ranges over the admissible subsets of $X_0$, into an isomorphism $\varphi:K'^\times \to K^\times$.
+Thus, we can combine the isomorphisms $\varphi^S$, where $S$ ranges over the admissible subsets of $X_0$, into an isomorphism $\varphi = \varinjlim \varphi^S: K'^\times \to K^\times$.
 $\varphi$ has the desired properties.

### 2015-10-16 Master Thesis/synopsis.tex
Update

@@
 \begin{lemma}\label{5.4}
-Let $S\subset X_0$ be admissible. Then there exists a finite subset $T\subset X_0$ with $T\cap S=\varnothing$ such that the following map is injective:
+Let $S\subset X_0$ be admissible. Then there exists a finite subset $T\subset X_0\setminus S$ such that the following map is injective:
 \[ r_{S,T}:K^S \to\prod_{P\in T}k(P)^\times,\ x\mapsto(r_P(x))_{P\in T} \]
@@
 Thus, it suffices to show that the composition $\varrho:K'^{S'}\to\prod k(\sigma P)^\times\to \prod k(P)^\times\to\operatorname{coker}(r_{S,T})$ is trivial.
-Consider the commutative diagram which results from tensoring $\mathbb{Z}/n\mathbb{Z}$ with $(\star)$:
+Consider the commutative diagram which results from tensoring $(\star)$ with $\mathbb{Z}/n\mathbb{Z}$:
 \[ \begin{tikzcd}
@@
 \end{tikzcd} \]
-According to Lemma~\ref{4.3} and Lemma~\ref{4.5} this diagram can now be completed by $\varphi_n: K'^{S'}/n\to K^S/n$.
+According to Lemma~\ref{4.3} and Lemma~\ref{4.5} (ii) this diagram can now be completed by $\varphi_n: K'^{S'}/n\to K^S/n$.
 This implies $\varrho/n=0$, i.e.
@@
 By Lemma~\ref{5.2} (iii) and Lemma~\ref{5.4} it follows $\varrho = 0$.
+By symmetry, $\varphi^S$ is an isomorphism.
 
-Let $T_1, T_2\subset X_0\setminus S$ be two $S$-admissible subsets and suppose $T_1\subset T_2$. We have:
+Let $T_1, T_2\subset X_0\setminus S$ be two $S$-admissible subsets and suppose $T_1\subset T_2$. By observing $(\star)$ we find:
 \[ r_{S,T_1}\circ\varphi^{S,T_2}=\Big(\prod_{P\in T_1}\varphi_P\Big)\circ r_{S',T_1'} = r_{S,T_1}\circ\varphi^{S,T_1} \]
@@
 Hence, $\varphi^{S,T}$ is actually independent of $T$, thus we can omit the $T$ from its notation.
-By symmetry, $\varphi^S$ is an isomorphism.
-
 Now, if $T$ ranges over all $S$-admissible subsets of $X_0$, then $(\star)$ shows:
@@
 Thus, we can combine the isomorphisms $\varphi^S$, where $S$ ranges over the admissible subsets of $X_0$, into an isomorphism $\varphi = \varinjlim \varphi^S: K'^\times \to K^\times$.
+Note that for every $x\in K^\times$ we have $v_P(x)\neq 0$ for finitely many $P\in X_0$; similarly for $K'$ and $X'_0$.
 $\varphi$ has the desired properties.

### 2015-10-17 Master Thesis/synopsis.tex
Clarify

@@
 
-\paragraph{} Let $k$ and $k'$ be the algebraic closures of $\mathbb{Q}$ in $K$ and $K'$ respectively, and let $\overline{k}$ and $\overline{k'}$ be the algebraic closure of $k$ and $k'$ in $\overline{K}$ and $\overline{K'}$ respectively.
+\paragraph{} Let $k$ and $k'$ be the algebraic closures of $\mathbb{Q}$ in $K$ and $K'$ respectively, i.e.
+the constant fields of $K$ and $K'$, and let $\overline{k}$ and $\overline{k'}$ be the algebraic closure of $k$ and $k'$ in $\overline{K}$ and $\overline{K'}$ respectively.
 
-Let $X$ and $X'$ be the unique smooth projective curve over $k$ and $k'$ with function field $K$ and $K'$ respectively.
+Let $X$ and $X'$ be the unique smooth projective curves over $k$ and $k'$ with function field $K$ and $K'$ respectively.
 We identify the set of closed points $X_0$ of $X$ with the set of valuations $v\in\operatorname{Spv}_\text{d}^{(1)}(K)$ with $\operatorname{char} k(v)=0$ and similarly identify $X'_0$ with the corresponding subset of $\operatorname{Spv}_\text{d}^{(1)}(K')$.
@@
 \begin{lemma}\label{5.4}
-Let $S\subset X_0$ be admissible. Then there exists a finite subset $T\subset X_0\setminus S$ such that the following map is injective:
+Let $S\subset X_0$ be admissible and $P_0\in X_0\setminus S$. Then there exists a finite subset $T\subset X_0\setminus S$ with $P_0\in T$ such that the following map is injective:
 \[ r_{S,T}:K^S \to\prod_{P\in T}k(P)^\times,\ x\mapsto(r_P(x))_{P\in T} \]
@@
 
-\begin{definition}
-For each admissible $S\subset X_0$ we can find a finite subset $T\subset X_0\setminus S$ such that both $r_{S,T}$ and $r_{S',T'}$ are injective.
+\begin{remark}
+For each admissible $S\subset X_0$ and $T_1\subset X_0\setminus S$ with injective $r_{S,T_1}$ we also have the injectivity of $r_{S,T_2}$ for every $T_2 \supset T_1$.
+Therefore,  we can find a finite subset $T\subset X_0\setminus S$ such that both $r_{S,T}$ and $r_{S',T'}$ are injective.
 Such a $T$ will be called $S$-\textit{admissible} and $(S,T)$ is called \textit{admissible pair}.
-\end{definition}
+\end{remark}
 

### 2015-10-17 Master Thesis/synopsis.tex
Finish proof of step 2

@@
 
+\begin{proof}
+Observe $\varphi(-1)^2 = \varphi(1) = 1$, and since $\varphi$ is bijective on $K'^\times$, it follows $\varphi(-1) \neq 1$ and hence $\varphi(-1) = -1$.
+Now, to prove the assertion, it suffices to show $\varphi(x + y) = \varphi(x) + \varphi(y)$ for all $x,y\in K'^\times$.
+\begin{itemize}
+\item \textit{Case 1:} If $x + y = 0$, we have:
+\[ \varphi(x + y) = 0 = \varphi(x) + \varphi(-1)\varphi(x) = \varphi(x) + \varphi(-x) = \varphi(x) + \varphi(y) \]
+\item \textit{Case 2:} Let $x + y\neq 0$. Looking at the commutative diagram in Step 1 we get
+\begin{align*}
+r_P(\varphi(x+ y)) &= \varphi_P(r_{P'}(x + y)) \\
+&= \varphi_P(r_{P'}(x)) + \varphi_P(r_{P'}(y)) = r_P(\varphi(x) + \varphi(y))
+\end{align*}
+for almost all $P\in X_0$.
+Note that $\varphi_P: k(P')\to k(P)$ is a field isomorphism and that the residue maps $r_{P}: \mathcal{O}_{P} \to k(P)$ are ring homomorphisms.
+Therefore, $\varphi(x+y) = \varphi(x) + \varphi(y)$.
+\end{itemize}
+\end{proof}
+
 \begin{step}

### 2015-10-18 Master Thesis/synopsis.tex
Fix qed
Clarify
Add end remark of section 3

@@
 \end{lemma}
+
+\begin{remark}
+If $L\subset M$ are intermediate fields of $\overline{K}/K$, which are finite over $K$, then $\sigma$ also induces bijections
+\[ \operatorname{Spv}_\text{d}^{(1)}(L) \to \operatorname{Spv}_\text{d}^{(1)}(L'),\qquad \operatorname{Spv}_\text{d}^{(1)}(M) \to \operatorname{Spv}_\text{d}^{(1)}(M') \]
+such that the following diagram commutes:
+\[\begin{tikzcd}
+\operatorname{Spv}_\text{d}^{(1)}(M) \rar["\sigma"]\dar & \operatorname{Spv}_\text{d}^{(1)}(M')\dar\\
+\operatorname{Spv}_\text{d}^{(1)}(L) \rar["\sigma"'] & \operatorname{Spv}_\text{d}^{(1)}(L')
+\end{tikzcd}\]
+where the vertical maps are restrictions. Let $w\in\operatorname{Spv}_\text{d}^{(1)}(M),\ v=w|_L$ and $w'=\sigma(w),\ v'=\sigma(v)$. Then the following diagram commutes:
+\[ \begin{tikzcd}
+k(w') \rar["\varphi_w"] & k(w)\\
+k(v') \rar["\varphi_v"']\uar[hook] & k(v)\uar[hook]
+\end{tikzcd} \]
+By a limit process we see that $\sigma$ induces a bijection $\sigma: \operatorname{Spv}_\text{d}^{(1)}(\overline{K})\to \operatorname{Spv}_\text{d}^{(1)}(\overline{K'})$.
+Let $\overline{v}\in \operatorname{Spv}_\text{d}^{(1)}(\overline{K})$ and $G_{\overline{v}}\subset G_K$ be the decomposition group of $\overline{v}$.
+Then we have $\sigma(G_{\overline{v}}) = G_{\sigma(\overline{v})}$.
+Now it is easy to see that for every $g\in G_K$:
+\[ \sigma(\overline{v}\circ g) = \sigma(\overline{v})\circ \sigma(g) \]
+\end{remark}
 
@@
 \[ \varphi(x + y) = 0 = \varphi(x) + \varphi(-1)\varphi(x) = \varphi(x) + \varphi(-x) = \varphi(x) + \varphi(y) \]
-\item \textit{Case 2:} Let $x + y\neq 0$. Looking at the commutative diagram in Step 1 we get
+\item \textit{Case 2:} Let $x + y\neq 0$ and $S\subset X_0$ be a finite subset such that $\varphi(x + y)$, $\varphi(x)$ and $\varphi(y)$ are contained in $K^S$.
+Looking at the commutative diagram in \mbox{Step 1} we get
 \begin{align*}
@@
 \end{align*}
-for almost all $P\in X_0$.
+for all $P\in X_0\setminus S$.
 Note that $\varphi_P: k(P')\to k(P)$ is a field isomorphism and that the residue maps $r_{P}: \mathcal{O}_{P} \to k(P)$ are ring homomorphisms.
-Therefore, $\varphi(x+y) = \varphi(x) + \varphi(y)$.
+Since $K^S\to \prod_{P\in X_0\setminus S} k(P)^\ast$ is injective, we get $\varphi(x+y) = \varphi(x) + \varphi(y)$.\qedhere
 \end{itemize}

### 2015-10-20 Master Thesis/synopsis.tex
Use group operation notation
Finalize chapter 1

@@
 Now it is easy to see that for every $g\in G_K$:
-\[ \sigma(\overline{v}\circ g) = \sigma(\overline{v})\circ \sigma(g) \]
+\[ \sigma(g\overline{v}) = \sigma(g)\sigma(\overline{v}) \]
 \end{remark}
@@
 Thus, we can combine the isomorphisms $\varphi^S$, where $S$ ranges over the admissible subsets of $X_0$, into an isomorphism $\varphi = \varinjlim \varphi^S: K'^\times \to K^\times$.
-Note that for every $x\in K^\times$ we have $v_P(x)\neq 0$ for finitely many $P\in X_0$; similarly for $K'$ and $X'_0$.
-$\varphi$ has the desired properties.
+Note that for every $x\in K^\times$ we have $v_P(x)= 0$ for almost all $P\in X_0$; similarly for $K'$ and $X'_0$.
+
+Now note that for any $x\in U_{P'}$ we can find an admissible set $S\subset X_0\setminus \{P\}$ such that $x\in K'^{S'}$.
 \end{proof}
@@
 \begin{step}
-$\varphi$ induces $\sigma:G_K\to G_{K'}$, i.e. $\sigma(g) = \varphi^{-1}g\varphi$ for all $g\in G_K$, and is uniquely determined by this property.
+$\varphi$ induces $\sigma:G_K\to G_{K'}$, i.e. $\sigma(g) = \varphi^{-1}\circ g\circ \varphi$ for all $g\in G_K$, and is uniquely determined by this property.
 \end{step}
+
+\begin{proof}
+Let $\tau:G_K\to G_{K'},\ g\mapsto \varphi^{-1}\circ g\circ \varphi$.
+Clearly, $\tau$ is a group isomorphism and we have $\tau(G_L)=G_{L'}=\sigma(G_L)$ for every finite subextension $L/K$ of $\overline{K}/K$.
+By Lemma \ref{3.2}, $\tau$ induces bijections $\tau: \operatorname{Spv}_\text{d}^{(1)}(L)\to \operatorname{Spv}_\text{d}^{(1)}(L')$ for any intermediate field $L$ of $\overline{K}/K$.
+\end{proof}

### 2015-11-21 Seminar Schnitttheorie/article.tex
Add template

\documentclass[10pt,a4paper]{article}
\usepackage{fontspec}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\author{Yichuan Shen}
\title{Vektorbündel \& Gysin-Homomorphismus}
\begin{document}
\maketitle

Hello World!
\end{document}

### 2015-11-22 Seminar Schnitttheorie/article.tex
Aussagenskelett

@@
 \usepackage{amssymb}
-\author{Yichuan Shen}
+\usepackage{amsthm}
+\usepackage{paralist}
+\linespread{1.1}
+
+\author{von \textsc{yichuan shen}}
 \title{Vektorbündel \& Gysin-Homomorphismus}
 \begin{document}
+
+\theoremstyle{plain}
+\theoremstyle{definition}
+\newtheorem{theorem}{Theorem}
+\newtheorem{lemma}[theorem]{Lemma}
+\newtheorem{proposition}[theorem]{Satz}
+\newtheorem{corollary}[theorem]{Korollar}
+\theoremstyle{definition}
+\newtheorem*{definition}{Definition}
+\newtheorem*{example}{Beispiel}
+\theoremstyle{remark}
+\newtheorem*{remark}{Bemerkung}
+
 \maketitle
 
-Hello World!
+\section{Definitionen}
+
+\begin{definition}
+\begin{enumerate}[(i)]
+\item Sei $p: E\to X$ ein Vektorbündel vom Rang $n$ und $X=\bigcup_\alpha U_\alpha$ eine offene Überdeckung mit $U_\alpha$-Isomorphismen $\varphi_j: f^{-1}(U_\alpha)\to \mathbf{A}_{U_\alpha}^n$.
+Wir bezeichnen den Nullschnitt mit $i_0: X\to E$.
+\item Für eine lokal freie Garbe $\mathcal{E}$ vom Rang $r$ auf $X$ definiert
+\[ \mathbf{V}(\mathcal{E}) = \mathbf{Spec}_X(\operatorname{Sym}^\bullet\mathcal{E})\to X \]
+ein Vektorbündel auf $X$.
+\item Für einen Vektorbündel $E\to X$ vom Rang $r$ definiert
+\[ \Gamma(E) = (U\mapsto \operatorname{Hom}_X(U, E)) \]
+eine lokal freie Garbe auf $X$ vom Rang $r$.
+\end{enumerate}
+\end{definition}
+
+\begin{definition}
+\begin{enumerate}[(i)]
+\item Das \textit{projektive Vektorbündel} $p: \mathbf{P}(\mathcal{E})\to X$ einer lokal freien Garbe $\mathcal{E}$ über $X$ ist gegeben durch:
+\[ \mathbf{P}(\mathcal{E}) = \mathbf{Proj}_X(\operatorname{Sym}^\bullet\mathcal{E}) \]
+Es gibt einen natürlichen surjektiven Morphismus $p^\ast\mathcal{E}\to \mathcal{O}_\mathcal{E}(1)$.
+\item Das \textit{projektive Vektorbündel} $p: P(E)\to X$ eines Vektorbündels $E\to X$ über $X$ ist gegeben durch:
+\[ P(E) = (E\setminus i_p(X))/\mathbb{G}_\text{m} \]
+\end{enumerate}
+\end{definition}
+
+\begin{definition}
+Die \textit{Segre-Klassen} $s_i(E)$ für $i\geq 1-r$ eines Vektorbündels $E\to X$ vom Rang $r$ auf einer Varietät $X$ sind die Homomorphismen:
+\begin{align*}
+s_i(E)\cap -: A_d(X)&\to A_{d-i}(X),\\ 
+\alpha&\mapsto p_\ast(c_1(\mathcal{O}_\mathcal{E}(1))^{(r-1)+i} \cap p^\ast\alpha)
+\end{align*}
+wobei $r-1$ die relative Dimension des projektiven Vektorbündels $p: P(E)\to X$ bezeichnet.
+\end{definition}
+
+\begin{proposition}
+Sei $\mathcal{E} = \Gamma(E)$ ein Vektorbündel auf einer Varietät $X$ und $\alpha\in A_\bullet(X)$.
+\begin{enumerate}[(i)]
+\item $s_i(\mathcal{E})\cap\alpha = 0$ für alle $1-r\leq i <0$.
+\item $s_0(\mathcal{E})\cap\alpha = \alpha$
+\item Sei $\mathcal{F} = \Gamma(F)$ ein weiteres Vektorbündel auf $X$. Dann gilt für alle $i,j\geq 0$:
+\[ s_i(\mathcal{E})\cap (s_j(\mathcal{F}) \cap \alpha) = s_j(\mathcal{F}) \cap (s_i(\mathcal{E})\cap \alpha) \]
+\item Sei $f:Y\to X$ ein eigentlicher Morphismus und $\beta\in A_\bullet(Y)$. Dann gilt die Projektionsformel:
+\[ f_\ast(s_i(f^\ast\mathcal{E})\cap\beta) = s_i(\mathcal{E})\cap f_\ast\beta \]
+\item Sei $f:Y\to X$ ein flacher Morphismus. Dann gilt:
+\[ f^\ast(s_i(\mathcal{E})\cap \alpha) = s_i(f^\ast\mathcal{E})\cap f^\ast\alpha \]
+\item Ist $\mathcal{E}$ vom Rang $1$, also ein Geradenbündel. Dann gilt:
+\[ s_1(\mathcal{E}) \cap\alpha = -c_1(\mathcal{E})\cap \alpha \]
+\end{enumerate}
+\end{proposition}
+
+\begin{corollary}
+Sei $p:E\to X$ ein Vektorbündel vom Rang $r$. Dann ist
+\[ p^\ast: A_q(X)\to A_{q+r-1}(P(E)) \]
+injektiv und besitzt den Schnitt:
+\[ \alpha\mapsto p_\ast(c_1(\mathcal{O}_E(1))^{r-1}\cap\alpha) \]
+\end{corollary}
+
+\begin{corollary}
+Sei $\mathcal{E}$ eine lokal freie Garbe vom Rang $r$ auf einer Varietät $X$. Dann gibt es einen flachen projektiven Morphismus $f:Y\to X$, so dass:
+\begin{enumerate}[(i)]
+\item Die induzierte Abbildung $f^\ast:A_\bullet(X)\to A_\bullet(Y)$ ist injektiv.
+\item Die Garbe $f^\ast\mathcal{E}$ besitzt eine Filtration von lokal freien Garben:
+\[ 0 = \mathcal{E}_0 \subset \mathcal{E}_1 \subset\ldots\subset \mathcal{E}_r = f^\ast\mathcal{E} \]
+wobei $\mathcal{E}_i$ vom Rang $i$ ist.
+\end{enumerate}
+\end{corollary}
+
+\begin{proposition}
+Sei $\mathcal{E}$ eine lokal freie Garbe vom Rang $r$ und $\mathcal{L}$ ein Geradenbündel auf $X$. Dann gilt:
+\[ s_i(\mathcal{E}\otimes\mathcal{L}) = \sum_{j=0}^i(-1)^{i-j} \binom{r-1+i}{r-1+j}s_j(\mathcal{E}) c_1(\mathcal{L})^{i-j} \]
+\end{proposition}
+
+\begin{definition}
+Das \textit{Segre-Polynom} einer lokal freien Garbe $\mathcal{E}$ einer Varietät $X$ ist definiert durch:
+\[ s_t(\mathcal{E}) = \sum_{i = 0}^\infty s_i(\mathcal{E})t^i \in\operatorname{End}(A_\bullet(X))[t] \]
+Es hat höchstens Grad $\dim(X)$. Die \textit{totale Segre-Klasse} ist:
+\[ s(\mathcal{E}) = s_t(\mathcal{E})|_{t = 1} \in\operatorname{End}(A_\bullet(X)) \]
+\end{definition}
+
+\begin{remark}
+Die Segre-Klassen $s_i(\mathcal{E})$ sind für $i>0$  nilpotent und liegen in einem kommutativen Teilring von $\operatorname{End}(A_\bullet(X))$. Da $s_0(\mathcal{E}) = \operatorname{id}$, folgt:
+\[ s_t(\mathcal{E}) \in\operatorname{End}(A_\bullet(X))[t]^\times \]
+\end{remark}
+
+\begin{definition}
+Sei $E\to X$ ein Vektorbündel auf $X$ und $\mathcal{E} = \Gamma(E)$. Das \textit{Chern-Polynom} von $E$ bzw. $\mathcal{E}$ ist definiert durch:
+\[ c_t(\mathcal{E}) = s_t(\mathcal{E})^{-1} \in\operatorname{End}(A_\bullet(X))[t]^\times \]
+Die Koeffizienten sind die \textit{Chern-Klassen}:   
+\[ c_t(\mathcal{E}) = \sum_{i =0}^\infty c_i(\mathcal{E}) t^i \]
+wobei $c_i(\mathcal{E})\cap - : A_q(X)\to A_{q-i}(X)$. Die \textit{totale Chern-Klasse} ist:
+\[ c(\mathcal{E}) = c_t(\mathcal{E})|_{t =1}\in\operatorname{End}(A_\bullet(X)) \]
+\end{definition}
+
+\begin{remark}
+\begin{enumerate}[(i)]
+\item Die zwei Definitionen von $c_1$ stimmen überein.
+\item Die Chern-Klassen liegen in denselben kommutativen Teilring wie die Segre-Klassen, kommutieren also mit Segre-Klassen.
+\item Explizite Formeln für Chern-Klassen sind gegeben durch:
+\begin{align*}
+c_0 &= 1\\
+c_1 &= -s_1\\
+c_2 &= s_1^2 - s_2\\
+&\vdots\\
+c_n &= -s_1c_{n-1}-s_2c_{n-2}-\ldots -s_{n-1}c_1-s_n\\
+&\vdots
+\end{align*}
+\end{enumerate}
+\end{remark}
+
+\begin{proposition}
+Sei $\mathcal{E} = \Gamma(E)$ ein Vektorbündel auf einer Varietät $X$ vom Rang $r$ und sei $\alpha\in A_\bullet(X)$.
+\begin{enumerate}[(i)]
+\item $c_i(\mathcal{E})\cap \alpha = 0$ für $i > r$.
+\item $c_0(\mathcal{E})\cap \alpha = \alpha$
+\item Sei $\mathcal{F} = \Gamma(F)$ ein weiteres Vektorbündel auf $X$. Dann gilt für alle $i,j\geq 0$:
+\[ c_i(\mathcal{E})\cap (c_j(\mathcal{F}) \cap \alpha) = c_j(\mathcal{F}) \cap (c_i(\mathcal{E})\cap \alpha) \]
+\item Sei $f:Y\to X$ ein eigentlicher Morphismus und $\beta\in A_\bullet(Y)$. Dann gilt die Projektionsformel:
+\[ f_\ast(c_i(f^\ast\mathcal{E})\cap\beta) = c_i(\mathcal{E})\cap f_\ast\beta \]
+\item Sei $f:Y\to X$ ein flacher Morphismus. Dann gilt:
+\[ f^\ast(c_i(\mathcal{E})\cap \alpha) = c_i(f^\ast\mathcal{E})\cap f^\ast\alpha \]
+\item \textit{(Whitneysche Summenformel)} Für eine kurze exakte Folge 
+\[ 0\longrightarrow\mathcal{E}'\longrightarrow\mathcal{E}\longrightarrow \mathcal{E}''\longrightarrow 0 \]
+von lokal freien Garben auf $X$ gilt:
+\[ c_n(\mathcal{E}) = \sum_{i+j = n}c_i(\mathcal{E}')c_j(\mathcal{E}'') \]
+\item Für einen Cartier-Divisor $D$ auf $X$ gilt:
+\[ c_1(\mathcal{O}_X(D))\cap [X] = [D] \]
+\end{enumerate}
+\end{proposition}
+
+\begin{remark}
+Chern-Klassen sind eindeutig durch die Eigenschaften (v), (vi) und (vii) bestimmt.
+\end{remark}
+
+\begin{proposition}
+Sei $X$ eine Varietät und $\mathcal{E}$ eine lokal freie Garbe vom Rang $r$ auf $X$.
+Sei $s\in\Gamma(X,\mathcal{E})$ und $Z=Z(s)\hookrightarrow X$ das geschlossene Unterschema wo $s$ verschwindet und $j:U=X\setminus Z\hookrightarrow X$ das offene Komplement.
+Dann gilt:
+\begin{enumerate}[(i)]
+\item Für alle $\alpha\in A_\bullet(X)$ gilt $c_r(\mathcal{E})\cap\alpha = \beta$, wobei $\beta$ einen Vertreter besitzt, dessen Träger in $Z$ liegt.
+\item $c_r(j^\ast\mathcal{E}) = 0$
+\item Sei $Z=\varnothing$ und $\mathcal{E}$ besitzt eine vollständige Filtrierung mit Geradenbündeln $\mathcal{L}_1,\ldots,\mathcal{L}_r$ als Quotienten. Dann gilt für alle $\alpha\in A_\bullet(X)$:
+\[ j^\ast\Big(\prod_{i=1}^r c_1(\mathcal{L}_i)\cap\alpha \Big) = 0 \]
+\end{enumerate}
+\end{proposition}
+
+\begin{theorem}
+Sei $p:E\to X$ ein Vektorbündel vom Rang $r$ und $q:P(E)\to X$ das assoziierte projektive Vektorbündel.
+\begin{enumerate}[(i)]
+\item Der flache Rückzug $p^\ast: A_d(X)\to A_{d+r}(E)$ ist ein Isomorphismus.
+\item Es gibt einen kanonischen Isomorphismus:
+\begin{align*}
+\vartheta_E: \bigoplus_{i=0}^{r-1}A_{d+i}(X) &\stackrel{\sim}{\longrightarrow} A_{d+r-1}(P(E))\\
+A_{d+i}(X)\ni \alpha_i &\longmapsto c_1(\mathcal{O}_E(1))^i\cap q^\ast(\alpha_i)
+\end{align*}
+\item Sei $\mathcal{E} = \Gamma(E)$. Die Chern-Klassen von $\mathcal{E}$ sind eindeutig durch die folgende Formeln bestimmt:
+\[ c_0(\mathcal{E}) = 1,\quad \sum_{i=0}^r c_1(\mathcal{O}_E(1))^i\cap q^\ast(c_{r-i}(\mathcal{E})\cap\alpha) = 0 \]
+\end{enumerate}
+\end{theorem}
+
+\begin{remark}
+Chern-Klassen sind eindeutig durch die Eigenschaft (iii) bestimmt.
+\end{remark}
+
+\begin{definition}
+Für ein Vektorbündel $p:E\to X$ vom Rang $r$ mit Nullschnitt $s:X\to E$ definieren wir den \textit{Gysin-Rückzug} entlang des Nullschnitts als:
+\[ s^\ast = (p^\ast)^{-1}: A_{d+r}(E)\to A_d(X) \]
+\end{definition}
+
 \end{document}

### 2015-11-22 Seminar Schnitttheorie/article.tex
Remove superfluous remark
Minor changes

@@
 \item Die induzierte Abbildung $f^\ast:A_\bullet(X)\to A_\bullet(Y)$ ist injektiv.
-\item Die Garbe $f^\ast\mathcal{E}$ besitzt eine Filtration von lokal freien Garben:
+\item Die Garbe $f^\ast\mathcal{E}$ besitzt eine \textit{vollständige Filtration}, d.h. eine Filtration von lokal freien Garben
 \[ 0 = \mathcal{E}_0 \subset \mathcal{E}_1 \subset\ldots\subset \mathcal{E}_r = f^\ast\mathcal{E} \]
@@
 
-\begin{remark}
-Chern-Klassen sind eindeutig durch die Eigenschaft (iii) bestimmt.
-\end{remark}
-
 \begin{definition}
-Für ein Vektorbündel $p:E\to X$ vom Rang $r$ mit Nullschnitt $s:X\to E$ definieren wir den \textit{Gysin-Rückzug} entlang des Nullschnitts als:
+Für ein Vektorbündel $p:E\to X$ vom Rang $r$ mit Nullschnitt $s:X\to E$ definieren wir den \textit{Gysin-Homomorphismus} entlang $s$ als:
 \[ s^\ast = (p^\ast)^{-1}: A_{d+r}(E)\to A_d(X) \]

### 2015-11-28 Seminar Schnitttheorie/article.tex
Remove superfluous title
Fix typos
Add some proofs

@@
 \usepackage{paralist}
+\usepackage{tikz-cd}
 \linespread{1.1}
@@
 
-\section{Definitionen}
-
 \begin{definition}
+Sei $X$ ein Schema.
 \begin{enumerate}[(i)]
@@
 
+\begin{proposition}
+Sei $X$ ein Schema. Für eine lokal freie Garbe $\mathcal{E}$ auf $X$ und einen Vektorbündel $E\to X$ gilt:
+\[ \Gamma(\mathbf{V}(\mathcal{E})) = \mathcal{E}^\vee \]
+\[ \mathbf{V}(\Gamma(E)) = E^\vee \]
+\end{proposition}
+
 \begin{definition}
+Sei $X$ ein Schema.
 \begin{enumerate}[(i)]
 \item Das \textit{projektive Vektorbündel} $p: \mathbf{P}(\mathcal{E})\to X$ einer lokal freien Garbe $\mathcal{E}$ über $X$ ist gegeben durch:
-\[ \mathbf{P}(\mathcal{E}) = \mathbf{Proj}_X(\operatorname{Sym}^\bullet\mathcal{E}) \]
-Es gibt einen natürlichen surjektiven Morphismus $p^\ast\mathcal{E}\to \mathcal{O}_\mathcal{E}(1)$.
+\[ \mathbf{P}(\mathcal{E}) = \mathbf{Proj}_X(\operatorname{Sym}^\bullet\mathcal{E})\to X \]
+Es gibt einen natürlichen surjektiven Morphismus $p^\ast\mathbb{E} \to \mathcal{O}_\mathbb{E}(1)$.
 \item Das \textit{projektive Vektorbündel} $p: P(E)\to X$ eines Vektorbündels $E\to X$ über $X$ ist gegeben durch:
-\[ P(E) = (E\setminus i_p(X))/\mathbb{G}_\text{m} \]
+\[ P(E) = (E\setminus i_0(X))/\mathbb{G}_\text{m}\to X \]
 \end{enumerate}
@@
 \begin{definition}
-Die \textit{Segre-Klassen} $s_i(E)$ für $i\geq 1-r$ eines Vektorbündels $E\to X$ vom Rang $r$ auf einer Varietät $X$ sind die Homomorphismen:
+Die \textit{Segre-Klassen} $s_i(E)$ für $i\geq 1-r$ eines Vektorbündels $E\to X$ vom Rang $r = \operatorname{rk}(E)$ auf einer Varietät $X$ sind die Homomorphismen:
 \begin{align*}
 s_i(E)\cap -: A_d(X)&\to A_{d-i}(X),\\ 
-\alpha&\mapsto p_\ast(c_1(\mathcal{O}_\mathcal{E}(1))^{(r-1)+i} \cap p^\ast\alpha)
+\alpha&\mapsto p_\ast(c_1(\mathcal{O}_E(1))^{(r-1)+i} \cap p^\ast\alpha)
 \end{align*}
@@
 
+\begin{proof}
+Projektive Vektorbündel sind stabil unter Basiswechsel. Wir haben das kartesische Quadrat:
+\[ \begin{tikzcd}
+P(f^\ast E) \rar["P(f)"]\dar["q"'] & P(E) \dar["p"] \\
+Y \rar["f"'] & X
+\end{tikzcd} \]
+und $P(f)^\ast\mathcal{O}_E(1) = \mathcal{O}_{f^\ast E}(1)$. Es gilt:
+\begin{align*}
+f_\ast(s_i(f^\ast\mathcal{E})\cap\beta) &= f_\ast q_\ast(c_1(\mathcal{O}_{f^\ast E}(1))^{r-1+i}\cap q^\ast \beta)\\
+&= p_\ast P(f)_\ast (c_1(P(f)^\ast\mathcal{O}_E(1))^{r-1+i}\cap q^\ast\beta)\\
+&= p_\ast(c_1(\mathcal{O}_E(1))^{r-1+i}\cap P(f)_\ast q^\ast\beta) \\
+&\phantom{==} \text{(Projektionsformel für $c_1$)}\\
+&= p_\ast(c_1(\mathcal{O}_E(1))^{r-1+i}\cap p^\ast f_\ast\beta)\\
+&= s_i(\mathcal{E})\cap f_\ast\beta
+\end{align*}
+\begin{align*}
+f^\ast(s_i(\mathcal{E})\cap\alpha) &= f^\ast p_\ast(c_1(\mathcal{O}_E(1))^{r-1+i}\cap p^\ast\alpha)\\
+&= q_\ast P(f)^\ast(c_1(\mathcal{O}_E(1))^{r-1+i}\cap p^\ast \alpha)\\
+&= q_\ast(c_1(P(f)^\ast\mathcal{O}_{E}(1))^{r-1+i}\cap P(f)^\ast p^\ast \alpha)\\
+&= q_\ast(c_1(\mathcal{O}_{f^\ast E}(1))^{r-1+i}\cap q^\ast f^\ast \alpha)\\
+&= s_i(f^\ast\mathcal{E})\cap f^\ast\alpha
+\end{align*}
+Dies zeigt (iv) und (v). Für (i) und (ii) sei $\alpha = [V]$ ein Primzykel und nach der Projektionsformel für $V\hookrightarrow X$ können wir o.B.d.A. annehmen, dass $V=X$ ganz ist. Dann gilt:
+\[s_i(\mathcal{E})\cap [X] \in \begin{cases}
+A_{\dim(X)-i}(X) = 0, & \text{wenn $i<0$}\\
+A_{\dim(X)}(X) = \mathbb{Z}[X], & \text{wenn $i = 0$}
+\end{cases}
+\]
+Um $s_1(\mathcal{E})\cap [X] = [X]$ zu zeigen, können wir nach (v) annehmen, dass $P(E)=X\times\mathbf{P}^r$ das triviale Bündel ist.
+Wieder nach (v) für $X\to k$ können wir sogar annehmen, dass $X=\operatorname{Spec}(k)$ und $P(E)=\mathbf{P}_k^{r-1}$.
+Dann gilt:
+\[ \deg(c_1(\mathcal{O}(1))^{r-1}\cap [\mathbf{P}_k^{r-1}]) = \deg([\mathrm{Pt}]) = 1 \]
+Für (iii) betrachte das kartesische Quadrat:
+\[ \begin{tikzcd}
+Y \rar["q'"]\dar["p'"']\ar[rd, "f", dashed] & P(E) \dar["p"]\\
+P(F)\rar["q"'] & X
+\end{tikzcd} \]
+Dann gilt:
+\begin{align*}
+s_i(\mathcal{E})\cap (s_j(\mathcal{F})\cap \alpha) &= p_\ast (c_1(\mathcal{O}_E(1))^{r-1+i} \cap p^\ast(q_\ast(c_1(\mathcal{O}_F(1))^{s-1+j}\cap q^\ast \alpha))\\
+&= p_\ast(c_1(\mathcal{O}_E(1))^{r-1+i}\cap q'_\ast(c_1(p'^\ast\mathcal{O}_F (1))^{s-1+j}\cap p'^\ast q^\ast\alpha))\\
+&= f_\ast(c_1(q'^\ast\mathcal{O}_E(1))^{r-1+i}\cap c_1(p'^\ast\mathcal{O}_F (1))^{s-1+j}\cap f^\ast\alpha)\\
+&\phantom{==} \text{(Projektionsformel)}
+\end{align*}
+Die Aussage folgt, da $c_1$ kommutativ ist.
+Für (vi) sei $E\to X$ ein Geradenbündel und $\mathcal{E}$ eine invertierbare Garbe mit $E=\mathbf{V}(\mathcal{E})$ und $P(E)=\mathbf{P}(E)=X$ und $\mathcal{O}_E(1)=\mathcal{E} = \Gamma(E)^\vee$.
+Dann gilt:
+\[ s_1(E)\cap \alpha = c_1(\mathcal{O}_E(1))\cap\alpha = -c_1(\Gamma(E))\cap\alpha = -c_1(E)\cap\alpha\qedhere \]
+\end{proof}
+
 \begin{corollary}
@@
 \end{corollary}
+
+\begin{proof}
+Die Komposition des Schnitts mit $p^\ast$ ist gerade $s_0(E)\cap - = \operatorname{id}$.
+\end{proof}
 
@@
 \end{corollary}
+
+\begin{proof}
+Betrachte das projektive Vektorbündel $f: \mathbf{P}(\mathcal{E})\to X$ und die natürliche surjektive Abbildung $f^\ast\mathcal{E}\to\mathcal{O}_\mathcal{E}(1)$.
+Alles folgt aus dem vorherigen Korollar.
+\end{proof}
 
@@
 
+\begin{proof}
+(ii) und (vii) sind klar. (iii), (iv) und (v) folgen aus den Aussagen für die Segre-Klassen, da Chern-Klassen Polynome in den Segre-Klassen sind.
+\end{proof}
+
 \begin{proposition}

### 2015-11-28 Seminar Schnitttheorie/article.tex
Add more proofs

@@
 \[ \mathbf{P}(\mathcal{E}) = \mathbf{Proj}_X(\operatorname{Sym}^\bullet\mathcal{E})\to X \]
-Es gibt einen natürlichen surjektiven Morphismus $p^\ast\mathbb{E} \to \mathcal{O}_\mathbb{E}(1)$.
+Es gibt einen natürlichen surjektiven Morphismus $p^\ast\mathcal{E} \to \mathcal{O}_\mathcal{E}(1)$.
 \item Das \textit{projektive Vektorbündel} $p: P(E)\to X$ eines Vektorbündels $E\to X$ über $X$ ist gegeben durch:
@@
 
-\begin{corollary}
+\begin{corollary}[\textit{Splitting-Prinzip}]
 Sei $\mathcal{E}$ eine lokal freie Garbe vom Rang $r$ auf einer Varietät $X$. Dann gibt es einen flachen projektiven Morphismus $f:Y\to X$, so dass:
@@
 \[ 0 = \mathcal{E}_0 \subset \mathcal{E}_1 \subset\ldots\subset \mathcal{E}_r = f^\ast\mathcal{E} \]
-wobei $\mathcal{E}_i$ vom Rang $i$ ist.
+so dass die Quotienten $\mathcal{E}_i/\mathcal{E}_{i-1}$ vom Rang $1$ sind.
 \end{enumerate}
@@
 Betrachte das projektive Vektorbündel $f: \mathbf{P}(\mathcal{E})\to X$ und die natürliche surjektive Abbildung $f^\ast\mathcal{E}\to\mathcal{O}_\mathcal{E}(1)$.
-Alles folgt aus dem vorherigen Korollar.
+Nach dem vorherigen Korollar ist $f^\ast$ injektiv.
 \end{proof}
@@
 (ii) und (vii) sind klar. (iii), (iv) und (v) folgen aus den Aussagen für die Segre-Klassen, da Chern-Klassen Polynome in den Segre-Klassen sind.
+
+Für (i) sei zunächst $E\to X$ ein Geradenbündel und $\mathcal{E} = \Gamma(E)$. Dann ist $P(E)=X$ und $\mathcal{O}_E(1)=\mathcal{E}^\vee$ und:
+\[ s_i(\mathcal{E})\cap\alpha = c_1(\mathcal{O}_E(1))^i\cap\alpha = (-1)^ic_1(\mathcal{E})^i\cap\alpha \]
+Somit gilt für die totale Segre-Klasse:
+\[ s(\mathcal{E}) = \sum_{i\geq 0}(-1)^ic_1(\mathcal{E})^i = \frac{1}{1+c_1(\mathcal{E})}\in\operatorname{End}(A_\bullet(X)) \]
+Dies zeigt $c_i(\mathcal{E}) = 0$ für $i> \operatorname{rk}(\mathcal{E}) = 1$.
+Für den allgemeinen Fall betrachte das Splitting-Prinzip und die Whitneysche Summenformel.
+Die Whitneysche Summenformel ist äquivalent zu $c(\mathcal{E}')c(\mathcal{E}'') = c(\mathcal{E})\in\operatorname{End}(A_\bullet(X))$.
+
+Für die Summenformel können wir nach dem Splitting-Prinzip und (v) annehmen, dass $\mathcal{E}'$ und $\mathcal{E}''$ vollständige Filtrationen besitzen.
+Diese induzieren eine vollständige Filtration auf $\mathcal{E}$ mit Geradenbündeln $\mathcal{L}_1,\ldots,\mathcal{L}_r$ als Quotienten und zu zeigen ist:
+\[ c(\mathcal{E}) = \prod_{i=1}^r c(\mathcal{L}_i) = \prod_{i=1}^r(1+c_1(\mathcal{L}_i)) \]
+Ist $\sigma_i\in\operatorname{End}(A_\bullet(X))$ das $i$-te elementar-symmetrische Polynom in $c_1(\mathcal{L}_1),\ldots,$ $c_1(\mathcal{L}_r)$.
+Dann ist die obige Gleichung äquivalent zu $c_i(\mathcal{E}) = \sigma_i$.
+
+Sei $p:P(E)\to X$ das zugehörige projektive Vektorbündel und $\tilde{\sigma}_i$ das $i$-te elementar-symmetrische Polynom in $c_1(p^\ast\mathcal{L}_1),\ldots,c_1(p^\ast\mathcal{L}_r)$.
+Setze $\zeta = c_1(\mathcal{O}_E(1))$.
+Die natürliche Surjektion $p^\ast\mathcal{E}^\vee\to \mathcal{O}_E(1)$ liefert einen Schnitt $s\in\mathrm{H}^0(P(E), p^\ast\mathcal{E}\otimes \mathcal{O}_E(1))$, das nirgendwo verschwindet.
+Die Garbe $p^\ast\mathcal{E}\otimes\mathcal{O}_E(1)$ hat eine vollständige Filtrierung mit Quotienten $p^\ast\mathcal{L}_i\otimes\mathcal{O}_E(1)$ und der nachfolgende Satz zeigt:
+\[ 0 = \prod_{i=1}^{r}(c_1(\mathcal{O}_E(1)) + c_1(p^\ast\mathcal{L}_i) = \sum_{i=0}^r\zeta^{r-i}\tilde{\sigma}_i\]
+Für ein $\alpha\in A_\bullet(X)$ und $\nu\geq 1$ gilt mit $s_j(\mathcal{E}) = 0$ für $-r<j<0$:
+\begin{align*}
+0 &= p_\ast\Big(\zeta^{\nu-1}\sum_{i=0}^r\zeta^{r-i}\tilde{\sigma}_i\cap p^\ast\alpha\Big)\\
+&= p_\ast\Big(\sum_{i=0}^r \zeta^{r-1+\nu-i}\cap p^\ast(\sigma_i\cap\alpha)\Big)\\
+&= (s_\nu(\mathcal{E})\sigma_0 + s_{\nu-1}(\mathcal{E})\sigma_1 + \ldots + s_{\nu-r}(\mathcal{E})\sigma_r)\cap \alpha
+\end{align*}
+und somit:
+\[ s(\mathcal{E})(\sigma_0+\sigma_1+\ldots+\sigma_r) = s_0(\mathcal{E})\sigma_0 = 1 \]
+Dies zeigt $c_i(\mathcal{E}) = \sigma_i$.
 \end{proof}
@@
 \item $c_r(j^\ast\mathcal{E}) = 0$
-\item Sei $Z=\varnothing$ und $\mathcal{E}$ besitzt eine vollständige Filtrierung mit Geradenbündeln $\mathcal{L}_1,\ldots,\mathcal{L}_r$ als Quotienten. Dann gilt für alle $\alpha\in A_\bullet(X)$:
-\[ j^\ast\Big(\prod_{i=1}^r c_1(\mathcal{L}_i)\cap\alpha \Big) = 0 \]
+\item Sei $Z=\varnothing$ und $\mathcal{E}$ besitze eine vollständige Filtrierung mit Geradenbündeln $\mathcal{L}_1,\ldots,\mathcal{L}_r$ als Quotienten. Dann gilt für alle $\alpha\in A_\bullet(X)$:
+\[ \prod_{i=1}^r c_1(\mathcal{L}_i)\cap\alpha = 0 \]
 \end{enumerate}
 \end{proposition}
+
+\begin{proof}
+(i) folgt aus (ii) durch Rückzug:
+\[ j^\ast(c_r(\mathcal{E})\cap\alpha) = c_r(j^\ast\mathcal{E})\cap j^\ast\alpha = 0 \]
+und Ausschneidung $A_\bullet(Z)\to A_\bullet(X) \stackrel{j^\ast}{\to} A_\bullet(U)\to 0$.
+(ii) folgt aus (iii) mit dem Splitting-Prinzip und der Whitney-Formel $c_r(\mathcal{E}) = \prod_{i=1}^rc_1(\mathcal{L}_i)$.
+
+(iii) zeigen wir per Induktion über $r$.
+Sei $s_r\in\mathrm{H}^0(X, \mathcal{L}_r)$ das Bild von $s$ unter $\mathcal{E}\to\mathcal{L}_r$, und sei $i_r: Z_r\hookrightarrow X$ die Nullstellenmenge von $s_r$, d.h.
+die Menge aller Punke $x\in X$, so dass das Bild von $s_r$ in $(\mathcal{L}_r)_x \otimes_{\mathcal{O}_{X,x}} k(x)$ verschwindet.
+Dann kann $\mathcal{L}_r$ durch einen Pseudo-Divisor $(\mathcal{L}_r,Z_r,s_r)$ dargestellt werden und nach Konstruktion gilt:
+\[ c_1(\mathcal{L}_r)\cap \alpha = i_{r,\ast}((\mathcal{L}_r, Z_r, s_r)\cdot\alpha) = i_{r,\ast}\beta \in i_{r,\ast}A_\bullet(Z_r)\subset A_\bullet(X) \]
+Nach der Induktionsvoraussetzung, angewendet auf $\mathcal{E}' = i_r^\ast \ker(\mathcal{E}\to\mathcal{L}_r)$, mit dem induzierten nicht-verschwindenden Schnitt $s|_{Z_r}\in\mathrm{H}^0(Z_r,\mathcal{E}')\subset \mathrm{H}^0(Z_r, i_r^\ast\mathcal{E})$ und Filtrationsquotienten $i^\ast_r\mathcal{L}_1,\ldots,i_r^\ast\mathcal{L}_{r-1}$, erhalten wir:
+\[ \prod_{i=1}^rc_1(\mathcal{L}_i)\cap\alpha = \prod_{i=1}^{r-1} c_1(\mathcal{L}_i)\cap (i_{r,\ast}\beta) = i_{r,\ast}\Big(\prod_{i=1}^{r-1} c_1(i_r^\ast\mathcal{L}_i)\cap\beta\Big) = 0 \]
+Der Induktionsanfang $r=1$ folgt aus der Tatsache, dass für den trivialen Geradenbündel $c_1(\mathcal{O}_X) = 0$ gilt.
+\end{proof}
 
@@
 
+\begin{proof}
+\begin{enumerate}[(i)]
+\item Die Surjektivität wurde im zweiten Vortrag bewiesen.
+\end{enumerate}
+\end{proof}
+
 \begin{definition}

### 2015-11-30 Seminar Schnitttheorie/article.tex
Fehlt noch Theorem 9

@@
 
+\begin{proposition}
+Sei $X$ ein Schema.
+\begin{enumerate}[(i)]
+\item Sei $\mathcal{E}$ eine lokal freie Garbe und $E=\mathbf{V}(\mathcal{E})\to X$ das zugehörige Vektorbündel. Dann gilt:
+\[ \mathbf{P}(\mathcal{E})\cong P(E) \]
+\end{enumerate}
+\end{proposition}
+
 \begin{definition}
@@
 \begin{align*}
-s_i(E)\cap -: A_d(X)&\to A_{d-i}(X),\\ 
+s_i(E)\cap -: A_k(X)&\to A_{k-i}(X),\\ 
 \alpha&\mapsto p_\ast(c_1(\mathcal{O}_E(1))^{(r-1)+i} \cap p^\ast\alpha)
 \end{align*}
-wobei $r-1$ die relative Dimension des projektiven Vektorbündels $p: P(E)\to X$ bezeichnet.
+wobei $r-1$ die relative Dimension des projektiven Vektorbündels $p: P(E)\to X$ bezeichnet. Für eine lokal freie Garbe $\mathcal{E}$ setzen wir:
+\[ s_i(\mathcal{E})= s_i(\mathbf{V}(\mathcal{E}^\vee)) \]
+Dies impliziert $s_i(E) = s_i(\Gamma(E))$.
 \end{definition}
@@
 Dies zeigt (iv) und (v). Für (i) und (ii) sei $\alpha = [V]$ ein Primzykel und nach der Projektionsformel für $V\hookrightarrow X$ können wir o.B.d.A. annehmen, dass $V=X$ ganz ist. Dann gilt:
-\[s_i(\mathcal{E})\cap [X] \in \begin{cases}
-A_{\dim(X)-i}(X) = 0, & \text{wenn $i<0$}\\
-A_{\dim(X)}(X) = \mathbb{Z}[X], & \text{wenn $i = 0$}
+\[s_i(\mathcal{E})\cap [X] \begin{cases}
+\in A_{\dim(X)-i}(X) = 0, & \text{wenn $i<0$}\\
+= m\cdot [X], & \text{für ein $m\in \mathbb{Z}$ wenn $i = 0$}
 \end{cases}
 \]
-Um $s_1(\mathcal{E})\cap [X] = [X]$ zu zeigen, können wir nach (v) annehmen, dass $P(E)=X\times\mathbf{P}^r$ das triviale Bündel ist.
+Um $m=1$ zu zeigen, können wir nach (v) auf eine offene Teilmenge einschränken, so dass $P(E)=X\times_k\mathbf{P}^{r-1}_k$ das triviale Bündel ist.
 Wieder nach (v) für $X\to k$ können wir sogar annehmen, dass $X=\operatorname{Spec}(k)$ und $P(E)=\mathbf{P}_k^{r-1}$.
-Dann gilt:
-\[ \deg(c_1(\mathcal{O}(1))^{r-1}\cap [\mathbf{P}_k^{r-1}]) = \deg([\mathrm{Pt}]) = 1 \]
-Für (iii) betrachte das kartesische Quadrat:
+Es gilt:
+\[ c_1(\mathcal{O}(1))\cap [\mathbf{P}_k^{r-1}] = [\mathbf{P}_k^{r-2}] \]
+Wendet man dies $(r-1)$-mal an, zeigt dies $m=1$. Für (iii) betrachte das kartesische Quadrat:
 \[ \begin{tikzcd}
@@
 Die Aussage folgt, da $c_1$ kommutativ ist.
-Für (vi) sei $E\to X$ ein Geradenbündel und $\mathcal{E}$ eine invertierbare Garbe mit $E=\mathbf{V}(\mathcal{E})$ und $P(E)=\mathbf{P}(E)=X$ und $\mathcal{O}_E(1)=\mathcal{E} = \Gamma(E)^\vee$.
+Für (vi) sei $E\to X$ ein Geradenbündel und $\mathcal{E}$ eine invertierbare Garbe mit $E=\mathbf{V}(\mathcal{E})$ und $P(E)=\mathbf{P}(\mathcal{E})=X$ und $\mathcal{O}_E(1)=\mathcal{E} = \Gamma(E)^\vee$.
 Dann gilt:
@@
 \begin{corollary}
-Sei $p:E\to X$ ein Vektorbündel vom Rang $r$. Dann ist
-\[ p^\ast: A_q(X)\to A_{q+r-1}(P(E)) \]
+Sei $E\to X$ ein Vektorbündel vom Rang $r$. Dann ist
+\[ p^\ast: A_k(X)\to A_{k+r-1}(P(E)) \]
 injektiv und besitzt den Schnitt:
@@
 \begin{proof}
-Betrachte das projektive Vektorbündel $f: \mathbf{P}(\mathcal{E})\to X$ und die natürliche surjektive Abbildung $f^\ast\mathcal{E}\to\mathcal{O}_\mathcal{E}(1)$.
-Nach dem vorherigen Korollar ist $f^\ast$ injektiv.
+Per Induktion über $r$.
+Ist $r=1$, so ist die Aussage trivial.
+Sei nun $r>1$ und $p: \mathbf{P}(\mathcal{E})\to X$ das projektive Vektorbündel.
+Nach dem vorherigen Korollar ist $p^\ast$ injektiv.
+Betrachte die exakte Folge:
+\[0 \longrightarrow \mathcal{K} \longrightarrow p^\ast\mathcal{E}\longrightarrow \mathcal{O}_\mathcal{E}(1) \longrightarrow 0\]
+Nach Induktionsvoraussetzung gibt es ein $f': Y\to\mathbf{P}(\mathcal{E})$ mit injektivem $f'^\ast$ und eine vollständige Filtration von $f'^\ast\mathcal{K} \subset f'^\ast p^\ast\mathcal{E}$.
+Setze $f = p\circ f'$.
 \end{proof}
@@
 \[ c_t(\mathcal{E}) = \sum_{i =0}^\infty c_i(\mathcal{E}) t^i \]
-wobei $c_i(\mathcal{E})\cap - : A_q(X)\to A_{q-i}(X)$. Die \textit{totale Chern-Klasse} ist:
+wobei $c_i(\mathcal{E})\cap - : A_k(X)\to A_{k-i}(X)$. Die \textit{totale Chern-Klasse} ist:
 \[ c(\mathcal{E}) = c_t(\mathcal{E})|_{t =1}\in\operatorname{End}(A_\bullet(X)) \]
@@
 \end{remark}
+
+\begin{lemma}
+Sei $X$ eine Varietät und $\mathcal{E}$ eine lokal freie Garbe vom Rang $r$ auf $X$.
+Sei $s\in\Gamma(X,\mathcal{E})$ und $Z=Z(s)\hookrightarrow X$ das geschlossene Unterschema wo $s$ verschwindet, d.h.
+$Z=\{x\in X\mid\text{Bild von $s$ verschwindet in $\mathcal{E}_x \otimes_{\mathcal{O}_{X,x}} k(x)$} \}$, und $j:U=X\setminus Z\hookrightarrow X$ das offene Komplement.
+Dann gilt:
+\begin{enumerate}[(i)]
+\item Für alle $\alpha\in A_\bullet(X)$ gilt $c_r(\mathcal{E})\cap\alpha = \beta$, wobei $\beta$ einen Vertreter besitzt, dessen Träger in $Z$ liegt.
+\item $c_r(j^\ast\mathcal{E}) = 0$
+\item Sei $Z=\varnothing$ und $\mathcal{E}$ besitze eine vollständige Filtrierung mit Geradenbündeln $\mathcal{L}_1,\ldots,\mathcal{L}_r$ als Quotienten. Dann gilt für alle $\alpha\in A_\bullet(X)$:
+\[ \prod_{i=1}^r c_1(\mathcal{L}_i)\cap\alpha = 0 \]
+\end{enumerate}
+\end{lemma}
 
@@
 \begin{proof}
-(ii) und (vii) sind klar. (iii), (iv) und (v) folgen aus den Aussagen für die Segre-Klassen, da Chern-Klassen Polynome in den Segre-Klassen sind.
+(ii), (iii) und (vii) sind klar. (iv) und (v) folgen aus den Aussagen für die Segre-Klassen, da Chern-Klassen Polynome in den Segre-Klassen sind.
 
@@
 Dies zeigt $c_i(\mathcal{E}) = 0$ für $i> \operatorname{rk}(\mathcal{E}) = 1$.
-Für den allgemeinen Fall betrachte das Splitting-Prinzip und die Whitneysche Summenformel.
+Für den allgemeinen Fall betrachte das Splitting-Prinzip und (v).
+Wir können annehmen, dass $\mathcal{E}$ eine vollständige Filtration besitzt mit Quotienten $\mathcal{L}_1,\ldots,\mathcal{L}_r$.
 Die Whitneysche Summenformel ist äquivalent zu $c(\mathcal{E}')c(\mathcal{E}'') = c(\mathcal{E})\in\operatorname{End}(A_\bullet(X))$.
+Daher gilt:
+\[ (\star) \qquad c(\mathcal{E}) = \prod_{i=1}^r c(\mathcal{L}_i) = \prod_{i=1}^r(1+c_1(\mathcal{L}_i)) \phantom{\qquad(\star)} \]
+Dies zeigt (a).
 
 Für die Summenformel können wir nach dem Splitting-Prinzip und (v) annehmen, dass $\mathcal{E}'$ und $\mathcal{E}''$ vollständige Filtrationen besitzen.
-Diese induzieren eine vollständige Filtration auf $\mathcal{E}$ mit Geradenbündeln $\mathcal{L}_1,\ldots,\mathcal{L}_r$ als Quotienten und zu zeigen ist:
-\[ c(\mathcal{E}) = \prod_{i=1}^r c(\mathcal{L}_i) = \prod_{i=1}^r(1+c_1(\mathcal{L}_i)) \]
-Ist $\sigma_i\in\operatorname{End}(A_\bullet(X))$ das $i$-te elementar-symmetrische Polynom in $c_1(\mathcal{L}_1),\ldots,$ $c_1(\mathcal{L}_r)$.
-Dann ist die obige Gleichung äquivalent zu $c_i(\mathcal{E}) = \sigma_i$.
+Diese induzieren eine vollständige Filtration auf $\mathcal{E}$ mit Geradenbündeln $\mathcal{L}_1,\ldots,\mathcal{L}_r$ als Quotienten und zu zeigen ist nun $(\star)$.
 
-Sei $p:P(E)\to X$ das zugehörige projektive Vektorbündel und $\tilde{\sigma}_i$ das $i$-te elementar-symmetrische Polynom in $c_1(p^\ast\mathcal{L}_1),\ldots,c_1(p^\ast\mathcal{L}_r)$.
+Ist $\sigma_i\in\operatorname{End}(A_\bullet(X))$ das $i$-te elementar-symmetrische Polynom in $c_1(\mathcal{L}_1),\ldots,$ $c_1(\mathcal{L}_r)$ und $\sigma_0 =1$.
+Dann ist $(\star)$ äquivalent zu $c_i(\mathcal{E}) = \sigma_i$.
+
+Sei $p:P(E)\to X$ das zugehörige projektive Vektorbündel und $\tilde{\sigma}_i$ das $i$-te elementar-symmetrische Polynom in $c_1(p^\ast\mathcal{L}_1),\ldots,c_1(p^\ast\mathcal{L}_r)$ und $\tilde{\sigma}_0 = 1$.
 Setze $\zeta = c_1(\mathcal{O}_E(1))$.
@@
 Die Garbe $p^\ast\mathcal{E}\otimes\mathcal{O}_E(1)$ hat eine vollständige Filtrierung mit Quotienten $p^\ast\mathcal{L}_i\otimes\mathcal{O}_E(1)$ und der nachfolgende Satz zeigt:
-\[ 0 = \prod_{i=1}^{r}(c_1(\mathcal{O}_E(1)) + c_1(p^\ast\mathcal{L}_i) = \sum_{i=0}^r\zeta^{r-i}\tilde{\sigma}_i\]
-Für ein $\alpha\in A_\bullet(X)$ und $\nu\geq 1$ gilt mit $s_j(\mathcal{E}) = 0$ für $-r<j<0$:
+\[ 0 = \prod_{i=1}^r c_1(p^\ast\mathcal{L}_i\otimes\mathcal{O}_E(1)) = \prod_{i=1}^{r}(\zeta + c_1(p^\ast\mathcal{L}_i)) = \sum_{i=0}^r\tilde{\sigma}_i\zeta^{r-i}\]
+Für ein $\alpha\in A_\bullet(X)$ und $\nu\geq 1$ gilt:
 \begin{align*}
 0 &= p_\ast\Big(\zeta^{\nu-1}\sum_{i=0}^r\zeta^{r-i}\tilde{\sigma}_i\cap p^\ast\alpha\Big)\\
-&= p_\ast\Big(\sum_{i=0}^r \zeta^{r-1+\nu-i}\cap p^\ast(\sigma_i\cap\alpha)\Big)\\
+&= p_\ast\Big(\sum_{i=0}^r c_1(\mathcal{O}_E(1))^{r-1+\nu-i}\cap p^\ast(\sigma_i\cap\alpha)\Big)\qquad \text{(Rückzug)}\\
 &= (s_\nu(\mathcal{E})\sigma_0 + s_{\nu-1}(\mathcal{E})\sigma_1 + \ldots + s_{\nu-r}(\mathcal{E})\sigma_r)\cap \alpha
 \end{align*}
-und somit:
+und somit folgt unter Beachtung von $s_j(\mathcal{E}) = 0$ für $j<0$:
 \[ s(\mathcal{E})(\sigma_0+\sigma_1+\ldots+\sigma_r) = s_0(\mathcal{E})\sigma_0 = 1 \]
@@
 
-\begin{proposition}
-Sei $X$ eine Varietät und $\mathcal{E}$ eine lokal freie Garbe vom Rang $r$ auf $X$.
-Sei $s\in\Gamma(X,\mathcal{E})$ und $Z=Z(s)\hookrightarrow X$ das geschlossene Unterschema wo $s$ verschwindet und $j:U=X\setminus Z\hookrightarrow X$ das offene Komplement.
-Dann gilt:
-\begin{enumerate}[(i)]
-\item Für alle $\alpha\in A_\bullet(X)$ gilt $c_r(\mathcal{E})\cap\alpha = \beta$, wobei $\beta$ einen Vertreter besitzt, dessen Träger in $Z$ liegt.
-\item $c_r(j^\ast\mathcal{E}) = 0$
-\item Sei $Z=\varnothing$ und $\mathcal{E}$ besitze eine vollständige Filtrierung mit Geradenbündeln $\mathcal{L}_1,\ldots,\mathcal{L}_r$ als Quotienten. Dann gilt für alle $\alpha\in A_\bullet(X)$:
-\[ \prod_{i=1}^r c_1(\mathcal{L}_i)\cap\alpha = 0 \]
-\end{enumerate}
-\end{proposition}
-
-\begin{proof}
+\begin{proof}[Beweis von Lemma.]
 (i) folgt aus (ii) durch Rückzug:
 \[ j^\ast(c_r(\mathcal{E})\cap\alpha) = c_r(j^\ast\mathcal{E})\cap j^\ast\alpha = 0 \]
-und Ausschneidung $A_\bullet(Z)\to A_\bullet(X) \stackrel{j^\ast}{\to} A_\bullet(U)\to 0$.
-(ii) folgt aus (iii) mit dem Splitting-Prinzip und der Whitney-Formel $c_r(\mathcal{E}) = \prod_{i=1}^rc_1(\mathcal{L}_i)$.
+und der Ausschneidungssequenz $A_\bullet(Z)\to A_\bullet(X) \stackrel{j^\ast}{\to} A_\bullet(U)\to 0$.
+(ii) folgt aus (iii) mit dem Splitting-Prinzip und der Whitney-Formel $c_r(\mathcal{E}) = \prod_{i=1}^rc_1(\mathcal{L}_i)$ für den höchsten Koeffizienten.
 
 (iii) zeigen wir per Induktion über $r$.
-Sei $s_r\in\mathrm{H}^0(X, \mathcal{L}_r)$ das Bild von $s$ unter $\mathcal{E}\to\mathcal{L}_r$, und sei $i_r: Z_r\hookrightarrow X$ die Nullstellenmenge von $s_r$, d.h.
-die Menge aller Punke $x\in X$, so dass das Bild von $s_r$ in $(\mathcal{L}_r)_x \otimes_{\mathcal{O}_{X,x}} k(x)$ verschwindet.
+Sei $s_r\in\mathrm{H}^0(X, \mathcal{L}_r)$ das Bild von $s$ unter $\mathcal{E}\twoheadrightarrow\mathcal{L}_r$, und sei $i_r: Z_r\hookrightarrow X$ das geschlossene Unterschema, wo $s_r$ verschwindet.
 Dann kann $\mathcal{L}_r$ durch einen Pseudo-Divisor $(\mathcal{L}_r,Z_r,s_r)$ dargestellt werden und nach Konstruktion gilt:
@@
 \begin{proof}
-\begin{enumerate}[(i)]
-\item Die Surjektivität wurde im zweiten Vortrag bewiesen.
-\end{enumerate}
+Die Surjektivität von $p^\ast$ wurde im zweiten Vortrag bewiesen.
 \end{proof}

### 2015-11-30 Seminar Schnitttheorie/article.tex
Seminarvortrag halbwegs fertig

@@
 \begin{enumerate}[(i)]
-\item Der flache Rückzug $p^\ast: A_d(X)\to A_{d+r}(E)$ ist ein Isomorphismus.
+\item Der flache Rückzug $p^\ast: A_k(X)\to A_{k+r}(E)$ ist ein Isomorphismus.
 \item Es gibt einen kanonischen Isomorphismus:
 \begin{align*}
-\vartheta_E: \bigoplus_{i=0}^{r-1}A_{d+i}(X) &\stackrel{\sim}{\longrightarrow} A_{d+r-1}(P(E))\\
-A_{d+i}(X)\ni \alpha_i &\longmapsto c_1(\mathcal{O}_E(1))^i\cap q^\ast(\alpha_i)
+\theta_E: \bigoplus_{i=0}^{r-1}A_{k +i}(X) &\stackrel{\sim}{\longrightarrow} A_{k +r-1}(P(E))\\
+A_{k+i}(X)\ni \alpha_i &\longmapsto c_1(\mathcal{O}_E(1))^i\cap q^\ast\alpha_i
 \end{align*}
-\item Sei $\mathcal{E} = \Gamma(E)$. Die Chern-Klassen von $\mathcal{E}$ sind eindeutig durch die folgende Formeln bestimmt:
-\[ c_0(\mathcal{E}) = 1,\quad \sum_{i=0}^r c_1(\mathcal{O}_E(1))^i\cap q^\ast(c_{r-i}(\mathcal{E})\cap\alpha) = 0 \]
 \end{enumerate}
@@
 \begin{proof}
-Die Surjektivität von $p^\ast$ wurde im zweiten Vortrag bewiesen.
+Die Surjektivität von $p^\ast$ wurde im zweiten Vortrag schon gezeigt.
+Um zu sehen, dass $\theta_E$ surjektiv ist, können wir per noethersche Induktion auf den Fall reduzieren, in dem $E=X\times\mathbf{A}^r$ trivial ist.
+Per Induktion über $r$, genügt es zu zeigen, dass $\theta_{E\oplus 1}$ surjektiv ist, wenn $\theta_E$ surjektiv ist.
+Der Fall $r=1$ ist trivial.
+
+Die Inklusion $E\hookrightarrow E\oplus 1$ induziert eine offene Einbettung $j: E\hookrightarrow P(E\oplus 1)$ mit Komplement $i: P(E)\hookrightarrow P(E\oplus 1)$ mit $\mathcal{O}_E(1)= i^\ast\mathcal{O}_{E\oplus 1}(1)$.
+Sei $q': P(E\oplus 1)\to X$ die Projektion.
+Es gilt:
+\[ j^\ast c_1(\mathcal{O}_{E\oplus 1}(1)) = 0 \]
+da $\mathcal{O}_{E\oplus 1}(1)$ die Idealgarbe zu $P(E)\hookrightarrow P(E\oplus 1)$ und daher trivial auf $E$ ist. Außerdem gilt:
+\[ i_\ast q^\ast\alpha = c_1(\mathcal{O}_{E\oplus 1}(1))\cap q'^\ast\alpha \]
+Ferner gilt:
+\begin{align*}
+i_\ast(c_1(\mathcal{O}_E(1))^\nu \cap q^\ast\alpha) &= i_\ast(c_1(i^\ast\mathcal{O}_{E\oplus 1}(1))^\nu \cap q^\ast\alpha)\\
+&= c_1(\mathcal{O}_{E\oplus 1}(1))^\nu\cap i_\ast q^\ast \alpha\\
+&= c_1(\mathcal{O}_{E\oplus 1}(1))^{\nu+1}\cap q'^\ast\alpha
+\end{align*}
+Es kommutiert daher:
+\[\begin{tikzcd}
+& A_{k + r}(P(E)) \rar["i_\ast"] & A_{k + r}(P(E\oplus 1)) \rar["j^\ast"] & A_{k + r}(E)\rar & 0\\
+0 \rar & \displaystyle\bigoplus_{i=1}^r A_{k + i}(X) \rar \uar["\theta_E{[1]}"] & \displaystyle\bigoplus_{i=0}^r A_{k + i}(X) \uar["\theta_{E\oplus 1}"] \rar["\operatorname{pr}_0"']  & A_k(X) \rar\uar["p^\ast"'] & 0
+\end{tikzcd}\]
+wobei die obere Zeile wegen der Ausschneidungssequenz exakt ist. Nach dem Fünfer-Lemma folgt (i) aus (ii). Per Diagrammjagd sehen wir die Surjektivität von $\theta_E[1]$.
+
+Für die Injektivität sei $\theta_E(\alpha_0,\ldots,\alpha_{r-1})=0$ und $\nu$ der größte Index mit $\alpha_\nu\neq 0$. Dann gilt:
+\[ 0 = q_\ast(c_1(\mathcal{O}_E(1))^{r-1-\nu}\cap \theta_E(\alpha_0,\ldots,\alpha_{r-1})) = \sum_{i=0}^\nu s_{i-\nu}(\mathcal{E})\cap\alpha_i = \alpha_\nu \]
+ein Widerspruch.
 \end{proof}

### 2015-11-30 Seminar Schnitttheorie/article.tex
Papierrand einstellen
Define vector bundles again

@@
-\documentclass[10pt,a4paper]{article}
+\documentclass[10pt,b5paper]{article}
+\usepackage[left=2.00cm, right=2.00cm, top=2.00cm, bottom=2.00cm]{geometry}
 \usepackage{fontspec}
@@
 \begin{enumerate}[(i)]
-\item Sei $p: E\to X$ ein Vektorbündel vom Rang $n$ und $X=\bigcup_\alpha U_\alpha$ eine offene Überdeckung mit $U_\alpha$-Isomorphismen $\varphi_j: f^{-1}(U_\alpha)\to \mathbf{A}_{U_\alpha}^n$.
-Wir bezeichnen den Nullschnitt mit $i_0: X\to E$.
-\item Für eine lokal freie Garbe $\mathcal{E}$ vom Rang $r$ auf $X$ definiert
+\item Ein Vektorbündel vom Rang $n$ ist ein Morphismus $p: E\to X$, so dass eine offene Überdeckung $X=\bigcup_\alpha U_\alpha$ existiert mit $U_\alpha$-Isomorphismen $\varphi_\alpha: f^{-1}(U_\alpha)\to \mathbf{A}_{U_\alpha}^n$, so dass die Übergangsabbildungen $\varphi_{\alpha\beta} = \varphi_\alpha\circ\varphi_\beta^{-1}: \mathbf{A}^n_{U_\alpha\cap U_\beta}\to\mathbf{A}^n_{U_\alpha\cap U_\beta}$ lineare Automorphismen sind.
+\item Für eine lokal freie Garbe $\mathcal{E}$ vom Rang $n$ auf $X$ definiert
 \[ \mathbf{V}(\mathcal{E}) = \mathbf{Spec}_X(\operatorname{Sym}^\bullet\mathcal{E})\to X \]
-ein Vektorbündel auf $X$.
-\item Für einen Vektorbündel $E\to X$ vom Rang $r$ definiert
+ein Vektorbündel vom Rang $n$ auf $X$.
+\item Für einen Vektorbündel $E\to X$ vom Rang $n$ definiert
 \[ \Gamma(E) = (U\mapsto \operatorname{Hom}_X(U, E)) \]
-eine lokal freie Garbe auf $X$ vom Rang $r$.
+eine lokal freie Garbe auf $X$ vom Rang $n$.
 \end{enumerate}

### 2016-02-15 Master Thesis/synopsis.tex
Small changes

@@
 Let $K$ be a field of finite dimension.
-We denote the set of equivalence classes of non-archimedean valuations of $K$ by $\operatorname{Spv}(K)$.
+We denote the set of equivalence classes of Krull valuations of $K$ by $\operatorname{Spv}(K)$.
 We often do not distinguish between a valuation and its equivalence class.
@@
 Then there exists a unique henselian $w\in|\operatorname{Spv}(K)|$.
-\item Assume that the hypothesis of (i) is satisfied for two different prime numbers. Then $w$ is defectless.
+\item Assume that (i) is satisfied for two different prime numbers. Then $w$ is defectless.
 \end{enumerate}
@@
 \end{tikzcd} \]
-By a limit process we see that $\sigma$ induces a bijection $\sigma: \operatorname{Spv}_\text{d}^{(1)}(\overline{K})\to \operatorname{Spv}_\text{d}^{(1)}(\overline{K'})$.
+By limit process we see that $\sigma$ induces a bijection $\sigma: \operatorname{Spv}_\text{d}^{(1)}(\overline{K})\to \operatorname{Spv}_\text{d}^{(1)}(\overline{K'})$.
 Let $\overline{v}\in \operatorname{Spv}_\text{d}^{(1)}(\overline{K})$ and $G_{\overline{v}}\subset G_K$ be the decomposition group of $\overline{v}$.
@@
 We denote the natural projection $U_P\to k(P)^\times,\ x\mapsto x\mod P$ by $r_P$.
-For $n\in\mathbb{N}$ the following sequence is exact:
-\[ 0\to U_P/n\to K^\times/n\to\mathbb{Z}/n\mathbb{Z}\to 0 \]
+For $n\in\mathbb{N}$ the following sequence is exact, since valuation rings are integrally closed:
+\[ 0\longrightarrow U_P/n\longrightarrow K^\times/n\longrightarrow\mathbb{Z}/n\mathbb{Z}\longrightarrow 0 \]
+We'll use the same notation for $P' = \sigma(P)\in X'_0$.
 

### 2016-02-15 Master Thesis/synopsis.tex
Turn remark into definition

@@
 
-\begin{remark}
+\begin{definition}
 For each admissible $S\subset X_0$ and $T_1\subset X_0\setminus S$ with injective $r_{S,T_1}$ we also have the injectivity of $r_{S,T_2}$ for every $T_2 \supset T_1$.
@@
 Such a $T$ will be called $S$-\textit{admissible} and $(S,T)$ is called \textit{admissible pair}.
-\end{remark}
+\end{definition}
 

### 2016-02-15 Master Thesis/synopsis.tex
WTF?
Remarks are numbered

@@
 
-\begin{remark}
+\begin{remark}\label{3.3}
 If $L\subset M$ are intermediate fields of $\overline{K}/K$, which are finite over $K$, then $\sigma$ also induces bijections
@@
 \begin{lemma}\label{5.4}
-Let $S\subset X_0$ be admissible and $P_0\in X_0\setminus S$. Then there exists a finite subset $T\subset X_0\setminus S$ with $P_0\in T$ such that the following map is injective:
+Let $S\subset X_0$ be admissible. Then there exists a finite subset $T\subset X_0\setminus S$ such that the following map is injective:
 \[ r_{S,T}:K^S \to\prod_{P\in T}k(P)^\times,\ x\mapsto(r_P(x))_{P\in T} \]

### 2016-02-15 Master Thesis/synopsis.tex
Completion of main proof.

@@
 Since any rational function on $X$ is uniquely determined by its values at those points where it is defined, $\varphi$ is unique.
-Furthermore, the following maps are injective for all finite subsets $S\subset X_0$:
+In other words, the following maps are injective for all finite subsets $S\subset X_0$:
 \[ r_S:K^S \to\prod_{P\in X_0\setminus S} k(P)^\times,\ x\mapsto (r_P(x))_P \]
@@
 Thus, we can combine the isomorphisms $\varphi^S$, where $S$ ranges over the admissible subsets of $X_0$, into an isomorphism $\varphi = \varinjlim \varphi^S: K'^\times \to K^\times$.
+
 Note that for every $x\in K^\times$ we have $v_P(x)= 0$ for almost all $P\in X_0$; similarly for $K'$ and $X'_0$.
-
-Now note that for any $x\in U_{P'}$ we can find an admissible set $S\subset X_0\setminus \{P\}$ such that $x\in K'^{S'}$.
+By Lemma \ref{4.5} (i) we can find an admissible set $S\subset X_0\setminus \{P\}$ such that $x\in K'^{S'}$ for any $x\in U_{P'}$.
 \end{proof}
@@
 
+\begin{remark}
+For any finite subextension $L/K$ of $\overline{K}/K$ there exists an isomorphism $\varphi_L: L'\to L$.
+If $K\subset L\subset M$ are finite subextensions, we have $\varphi_M|_{L'} = \varphi_{L'}$, since $r_Q(\varphi_M(x)) = r_Q(\varphi_L(x))$ for all $x\in L'^\times$ and almost all $Q\in (X_M)_0$.
+By passing to limits we obtain an isomorphism $\varphi: \overline{K}'\to \overline{K}$.
+\end{remark}
+
 \begin{step}
@@
 Clearly, $\tau$ is a group isomorphism and we have $\tau(G_L)=G_{L'}=\sigma(G_L)$ for every finite subextension $L/K$ of $\overline{K}/K$.
-By Lemma \ref{3.2}, $\tau$ induces bijections $\tau: \operatorname{Spv}_\text{d}^{(1)}(L)\to \operatorname{Spv}_\text{d}^{(1)}(L')$ for any intermediate field $L$ of $\overline{K}/K$.
+By applying Lemma \ref{3.2} on $\tau$, we get bijections $\tau: \operatorname{Spv}_\text{d}^{(1)}(L)\to \operatorname{Spv}_\text{d}^{(1)}(L')$ for any intermediate field $L$ of $\overline{K}/K$.
+
+We have $L_{\tau(v)} = (L_v)' = L_{\sigma(v)}$ for all $v\in\operatorname{Spv}_\text{d}^{(1)}(L)$, i.e.
+$\sigma$ and $\tau$ induce the same maps $\operatorname{Spv}_\text{d}^{(1)}(L)\to \operatorname{Spv}_\text{d}^{(1)}(L')$.
+By limit process, $\sigma$ and $\tau$ induce the same map $\operatorname{Spv}_\text{d}^{(1)}(\overline{K})\to \operatorname{Spv}_\text{d}^{(1)}(\overline{K}')$.
+
+Let $g\in G_K$ and $\overline{v}\in\operatorname{Spv}_\text{d}^{(1)}(\overline{K})$. By Remark \ref{3.3} we have:
+\[ \sigma(g)\sigma(\overline{v}) = \sigma(g\overline{v}) = \tau(g\overline{v}) = \tau(g)\tau(\overline{v}) = \tau(g) \sigma(\overline{v}) \]
+Therefore $\sigma(g)^{-1}\tau(g)$ is contained in the decomposition group $G_{\overline{v}}$ for all $\overline{v}\in\operatorname{Spv}_\text{d}^{(1)}(\overline{K})$.
+By the Theorem of F.K.
+Schmidt we have $G_{\overline{v}_1}\cap G_{\overline{v}_2} = 1$ for $\overline{v}_1\neq\overline{v}_2$.
+Therefore, $\sigma(g) = \tau(g) = \varphi^{-1}g\varphi$ for all $g\in G_K$.
+
+For the uniqueness of $\varphi$ let $\psi: \overline{K}'\to \overline{K}$ be another isomorphism with $\sigma(g) = \psi^{-1} g \psi$. Setting $h = \psi\varphi^{-1}\in G_K$ we observe
+\[ h^{-1}gh = \varphi\sigma(g)\varphi^{-1} = \sigma^{-1}\sigma(g) = g \]
+for all $g\in G_K$.
+For $\overline{v}\in\operatorname{Spv}_\text{d}^{(1)}(\overline{K})$ we have therefore $G_{\overline{v}} = h G_{\overline{v}} h^{-1} = G_{h\overline{v}}$, thus $h\in G_{\overline{v}}$ for all $\overline{v}$.
+Again, by F.K.
+Schmidt, we have $h=1$ and thus $\varphi = \psi$.
 \end{proof}

### 2016-02-15 Master Thesis/synopsis.tex
Fix typo

@@
 \begin{definition}\label{5.1}
-An abelian group is called \textit{free-by-finite} if its torsion subgroup $A_\text{tor}$ is finite and the quotien $A/A_\text{tor}$ is free abelian.
+An abelian group is called \textit{free-by-finite} if its torsion subgroup $A_\text{tor}$ is finite and the quotient $A/A_\text{tor}$ is free abelian.
 $A$ is free-by-finite if and only if $A$ is the direct sum of a finite group and a free abelian group.
@@
 \begin{enumerate}[(i)]
-\item Let $A\to B$ be a homomorphism of abelian groups with finite kernel and cokernel. Then $A$ is free-by-finite iff $B$ is free-by-finite.
+\item Let $f: A\to B$ be a homomorphism of abelian groups with finite kernel and cokernel. Then $A$ is free-by-finite iff $B$ is free-by-finite.
 \item Let $0\to A'\stackrel{i}{\to}A\stackrel{p}{\to}A''\to 0$ be an exact sequence of abelian groups.

### 2016-03-06 Master Thesis/synopsis.tex
Proof of main lemma

@@
 \end{lemma}
+
+\begin{proof}
+The degree $[K(\mu_\ell, \sqrt{-1}):K]$ is prime to $\ell$, thus we can assume that $\mu_\ell\subset K$ and $K$ is not formally real without loss of generality.
+Now, we can replace $\mathbb{Z}/\ell\mathbb{Z}$ by $\mathbb{Z}/\ell\mathbb{Z}(2) = \mu_\ell^{\otimes 2}$.
+
+We write $\mathrm{H}^3(L) = \mathrm{H}^3(L, \mathbb{Z}/\ell\mathbb{Z}(2))$.
+Let $\mathfrak{X}$ be a model of $K$.
+$\mathfrak{X}_i$ denotes the set of points $x\in\mathfrak{X}$ whose closure $\overline{\{x\}}$ is $i$-dimensional.
+For $x\in\mathfrak{X}_1$ the local ring $\mathcal{O}_{\mathfrak{X}, x}$ is a discrete valuation ring and defines a valuation $v_x\in\operatorname{Spv}_\text{d}^{(1)}(K)$.
+By [K1] Theorem 0.8, the following map
+\[ \mathrm{H}^3(K)\longrightarrow \prod_{x\in\mathfrak{X}_1}\mathrm{H}^3(K_{v_x}) \]
+is injective and its image is contained in $\bigoplus \mathrm{H}^3(K_{v_x})$. By [K1] Lemma 1.4, for a discrete valuation $v$ of $K$ the map
+\[ \mathrm{H}^3(K_v) \longrightarrow \prod_w\mathrm{H}^3(K_w) \]
+where $w$ ranges over all rank-$2$ valuations with $v(w) = v$, can be identified with the following map:
+\[ {}_\ell\operatorname{Br}(k) \longrightarrow \prod_{\tilde{w}\in\operatorname{Spv}^{(1)}(k)} {}_\ell\operatorname{Br}(k_{\tilde{w}}) \]
+where $k = k(v)$. This map is injective by the classical Hasse principle and its image is contained in $\bigoplus {}_\ell\operatorname{Br}(k_{\tilde{w}})$. Combining these maps yields the injection:
+\[ \mathrm{H}^3(K)\longrightarrow \bigoplus_{x\in\mathfrak{X}_1}\  \bigoplus_{\substack{w\in\operatorname{Spv}_\text{rd}^{(2)}(K) \\ v(w) = v_x}} \mathrm{H}^3(K_w) \]
+The direct sum ranges over a subset of $|\operatorname{Spv}(K)|$, thus concludes the proof.
+\end{proof}
 
@@
 \[ \sigma(g)\sigma(\overline{v}) = \sigma(g\overline{v}) = \tau(g\overline{v}) = \tau(g)\tau(\overline{v}) = \tau(g) \sigma(\overline{v}) \]
-Therefore $\sigma(g)^{-1}\tau(g)$ is contained in the decomposition group $G_{\overline{v}}$ for all $\overline{v}\in\operatorname{Spv}_\text{d}^{(1)}(\overline{K})$.
+Therefore $\sigma(g)^{-1}\tau(g)$ is contained in $G_{\overline{v}}$ for all $\overline{v}\in\operatorname{Spv}_\text{d}^{(1)}(\overline{K})$.
 By the Theorem of F.K.

### 2016-03-06 Master Thesis/synopsis.tex
Proof of 2.5

@@
 \end{lemma}
+
+\begin{proof}
+Let $G_\ell$ be an pro-$\ell$-Sylow group of $G_K$.
+By the ramification theory for general valuations ([E], §20) we have $G_\ell\cong\mathbb{Z}_\ell(1)^r\rtimes G_k(\ell)$ where the action is defined via the cyclotomic character and $r = \dim_{\mathbb{F}_\ell}(\Gamma/\ell\Gamma)$.
+In particular, $r\leq 2$ and $G_k(\ell)$ is the pro-$\ell$-Sylow group of $G_k$ hence $G_k(\ell)\cong \mathbb{Z}_\ell$ or $G_k(\ell) \cong 1$.
+
+Let $G_K(\ell)$ be the maximal pro-$\ell$-quotient of $G_K$.
+Since $\mu_\ell\subset K$, the action of $G_k$ on the pro-$\ell$-Sylow group of the inertia group of $w$ factors through $G_k(\ell)$.
+We see that the canonical projection $G_K\to G_K(\ell)$ is an isomorphism.
+Therefore we have:
+\[ \mathrm{H}^3(K,\mathbb{Z}/\ell\mathbb{Z}) \cong\mathrm{H}^3(G_K(\ell),\mathbb{Z}/\ell) \cong\mathrm{H}^3(G_\ell,\mathbb{Z}/\ell\mathbb{Z}) \]
+Now the equivalence of (i) and (ii) follows from [Se], Ch. I, §4.
+
+The second assertion follows from the first one, applied to $L$, by using the following equation of supernatural numbers ([E], Theorem 20.21):
+\[ [L:K] = [\Gamma':\Gamma][k':k]\cdot p^d \]
+for some $d \leq \infty$, where $\Gamma'$ denotes the value group and $k'$ the residue field of the unique prolongation of $w$ to $L$.
+\end{proof}
 

### 2016-03-06 Master Thesis/synopsis.tex
Fix group index notation

@@
 The degree $[K(\mu_\ell, \sqrt{-1}):K]$ is prime to $\ell$, thus we can assume that $\mu_\ell\subset K$ and $K$ is not formally real without loss of generality.
-Now, we can replace $\mathbb{Z}/\ell\mathbb{Z}$ by $\mathbb{Z}/\ell\mathbb{Z}(2) = \mu_\ell^{\otimes 2}$.
+Now we can replace $\mathbb{Z}/\ell\mathbb{Z}$ by $\mathbb{Z}/\ell\mathbb{Z}(2) = \mu_\ell^{\otimes 2}$.
 
@@
 The second assertion follows from the first one, applied to $L$, by using the following equation of supernatural numbers ([E], Theorem 20.21):
-\[ [L:K] = [\Gamma':\Gamma][k':k]\cdot p^d \]
+\[ [L:K] = (\Gamma':\Gamma) \cdot [k':k]\cdot p^d \]
 for some $d \leq \infty$, where $\Gamma'$ denotes the value group and $k'$ the residue field of the unique prolongation of $w$ to $L$.

### 2016-03-06 Master Thesis/synopsis.tex
Proof to Proposition 2.6

@@
 \end{proposition}
+
+\begin{proof}
+\begin{enumerate}[(i)]
+\item Since $\mathrm{H}^3(K_0, \mathbb{Z}/\ell\mathbb{Z})\neq 0$ by assumption, according to Lemma \ref{2.4} we can find a valuation $w_0\in|\operatorname{Spv}(K_0)|$ such that:
+\[ \mathrm{H}^3((K_0)_{w_0}, \mathbb{Z}/\ell\mathbb{Z}) \neq 0 \]
+By requirement $\ell^\infty \nmid [(K_0)_{w_0} : K_0]$. By replacing $K_0$ by $(K_0)_{w_0}$, we can assume that $w_0$ is henselian.
+
+Assume that $w = w_0|_K$ is not henselian.
+Then there exists a $w_1\in|\operatorname{Spv}(K_0)|$ with $w_1\neq w_0$ and such that $w = w_1|_K$, for otherwise $w$ would have a unique prolongation to $K_0$ and hence to $\overline{K}$ since $w_0$ is henselian.
+
+Le $(K_0)_{w_1}$ be a henselization of $K_0$ with respect to $w_1$ and let $K_w$ be a henselization of $K$ with respect to $w$ in $(K_0)_{w_1}$.
+Then $(K_0)_{w_1}$ is the composite of $K_0$ and $K_w$ and we have $\ell^\infty\nmid [(K_0)_{w_1} : K_w]$.
+
+Since $K_0$ is henselian with respect to $w_0$, a conjugate $\tau(K_w)$ for some $\tau\in G_K$ of $K_w$ lies in $K_0$, hence $\ell^\infty\nmid [K_w : K]$.
+Consequently, $\ell^\infty \nmid [(K_0)_{w_1} : K]$ and therefore $\ell^\infty \nmid [(K_0)_{w_1} : K_0]$.
+By requirement, $\mathrm{H}^3((K_0)_{w_1}, \mathbb{Z}/\ell\mathbb{Z}) \neq 0$.
+
+Now $(K_0)_{w_1}$ is henselian both with respect to $w_1$ and with respect to the prolongation of $w_0$.
+Lemma \ref{2.2} yields $\operatorname{cd}(G_{(K_0)_{w_1}})\leq 2$, a contradiction.
+The uniqueness of $w$ is again a consequence of Lemma \ref{2.2}.
+\item Choose a prime $\ell$ different from $\operatorname{char} k(w)$. According to Lemma \ref{2.5} we have $\operatorname{rrank}(w)\geq \dim_{\mathbb{F}_\ell}(\Gamma_w /\ell\Gamma_w) = 2 = \dim(K)$.
+\end{enumerate}
+\end{proof}
 
@@
 Hence, $\varphi^{S,T}$ is actually independent of $T$, thus we can omit the $T$ from its notation.
-Now, if $T$ ranges over all $S$-admissible subsets of $X_0$, then $(\star)$ shows:
+Now if $T$ ranges over all $S$-admissible subsets of $X_0$, then $(\star)$ shows:
 \[ r_S\circ\varphi^S = \Big(\prod_{P\in X_0\setminus S}\varphi_P \Big)\circ r_{S'} \]

### 2016-03-06 Master Thesis/synopsis.tex
Start proof to 2.2

@@
 Then $K$ is algebraically closed or $w_1,w_2\in\operatorname{Spv}_\text{rd}^{(2)}(K),\ v(w_1)=v(w_2)$ and $k(v(w_1))$ is separably closed.
-In the last case we have $\operatorname{cd}(G_K)\leq 2$.
+Furthermore, we have $\operatorname{cd}(G_K)\leq 2$.
 \end{lemma}
+
+\begin{proof}
+Assume that $K$ is not algebraically closed, i.e.
+$G_K\neq 1$ and set $v_i = v(w_i)$.
+If $v_1\neq v_2$, by the Theorem of F.K.
+Schmidt we have $G_{v_1}\cap G_{v_2} = 1 \neq G_K$, a contradiction.
+Therefore $v_1 = v_2$.
+
+Now $k(v_1) = k(v_2)$ is henselian with respect to both $w_1\mod v_1$ and $w_2\mod v_2$. Again by the Theorem of F.K. Schmidt we get $G_{k(v_1)} = 1$, hence $k(v_1)$ is separably closed.
+
+The second assertion follows from the first and from [K2], §5, Theorem 1.
+\end{proof}
 

### 2016-04-29 Master Thesis/synopsis.tex
Sounds less forceful
Add temporary proof for Lemma 3.1

@@
 Let $K,K'$ be two finitely generated fields of characteristic $0$ and dimension $2$ and $\sigma:G_K\to G_{K'}$ be an isomorphism of profinite groups.
-For intermediate fields $L$ of $\overline{K}/K$ the corresponding intermediate field of $\overline{K'}/K'$ will be denoted by $L'$.
+For intermediate fields $L$ of $\overline{K}/K$ we denote the corresponding intermediate field of $\overline{K'}/K'$ by $L'$.
 In other words $\sigma G_L=G_{L'}$.
@@
 \end{lemma}
+
+\begin{proof}
+Lemma \ref{3.1} shows that $\sigma$ induces a bijection $\sigma: \operatorname{Spv}_\text{d}^{(2)}(K)\to \operatorname{Spv}_\text{d}^{(2)}(K')$.
+For all $v\in\operatorname{Spv}(K)$ we choose a fixed henselization $K_v\subset\overline{K}$, such that if $v,w\in\operatorname{Spv}(K)$ and $w$ is a specialization of $v$ we have $K_v\subset K_w$.
+This means that if $v$ is nontrivial, we get $w\in\operatorname{Spv}_\text{rd}^{(2)}(K)$ and $v=v(w)\in\operatorname{Spv}_\text{d}^{(1)}(K)$.
+We set $K'_{\sigma(w)} = (K_w)'$ for $w\in\operatorname{Spv}_\text{d}^{(2)}(K)$.
+This implies that $K'_{\sigma(w)}$ is a henselization of $K'$ with respect to $\sigma(w)$.
+
+Let $w_1,w_2 \in\operatorname{Spv}_\text{d}^{(2)}(K)$ be two different valuations. By Lemma \ref{2.2} the following assertions are equivalent:
+\begin{enumerate}[(i)]
+\item The composite $K_{w_1}K_{w_2}$ is not algebraically closed.
+\item $w_1,w_2\in\operatorname{Spv}_\text{rd}^{(2)}(K)$ and $v(w_1) = v(w_2)$.
+\end{enumerate}
+In this case, $K_{w_1}K_{w_2}$ is the inertia field $K_v^\text{t}$ of $K_v$ where $v=v(w_1)$.
+Furthermore, $K'_{\sigma(w_1)}K'_{\sigma(w_2)} = (K_{w_1}K_{w_2})' \neq \overline{K'}$ and again by Lemma \ref{2.2} we get: \[\sigma(w_1),\sigma(w_2)\in\operatorname{Spv}_\text{rd}^{(2)}(K)\quad \text{and} \quad v(\sigma(w_1)) = v(\sigma(w_2))\]
+Thus $\sigma: \operatorname{Spv}_\text{d}^{(2)}(K)\to \operatorname{Spv}_\text{d}^{(2)}(K')$ maps $\operatorname{Spv}_\text{rd}^{(2)}(K)$ onto $\operatorname{Spv}_\text{rd}^{(2)}(K')$. Furthermore:
+\[ v(w_1)=v(w_2) \iff v(\sigma(w_1)) = v(\sigma(w_2)) \]
+This shows that $\sigma$ also induces a bijection $\sigma:\operatorname{Spv}_\text{d}^{(1)}(K)\to\operatorname{Spv}_\text{d}^{(1)}(K')$ and for all $w\in\operatorname{Spv}_\text{rd}^{(2)}(K)$ it holds $\sigma(v(w)) = v(\sigma(w))$.
+
+Let $v\in\operatorname{Spv}_\text{d}^{(1)}$ and let $N$ be the normal subgroup of $G_{K_v}$ generated by all subgroups $G_{K_w}$ for $w\in\operatorname{Spv}_\text{rd}^{(2)}(K)$ with $v(w) = v$.
+Since a global field has no non-trivial algebraic extension in which every prime splits completely we get $N=G_{K_v}$.
+This shows $K'_{v'} := (K_v)'$ is a henselization of $K'$ with respect to $v'=\sigma(v)$.
+Furthermore, let $w_1,w_2\in\operatorname{Spv}_\text{rd}^{(2)}(K)$ with $w_1\neq w_2$ and $v=v(w_1)=v(w_2)$, then:
+\[ \sigma(G_{K^\text{t}_v}) = \sigma(G_{K_{w_1}}\cap G_{K_{w_2}}) = G_{K'_{w'_1}}\cap G_{K'_{w'_2}} = G_{(K')^\text{t}_{v'}} \]
+Therefore, $\sigma$ induces an isomorphism:
+\[ G_{k(v)} \cong G_{K_v}/G_{K_v^\text{t}} \to G_{K'_{v'}}/G_{(K')_{v'}^\text{t}}\cong G_{k(v')} \]
+The rest follows from the Theorem of Neukirch-Uchida.
+\end{proof}
 

### 2016-04-29 Master Thesis/synopsis.tex
Copy of proof to 3.1

@@
 \end{lemma}
+
+\begin{proof}
+Let $\ell_1,\ell_2$ be two different prime numbers different from $\operatorname{char} k(w)$ and let $E_0 = E(\mu_{\ell_1\ell_2})$.
+Then $E_0/E$, and also $E'_0/E'$, satisfies the condition of \ref{2.6} (ii).
+Thus there exists a unique henselian defectless valuation $w'_{E'}$ of $E'$, so $E'$ contains a henselization $K'_{w'}$ of $w'_{E'}|_{K'}$.
+By applying the same argument to $\sigma^{-1}$ we find that $E' = K'_{w'}$.
+
+Let $F$ be another henselization of $w$. Then $E, F$ are $K$-isomorphic, hence $E',F'$ are $K'$-isomorphic and the corresponding henselian valuations of $E', F'$ respectively are mapped to each other.
+\end{proof}
 

### 2016-07-05 Seminar Graphentheorie/article.tex
Add first two sections of graph theory seminar

\documentclass[10pt,b5paper]{article}
\usepackage[left=2.00cm, right=2.00cm, top=2.00cm, bottom=2.00cm]{geometry}
\usepackage{fontspec}
\usepackage[ngerman]{babel}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{paralist}
\usepackage{tikz-cd}
\linespread{1.1}

\author{von \textsc{yichuan shen}}
\title{Ebene Graphen}
\begin{document}

\theoremstyle{plain}
\theoremstyle{definition}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Satz}
\newtheorem{corollary}[theorem]{Korollar}
\theoremstyle{definition}
\newtheorem*{definition}{Definition}
\newtheorem*{example}{Beispiel}
\theoremstyle{remark}
\newtheorem*{remark}{Bemerkung}

\maketitle

\section{Maximale Planarität}

\begin{definition}
Ein \textit{ebener Graph} ist ein Paar $G= (V, E)$ endlicher Mengen, so dass:
\begin{enumerate}[(i)]
\item $V$ ist Teilmenge von $\mathbb{R}^2$.
\item Jedes Element in $E$ ist ein Polygonzug zwischen zwei Knoten.
\item  Veschiedene Kanten haben verschiedene Menge von Endpunkten.
\item Das Innere einer Kante enthält keine Knote oder einen Punkt einer anderen Kante.
\end{enumerate}
Elemente von $V$ heißen \textit{Knoten}, Elemente von $E$ heißen \textit{Kanten}.
\end{definition}

\begin{remark}
\begin{itemize}
\item  Ein ebener Graph $G$ definiert in natürlicher Weise einen (abstrakten) Graphen, den wir ebenfalls mit $G$ bezeichnen.
\item Die unterliegende Punktmenge eines ebenen Graphens bezeichnen wir auch mit $G$:
\[ G = V \cup\bigcup_{e\in E}e \]
\end{itemize}
\end{remark}

\begin{definition}
\begin{itemize}
\item Sei $G = (V, E)$ ein ebener Graph. Die Zusammenhangskomponenten von $\mathbb{R}^2\setminus G$ heißen \textit{Gebiete} von $G$. 
\item Die Menge aller Gebiete von $G$ bezeichnen wir mit $F(G)$.
\item Sei $f\in F(G)$. Der Rand von $f$ bezeichnen wir mit $G[f]$. Wir können $G[f]$ als Teilgraph von $G$ auffassen.
\end{itemize}
\end{definition}

\begin{definition}
Sei $G$ ein ebener Graph.
\begin{itemize}
\item $G$ heißt \textit{maximal eben}, wenn durch das Hinzufügen einer Kante $G$ kein ebener Graph mehr ist.
\item $G$ heißt \textit{ebener Dreiecksgraph}, wenn jedes seiner Gebiete durch einen $K^3$ berandet ist.
\end{itemize}
\end{definition}

\begin{proposition}
Ein ebener Graph $G$ mit $|G|\geq 3$ ist genau dann maximal eben, wenn er ein ebener Dreiecksgraph ist.
\end{proposition}

\begin{proof}
Sei $G$ ein ebener Dreiecksgraph und $e$ eine zusätzliche Kante. Dann hat $e$ ihr Inneres in einem Gebiet $f$ von $G$ und ihre Endpunkte auf dem Rand von $f$. Per Definition ist $G[f] = K^3$ ein vollständiger Graph, also sind die Endpunkte von $e$ bereits in $G$ benachbart. Da Multikanten in einem ebenen Graphen nicht erlaubt sind, war $G$ schon maximal eben.

Sei nun umgekehrt $G$ ein maximal ebener Graph und $f$ ein Gebiet von $G$. Setze $H=G[f]$ und betrachte den induzierten Untergraphen $G[H]$. Angenommen, es gibt zwei Knoten $x,y$ in $G[H]$, die nicht benachbart sind. Aber dann könnten wir einen Polygonzug zwischen $x$ und $y$ in $f$ konstruieren und diese als ebene Kante zu $G$ hinzufügen, ein Widerspruch zur Maximalität von $G$. Also muss $G[H]$ vollständig sein.

Sei $n = |H|$. Angenommen, $H$ enthält keinen Kreis. Ist $n\geq 3$, so ist $K^3\subseteq G[H]\subseteq G$, d.h. $G$ enthält einen Kreis und daher $G\setminus H \neq\varnothing$. Für $n<3$ gilt auch $G\setminus H\neq\varnothing$, da $|G|\geq 3$. Andererseits ist $H$ ein Wald und hat daher genau einen Gebiet. $f$ ist ein Gebiet von $G[f] = H$, also ist $f$ das einzige Gebiet von $H$, d.h. $f\cup H = \mathbb{R}^2$. Insgesamt erhalten wir $G\setminus H \subseteq f$ und es folgt der Widerspruch:
\[ G\setminus H = G\setminus H \cap f \subseteq G\cap f =\varnothing \]
Also muss $H$ einen Kreis enthalten.

Es bleibt noch zu zeigen, dass $n\leq 3$ ist. Nehmen wir an, dass $n\geq 4$. Sei $C = v_1v_2v_3v_4v_1$ ein Kreis in $G[H]$. Wegen $C\subseteq G$ liegt $f$ in einem Gebiet $c\in F(C)$. Sei $c'\in F(C)$ das andere Gebiet. $v_1$ und $v_3$ liegen auf dem Rand von $f$. Wir können sie mit einem Polygonzug $P$ in $c$ verbinden, das sich mit $G$ nicht schneidet. Daher muss die ebene Kante zwischen $v_2$ und $v_4$ in $c'$ befinden, da sich diese mit $P$ nicht schneiden darf. Das gleiche Argument für $v_2$ und $v_4$ zeigt, dass die ebene Kante zwischen $v_1$ und $v_3$ in $c'$ befinden muss. Dies ein Widerspruch, da eine solche Kante mit der Kante zwischen $v_2$ und $v_4$ schneiden muss.
\end{proof}

\begin{corollary}
Ein ebener Graph $G$ der Ordnung $n\geq 3$ hat höchstens $3n-6$ Kanten.
\end{corollary}

\begin{proof}
Jeder ebene Dreiecksgraph mit $n$ Ecken hat $3n-6$ Kanten.
\end{proof}

\begin{corollary}\label{cor:kuratowski-motivation}
Kein ebener Graph enthält einen $K^5$ oder $K_{3,3}$ als einen topologischen Minor.
\end{corollary}

\begin{proof}
$K^5$ hat $10 > 3\cdot 5 - 6$ Kanten. Für $K_{3,3}$ kann man mithilfe der Euler-Charakteristik für ebene Graphen auch eine widersprüchliche Abschätzung für die Kantenzahl finden, siehe Korollar 3.2.11 in [Diestel: Graphentheorie]. Mit $K^5$ und $K_{3,3}$ können natürlich auch deren Unterteilungen nicht als ebene Graphen auftreten.
\end{proof}

\begin{proposition}
\begin{itemize}
\item Eine \textit{Einbettung in die Ebene} eines (abstrakten) Graphen $G$ ist ein abstrakter Graphenisomorphismus zwischen $G$ und einem ebenen Graphen $H$. 
\item $H$ nennen wir auch eine \textit{Zeichnung} von $G$.
\item Ein Graph $G$ heißt \textit{plättbar}, wenn es eine Einbettung in die Ebene für $G$ gibt.
\end{itemize}

\end{proposition}

\section{Satz von Kuratowski}

Interessanterweise gilt auch die Umkehrung von Korollar~\ref{cor:kuratowski-motivation}:

\begin{theorem}[\textit{Satz von Kuratowski, 1930}]
Die folgenden Aussagen sind für einen Graphen $G$ äquivalent:
\begin{enumerate}[(i)]
\item $G$ ist plättbar.
\item  $G$ enthält weder einen $K^5$ noch einen $K_{3,3}$ als Minor.
\item  $G$ enthält weder einen $K^5$ noch einen $K_{3,3}$ als topologischen Minor.
\end{enumerate}
\end{theorem}

Wir zeigen die Umkehrung zunächst für $3$-zusammenhängende Graphen. Dazu brauchen wir zunächst die folgenden Lemmata:

\begin{lemma}\label{lem:kuratowski-2lemma}
In einem $2$-zusammenhängenden ebenen Graphen ist jedes Gebiet durch einen Kreis berandet.
\end{lemma}

\begin{proof}
Siehe Lemma 3.2.6 in [Diestel: Graphentheorie].
\end{proof}

\begin{lemma}\label{lem:kuratowski-3lemma}
Ist $G$ $3$-zusammenhängend und $|G|>4$, so hat $G$ eine Kante $e$, so dass $G/e$ wieder $3$-zusammenhängend ist.
\end{lemma}

\begin{proof}
Siehe Lemma 2.2.1 in [Diestel: Graphentheorie].
\end{proof}

\begin{proposition}
Ist ein Graph $G$ $3$-zusammenhängend, und enthält $G$ weder einen $K^5$ noch einen $K_{3,3}$ als topologischen Minor, so ist $G$ plättbar.
\end{proposition}

\begin{proof}
Per Induktion nach $|G|$. Für $|G|=4$ ist $G = K^4$ und $G$ plättbar:
\[\begin{tikzcd}
& 1\ar[d, -]\ar[ldd, -]\ar[rdd, -] & K^4\\
& 2\ar[ld, -]\ar[rd, -] & \\
3\ar[rr, -] & & 4
\end{tikzcd}\]
Sei nun $|G|>4$ und die Aussage wahr für kleinere Graphen. Nach Lemma~\ref{lem:kuratowski-3lemma} haben wir eine Kante $xy$, so dass $G/xy$ wieder $3$-zusammenhängend ist. Nun enthält $G/xy$ ebenfalls weder $K^5$ noch einen $K_{3,3}$ als topologischen Minor. Nach Induktionsvoraussetzung ist $G/xy$ plättbar. Sei also $H$ eine Zeichnung von $G/xy$.

Sei $v$ der Knoten in $H$, das die Kante $xy$ repräsentiert. Betrachte das Gebiet $f$ von $H-v$, das den Punkt $v$ enthält und sei $C$ der Rand von $f$. Setze:
\[ X = N(x) \setminus{y},\quad Y = N(y)\setminus{x} \]
Dann gilt $X\cup Y\subseteq N(v)\subseteq C$. Da $H-v$ $2$-zusammenhängend ist, ist $C$ nach Lemma~\ref{lem:kuratowski-2lemma} ein Kreis. Seien $x_1,\ldots, x_k$ die Elemente in $X$ in natürlicher Reihenfolge entlang $C$ und $P_i$ der Verbindungsweg auf $C$ zwischen $x_i$ und $x_{i+1}$, wobei $x_{k+1} = x_1$. Betrachte den folgenden ebenen Graphen:
\[ H' = H - \{vw \mid w\in Y\setminus X\} \]
Wir können $H'$ auch als Zeichnung von $G-y$ deuten, indem wir den Knoten $v$ als $x$ auffassen. Ziel ist es nun, auch $y$ in der Zeichnung unterzubringen.

Dafür reicht es zu zeigen, dass ein $i$ existiert, so dass $Y\subseteq V(P_i)$. Dann können wir $y$ in dem Gebiet platzieren, der durch $x_i P_i x_{i+1} x x_i$ definiert ist. Angenommen, es existiert kein $i$ mit $Y\subseteq V(P_i)$. Wir unterscheiden drei Fälle:
\begin{enumerate}
\item Fall: Es gibt ein $y'\in Y\setminus X$. Sei etwa $y'\in P_i$ und $y''\in C\setminus P_i$ ein weiterer Nachbar von $y$. Setze $x'=x_i$ und $x''=x_{i+1}$. Dann werden $y'$ und $y''$ durch $x'$ und $x''$ in $C$ getrennt.
\item Fall: Es ist $Y\subseteq X$ und $|Y| \leq 2$, d.h. $y$ hat genau zwei Nachbarn $y'$ und $y''$ auf $C$, die nicht im gleichen $P_i$ liegen. Diese werden durch zwei $x',x''\in X$ in $C$ getrennt.
\item Fall: Es ist $Y\subseteq X$ und $|Y|\geq 3$.
\end{enumerate}
In den ersten beiden Fällen bilden $x,y',y''$ und $y,x',x''$ einen $\mathrm{T}K_{3,3}$ in $G$. Im dritten Fall haben $y$ und $x$ drei gemeinsame Nachbarn auf $C$. Diese bilden zusammen mit $x$ und $y$ einen $\mathrm{T}K^5$ in $G$.
\end{proof}

Um den Beweis von Satz von Kuratowski abzuschließen, muss man noch die folgenden Lemmata beweisen:

\begin{lemma}
Ein Graph enthält genau dann einen $\mathrm{T}K^5$ oder einen $\mathrm{T}K_{3,3}$, wenn er einen $K^5$ oder einen $K_{3,3}$ als Minor enthält.
\end{lemma}

\begin{proof}
Siehe Lemma 3.4.2 in [Diestel: Graphentheorie].
\end{proof}

\begin{lemma}
Ist $G$ ein Graph mit $|G|>4$, der kantenmaximal mit $\mathrm{T}K^5,\mathrm{T}K_{3,3}\not\subseteq G$ ist, so ist $G$ $3$-zusammenhängend.
\end{lemma}

\begin{proof}
Siehe Lemma 3.4.5 in [Diestel: Graphentheorie].
\end{proof}

\section{Algebraisches Plättbarkeitskriterium}

\end{document}

### 2016-07-05 Seminar Graphentheorie/article.tex
Add section 3 of graph theory seminar

@@
 
-\begin{theorem}[\textit{Satz von Kuratowski, 1930}]
+\begin{theorem}[\textit{Satz von Kuratowski, 1930}]\label{thm:kuratowski}
 Die folgenden Aussagen sind für einen Graphen $G$ äquivalent:
@@
 
+\begin{definition}
+Sei $G=(V, E)$ ein Graph.
+\begin{itemize}
+\item Der \textit{Kantenraum} von $G$ ist definiert als den $\mathbb{F}_2$-Vektorraum 
+\[\{\text{Abbildungen } h:E\to\mathbb{F}_2\}\] 
+mit komponentenweiser Addition.
+Wir identifizieren Vektoren darin mit Teilmengen von $E$.
+Somit ist die Addition von Teilmengen nichts anderes als das Bilden der symmetrischen Differenz der beiden Mengen:
+\[ E_1 + E_2 = (E_1 \cup E_2) \setminus (E_1\cap E_2) \]
+\item  Der \textit{Zyklenraum} $\mathcal{C}(G)$ von $G$ ist definiert als der $\mathbb{F}_2$-Untervektorraum des Kantenraums, der von den Kantenmengen von Kreisen in $G$ erzeugt wird.
+Vektoren in $\mathcal{C}(G)$ kann man als Summe von disjunkten Kreisen in $G$ schreiben.
+\item Eine Teilmenge $\mathcal{F}$ des Kantenraums von $G$ heißt \textit{schlicht}, wenn jede Kante in $G$ in höchstens zwei Mengen aus $\mathcal{F}$ liegt.
+\end{itemize}
+\end{definition}
+
+\begin{lemma}\label{lem:dimensionsformel}
+Sei $G$ ein zusammenhängender Graph mit $n$ Knoten und $m$ Kanten. Dann gilt für die Dimension des Zyklenraums:
+\[ \dim \mathcal{C}(G) = m-n+1 \]
+\end{lemma}
+
+\begin{proof}
+Siehe Satz 0.9.6 in [Diestel: Graphentheorie].
+\end{proof}
+
+\begin{theorem}[\textit{MacLane 1937}]\label{thm:maclane}
+Ein Graph $G$ ist genau dann plättbar, wenn sein Zyklenraum $\mathcal{C}(G)$ eine schlichte Basis besitzt.
+\end{theorem}
+
+\begin{proof}
+Für $|G|\leq 2$ ist die Aussage trivial.
+Sei $|G|\geq 3$.
+Sei $G$ zunächst einmal höchstens $1$-zusammenhängend, d.h.
+$G$ ist die Vereinigung zweier Untergraphen $G',G''\subset G$ mit $|G'\cap G''|\leq 1$.
+Ein Kreis in $G$ ist entweder ein Kreis in $G'$ oder ein Kreis in $G''$, also folgt:
+\[ \mathcal{C}(G) = \mathcal{C}(G')\oplus\mathcal{C}(G'') \]
+Somit hat $\mathcal{C}(G)$ genau dann eine schlichte Basis, wenn $\mathcal{C}(G')$ und $\mathcal{C}(G'')$ eine haben.
+Ferner ist $G$ genau dann plättbar, wenn $G'$ und $G''$ es sind.
+Somit folgt die Aussage induktiv.
+Sei ab jetzt $G$ $2$-zusammenhängend.
+
+Sei $G$ plättbar und wähle eine Zeichnung.
+Nach Lemma~\ref{lem:kuratowski-2lemma} sind alle Gebietsränder Kreise, liegen also in $\mathcal{C}(G)$.
+Wir zeigen, dass die Gebietsränder schon ganz $\mathcal{C}(G)$ erzeugen.
+Da eine ebene Kante auf dem Rand höchstens zweier Gebiete liegen kann, besitzt $\mathcal{C}(G)$ dann eine schlichte Basis.
+
+Sei $C\subset G$ ein Kreis und $f$ sein Innengebiet.
+Jede Kante $e$ liegt auf einem Kreis von $G$.
+Liegt das Innere von $e$ in $f$, so liegt $e$ auf dem Rand genau zweier Gebiete von $G$, die in $f$ enthalten sind.
+Liegt $e$ dagegen auf $C$, so liegt $e$ genau auf einem Gebiet, das in $f$ enthalten ist.
+Somit gilt:
+\[ C = \sum_{\substack{f'\in F(G)\\ f'\subseteq f}} E(G[f']) \]
+
+Sei nun umgekehrt $\{C_1,\ldots, C_k\}$ eine schlichte Basis von $\mathcal{C}(G)$. Für jede Kante $e$ besitzt $\mathcal{C}(G-e)$ auch eine schlichte Basis, denn:
+\begin{itemize}
+\item Liegt $e$ in nur einem Basiszyklus, etwa $C_1$, so ist $\{C_2,\ldots, C_k\}$ eine Basis von $\mathcal{C}(G-e)$.
+\item Liegt $e$ in zwei Basiszyklen, etwa $C_1$ und $C_2$, so ist $\{C_1+C_2,\ldots, C_k\}$ eine Basis von $\mathcal{C}(G-e)$.
+\end{itemize}
+Angenommen, $G$ ist nicht plättbar.
+Nach dem Satz von Kuratowski enthält $G$ einen $\mathrm{T}K^5$ oder einen $\mathrm{T}K_{3,3}$.
+Als Teilgraphen  von $G$ hat dann $\mathcal{C}(\mathrm{T}K^5)$ bzw.
+$\mathcal{C}(\mathrm{T}K_{3,3})$ eine schlichte Basis.
+Da topologische Minoren keine Kreise hinzufügen bzw.
+entfernen bleibt der Zyklenraum gleich, d.h.
+$\mathcal{C}(K^5)$ bzw.
+$\mathcal{C}(K_{3,3})$ hat eine schlichte Basis.
+Wir führen nun beide Fälle in den folgenden Lemmata zum Widerspruch.
+\end{proof}
+
+\begin{lemma}
+$\mathcal{C}(K^5)$ hat keine schlichte Basis.
+\end{lemma}
+
+\begin{proof}
+Nach der Dimensionsformel Lemma~\ref{lem:dimensionsformel} gilt $\dim\mathcal{C}(K^5) = 6$. Angenommen, $\mathcal{C}(K^5)$ habe eine schlichte Basis $B=\{C_1,\ldots,C_6\}$. Setze:
+\[ C_0 = C_1+\ldots + C_6 \]
+Keines der $C_0, C_1,\ldots, C_6$ sind leer und enthalten alle mindestens drei Kanten.
+Da eine Kante in höchstens zwei Zyklen aus $B$ liegt, liegt jede Kante in $C_0$ in nur einem der Basiszyklen in $B$.
+Daher ist die Menge $\{C_0,C_1,\ldots ,C_6\}$ ebenfalls schlicht, also folgt:
+\[ 21 = 7\cdot 3 \leq |C_0|+\ldots + |C_6| \leq 2\cdot \|K^5\| = 20 \qedhere \]
+\end{proof}
+
+\begin{lemma}
+$\mathcal{C}(K_{3,3})$ hat keine schlichte Basis.
+\end{lemma}
+
+\begin{proof}
+Nach der Dimensionsformel Lemma~\ref{lem:dimensionsformel} gilt $\dim\mathcal{C}(K_{3,3}) = 4$. Angenommen,  $\mathcal{C}(K_{3,3})$ habe eine schlichte Basis $B=\{C_1,\ldots, C_4\}$. Setze:
+\[ C_0 = C_1+\ldots + C_4 \]
+Keines der $C_0,C_1,\ldots,C_4$ sind leer und enthalten alle mindestens vier Kanten wegen der Bipartität.
+Mit dem gleichen Argument wie im vorherigen Lemma ist $\{C_0,C_1,\ldots,C_4\}$ ebenfalls schlicht, also folgt:
+\[ 20 = 5\cdot 4 \leq |C_0|+\ldots + |C_4|\leq 2\cdot \|K_{3,3}\| = 18\qedhere \]
+\end{proof}
+
 \end{document}

### 2016-07-05 Seminar Graphentheorie/article.tex
Add final section of graph theory seminar

@@
 \begin{definition}
-Ein \textit{ebener Graph} ist ein Paar $G= (V, E)$ endlicher Mengen, so dass:
+Ein \textit{ebener Graph} ist ein Paar $G= (V, E)$ endlicher Mengen (Elemente von $V$ heißen \textit{Knoten}, Elemente von $E$ heißen \textit{Kanten}), so dass:
 \begin{enumerate}[(i)]
@@
 \end{enumerate}
-Elemente von $V$ heißen \textit{Knoten}, Elemente von $E$ heißen \textit{Kanten}.
 \end{definition}
@@
 \[ E_1 + E_2 = (E_1 \cup E_2) \setminus (E_1\cap E_2) \]
+\item Der \textit{Schnittraum} $\mathcal{C}^\ast(G)$ von $G$ ist definiert als der $\mathbb{F}_2$-Untervektorraum des Kantenraums, der nur aus den Schnittmengen in $G$ besteht, d.h.
+Mengen der Form $E(V', V'')$ für eine Partition $\{V', V''\}$ in $G$.
 \item  Der \textit{Zyklenraum} $\mathcal{C}(G)$ von $G$ ist definiert als der $\mathbb{F}_2$-Untervektorraum des Kantenraums, der von den Kantenmengen von Kreisen in $G$ erzeugt wird.
@@
 
+\section{Plättbarkeit \& Dualität}
+
+\begin{definition}
+Ein \textit{ebener Multigraph} ist ein Paar $G=(V,E)$ endlicher Mengen (Elemente von $V$ heißen \textit{Knoten}, Elemente von $E$ heißen \textit{Kanten}), so dass:
+\begin{enumerate}[(i)]
+\item $V$ ist Teilmenge von $\mathbb{R}^2$.
+\item Jedes Element in $E$ ist ein Polygonzug zwischen zwei Knoten oder ein Polygon, das genau eine Knote enthält.
+\item Das Innere einer Kante enthält keine Knote oder einen Punkt einer anderen Kante.
+\end{enumerate}
+\end{definition}
+
+\begin{definition}
+Sei $G=(V, E)$ ein ebener Multigraph. Wir setzen in jedes ebiet von $G$ einen neuen Knoten und verbinden diese zu einem neuen ebenen Multigraphen $G^\ast$:
+\begin{itemize}
+\item Für jede Kante $e$ von $G$ verbinden wir die neuen Knoten in den beiden Gebieten, auf deren Rand $e$ liegt, durch eine neue Kante $e^\ast$.
+\item Liegt $e$ nur auf dem Rand eines Gebiets, so legen wir an dessen neue Knote eine Schlinge $e^\ast$ durch $e$.
+\end{itemize}
+Der neue Graph $G^\ast$ heißt \textit{topologisches Dual} zu $G$.
+\end{definition}
+
+Wir finden den folgenden einfachen Zusammenhang zwischen einem Graphen und sein topologisches Dual:
+
+\begin{proposition}\label{prop:kombdual-motivation}
+Sei $G$ ein zusammenhängender ebener Multigraph und $E\subseteq E(G)$ eine Kantenmenge.
+$E$ ist genau dann von einem Kreis induziert, wenn $E^\ast = \{e^\ast\mid e\in E\}$ ein minimaler Schnitt in $G^\ast$ ist.
+\end{proposition}
+
+\begin{proof}
+Siehe Proposition 3.6.1 in [Diestel: Graphentheorie].
+\end{proof}
+
+\begin{definition}
+Sei $G$ ein abstrakter Multigraph.
+Ein abstrakter Multigraph $G^\ast$ heißt zu $G$ \textit{kombinatorisch dual}, wenn $E(G^\ast) = E(G)$ und die Minimalschnitte von $G^\ast$ gerade die Kantenmengen von Kreise in $G$ sind.
+\end{definition}
+
+\begin{proposition}\label{prop:kombdual-zyklenraum}
+Sei $G^\ast$ kombinatorisch dual zu $G$. Dann gilt:
+\[ \mathcal{C}^\ast(G^\ast) =\mathcal{C}(G) \]
+\end{proposition}
+
+\begin{proof}
+$\mathcal{C}^\ast(G^\ast)$ wird erzeugt durch die Minimalschnitte von $G^\ast$, während $\mathcal{C}(G)$ von den Kantenmengen von Kreise in $G$ erzeugt wird.
+\end{proof}
+
+\begin{lemma}\label{lem:schnittraum-schlicht}
+Sei $G=(V, E)$ ein Graph. Dann wird $\mathcal{C}^\ast(G)$ erzeugt von Schnitten der Form $E(v) = \{vw\mid w\in V\setminus\{v\} \},\ v\in V$.
+\end{lemma}
+
+\begin{proof}
+Sei $\{V', V''\}$ eine Partition von $G$ und betrachte den Schnitt $E(V', V'')$. Da jede Kante $vw$ in genau $E(v)$ und $E(w)$ liegt, gilt:
+\[ E(V', V'') = \sum_{v\in V'} E(v) \qedhere \]
+\end{proof}
+
+\begin{corollary}\label{cor:schnittraum-schlicht}
+Sei $G$ ein Graph. Dann hat $\mathcal{C}^\ast(G)$ eine schlichte Basis.
+\end{corollary}
+
+\begin{theorem}[\textit{Whitney 1993}]
+Ein Graph $G$ ist genau dann plättbar, wenn ein zu ihm kombinatorisch dualer Multigraph existiert.
+\end{theorem}
+
+\begin{proof}
+Eine Richtung folgt aus Satz~\ref{prop:kombdual-motivation}.
+Sei $G^\ast$ ein kombinatorisches Dual zu $G$.
+Nach Satz~\ref{prop:kombdual-zyklenraum} gilt $\mathcal{C}^\ast(G^\ast) =\mathcal{C}(G)$.
+Nach MacLane Theorem~\ref{thm:maclane} reicht es zu zeigen, dass $\mathcal{C}^\ast(G^\ast)$ eine schlichte Basis besitzt.
+Dies folgt aus Korollar~\ref{cor:schnittraum-schlicht}.
+\end{proof}
+
 \end{document}

### 2016-07-05 Seminar Graphentheorie/article.tex
Fix structure
Fix definition specified as proposition
Fix punctuation

@@
 \begin{definition}
+Sei $G = (V, E)$ ein ebener Graph. 
 \begin{itemize}
-\item Sei $G = (V, E)$ ein ebener Graph. Die Zusammenhangskomponenten von $\mathbb{R}^2\setminus G$ heißen \textit{Gebiete} von $G$. 
+\item Die Zusammenhangskomponenten von $\mathbb{R}^2\setminus G$ heißen \textit{Gebiete} von $G$. 
 \item Die Menge aller Gebiete von $G$ bezeichnen wir mit $F(G)$.
@@
 
-\begin{proposition}
+\begin{definition}
 \begin{itemize}
@@
 
-\end{proposition}
+\end{definition}
 
@@
 
-Sei $v$ der Knoten in $H$, das die Kante $xy$ repräsentiert. Betrachte das Gebiet $f$ von $H-v$, das den Punkt $v$ enthält und sei $C$ der Rand von $f$. Setze:
+Sei $v$ der Knoten in $H$, das die Kante $xy$ repräsentiert. Betrachte das Gebiet $f$ von $H-v$, das den Punkt $v$ enthält, und sei $C$ der Rand von $f$. Setze:
 \[ X = N(x) \setminus{y},\quad Y = N(y)\setminus{x} \]

### 2016-07-05 Seminar Graphentheorie/article.tex
Fix typo
Fix definition of E(v)

@@
 \item Jedes Element in $E$ ist ein Polygonzug zwischen zwei Knoten.
-\item  Veschiedene Kanten haben verschiedene Menge von Endpunkten.
+\item Verschiedene Kanten haben verschiedene Mengen von Endpunkten.
 \item Das Innere einer Kante enthält keine Knote oder einen Punkt einer anderen Kante.
@@
 \begin{lemma}\label{lem:schnittraum-schlicht}
-Sei $G=(V, E)$ ein Graph. Dann wird $\mathcal{C}^\ast(G)$ erzeugt von Schnitten der Form $E(v) = \{vw\mid w\in V\setminus\{v\} \},\ v\in V$.
+Sei $G=(V, E)$ ein Graph. Dann wird $\mathcal{C}^\ast(G)$ erzeugt von Schnitten der Form $E(v) = \{vw\mid w\in N(v) \},\ v\in V$.
 \end{lemma}

### 2016-07-25 Seminar Graphentheorie/article.tex
Add link to presentation
Fix year of Whitney's theorem

@@
 \usepackage{tikz-cd}
+\usepackage{hyperref}
+\hypersetup{
+	bookmarksopen=true,
+	pdfstartview=FitH
+}
 \linespread{1.1}
@@
 \maketitle
+
+\paragraph{Präsentation} \url{http://slides.com/yichuanshen/ebene-graphen/}
 
@@
 \item Der \textit{Schnittraum} $\mathcal{C}^\ast(G)$ von $G$ ist definiert als der $\mathbb{F}_2$-Untervektorraum des Kantenraums, der nur aus den Schnittmengen in $G$ besteht, d.h.
-Mengen der Form $E(V', V'')$ für eine Partition $\{V', V''\}$ in $G$.
+Mengen der Form $E(V', V'')$ wobei $\{V', V''\}$ eine Partition von $G$ ist.
 \item  Der \textit{Zyklenraum} $\mathcal{C}(G)$ von $G$ ist definiert als der $\mathbb{F}_2$-Untervektorraum des Kantenraums, der von den Kantenmengen von Kreisen in $G$ erzeugt wird.
@@
 
-\begin{theorem}[\textit{Whitney 1993}]
+\begin{theorem}[\textit{Whitney 1933}]
 Ein Graph $G$ ist genau dann plättbar, wenn ein zu ihm kombinatorisch dualer Multigraph existiert.

### 2016-11-04 Master Thesis/synopsis.tex
Some typos

@@
 
-\begin{lemma}
+\begin{lemma}\label{2.2-pre}
 Let $v\in\operatorname{Spv}(K)$. There is a canonical bijection:
@@
 $G_K\neq 1$ and set $v_i = v(w_i)$.
+By Lemma~\ref{2.2-pre} $K$ is henselian with respect to both $v_1$ and $v_2$.
 If $v_1\neq v_2$, by the Theorem of F.K.
@@
 \begin{proof}
-The degree $[K(\mu_\ell, \sqrt{-1}):K]$ is prime to $\ell$, thus we can assume that $\mu_\ell\subset K$ and $K$ is not formally real without loss of generality.
+The degree $[K(\mu_\ell, \sqrt{-1}):K]$ is prime to $\ell$, thus we can assume that $\mu_\ell\subset K$ and $K$ is not formally real with a usual norm argument.
 Now we can replace $\mathbb{Z}/\ell\mathbb{Z}$ by $\mathbb{Z}/\ell\mathbb{Z}(2) = \mu_\ell^{\otimes 2}$.
@@
 
-Le $(K_0)_{w_1}$ be a henselization of $K_0$ with respect to $w_1$ and let $K_w$ be a henselization of $K$ with respect to $w$ in $(K_0)_{w_1}$.
+Let $(K_0)_{w_1}$ be a henselization of $K_0$ with respect to $w_1$ and let $K_w$ be a henselization of $K$ with respect to $w$ in $(K_0)_{w_1}$.
 Then $(K_0)_{w_1}$ is the composite of $K_0$ and $K_w$ and we have $\ell^\infty\nmid [(K_0)_{w_1} : K_w]$.
 
-Since $K_0$ is henselian with respect to $w_0$, a conjugate $\tau(K_w)$ for some $\tau\in G_K$ of $K_w$ lies in $K_0$, hence $\ell^\infty\nmid [K_w : K]$.
+Since $K_0$ is henselian with respect to $w_0$, a conjugate $\tau(K_w)$ for some $\tau\in G_K$ of $K_w$ lies in $K_0$, hence $\ell^\infty\nmid [K_w : K]$ as well.
 Consequently, $\ell^\infty \nmid [(K_0)_{w_1} : K]$ and therefore $\ell^\infty \nmid [(K_0)_{w_1} : K_0]$.
@@
 The uniqueness of $w$ is again a consequence of Lemma \ref{2.2}.
-\item Choose a prime $\ell$ different from $\operatorname{char} k(w)$. According to Lemma \ref{2.5} we have $\operatorname{rrank}(w)\geq \dim_{\mathbb{F}_\ell}(\Gamma_w /\ell\Gamma_w) = 2 = \dim(K)$.
+\item Choose the prime $\ell$ different from $\operatorname{char} k(w)$.
+According to Lemma \ref{2.5} we have $\operatorname{rrank}(w)=\dim_\mathbb{Q}(\Gamma_w\otimes_\mathbb{Z}\mathbb{Q} )\geq \dim_{\mathbb{F}_\ell}(\Gamma_w /\ell\Gamma_w) = 2 = \dim(K)$.
 \end{enumerate}

### 2016-11-26 Master Thesis/synopsis.tex
Update title
Proof of 4.2

@@
 \end{align*}
-Valuations in $\operatorname{Spv}_\text{d}$ are called \textit{defectless}\index{valuation!defectless}, those in $\operatorname{Spv}_\text{rd}$ are called \textit{rank-defectless}\index{valuation!rank-defectless}.
+Valuations in $\operatorname{Spv}_\text{d}$ are called \textit{defectless}\index{valuation!defectless}, whereas those in $\operatorname{Spv}_\text{rd}$ are called \textit{rank-defectless}\index{valuation!rank-defectless}.
 For $\ast\in\{\text{d},\text{rd}\}$ we set:
@@
 Then $E_0/E$, and also $E'_0/E'$, satisfies the condition of \ref{2.6} (ii).
-Thus there exists a unique henselian defectless valuation $w'_{E'}$ of $E'$, so $E'$ contains a henselization $K'_{w'}$ of $w'_{E'}|_{K'}$.
+Thus there exists a unique henselian defectless valuation $w'_{E'}$ of $E'$, so $E'$ contains a henselization $K'_{w'}$ of $w' = w'_{E'}|_{K'}$.
 By applying the same argument to $\sigma^{-1}$ we find that $E' = K'_{w'}$.
@@
 
+\begin{theorem}[\textit{Hilbert's Irreducibility Theorem}]\label{hilberts-irreducibility-theorem}
+Inhalt...
+\end{theorem}
+
 \begin{lemma}\label{4.1}
@@
 
+\begin{proof}
+Let $f: X\to\mathbf{P}^1_E$ be a non-constant morphism and let $f' =  f\circ \pi: X'\to\mathbf{P}^1_E$.
+By Hilbert's Irreducibility Theorem \ref{hilberts-irreducibility-theorem} there exists a closed $E$-rational point $P$ of $\mathbf{P}^1_E$ which has only one pre-image under $f'$.
+Consequently, $P$ only has one pre-image under $f$, say $P'$, and $P'$ has only one pre-image under $\pi$.
+\end{proof}
+
 \paragraph{} Let $k$ and $k'$ be the algebraic closures of $\mathbb{Q}$ in $K$ and $K'$ respectively, i.e.
@@
 We identify the set of closed points $X_0$ of $X$ with the set of valuations $v\in\operatorname{Spv}_\text{d}^{(1)}(K)$ with $\operatorname{char} k(v)=0$ and similarly identify $X'_0$ with the corresponding subset of $\operatorname{Spv}_\text{d}^{(1)}(K')$.
+
+\begin{lemma}\label{4.2-pre}
+We have $[k:\mathbb{Q}] = [k':\mathbb{Q}]$.
+\end{lemma}
+
+\begin{proof}
+By Lemma \ref{3.2} $\sigma$ induces a bijection $\sigma: X_0\to X'_0$ and for any $P\in X_0$ an isomorphism $\varphi_P: k(\sigma(P))\to k(P)$.
+
+First we suppose that $k/\mathbb{Q}$ is a Galois extension and $X$ has a $k$-rational point $P_0$. We can consider $k'$ as a subfield of $k$ via: 
+\[\begin{tikzcd}
+k'\ar[r] & k(\sigma(P_0))\ar[r, "\varphi_{P_0}"', "\cong"] & k(P_0) \ar[r, "\cong"] & k
+\end{tikzcd}\]
+For all $P' = \sigma(P)\in X'_0$ we also get an embedding:
+\[ \begin{tikzcd}
+k \ar[r] & k(P) \ar[r, "\varphi_P^{-1}"', "\cong"] & k(P')
+\end{tikzcd} \]
+We see any embedding $k\to \overline{k(P')}$ maps $k$ into $k(P')$ because $k/\mathbb{Q}$ is Galois. Thus we have for all $P'\in X'_0$:
+\[ k\otimes_{k'} k(P')\cong \prod_{\operatorname{Hom}_{k'}(k, k(P'))} k(P') \]
+in which case $X'_k\to X'$ splits completely. By Lemma \ref{4.1} we get $k\cong k'$, hence $[k:\mathbb{Q}]= [k':\mathbb{Q}]$.
+
+Now we drop the assumption that $k/\mathbb{Q}$ is Galois.
+There exists a finite extension $k_1/k$ such that $k_1/\mathbb{Q}$ is Galois and $X(k_1)\neq\varnothing$.
+Let $K_1 = Kk_1$ be the constant field extension of $K$ with $k_1$ and let $k'_1$ be the constant field of $K'_1$.
+Applying the above argument to $K_1$ and $K'_1$ yields $[k_1:\mathbb{Q}] = [k'_1:\mathbb{Q}]$.
+Therefore:
+\begin{align*}
+[k:\mathbb{Q}] &= [k_1:\mathbb{Q}]\cdot [K_1:K]^{-1} = [k'_1:\mathbb{Q}]\cdot [K'_1:K']^{-1}\\
+&\leq [k'_1:\mathbb{Q}]\cdot [k'_1:k']^{-1} = [k':\mathbb{Q}]
+\end{align*}
+By symmetry, we get $[k:\mathbb{Q}] = [k':\mathbb{Q}]$.
+\end{proof}
 
@@
 \end{lemma}
+
+\begin{proof}
+Let $L/K$ be a finite subextension of $\overline{K}/K$. Let $\ell$ and $\ell'$ be the constant fields of $L$ and $L'$ respectively. Applying Lemma \ref{4.2-pre} to $K$ and $L$ gives:
+\[ [\ell:k] = [\ell':k'] \]
+Therefore $L/K$ is a constant field extension iff $L'/K'$ is a constant field extension.
+Since $K\overline{k}/K$ and $K'\overline{k'}/K$ are the union of their finite subextensions, we obtain $(K\overline{k})' = K'\overline{k'}$.
+This implies that $\sigma$ induces an isomorphism $G_k\to G_{k'}$.
+Everything else follows from Neukirch-Uchida \ref{thm:neukirch-uchida}.
+\end{proof}
 

### 2016-11-26 Master Thesis/synopsis.tex
Clarify inequality in 4.2-pre
Clarify isomorphism in 4.2

@@
 [k:\mathbb{Q}] &= [k_1:\mathbb{Q}]\cdot [K_1:K]^{-1} = [k'_1:\mathbb{Q}]\cdot [K'_1:K']^{-1}\\
-&\leq [k'_1:\mathbb{Q}]\cdot [k'_1:k']^{-1} = [k':\mathbb{Q}]
+&= [k'_1:\mathbb{Q}]\cdot ([K'_1: K'k_1'] \cdot [K'k_1':K'])^{-1}\\
+&= [k'_1:\mathbb{Q}]\cdot  ([K'_1: K'k_1'] \cdot [k'_1:k'])^{-1} \leq [k'_1:\mathbb{Q}]\cdot [k'_1:k']^{-1} = [k':\mathbb{Q}]
 \end{align*}
@@
 \[ \begin{tikzcd}
-G_K \ar[r, "\sigma"] \ar[d, "\text{res}"'] & G_{K'}\ar[d, "\text{res}"]\\
+G_K \ar[r, "\sigma"] \ar[d] & G_{K'}\ar[d]\\
 G_k \ar[r, "\sigma"'] & G_{k'}
@@
 Since $K\overline{k}/K$ and $K'\overline{k'}/K$ are the union of their finite subextensions, we obtain $(K\overline{k})' = K'\overline{k'}$.
-This implies that $\sigma$ induces an isomorphism $G_k\to G_{k'}$.
+This implies that $\sigma$ induces an isomorphism:
+\[G_k \cong \operatorname{Gal}(K\overline{k}/K) \to \operatorname{Gal}(K'\overline{k'}/K')\cong G_{k'}\] 
 Everything else follows from Neukirch-Uchida \ref{thm:neukirch-uchida}.

### 2016-11-26 Master Thesis/synopsis.tex
Add remark after 4.2

@@
 
+\begin{remark}
+Consider $\overline{K}'' = \overline{K}'\otimes_{\overline{k'}}\overline{k}$ via the isomorphism $\varphi: \overline{k'}\to \overline{k}$. If we replace $\sigma: G_K\to G_{K'}$ with the composition
+\[ \begin{tikzcd}
+G_K \ar[r, "\sigma"] & G_{K'} \ar[r] & G_{K'}
+\end{tikzcd} \]
+where the second map is induced by the isomorphism $\overline{K''}\to\overline{K'},\ x\otimes y\mapsto \varphi^{-1}(y)\cdot x$, we may assume that $k=k'$, $\overline{k} = \overline{k'}$ and $\varphi:\overline{k}\to\overline{k}$, $\sigma: G_k\to G_k$ are identities.
+\end{remark}
+
 \paragraph{} Let $n\in\mathbb{N}$. For all intermediate fields $L$ of $\overline{K}/K$, $\sigma$ induces an isomorphism:

### 2016-12-07 Master Thesis/synopsis.tex
Remark on Kronecker dimension
Add some proof for 4.3

@@
 \end{definition}
+
+\begin{remark}
+A one-dimensional finitely generated field over its prime field is just a global field.
+A two-dimensional finitely generated field over $\mathbb{Q}$ is a function field in one variable over $\mathbb{Q}$.
+\end{remark}
 
@@
 
-\paragraph{} Let $n\in\mathbb{N}$. For all intermediate fields $L$ of $\overline{K}/K$, $\sigma$ induces an isomorphism:
+\paragraph{} Let $n\in\mathbb{N}$. In the previous remark, we have determined that the following diagram commutes: 
+\[ \begin{tikzcd}
+G_K \ar[r]\ar[d, "\sigma"'] & G_k\\
+G_{K'}\ar[ru]
+\end{tikzcd} \]
+Now the operation of $G_K$ and $G_{K'}$ on $\mu_n$ factors through $G_k$, thus we have $\sigma(g)\xi = g\xi$ for all $\xi\in\mu_n$ and $g\in G_K$.
+Therefore, for all intermediate fields $L$ of $\overline{K}/K$, $\sigma$ induces an isomorphism:
 \[ \varphi_n:L'^\times/n \cong\mathrm{H}^1(L', \mu_n) \to \mathrm{H}^1(L,\mu_n) \cong L^\times/n \]
@@
 \end{lemma}
+
+\begin{proof}
+\begin{enumerate}[(i)]
+\item  We have $(K^\text{t}_P)^\times/n\cong \mathbb{Z}/n\mathbb{Z}$ since $\operatorname{char}k(P) = 0$.
+The canonical map $K^\times\to (K^\text{t}_P)^\times/n$ can be identified with the map $K^\times /n\to\mathbb{Z}/n$ induced by $v_P$.
+Now the following commutative diagram gives us the result:
+\[ \begin{tikzcd}
+0 \ar[d] & 0 \ar[d]\\
+U_{P'}/n \ar[r, dashed]\ar[d] & U_P/n\ar[d]\\
+K'^\times /n \ar[r, "\varphi_n"]\ar[d] & K^\times/n\ar[d]\\
+(K^\text{t}_P)'^\times /n \ar[r, "\varphi_n"'] &  (K^\text{t}_P)^\times /n
+\end{tikzcd} \]
+\item Consider the field extension $K_P/K$
+\end{enumerate}
+\end{proof}
 

### 2016-12-14 Master Thesis/synopsis.tex
Add proof for 4.3.4
Add proof for 4.5

@@
 
-\section{Some Results on $\operatorname{Spv}(K)$}
+\section{Some Results on the valuation spectra}
 
@@
 
+By Lemma \ref{3.2} $\sigma$ induces a bijection $\sigma: X_0\to X'_0$ and for any $P\in X_0$ an isomorphism $\varphi_P: k(\sigma(P))\to k(P)$.
+
 \begin{lemma}\label{4.2-pre}
@@
 \begin{proof}
-By Lemma \ref{3.2} $\sigma$ induces a bijection $\sigma: X_0\to X'_0$ and for any $P\in X_0$ an isomorphism $\varphi_P: k(\sigma(P))\to k(P)$.
-
 First we suppose that $k/\mathbb{Q}$ is a Galois extension and $X$ has a $k$-rational point $P_0$. We can consider $k'$ as a subfield of $k$ via: 
@@
 We denote the natural projection $U_P\to k(P)^\times,\ x\mapsto x\mod P$ by $r_P$.
-For $n\in\mathbb{N}$ the following sequence is exact, since valuation rings are integrally closed:
+
+Let $n\in\mathbb{N}$. Tensoring the exact sequence $0\to U_P\to K^\times \to\mathbb{Z}\to 0$ with $\mathbb{Z}/n\mathbb{Z}$ preserves its exactness since $\mathbb{Z}$ is $\mathbb{Z}$-flat:
 \[ 0\longrightarrow U_P/n\longrightarrow K^\times/n\longrightarrow\mathbb{Z}/n\mathbb{Z}\longrightarrow 0 \]
@@
 (K^\text{t}_P)'^\times /n \ar[r, "\varphi_n"'] &  (K^\text{t}_P)^\times /n
-\end{tikzcd} \]
-\item Consider the field extension $K_P/K$
+\end{tikzcd}\]
+\item Consider the field extension $K_P/K$. Without loss of generality, we can assume $K=K_P$. Then $r_P/n$ is an isomorphism and the composition
+\[\begin{tikzcd}
+k(P)^\times / n\ar[rr, "(r_P/n)^{-1}"] && U_P/n \ar[r, hook] & K_P^\times/n
+\end{tikzcd}\]
+is the inflation $\mathrm{H}^1(k(P), \mu_n) \to \mathrm{H}^1(K_P,\mu_n)$.
+The same holds for $P'$ due to Lemma \ref{3.2}.
+Therefore, the assertion follows from the commutativity of the following diagram:
+\[ \begin{tikzcd}
+k(P')^\times/n \ar[r, "\cong"] \ar[d, "\varphi_P/n"'] & \mathrm{H}^1(k(P'),\mu_n) \ar[r, "\mathrm{inf}"] \ar[d]& \mathrm{H}^1(K'_{P'}, \mu_n) \ar[d, "\varphi_n"] \\
+k(P)^\times/n \ar[r, "\cong"'] & \mathrm{H}^1(k(P),\mu_n)  \ar[r, "\mathrm{inf}"'] & \mathrm{H}^1(K_{P}, \mu_n)
+\end{tikzcd}  \qedhere\]
 \end{enumerate}
 \end{proof}
+
+\begin{theorem}[\textit{Mordell-Weil Theorem}]\label{mordell-weil-theorem}
+Inhalt...
+\end{theorem}
 
@@
 \end{lemma}
+
+\begin{proof}
+\begin{enumerate}[(i)]
+\item By Mordell-Weil Theorem \ref{mordell-weil-theorem} the Picard group of $X$ is finitely generated.
+Let $D_1,\ldots, D_r$ be Weil divisors of $X$ such that the associated line bundles generate $\operatorname{Pic}(X)$ and similarly let $D'_1,\ldots,D'_s$ generate $\operatorname{Pic}(X')$.
+Clearly any finite $S\subset X_0$ such that
+\[ \bigcup_{i=1}^r \operatorname{supp}D_i\subset S\quad\text{and}\quad \bigcup_{j=1}^s\operatorname{supp}D'_j\subset S' \]
+is admissible. Now let $V\subset X$ be an open subscheme. By the approximation theorem, we can choose the $D_i$ and the $D'_j$ in $V_0$ and $\sigma (V_0)$ respectively.
+\item Observe that the following sequence is exact:
+\[ \begin{tikzcd}
+0 \ar[r] & K^S \ar[r]& K^\times \ar[r, "\operatorname{div}"]& \displaystyle \bigoplus_{P\in X_0\setminus S}\mathbb{Z} \ar[r]& 0
+\end{tikzcd} \]
+Since the rightmost module is $\mathbb{Z}$-flat, tensoring with $\mathbb{Z}/n$ yields another exact sequence and we can consider $K^S/n$ as a subgroup of $K^\times/n$. Then:
+\[ K^S/n = \bigcap_{P\in X_0\setminus X}U_P/n \]
+The analogous result holds for $K'^{S'}/n$. Thus the assertion follows from Lemma \ref{4.3} (i).\qedhere
+\end{enumerate}
+\end{proof}
 

### 2016-12-14 Master Thesis/synopsis.tex
Comment out 2.3

@@
 
+\iffalse
 \begin{lemma}\phantomsection\label{2.3}
@@
 \end{lemma}
+\fi
 
@@
 \begin{definition}
-For each admissible $S\subset X_0$ and $T_1\subset X_0\setminus S$ with injective $r_{S,T_1}$ we also have the injectivity of $r_{S,T_2}$ for every $T_2 \supset T_1$.
+For each admissible $S\subset X_0$ and $T_1\subset X_0\setminus S$ with injective $r_{S,T_1}$ we also have the injectivity of $r_{S,T_2}$ for every $T_1 \subset T_2$.
 Therefore,  we can find a finite subset $T\subset X_0\setminus S$ such that both $r_{S,T}$ and $r_{S',T'}$ are injective.

### 2016-12-17 Master Thesis/synopsis.tex
Add proof for 5.2 (iii)

@@
 \item Let $f: A\to B$ be a homomorphism of abelian groups with finite kernel and cokernel. Then $A$ is free-by-finite iff $B$ is free-by-finite.
-\item Let $0\to A'\stackrel{i}{\to}A\stackrel{p}{\to}A''\to 0$ be an exact sequence of abelian groups.
+\item Let $0\longrightarrow A'\stackrel{i}{\longrightarrow}A\stackrel{p}{\longrightarrow}A''\longrightarrow 0$ be an exact sequence of abelian groups.
 If $A'$ and $A''$ are free-by-finite, then $A$ is free-by-finite.
@@
 \end{lemma}
+
+\begin{proof}
+\begin{enumerate}[(i)]
+\item Let $A$ be free-by-finite, i.e. $A = A' \oplus F$ for some finite group $A'$ and free group $F$. Since $\ker(f)\subset A$ is finite, it's actually contained in $A'$.
+
+Now let $B$ be free-by-finite and $K = \ker(f)$.
+Since the injective map $A/K \to B$ induced by $f$ can be restricted to their torsion subgroups, we see that $(A/K)_\text{tor}$ is finite.
+Every $a\in A_\text{tor}$ can be written as $a = a' + k$ for some $a'\in A,\ k\in K$ such that $na'\in K$ for some $n\in\mathbb{N}$.
+Since both $(A/K)_\text{tor}$ and $K$ are finite, $A_\text{tor}$ is finite as well.
+TODO
+\item TODO
+\item It suffices to show this assertion for finite groups and for $A=\mathbb{Z}$.
+For $A$ finite we have $nA = 0$ for $n = \#A$.
+Since $\mathbb{Z}$ has no nilpotent elements, the intersection of all prime ideals of $\mathbb{Z}$ is already trivial.
+\qedhere
+\end{enumerate}
+\end{proof}
 

### 2016-12-17 Master Thesis/synopsis.tex
Add proof for 5.3

@@
 \item Let $F_1,\ldots,F_n$ be finite extensions of $F$.
-Then $\prod_{i=1}^{n}F_i^\times/F^\times$ is free-by-finite, where $F^\times$ is considered as a subgroup of $\prod_{i=1}^{n}F_i^\times$ via the diagonal embedding.
+Then $\big(\prod_{i=1}^{n}F_i^\times\big)/F^\times$ is free-by-finite, where $F^\times$ is considered as a subgroup of $\prod_{i=1}^{n}F_i^\times$ via the diagonal embedding.
 \end{enumerate}
 \end{lemma}
+
+\begin{proof}
+\begin{enumerate}[(i)]
+\item Let $\mathcal{P}(F)$ be the group of principal ideals of $F$. Consider the following exact sequence:
+\[ \begin{tikzcd}
+1 \ar[r] & \mathcal{O}_F^\times \ar[r] & F^\times \ar[r]& \mathcal{P}(F) \ar[r] & 1
+\end{tikzcd} \]
+By Dirichlet's Unit Theorem, $\mathcal{O}_F^\times$ is free-by-finite.
+$\mathcal{P}(F)$, as a subset of the ideal group of $F$, is free abelian, in particular free-by-finite.
+The assertion follows from Lemma \ref{5.2} (ii).
+\item Consider the following exact sequence:
+\[ \begin{tikzcd}
+1 \ar[r] & \displaystyle \prod_{i=1}^{n-1} F_i^\times \ar[r]& \displaystyle \Big(\prod_{i=1}^{n}F_i^\times\Big)/F^\times \ar[r]& F_n^\times/F^\times \ar[r]& 1
+\end{tikzcd} \]
+The left group is free-by-finite by (i), and by Lemma \ref{5.2} (ii), we can use induction to reduce to the case $n=1$.
+
+Let $\mathcal{P}(F)$ denote the group of principal ideals of $F$, $\mathcal{I}(F)$ the ideal group of $F$, and $\operatorname{Cl}(F) = \mathcal{I}(F)/\mathcal{P}(F)$ the ideal class group of $F$.
+We have the following exact sequence:
+\[ \begin{tikzcd}
+1 \ar[r]& \mathcal{O}_{F_1}^\times /\mathcal{O}_F^\times\ar[r] & F_1^\times / F^\times\ar[r] & \mathcal{P}(F_1)/\mathcal{P}(F)\ar[r] & 1
+\end{tikzcd} \]
+Again by Lemma \ref{5.2} (ii), it suffices to show that $\mathcal{P}(F_1)/\mathcal{P}(F)$ is free-by-finite. Using the snake lemma, we get the following exact sequence:
+\[
+0 \longrightarrow \ker(f)\longrightarrow \mathcal{P}(F_1)/\mathcal{P}(F)\longrightarrow \mathcal{I}(F_1)/\mathcal{I}(F)
+\longrightarrow \operatorname{coker}(f)\longrightarrow 0 \]
+where $f:\operatorname{Cl}(F)\to \operatorname{Cl}(F_1)$.
+Since the ideal class groups are finite, according to Lemma \ref{5.2} (i), it suffices to show that $\mathcal{I}(F_1)/\mathcal{I}(F)$ is free-by-finite.
+This follows from the fact that there are only finitely many primes of $F$ which are ramified in $F_1$.\qedhere
+\end{enumerate}
+\end{proof}
 

### 2016-12-17 Master Thesis/synopsis.tex
Add proof for 5.4

@@
 \end{tikzcd} \]
-Since the rightmost module is $\mathbb{Z}$-flat, tensoring with $\mathbb{Z}/n$ yields another exact sequence and we can consider $K^S/n$ as a subgroup of $K^\times/n$. Then:
+Since the right module is $\mathbb{Z}$-flat, tensoring with $\mathbb{Z}/n\mathbb{Z}$ yields another exact sequence and we can consider $K^S/n$ as a subgroup of $K^\times/n$. Then:
 \[ K^S/n = \bigcap_{P\in X_0\setminus X}U_P/n \]
@@
 \end{lemma}
+
+\begin{proof}
+Let $U = X\setminus S$ and $P_1\in U$ be a closed point. We set:
+\[ C_1 = \ker(r_{P_1}|_{K^S}: K^S \to k(P_1)^\times) \]
+Since $r_{P_1}|_{k^\times}:k^\times \to k(P_1)^\times$ is injective, we have $C_1\cap k^\times = 1$ and the following map is injective:
+\[ C_1 \to\prod_{P\in S}\mathbb{Z},\ x \mapsto (v_P(x))_{P\in S} \]
+Therefore, $C_1$ is a free abelian group of finite rank.
+If $C_1 = 1$, then we set $T = \{P_1\}$ and we are done.
+Otherwise, let $x\in C_1,\ x\neq 1$.
+Then there exists an $a\in k^\times$ with the following properties:
+\begin{enumerate}[(i)]
+\item There exists a point $P_2\in X_0\setminus(S\cup\{P_1\})$ which lies above the prime of $k(x)$ corresponding to the polynomial $x-a\in k[x]$.
+\item $a$ is not a root of unity.
+\end{enumerate}
+We set $C_2 = \ker(r_{P_2}|_{C_1})\subset C_1$.
+$C_2$ is free abelian of finite rank as well.
+By requirement $r_{P_2}(x)=a$ and $a\in k^\times$ is not of finite order.
+It follows $\operatorname{rank}(C_2) < \operatorname{rank}(C_1)$.
+By repeating this process until $C_r=1$, we get points $P_2,\ldots,P_r\in U_0$, such that the following map is injective:
+\[ C_1\to\prod_{i=2}^{r} k(P_i),\ x\mapsto(r_{P_i}(x))_i \]
+Finally for $T=\{P_1,\ldots,P_r\}$, the map $r_{S,T}$ is injective. For the second assertion, note that:
+\[ \operatorname{coker}(r_{S,T}) \cong\operatorname{coker}\Big(K^S/k^\times \to \prod_{P\in T}k(P)^\times /k^\times \Big) \]
+Since $K^S/k^\times$ is finitely generated, Lemma \ref{5.2} (ii) and Lemma \ref{5.3} (ii) show that the latter group is free-by-finite.
+\end{proof}
 

### 2016-12-18 Master Thesis/synopsis.tex
Minor edits
Fix
Add proof of 5.2 (iii)

@@
 
-\begin{lemma}
+\begin{lemma}\label{rank-dimension-inequality}
 Let $v\in\operatorname{Spv}(K)$. Then we have:
@@
 Then there exists a unique henselian $w\in|\operatorname{Spv}(K)|$.
-\item Assume that (i) is satisfied for two different prime numbers. Then $w$ is defectless.
+\item Assume that (i) is satisfied for two different prime numbers. Then $w$ is a defectless valuation.
 \end{enumerate}
@@
 According to Lemma \ref{2.5} we have $\operatorname{rrank}(w)=\dim_\mathbb{Q}(\Gamma_w\otimes_\mathbb{Z}\mathbb{Q} )\geq \dim_{\mathbb{F}_\ell}(\Gamma_w /\ell\Gamma_w) = 2 = \dim(K)$.
+By Lemma \ref{rank-dimension-inequality}, this inequality is actually an equality.
 \end{enumerate}
@@
 Let $v\in\operatorname{Spv}_\text{d}^{(1)}$ and let $N$ be the normal subgroup of $G_{K_v}$ generated by all subgroups $G_{K_w}$ for $w\in\operatorname{Spv}_\text{rd}^{(2)}(K)$ with $v(w) = v$.
-Since a global field has no non-trivial algebraic extension in which every prime splits completely we get $N=G_{K_v}$.
-This shows $K'_{v'} := (K_v)'$ is a henselization of $K'$ with respect to $v'=\sigma(v)$.
+Since a global field has no non-trivial algebraic extension in which every prime splits completely, we get $N=G_{K_v}$.
+This shows that $K'_{v'} = (K_v)'$ is a henselization of $K'$ with respect to $v'=\sigma(v)$.
 Furthermore, let $w_1,w_2\in\operatorname{Spv}_\text{rd}^{(2)}(K)$ with $w_1\neq w_2$ and $v=v(w_1)=v(w_2)$, then:
@@
 \end{tikzcd}\]
+Note that $(K^\text{t}_P)' = (K'^{\text{t}}_{P'})$ as proven in Lemma \ref{3.2}.
 \item Consider the field extension $K_P/K$. Without loss of generality, we can assume $K=K_P$. Then $r_P/n$ is an isomorphism and the composition
@@
 
+\begin{example}
+\begin{itemize}
+\item Free or finite abelian groups are free-by-finite.
+\item Finitely generated abelian groups are free-by-finite by the structure theorem for finitely generated $\mathbb{Z}$-modules.
+\end{itemize}
+\end{example}
+
 \begin{lemma}\phantomsection\label{5.2}
@@
 If $A'$ and $A''$ are free-by-finite, then $A$ is free-by-finite.
-Conversely, if $A$ is free-by-finite, then $A'$ is free-by-finite.
-If additionally, $A'$ is finitely generated, then $A''$ is free-by-finite.
+\item Let $0\longrightarrow A'\stackrel{i}{\longrightarrow}A\stackrel{p}{\longrightarrow}A''\longrightarrow 0$ be an exact sequence of abelian groups.
+If $A$ is free-by-finite, then $A'$ is free-by-finite.
+Additionally, if $A'$ is finitely generated, then $A''$ is free-by-finite.
 \item Let $A$ be free-by-finite. Then $\bigcap_{n\in\mathbb{N}}nA=0$.
@@
 \item TODO
+\item Subgroups of free-by-finite groups are free-by-finite.
+Now assume $A'$ is finitely generated.
+We consider $A'$ as a subgroup of $A$.
+Let $T=A_\text{tor}$.
+By replacing $A'$, $A$, $A''$ with $A'/(T\cap A')$, $A/T$, $A''/p(T)$ respectively, we can assume that $A$ is free and $A'$ is free of finite rank.
+Let $B$ be a basis of $A$.
+There exists a finite subset $C\subset B$ such that $A'\subset \bigoplus_{b\in C}\mathbb{Z}b$.
+We get an exact sequence:
+\[ 0 \longrightarrow \Big(\bigoplus_{b\in C} \mathbb{Z}b\Big)\Big/A' \longrightarrow A/A' \longrightarrow \bigoplus_{b\in B\setminus C}\mathbb{Z}b \longrightarrow 0 \]
+The left abelian group is finitely generated, particularly free-by-finite. The right group is free, hence also free-by-finite. By (ii), $A/A'\cong A''$ is free-by-finite as well.
 \item It suffices to show this assertion for finite groups and for $A=\mathbb{Z}$.
@@
 \[ \begin{tikzcd}
-1 \ar[r] & \displaystyle \prod_{i=1}^{n-1} F_i^\times \ar[r]& \displaystyle \Big(\prod_{i=1}^{n}F_i^\times\Big)/F^\times \ar[r]& F_n^\times/F^\times \ar[r]& 1
+1 \ar[r] & \displaystyle \prod_{i=1}^{n-1} F_i^\times \ar[r]& \displaystyle \Big(\prod_{i=1}^{n}F_i^\times\Big)\Big/F^\times \ar[r]& F_n^\times/F^\times \ar[r]& 1
 \end{tikzcd} \]
@@
 \[ \operatorname{coker}(r_{S,T}) \cong\operatorname{coker}\Big(K^S/k^\times \to \prod_{P\in T}k(P)^\times /k^\times \Big) \]
-Since $K^S/k^\times$ is finitely generated, Lemma \ref{5.2} (ii) and Lemma \ref{5.3} (ii) show that the latter group is free-by-finite.
+Since $K^S/k^\times$ is finitely generated, Lemma \ref{5.2} (iii) and Lemma \ref{5.3} (ii) show that the latter group is free-by-finite.
 \end{proof}
@@
 By Lemma~\ref{5.2} (iii) and Lemma~\ref{5.4} it follows $\varrho = 0$.
-By symmetry, $\varphi^S$ is an isomorphism.
+By symmetry, $\varphi^{S,T}$ is an isomorphism.
 
-Let $T_1, T_2\subset X_0\setminus S$ be two $S$-admissible subsets and suppose $T_1\subset T_2$. By observing $(\star)$ we find:
+Let $T_1, T_2\subset X_0\setminus S$ be two $S$-admissible subsets and suppose $T_1\subset T_2$. By considering $(\star)$ we find:
 \[ r_{S,T_1}\circ\varphi^{S,T_2}=\Big(\prod_{P\in T_1}\varphi_P\Big)\circ r_{S',T_1'} = r_{S,T_1}\circ\varphi^{S,T_1} \]

### 2016-12-18 Master Thesis/synopsis.tex
Fix
inertial fields -> inertia field

@@
 If $E=K_v\subset\overline{K}$ is a henselization of $K$ with respect to $v\in\operatorname{Spv}_\text{d}^{(1)}(K)$, then $E'$ is a henselization of $K'$ with respect to $v'=\sigma(v)$.
-Furthermore, if $E^{\text{t}}$ and $E'^{\text{t}}$ denote the inertial fields of $E$ and $E'$ respectively, then $\sigma$ induces an isomorphism:
+Furthermore, if $E^{\text{t}}$ and $E'^{\text{t}}$ denote the inertia fields of $E$ and $E'$ respectively, then $\sigma$ induces an isomorphism:
 \[ G_{k(v)}\cong\operatorname{Gal}(E^\text{t}/E) \to\operatorname{Gal}(E'^{\text{t}}/E')\cong G_{k(v')} \]
@@
 Furthermore, $K'_{\sigma(w_1)}K'_{\sigma(w_2)} = (K_{w_1}K_{w_2})' \neq \overline{K'}$ and again by Lemma \ref{2.2} we get: \[\sigma(w_1),\sigma(w_2)\in\operatorname{Spv}_\text{rd}^{(2)}(K)\quad \text{and} \quad v(\sigma(w_1)) = v(\sigma(w_2))\]
-Thus $\sigma: \operatorname{Spv}_\text{d}^{(2)}(K)\to \operatorname{Spv}_\text{d}^{(2)}(K')$ maps $\operatorname{Spv}_\text{rd}^{(2)}(K)$ onto $\operatorname{Spv}_\text{rd}^{(2)}(K')$. Furthermore:
+Thus $\sigma: \operatorname{Spv}_\text{d}^{(2)}(K)\to \operatorname{Spv}_\text{d}^{(2)}(K')$ maps $\operatorname{Spv}_\text{rd}^{(2)}(K)$ onto $\operatorname{Spv}_\text{rd}^{(2)}(K')$ such that:
 \[ v(w_1)=v(w_2) \iff v(\sigma(w_1)) = v(\sigma(w_2)) \]
@@
 \item  We have $(K^\text{t}_P)^\times/n\cong \mathbb{Z}/n\mathbb{Z}$ since $\operatorname{char}k(P) = 0$.
-The canonical map $K^\times\to (K^\text{t}_P)^\times/n$ can be identified with the map $K^\times /n\to\mathbb{Z}/n$ induced by $v_P$.
+The canonical map $K^\times/n \to (K^\text{t}_P)^\times/n$ can be identified with the map $K^\times /n\to\mathbb{Z}/n$ induced by $v_P$.
 Now the following commutative diagram gives us the result:

### 2016-12-29 Master Thesis/synopsis.tex
Finish proof 5.2 (i)

@@
 \begin{enumerate}[(i)]
-\item Let $A$ be free-by-finite, i.e. $A = A' \oplus F$ for some finite group $A'$ and free group $F$. Since $\ker(f)\subset A$ is finite, it's actually contained in $A'$.
+\item Let $B$ be free-by-finite and $K = \ker(f)$.
+Considering $A/K$ as a subgroup of $B$, we see that $A/K$ is free-by-finite.
+Since $(A/K)_\text{tor}$ and $K$ are finite, $K' = \{a\in A\mid na\in K \text{ for some }n\in\mathbb{N}\}$ is finite as well, in particular $A_\text{tor}$ is finite.
+In fact $K'=A_\text{tor}$, since $K\subset A_\text{tor}$.
+Now consider the surjection $A\to A/K\to (A/K)/(A/K)_\text{tor}$.
+This map has kernel $K'$, thus inducing an isomorphism from $A/A_\text{tor}$ to a free abelian group.
 
-Now let $B$ be free-by-finite and $K = \ker(f)$.
-Since the injective map $A/K \to B$ induced by $f$ can be restricted to their torsion subgroups, we see that $(A/K)_\text{tor}$ is finite.
-Every $a\in A_\text{tor}$ can be written as $a = a' + k$ for some $a'\in A,\ k\in K$ such that $na'\in K$ for some $n\in\mathbb{N}$.
-Since both $(A/K)_\text{tor}$ and $K$ are finite, $A_\text{tor}$ is finite as well.
-TODO
 \item TODO
+Now let $A$ be free-by-finite, i.e.
+$A = A_\text{tor} \oplus F$ for some free abelian group $F$, and $K=\ker(f)$.
+Since $K\subset A$ is finite, $K$ is actually contained in $A_\text{tor}$.
+Therefore, we see that $A/K$ is free by finite as well, and by extension, the image $I$ of $f$.
+Set $m = \#B/I$ and consider the canonical map $B\stackrel{\cdot m}{\to} I \to I/I_\text{tor}$.
+This map has kernel $\{b\in B \mid mb \in I_\text{tor}\} = B_\text{tor}$, making $B/B_\text{tor}$ isomorphic to a subgroup of the free abelian group $I/I_\text{tor}$.
+Now both $I_\text{tor}$ and $B_\text{tor}/I_\text{tor} \subset B/I$ are finite, hence $B_\text{tor}$ is finite as well.
 \item Subgroups of free-by-finite groups are free-by-finite.

### 2016-12-29 Master Thesis/synopsis.tex
Fix

@@
 \begin{definition}\label{5.1}
-An abelian group is called \textit{free-by-finite} if its torsion subgroup $A_\text{tor}$ is finite and the quotient $A/A_\text{tor}$ is free abelian.
-$A$ is free-by-finite if and only if $A$ is the direct sum of a finite group and a free abelian group.
+An abelian group $A$ is called \textit{free-by-finite} if its torsion subgroup $A_\text{tor}$ is finite and the quotient $A/A_\text{tor}$ is free abelian.
+$A$ is free-by-finite if and only if $A$ is the direct sum of a finite abelian group and a free abelian group.
 \end{definition}
@@
 The left abelian group is finitely generated, particularly free-by-finite. The right group is free, hence also free-by-finite. By (ii), $A/A'\cong A''$ is free-by-finite as well.
-\item It suffices to show this assertion for finite groups and for $A=\mathbb{Z}$.
+\item It suffices to prove this assertion for finite groups and for $A=\mathbb{Z}$.
 For $A$ finite we have $nA = 0$ for $n = \#A$.

### 2016-12-31 Master Thesis/synopsis.tex
Finish proof of 5.2

@@
 
-\item TODO
 Now let $A$ be free-by-finite, i.e.
@@
 Now both $I_\text{tor}$ and $B_\text{tor}/I_\text{tor} \subset B/I$ are finite, hence $B_\text{tor}$ is finite as well.
+\item The exact sequence induces the exact sequence $0 \longrightarrow A'_\text{tor} \longrightarrow A_\text{tor} \longrightarrow A''_\text{tor}$.
+Since the surrounding groups are finite, $A_\text{tor}$ is finite as well.
+Using (i) we see that it suffices to show that $A/A_\text{tor}$ is free-by-finite.
+Consider the following commutative diagram with exact rows:
+\[ \begin{tikzcd}
+0 \ar[r]& A'/A'_\text{tor}\ar[r]\ar[d, "f"'] & A/A_\text{tor}\ar[r]\ar[d, "\cong"] & A/(A'+A_\text{tor}) \ar[r]\ar[d, "g"]& 0\\
+0 \ar[r] & p^{-1}(A''_\text{tor})/A_\text{tor} \ar[r]& A/A_\text{tor}\ar[r] & A''/A''_\text{tor}\ar[r] & 0
+\end{tikzcd} \]
+Note that $f$ is injective and $g$ is surjective. Using the snake lemma, we get an isomorphism $\operatorname{coker}(f)\cong \ker(g)$. We have:
+\[\ker(g) = \ker( (A/A')/(A_\text{tor}/A'_\text{tor}) \to A''/A''_\text{tor} ) = A''_\text{tor}/(A_\text{tor}/A'_\text{tor})\]
+which implies that $\operatorname{coker}(f)\cong \ker(g)$ is finite.
+Applying (i) on $f$, we can see that $p^{-1}(A''_\text{tor})/A_\text{tor}$ is free-by-finite.
+Since $A''/A''_\text{tor}$ is free abelian, the short exact sequence in the second row splits, making $A/A_\text{tor}$ a direct sum of a free-by-finite group and a free group.
 \item Subgroups of free-by-finite groups are free-by-finite.
@@
 \[ \begin{tikzcd}
-1 \ar[r] & \displaystyle \prod_{i=1}^{n-1} F_i^\times \ar[r]& \displaystyle \Big(\prod_{i=1}^{n}F_i^\times\Big)\Big/F^\times \ar[r]& F_n^\times/F^\times \ar[r]& 1
+1 \ar[r] & \displaystyle \prod_{i=2}^{n} F_i^\times \ar[r]& \displaystyle \Big(\prod_{i=1}^{n}F_i^\times\Big)\Big/F^\times \ar[r]& F_1^\times/F^\times \ar[r]& 1
 \end{tikzcd} \]
-The left group is free-by-finite by (i), and by Lemma \ref{5.2} (ii), we can use induction to reduce to the case $n=1$.
+The left group is free-by-finite by (i), and by Lemma \ref{5.2} (ii), we only need to prove that the right group is free-by-finite.
 
@@
 where $f:\operatorname{Cl}(F)\to \operatorname{Cl}(F_1)$.
-Since the ideal class groups are finite, according to Lemma \ref{5.2} (i), it suffices to show that $\mathcal{I}(F_1)/\mathcal{I}(F)$ is free-by-finite.
+Since the class numbers of number fields are finite, according to Lemma \ref{5.2} (i), it suffices to show that $\mathcal{I}(F_1)/\mathcal{I}(F)$ is free-by-finite.
 This follows from the fact that there are only finitely many primes of $F$ which are ramified in $F_1$.\qedhere

### 2016-12-31 Master Thesis/synopsis.tex
Mention subgroup considerations

@@
 Using (i) we see that it suffices to show that $A/A_\text{tor}$ is free-by-finite.
-Consider the following commutative diagram with exact rows:
+
+Observe the following commutative diagram with exact rows, where we consider $A'$ as a subgroup of $A$:
 \[ \begin{tikzcd}
@@
 \[\ker(g) = \ker( (A/A')/(A_\text{tor}/A'_\text{tor}) \to A''/A''_\text{tor} ) = A''_\text{tor}/(A_\text{tor}/A'_\text{tor})\]
-which implies that $\operatorname{coker}(f)\cong \ker(g)$ is finite.
+Since $A''_\text{tor}$ is finite, $\operatorname{coker}(f)\cong \ker(g)$ is finite, too.
 Applying (i) on $f$, we can see that $p^{-1}(A''_\text{tor})/A_\text{tor}$ is free-by-finite.

### 2017-01-02 Master Thesis/synopsis.tex
Add Mordell-Weil Theorem

@@
 \end{tikzcd}\]
-For all $P' = \sigma(P)\in X'_0$ we also get an embedding:
+For all $P\in X_0$ and $P'=\sigma(P)\in X'_0$ we also get an embedding:
 \[ \begin{tikzcd}
@@
 
-\begin{theorem}[\textit{Mordell-Weil Theorem}]\label{mordell-weil-theorem}
-Inhalt...
-\end{theorem}
-
 \begin{definition}\label{4.4}
@@
 \end{definition}
+
+\begin{theorem}[\textit{Mordell-Weil Theorem}]\label{mordell-weil-theorem}
+Let $A$ be an abelian variety over a number field $F$. Then the abelian group $A(F)$ of $F$-rational points is finitely generated.
+\end{theorem}
+
+\begin{corollary}\label{mordell-weil-corollary}
+Let $C$ be a smooth projective curve over a number field $F$. Then the Picard group $\operatorname{Pic}(C)$ is finitely generated.
+\end{corollary}
+
+\begin{proof}
+With base changing to a Galois extension of $F$, we can assume that $C$ has $F$-rational points. We have the following exact sequence:
+\[ 0\longrightarrow \operatorname{Pic}^0(C) \longrightarrow\operatorname{Pic}(C) \longrightarrow \mathbb{Z} \longrightarrow 0  \]
+Now $\operatorname{Pic}^0(C)$ can be identified with the $F$-rational points of the Jacobian variety $J$ of $C$.
+Now $J(F)$ is finitely generated by the Mordell-Weil Theorem \ref{mordell-weil-theorem}, hence $\operatorname{Pic}(C)$ is finitely generated as well.
+\end{proof}
 
@@
 \begin{enumerate}[(i)]
-\item By Mordell-Weil Theorem \ref{mordell-weil-theorem} the Picard group of $X$ is finitely generated.
+\item By Corollary \ref{mordell-weil-corollary}, the Picard group of $X$ is finitely generated.
 Let $D_1,\ldots, D_r$ be Weil divisors of $X$ such that the associated line bundles generate $\operatorname{Pic}(X)$ and similarly let $D'_1,\ldots,D'_s$ generate $\operatorname{Pic}(X')$.

### 2017-01-02 Master Thesis/introduction.tex
"inertia field" is actually "maximal unramified field extension"

@@
 K^\text{s} & \text{separable closure of $K$}\\
+K^\text{nr} & \text{maximal unramified field extension of a henselian $K$}\\
 G_K & \text{absolute Galois group of $K$}

### 2017-01-02 Master Thesis/synopsis.tex
Reorder
"inertia field" is actually "maximal unramified field extension"

@@
 If $E=K_v\subset\overline{K}$ is a henselization of $K$ with respect to $v\in\operatorname{Spv}_\text{d}^{(1)}(K)$, then $E'$ is a henselization of $K'$ with respect to $v'=\sigma(v)$.
-Furthermore, if $E^{\text{t}}$ and $E'^{\text{t}}$ denote the inertia fields of $E$ and $E'$ respectively, then $\sigma$ induces an isomorphism:
-\[ G_{k(v)}\cong\operatorname{Gal}(E^\text{t}/E) \to\operatorname{Gal}(E'^{\text{t}}/E')\cong G_{k(v')} \]
+Furthermore, if $E^{\text{nr}}$ and $E'^{\text{nr}}$ denote the maximal unramified field extensions of $E$ and $E'$ respectively, then $\sigma$ induces an isomorphism:
+\[ G_{k(v)}\cong\operatorname{Gal}(E^\text{nr}/E) \to\operatorname{Gal}(E'^{\text{nr}}/E')\cong G_{k(v')} \]
 which is induced by a unique isomorphism $\varphi_v^\text{s}:k(v')^\text{s}\to k(v)^\text{s}$. In particular $\varphi_v=\varphi_v^\text{s}|_{k(v')}:k(v')\to k(v)$ is an isomorphism.
@@
 \end{enumerate}
-In this case, $K_{w_1}K_{w_2}$ is the inertia field $K_v^\text{t}$ of $K_v$ where $v=v(w_1)$.
+In this case, $K_{w_1}K_{w_2}$ is the maximal unramified field extension $K_v^\text{nr}$ of $K_v$ where $v=v(w_1)$.
 Furthermore, $K'_{\sigma(w_1)}K'_{\sigma(w_2)} = (K_{w_1}K_{w_2})' \neq \overline{K'}$ and again by Lemma \ref{2.2} we get: \[\sigma(w_1),\sigma(w_2)\in\operatorname{Spv}_\text{rd}^{(2)}(K)\quad \text{and} \quad v(\sigma(w_1)) = v(\sigma(w_2))\]
@@
 Furthermore, let $w_1,w_2\in\operatorname{Spv}_\text{rd}^{(2)}(K)$ with $w_1\neq w_2$ and $v=v(w_1)=v(w_2)$, then:
-\[ \sigma(G_{K^\text{t}_v}) = \sigma(G_{K_{w_1}}\cap G_{K_{w_2}}) = G_{K'_{w'_1}}\cap G_{K'_{w'_2}} = G_{(K')^\text{t}_{v'}} \]
+\[ \sigma(G_{K^\text{nr}_v}) = \sigma(G_{K_{w_1}}\cap G_{K_{w_2}}) = G_{K'_{w'_1}}\cap G_{K'_{w'_2}} = G_{(K')^\text{nr}_{v'}} \]
 Therefore, $\sigma$ induces an isomorphism:
-\[ G_{k(v)} \cong G_{K_v}/G_{K_v^\text{t}} \to G_{K'_{v'}}/G_{(K')_{v'}^\text{t}}\cong G_{k(v')} \]
+\[ G_{k(v)} \cong G_{K_v}/G_{K_v^\text{nr}} \to G_{K'_{v'}}/G_{(K')_{v'}^\text{nr}}\cong G_{k(v')} \]
 The rest follows from the Theorem of Neukirch-Uchida.
@@
 \begin{enumerate}[(i)]
-\item  We have $(K^\text{t}_P)^\times/n\cong \mathbb{Z}/n\mathbb{Z}$ since $\operatorname{char}k(P) = 0$.
-The canonical map $K^\times/n \to (K^\text{t}_P)^\times/n$ can be identified with the map $K^\times /n\to\mathbb{Z}/n$ induced by $v_P$.
+\item  We have $(K^\text{nr}_P)^\times/n\cong \mathbb{Z}/n\mathbb{Z}$ since $\operatorname{char}k(P) = 0$.
+The canonical map $K^\times/n \to (K^\text{nr}_P)^\times/n$ can be identified with the map $K^\times /n\to\mathbb{Z}/n\mathbb{Z}$ induced by $v_P$.
 Now the following commutative diagram gives us the result:
@@
 K'^\times /n \ar[r, "\varphi_n"]\ar[d] & K^\times/n\ar[d]\\
-(K^\text{t}_P)'^\times /n \ar[r, "\varphi_n"'] &  (K^\text{t}_P)^\times /n
+(K^\text{nr}_P)'^\times /n \ar[r, "\varphi_n"'] &  (K^\text{nr}_P)^\times /n
 \end{tikzcd}\]
-Note that $(K^\text{t}_P)' = (K'^{\text{t}}_{P'})$ as proven in Lemma \ref{3.2}.
+Note that $(K^\text{nr}_P)' = (K'^{\text{nr}}_{P'})$ as proven in Lemma \ref{3.2}.
 \item Consider the field extension $K_P/K$. Without loss of generality, we can assume $K=K_P$. Then $r_P/n$ is an isomorphism and the composition
 \[\begin{tikzcd}
-k(P)^\times / n\ar[rr, "(r_P/n)^{-1}"] && U_P/n \ar[r, hook] & K_P^\times/n
+k(P)^\times / n\ar[rr, "(r_P/n)^{-1}"] && U_P/n \ar[rr, hook] && K_P^\times/n
 \end{tikzcd}\]
@@
 \end{proof}
-
-\begin{definition}\label{4.4}
-For a finite subset $S\subset X_0$ set:
-\[ K^S=\{x\in K^\times \mid v_P(x)=0\text{ for all }P\in X_0\setminus S \} \]
-$K'^{S'}$ is defined similarly. For finite subsets $S$ of $X_0$, the corresponding subsets of $X_0'$ will be denoted by $S'$, i.e. $S'=\{\sigma P\mid P\in S \} $.
-\end{definition}
 
@@
 \end{proof}
+
+\begin{definition}\label{4.4}
+For a finite subset $S\subset X_0$ set:
+\[ K^S=\{x\in K^\times \mid v_P(x)=0\text{ for all }P\in X_0\setminus S \} \]
+$K'^{S'}$ is defined similarly. For finite subsets $S$ of $X_0$, the corresponding subsets of $X_0'$ will be denoted by $S'$, i.e. $S'=\{\sigma P\mid P\in S \} $.
+\end{definition}
 

### 2017-01-04 Master Thesis/synopsis.tex
Change phrasing of 2.5

@@
 
-\begin{lemma}
+\begin{lemma}\label{2.2-prepre}
 Let $v,w\in\operatorname{Spv}(K)$. We have:
 \[\mathcal{O}_w\subset\mathcal{O}_v \iff w\in\overline{\{v\}} \]
-Thus $v$ is a closed point if and only if $\dim k(v)=0$.
+Thus $v$ is a closed point if and only if it has no non-trivial specializations or iff $\dim k(v)=0$, i.e. $k(v)$ has positive characteristic and is algebraic over its prime field.
 \end{lemma}
@@
 Set $k=k(w)$ and $\Gamma = \Gamma_w$.
-Then $k$ is an algebraic extension of a finite field $\mathbb{F}_p$.
+Then $k$ is an algebraic extension of a finite field $\mathbb{F}_p$ by Lemma \ref{2.2-prepre}.
 Let $\ell\neq p$ be a prime and assume $\mu_\ell\subset K$.
-Then the following conditions are equivalent:
-\begin{enumerate}[(i)]
-\item $\mathrm{H}^3(K,\mathbb{Z}/\ell\mathbb{Z})=\mathbb{Z}/\ell\mathbb{Z}$
-\item $\ell^\infty\nmid[k:\mathbb{F}_p]$ and $\dim_{\mathbb{F}_\ell}(\Gamma/\ell\Gamma) = 2$
-\end{enumerate}
-Otherwise we have $\mathrm{H}^3(K,\mathbb{Z}/\ell\mathbb{Z})=0$. If $L/K$ is an algebraic extension and (i) holds, we have:
+Then:
+\[ \mathrm{H}^3(K, \mathbb{Z}/\ell\mathbb{Z}) = \begin{cases}
+\mathbb{Z}/\ell\mathbb{Z}, & \ell^\infty\nmid [k:\mathbb{F}_p] \text{ and } \dim_{\mathbb{F}_\ell}(\Gamma / \ell\Gamma) = 2\\
+0, & \text{otherwise}
+\end{cases} \]
+If $L/K$ is an algebraic extension and $\mathrm{H}^3(K, \mathbb{Z}/\ell\mathbb{Z}) = \mathbb{Z}/\ell\mathbb{Z}$ holds, we have:
 \[ \mathrm{H}^3(L,\mathbb{Z}/\ell\mathbb{Z}) = \begin{cases}
@@
 Therefore we have:
-\[ \mathrm{H}^3(K,\mathbb{Z}/\ell\mathbb{Z}) \cong\mathrm{H}^3(G_K(\ell),\mathbb{Z}/\ell) \cong\mathrm{H}^3(G_\ell,\mathbb{Z}/\ell\mathbb{Z}) \]
+\[ \mathrm{H}^3(K,\mathbb{Z}/\ell\mathbb{Z}) \cong\mathrm{H}^3(G_K(\ell),\mathbb{Z}/\ell\mathbb{Z}) \cong\mathrm{H}^3(G_\ell,\mathbb{Z}/\ell\mathbb{Z}) \]
 Now the equivalence of (i) and (ii) follows from [Se], Ch. I, §4.

### 2017-01-04 Master Thesis/synopsis.tex
Add Theorem of F.K. Schmidt

@@
 \[\{ w\in\operatorname{Spv}(K)\mid \mathcal{O}_w\subset\mathcal{O}_v \} \to \operatorname{Spv}k(v),\ w\mapsto w\mod v\]
-If $w\in\operatorname{Spv}_\text{rd}^{(i)}(K)$ for some $i\geq 1$ then there exists a unique $v(w)\in\operatorname{Spv}_\text{rd}^{(i-1)}(K)$ such that $\mathcal{O}_w\subset\mathcal{O}_v$.
-$w$ is henselian if and only if $v$ and $w\mod v$ are henselian.
+If $w\in\operatorname{Spv}_\text{rd}^{(i)}(K)$ for some $i\geq 1$ then there exists a unique $v(w)\in\operatorname{Spv}_\text{rd}^{(i-1)}(K)$ such that $\mathcal{O}_w\subset\mathcal{O}_{v(w)}$.
+$w$ is henselian if and only if $v(w)$ and $w\mod v(w)$ are henselian.
 \end{lemma}
@@
 
+\begin{theorem}[\textit{F.K. Schmidt}]\label{theorem-fkschmidt}
+Let $v_1,v_2\in\operatorname{Spv}(K)$ be two inequivalent henselian valuations of rank $1$ on a field $K$. Then $K$ is separably closed.
+\end{theorem}
+
+\begin{proof}
+See Valued Fields.
+\end{proof}
+
 \begin{lemma}\label{2.2}
-Let $K$ be henselian with respect to two different valuations $w_1,w_2\in |\operatorname{Spv}(K)|$.
-Then $K$ is algebraically closed or $w_1,w_2\in\operatorname{Spv}_\text{rd}^{(2)}(K),\ v(w_1)=v(w_2)$ and $k(v(w_1))$ is separably closed.
-Furthermore, we have $\operatorname{cd}(G_K)\leq 2$.
+Let $K$ be henselian with respect to two different valuations $w_1,w_2\in |\operatorname{Spv}(K)|$. Then the following statements are equivalent: 
+\begin{enumerate}[(i)]
+\item $K$ is not algebraically closed.
+\item $w_1,w_2\in\operatorname{Spv}_\text{rd}^{(2)}(K),\ v(w_1)=v(w_2)$ and $k(v(w_1))$ is separably closed
+\end{enumerate}
+In this case, we have $\operatorname{cd}(G_K)\leq 2$.
 \end{lemma}
@@
 \begin{proof}
-Assume that $K$ is not algebraically closed, i.e.
-$G_K\neq 1$ and set $v_i = v(w_i)$.
-By Lemma~\ref{2.2-pre} $K$ is henselian with respect to both $v_1$ and $v_2$.
-If $v_1\neq v_2$, by the Theorem of F.K.
-Schmidt we have $G_{v_1}\cap G_{v_2} = 1 \neq G_K$, a contradiction.
-Therefore $v_1 = v_2$.
-
-Now $k(v_1) = k(v_2)$ is henselian with respect to both $w_1\mod v_1$ and $w_2\mod v_2$. Again by the Theorem of F.K. Schmidt we get $G_{k(v_1)} = 1$, hence $k(v_1)$ is separably closed.
-
-The second assertion follows from the first and from [K2], §5, Theorem 1.
+\begin{itemize}
+\item Assume that $K$ is not algebraically closed.
+By the Theorem of F.K.
+Schmidt \ref{theorem-fkschmidt}, we can see that $w_1,w_2$ are not of rank $1$.
+By the dimension inequality \ref{rank-dimension-inequality} and considering $\dim(K) = 2$ and $\dim k(w_i) = 0$ it follows $w_1,w_2\in\operatorname{Spv}^{(2)}_\text{rd}(K)$.
+Set $v_i=v(w_i)$.
+By Lemma \ref{2.2-pre} $K$ is henselian with respect to both $v_1$ and $v_2$.
+Again, by the Theorem of F.K.
+Schmidt \ref{theorem-fkschmidt}, $v_1 = v_2$.
+Furthermore $k(v_1) = k(v_2)$ is henselian with respect to two different valuations of rank $1$, namely $w_1\mod v_1$ and $w_2\mod v_1$, making $k(v_1)$ separably closed.
+\item Now assume (ii) holds.
+\item The second assertion follows from the first and from [K2], §5, Theorem 1.\qedhere
+\end{itemize}
 \end{proof}
@@
 We set $K'_{\sigma(w)} = (K_w)'$ for $w\in\operatorname{Spv}_\text{d}^{(2)}(K)$.
-This implies that $K'_{\sigma(w)}$ is a henselization of $K'$ with respect to $\sigma(w)$.
+Lemma \ref{3.1} shows that $K'_{\sigma(w)}$ is a henselization of $K'$ with respect to $\sigma(w)$.
 

### 2017-01-04 Master Thesis/synopsis.tex
Preparing for non-finitely generatedness

@@
 The \textit{rank}\index{valuation!rank} of $v$ is defined as the Krull dimension of its valuation ring $\mathcal{O}_v$.
-The \textit{rational rank}\index{valuation!rational rank} of $v:K\to\Gamma_v$ is defined as the rank of its valuation group $\Gamma_v$ as a $\mathbb{Z}$-module:
+The \textit{rational rank}\index{valuation!rational rank} of $v:K^\times\to\Gamma_v$ is defined as the rank of its valuation group $\Gamma_v$ as a $\mathbb{Z}$-module:
 \[\operatorname{rrank}(v) = \dim_\mathbb{Q}(\Gamma_v\otimes_\mathbb{Z}\mathbb{Q}) \]
@@
 
-\begin{lemma}
+\begin{lemma}\label{dimension-equality}
 Let $K$ be a finitely generated field and $v\in\operatorname{Spv}(K)$ with:
@@
 \begin{lemma}
-$\operatorname{Spv}(K)$ with the topology defined above is a quasicompact $T_0$-space.
-Every closed subset $A\subset\operatorname{Spv}(K)$ is the closure of its closed points.
+$\operatorname{Spv}(K)$ with the topology defined above is a quasicompact $T_0$-space. Every closed subset $A\subset\operatorname{Spv}(K)$ is the closure of its closed points.
+
 In particular the set of closed points $|\operatorname{Spv}(K)|$ of $\operatorname{Spv}(K)$ is a quasicompact $T_1$-space, endowed with the induced topology.
+By the dimension inequality \ref{rank-dimension-inequality} we have $\operatorname{Spv}^{(n)}(K)\subset |\operatorname{Spv}(K)|$ where $n=\dim(K)$.
 \end{lemma}
@@
 
-\paragraph{} Let $K$ be a finitely generated field of characteristic $0$ and dimension $2$.
+\begin{lemma}\label{lemma-spvd}
+Let $K$ be a finitely generated field of dimension $2$. Then we have:
+\begin{enumerate}[(i)]
+\item $\operatorname{Spv}^{(2)}(K) = \operatorname{Spv}^{(2)}_\text{d}(K) = \operatorname{Spv}^{(2)}_\text{rd}(K)$
+\item $\operatorname{Spv}(K) = \operatorname{Spv}_\text{d}(K)$
+\end{enumerate}
+\end{lemma}
+
+\begin{proof}
+\begin{enumerate}[(i)]
+\item Let $v\in\operatorname{Spv}^{(2)}(K)\subset |\operatorname{Spv}(K)|$.
+By Lemma \ref{2.2-prepre} we have $\dim k(v) = 0$ and thus the equality of the dimension inequality.
+By Lemma \ref{dimension-equality} $\Gamma_v$ is a finitely generated abelian group of rank $2$, thus not a subgroup of $\mathbb{R}$, i.e.
+$\operatorname{rank}(v)> 1$.
+\item Let $v\in\operatorname{Spv}(K)$.
+If $\operatorname{rank}(v) = 2$, the assertion follows from Lemma \ref{rank-dimension-inequality}.
+If $\operatorname{rank}(v) = 0$, the value group $\Gamma_v$ is trivial, thus $\operatorname{rrank}(v) = 0$ as well.
+If $\operatorname{rank}(v) = 1$ and $\operatorname{rrank}(v) = 2$, we have $v\in\operatorname{Spv}^{(2)}_\text{d}(K)$ by (i), a contradiction.
+\qedhere
+\end{enumerate}
+\end{proof}
+
+\paragraph{} Let $K$ be a (not necessarily finitely generated) field of characteristic $0$ and dimension $2$.
 
@@
 Schmidt \ref{theorem-fkschmidt}, we can see that $w_1,w_2$ are not of rank $1$.
-By the dimension inequality \ref{rank-dimension-inequality} and considering $\dim(K) = 2$ and $\dim k(w_i) = 0$ it follows $w_1,w_2\in\operatorname{Spv}^{(2)}_\text{rd}(K)$.
+Thus they are of rank $2$ and by Lemma \ref{rank-dimension-inequality} we can see $w_1,w_2\in\operatorname{Spv}_\text{rd}^{(2)}(K)$.
 Set $v_i=v(w_i)$.
@@
 \[ \mathrm{H}^3(K)\longrightarrow \bigoplus_{x\in\mathfrak{X}_1}\  \bigoplus_{\substack{w\in\operatorname{Spv}_\text{rd}^{(2)}(K) \\ v(w) = v_x}} \mathrm{H}^3(K_w) \]
-The direct sum ranges over a subset of $|\operatorname{Spv}(K)|$, thus concludes the proof.
+The direct sum ranges over a subset of $|\operatorname{Spv}(K)|$, thus this concludes the proof.
 \end{proof}

### 2017-01-04 Master Thesis/synopsis.tex
Add proof for 2.4 for non-finitely-generated fields

@@
 
+\begin{lemma}\label{2.4-pre}
+Let $K$ be finitely generated, $\ell\neq 2$ be a prime and $f\in\mathrm{H}^3(K, \mathbb{Z}/\ell\mathbb{Z})$. Then the support of $f$ is closed in $|\operatorname{Spv}(K)|$:
+\[ \operatorname{supp}(f) = \{w\in |\operatorname{Spv}(K)|\mid \operatorname{res}_{K_w/K}(f)\neq 0 \} \]
+\end{lemma}
+
 \begin{lemma}\label{2.4}
@@
 Now we can replace $\mathbb{Z}/\ell\mathbb{Z}$ by $\mathbb{Z}/\ell\mathbb{Z}(2) = \mu_\ell^{\otimes 2}$.
+We write $\mathrm{H}^3(L) = \mathrm{H}^3(L, \mathbb{Z}/\ell\mathbb{Z}(2))$.
 
-We write $\mathrm{H}^3(L) = \mathrm{H}^3(L, \mathbb{Z}/\ell\mathbb{Z}(2))$.
+First, let's consider the case where $K$ is finitely generated.
 Let $\mathfrak{X}$ be a model of $K$.
@@
 \[ \mathrm{H}^3(K)\longrightarrow \bigoplus_{x\in\mathfrak{X}_1}\  \bigoplus_{\substack{w\in\operatorname{Spv}_\text{rd}^{(2)}(K) \\ v(w) = v_x}} \mathrm{H}^3(K_w) \]
-The direct sum ranges over a subset of $|\operatorname{Spv}(K)|$, thus this concludes the proof.
+The direct sum ranges over a subset of $|\operatorname{Spv}(K)|$, thus this shows the assertion. Now, we drop the assumption that $K$ is finitely generated.
+
+Let $f\in\mathrm{H}^3(K)$ such that $\operatorname{res}_{K_w/K}(f) = 0$ for all $w\in |\operatorname{Spv}(K)|$.
+There exists a finitely generated subfield $K_0$ of $K$ such that $f = \operatorname{res}_{K/K_0}(g)$ for some $g\in\mathrm{H}^3(K_0)$ and $K/K_0$ is algebraic.
+Let $\{K_i\mid i\in I\}$ be the set of all finite subextensions of $K/K_0$.
+For $w\in |\operatorname{Spv}(K)|$ we denote the restriction of $w$ to $K_i$ with $w$ as well.
+
+Let  $w\in |\operatorname{Spv}(K)|$.
+Since $\mathrm{H}^3(K_w) = \varinjlim_I \mathrm{H}^3((K_i)_w)$ there exists an $i = i(w)\in I$ such that $\operatorname{res}_{(K_i)_w/K}(g) = 0$.
+By Lemma \ref{2.4-pre} there is an open neighborhood $U_{i(w)}\subset |\operatorname{Spv}(K_i)|$ of $w$ such that $\operatorname{res}_{(K_i)_v/K}(g)=0$ for all $v\in U_{i(w)}$.
+For $K_i\subset K_j\subset K$ let
+\[ \varphi_i: |\operatorname{Spv}(K)| \to |\operatorname{Spv}(K_i)| \]
+\[ \varphi_{ij}: |\operatorname{Spv}(K_j)|\to |\operatorname{Spv}(K_i)| \]
+be the restriction maps. They are continuous. Since $|\operatorname{Spv}(K)|$ is quasicompact we can find a finite set $S\subset |\operatorname{Spv}(K)|$ such that:
+\[ \bigcup_{w\in S}\varphi^{-1}_{i(w)}(U_{i(w)}) = |\operatorname{Spv}(K)| \]
+Let $j\in I$ such that $K_j$ is the composite of the fields $K_{i(w)},\ w\in S$. We also have:
+\[ \bigcup_{w\in S}\varphi^{-1}_{i(w)j}(U_{i(w)}) = |\operatorname{Spv}(K_j)| \]
+Let $v\in |\operatorname{Spv}(K_j)|$ and let $i = i(w),\ w\in S$ such that $v\in\varphi_{ij}^{-1}(U_{i(w)})$. We have:
+\[ \operatorname{res}_{(K_j)_v/K_j}(\operatorname{res}_{K_j/K_0}(g)) = \operatorname{res}_{(K_j)_v/(K_i)_v}(\operatorname{res}_{(K_i)_v/K_0}(g)) = 0 \]
+Since $K_j$ is finitely generated and we proved the statement for finitely generated fields, we obtain $\operatorname{res}_{K_j/K_0}(g) = 0$, thus $f = \operatorname{res}_{K/K_0}(g) = 0$.
 \end{proof}

### 2017-01-04 Master Thesis/synopsis.tex
Show that the big-ass set is closed

@@
 A \textit{model} of $K$ is a regular, proper scheme $\mathfrak{X}$ over $\mathbb{Z}$ with function field $K$.
-A \textit{morphism of models} is a morphism of schemes inducing the identity on $K$.
-We denote the category of models of $K$ by $\mathfrak{M}(K)$; it is filtered.
+$\mathfrak{X}_i$ denotes the set of points $x\in\mathfrak{X}$ whose closure $\overline{\{x\}}$ is $i$-dimensional.
+
+A \textit{morphism of models} is a morphism of schemes inducing the identity on $K$. We denote the category of models of $K$ by $\mathfrak{M}(K)$; it is filtered.
 
@@
 
-\begin{lemma}\label{2.4-pre}
-Let $K$ be finitely generated, $\ell\neq 2$ be a prime and $f\in\mathrm{H}^3(K, \mathbb{Z}/\ell\mathbb{Z})$. Then the support of $f$ is closed in $|\operatorname{Spv}(K)|$:
-\[ \operatorname{supp}(f) = \{w\in |\operatorname{Spv}(K)|\mid \operatorname{res}_{K_w/K}(f)\neq 0 \} \]
-\end{lemma}
-
-\begin{lemma}\label{2.4}
-Let $\ell\neq 2$ be a prime. The following map, induced by the restrictions, is injective:
+\begin{lemma}\label{2.4-finitelygenerated}
+Let $K$ be finitely generated and $\ell\neq 2$ be a prime. The following map, induced by the restrictions, is injective:
 \[ \mathrm{H}^3(K,\mathbb{Z}/\ell\mathbb{Z})\to\prod_{w\in |\operatorname{Spv}(K)| }\mathrm{H}^3(K_w,\mathbb{Z}/\ell\mathbb{Z}) \]
@@
 
-First, let's consider the case where $K$ is finitely generated.
 Let $\mathfrak{X}$ be a model of $K$.
-$\mathfrak{X}_i$ denotes the set of points $x\in\mathfrak{X}$ whose closure $\overline{\{x\}}$ is $i$-dimensional.
 For $x\in\mathfrak{X}_1$ the local ring $\mathcal{O}_{\mathfrak{X}, x}$ is a discrete valuation ring and defines a valuation $v_x\in\operatorname{Spv}_\text{d}^{(1)}(K)$.
-By [K1] Theorem 0.8, the following map
+By [K1] Theorem 0.8, the map
 \[ \mathrm{H}^3(K)\longrightarrow \prod_{x\in\mathfrak{X}_1}\mathrm{H}^3(K_{v_x}) \]
@@
 where $k = k(v)$. This map is injective by the classical Hasse principle and its image is contained in $\bigoplus {}_\ell\operatorname{Br}(k_{\tilde{w}})$. Combining these maps yields the injection:
-\[ \mathrm{H}^3(K)\longrightarrow \bigoplus_{x\in\mathfrak{X}_1}\  \bigoplus_{\substack{w\in\operatorname{Spv}_\text{rd}^{(2)}(K) \\ v(w) = v_x}} \mathrm{H}^3(K_w) \]
-The direct sum ranges over a subset of $|\operatorname{Spv}(K)|$, thus this shows the assertion. Now, we drop the assumption that $K$ is finitely generated.
+\[\Phi: \mathrm{H}^3(K)\longrightarrow \bigoplus_{x\in\mathfrak{X}_1}\  \bigoplus_{\substack{w\in\operatorname{Spv}_\text{rd}^{(2)}(K) \\ v(w) = v_x}} \mathrm{H}^3(K_w) \]
+The direct sum ranges over a subset of $|\operatorname{Spv}(K)|$, thus this concludes the proof. 
+\end{proof}
+
+\begin{lemma}\label{2.4-closedset}
+Let $K$ be finitely generated, $\ell\neq 2$ be a prime and $f\in\mathrm{H}^3(K)$. Let $\mu_\ell\subset K$ and $K$ not formally real. Then the support of $f$ is closed in $|\operatorname{Spv}(K)|$:
+\[ \operatorname{supp}(f) = \{w\in |\operatorname{Spv}(K)|\mid \operatorname{res}_{K_w/K}(f)\neq 0 \} \]
+where $\operatorname{res}_{K_w/K}: \mathrm{H}^3(K)\to \mathrm{H}^3(K_w)$.
+\end{lemma}
+
+\begin{proof}
+We adopt the notation from the previous proof. First, we shall prove that the following map is injective for every model $\mathfrak{X}$ of $K$:
+\[ f: \mathrm{H}^3(K) \to \prod_{y\in\mathfrak{X}_0}\mathrm{H}^3(K_y) \]
+For $x\in\mathfrak{X}$ let $\mathcal{O}_x$ be a henselization of $\mathcal{O}_{\mathfrak{X}, x}$ and $K_x$ its quotient field.
+For $y\in\mathfrak{X}_0$ set $P_y=\operatorname{Spec}(\mathcal{O}_y)_1$.
+For $z\in P_y$ let $\mathcal{O}_z$ be a henselization of $\mathcal{O}_y$ at $z$ and $K_z$ its quotient field.
+For $y\in\mathfrak{X}_0$ and $x\in\mathfrak{X}_1$ let $P_{x,y}$ be the subset of $P_y$ consisting of all $z\in P_y$ which are mapped to $x$ under $\operatorname{Spec}(\mathcal{O}_y)\to\mathfrak{X}$.
+Consider the following commutative diagram:
+\[ \begin{tikzcd}
+\mathrm{H}^3(K)\ar[r, "\varphi"]\ar[d, "f"'] & \displaystyle\prod_{x\in\mathfrak{X}_1}\mathrm{H}^3(K_x)\ar[d, "\psi"]\\
+\displaystyle\prod_{y\in\mathfrak{X}_0}\mathrm{H}^3(K_y) \ar[r, "g"']& \displaystyle\prod_{\substack{x\in\mathfrak{X}_1 \\ y\in\mathfrak{X}_0}}\prod_{z\in P_{x,y}}\mathrm{H}^3(K_z)
+\end{tikzcd} \]
+According to [Sa], Theorem (2-15) $g$ is injective.
+Now, $\psi\circ\varphi$ can be identified with the injective map $\Phi$ from the proof of the previous Lemma.
+Thus its image is contained in $\bigoplus\mathrm{H}^3(K_w)$.
+Therefore, $f$ is injective as well and its image is contained in $\bigoplus\mathrm{H}^3(K_y)$.
+Now, one easily checks that
+\[ \operatorname{supp}(f) = \bigcap_{\mathfrak{M}_K} \pi_\mathfrak{X}^{-1}(\{ y\in\mathfrak{X}_0\mid \operatorname{res}_{K_y/K}(f)\neq 0 \}) \]
+is closed.
+\end{proof}
+
+\begin{lemma}\label{2.4}
+Let $K$ be not necessarily finitely generated and $\ell\neq 2$ be a prime. The following map, induced by the restrictions, is injective:
+\[ \mathrm{H}^3(K,\mathbb{Z}/\ell\mathbb{Z})\to\prod_{w\in |\operatorname{Spv}(K)| }\mathrm{H}^3(K_w,\mathbb{Z}/\ell\mathbb{Z}) \]
+\end{lemma}
+
+\begin{proof}
+Just like in Lemma \ref{2.4-finitelygenerated} we can assume that $\mu_\ell\subset K$ and $K$ is not formally real and replace $\mathbb{Z}/\ell\mathbb{Z}$ by $\mathbb{Z}/\ell\mathbb{Z}(2)$.
+We adopt the notation from the previous proof.
 
@@
 Since $\mathrm{H}^3(K_w) = \varinjlim_I \mathrm{H}^3((K_i)_w)$ there exists an $i = i(w)\in I$ such that $\operatorname{res}_{(K_i)_w/K}(g) = 0$.
-By Lemma \ref{2.4-pre} there is an open neighborhood $U_{i(w)}\subset |\operatorname{Spv}(K_i)|$ of $w$ such that $\operatorname{res}_{(K_i)_v/K}(g)=0$ for all $v\in U_{i(w)}$.
+By Lemma \ref{2.4-closedset} there is an open neighborhood $U_{i(w)}\subset |\operatorname{Spv}(K_i)|$ of $w$ such that $\operatorname{res}_{(K_i)_v/K}(g)=0$ for all $v\in U_{i(w)}$.
 For $K_i\subset K_j\subset K$ let
@@
 \[ \operatorname{res}_{(K_j)_v/K_j}(\operatorname{res}_{K_j/K_0}(g)) = \operatorname{res}_{(K_j)_v/(K_i)_v}(\operatorname{res}_{(K_i)_v/K_0}(g)) = 0 \]
-Since $K_j$ is finitely generated and we proved the statement for finitely generated fields, we obtain $\operatorname{res}_{K_j/K_0}(g) = 0$, thus $f = \operatorname{res}_{K/K_0}(g) = 0$.
+Since $K_j$ is finitely generated we obtain $\operatorname{res}_{K_j/K_0}(g) = 0$ with Lemma \ref{2.4-finitelygenerated}, thus $f = \operatorname{res}_{K/K_0}(g) = 0$.
 \end{proof}

### 2017-01-04 Master Thesis/synopsis.tex
Add back missing Lemma

@@
 
-\iffalse
 \begin{lemma}\phantomsection\label{2.3}
@@
 \[ \mathcal{O}_v = \varinjlim_{\mathfrak{X}\in\mathfrak{M}(K)}\mathcal{O}_{\mathfrak{X},\pi_\mathfrak{X}(v)} \]
-\item For $v\in\operatorname{Spv}(K)$ let $K_v$ be the quotient field of a henselization of $\mathcal{O}_v$ and for $\mathfrak{X}\in\mathfrak{M}(K),\ x\in\mathfrak{X}$ let $K_\mathfrak{X}$ be the quotient field of a henselization of $\mathcal{O}_{\mathfrak{X},x}$.
+\item For $v\in\operatorname{Spv}(K)$ let $K_v$ be the quotient field of a henselization of $\mathcal{O}_v$ and for $\mathfrak{X}\in\mathfrak{M}(K),\ x\in\mathfrak{X}$ let $K_x$ be the quotient field of a henselization of $\mathcal{O}_{\mathfrak{X},x}$.
 Then:
@@
 \end{lemma}
-\fi
+
+\begin{proof}
+(i) and (ii) follow from [HK], Lemma 2.4.3 and (iii) is a consequence of (ii) and [EGA IV] 18.6.14.
+\end{proof}
 
@@
 \begin{proof}
-Observe $\varphi(-1)^2 = \varphi(1) = 1$, and since $\varphi$ is bijective on $K'^\times$, it follows $\varphi(-1) \neq 1$ and hence $\varphi(-1) = -1$.
+Observe that $\varphi(-1)^2 = \varphi(1) = 1$, and since $\varphi$ is bijective on $K'^\times$, it follows $\varphi(-1) \neq 1$ and hence $\varphi(-1) = -1$.
 Now, to prove the assertion, it suffices to show $\varphi(x + y) = \varphi(x) + \varphi(y)$ for all $x,y\in K'^\times$.

### 2017-01-04 Master Thesis/synopsis.tex
Better helper lemma

@@
 \item $\operatorname{Spv}^{(2)}(K) = \operatorname{Spv}^{(2)}_\text{d}(K) = \operatorname{Spv}^{(2)}_\text{rd}(K)$
-\item $\operatorname{Spv}(K) = \operatorname{Spv}_\text{d}(K)$
+\item $\operatorname{Spv}^{(1)}(K) = \operatorname{Spv}^{(1)}_\text{d}(K)$
 \end{enumerate}
@@
 \item Let $v\in\operatorname{Spv}^{(2)}(K)\subset |\operatorname{Spv}(K)|$.
-By Lemma \ref{2.2-prepre} we have $\dim k(v) = 0$ and thus the equality of the dimension inequality.
+By Lemma \ref{2.2-prepre} we have $\dim k(v) = 0$ and thus the equality of the dimension inequality in Lemma \ref{rank-dimension-inequality}.
 By Lemma \ref{dimension-equality} $\Gamma_v$ is a finitely generated abelian group of rank $2$, thus not a subgroup of $\mathbb{R}$, i.e.
 $\operatorname{rank}(v)> 1$.
-\item Let $v\in\operatorname{Spv}(K)$.
-If $\operatorname{rank}(v) = 2$, the assertion follows from Lemma \ref{rank-dimension-inequality}.
-If $\operatorname{rank}(v) = 0$, the value group $\Gamma_v$ is trivial, thus $\operatorname{rrank}(v) = 0$ as well.
-If $\operatorname{rank}(v) = 1$ and $\operatorname{rrank}(v) = 2$, we have $v\in\operatorname{Spv}^{(2)}_\text{d}(K)$ by (i), a contradiction.
-\qedhere
+\item Let $v\in\operatorname{Spv}^{(1)}(K)$. If $\operatorname{rank}(v) = 0$, $v$ would be the trivial valuation with trivial value group $\Gamma_v = 0$ and rational rank $0$. \qedhere
 \end{enumerate}

### 2017-01-07 Master Thesis/synopsis.tex
Add reference to F.K. Schmidt
Remove wrong assertion

@@
 \end{lemma}
-
-\begin{lemma}\label{lemma-spvd}
-Let $K$ be a finitely generated field of dimension $2$. Then we have:
-\begin{enumerate}[(i)]
-\item $\operatorname{Spv}^{(2)}(K) = \operatorname{Spv}^{(2)}_\text{d}(K) = \operatorname{Spv}^{(2)}_\text{rd}(K)$
-\item $\operatorname{Spv}^{(1)}(K) = \operatorname{Spv}^{(1)}_\text{d}(K)$
-\end{enumerate}
-\end{lemma}
-
-\begin{proof}
-\begin{enumerate}[(i)]
-\item Let $v\in\operatorname{Spv}^{(2)}(K)\subset |\operatorname{Spv}(K)|$.
-By Lemma \ref{2.2-prepre} we have $\dim k(v) = 0$ and thus the equality of the dimension inequality in Lemma \ref{rank-dimension-inequality}.
-By Lemma \ref{dimension-equality} $\Gamma_v$ is a finitely generated abelian group of rank $2$, thus not a subgroup of $\mathbb{R}$, i.e.
-$\operatorname{rank}(v)> 1$.
-\item Let $v\in\operatorname{Spv}^{(1)}(K)$. If $\operatorname{rank}(v) = 0$, $v$ would be the trivial valuation with trivial value group $\Gamma_v = 0$ and rational rank $0$. \qedhere
-\end{enumerate}
-\end{proof}
 
@@
 By the Theorem of F.K.
-Schmidt we have $G_{\overline{v}_1}\cap G_{\overline{v}_2} = 1$ for $\overline{v}_1\neq\overline{v}_2$.
+Schmidt \ref{theorem-fkschmidt} we have $G_{\overline{v}_1}\cap G_{\overline{v}_2} = 1$ for $\overline{v}_1\neq\overline{v}_2$.
 Therefore, $\sigma(g) = \tau(g) = \varphi^{-1}g\varphi$ for all $g\in G_K$.

### 2017-01-07 Master Thesis/synopsis.tex
Add some references

@@
 
-\begin{lemma}\label{dimension-equality}
-Let $K$ be a finitely generated field and $v\in\operatorname{Spv}(K)$ with:
-\[\operatorname{rrank}(v)+\dim k(v)= \dim(K) \]
-Then $\Gamma_v$ is a finitely generated $\mathbb{Z}$-module and $k(v)$ is a finitely generated field.
-\end{lemma}
+\begin{proof}
+See Valued Fields 3.4.1 and 3.4.3.
+\end{proof}
 
@@
 
-\begin{lemma}
+\begin{lemma}\label{spv-topology}
 $\operatorname{Spv}(K)$ with the topology defined above is a quasicompact $T_0$-space. Every closed subset $A\subset\operatorname{Spv}(K)$ is the closure of its closed points.
@@
 In particular the set of closed points $|\operatorname{Spv}(K)|$ of $\operatorname{Spv}(K)$ is a quasicompact $T_1$-space, endowed with the induced topology. 
-By the dimension inequality \ref{rank-dimension-inequality} we have $\operatorname{Spv}^{(n)}(K)\subset |\operatorname{Spv}(K)|$ where $n=\dim(K)$.
 \end{lemma}
+
+\begin{proof}
+See [ZS] VI §7 Theorem 40.
+\end{proof}
 
@@
 \[\mathcal{O}_w\subset\mathcal{O}_v \iff w\in\overline{\{v\}} \]
-Thus $v$ is a closed point if and only if it has no non-trivial specializations or iff $\dim k(v)=0$, i.e. $k(v)$ has positive characteristic and is algebraic over its prime field.
+Thus $v$ is a closed point if and only if it has no non-trivial specializations or iff $\dim k(v)=0$, i.e.
+$k(v)$ has positive characteristic and is algebraic over its prime field.
+By the dimension inequality \ref{rank-dimension-inequality} we have $\operatorname{Spv}^{(n)}(K)\subset |\operatorname{Spv}(K)|$ where $n=\dim(K)$.
 \end{lemma}
@@
 Thus they are of rank $2$ and by Lemma \ref{rank-dimension-inequality} we can see $w_1,w_2\in\operatorname{Spv}_\text{rd}^{(2)}(K)$.
-Set $v_i=v(w_i)$.
+Set $v_i=v(w_i)\in\operatorname{Spv}^{(1)}_\text{rd}(K)$.
 By Lemma \ref{2.2-pre} $K$ is henselian with respect to both $v_1$ and $v_2$.
@@
 \begin{lemma}\label{2.4-closedset}
-Let $K$ be finitely generated, $\ell\neq 2$ be a prime and $f\in\mathrm{H}^3(K)$. Let $\mu_\ell\subset K$ and $K$ not formally real. Then the support of $f$ is closed in $|\operatorname{Spv}(K)|$:
+Let $f\in\mathrm{H}^3(K)$. Then the support of $f$ is closed in $|\operatorname{Spv}(K)|$:
 \[ \operatorname{supp}(f) = \{w\in |\operatorname{Spv}(K)|\mid \operatorname{res}_{K_w/K}(f)\neq 0 \} \]
@@
 Now, one easily checks that
-\[ \operatorname{supp}(f) = \bigcap_{\mathfrak{M}_K} \pi_\mathfrak{X}^{-1}(\{ y\in\mathfrak{X}_0\mid \operatorname{res}_{K_y/K}(f)\neq 0 \}) \]
+\[ \operatorname{supp}(f) = \bigcap_{\mathfrak{X}\in \mathfrak{M}(K)} \pi_\mathfrak{X}^{-1}(\{ y\in\mathfrak{X}_0\mid \operatorname{res}_{K_y/K}(f)\neq 0 \}) \]
 is closed.
@@
 Just like in Lemma \ref{2.4-finitelygenerated} we can assume that $\mu_\ell\subset K$ and $K$ is not formally real and replace $\mathbb{Z}/\ell\mathbb{Z}$ by $\mathbb{Z}/\ell\mathbb{Z}(2)$.
-We adopt the notation from the previous proof.
+We adopt the notation from the previous proofs.
 
@@
 \[ \varphi_{ij}: |\operatorname{Spv}(K_j)|\to |\operatorname{Spv}(K_i)| \]
-be the restriction maps. They are continuous. Since $|\operatorname{Spv}(K)|$ is quasicompact we can find a finite set $S\subset |\operatorname{Spv}(K)|$ such that:
+be the restriction maps. They are continuous. Since $|\operatorname{Spv}(K)|$ is quasicompact by Lemma \ref{spv-topology} we can find a finite set $S\subset |\operatorname{Spv}(K)|$ such that:
 \[ \bigcup_{w\in S}\varphi^{-1}_{i(w)}(U_{i(w)}) = |\operatorname{Spv}(K)| \]
@@
 The uniqueness of $w$ is again a consequence of Lemma \ref{2.2}.
-\item Choose the prime $\ell$ different from $\operatorname{char} k(w)$.
+\item Choose the prime $\ell$ different from $\operatorname{char} k(w)$ so that (i) holds.
 According to Lemma \ref{2.5} we have $\operatorname{rrank}(w)=\dim_\mathbb{Q}(\Gamma_w\otimes_\mathbb{Z}\mathbb{Q} )\geq \dim_{\mathbb{F}_\ell}(\Gamma_w /\ell\Gamma_w) = 2 = \dim(K)$.
@@
 If $E=K_v\subset\overline{K}$ is a henselization of $K$ with respect to $v\in\operatorname{Spv}_\text{d}^{(1)}(K)$, then $E'$ is a henselization of $K'$ with respect to $v'=\sigma(v)$.
+
 Furthermore, if $E^{\text{nr}}$ and $E'^{\text{nr}}$ denote the maximal unramified field extensions of $E$ and $E'$ respectively, then $\sigma$ induces an isomorphism:
@@
 \[ v(w_1)=v(w_2) \iff v(\sigma(w_1)) = v(\sigma(w_2)) \]
-This shows that $\sigma$ also induces a bijection $\sigma:\operatorname{Spv}_\text{d}^{(1)}(K)\to\operatorname{Spv}_\text{d}^{(1)}(K')$ and for all $w\in\operatorname{Spv}_\text{rd}^{(2)}(K)$ it holds $\sigma(v(w)) = v(\sigma(w))$.
+This shows that $\sigma$ also induces a bijection $\sigma:\operatorname{Spv}_\text{d}^{(1)}(K)\to\operatorname{Spv}_\text{d}^{(1)}(K')$ and for all $w\in\operatorname{Spv}_\text{rd}^{(2)}(K)$ we have $\sigma(v(w)) = v(\sigma(w))$.
 
@@
 
-By Lemma \ref{3.2} $\sigma$ induces a bijection $\sigma: X_0\to X'_0$ and for any $P\in X_0$ an isomorphism $\varphi_P: k(\sigma(P))\to k(P)$.
+By Lemma \ref{3.2} $\sigma$ induces a bijection $\sigma: X_0\to X'_0,\ P\mapsto P'$ and for any $P\in X_0$ an isomorphism $\varphi_P: k(P')\to k(P)$.
 
@@
 \begin{proof}
-First we suppose that $k/\mathbb{Q}$ is a Galois extension and $X$ has a $k$-rational point $P_0$. We can consider $k'$ as a subfield of $k$ via: 
+First we assume that $k/\mathbb{Q}$ is a Galois extension and $X$ has a $k$-rational point $P_0$. We can consider $k'$ as a subfield of $k$ via: 
 \[\begin{tikzcd}
-k'\ar[r] & k(\sigma(P_0))\ar[r, "\varphi_{P_0}"', "\cong"] & k(P_0) \ar[r, "\cong"] & k
+k'\ar[r] & k(P'_0)\ar[r, "\varphi_{P_0}"', "\cong"] & k(P_0) \ar[r, "\cong"] & k
 \end{tikzcd}\]

### 2017-01-09 Master Thesis/synopsis.tex
Add Hilbert's Irreducibility Theorem

@@
 \begin{theorem}[\textit{Hilbert's Irreducibility Theorem}]\label{hilberts-irreducibility-theorem}
-Inhalt...
+Let $E$ be a number field and $f_1,\ldots, f_n$ be irreducible polynomials in $E[X_1,\ldots,X_r,Y_1,\ldots,Y_s]$. Then there exists $a_1,\ldots,a_r\in E$ such that the specializations
+\[ f_1(a_1,\ldots,a_r,Y_1,\ldots,Y_s),\ \ldots,\ f_n(a_1,\ldots,a_r,Y_1,\ldots,Y_s) \]
+are irreducible in $E[Y_1,\ldots,Y_s]$.
 \end{theorem}

### 2017-01-10 Master Thesis/synopsis.tex
Change polynomial variables
Add proof of 4.1

@@
 \begin{theorem}[\textit{Hilbert's Irreducibility Theorem}]\label{hilberts-irreducibility-theorem}
-Let $E$ be a number field and $f_1,\ldots, f_n$ be irreducible polynomials in $E[X_1,\ldots,X_r,Y_1,\ldots,Y_s]$. Then there exists $a_1,\ldots,a_r\in E$ such that the specializations
-\[ f_1(a_1,\ldots,a_r,Y_1,\ldots,Y_s),\ \ldots,\ f_n(a_1,\ldots,a_r,Y_1,\ldots,Y_s) \]
+Let $E$ be a number field and $f_1,\ldots, f_n$ be irreducible polynomials in $E[X_1,\ldots,X_r,Y_1,\ldots,Y_s]$. Then there exists $e_1,\ldots,e_r\in E$ such that the specializations
+\[ f_1(e_1,\ldots,e_r,Y_1,\ldots,Y_s),\ \ldots,\ f_n(e_1,\ldots,e_r,Y_1,\ldots,Y_s) \]
 are irreducible in $E[Y_1,\ldots,Y_s]$.
@@
 Let $f: X\to\mathbf{P}^1_E$ be a non-constant morphism and let $f' =  f\circ \pi: X'\to\mathbf{P}^1_E$.
-By Hilbert's Irreducibility Theorem \ref{hilberts-irreducibility-theorem} there exists a closed $E$-rational point $P$ of $\mathbf{P}^1_E$ which has only one pre-image under $f'$.
-Consequently, $P$ only has one pre-image under $f$, say $P'$, and $P'$ has only one pre-image under $\pi$.
+This induces a finite field extension $F/E(T)$ of their function fields.
+Since we're in characteristic $0$, $F/E(T)$ is also separable, and thus we have a primitive element $a\in F$, i.e.
+$F = E(T,a)$.
+
+Consider the minimal polynomial of $a$ in $E(T)[A]$.
+Clearing out the denominators, we get an irreducible polynomial $f\in E[T,A]$.
+By Hilbert's Irreducibility Theorem \ref{hilberts-irreducibility-theorem} we can find an $e\in E$ such that $f(e,A)\in E[A]$ is irreducible.
+
+Now $(T-e)\in\mathbf{A}^1_E$ corresponds to a closed point $P\in \mathbf{P}^1_E$, which in turn corresponds to a valuation in $E(T)$.
+So the points in the fiber $f'^{-1}(P)$ corresponds to valuations in $F$ above $T-e$.
+But $T-e$ is inert in $F/E(T)$ since $a\in E[T,a]/(T-e)$ has minimal polynomial $f(e,A)$ over $E[T]/(T-e)$ and $\deg f(e,A) = [F:E(T)]$.
+
+This shows that $P$ has only one pre-image under $f'$. Therefore, it has only one pre-image under $f$, say $P'$, and $P'$ has only one pre-image under $\pi$. This concludes the proof.
 \end{proof}
@@
 We see any embedding $k\to \overline{k(P')}$ maps $k$ into $k(P')$ because $k/\mathbb{Q}$ is Galois. Thus we have for all $P'\in X'_0$:
-\[ k\otimes_{k'} k(P')\cong \prod_{\operatorname{Hom}_{k'}(k, k(P'))} k(P') \]
-in which case $X'_k\to X'$ splits completely. By Lemma \ref{4.1} we get $k\cong k'$, hence $[k:\mathbb{Q}]= [k':\mathbb{Q}]$.
+\[ k(P')\otimes_{k'} k\cong \prod_{\operatorname{Hom}_{k'}(k, k(P'))} k(P') \]
+in which case $X'\times_{k'}\operatorname{Spec} (k)\to X'$ splits completely. By Lemma \ref{4.1} we get $k\cong k'$, hence $[k:\mathbb{Q}]= [k':\mathbb{Q}]$.
 
@@
 Again, by F.K.
-Schmidt, we have $h=1$ and thus $\varphi = \psi$.
+Schmidt \ref{theorem-fkschmidt}, we have $h=1$ and thus $\varphi = \psi$.
 \end{proof}

### 2017-01-10 Master Thesis/synopsis.tex
Update

@@
 
-Now $(T-e)\in\mathbf{A}^1_E$ corresponds to a closed point $P\in \mathbf{P}^1_E$, which in turn corresponds to a valuation in $E(T)$.
+Now $T-e\in E[T]$ corresponds to a closed point $P\in \mathbf{P}^1_E$, which in turn corresponds to a valuation in $E(T)$.
 So the points in the fiber $f'^{-1}(P)$ corresponds to valuations in $F$ above $T-e$.
@@
 \begin{proof}
-With base changing to a Galois extension of $F$, we can assume that $C$ has $F$-rational points. We have the following exact sequence:
+We have the following exact sequence:
 \[ 0\longrightarrow \operatorname{Pic}^0(C) \longrightarrow\operatorname{Pic}(C) \longrightarrow \mathbb{Z} \longrightarrow 0  \]
 Now $\operatorname{Pic}^0(C)$ can be identified with the $F$-rational points of the Jacobian variety $J$ of $C$.
-Now $J(F)$ is finitely generated by the Mordell-Weil Theorem \ref{mordell-weil-theorem}, hence $\operatorname{Pic}(C)$ is finitely generated as well.
+$J(F)$ is finitely generated by the Mordell-Weil Theorem \ref{mordell-weil-theorem}, hence $\operatorname{Pic}(C)$ is finitely generated as well.
 \end{proof}
@@
 $C_2$ is free abelian of finite rank as well.
-By requirement $r_{P_2}(x)=a$ and $a\in k^\times$ is not of finite order.
+By requirement $r_{P_2}(x)=a$ and $a\in k^\times$ is not of finite order, thus $x\not\in C_2$.
 It follows $\operatorname{rank}(C_2) < \operatorname{rank}(C_1)$.
-By repeating this process until $C_r=1$, we get points $P_2,\ldots,P_r\in U_0$, such that the following map is injective:
+By repeating this process until $C_r=1$, we get points $P_2,\ldots,P_r\in U_0\setminus S$, such that the following map is injective:
 \[ C_1\to\prod_{i=2}^{r} k(P_i),\ x\mapsto(r_{P_i}(x))_i \]

### 2017-01-11 Master Thesis/synopsis.tex
Update

@@
 \begin{proof}
+By Lemma \ref{5.4}, the following maps are injective for all finite subsets $S\subset X_0$:
+\[ r_S:K^S \to\prod_{P\in X_0\setminus S} k(P)^\times,\ x\mapsto (r_P(x))_P \]
 Since any rational function on $X$ is uniquely determined by its values at those points where it is defined, $\varphi$ is unique.
-In other words, the following maps are injective for all finite subsets $S\subset X_0$:
-\[ r_S:K^S \to\prod_{P\in X_0\setminus S} k(P)^\times,\ x\mapsto (r_P(x))_P \]
+
 For existence, let $(S,T)$ be an admissible pair. First, we show that there is a unique homomorphism $\varphi^{S,T}$, making the following diagram commutative:
@@
 
-Note that for every $x\in K^\times$ we have $v_P(x)= 0$ for almost all $P\in X_0$; similarly for $K'$ and $X'_0$.
-By Lemma \ref{4.5} (i) we can find an admissible set $S\subset X_0\setminus \{P\}$ such that $x\in K'^{S'}$ for any $x\in U_{P'}$.
+The commutativity of the diagram follows from the commutativity of $(\star)$ with $T=X_0\setminus S$.
+Note that with Lemma \ref{4.5} (i), we can find for all $x\in U_{P'}$ an admissible set $S\subset X_0\setminus \{P\}$ such that $x\in K'^{S'}$.
 \end{proof}
@@
 For any finite subextension $L/K$ of $\overline{K}/K$ there exists an isomorphism $\varphi_L: L'\to L$.
-If $K\subset L\subset M$ are finite subextensions, we have $\varphi_M|_{L'} = \varphi_{L'}$, since $r_Q(\varphi_M(x)) = r_Q(\varphi_L(x))$ for all $x\in L'^\times$ and almost all $Q\in (X_M)_0$.
+If $K\subset L\subset M$ are finite subextensions, we have $\varphi_M|_{L'} = \varphi_{L'}$, since $r_Q(\varphi_M(x)) = r_Q(\varphi_L(x))$ for all $x\in L'^\times$ and almost all $Q\in (X_M)_0$ according to Remark \ref{3.3} and Step 1.
 By passing to limits we obtain an isomorphism $\varphi: \overline{K}'\to \overline{K}$.

### 2017-01-13 Extra/PlotFunction.m
Remove 'Extra'

@@ -1,12 +1,0 @@
-function PlotFunction(x)
-    t = [0.200, 0.200, 0.222, 0.222, 0.286, 0.286, 0.400, 0.400, 0.667, 0.667, 2.000, 2.000]';
-    y = [0.1024, 0.0952, 0.1116, 0.1012, 0.1169, 0.1066, 0.1693, 0.1803, 0.2428, 0.2366, 0.4846, 0.4452]';
-    h = @(t) x(1) .* t ./ (x(2) + t);
-    range = (0:1000) * 2 / 1000;
-    
-    hold on
-    plot(t, y, 'x');
-    plot(range, h(range), '-');
-    hold off
-end
-

### 2017-01-13 Extra/VGN.m

@@ -1,28 +1,0 @@
-function [xstar, iterations] = VGN(f, J, x0)
-    % f is a vector function consisting of the 
-    % residue functions with respect to x
-    %
-    % J is the Jacobian matrix function of f 
-    % with respect to x
-
-    iterations = 0;
-    xstar = x0;
-    
-    while 1
-        xold = xstar;
-        
-        % Calculate new xstar
-        D = J(xold);
-        r = f(xold);
-        xstar = xold - (D' * D) \ (D' * r);
-        
-        % Increment iterations
-        iterations = iterations + 1;
-
-        % Stop condition
-        if max(abs(xstar - xold)) <= 10^(-8)
-            return
-        end
-    end
-end
-

### 2017-01-13 Extra/main.m

@@ -1,9 +1,0 @@
-t = [0.200, 0.200, 0.222, 0.222, 0.286, 0.286, 0.400, 0.400, 0.667, 0.667, 2.000, 2.000]';
-y = [0.1024, 0.0952, 0.1116, 0.1012, 0.1169, 0.1066, 0.1693, 0.1803, 0.2428, 0.2366, 0.4846, 0.4452]';
-h = @(t, x) x(1) .* t ./ (x(2) + t);
-f = @(x) y - h(t, x);
-J = @(x) [-t ./ (x(2) + t), x(1) .* t ./ (x(2) + t).^2];
-x0 = [1, 2]';
-
-[xstar, iterations] = VGN(f, J, x0)
-PlotFunction(xstar);

### 2017-01-13 Master Thesis/synopsis.tex
Close 2.6 (ii)
Formulation

@@
 \begin{enumerate}[(i)]
-\item $\operatorname{Spv}(K)$, together with the maps $\pi_{\mathfrak{X}}$, is the projective limit of topological spaces: 
+\item $\operatorname{Spv}(K)$, together with the maps $\pi_{\mathfrak{X}}$, is the following projective limit in the category of topological spaces: 
 \[\operatorname{Spv}(K) =\varprojlim_{\mathfrak{X}\in\mathfrak{M}(K)}\mathfrak{X}\]
@@
 Therefore, $f$ is injective as well and its image is contained in $\bigoplus\mathrm{H}^3(K_y)$.
-Now, one easily checks that
+Now, one easily checks that:
 \[ \operatorname{supp}(f) = \bigcap_{\mathfrak{X}\in \mathfrak{M}(K)} \pi_\mathfrak{X}^{-1}(\{ y\in\mathfrak{X}_0\mid \operatorname{res}_{K_y/K}(f)\neq 0 \}) \]
-is closed.
+Note that the right set is closed in $\operatorname{Spv}(K)$.
 \end{proof}
@@
 \item Choose the prime $\ell$ different from $\operatorname{char} k(w)$ so that (i) holds.
-According to Lemma \ref{2.5} we have $\operatorname{rrank}(w)=\dim_\mathbb{Q}(\Gamma_w\otimes_\mathbb{Z}\mathbb{Q} )\geq \dim_{\mathbb{F}_\ell}(\Gamma_w /\ell\Gamma_w) = 2 = \dim(K)$.
+According to Lemma \ref{2.5} we have $\operatorname{rrank}(w)=\dim_\mathbb{Q}(\Gamma_w\otimes_\mathbb{Z}\mathbb{Q} )\leq \dim_{\mathbb{F}_\ell}(\Gamma_w /\ell\Gamma_w) = 2$.
 By Lemma \ref{rank-dimension-inequality}, this inequality is actually an equality.
+\qedhere
 \end{enumerate}
@@
 Let $\ell_1,\ell_2$ be two different prime numbers different from $\operatorname{char} k(w)$ and let $E_0 = E(\mu_{\ell_1\ell_2})$.
-Then $E_0/E$, and also $E'_0/E'$, satisfies the condition of \ref{2.6} (ii).
+Then, by Lemma \ref{2.5}, $E_0/E$, and also $E'_0/E'$, satisfy the condition of Proposition \ref{2.6} (ii).
 Thus there exists a unique henselian defectless valuation $w'_{E'}$ of $E'$, so $E'$ contains a henselization $K'_{w'}$ of $w' = w'_{E'}|_{K'}$.

### 2017-01-16 Master Thesis/synopsis.tex
Clarify 4.2-pre

@@
 \end{tikzcd} \]
-We see any embedding $k\to \overline{k(P')}$ maps $k$ into $k(P')$ because $k/\mathbb{Q}$ is Galois. Thus we have for all $P'\in X'_0$:
+Let us consider the morphism $f:X'\times_{k'} k\to X'$. We want to see that $f$ splits completely. For $P'\in X'_0$ we have as topological spaces:
+\[ f^{-1}(P') = \operatorname{Spec} k(P')\times_{k'} k \]
+Any embedding $k\to \overline{k(P')}$ maps $k$ into $k(P')$ because $k/\mathbb{Q}$ is Galois. Thus we have:
 \[ k(P')\otimes_{k'} k\cong \prod_{\operatorname{Hom}_{k'}(k, k(P'))} k(P') \]
-in which case $X'\times_{k'}\operatorname{Spec} (k)\to X'$ splits completely. By Lemma \ref{4.1} we get $k\cong k'$, hence $[k:\mathbb{Q}]= [k':\mathbb{Q}]$.
+Now we can see that $f$ splits completely indeed. By Lemma \ref{4.1} we get $k\cong k'$, hence $[k:\mathbb{Q}]= [k':\mathbb{Q}]$.
 
@@
 \begin{align*}
-[k:\mathbb{Q}] &= [k_1:\mathbb{Q}]\cdot [K_1:K]^{-1} = [k'_1:\mathbb{Q}]\cdot [K'_1:K']^{-1}\\
+[k:\mathbb{Q}] &= [k_1:\mathbb{Q}]\cdot [K_1:K]^{-1} \\
+&= [k'_1:\mathbb{Q}]\cdot [K'_1:K']^{-1}\\
 &= [k'_1:\mathbb{Q}]\cdot ([K'_1: K'k_1'] \cdot [K'k_1':K'])^{-1}\\
-&= [k'_1:\mathbb{Q}]\cdot  ([K'_1: K'k_1'] \cdot [k'_1:k'])^{-1} \leq [k'_1:\mathbb{Q}]\cdot [k'_1:k']^{-1} = [k':\mathbb{Q}]
+&= [k'_1:\mathbb{Q}]\cdot  ([K'_1: K'k_1'] \cdot [k'_1:k'])^{-1} \\
+&\leq [k'_1:\mathbb{Q}]\cdot [k'_1:k']^{-1} = [k':\mathbb{Q}]
 \end{align*}

### 2017-01-16 Master Thesis/synopsis.tex
Name chapters & sections
Update 2.5

@@
-\chapter{Synopsis}
+\chapter{Valuations}
 
-\section{Some Results on the valuation spectra}
+\section{The Valuation Spectra}
 
@@
 \end{lemma}
+
+\section{The Main Lemma}
 
@@
 \begin{proof}
-Let $G_\ell$ be an pro-$\ell$-Sylow group of $G_K$.
+Let $G_\ell$ be an pro-$\ell$ Sylow group of $G_K$.
 By the ramification theory for general valuations ([E], §20) we have $G_\ell\cong\mathbb{Z}_\ell(1)^r\rtimes G_k(\ell)$ where the action is defined via the cyclotomic character and $r = \dim_{\mathbb{F}_\ell}(\Gamma/\ell\Gamma)$.
@@
 \[ \mathrm{H}^3(K,\mathbb{Z}/\ell\mathbb{Z}) \cong\mathrm{H}^3(G_K(\ell),\mathbb{Z}/\ell\mathbb{Z}) \cong\mathrm{H}^3(G_\ell,\mathbb{Z}/\ell\mathbb{Z}) \]
-Now the equivalence of (i) and (ii) follows from [Se], Ch. I, §4.
+Now the first assertion follows from [Se], Ch. I, §4.
 
@@
 for some $d \leq \infty$, where $\Gamma'$ denotes the value group and $k'$ the residue field of the unique prolongation of $w$ to $L$.
+Note that $(\Gamma':\Gamma)< \infty$ and $\operatorname{rank}(w) = \operatorname{rank}(v)$ in an algebraic field extension ([ZS], Ch.
+VI, §11, Lemma 1 and 2).
 \end{proof}
@@
 
-\section{Local Correspondence}
+\chapter{Correspondences}
 
@@
 In other words $\sigma G_L=G_{L'}$.
+
+\section{Local Correspondence}
 
@@
 
+Let $k$ and $k'$ be the algebraic closures of $\mathbb{Q}$ in $K$ and $K'$ respectively, i.e.
+the constant fields of $K$ and $K'$, and let $\overline{k}$ and $\overline{k'}$ be the algebraic closure of $k$ and $k'$ in $\overline{K}$ and $\overline{K'}$ respectively.
+
+Let $X$ and $X'$ be the unique smooth projective curves over $k$ and $k'$ with function field $K$ and $K'$ respectively.
+We identify the set of closed points $X_0$ of $X$ with the set of valuations $v\in\operatorname{Spv}_\text{d}^{(1)}(K)$ with $\operatorname{char} k(v)=0$ and similarly identify $X'_0$ with the corresponding subset of $\operatorname{Spv}_\text{d}^{(1)}(K')$.
+
+By Lemma \ref{3.2} $\sigma$ induces a bijection $\sigma: X_0\to X'_0,\ P\mapsto P'$ and for any $P\in X_0$ an isomorphism $\varphi_P: k(P')\to k(P)$.
+
 \begin{theorem}[\textit{Hilbert's Irreducibility Theorem}]\label{hilberts-irreducibility-theorem}
@@
 \end{proof}
-
-\paragraph{} Let $k$ and $k'$ be the algebraic closures of $\mathbb{Q}$ in $K$ and $K'$ respectively, i.e.
-the constant fields of $K$ and $K'$, and let $\overline{k}$ and $\overline{k'}$ be the algebraic closure of $k$ and $k'$ in $\overline{K}$ and $\overline{K'}$ respectively.
-
-Let $X$ and $X'$ be the unique smooth projective curves over $k$ and $k'$ with function field $K$ and $K'$ respectively.
-We identify the set of closed points $X_0$ of $X$ with the set of valuations $v\in\operatorname{Spv}_\text{d}^{(1)}(K)$ with $\operatorname{char} k(v)=0$ and similarly identify $X'_0$ with the corresponding subset of $\operatorname{Spv}_\text{d}^{(1)}(K')$.
-
-By Lemma \ref{3.2} $\sigma$ induces a bijection $\sigma: X_0\to X'_0,\ P\mapsto P'$ and for any $P\in X_0$ an isomorphism $\varphi_P: k(P')\to k(P)$.
 
@@
 
-\begin{theorem}[\textit{Mordell-Weil Theorem}]\label{mordell-weil-theorem}
-Let $A$ be an abelian variety over a number field $F$. Then the abelian group $A(F)$ of $F$-rational points is finitely generated.
-\end{theorem}
+\chapter{Theorem}
 
-\begin{corollary}\label{mordell-weil-corollary}
-Let $C$ be a smooth projective curve over a number field $F$. Then the Picard group $\operatorname{Pic}(C)$ is finitely generated.
-\end{corollary}
-
-\begin{proof}
-We have the following exact sequence:
-\[ 0\longrightarrow \operatorname{Pic}^0(C) \longrightarrow\operatorname{Pic}(C) \longrightarrow \mathbb{Z} \longrightarrow 0  \]
-Now $\operatorname{Pic}^0(C)$ can be identified with the $F$-rational points of the Jacobian variety $J$ of $C$.
-$J(F)$ is finitely generated by the Mordell-Weil Theorem \ref{mordell-weil-theorem}, hence $\operatorname{Pic}(C)$ is finitely generated as well.
-\end{proof}
-
-\begin{definition}\label{4.4}
-For a finite subset $S\subset X_0$ set:
-\[ K^S=\{x\in K^\times \mid v_P(x)=0\text{ for all }P\in X_0\setminus S \} \]
-$K'^{S'}$ is defined similarly. For finite subsets $S$ of $X_0$, the corresponding subsets of $X_0'$ will be denoted by $S'$, i.e. $S'=\{\sigma P\mid P\in S \} $.
-\end{definition}
-
-\begin{lemma}\phantomsection\label{4.5}
-\begin{enumerate}[(i)]
-\item There exists a finite subset $S\subset X_0$ such that:
-\[ \operatorname{Pic}(X\setminus S)=0=\operatorname{Pic}(X'\setminus S') \]
-Such a set $S$ is called \textit{admissible}. If $V\subset X$ is an open subscheme, then one can choose $S$ in $V_0$.
-\item Let $S$ be admissible. For all $n\in\mathbb{N}$ the following canonical maps are injective:
-\[ K^S/n \to K^\times /n,\quad K'^{S'}/n\to K'^\times /n \]
-$\varphi_n$ induces isomorphisms $\varphi_n: K'^{S'}/n\to K^S/n$.
-\end{enumerate}
-\end{lemma}
-
-\begin{proof}
-\begin{enumerate}[(i)]
-\item By Corollary \ref{mordell-weil-corollary}, the Picard group of $X$ is finitely generated.
-Let $D_1,\ldots, D_r$ be Weil divisors of $X$ such that the associated line bundles generate $\operatorname{Pic}(X)$ and similarly let $D'_1,\ldots,D'_s$ generate $\operatorname{Pic}(X')$.
-Clearly any finite $S\subset X_0$ such that
-\[ \bigcup_{i=1}^r \operatorname{supp}D_i\subset S\quad\text{and}\quad \bigcup_{j=1}^s\operatorname{supp}D'_j\subset S' \]
-is admissible. Now let $V\subset X$ be an open subscheme. By the approximation theorem, we can choose the $D_i$ and the $D'_j$ in $V_0$ and $\sigma (V_0)$ respectively.
-\item Observe that the following sequence is exact:
-\[ \begin{tikzcd}
-0 \ar[r] & K^S \ar[r]& K^\times \ar[r, "\operatorname{div}"]& \displaystyle \bigoplus_{P\in X_0\setminus S}\mathbb{Z} \ar[r]& 0
-\end{tikzcd} \]
-Since the right module is $\mathbb{Z}$-flat, tensoring with $\mathbb{Z}/n\mathbb{Z}$ yields another exact sequence and we can consider $K^S/n$ as a subgroup of $K^\times/n$. Then:
-\[ K^S/n = \bigcap_{P\in X_0\setminus X}U_P/n \]
-The analogous result holds for $K'^{S'}/n$. Thus the assertion follows from Lemma \ref{4.3} (i).\qedhere
-\end{enumerate}
-\end{proof}
-
-\section{The Proof}
+\section{Free-By-Finite Groups}
 
@@
 
+\section{Admissible Sets}
+
+\begin{theorem}[\textit{Mordell-Weil Theorem}]\label{mordell-weil-theorem}
+Let $A$ be an abelian variety over a number field $F$. Then the abelian group $A(F)$ of $F$-rational points is finitely generated.
+\end{theorem}
+
+\begin{corollary}\label{mordell-weil-corollary}
+Let $C$ be a smooth projective curve over a number field $F$. Then the Picard group $\operatorname{Pic}(C)$ is finitely generated.
+\end{corollary}
+
+\begin{proof}
+We have the following exact sequence:
+\[ 0\longrightarrow \operatorname{Pic}^0(C) \longrightarrow\operatorname{Pic}(C) \longrightarrow \mathbb{Z} \longrightarrow 0  \]
+Now $\operatorname{Pic}^0(C)$ can be identified with the $F$-rational points of the Jacobian variety $J$ of $C$.
+$J(F)$ is finitely generated by the Mordell-Weil Theorem \ref{mordell-weil-theorem}, hence $\operatorname{Pic}(C)$ is finitely generated as well.
+\end{proof}
+
+\begin{definition}\label{4.4}
+For a finite subset $S\subset X_0$ set:
+\[ K^S=\{x\in K^\times \mid v_P(x)=0\text{ for all }P\in X_0\setminus S \} \]
+$K'^{S'}$ is defined similarly. For finite subsets $S$ of $X_0$, the corresponding subsets of $X_0'$ will be denoted by $S'$, i.e. $S'=\{\sigma P\mid P\in S \} $.
+\end{definition}
+
+\begin{lemma}\phantomsection\label{4.5}
+\begin{enumerate}[(i)]
+\item There exists a finite subset $S\subset X_0$ such that:
+\[ \operatorname{Pic}(X\setminus S)=0=\operatorname{Pic}(X'\setminus S') \]
+Such a set $S$ is called \textit{admissible}. If $V\subset X$ is an open subscheme, then one can choose $S$ in $V_0$.
+\item Let $S$ be admissible. For all $n\in\mathbb{N}$ the following canonical maps are injective:
+\[ K^S/n \to K^\times /n,\quad K'^{S'}/n\to K'^\times /n \]
+$\varphi_n$ induces isomorphisms $\varphi_n: K'^{S'}/n\to K^S/n$.
+\end{enumerate}
+\end{lemma}
+
+\begin{proof}
+\begin{enumerate}[(i)]
+\item By Corollary \ref{mordell-weil-corollary}, the Picard group of $X$ is finitely generated.
+Let $D_1,\ldots, D_r$ be Weil divisors of $X$ such that the associated line bundles generate $\operatorname{Pic}(X)$ and similarly let $D'_1,\ldots,D'_s$ generate $\operatorname{Pic}(X')$.
+Clearly any finite $S\subset X_0$ such that
+\[ \bigcup_{i=1}^r \operatorname{supp}D_i\subset S\quad\text{and}\quad \bigcup_{j=1}^s\operatorname{supp}D'_j\subset S' \]
+is admissible. Now let $V\subset X$ be an open subscheme. By the approximation theorem, we can choose the $D_i$ and the $D'_j$ in $V_0$ and $\sigma (V_0)$ respectively.
+\item Observe that the following sequence is exact:
+\[ \begin{tikzcd}
+0 \ar[r] & K^S \ar[r]& K^\times \ar[r, "\operatorname{div}"]& \displaystyle \bigoplus_{P\in X_0\setminus S}\mathbb{Z} \ar[r]& 0
+\end{tikzcd} \]
+Since the right module is $\mathbb{Z}$-flat, tensoring with $\mathbb{Z}/n\mathbb{Z}$ yields another exact sequence and we can consider $K^S/n$ as a subgroup of $K^\times/n$. Then:
+\[ K^S/n = \bigcap_{P\in X_0\setminus X}U_P/n \]
+The analogous result holds for $K'^{S'}/n$. Thus the assertion follows from Lemma \ref{4.3} (i).\qedhere
+\end{enumerate}
+\end{proof}
+
 \begin{lemma}\label{5.4}
@@
 \end{definition}
+
+\section{The Proof}
 

### 2017-01-17 Master Thesis/synopsis.tex
Clarify 5.4
Finish explaining proof of 4.3

@@
 \begin{enumerate}[(i)]
-\item  We have $(K^\text{nr}_P)^\times/n\cong \mathbb{Z}/n\mathbb{Z}$ since $\operatorname{char}k(P) = 0$.
-The canonical map $K^\times/n \to (K^\text{nr}_P)^\times/n$ can be identified with the map $K^\times /n\to\mathbb{Z}/n\mathbb{Z}$ induced by $v_P$.
-Now the following commutative diagram gives us the result:
+\item Consider the maximal unramified extension $L = K_P^\text{nr}$ of $K_P$ and let $w$ be the unique extension of $v_P$ in $L$.
+Now $k(w) = k(v_P)^\text{s}$.
+Since $w$ is unramified over $v$, $w$ is discrete as well.
+We have the following exact sequence:
+\[ 1 \longrightarrow \mathcal{O}_w^\times/n \longrightarrow L^\times /n\longrightarrow \mathbb{Z}/n\mathbb{Z}\longrightarrow 0 \]
+Since $k(w)$ is separably closed, with Hensel's lemma and $\operatorname{char} k(w)=0$ we can see that $\mathcal{O}_w^\times$ is a divisible group, hence we get an isomorphism $L^\times /n\cong \mathbb{Z}/n\mathbb{Z}$ induced by $w$.
+Therefore, the canonical map $K^\times/n\to L^\times/n$ can be identified with the map $K^\times/n \to\mathbb{Z}/n\mathbb{Z}$ induced by $v_P$.
+Now the following commutative diagram with exact columns gives us the result:
 \[ \begin{tikzcd}
@@
 K'^\times /n \ar[r, "\varphi_n"]\ar[d] & K^\times/n\ar[d]\\
-(K^\text{nr}_P)'^\times /n \ar[r, "\varphi_n"'] &  (K^\text{nr}_P)^\times /n
+L'^\times /n \ar[r, "\varphi_n"'] &  L^\times /n
 \end{tikzcd}\]
-Note that $(K^\text{nr}_P)' = (K'^{\text{nr}}_{P'})$ as proven in Lemma \ref{3.2}.
+Note that $(K^\text{nr}_P)' = K'^{\text{nr}}_{P'}$ as proven in Lemma \ref{3.2}.
 \item Consider the field extension $K_P/K$. Without loss of generality, we can assume $K=K_P$. Then $r_P/n$ is an isomorphism and the composition
@@
 It follows $\operatorname{rank}(C_2) < \operatorname{rank}(C_1)$.
-By repeating this process until $C_r=1$, we get points $P_2,\ldots,P_r\in U_0\setminus S$, such that the following map is injective:
+By repeating this process until $C_r=1$, we get points $P_2,\ldots,P_r\in U_0$, such that the following map is injective:
 \[ C_1\to\prod_{i=2}^{r} k(P_i),\ x\mapsto(r_{P_i}(x))_i \]
-Finally for $T=\{P_1,\ldots,P_r\}$, the map $r_{S,T}$ is injective. For the second assertion, note that:
+Finally for $T=\{P_1,\ldots,P_r\}\subset U_0$, the map $r_{S,T}$ is injective. For the second assertion, note the following isomorphism with the snake lemma:
 \[ \operatorname{coker}(r_{S,T}) \cong\operatorname{coker}\Big(K^S/k^\times \to \prod_{P\in T}k(P)^\times /k^\times \Big) \]
@@
 
-\section{The Proof}
+\section{Proof of Theorem}
 

### 2017-01-17 Master Thesis/synopsis.tex
Write 1 instead of 0

@@
 Let $n\in\mathbb{N}$. Tensoring the exact sequence $0\to U_P\to K^\times \to\mathbb{Z}\to 0$ with $\mathbb{Z}/n\mathbb{Z}$ preserves its exactness since $\mathbb{Z}$ is $\mathbb{Z}$-flat:
-\[ 0\longrightarrow U_P/n\longrightarrow K^\times/n\longrightarrow\mathbb{Z}/n\mathbb{Z}\longrightarrow 0 \]
+\[ 1\longrightarrow U_P/n\longrightarrow K^\times/n\longrightarrow\mathbb{Z}/n\mathbb{Z}\longrightarrow 0 \]
 We'll use the same notation for $P' = \sigma(P)\in X'_0$.
@@
 We have the following exact sequence:
-\[ 1 \longrightarrow \mathcal{O}_w^\times/n \longrightarrow L^\times /n\longrightarrow \mathbb{Z}/n\mathbb{Z}\longrightarrow 0 \]
-Since $k(w)$ is separably closed, with Hensel's lemma and $\operatorname{char} k(w)=0$ we can see that $\mathcal{O}_w^\times$ is a divisible group, hence we get an isomorphism $L^\times /n\cong \mathbb{Z}/n\mathbb{Z}$ induced by $w$.
+\[ 1 \longrightarrow U_w/n \longrightarrow L^\times /n\longrightarrow \mathbb{Z}/n\mathbb{Z}\longrightarrow 0 \]
+Since $k(w)$ is separably closed, with Hensel's lemma and $\operatorname{char} k(w)=0$ we can see that $U_w$ is a divisible group, hence we get an isomorphism $L^\times /n\cong \mathbb{Z}/n\mathbb{Z}$ induced by $w$.
 Therefore, the canonical map $K^\times/n\to L^\times/n$ can be identified with the map $K^\times/n \to\mathbb{Z}/n\mathbb{Z}$ induced by $v_P$.
@@
 \[ \begin{tikzcd}
-0 \ar[d] & 0 \ar[d]\\
+1 \ar[d] & 1 \ar[d]\\
 U_{P'}/n \ar[r, dashed]\ar[d] & U_P/n\ar[d]\\

### 2017-01-18 Master Thesis/synopsis.tex
Now really finished with explaining proof of 4.3

@@
 \[ 1 \longrightarrow U_w/n \longrightarrow L^\times /n\longrightarrow \mathbb{Z}/n\mathbb{Z}\longrightarrow 0 \]
-Since $k(w)$ is separably closed, with Hensel's lemma and $\operatorname{char} k(w)=0$ we can see that $U_w$ is a divisible group, hence we get an isomorphism $L^\times /n\cong \mathbb{Z}/n\mathbb{Z}$ induced by $w$.
+Since $k(w)$ is separably closed, with $\operatorname{char} k(w)=0$ and Hensel's lemma we can see that $U_w$ is a divisible group, hence we get an isomorphism $L^\times /n\cong \mathbb{Z}/n\mathbb{Z}$ induced by $w$.
 Therefore, the canonical map $K^\times/n\to L^\times/n$ can be identified with the map $K^\times/n \to\mathbb{Z}/n\mathbb{Z}$ induced by $v_P$.
@@
 Note that $(K^\text{nr}_P)' = K'^{\text{nr}}_{P'}$ as proven in Lemma \ref{3.2}.
-\item Consider the field extension $K_P/K$. Without loss of generality, we can assume $K=K_P$. Then $r_P/n$ is an isomorphism and the composition
+\item Consider the field extension $K_P/K$. Without loss of generality, we can assume $K=K_P$. Since $\operatorname{char}k(P)=0$, by Hensel's lemma, $r_P/n$ is an isomorphism and the composition
 \[\begin{tikzcd}
@@
 \end{tikzcd}\]
-is the inflation $\mathrm{H}^1(k(P), \mu_n) \to \mathrm{H}^1(K_P,\mu_n)$.
+is the inflation $\operatorname{inf}: \mathrm{H}^1(k(P), \mu_n) \to \mathrm{H}^1(K_P,\mu_n)$.
 The same holds for $P'$ due to Lemma \ref{3.2}.

### 2017-01-18 Master Thesis/synopsis.tex
Solve TeX warnings
Clarify proof of 4.5

@@
 
-\begin{lemma}\label{2.4}
+\begin{lemma}[\textit{Main Lemma}]\label{2.4}
 Let $K$ be not necessarily finitely generated and $\ell\neq 2$ be a prime. The following map, induced by the restrictions, is injective:
@@
 
-\section{Global Correspondence mod $n$}
+\section{Global Correspondence mod \texorpdfstring{$n$}{n}}
 
@@
 \[ k(P')\otimes_{k'} k\cong \prod_{\operatorname{Hom}_{k'}(k, k(P'))} k(P') \]
-Now we can see that $f$ splits completely indeed. By Lemma \ref{4.1} we get $k\cong k'$, hence $[k:\mathbb{Q}]= [k':\mathbb{Q}]$.
+Now we can see that $f$ splits completely. By Lemma \ref{4.1} we get $[k:k'] = 1$, hence $[k:\mathbb{Q}]= [k':\mathbb{Q}]$.
 
@@
 \[ \begin{tikzcd}
-0 \ar[r] & K^S \ar[r]& K^\times \ar[r, "\operatorname{div}"]& \displaystyle \bigoplus_{P\in X_0\setminus S}\mathbb{Z} \ar[r]& 0
+1 \ar[r] & K^S \ar[r]& K^\times \ar[r, "\operatorname{div}"]& \displaystyle \bigoplus_{P\in X_0\setminus S}\mathbb{Z} \ar[r]& \operatorname{Pic}(X\setminus S) = 0
 \end{tikzcd} \]
-Since the right module is $\mathbb{Z}$-flat, tensoring with $\mathbb{Z}/n\mathbb{Z}$ yields another exact sequence and we can consider $K^S/n$ as a subgroup of $K^\times/n$. Then:
+Since $\bigoplus\mathbb{Z}$ is $\mathbb{Z}$-flat, tensoring with $\mathbb{Z}/n\mathbb{Z}$ yields another exact sequence and we can consider $K^S/n$ as a subgroup of $K^\times/n$. Then:
 \[ K^S/n = \bigcap_{P\in X_0\setminus X}U_P/n \]

### 2017-01-18 Master Thesis/synopsis.tex
Fix typo

@@
 \end{tikzcd} \]
-where the second map is induced by the isomorphism $\overline{K''}\to\overline{K'},\ x\otimes y\mapsto \varphi^{-1}(y)\cdot x$, we may assume that $k=k'$, $\overline{k} = \overline{k'}$ and $\varphi:\overline{k}\to\overline{k}$, $\sigma: G_k\to G_k$ are identities.
+where the second map is induced by the isomorphism $\overline{K''}\to\overline{K'},\ x\otimes y\mapsto \varphi^{-1}(y)\cdot x$, we may assume that $k=k'$, $\overline{k} = \overline{k'}$ and that $\varphi:\overline{k}\to\overline{k}$, $\sigma: G_k\to G_k$ are identities.
 \end{remark}
@@
 Note that $\varphi_P: k(P')\to k(P)$ is a field isomorphism and that the residue maps $r_{P}: \mathcal{O}_{P} \to k(P)$ are ring homomorphisms.
-Since $K^S\to \prod_{P\in X_0\setminus S} k(P)^\ast$ is injective, we get $\varphi(x+y) = \varphi(x) + \varphi(y)$.\qedhere
+Since $K^S\to \prod_{P\in X_0\setminus S} k(P)^\times$ is injective, we get $\varphi(x+y) = \varphi(x) + \varphi(y)$.\qedhere
 \end{itemize}

### 2017-01-21 Master Thesis/synopsis.tex
Fix typo
Simplify end proof step 1

@@
 Since $\bigoplus\mathbb{Z}$ is $\mathbb{Z}$-flat, tensoring with $\mathbb{Z}/n\mathbb{Z}$ yields another exact sequence and we can consider $K^S/n$ as a subgroup of $K^\times/n$. Then:
-\[ K^S/n = \bigcap_{P\in X_0\setminus X}U_P/n \]
+\[ K^S/n = \bigcap_{P\in X_0\setminus S}U_P/n \]
 The analogous result holds for $K'^{S'}/n$. Thus the assertion follows from Lemma \ref{4.3} (i).\qedhere
@@
 \begin{proof}
-By Lemma \ref{5.4}, the following maps are injective for all finite subsets $S\subset X_0$:
+By Lemma \ref{5.4}, the following maps are injective for all admissible subsets $S\subset X_0$:
 \[ r_S:K^S \to\prod_{P\in X_0\setminus S} k(P)^\times,\ x\mapsto (r_P(x))_P \]
-Since any rational function on $X$ is uniquely determined by its values at those points where it is defined, $\varphi$ is unique.
-
-For existence, let $(S,T)$ be an admissible pair. First, we show that there is a unique homomorphism $\varphi^{S,T}$, making the following diagram commutative:
+Therefore, $\varphi$ is unique. For existence, let $(S,T)$ be an admissible pair. First, we show that there is a unique homomorphism $\varphi^{S,T}$, making the following diagram commutative:
 \[ (\star) \qquad \begin{tikzcd}

### 2017-01-21 Master Thesis/synopsis.tex
Fix typos

@@
 \end{tikzcd}\]
-is the inflation $\operatorname{inf}: \mathrm{H}^1(k(P), \mu_n) \to \mathrm{H}^1(K_P,\mu_n)$.
+coincides with the inflation $\operatorname{inf}: \mathrm{H}^1(k(P), \mu_n) \to \mathrm{H}^1(K_P,\mu_n)$.
 The same holds for $P'$ due to Lemma \ref{3.2}.
@@
 \[ \bigcup_{i=1}^r \operatorname{supp}D_i\subset S\quad\text{and}\quad \bigcup_{j=1}^s\operatorname{supp}D'_j\subset S' \]
-is admissible. Now let $V\subset X$ be an open subscheme. By the approximation theorem, we can choose the $D_i$ and the $D'_j$ in $V_0$ and $\sigma (V_0)$ respectively.
+is admissible. Now let $V\subset X$ be an open subscheme. By the approximation theorem, we can choose the $D_i$ and the $D'_j$ so that their support is in $V_0$ and $\sigma (V_0)$ respectively.
 \item Observe that the following sequence is exact:
@@
 \[ C_1\to\prod_{i=2}^{r} k(P_i),\ x\mapsto(r_{P_i}(x))_i \]
-Finally for $T=\{P_1,\ldots,P_r\}\subset U_0$, the map $r_{S,T}$ is injective. For the second assertion, note the following isomorphism with the snake lemma:
-\[ \operatorname{coker}(r_{S,T}) \cong\operatorname{coker}\Big(K^S/k^\times \to \prod_{P\in T}k(P)^\times /k^\times \Big) \]
+Finally for $T=\{P_1,\ldots,P_r\}\subset U_0$, the map $r_{S,T}$ is injective. For the second assertion, note the following isomorphism:
+\[ \operatorname{coker}(r_{S,T}) \cong\operatorname{coker}\Big(K^S/k^\times \to \Big(\prod_{P\in T}k(P)^\times \Big)\Big/k^\times \Big) \]
 Since $K^S/k^\times$ is finitely generated, Lemma \ref{5.2} (iii) and Lemma \ref{5.3} (ii) show that the latter group is free-by-finite.
@@
 \[ r_S:K^S \to\prod_{P\in X_0\setminus S} k(P)^\times,\ x\mapsto (r_P(x))_P \]
-Therefore, $\varphi$ is unique. For existence, let $(S,T)$ be an admissible pair. First, we show that there is a unique homomorphism $\varphi^{S,T}$, making the following diagram commutative:
+Now it's evident that $\varphi$ is unique.
+For existence, let $(S,T)$ be an admissible pair.
+First, we show that there is a unique homomorphism $\varphi^{S,T}$, making the following diagram commutative:
 \[ (\star) \qquad \begin{tikzcd}

### 2017-01-21 Master Thesis/synopsis.tex
Add half-reference for 2.2-pre
Fix typo

@@
 
+\begin{proof}
+See Valued Fields 4.1.4.
+\end{proof}
+
 \section{The Main Lemma}
 
-\paragraph{} Let $K$ be a (not necessarily finitely generated) field of characteristic $0$ and dimension $2$.
+Let $K$ be a (not necessarily finitely generated) field over $\mathbb{Q}$ and dimension $2$.
 
@@
 \[ \operatorname{Pic}(X\setminus S)=0=\operatorname{Pic}(X'\setminus S') \]
-Such a set $S$ is called \textit{admissible}. If $V\subset X$ is an open subscheme, then one can choose $S$ in $V_0$.
+Such a set $S$ is called \textit{admissible}. If $V\subset X$ is an open subscheme, then we can choose $S$ in $V_0$.
 \item Let $S$ be admissible. For all $n\in\mathbb{N}$ the following canonical maps are injective:
@@
 \end{enumerate}
+Clearly, (i) and (ii) hold for almost all $a\in k^\times$.
 We set $C_2 = \ker(r_{P_2}|_{C_1})\subset C_1$.
 $C_2$ is free abelian of finite rank as well.
-By requirement $r_{P_2}(x)=a$ and $a\in k^\times$ is not of finite order, thus $x\not\in C_2$.
-It follows $\operatorname{rank}(C_2) < \operatorname{rank}(C_1)$.
+By requirement $r_{P_2}(x)=a$ and $a\in k^\times$ is not of finite order, thus $x\not\in C_2$ and $\operatorname{rank}(C_2) < \operatorname{rank}(C_1)$.
 By repeating this process until $C_r=1$, we get points $P_2,\ldots,P_r\in U_0$, such that the following map is injective:
-\[ C_1\to\prod_{i=2}^{r} k(P_i),\ x\mapsto(r_{P_i}(x))_i \]
+\[ C_1\to\prod_{i=2}^{r} k(P_i)^\times,\ x\mapsto(r_{P_i}(x))_i \]
 Finally for $T=\{P_1,\ldots,P_r\}\subset U_0$, the map $r_{S,T}$ is injective. For the second assertion, note the following isomorphism:

### 2017-01-22 Master Thesis/synopsis.tex
Add weak approximation theorem

@@
 
+\begin{theorem}[\textit{Weak Approximation Theorem}]\label{weak-approximation}
+Let $P_1,\ldots,P_n\in X_0$ be pairwise distinct closed points, $x_1,\ldots,x_n\in K$ and $a_1,\ldots,a_n\in\mathbb{Z}$. Then there is some $x\in K$, such that for all $i$ we have:
+\[ v_{P_i}(x-x_i) = a_i \]
+\end{theorem}
+
 \begin{lemma}\phantomsection\label{4.5}
@@
 \[ \operatorname{Pic}(X\setminus S)=0=\operatorname{Pic}(X'\setminus S') \]
-Such a set $S$ is called \textit{admissible}. If $V\subset X$ is an open subscheme, then we can choose $S$ in $V_0$.
+Such a set $S$ is called \textit{admissible}. If $T\subset X_0$ is a finite set of closed points, then we can choose $S$ in $X_0\setminus T$.
 \item Let $S$ be admissible. For all $n\in\mathbb{N}$ the following canonical maps are injective:
@@
 \[ \bigcup_{i=1}^r \operatorname{supp}D_i\subset S\quad\text{and}\quad \bigcup_{j=1}^s\operatorname{supp}D'_j\subset S' \]
-is admissible. Now let $V\subset X$ be an open subscheme. By the approximation theorem, we can choose the $D_i$ and the $D'_j$ so that their support is in $V_0$ and $\sigma (V_0)$ respectively.
+is admissible.
+Now let $T\subset X_0$ be finite.
+By the Weak Approximation Theorem \ref{weak-approximation}, we can add a principal divisor to the $D_i$ and the $D'_j$ so that their support is in $X_0\setminus T$ and $X'_0\setminus T'$ respectively.
 \item Observe that the following sequence is exact:

### 2017-01-22 Master Thesis/synopsis.tex
Mention Néron-Severi group

@@
 We have the following exact sequence:
-\[ 0\longrightarrow \operatorname{Pic}^0(C) \longrightarrow\operatorname{Pic}(C) \longrightarrow \mathbb{Z} \longrightarrow 0  \]
+\[ 0\longrightarrow \operatorname{Pic}^0(C) \longrightarrow\operatorname{Pic}(C) \longrightarrow \operatorname{NS}(C) \longrightarrow 0  \]
+where $\operatorname{NS}(C)$ denotes the finitely generated Néron-Severi group of $C$.
 Now $\operatorname{Pic}^0(C)$ can be identified with the $F$-rational points of the Jacobian variety $J$ of $C$.

### 2017-01-25 Master Thesis/synopsis.tex
Merge 2.2-pre and 2.2-prepre

@@
 
-\begin{lemma}\label{2.2-prepre}
+\begin{lemma}\label{2.2-pre}
 Let $v,w\in\operatorname{Spv}(K)$. We have:
 \[\mathcal{O}_w\subset\mathcal{O}_v \iff w\in\overline{\{v\}} \]
-Thus $v$ is a closed point if and only if it has no non-trivial specializations or iff $\dim k(v)=0$, i.e.
-$k(v)$ has positive characteristic and is algebraic over its prime field.
+Thus $v$ is a closed point if and only if it has no non-trivial specializations. Furthermore, for a fixed $v\in \operatorname{Spv}(K)$ there is a canonical bijection: 
+\[\{ w\in\operatorname{Spv}(K)\mid \mathcal{O}_w\subset\mathcal{O}_v \} \to \operatorname{Spv}k(v),\ w\mapsto w\mod v\]
+Therefore $v$ is a closed point iff $k(v)$ only admits the trivial valuation, i.e.
+$k(v)$ has positive characteristic $p>0$ and is algebraic over $\mathbb{F}_p$.
+In other words, $v$ is a closed point iff $\dim k(v)=0$.
 By the dimension inequality \ref{rank-dimension-inequality} we have $\operatorname{Spv}^{(n)}(K)\subset |\operatorname{Spv}(K)|$ where $n=\dim(K)$.
-\end{lemma}
 
-\begin{lemma}\label{2.2-pre}
-Let $v\in\operatorname{Spv}(K)$. There is a canonical bijection:
-\[\{ w\in\operatorname{Spv}(K)\mid \mathcal{O}_w\subset\mathcal{O}_v \} \to \operatorname{Spv}k(v),\ w\mapsto w\mod v\]
 If $w\in\operatorname{Spv}_\text{rd}^{(i)}(K)$ for some $i\geq 1$ then there exists a unique $v(w)\in\operatorname{Spv}_\text{rd}^{(i-1)}(K)$ such that $\mathcal{O}_w\subset\mathcal{O}_{v(w)}$.
@@
 Set $k=k(w)$ and $\Gamma = \Gamma_w$.
-Then $k$ is an algebraic extension of a finite field $\mathbb{F}_p$ by Lemma \ref{2.2-prepre}.
+Then $k$ is an algebraic extension of a finite field $\mathbb{F}_p$ by Lemma \ref{2.2-pre}.
 Let $\ell\neq p$ be a prime and assume $\mu_\ell\subset K$.

### 2017-01-25 Master Thesis/synopsis.tex
Complete symbols

@@
 The \textit{rank}\index{valuation!rank} of $v$ is defined as the Krull dimension of its valuation ring $\mathcal{O}_v$.
-The \textit{rational rank}\index{valuation!rational rank} of $v:K^\times\to\Gamma_v$ is defined as the rank of its valuation group $\Gamma_v$ as a $\mathbb{Z}$-module:
+The \textit{rational rank}\index{valuation!rational rank} of $v:K\to\Gamma_v\cup\{\infty\}$ is defined as the rank of its valuation group $\Gamma_v$ as a $\mathbb{Z}$-module:
 \[\operatorname{rrank}(v) = \dim_\mathbb{Q}(\Gamma_v\otimes_\mathbb{Z}\mathbb{Q}) \]
@@
 \end{align*}
-Valuations in $\operatorname{Spv}_\text{d}$ are called \textit{defectless}\index{valuation!defectless}, whereas those in $\operatorname{Spv}_\text{rd}$ are called \textit{rank-defectless}\index{valuation!rank-defectless}.
+Valuations in $\operatorname{Spv}_\text{d}(K)$ are called \textit{defectless}\index{valuation!defectless}, whereas those in $\operatorname{Spv}_\text{rd}(K)$ are called \textit{rank-defectless}\index{valuation!rank-defectless}.
 For $\ast\in\{\text{d},\text{rd}\}$ we set:

### 2017-01-25 Master Thesis/synopsis.tex
Simplify 5.2

@@
 
-Let $n\in\mathbb{N}$. Tensoring the exact sequence $0\to U_P\to K^\times \to\mathbb{Z}\to 0$ with $\mathbb{Z}/n\mathbb{Z}$ preserves its exactness since $\mathbb{Z}$ is $\mathbb{Z}$-flat:
+Let $n\in\mathbb{N}$. Tensoring the exact sequence $1\to U_P\to K^\times \to\mathbb{Z}\to 0$ with $\mathbb{Z}/n\mathbb{Z}$ preserves its exactness since $\mathbb{Z}$ is $\mathbb{Z}$-flat:
 \[ 1\longrightarrow U_P/n\longrightarrow K^\times/n\longrightarrow\mathbb{Z}/n\mathbb{Z}\longrightarrow 0 \]
@@
 \item Finitely generated abelian groups are free-by-finite by the structure theorem for finitely generated $\mathbb{Z}$-modules.
+\item Any subgroup of a free-by-finite group is free-by-finite as well.
 \end{itemize}
@@
 \item Let $0\longrightarrow A'\stackrel{i}{\longrightarrow}A\stackrel{p}{\longrightarrow}A''\longrightarrow 0$ be an exact sequence of abelian groups.
-If $A$ is free-by-finite, then $A'$ is free-by-finite.
-Additionally, if $A'$ is finitely generated, then $A''$ is free-by-finite.
+If $A$ is free-by-finite and $A'$ is finitely generated, then $A''$ is free-by-finite.
 \item Let $A$ be free-by-finite. Then $\bigcap_{n\in\mathbb{N}}nA=0$.
@@
 Since $A''/A''_\text{tor}$ is free abelian, the short exact sequence in the second row splits, making $A/A_\text{tor}$ a direct sum of a free-by-finite group and a free group.
-\item Subgroups of free-by-finite groups are free-by-finite.
-Now assume $A'$ is finitely generated.
-We consider $A'$ as a subgroup of $A$.
+\item We consider $A'$ as a subgroup of $A$.
 Let $T=A_\text{tor}$.
@@
 \[ \operatorname{coker}(r_{S,T}) \cong\operatorname{coker}\Big(K^S/k^\times \to \Big(\prod_{P\in T}k(P)^\times \Big)\Big/k^\times \Big) \]
-Since $K^S/k^\times$ is finitely generated, Lemma \ref{5.2} (iii) and Lemma \ref{5.3} (ii) show that the latter group is free-by-finite.
+Since $K^S/k^\times$ is generated by the uniformizers of the corresponding valuations in $S$, Lemma \ref{5.2} (iii) and Lemma \ref{5.3} (ii) show that the latter group is free-by-finite.
 \end{proof}

### 2017-01-25 Master Thesis/synopsis.tex
Clarify

@@
 for all $g\in G_K$.
-For $\overline{v}\in\operatorname{Spv}_\text{d}^{(1)}(\overline{K})$ we have therefore $G_{\overline{v}} = h G_{\overline{v}} h^{-1} = G_{h\overline{v}}$, thus $h\in G_{\overline{v}}$ for all $\overline{v}$.
+For $\overline{v}\in\operatorname{Spv}_\text{d}^{(1)}(\overline{K})$ we have therefore $G_{\overline{v}} = h G_{\overline{v}} h^{-1} = G_{h\overline{v}}$, thus $h\in G_{\overline{v}}$ by F.K.
+Schmidt \ref{theorem-fkschmidt}.
+Since $\overline{v}$ was arbitrary, we get $h\in G_{\overline{v}}$ for all $\overline{v}$.
 Again, by F.K.

### 2017-01-26 Master Thesis/synopsis.tex
Add corollary-fkschmidt

@@
 \[\operatorname{rrank}(v) = \dim_\mathbb{Q}(\Gamma_v\otimes_\mathbb{Z}\mathbb{Q}) \]
+We write $K_v$ for the quotient field of a henselization of $\mathcal{O}_v$.
 \end{definition}
@@
 
+\begin{corollary}\label{corollary-fkschmidt}
+Let $v_1,v_2\in\operatorname{Spv}(K)$ be two inequivalent valuations of rank $1$ on a field $K$ with decomposition groups $G_{v_1},G_{v_2}\subset G_K$. Then $G_{v_1}\cap G_{v_2}=1$.
+\end{corollary}
+
+\begin{proof}
+The compositum $K_{v_1}K_{v_2}$ in a fixed separable closure of $K$ is henselian with respect to both $v_1$ and $v_2$, hence separably closed by Theorem \ref{theorem-fkschmidt}. Therefore:
+\[1=G_{K_{v_1}K_{v_2}} = G_{K_{v_1}}\cap G_{K_{v_2}} = G_{v_1}\cap G_{v_2} \qedhere\]
+\end{proof}
+
 \begin{lemma}\label{2.2}
@@
 \[ \mathcal{O}_v = \varinjlim_{\mathfrak{X}\in\mathfrak{M}(K)}\mathcal{O}_{\mathfrak{X},\pi_\mathfrak{X}(v)} \]
-\item For $v\in\operatorname{Spv}(K)$ let $K_v$ be the quotient field of a henselization of $\mathcal{O}_v$ and for $\mathfrak{X}\in\mathfrak{M}(K),\ x\in\mathfrak{X}$ let $K_x$ be the quotient field of a henselization of $\mathcal{O}_{\mathfrak{X},x}$.
-Then:
+\item For $\mathfrak{X}\in\mathfrak{M}(K),\ x\in\mathfrak{X}$ let $K_x$ be the quotient field of a henselization of $\mathcal{O}_{\mathfrak{X},x}$. Then we have:
 \[ K_v=\varinjlim_{\mathfrak{X}\in\mathfrak{M}(K)}K_{\pi_\mathfrak{X}(v)} \]
@@
 
-Let $K,K'$ be two finitely generated fields of characteristic $0$ and dimension $2$ and $\sigma:G_K\to G_{K'}$ be an isomorphism of profinite groups.
+Let $K,K'$ be two finitely generated fields of characteristic $0$ and dimension $2$ and let $\sigma:G_K\to G_{K'}$ be an isomorphism of profinite groups.
 For intermediate fields $L$ of $\overline{K}/K$ we denote the corresponding intermediate field of $\overline{K'}/K'$ by $L'$.
-In other words $\sigma G_L=G_{L'}$.
+In other words, $\sigma G_L=G_{L'}$.
 
@@
 By the Theorem of F.K.
-Schmidt \ref{theorem-fkschmidt} we have $G_{\overline{v}_1}\cap G_{\overline{v}_2} = 1$ for $\overline{v}_1\neq\overline{v}_2$.
+Schmidt \ref{corollary-fkschmidt} we have $G_{\overline{v}_1}\cap G_{\overline{v}_2} = 1$ for $\overline{v}_1\neq\overline{v}_2$.
 Therefore, $\sigma(g) = \tau(g) = \varphi^{-1}g\varphi$ for all $g\in G_K$.
@@
 For $\overline{v}\in\operatorname{Spv}_\text{d}^{(1)}(\overline{K})$ we have therefore $G_{\overline{v}} = h G_{\overline{v}} h^{-1} = G_{h\overline{v}}$, thus $h\in G_{\overline{v}}$ by F.K.
-Schmidt \ref{theorem-fkschmidt}.
+Schmidt \ref{corollary-fkschmidt}.
 Since $\overline{v}$ was arbitrary, we get $h\in G_{\overline{v}}$ for all $\overline{v}$.
 Again, by F.K.
-Schmidt \ref{theorem-fkschmidt}, we have $h=1$ and thus $\varphi = \psi$.
+Schmidt \ref{corollary-fkschmidt}, we have $h=1$ and thus $\varphi = \psi$.
 \end{proof}

### 2017-01-27 Master Thesis/synopsis.tex
Fix typos

@@
 \begin{proof}
-The compositum $K_{v_1}K_{v_2}$ in a fixed separable closure of $K$ is henselian with respect to both $v_1$ and $v_2$, hence separably closed by Theorem \ref{theorem-fkschmidt}. Therefore:
+Let $K_{v_i}$ be a henselization of $K$ with respect to $v_i$.
+The compositum $K_{v_1}K_{v_2}$ in a fixed separable closure of $K$ is henselian with respect to both $v_1$ and $v_2$, hence separably closed by Theorem \ref{theorem-fkschmidt}.
+Therefore:
 \[1=G_{K_{v_1}K_{v_2}} = G_{K_{v_1}}\cap G_{K_{v_2}} = G_{v_1}\cap G_{v_2} \qedhere\]
@@
 
-\paragraph{} Let $n\in\mathbb{N}$. In the previous remark, we have determined that the following diagram commutes: 
+\paragraph{} Let $n\in\mathbb{N}$. In the previous remark, we have reduced the problem, so that the following diagram commutes: 
 \[ \begin{tikzcd}
@@
 \end{tikzcd} \]
-Let $P\in X_0$ and $v_P:K\to\mathbb{Z}\cup\{\infty\}$ be the corresponding discrete normalized valuation and let $U_P$ be the group of units of the valuation ring $\mathcal{O}_P$.
+Let $P\in X_0$ and $v_P:K\to\mathbb{Z}\cup\{\infty\}$ be the corresponding normalized discrete valuation and let $U_P$ be the group of units of the valuation ring $\mathcal{O}_P$.
 We denote the natural projection $U_P\to k(P)^\times,\ x\mapsto x\mod P$ by $r_P$.
@@
 \[ \varphi(x + y) = 0 = \varphi(x) + \varphi(-1)\varphi(x) = \varphi(x) + \varphi(-x) = \varphi(x) + \varphi(y) \]
-\item \textit{Case 2:} Let $x + y\neq 0$ and $S\subset X_0$ be a finite subset such that $\varphi(x + y)$, $\varphi(x)$ and $\varphi(y)$ are contained in $K^S$.
+\item \textit{Case 2:} Let $x + y\neq 0$ and $S\subset X_0$ be an admissible finite subset such that $\varphi(x + y)$, $\varphi(x)$ and $\varphi(y)$ are contained in $K^S$.
 Looking at the commutative diagram in \mbox{Step 1} we get
@@
 for all $P\in X_0\setminus S$.
-Note that $\varphi_P: k(P')\to k(P)$ is a field isomorphism and that the residue maps $r_{P}: \mathcal{O}_{P} \to k(P)$ are ring homomorphisms.
+Note that $\varphi_P: k(P')\to k(P)$ is a field isomorphism and that the residue maps $r_{P'}: \mathcal{O}_{P'} \to k(P')$ are ring homomorphisms.
 Since $K^S\to \prod_{P\in X_0\setminus S} k(P)^\times$ is injective, we get $\varphi(x+y) = \varphi(x) + \varphi(y)$.\qedhere

### 2017-01-27 Master Thesis/synopsis.tex
Add Theorem about unramified extensions

@@
 \[\operatorname{rrank}(v) = \dim_\mathbb{Q}(\Gamma_v\otimes_\mathbb{Z}\mathbb{Q}) \]
-We write $K_v$ for the quotient field of a henselization of $\mathcal{O}_v$.
+We write $K_v$ for a henselization of $K$ with respect to $v$.
 \end{definition}
@@
 
-\section{The Main Lemma}
+\begin{theorem}\label{theorem-maxunramified}
+Let $K$ be a henselian field with respect to $v: K\to\Gamma\cup\{\infty\}$. Let $K^\text{nr}$ be the maximal unramified extension of $K$ with the unique extension $w$ of $v$. Then:
+\begin{enumerate}[(i)]
+\item $w$ has value group $\Gamma$ and residue class field $k(v)^\text{s}$.
+\item $K^\text{nr}$ is a Galois extension of $K$ and $\operatorname{Gal}(K^\text{nr}/K)\cong G_{k(v)}$. There exists an inclusion-preserving one-to-one correspondence: 
+\begin{align*}
+\{ \text{Intermediate extensions in $K^\text{nr}/K$} \} &\longrightarrow \{\text{Separable extensions of $k(v)$} \}\\ 
+L &\longmapsto k(v_L)  
+\end{align*}
+where $v_L$ is the unique extension of $v$ to $L$.
+\end{enumerate}
+\end{theorem}
 
-Let $K$ be a (not necessarily finitely generated) field over $\mathbb{Q}$ and dimension $2$.
+\begin{proof}
+See Valued Fields 5.2.7.
+\end{proof}
 
@@
 \begin{proof}
-See Valued Fields.
+See Valued Fields 4.4.1.
 \end{proof}
@@
 \end{proof}
+
+\section{The Main Lemma}
+
+Let $K$ be a (not necessarily finitely generated) field over $\mathbb{Q}$ and dimension $2$.
 
@@
 \[ \sigma(G_{K^\text{nr}_v}) = \sigma(G_{K_{w_1}}\cap G_{K_{w_2}}) = G_{K'_{w'_1}}\cap G_{K'_{w'_2}} = G_{(K')^\text{nr}_{v'}} \]
-Therefore, $\sigma$ induces an isomorphism:
+Therefore, $\sigma$ induces with Theorem \ref{theorem-maxunramified} an isomorphism:
 \[ G_{k(v)} \cong G_{K_v}/G_{K_v^\text{nr}} \to G_{K'_{v'}}/G_{(K')_{v'}^\text{nr}}\cong G_{k(v')} \]

### 2017-01-27 Master Thesis/synopsis.tex
Add some info to 3.2

@@
 
-Let $w_1,w_2 \in\operatorname{Spv}_\text{d}^{(2)}(K)$ be two different valuations. By Lemma \ref{2.2} the following assertions are equivalent:
+Let $w_1,w_2 \in\operatorname{Spv}_\text{d}^{(2)}(K)$ be two different valuations. By Lemma \ref{2.2} the following statements are equivalent:
 \begin{enumerate}[(i)]
@@
 \end{enumerate}
-In this case, $K_{w_1}K_{w_2}$ is the maximal unramified field extension $K_v^\text{nr}$ of $K_v$ where $v=v(w_1)$.
+In this case, $k(v')$ is separably closed where $v'$ is the unique extension of the valuation on $K_v$ to $K_{w_1}K_{w_2}$ where $v=v(w_1)$.
+By Theorem \ref{theorem-maxunramified} $K_{w_1}K_{w_2}$ is the maximal unramified field extension $K_v^\text{nr}$ of $K_v$.
 Furthermore, $K'_{\sigma(w_1)}K'_{\sigma(w_2)} = (K_{w_1}K_{w_2})' \neq \overline{K'}$ and again by Lemma \ref{2.2} we get: \[\sigma(w_1),\sigma(w_2)\in\operatorname{Spv}_\text{rd}^{(2)}(K)\quad \text{and} \quad v(\sigma(w_1)) = v(\sigma(w_2))\]
@@
 
-Let $v\in\operatorname{Spv}_\text{d}^{(1)}$ and let $N$ be the normal subgroup of $G_{K_v}$ generated by all subgroups $G_{K_w}$ for $w\in\operatorname{Spv}_\text{rd}^{(2)}(K)$ with $v(w) = v$.
+Let $v\in\operatorname{Spv}_\text{d}^{(1)}(K)$ and let $N$ be the normal subgroup of $G_{K_v}$ generated by all subgroups $G_{K_w}$ for $w\in\operatorname{Spv}_\text{rd}^{(2)}(K)$ with $v(w) = v$.
 Since a global field has no non-trivial algebraic extension in which every prime splits completely, we get $N=G_{K_v}$.
@@
 \[ \sigma(G_{K^\text{nr}_v}) = \sigma(G_{K_{w_1}}\cap G_{K_{w_2}}) = G_{K'_{w'_1}}\cap G_{K'_{w'_2}} = G_{(K')^\text{nr}_{v'}} \]
-Therefore, $\sigma$ induces with Theorem \ref{theorem-maxunramified} an isomorphism:
+Therefore, $\sigma$ induces with Theorem \ref{theorem-maxunramified} the following isomorphism:
 \[ G_{k(v)} \cong G_{K_v}/G_{K_v^\text{nr}} \to G_{K'_{v'}}/G_{(K')_{v'}^\text{nr}}\cong G_{k(v')} \]

### 2017-01-28 Master Thesis/synopsis.tex
Replace max unramified with absolute inertia

@@
 
-\begin{theorem}\label{theorem-maxunramified}
-Let $K$ be a henselian field with respect to $v: K\to\Gamma\cup\{\infty\}$. Let $K^\text{nr}$ be the maximal unramified extension of $K$ with the unique extension $w$ of $v$. Then:
+\begin{theorem}\label{theorem-absoluteinertia}
+Let $K$ be a henselian field with respect to $v: K\to\Gamma\cup\{\infty\}$. Let $K^\text{t}$ be the absolute inertia field of $K$ with the unique extension $w$ of $v$. Then:
 \begin{enumerate}[(i)]
 \item $w$ has value group $\Gamma$ and residue class field $k(v)^\text{s}$.
-\item $K^\text{nr}$ is a Galois extension of $K$ and $\operatorname{Gal}(K^\text{nr}/K)\cong G_{k(v)}$. There exists an inclusion-preserving one-to-one correspondence: 
+\item $K^\text{t}$ is a Galois extension of $K$ and $\operatorname{Gal}(K^\text{t}/K)\cong G_{k(v)}$. There exists an inclusion-preserving one-to-one correspondence: 
 \begin{align*}
-\{ \text{Intermediate extensions in $K^\text{nr}/K$} \} &\longrightarrow \{\text{Separable extensions of $k(v)$} \}\\ 
+\{ \text{Intermediate extensions in $K^\text{t}/K$} \} &\longrightarrow \{\text{Separable extensions of $k(v)$} \}\\ 
 L &\longmapsto k(v_L)  
@@
 
-Furthermore, if $E^{\text{nr}}$ and $E'^{\text{nr}}$ denote the maximal unramified field extensions of $E$ and $E'$ respectively, then $\sigma$ induces an isomorphism:
-\[ G_{k(v)}\cong\operatorname{Gal}(E^\text{nr}/E) \to\operatorname{Gal}(E'^{\text{nr}}/E')\cong G_{k(v')} \]
+Furthermore, if $E^{\text{t}}$ and $E'^{\text{t}}$ denote the absolute inertia field of $E$ and $E'$ respectively, then $\sigma$ induces an isomorphism:
+\[ G_{k(v)}\cong\operatorname{Gal}(E^\text{t}/E) \to\operatorname{Gal}(E'^{\text{t}}/E')\cong G_{k(v')} \]
 which is induced by a unique isomorphism $\varphi_v^\text{s}:k(v')^\text{s}\to k(v)^\text{s}$. In particular $\varphi_v=\varphi_v^\text{s}|_{k(v')}:k(v')\to k(v)$ is an isomorphism.
@@
 In this case, $k(v')$ is separably closed where $v'$ is the unique extension of the valuation on $K_v$ to $K_{w_1}K_{w_2}$ where $v=v(w_1)$.
-By Theorem \ref{theorem-maxunramified} $K_{w_1}K_{w_2}$ is the maximal unramified field extension $K_v^\text{nr}$ of $K_v$.
+By Theorem \ref{theorem-absoluteinertia} $K_{w_1}K_{w_2}$ is the absolute inertia field $K_v^\text{t}$ of $K_v$.
 Furthermore, $K'_{\sigma(w_1)}K'_{\sigma(w_2)} = (K_{w_1}K_{w_2})' \neq \overline{K'}$ and again by Lemma \ref{2.2} we get: \[\sigma(w_1),\sigma(w_2)\in\operatorname{Spv}_\text{rd}^{(2)}(K)\quad \text{and} \quad v(\sigma(w_1)) = v(\sigma(w_2))\]
@@
 Furthermore, let $w_1,w_2\in\operatorname{Spv}_\text{rd}^{(2)}(K)$ with $w_1\neq w_2$ and $v=v(w_1)=v(w_2)$, then:
-\[ \sigma(G_{K^\text{nr}_v}) = \sigma(G_{K_{w_1}}\cap G_{K_{w_2}}) = G_{K'_{w'_1}}\cap G_{K'_{w'_2}} = G_{(K')^\text{nr}_{v'}} \]
-Therefore, $\sigma$ induces with Theorem \ref{theorem-maxunramified} the following isomorphism:
-\[ G_{k(v)} \cong G_{K_v}/G_{K_v^\text{nr}} \to G_{K'_{v'}}/G_{(K')_{v'}^\text{nr}}\cong G_{k(v')} \]
+\[ \sigma(G_{K^\text{t}_v}) = \sigma(G_{K_{w_1}}\cap G_{K_{w_2}}) = G_{K'_{w'_1}}\cap G_{K'_{w'_2}} = G_{(K')^\text{t}_{v'}} \]
+Therefore, $\sigma$ induces with Theorem \ref{theorem-absoluteinertia} the following isomorphism:
+\[ G_{k(v)} \cong G_{K_v}/G_{K_v^\text{t}} \to G_{K'_{v'}}/G_{(K')_{v'}^\text{t}}\cong G_{k(v')} \]
 The rest follows from the Theorem of Neukirch-Uchida.
@@
 \begin{enumerate}[(i)]
-\item Consider the maximal unramified extension $L = K_P^\text{nr}$ of $K_P$ and let $w$ be the unique extension of $v_P$ in $L$.
+\item Consider the absolute inertia field $L = K_P^\text{t}$ of $K_P$ and let $w$ be the unique extension of $v_P$ in $L$.
 Now $k(w) = k(v_P)^\text{s}$.
@@
 \end{tikzcd}\]
-Note that $(K^\text{nr}_P)' = K'^{\text{nr}}_{P'}$ as proven in Lemma \ref{3.2}.
+Note that $(K^\text{t}_P)' = K'^{\text{t}}_{P'}$ as proven in Lemma \ref{3.2}.
 \item Consider the field extension $K_P/K$. Without loss of generality, we can assume $K=K_P$. Since $\operatorname{char}k(P)=0$, by Hensel's lemma, $r_P/n$ is an isomorphism and the composition

### 2017-01-28 Master Thesis/synopsis.tex
Add henselization and fix corollary-fkschmidt

@@
 \[\operatorname{rrank}(v) = \dim_\mathbb{Q}(\Gamma_v\otimes_\mathbb{Z}\mathbb{Q}) \]
-We write $K_v$ for a henselization of $K$ with respect to $v$.
 \end{definition}
@@
 Thus $v$ is a closed point if and only if it has no non-trivial specializations. Furthermore, for a fixed $v\in \operatorname{Spv}(K)$ there is a canonical bijection: 
-\[\{ w\in\operatorname{Spv}(K)\mid \mathcal{O}_w\subset\mathcal{O}_v \} \to \operatorname{Spv}k(v),\ w\mapsto w\mod v\]
+\begin{align*}
+\{ w\in\operatorname{Spv}(K)\mid \mathcal{O}_w\subset\mathcal{O}_v \} &\longrightarrow \operatorname{Spv}k(v)\\ 
+w&\longmapsto w\mod v
+\end{align*}
 Therefore $v$ is a closed point iff $k(v)$ only admits the trivial valuation, i.e.
@@
 See Valued Fields 4.1.4.
+\end{proof}
+
+\begin{theorem}\label{theorem-henselization}
+Let $K$ be a field and $v\in\operatorname{Spv}(K)$. Then there is a Galois extension $K_v/K$ with a henselian extension $\overline{v}$ of $v$ with the following characterization:
+
+If $K'/K$ is henselian with respect to an extension $w$ of $v$, then there exists a uniquely determined $K$-embedding $K_v\to K'$, so that $w$ extends $\overline{v}$.
+
+Furthermore, $\overline{v}$ has the same value group as $v$ and $k(\overline{v}) = k(v)$.
+$G_{K_v}$ is the decomposition group $G_{\overline{v}}\subset G_K$ of the unique extension of $\overline{v}$ to $K^\text{s}$.
+$K_v$ is called a \textit{henselization} of $K$ with respect to $v$.
+Any two henselization of $K$ with respect to $v$ are $K$-conjugate.
+\end{theorem}
+
+\begin{proof}
+See Valued Fields Lemma 5.2.1, Lemma 5.2.2, Theorem 5.2.5.
 \end{proof}
@@
 \begin{corollary}\label{corollary-fkschmidt}
-Let $v_1,v_2\in\operatorname{Spv}(K)$ be two inequivalent valuations of rank $1$ on a field $K$ with decomposition groups $G_{v_1},G_{v_2}\subset G_K$. Then $G_{v_1}\cap G_{v_2}=1$.
+Let $w_1,w_2\in\operatorname{Spv}(K^\text{s})$ be two inequivalent valuations of rank $1$ with decomposition groups $G_{{w_1}},G_{{w_2}}\subset G_K$. Then $G_{{w_1}}\cap G_{{w_2}}=1$.
 \end{corollary}
@@
 \begin{proof}
-Let $K_{v_i}$ be a henselization of $K$ with respect to $v_i$.
-The compositum $K_{v_1}K_{v_2}$ in a fixed separable closure of $K$ is henselian with respect to both $v_1$ and $v_2$, hence separably closed by Theorem \ref{theorem-fkschmidt}.
-Therefore:
-\[1=G_{K_{v_1}K_{v_2}} = G_{K_{v_1}}\cap G_{K_{v_2}} = G_{v_1}\cap G_{v_2} \qedhere\]
+Let $L$ be the fixed field of $G_{w_1}\cap G_{w_2}$ and set $v_i = w_i|_L$.
+Then $L$ is henselian with respect to both $v_1$ and $v_2$.
+Since $v_1$ and $v_2$ are unramified in $K^\text{s}/L$, they are inequivalent valuations of rank $1$.
+Now $L$ is separably closed by Theorem \ref{theorem-fkschmidt}.
+Therefore by Theorem \ref{theorem-henselization} we get $G_{{w_1}}\cap G_{{w_1}}=G_L =1$.
 \end{proof}
@@
 In this case, $k(v')$ is separably closed where $v'$ is the unique extension of the valuation on $K_v$ to $K_{w_1}K_{w_2}$ where $v=v(w_1)$.
-By Theorem \ref{theorem-absoluteinertia} $K_{w_1}K_{w_2}$ is the absolute inertia field $K_v^\text{t}$ of $K_v$.
+By Theorem \ref{theorem-absoluteinertia} (ii) $K_{w_1}K_{w_2}$ is the absolute inertia field $K_v^\text{t}$ of $K_v$.
 Furthermore, $K'_{\sigma(w_1)}K'_{\sigma(w_2)} = (K_{w_1}K_{w_2})' \neq \overline{K'}$ and again by Lemma \ref{2.2} we get: \[\sigma(w_1),\sigma(w_2)\in\operatorname{Spv}_\text{rd}^{(2)}(K)\quad \text{and} \quad v(\sigma(w_1)) = v(\sigma(w_2))\]
@@
 \item Consider the absolute inertia field $L = K_P^\text{t}$ of $K_P$ and let $w$ be the unique extension of $v_P$ in $L$.
-Now $k(w) = k(v_P)^\text{s}$.
-Since $w$ is unramified over $v$, $w$ is discrete as well.
+By Theorem \ref{theorem-absoluteinertia} (i), $w$ is discrete as well with $k(w) = k(v_P)^\text{s}$.
 We have the following exact sequence:
@@
 Therefore $\sigma(g)^{-1}\tau(g)$ is contained in $G_{\overline{v}}$ for all $\overline{v}\in\operatorname{Spv}_\text{d}^{(1)}(\overline{K})$.
-By the Theorem of F.K.
-Schmidt \ref{corollary-fkschmidt} we have $G_{\overline{v}_1}\cap G_{\overline{v}_2} = 1$ for $\overline{v}_1\neq\overline{v}_2$.
+By Corollary \ref{corollary-fkschmidt} we have $G_{\overline{v}_1}\cap G_{\overline{v}_2} = 1$ for $\overline{v}_1\neq\overline{v}_2$.
 Therefore, $\sigma(g) = \tau(g) = \varphi^{-1}g\varphi$ for all $g\in G_K$.
@@
 for all $g\in G_K$.
-For $\overline{v}\in\operatorname{Spv}_\text{d}^{(1)}(\overline{K})$ we have therefore $G_{\overline{v}} = h G_{\overline{v}} h^{-1} = G_{h\overline{v}}$, thus $h\in G_{\overline{v}}$ by F.K.
-Schmidt \ref{corollary-fkschmidt}.
+For $\overline{v}\in\operatorname{Spv}_\text{d}^{(1)}(\overline{K})$ we have therefore $G_{\overline{v}} = h G_{\overline{v}} h^{-1} = G_{h\overline{v}}$, thus $h\in G_{\overline{v}}$ by Corollary \ref{corollary-fkschmidt}.
 Since $\overline{v}$ was arbitrary, we get $h\in G_{\overline{v}}$ for all $\overline{v}$.
-Again, by F.K.
-Schmidt \ref{corollary-fkschmidt}, we have $h=1$ and thus $\varphi = \psi$.
+Again, by Corollary \ref{corollary-fkschmidt}, we have $h=1$ and thus $\varphi = \psi$.
 \end{proof}

### 2017-01-28 Master Thesis/introduction.tex
Add more notation

--- 2015-07-15 Master Thesis/introduction.tex
@@
 \mathbb{C} & \text{complex numbers}\\
-\mathbb{Z}_p & \text{$p$-adic integers}\\
-\mathbb{Q}_p & \text{$p$-adic numbers}\\
 \\
@@
 K^\text{s} & \text{separable closure of $K$}\\
-G_K & \text{absolute Galois group of $K$}
+K^\text{t} & \text{absolute inertia field of $K$}\\
+K_v & \text{henselization of $K$ with respect to $v$}\\
+G_K & \text{absolute Galois group of $K$}\\
+\operatorname{Gal}(L/K) & \text{Galois group of $L/K$}\\
+\operatorname{char}(K) & \text{characteristic of $K$}\\
+\operatorname{trdeg}(L/K) & \text{transcendence degree of $L/K$}
 \end{array}$

### 2017-01-28 Master Thesis/synopsis.tex
Formulation

@@
 
-If $K'/K$ is henselian with respect to an extension $w$ of $v$, then there exists a uniquely determined $K$-embedding $K_v\to K'$, so that $w$ extends $\overline{v}$.
+\textit{If $K'/K$ is henselian with respect to an extension $w$ of $v$, then there exists a uniquely determined $K$-embedding $K_v\to K'$, so that $w$ extends $\overline{v}$.}
 
@@
 \item $K$ is not algebraically closed.
-\item $w_1,w_2\in\operatorname{Spv}_\text{rd}^{(2)}(K),\ v(w_1)=v(w_2)$ and $k(v(w_1))$ is separably closed
+\item $w_1,w_2\in\operatorname{Spv}_\text{rd}^{(2)}(K),\ v(w_1)=v(w_2)$ and $k(v(w_1))$ is separably closed.
 \end{enumerate}
@@
 Again, by the Theorem of F.K.
-Schmidt \ref{theorem-fkschmidt}, $v_1 = v_2$.
+Schmidt \ref{theorem-fkschmidt}, we get $v_1 = v_2$.
 Furthermore $k(v_1) = k(v_2)$ is henselian with respect to two different valuations of rank $1$, namely $w_1\mod v_1$ and $w_2\mod v_1$, making $k(v_1)$ separably closed.

### 2017-01-28 Master Thesis/synopsis.tex
Fix typo

@@
 We denote the valuation ring of $v\in\operatorname{Spv}(K)$ by $(\mathcal{O}_v,\mathfrak{m}_v)$ and set $U_v=\mathcal{O}_v^\times$ and $k(v)=\mathcal{O}_v/\mathfrak{m}_v$.
-The \textit{rank}\index{valuation!rank} of $v$ is defined as the Krull dimension of its valuation ring $\mathcal{O}_v$.
-The \textit{rational rank}\index{valuation!rational rank} of $v:K\to\Gamma_v\cup\{\infty\}$ is defined as the rank of its valuation group $\Gamma_v$ as a $\mathbb{Z}$-module:
+The \textit{rank} of $v$ is defined as the Krull dimension of its valuation ring $\mathcal{O}_v$.
+The \textit{rational rank}\index{valuation!rational rank} of $v:K\to\Gamma_v\cup\{\infty\}$ is defined as the rank of its value group $\Gamma_v$ as a $\mathbb{Z}$-module:
 \[\operatorname{rrank}(v) = \dim_\mathbb{Q}(\Gamma_v\otimes_\mathbb{Z}\mathbb{Q}) \]

### 2017-02-01 Master Thesis/synopsis.tex
Add some references

@@
 \begin{lemma}\label{spv-topology}
-$\operatorname{Spv}(K)$ with the topology defined above is a quasicompact $T_0$-space. Every closed subset $A\subset\operatorname{Spv}(K)$ is the closure of its closed points.
-
+$\operatorname{Spv}(K)$ with the topology defined above is a quasicompact $T_0$-space.
 In particular the set of closed points $|\operatorname{Spv}(K)|$ of $\operatorname{Spv}(K)$ is a quasicompact $T_1$-space, endowed with the induced topology.
+Let $v,w\in\operatorname{Spv}(K)$.
+We have:
+\[\mathcal{O}_w\subset\mathcal{O}_v \iff w\in\overline{\{v\}} \]
+Thus $v$ is a closed point if and only if it has no non-trivial specializations.
 \end{lemma}
@@
 \begin{proof}
-See [ZS] VI §7 Theorem 40.
+See [ZS] VI §17 Theorem 38, §17 Theorem 40.
 \end{proof}
@@
 \begin{lemma}\label{2.2-pre}
-Let $v,w\in\operatorname{Spv}(K)$. We have:
-\[\mathcal{O}_w\subset\mathcal{O}_v \iff w\in\overline{\{v\}} \]
-Thus $v$ is a closed point if and only if it has no non-trivial specializations. Furthermore, for a fixed $v\in \operatorname{Spv}(K)$ there is a canonical bijection: 
+For a fixed $v\in \operatorname{Spv}(K)$ there is a canonical bijection: 
 \begin{align*}
@@
 Therefore $v$ is a closed point iff $k(v)$ only admits the trivial valuation, i.e.
-$k(v)$ has positive characteristic $p>0$ and is algebraic over $\mathbb{F}_p$.
+$k(v)$ has characteristic $p>0$ and is algebraic over $\mathbb{F}_p$.
 In other words, $v$ is a closed point iff $\dim k(v)=0$.
@@
 \begin{proof}
-See Valued Fields 4.1.4.
+See Valued Fields 4.1.4, [ZS] VI §3 Theorem 3.
 \end{proof}
@@
 
+\begin{proof}
+See [ZS] VI §10 Theorem 18
+\end{proof}
+
 \begin{lemma}\phantomsection\label{4.5}

### 2017-02-01 Master Thesis/synopsis.tex
Use \longrightarrow

@@
 
-Let $K$ be a (not necessarily finitely generated) field over $\mathbb{Q}$ and dimension $2$.
+Let $K$ be a (not necessarily finitely generated) field over $\mathbb{Q}$ and of dimension $2$.
 
@@
 Let $K$ be finitely generated and $\ell\neq 2$ be a prime. The following map, induced by the restrictions, is injective:
-\[ \mathrm{H}^3(K,\mathbb{Z}/\ell\mathbb{Z})\to\prod_{w\in |\operatorname{Spv}(K)| }\mathrm{H}^3(K_w,\mathbb{Z}/\ell\mathbb{Z}) \]
+\[ \mathrm{H}^3(K,\mathbb{Z}/\ell\mathbb{Z})\longrightarrow\prod_{w\in |\operatorname{Spv}(K)| }\mathrm{H}^3(K_w,\mathbb{Z}/\ell\mathbb{Z}) \]
 \end{lemma}
@@
 We adopt the notation from the previous proof. First, we shall prove that the following map is injective for every model $\mathfrak{X}$ of $K$:
-\[ f: \mathrm{H}^3(K) \to \prod_{y\in\mathfrak{X}_0}\mathrm{H}^3(K_y) \]
+\[ f: \mathrm{H}^3(K) \longrightarrow \prod_{y\in\mathfrak{X}_0}\mathrm{H}^3(K_y) \]
 For $x\in\mathfrak{X}$ let $\mathcal{O}_x$ be a henselization of $\mathcal{O}_{\mathfrak{X}, x}$ and $K_x$ its quotient field.
@@
 Let $K$ be not necessarily finitely generated and $\ell\neq 2$ be a prime. The following map, induced by the restrictions, is injective:
-\[ \mathrm{H}^3(K,\mathbb{Z}/\ell\mathbb{Z})\to\prod_{w\in |\operatorname{Spv}(K)| }\mathrm{H}^3(K_w,\mathbb{Z}/\ell\mathbb{Z}) \]
+\[ \mathrm{H}^3(K,\mathbb{Z}/\ell\mathbb{Z})\longrightarrow \prod_{w\in |\operatorname{Spv}(K)| }\mathrm{H}^3(K_w,\mathbb{Z}/\ell\mathbb{Z}) \]
 \end{lemma}

### 2017-02-01 Master Thesis/synopsis.tex
Small changes

@@
 for some $d \leq \infty$, where $\Gamma'$ denotes the value group and $k'$ the residue field of the unique prolongation of $w$ to $L$.
-Note that $(\Gamma':\Gamma)< \infty$ and $\operatorname{rank}(w) = \operatorname{rank}(v)$ in an algebraic field extension ([ZS], Ch.
+Note that $\operatorname{rank}(w) = \operatorname{rank}(v)$ in an algebraic field extension ([ZS], Ch.
 VI, §11, Lemma 1 and 2).
@@
 Assume that $w = w_0|_K$ is not henselian.
-Then there exists a $w_1\in|\operatorname{Spv}(K_0)|$ with $w_1\neq w_0$ and such that $w = w_1|_K$, for otherwise $w$ would have a unique prolongation to $K_0$ and hence to $\overline{K}$ since $w_0$ is henselian.
+Then there exists a $w_1\in|\operatorname{Spv}(K_0)|$ with $w_1\neq w_0$ and such that $w = w_1|_K$, for otherwise $w$ would have a unique extension to $K_0$ and hence to $\overline{K}$ since $w_0$ is henselian.
 
@@
 
-Now $(K_0)_{w_1}$ is henselian both with respect to $w_1$ and with respect to the prolongation of $w_0$.
+Now $(K_0)_{w_1}$ is henselian both with respect to $w_1$ and with respect to the extension of $w_0$.
 Lemma \ref{2.2} yields $\operatorname{cd}(G_{(K_0)_{w_1}})\leq 2$, a contradiction.

### 2017-02-10 Master Thesis/synopsis.tex
Add explanation for WLOG in 2.4

@@
 \begin{proof}
-The degree $[K(\mu_\ell, \sqrt{-1}):K]$ is prime to $\ell$, thus we can assume that $\mu_\ell\subset K$ and $K$ is not formally real with a usual norm argument.
-Now we can replace $\mathbb{Z}/\ell\mathbb{Z}$ by $\mathbb{Z}/\ell\mathbb{Z}(2) = \mu_\ell^{\otimes 2}$.
-We write $\mathrm{H}^3(L) = \mathrm{H}^3(L, \mathbb{Z}/\ell\mathbb{Z}(2))$.
+Consider the extension $L=K(\mu_{\ell},\sqrt{-1})/K$. We have the restriction and corestriction map:
+\[ \operatorname{res}: \mathrm{H}^3(K,\mathbb{Z}/\ell\mathbb{Z}) \to \mathrm{H}^3(L,\mathbb{Z}/\ell\mathbb{Z}) \qquad \operatorname{cor}:\mathrm{H}^3(L,\mathbb{Z}/\ell\mathbb{Z})\to \mathrm{H}^3(K,\mathbb{Z}/\ell\mathbb{Z}) \]
+We have $\operatorname{cor}\circ\operatorname{res} = [L:K]$.
+Since $[L:K]$ is coprime to $\ell$, we see that $\operatorname{res}$ is injective.
+The same argument can be made for $K_w(\mu_{\ell},\sqrt{-1})/K_w$.
+Thus we can assume that $K=L$, i.e.
+$\mu_{\ell}\subset K$ and $K$ is not formally real.
+Now we can replace $\mathbb{Z}/\ell\mathbb{Z}$ by $\mathbb{Z}/\ell\mathbb{Z}(2) = \mu_\ell^{\otimes 2}$ and let us write $\mathrm{H}^3(L) = \mathrm{H}^3(L, \mathbb{Z}/\ell\mathbb{Z}(2))$.
 

### 2017-03-07 Master Thesis/synopsis.tex
Update corollary-fkschmidt

@@
 \begin{proof}
-Let $L$ be the fixed field of $G_{w_1}\cap G_{w_2}$ and set $v_i = w_i|_L$.
+Let $L$ be the fixed field of $G_{w_1}\cap G_{w_2}\subset G_K$ and set $v_i = w_i|_L$.
 Then $L$ is henselian with respect to both $v_1$ and $v_2$.
-Since $v_1$ and $v_2$ are unramified in $K^\text{s}/L$, they are inequivalent valuations of rank $1$.
+Since $w_1,w_2$ are the only extensions of $v_1$ and $v_2$ in $K^\text{s}$ respectively, $v_1$ and $v_2$ are inequivalent valuations of rank $1$.
 Now $L$ is separably closed by Theorem \ref{theorem-fkschmidt}.

### 2017-03-07 Master Thesis/synopsis.tex
Add Lemma about henselization of specialization of valuations

@@
 \begin{proof}
-See Valued Fields 4.1.4, [ZS] VI §3 Theorem 3.
+See Valued Fields 2.3, 4.1.4, [ZS] VI §3 Theorem 3.
 \end{proof}
@@
 See Valued Fields 5.2.7.
+\end{proof}
+
+\begin{lemma}\label{lemma-henselization-specialization}
+Let $K$ be a field and $w\in\operatorname{Spv}(K)$ with a fixed henselization $K_w$ of $w$.
+If $v\in\operatorname{Spv}(K)$ such that $w$ is a specialization of $v$, then there is a henselization $K_v$ of $v$ such that $K_v\subset K_w\subset K_v^\text{t}$.
+\end{lemma}
+
+\begin{proof}
+Let $\overline{w}$ be the extension of $w\in\operatorname{Spv}(K_w)$ on $K^\text{s}$.
+By Valued Fields 3.1.5 we can find an extension $\overline{v}$ of $v$ on $K^\text{s}$ such that $\overline{w}$ is a specialization of $\overline{v}$.
+Let $K_v$ be the fixed field of $G_{\overline{v}}\subset G_K$.
+
+Let $\sigma\in G_{\overline{w}}$.
+Then both $\mathcal{O}_{\overline{v}}$ and $\sigma(\mathcal{O}_{\overline{v}})$ are overrings of $\mathcal{O}_{\overline{w}}$.
+Now the overrings of valuation rings are totally ordered by inclusion and correspond to localizations of $\mathcal{O}_{\overline{w}}$ at prime ideals (See Valued Fields 2.3).
+Since $\mathcal{O}_{\overline{v}}$ and $\sigma(\mathcal{O}_{\overline{v}})$ have the same Krull dimension, they must be the same, thus $\sigma\in G_{\overline{v}}$.
+This proves $K_v\subset K_w$.
+
+Now let $\sigma\in G_{K_v^\text{t}}$, i.e.
+$\sigma\in G_{\overline{v}}$ such that the induced automorphism on the residue field $\overline{\sigma}:k(\overline{v})\to k(\overline{v})$ is the identity.
+Let $\mathfrak{p}\subset\mathcal{O}_{\overline{w}}$ be a prime ideal with $\mathcal{O}_{\overline{v}} = (\mathcal{O}_{\overline{w}})_\mathfrak{p}$.
+In particular, the restriction $\overline{\sigma}|: \mathcal{O}_{\overline{w}}/\mathfrak{p}\to \mathcal{O}_{\overline{w}}/\mathfrak{p}$ is the identity, thus $\sigma(\mathcal{O}_{\overline{w}})=\mathcal{O}_{\overline{w}}$, i.e.
+$\sigma\in G_{\overline{w}}$.
+This proves $K_w\subset K_v^\text{t}$.
 \end{proof}
@@
 Lemma \ref{3.1} shows that $\sigma$ induces a bijection $\sigma: \operatorname{Spv}_\text{d}^{(2)}(K)\to \operatorname{Spv}_\text{d}^{(2)}(K')$.
-For all $v\in\operatorname{Spv}(K)$ we choose a fixed henselization $K_v\subset\overline{K}$, such that if $v,w\in\operatorname{Spv}(K)$ and $w$ is a specialization of $v$ we have $K_v\subset K_w$.
+For all $v\in\operatorname{Spv}(K)$ we choose a fixed henselization $K_v\subset\overline{K}$ with Lemma \ref{lemma-henselization-specialization}, such that if $v,w\in\operatorname{Spv}(K)$ and $w$ is a specialization of $v$ we have $K_v\subset K_w$.
 This means that if $v$ is nontrivial, we get $w\in\operatorname{Spv}_\text{rd}^{(2)}(K)$ and $v=v(w)\in\operatorname{Spv}_\text{d}^{(1)}(K)$.
@@
 In this case, $k(v')$ is separably closed where $v'$ is the unique extension of the valuation on $K_v$ to $K_{w_1}K_{w_2}$ where $v=v(w_1)$.
-By Theorem \ref{theorem-absoluteinertia} (ii) $K_{w_1}K_{w_2}$ is the absolute inertia field $K_v^\text{t}$ of $K_v$.
+By Lemma \ref{lemma-henselization-specialization} we have $K_{w_1}K_{w_2}\subset K_v^\text{t}$ and by Theorem \ref{theorem-absoluteinertia} (ii) $K_{w_1}K_{w_2}$ is the absolute inertia field $K_v^\text{t}$ of $K_v$.
 Furthermore, $K'_{\sigma(w_1)}K'_{\sigma(w_2)} = (K_{w_1}K_{w_2})' \neq \overline{K'}$ and again by Lemma \ref{2.2} we get: \[\sigma(w_1),\sigma(w_2)\in\operatorname{Spv}_\text{rd}^{(2)}(K)\quad \text{and} \quad v(\sigma(w_1)) = v(\sigma(w_2))\]

### 2017-03-08 Master Thesis/synopsis.tex
Some tweaks

@@
 
-We denote the valuation ring of $v\in\operatorname{Spv}(K)$ by $(\mathcal{O}_v,\mathfrak{m}_v)$ and set $U_v=\mathcal{O}_v^\times$ and $k(v)=\mathcal{O}_v/\mathfrak{m}_v$.
+We denote the valuation ring of $v\in\operatorname{Spv}(K)$ by $(\mathcal{O}_v,\mathfrak{m}_v)$ and set $U_v=\mathcal{O}_v^\times$ and let $k(v)=\mathcal{O}_v/\mathfrak{m}_v$ be its residue field.
 The \textit{rank} of $v$ is defined as the Krull dimension of its valuation ring $\mathcal{O}_v$.
@@
 \begin{proof}
-Consider the extension $L=K(\mu_{\ell},\sqrt{-1})/K$. We have the restriction and corestriction map:
+Let $L=K(\mu_{\ell},\sqrt{-1})/K$. Consider the restriction and corestriction maps:
 \[ \operatorname{res}: \mathrm{H}^3(K,\mathbb{Z}/\ell\mathbb{Z}) \to \mathrm{H}^3(L,\mathbb{Z}/\ell\mathbb{Z}) \qquad \operatorname{cor}:\mathrm{H}^3(L,\mathbb{Z}/\ell\mathbb{Z})\to \mathrm{H}^3(K,\mathbb{Z}/\ell\mathbb{Z}) \]
@@
 Since $[L:K]$ is coprime to $\ell$, we see that $\operatorname{res}$ is injective.
-The same argument can be made for $K_w(\mu_{\ell},\sqrt{-1})/K_w$.
 Thus we can assume that $K=L$, i.e.
@@
 \begin{definition}
-For each admissible $S\subset X_0$ and $T_1\subset X_0\setminus S$ with injective $r_{S,T_1}$ we also have the injectivity of $r_{S,T_2}$ for every $T_1 \subset T_2$.
+For each admissible $S\subset X_0$ and $T_1\subset X_0\setminus S$ with injective $r_{S,T_1}$ we also have the injectivity of $r_{S,T_2}$ for every $T_2 \supset T_1$.
 Therefore,  we can find a finite subset $T\subset X_0\setminus S$ such that both $r_{S,T}$ and $r_{S',T'}$ are injective.
@@
 \end{tikzcd} \phantom{\qquad (\star)} \]
-The vertical maps are injective by definition.
-Thus, it suffices to show that the composition $\varrho:K'^{S'}\to\prod k(\sigma P)^\times\to \prod k(P)^\times\to\operatorname{coker}(r_{S,T})$ is trivial.
+The vertical maps are injective by definition. Thus, it suffices to show that the following composition is trivial:
+\[\varrho:K'^{S'}\longrightarrow\prod_{P\in T} k(\sigma P)^\times \longrightarrow \prod_{P\in T} k(P)^\times\longrightarrow\operatorname{coker}(r_{S,T})\] 
 Consider the commutative diagram which results from tensoring $(\star)$ with $\mathbb{Z}/n\mathbb{Z}$:
@@
 \begin{step}
-Extending $\varphi$ to a map $K'\to K$ by setting $\varphi(0)=0$ yields a field isomorphism $\varphi:K'\to K$.
+Extending $\varphi$ to a map $K'\to K$ by setting $\varphi(0)=0$ yields an isomorphism of fields: 
+\[\varphi:K'\to K\]
 \end{step}

### 2017-03-08 Master Thesis/synopsis.tex
Add the fact that value group index is finite in algebraic extension

@@
 \end{tikzcd} \]
-According to [Sa], Theorem (2-15) $g$ is injective.
+According to [Sa] Theorem (2-15) $g$ is injective.
 Now, $\psi\circ\varphi$ can be identified with the injective map $\Phi$ from the proof of the previous Lemma.
@@
 for some $d \leq \infty$, where $\Gamma'$ denotes the value group and $k'$ the residue field of the unique prolongation of $w$ to $L$.
-Note that $\operatorname{rank}(w) = \operatorname{rank}(v)$ in an algebraic field extension ([ZS], Ch.
+Note that $(\Gamma':\Gamma)<\infty$ and $\operatorname{rank}(w) = \operatorname{rank}(v)$ in an algebraic field extension ([ZS] Ch.
 VI, §11, Lemma 1 and 2).

### 2017-03-08 Master Thesis/synopsis.tex
Update theorem on henselizations

--- 2017-03-08 Master Thesis/synopsis.tex
@@
 \begin{theorem}\label{theorem-henselization}
-Let $K$ be a field and $v\in\operatorname{Spv}(K)$. Then there is a Galois extension $K_v/K$ with a henselian extension $\overline{v}$ of $v$ with the following characterization:
+Let $K$ be a field and $v\in\operatorname{Spv}(K)$. Then there is a separable extension $K_v/K$ with a henselian extension $\overline{v}$ of $v$ with the following characterization:
 
@@
 
-Furthermore, $\overline{v}$ has the same value group as $v$ and $k(\overline{v}) = k(v)$.
+$K_v$ is called a \textit{henselization} of $K$ with respect to $v$.
+Furthermore, $\overline{v}$ has the same value group as $v$ and the same residue field $k(\overline{v}) = k(v)$.
 $G_{K_v}$ is the decomposition group $G_{\overline{v}}\subset G_K$ of the unique extension of $\overline{v}$ to $K^\text{s}$.
-$K_v$ is called a \textit{henselization} of $K$ with respect to $v$.
-Any two henselization of $K$ with respect to $v$ are $K$-conjugate.
+
+Conversely, taking the fixed field of the decomposition group of an extension $\overline{v}'$ of $v$ in $K^\text{s}$ yields a henselization of $K$ with respect to $v$.
+Any two of such henselizations are $K$-conjugate.
 \end{theorem}
@@
 Note that $\operatorname{rank}(w) = \operatorname{rank}(v)$ in an algebraic field extension ([ZS], Ch.
-VI, §11, Lemma 1 and 2).
+VI, §11, Lemma 2).
 \end{proof}

### 2017-03-13 Master Thesis/synopsis.tex
Add 'Moreover' when dimension inequality becomes an equallity

@@
 \begin{definition}
-Let $K$ be a field and $P$ its prime field. The \textit{(Kronecker) dimension}\index{dimension!Kronecker} $\dim(K)$ of $K$ is defined as:
+Let $K$ be a field and $P$ its prime field.
+$K$ is called \textit{finitely generated} if $K/P$ is finitely generated.
+The \textit{(Kronecker) dimension}\index{dimension!Kronecker} $\dim(K)$ of $K$ is defined as:
 \[\dim(K) = \begin{cases}
@@
 \begin{remark}
-A one-dimensional finitely generated field over its prime field is just a global field.
-A two-dimensional finitely generated field over $\mathbb{Q}$ is a function field in one variable over $\mathbb{Q}$.
+\begin{itemize}
+\item A one-dimensional finitely generated field is just a global field. 
+\item A two-dimensional finitely generated field over $\mathbb{Q}$ is a function field in one variable over $\mathbb{Q}$.
+\end{itemize}
 \end{remark}
@@
 Let $v\in\operatorname{Spv}(K)$. Then we have:
-\[\operatorname{rank}(v)\leq \operatorname{rrank}(v)\quad\text{and}\quad\operatorname{rrank}(v)+\dim k(v)\leq\dim(K) \]
+\begin{enumerate}[(i)]
+\item $\operatorname{rank}(v)\leq \operatorname{rrank}(v)$
+\item $\operatorname{rrank}(v)+\dim k(v)\leq\dim(K)$
+\end{enumerate}
+Moreover, if equality holds in (ii) and $K$ is finitely generated, then $k(v)$ is finitely generated as well, and $\Gamma_v$ is a finitely generated $\mathbb{Z}$-module.
 \end{lemma}
@@
 
-If $w\in\operatorname{Spv}_\text{rd}^{(i)}(K)$ for some $i\geq 1$ then there exists a unique $v(w)\in\operatorname{Spv}_\text{rd}^{(i-1)}(K)$ such that $\mathcal{O}_w\subset\mathcal{O}_{v(w)}$.
+If $w\in\operatorname{Spv}_\text{rd}^{(i)}(K)$ for some $i\geq 1$ then there exists a unique $v(w)\in\operatorname{Spv}_\text{d}^{(i-1)}(K)$ such that $\mathcal{O}_w\subset\mathcal{O}_{v(w)}$.
 $w$ is henselian if and only if $v(w)$ and $w\mod v(w)$ are henselian.

### 2017-03-13 Master Thesis/synopsis.tex
Add items to bibliography

@@
 \begin{proof}
-See Valued Fields 3.4.1 and 3.4.3.
+See \cite{EP05} 3.4.1 and 3.4.3.
 \end{proof}
@@
 \begin{proof}
-See [ZS] VI §17 Theorem 38, §17 Theorem 40.
+See \cite{ZS60} VI §17 Theorem 38, §17 Theorem 40.
 \end{proof}
@@
 \begin{proof}
-See Valued Fields 2.3, 4.1.4, [ZS] VI §3 Theorem 3.
+See \cite{EP05} 2.3, 4.1.4, \cite{ZS60} VI §3 Theorem 3.
 \end{proof}
@@
 \begin{proof}
-See Valued Fields Lemma 5.2.1, Lemma 5.2.2, Theorem 5.2.5.
+See \cite{EP05} Lemma 5.2.1, Lemma 5.2.2, Theorem 5.2.5.
 \end{proof}
@@
 \begin{proof}
-See Valued Fields 5.2.7.
+See \cite{EP05} 5.2.7.
 \end{proof}
@@
 Let $\overline{w}$ be the extension of $w\in\operatorname{Spv}(K_w)$ on $K^\text{s}$.
-By Valued Fields 3.1.5 we can find an extension $\overline{v}$ of $v$ on $K^\text{s}$ such that $\overline{w}$ is a specialization of $\overline{v}$.
+By \cite{EP05} 3.1.5 we can find an extension $\overline{v}$ of $v$ on $K^\text{s}$ such that $\overline{w}$ is a specialization of $\overline{v}$.
 Let $K_v$ be the fixed field of $G_{\overline{v}}\subset G_K$.
@@
 Then both $\mathcal{O}_{\overline{v}}$ and $\sigma(\mathcal{O}_{\overline{v}})$ are overrings of $\mathcal{O}_{\overline{w}}$.
-Now the overrings of valuation rings are totally ordered by inclusion and correspond to localizations of $\mathcal{O}_{\overline{w}}$ at prime ideals (See Valued Fields 2.3).
+Now the overrings of valuation rings are totally ordered by inclusion and correspond to localizations of $\mathcal{O}_{\overline{w}}$ at prime ideals (See \cite{EP05} 2.3).
 Since $\mathcal{O}_{\overline{v}}$ and $\sigma(\mathcal{O}_{\overline{v}})$ have the same Krull dimension, they must be the same, thus $\sigma\in G_{\overline{v}}$.
@@
 \begin{proof}
-See Valued Fields 4.4.1.
+See \cite{EP05} 4.4.1.
 \end{proof}
@@
 for some $d \leq \infty$, where $\Gamma'$ denotes the value group and $k'$ the residue field of the unique prolongation of $w$ to $L$.
-Note that $\operatorname{rank}(w) = \operatorname{rank}(v)$ in an algebraic field extension ([ZS], Ch.
+Note that $\operatorname{rank}(w) = \operatorname{rank}(v)$ in an algebraic field extension (\cite{ZS60}, Ch.
 VI, §11, Lemma 2).
@@
 \begin{proof}
-See [ZS] VI §10 Theorem 18
+See \cite{ZS60} VI §10 Theorem 18
 \end{proof}

### 2017-03-13 Master Thesis/synopsis.tex
Add remark for wlog assertion in 4.3

@@
 \[ \varphi_n:L'^\times/n \cong\mathrm{H}^1(L', \mu_n) \to \mathrm{H}^1(L,\mu_n) \cong L^\times/n \]
-Obviously, for intermediate fields $L\subset M$ of $\overline{K}/K$ the following diagram commutes:
+
+\begin{remark}\label{remark-localphi-extension}
+Obviously, for intermediate fields $L\subset M$ of $\overline{K}/K$, the following diagram commutes:
 \[ \begin{tikzcd}
@@
 \end{tikzcd} \]
+\end{remark}
+
 Let $P\in X_0$ and $v_P:K\to\mathbb{Z}\cup\{\infty\}$ be the corresponding normalized discrete valuation and let $U_P$ be the group of units of the valuation ring $\mathcal{O}_P$.
@@
 Note that $(K^\text{t}_P)' = K'^{\text{t}}_{P'}$ as proven in Lemma \ref{3.2}.
-\item Consider the field extension $K_P/K$. Without loss of generality, we can assume $K=K_P$. Since $\operatorname{char}k(P)=0$, by Hensel's lemma, $r_P/n$ is an isomorphism and the composition
+\item Consider the field extension $K_P/K$.
+By Remark \ref{remark-localphi-extension}, without loss of generality, we can assume $K=K_P$.
+Since $\operatorname{char}k(P)=0$, by Hensel's lemma, $r_P/n$ is an isomorphism and the composition
 \[\begin{tikzcd}

### 2017-03-14 Master Thesis/synopsis.tex
More cites
Remove unnecessary sentence
Add Theorem for Sylow groups of inertia group

@@
 Furthermore, $\overline{v}$ has the same value group as $v$ and the same residue field $k(\overline{v}) = k(v)$.
-$G_{K_v}$ is the decomposition group $G_{\overline{v}}\subset G_K$ of the unique extension of $\overline{v}$ to $K^\text{s}$.
-
+$G_{K_v}$ is the decomposition group of the unique extension of $\overline{v}$ to $K^\text{s}$:
+\[ G_{\overline{v}} = \{ \sigma\in G_K\mid \sigma(\mathcal{O}_{\overline{v}}) =\mathcal{O}_{\overline{v}} \} \subset G_K \]
 Conversely, taking the fixed field of the decomposition group of an extension $\overline{v}'$ of $v$ in $K^\text{s}$ yields a henselization of $K$ with respect to $v$.
@@
 \begin{theorem}\label{theorem-absoluteinertia}
-Let $K$ be a henselian field with respect to $v: K\to\Gamma\cup\{\infty\}$. Let $K^\text{t}$ be the absolute inertia field of $K$ with the unique extension $w$ of $v$. Then:
+Let $K$ be a henselian field with respect to $v: K\to\Gamma\cup\{\infty\}$ with unique extension $\overline{v}$ on $K^\text{s}$.
+Let $K^\text{t}$ be the absolute inertia field of $K$ with the unique extension $w$ of $v$, i.e.
+$K^\text{t}$ is the fixed field of the absolute inertia group $T$:
+\[ T = \{ \sigma\in G_K\mid \sigma(x)\equiv x\mod \mathfrak{m}_{\overline{v}} \quad \text{for all }x\in\mathcal{O}_{\overline{v}} \}\lhd G_K \]
+Then we have:
 \begin{enumerate}[(i)]
@@
 \begin{proof}
-See \cite{EP05} 5.2.7.
+See \cite{EP05} Theorem 5.2.7.
 \end{proof}
@@
 Let $\overline{w}$ be the extension of $w\in\operatorname{Spv}(K_w)$ on $K^\text{s}$.
-By \cite{EP05} 3.1.5 we can find an extension $\overline{v}$ of $v$ on $K^\text{s}$ such that $\overline{w}$ is a specialization of $\overline{v}$.
+By \cite{EP05} Lemma 3.1.5 we can find an extension $\overline{v}$ of $v$ on $K^\text{s}$ such that $\overline{w}$ is a specialization of $\overline{v}$.
 Let $K_v$ be the fixed field of $G_{\overline{v}}\subset G_K$.
@@
 
+\begin{theorem}\label{theorem-sylowgroups-inertiagroup}
+Let $K$ be a henselian field with respect to $v:K\to\Gamma\cup\{\infty\}$ with unique extension $\overline{v}$ on $K^\text{s}$, and $p=\operatorname{char}k(v)$. Then the absolute ramification group
+\[ V = \Big\{ \sigma\in G_K\mathop{\Big|} \frac{\sigma(x)}{x}-1\in\mathfrak{m}_{\overline{v}}\quad \text{for all }x\in {K^\text{s}}^\times \Big\} \lhd T \]
+is the unique $p$-Sylow group of the absolute inertia group $T$. $V$ is trivial if $p=0$. We have the following isomorphism of profinite groups:
+\[ T/V \cong \prod_{\ell\neq p\text{ prime}}\mathbb{Z}_\ell^{r_\ell} \]
+where $r_\ell = \dim_{\mathbb{F}_\ell}(\Gamma/\ell\Gamma)$.
+\end{theorem}
+
+\begin{proof}
+See \cite{EP05} Theorem 5.3.3.
+\end{proof}
+
 \begin{theorem}[\textit{F.K. Schmidt}]\label{theorem-fkschmidt}
@@
 \begin{proof}
-See \cite{EP05} 4.4.1.
+See \cite{EP05} Theorem 4.4.1.
 \end{proof}
@@
 For $x\in\mathfrak{X}_1$ the local ring $\mathcal{O}_{\mathfrak{X}, x}$ is a discrete valuation ring and defines a valuation $v_x\in\operatorname{Spv}_\text{d}^{(1)}(K)$.
-By [K1] Theorem 0.8, the map
+By \cite{Ka86} Theorem 0.8, the map
 \[ \mathrm{H}^3(K)\longrightarrow \prod_{x\in\mathfrak{X}_1}\mathrm{H}^3(K_{v_x}) \]
-is injective and its image is contained in $\bigoplus \mathrm{H}^3(K_{v_x})$. By [K1] Lemma 1.4, for a discrete valuation $v$ of $K$ the map
+is injective and its image is contained in $\bigoplus \mathrm{H}^3(K_{v_x})$. By \cite{Ka86} Lemma 1.4, for a discrete valuation $v$ of $K$ the map
 \[ \mathrm{H}^3(K_v) \longrightarrow \prod_w\mathrm{H}^3(K_w) \]
@@
 \end{tikzcd} \]
-Let us consider the morphism $f:X'\times_{k'} k\to X'$. We want to see that $f$ splits completely. For $P'\in X'_0$ we have as topological spaces:
+Let us consider the morphism $f:X'\times_{k'} k\to X'$. For $P'\in X'_0$ we have as topological spaces:
 \[ f^{-1}(P') = \operatorname{Spec} k(P')\times_{k'} k \]

### 2017-03-14 Master Thesis/synopsis.tex
Clarify semi-direct product in 2.5

@@
 \item $w$ has value group $\Gamma$ and residue class field $k(v)^\text{s}$.
-\item $K^\text{t}$ is a Galois extension of $K$ and $\operatorname{Gal}(K^\text{t}/K)\cong G_{k(v)}$. There exists an inclusion-preserving one-to-one correspondence: 
+\item $K^\text{t}$ is a Galois extension of $K$ and $\operatorname{Gal}(K^\text{t}/K)\cong G_{k(v)}$ as profinite groups. There exists an inclusion-preserving one-to-one correspondence: 
 \begin{align*}
@@
 \begin{proof}
-Let $G_\ell$ be an pro-$\ell$ Sylow group of $G_K$.
-By the ramification theory for general valuations ([E], §20) we have $G_\ell\cong\mathbb{Z}_\ell(1)^r\rtimes G_k(\ell)$ where the action is defined via the cyclotomic character and $r = \dim_{\mathbb{F}_\ell}(\Gamma/\ell\Gamma)$.
-In particular, $r\leq 2$ and $G_k(\ell)$ is the pro-$\ell$-Sylow group of $G_k$ hence $G_k(\ell)\cong \mathbb{Z}_\ell$ or $G_k(\ell) \cong 1$.
+Let $T$ be the absolute inertia group of $K$ and consider the short exact sequence from Theorem \ref{theorem-absoluteinertia}:
+\[ 1\longrightarrow T\longrightarrow G_K\longrightarrow G_k\longrightarrow 1 \]
+Let $G_\ell$ be an $\ell$-Sylow group of $G_K$.
+By Theorem \ref{theorem-sylowgroups-inertiagroup}, the $\ell$-Sylow group of $T$ isomorphic to $\mathbb{Z}_\ell^r$, where $r =\dim_{\mathbb{F}_\ell}(\Gamma/\ell\Gamma)$.
+Furthermore, $G_k\subset G_{\mathbb{F}_p}\cong\hat{\mathbb{Z}}$, thus the $\ell$-Sylow group $G_k(\ell)$ of $G_k$ is either isomorphic to $\mathbb{Z}_\ell$ or trivial.
+The exact sequence above gives rise to the short exact sequence of their $\ell$-Sylow subgroups:
+\[ 0\longrightarrow\mathbb{Z}_\ell^r \longrightarrow G_\ell \longrightarrow G_k(\ell)\longrightarrow 1 \]
+Since $G_k(\ell)$ is a free pro-$\ell$ group, this sequence splits, and we have an isomorphism $G_\ell \cong \mathbb{Z}_\ell^r\rtimes G_k(\ell)$ with $r =\dim_{\mathbb{F}_\ell}(\Gamma/\ell\Gamma)\leq \operatorname{rrank}(\Gamma)\leq 2$.
 
@@
 Since $\mu_\ell\subset K$, the action of $G_k$ on the pro-$\ell$-Sylow group of the inertia group of $w$ factors through $G_k(\ell)$.
-We see that the canonical projection $G_K\to G_K(\ell)$ is an isomorphism.
+We see that the canonical projection $G_K\to G_K(\ell)$ restricts to an isomorphism $G_\ell\to G_K(\ell)$.
 Therefore we have:

### 2017-04-06 Master Thesis/synopsis.tex
Fix 2.2

@@
 \begin{theorem}[\textit{F.K. Schmidt}]\label{theorem-fkschmidt}
-Let $v_1,v_2\in\operatorname{Spv}(K)$ be two inequivalent henselian valuations of rank $1$ on a field $K$. Then $K$ is separably closed.
+Let $v_1,v_2\in\operatorname{Spv}(K)$ be two independent henselian valuations on a field $K$, i.e. $\mathcal{O}_{v_1}\mathcal{O}_{v_2}=K$. Then $K$ is separably closed.
 \end{theorem}
@@
 \end{proof}
+
+\begin{remark}
+Two valuations of rank $1$ are independent iff they are inequivalent.
+\end{remark}
 
@@
 Now $L$ is separably closed by Theorem \ref{theorem-fkschmidt}.
-Therefore by Theorem \ref{theorem-henselization} we get $G_{{w_1}}\cap G_{{w_1}}=G_L =1$.
+Therefore, by Theorem \ref{theorem-henselization}, we get $G_{{w_1}}\cap G_{{w_1}}=G_L =1$.
 \end{proof}
@@
 \begin{lemma}\label{2.2}
-Let $K$ be henselian with respect to two different valuations $w_1,w_2\in |\operatorname{Spv}(K)|$. Then the following statements are equivalent: 
-\begin{enumerate}[(i)]
-\item $K$ is not algebraically closed.
-\item $w_1,w_2\in\operatorname{Spv}_\text{rd}^{(2)}(K),\ v(w_1)=v(w_2)$ and $k(v(w_1))$ is separably closed.
-\end{enumerate}
+Let $K$ be henselian with respect to two different valuations $w_1,w_2\in |\operatorname{Spv}(K)|$.
+Then $K$ is algebraically closed or $w_1,w_2\in\operatorname{Spv}_\text{rd}^{(2)}(K),\ v(w_1)=v(w_2)$ and $k(v(w_1))$ is separably closed.
 In this case, we have $\operatorname{cd}(G_K)\leq 2$.
@@
 \begin{proof}
-\begin{itemize}
-\item Assume that $K$ is not algebraically closed.
+Assume that $K$ is not algebraically closed.
+$w_1$ and $w_2$ can't be trivial, for otherwise $\dim k(w_i)=\dim (K)\neq 0$.
 By the Theorem of F.K.
-Schmidt \ref{theorem-fkschmidt}, we can see that $w_1,w_2$ are not of rank $1$.
-Thus they are of rank $2$ and by Lemma \ref{rank-dimension-inequality} we can see $w_1,w_2\in\operatorname{Spv}_\text{rd}^{(2)}(K)$.
-Set $v_i=v(w_i)\in\operatorname{Spv}^{(1)}_\text{rd}(K)$.
+Schmidt \ref{theorem-fkschmidt}, we can see that $w_1,w_2$ are dependent.
+Since two different valuations of rank $1$ are independent, at least one of $w_1$ and $w_2$ is of rank $2$.
+Now, let's assume that $w_1$ is of rank $1$ and $w_2$ is of rank $2$.
+Since they are dependent, it follows $v(w_2) = w_1$ and $w_1$ has a non-trivial specialization, a contradiction to $w_1\in |\operatorname{Spv}(K)|$.
+
+Thus $w_1,w_2$ are both of rank $2$ and by Lemma \ref{rank-dimension-inequality} we can see that $w_1,w_2\in\operatorname{Spv}_\text{rd}^{(2)}(K)$.
+Set $v_i=v(w_i)\in\operatorname{Spv}^{(1)}_\text{d}(K)$.
 By Lemma \ref{2.2-pre} $K$ is henselian with respect to both $v_1$ and $v_2$.
@@
 Furthermore $k(v_1) = k(v_2)$ is henselian with respect to two different valuations of rank $1$, namely $w_1\mod v_1$ and $w_2\mod v_1$, making $k(v_1)$ separably closed.
-\item Now assume (ii) holds.
-\item The second assertion follows from the first and from [K2], §5, Theorem 1.\qedhere
-\end{itemize}
+
+The second assertion follows from the first and from [K2], §5, Theorem 1.
 \end{proof}
@@
 According to Lemma \ref{2.5} we have $\operatorname{rrank}(w)=\dim_\mathbb{Q}(\Gamma_w\otimes_\mathbb{Z}\mathbb{Q} )\leq \dim_{\mathbb{F}_\ell}(\Gamma_w /\ell\Gamma_w) = 2$.
-By Lemma \ref{rank-dimension-inequality}, this inequality is actually an equality.
-\qedhere
+TODO \iffalse By Lemma \ref{rank-dimension-inequality}, this inequality is actually an equality.\fi \qedhere
 \end{enumerate}
@@
 \item It suffices to prove this assertion for finite groups and for $A=\mathbb{Z}$.
-For $A$ finite we have $nA = 0$ for $n = \#A$.
+For finite $A$ we have $nA = 0$ for $n = \#A$.
 Since $\mathbb{Z}$ has no nilpotent elements, the intersection of all prime ideals of $\mathbb{Z}$ is already trivial.

### 2017-04-06 Master Thesis/synopsis.tex
Kinda complete 2.5

@@
 Let $L=K(\mu_{\ell},\sqrt{-1})/K$. Consider the restriction and corestriction maps:
-\[ \operatorname{res}: \mathrm{H}^3(K,\mathbb{Z}/\ell\mathbb{Z}) \to \mathrm{H}^3(L,\mathbb{Z}/\ell\mathbb{Z}) \qquad \operatorname{cor}:\mathrm{H}^3(L,\mathbb{Z}/\ell\mathbb{Z})\to \mathrm{H}^3(K,\mathbb{Z}/\ell\mathbb{Z}) \]
+\[ \begin{tikzcd}
+\mathrm{H}^3(K,\mathbb{Z}/\ell\mathbb{Z}) \ar[r, "\operatorname{res}"] & \mathrm{H}^3(L,\mathbb{Z}/\ell\mathbb{Z})\ar[r, "\operatorname{cor}"]& \mathrm{H}^3(K,\mathbb{Z}/\ell\mathbb{Z}) 
+\end{tikzcd} \]
 We have $\operatorname{cor}\circ\operatorname{res} = [L:K]$.
@@
 By Theorem \ref{theorem-sylowgroups-inertiagroup}, the $\ell$-Sylow group of $T$ isomorphic to $\mathbb{Z}_\ell^r$, where $r =\dim_{\mathbb{F}_\ell}(\Gamma/\ell\Gamma)$.
-Furthermore, $G_k\subset G_{\mathbb{F}_p}\cong\hat{\mathbb{Z}}$, thus the $\ell$-Sylow group $G_k(\ell)$ of $G_k$ is either isomorphic to $\mathbb{Z}_\ell$ or trivial.
+Furthermore, $G_k\subset G_{\mathbb{F}_p}\cong\hat{\mathbb{Z}}$, thus the $\ell$-Sylow group $G_{k,\ell}$ of $G_k$ is either isomorphic to $\mathbb{Z}_\ell$ or trivial.
 The exact sequence above gives rise to the short exact sequence of their $\ell$-Sylow subgroups:
-\[ 0\longrightarrow\mathbb{Z}_\ell^r \longrightarrow G_\ell \longrightarrow G_k(\ell)\longrightarrow 1 \]
-Since $G_k(\ell)$ is a free pro-$\ell$ group, this sequence splits, and we have an isomorphism $G_\ell \cong \mathbb{Z}_\ell^r\rtimes G_k(\ell)$ with $r =\dim_{\mathbb{F}_\ell}(\Gamma/\ell\Gamma)\leq \operatorname{rrank}(\Gamma)\leq 2$.
+\[ 0\longrightarrow\mathbb{Z}_\ell^r \longrightarrow G_\ell \longrightarrow G_{k,\ell}\longrightarrow 1 \]
+Since $G_{k,\ell}$ is a free pro-$\ell$ group, this sequence splits and we have an isomorphism $G_\ell \cong \mathbb{Z}_\ell^r\rtimes G_{k,\ell}$ with $r =\dim_{\mathbb{F}_\ell}(\Gamma/\ell\Gamma)\leq \operatorname{rrank}(\Gamma)\leq 2$.
 
-Let $G_K(\ell)$ be the maximal pro-$\ell$-quotient of $G_K$.
-Since $\mu_\ell\subset K$, the action of $G_k$ on the pro-$\ell$-Sylow group of the inertia group of $w$ factors through $G_k(\ell)$.
-We see that the canonical projection $G_K\to G_K(\ell)$ restricts to an isomorphism $G_\ell\to G_K(\ell)$.
-Therefore we have:
-\[ \mathrm{H}^3(K,\mathbb{Z}/\ell\mathbb{Z}) \cong\mathrm{H}^3(G_K(\ell),\mathbb{Z}/\ell\mathbb{Z}) \cong\mathrm{H}^3(G_\ell,\mathbb{Z}/\ell\mathbb{Z}) \]
-Now the first assertion follows from [Se], Ch. I, §4.
+Let $G_K(\ell)$ be the maximal pro-$\ell$ quotient of $G_K$.
+Since $\mu_\ell\subset K$, the action of $G_k$ on the $\ell$-Sylow group of the inertia group of $w$ factors through the maximal pro-$\ell$ quotient $G_k(\ell)$ of $G_k$.
+Since $G_{k,\ell}\cong G_k(\ell)$, we also see that the canonical projection $G_K\to G_K(\ell)$ restricts to an isomorphism $G_\ell\to G_K(\ell)$.
+Now consider the following composition:
+\[ \begin{tikzcd}
+\mathrm{H}^3(G_K(\ell),\mathbb{Z}/\ell\mathbb{Z}) \ar[r, "\operatorname{inf}"] & \mathrm{H}^3(G_K,\mathbb{Z}/\ell\mathbb{Z}) \ar[r, "\operatorname{res}"] & \mathrm{H}^3(G_\ell,\mathbb{Z}/\ell\mathbb{Z})
+\end{tikzcd} \]
+Since $\operatorname{res}\circ\operatorname{inf}$ is an isomorphism, $\operatorname{inf}$ is injective and $\operatorname{res}$ is surjective.
+We have $\operatorname{cor}\circ\operatorname{res} = [G_K:G_\ell]$.
+However, $[G_K:G_\ell]$ is coprime to $\ell$, which makes $\operatorname{res}$ injective as well.
+Therefore, $\operatorname{inf}$ and $\operatorname{res}$ are both isomorphisms and it suffices to calculate $\mathrm{H}^3(G_\ell,\mathbb{Z}/\ell\mathbb{Z})$.
+By [Se], Ch.
+I, §4, Proposition 22 we have:
+\[\operatorname{cd}(G_\ell) = \operatorname{cd}(\mathbb{Z}_\ell^r\rtimes G_k(\ell)) = \begin{cases}
+r, & G_k(\ell) = 0 \\
+r + 1, & G_k(\ell)\cong\mathbb{Z}_\ell
+\end{cases} \] 
+since $\operatorname{cd}(\mathbb{Z}_\ell)=1$.
+Thus $\mathrm{H}^3(G_\ell, \mathbb{Z}/\ell\mathbb{Z}) =0$ if $G_k(\ell) = 0$ or $r < 2$.
+Now let $G_k(\ell) \cong\mathbb{Z}_\ell$ and $r = 2$.
+According to the proof of [Se], Ch.
+I, §4, Proposition 22, we have:
+\[ \mathrm{H}^3(G_\ell,\mathbb{Z}/\ell\mathbb{Z}) = \mathrm{H}^1(G_k(\ell), \mathrm{H}^2 (\mathbb{Z}_\ell^r, \mathbb{Z}/\ell\mathbb{Z})) \]
+By [NSW] Theorem 2.4.6, we have $\mathrm{H}^2(\mathbb{Z}_\ell^r,\mathbb{Z}/\ell\mathbb{Z}) \cong\mathrm{H}^1(\mathbb{Z}_\ell, \mathrm{H}^1(\mathbb{Z}_\ell, \mathbb{Z}/\ell\mathbb{Z})) = \mathbb{Z}/\ell\mathbb{Z}$, since $\mathrm{H}^1(\mathbb{Z}_\ell, \mathbb{Z}/\ell\mathbb{Z})=\operatorname{Hom}_\text{cont}(\mathbb{Z}_\ell, \mathbb{Z}/\ell\mathbb{Z})=\mathbb{Z}/\ell\mathbb{Z}$.
+Therefore, $\mathrm{H}^3(G_\ell,\mathbb{Z}/\ell\mathbb{Z})=\mathbb{Z}/\ell\mathbb{Z}$.
 

### 2017-04-06 Master Thesis/synopsis.tex
Fix mistake in original paper

@@
 \begin{align*}
-\operatorname{Spv}_\text{d}(K) &= \{v\in\operatorname{Spv}(K)\mid\operatorname{rank}(v)=\operatorname{rrank}(v) \} \\
-\operatorname{Spv}_\text{rd}(K) &= \{v\in\operatorname{Spv}_\text{d}(K)\mid \operatorname{rrank}(v)+\dim k(v)=\dim(K) \}
+\operatorname{Spv}_\text{d}(K) &= \{v\in\operatorname{Spv}(K)\mid \operatorname{rrank}(v)+\dim k(v)=\dim(K) \} \\
+\operatorname{Spv}_\text{rd}(K) &= \{v\in\operatorname{Spv}_\text{d}(K)\mid \operatorname{rank}(v)=\operatorname{rrank}(v) \}
 \end{align*}
@@
 The uniqueness of $w$ is again a consequence of Lemma \ref{2.2}.
-\item Choose the prime $\ell$ different from $\operatorname{char} k(w)$ so that (i) holds.
-According to Lemma \ref{2.5} we have $\operatorname{rrank}(w)=\dim_\mathbb{Q}(\Gamma_w\otimes_\mathbb{Z}\mathbb{Q} )\leq \dim_{\mathbb{F}_\ell}(\Gamma_w /\ell\Gamma_w) = 2$.
-TODO \iffalse By Lemma \ref{rank-dimension-inequality}, this inequality is actually an equality.\fi \qedhere
+\item Choose the prime $\ell$ different from $\operatorname{char} k(w)$ so that (i) holds. According to Lemma \ref{2.5} we have:
+\[\operatorname{rrank}(w)=\dim_\mathbb{Q}(\Gamma_w\otimes_\mathbb{Z}\mathbb{Q} )\geq \dim_{\mathbb{F}_\ell}(\Gamma_w /\ell\Gamma_w) = 2 = \dim(K)\qedhere\]
 \end{enumerate}
@@
 For all $v\in\operatorname{Spv}(K)$ we choose a fixed henselization $K_v\subset\overline{K}$ with Lemma \ref{lemma-henselization-specialization}, such that if $v,w\in\operatorname{Spv}(K)$ and $w$ is a specialization of $v$ we have $K_v\subset K_w$.
-This means that if $v$ is nontrivial, we get $w\in\operatorname{Spv}_\text{rd}^{(2)}(K)$ and $v=v(w)\in\operatorname{Spv}_\text{d}^{(1)}(K)$.
+If $v$ is nontrivial, then in this case we get $w\in\operatorname{Spv}_\text{rd}^{(2)}(K)$ and $v=v(w)\in\operatorname{Spv}_\text{d}^{(1)}(K)$.
 We set $K'_{\sigma(w)} = (K_w)'$ for $w\in\operatorname{Spv}_\text{d}^{(2)}(K)$.
@@
 
-Let $w_1,w_2 \in\operatorname{Spv}_\text{d}^{(2)}(K)$ be two different valuations. By Lemma \ref{2.2} the following statements are equivalent:
+Let $w_1,w_2 \in\operatorname{Spv}_\text{d}^{(2)}(K)$ be two different valuations. The following statements are equivalent:
 \begin{enumerate}[(i)]
@@
 \end{enumerate}
+The implication (i)$\implies$(ii) follows from Lemma \ref{2.2}. TODO
+
 In this case, $k(v')$ is separably closed where $v'$ is the unique extension of the valuation on $K_v$ to $K_{w_1}K_{w_2}$ where $v=v(w_1)$.
@@
 Since a global field has no non-trivial algebraic extension in which every prime splits completely, we get $N=G_{K_v}$.
-This shows that $K'_{v'} = (K_v)'$ is a henselization of $K'$ with respect to $v'=\sigma(v)$.
+This shows that $K'_{v'} := (K_v)'$ is a henselization of $K'$ with respect to $v'=\sigma(v)$.
 Furthermore, let $w_1,w_2\in\operatorname{Spv}_\text{rd}^{(2)}(K)$ with $w_1\neq w_2$ and $v=v(w_1)=v(w_2)$, then:
@@
 
-By Lemma \ref{3.2} $\sigma$ induces a bijection $\sigma: X_0\to X'_0,\ P\mapsto P'$ and for any $P\in X_0$ an isomorphism $\varphi_P: k(P')\to k(P)$.
+By Lemma \ref{3.2} $\sigma$ induces a bijection $\sigma: X^{\phantom{'}}_0\to X'_0,\ P\mapsto P'$ and for any $P\in X_0$ an isomorphism $\varphi_P: k(P')\to k(P)$.
 

### 2017-05-06 Master Thesis/bibliography.tex
Add Kato to bibliography
Add items to bibliography
Add introduction

@@
-\begin{thebibliography}{99999}
+\begin{thebibliography}{999999}
+
+\bibitem[EP05]{EP05}
+    A.J. Engler, A. Prestel
+    \textit{Valued Fields}.
+    Springer Monographs in Mathematics (2005).
+
+\bibitem[Ka86]{Ka86}
+    K. Kato
+    \textit{A Hasse principle for two dimensional global fields, with an appendix by J.-L. Colliot Thélène}.
+    J. reine angew. Math. \textbf{366} (1986), 142\---183.
+
+\bibitem[Ne69]{Ne69}
+    J. Neukirch
+    \textit{Kennzeichnung der $p$-adischen und der endlichen algebraischen Zahlkörper}.
+    Invent. Math. \textbf{6} (1969), 296\---314.
+
+\bibitem[Ne77]{Ne77}
+    J. Neukirch
+    \textit{Über die absoluten Galoisgruppen algebraischer Zahlkörper}. Asterisque, \textbf{41\---42}, Soc. Math. France, Paris (1977), 67\---79.
+
+\bibitem[Po90]{Po90}
+    F. Pop
+    \textit{On the Galois theory of function fields of one variable over number fields}.
+    J. reine angew. Math. \textbf{406} (1990), 200\---218.
+
+\bibitem[Po94]{Po94}
+    F. Pop
+    \textit{On Grothendieck's conjecture of birational anabelian geometry}.
+    Ann. Math. \textbf{138} (1994), 145\---182.
+
+\bibitem[Po95]{Po95}
+    F. Pop
+    \textit{On Grothendieck's conjecture of birational anabelian geometry II}, Preprint (1995).
 
 \bibitem[Sp96]{Sp96}
-	M. Spiess (1996)
+	M. Spiess
 	\textit{An arithmetic proof of Pop's Theorem concerning Galois groups of function fields over number fields}.
-    J. reine angew. Math. \textbf{478}, 107\---126.
+    J. reine angew. Math. \textbf{478} (1996), 107\---126.
+
+\bibitem[Uc76]{Uc76}
+    K. Uchida
+    \textit{Isomorphisms of Galois groups}.
+    J. Math. Soc. Japan \textbf{28} (1976), no. 4, 617\---620.
+
+\bibitem[Uc77]{Uc77}
+    K. Uchida
+    \textit{Isomorphisms of Galois groups of algebraic function fields}.
+    Ann. Math. (2) \textbf{106} (1977), no. 3, 589\---598.
+
+\bibitem[ZS60]{ZS60}
+    O. Zariski, P. Samuel
+    \textit{Commutative Algebra, Volume II} (1960). 
+    Van Nostrand.
 

### 2017-05-06 Master Thesis/introduction.tex
Update notations
Add n-th roots of unity to list of notations
Add introduction

@@
 \chapter{Introduction}
+
+Given an algebraic variety $X$, we can ask ourselves how much information about $X$ is contained in the knowledge of its étale fundamental group $\pi_1^{\text{ét}}(X)$.
+Especially interesting are schemes which should be reconstructible solely from its étale fundamental group.
+If $X$ is the spectrum of a field $K$, the étale fundamental group of $X$ is just its absolute Galois group $G_K$.
+
+Finite fields have an abelian absolute Galois group.
+In fact, they all have the same absolute Galois group $\hat{\mathbb{Z}}$, therefore not much information can be retrieved from these groups.
+The study of these types of questions belong in the area of \textit{anabelian geometry}.
+This term refers to the fact that the less abelian $\pi_1^\text{ét}(X)$ is, the more information it carries about $X$.
+
+Number fields stand in stark contrast to finite fields.
+Two number fields $k$, $k'$ with isomorphic Galois groups $G_k$, $G_{k'}$ are isomorphic as fields as well.
+It was J.
+Neukirch who investigated this question for the first time in the end of the 1960s (see \cite{Ne69}, \cite{Uc76}, \cite{Ne77}, \cite{Uc77}).
+Together with K.
+Uchida, their studies resulted in the following theorem for global fields:
+
+\begin{theorem}[\textit{Neukirch-Uchida}]\label{thm:neukirch-uchida}
+Let $k,k'$ be two global fields.
+Then any isomorphism $\sigma: G_k\to G_{k'}$ between their absolute Galois groups is induced by a unique field isomorphism $\varphi: k'^{\text{s}}\to k^\text{s}$, i.e:
+\[\sigma(g) =\varphi^{-1}g\varphi \quad \text{for all }g\in G_k\] 
+In particular $\varphi$ induces a field isomorphism $k'\to k$.
+\end{theorem}
+
+This means that global fields should be reconstructible from their Galois groups. Furthermore, the Galois group $G_k$ also contains important arithmetic information from $k$.
+
+It is believed that spectra of finitely generated fields are the building blocks of the anabelian world, which led Grothendieck to the conjecture that Theorem \ref{thm:neukirch-uchida} should also hold for arbitrary finitely generated fields $k$ and $k'$ of characteristic $0$.
+
+This has been proved by F.
+Pop (see \cite{Po90}, \cite{Po94}, \cite{Po95}) using model theory.
+In this thesis, we want to follow M.
+Spiess \cite{Sp96} to prove the statement for finitely generated fields with transcendence degree $1$ over $\mathbb{Q}$, using a purely arithmetical approach and a direct generalization of the original proof of Theorem \ref{thm:neukirch-uchida}:
+
+\pagebreak
 
@@
 Let $K,K'$ be two finite field extensions of $\mathbb{Q}(T)$.
-If there is an isomorphism $\sigma:G_K\to G_{K'}$ between their absolute Galois groups, then there exists a unique field isomorphism $\varphi:\overline{K'}\to\overline{K}$ with $\sigma(g)=\varphi^{-1}g\varphi$ for all $g\in G_K$.
+Then any isomorphism $\sigma:G_K\to G_{K'}$ between their absolute Galois groups is induced by a unique field isomorphism $\varphi:\overline{K'}\to\overline{K}$, i.e:
+\[\sigma(g) =\varphi^{-1}g\varphi \quad \text{for all }g\in G_k\]
 In particular $\varphi$ induces a field isomorphism $K'\to K$.
-\end{theorem}
-
-\begin{theorem}[\textit{Neukirch-Uchida}]\label{thm:neukirch-uchida}
-Let $k,k'$ be two global fields.
-Then any isomorphism $G_k\to G_{k'}$ is induced by a unique isomorphism $k'^{\text{s}}\to k^\text{s}$.
-In particular, if $k,k'$ have isomorphic absolute Galois groups, they are isomorphic as fields.
 \end{theorem}
@@
 \mathbb{C} & \text{complex numbers}\\
+\mathbb{F}_q & \text{finite field with $q$ elements}\\
+\mathbb{Z}_p & \text{$p$-adic integers}\\
+\hat{\mathbb{Z}} & \text{profinite completion of $\mathbb{Z}$}\\
 \\
@@
 \operatorname{char}(K) & \text{characteristic of $K$}\\
-\operatorname{trdeg}(L/K) & \text{transcendence degree of $L/K$}
+\operatorname{trdeg}(L/K) & \text{transcendence degree of $L/K$}\\
+\mu_n & \text{$n$-th roots of unity in a separable closure}
 \end{array}$

### 2017-05-06 Master Thesis/thesis.tex
Remarks are numbered
Add working title
Update title
Add proof for 4.5
Solve TeX warnings
Add introduction

@@
 \author{Yichuan Shen}
-\title{Title of the Master Thesis}
-\makeindex
+\title{Galois Characterization of Function Fields of \\ Transcendence Degree $1$ over Number Fields}
 
@@
 \theoremstyle{remark}
-\newtheorem*{remark}{Remark}
+\newtheorem{remark}[theorem]{Remark}
 \newtheorem{step}{Step}
@@
 \vspace{10mm}
+\linespread{2}\selectfont
 \begin{huge}\@title \end{huge}\\
+\linespread{1.1}\selectfont
 \vspace{10mm}
@@
 \vfill
-\paragraph{Declaration.} I declare that I carried out this master thesis independently, and only with the cited sources, literature and other professional sources.
+\paragraph{Declaration.} I declare that I carried out this master thesis independently, and only with the cited sources, literature, and other professional sources.
 This thesis was not used in the same or in a similar version to achieve an academic grading or is being published elsewhere.\\
@@
 \input{bibliography.tex}
-
-\printindex
 \end{document}

### 2017-05-06 Master Thesis/introduction.tex
Add chapter overview

@@
 
-Given an algebraic variety $X$, we can ask ourselves how much information about $X$ is contained in the knowledge of its étale fundamental group $\pi_1^{\text{ét}}(X)$.
-Especially interesting are schemes which should be reconstructible solely from its étale fundamental group.
+Given an algebraic variety $X$, we can ask ourselves how much information about $X$ is contained in its étale fundamental group $\pi_1^{\text{ét}}(X)$.
+Especially interesting are schemes which are reconstructible solely from its étale fundamental group.
 If $X$ is the spectrum of a field $K$, the étale fundamental group of $X$ is just its absolute Galois group $G_K$.
@@
 
+\section{Overview}
+
+In the first chapter, we want to recap some results from valuation theory.
+In the second part, we will prove some lemmata in order to be able to characterize henselian fields over $K$ of valuations of rational rank $2$ in terms of their Galois cohomology groups of dimension $3$.
+
+In the second chapter, we will show that an isomorphism $\sigma: G_K\to G_{K'}$ induces bijections on the equivalence classes of defectless valuations of $K$ and $K'$ using the cohomological characterization of henselian fields.
+With the Theorem of Neukirch-Uchida \ref{thm:neukirch-uchida}, we will also show that $\sigma$ induces a field isomorphism between the residue fields of corresponding defectless valuations of rational rank $1$.
+Using the local correspondences, we will proceed to show that $\sigma$ induces group isomorphisms $K^\times/n\to K'^\times/n$ for all $n\in\mathbb{N}$.
+
+In the third chapter, we want to use the correspondences established in the second chapter to prove Theorem \ref{thm:main-result}.
+In order to do so, we need to discuss so-called free-by-finite groups.
+Using the existence of a so-called admissible set, we can construct a group isomorphism $K^\times\to K'^\times$.
+Finally, we will show that the group isomorphism can be extended to a field isomorphism $K\to K'$.
+By passing to the limits, we obtain a field isomorphism $\varphi: \overline{K}\to\overline{K'}$ which induces $\sigma$.
+
+\section{Acknowledgments}
+
 \clearpage

### 2017-05-06 Master Thesis/synopsis.tex
Add chapter overview

@@
 \begin{lemma}\label{spv-topology}
-$\operatorname{Spv}(K)$ with the topology defined above is a quasicompact $T_0$-space.
-In particular the set of closed points $|\operatorname{Spv}(K)|$ of $\operatorname{Spv}(K)$ is a quasicompact $T_1$-space, endowed with the induced topology.
+$\operatorname{Spv}(K)$ with the topology defined above is a quasi-compact $T_0$-space.
+In particular the set of closed points $|\operatorname{Spv}(K)|$ of $\operatorname{Spv}(K)$ is a quasi-compact $T_1$-space, endowed with the induced topology.
 Let $v,w\in\operatorname{Spv}(K)$.
@@
 \[ \varphi_{ij}: |\operatorname{Spv}(K_j)|\to |\operatorname{Spv}(K_i)| \]
-be the restriction maps. They are continuous. Since $|\operatorname{Spv}(K)|$ is quasicompact by Lemma \ref{spv-topology} we can find a finite set $S\subset |\operatorname{Spv}(K)|$ such that:
+be the restriction maps. They are continuous. Since $|\operatorname{Spv}(K)|$ is quasi-compact by Lemma \ref{spv-topology} we can find a finite set $S\subset |\operatorname{Spv}(K)|$ such that:
 \[ \bigcup_{w\in S}\varphi^{-1}_{i(w)}(U_{i(w)}) = |\operatorname{Spv}(K)| \]
@@
 \[ G_{k(v)} \cong G_{K_v}/G_{K_v^\text{t}} \to G_{K'_{v'}}/G_{(K')_{v'}^\text{t}}\cong G_{k(v')} \]
-The rest follows from the Theorem of Neukirch-Uchida.
+The rest follows from the Theorem of Neukirch-Uchida \ref{thm:neukirch-uchida}.
 \end{proof}
@@
 
-This shows that $P$ has only one pre-image under $f'$. Therefore, it has only one pre-image under $f$, say $P'$, and $P'$ has only one pre-image under $\pi$. This concludes the proof.
+This shows that $P$ has only one preimage under $f'$. Therefore, it has only one preimage under $f$, say $P'$, and $P'$ has only one preimage under $\pi$. This concludes the proof.
 \end{proof}
@@
 
-\paragraph{} Let $n\in\mathbb{N}$. In the previous remark, we have reduced the problem, so that the following diagram commutes: 
+\paragraph{} Let $n\in\mathbb{N}$. In the previous remark, we have replaced $\sigma$, so that now the following diagram commutes: 
 \[ \begin{tikzcd}
@@
 \end{tikzcd} \]
-Now the operation of $G_K$ and $G_{K'}$ on $\mu_n$ factors through $G_k$, thus we have $\sigma(g)\xi = g\xi$ for all $\xi\in\mu_n$ and $g\in G_K$.
+The operation of $G_K$ and $G_{K'}$ on $\mu_n$ factors through $G_k$, thus we have $\sigma(g)\xi = g\xi$ for all $\xi\in\mu_n$ and $g\in G_K$.
 Therefore, for all intermediate fields $L$ of $\overline{K}/K$, $\sigma$ induces an isomorphism:

### 2017-05-06 Master Thesis/synopsis.tex
Fix typo

@@
 where $\operatorname{NS}(C)$ denotes the finitely generated Néron-Severi group of $C$.
-Now $\operatorname{Pic}^0(C)$ can be identified with the $F$-rational points of the Jacobian variety $J$ of $C$.
-$J(F)$ is finitely generated by the Mordell-Weil Theorem \ref{mordell-weil-theorem}, hence $\operatorname{Pic}(C)$ is finitely generated as well.
+Now $\operatorname{Pic}^0(C)$ can be identified with the $F$-rational points of the Jacobian variety $J(C)$ of $C$.
+$J(C)$ is finitely generated by the Mordell-Weil Theorem \ref{mordell-weil-theorem}, hence $\operatorname{Pic}(C)$ is finitely generated as well.
 \end{proof}

### 2017-05-10 Master Thesis/synopsis.tex
Add lemma for defectless valuations

@@
 
+\begin{lemma}\phantomsection\label{lemma-defectless-valuations}
+\begin{enumerate}[(i)]
+\item $\operatorname{Spv}_\text{d}^{(1)}(K) = \operatorname{Spv}_\text{rd}^{(1)}(K)$
+\item $\operatorname{Spv}^{(2)}(K) = \operatorname{Spv}_\text{d}^{(2)}(K)$
+\end{enumerate}
+\end{lemma}
+
+\begin{proof}
+\begin{enumerate}[(i)]
+\item Let $v$ be a defectless valuation of rational rank $1$.
+By Lemma \ref{rank-dimension-inequality}, $\operatorname{rank}(v) \leq \operatorname{rrank}(v) = 1$.
+Since $\Gamma_v$ is not trivial, $\operatorname{rank}(v) = 1=\operatorname{rrank}(v)$.
+\item Let $v$ be a valuation of rational rank $2$. Due to Lemma \ref{rank-dimension-inequality}, the dimension inequality is an equality. \qedhere
+\end{enumerate}
+\end{proof}
+
 \begin{lemma}\label{2.2}
@@
 Since two different valuations of rank $1$ are independent, at least one of $w_1$ and $w_2$ is of rank $2$.
-Now, let's assume that $w_1$ is of rank $1$ and $w_2$ is of rank $2$.
+Now, assume that $w_1$ is, say, of rank $1$ and $w_2$ is of rank $2$.
 Since they are dependent, it follows $v(w_2) = w_1$ and $w_1$ has a non-trivial specialization, a contradiction to $w_1\in |\operatorname{Spv}(K)|$.

### 2017-05-10 Master Thesis/synopsis.tex
Add abstracts

@@
 \begin{proof}
-(i) and (ii) follow from [HK], Lemma 2.4.3 and (iii) is a consequence of (ii) and [EGA IV] 18.6.14.
+(i) and (ii) follow from [HK], Lemma 2.4.3 and (iii) is a consequence of (ii) and the fact that henselizations commute with filtrated direct limits, see \cite{EGA4} 18.6.14.
 \end{proof}

### 2017-05-10 Master Thesis/thesis.tex
Add abstracts

@@
 \begin{abstract}
-Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Curabitur interdum, dui non molestie tincidunt, lectus enim aliquet arcu, suscipit pulvinar dui tortor vitae mi.
-Nam cursus mi sem, ac gravida tellus congue vitae.
-Nam elementum eleifend lorem dignissim rhoncus.
-Quisque ac nulla dui.
-Duis rutrum eros ornare faucibus congue.
+In this thesis we will follow M.
+Spiess' argument and present his purely arithmetic proof of the fact that isomorphisms between the absolute Galois groups of function fields of transcendence degree $1$ over number fields are induced by field isomorphisms between these fields.
 \end{abstract}
@@
 \begin{abstract}
-Integer congue euismod justo, quis venenatis tortor.
-Proin sem leo, accumsan eget pulvinar vitae, tincidunt ut elit.
-Morbi ut purus volutpat, efficitur quam id, consequat augue.
-Cras ullamcorper lacus eget massa vehicula, a elementum diam lacinia.
-Morbi efficitur elementum malesuada.
-Cras a aliquam tortor, et interdum neque.
+In dieser Arbeit wollen wir M.
+Spiess' rein arithmetischer Beweis nachgehen und zeigen, wieso Isomorphismen zwischen den absoluten Galoisgruppen von Funktionenkörpern vom Transzendenzgrad $1$ über Zahlkörper schon von Körperisomorphismen zwischen diesen Körpern induziert werden.
 \end{abstract}

### 2017-05-10 Master Thesis/synopsis.tex
Fix typos

@@
 For finite $A$ we have $nA = 0$ for $n = \#A$.
-Since $\mathbb{Z}$ has no nilpotent elements, the intersection of all prime ideals of $\mathbb{Z}$ is already trivial.
+Since $\mathbb{Z}$ has no nontrivial nilpotent elements, the intersection of all prime ideals of $\mathbb{Z}$ is already trivial.
 \qedhere
@@
 \begin{proof}
-See \cite{ZS60} VI §10 Theorem 18
+See \cite{ZS60} VI §10 Theorem 18.
 \end{proof}
@@
 Let $K,K'$ be two finitely generated fields of characteristic $0$ and dimension $2$.
-Suppose there is an isomorphism $\sigma:G_K\to G_{K'}$ between their absolute Galois groups, then there exists a unique field isomorphism $\varphi:\overline{K'}\to\overline{K}$ with $\sigma(g)=\varphi^{-1}g\varphi$ for all $g\in G_K$.
+Suppose there is an isomorphism $\sigma:G_K\to G_{K'}$ between their absolute Galois groups, then there exists a unique field isomorphism $\varphi:\overline{K'}\to\overline{K}$ such that:
+\[\sigma(g)=\varphi^{-1}g\varphi\quad \text{for all }g\in G_K\] 
 In particular $\varphi$ induces a field isomorphism $\varphi:K'\to K$.

### 2017-05-10 Master Thesis/synopsis.tex
Add some chapter comments

@@
 \section{The Valuation Spectra}
+
+This section we recall facts from valuation theory.
 
@@
 In other words, $v$ is a closed point iff $\dim k(v)=0$.
-By the dimension inequality \ref{rank-dimension-inequality} we have $\operatorname{Spv}^{(n)}(K)\subset |\operatorname{Spv}(K)|$ where $n=\dim(K)$.
-
-If $w\in\operatorname{Spv}_\text{rd}^{(i)}(K)$ for some $i\geq 1$ then there exists a unique $v(w)\in\operatorname{Spv}_\text{d}^{(i-1)}(K)$ such that $\mathcal{O}_w\subset\mathcal{O}_{v(w)}$.
+
+If $w\in\operatorname{Spv}_\text{rd}^{(i)}(K)$ for some $i\geq 1$ then there exists a unique $v(w)\in\operatorname{Spv}_\text{rd}^{(i-1)}(K)$ such that $\mathcal{O}_w\subset\mathcal{O}_{v(w)}$.
 $w$ is henselian if and only if $v(w)$ and $w\mod v(w)$ are henselian.
@@
 \item $\operatorname{Spv}_\text{d}^{(1)}(K) = \operatorname{Spv}_\text{rd}^{(1)}(K)$
-\item $\operatorname{Spv}^{(2)}(K) = \operatorname{Spv}_\text{d}^{(2)}(K)$
+\item $\operatorname{Spv}^{(2)}(K) = \operatorname{Spv}_\text{d}^{(2)}(K)\subset |\operatorname{Spv}(K)|$
 \end{enumerate}
@@
 Since $\Gamma_v$ is not trivial, $\operatorname{rank}(v) = 1=\operatorname{rrank}(v)$.
-\item Let $v$ be a valuation of rational rank $2$. Due to Lemma \ref{rank-dimension-inequality}, the dimension inequality is an equality. \qedhere
+\item Let $v$ be a valuation of rational rank $2$. Due to Lemma \ref{rank-dimension-inequality}, the dimension inequality is an equality and we have $\dim k(v) = 0$. \qedhere
 \end{enumerate}
@@
 \end{proof}
+
+The following lemma is the key result in proving Theorem \ref{thm:main-result}.
+First, we use K.
+Kato's two-dimensional local-global principle to show the assertion for finitely generated $K$.
+Then, we'll use a quasi-compactness argument to see that we can drop the assumption that $K$ is finitely generated.
 
@@
 
+In order to generalize Lemma \ref{2.4-finitelygenerated} for arbitrary, not necessarily finitely generated $K$, we need the following Lemma:
+
 \begin{lemma}\label{2.4-closedset}
@@
 \end{proof}
+
+Now the general case follows by a quasi-compactness argument:
 
@@
 \end{proof}
+
+Now we calculate the cohomology groups $\mathrm{H}^3(K_w,\mathbb{Z}/\ell\mathbb{Z})$ for $w\in |\operatorname{Spv}(K)|$.
+With the help of Lemma \ref{2.4} we can find a cohomological criterion of henselian fields over $K$ with respect to a $w\in\operatorname{Spv}_\text{d}^{(2)}(K)$.
 
@@
 \[ 1\longrightarrow T\longrightarrow G_K\longrightarrow G_k\longrightarrow 1 \]
+
 Let $G_\ell$ be an $\ell$-Sylow group of $G_K$.
@@
 Therefore, $\operatorname{inf}$ and $\operatorname{res}$ are both isomorphisms and it suffices to calculate $\mathrm{H}^3(G_\ell,\mathbb{Z}/\ell\mathbb{Z})$.
-By [Se], Ch.
+By \cite{Se64}, Ch.
 I, §4, Proposition 22 we have:
@@
 Now let $G_k(\ell) \cong\mathbb{Z}_\ell$ and $r = 2$.
-According to the proof of [Se], Ch.
+According to the proof of \cite{Se64}, Ch.
 I, §4, Proposition 22, we have:
 \[ \mathrm{H}^3(G_\ell,\mathbb{Z}/\ell\mathbb{Z}) = \mathrm{H}^1(G_k(\ell), \mathrm{H}^2 (\mathbb{Z}_\ell^r, \mathbb{Z}/\ell\mathbb{Z})) \]
-By [NSW] Theorem 2.4.6, we have $\mathrm{H}^2(\mathbb{Z}_\ell^r,\mathbb{Z}/\ell\mathbb{Z}) \cong\mathrm{H}^1(\mathbb{Z}_\ell, \mathrm{H}^1(\mathbb{Z}_\ell, \mathbb{Z}/\ell\mathbb{Z})) = \mathbb{Z}/\ell\mathbb{Z}$, since $\mathrm{H}^1(\mathbb{Z}_\ell, \mathbb{Z}/\ell\mathbb{Z})=\operatorname{Hom}_\text{cont}(\mathbb{Z}_\ell, \mathbb{Z}/\ell\mathbb{Z})=\mathbb{Z}/\ell\mathbb{Z}$.
+By [NSW] Theorem 2.4.6, we have $\mathrm{H}^2(\mathbb{Z}_\ell^r,\mathbb{Z}/\ell\mathbb{Z}) \cong\mathrm{H}^1(\mathbb{Z}_\ell, \mathrm{H}^1(\mathbb{Z}_\ell, \mathbb{Z}/\ell\mathbb{Z})) = \mathbb{Z}/\ell\mathbb{Z}$, since $\mathrm{H}^1(\mathbb{Z}_\ell, \mathbb{Z}/\ell\mathbb{Z})=\operatorname{Hom}_\text{cont.}(\mathbb{Z}_\ell, \mathbb{Z}/\ell\mathbb{Z})=\mathbb{Z}/\ell\mathbb{Z}$.
 Therefore, $\mathrm{H}^3(G_\ell,\mathbb{Z}/\ell\mathbb{Z})=\mathbb{Z}/\ell\mathbb{Z}$.
@@
 
+In this section we want to see that $\sigma$ induces bijections on defectless valuations.
+A direct consequence of Proposition \ref{2.6} is the bijection $\operatorname{Spv}_\text{d}^{(2)}(K)\to \operatorname{Spv}_\text{d}^{(2)}(K')$:
+
 \begin{lemma}\label{3.1}
@@
 \end{proof}
+
+Using the bijection $\operatorname{Spv}_\text{d}^{(2)}(K)\to \operatorname{Spv}_\text{d}^{(2)}(K')$ above we want to establish a bijection $\operatorname{Spv}_\text{rd}^{(2)}(K)\to \operatorname{Spv}_\text{rd}^{(2)}(K')$ in order to establish a bijection between the defectless rank-$1$ valuations:
 
@@
 
+In this section we shall show that $\sigma: G_K\to G_{K'}$ induces group isomorphisms $K'^\times/n\to K^\times/n$ for all $n\in\mathbb{N}$.
+
 Let $k$ and $k'$ be the algebraic closures of $\mathbb{Q}$ in $K$ and $K'$ respectively, i.e.
@@
 \end{proof}
+
+First of all, we want to show that $\sigma$ induces an isomorphism between the absolute Galois groups of the constant fields $G_k$ and $G_{k'}$. We need the following Lemma:
 
@@
 
-\paragraph{} Let $n\in\mathbb{N}$. In the previous remark, we have replaced $\sigma$, so that now the following diagram commutes: 
+\paragraph{} Let $n\in\mathbb{N}$. In the previous remark, we have replaced $\sigma$, so that the following diagram commutes now:
 \[ \begin{tikzcd}

### 2017-05-20 Master Thesis/synopsis.tex
Add more chapter comments

@@
 
-This section we recall facts from valuation theory.
+In this section we shall recall facts from valuation theory.
 
@@
 
-Using the bijection $\operatorname{Spv}_\text{d}^{(2)}(K)\to \operatorname{Spv}_\text{d}^{(2)}(K')$ above we want to establish a bijection $\operatorname{Spv}_\text{rd}^{(2)}(K)\to \operatorname{Spv}_\text{rd}^{(2)}(K')$ in order to establish a bijection between the defectless rank-$1$ valuations:
+We will see that the bijection $\operatorname{Spv}_\text{d}^{(2)}(K)\to \operatorname{Spv}_\text{d}^{(2)}(K')$ above restricts to a bijection $\operatorname{Spv}_\text{rd}^{(2)}(K)\to \operatorname{Spv}_\text{rd}^{(2)}(K')$.
+Using this, we shall establish a bijection between the defectless valuations of rational rank $1$:
 
@@
 
-In this section we shall show that $\sigma: G_K\to G_{K'}$ induces group isomorphisms $K'^\times/n\to K^\times/n$ for all $n\in\mathbb{N}$.
+In this section we shall show that $\sigma: G_K\to G_{K'}$ induces group isomorphisms $K'^\times/n\to K^\times/n$ and $U_{v'}/n\to U_v/n$ for all $n\in\mathbb{N}$ and defectless valuations $v$ of rational rank $1$ with $\operatorname{char}k(v) =0$.
 
@@
 \[ \begin{tikzcd}
-G_K \ar[r, "\sigma"] & G_{K'} \ar[r] & G_{K'}
-\end{tikzcd} \]
-where the second map is induced by the isomorphism $\overline{K''}\to\overline{K'},\ x\otimes y\mapsto \varphi^{-1}(y)\cdot x$, we may assume that $k=k'$, $\overline{k} = \overline{k'}$ and that $\varphi:\overline{k}\to\overline{k}$, $\sigma: G_k\to G_k$ are identities.
-\end{remark}
-
-\paragraph{} Let $n\in\mathbb{N}$. In the previous remark, we have replaced $\sigma$, so that the following diagram commutes now:
+G_K \ar[r, "\sigma"] & G_{K'} \ar[r, "\phi"] & G_{K'}
+\end{tikzcd} \]
+where $\phi$ is induced by the isomorphism $\overline{K''}\to\overline{K'},\ x\otimes y\mapsto \varphi^{-1}(y)\cdot x$, we may assume that $k=k'$, $\overline{k} = \overline{k'}$ and that $\varphi:\overline{k}\to\overline{k}$, $\sigma: G_k\to G_k$ are identities.
+Now the following diagram commutes:
 \[ \begin{tikzcd}
@@
 \end{tikzcd} \]
+\end{remark}
+
+\paragraph{} Let $n\in\mathbb{N}$.
 The operation of $G_K$ and $G_{K'}$ on $\mu_n$ factors through $G_k$, thus we have $\sigma(g)\xi = g\xi$ for all $\xi\in\mu_n$ and $g\in G_K$.
@@
 \item Consider the absolute inertia field $L = K_P^\text{t}$ of $K_P$ and let $w$ be the unique extension of $v_P$ in $L$.
-By Theorem \ref{theorem-absoluteinertia} (i), $w$ is discrete as well with $k(w) = k(v_P)^\text{s}$.
-We have the following exact sequence:
+By Theorem \ref{theorem-absoluteinertia} (i), $w$ is discrete as well and we have $k(w) = k(v_P)^\text{s}$.
+Observe the following exact sequence:
 \[ 1 \longrightarrow U_w/n \longrightarrow L^\times /n\longrightarrow \mathbb{Z}/n\mathbb{Z}\longrightarrow 0 \]
@@
 
+Before we start proving Pop's Theorem \ref{thm:main-result}, we need some tools that will help us constructing a group isomorphism $K'^\times\to K^\times$ from the modulo $n$ isomorphisms $K'^\times/n\to K^\times/n$.
+
 \section{Free-By-Finite Groups}
@@
 \begin{definition}\label{5.1}
-An abelian group $A$ is called \textit{free-by-finite} if its torsion subgroup $A_\text{tor}$ is finite and the quotient $A/A_\text{tor}$ is free abelian.
-$A$ is free-by-finite if and only if $A$ is the direct sum of a finite abelian group and a free abelian group.
+Let $A$ be an abelian group. We have the following short exact sequence:
+\[0\longrightarrow A_\text{tor} \longrightarrow A\longrightarrow A/A_\text{tor}\longrightarrow 0 \]
+$A$ is called \textit{free-by-finite} if its torsion subgroup $A_\text{tor}$ is finite and the quotient $A/A_\text{tor}$ is free abelian.
+If $A$ is free-by-finite, the above sequence splits and $A$ is the direct sum of a finite abelian group and a free abelian group.
+It's evident that the converse also holds.
 \end{definition}
@@
 Considering $A/K$ as a subgroup of $B$, we see that $A/K$ is free-by-finite.
-Since $(A/K)_\text{tor}$ and $K$ are finite, $K' = \{a\in A\mid na\in K \text{ for some }n\in\mathbb{N}\}$ is finite as well, in particular $A_\text{tor}$ is finite.
+Since $(A/K)_\text{tor}$ and $K$ are finite, $K' = \{a\in A\mid na\in K \text{ for some }n\in\mathbb{N}\}$ is finite as well, in particular $A_\text{tor}\subset K'$ is finite.
 In fact $K'=A_\text{tor}$, since $K\subset A_\text{tor}$.
@@
 Now let $A$ be free-by-finite, i.e.
-$A = A_\text{tor} \oplus F$ for some free abelian group $F$, and $K=\ker(f)$.
-Since $K\subset A$ is finite, $K$ is actually contained in $A_\text{tor}$.
-Therefore, we see that $A/K$ is free by finite as well, and by extension, the image $I$ of $f$.
+$A = A_\text{tor} \oplus F$ for some free abelian group $F$, and $K=\ker(f)\subset A_\text{tor}$.
+We see that $A/K$ is free by finite as well, and by extension, the image $I$ of $f$.
 Set $m = \#B/I$ and consider the canonical map $B\stackrel{\cdot m}{\to} I \to I/I_\text{tor}$.
@@
 Now both $I_\text{tor}$ and $B_\text{tor}/I_\text{tor} \subset B/I$ are finite, hence $B_\text{tor}$ is finite as well.
-\item The exact sequence induces the exact sequence $0 \longrightarrow A'_\text{tor} \longrightarrow A_\text{tor} \longrightarrow A''_\text{tor}$.
-Since the surrounding groups are finite, $A_\text{tor}$ is finite as well.
-Using (i) we see that it suffices to show that $A/A_\text{tor}$ is free-by-finite.
+\item The exact sequence induces the exact sequence:
+\[0 \longrightarrow A'_\text{tor} \longrightarrow A_\text{tor} \longrightarrow A''_\text{tor}\] 
+Since the surrounding groups are finite, $A_\text{tor}$ is finite as well. Using (i) we see that it suffices to show that $A/A_\text{tor}$ is free-by-finite.
 
@@
 \item $F^\times$ is free-by-finite.
-\item Let $F_1,\ldots,F_n$ be finite extensions of $F$.
-Then $\big(\prod_{i=1}^{n}F_i^\times\big)/F^\times$ is free-by-finite, where $F^\times$ is considered as a subgroup of $\prod_{i=1}^{n}F_i^\times$ via the diagonal embedding.
+\item Let $F_1,\ldots,F_n$ be finite extensions of $F$. Then 
+\[\Big(\prod_{i=1}^{n}F_i^\times\Big)/F^\times\] 
+is free-by-finite, where $F^\times$ is considered as a subgroup of $\prod_{i=1}^{n}F_i^\times$ via the diagonal embedding.
 \end{enumerate}
@@
 
-\section{Proof of Theorem}
+\section{Proof of Pop's Theorem}
 

### 2017-05-20 Master Thesis/synopsis.tex
Use - instead of \setminus

@@
 Valuations in $\operatorname{Spv}_\text{d}(K)$ are called \textit{defectless}\index{valuation!defectless}, whereas those in $\operatorname{Spv}_\text{rd}(K)$ are called \textit{rank-defectless}\index{valuation!rank-defectless}.
-For $\ast\in\{\text{d},\text{rd}\}$ we set:
-\[\operatorname{Spv}_\ast^{(i)}(K)=\{v\in\operatorname{Spv}_\ast(K)\mid \operatorname{rrank}(v)=i \} \]
+For $\bullet\in\{\text{d},\text{rd}\}$ we set:
+\[\operatorname{Spv}_\bullet^{(i)}(K)=\{v\in\operatorname{Spv}_\bullet(K)\mid \operatorname{rrank}(v)=i \} \]
 \end{definition}
@@
 We get an exact sequence:
-\[ 0 \longrightarrow \Big(\bigoplus_{b\in C} \mathbb{Z}b\Big)\Big/A' \longrightarrow A/A' \longrightarrow \bigoplus_{b\in B\setminus C}\mathbb{Z}b \longrightarrow 0 \]
+\[ 0 \longrightarrow \Big(\bigoplus_{b\in C} \mathbb{Z}b\Big)\Big/A' \longrightarrow A/A' \longrightarrow \bigoplus_{b\in B -  C}\mathbb{Z}b \longrightarrow 0 \]
 The left abelian group is finitely generated, particularly free-by-finite. The right group is free, hence also free-by-finite. By (ii), $A/A'\cong A''$ is free-by-finite as well.
@@
 For a finite subset $S\subset X_0$ set:
-\[ K^S=\{x\in K^\times \mid v_P(x)=0\text{ for all }P\in X_0\setminus S \} \]
+\[ K^S=\{x\in K^\times \mid v_P(x)=0\text{ for all }P\in X_0 -  S \} \]
 $K'^{S'}$ is defined similarly. For finite subsets $S$ of $X_0$, the corresponding subsets of $X_0'$ will be denoted by $S'$, i.e. $S'=\{\sigma P\mid P\in S \} $.
@@
 \item There exists a finite subset $S\subset X_0$ such that:
-\[ \operatorname{Pic}(X\setminus S)=0=\operatorname{Pic}(X'\setminus S') \]
-Such a set $S$ is called \textit{admissible}. If $T\subset X_0$ is a finite set of closed points, then we can choose $S$ in $X_0\setminus T$.
+\[ \operatorname{Pic}(X -  S)=0=\operatorname{Pic}(X' -  S') \]
+Such a set $S$ is called \textit{admissible}. If $T\subset X_0$ is a finite set of closed points, then we can choose $S$ in $X_0 -  T$.
 \item Let $S$ be admissible. For all $n\in\mathbb{N}$ the following canonical maps are injective:
@@
 Now let $T\subset X_0$ be finite.
-By the Weak Approximation Theorem \ref{weak-approximation}, we can add a principal divisor to the $D_i$ and the $D'_j$ so that their support is in $X_0\setminus T$ and $X'_0\setminus T'$ respectively.
+By the Weak Approximation Theorem \ref{weak-approximation}, we can add a principal divisor to the $D_i$ and the $D'_j$ so that their support is in $X_0 -  T$ and $X'_0 -  T'$ respectively.
 \item Observe that the following sequence is exact:
 \[ \begin{tikzcd}
-1 \ar[r] & K^S \ar[r]& K^\times \ar[r, "\operatorname{div}"]& \displaystyle \bigoplus_{P\in X_0\setminus S}\mathbb{Z} \ar[r]& \operatorname{Pic}(X\setminus S) = 0
+1 \ar[r] & K^S \ar[r]& K^\times \ar[r, "\operatorname{div}"]& \displaystyle \bigoplus_{P\in X_0 -  S}\mathbb{Z} \ar[r]& \operatorname{Pic}(X -  S) = 0
 \end{tikzcd} \]
 Since $\bigoplus\mathbb{Z}$ is $\mathbb{Z}$-flat, tensoring with $\mathbb{Z}/n\mathbb{Z}$ yields another exact sequence and we can consider $K^S/n$ as a subgroup of $K^\times/n$. Then:
-\[ K^S/n = \bigcap_{P\in X_0\setminus S}U_P/n \]
+\[ K^S/n = \bigcap_{P\in X_0 -  S}U_P/n \]
 The analogous result holds for $K'^{S'}/n$. Thus the assertion follows from Lemma \ref{4.3} (i).\qedhere
@@
 \begin{lemma}\label{5.4}
-Let $S\subset X_0$ be admissible. Then there exists a finite subset $T\subset X_0\setminus S$ such that the following map is injective:
+Let $S\subset X_0$ be admissible. Then there exists a finite subset $T\subset X_0 -  S$ such that the following map is injective:
 \[ r_{S,T}:K^S \to\prod_{P\in T}k(P)^\times,\ x\mapsto(r_P(x))_{P\in T} \]
@@
 \begin{proof}
-Let $U = X\setminus S$ and $P_1\in U$ be a closed point. We set:
+Let $U = X -  S$ and $P_1\in U$ be a closed point. We set:
 \[ C_1 = \ker(r_{P_1}|_{K^S}: K^S \to k(P_1)^\times) \]
@@
 \begin{enumerate}[(i)]
-\item There exists a point $P_2\in X_0\setminus(S\cup\{P_1\})$ which lies above the prime of $k(x)$ corresponding to the polynomial $x-a\in k[x]$.
+\item There exists a point $P_2\in X_0 - (S\cup\{P_1\})$ which lies above the prime of $k(x)$ corresponding to the polynomial $x-a\in k[x]$.
 \item $a$ is not a root of unity.
@@
 \begin{definition}
-For each admissible $S\subset X_0$ and $T_1\subset X_0\setminus S$ with injective $r_{S,T_1}$ we also have the injectivity of $r_{S,T_2}$ for every $T_2 \supset T_1$.
-Therefore,  we can find a finite subset $T\subset X_0\setminus S$ such that both $r_{S,T}$ and $r_{S',T'}$ are injective.
+For each admissible $S\subset X_0$ and $T_1\subset X_0 -  S$ with injective $r_{S,T_1}$ we also have the injectivity of $r_{S,T_2}$ for every $T_2 \supset T_1$.
+Therefore,  we can find a finite subset $T\subset X_0 -  S$ such that both $r_{S,T}$ and $r_{S',T'}$ are injective.
 Such a $T$ will be called $S$-\textit{admissible} and $(S,T)$ is called \textit{admissible pair}.
@@
 By Lemma \ref{5.4}, the following maps are injective for all admissible subsets $S\subset X_0$:
-\[ r_S:K^S \to\prod_{P\in X_0\setminus S} k(P)^\times,\ x\mapsto (r_P(x))_P \]
+\[ r_S:K^S \to\prod_{P\in X_0 -  S} k(P)^\times,\ x\mapsto (r_P(x))_P \]
 Now it's evident that $\varphi$ is unique.
@@
 
-Let $T_1, T_2\subset X_0\setminus S$ be two $S$-admissible subsets and suppose $T_1\subset T_2$. By considering $(\star)$ we find:
+Let $T_1, T_2\subset X_0 -  S$ be two $S$-admissible subsets and suppose $T_1\subset T_2$. By considering $(\star)$ we find:
 \[ r_{S,T_1}\circ\varphi^{S,T_2}=\Big(\prod_{P\in T_1}\varphi_P\Big)\circ r_{S',T_1'} = r_{S,T_1}\circ\varphi^{S,T_1} \]
@@
 Now if $T$ ranges over all $S$-admissible subsets of $X_0$, then $(\star)$ shows:
-\[ r_S\circ\varphi^S = \Big(\prod_{P\in X_0\setminus S}\varphi_P \Big)\circ r_{S'} \]
-In other words, we can replace $T$ by $X_0\setminus S$ in $(\star)$. If $S_1\subset S_2$ are two admissible subsets, then:
-\[ r_{S_2}\circ\varphi^{S_2}|_{K'^{S'_1}} = \Big(\prod_{P\in X_0\setminus S_2}\varphi_P \Big)\circ r_{S'_2}|_{K'^{S'_1}} = r_{S_2}\circ\varphi^{S_1} \]
+\[ r_S\circ\varphi^S = \Big(\prod_{P\in X_0 -  S}\varphi_P \Big)\circ r_{S'} \]
+In other words, we can replace $T$ by $X_0 -  S$ in $(\star)$. If $S_1\subset S_2$ are two admissible subsets, then:
+\[ r_{S_2}\circ\varphi^{S_2}|_{K'^{S'_1}} = \Big(\prod_{P\in X_0 -  S_2}\varphi_P \Big)\circ r_{S'_2}|_{K'^{S'_1}} = r_{S_2}\circ\varphi^{S_1} \]
 Hence, $\varphi^{S_1}=\varphi^{S_2}|_{K'^{S'_1}}$.
@@
 
-The commutativity of the diagram follows from the commutativity of $(\star)$ with $T=X_0\setminus S$.
-Note that with Lemma \ref{4.5} (i), we can find for all $x\in U_{P'}$ an admissible set $S\subset X_0\setminus \{P\}$ such that $x\in K'^{S'}$.
+The commutativity of the diagram follows from the commutativity of $(\star)$ with $T=X_0 -  S$.
+Note that with Lemma \ref{4.5} (i), we can find for all $x\in U_{P'}$ an admissible set $S\subset X_0 -  \{P\}$ such that $x\in K'^{S'}$.
 \end{proof}
@@
 \end{align*}
-for all $P\in X_0\setminus S$.
+for all $P\in X_0 -  S$.
 Note that $\varphi_P: k(P')\to k(P)$ is a field isomorphism and that the residue maps $r_{P'}: \mathcal{O}_{P'} \to k(P')$ are ring homomorphisms.
-Since $K^S\to \prod_{P\in X_0\setminus S} k(P)^\times$ is injective, we get $\varphi(x+y) = \varphi(x) + \varphi(y)$.\qedhere
+Since $K^S\to \prod_{P\in X_0 -  S} k(P)^\times$ is injective, we get $\varphi(x+y) = \varphi(x) + \varphi(y)$.\qedhere
 \end{itemize}

### 2017-05-20 Master Thesis/synopsis.tex
Use \widetilde instead of \tilde

@@
 where $w$ ranges over all rank-$2$ valuations with $v(w) = v$, can be identified with the following map:
-\[ {}_\ell\operatorname{Br}(k) \longrightarrow \prod_{\tilde{w}\in\operatorname{Spv}^{(1)}(k)} {}_\ell\operatorname{Br}(k_{\tilde{w}}) \]
-where $k = k(v)$. This map is injective by the classical Hasse principle and its image is contained in $\bigoplus {}_\ell\operatorname{Br}(k_{\tilde{w}})$. Combining these maps yields the injection:
+\[ {}_\ell\operatorname{Br}(k) \longrightarrow \prod_{\widetilde{w}\in\operatorname{Spv}^{(1)}(k)} {}_\ell\operatorname{Br}(k_{\widetilde{w}}) \]
+where $k = k(v)$.
+This map is injective by the classical Hasse principle and its image is contained in $\bigoplus {}_\ell\operatorname{Br}(k_{\widetilde{w}})$.
+Combining these maps yields the injection:
 \[\Phi: \mathrm{H}^3(K)\longrightarrow \bigoplus_{x\in\mathfrak{X}_1}\  \bigoplus_{\substack{w\in\operatorname{Spv}_\text{rd}^{(2)}(K) \\ v(w) = v_x}} \mathrm{H}^3(K_w) \]
@@
 
-First of all, we want to show that $\sigma$ induces an isomorphism between the absolute Galois groups of the constant fields $G_k$ and $G_{k'}$. We need the following Lemma:
+We start by showing that $\sigma$ induces an isomorphism between the absolute Galois groups of the constant fields $G_k$ and $G_{k'}$. We need the following Lemma:
 

### 2017-05-20 Master Seminar/article.tex
Add master seminar tex file

--- 2017-05-20 Master Thesis/synopsis.tex
@@
-\chapter{Valuations}
+\documentclass[10pt,b5paper]{article}
+\usepackage[left=2.00cm, right=2.00cm, top=2.00cm, bottom=2.00cm]{geometry}
+\usepackage{fontspec}
+\usepackage[ngerman]{babel}
+\usepackage{amsmath}
+\usepackage{amsfonts}
+\usepackage{amssymb}
+\usepackage{amsthm}
+\usepackage{paralist}
+\usepackage{tikz}
+\usetikzlibrary{cd, babel}
+\usepackage{hyperref}
+\hypersetup{
+	bookmarksopen=true,
+	pdfstartview=FitH
+}
+\linespread{1.1}
 
-\section{The Valuation Spectra}
+\author{von \textsc{yichuan shen}}
+\title{Galoissche Kennzeichnung von Funktionenkörper \\ vom Transzendenzgrad $1$ über Zahlkörper}
+\begin{document}
 
-In this section we shall recall facts from valuation theory.
+\theoremstyle{plain}
+\theoremstyle{definition}
+\newtheorem{theorem}{Theorem}
+\newtheorem{lemma}[theorem]{Lemma}
+\newtheorem{proposition}[theorem]{Satz}
+\newtheorem{corollary}[theorem]{Korollar}
+\theoremstyle{definition}
+\newtheorem*{definition}{Definition}
+\newtheorem*{example}{Beispiel}
+\theoremstyle{remark}
+\newtheorem*{remark}{Bemerkung}
+\newtheorem{step}{Schritt}
+
+\maketitle
+
+\section{Valuations}
 
@@
 
-\begin{remark}
-\begin{itemize}
-\item A one-dimensional finitely generated field is just a global field. 
-\item A two-dimensional finitely generated field over $\mathbb{Q}$ is a function field in one variable over $\mathbb{Q}$.
-\end{itemize}
-\end{remark}
-
 \begin{definition}\label{2.1}
@@
 
-We denote the valuation ring of $v\in\operatorname{Spv}(K)$ by $(\mathcal{O}_v,\mathfrak{m}_v)$ and set $U_v=\mathcal{O}_v^\times$ and let $k(v)=\mathcal{O}_v/\mathfrak{m}_v$ be its residue field.
-The \textit{rank} of $v$ is defined as the Krull dimension of its valuation ring $\mathcal{O}_v$.
-The \textit{rational rank}\index{valuation!rational rank} of $v:K\to\Gamma_v\cup\{\infty\}$ is defined as the rank of its value group $\Gamma_v$ as a $\mathbb{Z}$-module:
-\[\operatorname{rrank}(v) = \dim_\mathbb{Q}(\Gamma_v\otimes_\mathbb{Z}\mathbb{Q}) \]
+We can endow $\operatorname{Spv}(K)$ with a natural topology by declaring following sets as a basis of open sets:
+\[\{v\in\operatorname{Spv}(K)\mid v(x)\geq 0\text{ for all }x\in S\} \]
+where $S$ ranges over all finite subsets of $K$. Let $v,w\in\operatorname{Spv}(K)$. We have:
+\[\mathcal{O}_w\subset\mathcal{O}_v \iff w\in\overline{\{v\}} \]
+Thus $v$ is a closed point if and only if it has no non-trivial specializations.
 \end{definition}
@@
 \end{lemma}
-
-\begin{proof}
-See \cite{EP05} 3.4.1 and 3.4.3.
-\end{proof}
 
@@
 
-\begin{definition}
-We can endow $\operatorname{Spv}(K)$ with a natural topology by declaring following sets as a basis of open sets:
-\[\{v\in\operatorname{Spv}(K)\mid v(x)\geq 0\text{ for all }x\in S\} \]
-where $S$ ranges over all finite subsets of $K$. 
-\end{definition}
+\section{Main Lemma}
 
-\begin{lemma}\label{spv-topology}
-$\operatorname{Spv}(K)$ with the topology defined above is a quasi-compact $T_0$-space.
-In particular the set of closed points $|\operatorname{Spv}(K)|$ of $\operatorname{Spv}(K)$ is a quasi-compact $T_1$-space, endowed with the induced topology.
-Let $v,w\in\operatorname{Spv}(K)$.
-We have:
-\[\mathcal{O}_w\subset\mathcal{O}_v \iff w\in\overline{\{v\}} \]
-Thus $v$ is a closed point if and only if it has no non-trivial specializations.
-\end{lemma}
-
-\begin{proof}
-See \cite{ZS60} VI §17 Theorem 38, §17 Theorem 40.
-\end{proof}
-
-\begin{lemma}\label{2.2-pre}
-For a fixed $v\in \operatorname{Spv}(K)$ there is a canonical bijection: 
-\begin{align*}
-\{ w\in\operatorname{Spv}(K)\mid \mathcal{O}_w\subset\mathcal{O}_v \} &\longrightarrow \operatorname{Spv}k(v)\\ 
-w&\longmapsto w\mod v
-\end{align*}
-Therefore $v$ is a closed point iff $k(v)$ only admits the trivial valuation, i.e.
-$k(v)$ has characteristic $p>0$ and is algebraic over $\mathbb{F}_p$.
-In other words, $v$ is a closed point iff $\dim k(v)=0$.
-
-If $w\in\operatorname{Spv}_\text{rd}^{(i)}(K)$ for some $i\geq 1$ then there exists a unique $v(w)\in\operatorname{Spv}_\text{rd}^{(i-1)}(K)$ such that $\mathcal{O}_w\subset\mathcal{O}_{v(w)}$.
-$w$ is henselian if and only if $v(w)$ and $w\mod v(w)$ are henselian.
-\end{lemma}
-
-\begin{proof}
-See \cite{EP05} 2.3, 4.1.4, \cite{ZS60} VI §3 Theorem 3.
-\end{proof}
-
-\begin{theorem}\label{theorem-henselization}
-Let $K$ be a field and $v\in\operatorname{Spv}(K)$. Then there is a separable extension $K_v/K$ with a henselian extension $\overline{v}$ of $v$ with the following characterization:
-
-\textit{If $K'/K$ is henselian with respect to an extension $w$ of $v$, then there exists a uniquely determined $K$-embedding $K_v\to K'$, so that $w$ extends $\overline{v}$.}
-
-$K_v$ is called a \textit{henselization} of $K$ with respect to $v$.
-Furthermore, $\overline{v}$ has the same value group as $v$ and the same residue field $k(\overline{v}) = k(v)$.
-$G_{K_v}$ is the decomposition group of the unique extension of $\overline{v}$ to $K^\text{s}$:
-\[ G_{\overline{v}} = \{ \sigma\in G_K\mid \sigma(\mathcal{O}_{\overline{v}}) =\mathcal{O}_{\overline{v}} \} \subset G_K \]
-Conversely, taking the fixed field of the decomposition group of an extension $\overline{v}'$ of $v$ in $K^\text{s}$ yields a henselization of $K$ with respect to $v$.
-Any two of such henselizations are $K$-conjugate.
-\end{theorem}
-
-\begin{proof}
-See \cite{EP05} Lemma 5.2.1, Lemma 5.2.2, Theorem 5.2.5.
-\end{proof}
-
-\begin{theorem}\label{theorem-absoluteinertia}
-Let $K$ be a henselian field with respect to $v: K\to\Gamma\cup\{\infty\}$ with unique extension $\overline{v}$ on $K^\text{s}$.
-Let $K^\text{t}$ be the absolute inertia field of $K$ with the unique extension $w$ of $v$, i.e.
-$K^\text{t}$ is the fixed field of the absolute inertia group $T$:
-\[ T = \{ \sigma\in G_K\mid \sigma(x)\equiv x\mod \mathfrak{m}_{\overline{v}} \quad \text{for all }x\in\mathcal{O}_{\overline{v}} \}\lhd G_K \]
-Then we have:
-\begin{enumerate}[(i)]
-\item $w$ has value group $\Gamma$ and residue class field $k(v)^\text{s}$.
-\item $K^\text{t}$ is a Galois extension of $K$ and $\operatorname{Gal}(K^\text{t}/K)\cong G_{k(v)}$ as profinite groups. There exists an inclusion-preserving one-to-one correspondence: 
-\begin{align*}
-\{ \text{Intermediate extensions in $K^\text{t}/K$} \} &\longrightarrow \{\text{Separable extensions of $k(v)$} \}\\ 
-L &\longmapsto k(v_L)  
-\end{align*}
-where $v_L$ is the unique extension of $v$ to $L$.
-\end{enumerate}
-\end{theorem}
-
-\begin{proof}
-See \cite{EP05} Theorem 5.2.7.
-\end{proof}
-
-\begin{lemma}\label{lemma-henselization-specialization}
-Let $K$ be a field and $w\in\operatorname{Spv}(K)$ with a fixed henselization $K_w$ of $w$.
-If $v\in\operatorname{Spv}(K)$ such that $w$ is a specialization of $v$, then there is a henselization $K_v$ of $v$ such that $K_v\subset K_w\subset K_v^\text{t}$.
-\end{lemma}
-
-\begin{proof}
-Let $\overline{w}$ be the extension of $w\in\operatorname{Spv}(K_w)$ on $K^\text{s}$.
-By \cite{EP05} Lemma 3.1.5 we can find an extension $\overline{v}$ of $v$ on $K^\text{s}$ such that $\overline{w}$ is a specialization of $\overline{v}$.
-Let $K_v$ be the fixed field of $G_{\overline{v}}\subset G_K$.
-
-Let $\sigma\in G_{\overline{w}}$.
-Then both $\mathcal{O}_{\overline{v}}$ and $\sigma(\mathcal{O}_{\overline{v}})$ are overrings of $\mathcal{O}_{\overline{w}}$.
-Now the overrings of valuation rings are totally ordered by inclusion and correspond to localizations of $\mathcal{O}_{\overline{w}}$ at prime ideals (See \cite{EP05} 2.3).
-Since $\mathcal{O}_{\overline{v}}$ and $\sigma(\mathcal{O}_{\overline{v}})$ have the same Krull dimension, they must be the same, thus $\sigma\in G_{\overline{v}}$.
-This proves $K_v\subset K_w$.
-
-Now let $\sigma\in G_{K_v^\text{t}}$, i.e.
-$\sigma\in G_{\overline{v}}$ such that the induced automorphism on the residue field $\overline{\sigma}:k(\overline{v})\to k(\overline{v})$ is the identity.
-Let $\mathfrak{p}\subset\mathcal{O}_{\overline{w}}$ be a prime ideal with $\mathcal{O}_{\overline{v}} = (\mathcal{O}_{\overline{w}})_\mathfrak{p}$.
-In particular, the restriction $\overline{\sigma}|: \mathcal{O}_{\overline{w}}/\mathfrak{p}\to \mathcal{O}_{\overline{w}}/\mathfrak{p}$ is the identity, thus $\sigma(\mathcal{O}_{\overline{w}})=\mathcal{O}_{\overline{w}}$, i.e.
-$\sigma\in G_{\overline{w}}$.
-This proves $K_w\subset K_v^\text{t}$.
-\end{proof}
-
-\begin{theorem}\label{theorem-sylowgroups-inertiagroup}
-Let $K$ be a henselian field with respect to $v:K\to\Gamma\cup\{\infty\}$ with unique extension $\overline{v}$ on $K^\text{s}$, and $p=\operatorname{char}k(v)$. Then the absolute ramification group
-\[ V = \Big\{ \sigma\in G_K\mathop{\Big|} \frac{\sigma(x)}{x}-1\in\mathfrak{m}_{\overline{v}}\quad \text{for all }x\in {K^\text{s}}^\times \Big\} \lhd T \]
-is the unique $p$-Sylow group of the absolute inertia group $T$. $V$ is trivial if $p=0$. We have the following isomorphism of profinite groups:
-\[ T/V \cong \prod_{\ell\neq p\text{ prime}}\mathbb{Z}_\ell^{r_\ell} \]
-where $r_\ell = \dim_{\mathbb{F}_\ell}(\Gamma/\ell\Gamma)$.
-\end{theorem}
-
-\begin{proof}
-See \cite{EP05} Theorem 5.3.3.
-\end{proof}
-
-\begin{theorem}[\textit{F.K. Schmidt}]\label{theorem-fkschmidt}
-Let $v_1,v_2\in\operatorname{Spv}(K)$ be two independent henselian valuations on a field $K$, i.e. $\mathcal{O}_{v_1}\mathcal{O}_{v_2}=K$. Then $K$ is separably closed.
-\end{theorem}
-
-\begin{proof}
-See \cite{EP05} Theorem 4.4.1.
-\end{proof}
-
-\begin{remark}
-Two valuations of rank $1$ are independent iff they are inequivalent.
-\end{remark}
-
-\begin{corollary}\label{corollary-fkschmidt}
-Let $w_1,w_2\in\operatorname{Spv}(K^\text{s})$ be two inequivalent valuations of rank $1$ with decomposition groups $G_{{w_1}},G_{{w_2}}\subset G_K$. Then $G_{{w_1}}\cap G_{{w_2}}=1$.
-\end{corollary}
-
-\begin{proof}
-Let $L$ be the fixed field of $G_{w_1}\cap G_{w_2}\subset G_K$ and set $v_i = w_i|_L$.
-Then $L$ is henselian with respect to both $v_1$ and $v_2$.
-Since $w_1,w_2$ are the only extensions of $v_1$ and $v_2$ in $K^\text{s}$ respectively, $v_1$ and $v_2$ are inequivalent valuations of rank $1$.
-Now $L$ is separably closed by Theorem \ref{theorem-fkschmidt}.
-Therefore, by Theorem \ref{theorem-henselization}, we get $G_{{w_1}}\cap G_{{w_1}}=G_L =1$.
-\end{proof}
-
-\section{The Main Lemma}
-
-Let $K$ be a (not necessarily finitely generated) field over $\mathbb{Q}$ and of dimension $2$.
-
-\begin{lemma}\phantomsection\label{lemma-defectless-valuations}
-\begin{enumerate}[(i)]
-\item $\operatorname{Spv}_\text{d}^{(1)}(K) = \operatorname{Spv}_\text{rd}^{(1)}(K)$
-\item $\operatorname{Spv}^{(2)}(K) = \operatorname{Spv}_\text{d}^{(2)}(K)\subset |\operatorname{Spv}(K)|$
-\end{enumerate}
-\end{lemma}
-
-\begin{proof}
-\begin{enumerate}[(i)]
-\item Let $v$ be a defectless valuation of rational rank $1$.
-By Lemma \ref{rank-dimension-inequality}, $\operatorname{rank}(v) \leq \operatorname{rrank}(v) = 1$.
-Since $\Gamma_v$ is not trivial, $\operatorname{rank}(v) = 1=\operatorname{rrank}(v)$.
-\item Let $v$ be a valuation of rational rank $2$. Due to Lemma \ref{rank-dimension-inequality}, the dimension inequality is an equality and we have $\dim k(v) = 0$. \qedhere
-\end{enumerate}
-\end{proof}
-
-\begin{lemma}\label{2.2}
-Let $K$ be henselian with respect to two different valuations $w_1,w_2\in |\operatorname{Spv}(K)|$.
-Then $K$ is algebraically closed or $w_1,w_2\in\operatorname{Spv}_\text{rd}^{(2)}(K),\ v(w_1)=v(w_2)$ and $k(v(w_1))$ is separably closed.
-In this case, we have $\operatorname{cd}(G_K)\leq 2$.
-\end{lemma}
-
-\begin{proof}
-Assume that $K$ is not algebraically closed.
-$w_1$ and $w_2$ can't be trivial, for otherwise $\dim k(w_i)=\dim (K)\neq 0$.
-By the Theorem of F.K.
-Schmidt \ref{theorem-fkschmidt}, we can see that $w_1,w_2$ are dependent.
-Since two different valuations of rank $1$ are independent, at least one of $w_1$ and $w_2$ is of rank $2$.
-Now, assume that $w_1$ is, say, of rank $1$ and $w_2$ is of rank $2$.
-Since they are dependent, it follows $v(w_2) = w_1$ and $w_1$ has a non-trivial specialization, a contradiction to $w_1\in |\operatorname{Spv}(K)|$.
-
-Thus $w_1,w_2$ are both of rank $2$ and by Lemma \ref{rank-dimension-inequality} we can see that $w_1,w_2\in\operatorname{Spv}_\text{rd}^{(2)}(K)$.
-Set $v_i=v(w_i)\in\operatorname{Spv}^{(1)}_\text{d}(K)$.
-By Lemma \ref{2.2-pre} $K$ is henselian with respect to both $v_1$ and $v_2$.
-Again, by the Theorem of F.K.
-Schmidt \ref{theorem-fkschmidt}, we get $v_1 = v_2$.
-Furthermore $k(v_1) = k(v_2)$ is henselian with respect to two different valuations of rank $1$, namely $w_1\mod v_1$ and $w_2\mod v_1$, making $k(v_1)$ separably closed.
-
-The second assertion follows from the first and from [K2], §5, Theorem 1.
-\end{proof}
-
-\begin{definition}
-Let $K$ be finitely generated.
-A \textit{model} of $K$ is a regular, proper scheme $\mathfrak{X}$ over $\mathbb{Z}$ with function field $K$.
-$\mathfrak{X}_i$ denotes the set of points $x\in\mathfrak{X}$ whose closure $\overline{\{x\}}$ is $i$-dimensional.
-
-A \textit{morphism of models} is a morphism of schemes inducing the identity on $K$. We denote the category of models of $K$ by $\mathfrak{M}(K)$; it is filtered.
-
-For every model $\mathfrak{X}$ there is a canonical map $\pi_\mathfrak{X}:\operatorname{Spv}(K)\to\mathfrak{X}$.
-$\pi_\mathfrak{X}(v)$  is the unique point of $\mathfrak{X}$ such that $\mathcal{O}_{\mathfrak{X}, \pi_\mathfrak{X}(v)}\subset\mathcal{O}_v$ and the inclusion $\mathcal{O}_{\mathfrak{X}, \pi_\mathfrak{X}(v)}\hookrightarrow \mathcal{O}_v$ is a homomorphism of local rings.
-\end{definition}
-
-\begin{lemma}\phantomsection\label{2.3}
-\begin{enumerate}[(i)]
-\item $\operatorname{Spv}(K)$, together with the maps $\pi_{\mathfrak{X}}$, is the following projective limit in the category of topological spaces: 
-\[\operatorname{Spv}(K) =\varprojlim_{\mathfrak{X}\in\mathfrak{M}(K)}\mathfrak{X}\]
-In particular we have:
-\[ |\operatorname{Spv}(K)| =\varprojlim_{\mathfrak{X}\in\mathfrak{M}(K)}|\mathfrak{X}| \]
-\item For all $v\in\operatorname{Spv}(K)$ we have:
-\[ \mathcal{O}_v = \varinjlim_{\mathfrak{X}\in\mathfrak{M}(K)}\mathcal{O}_{\mathfrak{X},\pi_\mathfrak{X}(v)} \]
-\item For $\mathfrak{X}\in\mathfrak{M}(K),\ x\in\mathfrak{X}$ let $K_x$ be the quotient field of a henselization of $\mathcal{O}_{\mathfrak{X},x}$. Then we have:
-\[ K_v=\varinjlim_{\mathfrak{X}\in\mathfrak{M}(K)}K_{\pi_\mathfrak{X}(v)} \]
-\end{enumerate}
-\end{lemma}
-
-\begin{proof}
-(i) and (ii) follow from [HK], Lemma 2.4.3 and (iii) is a consequence of (ii) and the fact that henselizations commute with filtrated direct limits, see \cite{EGA4} 18.6.14.
-\end{proof}
-
-The following lemma is the key result in proving Theorem \ref{thm:main-result}.
-First, we use K.
-Kato's two-dimensional local-global principle to show the assertion for finitely generated $K$.
-Then, we'll use a quasi-compactness argument to see that we can drop the assumption that $K$ is finitely generated.
-
-\begin{lemma}\label{2.4-finitelygenerated}
-Let $K$ be finitely generated and $\ell\neq 2$ be a prime. The following map, induced by the restrictions, is injective:
-\[ \mathrm{H}^3(K,\mathbb{Z}/\ell\mathbb{Z})\longrightarrow\prod_{w\in |\operatorname{Spv}(K)| }\mathrm{H}^3(K_w,\mathbb{Z}/\ell\mathbb{Z}) \]
-\end{lemma}
-
-\begin{proof}
-Let $L=K(\mu_{\ell},\sqrt{-1})/K$. Consider the restriction and corestriction maps:
-\[ \begin{tikzcd}
-\mathrm{H}^3(K,\mathbb{Z}/\ell\mathbb{Z}) \ar[r, "\operatorname{res}"] & \mathrm{H}^3(L,\mathbb{Z}/\ell\mathbb{Z})\ar[r, "\operatorname{cor}"]& \mathrm{H}^3(K,\mathbb{Z}/\ell\mathbb{Z}) 
-\end{tikzcd} \]
-We have $\operatorname{cor}\circ\operatorname{res} = [L:K]$.
-Since $[L:K]$ is coprime to $\ell$, we see that $\operatorname{res}$ is injective.
-Thus we can assume that $K=L$, i.e.
-$\mu_{\ell}\subset K$ and $K$ is not formally real.
-Now we can replace $\mathbb{Z}/\ell\mathbb{Z}$ by $\mathbb{Z}/\ell\mathbb{Z}(2) = \mu_\ell^{\otimes 2}$ and let us write $\mathrm{H}^3(L) = \mathrm{H}^3(L, \mathbb{Z}/\ell\mathbb{Z}(2))$.
-
-Let $\mathfrak{X}$ be a model of $K$.
-For $x\in\mathfrak{X}_1$ the local ring $\mathcal{O}_{\mathfrak{X}, x}$ is a discrete valuation ring and defines a valuation $v_x\in\operatorname{Spv}_\text{d}^{(1)}(K)$.
-By \cite{Ka86} Theorem 0.8, the map
-\[ \mathrm{H}^3(K)\longrightarrow \prod_{x\in\mathfrak{X}_1}\mathrm{H}^3(K_{v_x}) \]
-is injective and its image is contained in $\bigoplus \mathrm{H}^3(K_{v_x})$. By \cite{Ka86} Lemma 1.4, for a discrete valuation $v$ of $K$ the map
-\[ \mathrm{H}^3(K_v) \longrightarrow \prod_w\mathrm{H}^3(K_w) \]
-where $w$ ranges over all rank-$2$ valuations with $v(w) = v$, can be identified with the following map:
-\[ {}_\ell\operatorname{Br}(k) \longrightarrow \prod_{\tilde{w}\in\operatorname{Spv}^{(1)}(k)} {}_\ell\operatorname{Br}(k_{\tilde{w}}) \]
-where $k = k(v)$. This map is injective by the classical Hasse principle and its image is contained in $\bigoplus {}_\ell\operatorname{Br}(k_{\tilde{w}})$. Combining these maps yields the injection:
-\[\Phi: \mathrm{H}^3(K)\longrightarrow \bigoplus_{x\in\mathfrak{X}_1}\  \bigoplus_{\substack{w\in\operatorname{Spv}_\text{rd}^{(2)}(K) \\ v(w) = v_x}} \mathrm{H}^3(K_w) \]
-The direct sum ranges over a subset of $|\operatorname{Spv}(K)|$, thus this concludes the proof. 
-\end{proof}
-
-In order to generalize Lemma \ref{2.4-finitelygenerated} for arbitrary, not necessarily finitely generated $K$, we need the following Lemma:
-
-\begin{lemma}\label{2.4-closedset}
-Let $f\in\mathrm{H}^3(K)$. Then the support of $f$ is closed in $|\operatorname{Spv}(K)|$:
-\[ \operatorname{supp}(f) = \{w\in |\operatorname{Spv}(K)|\mid \operatorname{res}_{K_w/K}(f)\neq 0 \} \]
-where $\operatorname{res}_{K_w/K}: \mathrm{H}^3(K)\to \mathrm{H}^3(K_w)$.
-\end{lemma}
-
-\begin{proof}
-We adopt the notation from the previous proof. First, we shall prove that the following map is injective for every model $\mathfrak{X}$ of $K$:
-\[ f: \mathrm{H}^3(K) \longrightarrow \prod_{y\in\mathfrak{X}_0}\mathrm{H}^3(K_y) \]
-For $x\in\mathfrak{X}$ let $\mathcal{O}_x$ be a henselization of $\mathcal{O}_{\mathfrak{X}, x}$ and $K_x$ its quotient field.
-For $y\in\mathfrak{X}_0$ set $P_y=\operatorname{Spec}(\mathcal{O}_y)_1$.
-For $z\in P_y$ let $\mathcal{O}_z$ be a henselization of $\mathcal{O}_y$ at $z$ and $K_z$ its quotient field.
-For $y\in\mathfrak{X}_0$ and $x\in\mathfrak{X}_1$ let $P_{x,y}$ be the subset of $P_y$ consisting of all $z\in P_y$ which are mapped to $x$ under $\operatorname{Spec}(\mathcal{O}_y)\to\mathfrak{X}$.
-Consider the following commutative diagram:
-\[ \begin{tikzcd}
-\mathrm{H}^3(K)\ar[r, "\varphi"]\ar[d, "f"'] & \displaystyle\prod_{x\in\mathfrak{X}_1}\mathrm{H}^3(K_x)\ar[d, "\psi"]\\
-\displaystyle\prod_{y\in\mathfrak{X}_0}\mathrm{H}^3(K_y) \ar[r, "g"']& \displaystyle\prod_{\substack{x\in\mathfrak{X}_1 \\ y\in\mathfrak{X}_0}}\prod_{z\in P_{x,y}}\mathrm{H}^3(K_z)
-\end{tikzcd} \]
-According to [Sa], Theorem (2-15) $g$ is injective.
-Now, $\psi\circ\varphi$ can be identified with the injective map $\Phi$ from the proof of the previous Lemma.
-Thus its image is contained in $\bigoplus\mathrm{H}^3(K_w)$.
-Therefore, $f$ is injective as well and its image is contained in $\bigoplus\mathrm{H}^3(K_y)$.
-Now, one easily checks that:
-\[ \operatorname{supp}(f) = \bigcap_{\mathfrak{X}\in \mathfrak{M}(K)} \pi_\mathfrak{X}^{-1}(\{ y\in\mathfrak{X}_0\mid \operatorname{res}_{K_y/K}(f)\neq 0 \}) \]
-Note that the right set is closed in $\operatorname{Spv}(K)$.
-\end{proof}
-
-Now the general case follows by a quasi-compactness argument:
+Let $K/\mathbb{Q}$ be a field of dimension $2$.
 
 \begin{lemma}[\textit{Main Lemma}]\label{2.4}
-Let $K$ be not necessarily finitely generated and $\ell\neq 2$ be a prime. The following map, induced by the restrictions, is injective:
+Let $\ell\neq 2$ be a prime. The following map, induced by the restrictions, is injective:
 \[ \mathrm{H}^3(K,\mathbb{Z}/\ell\mathbb{Z})\longrightarrow \prod_{w\in |\operatorname{Spv}(K)| }\mathrm{H}^3(K_w,\mathbb{Z}/\ell\mathbb{Z}) \]
 \end{lemma}
-
-\begin{proof}
-Just like in Lemma \ref{2.4-finitelygenerated} we can assume that $\mu_\ell\subset K$ and $K$ is not formally real and replace $\mathbb{Z}/\ell\mathbb{Z}$ by $\mathbb{Z}/\ell\mathbb{Z}(2)$.
-We adopt the notation from the previous proofs.
-
-Let $f\in\mathrm{H}^3(K)$ such that $\operatorname{res}_{K_w/K}(f) = 0$ for all $w\in |\operatorname{Spv}(K)|$.
-There exists a finitely generated subfield $K_0$ of $K$ such that $f = \operatorname{res}_{K/K_0}(g)$ for some $g\in\mathrm{H}^3(K_0)$ and $K/K_0$ is algebraic.
-Let $\{K_i\mid i\in I\}$ be the set of all finite subextensions of $K/K_0$.
-For $w\in |\operatorname{Spv}(K)|$ we denote the restriction of $w$ to $K_i$ with $w$ as well.
-
-Let  $w\in |\operatorname{Spv}(K)|$.
-Since $\mathrm{H}^3(K_w) = \varinjlim_I \mathrm{H}^3((K_i)_w)$ there exists an $i = i(w)\in I$ such that $\operatorname{res}_{(K_i)_w/K}(g) = 0$.
-By Lemma \ref{2.4-closedset} there is an open neighborhood $U_{i(w)}\subset |\operatorname{Spv}(K_i)|$ of $w$ such that $\operatorname{res}_{(K_i)_v/K}(g)=0$ for all $v\in U_{i(w)}$.
-For $K_i\subset K_j\subset K$ let
-\[ \varphi_i: |\operatorname{Spv}(K)| \to |\operatorname{Spv}(K_i)| \]
-\[ \varphi_{ij}: |\operatorname{Spv}(K_j)|\to |\operatorname{Spv}(K_i)| \]
-be the restriction maps. They are continuous. Since $|\operatorname{Spv}(K)|$ is quasi-compact by Lemma \ref{spv-topology} we can find a finite set $S\subset |\operatorname{Spv}(K)|$ such that:
-\[ \bigcup_{w\in S}\varphi^{-1}_{i(w)}(U_{i(w)}) = |\operatorname{Spv}(K)| \]
-Let $j\in I$ such that $K_j$ is the composite of the fields $K_{i(w)},\ w\in S$. We also have:
-\[ \bigcup_{w\in S}\varphi^{-1}_{i(w)j}(U_{i(w)}) = |\operatorname{Spv}(K_j)| \]
-Let $v\in |\operatorname{Spv}(K_j)|$ and let $i = i(w),\ w\in S$ such that $v\in\varphi_{ij}^{-1}(U_{i(w)})$. We have:
-\[ \operatorname{res}_{(K_j)_v/K_j}(\operatorname{res}_{K_j/K_0}(g)) = \operatorname{res}_{(K_j)_v/(K_i)_v}(\operatorname{res}_{(K_i)_v/K_0}(g)) = 0 \]
-Since $K_j$ is finitely generated we obtain $\operatorname{res}_{K_j/K_0}(g) = 0$ with Lemma \ref{2.4-finitelygenerated}, thus $f = \operatorname{res}_{K/K_0}(g) = 0$.
-\end{proof}
-
-Now we calculate the cohomology groups $\mathrm{H}^3(K_w,\mathbb{Z}/\ell\mathbb{Z})$ for $w\in |\operatorname{Spv}(K)|$.
-With the help of Lemma \ref{2.4} we can find a cohomological criterion of henselian fields over $K$ with respect to a $w\in\operatorname{Spv}_\text{d}^{(2)}(K)$.
 
@@
 Set $k=k(w)$ and $\Gamma = \Gamma_w$.
-Then $k$ is an algebraic extension of a finite field $\mathbb{F}_p$ by Lemma \ref{2.2-pre}.
+Then $k$ is an algebraic extension of a finite field $\mathbb{F}_p$.
 Let $\ell\neq p$ be a prime and assume $\mu_\ell\subset K$.
@@
 \end{lemma}
-
-\begin{proof}
-Let $T$ be the absolute inertia group of $K$ and consider the short exact sequence from Theorem \ref{theorem-absoluteinertia}:
-\[ 1\longrightarrow T\longrightarrow G_K\longrightarrow G_k\longrightarrow 1 \]
-
-Let $G_\ell$ be an $\ell$-Sylow group of $G_K$.
-By Theorem \ref{theorem-sylowgroups-inertiagroup}, the $\ell$-Sylow group of $T$ isomorphic to $\mathbb{Z}_\ell^r$, where $r =\dim_{\mathbb{F}_\ell}(\Gamma/\ell\Gamma)$.
-Furthermore, $G_k\subset G_{\mathbb{F}_p}\cong\hat{\mathbb{Z}}$, thus the $\ell$-Sylow group $G_{k,\ell}$ of $G_k$ is either isomorphic to $\mathbb{Z}_\ell$ or trivial.
-The exact sequence above gives rise to the short exact sequence of their $\ell$-Sylow subgroups:
-\[ 0\longrightarrow\mathbb{Z}_\ell^r \longrightarrow G_\ell \longrightarrow G_{k,\ell}\longrightarrow 1 \]
-Since $G_{k,\ell}$ is a free pro-$\ell$ group, this sequence splits and we have an isomorphism $G_\ell \cong \mathbb{Z}_\ell^r\rtimes G_{k,\ell}$ with $r =\dim_{\mathbb{F}_\ell}(\Gamma/\ell\Gamma)\leq \operatorname{rrank}(\Gamma)\leq 2$.
-
-Let $G_K(\ell)$ be the maximal pro-$\ell$ quotient of $G_K$.
-Since $\mu_\ell\subset K$, the action of $G_k$ on the $\ell$-Sylow group of the inertia group of $w$ factors through the maximal pro-$\ell$ quotient $G_k(\ell)$ of $G_k$.
-Since $G_{k,\ell}\cong G_k(\ell)$, we also see that the canonical projection $G_K\to G_K(\ell)$ restricts to an isomorphism $G_\ell\to G_K(\ell)$.
-Now consider the following composition:
-\[ \begin{tikzcd}
-\mathrm{H}^3(G_K(\ell),\mathbb{Z}/\ell\mathbb{Z}) \ar[r, "\operatorname{inf}"] & \mathrm{H}^3(G_K,\mathbb{Z}/\ell\mathbb{Z}) \ar[r, "\operatorname{res}"] & \mathrm{H}^3(G_\ell,\mathbb{Z}/\ell\mathbb{Z})
-\end{tikzcd} \]
-Since $\operatorname{res}\circ\operatorname{inf}$ is an isomorphism, $\operatorname{inf}$ is injective and $\operatorname{res}$ is surjective.
-We have $\operatorname{cor}\circ\operatorname{res} = [G_K:G_\ell]$.
-However, $[G_K:G_\ell]$ is coprime to $\ell$, which makes $\operatorname{res}$ injective as well.
-Therefore, $\operatorname{inf}$ and $\operatorname{res}$ are both isomorphisms and it suffices to calculate $\mathrm{H}^3(G_\ell,\mathbb{Z}/\ell\mathbb{Z})$.
-By \cite{Se64}, Ch.
-I, §4, Proposition 22 we have:
-\[\operatorname{cd}(G_\ell) = \operatorname{cd}(\mathbb{Z}_\ell^r\rtimes G_k(\ell)) = \begin{cases}
-r, & G_k(\ell) = 0 \\
-r + 1, & G_k(\ell)\cong\mathbb{Z}_\ell
-\end{cases} \] 
-since $\operatorname{cd}(\mathbb{Z}_\ell)=1$.
-Thus $\mathrm{H}^3(G_\ell, \mathbb{Z}/\ell\mathbb{Z}) =0$ if $G_k(\ell) = 0$ or $r < 2$.
-Now let $G_k(\ell) \cong\mathbb{Z}_\ell$ and $r = 2$.
-According to the proof of \cite{Se64}, Ch.
-I, §4, Proposition 22, we have:
-\[ \mathrm{H}^3(G_\ell,\mathbb{Z}/\ell\mathbb{Z}) = \mathrm{H}^1(G_k(\ell), \mathrm{H}^2 (\mathbb{Z}_\ell^r, \mathbb{Z}/\ell\mathbb{Z})) \]
-By [NSW] Theorem 2.4.6, we have $\mathrm{H}^2(\mathbb{Z}_\ell^r,\mathbb{Z}/\ell\mathbb{Z}) \cong\mathrm{H}^1(\mathbb{Z}_\ell, \mathrm{H}^1(\mathbb{Z}_\ell, \mathbb{Z}/\ell\mathbb{Z})) = \mathbb{Z}/\ell\mathbb{Z}$, since $\mathrm{H}^1(\mathbb{Z}_\ell, \mathbb{Z}/\ell\mathbb{Z})=\operatorname{Hom}_\text{cont.}(\mathbb{Z}_\ell, \mathbb{Z}/\ell\mathbb{Z})=\mathbb{Z}/\ell\mathbb{Z}$.
-Therefore, $\mathrm{H}^3(G_\ell,\mathbb{Z}/\ell\mathbb{Z})=\mathbb{Z}/\ell\mathbb{Z}$.
-
-The second assertion follows from the first one, applied to $L$, by using the following equation of supernatural numbers ([E], Theorem 20.21):
-\[ [L:K] = (\Gamma':\Gamma) \cdot [k':k]\cdot p^d \]
-for some $d \leq \infty$, where $\Gamma'$ denotes the value group and $k'$ the residue field of the unique prolongation of $w$ to $L$.
-Note that $\operatorname{rank}(w) = \operatorname{rank}(v)$ in an algebraic field extension (\cite{ZS60}, Ch.
-VI, §11, Lemma 2).
-\end{proof}
 
@@
 
-Now $(K_0)_{w_1}$ is henselian both with respect to $w_1$ and with respect to the extension of $w_0$.
-Lemma \ref{2.2} yields $\operatorname{cd}(G_{(K_0)_{w_1}})\leq 2$, a contradiction.
-The uniqueness of $w$ is again a consequence of Lemma \ref{2.2}.
+Now $(K_0)_{w_1}$ is henselian both with respect to $w_1$ and with respect to the extension of $w_0$. A contradiction to $\operatorname{cd}(G_{(K_0)_{w_1}})\leq 2$.
 \item Choose the prime $\ell$ different from $\operatorname{char} k(w)$ so that (i) holds. According to Lemma \ref{2.5} we have:
@@
 
-\chapter{Correspondences}
-
-Let $K,K'$ be two finitely generated fields of characteristic $0$ and dimension $2$ and let $\sigma:G_K\to G_{K'}$ be an isomorphism of profinite groups.
-For intermediate fields $L$ of $\overline{K}/K$ we denote the corresponding intermediate field of $\overline{K'}/K'$ by $L'$.
-In other words, $\sigma G_L=G_{L'}$.
-
 \section{Local Correspondence}
 
-In this section we want to see that $\sigma$ induces bijections on defectless valuations.
-A direct consequence of Proposition \ref{2.6} is the bijection $\operatorname{Spv}_\text{d}^{(2)}(K)\to \operatorname{Spv}_\text{d}^{(2)}(K')$:
+Let $K,K'/\mathbb{Q}$ be two finitely generated fields of dimension $2$.
 
@@
 
-We will see that the bijection $\operatorname{Spv}_\text{d}^{(2)}(K)\to \operatorname{Spv}_\text{d}^{(2)}(K')$ above restricts to a bijection $\operatorname{Spv}_\text{rd}^{(2)}(K)\to \operatorname{Spv}_\text{rd}^{(2)}(K')$.
-Using this, we shall establish a bijection between the defectless valuations of rational rank $1$:
+\begin{lemma}
+$\sigma$ induces a bijection $\sigma:\operatorname{Spv}_\text{rd}^{(2)}(K) \to\operatorname{Spv}_\text{rd}^{(2)}(K')$.
+\end{lemma}
 
@@
 
-\begin{proof}
-Lemma \ref{3.1} shows that $\sigma$ induces a bijection $\sigma: \operatorname{Spv}_\text{d}^{(2)}(K)\to \operatorname{Spv}_\text{d}^{(2)}(K')$.
-For all $v\in\operatorname{Spv}(K)$ we choose a fixed henselization $K_v\subset\overline{K}$ with Lemma \ref{lemma-henselization-specialization}, such that if $v,w\in\operatorname{Spv}(K)$ and $w$ is a specialization of $v$ we have $K_v\subset K_w$.
-If $v$ is nontrivial, then in this case we get $w\in\operatorname{Spv}_\text{rd}^{(2)}(K)$ and $v=v(w)\in\operatorname{Spv}_\text{d}^{(1)}(K)$.
-We set $K'_{\sigma(w)} = (K_w)'$ for $w\in\operatorname{Spv}_\text{d}^{(2)}(K)$.
-Lemma \ref{3.1} shows that $K'_{\sigma(w)}$ is a henselization of $K'$ with respect to $\sigma(w)$.
-
-Let $w_1,w_2 \in\operatorname{Spv}_\text{d}^{(2)}(K)$ be two different valuations. The following statements are equivalent:
-\begin{enumerate}[(i)]
-\item The composite $K_{w_1}K_{w_2}$ is not algebraically closed.
-\item $w_1,w_2\in\operatorname{Spv}_\text{rd}^{(2)}(K)$ and $v(w_1) = v(w_2)$.
-\end{enumerate}
-The implication (i)$\implies$(ii) follows from Lemma \ref{2.2}. TODO
-
-In this case, $k(v')$ is separably closed where $v'$ is the unique extension of the valuation on $K_v$ to $K_{w_1}K_{w_2}$ where $v=v(w_1)$.
-By Lemma \ref{lemma-henselization-specialization} we have $K_{w_1}K_{w_2}\subset K_v^\text{t}$ and by Theorem \ref{theorem-absoluteinertia} (ii) $K_{w_1}K_{w_2}$ is the absolute inertia field $K_v^\text{t}$ of $K_v$.
-Furthermore, $K'_{\sigma(w_1)}K'_{\sigma(w_2)} = (K_{w_1}K_{w_2})' \neq \overline{K'}$ and again by Lemma \ref{2.2} we get: \[\sigma(w_1),\sigma(w_2)\in\operatorname{Spv}_\text{rd}^{(2)}(K)\quad \text{and} \quad v(\sigma(w_1)) = v(\sigma(w_2))\]
-Thus $\sigma: \operatorname{Spv}_\text{d}^{(2)}(K)\to \operatorname{Spv}_\text{d}^{(2)}(K')$ maps $\operatorname{Spv}_\text{rd}^{(2)}(K)$ onto $\operatorname{Spv}_\text{rd}^{(2)}(K')$ such that:
-\[ v(w_1)=v(w_2) \iff v(\sigma(w_1)) = v(\sigma(w_2)) \]
-This shows that $\sigma$ also induces a bijection $\sigma:\operatorname{Spv}_\text{d}^{(1)}(K)\to\operatorname{Spv}_\text{d}^{(1)}(K')$ and for all $w\in\operatorname{Spv}_\text{rd}^{(2)}(K)$ we have $\sigma(v(w)) = v(\sigma(w))$.
-
-Let $v\in\operatorname{Spv}_\text{d}^{(1)}(K)$ and let $N$ be the normal subgroup of $G_{K_v}$ generated by all subgroups $G_{K_w}$ for $w\in\operatorname{Spv}_\text{rd}^{(2)}(K)$ with $v(w) = v$.
-Since a global field has no non-trivial algebraic extension in which every prime splits completely, we get $N=G_{K_v}$.
-This shows that $K'_{v'} := (K_v)'$ is a henselization of $K'$ with respect to $v'=\sigma(v)$.
-Furthermore, let $w_1,w_2\in\operatorname{Spv}_\text{rd}^{(2)}(K)$ with $w_1\neq w_2$ and $v=v(w_1)=v(w_2)$, then:
-\[ \sigma(G_{K^\text{t}_v}) = \sigma(G_{K_{w_1}}\cap G_{K_{w_2}}) = G_{K'_{w'_1}}\cap G_{K'_{w'_2}} = G_{(K')^\text{t}_{v'}} \]
-Therefore, $\sigma$ induces with Theorem \ref{theorem-absoluteinertia} the following isomorphism:
-\[ G_{k(v)} \cong G_{K_v}/G_{K_v^\text{t}} \to G_{K'_{v'}}/G_{(K')_{v'}^\text{t}}\cong G_{k(v')} \]
-The rest follows from the Theorem of Neukirch-Uchida \ref{thm:neukirch-uchida}.
-\end{proof}
-
-\begin{remark}\label{3.3}
-If $L\subset M$ are intermediate fields of $\overline{K}/K$, which are finite over $K$, then $\sigma$ also induces bijections
-\[ \operatorname{Spv}_\text{d}^{(1)}(L) \to \operatorname{Spv}_\text{d}^{(1)}(L'),\qquad \operatorname{Spv}_\text{d}^{(1)}(M) \to \operatorname{Spv}_\text{d}^{(1)}(M') \]
-such that the following diagram commutes:
-\[\begin{tikzcd}
-\operatorname{Spv}_\text{d}^{(1)}(M) \rar["\sigma"]\dar & \operatorname{Spv}_\text{d}^{(1)}(M')\dar\\
-\operatorname{Spv}_\text{d}^{(1)}(L) \rar["\sigma"'] & \operatorname{Spv}_\text{d}^{(1)}(L')
-\end{tikzcd}\]
-where the vertical maps are restrictions. Let $w\in\operatorname{Spv}_\text{d}^{(1)}(M),\ v=w|_L$ and $w'=\sigma(w),\ v'=\sigma(v)$. Then the following diagram commutes:
-\[ \begin{tikzcd}
-k(w') \rar["\varphi_w"] & k(w)\\
-k(v') \rar["\varphi_v"']\uar[hook] & k(v)\uar[hook]
-\end{tikzcd} \]
-By limit process we see that $\sigma$ induces a bijection $\sigma: \operatorname{Spv}_\text{d}^{(1)}(\overline{K})\to \operatorname{Spv}_\text{d}^{(1)}(\overline{K'})$.
-Let $\overline{v}\in \operatorname{Spv}_\text{d}^{(1)}(\overline{K})$ and $G_{\overline{v}}\subset G_K$ be the decomposition group of $\overline{v}$.
-Then we have $\sigma(G_{\overline{v}}) = G_{\sigma(\overline{v})}$.
-Now it is easy to see that for every $g\in G_K$:
-\[ \sigma(g\overline{v}) = \sigma(g)\sigma(\overline{v}) \]
-\end{remark}
-
-\section{Global Correspondence mod \texorpdfstring{$n$}{n}}
-
-In this section we shall show that $\sigma: G_K\to G_{K'}$ induces group isomorphisms $K'^\times/n\to K^\times/n$ and $U_{v'}/n\to U_v/n$ for all $n\in\mathbb{N}$ and defectless valuations $v$ of rational rank $1$ with $\operatorname{char}k(v) =0$.
+\section{Global Correspondence mod $n$}
 
@@
 
-\begin{theorem}[\textit{Hilbert's Irreducibility Theorem}]\label{hilberts-irreducibility-theorem}
-Let $E$ be a number field and $f_1,\ldots, f_n$ be irreducible polynomials in $E[X_1,\ldots,X_r,Y_1,\ldots,Y_s]$. Then there exists $e_1,\ldots,e_r\in E$ such that the specializations
-\[ f_1(e_1,\ldots,e_r,Y_1,\ldots,Y_s),\ \ldots,\ f_n(e_1,\ldots,e_r,Y_1,\ldots,Y_s) \]
-are irreducible in $E[Y_1,\ldots,Y_s]$.
-\end{theorem}
-
-\begin{lemma}\label{4.1}
-Let $E$ be a number field and let $X$ be a smooth projective geometrically connected curve over $E$.
-Let $\pi:X'\to X$ be a connected Galois covering of degree $d$ which splits completely, i.e.
-every closed point $P\in X_0$ has exactly $d$ pre-images in $X'$.
-Then $d=1$.
-\end{lemma}
-
-\begin{proof}
-Let $f: X\to\mathbf{P}^1_E$ be a non-constant morphism and let $f' =  f\circ \pi: X'\to\mathbf{P}^1_E$.
-This induces a finite field extension $F/E(T)$ of their function fields.
-Since we're in characteristic $0$, $F/E(T)$ is also separable, and thus we have a primitive element $a\in F$, i.e.
-$F = E(T,a)$.
-
-Consider the minimal polynomial of $a$ in $E(T)[A]$.
-Clearing out the denominators, we get an irreducible polynomial $f\in E[T,A]$.
-By Hilbert's Irreducibility Theorem \ref{hilberts-irreducibility-theorem} we can find an $e\in E$ such that $f(e,A)\in E[A]$ is irreducible.
-
-Now $T-e\in E[T]$ corresponds to a closed point $P\in \mathbf{P}^1_E$, which in turn corresponds to a valuation in $E(T)$.
-So the points in the fiber $f'^{-1}(P)$ corresponds to valuations in $F$ above $T-e$.
-But $T-e$ is inert in $F/E(T)$ since $a\in E[T,a]/(T-e)$ has minimal polynomial $f(e,A)$ over $E[T]/(T-e)$ and $\deg f(e,A) = [F:E(T)]$.
-
-This shows that $P$ has only one preimage under $f'$. Therefore, it has only one preimage under $f$, say $P'$, and $P'$ has only one preimage under $\pi$. This concludes the proof.
-\end{proof}
-
-First of all, we want to show that $\sigma$ induces an isomorphism between the absolute Galois groups of the constant fields $G_k$ and $G_{k'}$. We need the following Lemma:
-
-\begin{lemma}\label{4.2-pre}
-We have $[k:\mathbb{Q}] = [k':\mathbb{Q}]$.
-\end{lemma}
-
-\begin{proof}
-First we assume that $k/\mathbb{Q}$ is a Galois extension and $X$ has a $k$-rational point $P_0$. We can consider $k'$ as a subfield of $k$ via: 
-\[\begin{tikzcd}
-k'\ar[r] & k(P'_0)\ar[r, "\varphi_{P_0}"', "\cong"] & k(P_0) \ar[r, "\cong"] & k
-\end{tikzcd}\]
-For all $P\in X_0$ and $P'=\sigma(P)\in X'_0$ we also get an embedding:
-\[ \begin{tikzcd}
-k \ar[r] & k(P) \ar[r, "\varphi_P^{-1}"', "\cong"] & k(P')
-\end{tikzcd} \]
-Let us consider the morphism $f:X'\times_{k'} k\to X'$. For $P'\in X'_0$ we have as topological spaces:
-\[ f^{-1}(P') = \operatorname{Spec} k(P')\times_{k'} k \]
-Any embedding $k\to \overline{k(P')}$ maps $k$ into $k(P')$ because $k/\mathbb{Q}$ is Galois. Thus we have:
-\[ k(P')\otimes_{k'} k\cong \prod_{\operatorname{Hom}_{k'}(k, k(P'))} k(P') \]
-Now we can see that $f$ splits completely. By Lemma \ref{4.1} we get $[k:k'] = 1$, hence $[k:\mathbb{Q}]= [k':\mathbb{Q}]$.
-
-Now we drop the assumption that $k/\mathbb{Q}$ is Galois.
-There exists a finite extension $k_1/k$ such that $k_1/\mathbb{Q}$ is Galois and $X(k_1)\neq\varnothing$.
-Let $K_1 = Kk_1$ be the constant field extension of $K$ with $k_1$ and let $k'_1$ be the constant field of $K'_1$.
-Applying the above argument to $K_1$ and $K'_1$ yields $[k_1:\mathbb{Q}] = [k'_1:\mathbb{Q}]$.
-Therefore:
-\begin{align*}
-[k:\mathbb{Q}] &= [k_1:\mathbb{Q}]\cdot [K_1:K]^{-1} \\
-&= [k'_1:\mathbb{Q}]\cdot [K'_1:K']^{-1}\\
-&= [k'_1:\mathbb{Q}]\cdot ([K'_1: K'k_1'] \cdot [K'k_1':K'])^{-1}\\
-&= [k'_1:\mathbb{Q}]\cdot  ([K'_1: K'k_1'] \cdot [k'_1:k'])^{-1} \\
-&\leq [k'_1:\mathbb{Q}]\cdot [k'_1:k']^{-1} = [k':\mathbb{Q}]
-\end{align*}
-By symmetry, we get $[k:\mathbb{Q}] = [k':\mathbb{Q}]$.
-\end{proof}
-
 \begin{lemma}\label{4.2}
@@
 \end{lemma}
-
-\begin{proof}
-Let $L/K$ be a finite subextension of $\overline{K}/K$. Let $\ell$ and $\ell'$ be the constant fields of $L$ and $L'$ respectively. Applying Lemma \ref{4.2-pre} to $K$ and $L$ gives:
-\[ [\ell:k] = [\ell':k'] \]
-Therefore $L/K$ is a constant field extension iff $L'/K'$ is a constant field extension.
-Since $K\overline{k}/K$ and $K'\overline{k'}/K$ are the union of their finite subextensions, we obtain $(K\overline{k})' = K'\overline{k'}$.
-This implies that $\sigma$ induces an isomorphism:
-\[G_k \cong \operatorname{Gal}(K\overline{k}/K) \to \operatorname{Gal}(K'\overline{k'}/K')\cong G_{k'}\] 
-Everything else follows from Neukirch-Uchida \ref{thm:neukirch-uchida}.
-\end{proof}
 
@@
 
-\begin{remark}\label{remark-localphi-extension}
-Obviously, for intermediate fields $L\subset M$ of $\overline{K}/K$, the following diagram commutes:
-\[ \begin{tikzcd}
-L'^\times/n \ar[r, "\varphi_n"]\dar & L^\times/n \dar\\
-M'^\times/n \ar[r, "\varphi_n"'] & M^\times/n
-\end{tikzcd} \]
-\end{remark}
-
-Let $P\in X_0$ and $v_P:K\to\mathbb{Z}\cup\{\infty\}$ be the corresponding normalized discrete valuation and let $U_P$ be the group of units of the valuation ring $\mathcal{O}_P$.
-We denote the natural projection $U_P\to k(P)^\times,\ x\mapsto x\mod P$ by $r_P$.
-
-Let $n\in\mathbb{N}$. Tensoring the exact sequence $1\to U_P\to K^\times \to\mathbb{Z}\to 0$ with $\mathbb{Z}/n\mathbb{Z}$ preserves its exactness since $\mathbb{Z}$ is $\mathbb{Z}$-flat:
-\[ 1\longrightarrow U_P/n\longrightarrow K^\times/n\longrightarrow\mathbb{Z}/n\mathbb{Z}\longrightarrow 0 \]
-We'll use the same notation for $P' = \sigma(P)\in X'_0$.
-
 \begin{lemma}\phantomsection\label{4.3}
@@
 
-\begin{proof}
-\begin{enumerate}[(i)]
-\item Consider the absolute inertia field $L = K_P^\text{t}$ of $K_P$ and let $w$ be the unique extension of $v_P$ in $L$.
-By Theorem \ref{theorem-absoluteinertia} (i), $w$ is discrete as well and we have $k(w) = k(v_P)^\text{s}$.
-Observe the following exact sequence:
-\[ 1 \longrightarrow U_w/n \longrightarrow L^\times /n\longrightarrow \mathbb{Z}/n\mathbb{Z}\longrightarrow 0 \]
-Since $k(w)$ is separably closed, with $\operatorname{char} k(w)=0$ and Hensel's lemma we can see that $U_w$ is a divisible group, hence we get an isomorphism $L^\times /n\cong \mathbb{Z}/n\mathbb{Z}$ induced by $w$.
-Therefore, the canonical map $K^\times/n\to L^\times/n$ can be identified with the map $K^\times/n \to\mathbb{Z}/n\mathbb{Z}$ induced by $v_P$.
-Now the following commutative diagram with exact columns gives us the result:
-\[ \begin{tikzcd}
-1 \ar[d] & 1 \ar[d]\\
-U_{P'}/n \ar[r, dashed]\ar[d] & U_P/n\ar[d]\\
-K'^\times /n \ar[r, "\varphi_n"]\ar[d] & K^\times/n\ar[d]\\
-L'^\times /n \ar[r, "\varphi_n"'] &  L^\times /n
-\end{tikzcd}\]
-Note that $(K^\text{t}_P)' = K'^{\text{t}}_{P'}$ as proven in Lemma \ref{3.2}.
-\item Consider the field extension $K_P/K$.
-By Remark \ref{remark-localphi-extension}, without loss of generality, we can assume $K=K_P$.
-Since $\operatorname{char}k(P)=0$, by Hensel's lemma, $r_P/n$ is an isomorphism and the composition
-\[\begin{tikzcd}
-k(P)^\times / n\ar[rr, "(r_P/n)^{-1}"] && U_P/n \ar[rr, hook] && K_P^\times/n
-\end{tikzcd}\]
-coincides with the inflation $\operatorname{inf}: \mathrm{H}^1(k(P), \mu_n) \to \mathrm{H}^1(K_P,\mu_n)$.
-The same holds for $P'$ due to Lemma \ref{3.2}.
-Therefore, the assertion follows from the commutativity of the following diagram:
-\[ \begin{tikzcd}
-k(P')^\times/n \ar[r, "\cong"] \ar[d, "\varphi_P/n"'] & \mathrm{H}^1(k(P'),\mu_n) \ar[r, "\mathrm{inf}"] \ar[d]& \mathrm{H}^1(K'_{P'}, \mu_n) \ar[d, "\varphi_n"] \\
-k(P)^\times/n \ar[r, "\cong"'] & \mathrm{H}^1(k(P),\mu_n)  \ar[r, "\mathrm{inf}"'] & \mathrm{H}^1(K_{P}, \mu_n)
-\end{tikzcd}  \qedhere\]
-\end{enumerate}
-\end{proof}
-
-\chapter{Theorem}
-
-Before we start proving Pop's Theorem \ref{thm:main-result}, we need some tools that will help us constructing a group isomorphism $K'^\times\to K^\times$ from the modulo $n$ isomorphisms $K'^\times/n\to K^\times/n$.
-
-\section{Free-By-Finite Groups}
-
-\begin{definition}\label{5.1}
-Let $A$ be an abelian group. We have the following short exact sequence:
-\[0\longrightarrow A_\text{tor} \longrightarrow A\longrightarrow A/A_\text{tor}\longrightarrow 0 \]
-$A$ is called \textit{free-by-finite} if its torsion subgroup $A_\text{tor}$ is finite and the quotient $A/A_\text{tor}$ is free abelian.
-If $A$ is free-by-finite, the above sequence splits and $A$ is the direct sum of a finite abelian group and a free abelian group.
-It's evident that the converse also holds.
-\end{definition}
-
-\begin{example}
-\begin{itemize}
-\item Free or finite abelian groups are free-by-finite.
-\item Finitely generated abelian groups are free-by-finite by the structure theorem for finitely generated $\mathbb{Z}$-modules.
-\item Any subgroup of a free-by-finite group is free-by-finite as well.
-\end{itemize}
-\end{example}
-
-\begin{lemma}\phantomsection\label{5.2}
-\begin{enumerate}[(i)]
-\item Let $f: A\to B$ be a homomorphism of abelian groups with finite kernel and cokernel. Then $A$ is free-by-finite iff $B$ is free-by-finite.
-\item Let $0\longrightarrow A'\stackrel{i}{\longrightarrow}A\stackrel{p}{\longrightarrow}A''\longrightarrow 0$ be an exact sequence of abelian groups.
-If $A'$ and $A''$ are free-by-finite, then $A$ is free-by-finite.
-\item Let $0\longrightarrow A'\stackrel{i}{\longrightarrow}A\stackrel{p}{\longrightarrow}A''\longrightarrow 0$ be an exact sequence of abelian groups.
-If $A$ is free-by-finite and $A'$ is finitely generated, then $A''$ is free-by-finite.
-\item Let $A$ be free-by-finite. Then $\bigcap_{n\in\mathbb{N}}nA=0$.
-\end{enumerate}
-\end{lemma}
-
-\begin{proof}
-\begin{enumerate}[(i)]
-\item Let $B$ be free-by-finite and $K = \ker(f)$.
-Considering $A/K$ as a subgroup of $B$, we see that $A/K$ is free-by-finite.
-Since $(A/K)_\text{tor}$ and $K$ are finite, $K' = \{a\in A\mid na\in K \text{ for some }n\in\mathbb{N}\}$ is finite as well, in particular $A_\text{tor}\subset K'$ is finite.
-In fact $K'=A_\text{tor}$, since $K\subset A_\text{tor}$.
-Now consider the surjection $A\to A/K\to (A/K)/(A/K)_\text{tor}$.
-This map has kernel $K'$, thus inducing an isomorphism from $A/A_\text{tor}$ to a free abelian group.
-
-Now let $A$ be free-by-finite, i.e.
-$A = A_\text{tor} \oplus F$ for some free abelian group $F$, and $K=\ker(f)\subset A_\text{tor}$.
-We see that $A/K$ is free by finite as well, and by extension, the image $I$ of $f$.
-Set $m = \#B/I$ and consider the canonical map $B\stackrel{\cdot m}{\to} I \to I/I_\text{tor}$.
-This map has kernel $\{b\in B \mid mb \in I_\text{tor}\} = B_\text{tor}$, making $B/B_\text{tor}$ isomorphic to a subgroup of the free abelian group $I/I_\text{tor}$.
-Now both $I_\text{tor}$ and $B_\text{tor}/I_\text{tor} \subset B/I$ are finite, hence $B_\text{tor}$ is finite as well.
-\item The exact sequence induces the exact sequence:
-\[0 \longrightarrow A'_\text{tor} \longrightarrow A_\text{tor} \longrightarrow A''_\text{tor}\] 
-Since the surrounding groups are finite, $A_\text{tor}$ is finite as well. Using (i) we see that it suffices to show that $A/A_\text{tor}$ is free-by-finite.
-
-Observe the following commutative diagram with exact rows, where we consider $A'$ as a subgroup of $A$:
-\[ \begin{tikzcd}
-0 \ar[r]& A'/A'_\text{tor}\ar[r]\ar[d, "f"'] & A/A_\text{tor}\ar[r]\ar[d, "\cong"] & A/(A'+A_\text{tor}) \ar[r]\ar[d, "g"]& 0\\
-0 \ar[r] & p^{-1}(A''_\text{tor})/A_\text{tor} \ar[r]& A/A_\text{tor}\ar[r] & A''/A''_\text{tor}\ar[r] & 0
-\end{tikzcd} \]
-Note that $f$ is injective and $g$ is surjective. Using the snake lemma, we get an isomorphism $\operatorname{coker}(f)\cong \ker(g)$. We have:
-\[\ker(g) = \ker( (A/A')/(A_\text{tor}/A'_\text{tor}) \to A''/A''_\text{tor} ) = A''_\text{tor}/(A_\text{tor}/A'_\text{tor})\]
-Since $A''_\text{tor}$ is finite, $\operatorname{coker}(f)\cong \ker(g)$ is finite, too.
-Applying (i) on $f$, we can see that $p^{-1}(A''_\text{tor})/A_\text{tor}$ is free-by-finite.
-Since $A''/A''_\text{tor}$ is free abelian, the short exact sequence in the second row splits, making $A/A_\text{tor}$ a direct sum of a free-by-finite group and a free group.
-\item We consider $A'$ as a subgroup of $A$.
-Let $T=A_\text{tor}$.
-By replacing $A'$, $A$, $A''$ with $A'/(T\cap A')$, $A/T$, $A''/p(T)$ respectively, we can assume that $A$ is free and $A'$ is free of finite rank.
-Let $B$ be a basis of $A$.
-There exists a finite subset $C\subset B$ such that $A'\subset \bigoplus_{b\in C}\mathbb{Z}b$.
-We get an exact sequence:
-\[ 0 \longrightarrow \Big(\bigoplus_{b\in C} \mathbb{Z}b\Big)\Big/A' \longrightarrow A/A' \longrightarrow \bigoplus_{b\in B -  C}\mathbb{Z}b \longrightarrow 0 \]
-The left abelian group is finitely generated, particularly free-by-finite. The right group is free, hence also free-by-finite. By (ii), $A/A'\cong A''$ is free-by-finite as well.
-\item It suffices to prove this assertion for finite groups and for $A=\mathbb{Z}$.
-For finite $A$ we have $nA = 0$ for $n = \#A$.
-Since $\mathbb{Z}$ has no nontrivial nilpotent elements, the intersection of all prime ideals of $\mathbb{Z}$ is already trivial.
-\qedhere
-\end{enumerate}
-\end{proof}
-
-\begin{lemma}\label{5.3}
-Let $F$ be a number field. Then:
-\begin{enumerate}[(i)]
-\item $F^\times$ is free-by-finite.
-\item Let $F_1,\ldots,F_n$ be finite extensions of $F$. Then 
-\[\Big(\prod_{i=1}^{n}F_i^\times\Big)/F^\times\] 
-is free-by-finite, where $F^\times$ is considered as a subgroup of $\prod_{i=1}^{n}F_i^\times$ via the diagonal embedding.
-\end{enumerate}
-\end{lemma}
-
-\begin{proof}
-\begin{enumerate}[(i)]
-\item Let $\mathcal{P}(F)$ be the group of principal ideals of $F$. Consider the following exact sequence:
-\[ \begin{tikzcd}
-1 \ar[r] & \mathcal{O}_F^\times \ar[r] & F^\times \ar[r]& \mathcal{P}(F) \ar[r] & 1
-\end{tikzcd} \]
-By Dirichlet's Unit Theorem, $\mathcal{O}_F^\times$ is free-by-finite.
-$\mathcal{P}(F)$, as a subset of the ideal group of $F$, is free abelian, in particular free-by-finite.
-The assertion follows from Lemma \ref{5.2} (ii).
-\item Consider the following exact sequence:
-\[ \begin{tikzcd}
-1 \ar[r] & \displaystyle \prod_{i=2}^{n} F_i^\times \ar[r]& \displaystyle \Big(\prod_{i=1}^{n}F_i^\times\Big)\Big/F^\times \ar[r]& F_1^\times/F^\times \ar[r]& 1
-\end{tikzcd} \]
-The left group is free-by-finite by (i), and by Lemma \ref{5.2} (ii), we only need to prove that the right group is free-by-finite.
-
-Let $\mathcal{P}(F)$ denote the group of principal ideals of $F$, $\mathcal{I}(F)$ the ideal group of $F$, and $\operatorname{Cl}(F) = \mathcal{I}(F)/\mathcal{P}(F)$ the ideal class group of $F$.
-We have the following exact sequence:
-\[ \begin{tikzcd}
-1 \ar[r]& \mathcal{O}_{F_1}^\times /\mathcal{O}_F^\times\ar[r] & F_1^\times / F^\times\ar[r] & \mathcal{P}(F_1)/\mathcal{P}(F)\ar[r] & 1
-\end{tikzcd} \]
-Again by Lemma \ref{5.2} (ii), it suffices to show that $\mathcal{P}(F_1)/\mathcal{P}(F)$ is free-by-finite. Using the snake lemma, we get the following exact sequence:
-\[
-0 \longrightarrow \ker(f)\longrightarrow \mathcal{P}(F_1)/\mathcal{P}(F)\longrightarrow \mathcal{I}(F_1)/\mathcal{I}(F)
-\longrightarrow \operatorname{coker}(f)\longrightarrow 0 \]
-where $f:\operatorname{Cl}(F)\to \operatorname{Cl}(F_1)$.
-Since the class numbers of number fields are finite, according to Lemma \ref{5.2} (i), it suffices to show that $\mathcal{I}(F_1)/\mathcal{I}(F)$ is free-by-finite.
-This follows from the fact that there are only finitely many primes of $F$ which are ramified in $F_1$.\qedhere
-\end{enumerate}
-\end{proof}
-
-\section{Admissible Sets}
-
-\begin{theorem}[\textit{Mordell-Weil Theorem}]\label{mordell-weil-theorem}
-Let $A$ be an abelian variety over a number field $F$. Then the abelian group $A(F)$ of $F$-rational points is finitely generated.
-\end{theorem}
-
-\begin{corollary}\label{mordell-weil-corollary}
-Let $C$ be a smooth projective curve over a number field $F$. Then the Picard group $\operatorname{Pic}(C)$ is finitely generated.
-\end{corollary}
-
-\begin{proof}
-We have the following exact sequence:
-\[ 0\longrightarrow \operatorname{Pic}^0(C) \longrightarrow\operatorname{Pic}(C) \longrightarrow \operatorname{NS}(C) \longrightarrow 0  \]
-where $\operatorname{NS}(C)$ denotes the finitely generated Néron-Severi group of $C$.
-Now $\operatorname{Pic}^0(C)$ can be identified with the $F$-rational points of the Jacobian variety $J(C)$ of $C$.
-$J(C)$ is finitely generated by the Mordell-Weil Theorem \ref{mordell-weil-theorem}, hence $\operatorname{Pic}(C)$ is finitely generated as well.
-\end{proof}
+\section{Proof of Pop's Theorem}
 
@@
 For a finite subset $S\subset X_0$ set:
-\[ K^S=\{x\in K^\times \mid v_P(x)=0\text{ for all }P\in X_0 -  S \} \]
+\[ K^S=\bigcap_{P\in X_0-S}U_P \]
 $K'^{S'}$ is defined similarly. For finite subsets $S$ of $X_0$, the corresponding subsets of $X_0'$ will be denoted by $S'$, i.e. $S'=\{\sigma P\mid P\in S \} $.
 \end{definition}
-
-\begin{theorem}[\textit{Weak Approximation Theorem}]\label{weak-approximation}
-Let $P_1,\ldots,P_n\in X_0$ be pairwise distinct closed points, $x_1,\ldots,x_n\in K$ and $a_1,\ldots,a_n\in\mathbb{Z}$. Then there is some $x\in K$, such that for all $i$ we have:
-\[ v_{P_i}(x-x_i) = a_i \]
-\end{theorem}
-
-\begin{proof}
-See \cite{ZS60} VI §10 Theorem 18.
-\end{proof}
 
@@
 \begin{enumerate}[(i)]
-\item By Corollary \ref{mordell-weil-corollary}, the Picard group of $X$ is finitely generated.
+\item By Mordell-Weil, the Picard group of $X$ is finitely generated.
 Let $D_1,\ldots, D_r$ be Weil divisors of $X$ such that the associated line bundles generate $\operatorname{Pic}(X)$ and similarly let $D'_1,\ldots,D'_s$ generate $\operatorname{Pic}(X')$.
@@
 Now let $T\subset X_0$ be finite.
-By the Weak Approximation Theorem \ref{weak-approximation}, we can add a principal divisor to the $D_i$ and the $D'_j$ so that their support is in $X_0 -  T$ and $X'_0 -  T'$ respectively.
+By the Weak Approximation Theorem, we can add a principal divisor to the $D_i$ and the $D'_j$ so that their support is in $X_0 -  T$ and $X'_0 -  T'$ respectively.
 \item Observe that the following sequence is exact:
@@
 \[ r_{S,T}:K^S \to\prod_{P\in T}k(P)^\times,\ x\mapsto(r_P(x))_{P\in T} \]
-In this case, $\operatorname{coker}(r_{S,T})$ is free-by-finite.
+In this case, $\bigcap_{n\in\mathbb{N}} n\cdot\operatorname{coker}(r_{S,T}) =0$.
 \end{lemma}
@@
 \[ C_1\to\prod_{i=2}^{r} k(P_i)^\times,\ x\mapsto(r_{P_i}(x))_i \]
-Finally for $T=\{P_1,\ldots,P_r\}\subset U_0$, the map $r_{S,T}$ is injective. For the second assertion, note the following isomorphism:
-\[ \operatorname{coker}(r_{S,T}) \cong\operatorname{coker}\Big(K^S/k^\times \to \Big(\prod_{P\in T}k(P)^\times \Big)\Big/k^\times \Big) \]
-Since $K^S/k^\times$ is generated by the uniformizers of the corresponding valuations in $S$, Lemma \ref{5.2} (iii) and Lemma \ref{5.3} (ii) show that the latter group is free-by-finite.
+Finally for $T=\{P_1,\ldots,P_r\}\subset U_0$, the map $r_{S,T}$ is injective.
 \end{proof}
@@
 \end{definition}
-
-\section{Proof of Pop's Theorem}
 
@@
 $\operatorname{im}(\varrho)\subset n\operatorname{coker}(r_{S,T})$ for every $n\in\mathbb{N}$.
-By Lemma~\ref{5.2} (iii) and Lemma~\ref{5.4} it follows $\varrho = 0$.
+It follows $\varrho = 0$.
 By symmetry, $\varphi^{S,T}$ is an isomorphism.
 
-Let $T_1, T_2\subset X_0 -  S$ be two $S$-admissible subsets and suppose $T_1\subset T_2$. By considering $(\star)$ we find:
-\[ r_{S,T_1}\circ\varphi^{S,T_2}=\Big(\prod_{P\in T_1}\varphi_P\Big)\circ r_{S',T_1'} = r_{S,T_1}\circ\varphi^{S,T_1} \]
-Since $r_{S,T_1}$ is injective we get $\varphi^{S,T_2}=\varphi^{S,T_1}$.
-Hence, $\varphi^{S,T}$ is actually independent of $T$, thus we can omit the $T$ from its notation.
-Now if $T$ ranges over all $S$-admissible subsets of $X_0$, then $(\star)$ shows:
+$\varphi^{S,T}$ actually independent of $T$, thus we can omit the $T$ from its notation. Now if $T$ ranges over all $S$-admissible subsets of $X_0$, then $(\star)$ shows:
 \[ r_S\circ\varphi^S = \Big(\prod_{P\in X_0 -  S}\varphi_P \Big)\circ r_{S'} \]
-In other words, we can replace $T$ by $X_0 -  S$ in $(\star)$. If $S_1\subset S_2$ are two admissible subsets, then:
-\[ r_{S_2}\circ\varphi^{S_2}|_{K'^{S'_1}} = \Big(\prod_{P\in X_0 -  S_2}\varphi_P \Big)\circ r_{S'_2}|_{K'^{S'_1}} = r_{S_2}\circ\varphi^{S_1} \]
-Hence, $\varphi^{S_1}=\varphi^{S_2}|_{K'^{S'_1}}$.
+In other words, we can replace $T$ by $X_0 -  S$ in $(\star)$.
+
+We have $\varphi^{S_1}=\varphi^{S_2}|_{K'^{S'_1}}$.
 Thus, we can combine the isomorphisms $\varphi^S$, where $S$ ranges over the admissible subsets of $X_0$, into an isomorphism $\varphi = \varinjlim \varphi^S: K'^\times \to K^\times$.
-
-The commutativity of the diagram follows from the commutativity of $(\star)$ with $T=X_0 -  S$.
-Note that with Lemma \ref{4.5} (i), we can find for all $x\in U_{P'}$ an admissible set $S\subset X_0 -  \{P\}$ such that $x\in K'^{S'}$.
 \end{proof}
@@
 
-Let $g\in G_K$ and $\overline{v}\in\operatorname{Spv}_\text{d}^{(1)}(\overline{K})$. By Remark \ref{3.3} we have:
+Let $g\in G_K$ and $\overline{v}\in\operatorname{Spv}_\text{d}^{(1)}(\overline{K})$. We have:
 \[ \sigma(g)\sigma(\overline{v}) = \sigma(g\overline{v}) = \tau(g\overline{v}) = \tau(g)\tau(\overline{v}) = \tau(g) \sigma(\overline{v}) \]
 Therefore $\sigma(g)^{-1}\tau(g)$ is contained in $G_{\overline{v}}$ for all $\overline{v}\in\operatorname{Spv}_\text{d}^{(1)}(\overline{K})$.
-By Corollary \ref{corollary-fkschmidt} we have $G_{\overline{v}_1}\cap G_{\overline{v}_2} = 1$ for $\overline{v}_1\neq\overline{v}_2$.
+By the Theorem of F.K.
+Schmidt we have $G_{\overline{v}_1}\cap G_{\overline{v}_2} = 1$ for $\overline{v}_1\neq\overline{v}_2$.
 Therefore, $\sigma(g) = \tau(g) = \varphi^{-1}g\varphi$ for all $g\in G_K$.
@@
 for all $g\in G_K$.
-For $\overline{v}\in\operatorname{Spv}_\text{d}^{(1)}(\overline{K})$ we have therefore $G_{\overline{v}} = h G_{\overline{v}} h^{-1} = G_{h\overline{v}}$, thus $h\in G_{\overline{v}}$ by Corollary \ref{corollary-fkschmidt}.
+For $\overline{v}\in\operatorname{Spv}_\text{d}^{(1)}(\overline{K})$ we have therefore $G_{\overline{v}} = h G_{\overline{v}} h^{-1} = G_{h\overline{v}}$, thus $h\in G_{\overline{v}}$ by F.K.
+Schmidt.
 Since $\overline{v}$ was arbitrary, we get $h\in G_{\overline{v}}$ for all $\overline{v}$.
-Again, by Corollary \ref{corollary-fkschmidt}, we have $h=1$ and thus $\varphi = \psi$.
+Again, by the Theorem of F.K.
+Schmidt, we have $h=1$ and thus $\varphi = \psi$.
 \end{proof}
+
+\end{document}

### 2017-05-29 Master Seminar/article.tex
Cut some things

@@
 \begin{definition}\label{2.1}
-Let $K$ be a field of finite dimension.
-We denote the set of equivalence classes of Krull valuations of $K$ by $\operatorname{Spv}(K)$.
-We often do not distinguish between a valuation and its equivalence class.
+Let $K$ be a field of finite dimension. We denote the set of (equivalence classes) of Krull valuations of $K$ by $\operatorname{Spv}(K)$.
 
@@
 
-\begin{lemma}
-$\sigma$ induces a bijection $\sigma:\operatorname{Spv}_\text{rd}^{(2)}(K) \to\operatorname{Spv}_\text{rd}^{(2)}(K')$.
-\end{lemma}
-
 \begin{lemma}\label{3.2}
-$\sigma$ induces a bijection $\sigma:\operatorname{Spv}_\text{d}^{(1)}(K) \to\operatorname{Spv}_\text{d}^{(1)}(K')$.
-If $E=K_v\subset\overline{K}$ is a henselization of $K$ with respect to $v\in\operatorname{Spv}_\text{d}^{(1)}(K)$, then $E'$ is a henselization of $K'$ with respect to $v'=\sigma(v)$.
-
-Furthermore, if $E^{\text{t}}$ and $E'^{\text{t}}$ denote the absolute inertia field of $E$ and $E'$ respectively, then $\sigma$ induces an isomorphism:
+$\sigma$ induces a bijection $\sigma:\operatorname{Spv}_\text{d}^{(1)}(K) \to\operatorname{Spv}_\text{d}^{(1)}(K')$. $\sigma$ induces an isomorphism:
 \[ G_{k(v)}\cong\operatorname{Gal}(E^\text{t}/E) \to\operatorname{Gal}(E'^{\text{t}}/E')\cong G_{k(v')} \]
@@
 \begin{lemma}\label{4.2}
-We have $(K\overline{k})'=K'\overline{k'}$, i.e. $\sigma(G_{K\overline{k}})=G_{K'\overline{k'}}$. $\sigma$ induces an isomorphism $G_k\to G_{k'}$ such that the following diagram commutes:
+$\sigma$ induces an isomorphism $G_k\to G_{k'}$ such that the following diagram commutes:
 \[ \begin{tikzcd}
@@
 \begin{lemma}\phantomsection\label{4.3}
-\begin{enumerate}[(i)]
-\item $\varphi_n$ induces an isomorphism $\varphi_n: U_{P'}/n\to U_P/n$.
-\item The following diagram commutes:
+$\varphi_n$ induces an isomorphism $\varphi_n: U_{P'}/n\to U_P/n$. The following diagram commutes:
 \[ \begin{tikzcd}
 U_{P'}/n \rar["\varphi_n"] \dar["r_{P'}/n"'] & U_P/n \dar["r_P/n"]\\
-k(P')^\times/n \rar["\varphi_n"'] & k(P)^\times/n
+k(P')^\times/n \rar["\varphi_P/n"'] & k(P)^\times/n
 \end{tikzcd} \]
-\end{enumerate}
 \end{lemma}
@@
 Such a set $S$ is called \textit{admissible}. If $T\subset X_0$ is a finite set of closed points, then we can choose $S$ in $X_0 -  T$.
-\item Let $S$ be admissible. For all $n\in\mathbb{N}$ the following canonical maps are injective:
-\[ K^S/n \to K^\times /n,\quad K'^{S'}/n\to K'^\times /n \]
-$\varphi_n$ induces isomorphisms $\varphi_n: K'^{S'}/n\to K^S/n$.
+\item Let $S$ be admissible. For all $n\in\mathbb{N}$ the map $\varphi_n$ induces isomorphisms: \[\varphi_n: K'^{S'}/n\to K^S/n\]
 \end{enumerate}

### 2017-06-03 Master Thesis/synopsis.tex
Fix map label
Fix argument in 2.6 and 3.1

@@
 \begin{proposition}\label{2.6}
-Let $K_0/K$ be an algebraic extension.
-\begin{enumerate}[(i)]
-\item Suppose there exists a prime number $\ell$ with $\ell^\infty\nmid[K_0:K]$ and such that for every algebraic extension $L/K_0$ we have: 
+Let $K_0/K$ be an algebraic extension. Suppose there exists a prime number $\ell\neq 2$ with $\ell^\infty\nmid[K_0:K]$ and such that for every algebraic extension $L/K_0$ we have: 
 \[ \mathrm{H}^3(L,\mathbb{Z}/\ell\mathbb{Z}) = \begin{cases}
@@
 Then there exists a unique henselian $w\in|\operatorname{Spv}(K)|$.
-\item Assume that (i) is satisfied for two different prime numbers. Then $w$ is a defectless valuation.
-\end{enumerate}
 \end{proposition}
@@
 \begin{proof}
-\begin{enumerate}[(i)]
-\item Since $\mathrm{H}^3(K_0, \mathbb{Z}/\ell\mathbb{Z})\neq 0$ by assumption, according to Lemma \ref{2.4} we can find a valuation $w_0\in|\operatorname{Spv}(K_0)|$ such that:
+Since $\mathrm{H}^3(K_0, \mathbb{Z}/\ell\mathbb{Z})\neq 0$ by assumption, according to Lemma \ref{2.4} we can find a valuation $w_0\in|\operatorname{Spv}(K_0)|$ such that:
 \[ \mathrm{H}^3((K_0)_{w_0}, \mathbb{Z}/\ell\mathbb{Z}) \neq 0 \]
@@
 The uniqueness of $w$ is again a consequence of Lemma \ref{2.2}.
-\item Choose the prime $\ell$ different from $\operatorname{char} k(w)$ so that (i) holds. According to Lemma \ref{2.5} we have:
-\[\operatorname{rrank}(w)=\dim_\mathbb{Q}(\Gamma_w\otimes_\mathbb{Z}\mathbb{Q} )\geq \dim_{\mathbb{F}_\ell}(\Gamma_w /\ell\Gamma_w) = 2 = \dim(K)\qedhere\]
-\end{enumerate}
 \end{proof}
@@
 \begin{proof}
-Let $\ell_1,\ell_2$ be two different prime numbers different from $\operatorname{char} k(w)$ and let $E_0 = E(\mu_{\ell_1\ell_2})$.
-Then, by Lemma \ref{2.5}, $E_0/E$, and also $E'_0/E'$, satisfy the condition of Proposition \ref{2.6} (ii).
-Thus there exists a unique henselian defectless valuation $w'_{E'}$ of $E'$, so $E'$ contains a henselization $K'_{w'}$ of $w' = w'_{E'}|_{K'}$.
+Let $\ell_1,\ell_2$ be two different odd prime numbers different from $p=\operatorname{char} k(w)$ and let $E_0 = E(\mu_{\ell_1\ell_2})$.
+Since $K$ is finitely generated, by Lemma \ref{rank-dimension-inequality}, $k(w)$ and $\Gamma_w$ are finitely generated as well.
+In particular we have $\ell_i^\infty \nmid [k(w):\mathbb{F}_p]$ and, since $\Gamma_w$ is torsion free, $\dim_{\mathbb{F}_{\ell_i}}(\Gamma_w/{\ell_i}\Gamma_w) = \dim_{\mathbb{Q}}(\Gamma_w\otimes_{\mathbb{Z}}\mathbb{Q})= \operatorname{rrank}(w)=2$ for $i=1,2$.
+
+By Lemma \ref{2.5}, $E_0/E$, and also $E'_0/E'$, satisfy the condition of Proposition \ref{2.6} for both $\ell_1$ and $\ell_2$.
+Thus there exists a unique henselian valuation $w'_{E'}\in|\operatorname{Spv}(E')|$.
+Choose a prime $\ell\neq \operatorname{char} k(w'_{E'})$ from $\ell_1,\ell_2$.
+Since $\mathrm{H}^3(E',\mathbb{Z}/\ell\mathbb{Z})\cong \mathrm{H}^3(E,\mathbb{Z}/\ell\mathbb{Z})\neq 0$, we get from Lemma \ref{2.5}:
+\[\operatorname{rrank}(w'_{E'}) = \dim_{\mathbb{Q}}(\Gamma_{w'_{E'}}\otimes_\mathbb{Z}\mathbb{Q}) \geq \dim_{\mathbb{F}_\ell}(\Gamma_{w'_{E'}}/\ell\Gamma_{w'_{E'}})=2=\dim(E')\]
+which makes $w'_{E'}\in\operatorname{Spv}^{(2)}_\text{d}(E')$.
+Now $E'$ contains a henselization $K'_{w'}$ of $w' = w'_{E'}|_{K'}\in\operatorname{Spv}^{(2)}_\text{d}(K')$.
 By applying the same argument to $\sigma^{-1}$ we find that $E' = K'_{w'}$.
@@
 U_{P'}/n \rar["\varphi_n"] \dar["r_{P'}/n"'] & U_P/n \dar["r_P/n"]\\
-k(P')^\times/n \rar["\varphi_n"'] & k(P)^\times/n
+k(P')^\times/n \rar["\varphi_P/n"'] & k(P)^\times/n
 \end{tikzcd} \]

### 2017-06-03 Master Seminar/article.tex
More reduction

@@
 
-We can endow $\operatorname{Spv}(K)$ with a natural topology by declaring following sets as a basis of open sets:
-\[\{v\in\operatorname{Spv}(K)\mid v(x)\geq 0\text{ for all }x\in S\} \]
-where $S$ ranges over all finite subsets of $K$. Let $v,w\in\operatorname{Spv}(K)$. We have:
-\[\mathcal{O}_w\subset\mathcal{O}_v \iff w\in\overline{\{v\}} \]
-Thus $v$ is a closed point if and only if it has no non-trivial specializations.
+\[|\operatorname{Spv}(K)|=\{v\in\operatorname{Spv}(K)\mid \dim k(v)=0\}\]
 \end{definition}
@@
 \begin{lemma}\label{rank-dimension-inequality}
-Let $v\in\operatorname{Spv}(K)$. Then we have:
-\begin{enumerate}[(i)]
-\item $\operatorname{rank}(v)\leq \operatorname{rrank}(v)$
-\item $\operatorname{rrank}(v)+\dim k(v)\leq\dim(K)$
-\end{enumerate}
-Moreover, if equality holds in (ii) and $K$ is finitely generated, then $k(v)$ is finitely generated as well, and $\Gamma_v$ is a finitely generated $\mathbb{Z}$-module.
+Let $v\in\operatorname{Spv}(K)$.
+Then we have $\operatorname{rrank}(v)+\dim k(v)\leq\dim(K)$.
+If equality holds and $K$ is finitely generated, then $k(v)$ is finitely generated as well, and $\Gamma_v$ is a finitely generated $\mathbb{Z}$-module.
 \end{lemma}
@@
 \begin{align*}
-\operatorname{Spv}_\text{d}(K) &= \{v\in\operatorname{Spv}(K)\mid \operatorname{rrank}(v)+\dim k(v)=\dim(K) \} \\
-\operatorname{Spv}_\text{rd}(K) &= \{v\in\operatorname{Spv}_\text{d}(K)\mid \operatorname{rank}(v)=\operatorname{rrank}(v) \}
+\operatorname{Spv}_\text{d}(K) &= \{v\in\operatorname{Spv}(K)\mid \operatorname{rrank}(v)+\dim k(v)=\dim(K) \}
 \end{align*}
-Valuations in $\operatorname{Spv}_\text{d}(K)$ are called \textit{defectless}\index{valuation!defectless}, whereas those in $\operatorname{Spv}_\text{rd}(K)$ are called \textit{rank-defectless}\index{valuation!rank-defectless}.
-For $\bullet\in\{\text{d},\text{rd}\}$ we set:
-\[\operatorname{Spv}_\bullet^{(i)}(K)=\{v\in\operatorname{Spv}_\bullet(K)\mid \operatorname{rrank}(v)=i \} \]
+Valuations in $\operatorname{Spv}_\text{d}(K)$ are called \textit{defectless}\index{valuation!defectless}.
+\[\operatorname{Spv}_\text{d}^{(i)}(K)=\{v\in\operatorname{Spv}_\text{d}(K)\mid \operatorname{rrank}(v)=i \} \]
 \end{definition}
-
-\section{Main Lemma}
 
@@
 Let $K$ be henselian with respect to $w\in|\operatorname{Spv}(K)|$.
-Set $k=k(w)$ and $\Gamma = \Gamma_w$.
-Then $k$ is an algebraic extension of a finite field $\mathbb{F}_p$.
+Then $k(w)$ is an algebraic extension of a finite field $\mathbb{F}_p$.
 Let $\ell\neq p$ be a prime and assume $\mu_\ell\subset K$.
@@
 \[ \mathrm{H}^3(K, \mathbb{Z}/\ell\mathbb{Z}) = \begin{cases}
-\mathbb{Z}/\ell\mathbb{Z}, & \ell^\infty\nmid [k:\mathbb{F}_p] \text{ and } \dim_{\mathbb{F}_\ell}(\Gamma / \ell\Gamma) = 2\\
+\mathbb{Z}/\ell\mathbb{Z}, & \ell^\infty\nmid [k(w):\mathbb{F}_p] \text{ and } \dim_{\mathbb{F}_\ell}(\Gamma_w / \ell\Gamma_w) = 2\\
 0, & \text{otherwise}
@@
 Then there exists a unique henselian $w\in|\operatorname{Spv}(K)|$.
-\item Assume that (i) is satisfied for two different prime numbers. Then $w$ is a defectless valuation.
-\end{enumerate}
 \end{proposition}
-
-\begin{proof}
-\begin{enumerate}[(i)]
-\item Since $\mathrm{H}^3(K_0, \mathbb{Z}/\ell\mathbb{Z})\neq 0$ by assumption, according to Lemma \ref{2.4} we can find a valuation $w_0\in|\operatorname{Spv}(K_0)|$ such that:
-\[ \mathrm{H}^3((K_0)_{w_0}, \mathbb{Z}/\ell\mathbb{Z}) \neq 0 \]
-By requirement $\ell^\infty \nmid [(K_0)_{w_0} : K_0]$. By replacing $K_0$ by $(K_0)_{w_0}$, we can assume that $w_0$ is henselian.
-
-Assume that $w = w_0|_K$ is not henselian.
-Then there exists a $w_1\in|\operatorname{Spv}(K_0)|$ with $w_1\neq w_0$ and such that $w = w_1|_K$, for otherwise $w$ would have a unique extension to $K_0$ and hence to $\overline{K}$ since $w_0$ is henselian.
-
-Let $(K_0)_{w_1}$ be a henselization of $K_0$ with respect to $w_1$ and let $K_w$ be a henselization of $K$ with respect to $w$ in $(K_0)_{w_1}$.
-Then $(K_0)_{w_1}$ is the composite of $K_0$ and $K_w$ and we have $\ell^\infty\nmid [(K_0)_{w_1} : K_w]$.
-
-Since $K_0$ is henselian with respect to $w_0$, a conjugate $\tau(K_w)$ for some $\tau\in G_K$ of $K_w$ lies in $K_0$, hence $\ell^\infty\nmid [K_w : K]$ as well.
-Consequently, $\ell^\infty \nmid [(K_0)_{w_1} : K]$ and therefore $\ell^\infty \nmid [(K_0)_{w_1} : K_0]$.
-By requirement, $\mathrm{H}^3((K_0)_{w_1}, \mathbb{Z}/\ell\mathbb{Z}) \neq 0$.
-
-Now $(K_0)_{w_1}$ is henselian both with respect to $w_1$ and with respect to the extension of $w_0$. A contradiction to $\operatorname{cd}(G_{(K_0)_{w_1}})\leq 2$.
-\item Choose the prime $\ell$ different from $\operatorname{char} k(w)$ so that (i) holds. According to Lemma \ref{2.5} we have:
-\[\operatorname{rrank}(w)=\dim_\mathbb{Q}(\Gamma_w\otimes_\mathbb{Z}\mathbb{Q} )\geq \dim_{\mathbb{F}_\ell}(\Gamma_w /\ell\Gamma_w) = 2 = \dim(K)\qedhere\]
-\end{enumerate}
-\end{proof}
 
@@
 
-\begin{proof}
-Let $U = X -  S$ and $P_1\in U$ be a closed point. We set:
-\[ C_1 = \ker(r_{P_1}|_{K^S}: K^S \to k(P_1)^\times) \]
-Since $r_{P_1}|_{k^\times}:k^\times \to k(P_1)^\times$ is injective, we have $C_1\cap k^\times = 1$ and the following map is injective:
-\[ C_1 \to\prod_{P\in S}\mathbb{Z},\ x \mapsto (v_P(x))_{P\in S} \]
-Therefore, $C_1$ is a free abelian group of finite rank.
-If $C_1 = 1$, then we set $T = \{P_1\}$ and we are done.
-Otherwise, let $x\in C_1,\ x\neq 1$.
-Then there exists an $a\in k^\times$ with the following properties:
-\begin{enumerate}[(i)]
-\item There exists a point $P_2\in X_0 - (S\cup\{P_1\})$ which lies above the prime of $k(x)$ corresponding to the polynomial $x-a\in k[x]$.
-\item $a$ is not a root of unity.
-\end{enumerate}
-Clearly, (i) and (ii) hold for almost all $a\in k^\times$.
-We set $C_2 = \ker(r_{P_2}|_{C_1})\subset C_1$.
-$C_2$ is free abelian of finite rank as well.
-By requirement $r_{P_2}(x)=a$ and $a\in k^\times$ is not of finite order, thus $x\not\in C_2$ and $\operatorname{rank}(C_2) < \operatorname{rank}(C_1)$.
-By repeating this process until $C_r=1$, we get points $P_2,\ldots,P_r\in U_0$, such that the following map is injective:
-\[ C_1\to\prod_{i=2}^{r} k(P_i)^\times,\ x\mapsto(r_{P_i}(x))_i \]
-Finally for $T=\{P_1,\ldots,P_r\}\subset U_0$, the map $r_{S,T}$ is injective.
-\end{proof}
-
-\begin{definition}
-For each admissible $S\subset X_0$ and $T_1\subset X_0 -  S$ with injective $r_{S,T_1}$ we also have the injectivity of $r_{S,T_2}$ for every $T_2 \supset T_1$.
-Therefore,  we can find a finite subset $T\subset X_0 -  S$ such that both $r_{S,T}$ and $r_{S',T'}$ are injective.
-Such a $T$ will be called $S$-\textit{admissible} and $(S,T)$ is called \textit{admissible pair}.
-\end{definition}
-
 \begin{theorem}[\textit{Pop}]\label{1.1}
@@
 \begin{proof}
-By Lemma \ref{5.4}, the following maps are injective for all admissible subsets $S\subset X_0$:
-\[ r_S:K^S \to\prod_{P\in X_0 -  S} k(P)^\times,\ x\mapsto (r_P(x))_P \]
-Now it's evident that $\varphi$ is unique.
-For existence, let $(S,T)$ be an admissible pair.
+Let $S\subset X_0$ be admissible and $T\subset X_0-S$ such that $r_{S,T}$ and $r_{S',T'}$ are injective.
 First, we show that there is a unique homomorphism $\varphi^{S,T}$, making the following diagram commutative:
@@
 
-$\varphi^{S,T}$ actually independent of $T$, thus we can omit the $T$ from its notation. Now if $T$ ranges over all $S$-admissible subsets of $X_0$, then $(\star)$ shows:
-\[ r_S\circ\varphi^S = \Big(\prod_{P\in X_0 -  S}\varphi_P \Big)\circ r_{S'} \]
-In other words, we can replace $T$ by $X_0 -  S$ in $(\star)$.
-
-We have $\varphi^{S_1}=\varphi^{S_2}|_{K'^{S'_1}}$.
-Thus, we can combine the isomorphisms $\varphi^S$, where $S$ ranges over the admissible subsets of $X_0$, into an isomorphism $\varphi = \varinjlim \varphi^S: K'^\times \to K^\times$.
+$\varphi^{S,T}$ actually independent of $T$, thus we can omit the $T$ from its notation.
+Now if $T$ ranges over all $S$-admissible subsets of $X_0$, we can replace $T$ by $X_0 -  S$ in $(\star)$.
+We can combine the isomorphisms $\varphi^S$, where $S$ ranges over the admissible subsets of $X_0$, into an isomorphism $\varphi = \varinjlim \varphi^S: K'^\times \to K^\times$.
 \end{proof}
@@
 \begin{remark}
-For any finite subextension $L/K$ of $\overline{K}/K$ there exists an isomorphism $\varphi_L: L'\to L$.
-If $K\subset L\subset M$ are finite subextensions, we have $\varphi_M|_{L'} = \varphi_{L'}$, since $r_Q(\varphi_M(x)) = r_Q(\varphi_L(x))$ for all $x\in L'^\times$ and almost all $Q\in (X_M)_0$ according to Remark \ref{3.3} and Step 1.
-By passing to limits we obtain an isomorphism $\varphi: \overline{K}'\to \overline{K}$.
+For any finite subextension $L/K$ of $\overline{K}/K$ there exists an isomorphism $\varphi_L: L'\to L$. By passing to limits we obtain an isomorphism $\varphi: \overline{K}'\to \overline{K}$.
 \end{remark}

### 2017-06-04 Master Thesis/synopsis.tex
Add bibliography entry for O. Endler's valuation theory

@@
 
-The second assertion follows from the first one, applied to $L$, by using the following equation of supernatural numbers ([E], Theorem 20.21):
+The second assertion follows from the first one, applied to $L$, by using the following equation of supernatural numbers (\cite{En72}, Theorem 20.21):
 \[ [L:K] = (\Gamma':\Gamma) \cdot [k':k]\cdot p^d \]
 for some $d \leq \infty$, where $\Gamma'$ denotes the value group and $k'$ the residue field of the unique prolongation of $w$ to $L$.
-Note that $\operatorname{rank}(w) = \operatorname{rank}(v)$ in an algebraic field extension (\cite{ZS60}, Ch.
-VI, §11, Lemma 2).
+Note that $\operatorname{rrank}(w) = \operatorname{rrank}(v)$ in an algebraic field extension (\cite{ZS60}, Ch.
+VI, §11, Lemma 1).
 \end{proof}

### 2017-06-04 Master Thesis/synopsis.tex
Clarify uniqueness of phi in proof of Pop's Theorem

@@
 For a finite subset $S\subset X_0$ set:
-\[ K^S=\{x\in K^\times \mid v_P(x)=0\text{ for all }P\in X_0 -  S \} \]
+\[ K^S=\{x\in K^\times \mid v_P(x)=0\text{ for all }P\in X_0 -  S \} = \bigcap_{P\in X_0 -  S} U_P \]
 $K'^{S'}$ is defined similarly. For finite subsets $S$ of $X_0$, the corresponding subsets of $X_0'$ will be denoted by $S'$, i.e. $S'=\{\sigma P\mid P\in S \} $.
@@
 \[ r_S:K^S \to\prod_{P\in X_0 -  S} k(P)^\times,\ x\mapsto (r_P(x))_P \]
-Now it's evident that $\varphi$ is unique.
-For existence, let $(S,T)$ be an admissible pair.
-First, we show that there is a unique homomorphism $\varphi^{S,T}$, making the following diagram commutative:
+Using the desired commutativity of the diagram above and, with Lemma \ref{4.5} (i), the fact that for all $x\in K^\times$ there is an admissible set $S\subset X_0$ such that $x\in K^S$, it's evident that $\varphi$ is unique.
+
+For the existence of $\varphi$, let $(S,T)$ be an admissible pair. First, we show that there is a unique homomorphism $\varphi^{S,T}$, making the following diagram commutative:
 \[ (\star) \qquad \begin{tikzcd}
@@
 
-The commutativity of the diagram follows from the commutativity of $(\star)$ with $T=X_0 -  S$.
+The commutativity of the desired diagram follows from the commutativity of $(\star)$ with $T=X_0 -  S$.
 Note that with Lemma \ref{4.5} (i), we can find for all $x\in U_{P'}$ an admissible set $S\subset X_0 -  \{P\}$ such that $x\in K'^{S'}$.

### 2017-06-04 Master Thesis/synopsis.tex
Update

@@
 \[ 0\longrightarrow\mathbb{Z}_\ell^r \longrightarrow G_\ell \longrightarrow G_{k,\ell}\longrightarrow 1 \]
-Since $G_{k,\ell}$ is a free pro-$\ell$ group, this sequence splits and we have an isomorphism $G_\ell \cong \mathbb{Z}_\ell^r\rtimes G_{k,\ell}$ with $r =\dim_{\mathbb{F}_\ell}(\Gamma/\ell\Gamma)\leq \operatorname{rrank}(\Gamma)\leq 2$.
+Since $G_{k,\ell}$ is a free pro-$\ell$ group, this sequence splits and we have an isomorphism $G_\ell \cong \mathbb{Z}_\ell^r\rtimes G_{k,\ell}$ with $r =\dim_{\mathbb{F}_\ell}(\Gamma/\ell\Gamma)\leq \operatorname{rrank}(\Gamma)\leq \dim(K)= 2$.
 
@@
 since $\operatorname{cd}(\mathbb{Z}_\ell)=1$.
-Thus $\mathrm{H}^3(G_\ell, \mathbb{Z}/\ell\mathbb{Z}) =0$ if $G_k(\ell) = 0$ or $r < 2$.
+Thus $\mathrm{H}^3(G_\ell, \mathbb{Z}/\ell\mathbb{Z}) =0$ iff $G_k(\ell) = 0$ or $r < 2$.
 Now let $G_k(\ell) \cong\mathbb{Z}_\ell$ and $r = 2$.
@@
 
-We will see that the bijection $\operatorname{Spv}_\text{d}^{(2)}(K)\to \operatorname{Spv}_\text{d}^{(2)}(K')$ above restricts to a bijection $\operatorname{Spv}_\text{rd}^{(2)}(K)\to \operatorname{Spv}_\text{rd}^{(2)}(K')$.
-Using this, we shall establish a bijection between the defectless valuations of rational rank $1$:
+This lemma yields a bijection $\sigma:\operatorname{Spv}_\text{d}^{(2)}(K) \to\operatorname{Spv}_\text{d}^{(2)}(K')$.
+We will see that this bijection restricts to a bijection $\operatorname{Spv}_\text{rd}^{(2)}(K)\to \operatorname{Spv}_\text{rd}^{(2)}(K')$.
+With that, we shall establish a bijection between the defectless valuations of rational rank $1$:
 
@@
 
-Let $w_1,w_2 \in\operatorname{Spv}_\text{d}^{(2)}(K)$ be two different valuations. The following statements are equivalent:
+Let $w_1,w_2 \in\operatorname{Spv}_\text{d}^{(2)}(K)$ be two different valuations. By Lemma \ref{2.2}, the following statements are equivalent:
 \begin{enumerate}[(i)]
@@
 \end{enumerate}
-The implication (i)$\implies$(ii) follows from Lemma \ref{2.2}. TODO
-
-In this case, $k(v')$ is separably closed where $v'$ is the unique extension of the valuation on $K_v$ to $K_{w_1}K_{w_2}$ where $v=v(w_1)$.
+If (ii) holds, $k(v)$ is separably closed where $v=v(w_1)$.
 By Lemma \ref{lemma-henselization-specialization} we have $K_{w_1}K_{w_2}\subset K_v^\text{t}$ and by Theorem \ref{theorem-absoluteinertia} (ii) $K_{w_1}K_{w_2}$ is the absolute inertia field $K_v^\text{t}$ of $K_v$.
@@
 \[ v(w_1)=v(w_2) \iff v(\sigma(w_1)) = v(\sigma(w_2)) \]
-This shows that $\sigma$ also induces a bijection $\sigma:\operatorname{Spv}_\text{d}^{(1)}(K)\to\operatorname{Spv}_\text{d}^{(1)}(K')$ and for all $w\in\operatorname{Spv}_\text{rd}^{(2)}(K)$ we have $\sigma(v(w)) = v(\sigma(w))$.
+Let $v\in\operatorname{Spv}_\text{d}^{(1)}(K)$.
+Since $v\not\in |\operatorname{Spv}(K)|$, there exists a $w\in\operatorname{Spv}_\text{rd}^{(2)}(K)$ such that $v(w)=v$.
+We define $v'=v(\sigma(w))$.
+This shows that $\sigma$ also induces a bijection:
+\[\sigma:\operatorname{Spv}_\text{d}^{(1)}(K)\to\operatorname{Spv}_\text{d}^{(1)}(K'),\ v\mapsto v'\] 
+This map is well-defined and for all $w\in\operatorname{Spv}_\text{rd}^{(2)}(K)$ we have $\sigma(v(w)) = v(\sigma(w))$.
 
@@
 Since a global field has no non-trivial algebraic extension in which every prime splits completely, we get $N=G_{K_v}$.
-This shows that $K'_{v'} := (K_v)'$ is a henselization of $K'$ with respect to $v'=\sigma(v)$.
+This shows that $(K_v)'$ is a henselization of $K'$ with respect to $v'=\sigma(v)$.
+We write $K'_{v'} = (K_v)'$.
 Furthermore, let $w_1,w_2\in\operatorname{Spv}_\text{rd}^{(2)}(K)$ with $w_1\neq w_2$ and $v=v(w_1)=v(w_2)$, then:
@@
 \end{tikzcd} \]
-By Dirichlet's Unit Theorem, $\mathcal{O}_F^\times$ is free-by-finite.
+By Dirichlet's Unit Theorem, $\mathcal{O}_F^\times$ is finitely generated and thus free-by-finite.
 $\mathcal{P}(F)$, as a subset of the ideal group of $F$, is free abelian, in particular free-by-finite.

### 2017-06-07 Master Thesis/synopsis.tex
Fix 3.2 argumentation

@@
 \item $w$ has value group $\Gamma$ and residue class field $k(v)^\text{s}$.
-\item $K^\text{t}$ is a Galois extension of $K$ and $\operatorname{Gal}(K^\text{t}/K)\cong G_{k(v)}$ as profinite groups. There exists an inclusion-preserving one-to-one correspondence: 
+\item $K^\text{t}$ is a Galois extension of $K$ and $\operatorname{Gal}(K^\text{t}/K)\cong G_{k(v)}$ as profinite groups. This isomorphism induces an inclusion-preserving one-to-one correspondence: 
 \begin{align*}
@@
 
-This lemma yields a bijection $\sigma:\operatorname{Spv}_\text{d}^{(2)}(K) \to\operatorname{Spv}_\text{d}^{(2)}(K')$.
+This lemma yields a bijection $\sigma:\operatorname{Spv}_\text{d}^{(2)}(K) \to\operatorname{Spv}_\text{d}^{(2)}(K'),\ w\mapsto w'$.
 We will see that this bijection restricts to a bijection $\operatorname{Spv}_\text{rd}^{(2)}(K)\to \operatorname{Spv}_\text{rd}^{(2)}(K')$.
@@
 If $v$ is nontrivial, then in this case we get $w\in\operatorname{Spv}_\text{rd}^{(2)}(K)$ and $v=v(w)\in\operatorname{Spv}_\text{d}^{(1)}(K)$.
-We set $K'_{\sigma(w)} = (K_w)'$ for $w\in\operatorname{Spv}_\text{d}^{(2)}(K)$.
-Lemma \ref{3.1} shows that $K'_{\sigma(w)}$ is a henselization of $K'$ with respect to $\sigma(w)$.
+We set $K'_{w'} = (K_w)'$ for $w\in\operatorname{Spv}_\text{d}^{(2)}(K)$.
+Lemma \ref{3.1} shows that $K'_{w'}$ is a henselization of $K'$ with respect to $w'$.
 
@@
 By Lemma \ref{lemma-henselization-specialization} we have $K_{w_1}K_{w_2}\subset K_v^\text{t}$ and by Theorem \ref{theorem-absoluteinertia} (ii) $K_{w_1}K_{w_2}$ is the absolute inertia field $K_v^\text{t}$ of $K_v$.
-Furthermore, $K'_{\sigma(w_1)}K'_{\sigma(w_2)} = (K_{w_1}K_{w_2})' \neq \overline{K'}$ and again by Lemma \ref{2.2} we get: \[\sigma(w_1),\sigma(w_2)\in\operatorname{Spv}_\text{rd}^{(2)}(K)\quad \text{and} \quad v(\sigma(w_1)) = v(\sigma(w_2))\]
+Furthermore, $K'_{w'_1}K'_{w'_2} = (K_{w_1}K_{w_2})' \neq \overline{K'}$ and again by Lemma \ref{2.2} we get: \[w'_1,w'_2\in\operatorname{Spv}_\text{rd}^{(2)}(K)\quad \text{and} \quad v(w'_1) = v(w'_2)\]
 Thus $\sigma: \operatorname{Spv}_\text{d}^{(2)}(K)\to \operatorname{Spv}_\text{d}^{(2)}(K')$ maps $\operatorname{Spv}_\text{rd}^{(2)}(K)$ onto $\operatorname{Spv}_\text{rd}^{(2)}(K')$ such that:
-\[ v(w_1)=v(w_2) \iff v(\sigma(w_1)) = v(\sigma(w_2)) \]
+\[ v(w_1)=v(w_2) \iff v(w'_1) = v(w'_2) \]
 Let $v\in\operatorname{Spv}_\text{d}^{(1)}(K)$.
 Since $v\not\in |\operatorname{Spv}(K)|$, there exists a $w\in\operatorname{Spv}_\text{rd}^{(2)}(K)$ such that $v(w)=v$.
-We define $v'=v(\sigma(w))$.
+We define $v'=v(w')$.
 This shows that $\sigma$ also induces a bijection:
@@
 
-Let $v\in\operatorname{Spv}_\text{d}^{(1)}(K)$ and let $N$ be the normal subgroup of $G_{K_v}$ generated by all subgroups $G_{K_w}$ for $w\in\operatorname{Spv}_\text{rd}^{(2)}(K)$ with $v(w) = v$.
-Since a global field has no non-trivial algebraic extension in which every prime splits completely, we get $N=G_{K_v}$.
-This shows that $(K_v)'$ is a henselization of $K'$ with respect to $v'=\sigma(v)$.
+Let $v\in\operatorname{Spv}_\text{d}^{(1)}(K)$ and $w\in\operatorname{Spv}_\text{rd}^{(2)}(K)$ with $v(w)=v$.
+Then $k(v)$ is finitely generated of dimension $1$ by Lemma \ref{rank-dimension-inequality}, thus a global field.
+By Theorem \ref{theorem-absoluteinertia}, we have:
+\[G_{K_w}/H \cong \operatorname{Gal}(K_v^\text{t}/K_w)\cong G_{k(v_w)}\] 
+where $v_w$ is the unique extension of $v$ on $K_w$ and $H=G_{K_v^\text{t}}$.
+With Lemma \ref{2.2-pre} we can see that $k(v_w)$ is the henselization of $k(v)$ with respect to $w\mod v\in\operatorname{Spv}k(v)$.
+Let $N$ be the closed subgroup of $G_{K_v}$ generated by all subgroups $G_{K_w}$ for $w\in\operatorname{Spv}_\text{rd}^{(2)}(K)$ with $v(w) = v$.
+Since a global field has no non-trivial algebraic extension in which every prime splits completely, we get $N/H= G_{K_v}/H\cong G_{k(v)}$, thus $N=G_{K_v}$.
+Since $(K_w)'=K'_{w'}$, this shows that $(K_v)'$ is a henselization of $K'$ with respect to $v'=\sigma(v)$.
 We write $K'_{v'} = (K_v)'$.
+
 Furthermore, let $w_1,w_2\in\operatorname{Spv}_\text{rd}^{(2)}(K)$ with $w_1\neq w_2$ and $v=v(w_1)=v(w_2)$, then:

### 2017-06-07 Master Thesis/synopsis.tex
Refactor

@@
 By Theorem \ref{theorem-absoluteinertia}, we have:
-\[G_{K_w}/H \cong \operatorname{Gal}(K_v^\text{t}/K_w)\cong G_{k(v_w)}\] 
-where $v_w$ is the unique extension of $v$ on $K_w$ and $H=G_{K_v^\text{t}}$.
+\[G_{K_w}/N \cong \operatorname{Gal}(K_v^\text{t}/K_w)\cong G_{k(v_w)}\] 
+where $v_w$ is the unique extension of $v$ on $K_w$ and $N=G_{K_v^\text{t}}$.
 With Lemma \ref{2.2-pre} we can see that $k(v_w)$ is the henselization of $k(v)$ with respect to $w\mod v\in\operatorname{Spv}k(v)$.
-Let $N$ be the closed subgroup of $G_{K_v}$ generated by all subgroups $G_{K_w}$ for $w\in\operatorname{Spv}_\text{rd}^{(2)}(K)$ with $v(w) = v$.
-Since a global field has no non-trivial algebraic extension in which every prime splits completely, we get $N/H= G_{K_v}/H\cong G_{k(v)}$, thus $N=G_{K_v}$.
+Let $H$ be the closed subgroup of $G_{K_v}$ generated by all subgroups $G_{K_w}$ for $w\in\operatorname{Spv}_\text{rd}^{(2)}(K)$ with $v(w) = v$.
+Since a global field has no non-trivial algebraic extension in which every prime splits completely, we get $H/N= G_{K_v}/H\cong G_{k(v)}$, thus $H=G_{K_v}$.
 Since $(K_w)'=K'_{w'}$, this shows that $(K_v)'$ is a henselization of $K'$ with respect to $v'=\sigma(v)$.

### 2017-06-08 Master Thesis/synopsis.tex
Abuse of notation notice
Fix Lemma 3.1

@@
 Furthermore, $\overline{v}$ has the same value group as $v$ and the same residue field $k(\overline{v}) = k(v)$.
+By abuse of notation, we will also write $v$ for $\overline{v}$.
 $G_{K_v}$ is the decomposition group of the unique extension of $\overline{v}$ to $K^\text{s}$:
@@
 Thus there exists a unique henselian valuation $w'_{E'}\in|\operatorname{Spv}(E')|$.
-Choose a prime $\ell\neq \operatorname{char} k(w'_{E'})$ from $\ell_1,\ell_2$.
-Since $\mathrm{H}^3(E',\mathbb{Z}/\ell\mathbb{Z})\cong \mathrm{H}^3(E,\mathbb{Z}/\ell\mathbb{Z})\neq 0$, we get from Lemma \ref{2.5}:
-\[\operatorname{rrank}(w'_{E'}) = \dim_{\mathbb{Q}}(\Gamma_{w'_{E'}}\otimes_\mathbb{Z}\mathbb{Q}) \geq \dim_{\mathbb{F}_\ell}(\Gamma_{w'_{E'}}/\ell\Gamma_{w'_{E'}})=2=\dim(E')\]
-which makes $w'_{E'}\in\operatorname{Spv}^{(2)}_\text{d}(E')$.
-Now $E'$ contains a henselization $K'_{w'}$ of $w' = w'_{E'}|_{K'}\in\operatorname{Spv}^{(2)}_\text{d}(K')$.
+Now $E'$ contains a henselization $K'_{w'}$ with respect to $w' = w'_{E'}|_{K'}$.
 By applying the same argument to $\sigma^{-1}$ we find that $E' = K'_{w'}$.
+
+It remains to see that $\operatorname{rrank}(w')=2$.
+Choose a prime $\ell\neq \operatorname{char} k(w')$ from $\ell_1,\ell_2$.
+Since $\mathrm{H}^3(E'_0(\mu_{\ell}),\mathbb{Z}/\ell\mathbb{Z})\neq 0$ by Lemma \ref{2.6}, we get from Lemma \ref{2.5}:
+\[\operatorname{rrank}(w')= \operatorname{rrank}( \widetilde{w}') = \dim_{\mathbb{Q}}(\Gamma_{\widetilde{w}'}\otimes_\mathbb{Z}\mathbb{Q}) \geq \dim_{\mathbb{F}_\ell}(\Gamma_{\widetilde{w}'}/\ell\Gamma_{\widetilde{w}'})=2\]
+where $\widetilde{w}'$ is the extension of $w'_{E'}$ to $E'_0(\mu_{\ell})$. This makes $w'\in\operatorname{Spv}^{(2)}_\text{d}(K')$.
 

### 2017-06-08 Master Thesis/synopsis.tex
Finally fix proof of 3.2!

@@
 
-Let $w_1,w_2 \in\operatorname{Spv}_\text{d}^{(2)}(K)$ be two different valuations. By Lemma \ref{2.2}, the following statements are equivalent:
+Let $w_1,w_2 \in\operatorname{Spv}_\text{d}^{(2)}(K)$ be two different valuations. Then the following statements are equivalent:
 \begin{enumerate}[(i)]
@@
 \end{enumerate}
-If (ii) holds, $k(v)$ is separably closed where $v=v(w_1)$.
-By Lemma \ref{lemma-henselization-specialization} we have $K_{w_1}K_{w_2}\subset K_v^\text{t}$ and by Theorem \ref{theorem-absoluteinertia} (ii) $K_{w_1}K_{w_2}$ is the absolute inertia field $K_v^\text{t}$ of $K_v$.
-Furthermore, $K'_{w'_1}K'_{w'_2} = (K_{w_1}K_{w_2})' \neq \overline{K'}$ and again by Lemma \ref{2.2} we get: \[w'_1,w'_2\in\operatorname{Spv}_\text{rd}^{(2)}(K)\quad \text{and} \quad v(w'_1) = v(w'_2)\]
+The implication (i)$\implies$(ii) follows from Lemma \ref{2.2}.
+Now assume (ii) holds.
+Then $k(v)$ is separably closed where $v=v(w_1)$.
+By Lemma \ref{lemma-henselization-specialization} we have $K_{w_1}K_{w_2}\subset K_v^\text{t}$ and by Theorem \ref{theorem-absoluteinertia} (ii) $K_{w_1}K_{w_2}=K_v^\text{t}$.
+The extension of $v$ on $K_v^\text{t}$ has value group $\Gamma_v$, which is finitely generated.
+However, value groups of valuations on an algebraically closed field are divisible, hence $K_v^\text{t}$ can't be algebraically closed.
+
+Furthermore, in this case, $K'_{w'_1}K'_{w'_2} = (K_{w_1}K_{w_2})' \neq \overline{K'}$ and again by Lemma \ref{2.2} we get: \[w'_1,w'_2\in\operatorname{Spv}_\text{rd}^{(2)}(K)\quad \text{and} \quad v(w'_1) = v(w'_2)\]
 Thus $\sigma: \operatorname{Spv}_\text{d}^{(2)}(K)\to \operatorname{Spv}_\text{d}^{(2)}(K')$ maps $\operatorname{Spv}_\text{rd}^{(2)}(K)$ onto $\operatorname{Spv}_\text{rd}^{(2)}(K')$ such that:

### 2017-06-08 Master Thesis/synopsis.tex
Fix typo

@@
 Then $k(v)$ is separably closed where $v=v(w_1)$.
-By Lemma \ref{lemma-henselization-specialization} we have $K_{w_1}K_{w_2}\subset K_v^\text{t}$ and by Theorem \ref{theorem-absoluteinertia} (ii) $K_{w_1}K_{w_2}=K_v^\text{t}$.
+By Lemma \ref{lemma-henselization-specialization} we have $K_{w_1}K_{w_2}\subset K_v^\text{t}$ and by Theorem \ref{theorem-absoluteinertia} $K_{w_1}K_{w_2}=K_v^\text{t}$.
 The extension of $v$ on $K_v^\text{t}$ has value group $\Gamma_v$, which is finitely generated.
@@
 
-Furthermore, in this case, $K'_{w'_1}K'_{w'_2} = (K_{w_1}K_{w_2})' \neq \overline{K'}$ and again by Lemma \ref{2.2} we get: \[w'_1,w'_2\in\operatorname{Spv}_\text{rd}^{(2)}(K)\quad \text{and} \quad v(w'_1) = v(w'_2)\]
+Furthermore, in case (ii), $K'_{w'_1}K'_{w'_2} = (K_{w_1}K_{w_2})' \neq \overline{K'}$ follows from (i) and again by Lemma \ref{2.2} we get: \[w'_1,w'_2\in\operatorname{Spv}_\text{rd}^{(2)}(K)\quad \text{and} \quad v(w'_1) = v(w'_2)\]
 Thus $\sigma: \operatorname{Spv}_\text{d}^{(2)}(K)\to \operatorname{Spv}_\text{d}^{(2)}(K')$ maps $\operatorname{Spv}_\text{rd}^{(2)}(K)$ onto $\operatorname{Spv}_\text{rd}^{(2)}(K')$ such that:
@@
 Let $H$ be the closed subgroup of $G_{K_v}$ generated by all subgroups $G_{K_w}$ for $w\in\operatorname{Spv}_\text{rd}^{(2)}(K)$ with $v(w) = v$.
-Since a global field has no non-trivial algebraic extension in which every prime splits completely, we get $H/N= G_{K_v}/H\cong G_{k(v)}$, thus $H=G_{K_v}$.
+Since a global field has no non-trivial algebraic extension in which every prime splits completely, we get $H/N= G_{K_v}/N\cong G_{k(v)}$, thus $H=G_{K_v}$.
 Since $(K_w)'=K'_{w'}$, this shows that $(K_v)'$ is a henselization of $K'$ with respect to $v'=\sigma(v)$.
@@
 
-Furthermore, let $w_1,w_2\in\operatorname{Spv}_\text{rd}^{(2)}(K)$ with $w_1\neq w_2$ and $v=v(w_1)=v(w_2)$, then:
+Now let $w_1,w_2\in\operatorname{Spv}_\text{rd}^{(2)}(K)$ with $w_1\neq w_2$ and $v=v(w_1)=v(w_2)$. We have:
 \[ \sigma(G_{K^\text{t}_v}) = \sigma(G_{K_{w_1}}\cap G_{K_{w_2}}) = G_{K'_{w'_1}}\cap G_{K'_{w'_2}} = G_{(K')^\text{t}_{v'}} \]

### 2017-06-08 Master Thesis/synopsis.tex
Complete argumentation in remark 3.3
Prefer ' notation instead over sigma notation

@@
 Therefore, $f$ is injective as well and its image is contained in $\bigoplus\mathrm{H}^3(K_y)$.
-Now, one easily checks that:
+Now, one checks that:
 \[ \operatorname{supp}(f) = \bigcap_{\mathfrak{X}\in \mathfrak{M}(K)} \pi_\mathfrak{X}^{-1}(\{ y\in\mathfrak{X}_0\mid \operatorname{res}_{K_y/K}(f)\neq 0 \}) \]
@@
 Then we have $\sigma(G_{\overline{v}}) = G_{\sigma(\overline{v})}$.
-Now it is easy to see that for every $g\in G_K$:
-\[ \sigma(g\overline{v}) = \sigma(g)\sigma(\overline{v}) \]
+It follows for any $g\in G_K$:
+\[G_{\sigma(g\overline{v})} = \sigma(G_{g\overline{v}}) = \sigma(g^{-1} G_{\overline{v}}g) = \sigma(g)^{-1} G_{\sigma(\overline{v})}\sigma(g) = G_{\sigma(g)\sigma(\overline{v})}\]
+Therefore, $\sigma(g\overline{v})=\sigma(g)\sigma(\overline{v})$.
 \end{remark}
@@
 
-Before we start proving Pop's Theorem \ref{thm:main-result}, we need some tools that will help us constructing a group isomorphism $K'^\times\to K^\times$ from the modulo $n$ isomorphisms $K'^\times/n\to K^\times/n$.
+Before we start proving Pop's Theorem \ref{thm:main-result}, we need some tools that will help us constructing a group isomorphism $K'^\times\to K^\times$ from the modulo $n$ isomorphisms $\varphi_n: K'^\times/n\to K^\times/n$.
 
@@
 K'^{S'} \rar["\varphi^{S,T}", dashed] \dar["r_{S',T'}"'] & K^S \dar["r_{S,T}"]\\
-\displaystyle \prod_{P\in T}k(\sigma P)^\times \rar["\prod\varphi_P"'] & \displaystyle \prod_{P\in T}k(P)^\times
+\displaystyle \prod_{P\in T}k(P')^\times \rar["\prod\varphi_P"'] & \displaystyle \prod_{P\in T}k(P)^\times
 \end{tikzcd} \phantom{\qquad (\star)} \]
 The vertical maps are injective by definition. Thus, it suffices to show that the following composition is trivial:
-\[\varrho:K'^{S'}\longrightarrow\prod_{P\in T} k(\sigma P)^\times \longrightarrow \prod_{P\in T} k(P)^\times\longrightarrow\operatorname{coker}(r_{S,T})\] 
+\[\varrho:K'^{S'}\longrightarrow\prod_{P\in T} k(P')^\times \longrightarrow \prod_{P\in T} k(P)^\times\longrightarrow\operatorname{coker}(r_{S,T})\] 
 Consider the commutative diagram which results from tensoring $(\star)$ with $\mathbb{Z}/n\mathbb{Z}$:
@@
 K'^{S'}/n \rar[dashed]\dar & K^S/n\dar\\
-\displaystyle \prod_{P\in T}k(\sigma P)^\times/n \rar & \displaystyle \prod_{P\in T}k(P)^\times/n
+\displaystyle \prod_{P\in T}k(P')^\times/n \rar & \displaystyle \prod_{P\in T}k(P)^\times/n
 \end{tikzcd} \]

### 2017-06-11 Master Thesis/synopsis.tex
Update

@@
 
-Assume that $w = w_0|_K$ is not henselian.
+For the sake of contradiction, assume that $w = w_0|_K$ is not henselian.
 Then there exists a $w_1\in|\operatorname{Spv}(K_0)|$ with $w_1\neq w_0$ and such that $w = w_1|_K$, for otherwise $w$ would have a unique extension to $K_0$ and hence to $\overline{K}$ since $w_0$ is henselian.
 
-Let $(K_0)_{w_1}$ be a henselization of $K_0$ with respect to $w_1$ and let $K_w$ be a henselization of $K$ with respect to $w$ in $(K_0)_{w_1}$.
+Let $(K_0)_{w_1}$ be a henselization of $K_0$ with respect to $w_1$ and let $K_w\subset (K_0)_{w_1}$ be a henselization of $K$ with respect to $w$.
 Then $(K_0)_{w_1}$ is the composite of $K_0$ and $K_w$ and we have $\ell^\infty\nmid [(K_0)_{w_1} : K_w]$.
@@
 \[G_{\sigma(g\overline{v})} = \sigma(G_{g\overline{v}}) = \sigma(g^{-1} G_{\overline{v}}g) = \sigma(g)^{-1} G_{\sigma(\overline{v})}\sigma(g) = G_{\sigma(g)\sigma(\overline{v})}\]
-Therefore, $\sigma(g\overline{v})=\sigma(g)\sigma(\overline{v})$.
+Therefore, $\sigma(g\overline{v})=\sigma(g)\sigma(\overline{v})$ by Corollary \ref{corollary-fkschmidt}.
 \end{remark}
@@
 We denote the natural projection $U_P\to k(P)^\times,\ x\mapsto x\mod P$ by $r_P$.
+We shall also write $K_P$ for a henselization of $K$ with respect to $v_P$.
 
@@
 By Remark \ref{remark-localphi-extension}, without loss of generality, we can assume $K=K_P$.
-Since $\operatorname{char}k(P)=0$, by Hensel's lemma, $r_P/n$ is an isomorphism and the composition
+Since $\operatorname{char}k(P)=0$, by Hensel's lemma the map $r_P/n: U_P/n\to k(P)^\times/n$ is an isomorphism and the composition
 \[\begin{tikzcd}

### 2017-06-15 Master Thesis/synopsis.tex
Add big diagram for snake lemma

@@
 For intermediate fields $L$ of $\overline{K}/K$ we denote the corresponding intermediate field of $\overline{K'}/K'$ by $L'$.
-In other words, $\sigma G_L=G_{L'}$.
+In other words, $\sigma (G_L)=G_{L'}$.
 
@@
 \end{tikzcd} \]
-Again by Lemma \ref{5.2} (ii), it suffices to show that $\mathcal{P}(F_1)/\mathcal{P}(F)$ is free-by-finite. Using the snake lemma, we get the following exact sequence:
+Again by Lemma \ref{5.2} (ii), it suffices to show that $\mathcal{P}(F_1)/\mathcal{P}(F)$ is free-by-finite. 
+Consider the following diagram with exact rows and columns:
+\[ \begin{tikzcd}
+& 0 \ar[d] & 0 \ar[d] & \ker(f) \ar[d]& \\
+0\ar[r] & \mathcal{P}(F) \ar[d] \ar[r]& \mathcal{I}(F) \ar[r]\ar[d]& \operatorname{Cl}(F)\ar[r]\ar[d, "f"] & 0\\
+0 \ar[r]& \mathcal{P}(F_1)\ar[d] \ar[r]& \mathcal{I}(F_1) \ar[d]\ar[r]& \operatorname{Cl}(F_1)\ar[d] \ar[r]& 0 \\
+& \mathcal{P}(F_1)/\mathcal{P}(F) & \mathcal{I}(F_1)/\mathcal{I}(F) & \operatorname{coker}(f) &
+\end{tikzcd} \]
+Using the snake lemma, we get the following exact sequence:
 \[
@@
 \longrightarrow \operatorname{coker}(f)\longrightarrow 0 \]
-where $f:\operatorname{Cl}(F)\to \operatorname{Cl}(F_1)$.
 Since the class numbers of number fields are finite, according to Lemma \ref{5.2} (i), it suffices to show that $\mathcal{I}(F_1)/\mathcal{I}(F)$ is free-by-finite.
@@
 Clearly any finite $S\subset X_0$ such that
-\[ \bigcup_{i=1}^r \operatorname{supp}D_i\subset S\quad\text{and}\quad \bigcup_{j=1}^s\operatorname{supp}D'_j\subset S' \]
+\[ \bigcup_{i=1}^r \operatorname{supp}(D_i)\subset S\quad\text{and}\quad \bigcup_{j=1}^s\operatorname{supp}(D'_j)\subset S' \]
 is admissible.

### 2017-06-16 Master Thesis/synopsis.tex
Mention Lemma 2.3 (i)
Reformulation
Conjugation of henselizations

@@
 Therefore, $f$ is injective as well and its image is contained in $\bigoplus\mathrm{H}^3(K_y)$.
-Now, one checks that:
+Now, Lemma \ref{2.3} (i) yields:
 \[ \operatorname{supp}(f) = \bigcap_{\mathfrak{X}\in \mathfrak{M}(K)} \pi_\mathfrak{X}^{-1}(\{ y\in\mathfrak{X}_0\mid \operatorname{res}_{K_y/K}(f)\neq 0 \}) \]
@@
 
-Let $F$ be another henselization of $w$. Then $E, F$ are $K$-isomorphic, hence $E',F'$ are $K'$-isomorphic and the corresponding henselian valuations of $E', F'$ respectively are mapped to each other.
+Let $F$ be another henselization of $w$.
+Then $E, F$ are $K$-conjugate, hence $E',F'$ are $K'$-conjugate, thus $K'$-isomorphic and the corresponding henselian valuations of $E', F'$ are respectively mapped to each other.
 \end{proof}
@@
 \end{tikzcd} \]
-Moreover, $\sigma:G_k\to G_{k'}$ is induced by a unique isomorphism $\varphi:\overline{k'}\to\overline{k}$.
+Moreover, $\sigma:G_k\to G_{k'}$ is induced by a unique isomorphism $\varphi:\overline{k'}\to\overline{k}$ which restricts to an isomorphism $\varphi: k'\to k$.
 \end{lemma}
@@
 \begin{remark}
-Consider $\overline{K}'' = \overline{K}'\otimes_{\overline{k'}}\overline{k}$ via the isomorphism $\varphi: \overline{k'}\to \overline{k}$. If we replace $\sigma: G_K\to G_{K'}$ with the composition
-\[ \begin{tikzcd}
-G_K \ar[r, "\sigma"] & G_{K'} \ar[r, "\phi"] & G_{K'}
+Consider $\overline{K''} = \overline{K}'\otimes_{\overline{k'}}\overline{k}$ via the isomorphism $\varphi: \overline{k'}\to \overline{k}$. If we replace $\sigma: G_K\to G_{K'}$ with the composition
+\[ \begin{tikzcd}
+G_K \ar[r, "\sigma"] & G_{K'} \ar[r, "\phi", "\cong"'] & \operatorname{Gal}(\overline{K''}/K')
 \end{tikzcd} \]
@@
 \begin{enumerate}[(i)]
-\item $\varphi_n$ induces an isomorphism $\varphi_n: U_{P'}/n\to U_P/n$.
+\item $\varphi_n$ restricts to an isomorphism $\varphi_n: U_{P'}/n\to U_P/n$.
 \item The following diagram commutes:

### 2017-06-20 Master Thesis/synopsis.tex
Remove variable U from 5.4

@@
 \begin{proof}
-Let $U = X -  S$ and $P_1\in U$ be a closed point. We set:
+Let $P_1\in X_0-S$. We set:
 \[ C_1 = \ker(r_{P_1}|_{K^S}: K^S \to k(P_1)^\times) \]
@@
 By requirement $r_{P_2}(x)=a$ and $a\in k^\times$ is not of finite order, thus $x\not\in C_2$ and $\operatorname{rank}(C_2) < \operatorname{rank}(C_1)$.
-By repeating this process until $C_r=1$, we get points $P_2,\ldots,P_r\in U_0$, such that the following map is injective:
+By repeating this process until $C_r=1$, we get points $P_2,\ldots,P_r\in X_0-S$, such that the following map is injective:
 \[ C_1\to\prod_{i=2}^{r} k(P_i)^\times,\ x\mapsto(r_{P_i}(x))_i \]
-Finally for $T=\{P_1,\ldots,P_r\}\subset U_0$, the map $r_{S,T}$ is injective. For the second assertion, note the following isomorphism:
+Finally for $T=\{P_1,\ldots,P_r\}\subset X_0-S$, the map $r_{S,T}$ is injective. For the second assertion, note the following isomorphism:
 \[ \operatorname{coker}(r_{S,T}) \cong\operatorname{coker}\Big(K^S/k^\times \to \Big(\prod_{P\in T}k(P)^\times \Big)\Big/k^\times \Big) \]

### 2017-06-25 Master Thesis/synopsis.tex
Fix typos

@@
 
-In this section we shall show that $\sigma: G_K\to G_{K'}$ induces group isomorphisms $K'^\times/n\to K^\times/n$ and $U_{v'}/n\to U_v/n$ for all $n\in\mathbb{N}$ and defectless valuations $v$ of rational rank $1$ with $\operatorname{char}k(v) =0$.
+In this section we shall show that $\sigma: G_K\to G_{K'}$ induces group isomorphisms $\varphi_n: K'^\times/n\to K^\times/n$.
 
@@
 Let $X$ and $X'$ be the unique smooth projective curves over $k$ and $k'$ with function field $K$ and $K'$ respectively.
-We identify the set of closed points $X_0$ of $X$ with the set of valuations $v\in\operatorname{Spv}_\text{d}^{(1)}(K)$ with $\operatorname{char} k(v)=0$ and similarly identify $X'_0$ with the corresponding subset of $\operatorname{Spv}_\text{d}^{(1)}(K')$.
+We identify the set of closed points $X_0$ of $X$ with the set of discrete valuations $v$ with $v(k^\times)=0$.
+That is exactly all $v\in\operatorname{Spv}_\text{d}^{(1)}(K)$ with $\operatorname{char} k(v)=0$.
+Similarly, we identify $X'_0$ with the corresponding subset of $\operatorname{Spv}_\text{d}^{(1)}(K')$.
 
@@
 \begin{remark}
-Consider $\overline{K''} = \overline{K}'\otimes_{\overline{k'}}\overline{k}$ via the isomorphism $\varphi: \overline{k'}\to \overline{k}$. If we replace $\sigma: G_K\to G_{K'}$ with the composition
+Consider $\overline{K''} = \overline{K'}\otimes_{\overline{k'}}\overline{k}$ via the isomorphism $\varphi: \overline{k'}\to \overline{k}$. If we replace $\sigma: G_K\to G_{K'}$ with the composition
 \[ \begin{tikzcd}
@@
 
-Let $n\in\mathbb{N}$. Tensoring the exact sequence $1\to U_P\to K^\times \to\mathbb{Z}\to 0$ with $\mathbb{Z}/n\mathbb{Z}$ preserves its exactness since $\mathbb{Z}$ is $\mathbb{Z}$-flat:
+Tensoring the exact sequence $1\to U_P\to K^\times \to\mathbb{Z}\to 0$ with $\mathbb{Z}/n\mathbb{Z}$ preserves its exactness since $\mathbb{Z}$ is $\mathbb{Z}$-flat:
 \[ 1\longrightarrow U_P/n\longrightarrow K^\times/n\longrightarrow\mathbb{Z}/n\mathbb{Z}\longrightarrow 0 \]
-We'll use the same notation for $P' = \sigma(P)\in X'_0$.
+Thus we can consider $U_P/n$ as a subgroup of $K^\times/n$. We'll use the same notation for $P' = \sigma(P)\in X'_0$.
 

### 2017-06-26 Master Thesis/synopsis.tex
Recall Kummer theory facts

@@
 
-Let $K$ be a (not necessarily finitely generated) field over $\mathbb{Q}$ and of dimension $2$.
-
-\begin{lemma}\phantomsection\label{lemma-defectless-valuations}
-\begin{enumerate}[(i)]
-\item $\operatorname{Spv}_\text{d}^{(1)}(K) = \operatorname{Spv}_\text{rd}^{(1)}(K)$
-\item $\operatorname{Spv}^{(2)}(K) = \operatorname{Spv}_\text{d}^{(2)}(K)\subset |\operatorname{Spv}(K)|$
-\end{enumerate}
-\end{lemma}
-
-\begin{proof}
-\begin{enumerate}[(i)]
-\item Let $v$ be a defectless valuation of rational rank $1$.
-By Lemma \ref{rank-dimension-inequality}, $\operatorname{rank}(v) \leq \operatorname{rrank}(v) = 1$.
-Since $\Gamma_v$ is not trivial, $\operatorname{rank}(v) = 1=\operatorname{rrank}(v)$.
-\item Let $v$ be a valuation of rational rank $2$. Due to Lemma \ref{rank-dimension-inequality}, the dimension inequality is an equality and we have $\dim k(v) = 0$. \qedhere
-\end{enumerate}
-\end{proof}
+Let $K$ be a (not necessarily finitely generated) field over $\mathbb{Q}$ and of dimension $2$. First, let us recall some facts from Galois cohomology:
+
+\begin{proposition}[\textit{Kummer Theory}]\label{prop:kummer-theory}
+Let $F$ be a field and $n\in\mathbb{N}$ invertible in $F$. Let $\mu_n\subset F^\text{s}$ be the group of all $n$-th roots of unity. Consider the following short exact sequence:
+\[ 1\longrightarrow \mu_n\longrightarrow (F^\text{s})^\times \stackrel{n}{\longrightarrow} (F^\text{s})^\times \longrightarrow 1 \]
+This gives rise to the following long exact cohomology sequence:
+\[\begin{tikzcd}
+\cdots\longrightarrow F^\times \ar[r, "n"]& F^\times \ar[r]& \mathrm{H}^1(F, \mu_n)\ar[r] & \mathrm{H}^1(F, (F^\text{s})^\times)\ar[dlll]\\
+\mathrm{H}^1(F, (F^\text{s})^\times)\ar[r] & \mathrm{H}^2(F, \mu_n)\ar[r] & \operatorname{Br}(F)\ar[r, "n"] & \operatorname{Br}(F) \longrightarrow \cdots
+\end{tikzcd}\]
+Since $\mathrm{H}^1(F, (F^\text{s})^\times)=0$ by Hilbert's Theorem 90, we get isomorphisms:
+\begin{align*}
+\mathrm{H}^1(F,\mu_n) \cong F^\times/n, \qquad \mathrm{H}^2(F, \mu_n) \cong {}_n\operatorname{Br}(F)
+\end{align*}
+\end{proposition}
 

### 2017-06-28 Master Thesis/synopsis.tex
Better phrasing

@@
 
-We denote the valuation ring of $v\in\operatorname{Spv}(K)$ by $(\mathcal{O}_v,\mathfrak{m}_v)$ and set $U_v=\mathcal{O}_v^\times$ and let $k(v)=\mathcal{O}_v/\mathfrak{m}_v$ be its residue field.
+We denote the valuation ring of $v\in\operatorname{Spv}(K)$ by $(\mathcal{O}_v,\mathfrak{m}_v)$, let $U_v=\mathcal{O}_v^\times$ be the unit group of $v$, and $k(v)=\mathcal{O}_v/\mathfrak{m}_v$ be its residue field.
 The \textit{rank} of $v$ is defined as the Krull dimension of its valuation ring $\mathcal{O}_v$.
@@
 \begin{definition}
-We can endow $\operatorname{Spv}(K)$ with a natural topology by declaring following sets as a basis of open sets:
 \[\{v\in\operatorname{Spv}(K)\mid v(x)\geq 0\text{ for all }x\in S\} \]
+We can endow $\operatorname{Spv}(K)$ with a natural topology by declaring the following sets as a basis of open sets:
 where $S$ ranges over all finite subsets of $K$. 
@@
 
-Let $K$ be a (not necessarily finitely generated) field over $\mathbb{Q}$ and of dimension $2$. First, let us recall some facts from Galois cohomology:
+First, let us recall some facts from Galois cohomology:
 
@@
 
+\paragraph{} Now let $K$ be a (not necessarily finitely generated) field over $\mathbb{Q}$ of dimension $2$. In this section, we want to give a cohomological characterization of $K$ being henselian.
+
 \begin{lemma}\label{2.2}
@@
 Now we calculate the cohomology groups $\mathrm{H}^3(K_w,\mathbb{Z}/\ell\mathbb{Z})$ for $w\in |\operatorname{Spv}(K)|$.
-With the help of Lemma \ref{2.4} we can find a cohomological criterion of henselian fields over $K$ with respect to a $w\in\operatorname{Spv}_\text{d}^{(2)}(K)$.
+With the help of our Main Lemma \ref{2.4} we can find a cohomological criterion of henselian fields over $K$ with respect to a $w\in |\operatorname{Spv}(K)|$.
 
@@
 Since $\operatorname{res}\circ\operatorname{inf}$ is an isomorphism, $\operatorname{inf}$ is injective and $\operatorname{res}$ is surjective.
-We have $\operatorname{cor}\circ\operatorname{res} = [G_K:G_\ell]$.
-However, $[G_K:G_\ell]$ is coprime to $\ell$, which makes $\operatorname{res}$ injective as well.
+We have $\operatorname{cor}\circ\operatorname{res} = (G_K:G_\ell)$.
+However, $(G_K:G_\ell)$ is coprime to $\ell$, which makes $\operatorname{res}$ injective as well.
 Therefore, $\operatorname{inf}$ and $\operatorname{res}$ are both isomorphisms and it suffices to calculate $\mathrm{H}^3(G_\ell,\mathbb{Z}/\ell\mathbb{Z})$.

### 2017-07-12 Master Thesis/synopsis.tex
Fix typos

@@
 
-Before we start proving Pop's Theorem \ref{thm:main-result}, we need some tools that will help us constructing a group isomorphism $K'^\times\to K^\times$ from the modulo $n$ isomorphisms $\varphi_n: K'^\times/n\to K^\times/n$.
+Before we start proving Pop's Theorem \ref{thm:main-result}, we need some tools that will help us to construct a group isomorphism $K'^\times\to K^\times$ from the modulo $n$ isomorphisms $\varphi_n: K'^\times/n\to K^\times/n$.
 
@@
 
-Let $T_1, T_2\subset X_0 -  S$ be two $S$-admissible subsets and suppose $T_1\subset T_2$. By considering $(\star)$ we find:
+Let $T_1, T_2\subset X_0 -  S$ be two $S$-admissible subsets and suppose $T_1\subset T_2$. By inspecting $(\star)$ we find:
 \[ r_{S,T_1}\circ\varphi^{S,T_2}=\Big(\prod_{P\in T_1}\varphi_P\Big)\circ r_{S',T_1'} = r_{S,T_1}\circ\varphi^{S,T_1} \]

### 2017-07-15 Master Thesis/synopsis.tex
Add more bibliography items

@@
 
+\paragraph{} Let $K$ be a field of finite dimension.
+
 \begin{definition}\label{2.1}
-Let $K$ be a field of finite dimension.
-We denote the set of equivalence classes of Krull valuations of $K$ by $\operatorname{Spv}(K)$.
-We often do not distinguish between a valuation and its equivalence class.
+We denote the set of equivalence classes of Krull valuations of $K$ by $\operatorname{Spv}(K)$. We often do not distinguish between a valuation and its equivalence class.
 
@@
 \begin{definition}
-\[\{v\in\operatorname{Spv}(K)\mid v(x)\geq 0\text{ for all }x\in S\} \]
 We can endow $\operatorname{Spv}(K)$ with a natural topology by declaring the following sets as a basis of open sets:
+\[U_S =\{v\in\operatorname{Spv}(K)\mid v(x)\geq 0\text{ for all }x\in S\} \]
 where $S$ ranges over all finite subsets of $K$. 
@@
 \begin{lemma}\label{spv-topology}
-$\operatorname{Spv}(K)$ with the topology defined above is a quasi-compact $T_0$-space.
-In particular the set of closed points $|\operatorname{Spv}(K)|$ of $\operatorname{Spv}(K)$ is a quasi-compact $T_1$-space, endowed with the induced topology.
-Let $v,w\in\operatorname{Spv}(K)$.
-We have:
+Let $v,w\in\operatorname{Spv}(K)$. We have:
 \[\mathcal{O}_w\subset\mathcal{O}_v \iff w\in\overline{\{v\}} \]
 Thus $v$ is a closed point if and only if it has no non-trivial specializations.
-\end{lemma}
-
-\begin{proof}
-See \cite{ZS60} VI §17 Theorem 38, §17 Theorem 40.
+
+$\operatorname{Spv}(K)$ is a quasi-compact Jacobson space, i.e.
+every closed subset $A\subset\operatorname{Spv}(K)$ is the closure of its closed points.
+In particular, the set of closed points $|\operatorname{Spv}(K)|$ of $\operatorname{Spv}(K)$ is a quasi-compact $T_1$-space, endowed with the induced topology.
+\end{lemma}
+
+\begin{proof}
+The first assertion follows from \cite{ZS60} VI, §17 Theorem 38. The quasi-compactness of $\operatorname{Spv}(K)$ follows from \cite{ZS60} VI, §17 Theorem 40.
+
+Let $A\subset \operatorname{Spv}(K)$ be a closed set, $S\subset K$ finite, and assume $U_S\cap A\neq\varnothing$. Let $v\in U_S\cap A$ be a point which is not closed.
 \end{proof}
@@
 
-\begin{proposition}[\textit{Kummer Theory}]\label{prop:kummer-theory}
+\begin{proposition}\label{prop:kummer-theory}
 Let $F$ be a field and $n\in\mathbb{N}$ invertible in $F$. Let $\mu_n\subset F^\text{s}$ be the group of all $n$-th roots of unity. Consider the following short exact sequence:
@@
 
-The second assertion follows from the first and from [K2], §5, Theorem 1.
+The second assertion follows from the first and from \cite{Ka79}, §5, Theorem 1.
 \end{proof}
@@
 \begin{proof}
-(i) and (ii) follow from [HK], Lemma 2.4.3 and (iii) is a consequence of (ii) and the fact that henselizations commute with filtrated direct limits, see \cite{EGA4} 18.6.14.
+(i) and (ii) follow from \cite{HK94}, Lemma 2.4.3, and (iii) is a consequence of (ii) and the fact that henselizations commute with filtrated direct limits, e.g. see \cite{EGA4} 18.6.14.
 \end{proof}
@@
 
-In this section we shall show that $\sigma: G_K\to G_{K'}$ induces group isomorphisms $\varphi_n: K'^\times/n\to K^\times/n$.
-
+In this section we shall show that $\sigma: G_K\to G_{K'}$ induces group isomorphisms $\varphi_n: K'^\times/n\to K^\times/n$. Due to Proposition \ref{prop:kummer-theory}, we have isomorphisms:
+\[ \mathrm{H}^1(K',\mu_n)\cong K'^\times/n,\qquad \mathrm{H}^1(K,\mu_n)\cong K^\times/n \]
+Our goal is to establish an isomorphism between the corresponding cohomology groups.
+
+\begin{definition}
 Let $k$ and $k'$ be the algebraic closures of $\mathbb{Q}$ in $K$ and $K'$ respectively, i.e.
@@
 By Lemma \ref{3.2} $\sigma$ induces a bijection $\sigma: X^{\phantom{'}}_0\to X'_0,\ P\mapsto P'$ and for any $P\in X_0$ an isomorphism $\varphi_P: k(P')\to k(P)$.
+\end{definition}
 
@@
 For a finite subset $S\subset X_0$ set:
-\[ K^S=\{x\in K^\times \mid v_P(x)=0\text{ for all }P\in X_0 -  S \} = \bigcap_{P\in X_0 -  S} U_P \]
+\begin{align*}
+K^S &= \bigcap_{P\in X_0 -  S} U_P
+\end{align*}
 $K'^{S'}$ is defined similarly. For finite subsets $S$ of $X_0$, the corresponding subsets of $X_0'$ will be denoted by $S'$, i.e. $S'=\{\sigma P\mid P\in S \} $.
@@
 \begin{theorem}[\textit{Weak Approximation Theorem}]\label{weak-approximation}
-Let $P_1,\ldots,P_n\in X_0$ be pairwise distinct closed points, $x_1,\ldots,x_n\in K$ and $a_1,\ldots,a_n\in\mathbb{Z}$. Then there is some $x\in K$, such that for all $i$ we have:
-\[ v_{P_i}(x-x_i) = a_i \]
+Let $v_1,\ldots,v_n\in X_0$ be inequivalent discrete valuations, $x_1,\ldots,x_n\in K$ and $a_1,\ldots,a_n\in\mathbb{Z}$. Then there is some $x\in K$, such that for all $i$ we have:
+\[ v_i(x-x_i) = a_i \]
 \end{theorem}

### 2017-07-15 Master Thesis/synopsis.tex
Phrasing

@@
 \[ \mathrm{H}^3(K)\longrightarrow \prod_{x\in\mathfrak{X}_1}\mathrm{H}^3(K_{v_x}) \]
-is injective and its image is contained in $\bigoplus \mathrm{H}^3(K_{v_x})$. By \cite{Ka86} Lemma 1.4, for a discrete valuation $v$ of $K$ the map
+is injective and, in fact, the proof shows that its image is contained in $\bigoplus \mathrm{H}^3(K_{v_x})$. By \cite{Ka86} Lemma 1.4, for a discrete valuation $v$ of $K$ the map
 \[ \mathrm{H}^3(K_v) \longrightarrow \prod_w\mathrm{H}^3(K_w) \]
@@
 
-In this section we shall show that $\sigma: G_K\to G_{K'}$ induces group isomorphisms $\varphi_n: K'^\times/n\to K^\times/n$. Due to Proposition \ref{prop:kummer-theory}, we have isomorphisms:
+Let $n\in\mathbb{N}$.
+In this section we shall show that $\sigma: G_K\to G_{K'}$ induces group isomorphisms $\varphi_n: K'^\times/n\to K^\times/n$.
+Due to Proposition \ref{prop:kummer-theory}, we have isomorphisms:
 \[ \mathrm{H}^1(K',\mu_n)\cong K'^\times/n,\qquad \mathrm{H}^1(K,\mu_n)\cong K^\times/n \]
@@
 Now let $T\subset X_0$ be finite.
-By the Weak Approximation Theorem \ref{weak-approximation}, we can add a principal divisor to the $D_i$ and the $D'_j$ so that their support is in $X_0 -  T$ and $X'_0 -  T'$ respectively.
+By the Weak Approximation Theorem \ref{weak-approximation}, we can add a principal divisor to the $D_i$ and the $D'_j$ so that their support lies in $X_0 -  T$ and $X'_0 -  T'$ respectively.
 \item Observe that the following sequence is exact:

### 2017-07-21 Master Thesis/synopsis.tex
Complete proof of 2.2

@@
 
-The second assertion follows from the first and from \cite{Ka79}, §5, Theorem 1.
+For the second assertion, let $K$ be not algebraically closed and $v=v(w_1)=v(w_2)\in\operatorname{Spv}_\text{d}^{(1)}(K)$.
+The restriction $v|$ of $v$ on $k=\mathbb{Q}(T)\subset K$ is a discrete valuation.
+Choose a henselization $k_{v|}$ of $k$ such that $k_{v|}\subset K$ and consider the absolute inertia field $k_{v|}^\text{t}$ of $k_{v|}$.
+Since $k(v)$ is separably closed, we have $k_{v|}^\text{t}\subset K$.
+Since $v|\in \operatorname{Spv} ({k_{v|}^\text{t}})$ is discrete and henselian, \cite{Ka79} §5, Theorem 1 shows $\operatorname{cd}(k_{v|}^\text{t})\leq 2$.
+In particular, $\operatorname{cd}(K)\leq 2$.
 \end{proof}

### 2017-07-21 Master Thesis/synopsis.tex
Update Saito reference

@@
 \[ \mathrm{H}^3(K)\longrightarrow \prod_{x\in\mathfrak{X}_1}\mathrm{H}^3(K_{v_x}) \]
-is injective and, in fact, the proof shows that its image is contained in $\bigoplus \mathrm{H}^3(K_{v_x})$. By \cite{Ka86} Lemma 1.4, for a discrete valuation $v$ of $K$ the map
+is injective, and, in fact, the proof shows that its image is contained in $\bigoplus \mathrm{H}^3(K_{v_x})$. By \cite{Ka86} Lemma 1.4, for a discrete valuation $v$ of $K$ the map
 \[ \mathrm{H}^3(K_v) \longrightarrow \prod_w\mathrm{H}^3(K_w) \]
@@
 \end{tikzcd} \]
-According to [Sa], Theorem (2-15) $g$ is injective.
+According to \cite{Sa87}, Theorem 5.3, $g$ is injective.
 Now, $\psi\circ\varphi$ can be identified with the injective map $\Phi$ from the proof of the previous Lemma.

### 2017-07-23 Master Thesis/synopsis.tex
Finish second part of Lemma 2.2

@@
 $K^\text{t}$ is the fixed field of the absolute inertia group $T$:
-\[ T = \{ \sigma\in G_K\mid \sigma(x)\equiv x\mod \mathfrak{m}_{\overline{v}} \quad \text{for all }x\in\mathcal{O}_{\overline{v}} \}\lhd G_K \]
+\[ T = \{ \sigma\in G_K\mid \sigma(x)\equiv x\mod \mathfrak{m}_{\overline{v}} \quad \text{for all }x\in\mathcal{O}_{\overline{v}} \}\]
 Then we have:
@@
 Since $k(v)$ is separably closed, we have $k_{v|}^\text{t}\subset K$.
-Since $v|\in \operatorname{Spv} ({k_{v|}^\text{t}})$ is discrete and henselian, \cite{Ka79} §5, Theorem 1 shows $\operatorname{cd}(k_{v|}^\text{t})\leq 2$.
-In particular, $\operatorname{cd}(K)\leq 2$.
+Since $v|\in \operatorname{Spv} ({k_{v|}^\text{t}})$ is discrete and henselian, \cite{Ka79} §5, Theorem 1 shows $\operatorname{cd}\big(G_{k_{v|}^\text{t}}\big)\leq 2$.
+In particular, $\operatorname{cd}(G_K)\leq 2$.
 \end{proof}
@@
 \[ 1\longrightarrow T\longrightarrow G_K\longrightarrow G_k\longrightarrow 1 \]
-
 Let $G_\ell$ be an $\ell$-Sylow group of $G_K$.
 By Theorem \ref{theorem-sylowgroups-inertiagroup}, the $\ell$-Sylow group of $T$ isomorphic to $\mathbb{Z}_\ell^r$, where $r =\dim_{\mathbb{F}_\ell}(\Gamma/\ell\Gamma)$.
-Furthermore, $G_k\subset G_{\mathbb{F}_p}\cong\hat{\mathbb{Z}}$, thus the $\ell$-Sylow group $G_{k,\ell}$ of $G_k$ is either isomorphic to $\mathbb{Z}_\ell$ or trivial.
+Furthermore, $G_k\subset G_{\mathbb{F}_p}\cong\widehat{\mathbb{Z}}$, thus the $\ell$-Sylow group $G_{k,\ell}$ of $G_k$ is either isomorphic to $\mathbb{Z}_\ell$ or trivial.
 The exact sequence above gives rise to the short exact sequence of their $\ell$-Sylow subgroups:
@@
 By \cite{Se64}, Ch.
-I, §4, Proposition 22 we have:
+I, §4, Proposition 22 we have
 \[\operatorname{cd}(G_\ell) = \operatorname{cd}(\mathbb{Z}_\ell^r\rtimes G_k(\ell)) = \begin{cases}
@@
 
-The second assertion follows from the first one, applied to $L$, by using the following equation of supernatural numbers (\cite{En72}, Theorem 20.21):
+The second assertion follows from the first one, applied to $L$, by inspecting the following equation of supernatural numbers (\cite{En72}, Theorem 20.21):
 \[ [L:K] = (\Gamma':\Gamma) \cdot [k':k]\cdot p^d \]
 for some $d \leq \infty$, where $\Gamma'$ denotes the value group and $k'$ the residue field of the unique prolongation of $w$ to $L$.
-Note that $\operatorname{rrank}(w) = \operatorname{rrank}(v)$ in an algebraic field extension (\cite{ZS60}, Ch.
-VI, §11, Lemma 1).
+By using this equation, it suffices to show the following equivalences:
+\begin{enumerate}[(i)]
+\item $\ell^\infty \nmid [k':k] \iff \ell^\infty \nmid [k':\mathbb{F}_p]$
+\item $\ell^\infty\nmid (\Gamma':\Gamma) \iff \dim_{\mathbb{F}_\ell}(\Gamma'/\ell\Gamma')=2$
+\end{enumerate}
+(i) is clear.
+For (ii) let $T'$ be the absolute inertia group of $L$.
+By Theorem \ref{theorem-sylowgroups-inertiagroup}, the $\ell$-Sylow group of $T'$ is isomorphic to $\mathbb{Z}_\ell^{r'}$, where $r'=\dim_{\mathbb{F}_\ell}(\Gamma'/\ell\Gamma')$.
+Furthermore, we have $(\Gamma':\Gamma) = (T:T')$ and the following exact sequence of $\ell$-Sylow groups:
+\[ 0 \longrightarrow \mathbb{Z}_\ell^{r'} \longrightarrow \mathbb{Z}_\ell^{r}\longrightarrow (T/T')_\ell\longrightarrow 0 \]
+Now $(T/T')_\ell$ is finite if and only if $r'=r=2$.
 \end{proof}

### 2017-07-23 Master Thesis/synopsis.tex
Better proof for lemma-henselization-specialization

@@
 \begin{lemma}\label{lemma-henselization-specialization}
-Let $K$ be a field and $w\in\operatorname{Spv}(K)$ with a fixed henselization $K_w$ of $w$.
-If $v\in\operatorname{Spv}(K)$ such that $w$ is a specialization of $v$, then there is a henselization $K_v$ of $v$ such that $K_v\subset K_w\subset K_v^\text{t}$.
-\end{lemma}
-
-\begin{proof}
-Let $\overline{w}$ be the extension of $w\in\operatorname{Spv}(K_w)$ on $K^\text{s}$.
-By \cite{EP05} Lemma 3.1.5 we can find an extension $\overline{v}$ of $v$ on $K^\text{s}$ such that $\overline{w}$ is a specialization of $\overline{v}$.
-Let $K_v$ be the fixed field of $G_{\overline{v}}\subset G_K$.
-
-Let $\sigma\in G_{\overline{w}}$.
-Then both $\mathcal{O}_{\overline{v}}$ and $\sigma(\mathcal{O}_{\overline{v}})$ are overrings of $\mathcal{O}_{\overline{w}}$.
-Now the overrings of valuation rings are totally ordered by inclusion and correspond to localizations of $\mathcal{O}_{\overline{w}}$ at prime ideals (See \cite{EP05} 2.3).
-Since $\mathcal{O}_{\overline{v}}$ and $\sigma(\mathcal{O}_{\overline{v}})$ have the same Krull dimension, they must be the same, thus $\sigma\in G_{\overline{v}}$.
-This proves $K_v\subset K_w$.
-
-Now let $\sigma\in G_{K_v^\text{t}}$, i.e.
-$\sigma\in G_{\overline{v}}$ such that the induced automorphism on the residue field $\overline{\sigma}:k(\overline{v})\to k(\overline{v})$ is the identity.
-Let $\mathfrak{p}\subset\mathcal{O}_{\overline{w}}$ be a prime ideal with $\mathcal{O}_{\overline{v}} = (\mathcal{O}_{\overline{w}})_\mathfrak{p}$.
-In particular, the restriction $\overline{\sigma}|: \mathcal{O}_{\overline{w}}/\mathfrak{p}\to \mathcal{O}_{\overline{w}}/\mathfrak{p}$ is the identity, thus $\sigma(\mathcal{O}_{\overline{w}})=\mathcal{O}_{\overline{w}}$, i.e.
-$\sigma\in G_{\overline{w}}$.
-This proves $K_w\subset K_v^\text{t}$.
+Let $K$ be a field and $v\in\operatorname{Spv}(K)$ with a fixed henselization $K_v$ of $v$.
+If $w\in\operatorname{Spv}(K)$ such that $w$ is a specialization of $v$, then there is a henselization $K_w$ of $w$ such that $K_v\subset K_w\subset K_v^\text{t}$.
+\end{lemma}
+
+\begin{proof}
+By Lemma \ref{2.2-pre}, $w$ induces a valuation $\tilde{w}\in\operatorname{Spv}k(v)$.
+Let $k(v)_{\tilde{w}}$ be henselization of $k(v)$ with respect to $\tilde{w}$.
+By Theorem \ref{theorem-absoluteinertia}, $k(v)_{\tilde{w}}$ corresponds to an intermediate extension $K_v\subset L \subset K^\text{t}_v$.
+Let $v_L$ be the unique extension of $v$ on $L$.
+Then $k(v_L)=k(v)_{\tilde{w}}$.
+Let $w_L$ be the valuation on $L$ corresponding to $\tilde{w}\in \operatorname{Spv} k(v_L)$.
+By \cite{EP05} 4.1.4, $w_L$ is henselian iff both $v_L$ and $\tilde{w}$ are henselian, thus $L$ is a henselization of $w$.
 \end{proof}
@@
 Let $F$ be a field and $n\in\mathbb{N}$ invertible in $F$. Let $\mu_n\subset F^\text{s}$ be the group of all $n$-th roots of unity. Consider the following short exact sequence:
-\[ 1\longrightarrow \mu_n\longrightarrow (F^\text{s})^\times \stackrel{n}{\longrightarrow} (F^\text{s})^\times \longrightarrow 1 \]
+\[ 1\longrightarrow \mu_n\longrightarrow (F^\text{s})^\times \stackrel{\cdot n}{\longrightarrow} (F^\text{s})^\times \longrightarrow 1 \]
 This gives rise to the following long exact cohomology sequence:
 \[\begin{tikzcd}
-\cdots\longrightarrow F^\times \ar[r, "n"]& F^\times \ar[r]& \mathrm{H}^1(F, \mu_n)\ar[r] & \mathrm{H}^1(F, (F^\text{s})^\times)\ar[dlll]\\
-\mathrm{H}^1(F, (F^\text{s})^\times)\ar[r] & \mathrm{H}^2(F, \mu_n)\ar[r] & \operatorname{Br}(F)\ar[r, "n"] & \operatorname{Br}(F) \longrightarrow \cdots
+\cdots\longrightarrow F^\times \ar[r, "\cdot n"]& F^\times \ar[r]& \mathrm{H}^1(F, \mu_n)\ar[r] & \mathrm{H}^1(F, (F^\text{s})^\times)\ar[dlll]\\
+\mathrm{H}^1(F, (F^\text{s})^\times)\ar[r] & \mathrm{H}^2(F, \mu_n)\ar[r] & \operatorname{Br}(F)\ar[r, "\cdot n"] & \operatorname{Br}(F) \longrightarrow \cdots
 \end{tikzcd}\]

### 2017-07-23 Master Thesis/synopsis.tex
Improve phrasing in proof of 3.2

@@
 Then $k(v)$ is separably closed where $v=v(w_1)$.
-By Lemma \ref{lemma-henselization-specialization} we have $K_{w_1}K_{w_2}\subset K_v^\text{t}$ and by Theorem \ref{theorem-absoluteinertia} $K_{w_1}K_{w_2}=K_v^\text{t}$.
+By Lemma \ref{lemma-henselization-specialization} we have $K_{w_1}K_{w_2}\subset K_v^\text{t}$ and hence $K_{w_1}K_{w_2}=K_v^\text{t}$ by Theorem \ref{theorem-absoluteinertia}.
 The extension of $v$ on $K_v^\text{t}$ has value group $\Gamma_v$, which is finitely generated.
@@
 
-Furthermore, in case (ii), $K'_{w'_1}K'_{w'_2} = (K_{w_1}K_{w_2})' \neq \overline{K'}$ follows from (i) and again by Lemma \ref{2.2} we get: \[w'_1,w'_2\in\operatorname{Spv}_\text{rd}^{(2)}(K)\quad \text{and} \quad v(w'_1) = v(w'_2)\]
-Thus $\sigma: \operatorname{Spv}_\text{d}^{(2)}(K)\to \operatorname{Spv}_\text{d}^{(2)}(K')$ maps $\operatorname{Spv}_\text{rd}^{(2)}(K)$ onto $\operatorname{Spv}_\text{rd}^{(2)}(K')$ such that:
+We have analogous statements (i') and (ii') for $K'$, $w'_1$, $w'_2$ of (i) and (ii) respectively.
+The equivalence of the corresponding statements holds for $K'$ as well.
+Now (i) and (i') are equivalent, because:
+\[\sigma\big(G_{K_{w_1}K_{w_2}}\big) = \sigma\big(G_{K_{w_1}}\cap G_{K_{w_2}}\big) = G_{K'_{w'_1}}\cap G_{K'_{w'_2}} = G_{K'_{w'_1}K'_{w'_2}} \]
+Thus (ii) and (ii') are equivalent, hence we can see that $\sigma: \operatorname{Spv}_\text{d}^{(2)}(K)\to \operatorname{Spv}_\text{d}^{(2)}(K')$ maps $\operatorname{Spv}_\text{rd}^{(2)}(K)$ onto $\operatorname{Spv}_\text{rd}^{(2)}(K')$ such that:
 \[ v(w_1)=v(w_2) \iff v(w'_1) = v(w'_2) \]
@@
 Then $k(v)$ is finitely generated of dimension $1$ by Lemma \ref{rank-dimension-inequality}, thus a global field.
+Set $N=G_{K_v^\text{t}}$.
 By Theorem \ref{theorem-absoluteinertia}, we have:
 \[G_{K_w}/N \cong \operatorname{Gal}(K_v^\text{t}/K_w)\cong G_{k(v_w)}\] 
-where $v_w$ is the unique extension of $v$ on $K_w$ and $N=G_{K_v^\text{t}}$.
-With Lemma \ref{2.2-pre} we can see that $k(v_w)$ is the henselization of $k(v)$ with respect to $w\mod v\in\operatorname{Spv}k(v)$.
+where $v_w$ is the unique extension of $v$ on $K_w$.
+With Lemma \ref{lemma-henselization-specialization} we can see that $k(v_w)$ is the henselization of $k(v)$ with respect to $w\mod v\in\operatorname{Spv}k(v)$.
 Let $H$ be the closed subgroup of $G_{K_v}$ generated by all subgroups $G_{K_w}$ for $w\in\operatorname{Spv}_\text{rd}^{(2)}(K)$ with $v(w) = v$.
@@
 Now let $w_1,w_2\in\operatorname{Spv}_\text{rd}^{(2)}(K)$ with $w_1\neq w_2$ and $v=v(w_1)=v(w_2)$. We have:
-\[ \sigma(G_{K^\text{t}_v}) = \sigma(G_{K_{w_1}}\cap G_{K_{w_2}}) = G_{K'_{w'_1}}\cap G_{K'_{w'_2}} = G_{(K')^\text{t}_{v'}} \]
+\[ \sigma\big(G_{K^\text{t}_v}\big) = \sigma\big(G_{K_{w_1}K_{w_2}}\big) = G_{K'_{w'_1}K'_{w'_2}} = G_{(K')^\text{t}_{v'}} \]
 Therefore, $\sigma$ induces with Theorem \ref{theorem-absoluteinertia} the following isomorphism:

### 2017-07-23 Master Thesis/synopsis.tex
Fix typos

@@
 
-Furthermore, if $E^{\text{t}}$ and $E'^{\text{t}}$ denote the absolute inertia field of $E$ and $E'$ respectively, then $\sigma$ induces an isomorphism:
+Furthermore, if $E^{\text{t}}$ and $E'^{\text{t}}$ denote the absolute inertia fields of $E$ and $E'$ respectively, then $\sigma$ induces an isomorphism:
 \[ G_{k(v)}\cong\operatorname{Gal}(E^\text{t}/E) \to\operatorname{Gal}(E'^{\text{t}}/E')\cong G_{k(v')} \]
@@
 Let $k$ and $k'$ be the algebraic closures of $\mathbb{Q}$ in $K$ and $K'$ respectively, i.e.
-the constant fields of $K$ and $K'$, and let $\overline{k}$ and $\overline{k'}$ be the algebraic closure of $k$ and $k'$ in $\overline{K}$ and $\overline{K'}$ respectively.
-
-Let $X$ and $X'$ be the unique smooth projective curves over $k$ and $k'$ with function field $K$ and $K'$ respectively.
+the constant fields of $K$ and $K'$, and let $\overline{k}$ and $\overline{k'}$ be the algebraic closures of $k$ and $k'$ in $\overline{K}$ and $\overline{K'}$ respectively.
+
+Let $X$ and $X'$ be the unique smooth projective curves over $k$ and $k'$ with function fields $K$ and $K'$ respectively.
 We identify the set of closed points $X_0$ of $X$ with the set of discrete valuations $v$ with $v(k^\times)=0$.
@@
 Now let $T\subset X_0$ be finite.
-By the Weak Approximation Theorem \ref{weak-approximation}, we can add a principal divisor to the $D_i$ and the $D'_j$ so that their support lies in $X_0 -  T$ and $X'_0 -  T'$ respectively.
+By the Weak Approximation Theorem \ref{weak-approximation}, we can add a principal divisor to the $D_i$ and the $D'_j$ so that their supports lie in $X_0 -  T$ and $X'_0 -  T'$ respectively.
 \item Observe that the following sequence is exact:

### 2017-07-23 Master Thesis/synopsis.tex
Consistent citations

@@
 \begin{proof}
-See \cite{EP05} 3.4.1 and 3.4.3.
+See \cite{EP05} Proposition 3.4.1 and Theorem 3.4.3.
 \end{proof}
@@
 \begin{proof}
-See \cite{EP05} 2.3, 4.1.4, \cite{ZS60} VI §3 Theorem 3.
+See \cite{EP05} Section 2.3, Corollary 4.1.4, \cite{ZS60} VI, §3 Theorem 3.
 \end{proof}
@@
 Let $w_L$ be the valuation on $L$ corresponding to $\tilde{w}\in \operatorname{Spv} k(v_L)$.
-By \cite{EP05} 4.1.4, $w_L$ is henselian iff both $v_L$ and $\tilde{w}$ are henselian, thus $L$ is a henselization of $w$.
+By \cite{EP05} Corollary 4.1.4, $w_L$ is henselian iff both $v_L$ and $\tilde{w}$ are henselian, thus $L$ is a henselization of $w$.
 \end{proof}
@@
 \begin{proof}
-(i) and (ii) follow from \cite{HK94}, Lemma 2.4.3, and (iii) is a consequence of (ii) and the fact that henselizations commute with filtrated direct limits, e.g. see \cite{EGA4} 18.6.14.
+(i) and (ii) follow from \cite{HK94} Lemma 2.4.3, and (iii) is a consequence of (ii) and the fact that henselizations commute with filtrated direct limits, e.g. see \cite{EGA4} Proposition 18.6.14.
 \end{proof}
@@
 \end{tikzcd} \]
-According to \cite{Sa87}, Theorem 5.3, $g$ is injective.
+According to \cite{Sa87} Theorem 5.3, $g$ is injective.
 Now, $\psi\circ\varphi$ can be identified with the injective map $\Phi$ from the proof of the previous Lemma.
@@
 Therefore, $\operatorname{inf}$ and $\operatorname{res}$ are both isomorphisms and it suffices to calculate $\mathrm{H}^3(G_\ell,\mathbb{Z}/\ell\mathbb{Z})$.
-By \cite{Se64}, Ch.
-I, §4, Proposition 22 we have
+By \cite{Se64} Ch.
+I, §4 Proposition 22 we have
 \[\operatorname{cd}(G_\ell) = \operatorname{cd}(\mathbb{Z}_\ell^r\rtimes G_k(\ell)) = \begin{cases}
@@
 Now let $G_k(\ell) \cong\mathbb{Z}_\ell$ and $r = 2$.
-According to the proof of \cite{Se64}, Ch.
-I, §4, Proposition 22, we have:
+According to the proof of \cite{Se64} Ch.
+I, §4 Proposition 22, we have:
 \[ \mathrm{H}^3(G_\ell,\mathbb{Z}/\ell\mathbb{Z}) = \mathrm{H}^1(G_k(\ell), \mathrm{H}^2 (\mathbb{Z}_\ell^r, \mathbb{Z}/\ell\mathbb{Z})) \]
@@
 
-The second assertion follows from the first one, applied to $L$, by inspecting the following equation of supernatural numbers (\cite{En72}, Theorem 20.21):
+The second assertion follows from the first one, applied to $L$, by inspecting the following equation of supernatural numbers (\cite{En72} Theorem 20.21):
 \[ [L:K] = (\Gamma':\Gamma) \cdot [k':k]\cdot p^d \]
@@
 \begin{proof}
-See \cite{ZS60} VI §10 Theorem 18.
+See \cite{ZS60} VI, §10 Theorem 18.
 \end{proof}

### 2017-07-26 Master Thesis/synopsis.tex
Add Lemma before 2.4 and remove reference to [K1] 1.4

@@
 
+\begin{lemma}\label{2.4-pre}
+Let $K$ be henselian with respect to a discrete valuation $v$, $k=k(v)$ its residue field with $\operatorname{cd}(k)\leq 2$. Let $\ell\neq p$ be a prime where $p=\operatorname{char}(k)$. Then:
+\[ \mathrm{H}^3(K, \mathbb{Z}/\ell\mathbb{Z}(2)) \cong \mathrm{H}^2(k, \mu_{\ell}) \cong {}_\ell\operatorname{Br}(k) \]
+\end{lemma}
+
+\begin{proof}
+Let $T$ be the absolute inertia group of $K$ and $V\lhd T$ its absolute ramification group.
+By Theorem \ref{theorem-sylowgroups-inertiagroup}, $V$ is the $p$-Sylow group of $T$ and the $\ell$-Sylow group of $T/V$ is isomorphic to $\mathbb{Z}_\ell$.
+Therefore, by \cite{Se64} Ch.
+I, §3 Proposition 15:
+\[ \operatorname{cd}_\ell(T)\leq \operatorname{cd}_\ell(V) + \operatorname{cd}_\ell(\mathbb{Z}_\ell) = 1 \]
+Inspection of the Hochschild-Serre spectral sequence 
+\[E_2^{i,j}=\mathrm{H}^i(G_K/T, \mathrm{H}^j(T, \mathbb{Z}/\ell\mathbb{Z}(2)))\Longrightarrow\mathrm{H}^{i+j}(G_K, \mathbb{Z}/\ell\mathbb{Z}(2))\]
+yields an isomorphism $\mathrm{H}^2(G_K/T, \mathrm{H}^1(T, \mathbb{Z}/\ell\mathbb{Z}(2)))\cong \mathrm{H}^3(G_K, \mathbb{Z}/\ell\mathbb{Z}(2))$.
+By Theorem \ref{theorem-absoluteinertia}, we have  $G_K/T\cong G_k$.
+Since $\mu_{\ell}$ is contained in the absolute inertia field of $K$, $T$ acts trivially on $\mu_{\ell}$, hence:
+\[ \mathrm{H}^1(T, \mathbb{Z}/\ell\mathbb{Z}(2)) \cong \mathrm{H}^1(T, \mu_{\ell}) \cong \operatorname{Hom}_\text{cont.}(T, \mu_{\ell}) \cong \operatorname{Hom}_\text{cont.}(T/V, \mu_{\ell}) \]
+By Theorem \ref{theorem-sylowgroups-inertiagroup}, $T/V$ is the free pro-$\mathcal{C}$ group, where $\mathcal{C}$ is the class of all finite groups with order prime to $p$.
+Therefore $\operatorname{Hom}_\text{cont.}(T/V, \mu_{\ell})\cong \mu_{\ell}$.
+The rest follows from Proposition \ref{prop:kummer-theory}.
+\end{proof}
+
 \begin{lemma}\label{2.4-finitelygenerated}
@@
 \[ \mathrm{H}^3(K)\longrightarrow \prod_{x\in\mathfrak{X}_1}\mathrm{H}^3(K_{v_x}) \]
-is injective, and, in fact, the proof shows that its image is contained in $\bigoplus \mathrm{H}^3(K_{v_x})$. By \cite{Ka86} Lemma 1.4, for a discrete valuation $v$ of $K$ the map
+is injective, and, in fact, the proof shows that its image is contained in $\bigoplus \mathrm{H}^3(K_{v_x})$.
+
+Let $v$ be a discrete valuation of $K$ with residue field $k=k(v)$. By Lemma \ref{2.4-pre}, the map
 \[ \mathrm{H}^3(K_v) \longrightarrow \prod_w\mathrm{H}^3(K_w) \]
@@
 \[ {}_\ell\operatorname{Br}(k) \longrightarrow \prod_{\widetilde{w}\in\operatorname{Spv}^{(1)}(k)} {}_\ell\operatorname{Br}(k_{\widetilde{w}}) \]
-where $k = k(v)$.
-This map is injective by the classical Hasse principle and its image is contained in $\bigoplus {}_\ell\operatorname{Br}(k_{\widetilde{w}})$.
-Combining these maps yields the injection:
+This map is injective by the classical Hasse principle and its image is contained in $\bigoplus {}_\ell\operatorname{Br}(k_{\widetilde{w}})$. Combining these maps yields the injection:
 \[\Phi: \mathrm{H}^3(K)\longrightarrow \bigoplus_{x\in\mathfrak{X}_1}\  \bigoplus_{\substack{w\in\operatorname{Spv}_\text{rd}^{(2)}(K) \\ v(w) = v_x}} \mathrm{H}^3(K_w) \]
@@
 Now let $G_k(\ell) \cong\mathbb{Z}_\ell$ and $r = 2$.
-According to the proof of \cite{Se64} Ch.
-I, §4 Proposition 22, we have:
-\[ \mathrm{H}^3(G_\ell,\mathbb{Z}/\ell\mathbb{Z}) = \mathrm{H}^1(G_k(\ell), \mathrm{H}^2 (\mathbb{Z}_\ell^r, \mathbb{Z}/\ell\mathbb{Z})) \]
-By [NSW] Theorem 2.4.6, we have $\mathrm{H}^2(\mathbb{Z}_\ell^r,\mathbb{Z}/\ell\mathbb{Z}) \cong\mathrm{H}^1(\mathbb{Z}_\ell, \mathrm{H}^1(\mathbb{Z}_\ell, \mathbb{Z}/\ell\mathbb{Z})) = \mathbb{Z}/\ell\mathbb{Z}$, since $\mathrm{H}^1(\mathbb{Z}_\ell, \mathbb{Z}/\ell\mathbb{Z})=\operatorname{Hom}_\text{cont.}(\mathbb{Z}_\ell, \mathbb{Z}/\ell\mathbb{Z})=\mathbb{Z}/\ell\mathbb{Z}$.
-Therefore, $\mathrm{H}^3(G_\ell,\mathbb{Z}/\ell\mathbb{Z})=\mathbb{Z}/\ell\mathbb{Z}$.
-
-The second assertion follows from the first one, applied to $L$, by inspecting the following equation of supernatural numbers (\cite{En72} Theorem 20.21):
+Inspecting the Hochschild-Serre spectral sequence yields an isomorphism:
+\[ \mathrm{H}^3(G_\ell,\mathbb{Z}/\ell\mathbb{Z}) \cong \mathrm{H}^1(G_k(\ell), \mathrm{H}^2 (\mathbb{Z}_\ell^r, \mathbb{Z}/\ell\mathbb{Z})) \]
+Using $\mathrm{H}^1(\mathbb{Z}_\ell, \mathbb{Z}/\ell\mathbb{Z})\cong\operatorname{Hom}_\text{cont.}(\mathbb{Z}_\ell, \mathbb{Z}/\ell\mathbb{Z})=\mathbb{Z}/\ell\mathbb{Z}$ and \cite{NSW} Theorem 2.4.6, we can calculate $\mathrm{H}^2(\mathbb{Z}_\ell^r,\mathbb{Z}/\ell\mathbb{Z}) \cong\mathbb{Z}/\ell\mathbb{Z}$.
+Therefore, $\mathrm{H}^3(G_\ell,\mathbb{Z}/\ell\mathbb{Z})\cong\mathbb{Z}/\ell\mathbb{Z}$.
+
+For the second assertion, we assume $\ell^\infty\nmid [k:\mathbb{F}_p]$ and $\dim_{\mathbb{F}_\ell}(\Gamma / \ell\Gamma) = 2$.
+By \cite{En72} Theorem 20.21, we have the following equation of supernatural numbers:
 \[ [L:K] = (\Gamma':\Gamma) \cdot [k':k]\cdot p^d \]
-for some $d \leq \infty$, where $\Gamma'$ denotes the value group and $k'$ the residue field of the unique prolongation of $w$ to $L$.
-By using this equation, it suffices to show the following equivalences:
+for some $d \leq \infty$, where $\Gamma'$ denotes the value group and $k'$ the residue field of the unique extension of $w$ to $L$.
+By using this equation and the already proven statement from the first part, it suffices to show the following equivalences:
 \begin{enumerate}[(i)]
@@
 \end{enumerate}
-(i) is clear.
+(i) follows from $[k':\mathbb{F}_p]=[k':k][k:\mathbb{F}_p]$.
 For (ii) let $T'$ be the absolute inertia group of $L$.
@@
 Furthermore, we have $(\Gamma':\Gamma) = (T:T')$ and the following exact sequence of $\ell$-Sylow groups:
-\[ 0 \longrightarrow \mathbb{Z}_\ell^{r'} \longrightarrow \mathbb{Z}_\ell^{r}\longrightarrow (T/T')_\ell\longrightarrow 0 \]
+\[ 0 \longrightarrow \mathbb{Z}_\ell^{r'} \longrightarrow \mathbb{Z}_\ell^{r}\longrightarrow (T/T')_\ell\longrightarrow 1 \]
 Now $(T/T')_\ell$ is finite if and only if $r'=r=2$.

### 2017-07-26 Master Thesis/introduction.tex
Fix typos
Natural numbers start with 1
Finish second part of Lemma 2.2
Small fixes

@@
 Finite fields have an abelian absolute Galois group.
-In fact, they all have the same absolute Galois group $\hat{\mathbb{Z}}$, therefore not much information can be retrieved from these groups.
-The study of these types of questions belong in the area of \textit{anabelian geometry}.
+In fact, they all have the same absolute Galois group $\widehat{\mathbb{Z}}$, therefore not much information can be retrieved from these groups.
+The study of these types of questions belong to the area of \textit{anabelian geometry}.
 This term refers to the fact that the less abelian $\pi_1^\text{ét}(X)$ is, the more information it carries about $X$.
@@
 
-This has been proved by F.
+This has been proven by F.
 Pop (see \cite{Po90}, \cite{Po94}, \cite{Po95}) using model theory.
@@
 In the first chapter, we want to recap some results from valuation theory.
-In the second part, we will prove some lemmata in order to be able to characterize henselian fields over $K$ of valuations of rational rank $2$ in terms of their Galois cohomology groups of dimension $3$.
+In the second part, we will prove some lemmata in order to characterize henselian fields over $K$ of valuations of rational rank $2$ in terms of their Galois cohomology groups of dimension $3$.
 
@@
 $\begin{array}{ll}
-\mathbb{N} & \text{natural numbers} \\
+\mathbb{N} & \text{natural numbers starting from $1$} \\
 \mathbb{Z} & \text{integers}\\
@@
 \mathbb{Z}_p & \text{$p$-adic integers}\\
-\hat{\mathbb{Z}} & \text{profinite completion of $\mathbb{Z}$}\\
+\widehat{\mathbb{Z}} & \text{profinite completion of $\mathbb{Z}$}\\
 \\

### 2017-07-26 Master Thesis/synopsis.tex
Small fixes

@@
 
-In this section we shall recall facts from valuation theory.
+In this section we shall recall some facts from valuation theory.
 
@@
 \begin{proof}
-See \cite{EP05} Section 2.3, Corollary 4.1.4, \cite{ZS60} VI, §3 Theorem 3.
+See \cite{EP05} Section 2.3 and Corollary 4.1.4, \cite{ZS60} VI, §3 Theorem 3.
 \end{proof}
@@
 \begin{proof}
-See \cite{EP05} Lemma 5.2.1, Lemma 5.2.2, Theorem 5.2.5.
+See \cite{EP05} Lemma 5.2.1, Lemma 5.2.2, and Theorem 5.2.5.
 \end{proof}
@@
 Then $K$ is algebraically closed or $w_1,w_2\in\operatorname{Spv}_\text{rd}^{(2)}(K),\ v(w_1)=v(w_2)$ and $k(v(w_1))$ is separably closed.
-In this case, we have $\operatorname{cd}(G_K)\leq 2$.
+In this case, we have $\operatorname{cd}(K)\leq 2$.
 \end{lemma}
@@
 \begin{proof}
-(i) and (ii) follow from \cite{HK94} Lemma 2.4.3, and (iii) is a consequence of (ii) and the fact that henselizations commute with filtrated direct limits, e.g. see \cite{EGA4} Proposition 18.6.14.
+(i) and (ii) follow from \cite{HK94} Lemma 2.4.3 and the fact that singularities of two-dimensional $\mathbb{Z}$-schemes of finite type can be resolved.
+(iii) is a consequence of (ii) and the fact that henselizations commute with filtrated direct limits, e.g.
+see \cite{EGA4} Proposition 18.6.14.
 \end{proof}
@@
 Again by Lemma \ref{5.2} (ii), it suffices to show that $\mathcal{P}(F_1)/\mathcal{P}(F)$ is free-by-finite. 
-Consider the following diagram with exact rows and columns:
-\[ \begin{tikzcd}
-& 0 \ar[d] & 0 \ar[d] & \ker(f) \ar[d]& \\
+Consider the following diagram with exact rows:
+\[ \begin{tikzcd}
 0\ar[r] & \mathcal{P}(F) \ar[d] \ar[r]& \mathcal{I}(F) \ar[r]\ar[d]& \operatorname{Cl}(F)\ar[r]\ar[d, "f"] & 0\\
-0 \ar[r]& \mathcal{P}(F_1)\ar[d] \ar[r]& \mathcal{I}(F_1) \ar[d]\ar[r]& \operatorname{Cl}(F_1)\ar[d] \ar[r]& 0 \\
-& \mathcal{P}(F_1)/\mathcal{P}(F) & \mathcal{I}(F_1)/\mathcal{I}(F) & \operatorname{coker}(f) &
+0 \ar[r]& \mathcal{P}(F_1) \ar[r]& \mathcal{I}(F_1) \ar[r]& \operatorname{Cl}(F_1) \ar[r]& 0 
 \end{tikzcd} \]

### 2017-07-27 Master Thesis/introduction.tex
Some fixes

@@
 Let $k,k'$ be two global fields.
-Then any isomorphism $\sigma: G_k\to G_{k'}$ between their absolute Galois groups is induced by a unique field isomorphism $\varphi: k'^{\text{s}}\to k^\text{s}$, i.e:
+Then any isomorphism $\sigma: G_k\to G_{k'}$ between their absolute Galois groups is induced by a unique field isomorphism $\varphi: k'^{\text{s}}\to k^\text{s}$ between their separable closures, i.e:
 \[\sigma(g) =\varphi^{-1}g\varphi \quad \text{for all }g\in G_k\] 
@@
 Let $K,K'$ be two finite field extensions of $\mathbb{Q}(T)$.
-Then any isomorphism $\sigma:G_K\to G_{K'}$ between their absolute Galois groups is induced by a unique field isomorphism $\varphi:\overline{K'}\to\overline{K}$, i.e:
-\[\sigma(g) =\varphi^{-1}g\varphi \quad \text{for all }g\in G_k\]
+Then any isomorphism $\sigma:G_K\to G_{K'}$ between their absolute Galois groups is induced by a unique field isomorphism $\varphi:\overline{K'}\to\overline{K}$ between their algebraic closures, i.e:
+\[\sigma(g) =\varphi^{-1}g\varphi \quad \text{for all }g\in G_K\]
 In particular $\varphi$ induces a field isomorphism $K'\to K$.
@@
 K^\text{s} & \text{separable closure of $K$}\\
-K^\text{t} & \text{absolute inertia field of $K$}\\
+K^\text{t} & \text{absolute inertia field of a henselian $K$}\\
 K_v & \text{henselization of $K$ with respect to $v$}\\
+\mu_n & \text{$n$-th roots of unity in a separable closure}\\
+\\
 G_K & \text{absolute Galois group of $K$}\\
@@
 \operatorname{trdeg}(L/K) & \text{transcendence degree of $L/K$}\\
-\mu_n & \text{$n$-th roots of unity in a separable closure}
 \end{array}$

### 2017-07-27 Master Thesis/synopsis.tex
Some fixes

@@
 
-The following lemma is the key result in proving Theorem \ref{thm:main-result}.
-First, we use K.
-Kato's two-dimensional local-global principle to show the assertion for finitely generated $K$.
-Then, we'll use a quasi-compactness argument to see that we can drop the assumption that $K$ is finitely generated.
-
 \begin{lemma}\label{2.4-pre}
@@
 \[ \operatorname{cd}_\ell(T)\leq \operatorname{cd}_\ell(V) + \operatorname{cd}_\ell(\mathbb{Z}_\ell) = 1 \]
-Inspection of the Hochschild-Serre spectral sequence 
+Thus, the Hochschild-Serre spectral sequence degenerates on $E_2$:
 \[E_2^{i,j}=\mathrm{H}^i(G_K/T, \mathrm{H}^j(T, \mathbb{Z}/\ell\mathbb{Z}(2)))\Longrightarrow\mathrm{H}^{i+j}(G_K, \mathbb{Z}/\ell\mathbb{Z}(2))\]
-yields an isomorphism $\mathrm{H}^2(G_K/T, \mathrm{H}^1(T, \mathbb{Z}/\ell\mathbb{Z}(2)))\cong \mathrm{H}^3(G_K, \mathbb{Z}/\ell\mathbb{Z}(2))$.
+It yields an isomorphism $\mathrm{H}^2(G_K/T, \mathrm{H}^1(T, \mathbb{Z}/\ell\mathbb{Z}(2)))\cong \mathrm{H}^3(G_K, \mathbb{Z}/\ell\mathbb{Z}(2))$.
 By Theorem \ref{theorem-absoluteinertia}, we have  $G_K/T\cong G_k$.
@@
 \[ \mathrm{H}^1(T, \mathbb{Z}/\ell\mathbb{Z}(2)) \cong \mathrm{H}^1(T, \mu_{\ell}) \cong \operatorname{Hom}_\text{cont.}(T, \mu_{\ell}) \cong \operatorname{Hom}_\text{cont.}(T/V, \mu_{\ell}) \]
-By Theorem \ref{theorem-sylowgroups-inertiagroup}, $T/V$ is the free pro-$\mathcal{C}$ group, where $\mathcal{C}$ is the class of all finite groups with order prime to $p$.
-Therefore $\operatorname{Hom}_\text{cont.}(T/V, \mu_{\ell})\cong \mu_{\ell}$.
+By Theorem \ref{theorem-sylowgroups-inertiagroup}, $T/V$ is the free pro-$\mathcal{C}$ group of rank $1$, where $\mathcal{C}$ is the class of all finite groups with order prime to $p$.
+Therefore, $\operatorname{Hom}_\text{cont.}(T/V, \mu_{\ell})\cong \mu_{\ell}$.
 The rest follows from Proposition \ref{prop:kummer-theory}.
 \end{proof}
+
+The following lemma is the key result in proving Theorem \ref{thm:main-result}.
+First, we use K.
+Kato's two-dimensional local-global principle to show the assertion for finitely generated $K$.
+Then, we'll use a quasi-compactness argument to see that we can drop the assumption that $K$ is finitely generated.
 
@@
 The exact sequence above gives rise to the short exact sequence of their $\ell$-Sylow subgroups:
-\[ 0\longrightarrow\mathbb{Z}_\ell^r \longrightarrow G_\ell \longrightarrow G_{k,\ell}\longrightarrow 1 \]
+\[ (\star)\qquad 0\longrightarrow\mathbb{Z}_\ell^r \longrightarrow G_\ell \longrightarrow G_{k,\ell}\longrightarrow 1 \phantom{\qquad (\star)} \]
 Since $G_{k,\ell}$ is a free pro-$\ell$ group, this sequence splits and we have an isomorphism $G_\ell \cong \mathbb{Z}_\ell^r\rtimes G_{k,\ell}$ with $r =\dim_{\mathbb{F}_\ell}(\Gamma/\ell\Gamma)\leq \operatorname{rrank}(\Gamma)\leq \dim(K)= 2$.
@@
 Now let $G_k(\ell) \cong\mathbb{Z}_\ell$ and $r = 2$.
-Inspecting the Hochschild-Serre spectral sequence yields an isomorphism:
+The Hochschild-Serre spectral sequence to $(\star)$ degenerates on $E_2$ and yields an isomorphism:
 \[ \mathrm{H}^3(G_\ell,\mathbb{Z}/\ell\mathbb{Z}) \cong \mathrm{H}^1(G_k(\ell), \mathrm{H}^2 (\mathbb{Z}_\ell^r, \mathbb{Z}/\ell\mathbb{Z})) \]

### 2017-07-27 Master Thesis/synopsis.tex
More fixes

@@
 
-Now we calculate the cohomology groups $\mathrm{H}^3(K_w,\mathbb{Z}/\ell\mathbb{Z})$ for $w\in |\operatorname{Spv}(K)|$.
+In the next lemma, we're going to calculate the cohomology groups $\mathrm{H}^3(K_w,\mathbb{Z}/\ell\mathbb{Z})$ for $w\in |\operatorname{Spv}(K)|$.
 With the help of our Main Lemma \ref{2.4} we can find a cohomological criterion of henselian fields over $K$ with respect to a $w\in |\operatorname{Spv}(K)|$.
@@
 Now $(K_0)_{w_1}$ is henselian both with respect to $w_1$ and with respect to the extension of $w_0$.
-Lemma \ref{2.2} yields $\operatorname{cd}(G_{(K_0)_{w_1}})\leq 2$, a contradiction.
+Lemma \ref{2.2} yields $\operatorname{cd}((K_0)_{w_1})\leq 2$, a contradiction.
 The uniqueness of $w$ is again a consequence of Lemma \ref{2.2}.
@@
 Let $F$ be another henselization of $w$.
-Then $E, F$ are $K$-conjugate, hence $E',F'$ are $K'$-conjugate, thus $K'$-isomorphic and the corresponding henselian valuations of $E', F'$ are respectively mapped to each other.
+Then $E, F$ are $K$-conjugate, thus $E',F'$ are $K'$-conjugate, hence $K'$-isomorphic and the corresponding henselian valuations of $E', F'$ are respectively mapped to each other.
 \end{proof}
@@
 Lemma \ref{3.1} shows that $\sigma$ induces a bijection $\sigma: \operatorname{Spv}_\text{d}^{(2)}(K)\to \operatorname{Spv}_\text{d}^{(2)}(K')$.
-For all $v\in\operatorname{Spv}(K)$ we choose a fixed henselization $K_v\subset\overline{K}$ with Lemma \ref{lemma-henselization-specialization}, such that if $v,w\in\operatorname{Spv}(K)$ and $w$ is a specialization of $v$ we have $K_v\subset K_w$.
+For all $v\in\operatorname{Spv}(K)$ we choose a fixed henselization $K_v\subset\overline{K}$ with Lemma \ref{lemma-henselization-specialization}, such that if $v,w\in\operatorname{Spv}(K)$ and $w$ is a specialization of $v$ we have $K_v\subset K_w\subset K_v^\text{t}$.
 If $v$ is nontrivial, then in this case we get $w\in\operatorname{Spv}_\text{rd}^{(2)}(K)$ and $v=v(w)\in\operatorname{Spv}_\text{d}^{(1)}(K)$.
@@
 Then $k(v)$ is separably closed where $v=v(w_1)$.
-By Lemma \ref{lemma-henselization-specialization} we have $K_{w_1}K_{w_2}\subset K_v^\text{t}$ and hence $K_{w_1}K_{w_2}=K_v^\text{t}$ by Theorem \ref{theorem-absoluteinertia}.
+We have $K_{w_1}K_{w_2}\subset K_v^\text{t}$ and hence $K_{w_1}K_{w_2}=K_v^\text{t}$ by Theorem \ref{theorem-absoluteinertia}.
 The extension of $v$ on $K_v^\text{t}$ has value group $\Gamma_v$, which is finitely generated.
@@
 In this section we shall show that $\sigma: G_K\to G_{K'}$ induces group isomorphisms $\varphi_n: K'^\times/n\to K^\times/n$.
-Due to Proposition \ref{prop:kummer-theory}, we have isomorphisms:
+By Proposition \ref{prop:kummer-theory}, we have isomorphisms:
 \[ \mathrm{H}^1(K',\mu_n)\cong K'^\times/n,\qquad \mathrm{H}^1(K,\mu_n)\cong K^\times/n \]

### 2017-08-01 Master Thesis/synopsis.tex
Add reference for coh. dim. of number fields <= 2

@@
 \begin{lemma}\label{2.4-pre}
-Let $K$ be henselian with respect to a discrete valuation $v$, $k=k(v)$ its residue field with $\operatorname{cd}(k)\leq 2$. Let $\ell\neq p$ be a prime where $p=\operatorname{char}(k)$. Then:
+Let $K$ be henselian with respect to a discrete valuation $v$, $k=k(v)$ its residue field. Let $p=\operatorname{char}(k)$, $\ell\neq p$ be a prime, and $\operatorname{cd}_\ell(k)\leq 2$. Then:
 \[ \mathrm{H}^3(K, \mathbb{Z}/\ell\mathbb{Z}(2)) \cong \mathrm{H}^2(k, \mu_{\ell}) \cong {}_\ell\operatorname{Br}(k) \]
@@
 \[ \operatorname{cd}_\ell(T)\leq \operatorname{cd}_\ell(V) + \operatorname{cd}_\ell(\mathbb{Z}_\ell) = 1 \]
-Thus, the Hochschild-Serre spectral sequence degenerates on $E_2$:
+By Theorem \ref{theorem-absoluteinertia}, we have an isomorphism $G_K/T\cong G_k$. Furthermore, we have $\operatorname{cd}_\ell(G_K/T)=\operatorname{cd}_\ell(k)\leq 2$ by requirement,
+thus the following Hochschild-Serre spectral sequence degenerates on $E_2$:
 \[E_2^{i,j}=\mathrm{H}^i(G_K/T, \mathrm{H}^j(T, \mathbb{Z}/\ell\mathbb{Z}(2)))\Longrightarrow\mathrm{H}^{i+j}(G_K, \mathbb{Z}/\ell\mathbb{Z}(2))\]
-It yields an isomorphism $\mathrm{H}^2(G_K/T, \mathrm{H}^1(T, \mathbb{Z}/\ell\mathbb{Z}(2)))\cong \mathrm{H}^3(G_K, \mathbb{Z}/\ell\mathbb{Z}(2))$.
-By Theorem \ref{theorem-absoluteinertia}, we have  $G_K/T\cong G_k$.
+It yields an isomorphism $\mathrm{H}^2(k, \mathrm{H}^1(T, \mathbb{Z}/\ell\mathbb{Z}(2)))\cong \mathrm{H}^3(K, \mathbb{Z}/\ell\mathbb{Z}(2))$.
 Since $\mu_{\ell}$ is contained in the absolute inertia field of $K$, $T$ acts trivially on $\mu_{\ell}$, hence:
@@
 \begin{lemma}\label{2.4-finitelygenerated}
-Let $K$ be finitely generated and $\ell\neq 2$ be a prime. The following map, induced by the restrictions, is injective:
+Let $K$ be finitely generated and $\ell\neq 2$ be a prime. The following map, induced by the restriction maps, is injective:
 \[ \mathrm{H}^3(K,\mathbb{Z}/\ell\mathbb{Z})\longrightarrow\prod_{w\in |\operatorname{Spv}(K)| }\mathrm{H}^3(K_w,\mathbb{Z}/\ell\mathbb{Z}) \]
@@
 
-Let $v$ be a discrete valuation of $K$ with residue field $k=k(v)$. By Lemma \ref{2.4-pre}, the map
-\[ \mathrm{H}^3(K_v) \longrightarrow \prod_w\mathrm{H}^3(K_w) \]
-where $w$ ranges over all rank-$2$ valuations with $v(w) = v$, can be identified with the following map:
+Let $v$ be a discrete valuation of $K$ with residue field $k=k(v)$. Consider the following map induced by the restriction maps: 
+\[ \phi: \mathrm{H}^3(K_v) \longrightarrow \prod_w\mathrm{H}^3(K_w) \]
+where $w$ ranges over all rank-$2$ valuations with $v(w) = v$.
+By Lemma \ref{rank-dimension-inequality}, we have $\dim(k) = 1$, thus $k$ is a global field of characteristic $0$, hence a number field.
+\cite{Se64} Ch.
+II, §4 Prop.
+13 shows $\operatorname{cd}_\ell(k)\leq 2$.
+With Lemma \ref{2.4-pre}, $\phi$ can be identified with
 \[ {}_\ell\operatorname{Br}(k) \longrightarrow \prod_{\widetilde{w}\in\operatorname{Spv}^{(1)}(k)} {}_\ell\operatorname{Br}(k_{\widetilde{w}}) \]
-This map is injective by the classical Hasse principle and its image is contained in $\bigoplus {}_\ell\operatorname{Br}(k_{\widetilde{w}})$. Combining these maps yields the injection:
+which is injective by the classical Hasse principle.
+Furthermore, its image is contained in $\bigoplus {}_\ell\operatorname{Br}(k_{\widetilde{w}})$.
+Combining these maps yields the following injection:
 \[\Phi: \mathrm{H}^3(K)\longrightarrow \bigoplus_{x\in\mathfrak{X}_1}\  \bigoplus_{\substack{w\in\operatorname{Spv}_\text{rd}^{(2)}(K) \\ v(w) = v_x}} \mathrm{H}^3(K_w) \]
@@
 \begin{lemma}[\textit{Main Lemma}]\label{2.4}
-Let $K$ be not necessarily finitely generated and $\ell\neq 2$ be a prime. The following map, induced by the restrictions, is injective:
+Let $K$ be not necessarily finitely generated and $\ell\neq 2$ be a prime. The following map, induced by the restriction maps, is injective:
 \[ \mathrm{H}^3(K,\mathbb{Z}/\ell\mathbb{Z})\longrightarrow \prod_{w\in |\operatorname{Spv}(K)| }\mathrm{H}^3(K_w,\mathbb{Z}/\ell\mathbb{Z}) \]

### 2017-08-05 Master Thesis/introduction.tex
Mixed up direction of arrows

@@
 In the first chapter, we want to recap some results from valuation theory.
-In the second part, we will prove some lemmata in order to characterize henselian fields over $K$ of valuations of rational rank $2$ in terms of their Galois cohomology groups of dimension $3$.
+In the second part, we will introduce some lemmata in order to characterize henselian fields over $K$ of valuations of rational rank $2$ in terms of their Galois cohomology groups of dimension $3$.
 
 In the second chapter, we will show that an isomorphism $\sigma: G_K\to G_{K'}$ induces bijections on the equivalence classes of defectless valuations of $K$ and $K'$ using the cohomological characterization of henselian fields.
-With the Theorem of Neukirch-Uchida \ref{thm:neukirch-uchida}, we will also show that $\sigma$ induces a field isomorphism between the residue fields of corresponding defectless valuations of rational rank $1$.
-Using the local correspondences, we will proceed to show that $\sigma$ induces group isomorphisms $K^\times/n\to K'^\times/n$ for all $n\in\mathbb{N}$.
+Using the Theorem of Neukirch-Uchida \ref{thm:neukirch-uchida}, we will also see that $\sigma$ induces a field isomorphism between the residue fields of corresponding defectless valuations of rational rank $1$.
+We will proceed to show that $\sigma$ induces group isomorphisms $K'^\times/n\to K^\times/n$ for all $n\in\mathbb{N}$.
 
 In the third chapter, we want to use the correspondences established in the second chapter to prove Theorem \ref{thm:main-result}.
-In order to do so, we need to discuss so-called free-by-finite groups.
-Using the existence of a so-called admissible set, we can construct a group isomorphism $K^\times\to K'^\times$.
+Before we are able to do that, we need to discuss free-by-finite groups.
+Using the existence of a so-called admissible set, we can construct a group isomorphism $K'^\times\to K^\times$.
 Finally, we will show that the group isomorphism can be extended to a field isomorphism $K\to K'$.
-By passing to the limits, we obtain a field isomorphism $\varphi: \overline{K}\to\overline{K'}$ which induces $\sigma$.
+By passing to the limits, we obtain a field isomorphism $\varphi: \overline{K'}\to\overline{K}$ which induces $\sigma$.
 

### 2017-08-05 Master Thesis/synopsis.tex
Reference ZS60 VI, §4 Cor 2 to Thm 4
Phrasing

@@
 \begin{proof}
-See \cite{EP05} Section 2.3 and Corollary 4.1.4, \cite{ZS60} VI, §3 Theorem 3.
+See \cite{EP05} Section 2.3 and Corollary 4.1.4, \cite{ZS60} VI, §3 Theorem 3 and VI, §4 Corollary 2 to Theorem 4.
 \end{proof}
@@
 Let $K$ be a field and $v\in\operatorname{Spv}(K)$ with a fixed henselization $K_v$ of $v$.
-If $w\in\operatorname{Spv}(K)$ such that $w$ is a specialization of $v$, then there is a henselization $K_w$ of $w$ such that $K_v\subset K_w\subset K_v^\text{t}$.
+If $w\in\operatorname{Spv}(K)$ is a specialization of $v$, then there is a henselization $K_w$ of $w$ such that $K_v\subset K_w\subset K_v^\text{t}$.
 \end{lemma}
@@
 
-For the second assertion, we assume $\ell^\infty\nmid [k:\mathbb{F}_p]$ and $\dim_{\mathbb{F}_\ell}(\Gamma / \ell\Gamma) = 2$.
+For the second assertion, we assume $\ell^\infty\nmid [k:\mathbb{F}_p]$ and $r=\dim_{\mathbb{F}_\ell}(\Gamma / \ell\Gamma) = 2$.
 By \cite{En72} Theorem 20.21, we have the following equation of supernatural numbers:
 \[ [L:K] = (\Gamma':\Gamma) \cdot [k':k]\cdot p^d \]
-for some $d \leq \infty$, where $\Gamma'$ denotes the value group and $k'$ the residue field of the unique extension of $w$ to $L$.
-By using this equation and the already proven statement from the first part, it suffices to show the following equivalences:
+for some $d \leq \infty$, where $\Gamma'$ denotes the value group and $k'$ is the residue field of the unique extension of $w$ to $L$.
+By using this equation and the already proven assertion from the first part, it suffices to show the following equivalences:
 \begin{enumerate}[(i)]
@@
 The extension of $v$ on $K_v^\text{t}$ has value group $\Gamma_v$, which is finitely generated.
-However, value groups of valuations on an algebraically closed field are divisible, hence $K_v^\text{t}$ can't be algebraically closed.
+However, value groups of valuations on an algebraically closed field are divisible, hence $K_v^\text{t}$ can't be algebraically closed and neither can the composite $K_{w_1}K_{w_2}$.
 
@@
 \end{tikzcd} \]
-Let us consider the morphism $f:X'\times_{k'} k\to X'$. For $P'\in X'_0$ we have as topological spaces:
+Let us consider the morphism $f:X'\times_{k'} k\to X'$. For $P'\in X'_0$ we have:
 \[ f^{-1}(P') = \operatorname{Spec} k(P')\times_{k'} k \]

### 2017-08-05 Master Thesis/synopsis.tex
Phrasing

@@
 Let $K$ be a field and $v\in\operatorname{Spv}(K)$ with a fixed henselization $K_v$ of $v$.
-If $w\in\operatorname{Spv}(K)$ is a specialization of $v$, then there is a henselization $K_w$ of $w$ such that $K_v\subset K_w\subset K_v^\text{t}$.
+If $w\in\operatorname{Spv}(K)$ such that $w$ is a specialization of $v$, then there is a henselization $K_w$ of $w$ such that $K_v\subset K_w\subset K_v^\text{t}$.
 \end{lemma}
@@
 The extension of $v$ on $K_v^\text{t}$ has value group $\Gamma_v$, which is finitely generated.
-However, value groups of valuations on an algebraically closed field are divisible, hence $K_v^\text{t}$ can't be algebraically closed and neither can the composite $K_{w_1}K_{w_2}$.
+However, value groups of valuations on an algebraically closed field are divisible, hence $K_v^\text{t}$ can't be algebraically closed.
 

### 2017-08-06 Master Thesis/synopsis.tex
Finally show that Spv(K) is a Jacobson space

--- 2017-08-05 Master Thesis/synopsis.tex
@@
 
-\begin{lemma}\label{spv-topology}
+\begin{lemma}\label{2.2-pre}
 Let $v,w\in\operatorname{Spv}(K)$. We have:
 \[\mathcal{O}_w\subset\mathcal{O}_v \iff w\in\overline{\{v\}} \]
-Thus $v$ is a closed point if and only if it has no non-trivial specializations.
-
-$\operatorname{Spv}(K)$ is a quasi-compact Jacobson space, i.e.
-every closed subset $A\subset\operatorname{Spv}(K)$ is the closure of its closed points.
-In particular, the set of closed points $|\operatorname{Spv}(K)|$ of $\operatorname{Spv}(K)$ is a quasi-compact $T_1$-space, endowed with the induced topology.
-\end{lemma}
-
-\begin{proof}
-The first assertion follows from \cite{ZS60} VI, §17 Theorem 38. The quasi-compactness of $\operatorname{Spv}(K)$ follows from \cite{ZS60} VI, §17 Theorem 40.
-
-Let $A\subset \operatorname{Spv}(K)$ be a closed set, $S\subset K$ finite, and assume $U_S\cap A\neq\varnothing$. Let $v\in U_S\cap A$ be a point which is not closed.
-\end{proof}
-
-\begin{lemma}\label{2.2-pre}
-For a fixed $v\in \operatorname{Spv}(K)$ there is a canonical bijection: 
+Thus $v$ is a closed point if and only if it has no non-trivial specializations. For a fixed $v\in \operatorname{Spv}(K)$ there is a canonical bijection: 
 \begin{align*}
@@
 \end{align*}
-Therefore $v$ is a closed point iff $k(v)$ only admits the trivial valuation, i.e.
+Therefore $v$ is a closed point if and only if $k(v)$ only admits the trivial valuation, i.e.
 $k(v)$ has characteristic $p>0$ and is algebraic over $\mathbb{F}_p$.
-In other words, $v$ is a closed point iff $\dim k(v)=0$.
-
+In other words, if $|\operatorname{Spv}(K)|$ denotes the set of all closed points in $\operatorname{Spv}(K)$, we have:
+\[ |\operatorname{Spv}(K)|=\{v\in\operatorname{Spv}(K)\mid \dim k(v)=0\} \]
 If $w\in\operatorname{Spv}_\text{rd}^{(i)}(K)$ for some $i\geq 1$ then there exists a unique $v(w)\in\operatorname{Spv}_\text{rd}^{(i-1)}(K)$ such that $\mathcal{O}_w\subset\mathcal{O}_{v(w)}$.
@@
 \begin{proof}
-See \cite{EP05} Section 2.3 and Corollary 4.1.4, \cite{ZS60} VI, §3 Theorem 3 and VI, §4 Corollary 2 to Theorem 4.
+The first equivalence follows from \cite{ZS60} VI, §17 Theorem 38.
+For the other statements, see \cite{EP05} Section 2.3 and Corollary 4.1.4, \cite{ZS60} VI, §3 Theorem 3 and VI, §4 Corollary 2 to Lemma.
+\end{proof}
+
+\begin{lemma}\label{spv-topology}
+$\operatorname{Spv}(K)$ is a quasi-compact Jacobson space, i.e.
+every closed subset $A\subset\operatorname{Spv}(K)$ is the closure of its closed points.
+In particular, $|\operatorname{Spv}(K)|$ is a quasi-compact $T_1$-space, endowed with the induced topology.
+\end{lemma}
+
+\begin{proof}
+The quasi-compactness of $\operatorname{Spv}(K)$ follows from \cite{ZS60} VI, §17 Theorem 40.
+
+Let $A\subset \operatorname{Spv}(K)$ be a closed set, $S\subset K$ finite, and assume $U_S\cap A\neq\varnothing$.
+Let us assume there exist a point $v\in U_S\cap A$ which is not closed.
+Let $\overline{S}$ be the image of $S\subset \mathcal{O}_v$ in $k(v)$.
+If $\operatorname{char}k(v)=0$, then $R=\mathbb{Z}[\overline{S}]\subset k(v)$ is a subring that is not a field, since $\overline{S}$ is finite.
+If $\operatorname{char} k(v)=p>0$, $k(v)/\mathbb{F}_p$ is an algebraic extension, thus we can find an element $T\in k(v)$ which is transcendental over $\mathbb{F}_p$.
+In this case, $R=\mathbb{F}_p[T, \overline{S}]\subset k(v)$ is a subring that is not a field.
+In both cases, by \cite{ZS60} VI, §4 Corollary 1 to Lemma, there exist a non-trivial valuation $\widetilde{v_1}\in\operatorname{Spv}k(v)$ with $\overline{S}\subset R\subset \mathcal{O}_{\widetilde{v_1}}$.
+By Lemma \ref{2.2-pre}, $\widetilde{v_1}$ corresponds to a valuation $v_1\in\operatorname{Spv}(K)$ with $S\subset \mathcal{O}_{v_1}\subsetneq \mathcal{O}_v$, i.e.
+$v_1\in \overline{\{v\}}\cap U_S$ and $\operatorname{rank}(v_1)>\operatorname{rank}(v)$.
+Repeating this process for $v_1$ and so on, we get a specialization chain in $U_S$:
+\[ \mathcal{O}_{v_n}\subset \cdots \subset\mathcal{O}_{v_1}\subset\mathcal{O}_v \]
+Since $K$ is assumed to be finite dimensional, Lemma \ref{rank-dimension-inequality} assures that this sequence ends with a closed point $v_n\in \overline{\{v\}}\cap U_S$.
 \end{proof}

### 2017-08-06 Master Thesis/synopsis.tex
Almost done!

@@
 If $\operatorname{char}k(v)=0$, then $R=\mathbb{Z}[\overline{S}]\subset k(v)$ is a subring that is not a field, since $\overline{S}$ is finite.
-If $\operatorname{char} k(v)=p>0$, $k(v)/\mathbb{F}_p$ is an algebraic extension, thus we can find an element $T\in k(v)$ which is transcendental over $\mathbb{F}_p$.
+If $\operatorname{char} k(v)=p>0$, $k(v)/\mathbb{F}_p$ is not an algebraic extension, thus we can find an element $T\in k(v)$ which is transcendental over $\mathbb{F}_p$.
 In this case, $R=\mathbb{F}_p[T, \overline{S}]\subset k(v)$ is a subring that is not a field.
-In both cases, by \cite{ZS60} VI, §4 Corollary 1 to Lemma, there exist a non-trivial valuation $\widetilde{v_1}\in\operatorname{Spv}k(v)$ with $\overline{S}\subset R\subset \mathcal{O}_{\widetilde{v_1}}$.
+In both cases, \cite{ZS60} VI, §4 Corollary 1 to Lemma guarantees the existence of a non-trivial valuation $\widetilde{v_1}\in\operatorname{Spv}k(v)$ with $\overline{S}\subset R\subset \mathcal{O}_{\widetilde{v_1}}$.
 By Lemma \ref{2.2-pre}, $\widetilde{v_1}$ corresponds to a valuation $v_1\in\operatorname{Spv}(K)$ with $S\subset \mathcal{O}_{v_1}\subsetneq \mathcal{O}_v$, i.e.
 $v_1\in \overline{\{v\}}\cap U_S$ and $\operatorname{rank}(v_1)>\operatorname{rank}(v)$.
-Repeating this process for $v_1$ and so on, we get a specialization chain in $U_S$:
-\[ \mathcal{O}_{v_n}\subset \cdots \subset\mathcal{O}_{v_1}\subset\mathcal{O}_v \]
-Since $K$ is assumed to be finite dimensional, Lemma \ref{rank-dimension-inequality} assures that this sequence ends with a closed point $v_n\in \overline{\{v\}}\cap U_S$.
+Repeating this process for $v_1$ and so on, we get a specialization chain of valuations in $U_S$:
+\[\mathcal{O}_v\supset \mathcal{O}_{v_1}\supset\cdots \supset\mathcal{O}_{v_n} \]
+Since $K$ is assumed to be finite dimensional, Lemma \ref{rank-dimension-inequality} assures that this sequence ends with a closed point $v_n\in \overline{\{v\}}\cap U_S\subset A\cap U_S$.
+
+Let $(U_i)_i$ be an open cover of $|\operatorname{Spv}(K)|$ and $U=\bigcup_i U_i$.
+Consider the closed set $A=\operatorname{Spv}(K)-U$.
+Since $\operatorname{Spv}(K)$ is a Jacobian space, $A$ is the closure of its closed points.
+However, $A$ doesn't contain any closed points, hence $A$ is empty.
+Therefore, $(U_i)_i$ already covers $\operatorname{Spv}(K)$ entirely and $|\operatorname{Spv}(K)|$ inherits the quasi-compactness from $\operatorname{Spv}(K)$.
 \end{proof}
@@
 
-Before we start proving Pop's Theorem \ref{thm:main-result}, we need some tools that will help us to construct a group isomorphism $K'^\times\to K^\times$ from the modulo $n$ isomorphisms $\varphi_n: K'^\times/n\to K^\times/n$.
+Before we start proving Pop's Theorem \ref{thm:main-result}, we need some tools that will help us construct a group isomorphism $\varphi: K'^\times\to K^\times$ from the modulo $n$ isomorphisms $\varphi_n: K'^\times/n\to K^\times/n$ we constructed in the last chapter.
 
@@
 \item Free or finite abelian groups are free-by-finite.
-\item Finitely generated abelian groups are free-by-finite by the structure theorem for finitely generated $\mathbb{Z}$-modules.
+\item Finitely generated abelian groups are free-by-finite by the structure theorem.
 \item Any subgroup of a free-by-finite group is free-by-finite as well.
@@
 The left abelian group is finitely generated, particularly free-by-finite. The right group is free, hence also free-by-finite. By (ii), $A/A'\cong A''$ is free-by-finite as well.
-\item It suffices to prove this assertion for finite groups and for $A=\mathbb{Z}$.
+\item It suffices to show this assertion for finite groups and for $A=\mathbb{Z}$.
 For finite $A$ we have $nA = 0$ for $n = \#A$.
@@
 \end{tikzcd} \]
-The left group is free-by-finite by (i), and by Lemma \ref{5.2} (ii), we only need to prove that the right group is free-by-finite.
+The left group is free-by-finite by (i), and by Lemma \ref{5.2} (ii), we only need to show that $F_1^\times/F^\times$ is free-by-finite.
 

### 2017-08-11 Master Thesis/synopsis.tex
Fix typos during reread

@@
 
-Let $A\subset \operatorname{Spv}(K)$ be a closed set, $S\subset K$ finite, and assume $U_S\cap A\neq\varnothing$.
-Let us assume there exist a point $v\in U_S\cap A$ which is not closed.
+Let $A\subset \operatorname{Spv}(K)$ be a closed set, $S\subset K$ finite, and let $U_S\cap A\neq\varnothing$.
+Assume there exist a point $v\in U_S\cap A$ which is not closed.
 Let $\overline{S}$ be the image of $S\subset \mathcal{O}_v$ in $k(v)$.
@@
 Repeating this process for $v_1$ and so on, we get a specialization chain of valuations in $U_S$:
-\[\mathcal{O}_v\supset \mathcal{O}_{v_1}\supset\cdots \supset\mathcal{O}_{v_n} \]
+\[\mathcal{O}_v\supsetneq \mathcal{O}_{v_1}\supsetneq\cdots \supsetneq\mathcal{O}_{v_n} \]
 Since $K$ is assumed to be finite dimensional, Lemma \ref{rank-dimension-inequality} assures that this sequence ends with a closed point $v_n\in \overline{\{v\}}\cap U_S\subset A\cap U_S$.
@@
 Conversely, taking the fixed field of the decomposition group of an extension $\overline{v}'$ of $v$ in $K^\text{s}$ yields a henselization of $K$ with respect to $v$.
-Any two of such henselizations are $K$-conjugate.
+Any two of such henselizations are conjugate.
 \end{theorem}
@@
 Choose a henselization $k_{v|}$ of $k$ such that $k_{v|}\subset K$ and consider the absolute inertia field $k_{v|}^\text{t}$ of $k_{v|}$.
-Since $k(v)$ is separably closed, we have $k_{v|}^\text{t}\subset K$.
+Since $k(v)$ is separably closed, we have $k_{v|}^\text{t}\subset K$, i.e.
+$G_K\subset G_{k_{v|}^\text{t}}$.
 Since $v|\in \operatorname{Spv} ({k_{v|}^\text{t}})$ is discrete and henselian, \cite{Ka79} §5, Theorem 1 shows $\operatorname{cd}\big(G_{k_{v|}^\text{t}}\big)\leq 2$.
@@
 \begin{lemma}\label{2.4-closedset}
-Let $f\in\mathrm{H}^3(K)$. Then the support of $f$ is closed in $|\operatorname{Spv}(K)|$:
-\[ \operatorname{supp}(f) = \{w\in |\operatorname{Spv}(K)|\mid \operatorname{res}_{K_w/K}(f)\neq 0 \} \]
+Let $\alpha\in\mathrm{H}^3(K)$. Then the support of $\alpha$ is closed in $|\operatorname{Spv}(K)|$:
+\[ \operatorname{supp}(\alpha) = \{w\in |\operatorname{Spv}(K)|\mid \operatorname{res}_{K_w/K}(\alpha)\neq 0 \} \]
 where $\operatorname{res}_{K_w/K}: \mathrm{H}^3(K)\to \mathrm{H}^3(K_w)$.
@@
 Therefore, $f$ is injective as well and its image is contained in $\bigoplus\mathrm{H}^3(K_y)$.
-Now, Lemma \ref{2.3} (i) yields:
-\[ \operatorname{supp}(f) = \bigcap_{\mathfrak{X}\in \mathfrak{M}(K)} \pi_\mathfrak{X}^{-1}(\{ y\in\mathfrak{X}_0\mid \operatorname{res}_{K_y/K}(f)\neq 0 \}) \]
-Note that the right set is closed in $\operatorname{Spv}(K)$.
+Therefore, the set
+\[ A_\mathfrak{X} = \{ y\in\mathfrak{X}_0\mid \operatorname{res}_{K_y/K}(\alpha)\neq 0 \} \]
+is finite, thus closed in $\mathfrak{X}$. Now Lemma \ref{2.3} (i) yields
+\[ \operatorname{supp}(\alpha) = \bigcap_{\mathfrak{X}\in \mathfrak{M}(K)} \pi_\mathfrak{X}^{-1}(A_\mathfrak{X}) \]
+and the right-hand side is closed in $\operatorname{Spv}(K)$.
 \end{proof}
@@
 
-Let $f\in\mathrm{H}^3(K)$ such that $\operatorname{res}_{K_w/K}(f) = 0$ for all $w\in |\operatorname{Spv}(K)|$.
-There exists a finitely generated subfield $K_0$ of $K$ such that $f = \operatorname{res}_{K/K_0}(g)$ for some $g\in\mathrm{H}^3(K_0)$ and $K/K_0$ is algebraic.
-Let $\{K_i\mid i\in I\}$ be the set of all finite subextensions of $K/K_0$.
-For $w\in |\operatorname{Spv}(K)|$ we denote the restriction of $w$ to $K_i$ with $w$ as well.
+Let $\alpha\in\mathrm{H}^3(K)$ such that $\operatorname{res}_{K_w/K}(\alpha) = 0$ for all $w\in |\operatorname{Spv}(K)|$.
+There exists a finitely generated subfield $K_0$ of $K$ such that $\alpha = \operatorname{res}_{K/K_0}(\beta)$ for some $\beta\in\mathrm{H}^3(K_0)$ and $K/K_0$ is algebraic.
+Let $\{K_i\mid i\in I\}$ be the set of all finite extensions of $K_0$ contained in $K$.
 
 Let $w\in |\operatorname{Spv}(K)|$.
+We denote the restriction of $w$ to $K_i$ with $w$ as well.
 Since $\mathrm{H}^3(K_w) = \varinjlim_I \mathrm{H}^3((K_i)_w)$ there exists an $i = i(w)\in I$ such that $\operatorname{res}_{(K_i)_w/K}(g) = 0$.
@@
 \[ \bigcup_{w\in S}\varphi^{-1}_{i(w)}(U_{i(w)}) = |\operatorname{Spv}(K)| \]
-Let $j\in I$ such that $K_j$ is the composite of the fields $K_{i(w)},\ w\in S$. We also have:
+Let $j\in I$ such that $K_j$ is the composite of the fields $K_{i(w)},\ w\in S$. Then we have:
 \[ \bigcup_{w\in S}\varphi^{-1}_{i(w)j}(U_{i(w)}) = |\operatorname{Spv}(K_j)| \]
 Let $v\in |\operatorname{Spv}(K_j)|$ and let $i = i(w),\ w\in S$ such that $v\in\varphi_{ij}^{-1}(U_{i(w)})$. We have:
-\[ \operatorname{res}_{(K_j)_v/K_j}(\operatorname{res}_{K_j/K_0}(g)) = \operatorname{res}_{(K_j)_v/(K_i)_v}(\operatorname{res}_{(K_i)_v/K_0}(g)) = 0 \]
-Since $K_j$ is finitely generated we obtain $\operatorname{res}_{K_j/K_0}(g) = 0$ with Lemma \ref{2.4-finitelygenerated}, thus $f = \operatorname{res}_{K/K_0}(g) = 0$.
+\[ \operatorname{res}_{(K_j)_v/K_j}(\operatorname{res}_{K_j/K_0}(\beta)) = \operatorname{res}_{(K_j)_v/(K_i)_v}(\operatorname{res}_{(K_i)_v/K_0}(\beta)) = 0 \]
+Since $K_j$ is finitely generated we obtain $\operatorname{res}_{K_j/K_0}(\beta) = 0$ with Lemma \ref{2.4-finitelygenerated}, thus $\alpha = \operatorname{res}_{K/K_0}(\beta) = 0$.
 \end{proof}
@@
 
-\begin{lemma}\label{2.5}
+\begin{proposition}\label{2.5}
 Let $K$ be henselian with respect to $w\in|\operatorname{Spv}(K)|$.
@@
 \end{cases} \]
-\end{lemma}
+\end{proposition}
 
@@
 Let $G_\ell$ be an $\ell$-Sylow group of $G_K$.
-By Theorem \ref{theorem-sylowgroups-inertiagroup}, the $\ell$-Sylow group of $T$ isomorphic to $\mathbb{Z}_\ell^r$, where $r =\dim_{\mathbb{F}_\ell}(\Gamma/\ell\Gamma)$.
+By Theorem \ref{theorem-sylowgroups-inertiagroup}, the $\ell$-Sylow group of $T$ isomorphic to $\mathbb{Z}_\ell^r$ where $r =\dim_{\mathbb{F}_\ell}(\Gamma/\ell\Gamma)$.
 Furthermore, $G_k\subset G_{\mathbb{F}_p}\cong\widehat{\mathbb{Z}}$, thus the $\ell$-Sylow group $G_{k,\ell}$ of $G_k$ is either isomorphic to $\mathbb{Z}_\ell$ or trivial.
@@
 since $\operatorname{cd}(\mathbb{Z}_\ell)=1$.
-Thus $\mathrm{H}^3(G_\ell, \mathbb{Z}/\ell\mathbb{Z}) =0$ iff $G_k(\ell) = 0$ or $r < 2$.
+Thus $\mathrm{H}^3(G_\ell, \mathbb{Z}/\ell\mathbb{Z}) =0$ if and only if $G_k(\ell) = 0$ or $r < 2$.
 Now let $G_k(\ell) \cong\mathbb{Z}_\ell$ and $r = 2$.
@@
 \[ [L:K] = (\Gamma':\Gamma) \cdot [k':k]\cdot p^d \]
-for some $d \leq \infty$, where $\Gamma'$ denotes the value group and $k'$ is the residue field of the unique extension of $w$ to $L$.
+for some $d \leq \infty$, where $\Gamma'$ denotes the value group and $k'$ denotes the residue field of the unique extension of $w$ to $L$.
 By using this equation and the already proven assertion from the first part, it suffices to show the following equivalences:
@@
 For (ii) let $T'$ be the absolute inertia group of $L$.
-By Theorem \ref{theorem-sylowgroups-inertiagroup}, the $\ell$-Sylow group of $T'$ is isomorphic to $\mathbb{Z}_\ell^{r'}$, where $r'=\dim_{\mathbb{F}_\ell}(\Gamma'/\ell\Gamma')$.
+By Theorem \ref{theorem-sylowgroups-inertiagroup}, the $\ell$-Sylow group of $T'$ is isomorphic to $\mathbb{Z}_\ell^{r'}$ where $r'=\dim_{\mathbb{F}_\ell}(\Gamma'/\ell\Gamma')$.
 Furthermore, we have $(\Gamma':\Gamma) = (T:T')$ and the following exact sequence of $\ell$-Sylow groups:
@@
 
-By Lemma \ref{2.5}, $E_0/E$, and also $E'_0/E'$, satisfy the condition of Proposition \ref{2.6} for both $\ell_1$ and $\ell_2$.
+By Proposition \ref{2.5}, $E_0/E$, and also $E'_0/E'$, satisfy the condition of Proposition \ref{2.6} for both $\ell_1$ and $\ell_2$.
 Thus there exists a unique henselian valuation $w'_{E'}\in|\operatorname{Spv}(E')|$.
@@
 Choose a prime $\ell\neq \operatorname{char} k(w')$ from $\ell_1,\ell_2$.
-Since $\mathrm{H}^3(E'_0(\mu_{\ell}),\mathbb{Z}/\ell\mathbb{Z})\neq 0$ by Lemma \ref{2.6}, we get from Lemma \ref{2.5}:
+Since $\mathrm{H}^3(E'_0(\mu_{\ell}),\mathbb{Z}/\ell\mathbb{Z})\neq 0$ by Lemma \ref{2.6}, we get from Proposition \ref{2.5}:
 \[\operatorname{rrank}(w')= \operatorname{rrank}( \widetilde{w}') = \dim_{\mathbb{Q}}(\Gamma_{\widetilde{w}'}\otimes_\mathbb{Z}\mathbb{Q}) \geq \dim_{\mathbb{F}_\ell}(\Gamma_{\widetilde{w}'}/\ell\Gamma_{\widetilde{w}'})=2\]
@@
 
-Let $F$ be another henselization of $w$.
-Then $E, F$ are $K$-conjugate, thus $E',F'$ are $K'$-conjugate, hence $K'$-isomorphic and the corresponding henselian valuations of $E', F'$ are respectively mapped to each other.
+Let $F$ be another henselization of $w$. Then $E, F$ are conjugate, thus $E',F'$ are conjugate, hence $K'$-isomorphic and the corresponding henselian valuations of $E', F'$ are equivalent.
 \end{proof}
@@
 
-\begin{lemma}\label{3.2}
+\begin{proposition}\label{3.2}
 $\sigma$ induces a bijection $\sigma:\operatorname{Spv}_\text{d}^{(1)}(K) \to\operatorname{Spv}_\text{d}^{(1)}(K')$.
-If $E=K_v\subset\overline{K}$ is a henselization of $K$ with respect to $v\in\operatorname{Spv}_\text{d}^{(1)}(K)$, then $E'$ is a henselization of $K'$ with respect to $v'=\sigma(v)$.
+If $E=K_v$ is a henselization of $K$ with respect to $v\in\operatorname{Spv}_\text{d}^{(1)}(K)$, then $E'$ is a henselization of $K'$ with respect to $v'=\sigma(v)$.
 
@@
 which is induced by a unique isomorphism $\varphi_v^\text{s}:k(v')^\text{s}\to k(v)^\text{s}$. In particular $\varphi_v=\varphi_v^\text{s}|_{k(v')}:k(v')\to k(v)$ is an isomorphism.
-\end{lemma}
+\end{proposition}
 
@@
 Lemma \ref{3.1} shows that $\sigma$ induces a bijection $\sigma: \operatorname{Spv}_\text{d}^{(2)}(K)\to \operatorname{Spv}_\text{d}^{(2)}(K')$.
-For all $v\in\operatorname{Spv}(K)$ we choose a fixed henselization $K_v\subset\overline{K}$ with Lemma \ref{lemma-henselization-specialization}, such that if $v,w\in\operatorname{Spv}(K)$ and $w$ is a specialization of $v$ we have $K_v\subset K_w\subset K_v^\text{t}$.
-If $v$ is nontrivial, then in this case we get $w\in\operatorname{Spv}_\text{rd}^{(2)}(K)$ and $v=v(w)\in\operatorname{Spv}_\text{d}^{(1)}(K)$.
+For all $v\in\operatorname{Spv}(K)$ we choose a fixed henselization $K_v$ with Lemma \ref{lemma-henselization-specialization} such that if $v,w\in\operatorname{Spv}(K)$ and $w$ is a specialization of $v$, we have $K_v\subset K_w\subset K_v^\text{t}$.
+If $v$ is nontrivial, then $w\in\operatorname{Spv}_\text{rd}^{(2)}(K)$ and $v=v(w)\in\operatorname{Spv}_\text{d}^{(1)}(K)$.
 We set $K'_{w'} = (K_w)'$ for $w\in\operatorname{Spv}_\text{d}^{(2)}(K)$.
-Lemma \ref{3.1} shows that $K'_{w'}$ is a henselization of $K'$ with respect to $w'$.
+Lemma \ref{3.1} shows that $K'_{w'}$ is indeed a henselization of $K'$ with respect to $w'$.
 
@@
 Now assume (ii) holds.
-Then $k(v)$ is separably closed where $v=v(w_1)$.
+Let $v=v(w_1)\in \operatorname{Spv}_\text{d}^{(1)}(K_{w_1}K_{w_2})$.
+Then $k(v)$ is separably closed.
 We have $K_{w_1}K_{w_2}\subset K_v^\text{t}$ and hence $K_{w_1}K_{w_2}=K_v^\text{t}$ by Theorem \ref{theorem-absoluteinertia}.
-The extension of $v$ on $K_v^\text{t}$ has value group $\Gamma_v$, which is finitely generated.
+$\Gamma_v$ is equal to the value group of $v|_K\in\operatorname{Spv}_\text{d}^{(1)}(K)$, which is finitely generated.
 However, value groups of valuations on an algebraically closed field are divisible, hence $K_v^\text{t}$ can't be algebraically closed and neither can the composite $K_{w_1}K_{w_2}$.
@@
 
-By Lemma \ref{3.2} $\sigma$ induces a bijection $\sigma: X^{\phantom{'}}_0\to X'_0,\ P\mapsto P'$ and for any $P\in X_0$ an isomorphism $\varphi_P: k(P')\to k(P)$.
+By Proposition \ref{3.2} $\sigma$ induces a bijection $\sigma: X^{\phantom{'}}_0\to X'_0,\ P\mapsto P'$ and for any $P\in X_0$ an isomorphism $\varphi_P: k(P')\to k(P)$.
 \end{definition}
@@
 \begin{lemma}\label{4.1}
-Let $E$ be a number field and let $X$ be a smooth projective geometrically connected curve over $E$.
+Let $E$ be a number field and let $X$ be a smooth, projective, geometrically connected curve over $E$.
 Let $\pi:X'\to X$ be a connected Galois covering of degree $d$ which splits completely, i.e.
@@
 \begin{remark}
-Consider $\overline{K''} = \overline{K'}\otimes_{\overline{k'}}\overline{k}$ via the isomorphism $\varphi: \overline{k'}\to \overline{k}$. If we replace $\sigma: G_K\to G_{K'}$ with the composition
+Let $\overline{K''} = \overline{K'}\otimes_{\overline{k'}}\overline{k}$ via the isomorphism $\varphi: \overline{k'}\to \overline{k}$. If we replace $\sigma: G_K\to G_{K'}$ with the composition
 \[ \begin{tikzcd}
@@
 U_{P'}/n \rar["\varphi_n"] \dar["r_{P'}/n"'] & U_P/n \dar["r_P/n"]\\
-k(P')^\times/n \rar["\varphi_P/n"'] & k(P)^\times/n
+k(P')^\times/n \rar["\varphi_P^\times/n"'] & k(P)^\times/n
 \end{tikzcd} \]
@@
 \item Consider the field extension $K_P/K$.
-By Remark \ref{remark-localphi-extension}, without loss of generality, we can assume $K=K_P$.
+By Remark \ref{remark-localphi-extension}, we can assume $K=K_P$ without loss of generality.
 Since $\operatorname{char}k(P)=0$, by Hensel's lemma the map $r_P/n: U_P/n\to k(P)^\times/n$ is an isomorphism and the composition
@@
 \[ \begin{tikzcd}
-k(P')^\times/n \ar[r, "\cong"] \ar[d, "\varphi_P/n"'] & \mathrm{H}^1(k(P'),\mu_n) \ar[r, "\mathrm{inf}"] \ar[d]& \mathrm{H}^1(K'_{P'}, \mu_n) \ar[d, "\varphi_n"] \\
+k(P')^\times/n \ar[r, "\cong"] \ar[d, "\varphi_P^\times/n"'] & \mathrm{H}^1(k(P'),\mu_n) \ar[r, "\mathrm{inf}"] \ar[d]& \mathrm{H}^1(K'_{P'}, \mu_n) \ar[d, "\varphi_n"] \\
 k(P)^\times/n \ar[r, "\cong"'] & \mathrm{H}^1(k(P),\mu_n)  \ar[r, "\mathrm{inf}"'] & \mathrm{H}^1(K_{P}, \mu_n)
@@
 U_{P'} \ar[r, "\varphi"]\ar[d, "r_{P'}"'] & U_P \ar[d, "r_P"]\\
-k(P')^\times \ar[r, "\varphi_P"'] & k(P)^\times
+k(P')^\times \ar[r, "\varphi_P^\times"'] & k(P)^\times
 \end{tikzcd} \]
@@
 K'^{S'} \rar["\varphi^{S,T}", dashed] \dar["r_{S',T'}"'] & K^S \dar["r_{S,T}"]\\
-\displaystyle \prod_{P\in T}k(P')^\times \rar["\prod\varphi_P"'] & \displaystyle \prod_{P\in T}k(P)^\times
+\displaystyle \prod_{P\in T}k(P')^\times \rar["\prod\varphi_P^\times"'] & \displaystyle \prod_{P\in T}k(P)^\times
 \end{tikzcd} \phantom{\qquad (\star)} \]
@@
 Let $T_1, T_2\subset X_0 -  S$ be two $S$-admissible subsets and suppose $T_1\subset T_2$. By inspecting $(\star)$ we find:
-\[ r_{S,T_1}\circ\varphi^{S,T_2}=\Big(\prod_{P\in T_1}\varphi_P\Big)\circ r_{S',T_1'} = r_{S,T_1}\circ\varphi^{S,T_1} \]
+\[ r_{S,T_1}\circ\varphi^{S,T_2}=\Big(\prod_{P\in T_1}\varphi_P^\times\Big)\circ r_{S',T_1'} = r_{S,T_1}\circ\varphi^{S,T_1} \]
 Since $r_{S,T_1}$ is injective we get $\varphi^{S,T_2}=\varphi^{S,T_1}$.
@@
 Now if $T$ ranges over all $S$-admissible subsets of $X_0$, then $(\star)$ shows:
-\[ r_S\circ\varphi^S = \Big(\prod_{P\in X_0 -  S}\varphi_P \Big)\circ r_{S'} \]
+\[ r_S\circ\varphi^S = \Big(\prod_{P\in X_0 -  S}\varphi_P^\times \Big)\circ r_{S'} \]
 In other words, we can replace $T$ by $X_0 -  S$ in $(\star)$. If $S_1\subset S_2$ are two admissible subsets, then:
-\[ r_{S_2}\circ\varphi^{S_2}|_{K'^{S'_1}} = \Big(\prod_{P\in X_0 -  S_2}\varphi_P \Big)\circ r_{S'_2}|_{K'^{S'_1}} = r_{S_2}\circ\varphi^{S_1} \]
+\[ r_{S_2}\circ\varphi^{S_2}|_{K'^{S'_1}} = \Big(\prod_{P\in X_0 -  S_2}\varphi_P^\times \Big)\circ r_{S'_2}|_{K'^{S'_1}} = r_{S_2}\circ\varphi^{S_1} \]
 Hence, $\varphi^{S_1}=\varphi^{S_2}|_{K'^{S'_1}}$.

### 2017-08-11 Master Thesis/synopsis.tex
Fixing comma mistakes

@@
 \end{align*}
-Therefore $v$ is a closed point if and only if $k(v)$ only admits the trivial valuation, i.e.
+Therefore, $v$ is a closed point if and only if $k(v)$ only admits the trivial valuation, i.e.
 $k(v)$ has characteristic $p>0$ and is algebraic over $\mathbb{F}_p$.
@@
 \[ |\operatorname{Spv}(K)|=\{v\in\operatorname{Spv}(K)\mid \dim k(v)=0\} \]
-If $w\in\operatorname{Spv}_\text{rd}^{(i)}(K)$ for some $i\geq 1$ then there exists a unique $v(w)\in\operatorname{Spv}_\text{rd}^{(i-1)}(K)$ such that $\mathcal{O}_w\subset\mathcal{O}_{v(w)}$.
+If $w\in\operatorname{Spv}_\text{rd}^{(i)}(K)$ for some $i\geq 1$, then there exists a unique $v(w)\in\operatorname{Spv}_\text{rd}^{(i-1)}(K)$ such that $\mathcal{O}_w\subset\mathcal{O}_{v(w)}$.
 $w$ is henselian if and only if $v(w)$ and $w\mod v(w)$ are henselian.
@@
 Let $A\subset \operatorname{Spv}(K)$ be a closed set, $S\subset K$ finite, and let $U_S\cap A\neq\varnothing$.
-Assume there exist a point $v\in U_S\cap A$ which is not closed.
+Assume there exists a point $v\in U_S\cap A$ which is not closed.
 Let $\overline{S}$ be the image of $S\subset \mathcal{O}_v$ in $k(v)$.
-If $\operatorname{char}k(v)=0$, then $R=\mathbb{Z}[\overline{S}]\subset k(v)$ is a subring that is not a field, since $\overline{S}$ is finite.
+If $\operatorname{char}k(v)=0$, then $R=\mathbb{Z}[\overline{S}]\subset k(v)$ is a subring that is not a field since $\overline{S}$ is finite.
 If $\operatorname{char} k(v)=p>0$, $k(v)/\mathbb{F}_p$ is not an algebraic extension, thus we can find an element $T\in k(v)$ which is transcendental over $\mathbb{F}_p$.
@@
 Since two different valuations of rank $1$ are independent, at least one of $w_1$ and $w_2$ is of rank $2$.
-Now, assume that $w_1$ is, say, of rank $1$ and $w_2$ is of rank $2$.
+Now assume that $w_1$ is, say, of rank $1$ and $w_2$ is of rank $2$.
 Since they are dependent, it follows $v(w_2) = w_1$ and $w_1$ has a non-trivial specialization, a contradiction to $w_1\in |\operatorname{Spv}(K)|$.
@@
 Schmidt \ref{theorem-fkschmidt}, we get $v_1 = v_2$.
-Furthermore $k(v_1) = k(v_2)$ is henselian with respect to two different valuations of rank $1$, namely $w_1\mod v_1$ and $w_2\mod v_1$, making $k(v_1)$ separably closed.
+Furthermore, $k(v_1) = k(v_2)$ is henselian with respect to two different valuations of rank $1$, namely $w_1\mod v_1$ and $w_2\mod v_1$, making $k(v_1)$ separably closed.
 
@@
 $G_K\subset G_{k_{v|}^\text{t}}$.
-Since $v|\in \operatorname{Spv} ({k_{v|}^\text{t}})$ is discrete and henselian, \cite{Ka79} §5, Theorem 1 shows $\operatorname{cd}\big(G_{k_{v|}^\text{t}}\big)\leq 2$.
+Since $v|\in \operatorname{Spv}_\text{d}^{(1)} ({k_{v|}^\text{t}})$ is discrete and henselian, \cite{Ka79} §5, Theorem 1 shows $\operatorname{cd}\big(G_{k_{v|}^\text{t}}\big)\leq 2$.
 In particular, $\operatorname{cd}(G_K)\leq 2$.
@@
 \[ \mathrm{H}^1(T, \mathbb{Z}/\ell\mathbb{Z}(2)) \cong \mathrm{H}^1(T, \mu_{\ell}) \cong \operatorname{Hom}_\text{cont.}(T, \mu_{\ell}) \cong \operatorname{Hom}_\text{cont.}(T/V, \mu_{\ell}) \]
-By Theorem \ref{theorem-sylowgroups-inertiagroup}, $T/V$ is the free pro-$\mathcal{C}$ group of rank $1$, where $\mathcal{C}$ is the class of all finite groups with order prime to $p$.
+By Theorem \ref{theorem-sylowgroups-inertiagroup}, $T/V$ is the free pro-$\mathcal{C}$ group of rank $1$ where $\mathcal{C}$ is the class of all finite groups with order prime to $p$.
 Therefore, $\operatorname{Hom}_\text{cont.}(T/V, \mu_{\ell})\cong \mu_{\ell}$.
@@
 Kato's two-dimensional local-global principle to show the assertion for finitely generated $K$.
-Then, we'll use a quasi-compactness argument to see that we can drop the assumption that $K$ is finitely generated.
+Then we'll use a quasi-compactness argument to drop the assumption that $K$ is finitely generated.
 
@@
 $\mu_{\ell}\subset K$ and $K$ is not formally real.
-Now we can replace $\mathbb{Z}/\ell\mathbb{Z}$ by $\mathbb{Z}/\ell\mathbb{Z}(2) = \mu_\ell^{\otimes 2}$ and let us write $\mathrm{H}^3(L) = \mathrm{H}^3(L, \mathbb{Z}/\ell\mathbb{Z}(2))$.
+Now we can replace $\mathbb{Z}/\ell\mathbb{Z}$ by $\mathbb{Z}/\ell\mathbb{Z}(2)$ and let us write $\mathrm{H}^3(L) = \mathrm{H}^3(L, \mathbb{Z}/\ell\mathbb{Z}(2))$.
 
@@
 \[ {}_\ell\operatorname{Br}(k) \longrightarrow \prod_{\widetilde{w}\in\operatorname{Spv}^{(1)}(k)} {}_\ell\operatorname{Br}(k_{\widetilde{w}}) \]
-which is injective by the classical Hasse principle.
-Furthermore, its image is contained in $\bigoplus {}_\ell\operatorname{Br}(k_{\widetilde{w}})$.
-Combining these maps yields the following injection:
+which is injective by the classical Hasse principle. Moreover, its image is contained in $\bigoplus {}_\ell\operatorname{Br}(k_{\widetilde{w}})$. Combining these maps yields the following injection:
 \[\Phi: \mathrm{H}^3(K)\longrightarrow \bigoplus_{x\in\mathfrak{X}_1}\  \bigoplus_{\substack{w\in\operatorname{Spv}_\text{rd}^{(2)}(K) \\ v(w) = v_x}} \mathrm{H}^3(K_w) \]
@@
 According to \cite{Sa87} Theorem 5.3, $g$ is injective.
-Now, $\psi\circ\varphi$ can be identified with the injective map $\Phi$ from the proof of the previous Lemma.
+Now $\psi\circ\varphi$ can be identified with the injective map $\Phi$ from the proof of the previous Lemma.
 Thus its image is contained in $\bigoplus\mathrm{H}^3(K_w)$.
 Therefore, $f$ is injective as well and its image is contained in $\bigoplus\mathrm{H}^3(K_y)$.
-Therefore, the set
+This shows that the set
 \[ A_\mathfrak{X} = \{ y\in\mathfrak{X}_0\mid \operatorname{res}_{K_y/K}(\alpha)\neq 0 \} \]
@@
 We denote the restriction of $w$ to $K_i$ with $w$ as well.
-Since $\mathrm{H}^3(K_w) = \varinjlim_I \mathrm{H}^3((K_i)_w)$ there exists an $i = i(w)\in I$ such that $\operatorname{res}_{(K_i)_w/K}(g) = 0$.
-By Lemma \ref{2.4-closedset} there is an open neighborhood $U_{i(w)}\subset |\operatorname{Spv}(K_i)|$ of $w$ such that $\operatorname{res}_{(K_i)_v/K}(g)=0$ for all $v\in U_{i(w)}$.
+Since $\mathrm{H}^3(K_w) = \varinjlim_I \mathrm{H}^3((K_i)_w)$, there exists an $i = i(w)\in I$ such that $\operatorname{res}_{(K_i)_w/K}(g) = 0$.
+By Lemma \ref{2.4-closedset}, there is an open neighborhood $U_{i(w)}\subset |\operatorname{Spv}(K_i)|$ of $w$ such that $\operatorname{res}_{(K_i)_v/K}(g)=0$ for all $v\in U_{i(w)}$.
 For $K_i\subset K_j\subset K$ let
@@
 \[ \varphi_{ij}: |\operatorname{Spv}(K_j)|\to |\operatorname{Spv}(K_i)| \]
-be the restriction maps. They are continuous. Since $|\operatorname{Spv}(K)|$ is quasi-compact by Lemma \ref{spv-topology} we can find a finite set $S\subset |\operatorname{Spv}(K)|$ such that:
+be the restriction maps. They are continuous. Since $|\operatorname{Spv}(K)|$ is quasi-compact by Lemma \ref{spv-topology}, we can find a finite set $S\subset |\operatorname{Spv}(K)|$ such that:
 \[ \bigcup_{w\in S}\varphi^{-1}_{i(w)}(U_{i(w)}) = |\operatorname{Spv}(K)| \]
@@
 Now $E'$ contains a henselization $K'_{w'}$ with respect to $w' = w'_{E'}|_{K'}$.
-By applying the same argument to $\sigma^{-1}$ we find that $E' = K'_{w'}$.
+By applying the same argument to $\sigma^{-1}$ we see that $E' = K'_{w'}$.
 
@@
 $\Gamma_v$ is equal to the value group of $v|_K\in\operatorname{Spv}_\text{d}^{(1)}(K)$, which is finitely generated.
-However, value groups of valuations on an algebraically closed field are divisible, hence $K_v^\text{t}$ can't be algebraically closed and neither can the composite $K_{w_1}K_{w_2}$.
+However, value groups of valuations on an algebraically closed field are divisible and in particular not finitely generated, hence $K_v^\text{t}$ can't be algebraically closed and neither can the composite $K_{w_1}K_{w_2}$.
 
@@
 \end{tikzcd} \]
-By limit process we see that $\sigma$ induces a bijection $\sigma: \operatorname{Spv}_\text{d}^{(1)}(\overline{K})\to \operatorname{Spv}_\text{d}^{(1)}(\overline{K'})$.
-Let $\overline{v}\in \operatorname{Spv}_\text{d}^{(1)}(\overline{K})$ and $G_{\overline{v}}\subset G_K$ be the decomposition group of $\overline{v}$.
-Then we have $\sigma(G_{\overline{v}}) = G_{\sigma(\overline{v})}$.
-It follows for any $g\in G_K$:
-\[G_{\sigma(g\overline{v})} = \sigma(G_{g\overline{v}}) = \sigma(g^{-1} G_{\overline{v}}g) = \sigma(g)^{-1} G_{\sigma(\overline{v})}\sigma(g) = G_{\sigma(g)\sigma(\overline{v})}\]
-Therefore, $\sigma(g\overline{v})=\sigma(g)\sigma(\overline{v})$ by Corollary \ref{corollary-fkschmidt}.
+By limit process we see that $\sigma$ induces a bijection: 
+\[\sigma: \operatorname{Spv}_\text{d}^{(1)}(\overline{K})\to \operatorname{Spv}_\text{d}^{(1)}(\overline{K'}),\ \overline{v}\mapsto\overline{v}'\]
 \end{remark}
+
+\begin{lemma}\label{lemma-sigma-groupaction-valuation}
+Let $\overline{v}\in \operatorname{Spv}_\text{d}^{(1)}(\overline{K})$, $g\in G_K$, and $G_{\overline{v}}\subset G_K$ be the decomposition group of $\overline{v}$.
+Then we have $\sigma(G_{\overline{v}}) = G_{\sigma(\overline{v})}$ and $\sigma(g\overline{v})=\sigma(g)\sigma(\overline{v})$.
+\end{lemma}
+
+\begin{proof}
+The first equality is just a rephrasing of $(K_{\overline{v}})' = K'_{\overline{v}'}$. We have:
+\[G_{\sigma(g\overline{v})} = \sigma(G_{g\overline{v}}) = \sigma(g G_{\overline{v}}g^{-1}) = \sigma(g) G_{\sigma(\overline{v})}\sigma(g)^{-1} = G_{\sigma(g)\sigma(\overline{v})}\]
+Now the second assertion follows by Corollary \ref{corollary-fkschmidt}.
+\end{proof}
 
@@
 \[ \mathrm{H}^1(K',\mu_n)\cong K'^\times/n,\qquad \mathrm{H}^1(K,\mu_n)\cong K^\times/n \]
-Our goal is to establish an isomorphism between the corresponding cohomology groups.
+In this section, our goal is to establish an isomorphism between the corresponding cohomology groups.
 
@@
 \begin{proof}
-First we assume that $k/\mathbb{Q}$ is a Galois extension and $X$ has a $k$-rational point $P_0$. We can consider $k'$ as a subfield of $k$ via: 
+First, we assume that $k/\mathbb{Q}$ is a Galois extension and $X$ has a $k$-rational point $P_0$. We can consider $k'$ as a subfield of $k$ via: 
 \[\begin{tikzcd}
@@
 \end{tikzcd} \]
-Moreover, $\sigma:G_k\to G_{k'}$ is induced by a unique isomorphism $\varphi:\overline{k'}\to\overline{k}$ which restricts to an isomorphism $\varphi: k'\to k$.
+Furthermore, $\sigma:G_k\to G_{k'}$ is induced by a unique isomorphism $\varphi:\overline{k'}\to\overline{k}$ which restricts to an isomorphism $\varphi: k'\to k$.
 \end{lemma}
@@
 \[ \begin{tikzcd}
-G_K \ar[r]\ar[d, "\sigma"'] & G_k\\
-G_{K'}\ar[ru]
+G_K \ar[d]\ar[r, "\sigma"] & G_{K'} \ar[ld]\\
+G_{k} &
 \end{tikzcd} \]
@@
 \[ 1 \longrightarrow U_w/n \longrightarrow L^\times /n\longrightarrow \mathbb{Z}/n\mathbb{Z}\longrightarrow 0 \]
-Since $k(w)$ is separably closed, with $\operatorname{char} k(w)=0$ and Hensel's lemma we can see that $U_w$ is a divisible group, hence we get an isomorphism $L^\times /n\cong \mathbb{Z}/n\mathbb{Z}$ induced by $w$.
+Since $k(w)$ is separably closed, with $\operatorname{char} k(w)=0$ and Hensel's lemma, we can see that $U_w$ is a divisible group, hence we get an isomorphism $L^\times /n\cong \mathbb{Z}/n\mathbb{Z}$ induced by $w$.
 Therefore, the canonical map $K^\times/n\to L^\times/n$ can be identified with the map $K^\times/n \to\mathbb{Z}/n\mathbb{Z}$ induced by $v_P$.
@@
 By Remark \ref{remark-localphi-extension}, we can assume $K=K_P$ without loss of generality.
-Since $\operatorname{char}k(P)=0$, by Hensel's lemma the map $r_P/n: U_P/n\to k(P)^\times/n$ is an isomorphism and the composition
+Since $\operatorname{char}k(P)=0$, by Hensel's lemma, the map $r_P/n: U_P/n\to k(P)^\times/n$ is an isomorphism and the composition
 \[\begin{tikzcd}
@@
 \[\ker(g) = \ker( (A/A')/(A_\text{tor}/A'_\text{tor}) \to A''/A''_\text{tor} ) = A''_\text{tor}/(A_\text{tor}/A'_\text{tor})\]
-Since $A''_\text{tor}$ is finite, $\operatorname{coker}(f)\cong \ker(g)$ is finite, too.
+Since $A''_\text{tor}$ is finite, $\operatorname{coker}(f)\cong \ker(g)$ is finite too.
 Applying (i) on $f$, we can see that $p^{-1}(A''_\text{tor})/A_\text{tor}$ is free-by-finite.
@@
 \end{align*}
-$K'^{S'}$ is defined similarly. For finite subsets $S$ of $X_0$, the corresponding subsets of $X_0'$ will be denoted by $S'$, i.e. $S'=\{\sigma P\mid P\in S \} $.
+$K'^{S'}$ is defined similarly. For finite subsets $S\subset X_0$ the corresponding subset of $X_0'$ will be denoted by $S'$, i.e. $S'=\{\sigma P\mid P\in S \} $.
 \end{definition}
@@
 \begin{theorem}[\textit{Weak Approximation Theorem}]\label{weak-approximation}
-Let $v_1,\ldots,v_n\in X_0$ be inequivalent discrete valuations, $x_1,\ldots,x_n\in K$ and $a_1,\ldots,a_n\in\mathbb{Z}$. Then there is some $x\in K$, such that for all $i$ we have:
+Let $v_1,\ldots,v_n\in X_0$ be inequivalent discrete valuations, $x_1,\ldots,x_n\in K$ and $a_1,\ldots,a_n\in\mathbb{Z}$. Then there exist some $x\in K$ such that for all $i=1,\ldots,n$ we have:
 \[ v_i(x-x_i) = a_i \]
@@
 \item By Corollary \ref{mordell-weil-corollary}, the Picard group of $X$ is finitely generated.
-Let $D_1,\ldots, D_r$ be Weil divisors of $X$ such that the associated line bundles generate $\operatorname{Pic}(X)$ and similarly let $D'_1,\ldots,D'_s$ generate $\operatorname{Pic}(X')$.
-Clearly any finite $S\subset X_0$ such that
+Let $D_1,\ldots, D_r$ be Weil divisors of $X$ such that the associated line bundles generate $\operatorname{Pic}(X)$ and, similarly, let $D'_1,\ldots,D'_s$ generate $\operatorname{Pic}(X')$.
+Clearly, any finite subset $S\subset X_0$ such that
 \[ \bigcup_{i=1}^r \operatorname{supp}(D_i)\subset S\quad\text{and}\quad \bigcup_{j=1}^s\operatorname{supp}(D'_j)\subset S' \]
@@
 By requirement $r_{P_2}(x)=a$ and $a\in k^\times$ is not of finite order, thus $x\not\in C_2$ and $\operatorname{rank}(C_2) < \operatorname{rank}(C_1)$.
-By repeating this process until $C_r=1$, we get points $P_2,\ldots,P_r\in X_0-S$, such that the following map is injective:
+By repeating this process until $C_r=1$, we get points $P_2,\ldots,P_r\in X_0-S$ such that the following map is injective:
 \[ C_1\to\prod_{i=2}^{r} k(P_i)^\times,\ x\mapsto(r_{P_i}(x))_i \]
-Finally for $T=\{P_1,\ldots,P_r\}\subset X_0-S$, the map $r_{S,T}$ is injective. For the second assertion, note the following isomorphism:
+Finally for $T=\{P_1,\ldots,P_r\}\subset X_0-S$ the map $r_{S,T}$ is injective. For the second assertion note the following isomorphism:
 \[ \operatorname{coker}(r_{S,T}) \cong\operatorname{coker}\Big(K^S/k^\times \to \Big(\prod_{P\in T}k(P)^\times \Big)\Big/k^\times \Big) \]
@@
 
-For the existence of $\varphi$, let $(S,T)$ be an admissible pair. First, we show that there is a unique homomorphism $\varphi^{S,T}$, making the following diagram commutative:
+For the existence of $\varphi$ let $(S,T)$ be an admissible pair. First, we show that there is a unique homomorphism $\varphi^{S,T}$ which makes the following diagram commutative:
 \[ (\star) \qquad \begin{tikzcd}
@@
 \end{tikzcd} \phantom{\qquad (\star)} \]
-The vertical maps are injective by definition. Thus, it suffices to show that the following composition is trivial:
+The vertical maps are injective by definition. Thus it suffices to show that the following composition is trivial:
 \[\varrho:K'^{S'}\longrightarrow\prod_{P\in T} k(P')^\times \longrightarrow \prod_{P\in T} k(P)^\times\longrightarrow\operatorname{coker}(r_{S,T})\] 
@@
 Since $r_{S,T_1}$ is injective we get $\varphi^{S,T_2}=\varphi^{S,T_1}$.
-Hence, $\varphi^{S,T}$ is actually independent of $T$, thus we can omit the $T$ from its notation.
+Hence $\varphi^{S,T}$ is actually independent of $T$, thus we can omit the $T$ from its notation.
 Now if $T$ ranges over all $S$-admissible subsets of $X_0$, then $(\star)$ shows:
@@
 \[ r_{S_2}\circ\varphi^{S_2}|_{K'^{S'_1}} = \Big(\prod_{P\in X_0 -  S_2}\varphi_P^\times \Big)\circ r_{S'_2}|_{K'^{S'_1}} = r_{S_2}\circ\varphi^{S_1} \]
-Hence, $\varphi^{S_1}=\varphi^{S_2}|_{K'^{S'_1}}$.
-Thus, we can combine the isomorphisms $\varphi^S$, where $S$ ranges over the admissible subsets of $X_0$, into an isomorphism $\varphi = \varinjlim \varphi^S: K'^\times \to K^\times$.
+Thus $\varphi^{S_1}=\varphi^{S_2}|_{K'^{S'_1}}$.
+Hence we can combine the isomorphisms $\varphi^S$, where $S$ ranges over the admissible subsets of $X_0$, into an isomorphism $\varphi = \varinjlim \varphi^S: K'^\times \to K^\times$.
 
 The commutativity of the desired diagram follows from the commutativity of $(\star)$ with $T=X_0 -  S$.
-Note that with Lemma \ref{4.5} (i), we can find for all $x\in U_{P'}$ an admissible set $S\subset X_0 -  \{P\}$ such that $x\in K'^{S'}$.
+Note that with Lemma \ref{4.5} (i) we can find for all $x\in U_{P'}$ an admissible set $S\subset X_0 -  \{P\}$ such that $x\in K'^{S'}$.
 \end{proof}
@@
 \item \textit{Case 2:} Let $x + y\neq 0$ and $S\subset X_0$ be an admissible finite subset such that $\varphi(x + y)$, $\varphi(x)$ and $\varphi(y)$ are contained in $K^S$.
-Looking at the commutative diagram in \mbox{Step 1} we get
+Looking at the commutative diagram in \mbox{Step 1} we get for all $P\in X_0 -  S$:
 \begin{align*}
@@
 \end{align*}
-for all $P\in X_0 -  S$.
 Note that $\varphi_P: k(P')\to k(P)$ is a field isomorphism and that the residue maps $r_{P'}: \mathcal{O}_{P'} \to k(P')$ are ring homomorphisms.
@@
 For any finite subextension $L/K$ of $\overline{K}/K$ there exists an isomorphism $\varphi_L: L'\to L$.
-If $K\subset L\subset M$ are finite subextensions, we have $\varphi_M|_{L'} = \varphi_{L'}$, since $r_Q(\varphi_M(x)) = r_Q(\varphi_L(x))$ for all $x\in L'^\times$ and almost all $Q\in (X_M)_0$ according to Remark \ref{3.3} and Step 1.
+If $K\subset L\subset M$ are finite subextensions, we have $\varphi_M|_{L'} = \varphi_{L'}$ since $r_Q(\varphi_M(x)) = r_Q(\varphi_L(x))$ for all $x\in L'^\times$ and almost all $Q\in (X_M)_0$ according to Remark \ref{3.3} and Step 1.
 By passing to limits we obtain an isomorphism $\varphi: \overline{K}'\to \overline{K}$.
@@
 
-Let $g\in G_K$ and $\overline{v}\in\operatorname{Spv}_\text{d}^{(1)}(\overline{K})$. By Remark \ref{3.3} we have:
+Let $g\in G_K$ and $\overline{v}\in\operatorname{Spv}_\text{d}^{(1)}(\overline{K})$. By Lemma \ref{lemma-sigma-groupaction-valuation} we have:
 \[ \sigma(g)\sigma(\overline{v}) = \sigma(g\overline{v}) = \tau(g\overline{v}) = \tau(g)\tau(\overline{v}) = \tau(g) \sigma(\overline{v}) \]
@@
 for all $g\in G_K$.
-For $\overline{v}\in\operatorname{Spv}_\text{d}^{(1)}(\overline{K})$ we have therefore $G_{\overline{v}} = h G_{\overline{v}} h^{-1} = G_{h\overline{v}}$, thus $h\in G_{\overline{v}}$ by Corollary \ref{corollary-fkschmidt}.
+For $\overline{v}\in\operatorname{Spv}_\text{d}^{(1)}(\overline{K})$ we have $G_{\overline{v}} = h G_{\overline{v}} h^{-1} = G_{h\overline{v}}$, thus $h\in G_{\overline{v}}$ by Corollary \ref{corollary-fkschmidt}.
 Since $\overline{v}$ was arbitrary, we get $h\in G_{\overline{v}}$ for all $\overline{v}$.
-Again, by Corollary \ref{corollary-fkschmidt}, we have $h=1$ and thus $\varphi = \psi$.
-\end{proof}
+Again, by Corollary \ref{corollary-fkschmidt}, we get $h=1$ and thus $\varphi = \psi$.
+\end{proof}

### 2017-08-11 Master Thesis/introduction.tex
Fixing comma mistakes
Fix

@@
 In the second chapter, we will show that an isomorphism $\sigma: G_K\to G_{K'}$ induces bijections on the equivalence classes of defectless valuations of $K$ and $K'$ using the cohomological characterization of henselian fields.
-Using the Theorem of Neukirch-Uchida \ref{thm:neukirch-uchida}, we will also see that $\sigma$ induces a field isomorphism between the residue fields of corresponding defectless valuations of rational rank $1$.
+Using the Theorem of Neukirch-Uchida \ref{thm:neukirch-uchida}, we will also see that $\sigma$ induces a field isomorphism between the residue fields of corresponding discrete valuations.
 We will proceed to show that $\sigma$ induces group isomorphisms $K'^\times/n\to K^\times/n$ for all $n\in\mathbb{N}$.
@@
 Using the existence of a so-called admissible set, we can construct a group isomorphism $K'^\times\to K^\times$.
-Finally, we will show that the group isomorphism can be extended to a field isomorphism $K\to K'$.
+Finally, we will show that this group isomorphism can be extended to a field isomorphism $K'\to K$.
 By passing to the limits, we obtain a field isomorphism $\varphi: \overline{K'}\to\overline{K}$ which induces $\sigma$.

### 2017-08-14 Master Thesis/synopsis.tex
Fix
Avoid iff

@@
 
-\section{The Valuation Spectra}
+\section{The Valuation Spectrum}
 
@@
 Let $w_L$ be the valuation on $L$ corresponding to $\tilde{w}\in \operatorname{Spv} k(v_L)$.
-By \cite{EP05} Corollary 4.1.4, $w_L$ is henselian iff both $v_L$ and $\tilde{w}$ are henselian, thus $L$ is a henselization of $w$.
+By \cite{EP05} Corollary 4.1.4, $w_L$ is henselian if and only if both $v_L$ and $\tilde{w}$ are henselian, thus $L$ is a henselization of $w$.
 \end{proof}
@@
 \begin{remark}
-Two valuations of rank $1$ are independent iff they are inequivalent.
+Two valuations of rank $1$ are independent if and only if they are inequivalent.
 \end{remark}
@@
 \begin{corollary}\label{corollary-fkschmidt}
-Let $w_1,w_2\in\operatorname{Spv}(K^\text{s})$ be two inequivalent valuations of rank $1$ with decomposition groups $G_{{w_1}},G_{{w_2}}\subset G_K$. Then $G_{{w_1}}\cap G_{{w_2}}=1$.
+Let $v_1,v_2\in\operatorname{Spv}(K^\text{s})$ be two inequivalent valuations of rank $1$ with decomposition groups $G_{{v_1}},G_{{v_2}}\subset G_K$. Then $G_{{v_1}}\cap G_{{v_2}}=1$.
 \end{corollary}
@@
 \begin{proof}
-Let $L$ be the fixed field of $G_{w_1}\cap G_{w_2}\subset G_K$ and set $v_i = w_i|_L$.
-Then $L$ is henselian with respect to both $v_1$ and $v_2$.
-Since $w_1,w_2$ are the only extensions of $v_1$ and $v_2$ in $K^\text{s}$ respectively, $v_1$ and $v_2$ are inequivalent valuations of rank $1$.
+Let $L$ be the fixed field of $G_{v_1}\cap G_{v_2}\subset G_K$.
+Then $L$ is henselian with respect to both $v_1|_L$ and $v_2|_L$.
+Since $v_1,v_2$ are the only extensions of $v_1|_L$ and $v_2|_L$ in $K^\text{s}$ respectively, $v_1$ and $v_2$ are inequivalent valuations of rank $1$ as well.
 Now $L$ is separably closed by Theorem \ref{theorem-fkschmidt}.
-Therefore, by Theorem \ref{theorem-henselization}, we get $G_{{w_1}}\cap G_{{w_1}}=G_L =1$.
+Therefore, by Theorem \ref{theorem-henselization}, we get $G_{{v_1}}\cap G_{{v_1}}=G_L =1$.
 \end{proof}
@@
 Let $K$ be henselian with respect to two different valuations $w_1,w_2\in |\operatorname{Spv}(K)|$.
-Then $K$ is algebraically closed or $w_1,w_2\in\operatorname{Spv}_\text{rd}^{(2)}(K),\ v(w_1)=v(w_2)$ and $k(v(w_1))$ is separably closed.
+Then $K$ is algebraically closed or $w_1,w_2\in\operatorname{Spv}_\text{rd}^{(2)}(K),\ v(w_1)=v(w_2)$, and $k(v(w_1))$ is separably closed.
 In this case, we have $\operatorname{cd}(K)\leq 2$.
@@
 \[ [\ell:k] = [\ell':k'] \]
-Therefore $L/K$ is a constant field extension iff $L'/K'$ is a constant field extension.
+Therefore $L/K$ is a constant field extension if and only if $L'/K'$ is a constant field extension.
 Since $K\overline{k}/K$ and $K'\overline{k'}/K$ are the union of their finite subextensions, we obtain $(K\overline{k})' = K'\overline{k'}$.
@@
 \begin{enumerate}[(i)]
-\item Let $f: A\to B$ be a homomorphism of abelian groups with finite kernel and cokernel. Then $A$ is free-by-finite iff $B$ is free-by-finite.
+\item Let $f: A\to B$ be a homomorphism of abelian groups with finite kernel and cokernel. Then $A$ is free-by-finite if and only if $B$ is free-by-finite.
 \item Let $0\longrightarrow A'\stackrel{i}{\longrightarrow}A\stackrel{p}{\longrightarrow}A''\longrightarrow 0$ be an exact sequence of abelian groups.
@@
 \begin{definition}\label{4.4}
-For a finite subset $S\subset X_0$ set:
+For a finite subset $S\subset X_0$ we set:
 \begin{align*}
-K^S &= \bigcap_{P\in X_0 -  S} U_P
+K^S &= \{ x\in K^\times \mid v_P(x)=0\quad\text{for all }P\in X_0-S \}
 \end{align*}

### 2017-08-15 Master Thesis/synopsis.tex
Correct characteristic issue in 2.4

@@
 \begin{lemma}\label{2.4-pre}
-Let $K$ be henselian with respect to a discrete valuation $v$, $k=k(v)$ its residue field. Let $p=\operatorname{char}(k)$, $\ell\neq p$ be a prime, and $\operatorname{cd}_\ell(k)\leq 2$. Then:
-\[ \mathrm{H}^3(K, \mathbb{Z}/\ell\mathbb{Z}(2)) \cong \mathrm{H}^2(k, \mu_{\ell}) \cong {}_\ell\operatorname{Br}(k) \]
-\end{lemma}
-
-\begin{proof}
+Let $K$ be henselian with respect to a discrete valuation $v$, $k=k(v)$ its residue field. Let $\ell$ be a prime such that $\ell=\operatorname{char}(k)$ or $\operatorname{cd}_\ell(k)\leq 2$. Then:
+\[ \mathrm{H}^3(K, \mathbb{Z}/\ell\mathbb{Z}(2)) \cong \mathrm{H}^2(k, \mathbb{Z}/\ell\mathbb{Z}(1)) \cong {}_\ell\operatorname{Br}(k) \]
+\end{lemma}
+
+\begin{proof}
+Let $p=\operatorname{char}(k)$.
+If $\ell = p$, the first isomorphism follows from \cite{Ka86} Lemma 1.4 (3) and the second one from \cite{Ka86} §0.
+Now assume $\ell\neq p$ and $\operatorname{cd}_\ell(k)\leq 2$.
+
 Let $T$ be the absolute inertia group of $K$ and $V\lhd T$ its absolute ramification group.
@@
 It yields an isomorphism $\mathrm{H}^2(k, \mathrm{H}^1(T, \mathbb{Z}/\ell\mathbb{Z}(2)))\cong \mathrm{H}^3(K, \mathbb{Z}/\ell\mathbb{Z}(2))$.
-Since $\mu_{\ell}$ is contained in the absolute inertia field of $K$, $T$ acts trivially on $\mu_{\ell}$, hence:
-\[ \mathrm{H}^1(T, \mathbb{Z}/\ell\mathbb{Z}(2)) \cong \mathrm{H}^1(T, \mu_{\ell}) \cong \operatorname{Hom}_\text{cont.}(T, \mu_{\ell}) \cong \operatorname{Hom}_\text{cont.}(T/V, \mu_{\ell}) \]
+Since $\mu_{\ell}$ is contained in the absolute inertia field of $K$, $T$ acts trivially on $\mathbb{Z}/\ell\mathbb{Z}(1)= \mu_{\ell}$ and $\mathbb{Z}/\ell\mathbb{Z}(2)=\mu_{\ell}^{\otimes 2}$, hence:
+\begin{align*}
+\mathrm{H}^1(T, \mathbb{Z}/\ell\mathbb{Z}(2)) &\cong \mathrm{H}^1(T, \mathbb{Z}/\ell\mathbb{Z}(1)) \\
+&\cong \operatorname{Hom}_\text{cont.}(T, \mathbb{Z}/\ell\mathbb{Z}(1)) \\
+&\cong \operatorname{Hom}_\text{cont.}(T/V, \mathbb{Z}/\ell\mathbb{Z}(1))
+\end{align*}
 By Theorem \ref{theorem-sylowgroups-inertiagroup}, $T/V$ is the free pro-$\mathcal{C}$ group of rank $1$ where $\mathcal{C}$ is the class of all finite groups with order prime to $p$.
-Therefore, $\operatorname{Hom}_\text{cont.}(T/V, \mu_{\ell})\cong \mu_{\ell}$.
-The rest follows from Proposition \ref{prop:kummer-theory}.
+Therefore, $\operatorname{Hom}_\text{cont.}(T/V, \mathbb{Z}/\ell\mathbb{Z}(1))\cong \mathbb{Z}/\ell\mathbb{Z}(1)$.
+The second isomorphism follows from Proposition \ref{prop:kummer-theory}
 \end{proof}
@@
 $\mu_{\ell}\subset K$ and $K$ is not formally real.
-Now we can replace $\mathbb{Z}/\ell\mathbb{Z}$ by $\mathbb{Z}/\ell\mathbb{Z}(2)$ and let us write $\mathrm{H}^3(L) = \mathrm{H}^3(L, \mathbb{Z}/\ell\mathbb{Z}(2))$.
+Now we can replace $\mathbb{Z}/\ell\mathbb{Z}$ by $\mathbb{Z}/\ell\mathbb{Z}(2) = \mu_{\ell}^{\otimes 2}$ and let us write $\mathrm{H}^3(L) = \mathrm{H}^3(L, \mathbb{Z}/\ell\mathbb{Z}(2))$.
 
@@
 where $w$ ranges over all rank-$2$ valuations with $v(w) = v$.
-By Lemma \ref{rank-dimension-inequality}, we have $\dim(k) = 1$, thus $k$ is a global field of characteristic $0$, hence a number field.
+By Lemma \ref{rank-dimension-inequality}, we have $\dim(k) = 1$, thus $k$ is a global field.
 \cite{Se64} Ch.
 II, §4 Prop.
-13 shows $\operatorname{cd}_\ell(k)\leq 2$.
+11 and \cite{Se64} Ch.
+II, §4 Prop.
+13 show $\operatorname{cd}_\ell(k)\leq 2$.
 With Lemma \ref{2.4-pre}, $\phi$ can be identified with
@@
 \[ [\ell:k] = [\ell':k'] \]
-Therefore $L/K$ is a constant field extension if and only if $L'/K'$ is a constant field extension.
+Therefore, $L/K$ is a constant field extension if and only if $L'/K'$ is a constant field extension.
 Since $K\overline{k}/K$ and $K'\overline{k'}/K$ are the union of their finite subextensions, we obtain $(K\overline{k})' = K'\overline{k'}$.

### 2017-08-15 Master Thesis/synopsis.tex
Roots of unity are contained in absolute inertia field

@@
 It yields an isomorphism $\mathrm{H}^2(k, \mathrm{H}^1(T, \mathbb{Z}/\ell\mathbb{Z}(2)))\cong \mathrm{H}^3(K, \mathbb{Z}/\ell\mathbb{Z}(2))$.
-Since $\mu_{\ell}$ is contained in the absolute inertia field of $K$, $T$ acts trivially on $\mathbb{Z}/\ell\mathbb{Z}(1)= \mu_{\ell}$ and $\mathbb{Z}/\ell\mathbb{Z}(2)=\mu_{\ell}^{\otimes 2}$, hence:
+Let $K^\text{t}/K$ be the absolute inertia field of $K$ and $w\in\operatorname{Spv}(K^\text{t})$ the extension of $v$.
+Since $k(w)$ is separably closed by Theorem \ref{theorem-absoluteinertia}, $\mu_{\ell}$ is contained in $K^\text{t}$ by Hensel's lemma.
+Therefore, $T = G_{K^\text{t}}$ acts trivially on $\mathbb{Z}/\ell\mathbb{Z}(1)= \mu_{\ell}$ and $\mathbb{Z}/\ell\mathbb{Z}(2)=\mu_{\ell}^{\otimes 2}$, hence:
 \begin{align*}
@@
 \[ 1 \longrightarrow U_w/n \longrightarrow L^\times /n\longrightarrow \mathbb{Z}/n\mathbb{Z}\longrightarrow 0 \]
-Since $k(w)$ is separably closed, with $\operatorname{char} k(w)=0$ and Hensel's lemma, we can see that $U_w$ is a divisible group, hence we get an isomorphism $L^\times /n\cong \mathbb{Z}/n\mathbb{Z}$ induced by $w$.
+Since $k(w)$ is separably closed, with Hensel's lemma, we can see that $U_w$ is a divisible group, hence we get an isomorphism $L^\times /n\cong \mathbb{Z}/n\mathbb{Z}$ induced by $w$.
 Therefore, the canonical map $K^\times/n\to L^\times/n$ can be identified with the map $K^\times/n \to\mathbb{Z}/n\mathbb{Z}$ induced by $v_P$.
@@
 By Remark \ref{remark-localphi-extension}, we can assume $K=K_P$ without loss of generality.
-Since $\operatorname{char}k(P)=0$, by Hensel's lemma, the map $r_P/n: U_P/n\to k(P)^\times/n$ is an isomorphism and the composition
+By Hensel's lemma, the map $r_P/n: U_P/n\to k(P)^\times/n$ is an isomorphism and the composition
 \[\begin{tikzcd}

### 2017-08-21 Master Thesis/synopsis.tex
Fix second batch typo

@@
 For the second assertion, let $K$ be not algebraically closed and $v=v(w_1)=v(w_2)\in\operatorname{Spv}_\text{d}^{(1)}(K)$.
-The restriction $v|$ of $v$ on $k=\mathbb{Q}(T)\subset K$ is a discrete valuation.
+The restriction $v|$ of $v$ on $k=\mathbb{Q}(T)\subset K$ is a discrete valuation since $k$ is finitely generated.
 Choose a henselization $k_{v|}$ of $k$ such that $k_{v|}\subset K$ and consider the absolute inertia field $k_{v|}^\text{t}$ of $k_{v|}$.
@@
 \begin{proof}
-Just like in Lemma \ref{2.4-finitelygenerated} we can assume that $\mu_\ell\subset K$ and $K$ is not formally real and replace $\mathbb{Z}/\ell\mathbb{Z}$ by $\mathbb{Z}/\ell\mathbb{Z}(2)$.
+Just like in Lemma \ref{2.4-finitelygenerated} we can assume that $\mu_\ell\subset K$ and $K$ is not formally real.
+Furthermore, we replace $\mathbb{Z}/\ell\mathbb{Z}$ by $\mathbb{Z}/\ell\mathbb{Z}(2)$.
 We adopt the notation from the previous proofs.
@@
 The implication (i)$\implies$(ii) follows from Lemma \ref{2.2}.
-Now assume (ii) holds.
-Let $v=v(w_1)\in \operatorname{Spv}_\text{d}^{(1)}(K_{w_1}K_{w_2})$.
-Then $k(v)$ is separably closed.
-We have $K_{w_1}K_{w_2}\subset K_v^\text{t}$ and hence $K_{w_1}K_{w_2}=K_v^\text{t}$ by Theorem \ref{theorem-absoluteinertia}.
-$\Gamma_v$ is equal to the value group of $v|_K\in\operatorname{Spv}_\text{d}^{(1)}(K)$, which is finitely generated.
+Now assume (ii) holds and let $v=v(w_1)\in\operatorname{Spv}^{(1)}_\text{d}(K)$.
+We have $K_{w_1}K_{w_2}\subset K_v^\text{t}$.
+By Theorem \ref{theorem-absoluteinertia}, the value group of $v$ on $K_v^\text{t}$ is equal to the value group of $v$ on $K$, which is finitely generated by Lemma \ref{rank-dimension-inequality}.
 However, value groups of valuations on an algebraically closed field are divisible and in particular not finitely generated, hence $K_v^\text{t}$ can't be algebraically closed and neither can the composite $K_{w_1}K_{w_2}$.
+
+Furthermore, if (ii) holds, the extension of $v$ on $K_{w_1}K_{w_2}$ has a separably closed residue field. By Theorem \ref{theorem-absoluteinertia}, we have $K_{w_1}K_{w_2}=K_v^\text{t}$.
 
@@
 Again by Lemma \ref{5.2} (ii), it suffices to show that $\mathcal{P}(F_1)/\mathcal{P}(F)$ is free-by-finite. 
-Consider the following diagram with exact rows:
+Consider the following commutative diagram with exact rows:
 \[ \begin{tikzcd}
@@
 \begin{corollary}\label{mordell-weil-corollary}
-Let $C$ be a smooth projective curve over a number field $F$. Then the Picard group $\operatorname{Pic}(C)$ is finitely generated.
+Let $C$ be a smooth projective curve over a number field $F$. Then its Picard group $\operatorname{Pic}(C)$ is finitely generated.
 \end{corollary}
@@
 Note that $\varphi_P: k(P')\to k(P)$ is a field isomorphism and that the residue maps $r_{P'}: \mathcal{O}_{P'} \to k(P')$ are ring homomorphisms.
-Since $K^S\to \prod_{P\in X_0 -  S} k(P)^\times$ is injective, we get $\varphi(x+y) = \varphi(x) + \varphi(y)$.\qedhere
+Since $K^S\to \prod_{P\in X_0 -  S} k(P)^\times$ is injective by Lemma \ref{5.4}, we get $\varphi(x+y) = \varphi(x) + \varphi(y)$.\qedhere
 \end{itemize}

### 2017-08-23 Master Thesis/introduction.tex
Add acknowledgements

@@
 
+I would like to thank my supervisor, Prof.
+Dr.
+Alexander Schmidt, for giving me such an interesting topic to pursue as my master thesis, and for all the insights and patient answers to all the countless questions I had.
+
+I also wish to offer my thanks to all my friends, who supported me in this endeavor and encouraged me to go on. Last but not least, I am grateful for my parents who never ran out of patience with me.
+
+This thesis was hard, arduous work, but, in the end, also very satisfying. And I couldn't have done it without you all. Thank you.
+
 \clearpage

### 2017-08-25 Master Thesis/synopsis.tex
'assures' > 'ensures'

@@
 \[\mathcal{O}_v\supsetneq \mathcal{O}_{v_1}\supsetneq\cdots \supsetneq\mathcal{O}_{v_n} \]
-Since $K$ is assumed to be finite dimensional, Lemma \ref{rank-dimension-inequality} assures that this sequence ends with a closed point $v_n\in \overline{\{v\}}\cap U_S\subset A\cap U_S$.
+Since $K$ is assumed to be finite dimensional, Lemma \ref{rank-dimension-inequality} ensures that this sequence ends with a closed point $v_n\in \overline{\{v\}}\cap U_S\subset A\cap U_S$.
 

### 2017-08-30 Master Thesis/introduction.tex
Rewrite of abstract and introduction

@@
 In fact, they all have the same absolute Galois group $\widehat{\mathbb{Z}}$, therefore not much information can be retrieved from these groups.
-The study of these types of questions belong to the area of \textit{anabelian geometry}.
+The study of these type of questions belong to the area of \textit{anabelian geometry}.
 This term refers to the fact that the less abelian $\pi_1^\text{ét}(X)$ is, the more information it carries about $X$.
@@
 Number fields stand in stark contrast to finite fields.
-Two number fields $k$, $k'$ with isomorphic Galois groups $G_k$, $G_{k'}$ are isomorphic as fields as well.
-It was J.
-Neukirch who investigated this question for the first time in the end of the 1960s (see \cite{Ne69}, \cite{Uc76}, \cite{Ne77}, \cite{Uc77}).
-Together with K.
-Uchida, their studies resulted in the following theorem for global fields:
+Two number fields $k$, $k'$ with isomorphic absolute Galois groups $G_k$, $G_{k'}$ are isomorphic as fields as well.
+This result is due to J.
+Neukirch who investigated these type of questions for the first time in the end of the 1960s (see \cite{Ne69}, \cite{Ne77}).
+With K.
+Uchida's studies (see \cite{Uc77}), it resulted in the following theorem for global fields:
 
@@
 
-It is believed that spectra of finitely generated fields are the building blocks of the anabelian world, which led Grothendieck to the conjecture that Theorem \ref{thm:neukirch-uchida} should also hold for arbitrary finitely generated fields $k$ and $k'$ of characteristic $0$.
+It is believed that spectra of finitely generated fields are the building blocks of the anabelian world, which led to the conjecture that Theorem \ref{thm:neukirch-uchida} should also hold for arbitrary infinite fields $k$ and $k'$ which are finitely generated over their prime fields.
+This has been proven by F.
+Pop using model theory (see \cite{Po90}, \cite{Po94}, \cite{Po95}).
 
-This has been proven by F.
-Pop (see \cite{Po90}, \cite{Po94}, \cite{Po95}) using model theory.
 In this thesis, we want to follow M.
-Spiess \cite{Sp96} to prove the statement for finitely generated fields with transcendence degree $1$ over $\mathbb{Q}$, using a purely arithmetical approach and a direct generalization of the original proof of Theorem \ref{thm:neukirch-uchida}:
+Spieß \cite{Sp96} to prove the statement for finitely generated fields with transcendence degree $1$ over $\mathbb{Q}$, using a purely arithmetical approach, which is a direct generalization of the original proof of Theorem \ref{thm:neukirch-uchida}:
 
@@
 
-I also wish to offer my thanks to all my friends, who supported me in this endeavor and encouraged me to go on. Last but not least, I am grateful for my parents who never ran out of patience with me.
+I also wish to offer my sincerest thanks to all my friends, who supported me in this endeavor and encouraged me to go on.
+Last but not least, I am grateful for my parents who never ran out of patience with me.
 
-This thesis was hard, arduous work, but, in the end, also very satisfying. And I couldn't have done it without you all. Thank you.
+This thesis was hard, arduous work, but, in the end, also very satisfying. And I couldn't have done it without all of you. Thank you.
 

### 2017-08-30 Master Thesis/synopsis.tex
Rewrite of abstract and introduction

@@
 
-\chapter{Correspondences}
-
-Let $K,K'$ be two finitely generated fields of characteristic $0$ and dimension $2$ and let $\sigma:G_K\to G_{K'}$ be an isomorphism of profinite groups.
+\chapter{Partial Correspondences}
+
+Let $K,K'$ be two finitely generated fields of characteristic $0$ and dimension $2$, and let $\sigma:G_K\to G_{K'}$ be an isomorphism of profinite groups.
 For intermediate fields $L$ of $\overline{K}/K$ we denote the corresponding intermediate field of $\overline{K'}/K'$ by $L'$.

### 2017-08-30 Master Thesis/thesis.tex
Slightly increase linespread
Rewrite of abstract and introduction

@@
 \hypersetup{bookmarksopen=true, pdfstartview=FitH}
-\linespread{1.1}
+\linespread{1.15}
 \author{Yichuan Shen}
@@
 In this thesis we will follow M.
-Spiess' argument and present his purely arithmetic proof of the fact that isomorphisms between the absolute Galois groups of function fields of transcendence degree $1$ over number fields are induced by field isomorphisms between these fields.
+Spieß' argument and present his purely arithmetic proof of the fact that isomorphisms between the absolute Galois groups of function fields of transcendence degree $1$ over number fields are induced by field isomorphisms between these fields.
 \end{abstract}
@@
 \begin{abstract}
-In dieser Arbeit wollen wir M.
-Spiess' rein arithmetischer Beweis nachgehen und zeigen, wieso Isomorphismen zwischen den absoluten Galoisgruppen von Funktionenkörpern vom Transzendenzgrad $1$ über Zahlkörper schon von Körperisomorphismen zwischen diesen Körpern induziert werden.
+In dieser Arbeit wollen wir zeigen, wieso Isomorphismen zwischen den absoluten Galoisgruppen von Funktionenkörpern vom Transzendenzgrad $1$ über Zahlkörper schon von Körperisomorphismen zwischen diesen Körpern induziert werden, indem wir M.
+Spieß' rein arithmetischen Beweis nachgehen.
 \end{abstract}

### 2017-08-31 Master Thesis/introduction.tex
Chapter restructuring

@@
 In the first chapter, we want to recap some results from valuation theory.
-In the second part, we will introduce some lemmata in order to characterize henselian fields over $K$ of valuations of rational rank $2$ in terms of their Galois cohomology groups of dimension $3$.
+Then we will introduce some lemmata in order to be able to characterize henselian fields over $K$ of valuations of rational rank $2$ in terms of their Galois cohomology groups of dimension $3$.
 
 In the second chapter, we will show that an isomorphism $\sigma: G_K\to G_{K'}$ induces bijections on the equivalence classes of defectless valuations of $K$ and $K'$ using the cohomological characterization of henselian fields.
-Using the Theorem of Neukirch-Uchida \ref{thm:neukirch-uchida}, we will also see that $\sigma$ induces a field isomorphism between the residue fields of corresponding discrete valuations.
+Using the theorem of Neukirch-Uchida \ref{thm:neukirch-uchida}, we will also see that $\sigma$ induces a field isomorphism between the residue fields of corresponding discrete valuations.
+
+In the third chapter, we will establish a field isomorphism between the constant fields of $K$ and $K'$ respectively.
 We will proceed to show that $\sigma$ induces group isomorphisms $K'^\times/n\to K^\times/n$ for all $n\in\mathbb{N}$.
 
-In the third chapter, we want to use the correspondences established in the second chapter to prove Theorem \ref{thm:main-result}.
+In the fourth chapter, we want to use the correspondences established in the third chapter to prove Theorem \ref{thm:main-result}.
 Before we are able to do that, we need to discuss free-by-finite groups.

### 2017-08-31 Master Thesis/synopsis.tex
Chapter restructuring

@@
 
-\chapter{Partial Correspondences}
+\chapter{Local Correspondence}
 
@@
 
-\section{Local Correspondence}
-
-In this section we want to see that $\sigma$ induces bijections on defectless valuations.
+In this chapter we want to see that $\sigma$ induces bijections on defectless valuations.
 A direct consequence of Proposition \ref{2.6} is the bijection $\operatorname{Spv}_\text{d}^{(2)}(K)\to \operatorname{Spv}_\text{d}^{(2)}(K')$:
@@
 
-\section{Global Correspondence mod \texorpdfstring{$n$}{n}}
-
-Let $n\in\mathbb{N}$.
-In this section we shall show that $\sigma: G_K\to G_{K'}$ induces group isomorphisms $\varphi_n: K'^\times/n\to K^\times/n$.
-By Proposition \ref{prop:kummer-theory}, we have isomorphisms:
-\[ \mathrm{H}^1(K',\mu_n)\cong K'^\times/n,\qquad \mathrm{H}^1(K,\mu_n)\cong K^\times/n \]
-In this section, our goal is to establish an isomorphism between the corresponding cohomology groups.
-
-\begin{definition}
+\chapter{Global Correspondence}
+
 Let $k$ and $k'$ be the algebraic closures of $\mathbb{Q}$ in $K$ and $K'$ respectively, i.e.
@@
 By Proposition \ref{3.2} $\sigma$ induces a bijection $\sigma: X^{\phantom{'}}_0\to X'_0,\ P\mapsto P'$ and for any $P\in X_0$ an isomorphism $\varphi_P: k(P')\to k(P)$.
-\end{definition}
+
+\section{Isomorphism of Constant Fields}
+
+In this section we shall establish an isomorphism between the absolute Galois groups of the constant fields $k$ and $k'$.
+In particular, since both $k$ and $k'$ are number fields, the theorem of Neukirch-Uchida \ref{thm:neukirch-uchida} will yield an isomorphism between $k$ and $k'$.
 
@@
 \end{proof}
-
-We start by showing that $\sigma$ induces an isomorphism between the absolute Galois groups of the constant fields $G_k$ and $G_{k'}$. We need the following Lemma:
 
@@
 \begin{lemma}\label{4.2}
-We have $(K\overline{k})'=K'\overline{k'}$, i.e. $\sigma(G_{K\overline{k}})=G_{K'\overline{k'}}$. $\sigma$ induces an isomorphism $G_k\to G_{k'}$ such that the following diagram commutes:
+We have $(K\overline{k})'=K'\overline{k'}$, i.e. $\sigma(G_{K\overline{k}})=G_{K'\overline{k'}}$. $\sigma$ induces an isomorphism $\sigma: G_k\to G_{k'}$ such that the following diagram commutes:
 \[ \begin{tikzcd}
@@
 \[G_k \cong \operatorname{Gal}(K\overline{k}/K) \to \operatorname{Gal}(K'\overline{k'}/K')\cong G_{k'}\] 
-Everything else follows from Neukirch-Uchida \ref{thm:neukirch-uchida}.
-\end{proof}
-
-\begin{remark}
+Everything else follows from the theorem of Neukirch-Uchida \ref{thm:neukirch-uchida}.
+\end{proof}
+
+\section{Global Correspondence mod \texorpdfstring{$n$}{n}}
+
+Let $n\in\mathbb{N}$.
+In this section we shall show that $\sigma: G_K\to G_{K'}$ induces group isomorphisms $\varphi_n: K'^\times/n\to K^\times/n$.
+By Proposition \ref{prop:kummer-theory}, we have isomorphisms:
+\[ \mathrm{H}^1(K',\mu_n)\cong K'^\times/n,\qquad \mathrm{H}^1(K,\mu_n)\cong K^\times/n \]
+Our goal is to establish an isomorphism between these cohomology groups.
+
 Let $\overline{K''} = \overline{K'}\otimes_{\overline{k'}}\overline{k}$ via the isomorphism $\varphi: \overline{k'}\to \overline{k}$. If we replace $\sigma: G_K\to G_{K'}$ with the composition
@@
 \end{tikzcd} \]
-\end{remark}
-
-\paragraph{} Let $n\in\mathbb{N}$.
-The operation of $G_K$ and $G_{K'}$ on $\mu_n$ factors through $G_k$, thus we have $\sigma(g)\xi = g\xi$ for all $\xi\in\mu_n$ and $g\in G_K$.
+
+The operation of $G_K$ and $G_{K'}$ on $\mu_n$ factors through $G_k$, thus we have $\sigma(g)\zeta = g\zeta$ for all $\zeta\in\mu_n$ and $g\in G_K$.
 Therefore, for all intermediate fields $L$ of $\overline{K}/K$, $\sigma$ induces an isomorphism:
@@
 
-\chapter{Theorem}
-
-Before we start proving Pop's Theorem \ref{thm:main-result}, we need some tools that will help us construct a group isomorphism $\varphi: K'^\times\to K^\times$ from the modulo $n$ isomorphisms $\varphi_n: K'^\times/n\to K^\times/n$ we constructed in the last chapter.
-
-\section{Free-By-Finite Groups}
+\chapter{Proof of Pop's Theorem}
+
+Before we start proving Pop's Theorem \ref{thm:main-result}, we need some tools that will help us construct a group isomorphism $\varphi: K'^\times\to K^\times$ from the modulo $n$ isomorphisms $\varphi_n: K'^\times/n\to K^\times/n$ which we constructed in the previous chapter.
+
+\section{Free-by-Finite Groups}
 
@@
 $C_2$ is free abelian of finite rank as well.
-By requirement $r_{P_2}(x)=a$ and $a\in k^\times$ is not of finite order, thus $x\not\in C_2$ and $\operatorname{rank}(C_2) < \operatorname{rank}(C_1)$.
+By requirement, $r_{P_2}(x)=a$ and $a\in k^\times$ is not of finite order, thus $x\not\in C_2$ and $\operatorname{rank}(C_2) < \operatorname{rank}(C_1)$.
 By repeating this process until $C_r=1$, we get points $P_2,\ldots,P_r\in X_0-S$ such that the following map is injective:
@@
 
-\section{Proof of Pop's Theorem}
+\section{Proof}
 

### 2017-08-31 Master Thesis/introduction.tex
Fix some typos

@@
 Given an algebraic variety $X$, we can ask ourselves how much information about $X$ is contained in its étale fundamental group $\pi_1^{\text{ét}}(X)$.
-Especially interesting are schemes which are reconstructible solely from its étale fundamental group.
+Especially interesting are schemes which are reconstructible solely from their étale fundamental group.
 If $X$ is the spectrum of a field $K$, the étale fundamental group of $X$ is just its absolute Galois group $G_K$.
@@
 In fact, they all have the same absolute Galois group $\widehat{\mathbb{Z}}$, therefore not much information can be retrieved from these groups.
-The study of these type of questions belong to the area of \textit{anabelian geometry}.
+The study of these types of questions belong to the area of \textit{anabelian geometry}.
 This term refers to the fact that the less abelian $\pi_1^\text{ét}(X)$ is, the more information it carries about $X$.
@@
 This result is due to J.
-Neukirch who investigated these type of questions for the first time in the end of the 1960s (see \cite{Ne69}, \cite{Ne77}).
+Neukirch who investigated these types of questions for the first time in the end of the 1960s (see \cite{Ne69}, \cite{Ne77}).
 With K.
@@
 
-This means that global fields should be reconstructible from their Galois groups. Furthermore, the Galois group $G_k$ also contains important arithmetic information from $k$.
+This means that global fields should be reconstructible from their absolute Galois groups. Furthermore, the Galois group $G_k$ also contains important arithmetic information from $k$.
 
@@
 
-In the second chapter, we will show that an isomorphism $\sigma: G_K\to G_{K'}$ induces bijections on the equivalence classes of defectless valuations of $K$ and $K'$ using the cohomological characterization of henselian fields.
+In the second chapter, we will show that an isomorphism $\sigma: G_K\to G_{K'}$ induces bijections on the equivalence classes of defectless valuations of $K$ and $K'$ using the cohomological characterization of henselian fields developed in the first chapter.
 Using the theorem of Neukirch-Uchida \ref{thm:neukirch-uchida}, we will also see that $\sigma$ induces a field isomorphism between the residue fields of corresponding discrete valuations.
@@
 
-In the fourth chapter, we want to use the correspondences established in the third chapter to prove Theorem \ref{thm:main-result}.
+In the fourth chapter, we want to use the correspondences established in the previous chapters to prove Theorem \ref{thm:main-result}.
 Before we are able to do that, we need to discuss free-by-finite groups.

### 2017-08-31 Master Thesis/synopsis.tex
Fix some typos

@@
 $w_1$ and $w_2$ can't be trivial, for otherwise $\dim k(w_i)=\dim (K)\neq 0$.
-By the Theorem of F.K.
+By the theorem of F.K.
 Schmidt \ref{theorem-fkschmidt}, we can see that $w_1,w_2$ are dependent.
@@
 By Lemma \ref{2.2-pre} $K$ is henselian with respect to both $v_1$ and $v_2$.
-Again, by the Theorem of F.K.
+Again, by the theorem of F.K.
 Schmidt \ref{theorem-fkschmidt}, we get $v_1 = v_2$.
@@
 \[ G_{k(v)} \cong G_{K_v}/G_{K_v^\text{t}} \to G_{K'_{v'}}/G_{(K')_{v'}^\text{t}}\cong G_{k(v')} \]
-The rest follows from the Theorem of Neukirch-Uchida \ref{thm:neukirch-uchida}.
+The rest follows from the theorem of Neukirch-Uchida \ref{thm:neukirch-uchida}.
 \end{proof}
@@
 Now $\operatorname{Pic}^0(C)$ can be identified with the $F$-rational points of the Jacobian variety $J(C)$ of $C$.
-$J(C)$ is finitely generated by the Mordell-Weil Theorem \ref{mordell-weil-theorem}, hence $\operatorname{Pic}(C)$ is finitely generated as well.
+$J(C)$ is finitely generated by the Mordell-Weil theorem \ref{mordell-weil-theorem}, hence $\operatorname{Pic}(C)$ is finitely generated as well.
 \end{proof}
